### 2010-09-14 proposal/preProposal.tex
Skeleton pre-proposal

\section{Team}

1.The Team. Names of the people working on the project. We suggest
working in groups of about three, but can be convinced otherwise if
you have a good reason. If you can’t find a group, email the TAs and
we will assign you to a group.

\section{Problem statement}

2.Problem statement and motivation. A statement of the problem, issue,
or task that you’re interested in studying, and why it is interesting
or important.

\section{I/O specification}

3.I/O Specification. A clear, concise description of what the final
system will do in terms of I/O behavior: What will it take in, and
what it will produce.

\section{Background reading (optional)}

4.Background Reading (OPTIONAL in pre-proposal stage). A list of
relevant readings that you’ll use to gain some background in your
selected topic.

\section{General approach (optional)}

5.General Approach (OPTIONAL in pre-proposal stage). A high-level
description of the general approach you’ll use (e.g. heuristic search,
learning, rules, belief networks). This section should have a
subsection entitled “Where’s the AI” where you explicitly articulate
the AI component of this project.

\section{System architecture and work plan (optional)}

6.System Architecture and work plan (OPTIONAL in pre-proposal
stage). Explain the main components of the system, how they can be
independently developed and independently tested, and who will do
what.

\section{Data sources (optional)}

7.Data sources (OPTIONAL in pre-proposal stage). If your project
involves analyzing real data (e.g. stock market data, sport data, user
behavior), identify potential sources for this data.

\section{Evaluation plans (optional)}

8.Evaluation Plans (OPTIONAL in pre-proposal stage). An explicit,
coherent plan for quantitatively and/or qualitatively evaluating the
system. In particular, identify (a) metrics that will help you measure
performance, e.g. time or space to solve a problem, quality of a
solution, or competition or comparison against a standard approach (b)
a simple “toy” problem that you can use for early testing of your
system – something that should be trivial to solve, (b) a “hard” test
case, that will serve as you ultimate test.

\section{Schedule}

9.Schedule. A schedule of work indicating the dates by which you plan
to complete components of the system. This should be presented as a
table listing dates and milestones.

### 2010-09-14 proposal/preProposal.tex
Schedule done, skeleton less skeletony :).

@@
 
-1.The Team. Names of the people working on the project. We suggest
-working in groups of about three, but can be convinced otherwise if
-you have a good reason. If you can’t find a group, email the TAs and
-we will assign you to a group.
+The human members of our team are Diana Hidalgo, Sarah Nguyen, and
+Jason Yosinski.  The sole robotic member of our team is a 9 DoF
+starfish shaped robot from Hod Lipson's lab.
 
 \section{Problem statement}
+
+\edit{todo}
 
@@
 
+\edit{todo}
+
 3.I/O Specification. A clear, concise description of what the final
@@
 
+\edit{todo}
+
 4.Background Reading (OPTIONAL in pre-proposal stage). A list of
@@
 \section{General approach (optional)}
+
+\edit{todo}
 
@@
 
+\edit{todo}
+
 6.System Architecture and work plan (OPTIONAL in pre-proposal
@@
 
+\edit{todo}
+
 7.Data sources (OPTIONAL in pre-proposal stage). If your project
@@
 \section{Evaluation plans (optional)}
+
+\edit{todo}
 
@@
 
-9.Schedule. A schedule of work indicating the dates by which you plan
-to complete components of the system. This should be presented as a
-table listing dates and milestones.
+Below is a rough schedule of fixed external deadlines and anticipated
+team milestones.
+
+\begin{center}
+\begin{tabular}{|l||p{2.5in}|p{1.5in}|}
+\hline
+                        & {\bf Milestones} & {\bf Deadlines}
+\\ \hline
+
+Week 1 (9/13-9/19)    & Read papers, get lab access, talk to relevant other researchers & 9/17 Final proposals due
+\\ \hline
+
+Week 2 (9/20-9/26)    & Continue reading, get robot to move &
+\\ \hline
+
+Week 3 (9/27-10/03)   & Implement parametrized gait and determine proposed coding schedule for more advanced algorithms in time for Code Review \#1. &
+\\ \hline
+
+Week 4 (10/04-10/10)  & Begin main algorithm dev/testing effort & 10/5 Code Review \#1
+\\ \hline
+
+Week 5 (10/11-10/17)  & Algorithm dev/testing &
+\\ \hline
+
+Week 6 (10/18-10/24)  & Algorithm dev/testing &
+\\ \hline
+
+Week 7 (10/25-10/31)  & Algorithm dev/testing, quantify/solidify current results for Code Review \#2 &
+\\ \hline
+
+Week 8 (11/1-11/7)    & Finish collecting results, begin writing & 11/2 Code RReview \#2
+\\ \hline
+
+Week 9 (11/8-11/14)   & Finish collecting results, writing &
+\\ \hline
+
+Week 10 (11/15-11/21) & Finish collecting results, writing, get final demo ready &
+\\ \hline
+
+Week 11 (11/22-11/28) & Finish collecting results, writing, get final demo ready  &
+\\ \hline
+
+Week 12 (11/29-11/30) & Final demo  & 11/30 Final presentation
+\\ \hline
+\end{tabular}
+\end{center}

### 2010-09-14 proposal/preProposal.tex
minor
Version presented/sent as pre-proposal

@@
 The human members of our team are Diana Hidalgo, Sarah Nguyen, and
-Jason Yosinski.  The sole robotic member of our team is a 9 DoF
-starfish shaped robot from Hod Lipson's lab.
+Jason Yosinski.  The sole robotic member of our team is a 9 degree of
+freedom starfish shaped robot from Hod Lipson's lab.
 
@@
 
-\edit{todo}
+There has been much interest in and research surrounding legged
+robots.  Such robots exhibit both challenges and opportunities not
+present in more traditional wheeled robots.  One of these challenges
+is getting the robot to move as quickly and efficiently as possible.
+Past research has shown that machine learning methods of optimizing a
+walking robot's gait can outperform even the best hand coded gaits
+(Chernova 2004).
 
-2.Problem statement and motivation. A statement of the problem, issue,
-or task that you’re interested in studying, and why it is interesting
-or important.
+To this end, we will use machine learning to design a gait for a
+quadraped robot, available from Hod Lipson's lab.  We anticipate using
+speed and efficiency as metrics for evaluation of the designed gait;
+these metrics are presented further in Section~\ref{sec:evaluation}.
+
+%2.Problem statement and motivation. A statement of the problem, issue,
+%or task that you’re interested in studying, and why it is interesting
+%or important.
 
@@
 
-\edit{todo}
+Input:
 
-3.I/O Specification. A clear, concise description of what the final
-system will do in terms of I/O behavior: What will it take in, and
-what it will produce.
+\begin{itemize}
+\item Motor encoder position (optional, likely)
+\item Motor force feedback (optional, likely)
+\item Robot position estimate via external pose estimation system (optional, less likely)
+\end{itemize}
 
-\section{Background reading (optional)}
+\noindent Output:
 
-\edit{todo}
+\begin{itemize}
+\item Motor position commands over time
+\end{itemize}
 
-4.Background Reading (OPTIONAL in pre-proposal stage). A list of
-relevant readings that you’ll use to gain some background in your
-selected topic.
+\section{Background reading}
 
-\section{General approach (optional)}
+We have selected three papers to read for background information on
+walking robots. The papers are \emph{An Evolutionary Approach to Gait
+  Learning for Four-Legged Robots} by Sonia Chenova and Manuela
+Veloso, \emph{Policy Gradient Reinforcement Learning for Fast
+  Quadrepedal Locomotion} by Nate Kohl and Peter Stone, and
+\emph{Evolving Dynamic Gaits on a Physical Robot} by Viktor Zykov,
+Josh Bongard, and Hod Lipson. Hopefully these papers will give us
+insights on what kinds of algorithms and techniques to employ in our
+learning problem.
 
-\edit{todo}
+\section{General approach}
 
-5.General Approach (OPTIONAL in pre-proposal stage). A high-level
-description of the general approach you’ll use (e.g. heuristic search,
-learning, rules, belief networks). This section should have a
-subsection entitled “Where’s the AI” where you explicitly articulate
-the AI component of this project.
+\subsection{Where's the AI?}
 
-\section{System architecture and work plan (optional)}
+Our general approach is still under development.  We are currently
+evaluating the following AI methods for possible inclusion:
 
-\edit{todo}
+\begin{itemize}
+\item Optimization (gradient descent, Levenberg-Marquardt, ...)
+\item Evolutionary algorithm / Genetic algorithm
+\item Reinforcement learning
+\item Supervised learning?
+\end{itemize}
 
-6.System Architecture and work plan (OPTIONAL in pre-proposal
-stage). Explain the main components of the system, how they can be
-independently developed and independently tested, and who will do
-what.
+In addition to these more general methods, the following issues will
+also need to be investigated:
 
-\section{Data sources (optional)}
+\begin{itemize}
+\item Reduced dimensionality of parameter space (periodic, symmetric, ...)
+\item Parametrized gait vs. non-parametrized gait (?)
+\item Geometric constraints
+\end{itemize}
 
-\edit{todo}
+\section{System architecture and work plan}
 
-7.Data sources (OPTIONAL in pre-proposal stage). If your project
-involves analyzing real data (e.g. stock market data, sport data, user
-behavior), identify potential sources for this data.
+We will be working with a robot that has an on-board computer running Linux. The lower level drivers are in C and we hope to implement the system in Python.
 
-\section{Evaluation plans (optional)}
+Once we begin working with the robot and have an idea of the algorithm we will use, we will divide the work.
 
-\edit{todo}
+% \section{Data sources (optional)}
+% 
+% \edit{todo}
+% 
+% 7.Data sources (OPTIONAL in pre-proposal stage). If your project
+% involves analyzing real data (e.g. stock market data, sport data, user
+% behavior), identify potential sources for this data.
+%
 
-8.Evaluation Plans (OPTIONAL in pre-proposal stage). An explicit,
-coherent plan for quantitatively and/or qualitatively evaluating the
-system. In particular, identify (a) metrics that will help you measure
-performance, e.g. time or space to solve a problem, quality of a
-solution, or competition or comparison against a standard approach (b)
-a simple “toy” problem that you can use for early testing of your
-system – something that should be trivial to solve, (b) a “hard” test
-case, that will serve as you ultimate test.
+\section{Evaluation plans}
+\label{sec:evaluation}
+
+To evaluate our system, we will have three main metrics: speed, number
+of failures, and efficiency, which will be compared to its initial
+hard-coded walking, and compared to other quadraped robots in the lab:
+
+\begin{itemize}
+\item Speed taken to walk a certain distance
+\item Number of failures (falling down, getting stuck) over a certain number of attempts
+\item Efficiency and power consumption, estimated through a model of the motor
+\end{itemize}
 

### 2010-09-17 proposal/preProposal.tex
edits with S and D

@@
 these metrics are presented further in Section~\ref{sec:evaluation}.
+The robot learning will be accomplished via one of several machine
+learning algorithms, discussed in Section~\ref{sec:approach}.
 
@@
 \section{General approach}
+\label{sec:approach}
 
@@
 
-We will be working with a robot that has an on-board computer running Linux. The lower level drivers are in C and we hope to implement the system in Python.
+We will be working with a robot that has an on-board computer running
+either Windows or Linux. The lower level drivers are in C and we hope
+to implement the system in Python.
 
-Once we begin working with the robot and have an idea of the algorithm we will use, we will divide the work.
+Once we begin working with the robot and have an idea of the algorithm
+we will use, we will divide the work among us.
 

### 2010-09-17 proposal/preProposal.tex
Changed workplan.

@@
 
-Once we begin working with the robot and have an idea of the algorithm
-we will use, we will divide the work among us.
+The first week we will work together on making the robot move.
+Until we get the platrom ready to run our code, it will be difficult to separate the tasks.
+However, once we begin implementing our chosen algorithms, this division will become more clear.
 

### 2010-09-17 proposal/proposal.tex
Moved to more appropriate name

--- 2010-09-17 proposal/preProposal.tex
@@
 
-To evaluate our system, we will have three main metrics: speed, number
-of failures, and efficiency, which will be compared to its initial
-hard-coded walking, and compared to other quadraped robots in the lab:
+We will need to evaluate our system for two separate reasons.  First,
+our learning methods depend on knowing the current performance of the
+robot.  In order to optimize this performance, the learning algorithms
+themselves will need access to the metrics of interest.  Of course,
+the other reason for evaluation is to assess the performance of the
+system as a whole.
+
+For this latter purpose, we propose the use of three main metrics:
 
@@
 \end{itemize}
+
+\noindent We anticipate comparing our results along these dimensions
+to a hard-coded walking configuration, as well as possibly to other
+quadraped robots in the lab:
 

### 2010-10-04 code/Robot.py
Halfway through Robot.py refactor

@@ -2,116 +2,206 @@ import os, dynamixel, time, datetime
 from RobotParams import *
 from Motion import lInterp
 
-def initialize():
-    '''Creates a network and scans it for servos.
-
-    Returns (network, actuatorList)
-    '''
-    # The number of Dynamixels on our bus.
-    nServos = 11
-
-    # Set your serial port accordingly.
-    if os.name == "posix":
-        portName = "/dev/ttyUSB0"
-    else:
-        portName = "COM11"
-
-    # Default baud rate of the USB2Dynamixel device.
-    baudRate = 1000000
-
-    serial = dynamixel.SerialStream( port=portName, baudrate=baudRate, timeout=1)
-    net = dynamixel.DynamixelNetwork( serial )
-    net.scan( 0, nServos )
-
-    actuators = list()
-
-    print "Scanning for Dynamixels...",
-    for dyn in net.get_dynamixels():
-        print dyn.id,
-        actuators.append(net[dyn.id])
-    print "...Done"
-
-    return net, actuators
+'''
+Much inspiration taken from http://code.google.com/p/pydynamixel/
+'''
 
 
-def runRobotWith(paramFunction, runSeconds = 10):
-    net, actuators = initialize()
 
-    for actuator in actuators:
-        actuator.moving_speed = 90
-        actuator.synchronized = True
-        actuator.torque_enable = True
-        actuator.torque_limit = 1000
-        actuator.max_torque = 1000
+class Robot():
+    ''''''
 
-    print 'Starting in 0 sec... ',
-    #time.sleep(1)
-    print 'go.'
+    def __init__(self, silentNetFail = False, nServos = 9):
+        '''Initialize the robot.
+        
+        Keyword arguments:
 
-    time0 = datetime.datetime.now()
-    startingPos = None
-    while True:
-        if startingPos == None:
-            startingPos = currentPositions(actuators)
-        timeDiff = datetime.datetime.now() - time0
-        seconds  = timeDiff.seconds + timeDiff.microseconds/1e6
+        silentNetworkFail -- Whether or not to fail silently if the
+                             network does not find all the dynamixel
+                             servos.
 
-        if seconds < 3:
-            goal = lInterp(seconds, [0, 3], startingPos, POS_FLAT)
-        elif seconds < 6:
-            goal = lInterp(seconds, [3, 6], POS_FLAT, POS_READY)
-        #elif seconds < 10:
[…]

### 2010-10-04 code/optimize.py
Cleaned up and commented code. Created slightNeighbor(). Added check to doRun() that checks whether the current state has already been tested or not

@@ -2,7 +2,7 @@
 
 '''
 
-@date: 1 Octoberx 2010
+@date: 4 October 2010
 
 '''
 
@@ -18,28 +18,28 @@ from copy import copy
 from Robot import runRobotWith, cropPositions
 from SineModel import sineModel
 
-""" 
-Given the ranges of the different parameters, chooses random values for
-each parameter. The ranges of parameters are in a list of tuples.
-
-"""
 def initialState(ranges):
+    """ 
+    Given the ranges of the different parameters, chooses random values for
+    each parameter. The ranges of parameters are in a list of tuples.
+
+    """
     parameters = []  # List of the chosen values for the parameters
     for rang in ranges:
         # Chooses random values for each parameter (initial state)
-        # TODO: Sometimes we need floats and sometimes ints. Hrmm. Need to deal
-        if isinstance(rang[0], bool):
+        if isinstance(rang[0], bool):  # If range is (true, false),
+				       # choose true or false
             parameters.append(random.uniform(0,1) > .5)
         else:
             parameters.append(random.uniform(rang[0], rang[1]))
     return parameters
 
-"""
-Given a list of parameters, picks a random parameter to change, randomly
-changes it (TODO: Something more sophisticated), and returns a new list.
-
-"""
 def neighbor(ranges, parameters):
+    """
+    Given a list of parameters, picks a random parameter to change, randomly
+    changes it, and returns a new list.
+
+    """
     ret = copy(parameters)
     print '  ** Neighbor old', ret
     index = random.randint(0, len(parameters) - 1)
@@ -53,11 +53,34 @@ def neighbor(ranges, parameters):
     print '  ** Neighbor new', ret
     return ret
 
+def slightNeighbor(ranges, parameters):
+    """
+    Given a list of parameters, picks a random parameter to change, and
+    decreases or increases it slightly.
 
+    """
+    ret = copy(parameters)
+    print '  ** Neighbor old', ret
+    index = random.randint(0, len(parameters) - 1)
+
+    print ranges
+    if isinstance(ranges[index][0], bool):
+        ret[index] = (random.uniform(0,1) > .5)
+    else:
+        if random.randint(0, 1) == 0:  # decrease slightly
+	    ret[index] = ret[index] - (.1 * (ranges[index][1] - ranges[index][0]))
+        else:  # increase slightly
[…]

### 2010-10-04 code/Robot.py
More work, probably still broken

@@ -8,6 +8,27 @@ Much inspiration taken from http://code.google.com/p/pydynamixel/
 
 
 
+'''Min and max values for the QuadraTot robot, based on some tests.
+
+Note that these values will avoid collisions only for each servo
+individually.  More complex collisions are still possible given
+certain vectors of motor position.
+'''
+
+MIN_INNER = 150
+MAX_INNER = 770
+MIN_OUTER = 30
+MAX_OUTER = 940
+MIN_CENTER = 512 - 180
+MAX_CENTER = 512 + 180
+
+POS_FLAT      = [512] * 9
+POS_READY     = [800,  40] * 4 + [512]
+POS_HALFSTAND = [700, 100] * 4 + [512]
+POS_STAND     = [512, 150] * 4 + [512]
+
+
+
 class Robot():
     ''''''
 
@@ -25,7 +46,10 @@ class Robot():
         '''
 
         # The number of Dynamixels on our bus.
-        self.nServos = 9
+        self.nServos = nServos
+
+        if self.nServos != 9:
+            raise Exception('Unfortunately, the Robot class currently assumes 9 servos.')
 
         # Set your serial port accordingly.
         if os.name == "posix":
@@ -142,12 +166,7 @@ class Robot():
 
             #print 'At %3.3f, commanding:' % seconds,
 
-            for ii,actuator in enumerate(actuators):
-                gg = int(round(goal[ii],0))
-                #gg = max(min(gg, servoMax), servoMin)
-                actuator.goal_position = gg
-                #print gg,
-            net.synchronize()
+            self.commandPosition(........)
             #print
 
             #for actuator in actuators[:8]:
@@ -172,23 +191,37 @@ class Robot():
                        positions are cropped.  Default: False.
         '''
 
+        goalPosition = self.cropPosition(position, cropWarning)
+        
+        for ii,actuator in enumerate(self.actuators):
+            gg = int(round(goal[ii],0))
+            #gg = max(min(gg, servoMax), servoMin)
+            actuator.goal_position = gg
+            #print gg,
+        net.synchronize()
         
 
 
 
-    def cropPositions(pos):
+    def cropPosition(position, cropWarning = False):
         '''Crops the given positions to their appropriate min/max values.
+        
         Requires a vector of length 9 to be sure the IDs are in the
         assumed order.'''
 
-        if len(pos) != 9:
[…]

### 2010-10-05 code/optimize.py
doRun() appends data to a log file.

@@ -9,7 +9,7 @@
 """
 Employs random hill-climbing to choose and evaluate the parameters of the robots'
 motion, either by changing one parameter completely randomly, or changing one
-parameter slightly. Evaluates each neighbor using user-inputted distance
+parameter slightly. Evaluates each neighbor using user-inputed distance
 walked.
 
 """
@@ -118,8 +118,13 @@ def doRun():
                 bestDistance = copy(currentDistance)
 
             print "best so far", bestState, bestDistance  # Prints best state and distance so far
-
-        currentState = neighbor(ranges, bestState)  # Pick some neighbor
+        
+	# Writes to log file that keeps track of tests so far
+	for a in currentState
+        stats += a + " "
+    logFile = open('log.txt', 'a')
+	logFile.write(stats + ", " + str(currentDistance))
+	logFile.close()
 
     return bestState  # Return the best solution found (a list of params)
 

### 2010-10-05 code/optimize.py
added new line to logfile stuff

@@ -123,7 +123,7 @@ def doRun():
 	for a in currentState
         stats += a + " "
     logFile = open('log.txt', 'a')
-	logFile.write(stats + ", " + str(currentDistance))
+	logFile.write(stats + ", " + str(currentDistance) + "\n")
 	logFile.close()
 
     return bestState  # Return the best solution found (a list of params)

### 2010-10-05 code/optimize.py
minor

@@ -17,7 +17,7 @@ walked.
 import math, pdb
 import random
 from copy import copy
-from Robot import runRobotWith, cropPositions
+from Robot import Robot
 from SineModel import sineModel
 
 def initialState(ranges):
@@ -120,11 +120,11 @@ def doRun():
             print "best so far", bestState, bestDistance  # Prints best state and distance so far
         
 	# Writes to log file that keeps track of tests so far
-	for a in currentState
+    for a in currentState:
         stats += a + " "
     logFile = open('log.txt', 'a')
-	logFile.write(stats + ", " + str(currentDistance) + "\n")
-	logFile.close()
+    logFile.write(stats + ", " + str(currentDistance) + "\n")
+    logFile.close()
 
     return bestState  # Return the best solution found (a list of params)
 

### 2010-10-05 code/Robot.py
Bug fixes

@@ -1,8 +1,11 @@
-import os, dynamixel, time, datetime
-from RobotParams import *
-from Motion import lInterp
+import os
+from datetime import datetime
 from time import sleep
 from types import FunctionType
+from copy import copy
+
+import dynamixel
+from Motion import lInterp
 
 '''
 Much inspiration taken from http://code.google.com/p/pydynamixel/
@@ -67,7 +70,9 @@ class Robot():
         # Default baud rate of the USB2Dynamixel device.
         self.baudRate = 1000000
 
-        serial = dynamixel.SerialStream(port=portName, baudrate=baudRate, timeout=1)
+        serial = dynamixel.SerialStream(port = portName,
+                                        baudrate = self.baudRate,
+                                        timeout = 1)
         self.net = dynamixel.DynamixelNetwork(serial)
 
         print "Scanning for Dynamixels...",
@@ -75,9 +80,9 @@ class Robot():
 
         self.actuators = list()
 
-        for dyn in net.get_dynamixels():
+        for dyn in self.net.get_dynamixels():
             print dyn.id,
-            self.actuators.append(net[dyn.id])
+            self.actuators.append(self.net[dyn.id])
         print "...Done"
 
         if len(self.actuators) != self.nServos and not silentNetFail:
@@ -92,7 +97,7 @@ class Robot():
             actuator.max_torque = 1000
 
         self.currentPos = None
-        self.resetTime()
+        self.resetClock()
         
     def run(self, motionFunction, runSeconds = 10, resetFirst = True,
             interpBegin = 0, interpEnd = 0):
@@ -143,16 +148,16 @@ class Robot():
 
         print 'Starting motion.'
 
-        self.resetTime()
+        self.resetClock()
         self.currentPos = self.readCurrentPosition()
 
         # Reset the robot position, if desired
         if resetFirst:
-            self.interpMove(self.currentPosition(), POS_FLAT, 3)
+            self.interpMove(self.readCurrentPosition(), POS_FLAT, 3)
             self.interpMove(POS_FLAT, POS_READY, 3)
             #self.interpMove(POS_READY, POS_HALFSTAND, 4)
             self.currentPos = POS_READY
-            self.resetTime()
+            self.resetClock()
 
         # Begin with a segment smoothly interpolated between the
         # current position and the motion model.
@@ -181,7 +186,7 @@ class Robot():
         while self.time < timeEnd:
             posS = start(seconds) if isinstance(start, FunctionType) else start
[…]

### 2010-10-05 code/optimize.py

@@ -99,15 +99,15 @@ def doRun():
     
     bestDistance = -1e100
 
+    stateSoFar = set()
     for ii in range(10000):
-	if currentState not in statesSoFar:  # Make sure this state is new
+	if tuple(currentState) not in statesSoFar:  # Make sure this state is new
 	    stateSoFar.add(tuple(currentState))
             print
             print 'Iteration', ii, 'params', currentState
         
             motionModel = lambda time: sineModel(time,
-                                                 parameters = currentState,
-                                                 croppingFunction = cropPositions)
+                                                 parameters = currentState)
 
             android.run(motionModel)
 

### 2010-10-05 code/SineModel.py
Reverted to older SineModel (newer code in branch), reenabled multFB and multLR

@@ -10,8 +10,7 @@ def sineModel(time, parameters, croppingFunction = None):
 
     pFloat = [float(xx) for xx in parameters]
     #amp, tau, scaleInOut, flipFB, flipLR = pFloat
-    #amp, tau, scaleInOut, multFB, multLR = pFloat
-    amp, tau, scaleInOut, multFB, multLR, shiftFB, shiftLR = pFloat
+    amp, tau, scaleInOut, multFB, multLR = pFloat
 
     centerConst = 512
     ret = [centerConst] * 9
@@ -21,29 +20,15 @@ def sineModel(time, parameters, croppingFunction = None):
     offsetOuter = 40
     
     # Compute base sine wave
-    angle0    = 2*pi*time/tau
-    angle1    = 2*pi*time/tau
-    angle2    = 2*pi*time/tau
-    angle3    = 2*pi*time/tau
-
-    angle0    += 2*pi*shiftFB
-    angle1    += 2*pi*shiftFB
-
-    angle1    += 2*pi*shiftLR
-    angle2    += 2*pi*shiftLR
-    
-    tBase = time
-    tFB   = tBase + shiftFB
     base = amp * sin(2*pi*time/tau)
 
     idxInner = [0, 2, 4, 6]
     idxOuter = [1, 3, 5, 7]
-    idxFB    = [0, 1, 2, 3]
 
     idxLR    = [6, 7, 0, 1]
     idxFront = [0, 1, 2, 3]
     idxBack  = [4, 5, 6, 7]
-
+    
     idxLR    = [6, 7, 0, 1]
     idxLeft  = [6, 7, 0, 1]
     idxRight = [2, 3, 4, 5]
@@ -63,7 +48,7 @@ def sineModel(time, parameters, croppingFunction = None):
         ret[idxFront[ii]] = multFB * ret[idxBack[ii]]
     for ii in range(len(idxLeft)):
         ret[idxLeft[ii]] = multLR * ret[idxRight[ii]]
-
+                    
     for ii in idxInner:
         ret[ii] += offsetInner
     for ii in idxOuter:

### 2010-10-05 code/Robot.py
Fixed a few bugs, made optimize.py printouts prettier, gave Robot class verbosity control

@@ -37,7 +37,8 @@ POS_STAND     = [512, 150] * 4 + [512]
 class Robot():
     ''''''
 
-    def __init__(self, silentNetFail = False, nServos = 9, commandRate = 40):
+    def __init__(self, silentNetFail = False, nServos = 9, commandRate = 40,
+                 loud = False):
         '''Initialize the robot.
         
         Keyword arguments:
@@ -57,6 +58,7 @@ class Robot():
         self.nServos = nServos
 
         self.sleep = 1. / float(commandRate)
+        self.loud  = loud
 
         if self.nServos != 9:
             raise Exception('Unfortunately, the Robot class currently assumes 9 servos.')
@@ -146,7 +148,8 @@ class Robot():
         #    interpBegin = 0, interpEnd = 0):
 
 
-        print 'Starting motion.'
+        if self.loud:
+            print 'Starting motion.'
 
         self.resetClock()
         self.currentPos = self.readCurrentPosition()
@@ -184,9 +187,9 @@ class Robot():
         timeEnd   = self.time + seconds
         
         while self.time < timeEnd:
-            posS = start(seconds) if isinstance(start, FunctionType) else start
-            posE =   end(seconds) if isinstance(end,   FunctionType) else end
-            goal = lInterp(self.time, [timeStart, timeEnd], start, end)
+            posS = start(self.time) if isinstance(start, FunctionType) else start
+            posE =   end(self.time) if isinstance(end,   FunctionType) else end
+            goal = lInterp(self.time, [timeStart, timeEnd], posS, posE)
 
             self.commandPosition(goal)
             sleep(self.sleep)
@@ -265,6 +268,10 @@ class Robot():
                             % (self.nServos, repr(position)))
         
         goalPosition = self.cropPosition([int(xx) for xx in position], cropWarning)
+
+        if self.loud:
+            posstr = ', '.join(['%4d' % xx for xx in goalPosition])
+            print '%.2fs -> %s' % (self.time, posstr)
         
         for ii,actuator in enumerate(self.actuators):
             actuator.goal_position = goalPosition[ii]

### 2010-10-05 code/optimize.py

@@ -14,8 +14,9 @@ walked.
 
 """
 
-import math, pdb
+import math, pdb, sys
 import random
+from datetime import datetime
 from copy import copy
 from Robot import Robot
 from SineModel import sineModel
@@ -30,7 +31,7 @@ def initialState(ranges):
     for rang in ranges:
         # Chooses random values for each parameter (initial state)
         if isinstance(rang[0], bool):  # If range is (true, false),
-				       # choose true or false
+                       # choose true or false
             parameters.append(random.uniform(0,1) > .5)
         else:
             parameters.append(random.uniform(rang[0], rang[1]))
@@ -43,16 +44,16 @@ def neighbor(ranges, parameters):
 
     """
     ret = copy(parameters)
-    print '  ** Neighbor old', ret
+    #print '  ** Neighbor old', ret
     index = random.randint(0, len(parameters) - 1)
 
-    print ranges
+    #print ranges
     if isinstance(ranges[index][0], bool):
         ret[index] = (random.uniform(0,1) > .5)
     else:
         ret[index] = random.uniform(ranges[index][0], ranges[index][1])
 
-    print '  ** Neighbor new', ret
+    #print '  ** Neighbor new', ret
     return ret
 
 def slightNeighbor(ranges, parameters):
@@ -70,61 +71,82 @@ def slightNeighbor(ranges, parameters):
         ret[index] = (random.uniform(0,1) > .5)
     else:
         if random.randint(0, 1) == 0:  # decrease slightly
-	    ret[index] = ret[index] - (.1 * (ranges[index][1] - ranges[index][0]))
+            ret[index] = ret[index] - (.1 * (ranges[index][1] - ranges[index][0]))
         else:  # increase slightly
-	    ret[index] = ret[index] - (.1 * (ranges[index][1] - ranges[index][0]))
+            ret[index] = ret[index] - (.1 * (ranges[index][1] - ranges[index][0]))
 
     print '  ** Neighbor new', ret
     return ret
 
+def prettyVec(vec):
+    return ('[' +
+            ' '.join(['%4f' % xx if isinstance(xx,float) else repr(xx) for xx in vec]) +
+            ']')
+    
 def doRun():
     # Parameters are: amp, tau, scaleInOut, flipFB, flipLR
     
-    android = Robot()
+    android = Robot(commandRate = 40, loud = False)
     # TODO: motion = Motion()
     # TODO: Move ranges below 
-    ranges = [(0, 400),
[…]

### 2010-10-05 report/report.tex
Report copied from Google Docs, formatted for Latex, and nearly finished.

\section{Executive Summary}

We are using machine learning to design a gait for a quadruped
robot. The robot has an on-board computer, drivers for the motors, and
will have camera feedback. The code so far consists of the robot
class, optimization class, parameterized model, sine model, and camera
feedback.  Both the hardware and software are progressing on schedule.
Some promising gaits have been generated, and we anticipate a final
evaluation of the robot’s gait in November.

\section{Proposal Summary}
% A short (one or two sentence) summary of the project
% goal/topic. This is just to remind us what you’re working on.

We are using machine learning to design a gait for a quadraped robot
from Hod Lipson’s lab. We anticipate using speed and number of
failures as metrics for evaluation of the designed gait.

\section{System Architecture Summary}
% A summary of the main components of the system, and how they
% interact with each other to achieve the goal. This is just to remind
% us what you are planning to develop.

The quadraped robot has an on-board computer running Linux. The lower
level drivers are in C and we are implementing the system in
Python. Feedback about distance travelled is currently input manually,
but we eventually plan to provide this via an Optitrack positioning
system.

\section{Implementation Status}
\seclabel{implement}
% A list of the major components of the system as well as the current
% status of each (e.g. designed, implemented, tested, finished, almost
% finished with coding). For the pieces not yet finished, indicate
% tentative completion dates. The schedule doesn’t have to be
% incredibly detailed, but it should include entries for all
% substantial modules/submodules of your system. Include also who in
% the team is responsible for completing this module.

\subsection*{\code{Robot} class}

\begin{itemize}
\item \emph{Team member:} Jason.
\item \emph{Current implementation:} Class wrapper for commanding
  motion of the robot.  The \code{Robot} class takes care of the robot
  initialization, communication with the servos, and timing of the
  runs.  In addition, it prevents the servos from ever being commanded
  to a point outside their normal range (0 - 1023) as well as beyond
  points where limbs would collide with parts of the robot body.  The
  main class function, \code{run}, accepts a motion model (any
  function that takes a time argument and outputs a 9 dimensional
  position) and will run the robot using this motion model, including,
  if desired, smooth interpolation over time for the beginning and end
  of the run.
\item \emph{Future improvements:} None planned, though some will
  probably arise.
\item \emph{Status:} 90\% complete.
\end{itemize}

\subsection*{\code{Optimization} class}

\begin{itemize}
\item \emph{Team member:} Sarah.
\item \emph{Current implementation:} Creates a random neighbor by
  randomly choosing one parameter to adjust, and changing either
  completely randomly in neighbor(), or slightly in
  slightNeighbor(). Then evaluates this neighbor by running the robot
  with the newly chosen parameters. If this neighbor results in a
  longer distance walked then the previous best neighbor, we save this
  neighbor as the new best neighbor. We then repeat the whole process
  using the best neighbor as a base.
\item \emph{Future improvements:} Gradient descent.
\item \emph{Status:} Initial algorithm coded. Improved algorithm
  slightNeighbor() coded. Gradient descent algorithm to be completed
  by Sarah.
\end{itemize}

\subsection*{\code{Parameterized Motion Model} abstract base class}

\begin{itemize}
\item \emph{Team member:} Jason.
\item \emph{Current implementation:} Several functions exist, but none are within a class framework.
\item \emph{Future improvements:} Make into class.
\item \emph{Status:} 40\% complete.
\end{itemize}

\subsection*{\code{SineModel} classes}

\begin{itemize}
\item \emph{Team member:} Jason.
\item \emph{Current implementation:} Commands motors to positions
  based on a sine wave, creating a periodic pattern. Parameters are:
  amplitude, wavelength, scale inner vs outer motors, multiply
  left-right, multiply back-front, shift front-back, shift right-left.
  Currently a function, not a class.
\item \emph{Future improvements:} Make class, add time offset
  parameters.  Continued tweaking of this class and other motion
  models will be a main goal of the next month.
\item \emph{Status:} 30\% complete.
\end{itemize}

\subsection*{Other derived motion model classes}

\begin{itemize}
\item \emph{Team member:} All.
\item \emph{Current implementation:} Several versions of a SineModel
  exist, each with different parameters, though solely as functions
  (not as derived classes of \code{ParameterizedMotionModel}).
\item \emph{Future improvements:} Implementing and tweaking current
  and new motion model classes will be our primary effort over the
  next month.
\item \emph{Status:} 20\% complete.
\end{itemize}

\subsection*{\code{CameraFeedback} class and hardware}

\begin{itemize}
\item \emph{Team member:} Diana.
\item \emph{Current implementation:} No software, still configuring
  camera hardware.
\item \emph{Future improvements:} Write python wrapper class, get
  positional information from the Optitrack camera system to the program
\item \emph{Status:} 20\% complete, goal to have cameras working by October 8.
\end{itemize}

\section{Evaluation status}
% A list of the major steps in your planned evaluation of the system
% and the status of each. For some projects it may be easier to merge
% this list with the component/module status list above of your
% system. Include also who in the team is responsible for completing
% this module.

After running our program and finding a good gait, we will then
evaluate our system in two ways: the time taken to walk a certain
distance and the number of failures (i.e. falling down, getting
stuck). Both of these evaluation metrics depend on the result of
running our program for many iterations. Neither have been
completed. The evaluation for our particular project is an ongoing
process, but we anticipate a final evaluation taking place
early-to-mid November. The evaluation process will be completed by
Jason, Sarah, and Diana.

\section{Project schedule timeline}
% Tell us if you are meeting your project timeline or you are
% exceeding it or falling behind. If you are falling behind, explain
% how you will catch up or adjust the goals.

We are meeting our project timeline and are, in fact, slightly ahead
of schedule. We have read the relevant papers, gotten lab access, and
talked with the appropriate people to learn the background information
necessary for working with the robot. We learned how to make the robot
move and have implemented an initial parameter optimization algorithm,
which has worked better than we anticipated.

\section{Schedule}

Below is a rough schedule of fixed external deadlines and anticipated
team milestones.

\definecolor{doneColor}{rgb}{0.0,0.7,0.0}
\newcommand{\done}[1]{{\color{doneColor}\sout{#1}}}
%\newcommand{\done}[1]{#1}

\begin{center}
\begin{tabular}{|l||p{2.5in}|p{1.5in}|}
\hline
                        & {\bf Milestones} & {\bf Deadlines}
\\ \hline

Week 1 (9/13-9/19)    & \done{Read papers, get lab access, talk to relevant other researchers} & 9/17 Final proposals due
\\ \hline

Week 2 (9/20-9/26)    & \done{Continue reading, get robot to move} &
\\ \hline

Week 3 (9/27-10/03)   & \done{Implement parametrized gait and determine proposed coding schedule for more advanced algorithms in time for Code Review \#1.} &
\\ \hline

Week 4 (10/04-10/10)  & \done{Begin main algorithm dev/testing effort} & 10/5 Code Review \#1
\\ \hline

Week 5 (10/11-10/17)  & Algorithm dev/testing &
\\ \hline

Week 6 (10/18-10/24)  & Algorithm dev/testing &
\\ \hline

Week 7 (10/25-10/31)  & Algorithm dev/testing, quantify/solidify current results for Code Review \#2 &
\\ \hline

Week 8 (11/1-11/7)    & Finish collecting results, begin writing & 11/2 Code Review \#2
\\ \hline

Week 9 (11/8-11/14)   & Finish collecting results, writing &
\\ \hline

Week 10 (11/15-11/21) & Finish collecting results, writing, get final demo ready &
\\ \hline

Week 11 (11/22-11/28) & Finish collecting results, writing, get final demo ready  &
\\ \hline

Week 12 (11/29-11/30) & Final demo  & 11/30 Final presentation
\\ \hline
\end{tabular}
\end{center}

\section{Activity log (Optional)}

% an informal log your efforts. If you’ve been keeping notes on your
% progress, feel free to turn these in as well. These might include:
% (1) notes on what aspect of the project you worked on and when, (2)
% what was accomplished in each “session”, (3) design questions, (4)
% implementation questions, (5) descriptions of experiments and
% results, (6) problems that arose, (7) anything else that you think
% is relevant to the project.

\begin{itemize}
\item \emph{9/13-9/19:} Set up. Set up trac, worked on pre-proposal
  and presentation, got lab access, met with Jim Torresen, contacted
  Juan Cristobal Zagal, read papers, got the robot to walk using
  previously coded program. Questions -- How do we move the robot? How
  do we command the motors?
\item \emph{9/20-9/26:} Moving the robot. Installed Linux on the
  robot, figured out how to command the motors, coded a program to
  make the robot stand up, coded a simple motion pattern. Questions --
  For initial program, what parameters and model should be used? How
  do we limit the search space? What algorithm should we start with?
  How will the robot get information about how far it moved?
\item \emph{9/27-10/3:} Initial program. Chose several parameters and
  created a simple parameterized sine model, implemented a random hill
  climbing algorithm to optimize the parameters, ran some trial
  iterations, began setting up the cameras for feedback. Questions --
  Did the initial program work? What parameters are important? Are
  additional parameters necessary? What algorithm should we implement
  for the final program?
\item \emph{10/4-10/5:} Plan for final implementation. Discussed
  possible algorithms to use and new parameters to add. Worked on
  report for code review \#1. Questions -- Advantages for gradient
  descent? Should we start with a completely random parameter vector
  or should we start with one that results in motion and let the
  program optimize it?
\end{itemize}

\section{Proposal Modifications}
% Let us know if you are deviating from your original proposal.

We are not deviating from our original proposal.

\section{Appendix}
% Describe the code files you uploaded into CMS

A brief description of the code uploaded to the CMS follows:

\begin{itemize}
\item \code{optimize.py}: Employs random hill-climbing to choose and
  evaluate the parameters of the robots’ motion, either by changing
  one parameter completely randomly, or changing one parameter
  slightly. Evaluates each neighbor using manually input distance
  walked.
\item \code{Robot.py}: Implements the \code{Robot} class, described in
  \secref{implement}.
\item \code{SineModel.py}: Implements a sine based motion model,
  described in \secref{implement}.
\item \code{Motion.py}: Motion helper functions.
\end{itemize}

### 2010-11-02 code/plot/load_run_data.m
ploting scripts

@@ -1,0 +1,16 @@
+function [N,iteration,parameters,fitness,best] = load_run_data(filename)
+  % Load raw data from file
+  %data = dlmread(filename);
+  data = load(filename);
+
+  % Separate into variables
+  N          = size(data,1);
+  iteration  = 1:N;
+  parameters = data(:,1:5);
+  fitness    = data(:,6);  
+  best = zeros(50,1);
+  best(1) = fitness(1);
+  for ii=2:N
+    best(ii) = max(best(ii-1), fitness(ii));
+  end
+end

### 2010-11-02 code/plot/plot_all.m

@@ -1,0 +1,15 @@
+plot_run('log_uniform_a.txt');
+plot_run('log_uniform_b.txt');
+plot_run('log_uniform_c.txt');
+plot_run('log_gaussian_a.txt');
+plot_run('log_gaussian_b.txt');
+plot_run('log_gaussian_c.txt');
+
+plot_several_runs({'log_uniform_a.txt', 'log_uniform_b.txt', 'log_uniform_c.txt'})
+plot_several_runs({'log_gaussian_a.txt', 'log_gaussian_b.txt', 'log_gaussian_c.txt'})
+
+plot_several_runs({'log_uniform_a.txt', 'log_gaussian_a.txt'});
+plot_several_runs({'log_uniform_b.txt', 'log_gaussian_b.txt'});
+plot_several_runs({'log_uniform_c.txt', 'log_gaussian_c.txt'});
+
+

### 2010-11-02 code/plot/plot_run.m

@@ -1,15 +1,20 @@
 function plot_run(filename)
   % Plots the fitness over a run from a log file
 
-  data = dlmread(filename);
-  best = zeros(50,1);
-  for i=2:50
-    best(i,1) = max(best(i-1,1), data(i,6));
-  end
-  idx = find(data(:,6) == best(:, 1));
-  plot(data(:,6), '-o', 'MarkerSize', 5);
-  hold on;
+  [N,iteration,parameters,fitness,best] = load_run_data(filename);
+
+  % Plot
+  figure(1); clf; hold on;
+  plot(iteration, fitness, '-o', 'MarkerSize', 5);
   plot(best(:,1), 'r');
-  plot(idx, best(idx,1), 'ro', 'MarkerSize', 5);
-  title(filename);
+  idx = find(fitness == best);
+  plot(idx, best(idx), 'ro', 'MarkerSize', 5);
+  title(filename, 'Interpreter', 'none');
+  legend(filename);
+
+  % Save
+  print('-dpng', strcat(filename, '.png'));
+  print('-depsc', strcat(filename, '.eps'));
+
+end
 

### 2010-11-02 code/Strategy.py
Learning strategy skeleton code

@@ -262,6 +262,36 @@ class GradientSampleStrategy(Strategy):
         nextState = [center, adjustment]
         return [sum(value) for value in zip(*nextState)]
 
+
+
+class LearningStrategy(Strategy):
+    '''
+    A strategy that uses supervised learning to guess which parameter vector woudl be good to try next.
+    '''
+
+    def __init__(self, *args, **kwargs):
+        super(LearningStrategy, self).__init__(*args, **kwargs)
+        self.X = []
+        self.y = []
+
+    def getNext(self, ranges):
+        '''Learn model on X and y...'''
+
+        # 1. Learn
+        # 2. Try some nearby values
+        # 3. Pick best one
+
+
+    def updateResults(self, dist, ranges):
+        '''This must be called for the last point that was handed out!'''
+
+        # about the same...
+        self.triedSoFar.append(self.stillToTry.pop(0))
+        self.triedSoFar[-1].append(dist)
+        print '        Got update, policy is now', self.triedSoFar[-1]
+
+
+
 if __name__ == "__main__":
     import doctest
     doctest.testmod()

### 2010-11-02 report/report.tex
Edited report.

@@
 \section{Executive Summary}
-
+% A few sentences summarizing each of the items below
 We are using machine learning to design a gait for a quadruped
@@
 We are using machine learning to design a gait for a quadraped robot
-from Hod Lipson’s lab. We anticipate using speed and number of
-failures as metrics for evaluation of the designed gait.
+from Hod Lipson’s lab. The metric for evaluation of the designed gait
+is speed.
 
@@
 level drivers are in C and we are implementing the system in
-Python. Feedback about distance travelled is currently input manually,
-but we eventually plan to provide this via an Optitrack positioning
-system.
+Python. Feedback about distance travelled is provided via a Wii remote.
 
@@
 
-\subsection*{\code{Optimization} class}
+\subsection*{\code{Strategy} and \code{RunManager} class}
 
 \begin{itemize}
-\item \emph{Team member:} Sarah.
-\item \emph{Current implementation:} Creates a random neighbor by
+\item \emph{Team member:} Jason and Sarah.
+\item \emph{Current implementation:} The user has a choice between four
+  different learning strategies: uniform random hill climbing, gaussian random
+  hill climbing, and N-dimensional policy gradient algorithm descent.
+  \begin{itemize} 
+  \item \emph{Uniform random hill climbing:} Creates a random neighbor by
   randomly choosing one parameter to adjust, and changing either
-  completely randomly in neighbor(), or slightly in
-  slightNeighbor(). Then evaluates this neighbor by running the robot
+  completely randomly in UniformStrategy(), or slightly in
+  UniformSlightStrategy(). Then evaluates this neighbor by running the robot
   with the newly chosen parameters. If this neighbor results in a
@@
   using the best neighbor as a base.
-\item \emph{Future improvements:} Gradient descent.
-\item \emph{Status:} Initial algorithm coded. Improved algorithm
-  slightNeighbor() coded. Gradient descent algorithm to be completed
-  by Sarah.
+  \item \emph{Gaussian random hill climbing} Creates a random neighbor byi
+  changing each parameter randomly based on a normal distribution. Then
+  evaluates this neighbor similarly to uniform random hill climbing.
+  \item \emph{N-dimensional policy gradient descent} Estimates the policy
+  gradient by evaluating \emph{t} randomly generated policies
+  near an initial parameter vector. Then computes the average score 
+  \end{itemize}
+\item \emph{Future improvements:} Machine learning of objective function.
+\item \emph{Status:} Described algorithms coded. Machine learning of objective
+  function to be coded by Jason.
 \end{itemize}
@@
 
-\subsection*{\code{CameraFeedback} class and hardware}
+\subsection*{\code{WiiTrackClient} and \code{WiiTrackServer} classes
+ and hardware}
 
 \begin{itemize}
-\item \emph{Team member:} Diana.
-\item \emph{Current implementation:} No software, still configuring
-  camera hardware.
-\item \emph{Future improvements:} Write python wrapper class, get
-  positional information from the Optitrack camera system to the program
-\item \emph{Status:} 20\% complete, goal to have cameras working by October 8.
+\item \emph{Team member:} Jason and Diana.
+\item \emph{Current implementation:} A wii remote keeps track of the
+position of the robot by tracking an infrared light on the robot. 
+There is a motion loop that uses own position...(Jason and Diana, write
+more about this.
+\item \emph{Status:} Complete.
 \end{itemize}
@@
 
-After running our program and finding a good gait, we will then
-evaluate our system in two ways: the time taken to walk a certain
-distance and the number of failures (i.e. falling down, getting
-stuck). Both of these evaluation metrics depend on the result of
-running our program for many iterations. Neither have been
-completed. The evaluation for our particular project is an ongoing
+\begin{itemize}
+\item After running our program and finding a good gait, we will then
+evaluate our system by the distance walked in a certain amount
+of time. This evaluation metric depend on the result of
+running our program for many iterations. We have done 50 runs of 
+3 different inital parameter vectors for random hill climbing, 
+gaussian random hill climbing, and policy gradient descent.
+The evaluation for our particular project is an ongoing
 process, but we anticipate a final evaluation taking place
@@
 Jason, Sarah, and Diana.
+
+\item We also coded \code{explore\_dimensions} to collect data for dimension
+varying plots. We intend to do further explore this by running
+\code{explore\_dimension} around a parameter that produces a medium-good gait.
+\end{itemize}
 
@@
 
-We are meeting our project timeline and are, in fact, slightly ahead
-of schedule. We have read the relevant papers, gotten lab access, and
-talked with the appropriate people to learn the background information
-necessary for working with the robot. We learned how to make the robot
-move and have implemented an initial parameter optimization algorithm,
-which has worked better than we anticipated.
+We are meeting our project timeline. We have developed two new algorithms,
+gaussian hill climbing and policy gradient descent and tested all three of
+our major algorithms on the robot. We have also quantified our current results. 
+In the next month we hope to develop one more learning strategy, machine
+learning of an objective function.
 
@@
 
-Week 5 (10/11-10/17)  & Algorithm dev/testing &
+Week 5 (10/11-10/17)  & \done{Algorithm dev/testing} &
 \\ \hline
 
-Week 6 (10/18-10/24)  & Algorithm dev/testing &
+Week 6 (10/18-10/24)  & \done{Algorithm dev/testing} &
 \\ \hline
 
-Week 7 (10/25-10/31)  & Algorithm dev/testing, quantify/solidify current results for Code Review \#2 &
+Week 7 (10/25-10/31)  & \done{Algorithm dev/testing, quantify/solidify current results for Code Review \#2} &
 \\ \hline
 
-Week 8 (11/1-11/7)    & Finish collecting results, begin writing & 11/2 Code Review \#2
+Week 8 (11/1-11/7)    & \done{Finish collecting results, begin writing} & 11/2 Code Review \#2
 \\ \hline
 
-Week 9 (11/8-11/14)   & Finish collecting results, writing &
+Week 9 (11/8-11/14)   & Algorithm dev/testing &
 \\ \hline
@@
 \begin{itemize}
-\item \code{optimize.py}: Employs random hill-climbing to choose and
-  evaluate the parameters of the robots’ motion, either by changing
-  one parameter completely randomly, or changing one parameter
-  slightly. Evaluates each neighbor using manually input distance
-  walked.
+\item \code{optimize.py}: Determines a strategy to try and runs the robot
+  with that strategy.
+\item \code{RunManager.py}: Deals with all the details of running the robot,
+  including choosing an initial parameter, running the robot multiple times, 
+  tracking distance walked, and writing to the log file. Also includes the
+  \code{explore\_dimensions} method.
+\item \code{Strategy.py}: Contains all the different possible strategies, which
+  will be passed as objects in \code{optimize.py}. 
 \item \code{Robot.py}: Implements the \code{Robot} class, described in
@@
 \item \code{Motion.py}: Motion helper functions.
+\item \code{WiiTrackServer.py, WiiTrackClient.py}: ? Jason, write this.
 \end{itemize}

### 2010-11-02 report/report.tex
edited report more.

@@
 \item \emph{Team member:} Jason and Sarah.
-\item \emph{Current implementation:} The user has a choice between four
+\item \emph{Current implementation:} The user has a choice between three
   different learning strategies: uniform random hill climbing, gaussian random
@@
 \item \emph{Team member:} Jason and Diana.
-\item \emph{Current implementation:} A wii remote keeps track of the
-position of the robot by tracking an infrared light on the robot. 
-There is a motion loop that uses own position...(Jason and Diana, write
-more about this.
+\item \emph{Current implementation:} A Wii remote tracks the location of the robot through an infrared LED mounted on top of the robot.
+We used CWiid library which is able to interface with the remote via bluetooth to determine the location of the infrared LED.
+This information is accessed through some functions we wrote and passed to the program.
+The program gets the robot's position at the beginning of each run and then again at the end.
+The program then calculates the net change in position and uses that as the distance walked.
 \item \emph{Status:} Complete.
@@
   for the final program?
-\item \emph{10/4-10/5:} Plan for final implementation. Discussed
+\item \emph{10/4-10/10:} Plan for final implementation. Discussed
   possible algorithms to use and new parameters to add. Worked on
-  report for code review \#1. Questions -- Advantages for gradient
+  report for code review. Discussed feedback from code review.  Questions -- Advantages for gradient
   descent? Should we start with a completely random parameter vector
   or should we start with one that results in motion and let the
-  program optimize it?
+  program optimize it? How does the camera system work?
+\item \emph{10/11-10/17:} Discussed how we were going to proceed with the project and set up a more detailed schedule.
+Got the cameras working and calibrated them.
+Researched the 3 available streaming protocols.
+Questions -- Which streaming protocol would work best?
+Which improvements should we start with?
+\item \emph{10/18-10/24:} Created a variation on the random hill climbing algorithm to choose the next neighbor based on a Gaussian distribution.
+Tried to work with two of the streaming protocols for the camera system, but were unsuccessful.
+Decided to use a Wii remote for positional information instead.
+Wrote the integration code for the Wii remote.
+Refactored our code to separate the different algorithms.
+Began planning the algorithm for a modified gradient descent.
+Questions -- What values should we use in the gradient descent algorithm?
+How much should the each parameter be changed?
+How many neighbors should be sampled before choosing the next state?
+How can we compare the results of our different algorithms?
+How repeatable is the distance traveled for a given parameter vector?
+\item \emph{10/25-11/2}: Created a modified gradient descent algorithm.
+Ran 150 trials of the three algorithms.
+Each algorithm started with a specific parameter vector and ran for 50 iterations.
+Created and ran a program to get a feel for the overall look of the search space.
+Ran some trials to see how reproducible the distance traveled was.
+Prepared for code review 2.
+Questions -- What does the search space look like?
+Are there a lot of spikes, or is it fairly smooth?
+Based on the results of the algorithms, what conclusions can we draw?
 \end{itemize}
@@
 \item \code{Motion.py}: Motion helper functions.
-\item \code{WiiTrackServer.py, WiiTrackClient.py}: ? Jason, write this.
+\item \code{WiiTrackServer.py}: Broadcasts the position of the infrared LED>
+\item \code{WiiTrackClient.py}: Connects to the WiiTrackServer to get the current position information.
 \end{itemize}

### 2010-11-02 code/util.py
util

@@ -1,0 +1,8 @@
+#! /usr/bin/env python
+
+'''Just a few utility functions'''
+
+def prettyVec(vec):
+    return ('[' +
+            ' '.join(['%4f' % xx if isinstance(xx,float) else repr(xx) for xx in vec]) +
+            ']')

### 2010-12-01 report/report.tex
Added 1.title,authors, and affiliations, and 2. abstract

@@
+\section{Abstract}
+% A brief summary of the report. This is not an introduction -
+% it should be complete. Include 1-2 sentences for each of the items
+% below, up to and including conclusions.
+This paper presents several approaches to optimizing a quadrupedal
+trot gait for forward speed. Given a parameterized walk designed for
+a quadruped robot designed and printed in Hod Lipson's lab, 
+we implement, test, and compare different learning strategies, 
+including uniform and Gaussian random hill climbing, a form of policy 
+gradient reinforcement learning\cite{1}, Simplex (Nelder-Mead), 
+linear regression, SVM regression, and an evolutionary neural 
+network (HyperNEAT)\cite{2}. Because the fastest learned walk was not
+significantly faster than the fastest randomly generated walk, we
+conjecture that the motion representation for the robot is more
+integral to forward speed than the learning algorithm.
+
 \section{Executive Summary}

### 2010-12-01 report/report.tex
Edited methods, system architecture & implementation, added related work section

@@
 trot gait for forward speed. Given a parameterized walk designed for
-a quadruped robot designed and printed in Hod Lipson's lab, 
-we implement, test, and compare different learning strategies, 
+a quadruped robot designed and printed in the Cornell Computational
+Synthesis Lab, we implement, test, and compare different learning strategies, 
 including uniform and Gaussian random hill climbing, a form of policy 
@@
 
-\section{Executive Summary}
-% A few sentences summarizing each of the items below
+Keywords: Learning Control, Walking Robots, Multi Legged Robots
+
+\section{Introduction}
+% Motivate and abstractly describe the problem you are addressing and
+% how you are addressing it. What is the problem? Why is it important?
+% What is your basic approach? A short discussion of how it fits into
+% related work in the area is also desirable. Summarize the basic results
+% and conclusions that you will present.
 We are using machine learning to design a gait for a quadruped
@@
 
-\section{Proposal Summary}
-% A short (one or two sentence) summary of the project
-% goal/topic. This is just to remind us what you’re working on.
+\section{Problem definition}
+% Precisely define the problem you are addressing (i.e. formally specify
+% the inputs and outputs). Elaborate on why this is an interesting and
+% important problem.
+We are testing several different learning methods to design a gait for 
+a quadraped robot from the Cornell Computational Synthesis Lab. The metric 
+for evaluation of the designed gait is speed. A comparison and evaluation 
+of the many different methods available for optimizing the gait of legged 
+robots will be useful for future work on this challenging multidimensional 
+control problem.
 
-We are using machine learning to design a gait for a quadraped robot
-from Hod Lipson’s lab. The metric for evaluation of the designed gait
-is speed.
+\section{Method}
+% Describe in reasonable detail the algorithm you are using to address this 
+% problem. A pseudo-code description of the algorithm you are using is 
+% frequently useful. If it makes sense for your project, trace through a 
+% concrete example, showing how your algorithm processes this example. The 
+% example should be complex enough to illustrate all of the important aspects 
+% of the problem but simple enough to be easily understood. If possible, 
+% an intuitively meaningful example is better than one with meaningless 
+% symbols.
+We implemented and tested 8 different learning strategies:
+\begin{itemize}
+\item \emph{Uniform random hill climbing}: Creates a random neighbor by
+  randomly choosing one parameter to adjust, and changing it completely 
+  randomly in UniformStrategy(). Then evaluates this neighbor by running 
+  the robot with the newly chosen parameters. If this neighbor results in a
+  longer distance walked then the previous best neighbor, we save this
+  neighbor as the new best neighbor. We then repeat the whole process
+  using the best neighbor as a base.
+\item \emph{Gaussian random hill climbing}: Creates a random neighbor by
+  changing each parameter randomly based on a normal distribution. Then
+  evaluates this neighbor similarly to uniform random hill climbing.
+\item \emph{N-dimensional policy gradient descent}: Estimates the policy
+  gradient by evaluating \emph{t} randomly generated policies
+  near an initial parameter vector. Then computes the average score for
+  each parameter and adjusts the base policy according to the estimated
+  gradient.
+\item \emph{Random}: Randomly generates policies in the range of the motion
+representation.
+\item \emph{Simplex (Nelder-Mead) Method}
+\item \emph{Linear regression}: To initialize, chooses five random points
+and trains on them using least-squares. Then in a loop, takes a fixed-size
+step in the direction of the gradient, and retrains on all points so far.
+\item \emph{SVM regression}
+\item \emph{Evolutionary Neural Network (HYPERNEAT)\cite{2}}
+\end{itemize}
 
-\section{System Architecture Summary}
-% A summary of the main components of the system, and how they
-% interact with each other to achieve the goal. This is just to remind
-% us what you are planning to develop.
+\section{Related work}
+% Briefly explain who else worked on related problems in the past and what
+% methods they used. Explain if you are using similar methods, or if your
+% approach is different and if so - how (either is ok).
+
+\section{System Architecture and Implementation}
+% Describe how you implemented your system and how you structured it. 
+% This should give an overview of the system, not a detailed 
+% documentation of the code. The documentation of the code is part of 
+% the code you hand in. You might want to comment on high-level design 
+% decisions that you made. Also explain how you obtained your
+% data and any pre-processing you did to it.
 
@@
 
-\section{Implementation Status}
-\seclabel{implement}
-% A list of the major components of the system as well as the current
-% status of each (e.g. designed, implemented, tested, finished, almost
-% finished with coding). For the pieces not yet finished, indicate
-% tentative completion dates. The schedule doesn’t have to be
-% incredibly detailed, but it should include entries for all
-% substantial modules/submodules of your system. Include also who in
-% the team is responsible for completing this module.
-
-\subsection*{\code{Robot} class}
-
 \begin{itemize}
-\item \emph{Team member:} Jason.
-\item \emph{Current implementation:} Class wrapper for commanding
+\item \code{Robot} class: Class wrapper for commanding
   motion of the robot.  The \code{Robot} class takes care of the robot
@@
   of the run.
-\item \emph{Future improvements:} None planned, though some will
-  probably arise.
-\item \emph{Status:} 90\% complete.
-\end{itemize}
 
-\subsection*{\code{Strategy} and \code{RunManager} class}
+\item \code{Strategy} and \code{RunManager} class: The user has a choice between seven
+  different learning strategies: uniform random hill climbing, Gaussian random
+  hill climbing, N-dimensional policy gradient descent, random, 
+  simplex, linear regression/prediction, and SVM regression/prediction.
 
-\begin{itemize}
-\item \emph{Team member:} Jason and Sarah.
-\item \emph{Current implementation:} The user has a choice between three
-  different learning strategies: uniform random hill climbing, gaussian random
-  hill climbing, and N-dimensional policy gradient algorithm descent.
-  \begin{itemize} 
-  \item \emph{Uniform random hill climbing:} Creates a random neighbor by
-  randomly choosing one parameter to adjust, and changing either
-  completely randomly in UniformStrategy(), or slightly in
-  UniformSlightStrategy(). Then evaluates this neighbor by running the robot
-  with the newly chosen parameters. If this neighbor results in a
-  longer distance walked then the previous best neighbor, we save this
-  neighbor as the new best neighbor. We then repeat the whole process
-  using the best neighbor as a base.
-  \item \emph{Gaussian random hill climbing} Creates a random neighbor byi
-  changing each parameter randomly based on a normal distribution. Then
-  evaluates this neighbor similarly to uniform random hill climbing.
-  \item \emph{N-dimensional policy gradient descent} Estimates the policy
-  gradient by evaluating \emph{t} randomly generated policies
-  near an initial parameter vector. Then computes the average score 
-  \end{itemize}
-\item \emph{Future improvements:} Machine learning of objective function.
-\item \emph{Status:} Described algorithms coded. Machine learning of objective
-  function to be coded by Jason.
-\end{itemize}
+\item \code{Parameterized Motion Model} abstract base class: Several functions exist, but none are within a class framework.
 
-\subsection*{\code{Parameterized Motion Model} abstract base class}
+\item \code{SineModel} classes: Commands motors to positions
+  based on a sine wave, creating a periodic pattern. Parameters are:
+  amplitude, wavelength, scale inner vs outer motors, scale left vs right
+  motors, scale back vs front motors. Currently a function, not a class.
 
-\begin{itemize}
-\item \emph{Team member:} Jason.
-\item \emph{Current implementation:} Several functions exist, but none are within a class framework.
-\item \emph{Future improvements:} Make into class.
-\item \emph{Status:} 40\% complete.
-\end{itemize}
+\item Other derived motion model classes: Several versions of a SineModel
+  exist, each with different parameters, though solely as functions.
 
-\subsection*{\code{SineModel} classes}
-
-\begin{itemize}
-\item \emph{Team member:} Jason.
-\item \emph{Current implementation:} Commands motors to positions
-  based on a sine wave, creating a periodic pattern. Parameters are:
-  amplitude, wavelength, scale inner vs outer motors, multiply
-  left-right, multiply back-front, shift front-back, shift right-left.
-  Currently a function, not a class.
-\item \emph{Future improvements:} Make class, add time offset
-  parameters.  Continued tweaking of this class and other motion
-  models will be a main goal of the next month.
-\item \emph{Status:} 30\% complete.
-\end{itemize}
-
-\subsection*{Other derived motion model classes}
-
-\begin{itemize}
-\item \emph{Team member:} All.
-\item \emph{Current implementation:} Several versions of a SineModel
-  exist, each with different parameters, though solely as functions
-  (not as derived classes of \code{ParameterizedMotionModel}).
-\item \emph{Future improvements:} Implementing and tweaking current
-  and new motion model classes will be our primary effort over the
-  next month.
-\item \emph{Status:} 20\% complete.
-\end{itemize}
-
-\subsection*{\code{WiiTrackClient} and \code{WiiTrackServer} classes
- and hardware}
-
-\begin{itemize}
-\item \emph{Team member:} Jason and Diana.
-\item \emph{Current implementation:} A Wii remote tracks the location of the robot through an infrared LED mounted on top of the robot.
+\item \code{WiiTrackClient} and \code{WiiTrackServer} classes
+ and hardware: A Wii remote tracks the location of the robot through an infrared LED mounted on top of the robot.
 We used CWiid library which is able to interface with the remote via bluetooth to determine the location of the infrared LED.
@@
 The program then calculates the net change in position and uses that as the distance walked.
-\item \emph{Status:} Complete.
+
 \end{itemize}

### 2010-12-01 report/report.tex
Edited report more...

@@
 
-\section{Evaluation status}
-% A list of the major steps in your planned evaluation of the system
-% and the status of each. For some projects it may be easier to merge
-% this list with the component/module status list above of your
-% system. Include also who in the team is responsible for completing
-% this module.
+\section{Experimental (or Theoretical) Evaluation}
 
+\subsection{Methodology}
+% What are the criteria you are using to evaluate your method? What
+% specific hypotheses does your experiment test? Describe the 
+% experimental methodology that you used. What are the dependent and 
+% independent variables? For projects in machine learning, what is 
+% the training/test data that was used, and why is it realistic or
+% interesting? Exactly what performance data did you collect and how 
+% are you presenting and analyzing it? Comparisons to competing methods 
+% that address the same problem or to variations of your own algorithm 
+% are particularly useful.
 \begin{itemize}
-\item After running our program and finding a good gait, we will then
-evaluate our system by the distance walked in a certain amount
-of time. This evaluation metric depend on the result of
-running our program for many iterations. We have done 50 runs of 
+\item Each algorithm was run on 3 different initial parameter vectors on the
+physical robot. We stop each run after plateauing results (no improvement 
+for half the policies seen so far). We evaluate each method based on the
+amount of improvement seen from the initial parameter vectors, and on the
+fastest speeds achieved during runs.
+\end{itemize}
+
+\subsection{Results}
+% Present the quantitative results of your experiments. Graphical data
+% presentation such as graphs and histograms are frequently better 
+% than tables. Explain what are the basic differences revealed in the 
+% data. If your method is stochastic or you analyze multiple datasets, 
+% calculate mean and standard error for every quantitative
+% performance result you show based on multiple runs.
+\begin{itemize}
+\item We have done runs of 
 3 different inital parameter vectors for random hill climbing, 
-gaussian random hill climbing, and policy gradient descent.
-The evaluation for our particular project is an ongoing
-process, but we anticipate a final evaluation taking place
-early-to-mid November. The evaluation process will be completed by
-Jason, Sarah, and Diana.
+gaussian random hill climbing, policy gradient descent, random,
+simplex, and linear regression. We developed several
+gaits that were about 4 times faster than the original hand-coded gait.
 
@@
 
-\section{Project schedule timeline}
-% Tell us if you are meeting your project timeline or you are
-% exceeding it or falling behind. If you are falling behind, explain
-% how you will catch up or adjust the goals.
+\subsection{Discussion}
+% Is your hypothesis supported? What conclusions do the results support
+% about the strengths and weaknesses of your method compared to other 
+% methods? How can the results be explained in terms of the underlying 
+% properties of the algorithm and/or the data.
 
-We are meeting our project timeline. We have developed two new algorithms,
-gaussian hill climbing and policy gradient descent and tested all three of
-our major algorithms on the robot. We have also quantified our current results. 
-In the next month we hope to develop one more learning strategy, machine
-learning of an objective function.
-
-\section{Schedule}
-
-Below is a rough schedule of fixed external deadlines and anticipated
-team milestones.
-
-\definecolor{doneColor}{rgb}{0.0,0.7,0.0}
-\newcommand{\done}[1]{{\color{doneColor}\sout{#1}}}
-%\newcommand{\done}[1]{#1}
-
-\begin{center}
-\begin{tabular}{|l||p{2.5in}|p{1.5in}|}
-\hline
-                        & {\bf Milestones} & {\bf Deadlines}
-\\ \hline
-
-Week 1 (9/13-9/19)    & \done{Read papers, get lab access, talk to relevant other researchers} & 9/17 Final proposals due
-\\ \hline
-
-Week 2 (9/20-9/26)    & \done{Continue reading, get robot to move} &
-\\ \hline
-
-Week 3 (9/27-10/03)   & \done{Implement parametrized gait and determine proposed coding schedule for more advanced algorithms in time for Code Review \#1.} &
-\\ \hline
-
-Week 4 (10/04-10/10)  & \done{Begin main algorithm dev/testing effort} & 10/5 Code Review \#1
-\\ \hline
-
-Week 5 (10/11-10/17)  & \done{Algorithm dev/testing} &
-\\ \hline
-
-Week 6 (10/18-10/24)  & \done{Algorithm dev/testing} &
-\\ \hline
-
-Week 7 (10/25-10/31)  & \done{Algorithm dev/testing, quantify/solidify current results for Code Review \#2} &
-\\ \hline
-
-Week 8 (11/1-11/7)    & \done{Finish collecting results, begin writing} & 11/2 Code Review \#2
-\\ \hline
-
-Week 9 (11/8-11/14)   & Algorithm dev/testing &
-\\ \hline
-
-Week 10 (11/15-11/21) & Finish collecting results, writing, get final demo ready &
-\\ \hline
-
-Week 11 (11/22-11/28) & Finish collecting results, writing, get final demo ready  &
-\\ \hline
-
-Week 12 (11/29-11/30) & Final demo  & 11/30 Final presentation
-\\ \hline
-\end{tabular}
-\end{center}
-
-\section{Activity log (Optional)}
-
-% an informal log your efforts. If you’ve been keeping notes on your
-% progress, feel free to turn these in as well. These might include:
-% (1) notes on what aspect of the project you worked on and when, (2)
-% what was accomplished in each “session”, (3) design questions, (4)
-% implementation questions, (5) descriptions of experiments and
-% results, (6) problems that arose, (7) anything else that you think
-% is relevant to the project.
-
+\section{Future work}
+% What are the major shortcomings of your current method? For each 
+% shortcoming, propose additions or enhancements that would help overcome it.
 \begin{itemize}
-\item \emph{9/13-9/19:} Set up. Set up trac, worked on pre-proposal
-  and presentation, got lab access, met with Jim Torresen, contacted
-  Juan Cristobal Zagal, read papers, got the robot to walk using
-  previously coded program. Questions -- How do we move the robot? How
-  do we command the motors?
-\item \emph{9/20-9/26:} Moving the robot. Installed Linux on the
-  robot, figured out how to command the motors, coded a program to
-  make the robot stand up, coded a simple motion pattern. Questions --
-  For initial program, what parameters and model should be used? How
-  do we limit the search space? What algorithm should we start with?
-  How will the robot get information about how far it moved?
-\item \emph{9/27-10/3:} Initial program. Chose several parameters and
-  created a simple parameterized sine model, implemented a random hill
-  climbing algorithm to optimize the parameters, ran some trial
-  iterations, began setting up the cameras for feedback. Questions --
-  Did the initial program work? What parameters are important? Are
-  additional parameters necessary? What algorithm should we implement
-  for the final program?
-\item \emph{10/4-10/10:} Plan for final implementation. Discussed
-  possible algorithms to use and new parameters to add. Worked on
-  report for code review. Discussed feedback from code review.  Questions -- Advantages for gradient
-  descent? Should we start with a completely random parameter vector
-  or should we start with one that results in motion and let the
-  program optimize it? How does the camera system work?
-\item \emph{10/11-10/17:} Discussed how we were going to proceed with the project and set up a more detailed schedule.
-Got the cameras working and calibrated them.
-Researched the 3 available streaming protocols.
-Questions -- Which streaming protocol would work best?
-Which improvements should we start with?
-\item \emph{10/18-10/24:} Created a variation on the random hill climbing algorithm to choose the next neighbor based on a Gaussian distribution.
-Tried to work with two of the streaming protocols for the camera system, but were unsuccessful.
-Decided to use a Wii remote for positional information instead.
-Wrote the integration code for the Wii remote.
-Refactored our code to separate the different algorithms.
-Began planning the algorithm for a modified gradient descent.
-Questions -- What values should we use in the gradient descent algorithm?
-How much should the each parameter be changed?
-How many neighbors should be sampled before choosing the next state?
-How can we compare the results of our different algorithms?
-How repeatable is the distance traveled for a given parameter vector?
-\item \emph{10/25-11/2}: Created a modified gradient descent algorithm.
-Ran 150 trials of the three algorithms.
-Each algorithm started with a specific parameter vector and ran for 50 iterations.
-Created and ran a program to get a feel for the overall look of the search space.
-Ran some trials to see how reproducible the distance traveled was.
-Prepared for code review 2.
-Questions -- What does the search space look like?
-Are there a lot of spikes, or is it fairly smooth?
-Based on the results of the algorithms, what conclusions can we draw?
+\item Different motion representations
+\item Evolutionary algorithms/HyperNEAT\cite{2}
+\item Learning how to turn
+\item Open sourcing code for others to extend
 \end{itemize}
 
-\section{Proposal Modifications}
-% Let us know if you are deviating from your original proposal.
+\section{Conclusion}
+% Briefly summarize the important results and conclusions presented in 
+% the paper. What are the most important points illustrated by your 
+% work? How will your results improve future research and applications 
+% in the area?
+Because the fastest learned walk was not
+significantly faster than the fastest randomly generated walk, we
+conjecture that the motion representation for the robot is more
+integral to forward speed than the learning algorithm.
 
-We are not deviating from our original proposal.
+\section{References}
+% A numbered list of all books, articles, and web pages you used in your 
+% work or that you cited in your report. Each item must be cited from 
+% somewhere within the text. When citing a URL, include date visited. 
+% When citing a commercial product, list the company.
+
+\section{Acknowledgments}
+% List any people not on the team who helped you with your project: Your
+% TA, other people you consulted with or had useful discussions (say in 
+% a word or two what they did), people who proofread your report, and 
+% any external code you used (libraries etc).
 

### 2010-12-01 report/report.tex
edited the team and acknowledgments

@@
 \item \emph{SVM regression}
-\item \emph{Evolutionary Neural Network (HYPERNEAT)\cite{2}}
+\item \emph{Evolutionary Neural Network (HyperNEAT)\cite{2}}
 \end{itemize}
@@
 % any external code you used (libraries etc).
+\begin{itemize}
+\item Hod Lipson, Cornell Computational Synthesis Lab: advisor
+\item Jim T\o rreson, University of Oslo: advisor
+\item Juan Zagal, University of Chile: designed and printed robot and provided assistance
+\item Jeff Clune, CCSL Red Couch: collaborated on HyperNEAT implementation/testing
+\item Cooper Bills, Cornell University: assisted with Wii tracker development
+\item Anshumali Srivastava, Cornell University: Teaching Assistant
+\end{itemize}
+
+\section{The Team}
+% A table listing the names of the people who worked on the project, and who
+% did what.
+\begin{itemize}
+\item Diana Hidalgo: 
+\begin{itemize}
+\item Simplex (Nelder-Mead)
+\item \code{WiiTrackClient} and \code{WiiTrackServer} classes and hardware
+\item Plotting and evaluating results
+\end{itemize}
+\item Sarah Nguyen
+\begin{itemize}
+\item Random
+\item Uniform random hill climbing
+\item Gaussian random hill climbing
+\item N-dimensional policy gradient descent
+\item Linear regression and prediction
+\item \code{RunManager} and \code{Strategy} classes
+\item Video editing
+\end{itemize}
+\item Jason Yosinski
+\begin{itemize}
+\item SVM regression and prediction
+\item Evolutionary Neural Network (HyperNEAT)
+\item \code{WiiTrackClient} and \code{WiiTrackServer} classes and hardware
+\item \code{Robot}, \code{RunManager}, and \code{Strategy} classes
+\item \code{Parameterized Motion Model} abstract base class
+\item \code{SineModel} classes
+\item \code{explore\_dimensions}
+\item Plotting and evaluating results
+\end{itemize}
+\end{itemize}
 

### 2010-12-01 code/plot/plot_explore.m
Dim explore plots work, version used for presentation yesterday

@@ -26,7 +26,6 @@ function plot_explore(file)
   end
 
   for ii = 1:dimensions
-    figure(ii); clf;
     pIdx = ((ii-1)*(pointsPer*duplications)+1) : ((ii)*(pointsPer*duplications));
     thisParams  = params(pIdx,:);
     thisFitness = fitness(pIdx,:);
@@ -36,7 +35,11 @@ function plot_explore(file)
     %std(fits, 0, 2)
 
     % HACK: only works for 2 dups
+    figure(ii); clf; hold on;
     errorbar(thisParams(1:2:end,ii), mean(fits, 2), std(fits, 0, 2), ">");
+    
+    centerIdx = find(all((thisParams == repmat(centerPoint, 20, 1))'));
+    plot(centerPoint(ii), mean(thisFitness(centerIdx)), 'ro', 'markersize', 10, 'linewidth', 2);
     title(sprintf('varying param %d', ii));
 
     %forplot = []

### 2010-12-03 report/report.tex
report update

@@
 Some promising gaits have been generated, and we anticipate a final
-evaluation of the robot’s gait in November.
+evaluation of the robots gait in November.
 
@@
 We are testing several different learning methods to design a gait for 
-a quadraped robot from the Cornell Computational Synthesis Lab. The metric 
+a quadruped robot from the Cornell Computational Synthesis Lab. The metric 
 for evaluation of the designed gait is speed. A comparison and evaluation 
@@
 representation.
-\item \emph{Simplex (Nelder-Mead) Method}
+\item \emph{Simplex (Nelder-Mead) Method}: Creates a simplex with 6 vertices and tests the fitness of each vertex.
+The worst vertex is reflected over the centroid.
+If the reflected point is better than the second worst point and worse than the best point, then the reflected point replaces the worst.
+If the reflected point is better than the best point, the simplex is expanded in the direction of the reflected point.
+The better of the reflected and the expanded point replaces the worst point.
+If the reflected point is worse than the second worst point, then the simplex is contracted away from the reflected point.
+If the contracted point is better than the reflected point, the contracted point replaces the worst point.
+If the contracted point is worse than the reflected point, the entire simplex is shrunk.
 \item \emph{Linear regression}: To initialize, chooses five random points
@@
 
-The quadraped robot has an on-board computer running Linux. The lower
-level drivers are in C and we are implementing the system in
-Python. Feedback about distance travelled is provided via a Wii remote.
+The quadruped robot has an on-board computer running Linux. The lower
+level drivers are in C and the system is implemented in Python. Feedback about distance travelled is provided via an infrared LED and a Wii remote.
 
@@
 
-\section{Experimental (or Theoretical) Evaluation}
+\section{Experimental Evaluation}
 
@@
 physical robot. We stop each run after plateauing results (no improvement 
-for half the policies seen so far). We evaluate each method based on the
+for one third of the policies seen so far). We evaluate each method based on the
 amount of improvement seen from the initial parameter vectors, and on the
@@
 
+\item For vector A, linear regression worked significantly better than the other algorithms, resulting in a gait (27.58 body lengths/minute) that walked over twice as fast as the next best gait (uniform random hill climbing - 11.37 body lengths/minute).
+
+\item Vector B resulted in similarly performing gaits with all the algorithms.
+Random produced the best gait (17.26 body lengths/minute) and uniform random hill climbing produced the worst (9.44 body lengths/minute).
+The remaining algorithms each produced a gait around 13 body lengths/minute.
+
+\item For vector C, simplex and Gaussian random hill climbing each produced a gait that substantially outperformed the other algorithms.
+Simplex resulted in a gait of nearly 15 body lengths/minute  and Gaussian random hill climbing produced a gait of just over 13 body lengths/minute, whereas the other algorithms returned gaits of less than 5 body lengths/minute.
+
+\item Based on the three trials, linear regression worked the best, followed by simplex.
+Gaussian random hill climbing performed about as well as random, but it is unclear whether Gaussian would continue to improve and become much better than random if allowed to run for more iterations.
+Uniform random hill climbing and policy gradient descent performed similarly.
+
 \item We also coded \code{explore\_dimensions} to collect data for dimension
@@
 % properties of the algorithm and/or the data.
+\begin{itemize}
+\item  Since only three trials were tested with each algorithm and no algorithm consistently outperformed the others, there is a large standard error for each algorithm.
+For this reason, it is unclear if any algorithm is superior to another in this application.
+More trials with each algorithm would be necessary to reach a definitive ranking of the performance of the algorithms.
+
+\item Each algorithm discovered at least one gait of over 10 body lengths/minute, including random.
+For this reason, we speculate that the motion model may be more critical than the learning algorithm.
+\end{itemize}
 

### 2010-12-03 report/abstract.tex
Older changes, maybe too old?

We herein propose a system...
\edit{Write abstract}

### 2010-12-03 report/report.tex
Older changes, maybe too old?

--- 2010-12-01 report/report.tex
@@
-\section{Abstract}
-% A brief summary of the report. This is not an introduction -
-% it should be complete. Include 1-2 sentences for each of the items
-% below, up to and including conclusions.
-This paper presents several approaches to optimizing a quadrupedal
-trot gait for forward speed. Given a parameterized walk designed for
-a quadruped robot designed and printed in the Cornell Computational
-Synthesis Lab, we implement, test, and compare different learning strategies, 
-including uniform and Gaussian random hill climbing, a form of policy 
-gradient reinforcement learning\cite{1}, Simplex (Nelder-Mead), 
-linear regression, SVM regression, and an evolutionary neural 
-network (HyperNEAT)\cite{2}. Because the fastest learned walk was not
-significantly faster than the fastest randomly generated walk, we
-conjecture that the motion representation for the robot is more
-integral to forward speed than the learning algorithm.
+\section{Executive Summary}
 
-Keywords: Learning Control, Walking Robots, Multi Legged Robots
+\edit{Write this}
 
-\section{Introduction}
-% Motivate and abstractly describe the problem you are addressing and
-% how you are addressing it. What is the problem? Why is it important?
-% What is your basic approach? A short discussion of how it fits into
-% related work in the area is also desirable. Summarize the basic results
-% and conclusions that you will present.
+\section{Problem Definition}
+
+\edit{Write this}
+
+\section{Method}
+
+\edit{Write this}
+
+\section{Related Work}
+
+\edit{Write this}
+
+\section{System Architecture and Implementation}
+
+\edit{Write this}
+
+\section{Experimental or (or Theoretical) Evaluation}
+
+\edit{Write this}
+
+\section{Future Work}
+
+\edit{Write this}
+
+\section{Conclusion}
+
+\edit{Write this}
+
+\section{References}
+
+\edit{Make this generated by citations}
+
+\section{Acknowledgements}
+
+\edit{Write this}
+
+\section{The Team}
+
+\edit{Write this}
+
+\section{Code Description Appendix}
+
+\edit{Write this}
+
+\edit{OLD REPORT AFTER THIS POINT}
+
+\section{Executive Summary}
+% A few sentences summarizing each of the items below
 We are using machine learning to design a gait for a quadruped
@@
 
-\section{Problem definition}
-% Precisely define the problem you are addressing (i.e. formally specify
-% the inputs and outputs). Elaborate on why this is an interesting and
-% important problem.
-We are testing several different learning methods to design a gait for 
-a quadraped robot from the Cornell Computational Synthesis Lab. The metric 
-for evaluation of the designed gait is speed. A comparison and evaluation 
-of the many different methods available for optimizing the gait of legged 
-robots will be useful for future work on this challenging multidimensional 
-control problem.
+\section{Proposal Summary}
+% A short (one or two sentence) summary of the project
+% goal/topic. This is just to remind us what you’re working on.
 
-\section{Method}
-% Describe in reasonable detail the algorithm you are using to address this 
-% problem. A pseudo-code description of the algorithm you are using is 
-% frequently useful. If it makes sense for your project, trace through a 
-% concrete example, showing how your algorithm processes this example. The 
-% example should be complex enough to illustrate all of the important aspects 
-% of the problem but simple enough to be easily understood. If possible, 
-% an intuitively meaningful example is better than one with meaningless 
-% symbols.
-We implemented and tested 8 different learning strategies:
-\begin{itemize}
-\item \emph{Uniform random hill climbing}: Creates a random neighbor by
-  randomly choosing one parameter to adjust, and changing it completely 
-  randomly in UniformStrategy(). Then evaluates this neighbor by running 
-  the robot with the newly chosen parameters. If this neighbor results in a
-  longer distance walked then the previous best neighbor, we save this
-  neighbor as the new best neighbor. We then repeat the whole process
-  using the best neighbor as a base.
-\item \emph{Gaussian random hill climbing}: Creates a random neighbor by
-  changing each parameter randomly based on a normal distribution. Then
-  evaluates this neighbor similarly to uniform random hill climbing.
-\item \emph{N-dimensional policy gradient descent}: Estimates the policy
-  gradient by evaluating \emph{t} randomly generated policies
-  near an initial parameter vector. Then computes the average score for
-  each parameter and adjusts the base policy according to the estimated
-  gradient.
-\item \emph{Random}: Randomly generates policies in the range of the motion
-representation.
-\item \emph{Simplex (Nelder-Mead) Method}
-\item \emph{Linear regression}: To initialize, chooses five random points
-and trains on them using least-squares. Then in a loop, takes a fixed-size
-step in the direction of the gradient, and retrains on all points so far.
-\item \emph{SVM regression}
-\item \emph{Evolutionary Neural Network (HYPERNEAT)\cite{2}}
-\end{itemize}
+We are using machine learning to design a gait for a quadraped robot
+from Hod Lipson’s lab. The metric for evaluation of the designed gait
+is speed.
 
-\section{Related work}
-% Briefly explain who else worked on related problems in the past and what
-% methods they used. Explain if you are using similar methods, or if your
-% approach is different and if so - how (either is ok).
-
-\section{System Architecture and Implementation}
-% Describe how you implemented your system and how you structured it. 
-% This should give an overview of the system, not a detailed 
-% documentation of the code. The documentation of the code is part of 
-% the code you hand in. You might want to comment on high-level design 
-% decisions that you made. Also explain how you obtained your
-% data and any pre-processing you did to it.
+\section{System Architecture Summary}
+% A summary of the main components of the system, and how they
+% interact with each other to achieve the goal. This is just to remind
+% us what you are planning to develop.
 
@@
 
+\section{Implementation Status}
+\seclabel{implement}
+% A list of the major components of the system as well as the current
+% status of each (e.g. designed, implemented, tested, finished, almost
+% finished with coding). For the pieces not yet finished, indicate
+% tentative completion dates. The schedule doesn’t have to be
+% incredibly detailed, but it should include entries for all
+% substantial modules/submodules of your system. Include also who in
+% the team is responsible for completing this module.
+
+\subsection*{\code{Robot} class}
+
 \begin{itemize}
-\item \code{Robot} class: Class wrapper for commanding
+\item \emph{Team member:} Jason.
+\item \emph{Current implementation:} Class wrapper for commanding
   motion of the robot.  The \code{Robot} class takes care of the robot
@@
   of the run.
+\item \emph{Future improvements:} None planned, though some will
+  probably arise.
+\item \emph{Status:} 90\% complete.
+\end{itemize}
 
-\item \code{Strategy} and \code{RunManager} class: The user has a choice between seven
-  different learning strategies: uniform random hill climbing, Gaussian random
-  hill climbing, N-dimensional policy gradient descent, random, 
-  simplex, linear regression/prediction, and SVM regression/prediction.
+\subsection*{\code{Strategy} and \code{RunManager} class}
 
-\item \code{Parameterized Motion Model} abstract base class: Several functions exist, but none are within a class framework.
+\begin{itemize}
+\item \emph{Team member:} Jason and Sarah.
+\item \emph{Current implementation:} The user has a choice between three
+  different learning strategies: uniform random hill climbing, gaussian random
+  hill climbing, and N-dimensional policy gradient algorithm descent.
+  \begin{itemize} 
+  \item \emph{Uniform random hill climbing:} Creates a random neighbor by
+  randomly choosing one parameter to adjust, and changing either
+  completely randomly in UniformStrategy(), or slightly in
+  UniformSlightStrategy(). Then evaluates this neighbor by running the robot
+  with the newly chosen parameters. If this neighbor results in a
+  longer distance walked then the previous best neighbor, we save this
+  neighbor as the new best neighbor. We then repeat the whole process
+  using the best neighbor as a base.
+  \item \emph{Gaussian random hill climbing} Creates a random neighbor byi
+  changing each parameter randomly based on a normal distribution. Then
+  evaluates this neighbor similarly to uniform random hill climbing.
+  \item \emph{N-dimensional policy gradient descent} Estimates the policy
+  gradient by evaluating \emph{t} randomly generated policies
+  near an initial parameter vector. Then computes the average score 
+  \end{itemize}
+\item \emph{Future improvements:} Machine learning of objective function.
+\item \emph{Status:} Described algorithms coded. Machine learning of objective
+  function to be coded by Jason.
+\end{itemize}
 
-\item \code{SineModel} classes: Commands motors to positions
+\subsection*{\code{Parameterized Motion Model} abstract base class}
+
+\begin{itemize}
+\item \emph{Team member:} Jason.
+\item \emph{Current implementation:} Several functions exist, but none are within a class framework.
+\item \emph{Future improvements:} Make into class.
+\item \emph{Status:} 40\% complete.
+\end{itemize}
+
+\subsection*{\code{SineModel} classes}
+
+\begin{itemize}
+\item \emph{Team member:} Jason.
+\item \emph{Current implementation:} Commands motors to positions
   based on a sine wave, creating a periodic pattern. Parameters are:
-  amplitude, wavelength, scale inner vs outer motors, scale left vs right
-  motors, scale back vs front motors. Currently a function, not a class.
+  amplitude, wavelength, scale inner vs outer motors, multiply
+  left-right, multiply back-front, shift front-back, shift right-left.
+  Currently a function, not a class.
+\item \emph{Future improvements:} Make class, add time offset
+  parameters.  Continued tweaking of this class and other motion
+  models will be a main goal of the next month.
+\item \emph{Status:} 30\% complete.
+\end{itemize}
 
-\item Other derived motion model classes: Several versions of a SineModel
-  exist, each with different parameters, though solely as functions.
+\subsection*{Other derived motion model classes}
 
-\item \code{WiiTrackClient} and \code{WiiTrackServer} classes
- and hardware: A Wii remote tracks the location of the robot through an infrared LED mounted on top of the robot.
+\begin{itemize}
+\item \emph{Team member:} All.
+\item \emph{Current implementation:} Several versions of a SineModel
+  exist, each with different parameters, though solely as functions
+  (not as derived classes of \code{ParameterizedMotionModel}).
+\item \emph{Future improvements:} Implementing and tweaking current
+  and new motion model classes will be our primary effort over the
+  next month.
+\item \emph{Status:} 20\% complete.
+\end{itemize}
+
+\subsection*{\code{WiiTrackClient} and \code{WiiTrackServer} classes
+ and hardware}
+
+\begin{itemize}
+\item \emph{Team member:} Jason and Diana.
+\item \emph{Current implementation:} A Wii remote tracks the location of the robot through an infrared LED mounted on top of the robot.
 We used CWiid library which is able to interface with the remote via bluetooth to determine the location of the infrared LED.
@@
 The program then calculates the net change in position and uses that as the distance walked.
-
+\item \emph{Status:} Complete.
 \end{itemize}
 
-\section{Experimental (or Theoretical) Evaluation}
+\section{Evaluation status}
+% A list of the major steps in your planned evaluation of the system
+% and the status of each. For some projects it may be easier to merge
+% this list with the component/module status list above of your
+% system. Include also who in the team is responsible for completing
+% this module.
 
-\subsection{Methodology}
-% What are the criteria you are using to evaluate your method? What
-% specific hypotheses does your experiment test? Describe the 
-% experimental methodology that you used. What are the dependent and 
-% independent variables? For projects in machine learning, what is 
-% the training/test data that was used, and why is it realistic or
-% interesting? Exactly what performance data did you collect and how 
-% are you presenting and analyzing it? Comparisons to competing methods 
-% that address the same problem or to variations of your own algorithm 
-% are particularly useful.
 \begin{itemize}
-\item Each algorithm was run on 3 different initial parameter vectors on the
-physical robot. We stop each run after plateauing results (no improvement 
-for half the policies seen so far). We evaluate each method based on the
-amount of improvement seen from the initial parameter vectors, and on the
-fastest speeds achieved during runs.
-\end{itemize}
-
-\subsection{Results}
-% Present the quantitative results of your experiments. Graphical data
-% presentation such as graphs and histograms are frequently better 
-% than tables. Explain what are the basic differences revealed in the 
-% data. If your method is stochastic or you analyze multiple datasets, 
-% calculate mean and standard error for every quantitative
-% performance result you show based on multiple runs.
-\begin{itemize}
-\item We have done runs of 
+\item After running our program and finding a good gait, we will then
+evaluate our system by the distance walked in a certain amount
+of time. This evaluation metric depend on the result of
+running our program for many iterations. We have done 50 runs of 
 3 different inital parameter vectors for random hill climbing, 
-gaussian random hill climbing, policy gradient descent, random,
-simplex, and linear regression. We developed several
-gaits that were about 4 times faster than the original hand-coded gait.
+gaussian random hill climbing, and policy gradient descent.
+The evaluation for our particular project is an ongoing
+process, but we anticipate a final evaluation taking place
+early-to-mid November. The evaluation process will be completed by
+Jason, Sarah, and Diana.
 
@@
 
-\subsection{Discussion}
-% Is your hypothesis supported? What conclusions do the results support
-% about the strengths and weaknesses of your method compared to other 
-% methods? How can the results be explained in terms of the underlying 
-% properties of the algorithm and/or the data.
+\section{Project schedule timeline}
+% Tell us if you are meeting your project timeline or you are
+% exceeding it or falling behind. If you are falling behind, explain
+% how you will catch up or adjust the goals.
 
-\section{Future work}
-% What are the major shortcomings of your current method? For each 
-% shortcoming, propose additions or enhancements that would help overcome it.
+We are meeting our project timeline. We have developed two new algorithms,
+gaussian hill climbing and policy gradient descent and tested all three of
+our major algorithms on the robot. We have also quantified our current results. 
+In the next month we hope to develop one more learning strategy, machine
+learning of an objective function.
+
+\section{Schedule}
+
+Below is a rough schedule of fixed external deadlines and anticipated
+team milestones.
+
+\definecolor{doneColor}{rgb}{0.0,0.7,0.0}
+\newcommand{\done}[1]{{\color{doneColor}\sout{#1}}}
+%\newcommand{\done}[1]{#1}
+
+\begin{center}
+\begin{tabular}{|l||p{2.5in}|p{1.5in}|}
+\hline
+                        & {\bf Milestones} & {\bf Deadlines}
+\\ \hline
+
+Week 1 (9/13-9/19)    & \done{Read papers, get lab access, talk to relevant other researchers} & 9/17 Final proposals due
+\\ \hline
+
+Week 2 (9/20-9/26)    & \done{Continue reading, get robot to move} &
+\\ \hline
+
+Week 3 (9/27-10/03)   & \done{Implement parametrized gait and determine proposed coding schedule for more advanced algorithms in time for Code Review \#1.} &
+\\ \hline
+
+Week 4 (10/04-10/10)  & \done{Begin main algorithm dev/testing effort} & 10/5 Code Review \#1
+\\ \hline
+
+Week 5 (10/11-10/17)  & \done{Algorithm dev/testing} &
+\\ \hline
+
+Week 6 (10/18-10/24)  & \done{Algorithm dev/testing} &
+\\ \hline
+
+Week 7 (10/25-10/31)  & \done{Algorithm dev/testing, quantify/solidify current results for Code Review \#2} &
+\\ \hline
+
+Week 8 (11/1-11/7)    & \done{Finish collecting results, begin writing} & 11/2 Code Review \#2
+\\ \hline
+
+Week 9 (11/8-11/14)   & \done{Algorithm dev/testing} &
+\\ \hline
+
+Week 10 (11/15-11/21) & \done{Finish collecting results, writing, get final demo ready} &
+\\ \hline
+
+Week 11 (11/22-11/28) & \done{Finish collecting results, writing, get final demo ready}  &
+\\ \hline
+
+Week 12 (11/29-11/30) & \done{Final demo}  & 11/30 Final presentation
+\\ \hline
+\end{tabular}
+\end{center}
+
+\section{Activity log (Optional)}
+
+% an informal log your efforts. If you’ve been keeping notes on your
+% progress, feel free to turn these in as well. These might include:
+% (1) notes on what aspect of the project you worked on and when, (2)
+% what was accomplished in each “session”, (3) design questions, (4)
+% implementation questions, (5) descriptions of experiments and
+% results, (6) problems that arose, (7) anything else that you think
+% is relevant to the project.
+
 \begin{itemize}
-\item Different motion representations
-\item Evolutionary algorithms/HyperNEAT\cite{2}
-\item Learning how to turn
-\item Open sourcing code for others to extend
+\item \emph{9/13-9/19:} Set up. Set up trac, worked on pre-proposal
+  and presentation, got lab access, met with Jim Torresen, contacted
+  Juan Cristobal Zagal, read papers, got the robot to walk using
+  previously coded program. Questions -- How do we move the robot? How
+  do we command the motors?
+\item \emph{9/20-9/26:} Moving the robot. Installed Linux on the
+  robot, figured out how to command the motors, coded a program to
+  make the robot stand up, coded a simple motion pattern. Questions --
+  For initial program, what parameters and model should be used? How
+  do we limit the search space? What algorithm should we start with?
+  How will the robot get information about how far it moved?
+\item \emph{9/27-10/3:} Initial program. Chose several parameters and
+  created a simple parameterized sine model, implemented a random hill
+  climbing algorithm to optimize the parameters, ran some trial
+  iterations, began setting up the cameras for feedback. Questions --
+  Did the initial program work? What parameters are important? Are
+  additional parameters necessary? What algorithm should we implement
+  for the final program?
+\item \emph{10/4-10/10:} Plan for final implementation. Discussed
+  possible algorithms to use and new parameters to add. Worked on
+  report for code review. Discussed feedback from code review.  Questions -- Advantages for gradient
+  descent? Should we start with a completely random parameter vector
+  or should we start with one that results in motion and let the
+  program optimize it? How does the camera system work?
+\item \emph{10/11-10/17:} Discussed how we were going to proceed with the project and set up a more detailed schedule.
+Got the cameras working and calibrated them.
+Researched the 3 available streaming protocols.
+Questions -- Which streaming protocol would work best?
+Which improvements should we start with?
+\item \emph{10/18-10/24:} Created a variation on the random hill climbing algorithm to choose the next neighbor based on a Gaussian distribution.
+Tried to work with two of the streaming protocols for the camera system, but were unsuccessful.
+Decided to use a Wii remote for positional information instead.
+Wrote the integration code for the Wii remote.
+Refactored our code to separate the different algorithms.
+Began planning the algorithm for a modified gradient descent.
+Questions -- What values should we use in the gradient descent algorithm?
+How much should the each parameter be changed?
+How many neighbors should be sampled before choosing the next state?
+How can we compare the results of our different algorithms?
+How repeatable is the distance traveled for a given parameter vector?
+\item \emph{10/25-11/2}: Created a modified gradient descent algorithm.
+Ran 150 trials of the three algorithms.
+Each algorithm started with a specific parameter vector and ran for 50 iterations.
+Created and ran a program to get a feel for the overall look of the search space.
+Ran some trials to see how reproducible the distance traveled was.
+Prepared for code review 2.
+Questions -- What does the search space look like?
+Are there a lot of spikes, or is it fairly smooth?
+Based on the results of the algorithms, what conclusions can we draw?
 \end{itemize}
 
-\section{Conclusion}
-% Briefly summarize the important results and conclusions presented in 
-% the paper. What are the most important points illustrated by your 
-% work? How will your results improve future research and applications 
-% in the area?
-Because the fastest learned walk was not
-significantly faster than the fastest randomly generated walk, we
-conjecture that the motion representation for the robot is more
-integral to forward speed than the learning algorithm.
+\section{Proposal Modifications}
+% Let us know if you are deviating from your original proposal.
 
-\section{References}
-% A numbered list of all books, articles, and web pages you used in your 
-% work or that you cited in your report. Each item must be cited from 
-% somewhere within the text. When citing a URL, include date visited. 
-% When citing a commercial product, list the company.
-
-\section{Acknowledgments}
-% List any people not on the team who helped you with your project: Your
-% TA, other people you consulted with or had useful discussions (say in 
-% a word or two what they did), people who proofread your report, and 
-% any external code you used (libraries etc).
+We are not deviating from our original proposal.
 

### 2010-12-03 report/report.tex

--- 2010-12-03 report/report.tex
@@
 % and conclusions that you will present.
+
 We are using machine learning to design a gait for a quadruped
@@
 \item We have done runs of 
-3 different inital parameter vectors for random hill climbing, 
-gaussian random hill climbing, policy gradient descent, random,
+3 different initial parameter vectors for random hill climbing, 
+Gaussian random hill climbing, policy gradient descent, random,
 simplex, and linear regression. We developed several
@@
 % shortcoming, propose additions or enhancements that would help overcome it.
+
 \begin{itemize}

### 2010-12-03 report/abstract.tex
Moved abstract to separate file, added ownstyles

@@
-We herein propose a system...
-\edit{Write abstract}
+% A brief summary of the report. This is not an introduction -
+% it should be complete. Include 1-2 sentences for each of the items
+% below, up to and including conclusions.
+This paper presents several approaches to optimizing a quadrupedal
+trot gait for forward speed. Given a parameterized walk designed for
+a quadruped robot designed and printed in the Cornell Computational
+Synthesis Lab, we implement, test, and compare different learning strategies, 
+including uniform and Gaussian random hill climbing, a form of policy 
+gradient reinforcement learning\cite{1}, Simplex (Nelder-Mead), 
+linear regression, SVM regression, and an evolutionary neural 
+network (HyperNEAT)\cite{2}. Because the fastest learned walk was not
+significantly faster than the fastest randomly generated walk, we
+conjecture that the motion representation for the robot is more
+integral to forward speed than the learning algorithm.
+
+Keywords: Learning Control, Walking Robots, Multi Legged Robots

### 2010-12-03 report/report.tex
Moved abstract to separate file, added ownstyles

@@
-\section{Abstract}
-% A brief summary of the report. This is not an introduction -
-% it should be complete. Include 1-2 sentences for each of the items
-% below, up to and including conclusions.
-This paper presents several approaches to optimizing a quadrupedal
-trot gait for forward speed. Given a parameterized walk designed for
-a quadruped robot designed and printed in the Cornell Computational
-Synthesis Lab, we implement, test, and compare different learning strategies, 
-including uniform and Gaussian random hill climbing, a form of policy 
-gradient reinforcement learning\cite{1}, Simplex (Nelder-Mead), 
-linear regression, SVM regression, and an evolutionary neural 
-network (HyperNEAT)\cite{2}. Because the fastest learned walk was not
-significantly faster than the fastest randomly generated walk, we
-conjecture that the motion representation for the robot is more
-integral to forward speed than the learning algorithm.
-
-Keywords: Learning Control, Walking Robots, Multi Legged Robots
-
 \section{Introduction}

### 2010-12-03 code/plot/plot_several_runs.m
Fixed eps file generation error (while using title)

@@ -34,8 +34,12 @@ function plot_several_runs(files)
     %plot(idx, best(idx), 'o', 'Color', clr{mod(ii,6)+1}, 'MarkerSize', 5);
   end
 
-  title(titl, 'Interpreter', 'none');
+  % [JBY] This line (the title) causes eps files which cause errors
+  %title(titl, 'Interpreter', 'none');   
+
   legend(leg);
+  xlabel('Iteration');
+  ylabel('Body lengths/minute');
 
   for ii = 1:length(files)
     file = files{ii};

### 2010-12-03 code/plot/plot_all.m
few more plots

@@ -28,11 +28,11 @@ plot_several_runs({'log_linear_regression_a.txt', 'log_linear_regression_b.txt',
 plot_several_runs({'log_simplex_a.txt', 'log_simplex_b.txt', 'log_simplex_c.txt'})
 
 plot_several_runs({'log_uniform_a.txt', 'log_gaussian_a.txt', 'log_gradient_a.txt', 'log_random_a.txt', ...
-                   'log_linear_regression_a.txt', 'log_simplex_a.txt'});
+                   'log_linear_regression_a.txt', 'log_simplex_a.txt'}, 'vectorA');
 plot_several_runs({'log_uniform_b.txt', 'log_gaussian_b.txt', 'log_gradient_b.txt', 'log_random_b.txt', ...
-                   'log_linear_regression_b.txt', 'log_simplex_b.txt'});
+                   'log_linear_regression_b.txt', 'log_simplex_b.txt'}, 'vectorB');
 plot_several_runs({'log_uniform_c.txt', 'log_gaussian_c.txt', 'log_gradient_c.txt', 'log_random_c.txt', ...
-                   'log_linear_regression_c.txt', 'log_simplex_c.txt'});
+                   'log_linear_regression_c.txt', 'log_simplex_c.txt'}, 'vectorC');
 
 
 plot_several_runs({'log_gaussian_a.txt', 'log_uniform_a.txt', 'log_gradient_a.txt', ...
@@ -41,3 +41,13 @@ plot_several_runs({'log_gaussian_a.txt', 'log_uniform_a.txt', 'log_gradient_a.tx
                    'log_random_b.txt', 'log_linear_regression_b.txt', 'log_simplex_b.txt', ...
                    'log_gaussian_c.txt', 'log_uniform_c.txt', 'log_gradient_c.txt', ...
                    'log_random_c.txt', 'log_linear_regression_c.txt', 'log_simplex_c.txt'})
+
+
+% error plots
+files = {'log_uniform_a.txt', 'log_uniform_b.txt', 'log_uniform_c.txt', ...
+         'log_gaussian_a.txt', 'log_gaussian_b.txt', 'log_gaussian_c.txt', ...
+         'log_gradient_a.txt', 'log_gradient_b.txt', 'log_gradient_c.txt', ...
+         'log_random_a.txt', 'log_random_b.txt', 'log_random_c.txt', ...
[…]

### 2010-12-03 code/plot/plot_error.m

@@ -1,9 +1,4 @@
-files = {'log_uniform_a.txt', 'log_uniform_b.txt', 'log_uniform_c.txt', ...
-         'log_gaussian_a.txt', 'log_gaussian_b.txt', 'log_gaussian_c.txt', ...
-         'log_gradient_a.txt', 'log_gradient_b.txt', 'log_gradient_c.txt', ...
-         'log_random_a.txt', 'log_random_b.txt', 'log_random_c.txt', ...
-         'log_linear_regression_a.txt', 'log_linear_regression_b.txt', 'log_linear_regression_c.txt', ...
-         'log_simplex_a.txt', 'log_simplex_b.txt', 'log_simplex_c.txt'};
+function plot_error(files, saveFileName)
 
 clr = {};
   clr{end+1} = 'k';
@@ -14,7 +9,7 @@ clr = {};
   clr{end+1} = 'm';
 %  clr{end+1} = 'y';
 
-figure(1); clf; hold on;
+figure(2); clf; hold on;
 ax = zeros(4, 1);
 ax(1) = 0; ax(2) = 80; ax(3) = 0; ax(4) = 20;
 axis(ax);
@@ -62,12 +57,20 @@ for ii = 0:5
   %plot(iteration, best, '-', 'color', clr{mod(ii,6)+1});
   e = errorbar(x, means, stdev);
   set(e, "color", clr{mod(ii,6)+1});
+
+  %e = plot(x, means, 'linewidth', 2);
+  %set(e, "color", clr{mod(ii,6)+1});
+  %e = plot(x, means + stdev);
+  %set(e, "color", clr{mod(ii,6)+1});
+  %e = plot(x, means - stdev);
+  %set(e, "color", clr{mod(ii,6)+1});
 end
 
 legend(leg);
 
-title('Mean and Standard Error', 'Interpreter', 'none');
+% I think this causes problems...
+%title('Mean and Standard Error', 'Interpreter', 'none');
 
 % Save
-print('-dpng', strcat('error_plot', '.png'));
-print('-depsc', strcat('error_plot', '.eps'));
+print('-dpng', strcat(saveFileName, '.png'));
+print('-depsc', strcat(saveFileName, '.eps'));

### 2010-12-03 code/plot/plot_several_runs.m

@@ -1,4 +1,4 @@
-function plot_several_runs(files)
+function plot_several_runs(files, saveFileName)
   % Plots the fitness of two runs together
 
   clr = {};
@@ -55,8 +55,11 @@ function plot_several_runs(files)
   end
   
   % Save
-  disp(sprintf('Saving as %s', titl));
-  print('-dpng', strcat(titl, '.png'));
-  print('-depsc', strcat(titl, '.eps'));
+  if ~exist('saveFileName', 'var')
+    saveFileName = titl
+  end
+  disp(sprintf('Saving as %s', saveFileName));
+  print('-dpng', strcat(saveFileName, '.png'));
+  print('-depsc', strcat(saveFileName, '.eps'));
 
 end

### 2010-12-03 report/report.tex
few more plots
Edits with everyone

@@
 
+\edit{write this}
+
 \section{Problem definition}
@@
 
+\edit{flesh this out... more details}
+
 \section{Method}
@@
 % symbols.
+
+\edit{also describe parameterized motion model}
+
 We implemented and tested 8 different learning strategies:
@@
 If the contracted point is better than the reflected point, the contracted point replaces the worst point.
-If the contracted point is worse than the reflected point, the entire simplex is shrunk.
+If the contracted point is worse than the reflected point, the entire simplex is shrunk \cite{NMWebsite}.
 \item \emph{Linear regression}: To initialize, chooses five random points
@@
 
+\edit{Actually write this...}
+
 \section{System Architecture and Implementation}
+\seclabel{implement}
 % Describe how you implemented your system and how you structured it. 
@@
 \begin{itemize}
+
 \item \code{Robot} class: Class wrapper for commanding
@@
 
-\item \code{Strategy} and \code{RunManager} class: The user has a choice between seven
+\item \code{RunManager} class \edit{write this}
+
+\item \code{Strategy} class: The user has a choice between seven
   different learning strategies: uniform random hill climbing, Gaussian random
@@
 
-\item \code{Parameterized Motion Model} abstract base class: Several functions exist, but none are within a class framework.
+\item \code{MotionModel} abstract base class: Several functions exist,
+  but none are within a class framework. ... ??? ... \code{SineModel}
+  classes: Commands motors to positions based on a sine wave, creating
+  a periodic pattern. Parameters are: amplitude, wavelength, scale
+  inner vs outer motors, scale left vs right motors, scale back vs
+  front motors. Currently a function, not a class. ... ??? ... THere
+  are also other motion models... Other derived motion model classes:
+  Several versions of a SineModel exist, each with different
+  parameters, though solely as functions.
 
-\item \code{SineModel} classes: Commands motors to positions
-  based on a sine wave, creating a periodic pattern. Parameters are:
-  amplitude, wavelength, scale inner vs outer motors, scale left vs right
-  motors, scale back vs front motors. Currently a function, not a class.
-
-\item Other derived motion model classes: Several versions of a SineModel
-  exist, each with different parameters, though solely as functions.
-
-\item \code{WiiTrackClient} and \code{WiiTrackServer} classes
- and hardware: A Wii remote tracks the location of the robot through an infrared LED mounted on top of the robot.
-We used CWiid library which is able to interface with the remote via bluetooth to determine the location of the infrared LED.
-This information is accessed through some functions we wrote and passed to the program.
-The program gets the robot's position at the beginning of each run and then again at the end.
-The program then calculates the net change in position and uses that as the distance walked.
+\item \code{WiiTrackClient} and \code{WiiTrackServer} classes and
+  hardware: A Wii remote tracks the location of the robot through an
+  infrared LED mounted on top of the robot. We used the CWiid library
+  which is able to interface with the remote via bluetooth to
+  determine the location of the infrared LED. This information is
+  accessed through some functions we wrote and passed to the
+  program. The program gets the robot's position at the beginning of
+  each run and then again at the end. The program then calculates the
+  net change in position and uses that as the distance
+  walked. \edit{Jason: add details}.
 
@@
 % are particularly useful.
-\begin{itemize}
-\item Each algorithm was run on 3 different initial parameter vectors on the
-physical robot. We stop each run after plateauing results (no improvement 
-for one third of the policies seen so far). We evaluate each method based on the
-amount of improvement seen from the initial parameter vectors, and on the
-fastest speeds achieved during runs.
-\end{itemize}
+
+\editbox{mention using speed as evaluating factor, ``we started the
+  robot, and it would collect data on its own.  if it walks otusdie fo
+  the viewable area, it tells us and we move it back inside and
+  restart it (but the data collection continues)''}
+
+We stop each run after plateauing results (no improvement 
+for one third of the policies seen so far)  (WHY??)
+
+Each algorithm was run on 3 different initial parameter vectors on the
+physical robot.  ``We decided to evaluate all methods starting at the
+same three vectors, to have more of a fair comparison...'' We evaluate
+each method based on the amount of improvement seen from the initial
+parameter vectors, and on the fastest speeds achieved during runs.
+
+Compare to Juan's run too.
 
@@
 % performance result you show based on multiple runs.
-\begin{itemize}
-\item We have done runs of 
+
+We have done runs of 
 3 different initial parameter vectors for random hill climbing, 
@@
 gaits that were about 4 times faster than the original hand-coded gait.
+
+\editbox{Pull out vectors A, B, and C into separte bulletted list, and
+  make table, like below}
+
+\begin{tabular}{|r|c|c|c|}
+\hline
+                       & A     & B     & C    \\
+\hline
+Solution depth ???        & 14    & 54    & 21   \\
+\hline
+Nodes visited          & 992   & 2405  & 383  \\
+\hline
+Time to find solution  & 139ms & 451ms & 54ms \\
+\hline
+\end{tabular}
+
+\begin{itemize}
 
@@
 
-\item Based on the three trials, linear regression worked the best, followed by simplex.
+\end{itemize}
+
+Based on the three trials, linear regression worked the best, followed by simplex.
 Gaussian random hill climbing performed about as well as random, but it is unclear whether Gaussian would continue to improve and become much better than random if allowed to run for more iterations.
@@
 
-\item We also coded \code{explore\_dimensions} to collect data for dimension
+We also coded \code{explore\_dimensions} to collect data for dimension
 varying plots. We intend to do further explore this by running
 \code{explore\_dimension} around a parameter that produces a medium-good gait.
-\end{itemize}
+
+\edit{fix captions in this section too!}
+
+\figvarp{vectorA}{.65}{Vector A.... more here....}{}
+
+\figvarp{vectorB}{.65}{Vector B.... more here....}{}
+
+\figvarp{vectorC}{.65}{Vector C.... more here....}{}
+
+\figvarp{std_error}{1}{Standard error....}{}
+
+\edit{Jason: add SVM and HyperNEAT results here!}
 
@@
 % properties of the algorithm and/or the data.
+
+\edit{important points... talk about these more :)}
+
 \begin{itemize}
@@
 
+\edit{Intro sentence here...}
+
 \begin{itemize}
+\item More runs and/or longer runs
 \item Different motion representations
@@
 % in the area?
+
+\edit{make this more awesome}
+
 Because the fastest learned walk was not
@@
 integral to forward speed than the learning algorithm.
-
-\section{References}
-% A numbered list of all books, articles, and web pages you used in your 
-% work or that you cited in your report. Each item must be cited from 
-% somewhere within the text. When citing a URL, include date visited. 
-% When citing a commercial product, list the company.
 
@@
 \item \code{Motion.py}: Motion helper functions.
-\item \code{WiiTrackServer.py}: Broadcasts the position of the infrared LED>
+\item \code{WiiTrackServer.py}: Broadcasts the position of the infrared LED.
 \item \code{WiiTrackClient.py}: Connects to the WiiTrackServer to get the current position information.

### 2010-12-03 report/report.tex
Edited RunManager description.
Added link.

@@
 
+A video presentation of this project should be viewed at http://www.youtube.com/watch?v=ODoiOj9DdGg.
 \edit{write this}
@@
 
-\item \code{RunManager} class \edit{write this}
+\item \code{RunManager} class: Deals with all the details of running the robot,
+  including choosing an initial parameter, running the robot multiple times, 
+  tracking distance walked, and writing to the log file. Also includes the
+  \code{explore\_dimensions} method.
 
@@
 
-We stop each run after plateauing results (no improvement 
+The metric for evaluation of the designed gait is speed. A comparison and evaluation. We stop each run after plateauing results (no improvement 
 for one third of the policies seen so far)  (WHY??)

### 2010-12-03 report/report.tex
report

@@
 
-\editbox{mention using speed as evaluating factor, ``we started the
+\editbox{``we started the
   robot, and it would collect data on its own.  if it walks otusdie fo
@@
 
-The metric for evaluation of the designed gait is speed. A comparison and evaluation. We stop each run after plateauing results (no improvement 
-for one third of the policies seen so far)  (WHY??)
+The metric for evaluation of the designed gait is speed. We stop each 
+run after plateauing results (no improvement for one third of the 
+policies seen so far). The standard length of a run designates that it
+should be stopped
 

### 2010-12-06 report/report.tex
update to report
report - edited problem statemetn

@@
 We are using machine learning to design a gait for a quadruped
-robot. The robot has an on-board computer, drivers for the motors, and
-will have camera feedback. The code so far consists of the robot
+robot. The code so far consists of the robot
 class, optimization class, parameterized model, sine model, and camera
-feedback.  Both the hardware and software are progressing on schedule.
-Some promising gaits have been generated, and we anticipate a final
-evaluation of the robots gait in November.
+feedback.
 
-A video presentation of this project should be viewed at http://www.youtube.com/watch?v=ODoiOj9DdGg.
+A video presentation of this project should be viewed at
+
+http://www.youtube.com/watch?v=ODoiOj9DdGg.
 \edit{write this}
@@
 % important problem.
-We are testing several different learning methods to design a gait for 
-a quadruped robot from the Cornell Computational Synthesis Lab. The metric 
-for evaluation of the designed gait is speed. A comparison and evaluation 
+We are testing several different learning methods to design a parametrized 
+gait for a quadruped robot from the Cornell Computational Synthesis Lab. 
+The metric for evaluation of the designed gait is speed. Given parameters 
+for the current walk and the current time, our code outputs the position
+of each of the robot's motors. A comparison and evaluation 
 of the many different methods available for optimizing the gait of legged 
@@
 control problem.
-
-\edit{flesh this out... more details}
 
@@
 \hline
-Solution depth ???        & 14    & 54    & 21   \\
+Uniform Random Hill Climbing             & 11.37   & 9.44   & 2.69   \\
 \hline
-Nodes visited          & 992   & 2405  & 383  \\
+Guassian Random Hill Climbing            & 3.10    & 13.59  & 13.40  \\
 \hline
-Time to find solution  & 139ms & 451ms & 54ms \\
+Policy Gradient Descent                  & 0.68    & 14.69  & 3.60   \\
+\hline
+Random                                   & 6.04    & 17.26  & 4.90   \\
+\hline
+Simplex (Nelder-Mead)                    & 8.51    & 13.62  & 14.83  \\
+\hline
+Linear Regression                        & 27.58   & 12.51  & 1.95   \\
+\hline
+SVM Regression                           & ???     & ???    & ???    \\
+\hline
+Evolutionary Neural Network (HyperNEAT)  & ???     & ???    & ???    \\
 \hline

### 2010-12-06 report/report.tex
edited future work, etc.

@@
 % how you are addressing it. What is the problem? Why is it important?
-% What is your basic approach? A short discussion of how it fits into
-% related work in the area is also desirable. Summarize the basic results
+% What is your basic approach? A short discussion of how it fits into % related work in the area is also desirable. Summarize the basic results
 % and conclusions that you will present.
@@
 
-\edit{also describe parameterized motion model}
+We have a parameterized motion model that commands motors to positions
+based on a sine wave, creating a periodic pattern. Parameters are: 
+amplitude, wavelength, scale inner vs outer motors, scale left vs right 
+motors, scale back vs front motors. Each strategy below attempts to choose
+the best possible parameters for this motion model.
 
@@
   gradient.
-\item \emph{Random}: Randomly generates policies in the range of the motion
-representation.
+\item \emph{Random}: Randomly generates parameter vector in the allowable
+  range. This strategy is used only as base line.
 \item \emph{Simplex (Nelder-Mead) Method}: Creates a simplex with 6 vertices and tests the fitness of each vertex.
@@
 \item \code{MotionModel} abstract base class: Several functions exist,
-  but none are within a class framework. ... ??? ... \code{SineModel}
+  but none are within a class framework. ... ??? ...
+
+  \begin{itemize} 
+  \item \code{SineModel}
   classes: Commands motors to positions based on a sine wave, creating
@@
   inner vs outer motors, scale left vs right motors, scale back vs
-  front motors. Currently a function, not a class. ... ??? ... THere
-  are also other motion models... Other derived motion model classes:
+  front motors. Currently a function, not a class. ... ??? ... There
+  are also other motion models...
+
+  \item Other derived motion model classes:
   Several versions of a SineModel exist, each with different
   parameters, though solely as functions.
+  \end{itemize}
 
@@
 \edit{Intro sentence here...}
+There are several additions we would make to flesh out our project. First,
+the error margins for our runs was quite large, so reducing them would lead
+to more conclusive data. It would also be insightful to run our algorithms
+on different motion models to get a better idea of how well different
+algorithms perform overall, and not just for our current motion model. We
+also intend to experiment more with evolutionary algorithms/HyperNEAT. 
+Some parameter vectors resulted in the robot turning, as opposed to it
+walking long distances. This could be an interesting learning goal in
+future projects. To these ends, we propose the following additions
+and enhancements:
 
@@
 \item Learning how to turn
-\item Open sourcing code for others to extend
 \end{itemize}

### 2010-12-06 report/report.tex
future work
methodology

@@
 
-\editbox{``we started the
-  robot, and it would collect data on its own.  if it walks otusdie fo
-  the viewable area, it tells us and we move it back inside and
-  restart it (but the data collection continues)''}
-
 The metric for evaluation of the designed gait is speed. We stop each 
@@
 policies seen so far). The standard length of a run designates that it
-should be stopped
+should be stopped after there is no improvement for one half of the policies
+seen so far, but since all runs took place on the actual robot, without use
+of a simulator, certain time limitations were imposed on the learning process.
+
+We controlled our experiments from a computer that was connected via a 
+wireless ethernet to the robot. The robot collected data about distance
+walked automatically on its own. If it walked outside of the Wii remote's
+viewable area, it informed the user, so the only human intervention
+required during an experiment involved moving the robot back inside the 
+viewable area and restarting it and dealing with various mechanical failures,
+which did not interrupt the learning process or result in the loss of data.
+
+We evaluated the efficacy of a set of parameters by sending these
+parameters to the robot and instructing it to walk for a certain length
+of time. The robot always begins from the same position and returns to the
+starting position at the end of the run in order to allow for uniformity
+in measuring. More efficient parameters resulted in a faster gait, which
+translated into a longer distance walked and a better score. After completing
+an evluation, the robot sent the resulting distance walked back to the
+host computer and prepared itself for a new set of parameters to evaluate.
 
 Each algorithm was run on 3 different initial parameter vectors on the
-physical robot.  ``We decided to evaluate all methods starting at the
-same three vectors, to have more of a fair comparison...'' We evaluate
+physical robot. We decided to evaluate all methods starting at the
+same three vector in order to allow for the fair comparison of each
+algorihtm.  We evaluate
 each method based on the amount of improvement seen from the initial
@@
 
-Compare to Juan's run too.
+The resulting gaits from our algorithms quickly outperformed the original
+hand-coded walk designed for this robot. The fasted walk, for example, was
+4 times faster.
 
@@
 
-\edit{Intro sentence here...}
 There are several additions we would make to flesh out our project. First,

### 2010-12-06 report/report.tex
few spelling changes

@@
 translated into a longer distance walked and a better score. After completing
-an evluation, the robot sent the resulting distance walked back to the
+an evaluation, the robot sent the resulting distance walked back to the
 host computer and prepared itself for a new set of parameters to evaluate.
@@
 same three vector in order to allow for the fair comparison of each
-algorihtm.  We evaluate
+algorithm.  We evaluate
 each method based on the amount of improvement seen from the initial
@@
 
-\editbox{Pull out vectors A, B, and C into separte bulletted list, and
+\editbox{Pull out vectors A, B, and C into separate bulletted list, and
   make table, like below}
@@
 \hline
-Guassian Random Hill Climbing            & 3.10    & 13.59  & 13.40  \\
+Gaussian Random Hill Climbing            & 3.10    & 13.59  & 13.40  \\
 \hline
@@
 \begin{itemize}
-\item Hod Lipson, Cornell Computational Synthesis Lab: advisor
-\item Jim T\o rreson, University of Oslo: advisor
+\item Hod Lipson, Cornell Computational Synthesis Lab: adviser
+\item Jim T\orreson, University of Oslo: adviser
 \item Juan Zagal, University of Chile: designed and printed robot and provided assistance
@@
 \end{itemize}
+
+% LocalWords:  vertices

### 2010-12-06 report/report.tex
minor
related work

--- 2010-12-06 report/report.tex
@@
 
-\edit{Actually write this...}
+Various maching learning techniques have proven to be useful in finding
+control policies for a wide variety of robots. Kohl and Stone\cite{1}
+presented a policy gradient reinforcement learning approach for generating
+a fast walk on legged robots. We experimented with this method to create
+a walk for our robot (Policy Gradient Descent). Chernova and Velosa\cite{3}
+took an evolutionary approach to this problem which we did not implement.
+Zykov, Bongard, and Lipson\cite{4} describe the evolution of dynamic gaits 
+on a physical robot requiring no prior
+assumptions about the locomotion pattern beyond the fact that it should be
+rhythmic.
 

### 2010-12-06 report/report.tex

--- 2010-12-06 report/report.tex
@@
 \item Hod Lipson, Cornell Computational Synthesis Lab: adviser
-\item Jim T\orreson, University of Oslo: adviser
+\item Jim T\o rreson, University of Oslo: adviser
 \item Juan Zagal, University of Chile: designed and printed robot and provided assistance

### 2010-12-06 report/abstract.tex
Edits with everyone
Added references section

@@
 including uniform and Gaussian random hill climbing, a form of policy 
-gradient reinforcement learning\cite{1}, Simplex (Nelder-Mead), 
+gradient reinforcement learning\cite{kohl}, Simplex (Nelder-Mead), 
 linear regression, SVM regression, and an evolutionary neural 
@@
 
+\edit{fix citations}
+
 Keywords: Learning Control, Walking Robots, Multi Legged Robots

### 2010-12-06 report/report.tex
intro edit

@@
 
-We are using machine learning to design a gait for a quadruped
-robot. The code so far consists of the robot
-class, optimization class, parameterized model, sine model, and camera
-feedback.
+This paper presents a comparison of varied learning methods to
+design a gait for a quadruped robot. Using these methods we have
+created walks that are several times faster than the original
+hand-tuned gait.
 

### 2010-12-06 report/report.tex
in progress

--- 2010-12-06 report/report.tex
@@
 \section{Introduction}
+
 % Motivate and abstractly describe the problem you are addressing and
 % how you are addressing it. What is the problem? Why is it important?
-% What is your basic approach? A short discussion of how it fits into % related work in the area is also desirable. Summarize the basic results
-% and conclusions that you will present.
+% What is your basic approach? A short discussion of how it fits into
+% % related work in the area is also desirable. Summarize the basic
+% results and conclusions that you will present.
+
+Applications of walking robots often call for the ability to walk as
+quickly, efficiently, or with as little power as possible.
+Oftentimes, these optimizations are done manually by an expert who
+designs a gait to be, for example, as fast as possible \cite{someone}.  Some groups (Kohl and Stone) have
+
+Previous
+Walking robots often have
+
+Many previous walking robots have had walking motions -- or gaits --
+designed and tweaked by hand \cite{someone,else}, or in some cases, beginning with a
 
 We are using machine learning to design a gait for a quadruped
-robot. The code so far consists of the robot
-class, optimization class, parameterized model, sine model, and camera
-feedback.
+robot. The code so far consists of the robot class, optimization
+class, parameterized model, sine model, and camera feedback.
 

### 2010-12-06 report/report.tex
Intro done?

--- 2010-12-06 report/report.tex
@@
 \section{Introduction}
+
 % Motivate and abstractly describe the problem you are addressing and
 % how you are addressing it. What is the problem? Why is it important?
-% What is your basic approach? A short discussion of how it fits into % related work in the area is also desirable. Summarize the basic results
-% and conclusions that you will present.
+% What is your basic approach? A short discussion of how it fits into
+% % related work in the area is also desirable. Summarize the basic
+% results and conclusions that you will present.
 
-This paper presents a comparison of varied learning methods to
-design a gait for a quadruped robot. Using these methods we have
-created walks that are several times faster than the original
-hand-tuned gait.
+Applications of walking robots often call for the ability to walk as
+quickly, efficiently, or with as little power as possible.  Often
+these optimizations are done manually by an expert who designs and
+tweaks a gait specifically for a given objective.  Other groups have
+used learning methods to generate gaits optimized for some metric.
+Approaches differ in their starting assumptions, some essentially
+tweaking the parameters of a hand-tuned model \cite{chernova}, others
+exploring a reasonably compact parameter space \cite{kohl}, and still
+others beginning with few assumptions besides periodicity
+\cite{zykov}.
 
-A video presentation of this project should be viewed at
+We aimed to strike a middle ground between these approaches.  Our
+motion generator did not relying on hand-tweaked gaits, but it did use
+a parameterized gait which, by its nature, imposes some assumptions on
+the answers produced.  From this model, we used machine learning to
+design a gait for a quadruped robot.
 
-http://www.youtube.com/watch?v=ODoiOj9DdGg.
-\edit{write this}
+This paper presents a comparison of varied learning methods to design
+a gait for a quadruped robot. Using these methods we have created
+walks that are several times faster than the original hand-tuned gait.
+We invite readers with short attention spans to view a video of some
+of our results online here:
+
+\url{http://www.youtube.com/watch?v=ODoiOj9DdGg}
+
+%The code so far consists of the robot class, optimization class,
+%parameterized model, sine model, and camera feedback.
+
+\edit{proofread}
 

### 2010-12-07 report/abstract.tex
much editing with sarah, diana

--- 2010-12-03 report/abstract.tex
@@
-We herein propose a system...
-\edit{Write abstract}
+% A brief summary of the report. This is not an introduction -
+% it should be complete. Include 1-2 sentences for each of the items
+% below, up to and including conclusions.
+
+This paper presents an array of approaches to optimizing a quadrupedal
+gait for forward speed.  We implement, test, and compare different
+learning strategies including uniform and Gaussian random hill
+climbing, policy gradient reinforcement learning\cite{kohl},
+Nelder-Mead simplex\cite{nm}, new predictive methods based on linear
+and support vector regression, and an evolved neural network
+(HyperNEAT)\cite{2}.  We compare results to a baseline random search
+method.  Many of the methods resulted in walks significantly faster
+than previously hand-tuned gaits.
+
+%Because the fastest learned walk
+%was not significantly faster than the fastest randomly generated walk,
+%we conjecture that the motion representation for the robot is more
+%integral to forward speed than the learning algorithm.
+
+\edit{fix citations}
+
+Keywords: Learning Control, Walking Robots, Gait Learning, Quadruped

### 2010-12-07 report/report.tex
much editing with sarah, diana

@@
 We aimed to strike a middle ground between these approaches.  Our
-motion generator did not relying on hand-tweaked gaits, but it did use
-a parameterized gait which, by its nature, imposes some assumptions on
-the answers produced.  From this model, we used machine learning to
-design a gait for a quadruped robot.
-
-This paper presents a comparison of varied learning methods to design
-a gait for a quadruped robot. Using these methods we have created
-walks that are several times faster than the original hand-tuned gait.
-We invite readers with short attention spans to view a video of some
-of our results online here:
+motion generator did not rely on hand-tweaked gaits, but it did use
+parameterized gaits which, by their nature, impose some assumptions on
+the answers produced.  We then used machine learning to design gaits
+for a quadruped robot with these models.  This paper presents a
+comparison of the different learning methods implemented.  Most
+methods created walks that are several times faster than the original
+hand-tuned gait.  We invite readers with short attention spans to view
+a video of some of our results online here:
 
@@
 % important problem.
-We are testing several different learning methods to design a parametrized 
-gait for a quadruped robot from the Cornell Computational Synthesis Lab. 
-The metric for evaluation of the designed gait is speed. Given parameters 
-for the current walk and the current time, our code outputs the position
-of each of the robot's motors. A comparison and evaluation 
-of the many different methods available for optimizing the gait of legged 
-robots will be useful for future work on this challenging multidimensional 
-control problem.
+
+We are testing several different learning methods to design a
+parametrized gait for a quadruped robot from the Cornell Computational
+Synthesis Lab.
+
+The output each of the learning algorithms is a function of time,
+$f(t)$, that outputs a vector of commanded motor positions.  This
+function is generated using a parametrized motion model, described in
+\secref{motion}.
+
+The robot executes these commands and measures its change in location
+using the tracking system described in \secref{implement}.  The input
+to the learning algorithms is this measured displacement, which the
+algorithms attempt to maximize. This displacement is measured for each
+gait over a constant length run, usually 12 seconds.
+
+A comparison and evaluation of the many different methods available
+for optimizing the gait of legged robots will be useful for future
+work on this challenging multidimensional control problem.
 
 \section{Method}
-% Describe in reasonable detail the algorithm you are using to address this 
-% problem. A pseudo-code description of the algorithm you are using is 
-% frequently useful. If it makes sense for your project, trace through a 
-% concrete example, showing how your algorithm processes this example. The 
-% example should be complex enough to illustrate all of the important aspects 
-% of the problem but simple enough to be easily understood. If possible, 
-% an intuitively meaningful example is better than one with meaningless 
-% symbols.
+\seclabel{method}
 
-We have a parameterized motion model that commands motors to positions
-based on a sine wave, creating a periodic pattern. Parameters are: 
-amplitude, wavelength, scale inner vs outer motors, scale left vs right 
-motors, scale back vs front motors. Each strategy below attempts to choose
-the best possible parameters for this motion model.
+% Describe in reasonable detail the algorithm you are using to address
+% this problem. A pseudo-code description of the algorithm you are
+% using is frequently useful. If it makes sense for your project,
+% trace through a concrete example, showing how your algorithm
+% processes this example. The example should be complex enough to
+% illustrate all of the important aspects of the problem but simple
+% enough to be easily understood. If possible, an intuitively
+% meaningful example is better than one with meaningless symbols.
 
-We implemented and tested 8 different learning strategies:
+We use several parameterized motion models that command motors to
+positions based on a sine wave, creating a periodic pattern.  While we
+investigated several models, for the bulk of our experiments, we used
+a model whose five parameters are: amplitude, wavelength, scale inner
+vs outer motors, scale left vs right motors, scale back vs front
+motors. Each strategy below attempts to choose the best possible
+parameters for this motion model.
+
+We implemented and tested 8 different learning strategies.  All
+strategies except for the HyperNEAT method\cite{clune} were
+constrained to pick parameters from within predetermined ranges.
+
 \begin{itemize}
-\item \emph{Uniform random hill climbing}: Creates a random neighbor by
-  randomly choosing one parameter to adjust, and changing it completely 
-  randomly in UniformStrategy(). Then evaluates this neighbor by running 
-  the robot with the newly chosen parameters. If this neighbor results in a
-  longer distance walked then the previous best neighbor, we save this
-  neighbor as the new best neighbor. We then repeat the whole process
-  using the best neighbor as a base.
-\item \emph{Gaussian random hill climbing}: Creates a random neighbor by
-  changing each parameter randomly based on a normal distribution. Then
-  evaluates this neighbor similarly to uniform random hill climbing.
-\item \emph{N-dimensional policy gradient descent}: Estimates the policy
-  gradient by evaluating \emph{t} randomly generated policies
-  near an initial parameter vector. Then computes the average score for
-  each parameter and adjusts the base policy according to the estimated
-  gradient.
-\item \emph{Random}: Randomly generates parameter vector in the allowable
-  range. This strategy is used only as base line.
-\item \emph{Simplex (Nelder-Mead) Method}: Creates a simplex with 6 vertices and tests the fitness of each vertex.
-The worst vertex is reflected over the centroid.
-If the reflected point is better than the second worst point and worse than the best point, then the reflected point replaces the worst.
-If the reflected point is better than the best point, the simplex is expanded in the direction of the reflected point.
-The better of the reflected and the expanded point replaces the worst point.
-If the reflected point is worse than the second worst point, then the simplex is contracted away from the reflected point.
-If the contracted point is better than the reflected point, the contracted point replaces the worst point.
-If the contracted point is worse than the reflected point, the entire simplex is shrunk \cite{NMWebsite}.
-\item \emph{Linear regression}: To initialize, chooses five random points
-and trains on them using least-squares. Then in a loop, takes a fixed-size
-step in the direction of the gradient, and retrains on all points so far.
-\item \emph{SVM regression}
-\item \emph{Evolutionary Neural Network (HyperNEAT)\cite{2}}
+
+\item \emph{Random}: This method randomly generates parameter vectors
+  in the allowable range. This strategy is used only as baseline.
+
+\item \emph{Uniform random hill climbing}: This method begins by
+  selecting a single random parameter vector.  Subsequent iterations
+  generate a neighbor by randomly choosing one parameter to adjust and
+  replacing it with a new value chosen with uniform probability in the
+  allowable range for that parameter. The neighbor is evaluated by
+  running the robot with the newly chosen parameters. If this neighbor
+  results in a longer distance walked than the previous best gait, it
+  is saved as the new best gait. The process is then repeated, always
+  starting with the best gait.
+
+\item \emph{Gaussian random hill climbing}: This method works
+  similarly to Uniform random hill climbing, except neighbors are
+  generated by adding random Gaussian noise to the current best gait.
+  This results in all parameters being changed at once, but the
+  resulting vector is always fairly close to the previous best gait.
+  We used independently selected noise in each dimension, scaled such
+  that the standard deviation of the noise was 5\% of the range of
+  that dimension.
+
+\item \emph{N-dimensional policy gradient descent}: As opposed to the
+  previous methods, this method explicitly estimates the gradient for
+  the objective function. It does this by first evaluating \emph{t}
+  randomly generated parameter vectors near the initial vector, each
+  dimension of these vectors being perturbed by either $-\epsilon$,
+  $0$, or $\epsilon$. Then, for each dimension, it groups vectors into
+  three groups: $-\epsilon$, $0$, and $\epsilon$.  The gradient along
+  this dimension is then estimated as the average score for the
+  $\epsilon$ group minus the average score for the $-\epsilon$
+  group. Finally, the method creates a new vector by changing all
+  parameters by a fixed-size step in the direction of the gradient.
+
+\item \emph{Nelder-Mead simplex method}\cite{nm}: The Nelder-Mead
+  simplex method creates an initial simplex with 6 vertices. The
+  initial parameter vector is stored as the first vertex and the other
+  five vertices are created by adding to one dimension at a time one
+  tenth of the allowable range for that parameter. It then tests the
+  fitness of each vertex and based on these fitnesses, it reflects the
+  worst point over the centroid in an attempt to improve it.  However,
+  to prevent cycles and becoming stuck in local minima, several other
+  rules are used.  In general, the worst vertex is reflected over the
+  centroid. If the reflected point is better than the second worst
+  point and worse than the best point, then the reflected point
+  replaces the worst. If the reflected point is better than the best
+  point, the simplex is expanded in the direction of the reflected
+  point. The better of the reflected and the expanded point replaces
+  the worst point. If the reflected point is worse than the second
+  worst point, then the simplex is contracted away from the reflected
+  point. If the contracted point is better than the reflected point,
+  the contracted point replaces the worst point. If the contracted
+  point is worse than the reflected point, the entire simplex is
+  shrunk \cite{NMWebsite}.
+
+\item \emph{Linear regression}: To initialize, this method chooses and
+  evaluates five random parameter vectors. It then fits a linear model
+  from parameter vector to fitness. In a loop, the method chooses and
+  evaluates a new parameter vector generated by taking a fixed-size
+  step in the direction of the gradient for each parameter, and fits a
+  new linear model to all vectors evaluated so far, choosing the model
+  to minimize the sum of squared errors.
+
+\item \emph{SVM regression}: Similarly to linear regression, this
+  model starts with several random vectors, but this time they are
+  chosen in a small neighborhood about some initial random vector.
+  These vectors (generally 8) are evaluated, and a support vector
+  regression model is fit to the observed fitnesses.  To choose the
+  next vector for evaluation, we randomly generate some number
+  (typically 100) of vectors in the neighborhood of the best observed
+  gait, and select for evaluation the vector with the best predicted
+  performance.
+
+  \edit{Has many parameters... in simulation... we found that it's
+    better to bump points...}
+
+\item \emph{Evolutionary Neural Network (HyperNEAT)\cite{clune}}
+
+  \edit{do this}
+
 \end{itemize}
@@
 
-Various maching learning techniques have proven to be useful in finding
-control policies for a wide variety of robots. Kohl and Stone\cite{1}
-presented a policy gradient reinforcement learning approach for generating
-a fast walk on legged robots. We experimented with this method to create
-a walk for our robot (Policy Gradient Descent). Chernova and Velosa\cite{3}
-took an evolutionary approach to this problem which we did not implement.
-Zykov, Bongard, and Lipson\cite{4} describe the evolution of dynamic gaits 
-on a physical robot requiring no prior
-assumptions about the locomotion pattern beyond the fact that it should be
-rhythmic.
+Various maching learning techniques have proven to be useful in
+finding control policies for a wide variety of robots. Kohl and
+Stone\cite{kohl} presented a policy gradient reinforcement learning
+approach for generating a fast walk on legged robots. We experimented
+with this method to create a walk for our robot (Policy Gradient
+Descent). Chernova and Velosa\cite{chernova} took an evolutionary
+approach to this problem which we did not implement.  Zykov, Bongard,
+and Lipson\cite{zykov} describe the evolution of dynamic gaits on a
+physical robot requiring no prior assumptions about the locomotion
+pattern beyond the fact that it should be rhythmic.
 
@@
 The quadruped robot has an on-board computer running Linux. The lower
-level drivers are in C and the system is implemented in Python. Feedback about distance travelled is provided via an infrared LED and a Wii remote.
+level drivers are in C and the system is implemented in
+Python. Feedback about distance travelled is provided via an infrared
+LED mounted on the robot and a Wii remote fixed to the ceiling.
+
+An overview of the code follows.
 
@@
 
-\item \code{RunManager} class: Deals with all the details of running the robot,
-  including choosing an initial parameter, running the robot multiple times, 
-  tracking distance walked, and writing to the log file. Also includes the
-  \code{explore\_dimensions} method.
+\item \code{RunManager} class: Deals with all the details of running
+  the robot, including running the robot multiple times, tracking
+  distance walked via a \code{WiiTrackClient} member object, and
+  writing results to the log file. Also includes the
+  \code{explore\_dimensions} method to generate plots by varying each
+  parameter independently.
 
-\item \code{Strategy} class: The user has a choice between seven
-  different learning strategies: uniform random hill climbing, Gaussian random
-  hill climbing, N-dimensional policy gradient descent, random, 
-  simplex, linear regression/prediction, and SVM regression/prediction.
+\item \code{Strategy} class: The user has a choice between eight
+  different learning strategies: Random search, uniform random hill
+  climbing, Gaussian random hill climbing, N-dimensional policy
+  gradient descent, Nelder-Mead simplex, linear regression/prediction,
+  SVM regression/prediction, and HyperNEAT evolution.  Each strategy
+  must derive from the base \code{Strategy} class and must implement
+  two methods: \code{getNext}, for getting the next parameter vector
+  to try, and \code{updateResults}, for communicating results of a run
+  back to the strategy.
 
-\item \code{MotionModel} abstract base class: Several functions exist,
-  but none are within a class framework. ... ??? ...
-
-  \begin{itemize} 
-  \item \code{SineModel}
-  classes: Commands motors to positions based on a sine wave, creating
-  a periodic pattern. Parameters are: amplitude, wavelength, scale
-  inner vs outer motors, scale left vs right motors, scale back vs
-  front motors. Currently a function, not a class. ... ??? ... There
-  are also other motion models...
-
-  \item Other derived motion model classes:
-  Several versions of a SineModel exist, each with different
-  parameters, though solely as functions.
-  \end{itemize}
+\item \code{MotionModel} class: We implemented several motion models,
+  the main being the \code{SineModel5} class.  All motion models take
+  as input a parameter vector and produce as output a motion model,
+  which is simply a function mapping from time to nine motor
+  positions.  The \code{SineModel5} model commands positions based on
+  a sine wave, creating a periodic pattern. The parameters are:
+  amplitude, wavelength, a multiplier for the inner vs outer motors,
+  multiplier for left vs right motors, and multiplier for back vs
+  front motors.  Other similar models were tested, including a seven
+  parameter model which allowed sine waves shifted in time, but these
+  were not used as extensively in our experiments as
+  \code{SineModel5}.
 
@@
   hardware: A Wii remote tracks the location of the robot through an
-  infrared LED mounted on top of the robot. We used the CWiid library
-  which is able to interface with the remote via bluetooth to
-  determine the location of the infrared LED. This information is
-  accessed through some functions we wrote and passed to the
-  program. The program gets the robot's position at the beginning of
-  each run and then again at the end. The program then calculates the
-  net change in position and uses that as the distance
-  walked. \edit{Jason: add details}.
+  infrared LED mounted on top of the robot. A \code{WiiTrackServer} is
+  run on the robot and continuously tracks its position using the
+  CWiid library\cite{cwiid} to interface with the remote via
+  bluetooth.  The \code{RunManager} then makes a
+  \code{WiiTrackClient}, which connects via socket to the tracking
+  server and requests position updates periodically.
+  \code{RunManager} currently gets the robot's position at the
+  beginning of each run and then again at the end and uses this to
+  calculate the net change in position.
 
@@
 We controlled our experiments from a computer that was connected via a
-wireless ethernet to the robot. The robot collected data about distance
-walked automatically on its own. If it walked outside of the Wii remote's
-viewable area, it informed the user, so the only human intervention
-required during an experiment involved moving the robot back inside the 
-viewable area and restarting it and dealing with various mechanical failures,
-which did not interrupt the learning process or result in the loss of data.
+wireless ethernet to the robot. The robot collected data about
+distance walked automatically on its own. If it walked outside of the
+Wii remote's viewable area, it informed the user, so the only human
+intervention required during an experiment was to move the robot back
+inside the viewable area and to resume the run, which did not
+interrupt the learning process or result in the loss of data.
 
 We evaluated the efficacy of a set of parameters by sending these
-parameters to the robot and instructing it to walk for a certain length
-of time. The robot always begins from the same position and returns to the
-starting position at the end of the run in order to allow for uniformity
-in measuring. More efficient parameters resulted in a faster gait, which
+parameters to the robot and instructing it to walk for a certain
+length of time. The robot always began from the same position and
+returned to the starting position at the end of the run in order to
+measure true displacement without giving credit for ending in a leaned
+position. More efficient parameters resulted in a faster gait, which
 translated into a longer distance walked and a better score. After completing
@@
 same three vector in order to allow for the fair comparison of each
-algorithm.  We evaluate
-each method based on the amount of improvement seen from the initial
-parameter vectors, and on the fastest speeds achieved during runs.
+algorithm.  We evaluate each method based on the amount of improvement
+seen from the initial parameter vectors, and on the fastest speeds
+achieved during runs.
 
 The resulting gaits from our algorithms quickly outperformed the original
-hand-coded walk designed for this robot. The fasted walk, for example, was
+hand-coded walk designed for this robot. The fastest walk, for example, was
 4 times faster.
@@
 
-We have done runs of 
-3 different initial parameter vectors for random hill climbing, 
-Gaussian random hill climbing, policy gradient descent, random,
-simplex, and linear regression. We developed several
-gaits that were about 4 times faster than the original hand-coded gait.
+We have done complete runs of 3 different initial parameter vectors
+for random search, uniform random hill climbing, Gaussian random hill
+climbing, policy gradient descent, Nelder-Mead simplex, and linear
+regression.  We also evaluated the SVM regression and HyperNEAT
+methods, but these methods were more experimental, and thus we do not
+yet have the same volume of data for these runs.  We developed several
+gaits that were about 4 times faster than the original hand-coded
+gait.
 
-\editbox{Pull out vectors A, B, and C into separate bulletted list, and
-  make table, like below}
+\edit{Convert to Table}
 
-\begin{tabular}{|r|c|c|c|}
+\begin{tabular}{|r|c|c|c||c|}
 \hline
-                                         & A       & B      & C      \\
+                                         & A       & B      & C      &  Average \\
 \hline                                                               
-Uniform Random Hill Climbing             & 11.37   & 9.44   & 2.69   \\
 \hline                                                               
-Gaussian Random Hill Climbing            & 3.10    & 13.59  & 13.40  \\
+Previous hand-coded gait                 & --      & --     & --     &  5.16 \\
 \hline                                                                 
-Policy Gradient Descent                  & 0.68    & 14.69  & 3.60   \\
+Random search                            & 6.04    & 17.26  & 4.90   &  9.40 \\
 \hline                                                                 
-Random                                   & 6.04    & 17.26  & 4.90   \\
+Uniform Random Hill Climbing             & 11.37   & 9.44   & 2.69   &  7.83 \\
 \hline                                                                 
-Simplex (Nelder-Mead)                    & 8.51    & 13.62  & 14.83  \\
+Gaussian Random Hill Climbing            & 3.10    & 13.59  & 13.40  &  10.03 \\
 \hline                                                                 
-Linear Regression                        & 27.58   & 12.51  & 1.95   \\
+Policy Gradient Descent                  & 0.68    & 14.69  & 3.60   &  6.32 \\
 \hline                                                                 
-SVM Regression                           & ???     & ???    & ???    \\
+Nelder-Mead simplex                      & 8.51    & 13.62  & 14.83  &  12.32 \\
 \hline                                                                 
-Evolutionary Neural Network (HyperNEAT)  & ???     & ???    & ???    \\
+Linear Regression                        & 27.58   & 12.51  & 1.95   &  14.01 \\
+%\hline                                                                
+%SVM Regression                           & ???     & ???    & ???   &   \\
+%\hline                                                                
+%Evolutionary Neural Network (HyperNEAT)  & ???     & ???    & ???   &   \\
 \hline
@@
 
+\edit{Caption: The best gaits found for each starting vector and
+  algorithm, in body lengths per minute.}
+
 \begin{itemize}
 
-\item For vector A, linear regression worked significantly better than the other algorithms, resulting in a gait (27.58 body lengths/minute) that walked over twice as fast as the next best gait (uniform random hill climbing - 11.37 body lengths/minute).
+\item For vector A, shown in \figref{vectorA}, linear regression
+  worked significantly better than the other algorithms, resulting in
+  a gait (27.58 body lengths/minute) that walked over twice as fast as
+  the next best gait -- uniform random hill climbing at 11.37 body
+  lengths/minute -- and 5.3 times better than the previous hand-coded
+  gait.
 
-\item Vector B resulted in similarly performing gaits with all the algorithms.
-Random produced the best gait (17.26 body lengths/minute) and uniform random hill climbing produced the worst (9.44 body lengths/minute).
-The remaining algorithms each produced a gait around 13 body lengths/minute.
+\item Vector B, depicted in \figref{vectorB}, resulted in similarly
+  performing gaits with all the algorithms. The random method got
+  lucky and produced the best gait (17.26 body lengths/minute) and
+  uniform random hill climbing produced the worst (9.44 body
+  lengths/minute). The remaining algorithms each produced a gait
+  around 13 body lengths/minute.
 
-\item For vector C, simplex and Gaussian random hill climbing each produced a gait that substantially outperformed the other algorithms.
-Simplex resulted in a gait of nearly 15 body lengths/minute  and Gaussian random hill climbing produced a gait of just over 13 body lengths/minute, whereas the other algorithms returned gaits of less than 5 body lengths/minute.
+\item For vector C, shown in \figref{vectorC}, simplex and Gaussian
+  random hill climbing each produced a gait that substantially
+  outperformed the other algorithms. Simplex resulted in a gait of
+  nearly 15 body lengths/minute and Gaussian random hill climbing
+  produced a gait of just over 13 body lengths/minute, whereas the
+  other algorithms returned gaits of less than 5 body lengths/minute.
 
@@
 
-Based on the three trials, linear regression worked the best, followed by simplex.
-Gaussian random hill climbing performed about as well as random, but it is unclear whether Gaussian would continue to improve and become much better than random if allowed to run for more iterations.
-Uniform random hill climbing and policy gradient descent performed similarly.
+\figvarp{vectorA}{1}{Results for runs beginning with vector A.
+  Linear regression worked significantly better than the other
+  algorithms, resulting in a gait (27.58 body lengths/minute) that
+  walked over twice as fast as the next best gait -- uniform random
+  hill climbing at 11.37 body lengths/minute -- and 5.3 times better
+  than the previous hand-coded gait.}{}
+
+\figvarp{vectorB}{1}{Results for run starting with vector B.  All
+  algorithms performed similarly.  The random method actually got
+  lucky this time and produced the best gait (17.26 body
+  lengths/minute) and uniform random hill climbing produced the worst
+  (9.44 body lengths/minute). The remaining algorithms each produced a
+  gait around 13 body lengths/minute.}{}
+
+\figvarp{vectorC}{1}{Results for runs starting with vector C.
+  Simplex and Gaussian random hill climbing each produced a gait that
+  substantially outperformed the other algorithms, with around 15 body
+  lengths/minute and 13 body lengths/minute, respectively, whereas the
+  other algorithms returned gaits of less than 5 body
+  lengths/minute.}{}
+
+Based on the three trials, as shown in \figref{std_error}, linear
+regression worked the best, followed by simplex. Gaussian random hill
+climbing performed about as well as random, but it is unclear whether
+Gaussian would continue to improve and become much better than random
+search if allowed to run for more iterations. Uniform random hill
+climbing and policy gradient descent performed similarly to each other
+on average, but differed greatly on individual runs.
+
+\figvarp{std_error}{1}{Average results for each method starting with
+  parameter vectors A, B, and C.  Error bars are plotted showing the
+  standard error.  Linear regression outperformed other methods,
+  followed by simplex. Gaussian random hill climbing performed about
+  as well as random, but it is unclear whether Gaussian would continue
+  to improve and become much better than random search if allowed to
+  run for more iterations. Uniform random hill climbing and policy
+  gradient descent performed similarly to each other on average, but
+  differed greatly on individual runs.}{}
+
+In addition to running strategies to optimize parameterized gaits, we
+also ......
 
@@
 
-\edit{fix captions in this section too!}
-
-\figvarp{vectorA}{.65}{Vector A.... more here....}{}
-
-\figvarp{vectorB}{.65}{Vector B.... more here....}{}
-
-\figvarp{vectorC}{.65}{Vector C.... more here....}{}
-
-\figvarp{std_error}{1}{Standard error....}{}
+\edit{Jason: add exploration runs}
 
@@
 
-\edit{important points... talk about these more :)}
+Because only three trials were tested with each algorithm and no
+algorithm consistently outperformed the others, there is a large
+standard error for each algorithm, as show in \figref{std_error}. For
+this reason, it is unclear if any algorithm is superior to another in
+this application. More trials with each algorithm would be necessary
+to reach a definitive ranking of the performance of the algorithms.
 
-\begin{itemize}
-\item  Since only three trials were tested with each algorithm and no algorithm consistently outperformed the others, there is a large standard error for each algorithm.
-For this reason, it is unclear if any algorithm is superior to another in this application.
-More trials with each algorithm would be necessary to reach a definitive ranking of the performance of the algorithms.
-
-\item Each algorithm discovered at least one gait of over 10 body lengths/minute, including random.
-For this reason, we speculate that the motion model may be more critical than the learning algorithm.
-\end{itemize}
+However, each algorithm discovered at least one gait of over 10 body
+lengths/minute, including random. For this reason, we speculate that
+the motion model is at least as critical as the learning algorithm
+used.
 
@@
 
-There are several additions we would make to flesh out our project. First,
-the error margins for our runs was quite large, so reducing them would lead
-to more conclusive data. It would also be insightful to run our algorithms
-on different motion models to get a better idea of how well different
-algorithms perform overall, and not just for our current motion model. We
-also intend to experiment more with evolutionary algorithms/HyperNEAT. 
+There are several directions in which we could continue our
+work. First, the error margins for our runs were large, so reducing
+them by running more trials would lead to more conclusive data. It
+would also be insightful to run our algorithms on different motion
+models.  We suspect that our choice of motion model influenced the
+results greatly, as even random choices in the space produced gaits
+that moved a significant fraction of the time (> 5\%).  It would be
+interesting to see how learning methods would perform using a model
+that included a much higher percentage of unproductive gaits. We also
+intend to experiment more with evolutionary algorithms/HyperNEAT.
 Some parameter vectors resulted in the robot turning, as opposed to it
 walking long distances. This could be an interesting learning goal in
-future projects. To these ends, we propose the following additions
-and enhancements:
+future projects. To these ends, we propose the following additions and
+enhancements:
 
@@
 \item Different motion representations
-\item Evolutionary algorithms/HyperNEAT\cite{2}
+\item Evolutionary algorithms/HyperNEAT\cite{clune}
 \item Learning how to turn
@@
 
-\edit{make this more awesome}
+We have presented an array of approaches to optimizing a quadrupedal
+gait for forward speed.  We have implemented and tested different
+learning strategies, including uniform and Gaussian random hill
+climbing, policy gradient reinforcement learning, Nelder-Mead simplex,
+several new predictive methods based on linear and support vector
+regression, and an evolved neural network (HyperNEAT).  We have also
+compared these approaches to random search as a baseline. Many of the
+methods resulted in walks significantly faster than previously
+hand-tuned gaits.
 
-Because the fastest learned walk was not
-significantly faster than the fastest randomly generated walk, we
-conjecture that the motion representation for the robot is more
-integral to forward speed than the learning algorithm.
+Because only three trials were tested with each algorithm and no
+algorithm consistently outperformed the others, there was a large
+standard error for each method, as shown in \figref{std_error}. Thus
+it was unclear if any algorithm was superior to another in this
+application. More trials with each algorithm would be necessary to
+reach a definitive ranking. Because each algorithm discovered at least one gait of
+over 10 body lengths/minute, including random search, we also
+conjectured that the motion representation for the robot is more
+integral to forward speed than the learning algorithm.  How to learn
+the motion representation, in addition to its parameters, remains an
+open problem.
 
@@
 \begin{itemize}
-\item Hod Lipson, Cornell Computational Synthesis Lab: adviser
-\item Jim T\o rreson, University of Oslo: adviser
-\item Juan Zagal, University of Chile: designed and printed robot and provided assistance
-\item Jeff Clune, CCSL Red Couch: collaborated on HyperNEAT implementation/testing
-\item Cooper Bills, Cornell University: assisted with Wii tracker development
-\item Anshumali Srivastava, Cornell University: Teaching Assistant
+\item Hod Lipson, Cornell Computational Synthesis Lab: adviser.
+\item Jim T\o rreson, University of Oslo: adviser.
+\item Juan Zagal, University of Chile: designed and printed robot and provided code for hand tuned gait.
+\item Jeff Clune, CCSL Red Couch: collaborated on HyperNEAT implementation/testing.
+\item Cooper Bills, Cornell University: assisted with Wii tracker development.
+\item Anshumali Srivastava, Cornell University: Teaching Assistant.
 \end{itemize}
@@
 % did what.
+
 \begin{itemize}
+
 \item Diana Hidalgo: 
@@
 \item Plotting and evaluating results
+\item Investigation into using Optitrack IR system for tracking robot
 \end{itemize}
+
 \item Sarah Nguyen
 \begin{itemize}
-\item Random
+\item Random search
 \item Uniform random hill climbing
@@
 \end{itemize}
+
 \item Jason Yosinski
@@
 \end{itemize}
+
 \end{itemize}
@@
 \begin{itemize}
-\item \code{optimize.py}: Determines a strategy to try and runs the robot
+\item \code{optimize.py}: Main program, determines a strategy to try and runs the robot
   with that strategy.

### 2010-12-07 report/report.tex
fixed a few things
explore plots in

@@
 %parameterized model, sine model, and camera feedback.
-
-\edit{proofread}
 
@@
 gaits that were about 4 times faster than the original hand-coded
-gait.
+gait.  Results are shown in \tabref{results_table}.
 
-\edit{Convert to Table}
-
+\begin{table}
+\begin{center}
 \begin{tabular}{|r|c|c|c||c|}
@@
 \end{tabular}
-
-\edit{Caption: The best gaits found for each starting vector and
-  algorithm, in body lengths per minute.}
+\caption{The best gaits found for each starting vector and algorithm,
+  in body lengths per minute.}
+\tablabel{results_table}
+\end{center}
+\end{table}
 
@@
 In addition to running strategies to optimize parameterized gaits, we
-also ......
+also wanted to investigate the space of possible gaits.  To accomplish
+this, we selected a parameter vector that resulted in motion, but not
+an exceptional gait, and plotted performance along each dimension
+individually.  In addition, we duplicated each measurement to be able
+to estimate the measurement noise at each point.  Results for this
+exploration are shown in \figref{explore_dim_1} through
+\figref{explore_dim_5}.
 
-We also coded \code{explore\_dimensions} to collect data for dimension
-varying plots. We intend to do further explore this by running
-\code{explore\_dimension} around a parameter that produces a medium-good gait.
+\figvarp{explore_dim_1}{.5}{Motion for varying just dimension 1}{}
 
-\edit{Jason: add exploration runs}
+\figvarp{explore_dim_2}{.5}{Motion for varying just dimension 2}{}
+
+\figvarp{explore_dim_3}{.5}{Motion for varying just dimension 3}{}
+
+\figvarp{explore_dim_4}{.5}{Motion for varying just dimension 4}{}
+
+\figvarp{explore_dim_5}{.5}{Motion for varying just dimension 5}{}
 

### 2010-12-07 report/report.tex
SVM section roughly done

@@
   gait, and select for evaluation the vector with the best predicted
-  performance.
+  performance.  We suspected that if we always chose the best
+  predicted point out of 100, we may end up progressing along a narrow
+  subspace, prohibiting learning of the true local fitness function.
+  Alternately, we would always choosing exploitation of knowledge
+  vs. exploration of the space.  To address this concern, we added a
+  parameter dubbed \code{bumpBy} that added noise to the final
+  selected point before it was evaluated.
 
-  \edit{Has many parameters... in simulation... we found that it's
-    better to bump points...}
+  Such a method naturally has many tunable parameters, and we
+  endeavored to select these parameteres by tuning the method in
+  simulation.  To estimate the performance of the algorithm, we ran it
+  agains a simulation with a known optimum.  The simulated function
+  was in the same five dimensional parameter space, and simply
+  returned a fitness determined as the height of a Gaussian with a
+  random mean.  The width of the Gaussian in each dimension was 20\%
+  of the range of each dimension, and the maximum value at the peak
+  was 100.  \figref{svm_sim_results} shows the learning results on
+  this simulated model using the ultimately selected SVM parameters.
+  Interestingly, a non-zero value of \code{bumpBy} resulted in better
+  learning than a zero valued, noise free (exploration free) learning.
 
-\item \emph{Evolutionary Neural Network (HyperNEAT)\cite{clune}}
+  Ultimately, however, the version of SVM tuned for simulation did not
+  yet show competitive performance on the real robot.  We tried tuning
+  some parameters on the real robot, but after some amount of tuning,
+  the method still exibited too little exploration and easily became
+  stuck in local minima.
 
-  \edit{do this}
+\item \emph{Evolutionary Neural Network (HyperNEAT)\cite{clune}}: Near
+  the end of term, we hacked together an interface between HyperNEAT
+  and the robot, requiring a slightly modified strategy interface.
+  Preliminary HyperNEAT runs were promising and resulted in serveral
+  interesting gaits.  Unfortunately, the gaits generated by HyperNEAT
+  also tended to stress the robot more than typical gaits had before,
+  and the servos would often overheat and malfunction, requiring
+  restart.  We think these issues may be addressed by adding a small
+  layer between the HyperNEAT strategy and the robot that disallows
+  quickly shifting commanded positions, and we hope to be able to test
+  these methods further once this filter is in place.
 
 \end{itemize}
+
+\figvarp{svm_sim_results}{.6}{Results for the SVM regression strategy
+  in simulation.  This simulation was used to tune the SVM strategy's
+  parameters before trying it on the physical robot.  The strategy
+  quickly approaches the maximum simulated fitness of 100.}{}
 
@@
 
-\edit{Jason: add SVM and HyperNEAT results here!}
-
 \subsection{Discussion}
@@
 that included a much higher percentage of unproductive gaits. We also
-intend to experiment more with evolutionary algorithms/HyperNEAT.
-Some parameter vectors resulted in the robot turning, as opposed to it
-walking long distances. This could be an interesting learning goal in
-future projects. To these ends, we propose the following additions and
-enhancements:
+intend to experiment further with SVM regression and evolutionary
+algorithms/HyperNEAT.  Some parameter vectors resulted in the robot
+turning, as opposed to it walking long distances. This could be an
+interesting learning goal in future projects. To these ends, we
+propose the following additions and enhancements:
 
@@
 \item Different motion representations
+\item Better tuning of SVM regression
 \item Evolutionary algorithms/HyperNEAT\cite{clune}

### 2010-12-07 code/SVMStrategy.py

@@ -242,8 +242,12 @@ class SVMLearningStrategy(OneStepStrategy):
 
 
     def plot(self):
-        from matplotlib.pyplot import plot, show
+        from matplotlib.pyplot import plot, show, savefig, xlabel, ylabel
         plot(self.y)
+        xlabel('Iteration')
+        ylabel('Fitness (arbitrary units)')
+        savefig('svm_sim_results.eps')
+        savefig('svm_sim_results.png')
         show()
 
     def saveAndCleanup(self):

### 2010-12-07 code/RunManager.py
removed comment

@@ -15,7 +15,6 @@ class RunManager:
         self.robot = Robot(commandRate = 40, loud = False)
         self.statesSoFar = set()  # Keeps track of the states tested so far
         
-    # TODO: Am I calling stuff in here correctly..?
     def run_robot(self, currentState):
         '''
         Runs the robot with currentState parameters and returns the

### 2011-01-12 paper/abstract.tex
Added copy of report as paper

--- 2010-12-07 report/abstract.tex
@@
 and support vector regression, and an evolved neural network
-(HyperNEAT)\cite{2}.  We compare results to a baseline random search
-method.  Many of the methods resulted in walks significantly faster
-than previously hand-tuned gaits.
+(HyperNEAT)\cite{clune}.  We compare results to a baseline random
+search method.  Many of the methods resulted in walks significantly
+faster than previously hand-tuned gaits.
 
@@
 
-\edit{fix citations}
-
 Keywords: Learning Control, Walking Robots, Gait Learning, Quadruped

### 2011-01-12 paper/report.tex
Added copy of report as paper

--- 2010-12-07 report/report.tex
@@
 function is generated using a parametrized motion model, described in
-\secref{motion}.
+\secref{implement}.
 
@@
   point is worse than the reflected point, the entire simplex is
-  shrunk \cite{NMWebsite}.
+  shrunk \cite{nm}.
 
@@
   subspace, prohibiting learning of the true local fitness function.
-  Alternately, we would always choosing exploitation of knowledge
+  Put another way, we would always choose exploitation of knowledge
   vs. exploration of the space.  To address this concern, we added a
@@
   Such a method naturally has many tunable parameters, and we
-  endeavored to select these parameteres by tuning the method in
+  endeavored to select these parameters by tuning the method in
   simulation.  To estimate the performance of the algorithm, we ran it
-  agains a simulation with a known optimum.  The simulated function
+  against a simulation with a known optimum.  The simulated function
   was in the same five dimensional parameter space, and simply
@@
   Interestingly, a non-zero value of \code{bumpBy} resulted in better
-  learning than a zero valued, noise free (exploration free) learning.
+  learning than noise free (exploration free) learning.
 
-  Ultimately, however, the version of SVM tuned for simulation did not
-  yet show competitive performance on the real robot.  We tried tuning
-  some parameters on the real robot, but after some amount of tuning,
-  the method still exibited too little exploration and easily became
-  stuck in local minima.
+  Ultimately, however, the version of SVM tuned for simulation still
+  did not show competitive performance on the real robot.  We tried
+  tuning some parameters on the real robot, but after some amount of
+  tuning, the method still exhibited too little exploration and easily
+  became stuck in local minima.
 
@@
   the end of term, we hacked together an interface between HyperNEAT
-  and the robot, requiring a slightly modified strategy interface.
-  Preliminary HyperNEAT runs were promising and resulted in serveral
+  -- an implementation of a method for evolving neural networks -- and
+  the robot, requiring a slightly modified strategy interface.
+  Preliminary HyperNEAT runs were promising and resulted in several
   interesting gaits.  Unfortunately, the gaits generated by HyperNEAT
@@
   and the servos would often overheat and malfunction, requiring
-  restart.  We think these issues may be addressed by adding a small
+  restarts.  We think these issues may be addressed by adding a small
   layer between the HyperNEAT strategy and the robot that disallows
@@
 exploration are shown in \figref{explore_dim_1} through
-\figref{explore_dim_5}.
+\figref{explore_dim_5}.  As is shown in the figures, some dimensions
+are smoother than others, and some measurements are fairly noisy.
 
-\figvarp{explore_dim_1}{.5}{Motion for varying just dimension 1}{}
-
-\figvarp{explore_dim_2}{.5}{Motion for varying just dimension 2}{}
-
-\figvarp{explore_dim_3}{.5}{Motion for varying just dimension 3}{}
-
-\figvarp{explore_dim_4}{.5}{Motion for varying just dimension 4}{}
-
-\figvarp{explore_dim_5}{.5}{Motion for varying just dimension 5}{}
+\figvarp{explore_dim_1}{.65}{Fitness mean and standard deviation
+  vs. dimension 1.  The circle is a common point in
+  \figref{explore_dim_1} through \figref{explore_dim_5}}{}
+\figvarp{explore_dim_2}{.65}{Fitness mean and standard deviation
+  vs. dimension 2.  The circle is a common point in
+  \figref{explore_dim_1} through \figref{explore_dim_5}}{}
+\figvarp{explore_dim_3}{.65}{Fitness mean and standard deviation
+  vs. dimension 3.  The circle is a common point in
+  \figref{explore_dim_1} through \figref{explore_dim_5}}{}
+\figvarp{explore_dim_4}{.65}{Fitness mean and standard deviation
+  vs. dimension 4.  The circle is a common point in
+  \figref{explore_dim_1} through \figref{explore_dim_5}}{}
+\figvarp{explore_dim_5}{.65}{Fitness mean and standard deviation
+  vs. dimension 5.  The circle is a common point in
+  \figref{explore_dim_1} through \figref{explore_dim_5}}{}
 

### 2011-01-14 paper/report.tex
First ugly version of report in acm style.  Figures ugly, bibliography commented out (not yet working)
edited methods - hyperNEAT a little...need to figure out details

@@
 
-\url{http://www.youtube.com/watch?v=ODoiOj9DdGg}
+%\url{http://www.youtube.com/watch?v=ODoiOj9DdGg}
+\texttt{http://www.youtube.com/watch?v=ODoiOj9DdGg}
 
@@
 
-\item \emph{Evolutionary Neural Network (HyperNEAT)\cite{clune}}: Near
-  the end of term, we hacked together an interface between HyperNEAT
+\item \emph{Evolutionary Neural Network (HyperNEAT)\cite{clune}}: We 
+  put together an interface between HyperNEAT
   -- an implementation of a method for evolving neural networks -- and
   the robot, requiring a slightly modified strategy interface.
-  Preliminary HyperNEAT runs were promising and resulted in several
-  interesting gaits.  Unfortunately, the gaits generated by HyperNEAT
-  also tended to stress the robot more than typical gaits had before,
-  and the servos would often overheat and malfunction, requiring
-  restarts.  We think these issues may be addressed by adding a small
-  layer between the HyperNEAT strategy and the robot that disallows
-  quickly shifting commanded positions, and we hope to be able to test
-  these methods further once this filter is in place.
+
+  % Preliminary HyperNEAT runs were promising and resulted in several
+  % interesting gaits.
+
+  % Unfortunately, the gaits generated by HyperNEAT
+  % also tended to stress the robot more than typical gaits had before,
+  % and the servos would often overheat and malfunction, requiring
+  % restarts.  We think these issues may be addressed by adding a small
+  % layer between the HyperNEAT strategy and the robot that disallows
+  % quickly shifting commanded positions, and we hope to be able to test
+  % these methods further once this filter is in place.
 

### 2011-01-20 paper/acknowledgments.tex
Restructuring of report into multiple files to minimize conflicts, references working, dealt old sections to appropriate locations in paper framework.

--- 2010-12-07 report/report.tex
@@
-\section{Introduction}
+\section{Acknowledgments}
 
-% Motivate and abstractly describe the problem you are addressing and
-% how you are addressing it. What is the problem? Why is it important?
-% What is your basic approach? A short discussion of how it fits into
-% % related work in the area is also desirable. Summarize the basic
-% results and conclusions that you will present.
+Thanks to ..., some grant maybe, Jim T\o rreson, ...and the reviewers.
 
-Applications of walking robots often call for the ability to walk as
-quickly, efficiently, or with as little power as possible.  Often
-these optimizations are done manually by an expert who designs and
-tweaks a gait specifically for a given objective.  Other groups have
-used learning methods to generate gaits optimized for some metric.
-Approaches differ in their starting assumptions, some essentially
-tweaking the parameters of a hand-tuned model \cite{chernova}, others
-exploring a reasonably compact parameter space \cite{kohl}, and still
-others beginning with few assumptions besides periodicity
-\cite{zykov}.
+\edit{write this, old below}
 
-We aimed to strike a middle ground between these approaches.  Our
-motion generator did not rely on hand-tweaked gaits, but it did use
-parameterized gaits which, by their nature, impose some assumptions on
-the answers produced.  We then used machine learning to design gaits
-for a quadruped robot with these models.  This paper presents a
-comparison of the different learning methods implemented.  Most
-methods created walks that are several times faster than the original
-hand-tuned gait.  We invite readers with short attention spans to view
-a video of some of our results online here:
-
-\url{http://www.youtube.com/watch?v=ODoiOj9DdGg}
-
-%The code so far consists of the robot class, optimization class,
-%parameterized model, sine model, and camera feedback.
-
-\section{Problem definition}
-% Precisely define the problem you are addressing (i.e. formally specify
-% the inputs and outputs). Elaborate on why this is an interesting and
-% important problem.
-
-We are testing several different learning methods to design a
-parametrized gait for a quadruped robot from the Cornell Computational
-Synthesis Lab.
-
-The output each of the learning algorithms is a function of time,
-$f(t)$, that outputs a vector of commanded motor positions.  This
-function is generated using a parametrized motion model, described in
-\secref{motion}.
-
-The robot executes these commands and measures its change in location
-using the tracking system described in \secref{implement}.  The input
-to the learning algorithms is this measured displacement, which the
-algorithms attempt to maximize. This displacement is measured for each
-gait over a constant length run, usually 12 seconds.
-
-A comparison and evaluation of the many different methods available
-for optimizing the gait of legged robots will be useful for future
-work on this challenging multidimensional control problem.
-
-\section{Method}
-\seclabel{method}
-
-% Describe in reasonable detail the algorithm you are using to address
-% this problem. A pseudo-code description of the algorithm you are
-% using is frequently useful. If it makes sense for your project,
-% trace through a concrete example, showing how your algorithm
-% processes this example. The example should be complex enough to
-% illustrate all of the important aspects of the problem but simple
-% enough to be easily understood. If possible, an intuitively
-% meaningful example is better than one with meaningless symbols.
-
-We use several parameterized motion models that command motors to
-positions based on a sine wave, creating a periodic pattern.  While we
-investigated several models, for the bulk of our experiments, we used
-a model whose five parameters are: amplitude, wavelength, scale inner
-vs outer motors, scale left vs right motors, scale back vs front
-motors. Each strategy below attempts to choose the best possible
-parameters for this motion model.
-
-We implemented and tested 8 different learning strategies.  All
-strategies except for the HyperNEAT method\cite{clune} were
-constrained to pick parameters from within predetermined ranges.
-
-\begin{itemize}
-
-\item \emph{Random}: This method randomly generates parameter vectors
-  in the allowable range. This strategy is used only as baseline.
-
-\item \emph{Uniform random hill climbing}: This method begins by
-  selecting a single random parameter vector.  Subsequent iterations
-  generate a neighbor by randomly choosing one parameter to adjust and
-  replacing it with a new value chosen with uniform probability in the
-  allowable range for that parameter. The neighbor is evaluated by
-  running the robot with the newly chosen parameters. If this neighbor
-  results in a longer distance walked than the previous best gait, it
-  is saved as the new best gait. The process is then repeated, always
-  starting with the best gait.
-
-\item \emph{Gaussian random hill climbing}: This method works
-  similarly to Uniform random hill climbing, except neighbors are
-  generated by adding random Gaussian noise to the current best gait.
-  This results in all parameters being changed at once, but the
-  resulting vector is always fairly close to the previous best gait.
-  We used independently selected noise in each dimension, scaled such
-  that the standard deviation of the noise was 5\% of the range of
-  that dimension.
-
-\item \emph{N-dimensional policy gradient descent}: As opposed to the
-  previous methods, this method explicitly estimates the gradient for
-  the objective function. It does this by first evaluating \emph{t}
-  randomly generated parameter vectors near the initial vector, each
-  dimension of these vectors being perturbed by either $-\epsilon$,
-  $0$, or $\epsilon$. Then, for each dimension, it groups vectors into
-  three groups: $-\epsilon$, $0$, and $\epsilon$.  The gradient along
-  this dimension is then estimated as the average score for the
-  $\epsilon$ group minus the average score for the $-\epsilon$
-  group. Finally, the method creates a new vector by changing all
-  parameters by a fixed-size step in the direction of the gradient.
-
-\item \emph{Nelder-Mead simplex method}\cite{nm}: The Nelder-Mead
-  simplex method creates an initial simplex with 6 vertices. The
-  initial parameter vector is stored as the first vertex and the other
-  five vertices are created by adding to one dimension at a time one
-  tenth of the allowable range for that parameter. It then tests the
-  fitness of each vertex and based on these fitnesses, it reflects the
-  worst point over the centroid in an attempt to improve it.  However,
-  to prevent cycles and becoming stuck in local minima, several other
-  rules are used.  In general, the worst vertex is reflected over the
-  centroid. If the reflected point is better than the second worst
-  point and worse than the best point, then the reflected point
-  replaces the worst. If the reflected point is better than the best
-  point, the simplex is expanded in the direction of the reflected
-  point. The better of the reflected and the expanded point replaces
-  the worst point. If the reflected point is worse than the second
-  worst point, then the simplex is contracted away from the reflected
-  point. If the contracted point is better than the reflected point,
-  the contracted point replaces the worst point. If the contracted
-  point is worse than the reflected point, the entire simplex is
-  shrunk \cite{NMWebsite}.
-
-\item \emph{Linear regression}: To initialize, this method chooses and
-  evaluates five random parameter vectors. It then fits a linear model
-  from parameter vector to fitness. In a loop, the method chooses and
-  evaluates a new parameter vector generated by taking a fixed-size
-  step in the direction of the gradient for each parameter, and fits a
-  new linear model to all vectors evaluated so far, choosing the model
-  to minimize the sum of squared errors.
-
-\item \emph{SVM regression}: Similarly to linear regression, this
-  model starts with several random vectors, but this time they are
-  chosen in a small neighborhood about some initial random vector.
-  These vectors (generally 8) are evaluated, and a support vector
-  regression model is fit to the observed fitnesses.  To choose the
-  next vector for evaluation, we randomly generate some number
-  (typically 100) of vectors in the neighborhood of the best observed
-  gait, and select for evaluation the vector with the best predicted
-  performance.  We suspected that if we always chose the best
-  predicted point out of 100, we may end up progressing along a narrow
-  subspace, prohibiting learning of the true local fitness function.
-  Alternately, we would always choosing exploitation of knowledge
-  vs. exploration of the space.  To address this concern, we added a
-  parameter dubbed \code{bumpBy} that added noise to the final
-  selected point before it was evaluated.
-
-  Such a method naturally has many tunable parameters, and we
-  endeavored to select these parameteres by tuning the method in
-  simulation.  To estimate the performance of the algorithm, we ran it
-  agains a simulation with a known optimum.  The simulated function
-  was in the same five dimensional parameter space, and simply
-  returned a fitness determined as the height of a Gaussian with a
-  random mean.  The width of the Gaussian in each dimension was 20\%
-  of the range of each dimension, and the maximum value at the peak
-  was 100.  \figref{svm_sim_results} shows the learning results on
-  this simulated model using the ultimately selected SVM parameters.
-  Interestingly, a non-zero value of \code{bumpBy} resulted in better
-  learning than a zero valued, noise free (exploration free) learning.
-
-  Ultimately, however, the version of SVM tuned for simulation did not
-  yet show competitive performance on the real robot.  We tried tuning
-  some parameters on the real robot, but after some amount of tuning,
-  the method still exibited too little exploration and easily became
-  stuck in local minima.
-
-\item \emph{Evolutionary Neural Network (HyperNEAT)\cite{clune}}: Near
-  the end of term, we hacked together an interface between HyperNEAT
-  and the robot, requiring a slightly modified strategy interface.
-  Preliminary HyperNEAT runs were promising and resulted in serveral
-  interesting gaits.  Unfortunately, the gaits generated by HyperNEAT
-  also tended to stress the robot more than typical gaits had before,
-  and the servos would often overheat and malfunction, requiring
-  restart.  We think these issues may be addressed by adding a small
-  layer between the HyperNEAT strategy and the robot that disallows
-  quickly shifting commanded positions, and we hope to be able to test
-  these methods further once this filter is in place.
-
-\end{itemize}
-
-\figvarp{svm_sim_results}{.6}{Results for the SVM regression strategy
-  in simulation.  This simulation was used to tune the SVM strategy's
-  parameters before trying it on the physical robot.  The strategy
-  quickly approaches the maximum simulated fitness of 100.}{}
-
-\section{Related work}
-% Briefly explain who else worked on related problems in the past and what
-% methods they used. Explain if you are using similar methods, or if your
-% approach is different and if so - how (either is ok).
-
-Various maching learning techniques have proven to be useful in
-finding control policies for a wide variety of robots. Kohl and
-Stone\cite{kohl} presented a policy gradient reinforcement learning
-approach for generating a fast walk on legged robots. We experimented
-with this method to create a walk for our robot (Policy Gradient
-Descent). Chernova and Velosa\cite{chernova} took an evolutionary
-approach to this problem which we did not implement.  Zykov, Bongard,
-and Lipson\cite{zykov} describe the evolution of dynamic gaits on a
-physical robot requiring no prior assumptions about the locomotion
-pattern beyond the fact that it should be rhythmic.
-
-\section{System Architecture and Implementation}
-\seclabel{implement}
-% Describe how you implemented your system and how you structured it. 
-% This should give an overview of the system, not a detailed 
-% documentation of the code. The documentation of the code is part of 
-% the code you hand in. You might want to comment on high-level design 
-% decisions that you made. Also explain how you obtained your
-% data and any pre-processing you did to it.
-
-The quadruped robot has an on-board computer running Linux. The lower
-level drivers are in C and the system is implemented in
-Python. Feedback about distance travelled is provided via an infrared
-LED mounted on the robot and a Wii remote fixed to the ceiling.
-
-An overview of the code follows.
-
-\begin{itemize}
-
-\item \code{Robot} class: Class wrapper for commanding
-  motion of the robot.  The \code{Robot} class takes care of the robot
-  initialization, communication with the servos, and timing of the
-  runs.  In addition, it prevents the servos from ever being commanded
-  to a point outside their normal range (0 - 1023) as well as beyond
-  points where limbs would collide with parts of the robot body.  The
-  main class function, \code{run}, accepts a motion model (any
-  function that takes a time argument and outputs a 9 dimensional
-  position) and will run the robot using this motion model, including,
-  if desired, smooth interpolation over time for the beginning and end
-  of the run.
-
-\item \code{RunManager} class: Deals with all the details of running
-  the robot, including running the robot multiple times, tracking
-  distance walked via a \code{WiiTrackClient} member object, and
-  writing results to the log file. Also includes the
-  \code{explore\_dimensions} method to generate plots by varying each
-  parameter independently.
-
-\item \code{Strategy} class: The user has a choice between eight
-  different learning strategies: Random search, uniform random hill
-  climbing, Gaussian random hill climbing, N-dimensional policy
-  gradient descent, Nelder-Mead simplex, linear regression/prediction,
-  SVM regression/prediction, and HyperNEAT evolution.  Each strategy
-  must derive from the base \code{Strategy} class and must implement
-  two methods: \code{getNext}, for getting the next parameter vector
-  to try, and \code{updateResults}, for communicating results of a run
-  back to the strategy.
-
-\item \code{MotionModel} class: We implemented several motion models,
-  the main being the \code{SineModel5} class.  All motion models take
-  as input a parameter vector and produce as output a motion model,
-  which is simply a function mapping from time to nine motor
-  positions.  The \code{SineModel5} model commands positions based on
-  a sine wave, creating a periodic pattern. The parameters are:
-  amplitude, wavelength, a multiplier for the inner vs outer motors,
-  multiplier for left vs right motors, and multiplier for back vs
-  front motors.  Other similar models were tested, including a seven
-  parameter model which allowed sine waves shifted in time, but these
-  were not used as extensively in our experiments as
-  \code{SineModel5}.
-
-\item \code{WiiTrackClient} and \code{WiiTrackServer} classes and
-  hardware: A Wii remote tracks the location of the robot through an
-  infrared LED mounted on top of the robot. A \code{WiiTrackServer} is
-  run on the robot and continuously tracks its position using the
-  CWiid library\cite{cwiid} to interface with the remote via
-  bluetooth.  The \code{RunManager} then makes a
-  \code{WiiTrackClient}, which connects via socket to the tracking
-  server and requests position updates periodically.
-  \code{RunManager} currently gets the robot's position at the
-  beginning of each run and then again at the end and uses this to
-  calculate the net change in position.
-
-\end{itemize}
-
-\section{Experimental Evaluation}
-
-\subsection{Methodology}
-% What are the criteria you are using to evaluate your method? What
-% specific hypotheses does your experiment test? Describe the 
-% experimental methodology that you used. What are the dependent and 
-% independent variables? For projects in machine learning, what is 
-% the training/test data that was used, and why is it realistic or
-% interesting? Exactly what performance data did you collect and how 
-% are you presenting and analyzing it? Comparisons to competing methods 
-% that address the same problem or to variations of your own algorithm 
-% are particularly useful.
-
-The metric for evaluation of the designed gait is speed. We stop each 
-run after plateauing results (no improvement for one third of the 
-policies seen so far). The standard length of a run designates that it
-should be stopped after there is no improvement for one half of the policies
-seen so far, but since all runs took place on the actual robot, without use
-of a simulator, certain time limitations were imposed on the learning process.
-
-We controlled our experiments from a computer that was connected via a
-wireless ethernet to the robot. The robot collected data about
-distance walked automatically on its own. If it walked outside of the
-Wii remote's viewable area, it informed the user, so the only human
-intervention required during an experiment was to move the robot back
-inside the viewable area and to resume the run, which did not
-interrupt the learning process or result in the loss of data.
-
-We evaluated the efficacy of a set of parameters by sending these
-parameters to the robot and instructing it to walk for a certain
-length of time. The robot always began from the same position and
-returned to the starting position at the end of the run in order to
-measure true displacement without giving credit for ending in a leaned
-position. More efficient parameters resulted in a faster gait, which
-translated into a longer distance walked and a better score. After completing
-an evaluation, the robot sent the resulting distance walked back to the
-host computer and prepared itself for a new set of parameters to evaluate.
-
-Each algorithm was run on 3 different initial parameter vectors on the
-physical robot. We decided to evaluate all methods starting at the
-same three vector in order to allow for the fair comparison of each
-algorithm.  We evaluate each method based on the amount of improvement
-seen from the initial parameter vectors, and on the fastest speeds
-achieved during runs.
-
-The resulting gaits from our algorithms quickly outperformed the original
-hand-coded walk designed for this robot. The fastest walk, for example, was
-4 times faster.
-
-\subsection{Results}
-% Present the quantitative results of your experiments. Graphical data
-% presentation such as graphs and histograms are frequently better 
-% than tables. Explain what are the basic differences revealed in the 
-% data. If your method is stochastic or you analyze multiple datasets, 
-% calculate mean and standard error for every quantitative
-% performance result you show based on multiple runs.
-
-We have done complete runs of 3 different initial parameter vectors
-for random search, uniform random hill climbing, Gaussian random hill
-climbing, policy gradient descent, Nelder-Mead simplex, and linear
-regression.  We also evaluated the SVM regression and HyperNEAT
-methods, but these methods were more experimental, and thus we do not
-yet have the same volume of data for these runs.  We developed several
-gaits that were about 4 times faster than the original hand-coded
-gait.  Results are shown in \tabref{results_table}.
-
-\begin{table}
-\begin{center}
-\begin{tabular}{|r|c|c|c||c|}
-\hline
-                                         & A       & B      & C      &  Average \\
-\hline                                                               
-\hline                                                               
-Previous hand-coded gait                 & --      & --     & --     &  5.16 \\
-\hline                                                                 
-Random search                            & 6.04    & 17.26  & 4.90   &  9.40 \\
-\hline                                                                 
-Uniform Random Hill Climbing             & 11.37   & 9.44   & 2.69   &  7.83 \\
-\hline                                                                 
-Gaussian Random Hill Climbing            & 3.10    & 13.59  & 13.40  &  10.03 \\
-\hline                                                                 
-Policy Gradient Descent                  & 0.68    & 14.69  & 3.60   &  6.32 \\
-\hline                                                                 
-Nelder-Mead simplex                      & 8.51    & 13.62  & 14.83  &  12.32 \\
-\hline                                                                 
-Linear Regression                        & 27.58   & 12.51  & 1.95   &  14.01 \\
-%\hline                                                                
-%SVM Regression                           & ???     & ???    & ???   &   \\
-%\hline                                                                
-%Evolutionary Neural Network (HyperNEAT)  & ???     & ???    & ???   &   \\
-\hline
-\end{tabular}
-\caption{The best gaits found for each starting vector and algorithm,
-  in body lengths per minute.}
-\tablabel{results_table}
-\end{center}
-\end{table}
-
-\begin{itemize}
-
-\item For vector A, shown in \figref{vectorA}, linear regression
-  worked significantly better than the other algorithms, resulting in
-  a gait (27.58 body lengths/minute) that walked over twice as fast as
-  the next best gait -- uniform random hill climbing at 11.37 body
-  lengths/minute -- and 5.3 times better than the previous hand-coded
-  gait.
-
-\item Vector B, depicted in \figref{vectorB}, resulted in similarly
-  performing gaits with all the algorithms. The random method got
-  lucky and produced the best gait (17.26 body lengths/minute) and
-  uniform random hill climbing produced the worst (9.44 body
-  lengths/minute). The remaining algorithms each produced a gait
-  around 13 body lengths/minute.
-
-\item For vector C, shown in \figref{vectorC}, simplex and Gaussian
-  random hill climbing each produced a gait that substantially
-  outperformed the other algorithms. Simplex resulted in a gait of
-  nearly 15 body lengths/minute and Gaussian random hill climbing
-  produced a gait of just over 13 body lengths/minute, whereas the
-  other algorithms returned gaits of less than 5 body lengths/minute.
-
-\end{itemize}
-
-\figvarp{vectorA}{1}{Results for runs beginning with vector A.
-  Linear regression worked significantly better than the other
-  algorithms, resulting in a gait (27.58 body lengths/minute) that
-  walked over twice as fast as the next best gait -- uniform random
-  hill climbing at 11.37 body lengths/minute -- and 5.3 times better
-  than the previous hand-coded gait.}{}
-
-\figvarp{vectorB}{1}{Results for run starting with vector B.  All
-  algorithms performed similarly.  The random method actually got
-  lucky this time and produced the best gait (17.26 body
-  lengths/minute) and uniform random hill climbing produced the worst
-  (9.44 body lengths/minute). The remaining algorithms each produced a
-  gait around 13 body lengths/minute.}{}
-
-\figvarp{vectorC}{1}{Results for runs starting with vector C.
-  Simplex and Gaussian random hill climbing each produced a gait that
-  substantially outperformed the other algorithms, with around 15 body
-  lengths/minute and 13 body lengths/minute, respectively, whereas the
-  other algorithms returned gaits of less than 5 body
-  lengths/minute.}{}
-
-Based on the three trials, as shown in \figref{std_error}, linear
-regression worked the best, followed by simplex. Gaussian random hill
-climbing performed about as well as random, but it is unclear whether
-Gaussian would continue to improve and become much better than random
-search if allowed to run for more iterations. Uniform random hill
-climbing and policy gradient descent performed similarly to each other
-on average, but differed greatly on individual runs.
-
-\figvarp{std_error}{1}{Average results for each method starting with
-  parameter vectors A, B, and C.  Error bars are plotted showing the
-  standard error.  Linear regression outperformed other methods,
-  followed by simplex. Gaussian random hill climbing performed about
-  as well as random, but it is unclear whether Gaussian would continue
-  to improve and become much better than random search if allowed to
-  run for more iterations. Uniform random hill climbing and policy
-  gradient descent performed similarly to each other on average, but
-  differed greatly on individual runs.}{}
-
-In addition to running strategies to optimize parameterized gaits, we
-also wanted to investigate the space of possible gaits.  To accomplish
-this, we selected a parameter vector that resulted in motion, but not
-an exceptional gait, and plotted performance along each dimension
-individually.  In addition, we duplicated each measurement to be able
-to estimate the measurement noise at each point.  Results for this
-exploration are shown in \figref{explore_dim_1} through
-\figref{explore_dim_5}.
-
-\figvarp{explore_dim_1}{.5}{Motion for varying just dimension 1}{}
-
-\figvarp{explore_dim_2}{.5}{Motion for varying just dimension 2}{}
-
-\figvarp{explore_dim_3}{.5}{Motion for varying just dimension 3}{}
-
-\figvarp{explore_dim_4}{.5}{Motion for varying just dimension 4}{}
-
-\figvarp{explore_dim_5}{.5}{Motion for varying just dimension 5}{}
-
-\subsection{Discussion}
-% Is your hypothesis supported? What conclusions do the results support
-% about the strengths and weaknesses of your method compared to other 
-% methods? How can the results be explained in terms of the underlying 
-% properties of the algorithm and/or the data.
-
-Because only three trials were tested with each algorithm and no
-algorithm consistently outperformed the others, there is a large
-standard error for each algorithm, as show in \figref{std_error}. For
-this reason, it is unclear if any algorithm is superior to another in
-this application. More trials with each algorithm would be necessary
-to reach a definitive ranking of the performance of the algorithms.
-
-However, each algorithm discovered at least one gait of over 10 body
-lengths/minute, including random. For this reason, we speculate that
-the motion model is at least as critical as the learning algorithm
-used.
-
-\section{Future work}
-% What are the major shortcomings of your current method? For each 
-% shortcoming, propose additions or enhancements that would help overcome it.
-
-There are several directions in which we could continue our
-work. First, the error margins for our runs were large, so reducing
-them by running more trials would lead to more conclusive data. It
-would also be insightful to run our algorithms on different motion
-models.  We suspect that our choice of motion model influenced the
-results greatly, as even random choices in the space produced gaits
-that moved a significant fraction of the time (> 5\%).  It would be
-interesting to see how learning methods would perform using a model
-that included a much higher percentage of unproductive gaits. We also
-intend to experiment further with SVM regression and evolutionary
-algorithms/HyperNEAT.  Some parameter vectors resulted in the robot
-turning, as opposed to it walking long distances. This could be an
-interesting learning goal in future projects. To these ends, we
-propose the following additions and enhancements:
-
-\begin{itemize}
-\item More runs and/or longer runs
-\item Different motion representations
-\item Better tuning of SVM regression
-\item Evolutionary algorithms/HyperNEAT\cite{clune}
-\item Learning how to turn
-\end{itemize}
-
-\section{Conclusion}
-% Briefly summarize the important results and conclusions presented in 
-% the paper. What are the most important points illustrated by your 
-% work? How will your results improve future research and applications 
-% in the area?
-
-We have presented an array of approaches to optimizing a quadrupedal
-gait for forward speed.  We have implemented and tested different
-learning strategies, including uniform and Gaussian random hill
-climbing, policy gradient reinforcement learning, Nelder-Mead simplex,
-several new predictive methods based on linear and support vector
-regression, and an evolved neural network (HyperNEAT).  We have also
-compared these approaches to random search as a baseline. Many of the
-methods resulted in walks significantly faster than previously
-hand-tuned gaits.
-
-Because only three trials were tested with each algorithm and no
-algorithm consistently outperformed the others, there was a large
-standard error for each method, as shown in \figref{std_error}. Thus
-it was unclear if any algorithm was superior to another in this
-application. More trials with each algorithm would be necessary to
-reach a definitive ranking. Because each algorithm discovered at least one gait of
-over 10 body lengths/minute, including random search, we also
-conjectured that the motion representation for the robot is more
-integral to forward speed than the learning algorithm.  How to learn
-the motion representation, in addition to its parameters, remains an
-open problem.
-
-\section{Acknowledgments}
 % List any people not on the team who helped you with your project: Your
@@
 \end{itemize}
-
-\section{The Team}
-% A table listing the names of the people who worked on the project, and who
-% did what.
-
-\begin{itemize}
-
-\item Diana Hidalgo: 
-\begin{itemize}
-\item Simplex (Nelder-Mead)
-\item \code{WiiTrackClient} and \code{WiiTrackServer} classes and hardware
-\item Plotting and evaluating results
-\item Investigation into using Optitrack IR system for tracking robot
-\end{itemize}
-
-\item Sarah Nguyen
-\begin{itemize}
-\item Random search
-\item Uniform random hill climbing
-\item Gaussian random hill climbing
-\item N-dimensional policy gradient descent
-\item Linear regression and prediction
-\item \code{RunManager} and \code{Strategy} classes
-\item Video editing
-\end{itemize}
-
-\item Jason Yosinski
-\begin{itemize}
-\item SVM regression and prediction
-\item Evolutionary Neural Network (HyperNEAT)
-\item \code{WiiTrackClient} and \code{WiiTrackServer} classes and hardware
-\item \code{Robot}, \code{RunManager}, and \code{Strategy} classes
-\item \code{Parameterized Motion Model} abstract base class
-\item \code{SineModel} classes
-\item \code{explore\_dimensions}
-\item Plotting and evaluating results
-\end{itemize}
-
-\end{itemize}
-
-\section{Appendix}
-% Describe the code files you uploaded into CMS
-
-A brief description of the code uploaded to the CMS follows:
-
-\begin{itemize}
-\item \code{optimize.py}: Main program, determines a strategy to try and runs the robot
-  with that strategy.
-\item \code{RunManager.py}: Deals with all the details of running the robot,
-  including choosing an initial parameter, running the robot multiple times, 
-  tracking distance walked, and writing to the log file. Also includes the
-  \code{explore\_dimensions} method.
-\item \code{Strategy.py}: Contains all the different possible strategies, which
-  will be passed as objects in \code{optimize.py}. 
-\item \code{Robot.py}: Implements the \code{Robot} class, described in
-  \secref{implement}.
-\item \code{SineModel.py}: Implements a sine based motion model,
-  described in \secref{implement}.
-\item \code{Motion.py}: Motion helper functions.
-\item \code{WiiTrackServer.py}: Broadcasts the position of the infrared LED.
-\item \code{WiiTrackClient.py}: Connects to the WiiTrackServer to get the current position information.
-\end{itemize}
-
-% LocalWords:  vertices

### 2011-01-20 paper/approach.tex
Restructuring of report into multiple files to minimize conflicts, references working, dealt old sections to appropriate locations in paper framework.

--- 2011-01-14 paper/report.tex
@@
-\section{Introduction}
+\section{Experimental Approach}
 
-% Motivate and abstractly describe the problem you are addressing and
-% how you are addressing it. What is the problem? Why is it important?
-% What is your basic approach? A short discussion of how it fits into
-% % related work in the area is also desirable. Summarize the basic
-% results and conclusions that you will present.
+\subsection{Learning Methods}
 
-Applications of walking robots often call for the ability to walk as
-quickly, efficiently, or with as little power as possible.  Often
-these optimizations are done manually by an expert who designs and
-tweaks a gait specifically for a given objective.  Other groups have
-used learning methods to generate gaits optimized for some metric.
-Approaches differ in their starting assumptions, some essentially
-tweaking the parameters of a hand-tuned model \cite{chernova}, others
-exploring a reasonably compact parameter space \cite{kohl}, and still
-others beginning with few assumptions besides periodicity
-\cite{zykov}.
+\edit{write this.  Old quadratot Method section below}
 
-We aimed to strike a middle ground between these approaches.  Our
-motion generator did not rely on hand-tweaked gaits, but it did use
-parameterized gaits which, by their nature, impose some assumptions on
-the answers produced.  We then used machine learning to design gaits
-for a quadruped robot with these models.  This paper presents a
-comparison of the different learning methods implemented.  Most
-methods created walks that are several times faster than the original
-hand-tuned gait.  We invite readers with short attention spans to view
-a video of some of our results online here:
-
-%\url{http://www.youtube.com/watch?v=ODoiOj9DdGg}
-\texttt{http://www.youtube.com/watch?v=ODoiOj9DdGg}
-
-%The code so far consists of the robot class, optimization class,
-%parameterized model, sine model, and camera feedback.
-
-\section{Problem definition}
-% Precisely define the problem you are addressing (i.e. formally specify
-% the inputs and outputs). Elaborate on why this is an interesting and
-% important problem.
-
-We are testing several different learning methods to design a
-parametrized gait for a quadruped robot from the Cornell Computational
-Synthesis Lab.
-
-The output each of the learning algorithms is a function of time,
-$f(t)$, that outputs a vector of commanded motor positions.  This
-function is generated using a parametrized motion model, described in
-\secref{implement}.
-
-The robot executes these commands and measures its change in location
-using the tracking system described in \secref{implement}.  The input
-to the learning algorithms is this measured displacement, which the
-algorithms attempt to maximize. This displacement is measured for each
-gait over a constant length run, usually 12 seconds.
-
-A comparison and evaluation of the many different methods available
-for optimizing the gait of legged robots will be useful for future
-work on this challenging multidimensional control problem.
-
-\section{Method}
+%\section{Method}
 \seclabel{method}
@@
 
-\section{Related work}
-% Briefly explain who else worked on related problems in the past and what
-% methods they used. Explain if you are using similar methods, or if your
-% approach is different and if so - how (either is ok).
+\subsection{Platform details}
 
-Various maching learning techniques have proven to be useful in
-finding control policies for a wide variety of robots. Kohl and
-Stone\cite{kohl} presented a policy gradient reinforcement learning
-approach for generating a fast walk on legged robots. We experimented
-with this method to create a walk for our robot (Policy Gradient
-Descent). Chernova and Velosa\cite{chernova} took an evolutionary
-approach to this problem which we did not implement.  Zykov, Bongard,
-and Lipson\cite{zykov} describe the evolution of dynamic gaits on a
-physical robot requiring no prior assumptions about the locomotion
-pattern beyond the fact that it should be rhythmic.
+\edit{write this.  Old quadratot System Architecture and Implementation section below.}
 
-\section{System Architecture and Implementation}
+%\section{System Architecture and Implementation}
 \seclabel{implement}
+
 % Describe how you implemented your system and how you structured it. 
@@
 
-\section{Experimental Evaluation}
+\subsection{Fitness evaluation details}
 
-\subsection{Methodology}
+\edit{write this.  Old quadratot Experimental Evaluation section below.}
+
+%\section{Experimental Evaluation}
+
+%\subsection{Methodology}
 % What are the criteria you are using to evaluate your method? What
@@
 4 times faster.
-
-\subsection{Results}
-% Present the quantitative results of your experiments. Graphical data
-% presentation such as graphs and histograms are frequently better 
-% than tables. Explain what are the basic differences revealed in the 
-% data. If your method is stochastic or you analyze multiple datasets, 
-% calculate mean and standard error for every quantitative
-% performance result you show based on multiple runs.
-
-We have done complete runs of 3 different initial parameter vectors
-for random search, uniform random hill climbing, Gaussian random hill
-climbing, policy gradient descent, Nelder-Mead simplex, and linear
-regression.  We also evaluated the SVM regression and HyperNEAT
-methods, but these methods were more experimental, and thus we do not
-yet have the same volume of data for these runs.  We developed several
-gaits that were about 4 times faster than the original hand-coded
-gait.  Results are shown in \tabref{results_table}.
-
-\begin{table}
-\begin{center}
-\begin{tabular}{|r|c|c|c||c|}
-\hline
-                                         & A       & B      & C      &  Average \\
-\hline                                                               
-\hline                                                               
-Previous hand-coded gait                 & --      & --     & --     &  5.16 \\
-\hline                                                                 
-Random search                            & 6.04    & 17.26  & 4.90   &  9.40 \\
-\hline                                                                 
-Uniform Random Hill Climbing             & 11.37   & 9.44   & 2.69   &  7.83 \\
-\hline                                                                 
-Gaussian Random Hill Climbing            & 3.10    & 13.59  & 13.40  &  10.03 \\
-\hline                                                                 
-Policy Gradient Descent                  & 0.68    & 14.69  & 3.60   &  6.32 \\
-\hline                                                                 
-Nelder-Mead simplex                      & 8.51    & 13.62  & 14.83  &  12.32 \\
-\hline                                                                 
-Linear Regression                        & 27.58   & 12.51  & 1.95   &  14.01 \\
-%\hline                                                                
-%SVM Regression                           & ???     & ???    & ???   &   \\
-%\hline                                                                
-%Evolutionary Neural Network (HyperNEAT)  & ???     & ???    & ???   &   \\
-\hline
-\end{tabular}
-\caption{The best gaits found for each starting vector and algorithm,
-  in body lengths per minute.}
-\tablabel{results_table}
-\end{center}
-\end{table}
-
-\begin{itemize}
-
-\item For vector A, shown in \figref{vectorA}, linear regression
-  worked significantly better than the other algorithms, resulting in
-  a gait (27.58 body lengths/minute) that walked over twice as fast as
-  the next best gait -- uniform random hill climbing at 11.37 body
-  lengths/minute -- and 5.3 times better than the previous hand-coded
-  gait.
-
-\item Vector B, depicted in \figref{vectorB}, resulted in similarly
-  performing gaits with all the algorithms. The random method got
-  lucky and produced the best gait (17.26 body lengths/minute) and
-  uniform random hill climbing produced the worst (9.44 body
-  lengths/minute). The remaining algorithms each produced a gait
-  around 13 body lengths/minute.
-
-\item For vector C, shown in \figref{vectorC}, simplex and Gaussian
-  random hill climbing each produced a gait that substantially
-  outperformed the other algorithms. Simplex resulted in a gait of
-  nearly 15 body lengths/minute and Gaussian random hill climbing
-  produced a gait of just over 13 body lengths/minute, whereas the
-  other algorithms returned gaits of less than 5 body lengths/minute.
-
-\end{itemize}
-
-\figvarp{vectorA}{1}{Results for runs beginning with vector A.
-  Linear regression worked significantly better than the other
-  algorithms, resulting in a gait (27.58 body lengths/minute) that
-  walked over twice as fast as the next best gait -- uniform random
-  hill climbing at 11.37 body lengths/minute -- and 5.3 times better
-  than the previous hand-coded gait.}{}
-
-\figvarp{vectorB}{1}{Results for run starting with vector B.  All
-  algorithms performed similarly.  The random method actually got
-  lucky this time and produced the best gait (17.26 body
-  lengths/minute) and uniform random hill climbing produced the worst
-  (9.44 body lengths/minute). The remaining algorithms each produced a
-  gait around 13 body lengths/minute.}{}
-
-\figvarp{vectorC}{1}{Results for runs starting with vector C.
-  Simplex and Gaussian random hill climbing each produced a gait that
-  substantially outperformed the other algorithms, with around 15 body
-  lengths/minute and 13 body lengths/minute, respectively, whereas the
-  other algorithms returned gaits of less than 5 body
-  lengths/minute.}{}
-
-Based on the three trials, as shown in \figref{std_error}, linear
-regression worked the best, followed by simplex. Gaussian random hill
-climbing performed about as well as random, but it is unclear whether
-Gaussian would continue to improve and become much better than random
-search if allowed to run for more iterations. Uniform random hill
-climbing and policy gradient descent performed similarly to each other
-on average, but differed greatly on individual runs.
-
-\figvarp{std_error}{1}{Average results for each method starting with
-  parameter vectors A, B, and C.  Error bars are plotted showing the
-  standard error.  Linear regression outperformed other methods,
-  followed by simplex. Gaussian random hill climbing performed about
-  as well as random, but it is unclear whether Gaussian would continue
-  to improve and become much better than random search if allowed to
-  run for more iterations. Uniform random hill climbing and policy
-  gradient descent performed similarly to each other on average, but
-  differed greatly on individual runs.}{}
-
-In addition to running strategies to optimize parameterized gaits, we
-also wanted to investigate the space of possible gaits.  To accomplish
-this, we selected a parameter vector that resulted in motion, but not
-an exceptional gait, and plotted performance along each dimension
-individually.  In addition, we duplicated each measurement to be able
-to estimate the measurement noise at each point.  Results for this
-exploration are shown in \figref{explore_dim_1} through
-\figref{explore_dim_5}.  As is shown in the figures, some dimensions
-are smoother than others, and some measurements are fairly noisy.
-
-\figvarp{explore_dim_1}{.65}{Fitness mean and standard deviation
-  vs. dimension 1.  The circle is a common point in
-  \figref{explore_dim_1} through \figref{explore_dim_5}}{}
-\figvarp{explore_dim_2}{.65}{Fitness mean and standard deviation
-  vs. dimension 2.  The circle is a common point in
-  \figref{explore_dim_1} through \figref{explore_dim_5}}{}
-\figvarp{explore_dim_3}{.65}{Fitness mean and standard deviation
-  vs. dimension 3.  The circle is a common point in
-  \figref{explore_dim_1} through \figref{explore_dim_5}}{}
-\figvarp{explore_dim_4}{.65}{Fitness mean and standard deviation
-  vs. dimension 4.  The circle is a common point in
-  \figref{explore_dim_1} through \figref{explore_dim_5}}{}
-\figvarp{explore_dim_5}{.65}{Fitness mean and standard deviation
-  vs. dimension 5.  The circle is a common point in
-  \figref{explore_dim_1} through \figref{explore_dim_5}}{}
-
-\subsection{Discussion}
-% Is your hypothesis supported? What conclusions do the results support
-% about the strengths and weaknesses of your method compared to other 
-% methods? How can the results be explained in terms of the underlying 
-% properties of the algorithm and/or the data.
-
-Because only three trials were tested with each algorithm and no
-algorithm consistently outperformed the others, there is a large
-standard error for each algorithm, as show in \figref{std_error}. For
-this reason, it is unclear if any algorithm is superior to another in
-this application. More trials with each algorithm would be necessary
-to reach a definitive ranking of the performance of the algorithms.
-
-However, each algorithm discovered at least one gait of over 10 body
-lengths/minute, including random. For this reason, we speculate that
-the motion model is at least as critical as the learning algorithm
-used.
-
-\section{Future work}
-% What are the major shortcomings of your current method? For each 
-% shortcoming, propose additions or enhancements that would help overcome it.
-
-There are several directions in which we could continue our
-work. First, the error margins for our runs were large, so reducing
-them by running more trials would lead to more conclusive data. It
-would also be insightful to run our algorithms on different motion
-models.  We suspect that our choice of motion model influenced the
-results greatly, as even random choices in the space produced gaits
-that moved a significant fraction of the time (> 5\%).  It would be
-interesting to see how learning methods would perform using a model
-that included a much higher percentage of unproductive gaits. We also
-intend to experiment further with SVM regression and evolutionary
-algorithms/HyperNEAT.  Some parameter vectors resulted in the robot
-turning, as opposed to it walking long distances. This could be an
-interesting learning goal in future projects. To these ends, we
-propose the following additions and enhancements:
-
-\begin{itemize}
-\item More runs and/or longer runs
-\item Different motion representations
-\item Better tuning of SVM regression
-\item Evolutionary algorithms/HyperNEAT\cite{clune}
-\item Learning how to turn
-\end{itemize}
-
-\section{Conclusion}
-% Briefly summarize the important results and conclusions presented in 
-% the paper. What are the most important points illustrated by your 
-% work? How will your results improve future research and applications 
-% in the area?
-
-We have presented an array of approaches to optimizing a quadrupedal
-gait for forward speed.  We have implemented and tested different
-learning strategies, including uniform and Gaussian random hill
-climbing, policy gradient reinforcement learning, Nelder-Mead simplex,
-several new predictive methods based on linear and support vector
-regression, and an evolved neural network (HyperNEAT).  We have also
-compared these approaches to random search as a baseline. Many of the
-methods resulted in walks significantly faster than previously
-hand-tuned gaits.
-
-Because only three trials were tested with each algorithm and no
-algorithm consistently outperformed the others, there was a large
-standard error for each method, as shown in \figref{std_error}. Thus
-it was unclear if any algorithm was superior to another in this
-application. More trials with each algorithm would be necessary to
-reach a definitive ranking. Because each algorithm discovered at least one gait of
-over 10 body lengths/minute, including random search, we also
-conjectured that the motion representation for the robot is more
-integral to forward speed than the learning algorithm.  How to learn
-the motion representation, in addition to its parameters, remains an
-open problem.
-
-\section{Acknowledgments}
-% List any people not on the team who helped you with your project: Your
-% TA, other people you consulted with or had useful discussions (say in 
-% a word or two what they did), people who proofread your report, and 
-% any external code you used (libraries etc).
-\begin{itemize}
-\item Hod Lipson, Cornell Computational Synthesis Lab: adviser.
-\item Jim T\o rreson, University of Oslo: adviser.
-\item Juan Zagal, University of Chile: designed and printed robot and provided code for hand tuned gait.
-\item Jeff Clune, CCSL Red Couch: collaborated on HyperNEAT implementation/testing.
-\item Cooper Bills, Cornell University: assisted with Wii tracker development.
-\item Anshumali Srivastava, Cornell University: Teaching Assistant.
-\end{itemize}
-
-\section{The Team}
-% A table listing the names of the people who worked on the project, and who
-% did what.
-
-\begin{itemize}
-
-\item Diana Hidalgo: 
-\begin{itemize}
-\item Simplex (Nelder-Mead)
-\item \code{WiiTrackClient} and \code{WiiTrackServer} classes and hardware
-\item Plotting and evaluating results
-\item Investigation into using Optitrack IR system for tracking robot
-\end{itemize}
-
-\item Sarah Nguyen
-\begin{itemize}
-\item Random search
-\item Uniform random hill climbing
-\item Gaussian random hill climbing
-\item N-dimensional policy gradient descent
-\item Linear regression and prediction
-\item \code{RunManager} and \code{Strategy} classes
-\item Video editing
-\end{itemize}
-
-\item Jason Yosinski
-\begin{itemize}
-\item SVM regression and prediction
-\item Evolutionary Neural Network (HyperNEAT)
-\item \code{WiiTrackClient} and \code{WiiTrackServer} classes and hardware
-\item \code{Robot}, \code{RunManager}, and \code{Strategy} classes
-\item \code{Parameterized Motion Model} abstract base class
-\item \code{SineModel} classes
-\item \code{explore\_dimensions}
-\item Plotting and evaluating results
-\end{itemize}
-
-\end{itemize}
-
-\section{Appendix}
-% Describe the code files you uploaded into CMS
-
-A brief description of the code uploaded to the CMS follows:
-
-\begin{itemize}
-\item \code{optimize.py}: Main program, determines a strategy to try and runs the robot
-  with that strategy.
-\item \code{RunManager.py}: Deals with all the details of running the robot,
-  including choosing an initial parameter, running the robot multiple times, 
-  tracking distance walked, and writing to the log file. Also includes the
-  \code{explore\_dimensions} method.
-\item \code{Strategy.py}: Contains all the different possible strategies, which
-  will be passed as objects in \code{optimize.py}. 
-\item \code{Robot.py}: Implements the \code{Robot} class, described in
-  \secref{implement}.
-\item \code{SineModel.py}: Implements a sine based motion model,
-  described in \secref{implement}.
-\item \code{Motion.py}: Motion helper functions.
-\item \code{WiiTrackServer.py}: Broadcasts the position of the infrared LED.
-\item \code{WiiTrackClient.py}: Connects to the WiiTrackServer to get the current position information.
-\end{itemize}
-
-% LocalWords:  vertices

### 2011-01-20 paper/conclusion.tex
Restructuring of report into multiple files to minimize conflicts, references working, dealt old sections to appropriate locations in paper framework.

--- 2010-12-07 report/report.tex
@@
-\section{Introduction}
+\section{Conclusion and Future Work}
 
-% Motivate and abstractly describe the problem you are addressing and
-% how you are addressing it. What is the problem? Why is it important?
-% What is your basic approach? A short discussion of how it fits into
-% % related work in the area is also desirable. Summarize the basic
-% results and conclusions that you will present.
+\edit{write this, old section below}
 
-Applications of walking robots often call for the ability to walk as
-quickly, efficiently, or with as little power as possible.  Often
-these optimizations are done manually by an expert who designs and
-tweaks a gait specifically for a given objective.  Other groups have
-used learning methods to generate gaits optimized for some metric.
-Approaches differ in their starting assumptions, some essentially
-tweaking the parameters of a hand-tuned model \cite{chernova}, others
-exploring a reasonably compact parameter space \cite{kohl}, and still
-others beginning with few assumptions besides periodicity
-\cite{zykov}.
-
-We aimed to strike a middle ground between these approaches.  Our
-motion generator did not rely on hand-tweaked gaits, but it did use
-parameterized gaits which, by their nature, impose some assumptions on
-the answers produced.  We then used machine learning to design gaits
-for a quadruped robot with these models.  This paper presents a
-comparison of the different learning methods implemented.  Most
-methods created walks that are several times faster than the original
-hand-tuned gait.  We invite readers with short attention spans to view
-a video of some of our results online here:
-
-\url{http://www.youtube.com/watch?v=ODoiOj9DdGg}
-
-%The code so far consists of the robot class, optimization class,
-%parameterized model, sine model, and camera feedback.
-
-\section{Problem definition}
-% Precisely define the problem you are addressing (i.e. formally specify
-% the inputs and outputs). Elaborate on why this is an interesting and
-% important problem.
-
-We are testing several different learning methods to design a
-parametrized gait for a quadruped robot from the Cornell Computational
-Synthesis Lab.
-
-The output each of the learning algorithms is a function of time,
-$f(t)$, that outputs a vector of commanded motor positions.  This
-function is generated using a parametrized motion model, described in
-\secref{motion}.
-
-The robot executes these commands and measures its change in location
-using the tracking system described in \secref{implement}.  The input
-to the learning algorithms is this measured displacement, which the
-algorithms attempt to maximize. This displacement is measured for each
-gait over a constant length run, usually 12 seconds.
-
-A comparison and evaluation of the many different methods available
-for optimizing the gait of legged robots will be useful for future
-work on this challenging multidimensional control problem.
-
-\section{Method}
-\seclabel{method}
-
-% Describe in reasonable detail the algorithm you are using to address
-% this problem. A pseudo-code description of the algorithm you are
-% using is frequently useful. If it makes sense for your project,
-% trace through a concrete example, showing how your algorithm
-% processes this example. The example should be complex enough to
-% illustrate all of the important aspects of the problem but simple
-% enough to be easily understood. If possible, an intuitively
-% meaningful example is better than one with meaningless symbols.
-
-We use several parameterized motion models that command motors to
-positions based on a sine wave, creating a periodic pattern.  While we
-investigated several models, for the bulk of our experiments, we used
-a model whose five parameters are: amplitude, wavelength, scale inner
-vs outer motors, scale left vs right motors, scale back vs front
-motors. Each strategy below attempts to choose the best possible
-parameters for this motion model.
-
-We implemented and tested 8 different learning strategies.  All
-strategies except for the HyperNEAT method\cite{clune} were
-constrained to pick parameters from within predetermined ranges.
-
-\begin{itemize}
-
-\item \emph{Random}: This method randomly generates parameter vectors
-  in the allowable range. This strategy is used only as baseline.
-
-\item \emph{Uniform random hill climbing}: This method begins by
-  selecting a single random parameter vector.  Subsequent iterations
-  generate a neighbor by randomly choosing one parameter to adjust and
-  replacing it with a new value chosen with uniform probability in the
-  allowable range for that parameter. The neighbor is evaluated by
-  running the robot with the newly chosen parameters. If this neighbor
-  results in a longer distance walked than the previous best gait, it
-  is saved as the new best gait. The process is then repeated, always
-  starting with the best gait.
-
-\item \emph{Gaussian random hill climbing}: This method works
-  similarly to Uniform random hill climbing, except neighbors are
-  generated by adding random Gaussian noise to the current best gait.
-  This results in all parameters being changed at once, but the
-  resulting vector is always fairly close to the previous best gait.
-  We used independently selected noise in each dimension, scaled such
-  that the standard deviation of the noise was 5\% of the range of
-  that dimension.
-
-\item \emph{N-dimensional policy gradient descent}: As opposed to the
-  previous methods, this method explicitly estimates the gradient for
-  the objective function. It does this by first evaluating \emph{t}
-  randomly generated parameter vectors near the initial vector, each
-  dimension of these vectors being perturbed by either $-\epsilon$,
-  $0$, or $\epsilon$. Then, for each dimension, it groups vectors into
-  three groups: $-\epsilon$, $0$, and $\epsilon$.  The gradient along
-  this dimension is then estimated as the average score for the
-  $\epsilon$ group minus the average score for the $-\epsilon$
-  group. Finally, the method creates a new vector by changing all
-  parameters by a fixed-size step in the direction of the gradient.
-
-\item \emph{Nelder-Mead simplex method}\cite{nm}: The Nelder-Mead
-  simplex method creates an initial simplex with 6 vertices. The
-  initial parameter vector is stored as the first vertex and the other
-  five vertices are created by adding to one dimension at a time one
-  tenth of the allowable range for that parameter. It then tests the
-  fitness of each vertex and based on these fitnesses, it reflects the
-  worst point over the centroid in an attempt to improve it.  However,
-  to prevent cycles and becoming stuck in local minima, several other
-  rules are used.  In general, the worst vertex is reflected over the
-  centroid. If the reflected point is better than the second worst
-  point and worse than the best point, then the reflected point
-  replaces the worst. If the reflected point is better than the best
-  point, the simplex is expanded in the direction of the reflected
-  point. The better of the reflected and the expanded point replaces
-  the worst point. If the reflected point is worse than the second
-  worst point, then the simplex is contracted away from the reflected
-  point. If the contracted point is better than the reflected point,
-  the contracted point replaces the worst point. If the contracted
-  point is worse than the reflected point, the entire simplex is
-  shrunk \cite{NMWebsite}.
-
-\item \emph{Linear regression}: To initialize, this method chooses and
-  evaluates five random parameter vectors. It then fits a linear model
-  from parameter vector to fitness. In a loop, the method chooses and
-  evaluates a new parameter vector generated by taking a fixed-size
-  step in the direction of the gradient for each parameter, and fits a
-  new linear model to all vectors evaluated so far, choosing the model
-  to minimize the sum of squared errors.
-
-\item \emph{SVM regression}: Similarly to linear regression, this
-  model starts with several random vectors, but this time they are
-  chosen in a small neighborhood about some initial random vector.
-  These vectors (generally 8) are evaluated, and a support vector
-  regression model is fit to the observed fitnesses.  To choose the
-  next vector for evaluation, we randomly generate some number
-  (typically 100) of vectors in the neighborhood of the best observed
-  gait, and select for evaluation the vector with the best predicted
-  performance.  We suspected that if we always chose the best
-  predicted point out of 100, we may end up progressing along a narrow
-  subspace, prohibiting learning of the true local fitness function.
-  Alternately, we would always choosing exploitation of knowledge
-  vs. exploration of the space.  To address this concern, we added a
-  parameter dubbed \code{bumpBy} that added noise to the final
-  selected point before it was evaluated.
-
-  Such a method naturally has many tunable parameters, and we
-  endeavored to select these parameteres by tuning the method in
-  simulation.  To estimate the performance of the algorithm, we ran it
-  agains a simulation with a known optimum.  The simulated function
-  was in the same five dimensional parameter space, and simply
-  returned a fitness determined as the height of a Gaussian with a
-  random mean.  The width of the Gaussian in each dimension was 20\%
-  of the range of each dimension, and the maximum value at the peak
-  was 100.  \figref{svm_sim_results} shows the learning results on
-  this simulated model using the ultimately selected SVM parameters.
-  Interestingly, a non-zero value of \code{bumpBy} resulted in better
-  learning than a zero valued, noise free (exploration free) learning.
-
-  Ultimately, however, the version of SVM tuned for simulation did not
-  yet show competitive performance on the real robot.  We tried tuning
-  some parameters on the real robot, but after some amount of tuning,
-  the method still exibited too little exploration and easily became
-  stuck in local minima.
-
-\item \emph{Evolutionary Neural Network (HyperNEAT)\cite{clune}}: Near
-  the end of term, we hacked together an interface between HyperNEAT
-  and the robot, requiring a slightly modified strategy interface.
-  Preliminary HyperNEAT runs were promising and resulted in serveral
-  interesting gaits.  Unfortunately, the gaits generated by HyperNEAT
-  also tended to stress the robot more than typical gaits had before,
-  and the servos would often overheat and malfunction, requiring
-  restart.  We think these issues may be addressed by adding a small
-  layer between the HyperNEAT strategy and the robot that disallows
-  quickly shifting commanded positions, and we hope to be able to test
-  these methods further once this filter is in place.
-
-\end{itemize}
-
-\figvarp{svm_sim_results}{.6}{Results for the SVM regression strategy
-  in simulation.  This simulation was used to tune the SVM strategy's
-  parameters before trying it on the physical robot.  The strategy
-  quickly approaches the maximum simulated fitness of 100.}{}
-
-\section{Related work}
-% Briefly explain who else worked on related problems in the past and what
-% methods they used. Explain if you are using similar methods, or if your
-% approach is different and if so - how (either is ok).
-
-Various maching learning techniques have proven to be useful in
-finding control policies for a wide variety of robots. Kohl and
-Stone\cite{kohl} presented a policy gradient reinforcement learning
-approach for generating a fast walk on legged robots. We experimented
-with this method to create a walk for our robot (Policy Gradient
-Descent). Chernova and Velosa\cite{chernova} took an evolutionary
-approach to this problem which we did not implement.  Zykov, Bongard,
-and Lipson\cite{zykov} describe the evolution of dynamic gaits on a
-physical robot requiring no prior assumptions about the locomotion
-pattern beyond the fact that it should be rhythmic.
-
-\section{System Architecture and Implementation}
-\seclabel{implement}
-% Describe how you implemented your system and how you structured it. 
-% This should give an overview of the system, not a detailed 
-% documentation of the code. The documentation of the code is part of 
-% the code you hand in. You might want to comment on high-level design 
-% decisions that you made. Also explain how you obtained your
-% data and any pre-processing you did to it.
-
-The quadruped robot has an on-board computer running Linux. The lower
-level drivers are in C and the system is implemented in
-Python. Feedback about distance travelled is provided via an infrared
-LED mounted on the robot and a Wii remote fixed to the ceiling.
-
-An overview of the code follows.
-
-\begin{itemize}
-
-\item \code{Robot} class: Class wrapper for commanding
-  motion of the robot.  The \code{Robot} class takes care of the robot
-  initialization, communication with the servos, and timing of the
-  runs.  In addition, it prevents the servos from ever being commanded
-  to a point outside their normal range (0 - 1023) as well as beyond
-  points where limbs would collide with parts of the robot body.  The
-  main class function, \code{run}, accepts a motion model (any
-  function that takes a time argument and outputs a 9 dimensional
-  position) and will run the robot using this motion model, including,
-  if desired, smooth interpolation over time for the beginning and end
-  of the run.
-
-\item \code{RunManager} class: Deals with all the details of running
-  the robot, including running the robot multiple times, tracking
-  distance walked via a \code{WiiTrackClient} member object, and
-  writing results to the log file. Also includes the
-  \code{explore\_dimensions} method to generate plots by varying each
-  parameter independently.
-
-\item \code{Strategy} class: The user has a choice between eight
-  different learning strategies: Random search, uniform random hill
-  climbing, Gaussian random hill climbing, N-dimensional policy
-  gradient descent, Nelder-Mead simplex, linear regression/prediction,
-  SVM regression/prediction, and HyperNEAT evolution.  Each strategy
-  must derive from the base \code{Strategy} class and must implement
-  two methods: \code{getNext}, for getting the next parameter vector
-  to try, and \code{updateResults}, for communicating results of a run
-  back to the strategy.
-
-\item \code{MotionModel} class: We implemented several motion models,
-  the main being the \code{SineModel5} class.  All motion models take
-  as input a parameter vector and produce as output a motion model,
-  which is simply a function mapping from time to nine motor
-  positions.  The \code{SineModel5} model commands positions based on
-  a sine wave, creating a periodic pattern. The parameters are:
-  amplitude, wavelength, a multiplier for the inner vs outer motors,
-  multiplier for left vs right motors, and multiplier for back vs
-  front motors.  Other similar models were tested, including a seven
-  parameter model which allowed sine waves shifted in time, but these
-  were not used as extensively in our experiments as
-  \code{SineModel5}.
-
-\item \code{WiiTrackClient} and \code{WiiTrackServer} classes and
-  hardware: A Wii remote tracks the location of the robot through an
-  infrared LED mounted on top of the robot. A \code{WiiTrackServer} is
-  run on the robot and continuously tracks its position using the
-  CWiid library\cite{cwiid} to interface with the remote via
-  bluetooth.  The \code{RunManager} then makes a
-  \code{WiiTrackClient}, which connects via socket to the tracking
-  server and requests position updates periodically.
-  \code{RunManager} currently gets the robot's position at the
-  beginning of each run and then again at the end and uses this to
-  calculate the net change in position.
-
-\end{itemize}
-
-\section{Experimental Evaluation}
-
-\subsection{Methodology}
-% What are the criteria you are using to evaluate your method? What
-% specific hypotheses does your experiment test? Describe the 
-% experimental methodology that you used. What are the dependent and 
-% independent variables? For projects in machine learning, what is 
-% the training/test data that was used, and why is it realistic or
-% interesting? Exactly what performance data did you collect and how 
-% are you presenting and analyzing it? Comparisons to competing methods 
-% that address the same problem or to variations of your own algorithm 
-% are particularly useful.
-
-The metric for evaluation of the designed gait is speed. We stop each 
-run after plateauing results (no improvement for one third of the 
-policies seen so far). The standard length of a run designates that it
-should be stopped after there is no improvement for one half of the policies
-seen so far, but since all runs took place on the actual robot, without use
-of a simulator, certain time limitations were imposed on the learning process.
-
-We controlled our experiments from a computer that was connected via a
-wireless ethernet to the robot. The robot collected data about
-distance walked automatically on its own. If it walked outside of the
-Wii remote's viewable area, it informed the user, so the only human
-intervention required during an experiment was to move the robot back
-inside the viewable area and to resume the run, which did not
-interrupt the learning process or result in the loss of data.
-
-We evaluated the efficacy of a set of parameters by sending these
-parameters to the robot and instructing it to walk for a certain
-length of time. The robot always began from the same position and
-returned to the starting position at the end of the run in order to
-measure true displacement without giving credit for ending in a leaned
-position. More efficient parameters resulted in a faster gait, which
-translated into a longer distance walked and a better score. After completing
-an evaluation, the robot sent the resulting distance walked back to the
-host computer and prepared itself for a new set of parameters to evaluate.
-
-Each algorithm was run on 3 different initial parameter vectors on the
-physical robot. We decided to evaluate all methods starting at the
-same three vector in order to allow for the fair comparison of each
-algorithm.  We evaluate each method based on the amount of improvement
-seen from the initial parameter vectors, and on the fastest speeds
-achieved during runs.
-
-The resulting gaits from our algorithms quickly outperformed the original
-hand-coded walk designed for this robot. The fastest walk, for example, was
-4 times faster.
-
-\subsection{Results}
-% Present the quantitative results of your experiments. Graphical data
-% presentation such as graphs and histograms are frequently better 
-% than tables. Explain what are the basic differences revealed in the 
-% data. If your method is stochastic or you analyze multiple datasets, 
-% calculate mean and standard error for every quantitative
-% performance result you show based on multiple runs.
-
-We have done complete runs of 3 different initial parameter vectors
-for random search, uniform random hill climbing, Gaussian random hill
-climbing, policy gradient descent, Nelder-Mead simplex, and linear
-regression.  We also evaluated the SVM regression and HyperNEAT
-methods, but these methods were more experimental, and thus we do not
-yet have the same volume of data for these runs.  We developed several
-gaits that were about 4 times faster than the original hand-coded
-gait.  Results are shown in \tabref{results_table}.
-
-\begin{table}
-\begin{center}
-\begin{tabular}{|r|c|c|c||c|}
-\hline
-                                         & A       & B      & C      &  Average \\
-\hline                                                               
-\hline                                                               
-Previous hand-coded gait                 & --      & --     & --     &  5.16 \\
-\hline                                                                 
-Random search                            & 6.04    & 17.26  & 4.90   &  9.40 \\
-\hline                                                                 
-Uniform Random Hill Climbing             & 11.37   & 9.44   & 2.69   &  7.83 \\
-\hline                                                                 
-Gaussian Random Hill Climbing            & 3.10    & 13.59  & 13.40  &  10.03 \\
-\hline                                                                 
-Policy Gradient Descent                  & 0.68    & 14.69  & 3.60   &  6.32 \\
-\hline                                                                 
-Nelder-Mead simplex                      & 8.51    & 13.62  & 14.83  &  12.32 \\
-\hline                                                                 
-Linear Regression                        & 27.58   & 12.51  & 1.95   &  14.01 \\
-%\hline                                                                
-%SVM Regression                           & ???     & ???    & ???   &   \\
-%\hline                                                                
-%Evolutionary Neural Network (HyperNEAT)  & ???     & ???    & ???   &   \\
-\hline
-\end{tabular}
-\caption{The best gaits found for each starting vector and algorithm,
-  in body lengths per minute.}
-\tablabel{results_table}
-\end{center}
-\end{table}
-
-\begin{itemize}
-
-\item For vector A, shown in \figref{vectorA}, linear regression
-  worked significantly better than the other algorithms, resulting in
-  a gait (27.58 body lengths/minute) that walked over twice as fast as
-  the next best gait -- uniform random hill climbing at 11.37 body
-  lengths/minute -- and 5.3 times better than the previous hand-coded
-  gait.
-
-\item Vector B, depicted in \figref{vectorB}, resulted in similarly
-  performing gaits with all the algorithms. The random method got
-  lucky and produced the best gait (17.26 body lengths/minute) and
-  uniform random hill climbing produced the worst (9.44 body
-  lengths/minute). The remaining algorithms each produced a gait
-  around 13 body lengths/minute.
-
-\item For vector C, shown in \figref{vectorC}, simplex and Gaussian
-  random hill climbing each produced a gait that substantially
-  outperformed the other algorithms. Simplex resulted in a gait of
-  nearly 15 body lengths/minute and Gaussian random hill climbing
-  produced a gait of just over 13 body lengths/minute, whereas the
-  other algorithms returned gaits of less than 5 body lengths/minute.
-
-\end{itemize}
-
-\figvarp{vectorA}{1}{Results for runs beginning with vector A.
-  Linear regression worked significantly better than the other
-  algorithms, resulting in a gait (27.58 body lengths/minute) that
-  walked over twice as fast as the next best gait -- uniform random
-  hill climbing at 11.37 body lengths/minute -- and 5.3 times better
-  than the previous hand-coded gait.}{}
-
-\figvarp{vectorB}{1}{Results for run starting with vector B.  All
-  algorithms performed similarly.  The random method actually got
-  lucky this time and produced the best gait (17.26 body
-  lengths/minute) and uniform random hill climbing produced the worst
-  (9.44 body lengths/minute). The remaining algorithms each produced a
-  gait around 13 body lengths/minute.}{}
-
-\figvarp{vectorC}{1}{Results for runs starting with vector C.
-  Simplex and Gaussian random hill climbing each produced a gait that
-  substantially outperformed the other algorithms, with around 15 body
-  lengths/minute and 13 body lengths/minute, respectively, whereas the
-  other algorithms returned gaits of less than 5 body
-  lengths/minute.}{}
-
-Based on the three trials, as shown in \figref{std_error}, linear
-regression worked the best, followed by simplex. Gaussian random hill
-climbing performed about as well as random, but it is unclear whether
-Gaussian would continue to improve and become much better than random
-search if allowed to run for more iterations. Uniform random hill
-climbing and policy gradient descent performed similarly to each other
-on average, but differed greatly on individual runs.
-
-\figvarp{std_error}{1}{Average results for each method starting with
-  parameter vectors A, B, and C.  Error bars are plotted showing the
-  standard error.  Linear regression outperformed other methods,
-  followed by simplex. Gaussian random hill climbing performed about
-  as well as random, but it is unclear whether Gaussian would continue
-  to improve and become much better than random search if allowed to
-  run for more iterations. Uniform random hill climbing and policy
-  gradient descent performed similarly to each other on average, but
-  differed greatly on individual runs.}{}
-
-In addition to running strategies to optimize parameterized gaits, we
-also wanted to investigate the space of possible gaits.  To accomplish
-this, we selected a parameter vector that resulted in motion, but not
-an exceptional gait, and plotted performance along each dimension
-individually.  In addition, we duplicated each measurement to be able
-to estimate the measurement noise at each point.  Results for this
-exploration are shown in \figref{explore_dim_1} through
-\figref{explore_dim_5}.
-
-\figvarp{explore_dim_1}{.5}{Motion for varying just dimension 1}{}
-
-\figvarp{explore_dim_2}{.5}{Motion for varying just dimension 2}{}
-
-\figvarp{explore_dim_3}{.5}{Motion for varying just dimension 3}{}
-
-\figvarp{explore_dim_4}{.5}{Motion for varying just dimension 4}{}
-
-\figvarp{explore_dim_5}{.5}{Motion for varying just dimension 5}{}
-
-\subsection{Discussion}
-% Is your hypothesis supported? What conclusions do the results support
-% about the strengths and weaknesses of your method compared to other 
-% methods? How can the results be explained in terms of the underlying 
-% properties of the algorithm and/or the data.
-
-Because only three trials were tested with each algorithm and no
-algorithm consistently outperformed the others, there is a large
-standard error for each algorithm, as show in \figref{std_error}. For
-this reason, it is unclear if any algorithm is superior to another in
-this application. More trials with each algorithm would be necessary
-to reach a definitive ranking of the performance of the algorithms.
-
-However, each algorithm discovered at least one gait of over 10 body
-lengths/minute, including random. For this reason, we speculate that
-the motion model is at least as critical as the learning algorithm
-used.
-
-\section{Future work}
-% What are the major shortcomings of your current method? For each 
-% shortcoming, propose additions or enhancements that would help overcome it.
-
-There are several directions in which we could continue our
-work. First, the error margins for our runs were large, so reducing
-them by running more trials would lead to more conclusive data. It
-would also be insightful to run our algorithms on different motion
-models.  We suspect that our choice of motion model influenced the
-results greatly, as even random choices in the space produced gaits
-that moved a significant fraction of the time (> 5\%).  It would be
-interesting to see how learning methods would perform using a model
-that included a much higher percentage of unproductive gaits. We also
-intend to experiment further with SVM regression and evolutionary
-algorithms/HyperNEAT.  Some parameter vectors resulted in the robot
-turning, as opposed to it walking long distances. This could be an
-interesting learning goal in future projects. To these ends, we
-propose the following additions and enhancements:
-
-\begin{itemize}
-\item More runs and/or longer runs
-\item Different motion representations
-\item Better tuning of SVM regression
-\item Evolutionary algorithms/HyperNEAT\cite{clune}
-\item Learning how to turn
-\end{itemize}
-
-\section{Conclusion}
 % Briefly summarize the important results and conclusions presented in 
@@
 open problem.
-
-\section{Acknowledgments}
-% List any people not on the team who helped you with your project: Your
-% TA, other people you consulted with or had useful discussions (say in 
-% a word or two what they did), people who proofread your report, and 
-% any external code you used (libraries etc).
-\begin{itemize}
-\item Hod Lipson, Cornell Computational Synthesis Lab: adviser.
-\item Jim T\o rreson, University of Oslo: adviser.
-\item Juan Zagal, University of Chile: designed and printed robot and provided code for hand tuned gait.
-\item Jeff Clune, CCSL Red Couch: collaborated on HyperNEAT implementation/testing.
-\item Cooper Bills, Cornell University: assisted with Wii tracker development.
-\item Anshumali Srivastava, Cornell University: Teaching Assistant.
-\end{itemize}
-
-\section{The Team}
-% A table listing the names of the people who worked on the project, and who
-% did what.
-
-\begin{itemize}
-
-\item Diana Hidalgo: 
-\begin{itemize}
-\item Simplex (Nelder-Mead)
-\item \code{WiiTrackClient} and \code{WiiTrackServer} classes and hardware
-\item Plotting and evaluating results
-\item Investigation into using Optitrack IR system for tracking robot
-\end{itemize}
-
-\item Sarah Nguyen
-\begin{itemize}
-\item Random search
-\item Uniform random hill climbing
-\item Gaussian random hill climbing
-\item N-dimensional policy gradient descent
-\item Linear regression and prediction
-\item \code{RunManager} and \code{Strategy} classes
-\item Video editing
-\end{itemize}
-
-\item Jason Yosinski
-\begin{itemize}
-\item SVM regression and prediction
-\item Evolutionary Neural Network (HyperNEAT)
-\item \code{WiiTrackClient} and \code{WiiTrackServer} classes and hardware
-\item \code{Robot}, \code{RunManager}, and \code{Strategy} classes
-\item \code{Parameterized Motion Model} abstract base class
-\item \code{SineModel} classes
-\item \code{explore\_dimensions}
-\item Plotting and evaluating results
-\end{itemize}
-
-\end{itemize}
-
-\section{Appendix}
-% Describe the code files you uploaded into CMS
-
-A brief description of the code uploaded to the CMS follows:
-
-\begin{itemize}
-\item \code{optimize.py}: Main program, determines a strategy to try and runs the robot
-  with that strategy.
-\item \code{RunManager.py}: Deals with all the details of running the robot,
-  including choosing an initial parameter, running the robot multiple times, 
-  tracking distance walked, and writing to the log file. Also includes the
-  \code{explore\_dimensions} method.
-\item \code{Strategy.py}: Contains all the different possible strategies, which
-  will be passed as objects in \code{optimize.py}. 
-\item \code{Robot.py}: Implements the \code{Robot} class, described in
-  \secref{implement}.
-\item \code{SineModel.py}: Implements a sine based motion model,
-  described in \secref{implement}.
-\item \code{Motion.py}: Motion helper functions.
-\item \code{WiiTrackServer.py}: Broadcasts the position of the infrared LED.
-\item \code{WiiTrackClient.py}: Connects to the WiiTrackServer to get the current position information.
-\end{itemize}
-
-% LocalWords:  vertices

### 2011-01-20 paper/futureWork.tex
Restructuring of report into multiple files to minimize conflicts, references working, dealt old sections to appropriate locations in paper framework.

--- 2010-12-07 report/report.tex
@@
-\section{Introduction}
+%\section{Future work}
 
-% Motivate and abstractly describe the problem you are addressing and
-% how you are addressing it. What is the problem? Why is it important?
-% What is your basic approach? A short discussion of how it fits into
-% % related work in the area is also desirable. Summarize the basic
-% results and conclusions that you will present.
+\edit{write this, old section below}
 
-Applications of walking robots often call for the ability to walk as
-quickly, efficiently, or with as little power as possible.  Often
-these optimizations are done manually by an expert who designs and
-tweaks a gait specifically for a given objective.  Other groups have
-used learning methods to generate gaits optimized for some metric.
-Approaches differ in their starting assumptions, some essentially
-tweaking the parameters of a hand-tuned model \cite{chernova}, others
-exploring a reasonably compact parameter space \cite{kohl}, and still
-others beginning with few assumptions besides periodicity
-\cite{zykov}.
-
-We aimed to strike a middle ground between these approaches.  Our
-motion generator did not rely on hand-tweaked gaits, but it did use
-parameterized gaits which, by their nature, impose some assumptions on
-the answers produced.  We then used machine learning to design gaits
-for a quadruped robot with these models.  This paper presents a
-comparison of the different learning methods implemented.  Most
-methods created walks that are several times faster than the original
-hand-tuned gait.  We invite readers with short attention spans to view
-a video of some of our results online here:
-
-\url{http://www.youtube.com/watch?v=ODoiOj9DdGg}
-
-%The code so far consists of the robot class, optimization class,
-%parameterized model, sine model, and camera feedback.
-
-\section{Problem definition}
-% Precisely define the problem you are addressing (i.e. formally specify
-% the inputs and outputs). Elaborate on why this is an interesting and
-% important problem.
-
-We are testing several different learning methods to design a
-parametrized gait for a quadruped robot from the Cornell Computational
-Synthesis Lab.
-
-The output each of the learning algorithms is a function of time,
-$f(t)$, that outputs a vector of commanded motor positions.  This
-function is generated using a parametrized motion model, described in
-\secref{motion}.
-
-The robot executes these commands and measures its change in location
-using the tracking system described in \secref{implement}.  The input
-to the learning algorithms is this measured displacement, which the
-algorithms attempt to maximize. This displacement is measured for each
-gait over a constant length run, usually 12 seconds.
-
-A comparison and evaluation of the many different methods available
-for optimizing the gait of legged robots will be useful for future
-work on this challenging multidimensional control problem.
-
-\section{Method}
-\seclabel{method}
-
-% Describe in reasonable detail the algorithm you are using to address
-% this problem. A pseudo-code description of the algorithm you are
-% using is frequently useful. If it makes sense for your project,
-% trace through a concrete example, showing how your algorithm
-% processes this example. The example should be complex enough to
-% illustrate all of the important aspects of the problem but simple
-% enough to be easily understood. If possible, an intuitively
-% meaningful example is better than one with meaningless symbols.
-
-We use several parameterized motion models that command motors to
-positions based on a sine wave, creating a periodic pattern.  While we
-investigated several models, for the bulk of our experiments, we used
-a model whose five parameters are: amplitude, wavelength, scale inner
-vs outer motors, scale left vs right motors, scale back vs front
-motors. Each strategy below attempts to choose the best possible
-parameters for this motion model.
-
-We implemented and tested 8 different learning strategies.  All
-strategies except for the HyperNEAT method\cite{clune} were
-constrained to pick parameters from within predetermined ranges.
-
-\begin{itemize}
-
-\item \emph{Random}: This method randomly generates parameter vectors
-  in the allowable range. This strategy is used only as baseline.
-
-\item \emph{Uniform random hill climbing}: This method begins by
-  selecting a single random parameter vector.  Subsequent iterations
-  generate a neighbor by randomly choosing one parameter to adjust and
-  replacing it with a new value chosen with uniform probability in the
-  allowable range for that parameter. The neighbor is evaluated by
-  running the robot with the newly chosen parameters. If this neighbor
-  results in a longer distance walked than the previous best gait, it
-  is saved as the new best gait. The process is then repeated, always
-  starting with the best gait.
-
-\item \emph{Gaussian random hill climbing}: This method works
-  similarly to Uniform random hill climbing, except neighbors are
-  generated by adding random Gaussian noise to the current best gait.
-  This results in all parameters being changed at once, but the
-  resulting vector is always fairly close to the previous best gait.
-  We used independently selected noise in each dimension, scaled such
-  that the standard deviation of the noise was 5\% of the range of
-  that dimension.
-
-\item \emph{N-dimensional policy gradient descent}: As opposed to the
-  previous methods, this method explicitly estimates the gradient for
-  the objective function. It does this by first evaluating \emph{t}
-  randomly generated parameter vectors near the initial vector, each
-  dimension of these vectors being perturbed by either $-\epsilon$,
-  $0$, or $\epsilon$. Then, for each dimension, it groups vectors into
-  three groups: $-\epsilon$, $0$, and $\epsilon$.  The gradient along
-  this dimension is then estimated as the average score for the
-  $\epsilon$ group minus the average score for the $-\epsilon$
-  group. Finally, the method creates a new vector by changing all
-  parameters by a fixed-size step in the direction of the gradient.
-
-\item \emph{Nelder-Mead simplex method}\cite{nm}: The Nelder-Mead
-  simplex method creates an initial simplex with 6 vertices. The
-  initial parameter vector is stored as the first vertex and the other
-  five vertices are created by adding to one dimension at a time one
-  tenth of the allowable range for that parameter. It then tests the
-  fitness of each vertex and based on these fitnesses, it reflects the
-  worst point over the centroid in an attempt to improve it.  However,
-  to prevent cycles and becoming stuck in local minima, several other
-  rules are used.  In general, the worst vertex is reflected over the
-  centroid. If the reflected point is better than the second worst
-  point and worse than the best point, then the reflected point
-  replaces the worst. If the reflected point is better than the best
-  point, the simplex is expanded in the direction of the reflected
-  point. The better of the reflected and the expanded point replaces
-  the worst point. If the reflected point is worse than the second
-  worst point, then the simplex is contracted away from the reflected
-  point. If the contracted point is better than the reflected point,
-  the contracted point replaces the worst point. If the contracted
-  point is worse than the reflected point, the entire simplex is
-  shrunk \cite{NMWebsite}.
-
-\item \emph{Linear regression}: To initialize, this method chooses and
-  evaluates five random parameter vectors. It then fits a linear model
-  from parameter vector to fitness. In a loop, the method chooses and
-  evaluates a new parameter vector generated by taking a fixed-size
-  step in the direction of the gradient for each parameter, and fits a
-  new linear model to all vectors evaluated so far, choosing the model
-  to minimize the sum of squared errors.
-
-\item \emph{SVM regression}: Similarly to linear regression, this
-  model starts with several random vectors, but this time they are
-  chosen in a small neighborhood about some initial random vector.
-  These vectors (generally 8) are evaluated, and a support vector
-  regression model is fit to the observed fitnesses.  To choose the
-  next vector for evaluation, we randomly generate some number
-  (typically 100) of vectors in the neighborhood of the best observed
-  gait, and select for evaluation the vector with the best predicted
-  performance.  We suspected that if we always chose the best
-  predicted point out of 100, we may end up progressing along a narrow
-  subspace, prohibiting learning of the true local fitness function.
-  Alternately, we would always choosing exploitation of knowledge
-  vs. exploration of the space.  To address this concern, we added a
-  parameter dubbed \code{bumpBy} that added noise to the final
-  selected point before it was evaluated.
-
-  Such a method naturally has many tunable parameters, and we
-  endeavored to select these parameteres by tuning the method in
-  simulation.  To estimate the performance of the algorithm, we ran it
-  agains a simulation with a known optimum.  The simulated function
-  was in the same five dimensional parameter space, and simply
-  returned a fitness determined as the height of a Gaussian with a
-  random mean.  The width of the Gaussian in each dimension was 20\%
-  of the range of each dimension, and the maximum value at the peak
-  was 100.  \figref{svm_sim_results} shows the learning results on
-  this simulated model using the ultimately selected SVM parameters.
-  Interestingly, a non-zero value of \code{bumpBy} resulted in better
-  learning than a zero valued, noise free (exploration free) learning.
-
-  Ultimately, however, the version of SVM tuned for simulation did not
-  yet show competitive performance on the real robot.  We tried tuning
-  some parameters on the real robot, but after some amount of tuning,
-  the method still exibited too little exploration and easily became
-  stuck in local minima.
-
-\item \emph{Evolutionary Neural Network (HyperNEAT)\cite{clune}}: Near
-  the end of term, we hacked together an interface between HyperNEAT
-  and the robot, requiring a slightly modified strategy interface.
-  Preliminary HyperNEAT runs were promising and resulted in serveral
-  interesting gaits.  Unfortunately, the gaits generated by HyperNEAT
-  also tended to stress the robot more than typical gaits had before,
-  and the servos would often overheat and malfunction, requiring
-  restart.  We think these issues may be addressed by adding a small
-  layer between the HyperNEAT strategy and the robot that disallows
-  quickly shifting commanded positions, and we hope to be able to test
-  these methods further once this filter is in place.
-
-\end{itemize}
-
-\figvarp{svm_sim_results}{.6}{Results for the SVM regression strategy
-  in simulation.  This simulation was used to tune the SVM strategy's
-  parameters before trying it on the physical robot.  The strategy
-  quickly approaches the maximum simulated fitness of 100.}{}
-
-\section{Related work}
-% Briefly explain who else worked on related problems in the past and what
-% methods they used. Explain if you are using similar methods, or if your
-% approach is different and if so - how (either is ok).
-
-Various maching learning techniques have proven to be useful in
-finding control policies for a wide variety of robots. Kohl and
-Stone\cite{kohl} presented a policy gradient reinforcement learning
-approach for generating a fast walk on legged robots. We experimented
-with this method to create a walk for our robot (Policy Gradient
-Descent). Chernova and Velosa\cite{chernova} took an evolutionary
-approach to this problem which we did not implement.  Zykov, Bongard,
-and Lipson\cite{zykov} describe the evolution of dynamic gaits on a
-physical robot requiring no prior assumptions about the locomotion
-pattern beyond the fact that it should be rhythmic.
-
-\section{System Architecture and Implementation}
-\seclabel{implement}
-% Describe how you implemented your system and how you structured it. 
-% This should give an overview of the system, not a detailed 
-% documentation of the code. The documentation of the code is part of 
-% the code you hand in. You might want to comment on high-level design 
-% decisions that you made. Also explain how you obtained your
-% data and any pre-processing you did to it.
-
-The quadruped robot has an on-board computer running Linux. The lower
-level drivers are in C and the system is implemented in
-Python. Feedback about distance travelled is provided via an infrared
-LED mounted on the robot and a Wii remote fixed to the ceiling.
-
-An overview of the code follows.
-
-\begin{itemize}
-
-\item \code{Robot} class: Class wrapper for commanding
-  motion of the robot.  The \code{Robot} class takes care of the robot
-  initialization, communication with the servos, and timing of the
-  runs.  In addition, it prevents the servos from ever being commanded
-  to a point outside their normal range (0 - 1023) as well as beyond
-  points where limbs would collide with parts of the robot body.  The
-  main class function, \code{run}, accepts a motion model (any
-  function that takes a time argument and outputs a 9 dimensional
-  position) and will run the robot using this motion model, including,
-  if desired, smooth interpolation over time for the beginning and end
-  of the run.
-
-\item \code{RunManager} class: Deals with all the details of running
-  the robot, including running the robot multiple times, tracking
-  distance walked via a \code{WiiTrackClient} member object, and
-  writing results to the log file. Also includes the
-  \code{explore\_dimensions} method to generate plots by varying each
-  parameter independently.
-
-\item \code{Strategy} class: The user has a choice between eight
-  different learning strategies: Random search, uniform random hill
-  climbing, Gaussian random hill climbing, N-dimensional policy
-  gradient descent, Nelder-Mead simplex, linear regression/prediction,
-  SVM regression/prediction, and HyperNEAT evolution.  Each strategy
-  must derive from the base \code{Strategy} class and must implement
-  two methods: \code{getNext}, for getting the next parameter vector
-  to try, and \code{updateResults}, for communicating results of a run
-  back to the strategy.
-
-\item \code{MotionModel} class: We implemented several motion models,
-  the main being the \code{SineModel5} class.  All motion models take
-  as input a parameter vector and produce as output a motion model,
-  which is simply a function mapping from time to nine motor
-  positions.  The \code{SineModel5} model commands positions based on
-  a sine wave, creating a periodic pattern. The parameters are:
-  amplitude, wavelength, a multiplier for the inner vs outer motors,
-  multiplier for left vs right motors, and multiplier for back vs
-  front motors.  Other similar models were tested, including a seven
-  parameter model which allowed sine waves shifted in time, but these
-  were not used as extensively in our experiments as
-  \code{SineModel5}.
-
-\item \code{WiiTrackClient} and \code{WiiTrackServer} classes and
-  hardware: A Wii remote tracks the location of the robot through an
-  infrared LED mounted on top of the robot. A \code{WiiTrackServer} is
-  run on the robot and continuously tracks its position using the
-  CWiid library\cite{cwiid} to interface with the remote via
-  bluetooth.  The \code{RunManager} then makes a
-  \code{WiiTrackClient}, which connects via socket to the tracking
-  server and requests position updates periodically.
-  \code{RunManager} currently gets the robot's position at the
-  beginning of each run and then again at the end and uses this to
-  calculate the net change in position.
-
-\end{itemize}
-
-\section{Experimental Evaluation}
-
-\subsection{Methodology}
-% What are the criteria you are using to evaluate your method? What
-% specific hypotheses does your experiment test? Describe the 
-% experimental methodology that you used. What are the dependent and 
-% independent variables? For projects in machine learning, what is 
-% the training/test data that was used, and why is it realistic or
-% interesting? Exactly what performance data did you collect and how 
-% are you presenting and analyzing it? Comparisons to competing methods 
-% that address the same problem or to variations of your own algorithm 
-% are particularly useful.
-
-The metric for evaluation of the designed gait is speed. We stop each 
-run after plateauing results (no improvement for one third of the 
-policies seen so far). The standard length of a run designates that it
-should be stopped after there is no improvement for one half of the policies
-seen so far, but since all runs took place on the actual robot, without use
-of a simulator, certain time limitations were imposed on the learning process.
-
-We controlled our experiments from a computer that was connected via a
-wireless ethernet to the robot. The robot collected data about
-distance walked automatically on its own. If it walked outside of the
-Wii remote's viewable area, it informed the user, so the only human
-intervention required during an experiment was to move the robot back
-inside the viewable area and to resume the run, which did not
-interrupt the learning process or result in the loss of data.
-
-We evaluated the efficacy of a set of parameters by sending these
-parameters to the robot and instructing it to walk for a certain
-length of time. The robot always began from the same position and
-returned to the starting position at the end of the run in order to
-measure true displacement without giving credit for ending in a leaned
-position. More efficient parameters resulted in a faster gait, which
-translated into a longer distance walked and a better score. After completing
-an evaluation, the robot sent the resulting distance walked back to the
-host computer and prepared itself for a new set of parameters to evaluate.
-
-Each algorithm was run on 3 different initial parameter vectors on the
-physical robot. We decided to evaluate all methods starting at the
-same three vector in order to allow for the fair comparison of each
-algorithm.  We evaluate each method based on the amount of improvement
-seen from the initial parameter vectors, and on the fastest speeds
-achieved during runs.
-
-The resulting gaits from our algorithms quickly outperformed the original
-hand-coded walk designed for this robot. The fastest walk, for example, was
-4 times faster.
-
-\subsection{Results}
-% Present the quantitative results of your experiments. Graphical data
-% presentation such as graphs and histograms are frequently better 
-% than tables. Explain what are the basic differences revealed in the 
-% data. If your method is stochastic or you analyze multiple datasets, 
-% calculate mean and standard error for every quantitative
-% performance result you show based on multiple runs.
-
-We have done complete runs of 3 different initial parameter vectors
-for random search, uniform random hill climbing, Gaussian random hill
-climbing, policy gradient descent, Nelder-Mead simplex, and linear
-regression.  We also evaluated the SVM regression and HyperNEAT
-methods, but these methods were more experimental, and thus we do not
-yet have the same volume of data for these runs.  We developed several
-gaits that were about 4 times faster than the original hand-coded
-gait.  Results are shown in \tabref{results_table}.
-
-\begin{table}
-\begin{center}
-\begin{tabular}{|r|c|c|c||c|}
-\hline
-                                         & A       & B      & C      &  Average \\
-\hline                                                               
-\hline                                                               
-Previous hand-coded gait                 & --      & --     & --     &  5.16 \\
-\hline                                                                 
-Random search                            & 6.04    & 17.26  & 4.90   &  9.40 \\
-\hline                                                                 
-Uniform Random Hill Climbing             & 11.37   & 9.44   & 2.69   &  7.83 \\
-\hline                                                                 
-Gaussian Random Hill Climbing            & 3.10    & 13.59  & 13.40  &  10.03 \\
-\hline                                                                 
-Policy Gradient Descent                  & 0.68    & 14.69  & 3.60   &  6.32 \\
-\hline                                                                 
-Nelder-Mead simplex                      & 8.51    & 13.62  & 14.83  &  12.32 \\
-\hline                                                                 
-Linear Regression                        & 27.58   & 12.51  & 1.95   &  14.01 \\
-%\hline                                                                
-%SVM Regression                           & ???     & ???    & ???   &   \\
-%\hline                                                                
-%Evolutionary Neural Network (HyperNEAT)  & ???     & ???    & ???   &   \\
-\hline
-\end{tabular}
-\caption{The best gaits found for each starting vector and algorithm,
-  in body lengths per minute.}
-\tablabel{results_table}
-\end{center}
-\end{table}
-
-\begin{itemize}
-
-\item For vector A, shown in \figref{vectorA}, linear regression
-  worked significantly better than the other algorithms, resulting in
-  a gait (27.58 body lengths/minute) that walked over twice as fast as
-  the next best gait -- uniform random hill climbing at 11.37 body
-  lengths/minute -- and 5.3 times better than the previous hand-coded
-  gait.
-
-\item Vector B, depicted in \figref{vectorB}, resulted in similarly
-  performing gaits with all the algorithms. The random method got
-  lucky and produced the best gait (17.26 body lengths/minute) and
-  uniform random hill climbing produced the worst (9.44 body
-  lengths/minute). The remaining algorithms each produced a gait
-  around 13 body lengths/minute.
-
-\item For vector C, shown in \figref{vectorC}, simplex and Gaussian
-  random hill climbing each produced a gait that substantially
-  outperformed the other algorithms. Simplex resulted in a gait of
-  nearly 15 body lengths/minute and Gaussian random hill climbing
-  produced a gait of just over 13 body lengths/minute, whereas the
-  other algorithms returned gaits of less than 5 body lengths/minute.
-
-\end{itemize}
-
-\figvarp{vectorA}{1}{Results for runs beginning with vector A.
-  Linear regression worked significantly better than the other
-  algorithms, resulting in a gait (27.58 body lengths/minute) that
-  walked over twice as fast as the next best gait -- uniform random
-  hill climbing at 11.37 body lengths/minute -- and 5.3 times better
-  than the previous hand-coded gait.}{}
-
-\figvarp{vectorB}{1}{Results for run starting with vector B.  All
-  algorithms performed similarly.  The random method actually got
-  lucky this time and produced the best gait (17.26 body
-  lengths/minute) and uniform random hill climbing produced the worst
-  (9.44 body lengths/minute). The remaining algorithms each produced a
-  gait around 13 body lengths/minute.}{}
-
-\figvarp{vectorC}{1}{Results for runs starting with vector C.
-  Simplex and Gaussian random hill climbing each produced a gait that
-  substantially outperformed the other algorithms, with around 15 body
-  lengths/minute and 13 body lengths/minute, respectively, whereas the
-  other algorithms returned gaits of less than 5 body
-  lengths/minute.}{}
-
-Based on the three trials, as shown in \figref{std_error}, linear
-regression worked the best, followed by simplex. Gaussian random hill
-climbing performed about as well as random, but it is unclear whether
-Gaussian would continue to improve and become much better than random
-search if allowed to run for more iterations. Uniform random hill
-climbing and policy gradient descent performed similarly to each other
-on average, but differed greatly on individual runs.
-
-\figvarp{std_error}{1}{Average results for each method starting with
-  parameter vectors A, B, and C.  Error bars are plotted showing the
-  standard error.  Linear regression outperformed other methods,
-  followed by simplex. Gaussian random hill climbing performed about
-  as well as random, but it is unclear whether Gaussian would continue
-  to improve and become much better than random search if allowed to
-  run for more iterations. Uniform random hill climbing and policy
-  gradient descent performed similarly to each other on average, but
-  differed greatly on individual runs.}{}
-
-In addition to running strategies to optimize parameterized gaits, we
-also wanted to investigate the space of possible gaits.  To accomplish
-this, we selected a parameter vector that resulted in motion, but not
-an exceptional gait, and plotted performance along each dimension
-individually.  In addition, we duplicated each measurement to be able
-to estimate the measurement noise at each point.  Results for this
-exploration are shown in \figref{explore_dim_1} through
-\figref{explore_dim_5}.
-
-\figvarp{explore_dim_1}{.5}{Motion for varying just dimension 1}{}
-
-\figvarp{explore_dim_2}{.5}{Motion for varying just dimension 2}{}
-
-\figvarp{explore_dim_3}{.5}{Motion for varying just dimension 3}{}
-
-\figvarp{explore_dim_4}{.5}{Motion for varying just dimension 4}{}
-
-\figvarp{explore_dim_5}{.5}{Motion for varying just dimension 5}{}
-
-\subsection{Discussion}
-% Is your hypothesis supported? What conclusions do the results support
-% about the strengths and weaknesses of your method compared to other 
-% methods? How can the results be explained in terms of the underlying 
-% properties of the algorithm and/or the data.
-
-Because only three trials were tested with each algorithm and no
-algorithm consistently outperformed the others, there is a large
-standard error for each algorithm, as show in \figref{std_error}. For
-this reason, it is unclear if any algorithm is superior to another in
-this application. More trials with each algorithm would be necessary
-to reach a definitive ranking of the performance of the algorithms.
-
-However, each algorithm discovered at least one gait of over 10 body
-lengths/minute, including random. For this reason, we speculate that
-the motion model is at least as critical as the learning algorithm
-used.
-
-\section{Future work}
 % What are the major shortcomings of your current method? For each 
@@
 \end{itemize}
-
-\section{Conclusion}
-% Briefly summarize the important results and conclusions presented in 
-% the paper. What are the most important points illustrated by your 
-% work? How will your results improve future research and applications 
-% in the area?
-
-We have presented an array of approaches to optimizing a quadrupedal
-gait for forward speed.  We have implemented and tested different
-learning strategies, including uniform and Gaussian random hill
-climbing, policy gradient reinforcement learning, Nelder-Mead simplex,
-several new predictive methods based on linear and support vector
-regression, and an evolved neural network (HyperNEAT).  We have also
-compared these approaches to random search as a baseline. Many of the
-methods resulted in walks significantly faster than previously
-hand-tuned gaits.
-
-Because only three trials were tested with each algorithm and no
-algorithm consistently outperformed the others, there was a large
-standard error for each method, as shown in \figref{std_error}. Thus
-it was unclear if any algorithm was superior to another in this
-application. More trials with each algorithm would be necessary to
-reach a definitive ranking. Because each algorithm discovered at least one gait of
-over 10 body lengths/minute, including random search, we also
-conjectured that the motion representation for the robot is more
-integral to forward speed than the learning algorithm.  How to learn
-the motion representation, in addition to its parameters, remains an
-open problem.
-
-\section{Acknowledgments}
-% List any people not on the team who helped you with your project: Your
-% TA, other people you consulted with or had useful discussions (say in 
-% a word or two what they did), people who proofread your report, and 
-% any external code you used (libraries etc).
-\begin{itemize}
-\item Hod Lipson, Cornell Computational Synthesis Lab: adviser.
-\item Jim T\o rreson, University of Oslo: adviser.
-\item Juan Zagal, University of Chile: designed and printed robot and provided code for hand tuned gait.
-\item Jeff Clune, CCSL Red Couch: collaborated on HyperNEAT implementation/testing.
-\item Cooper Bills, Cornell University: assisted with Wii tracker development.
-\item Anshumali Srivastava, Cornell University: Teaching Assistant.
-\end{itemize}
-
-\section{The Team}
-% A table listing the names of the people who worked on the project, and who
-% did what.
-
-\begin{itemize}
-
-\item Diana Hidalgo: 
-\begin{itemize}
-\item Simplex (Nelder-Mead)
-\item \code{WiiTrackClient} and \code{WiiTrackServer} classes and hardware
-\item Plotting and evaluating results
-\item Investigation into using Optitrack IR system for tracking robot
-\end{itemize}
-
-\item Sarah Nguyen
-\begin{itemize}
-\item Random search
-\item Uniform random hill climbing
-\item Gaussian random hill climbing
-\item N-dimensional policy gradient descent
-\item Linear regression and prediction
-\item \code{RunManager} and \code{Strategy} classes
-\item Video editing
-\end{itemize}
-
-\item Jason Yosinski
-\begin{itemize}
-\item SVM regression and prediction
-\item Evolutionary Neural Network (HyperNEAT)
-\item \code{WiiTrackClient} and \code{WiiTrackServer} classes and hardware
-\item \code{Robot}, \code{RunManager}, and \code{Strategy} classes
-\item \code{Parameterized Motion Model} abstract base class
-\item \code{SineModel} classes
-\item \code{explore\_dimensions}
-\item Plotting and evaluating results
-\end{itemize}
-
-\end{itemize}
-
-\section{Appendix}
-% Describe the code files you uploaded into CMS
-
-A brief description of the code uploaded to the CMS follows:
-
-\begin{itemize}
-\item \code{optimize.py}: Main program, determines a strategy to try and runs the robot
-  with that strategy.
-\item \code{RunManager.py}: Deals with all the details of running the robot,
-  including choosing an initial parameter, running the robot multiple times, 
-  tracking distance walked, and writing to the log file. Also includes the
-  \code{explore\_dimensions} method.
-\item \code{Strategy.py}: Contains all the different possible strategies, which
-  will be passed as objects in \code{optimize.py}. 
-\item \code{Robot.py}: Implements the \code{Robot} class, described in
-  \secref{implement}.
-\item \code{SineModel.py}: Implements a sine based motion model,
-  described in \secref{implement}.
-\item \code{Motion.py}: Motion helper functions.
-\item \code{WiiTrackServer.py}: Broadcasts the position of the infrared LED.
-\item \code{WiiTrackClient.py}: Connects to the WiiTrackServer to get the current position information.
-\end{itemize}
-
-% LocalWords:  vertices

### 2011-01-20 paper/introduction.tex
Restructuring of report into multiple files to minimize conflicts, references working, dealt old sections to appropriate locations in paper framework.

--- 2010-12-07 report/report.tex
@@
-\section{Introduction}
+\section{Introduction and Background}
+
+\edit{Write this}
+
+\editbox{Outline \\
+ - Why gait learning is cool \\
+ - Gait learning has been done before, in simulation and on hardware (cite, cite, cite) \\
+ - HyperNEAT has been used to evolve gaits before in sim (cite, ...) \\
+ - Here, we do two cool things for the first time: \\
+ - - evolve HN in hardware \\
+ - - compare it to other learning methods on the same hardware.}
+
+\edit{Old Quadratot intro follows}
 
@@
 %parameterized model, sine model, and camera feedback.
-
-\section{Problem definition}
-% Precisely define the problem you are addressing (i.e. formally specify
-% the inputs and outputs). Elaborate on why this is an interesting and
-% important problem.
-
-We are testing several different learning methods to design a
-parametrized gait for a quadruped robot from the Cornell Computational
-Synthesis Lab.
-
-The output each of the learning algorithms is a function of time,
-$f(t)$, that outputs a vector of commanded motor positions.  This
-function is generated using a parametrized motion model, described in
-\secref{motion}.
-
-The robot executes these commands and measures its change in location
-using the tracking system described in \secref{implement}.  The input
-to the learning algorithms is this measured displacement, which the
-algorithms attempt to maximize. This displacement is measured for each
-gait over a constant length run, usually 12 seconds.
-
-A comparison and evaluation of the many different methods available
-for optimizing the gait of legged robots will be useful for future
-work on this challenging multidimensional control problem.
-
-\section{Method}
-\seclabel{method}
-
-% Describe in reasonable detail the algorithm you are using to address
-% this problem. A pseudo-code description of the algorithm you are
-% using is frequently useful. If it makes sense for your project,
-% trace through a concrete example, showing how your algorithm
-% processes this example. The example should be complex enough to
-% illustrate all of the important aspects of the problem but simple
-% enough to be easily understood. If possible, an intuitively
-% meaningful example is better than one with meaningless symbols.
-
-We use several parameterized motion models that command motors to
-positions based on a sine wave, creating a periodic pattern.  While we
-investigated several models, for the bulk of our experiments, we used
-a model whose five parameters are: amplitude, wavelength, scale inner
-vs outer motors, scale left vs right motors, scale back vs front
-motors. Each strategy below attempts to choose the best possible
-parameters for this motion model.
-
-We implemented and tested 8 different learning strategies.  All
-strategies except for the HyperNEAT method\cite{clune} were
-constrained to pick parameters from within predetermined ranges.
-
-\begin{itemize}
-
-\item \emph{Random}: This method randomly generates parameter vectors
-  in the allowable range. This strategy is used only as baseline.
-
-\item \emph{Uniform random hill climbing}: This method begins by
-  selecting a single random parameter vector.  Subsequent iterations
-  generate a neighbor by randomly choosing one parameter to adjust and
-  replacing it with a new value chosen with uniform probability in the
-  allowable range for that parameter. The neighbor is evaluated by
-  running the robot with the newly chosen parameters. If this neighbor
-  results in a longer distance walked than the previous best gait, it
-  is saved as the new best gait. The process is then repeated, always
-  starting with the best gait.
-
-\item \emph{Gaussian random hill climbing}: This method works
-  similarly to Uniform random hill climbing, except neighbors are
-  generated by adding random Gaussian noise to the current best gait.
-  This results in all parameters being changed at once, but the
-  resulting vector is always fairly close to the previous best gait.
-  We used independently selected noise in each dimension, scaled such
-  that the standard deviation of the noise was 5\% of the range of
-  that dimension.
-
-\item \emph{N-dimensional policy gradient descent}: As opposed to the
-  previous methods, this method explicitly estimates the gradient for
-  the objective function. It does this by first evaluating \emph{t}
-  randomly generated parameter vectors near the initial vector, each
-  dimension of these vectors being perturbed by either $-\epsilon$,
-  $0$, or $\epsilon$. Then, for each dimension, it groups vectors into
-  three groups: $-\epsilon$, $0$, and $\epsilon$.  The gradient along
-  this dimension is then estimated as the average score for the
-  $\epsilon$ group minus the average score for the $-\epsilon$
-  group. Finally, the method creates a new vector by changing all
-  parameters by a fixed-size step in the direction of the gradient.
-
-\item \emph{Nelder-Mead simplex method}\cite{nm}: The Nelder-Mead
-  simplex method creates an initial simplex with 6 vertices. The
-  initial parameter vector is stored as the first vertex and the other
-  five vertices are created by adding to one dimension at a time one
-  tenth of the allowable range for that parameter. It then tests the
-  fitness of each vertex and based on these fitnesses, it reflects the
-  worst point over the centroid in an attempt to improve it.  However,
-  to prevent cycles and becoming stuck in local minima, several other
-  rules are used.  In general, the worst vertex is reflected over the
-  centroid. If the reflected point is better than the second worst
-  point and worse than the best point, then the reflected point
-  replaces the worst. If the reflected point is better than the best
-  point, the simplex is expanded in the direction of the reflected
-  point. The better of the reflected and the expanded point replaces
-  the worst point. If the reflected point is worse than the second
-  worst point, then the simplex is contracted away from the reflected
-  point. If the contracted point is better than the reflected point,
-  the contracted point replaces the worst point. If the contracted
-  point is worse than the reflected point, the entire simplex is
-  shrunk \cite{NMWebsite}.
-
-\item \emph{Linear regression}: To initialize, this method chooses and
-  evaluates five random parameter vectors. It then fits a linear model
-  from parameter vector to fitness. In a loop, the method chooses and
-  evaluates a new parameter vector generated by taking a fixed-size
-  step in the direction of the gradient for each parameter, and fits a
-  new linear model to all vectors evaluated so far, choosing the model
-  to minimize the sum of squared errors.
-
-\item \emph{SVM regression}: Similarly to linear regression, this
-  model starts with several random vectors, but this time they are
-  chosen in a small neighborhood about some initial random vector.
-  These vectors (generally 8) are evaluated, and a support vector
-  regression model is fit to the observed fitnesses.  To choose the
-  next vector for evaluation, we randomly generate some number
-  (typically 100) of vectors in the neighborhood of the best observed
-  gait, and select for evaluation the vector with the best predicted
-  performance.  We suspected that if we always chose the best
-  predicted point out of 100, we may end up progressing along a narrow
-  subspace, prohibiting learning of the true local fitness function.
-  Alternately, we would always choosing exploitation of knowledge
-  vs. exploration of the space.  To address this concern, we added a
-  parameter dubbed \code{bumpBy} that added noise to the final
-  selected point before it was evaluated.
-
-  Such a method naturally has many tunable parameters, and we
-  endeavored to select these parameteres by tuning the method in
-  simulation.  To estimate the performance of the algorithm, we ran it
-  agains a simulation with a known optimum.  The simulated function
-  was in the same five dimensional parameter space, and simply
-  returned a fitness determined as the height of a Gaussian with a
-  random mean.  The width of the Gaussian in each dimension was 20\%
-  of the range of each dimension, and the maximum value at the peak
-  was 100.  \figref{svm_sim_results} shows the learning results on
-  this simulated model using the ultimately selected SVM parameters.
-  Interestingly, a non-zero value of \code{bumpBy} resulted in better
-  learning than a zero valued, noise free (exploration free) learning.
-
-  Ultimately, however, the version of SVM tuned for simulation did not
-  yet show competitive performance on the real robot.  We tried tuning
-  some parameters on the real robot, but after some amount of tuning,
-  the method still exibited too little exploration and easily became
-  stuck in local minima.
-
-\item \emph{Evolutionary Neural Network (HyperNEAT)\cite{clune}}: Near
-  the end of term, we hacked together an interface between HyperNEAT
-  and the robot, requiring a slightly modified strategy interface.
-  Preliminary HyperNEAT runs were promising and resulted in serveral
-  interesting gaits.  Unfortunately, the gaits generated by HyperNEAT
-  also tended to stress the robot more than typical gaits had before,
-  and the servos would often overheat and malfunction, requiring
-  restart.  We think these issues may be addressed by adding a small
-  layer between the HyperNEAT strategy and the robot that disallows
-  quickly shifting commanded positions, and we hope to be able to test
-  these methods further once this filter is in place.
-
-\end{itemize}
-
-\figvarp{svm_sim_results}{.6}{Results for the SVM regression strategy
-  in simulation.  This simulation was used to tune the SVM strategy's
-  parameters before trying it on the physical robot.  The strategy
-  quickly approaches the maximum simulated fitness of 100.}{}
-
-\section{Related work}
-% Briefly explain who else worked on related problems in the past and what
-% methods they used. Explain if you are using similar methods, or if your
-% approach is different and if so - how (either is ok).
-
-Various maching learning techniques have proven to be useful in
-finding control policies for a wide variety of robots. Kohl and
-Stone\cite{kohl} presented a policy gradient reinforcement learning
-approach for generating a fast walk on legged robots. We experimented
-with this method to create a walk for our robot (Policy Gradient
-Descent). Chernova and Velosa\cite{chernova} took an evolutionary
-approach to this problem which we did not implement.  Zykov, Bongard,
-and Lipson\cite{zykov} describe the evolution of dynamic gaits on a
-physical robot requiring no prior assumptions about the locomotion
-pattern beyond the fact that it should be rhythmic.
-
-\section{System Architecture and Implementation}
-\seclabel{implement}
-% Describe how you implemented your system and how you structured it. 
-% This should give an overview of the system, not a detailed 
-% documentation of the code. The documentation of the code is part of 
-% the code you hand in. You might want to comment on high-level design 
-% decisions that you made. Also explain how you obtained your
-% data and any pre-processing you did to it.
-
-The quadruped robot has an on-board computer running Linux. The lower
-level drivers are in C and the system is implemented in
-Python. Feedback about distance travelled is provided via an infrared
-LED mounted on the robot and a Wii remote fixed to the ceiling.
-
-An overview of the code follows.
-
-\begin{itemize}
-
-\item \code{Robot} class: Class wrapper for commanding
-  motion of the robot.  The \code{Robot} class takes care of the robot
-  initialization, communication with the servos, and timing of the
-  runs.  In addition, it prevents the servos from ever being commanded
-  to a point outside their normal range (0 - 1023) as well as beyond
-  points where limbs would collide with parts of the robot body.  The
-  main class function, \code{run}, accepts a motion model (any
-  function that takes a time argument and outputs a 9 dimensional
-  position) and will run the robot using this motion model, including,
-  if desired, smooth interpolation over time for the beginning and end
-  of the run.
-
-\item \code{RunManager} class: Deals with all the details of running
-  the robot, including running the robot multiple times, tracking
-  distance walked via a \code{WiiTrackClient} member object, and
-  writing results to the log file. Also includes the
-  \code{explore\_dimensions} method to generate plots by varying each
-  parameter independently.
-
-\item \code{Strategy} class: The user has a choice between eight
-  different learning strategies: Random search, uniform random hill
-  climbing, Gaussian random hill climbing, N-dimensional policy
-  gradient descent, Nelder-Mead simplex, linear regression/prediction,
-  SVM regression/prediction, and HyperNEAT evolution.  Each strategy
-  must derive from the base \code{Strategy} class and must implement
-  two methods: \code{getNext}, for getting the next parameter vector
-  to try, and \code{updateResults}, for communicating results of a run
-  back to the strategy.
-
-\item \code{MotionModel} class: We implemented several motion models,
-  the main being the \code{SineModel5} class.  All motion models take
-  as input a parameter vector and produce as output a motion model,
-  which is simply a function mapping from time to nine motor
-  positions.  The \code{SineModel5} model commands positions based on
-  a sine wave, creating a periodic pattern. The parameters are:
-  amplitude, wavelength, a multiplier for the inner vs outer motors,
-  multiplier for left vs right motors, and multiplier for back vs
-  front motors.  Other similar models were tested, including a seven
-  parameter model which allowed sine waves shifted in time, but these
-  were not used as extensively in our experiments as
-  \code{SineModel5}.
-
-\item \code{WiiTrackClient} and \code{WiiTrackServer} classes and
-  hardware: A Wii remote tracks the location of the robot through an
-  infrared LED mounted on top of the robot. A \code{WiiTrackServer} is
-  run on the robot and continuously tracks its position using the
-  CWiid library\cite{cwiid} to interface with the remote via
-  bluetooth.  The \code{RunManager} then makes a
-  \code{WiiTrackClient}, which connects via socket to the tracking
-  server and requests position updates periodically.
-  \code{RunManager} currently gets the robot's position at the
-  beginning of each run and then again at the end and uses this to
-  calculate the net change in position.
-
-\end{itemize}
-
-\section{Experimental Evaluation}
-
-\subsection{Methodology}
-% What are the criteria you are using to evaluate your method? What
-% specific hypotheses does your experiment test? Describe the 
-% experimental methodology that you used. What are the dependent and 
-% independent variables? For projects in machine learning, what is 
-% the training/test data that was used, and why is it realistic or
-% interesting? Exactly what performance data did you collect and how 
-% are you presenting and analyzing it? Comparisons to competing methods 
-% that address the same problem or to variations of your own algorithm 
-% are particularly useful.
-
-The metric for evaluation of the designed gait is speed. We stop each 
-run after plateauing results (no improvement for one third of the 
-policies seen so far). The standard length of a run designates that it
-should be stopped after there is no improvement for one half of the policies
-seen so far, but since all runs took place on the actual robot, without use
-of a simulator, certain time limitations were imposed on the learning process.
-
-We controlled our experiments from a computer that was connected via a
-wireless ethernet to the robot. The robot collected data about
-distance walked automatically on its own. If it walked outside of the
-Wii remote's viewable area, it informed the user, so the only human
-intervention required during an experiment was to move the robot back
-inside the viewable area and to resume the run, which did not
-interrupt the learning process or result in the loss of data.
-
-We evaluated the efficacy of a set of parameters by sending these
-parameters to the robot and instructing it to walk for a certain
-length of time. The robot always began from the same position and
-returned to the starting position at the end of the run in order to
-measure true displacement without giving credit for ending in a leaned
-position. More efficient parameters resulted in a faster gait, which
-translated into a longer distance walked and a better score. After completing
-an evaluation, the robot sent the resulting distance walked back to the
-host computer and prepared itself for a new set of parameters to evaluate.
-
-Each algorithm was run on 3 different initial parameter vectors on the
-physical robot. We decided to evaluate all methods starting at the
-same three vector in order to allow for the fair comparison of each
-algorithm.  We evaluate each method based on the amount of improvement
-seen from the initial parameter vectors, and on the fastest speeds
-achieved during runs.
-
-The resulting gaits from our algorithms quickly outperformed the original
-hand-coded walk designed for this robot. The fastest walk, for example, was
-4 times faster.
-
-\subsection{Results}
-% Present the quantitative results of your experiments. Graphical data
-% presentation such as graphs and histograms are frequently better 
-% than tables. Explain what are the basic differences revealed in the 
-% data. If your method is stochastic or you analyze multiple datasets, 
-% calculate mean and standard error for every quantitative
-% performance result you show based on multiple runs.
-
-We have done complete runs of 3 different initial parameter vectors
-for random search, uniform random hill climbing, Gaussian random hill
-climbing, policy gradient descent, Nelder-Mead simplex, and linear
-regression.  We also evaluated the SVM regression and HyperNEAT
-methods, but these methods were more experimental, and thus we do not
-yet have the same volume of data for these runs.  We developed several
-gaits that were about 4 times faster than the original hand-coded
-gait.  Results are shown in \tabref{results_table}.
-
-\begin{table}
-\begin{center}
-\begin{tabular}{|r|c|c|c||c|}
-\hline
-                                         & A       & B      & C      &  Average \\
-\hline                                                               
-\hline                                                               
-Previous hand-coded gait                 & --      & --     & --     &  5.16 \\
-\hline                                                                 
-Random search                            & 6.04    & 17.26  & 4.90   &  9.40 \\
-\hline                                                                 
-Uniform Random Hill Climbing             & 11.37   & 9.44   & 2.69   &  7.83 \\
-\hline                                                                 
-Gaussian Random Hill Climbing            & 3.10    & 13.59  & 13.40  &  10.03 \\
-\hline                                                                 
-Policy Gradient Descent                  & 0.68    & 14.69  & 3.60   &  6.32 \\
-\hline                                                                 
-Nelder-Mead simplex                      & 8.51    & 13.62  & 14.83  &  12.32 \\
-\hline                                                                 
-Linear Regression                        & 27.58   & 12.51  & 1.95   &  14.01 \\
-%\hline                                                                
-%SVM Regression                           & ???     & ???    & ???   &   \\
-%\hline                                                                
-%Evolutionary Neural Network (HyperNEAT)  & ???     & ???    & ???   &   \\
-\hline
-\end{tabular}
-\caption{The best gaits found for each starting vector and algorithm,
-  in body lengths per minute.}
-\tablabel{results_table}
-\end{center}
-\end{table}
-
-\begin{itemize}
-
-\item For vector A, shown in \figref{vectorA}, linear regression
-  worked significantly better than the other algorithms, resulting in
-  a gait (27.58 body lengths/minute) that walked over twice as fast as
-  the next best gait -- uniform random hill climbing at 11.37 body
-  lengths/minute -- and 5.3 times better than the previous hand-coded
-  gait.
-
-\item Vector B, depicted in \figref{vectorB}, resulted in similarly
-  performing gaits with all the algorithms. The random method got
-  lucky and produced the best gait (17.26 body lengths/minute) and
-  uniform random hill climbing produced the worst (9.44 body
-  lengths/minute). The remaining algorithms each produced a gait
-  around 13 body lengths/minute.
-
-\item For vector C, shown in \figref{vectorC}, simplex and Gaussian
-  random hill climbing each produced a gait that substantially
-  outperformed the other algorithms. Simplex resulted in a gait of
-  nearly 15 body lengths/minute and Gaussian random hill climbing
-  produced a gait of just over 13 body lengths/minute, whereas the
-  other algorithms returned gaits of less than 5 body lengths/minute.
-
-\end{itemize}
-
-\figvarp{vectorA}{1}{Results for runs beginning with vector A.
-  Linear regression worked significantly better than the other
-  algorithms, resulting in a gait (27.58 body lengths/minute) that
-  walked over twice as fast as the next best gait -- uniform random
-  hill climbing at 11.37 body lengths/minute -- and 5.3 times better
-  than the previous hand-coded gait.}{}
-
-\figvarp{vectorB}{1}{Results for run starting with vector B.  All
-  algorithms performed similarly.  The random method actually got
-  lucky this time and produced the best gait (17.26 body
-  lengths/minute) and uniform random hill climbing produced the worst
-  (9.44 body lengths/minute). The remaining algorithms each produced a
-  gait around 13 body lengths/minute.}{}
-
-\figvarp{vectorC}{1}{Results for runs starting with vector C.
-  Simplex and Gaussian random hill climbing each produced a gait that
-  substantially outperformed the other algorithms, with around 15 body
-  lengths/minute and 13 body lengths/minute, respectively, whereas the
-  other algorithms returned gaits of less than 5 body
-  lengths/minute.}{}
-
-Based on the three trials, as shown in \figref{std_error}, linear
-regression worked the best, followed by simplex. Gaussian random hill
-climbing performed about as well as random, but it is unclear whether
-Gaussian would continue to improve and become much better than random
-search if allowed to run for more iterations. Uniform random hill
-climbing and policy gradient descent performed similarly to each other
-on average, but differed greatly on individual runs.
-
-\figvarp{std_error}{1}{Average results for each method starting with
-  parameter vectors A, B, and C.  Error bars are plotted showing the
-  standard error.  Linear regression outperformed other methods,
-  followed by simplex. Gaussian random hill climbing performed about
-  as well as random, but it is unclear whether Gaussian would continue
-  to improve and become much better than random search if allowed to
-  run for more iterations. Uniform random hill climbing and policy
-  gradient descent performed similarly to each other on average, but
-  differed greatly on individual runs.}{}
-
-In addition to running strategies to optimize parameterized gaits, we
-also wanted to investigate the space of possible gaits.  To accomplish
-this, we selected a parameter vector that resulted in motion, but not
-an exceptional gait, and plotted performance along each dimension
-individually.  In addition, we duplicated each measurement to be able
-to estimate the measurement noise at each point.  Results for this
-exploration are shown in \figref{explore_dim_1} through
-\figref{explore_dim_5}.
-
-\figvarp{explore_dim_1}{.5}{Motion for varying just dimension 1}{}
-
-\figvarp{explore_dim_2}{.5}{Motion for varying just dimension 2}{}
-
-\figvarp{explore_dim_3}{.5}{Motion for varying just dimension 3}{}
-
-\figvarp{explore_dim_4}{.5}{Motion for varying just dimension 4}{}
-
-\figvarp{explore_dim_5}{.5}{Motion for varying just dimension 5}{}
-
-\subsection{Discussion}
-% Is your hypothesis supported? What conclusions do the results support
-% about the strengths and weaknesses of your method compared to other 
-% methods? How can the results be explained in terms of the underlying 
-% properties of the algorithm and/or the data.
-
-Because only three trials were tested with each algorithm and no
-algorithm consistently outperformed the others, there is a large
-standard error for each algorithm, as show in \figref{std_error}. For
-this reason, it is unclear if any algorithm is superior to another in
-this application. More trials with each algorithm would be necessary
-to reach a definitive ranking of the performance of the algorithms.
-
-However, each algorithm discovered at least one gait of over 10 body
-lengths/minute, including random. For this reason, we speculate that
-the motion model is at least as critical as the learning algorithm
-used.
-
-\section{Future work}
-% What are the major shortcomings of your current method? For each 
-% shortcoming, propose additions or enhancements that would help overcome it.
-
-There are several directions in which we could continue our
-work. First, the error margins for our runs were large, so reducing
-them by running more trials would lead to more conclusive data. It
-would also be insightful to run our algorithms on different motion
-models.  We suspect that our choice of motion model influenced the
-results greatly, as even random choices in the space produced gaits
-that moved a significant fraction of the time (> 5\%).  It would be
-interesting to see how learning methods would perform using a model
-that included a much higher percentage of unproductive gaits. We also
-intend to experiment further with SVM regression and evolutionary
-algorithms/HyperNEAT.  Some parameter vectors resulted in the robot
-turning, as opposed to it walking long distances. This could be an
-interesting learning goal in future projects. To these ends, we
-propose the following additions and enhancements:
-
-\begin{itemize}
-\item More runs and/or longer runs
-\item Different motion representations
-\item Better tuning of SVM regression
-\item Evolutionary algorithms/HyperNEAT\cite{clune}
-\item Learning how to turn
-\end{itemize}
-
-\section{Conclusion}
-% Briefly summarize the important results and conclusions presented in 
-% the paper. What are the most important points illustrated by your 
-% work? How will your results improve future research and applications 
-% in the area?
-
-We have presented an array of approaches to optimizing a quadrupedal
-gait for forward speed.  We have implemented and tested different
-learning strategies, including uniform and Gaussian random hill
-climbing, policy gradient reinforcement learning, Nelder-Mead simplex,
-several new predictive methods based on linear and support vector
-regression, and an evolved neural network (HyperNEAT).  We have also
-compared these approaches to random search as a baseline. Many of the
-methods resulted in walks significantly faster than previously
-hand-tuned gaits.
-
-Because only three trials were tested with each algorithm and no
-algorithm consistently outperformed the others, there was a large
-standard error for each method, as shown in \figref{std_error}. Thus
-it was unclear if any algorithm was superior to another in this
-application. More trials with each algorithm would be necessary to
-reach a definitive ranking. Because each algorithm discovered at least one gait of
-over 10 body lengths/minute, including random search, we also
-conjectured that the motion representation for the robot is more
-integral to forward speed than the learning algorithm.  How to learn
-the motion representation, in addition to its parameters, remains an
-open problem.
-
-\section{Acknowledgments}
-% List any people not on the team who helped you with your project: Your
-% TA, other people you consulted with or had useful discussions (say in 
-% a word or two what they did), people who proofread your report, and 
-% any external code you used (libraries etc).
-\begin{itemize}
-\item Hod Lipson, Cornell Computational Synthesis Lab: adviser.
-\item Jim T\o rreson, University of Oslo: adviser.
-\item Juan Zagal, University of Chile: designed and printed robot and provided code for hand tuned gait.
-\item Jeff Clune, CCSL Red Couch: collaborated on HyperNEAT implementation/testing.
-\item Cooper Bills, Cornell University: assisted with Wii tracker development.
-\item Anshumali Srivastava, Cornell University: Teaching Assistant.
-\end{itemize}
-
-\section{The Team}
-% A table listing the names of the people who worked on the project, and who
-% did what.
-
-\begin{itemize}
-
-\item Diana Hidalgo: 
-\begin{itemize}
-\item Simplex (Nelder-Mead)
-\item \code{WiiTrackClient} and \code{WiiTrackServer} classes and hardware
-\item Plotting and evaluating results
-\item Investigation into using Optitrack IR system for tracking robot
-\end{itemize}
-
-\item Sarah Nguyen
-\begin{itemize}
-\item Random search
-\item Uniform random hill climbing
-\item Gaussian random hill climbing
-\item N-dimensional policy gradient descent
-\item Linear regression and prediction
-\item \code{RunManager} and \code{Strategy} classes
-\item Video editing
-\end{itemize}
-
-\item Jason Yosinski
-\begin{itemize}
-\item SVM regression and prediction
-\item Evolutionary Neural Network (HyperNEAT)
-\item \code{WiiTrackClient} and \code{WiiTrackServer} classes and hardware
-\item \code{Robot}, \code{RunManager}, and \code{Strategy} classes
-\item \code{Parameterized Motion Model} abstract base class
-\item \code{SineModel} classes
-\item \code{explore\_dimensions}
-\item Plotting and evaluating results
-\end{itemize}
-
-\end{itemize}
-
-\section{Appendix}
-% Describe the code files you uploaded into CMS
-
-A brief description of the code uploaded to the CMS follows:
-
-\begin{itemize}
-\item \code{optimize.py}: Main program, determines a strategy to try and runs the robot
-  with that strategy.
-\item \code{RunManager.py}: Deals with all the details of running the robot,
-  including choosing an initial parameter, running the robot multiple times, 
-  tracking distance walked, and writing to the log file. Also includes the
-  \code{explore\_dimensions} method.
-\item \code{Strategy.py}: Contains all the different possible strategies, which
-  will be passed as objects in \code{optimize.py}. 
-\item \code{Robot.py}: Implements the \code{Robot} class, described in
-  \secref{implement}.
-\item \code{SineModel.py}: Implements a sine based motion model,
-  described in \secref{implement}.
-\item \code{Motion.py}: Motion helper functions.
-\item \code{WiiTrackServer.py}: Broadcasts the position of the infrared LED.
-\item \code{WiiTrackClient.py}: Connects to the WiiTrackServer to get the current position information.
-\end{itemize}
-
-% LocalWords:  vertices

### 2011-01-20 paper/paperOutline.tex
Restructuring of report into multiple files to minimize conflicts, references working, dealt old sections to appropriate locations in paper framework.

\subsection{Outline of Sections}

\edit{write this}

The remainder of the paper is organized as follows. In \secref{foo}, we
introduce,  \secref{bar} 3 describes.  Finally, we describe future work
in \secref{future}

### 2011-01-20 paper/problemDefinition.tex
Restructuring of report into multiple files to minimize conflicts, references working, dealt old sections to appropriate locations in paper framework.

--- 2010-12-07 report/report.tex
@@
-\section{Introduction}
-
-% Motivate and abstractly describe the problem you are addressing and
-% how you are addressing it. What is the problem? Why is it important?
-% What is your basic approach? A short discussion of how it fits into
-% % related work in the area is also desirable. Summarize the basic
-% results and conclusions that you will present.
-
-Applications of walking robots often call for the ability to walk as
-quickly, efficiently, or with as little power as possible.  Often
-these optimizations are done manually by an expert who designs and
-tweaks a gait specifically for a given objective.  Other groups have
-used learning methods to generate gaits optimized for some metric.
-Approaches differ in their starting assumptions, some essentially
-tweaking the parameters of a hand-tuned model \cite{chernova}, others
-exploring a reasonably compact parameter space \cite{kohl}, and still
-others beginning with few assumptions besides periodicity
-\cite{zykov}.
-
-We aimed to strike a middle ground between these approaches.  Our
-motion generator did not rely on hand-tweaked gaits, but it did use
-parameterized gaits which, by their nature, impose some assumptions on
-the answers produced.  We then used machine learning to design gaits
-for a quadruped robot with these models.  This paper presents a
-comparison of the different learning methods implemented.  Most
-methods created walks that are several times faster than the original
-hand-tuned gait.  We invite readers with short attention spans to view
-a video of some of our results online here:
-
-\url{http://www.youtube.com/watch?v=ODoiOj9DdGg}
-
-%The code so far consists of the robot class, optimization class,
-%parameterized model, sine model, and camera feedback.
-
 \section{Problem definition}
@@
 % important problem.
+
+\edit{write this.  Quadratot report section below}
 
@@
 work on this challenging multidimensional control problem.
-
-\section{Method}
-\seclabel{method}
-
-% Describe in reasonable detail the algorithm you are using to address
-% this problem. A pseudo-code description of the algorithm you are
-% using is frequently useful. If it makes sense for your project,
-% trace through a concrete example, showing how your algorithm
-% processes this example. The example should be complex enough to
-% illustrate all of the important aspects of the problem but simple
-% enough to be easily understood. If possible, an intuitively
-% meaningful example is better than one with meaningless symbols.
-
-We use several parameterized motion models that command motors to
-positions based on a sine wave, creating a periodic pattern.  While we
-investigated several models, for the bulk of our experiments, we used
-a model whose five parameters are: amplitude, wavelength, scale inner
-vs outer motors, scale left vs right motors, scale back vs front
-motors. Each strategy below attempts to choose the best possible
-parameters for this motion model.
-
-We implemented and tested 8 different learning strategies.  All
-strategies except for the HyperNEAT method\cite{clune} were
-constrained to pick parameters from within predetermined ranges.
-
-\begin{itemize}
-
-\item \emph{Random}: This method randomly generates parameter vectors
-  in the allowable range. This strategy is used only as baseline.
-
-\item \emph{Uniform random hill climbing}: This method begins by
-  selecting a single random parameter vector.  Subsequent iterations
-  generate a neighbor by randomly choosing one parameter to adjust and
-  replacing it with a new value chosen with uniform probability in the
-  allowable range for that parameter. The neighbor is evaluated by
-  running the robot with the newly chosen parameters. If this neighbor
-  results in a longer distance walked than the previous best gait, it
-  is saved as the new best gait. The process is then repeated, always
-  starting with the best gait.
-
-\item \emph{Gaussian random hill climbing}: This method works
-  similarly to Uniform random hill climbing, except neighbors are
-  generated by adding random Gaussian noise to the current best gait.
-  This results in all parameters being changed at once, but the
-  resulting vector is always fairly close to the previous best gait.
-  We used independently selected noise in each dimension, scaled such
-  that the standard deviation of the noise was 5\% of the range of
-  that dimension.
-
-\item \emph{N-dimensional policy gradient descent}: As opposed to the
-  previous methods, this method explicitly estimates the gradient for
-  the objective function. It does this by first evaluating \emph{t}
-  randomly generated parameter vectors near the initial vector, each
-  dimension of these vectors being perturbed by either $-\epsilon$,
-  $0$, or $\epsilon$. Then, for each dimension, it groups vectors into
-  three groups: $-\epsilon$, $0$, and $\epsilon$.  The gradient along
-  this dimension is then estimated as the average score for the
-  $\epsilon$ group minus the average score for the $-\epsilon$
-  group. Finally, the method creates a new vector by changing all
-  parameters by a fixed-size step in the direction of the gradient.
-
-\item \emph{Nelder-Mead simplex method}\cite{nm}: The Nelder-Mead
-  simplex method creates an initial simplex with 6 vertices. The
-  initial parameter vector is stored as the first vertex and the other
-  five vertices are created by adding to one dimension at a time one
-  tenth of the allowable range for that parameter. It then tests the
-  fitness of each vertex and based on these fitnesses, it reflects the
-  worst point over the centroid in an attempt to improve it.  However,
-  to prevent cycles and becoming stuck in local minima, several other
-  rules are used.  In general, the worst vertex is reflected over the
-  centroid. If the reflected point is better than the second worst
-  point and worse than the best point, then the reflected point
-  replaces the worst. If the reflected point is better than the best
-  point, the simplex is expanded in the direction of the reflected
-  point. The better of the reflected and the expanded point replaces
-  the worst point. If the reflected point is worse than the second
-  worst point, then the simplex is contracted away from the reflected
-  point. If the contracted point is better than the reflected point,
-  the contracted point replaces the worst point. If the contracted
-  point is worse than the reflected point, the entire simplex is
-  shrunk \cite{NMWebsite}.
-
-\item \emph{Linear regression}: To initialize, this method chooses and
-  evaluates five random parameter vectors. It then fits a linear model
-  from parameter vector to fitness. In a loop, the method chooses and
-  evaluates a new parameter vector generated by taking a fixed-size
-  step in the direction of the gradient for each parameter, and fits a
-  new linear model to all vectors evaluated so far, choosing the model
-  to minimize the sum of squared errors.
-
-\item \emph{SVM regression}: Similarly to linear regression, this
-  model starts with several random vectors, but this time they are
-  chosen in a small neighborhood about some initial random vector.
-  These vectors (generally 8) are evaluated, and a support vector
-  regression model is fit to the observed fitnesses.  To choose the
-  next vector for evaluation, we randomly generate some number
-  (typically 100) of vectors in the neighborhood of the best observed
-  gait, and select for evaluation the vector with the best predicted
-  performance.  We suspected that if we always chose the best
-  predicted point out of 100, we may end up progressing along a narrow
-  subspace, prohibiting learning of the true local fitness function.
-  Alternately, we would always choosing exploitation of knowledge
-  vs. exploration of the space.  To address this concern, we added a
-  parameter dubbed \code{bumpBy} that added noise to the final
-  selected point before it was evaluated.
-
-  Such a method naturally has many tunable parameters, and we
-  endeavored to select these parameteres by tuning the method in
-  simulation.  To estimate the performance of the algorithm, we ran it
-  agains a simulation with a known optimum.  The simulated function
-  was in the same five dimensional parameter space, and simply
-  returned a fitness determined as the height of a Gaussian with a
-  random mean.  The width of the Gaussian in each dimension was 20\%
-  of the range of each dimension, and the maximum value at the peak
-  was 100.  \figref{svm_sim_results} shows the learning results on
-  this simulated model using the ultimately selected SVM parameters.
-  Interestingly, a non-zero value of \code{bumpBy} resulted in better
-  learning than a zero valued, noise free (exploration free) learning.
-
-  Ultimately, however, the version of SVM tuned for simulation did not
-  yet show competitive performance on the real robot.  We tried tuning
-  some parameters on the real robot, but after some amount of tuning,
-  the method still exibited too little exploration and easily became
-  stuck in local minima.
-
-\item \emph{Evolutionary Neural Network (HyperNEAT)\cite{clune}}: Near
-  the end of term, we hacked together an interface between HyperNEAT
-  and the robot, requiring a slightly modified strategy interface.
-  Preliminary HyperNEAT runs were promising and resulted in serveral
-  interesting gaits.  Unfortunately, the gaits generated by HyperNEAT
-  also tended to stress the robot more than typical gaits had before,
-  and the servos would often overheat and malfunction, requiring
-  restart.  We think these issues may be addressed by adding a small
-  layer between the HyperNEAT strategy and the robot that disallows
-  quickly shifting commanded positions, and we hope to be able to test
-  these methods further once this filter is in place.
-
-\end{itemize}
-
-\figvarp{svm_sim_results}{.6}{Results for the SVM regression strategy
-  in simulation.  This simulation was used to tune the SVM strategy's
-  parameters before trying it on the physical robot.  The strategy
-  quickly approaches the maximum simulated fitness of 100.}{}
-
-\section{Related work}
-% Briefly explain who else worked on related problems in the past and what
-% methods they used. Explain if you are using similar methods, or if your
-% approach is different and if so - how (either is ok).
-
-Various maching learning techniques have proven to be useful in
-finding control policies for a wide variety of robots. Kohl and
-Stone\cite{kohl} presented a policy gradient reinforcement learning
-approach for generating a fast walk on legged robots. We experimented
-with this method to create a walk for our robot (Policy Gradient
-Descent). Chernova and Velosa\cite{chernova} took an evolutionary
-approach to this problem which we did not implement.  Zykov, Bongard,
-and Lipson\cite{zykov} describe the evolution of dynamic gaits on a
-physical robot requiring no prior assumptions about the locomotion
-pattern beyond the fact that it should be rhythmic.
-
-\section{System Architecture and Implementation}
-\seclabel{implement}
-% Describe how you implemented your system and how you structured it. 
-% This should give an overview of the system, not a detailed 
-% documentation of the code. The documentation of the code is part of 
-% the code you hand in. You might want to comment on high-level design 
-% decisions that you made. Also explain how you obtained your
-% data and any pre-processing you did to it.
-
-The quadruped robot has an on-board computer running Linux. The lower
-level drivers are in C and the system is implemented in
-Python. Feedback about distance travelled is provided via an infrared
-LED mounted on the robot and a Wii remote fixed to the ceiling.
-
-An overview of the code follows.
-
-\begin{itemize}
-
-\item \code{Robot} class: Class wrapper for commanding
-  motion of the robot.  The \code{Robot} class takes care of the robot
-  initialization, communication with the servos, and timing of the
-  runs.  In addition, it prevents the servos from ever being commanded
-  to a point outside their normal range (0 - 1023) as well as beyond
-  points where limbs would collide with parts of the robot body.  The
-  main class function, \code{run}, accepts a motion model (any
-  function that takes a time argument and outputs a 9 dimensional
-  position) and will run the robot using this motion model, including,
-  if desired, smooth interpolation over time for the beginning and end
-  of the run.
-
-\item \code{RunManager} class: Deals with all the details of running
-  the robot, including running the robot multiple times, tracking
-  distance walked via a \code{WiiTrackClient} member object, and
-  writing results to the log file. Also includes the
-  \code{explore\_dimensions} method to generate plots by varying each
-  parameter independently.
-
-\item \code{Strategy} class: The user has a choice between eight
-  different learning strategies: Random search, uniform random hill
-  climbing, Gaussian random hill climbing, N-dimensional policy
-  gradient descent, Nelder-Mead simplex, linear regression/prediction,
-  SVM regression/prediction, and HyperNEAT evolution.  Each strategy
-  must derive from the base \code{Strategy} class and must implement
-  two methods: \code{getNext}, for getting the next parameter vector
-  to try, and \code{updateResults}, for communicating results of a run
-  back to the strategy.
-
-\item \code{MotionModel} class: We implemented several motion models,
-  the main being the \code{SineModel5} class.  All motion models take
-  as input a parameter vector and produce as output a motion model,
-  which is simply a function mapping from time to nine motor
-  positions.  The \code{SineModel5} model commands positions based on
-  a sine wave, creating a periodic pattern. The parameters are:
-  amplitude, wavelength, a multiplier for the inner vs outer motors,
-  multiplier for left vs right motors, and multiplier for back vs
-  front motors.  Other similar models were tested, including a seven
-  parameter model which allowed sine waves shifted in time, but these
-  were not used as extensively in our experiments as
-  \code{SineModel5}.
-
-\item \code{WiiTrackClient} and \code{WiiTrackServer} classes and
-  hardware: A Wii remote tracks the location of the robot through an
-  infrared LED mounted on top of the robot. A \code{WiiTrackServer} is
-  run on the robot and continuously tracks its position using the
-  CWiid library\cite{cwiid} to interface with the remote via
-  bluetooth.  The \code{RunManager} then makes a
-  \code{WiiTrackClient}, which connects via socket to the tracking
-  server and requests position updates periodically.
-  \code{RunManager} currently gets the robot's position at the
-  beginning of each run and then again at the end and uses this to
-  calculate the net change in position.
-
-\end{itemize}
-
-\section{Experimental Evaluation}
-
-\subsection{Methodology}
-% What are the criteria you are using to evaluate your method? What
-% specific hypotheses does your experiment test? Describe the 
-% experimental methodology that you used. What are the dependent and 
-% independent variables? For projects in machine learning, what is 
-% the training/test data that was used, and why is it realistic or
-% interesting? Exactly what performance data did you collect and how 
-% are you presenting and analyzing it? Comparisons to competing methods 
-% that address the same problem or to variations of your own algorithm 
-% are particularly useful.
-
-The metric for evaluation of the designed gait is speed. We stop each 
-run after plateauing results (no improvement for one third of the 
-policies seen so far). The standard length of a run designates that it
-should be stopped after there is no improvement for one half of the policies
-seen so far, but since all runs took place on the actual robot, without use
-of a simulator, certain time limitations were imposed on the learning process.
-
-We controlled our experiments from a computer that was connected via a
-wireless ethernet to the robot. The robot collected data about
-distance walked automatically on its own. If it walked outside of the
-Wii remote's viewable area, it informed the user, so the only human
-intervention required during an experiment was to move the robot back
-inside the viewable area and to resume the run, which did not
-interrupt the learning process or result in the loss of data.
-
-We evaluated the efficacy of a set of parameters by sending these
-parameters to the robot and instructing it to walk for a certain
-length of time. The robot always began from the same position and
-returned to the starting position at the end of the run in order to
-measure true displacement without giving credit for ending in a leaned
-position. More efficient parameters resulted in a faster gait, which
-translated into a longer distance walked and a better score. After completing
-an evaluation, the robot sent the resulting distance walked back to the
-host computer and prepared itself for a new set of parameters to evaluate.
-
-Each algorithm was run on 3 different initial parameter vectors on the
-physical robot. We decided to evaluate all methods starting at the
-same three vector in order to allow for the fair comparison of each
-algorithm.  We evaluate each method based on the amount of improvement
-seen from the initial parameter vectors, and on the fastest speeds
-achieved during runs.
-
-The resulting gaits from our algorithms quickly outperformed the original
-hand-coded walk designed for this robot. The fastest walk, for example, was
-4 times faster.
-
-\subsection{Results}
-% Present the quantitative results of your experiments. Graphical data
-% presentation such as graphs and histograms are frequently better 
-% than tables. Explain what are the basic differences revealed in the 
-% data. If your method is stochastic or you analyze multiple datasets, 
-% calculate mean and standard error for every quantitative
-% performance result you show based on multiple runs.
-
-We have done complete runs of 3 different initial parameter vectors
-for random search, uniform random hill climbing, Gaussian random hill
-climbing, policy gradient descent, Nelder-Mead simplex, and linear
-regression.  We also evaluated the SVM regression and HyperNEAT
-methods, but these methods were more experimental, and thus we do not
-yet have the same volume of data for these runs.  We developed several
-gaits that were about 4 times faster than the original hand-coded
-gait.  Results are shown in \tabref{results_table}.
-
-\begin{table}
-\begin{center}
-\begin{tabular}{|r|c|c|c||c|}
-\hline
-                                         & A       & B      & C      &  Average \\
-\hline                                                               
-\hline                                                               
-Previous hand-coded gait                 & --      & --     & --     &  5.16 \\
-\hline                                                                 
-Random search                            & 6.04    & 17.26  & 4.90   &  9.40 \\
-\hline                                                                 
-Uniform Random Hill Climbing             & 11.37   & 9.44   & 2.69   &  7.83 \\
-\hline                                                                 
-Gaussian Random Hill Climbing            & 3.10    & 13.59  & 13.40  &  10.03 \\
-\hline                                                                 
-Policy Gradient Descent                  & 0.68    & 14.69  & 3.60   &  6.32 \\
-\hline                                                                 
-Nelder-Mead simplex                      & 8.51    & 13.62  & 14.83  &  12.32 \\
-\hline                                                                 
-Linear Regression                        & 27.58   & 12.51  & 1.95   &  14.01 \\
-%\hline                                                                
-%SVM Regression                           & ???     & ???    & ???   &   \\
-%\hline                                                                
-%Evolutionary Neural Network (HyperNEAT)  & ???     & ???    & ???   &   \\
-\hline
-\end{tabular}
-\caption{The best gaits found for each starting vector and algorithm,
-  in body lengths per minute.}
-\tablabel{results_table}
-\end{center}
-\end{table}
-
-\begin{itemize}
-
-\item For vector A, shown in \figref{vectorA}, linear regression
-  worked significantly better than the other algorithms, resulting in
-  a gait (27.58 body lengths/minute) that walked over twice as fast as
-  the next best gait -- uniform random hill climbing at 11.37 body
-  lengths/minute -- and 5.3 times better than the previous hand-coded
-  gait.
-
-\item Vector B, depicted in \figref{vectorB}, resulted in similarly
-  performing gaits with all the algorithms. The random method got
-  lucky and produced the best gait (17.26 body lengths/minute) and
-  uniform random hill climbing produced the worst (9.44 body
-  lengths/minute). The remaining algorithms each produced a gait
-  around 13 body lengths/minute.
-
-\item For vector C, shown in \figref{vectorC}, simplex and Gaussian
-  random hill climbing each produced a gait that substantially
-  outperformed the other algorithms. Simplex resulted in a gait of
-  nearly 15 body lengths/minute and Gaussian random hill climbing
-  produced a gait of just over 13 body lengths/minute, whereas the
-  other algorithms returned gaits of less than 5 body lengths/minute.
-
-\end{itemize}
-
-\figvarp{vectorA}{1}{Results for runs beginning with vector A.
-  Linear regression worked significantly better than the other
-  algorithms, resulting in a gait (27.58 body lengths/minute) that
-  walked over twice as fast as the next best gait -- uniform random
-  hill climbing at 11.37 body lengths/minute -- and 5.3 times better
-  than the previous hand-coded gait.}{}
-
-\figvarp{vectorB}{1}{Results for run starting with vector B.  All
-  algorithms performed similarly.  The random method actually got
-  lucky this time and produced the best gait (17.26 body
-  lengths/minute) and uniform random hill climbing produced the worst
-  (9.44 body lengths/minute). The remaining algorithms each produced a
-  gait around 13 body lengths/minute.}{}
-
-\figvarp{vectorC}{1}{Results for runs starting with vector C.
-  Simplex and Gaussian random hill climbing each produced a gait that
-  substantially outperformed the other algorithms, with around 15 body
-  lengths/minute and 13 body lengths/minute, respectively, whereas the
-  other algorithms returned gaits of less than 5 body
-  lengths/minute.}{}
-
-Based on the three trials, as shown in \figref{std_error}, linear
-regression worked the best, followed by simplex. Gaussian random hill
-climbing performed about as well as random, but it is unclear whether
-Gaussian would continue to improve and become much better than random
-search if allowed to run for more iterations. Uniform random hill
-climbing and policy gradient descent performed similarly to each other
-on average, but differed greatly on individual runs.
-
-\figvarp{std_error}{1}{Average results for each method starting with
-  parameter vectors A, B, and C.  Error bars are plotted showing the
-  standard error.  Linear regression outperformed other methods,
-  followed by simplex. Gaussian random hill climbing performed about
-  as well as random, but it is unclear whether Gaussian would continue
-  to improve and become much better than random search if allowed to
-  run for more iterations. Uniform random hill climbing and policy
-  gradient descent performed similarly to each other on average, but
-  differed greatly on individual runs.}{}
-
-In addition to running strategies to optimize parameterized gaits, we
-also wanted to investigate the space of possible gaits.  To accomplish
-this, we selected a parameter vector that resulted in motion, but not
-an exceptional gait, and plotted performance along each dimension
-individually.  In addition, we duplicated each measurement to be able
-to estimate the measurement noise at each point.  Results for this
-exploration are shown in \figref{explore_dim_1} through
-\figref{explore_dim_5}.
-
-\figvarp{explore_dim_1}{.5}{Motion for varying just dimension 1}{}
-
-\figvarp{explore_dim_2}{.5}{Motion for varying just dimension 2}{}
-
-\figvarp{explore_dim_3}{.5}{Motion for varying just dimension 3}{}
-
-\figvarp{explore_dim_4}{.5}{Motion for varying just dimension 4}{}
-
-\figvarp{explore_dim_5}{.5}{Motion for varying just dimension 5}{}
-
-\subsection{Discussion}
-% Is your hypothesis supported? What conclusions do the results support
-% about the strengths and weaknesses of your method compared to other 
-% methods? How can the results be explained in terms of the underlying 
-% properties of the algorithm and/or the data.
-
-Because only three trials were tested with each algorithm and no
-algorithm consistently outperformed the others, there is a large
-standard error for each algorithm, as show in \figref{std_error}. For
-this reason, it is unclear if any algorithm is superior to another in
-this application. More trials with each algorithm would be necessary
-to reach a definitive ranking of the performance of the algorithms.
-
-However, each algorithm discovered at least one gait of over 10 body
-lengths/minute, including random. For this reason, we speculate that
-the motion model is at least as critical as the learning algorithm
-used.
-
-\section{Future work}
-% What are the major shortcomings of your current method? For each 
-% shortcoming, propose additions or enhancements that would help overcome it.
-
-There are several directions in which we could continue our
-work. First, the error margins for our runs were large, so reducing
-them by running more trials would lead to more conclusive data. It
-would also be insightful to run our algorithms on different motion
-models.  We suspect that our choice of motion model influenced the
-results greatly, as even random choices in the space produced gaits
-that moved a significant fraction of the time (> 5\%).  It would be
-interesting to see how learning methods would perform using a model
-that included a much higher percentage of unproductive gaits. We also
-intend to experiment further with SVM regression and evolutionary
-algorithms/HyperNEAT.  Some parameter vectors resulted in the robot
-turning, as opposed to it walking long distances. This could be an
-interesting learning goal in future projects. To these ends, we
-propose the following additions and enhancements:
-
-\begin{itemize}
-\item More runs and/or longer runs
-\item Different motion representations
-\item Better tuning of SVM regression
-\item Evolutionary algorithms/HyperNEAT\cite{clune}
-\item Learning how to turn
-\end{itemize}
-
-\section{Conclusion}
-% Briefly summarize the important results and conclusions presented in 
-% the paper. What are the most important points illustrated by your 
-% work? How will your results improve future research and applications 
-% in the area?
-
-We have presented an array of approaches to optimizing a quadrupedal
-gait for forward speed.  We have implemented and tested different
-learning strategies, including uniform and Gaussian random hill
-climbing, policy gradient reinforcement learning, Nelder-Mead simplex,
-several new predictive methods based on linear and support vector
-regression, and an evolved neural network (HyperNEAT).  We have also
-compared these approaches to random search as a baseline. Many of the
-methods resulted in walks significantly faster than previously
-hand-tuned gaits.
-
-Because only three trials were tested with each algorithm and no
-algorithm consistently outperformed the others, there was a large
-standard error for each method, as shown in \figref{std_error}. Thus
-it was unclear if any algorithm was superior to another in this
-application. More trials with each algorithm would be necessary to
-reach a definitive ranking. Because each algorithm discovered at least one gait of
-over 10 body lengths/minute, including random search, we also
-conjectured that the motion representation for the robot is more
-integral to forward speed than the learning algorithm.  How to learn
-the motion representation, in addition to its parameters, remains an
-open problem.
-
-\section{Acknowledgments}
-% List any people not on the team who helped you with your project: Your
-% TA, other people you consulted with or had useful discussions (say in 
-% a word or two what they did), people who proofread your report, and 
-% any external code you used (libraries etc).
-\begin{itemize}
-\item Hod Lipson, Cornell Computational Synthesis Lab: adviser.
-\item Jim T\o rreson, University of Oslo: adviser.
-\item Juan Zagal, University of Chile: designed and printed robot and provided code for hand tuned gait.
-\item Jeff Clune, CCSL Red Couch: collaborated on HyperNEAT implementation/testing.
-\item Cooper Bills, Cornell University: assisted with Wii tracker development.
-\item Anshumali Srivastava, Cornell University: Teaching Assistant.
-\end{itemize}
-
-\section{The Team}
-% A table listing the names of the people who worked on the project, and who
-% did what.
-
-\begin{itemize}
-
-\item Diana Hidalgo: 
-\begin{itemize}
-\item Simplex (Nelder-Mead)
-\item \code{WiiTrackClient} and \code{WiiTrackServer} classes and hardware
-\item Plotting and evaluating results
-\item Investigation into using Optitrack IR system for tracking robot
-\end{itemize}
-
-\item Sarah Nguyen
-\begin{itemize}
-\item Random search
-\item Uniform random hill climbing
-\item Gaussian random hill climbing
-\item N-dimensional policy gradient descent
-\item Linear regression and prediction
-\item \code{RunManager} and \code{Strategy} classes
-\item Video editing
-\end{itemize}
-
-\item Jason Yosinski
-\begin{itemize}
-\item SVM regression and prediction
-\item Evolutionary Neural Network (HyperNEAT)
-\item \code{WiiTrackClient} and \code{WiiTrackServer} classes and hardware
-\item \code{Robot}, \code{RunManager}, and \code{Strategy} classes
-\item \code{Parameterized Motion Model} abstract base class
-\item \code{SineModel} classes
-\item \code{explore\_dimensions}
-\item Plotting and evaluating results
-\end{itemize}
-
-\end{itemize}
-
-\section{Appendix}
-% Describe the code files you uploaded into CMS
-
-A brief description of the code uploaded to the CMS follows:
-
-\begin{itemize}
-\item \code{optimize.py}: Main program, determines a strategy to try and runs the robot
-  with that strategy.
-\item \code{RunManager.py}: Deals with all the details of running the robot,
-  including choosing an initial parameter, running the robot multiple times, 
-  tracking distance walked, and writing to the log file. Also includes the
-  \code{explore\_dimensions} method.
-\item \code{Strategy.py}: Contains all the different possible strategies, which
-  will be passed as objects in \code{optimize.py}. 
-\item \code{Robot.py}: Implements the \code{Robot} class, described in
-  \secref{implement}.
-\item \code{SineModel.py}: Implements a sine based motion model,
-  described in \secref{implement}.
-\item \code{Motion.py}: Motion helper functions.
-\item \code{WiiTrackServer.py}: Broadcasts the position of the infrared LED.
-\item \code{WiiTrackClient.py}: Connects to the WiiTrackServer to get the current position information.
-\end{itemize}
-
-% LocalWords:  vertices

### 2011-01-20 paper/relatedWork.tex
Restructuring of report into multiple files to minimize conflicts, references working, dealt old sections to appropriate locations in paper framework.

--- 2010-12-07 report/report.tex
@@
-\section{Introduction}
+\subsection{Related Work}
 
-% Motivate and abstractly describe the problem you are addressing and
-% how you are addressing it. What is the problem? Why is it important?
-% What is your basic approach? A short discussion of how it fits into
-% % related work in the area is also desirable. Summarize the basic
-% results and conclusions that you will present.
+\editbox{If we talk about past gait-learning work in here, is it a bit redundant?}
 
-Applications of walking robots often call for the ability to walk as
-quickly, efficiently, or with as little power as possible.  Often
-these optimizations are done manually by an expert who designs and
-tweaks a gait specifically for a given objective.  Other groups have
-used learning methods to generate gaits optimized for some metric.
-Approaches differ in their starting assumptions, some essentially
-tweaking the parameters of a hand-tuned model \cite{chernova}, others
-exploring a reasonably compact parameter space \cite{kohl}, and still
-others beginning with few assumptions besides periodicity
-\cite{zykov}.
+\edit{write this.  Old quadratot section below}
 
-We aimed to strike a middle ground between these approaches.  Our
-motion generator did not rely on hand-tweaked gaits, but it did use
-parameterized gaits which, by their nature, impose some assumptions on
-the answers produced.  We then used machine learning to design gaits
-for a quadruped robot with these models.  This paper presents a
-comparison of the different learning methods implemented.  Most
-methods created walks that are several times faster than the original
-hand-tuned gait.  We invite readers with short attention spans to view
-a video of some of our results online here:
-
-\url{http://www.youtube.com/watch?v=ODoiOj9DdGg}
-
-%The code so far consists of the robot class, optimization class,
-%parameterized model, sine model, and camera feedback.
-
-\section{Problem definition}
-% Precisely define the problem you are addressing (i.e. formally specify
-% the inputs and outputs). Elaborate on why this is an interesting and
-% important problem.
-
-We are testing several different learning methods to design a
-parametrized gait for a quadruped robot from the Cornell Computational
-Synthesis Lab.
-
-The output each of the learning algorithms is a function of time,
-$f(t)$, that outputs a vector of commanded motor positions.  This
-function is generated using a parametrized motion model, described in
-\secref{motion}.
-
-The robot executes these commands and measures its change in location
-using the tracking system described in \secref{implement}.  The input
-to the learning algorithms is this measured displacement, which the
-algorithms attempt to maximize. This displacement is measured for each
-gait over a constant length run, usually 12 seconds.
-
-A comparison and evaluation of the many different methods available
-for optimizing the gait of legged robots will be useful for future
-work on this challenging multidimensional control problem.
-
-\section{Method}
-\seclabel{method}
-
-% Describe in reasonable detail the algorithm you are using to address
-% this problem. A pseudo-code description of the algorithm you are
-% using is frequently useful. If it makes sense for your project,
-% trace through a concrete example, showing how your algorithm
-% processes this example. The example should be complex enough to
-% illustrate all of the important aspects of the problem but simple
-% enough to be easily understood. If possible, an intuitively
-% meaningful example is better than one with meaningless symbols.
-
-We use several parameterized motion models that command motors to
-positions based on a sine wave, creating a periodic pattern.  While we
-investigated several models, for the bulk of our experiments, we used
-a model whose five parameters are: amplitude, wavelength, scale inner
-vs outer motors, scale left vs right motors, scale back vs front
-motors. Each strategy below attempts to choose the best possible
-parameters for this motion model.
-
-We implemented and tested 8 different learning strategies.  All
-strategies except for the HyperNEAT method\cite{clune} were
-constrained to pick parameters from within predetermined ranges.
-
-\begin{itemize}
-
-\item \emph{Random}: This method randomly generates parameter vectors
-  in the allowable range. This strategy is used only as baseline.
-
-\item \emph{Uniform random hill climbing}: This method begins by
-  selecting a single random parameter vector.  Subsequent iterations
-  generate a neighbor by randomly choosing one parameter to adjust and
-  replacing it with a new value chosen with uniform probability in the
-  allowable range for that parameter. The neighbor is evaluated by
-  running the robot with the newly chosen parameters. If this neighbor
-  results in a longer distance walked than the previous best gait, it
-  is saved as the new best gait. The process is then repeated, always
-  starting with the best gait.
-
-\item \emph{Gaussian random hill climbing}: This method works
-  similarly to Uniform random hill climbing, except neighbors are
-  generated by adding random Gaussian noise to the current best gait.
-  This results in all parameters being changed at once, but the
-  resulting vector is always fairly close to the previous best gait.
-  We used independently selected noise in each dimension, scaled such
-  that the standard deviation of the noise was 5\% of the range of
-  that dimension.
-
-\item \emph{N-dimensional policy gradient descent}: As opposed to the
-  previous methods, this method explicitly estimates the gradient for
-  the objective function. It does this by first evaluating \emph{t}
-  randomly generated parameter vectors near the initial vector, each
-  dimension of these vectors being perturbed by either $-\epsilon$,
-  $0$, or $\epsilon$. Then, for each dimension, it groups vectors into
-  three groups: $-\epsilon$, $0$, and $\epsilon$.  The gradient along
-  this dimension is then estimated as the average score for the
-  $\epsilon$ group minus the average score for the $-\epsilon$
-  group. Finally, the method creates a new vector by changing all
-  parameters by a fixed-size step in the direction of the gradient.
-
-\item \emph{Nelder-Mead simplex method}\cite{nm}: The Nelder-Mead
-  simplex method creates an initial simplex with 6 vertices. The
-  initial parameter vector is stored as the first vertex and the other
-  five vertices are created by adding to one dimension at a time one
-  tenth of the allowable range for that parameter. It then tests the
-  fitness of each vertex and based on these fitnesses, it reflects the
-  worst point over the centroid in an attempt to improve it.  However,
-  to prevent cycles and becoming stuck in local minima, several other
-  rules are used.  In general, the worst vertex is reflected over the
-  centroid. If the reflected point is better than the second worst
-  point and worse than the best point, then the reflected point
-  replaces the worst. If the reflected point is better than the best
-  point, the simplex is expanded in the direction of the reflected
-  point. The better of the reflected and the expanded point replaces
-  the worst point. If the reflected point is worse than the second
-  worst point, then the simplex is contracted away from the reflected
-  point. If the contracted point is better than the reflected point,
-  the contracted point replaces the worst point. If the contracted
-  point is worse than the reflected point, the entire simplex is
-  shrunk \cite{NMWebsite}.
-
-\item \emph{Linear regression}: To initialize, this method chooses and
-  evaluates five random parameter vectors. It then fits a linear model
-  from parameter vector to fitness. In a loop, the method chooses and
-  evaluates a new parameter vector generated by taking a fixed-size
-  step in the direction of the gradient for each parameter, and fits a
-  new linear model to all vectors evaluated so far, choosing the model
-  to minimize the sum of squared errors.
-
-\item \emph{SVM regression}: Similarly to linear regression, this
-  model starts with several random vectors, but this time they are
-  chosen in a small neighborhood about some initial random vector.
-  These vectors (generally 8) are evaluated, and a support vector
-  regression model is fit to the observed fitnesses.  To choose the
-  next vector for evaluation, we randomly generate some number
-  (typically 100) of vectors in the neighborhood of the best observed
-  gait, and select for evaluation the vector with the best predicted
-  performance.  We suspected that if we always chose the best
-  predicted point out of 100, we may end up progressing along a narrow
-  subspace, prohibiting learning of the true local fitness function.
-  Alternately, we would always choosing exploitation of knowledge
-  vs. exploration of the space.  To address this concern, we added a
-  parameter dubbed \code{bumpBy} that added noise to the final
-  selected point before it was evaluated.
-
-  Such a method naturally has many tunable parameters, and we
-  endeavored to select these parameteres by tuning the method in
-  simulation.  To estimate the performance of the algorithm, we ran it
-  agains a simulation with a known optimum.  The simulated function
-  was in the same five dimensional parameter space, and simply
-  returned a fitness determined as the height of a Gaussian with a
-  random mean.  The width of the Gaussian in each dimension was 20\%
-  of the range of each dimension, and the maximum value at the peak
-  was 100.  \figref{svm_sim_results} shows the learning results on
-  this simulated model using the ultimately selected SVM parameters.
-  Interestingly, a non-zero value of \code{bumpBy} resulted in better
-  learning than a zero valued, noise free (exploration free) learning.
-
-  Ultimately, however, the version of SVM tuned for simulation did not
-  yet show competitive performance on the real robot.  We tried tuning
-  some parameters on the real robot, but after some amount of tuning,
-  the method still exibited too little exploration and easily became
-  stuck in local minima.
-
-\item \emph{Evolutionary Neural Network (HyperNEAT)\cite{clune}}: Near
-  the end of term, we hacked together an interface between HyperNEAT
-  and the robot, requiring a slightly modified strategy interface.
-  Preliminary HyperNEAT runs were promising and resulted in serveral
-  interesting gaits.  Unfortunately, the gaits generated by HyperNEAT
-  also tended to stress the robot more than typical gaits had before,
-  and the servos would often overheat and malfunction, requiring
-  restart.  We think these issues may be addressed by adding a small
-  layer between the HyperNEAT strategy and the robot that disallows
-  quickly shifting commanded positions, and we hope to be able to test
-  these methods further once this filter is in place.
-
-\end{itemize}
-
-\figvarp{svm_sim_results}{.6}{Results for the SVM regression strategy
-  in simulation.  This simulation was used to tune the SVM strategy's
-  parameters before trying it on the physical robot.  The strategy
-  quickly approaches the maximum simulated fitness of 100.}{}
-
-\section{Related work}
 % Briefly explain who else worked on related problems in the past and what
@@
 pattern beyond the fact that it should be rhythmic.
-
-\section{System Architecture and Implementation}
-\seclabel{implement}
-% Describe how you implemented your system and how you structured it. 
-% This should give an overview of the system, not a detailed 
-% documentation of the code. The documentation of the code is part of 
-% the code you hand in. You might want to comment on high-level design 
-% decisions that you made. Also explain how you obtained your
-% data and any pre-processing you did to it.
-
-The quadruped robot has an on-board computer running Linux. The lower
-level drivers are in C and the system is implemented in
-Python. Feedback about distance travelled is provided via an infrared
-LED mounted on the robot and a Wii remote fixed to the ceiling.
-
-An overview of the code follows.
-
-\begin{itemize}
-
-\item \code{Robot} class: Class wrapper for commanding
-  motion of the robot.  The \code{Robot} class takes care of the robot
-  initialization, communication with the servos, and timing of the
-  runs.  In addition, it prevents the servos from ever being commanded
-  to a point outside their normal range (0 - 1023) as well as beyond
-  points where limbs would collide with parts of the robot body.  The
-  main class function, \code{run}, accepts a motion model (any
-  function that takes a time argument and outputs a 9 dimensional
-  position) and will run the robot using this motion model, including,
-  if desired, smooth interpolation over time for the beginning and end
-  of the run.
-
-\item \code{RunManager} class: Deals with all the details of running
-  the robot, including running the robot multiple times, tracking
-  distance walked via a \code{WiiTrackClient} member object, and
-  writing results to the log file. Also includes the
-  \code{explore\_dimensions} method to generate plots by varying each
-  parameter independently.
-
-\item \code{Strategy} class: The user has a choice between eight
-  different learning strategies: Random search, uniform random hill
-  climbing, Gaussian random hill climbing, N-dimensional policy
-  gradient descent, Nelder-Mead simplex, linear regression/prediction,
-  SVM regression/prediction, and HyperNEAT evolution.  Each strategy
-  must derive from the base \code{Strategy} class and must implement
-  two methods: \code{getNext}, for getting the next parameter vector
-  to try, and \code{updateResults}, for communicating results of a run
-  back to the strategy.
-
-\item \code{MotionModel} class: We implemented several motion models,
-  the main being the \code{SineModel5} class.  All motion models take
-  as input a parameter vector and produce as output a motion model,
-  which is simply a function mapping from time to nine motor
-  positions.  The \code{SineModel5} model commands positions based on
-  a sine wave, creating a periodic pattern. The parameters are:
-  amplitude, wavelength, a multiplier for the inner vs outer motors,
-  multiplier for left vs right motors, and multiplier for back vs
-  front motors.  Other similar models were tested, including a seven
-  parameter model which allowed sine waves shifted in time, but these
-  were not used as extensively in our experiments as
-  \code{SineModel5}.
-
-\item \code{WiiTrackClient} and \code{WiiTrackServer} classes and
-  hardware: A Wii remote tracks the location of the robot through an
-  infrared LED mounted on top of the robot. A \code{WiiTrackServer} is
-  run on the robot and continuously tracks its position using the
-  CWiid library\cite{cwiid} to interface with the remote via
-  bluetooth.  The \code{RunManager} then makes a
-  \code{WiiTrackClient}, which connects via socket to the tracking
-  server and requests position updates periodically.
-  \code{RunManager} currently gets the robot's position at the
-  beginning of each run and then again at the end and uses this to
-  calculate the net change in position.
-
-\end{itemize}
-
-\section{Experimental Evaluation}
-
-\subsection{Methodology}
-% What are the criteria you are using to evaluate your method? What
-% specific hypotheses does your experiment test? Describe the 
-% experimental methodology that you used. What are the dependent and 
-% independent variables? For projects in machine learning, what is 
-% the training/test data that was used, and why is it realistic or
-% interesting? Exactly what performance data did you collect and how 
-% are you presenting and analyzing it? Comparisons to competing methods 
-% that address the same problem or to variations of your own algorithm 
-% are particularly useful.
-
-The metric for evaluation of the designed gait is speed. We stop each 
-run after plateauing results (no improvement for one third of the 
-policies seen so far). The standard length of a run designates that it
-should be stopped after there is no improvement for one half of the policies
-seen so far, but since all runs took place on the actual robot, without use
-of a simulator, certain time limitations were imposed on the learning process.
-
-We controlled our experiments from a computer that was connected via a
-wireless ethernet to the robot. The robot collected data about
-distance walked automatically on its own. If it walked outside of the
-Wii remote's viewable area, it informed the user, so the only human
-intervention required during an experiment was to move the robot back
-inside the viewable area and to resume the run, which did not
-interrupt the learning process or result in the loss of data.
-
-We evaluated the efficacy of a set of parameters by sending these
-parameters to the robot and instructing it to walk for a certain
-length of time. The robot always began from the same position and
-returned to the starting position at the end of the run in order to
-measure true displacement without giving credit for ending in a leaned
-position. More efficient parameters resulted in a faster gait, which
-translated into a longer distance walked and a better score. After completing
-an evaluation, the robot sent the resulting distance walked back to the
-host computer and prepared itself for a new set of parameters to evaluate.
-
-Each algorithm was run on 3 different initial parameter vectors on the
-physical robot. We decided to evaluate all methods starting at the
-same three vector in order to allow for the fair comparison of each
-algorithm.  We evaluate each method based on the amount of improvement
-seen from the initial parameter vectors, and on the fastest speeds
-achieved during runs.
-
-The resulting gaits from our algorithms quickly outperformed the original
-hand-coded walk designed for this robot. The fastest walk, for example, was
-4 times faster.
-
-\subsection{Results}
-% Present the quantitative results of your experiments. Graphical data
-% presentation such as graphs and histograms are frequently better 
-% than tables. Explain what are the basic differences revealed in the 
-% data. If your method is stochastic or you analyze multiple datasets, 
-% calculate mean and standard error for every quantitative
-% performance result you show based on multiple runs.
-
-We have done complete runs of 3 different initial parameter vectors
-for random search, uniform random hill climbing, Gaussian random hill
-climbing, policy gradient descent, Nelder-Mead simplex, and linear
-regression.  We also evaluated the SVM regression and HyperNEAT
-methods, but these methods were more experimental, and thus we do not
-yet have the same volume of data for these runs.  We developed several
-gaits that were about 4 times faster than the original hand-coded
-gait.  Results are shown in \tabref{results_table}.
-
-\begin{table}
-\begin{center}
-\begin{tabular}{|r|c|c|c||c|}
-\hline
-                                         & A       & B      & C      &  Average \\
-\hline                                                               
-\hline                                                               
-Previous hand-coded gait                 & --      & --     & --     &  5.16 \\
-\hline                                                                 
-Random search                            & 6.04    & 17.26  & 4.90   &  9.40 \\
-\hline                                                                 
-Uniform Random Hill Climbing             & 11.37   & 9.44   & 2.69   &  7.83 \\
-\hline                                                                 
-Gaussian Random Hill Climbing            & 3.10    & 13.59  & 13.40  &  10.03 \\
-\hline                                                                 
-Policy Gradient Descent                  & 0.68    & 14.69  & 3.60   &  6.32 \\
-\hline                                                                 
-Nelder-Mead simplex                      & 8.51    & 13.62  & 14.83  &  12.32 \\
-\hline                                                                 
-Linear Regression                        & 27.58   & 12.51  & 1.95   &  14.01 \\
-%\hline                                                                
-%SVM Regression                           & ???     & ???    & ???   &   \\
-%\hline                                                                
-%Evolutionary Neural Network (HyperNEAT)  & ???     & ???    & ???   &   \\
-\hline
-\end{tabular}
-\caption{The best gaits found for each starting vector and algorithm,
-  in body lengths per minute.}
-\tablabel{results_table}
-\end{center}
-\end{table}
-
-\begin{itemize}
-
-\item For vector A, shown in \figref{vectorA}, linear regression
-  worked significantly better than the other algorithms, resulting in
-  a gait (27.58 body lengths/minute) that walked over twice as fast as
-  the next best gait -- uniform random hill climbing at 11.37 body
-  lengths/minute -- and 5.3 times better than the previous hand-coded
-  gait.
-
-\item Vector B, depicted in \figref{vectorB}, resulted in similarly
-  performing gaits with all the algorithms. The random method got
-  lucky and produced the best gait (17.26 body lengths/minute) and
-  uniform random hill climbing produced the worst (9.44 body
-  lengths/minute). The remaining algorithms each produced a gait
-  around 13 body lengths/minute.
-
-\item For vector C, shown in \figref{vectorC}, simplex and Gaussian
-  random hill climbing each produced a gait that substantially
-  outperformed the other algorithms. Simplex resulted in a gait of
-  nearly 15 body lengths/minute and Gaussian random hill climbing
-  produced a gait of just over 13 body lengths/minute, whereas the
-  other algorithms returned gaits of less than 5 body lengths/minute.
-
-\end{itemize}
-
-\figvarp{vectorA}{1}{Results for runs beginning with vector A.
-  Linear regression worked significantly better than the other
-  algorithms, resulting in a gait (27.58 body lengths/minute) that
-  walked over twice as fast as the next best gait -- uniform random
-  hill climbing at 11.37 body lengths/minute -- and 5.3 times better
-  than the previous hand-coded gait.}{}
-
-\figvarp{vectorB}{1}{Results for run starting with vector B.  All
-  algorithms performed similarly.  The random method actually got
-  lucky this time and produced the best gait (17.26 body
-  lengths/minute) and uniform random hill climbing produced the worst
-  (9.44 body lengths/minute). The remaining algorithms each produced a
-  gait around 13 body lengths/minute.}{}
-
-\figvarp{vectorC}{1}{Results for runs starting with vector C.
-  Simplex and Gaussian random hill climbing each produced a gait that
-  substantially outperformed the other algorithms, with around 15 body
-  lengths/minute and 13 body lengths/minute, respectively, whereas the
-  other algorithms returned gaits of less than 5 body
-  lengths/minute.}{}
-
-Based on the three trials, as shown in \figref{std_error}, linear
-regression worked the best, followed by simplex. Gaussian random hill
-climbing performed about as well as random, but it is unclear whether
-Gaussian would continue to improve and become much better than random
-search if allowed to run for more iterations. Uniform random hill
-climbing and policy gradient descent performed similarly to each other
-on average, but differed greatly on individual runs.
-
-\figvarp{std_error}{1}{Average results for each method starting with
-  parameter vectors A, B, and C.  Error bars are plotted showing the
-  standard error.  Linear regression outperformed other methods,
-  followed by simplex. Gaussian random hill climbing performed about
-  as well as random, but it is unclear whether Gaussian would continue
-  to improve and become much better than random search if allowed to
-  run for more iterations. Uniform random hill climbing and policy
-  gradient descent performed similarly to each other on average, but
-  differed greatly on individual runs.}{}
-
-In addition to running strategies to optimize parameterized gaits, we
-also wanted to investigate the space of possible gaits.  To accomplish
-this, we selected a parameter vector that resulted in motion, but not
-an exceptional gait, and plotted performance along each dimension
-individually.  In addition, we duplicated each measurement to be able
-to estimate the measurement noise at each point.  Results for this
-exploration are shown in \figref{explore_dim_1} through
-\figref{explore_dim_5}.
-
-\figvarp{explore_dim_1}{.5}{Motion for varying just dimension 1}{}
-
-\figvarp{explore_dim_2}{.5}{Motion for varying just dimension 2}{}
-
-\figvarp{explore_dim_3}{.5}{Motion for varying just dimension 3}{}
-
-\figvarp{explore_dim_4}{.5}{Motion for varying just dimension 4}{}
-
-\figvarp{explore_dim_5}{.5}{Motion for varying just dimension 5}{}
-
-\subsection{Discussion}
-% Is your hypothesis supported? What conclusions do the results support
-% about the strengths and weaknesses of your method compared to other 
-% methods? How can the results be explained in terms of the underlying 
-% properties of the algorithm and/or the data.
-
-Because only three trials were tested with each algorithm and no
-algorithm consistently outperformed the others, there is a large
-standard error for each algorithm, as show in \figref{std_error}. For
-this reason, it is unclear if any algorithm is superior to another in
-this application. More trials with each algorithm would be necessary
-to reach a definitive ranking of the performance of the algorithms.
-
-However, each algorithm discovered at least one gait of over 10 body
-lengths/minute, including random. For this reason, we speculate that
-the motion model is at least as critical as the learning algorithm
-used.
-
-\section{Future work}
-% What are the major shortcomings of your current method? For each 
-% shortcoming, propose additions or enhancements that would help overcome it.
-
-There are several directions in which we could continue our
-work. First, the error margins for our runs were large, so reducing
-them by running more trials would lead to more conclusive data. It
-would also be insightful to run our algorithms on different motion
-models.  We suspect that our choice of motion model influenced the
-results greatly, as even random choices in the space produced gaits
-that moved a significant fraction of the time (> 5\%).  It would be
-interesting to see how learning methods would perform using a model
-that included a much higher percentage of unproductive gaits. We also
-intend to experiment further with SVM regression and evolutionary
-algorithms/HyperNEAT.  Some parameter vectors resulted in the robot
-turning, as opposed to it walking long distances. This could be an
-interesting learning goal in future projects. To these ends, we
-propose the following additions and enhancements:
-
-\begin{itemize}
-\item More runs and/or longer runs
-\item Different motion representations
-\item Better tuning of SVM regression
-\item Evolutionary algorithms/HyperNEAT\cite{clune}
-\item Learning how to turn
-\end{itemize}
-
-\section{Conclusion}
-% Briefly summarize the important results and conclusions presented in 
-% the paper. What are the most important points illustrated by your 
-% work? How will your results improve future research and applications 
-% in the area?
-
-We have presented an array of approaches to optimizing a quadrupedal
-gait for forward speed.  We have implemented and tested different
-learning strategies, including uniform and Gaussian random hill
-climbing, policy gradient reinforcement learning, Nelder-Mead simplex,
-several new predictive methods based on linear and support vector
-regression, and an evolved neural network (HyperNEAT).  We have also
-compared these approaches to random search as a baseline. Many of the
-methods resulted in walks significantly faster than previously
-hand-tuned gaits.
-
-Because only three trials were tested with each algorithm and no
-algorithm consistently outperformed the others, there was a large
-standard error for each method, as shown in \figref{std_error}. Thus
-it was unclear if any algorithm was superior to another in this
-application. More trials with each algorithm would be necessary to
-reach a definitive ranking. Because each algorithm discovered at least one gait of
-over 10 body lengths/minute, including random search, we also
-conjectured that the motion representation for the robot is more
-integral to forward speed than the learning algorithm.  How to learn
-the motion representation, in addition to its parameters, remains an
-open problem.
-
-\section{Acknowledgments}
-% List any people not on the team who helped you with your project: Your
-% TA, other people you consulted with or had useful discussions (say in 
-% a word or two what they did), people who proofread your report, and 
-% any external code you used (libraries etc).
-\begin{itemize}
-\item Hod Lipson, Cornell Computational Synthesis Lab: adviser.
-\item Jim T\o rreson, University of Oslo: adviser.
-\item Juan Zagal, University of Chile: designed and printed robot and provided code for hand tuned gait.
-\item Jeff Clune, CCSL Red Couch: collaborated on HyperNEAT implementation/testing.
-\item Cooper Bills, Cornell University: assisted with Wii tracker development.
-\item Anshumali Srivastava, Cornell University: Teaching Assistant.
-\end{itemize}
-
-\section{The Team}
-% A table listing the names of the people who worked on the project, and who
-% did what.
-
-\begin{itemize}
-
-\item Diana Hidalgo: 
-\begin{itemize}
-\item Simplex (Nelder-Mead)
-\item \code{WiiTrackClient} and \code{WiiTrackServer} classes and hardware
-\item Plotting and evaluating results
-\item Investigation into using Optitrack IR system for tracking robot
-\end{itemize}
-
-\item Sarah Nguyen
-\begin{itemize}
-\item Random search
-\item Uniform random hill climbing
-\item Gaussian random hill climbing
-\item N-dimensional policy gradient descent
-\item Linear regression and prediction
-\item \code{RunManager} and \code{Strategy} classes
-\item Video editing
-\end{itemize}
-
-\item Jason Yosinski
-\begin{itemize}
-\item SVM regression and prediction
-\item Evolutionary Neural Network (HyperNEAT)
-\item \code{WiiTrackClient} and \code{WiiTrackServer} classes and hardware
-\item \code{Robot}, \code{RunManager}, and \code{Strategy} classes
-\item \code{Parameterized Motion Model} abstract base class
-\item \code{SineModel} classes
-\item \code{explore\_dimensions}
-\item Plotting and evaluating results
-\end{itemize}
-
-\end{itemize}
-
-\section{Appendix}
-% Describe the code files you uploaded into CMS
-
-A brief description of the code uploaded to the CMS follows:
-
-\begin{itemize}
-\item \code{optimize.py}: Main program, determines a strategy to try and runs the robot
-  with that strategy.
-\item \code{RunManager.py}: Deals with all the details of running the robot,
-  including choosing an initial parameter, running the robot multiple times, 
-  tracking distance walked, and writing to the log file. Also includes the
-  \code{explore\_dimensions} method.
-\item \code{Strategy.py}: Contains all the different possible strategies, which
-  will be passed as objects in \code{optimize.py}. 
-\item \code{Robot.py}: Implements the \code{Robot} class, described in
-  \secref{implement}.
-\item \code{SineModel.py}: Implements a sine based motion model,
-  described in \secref{implement}.
-\item \code{Motion.py}: Motion helper functions.
-\item \code{WiiTrackServer.py}: Broadcasts the position of the infrared LED.
-\item \code{WiiTrackClient.py}: Connects to the WiiTrackServer to get the current position information.
-\end{itemize}
-
-% LocalWords:  vertices

### 2011-01-20 paper/report.tex
Restructuring of report into multiple files to minimize conflicts, references working, dealt old sections to appropriate locations in paper framework.

--- 2010-12-07 report/report.tex
@@
-\section{Introduction}
-
-% Motivate and abstractly describe the problem you are addressing and
-% how you are addressing it. What is the problem? Why is it important?
-% What is your basic approach? A short discussion of how it fits into
-% % related work in the area is also desirable. Summarize the basic
-% results and conclusions that you will present.
-
-Applications of walking robots often call for the ability to walk as
-quickly, efficiently, or with as little power as possible.  Often
-these optimizations are done manually by an expert who designs and
-tweaks a gait specifically for a given objective.  Other groups have
-used learning methods to generate gaits optimized for some metric.
-Approaches differ in their starting assumptions, some essentially
-tweaking the parameters of a hand-tuned model \cite{chernova}, others
-exploring a reasonably compact parameter space \cite{kohl}, and still
-others beginning with few assumptions besides periodicity
-\cite{zykov}.
-
-We aimed to strike a middle ground between these approaches.  Our
-motion generator did not rely on hand-tweaked gaits, but it did use
-parameterized gaits which, by their nature, impose some assumptions on
-the answers produced.  We then used machine learning to design gaits
-for a quadruped robot with these models.  This paper presents a
-comparison of the different learning methods implemented.  Most
-methods created walks that are several times faster than the original
-hand-tuned gait.  We invite readers with short attention spans to view
-a video of some of our results online here:
-
-\url{http://www.youtube.com/watch?v=ODoiOj9DdGg}
-
-%The code so far consists of the robot class, optimization class,
-%parameterized model, sine model, and camera feedback.
-
-\section{Problem definition}
-% Precisely define the problem you are addressing (i.e. formally specify
-% the inputs and outputs). Elaborate on why this is an interesting and
-% important problem.
-
-We are testing several different learning methods to design a
-parametrized gait for a quadruped robot from the Cornell Computational
-Synthesis Lab.
-
-The output each of the learning algorithms is a function of time,
-$f(t)$, that outputs a vector of commanded motor positions.  This
-function is generated using a parametrized motion model, described in
-\secref{motion}.
-
-The robot executes these commands and measures its change in location
-using the tracking system described in \secref{implement}.  The input
-to the learning algorithms is this measured displacement, which the
-algorithms attempt to maximize. This displacement is measured for each
-gait over a constant length run, usually 12 seconds.
-
-A comparison and evaluation of the many different methods available
-for optimizing the gait of legged robots will be useful for future
-work on this challenging multidimensional control problem.
-
-\section{Method}
-\seclabel{method}
-
-% Describe in reasonable detail the algorithm you are using to address
-% this problem. A pseudo-code description of the algorithm you are
-% using is frequently useful. If it makes sense for your project,
-% trace through a concrete example, showing how your algorithm
-% processes this example. The example should be complex enough to
-% illustrate all of the important aspects of the problem but simple
-% enough to be easily understood. If possible, an intuitively
-% meaningful example is better than one with meaningless symbols.
-
-We use several parameterized motion models that command motors to
-positions based on a sine wave, creating a periodic pattern.  While we
-investigated several models, for the bulk of our experiments, we used
-a model whose five parameters are: amplitude, wavelength, scale inner
-vs outer motors, scale left vs right motors, scale back vs front
-motors. Each strategy below attempts to choose the best possible
-parameters for this motion model.
-
-We implemented and tested 8 different learning strategies.  All
-strategies except for the HyperNEAT method\cite{clune} were
-constrained to pick parameters from within predetermined ranges.
-
-\begin{itemize}
-
-\item \emph{Random}: This method randomly generates parameter vectors
-  in the allowable range. This strategy is used only as baseline.
-
-\item \emph{Uniform random hill climbing}: This method begins by
-  selecting a single random parameter vector.  Subsequent iterations
-  generate a neighbor by randomly choosing one parameter to adjust and
-  replacing it with a new value chosen with uniform probability in the
-  allowable range for that parameter. The neighbor is evaluated by
-  running the robot with the newly chosen parameters. If this neighbor
-  results in a longer distance walked than the previous best gait, it
-  is saved as the new best gait. The process is then repeated, always
-  starting with the best gait.
-
-\item \emph{Gaussian random hill climbing}: This method works
-  similarly to Uniform random hill climbing, except neighbors are
-  generated by adding random Gaussian noise to the current best gait.
-  This results in all parameters being changed at once, but the
-  resulting vector is always fairly close to the previous best gait.
-  We used independently selected noise in each dimension, scaled such
-  that the standard deviation of the noise was 5\% of the range of
-  that dimension.
-
-\item \emph{N-dimensional policy gradient descent}: As opposed to the
-  previous methods, this method explicitly estimates the gradient for
-  the objective function. It does this by first evaluating \emph{t}
-  randomly generated parameter vectors near the initial vector, each
-  dimension of these vectors being perturbed by either $-\epsilon$,
-  $0$, or $\epsilon$. Then, for each dimension, it groups vectors into
-  three groups: $-\epsilon$, $0$, and $\epsilon$.  The gradient along
-  this dimension is then estimated as the average score for the
-  $\epsilon$ group minus the average score for the $-\epsilon$
-  group. Finally, the method creates a new vector by changing all
-  parameters by a fixed-size step in the direction of the gradient.
-
-\item \emph{Nelder-Mead simplex method}\cite{nm}: The Nelder-Mead
-  simplex method creates an initial simplex with 6 vertices. The
-  initial parameter vector is stored as the first vertex and the other
-  five vertices are created by adding to one dimension at a time one
-  tenth of the allowable range for that parameter. It then tests the
-  fitness of each vertex and based on these fitnesses, it reflects the
-  worst point over the centroid in an attempt to improve it.  However,
-  to prevent cycles and becoming stuck in local minima, several other
-  rules are used.  In general, the worst vertex is reflected over the
-  centroid. If the reflected point is better than the second worst
-  point and worse than the best point, then the reflected point
-  replaces the worst. If the reflected point is better than the best
-  point, the simplex is expanded in the direction of the reflected
-  point. The better of the reflected and the expanded point replaces
-  the worst point. If the reflected point is worse than the second
-  worst point, then the simplex is contracted away from the reflected
-  point. If the contracted point is better than the reflected point,
-  the contracted point replaces the worst point. If the contracted
-  point is worse than the reflected point, the entire simplex is
-  shrunk \cite{NMWebsite}.
-
-\item \emph{Linear regression}: To initialize, this method chooses and
-  evaluates five random parameter vectors. It then fits a linear model
-  from parameter vector to fitness. In a loop, the method chooses and
-  evaluates a new parameter vector generated by taking a fixed-size
-  step in the direction of the gradient for each parameter, and fits a
-  new linear model to all vectors evaluated so far, choosing the model
-  to minimize the sum of squared errors.
-
-\item \emph{SVM regression}: Similarly to linear regression, this
-  model starts with several random vectors, but this time they are
-  chosen in a small neighborhood about some initial random vector.
-  These vectors (generally 8) are evaluated, and a support vector
-  regression model is fit to the observed fitnesses.  To choose the
-  next vector for evaluation, we randomly generate some number
-  (typically 100) of vectors in the neighborhood of the best observed
-  gait, and select for evaluation the vector with the best predicted
-  performance.  We suspected that if we always chose the best
-  predicted point out of 100, we may end up progressing along a narrow
-  subspace, prohibiting learning of the true local fitness function.
-  Alternately, we would always choosing exploitation of knowledge
-  vs. exploration of the space.  To address this concern, we added a
-  parameter dubbed \code{bumpBy} that added noise to the final
-  selected point before it was evaluated.
-
-  Such a method naturally has many tunable parameters, and we
-  endeavored to select these parameteres by tuning the method in
-  simulation.  To estimate the performance of the algorithm, we ran it
-  agains a simulation with a known optimum.  The simulated function
-  was in the same five dimensional parameter space, and simply
-  returned a fitness determined as the height of a Gaussian with a
-  random mean.  The width of the Gaussian in each dimension was 20\%
-  of the range of each dimension, and the maximum value at the peak
-  was 100.  \figref{svm_sim_results} shows the learning results on
-  this simulated model using the ultimately selected SVM parameters.
-  Interestingly, a non-zero value of \code{bumpBy} resulted in better
-  learning than a zero valued, noise free (exploration free) learning.
-
-  Ultimately, however, the version of SVM tuned for simulation did not
-  yet show competitive performance on the real robot.  We tried tuning
-  some parameters on the real robot, but after some amount of tuning,
-  the method still exibited too little exploration and easily became
-  stuck in local minima.
-
-\item \emph{Evolutionary Neural Network (HyperNEAT)\cite{clune}}: Near
-  the end of term, we hacked together an interface between HyperNEAT
-  and the robot, requiring a slightly modified strategy interface.
-  Preliminary HyperNEAT runs were promising and resulted in serveral
-  interesting gaits.  Unfortunately, the gaits generated by HyperNEAT
-  also tended to stress the robot more than typical gaits had before,
-  and the servos would often overheat and malfunction, requiring
-  restart.  We think these issues may be addressed by adding a small
-  layer between the HyperNEAT strategy and the robot that disallows
-  quickly shifting commanded positions, and we hope to be able to test
-  these methods further once this filter is in place.
-
-\end{itemize}
-
-\figvarp{svm_sim_results}{.6}{Results for the SVM regression strategy
-  in simulation.  This simulation was used to tune the SVM strategy's
-  parameters before trying it on the physical robot.  The strategy
-  quickly approaches the maximum simulated fitness of 100.}{}
-
-\section{Related work}
-% Briefly explain who else worked on related problems in the past and what
-% methods they used. Explain if you are using similar methods, or if your
-% approach is different and if so - how (either is ok).
-
-Various maching learning techniques have proven to be useful in
-finding control policies for a wide variety of robots. Kohl and
-Stone\cite{kohl} presented a policy gradient reinforcement learning
-approach for generating a fast walk on legged robots. We experimented
-with this method to create a walk for our robot (Policy Gradient
-Descent). Chernova and Velosa\cite{chernova} took an evolutionary
-approach to this problem which we did not implement.  Zykov, Bongard,
-and Lipson\cite{zykov} describe the evolution of dynamic gaits on a
-physical robot requiring no prior assumptions about the locomotion
-pattern beyond the fact that it should be rhythmic.
-
-\section{System Architecture and Implementation}
-\seclabel{implement}
-% Describe how you implemented your system and how you structured it. 
-% This should give an overview of the system, not a detailed 
-% documentation of the code. The documentation of the code is part of 
-% the code you hand in. You might want to comment on high-level design 
-% decisions that you made. Also explain how you obtained your
-% data and any pre-processing you did to it.
-
-The quadruped robot has an on-board computer running Linux. The lower
-level drivers are in C and the system is implemented in
-Python. Feedback about distance travelled is provided via an infrared
-LED mounted on the robot and a Wii remote fixed to the ceiling.
-
-An overview of the code follows.
-
-\begin{itemize}
-
-\item \code{Robot} class: Class wrapper for commanding
-  motion of the robot.  The \code{Robot} class takes care of the robot
-  initialization, communication with the servos, and timing of the
-  runs.  In addition, it prevents the servos from ever being commanded
-  to a point outside their normal range (0 - 1023) as well as beyond
-  points where limbs would collide with parts of the robot body.  The
-  main class function, \code{run}, accepts a motion model (any
-  function that takes a time argument and outputs a 9 dimensional
-  position) and will run the robot using this motion model, including,
-  if desired, smooth interpolation over time for the beginning and end
-  of the run.
-
-\item \code{RunManager} class: Deals with all the details of running
-  the robot, including running the robot multiple times, tracking
-  distance walked via a \code{WiiTrackClient} member object, and
-  writing results to the log file. Also includes the
-  \code{explore\_dimensions} method to generate plots by varying each
-  parameter independently.
-
-\item \code{Strategy} class: The user has a choice between eight
-  different learning strategies: Random search, uniform random hill
-  climbing, Gaussian random hill climbing, N-dimensional policy
-  gradient descent, Nelder-Mead simplex, linear regression/prediction,
-  SVM regression/prediction, and HyperNEAT evolution.  Each strategy
-  must derive from the base \code{Strategy} class and must implement
-  two methods: \code{getNext}, for getting the next parameter vector
-  to try, and \code{updateResults}, for communicating results of a run
-  back to the strategy.
-
-\item \code{MotionModel} class: We implemented several motion models,
-  the main being the \code{SineModel5} class.  All motion models take
-  as input a parameter vector and produce as output a motion model,
-  which is simply a function mapping from time to nine motor
-  positions.  The \code{SineModel5} model commands positions based on
-  a sine wave, creating a periodic pattern. The parameters are:
-  amplitude, wavelength, a multiplier for the inner vs outer motors,
-  multiplier for left vs right motors, and multiplier for back vs
-  front motors.  Other similar models were tested, including a seven
-  parameter model which allowed sine waves shifted in time, but these
-  were not used as extensively in our experiments as
-  \code{SineModel5}.
-
-\item \code{WiiTrackClient} and \code{WiiTrackServer} classes and
-  hardware: A Wii remote tracks the location of the robot through an
-  infrared LED mounted on top of the robot. A \code{WiiTrackServer} is
-  run on the robot and continuously tracks its position using the
-  CWiid library\cite{cwiid} to interface with the remote via
-  bluetooth.  The \code{RunManager} then makes a
-  \code{WiiTrackClient}, which connects via socket to the tracking
-  server and requests position updates periodically.
-  \code{RunManager} currently gets the robot's position at the
-  beginning of each run and then again at the end and uses this to
-  calculate the net change in position.
-
-\end{itemize}
-
-\section{Experimental Evaluation}
-
-\subsection{Methodology}
-% What are the criteria you are using to evaluate your method? What
-% specific hypotheses does your experiment test? Describe the 
-% experimental methodology that you used. What are the dependent and 
-% independent variables? For projects in machine learning, what is 
-% the training/test data that was used, and why is it realistic or
-% interesting? Exactly what performance data did you collect and how 
-% are you presenting and analyzing it? Comparisons to competing methods 
-% that address the same problem or to variations of your own algorithm 
-% are particularly useful.
-
-The metric for evaluation of the designed gait is speed. We stop each 
-run after plateauing results (no improvement for one third of the 
-policies seen so far). The standard length of a run designates that it
-should be stopped after there is no improvement for one half of the policies
-seen so far, but since all runs took place on the actual robot, without use
-of a simulator, certain time limitations were imposed on the learning process.
-
-We controlled our experiments from a computer that was connected via a
-wireless ethernet to the robot. The robot collected data about
-distance walked automatically on its own. If it walked outside of the
-Wii remote's viewable area, it informed the user, so the only human
-intervention required during an experiment was to move the robot back
-inside the viewable area and to resume the run, which did not
-interrupt the learning process or result in the loss of data.
-
-We evaluated the efficacy of a set of parameters by sending these
-parameters to the robot and instructing it to walk for a certain
-length of time. The robot always began from the same position and
-returned to the starting position at the end of the run in order to
-measure true displacement without giving credit for ending in a leaned
-position. More efficient parameters resulted in a faster gait, which
-translated into a longer distance walked and a better score. After completing
-an evaluation, the robot sent the resulting distance walked back to the
-host computer and prepared itself for a new set of parameters to evaluate.
-
-Each algorithm was run on 3 different initial parameter vectors on the
-physical robot. We decided to evaluate all methods starting at the
-same three vector in order to allow for the fair comparison of each
-algorithm.  We evaluate each method based on the amount of improvement
-seen from the initial parameter vectors, and on the fastest speeds
-achieved during runs.
-
-The resulting gaits from our algorithms quickly outperformed the original
-hand-coded walk designed for this robot. The fastest walk, for example, was
-4 times faster.
-
-\subsection{Results}
-% Present the quantitative results of your experiments. Graphical data
-% presentation such as graphs and histograms are frequently better 
-% than tables. Explain what are the basic differences revealed in the 
-% data. If your method is stochastic or you analyze multiple datasets, 
-% calculate mean and standard error for every quantitative
-% performance result you show based on multiple runs.
-
-We have done complete runs of 3 different initial parameter vectors
-for random search, uniform random hill climbing, Gaussian random hill
-climbing, policy gradient descent, Nelder-Mead simplex, and linear
-regression.  We also evaluated the SVM regression and HyperNEAT
-methods, but these methods were more experimental, and thus we do not
-yet have the same volume of data for these runs.  We developed several
-gaits that were about 4 times faster than the original hand-coded
-gait.  Results are shown in \tabref{results_table}.
-
-\begin{table}
-\begin{center}
-\begin{tabular}{|r|c|c|c||c|}
-\hline
-                                         & A       & B      & C      &  Average \\
-\hline                                                               
-\hline                                                               
-Previous hand-coded gait                 & --      & --     & --     &  5.16 \\
-\hline                                                                 
-Random search                            & 6.04    & 17.26  & 4.90   &  9.40 \\
-\hline                                                                 
-Uniform Random Hill Climbing             & 11.37   & 9.44   & 2.69   &  7.83 \\
-\hline                                                                 
-Gaussian Random Hill Climbing            & 3.10    & 13.59  & 13.40  &  10.03 \\
-\hline                                                                 
-Policy Gradient Descent                  & 0.68    & 14.69  & 3.60   &  6.32 \\
-\hline                                                                 
-Nelder-Mead simplex                      & 8.51    & 13.62  & 14.83  &  12.32 \\
-\hline                                                                 
-Linear Regression                        & 27.58   & 12.51  & 1.95   &  14.01 \\
-%\hline                                                                
-%SVM Regression                           & ???     & ???    & ???   &   \\
-%\hline                                                                
-%Evolutionary Neural Network (HyperNEAT)  & ???     & ???    & ???   &   \\
-\hline
-\end{tabular}
-\caption{The best gaits found for each starting vector and algorithm,
-  in body lengths per minute.}
-\tablabel{results_table}
-\end{center}
-\end{table}
-
-\begin{itemize}
-
-\item For vector A, shown in \figref{vectorA}, linear regression
-  worked significantly better than the other algorithms, resulting in
-  a gait (27.58 body lengths/minute) that walked over twice as fast as
-  the next best gait -- uniform random hill climbing at 11.37 body
-  lengths/minute -- and 5.3 times better than the previous hand-coded
-  gait.
-
-\item Vector B, depicted in \figref{vectorB}, resulted in similarly
-  performing gaits with all the algorithms. The random method got
-  lucky and produced the best gait (17.26 body lengths/minute) and
-  uniform random hill climbing produced the worst (9.44 body
-  lengths/minute). The remaining algorithms each produced a gait
-  around 13 body lengths/minute.
-
-\item For vector C, shown in \figref{vectorC}, simplex and Gaussian
-  random hill climbing each produced a gait that substantially
-  outperformed the other algorithms. Simplex resulted in a gait of
-  nearly 15 body lengths/minute and Gaussian random hill climbing
-  produced a gait of just over 13 body lengths/minute, whereas the
-  other algorithms returned gaits of less than 5 body lengths/minute.
-
-\end{itemize}
-
-\figvarp{vectorA}{1}{Results for runs beginning with vector A.
-  Linear regression worked significantly better than the other
-  algorithms, resulting in a gait (27.58 body lengths/minute) that
-  walked over twice as fast as the next best gait -- uniform random
-  hill climbing at 11.37 body lengths/minute -- and 5.3 times better
-  than the previous hand-coded gait.}{}
-
-\figvarp{vectorB}{1}{Results for run starting with vector B.  All
-  algorithms performed similarly.  The random method actually got
-  lucky this time and produced the best gait (17.26 body
-  lengths/minute) and uniform random hill climbing produced the worst
-  (9.44 body lengths/minute). The remaining algorithms each produced a
-  gait around 13 body lengths/minute.}{}
-
-\figvarp{vectorC}{1}{Results for runs starting with vector C.
-  Simplex and Gaussian random hill climbing each produced a gait that
-  substantially outperformed the other algorithms, with around 15 body
-  lengths/minute and 13 body lengths/minute, respectively, whereas the
-  other algorithms returned gaits of less than 5 body
-  lengths/minute.}{}
-
-Based on the three trials, as shown in \figref{std_error}, linear
-regression worked the best, followed by simplex. Gaussian random hill
-climbing performed about as well as random, but it is unclear whether
-Gaussian would continue to improve and become much better than random
-search if allowed to run for more iterations. Uniform random hill
-climbing and policy gradient descent performed similarly to each other
-on average, but differed greatly on individual runs.
-
-\figvarp{std_error}{1}{Average results for each method starting with
-  parameter vectors A, B, and C.  Error bars are plotted showing the
-  standard error.  Linear regression outperformed other methods,
-  followed by simplex. Gaussian random hill climbing performed about
-  as well as random, but it is unclear whether Gaussian would continue
-  to improve and become much better than random search if allowed to
-  run for more iterations. Uniform random hill climbing and policy
-  gradient descent performed similarly to each other on average, but
-  differed greatly on individual runs.}{}
-
-In addition to running strategies to optimize parameterized gaits, we
-also wanted to investigate the space of possible gaits.  To accomplish
-this, we selected a parameter vector that resulted in motion, but not
-an exceptional gait, and plotted performance along each dimension
-individually.  In addition, we duplicated each measurement to be able
-to estimate the measurement noise at each point.  Results for this
-exploration are shown in \figref{explore_dim_1} through
-\figref{explore_dim_5}.
-
-\figvarp{explore_dim_1}{.5}{Motion for varying just dimension 1}{}
-
-\figvarp{explore_dim_2}{.5}{Motion for varying just dimension 2}{}
-
-\figvarp{explore_dim_3}{.5}{Motion for varying just dimension 3}{}
-
-\figvarp{explore_dim_4}{.5}{Motion for varying just dimension 4}{}
-
-\figvarp{explore_dim_5}{.5}{Motion for varying just dimension 5}{}
-
-\subsection{Discussion}
-% Is your hypothesis supported? What conclusions do the results support
-% about the strengths and weaknesses of your method compared to other 
-% methods? How can the results be explained in terms of the underlying 
-% properties of the algorithm and/or the data.
-
-Because only three trials were tested with each algorithm and no
-algorithm consistently outperformed the others, there is a large
-standard error for each algorithm, as show in \figref{std_error}. For
-this reason, it is unclear if any algorithm is superior to another in
-this application. More trials with each algorithm would be necessary
-to reach a definitive ranking of the performance of the algorithms.
-
-However, each algorithm discovered at least one gait of over 10 body
-lengths/minute, including random. For this reason, we speculate that
-the motion model is at least as critical as the learning algorithm
-used.
-
-\section{Future work}
-% What are the major shortcomings of your current method? For each 
-% shortcoming, propose additions or enhancements that would help overcome it.
-
-There are several directions in which we could continue our
-work. First, the error margins for our runs were large, so reducing
-them by running more trials would lead to more conclusive data. It
-would also be insightful to run our algorithms on different motion
-models.  We suspect that our choice of motion model influenced the
-results greatly, as even random choices in the space produced gaits
-that moved a significant fraction of the time (> 5\%).  It would be
-interesting to see how learning methods would perform using a model
-that included a much higher percentage of unproductive gaits. We also
-intend to experiment further with SVM regression and evolutionary
-algorithms/HyperNEAT.  Some parameter vectors resulted in the robot
-turning, as opposed to it walking long distances. This could be an
-interesting learning goal in future projects. To these ends, we
-propose the following additions and enhancements:
-
-\begin{itemize}
-\item More runs and/or longer runs
-\item Different motion representations
-\item Better tuning of SVM regression
-\item Evolutionary algorithms/HyperNEAT\cite{clune}
-\item Learning how to turn
-\end{itemize}
-
-\section{Conclusion}
-% Briefly summarize the important results and conclusions presented in 
-% the paper. What are the most important points illustrated by your 
-% work? How will your results improve future research and applications 
-% in the area?
-
-We have presented an array of approaches to optimizing a quadrupedal
-gait for forward speed.  We have implemented and tested different
-learning strategies, including uniform and Gaussian random hill
-climbing, policy gradient reinforcement learning, Nelder-Mead simplex,
-several new predictive methods based on linear and support vector
-regression, and an evolved neural network (HyperNEAT).  We have also
-compared these approaches to random search as a baseline. Many of the
-methods resulted in walks significantly faster than previously
-hand-tuned gaits.
-
-Because only three trials were tested with each algorithm and no
-algorithm consistently outperformed the others, there was a large
-standard error for each method, as shown in \figref{std_error}. Thus
-it was unclear if any algorithm was superior to another in this
-application. More trials with each algorithm would be necessary to
-reach a definitive ranking. Because each algorithm discovered at least one gait of
-over 10 body lengths/minute, including random search, we also
-conjectured that the motion representation for the robot is more
-integral to forward speed than the learning algorithm.  How to learn
-the motion representation, in addition to its parameters, remains an
-open problem.
-
-\section{Acknowledgments}
-% List any people not on the team who helped you with your project: Your
-% TA, other people you consulted with or had useful discussions (say in 
-% a word or two what they did), people who proofread your report, and 
-% any external code you used (libraries etc).
-\begin{itemize}
-\item Hod Lipson, Cornell Computational Synthesis Lab: adviser.
-\item Jim T\o rreson, University of Oslo: adviser.
-\item Juan Zagal, University of Chile: designed and printed robot and provided code for hand tuned gait.
-\item Jeff Clune, CCSL Red Couch: collaborated on HyperNEAT implementation/testing.
-\item Cooper Bills, Cornell University: assisted with Wii tracker development.
-\item Anshumali Srivastava, Cornell University: Teaching Assistant.
-\end{itemize}
+\edit{all this stuff has been moved to individual section files}
 

### 2011-01-20 paper/results.tex
Restructuring of report into multiple files to minimize conflicts, references working, dealt old sections to appropriate locations in paper framework.

--- 2010-12-07 report/report.tex
@@
-\section{Introduction}
+\section{Results and Discussion}
 
-% Motivate and abstractly describe the problem you are addressing and
-% how you are addressing it. What is the problem? Why is it important?
-% What is your basic approach? A short discussion of how it fits into
-% % related work in the area is also desirable. Summarize the basic
-% results and conclusions that you will present.
+\edit{write this.}
 
-Applications of walking robots often call for the ability to walk as
-quickly, efficiently, or with as little power as possible.  Often
-these optimizations are done manually by an expert who designs and
-tweaks a gait specifically for a given objective.  Other groups have
-used learning methods to generate gaits optimized for some metric.
-Approaches differ in their starting assumptions, some essentially
-tweaking the parameters of a hand-tuned model \cite{chernova}, others
-exploring a reasonably compact parameter space \cite{kohl}, and still
-others beginning with few assumptions besides periodicity
-\cite{zykov}.
-
-We aimed to strike a middle ground between these approaches.  Our
-motion generator did not rely on hand-tweaked gaits, but it did use
-parameterized gaits which, by their nature, impose some assumptions on
-the answers produced.  We then used machine learning to design gaits
-for a quadruped robot with these models.  This paper presents a
-comparison of the different learning methods implemented.  Most
-methods created walks that are several times faster than the original
-hand-tuned gait.  We invite readers with short attention spans to view
-a video of some of our results online here:
-
-\url{http://www.youtube.com/watch?v=ODoiOj9DdGg}
-
-%The code so far consists of the robot class, optimization class,
-%parameterized model, sine model, and camera feedback.
-
-\section{Problem definition}
-% Precisely define the problem you are addressing (i.e. formally specify
-% the inputs and outputs). Elaborate on why this is an interesting and
-% important problem.
-
-We are testing several different learning methods to design a
-parametrized gait for a quadruped robot from the Cornell Computational
-Synthesis Lab.
-
-The output each of the learning algorithms is a function of time,
-$f(t)$, that outputs a vector of commanded motor positions.  This
-function is generated using a parametrized motion model, described in
-\secref{motion}.
-
-The robot executes these commands and measures its change in location
-using the tracking system described in \secref{implement}.  The input
-to the learning algorithms is this measured displacement, which the
-algorithms attempt to maximize. This displacement is measured for each
-gait over a constant length run, usually 12 seconds.
-
-A comparison and evaluation of the many different methods available
-for optimizing the gait of legged robots will be useful for future
-work on this challenging multidimensional control problem.
-
-\section{Method}
-\seclabel{method}
-
-% Describe in reasonable detail the algorithm you are using to address
-% this problem. A pseudo-code description of the algorithm you are
-% using is frequently useful. If it makes sense for your project,
-% trace through a concrete example, showing how your algorithm
-% processes this example. The example should be complex enough to
-% illustrate all of the important aspects of the problem but simple
-% enough to be easily understood. If possible, an intuitively
-% meaningful example is better than one with meaningless symbols.
-
-We use several parameterized motion models that command motors to
-positions based on a sine wave, creating a periodic pattern.  While we
-investigated several models, for the bulk of our experiments, we used
-a model whose five parameters are: amplitude, wavelength, scale inner
-vs outer motors, scale left vs right motors, scale back vs front
-motors. Each strategy below attempts to choose the best possible
-parameters for this motion model.
-
-We implemented and tested 8 different learning strategies.  All
-strategies except for the HyperNEAT method\cite{clune} were
-constrained to pick parameters from within predetermined ranges.
-
-\begin{itemize}
-
-\item \emph{Random}: This method randomly generates parameter vectors
-  in the allowable range. This strategy is used only as baseline.
-
-\item \emph{Uniform random hill climbing}: This method begins by
-  selecting a single random parameter vector.  Subsequent iterations
-  generate a neighbor by randomly choosing one parameter to adjust and
-  replacing it with a new value chosen with uniform probability in the
-  allowable range for that parameter. The neighbor is evaluated by
-  running the robot with the newly chosen parameters. If this neighbor
-  results in a longer distance walked than the previous best gait, it
-  is saved as the new best gait. The process is then repeated, always
-  starting with the best gait.
-
-\item \emph{Gaussian random hill climbing}: This method works
-  similarly to Uniform random hill climbing, except neighbors are
-  generated by adding random Gaussian noise to the current best gait.
-  This results in all parameters being changed at once, but the
-  resulting vector is always fairly close to the previous best gait.
-  We used independently selected noise in each dimension, scaled such
-  that the standard deviation of the noise was 5\% of the range of
-  that dimension.
-
-\item \emph{N-dimensional policy gradient descent}: As opposed to the
-  previous methods, this method explicitly estimates the gradient for
-  the objective function. It does this by first evaluating \emph{t}
-  randomly generated parameter vectors near the initial vector, each
-  dimension of these vectors being perturbed by either $-\epsilon$,
-  $0$, or $\epsilon$. Then, for each dimension, it groups vectors into
-  three groups: $-\epsilon$, $0$, and $\epsilon$.  The gradient along
-  this dimension is then estimated as the average score for the
-  $\epsilon$ group minus the average score for the $-\epsilon$
-  group. Finally, the method creates a new vector by changing all
-  parameters by a fixed-size step in the direction of the gradient.
-
-\item \emph{Nelder-Mead simplex method}\cite{nm}: The Nelder-Mead
-  simplex method creates an initial simplex with 6 vertices. The
-  initial parameter vector is stored as the first vertex and the other
-  five vertices are created by adding to one dimension at a time one
-  tenth of the allowable range for that parameter. It then tests the
-  fitness of each vertex and based on these fitnesses, it reflects the
-  worst point over the centroid in an attempt to improve it.  However,
-  to prevent cycles and becoming stuck in local minima, several other
-  rules are used.  In general, the worst vertex is reflected over the
-  centroid. If the reflected point is better than the second worst
-  point and worse than the best point, then the reflected point
-  replaces the worst. If the reflected point is better than the best
-  point, the simplex is expanded in the direction of the reflected
-  point. The better of the reflected and the expanded point replaces
-  the worst point. If the reflected point is worse than the second
-  worst point, then the simplex is contracted away from the reflected
-  point. If the contracted point is better than the reflected point,
-  the contracted point replaces the worst point. If the contracted
-  point is worse than the reflected point, the entire simplex is
-  shrunk \cite{NMWebsite}.
-
-\item \emph{Linear regression}: To initialize, this method chooses and
-  evaluates five random parameter vectors. It then fits a linear model
-  from parameter vector to fitness. In a loop, the method chooses and
-  evaluates a new parameter vector generated by taking a fixed-size
-  step in the direction of the gradient for each parameter, and fits a
-  new linear model to all vectors evaluated so far, choosing the model
-  to minimize the sum of squared errors.
-
-\item \emph{SVM regression}: Similarly to linear regression, this
-  model starts with several random vectors, but this time they are
-  chosen in a small neighborhood about some initial random vector.
-  These vectors (generally 8) are evaluated, and a support vector
-  regression model is fit to the observed fitnesses.  To choose the
-  next vector for evaluation, we randomly generate some number
-  (typically 100) of vectors in the neighborhood of the best observed
-  gait, and select for evaluation the vector with the best predicted
-  performance.
-
-  \edit{Has many parameters... in simulation... we found that it's
-    better to bump points...}
-
-\item \emph{Evolutionary Neural Network (HyperNEAT)\cite{clune}}
-
-  \edit{do this}
-
-\end{itemize}
-
-\section{Related work}
-% Briefly explain who else worked on related problems in the past and what
-% methods they used. Explain if you are using similar methods, or if your
-% approach is different and if so - how (either is ok).
-
-Various maching learning techniques have proven to be useful in
-finding control policies for a wide variety of robots. Kohl and
-Stone\cite{kohl} presented a policy gradient reinforcement learning
-approach for generating a fast walk on legged robots. We experimented
-with this method to create a walk for our robot (Policy Gradient
-Descent). Chernova and Velosa\cite{chernova} took an evolutionary
-approach to this problem which we did not implement.  Zykov, Bongard,
-and Lipson\cite{zykov} describe the evolution of dynamic gaits on a
-physical robot requiring no prior assumptions about the locomotion
-pattern beyond the fact that it should be rhythmic.
-
-\section{System Architecture and Implementation}
-\seclabel{implement}
-% Describe how you implemented your system and how you structured it. 
-% This should give an overview of the system, not a detailed 
-% documentation of the code. The documentation of the code is part of 
-% the code you hand in. You might want to comment on high-level design 
-% decisions that you made. Also explain how you obtained your
-% data and any pre-processing you did to it.
-
-The quadruped robot has an on-board computer running Linux. The lower
-level drivers are in C and the system is implemented in
-Python. Feedback about distance travelled is provided via an infrared
-LED mounted on the robot and a Wii remote fixed to the ceiling.
-
-An overview of the code follows.
-
-\begin{itemize}
-
-\item \code{Robot} class: Class wrapper for commanding
-  motion of the robot.  The \code{Robot} class takes care of the robot
-  initialization, communication with the servos, and timing of the
-  runs.  In addition, it prevents the servos from ever being commanded
-  to a point outside their normal range (0 - 1023) as well as beyond
-  points where limbs would collide with parts of the robot body.  The
-  main class function, \code{run}, accepts a motion model (any
-  function that takes a time argument and outputs a 9 dimensional
-  position) and will run the robot using this motion model, including,
-  if desired, smooth interpolation over time for the beginning and end
-  of the run.
-
-\item \code{RunManager} class: Deals with all the details of running
-  the robot, including running the robot multiple times, tracking
-  distance walked via a \code{WiiTrackClient} member object, and
-  writing results to the log file. Also includes the
-  \code{explore\_dimensions} method to generate plots by varying each
-  parameter independently.
-
-\item \code{Strategy} class: The user has a choice between eight
-  different learning strategies: Random search, uniform random hill
-  climbing, Gaussian random hill climbing, N-dimensional policy
-  gradient descent, Nelder-Mead simplex, linear regression/prediction,
-  SVM regression/prediction, and HyperNEAT evolution.  Each strategy
-  must derive from the base \code{Strategy} class and must implement
-  two methods: \code{getNext}, for getting the next parameter vector
-  to try, and \code{updateResults}, for communicating results of a run
-  back to the strategy.
-
-\item \code{MotionModel} class: We implemented several motion models,
-  the main being the \code{SineModel5} class.  All motion models take
-  as input a parameter vector and produce as output a motion model,
-  which is simply a function mapping from time to nine motor
-  positions.  The \code{SineModel5} model commands positions based on
-  a sine wave, creating a periodic pattern. The parameters are:
-  amplitude, wavelength, a multiplier for the inner vs outer motors,
-  multiplier for left vs right motors, and multiplier for back vs
-  front motors.  Other similar models were tested, including a seven
-  parameter model which allowed sine waves shifted in time, but these
-  were not used as extensively in our experiments as
-  \code{SineModel5}.
-
-\item \code{WiiTrackClient} and \code{WiiTrackServer} classes and
-  hardware: A Wii remote tracks the location of the robot through an
-  infrared LED mounted on top of the robot. A \code{WiiTrackServer} is
-  run on the robot and continuously tracks its position using the
-  CWiid library\cite{cwiid} to interface with the remote via
-  bluetooth.  The \code{RunManager} then makes a
-  \code{WiiTrackClient}, which connects via socket to the tracking
-  server and requests position updates periodically.
-  \code{RunManager} currently gets the robot's position at the
-  beginning of each run and then again at the end and uses this to
-  calculate the net change in position.
-
-\end{itemize}
-
-\section{Experimental Evaluation}
-
-\subsection{Methodology}
-% What are the criteria you are using to evaluate your method? What
-% specific hypotheses does your experiment test? Describe the 
-% experimental methodology that you used. What are the dependent and 
-% independent variables? For projects in machine learning, what is 
-% the training/test data that was used, and why is it realistic or
-% interesting? Exactly what performance data did you collect and how 
-% are you presenting and analyzing it? Comparisons to competing methods 
-% that address the same problem or to variations of your own algorithm 
-% are particularly useful.
-
-The metric for evaluation of the designed gait is speed. We stop each 
-run after plateauing results (no improvement for one third of the 
-policies seen so far). The standard length of a run designates that it
-should be stopped after there is no improvement for one half of the policies
-seen so far, but since all runs took place on the actual robot, without use
-of a simulator, certain time limitations were imposed on the learning process.
-
-We controlled our experiments from a computer that was connected via a
-wireless ethernet to the robot. The robot collected data about
-distance walked automatically on its own. If it walked outside of the
-Wii remote's viewable area, it informed the user, so the only human
-intervention required during an experiment was to move the robot back
-inside the viewable area and to resume the run, which did not
-interrupt the learning process or result in the loss of data.
-
-We evaluated the efficacy of a set of parameters by sending these
-parameters to the robot and instructing it to walk for a certain
-length of time. The robot always began from the same position and
-returned to the starting position at the end of the run in order to
-measure true displacement without giving credit for ending in a leaned
-position. More efficient parameters resulted in a faster gait, which
-translated into a longer distance walked and a better score. After completing
-an evaluation, the robot sent the resulting distance walked back to the
-host computer and prepared itself for a new set of parameters to evaluate.
-
-Each algorithm was run on 3 different initial parameter vectors on the
-physical robot. We decided to evaluate all methods starting at the
-same three vector in order to allow for the fair comparison of each
-algorithm.  We evaluate each method based on the amount of improvement
-seen from the initial parameter vectors, and on the fastest speeds
-achieved during runs.
-
-The resulting gaits from our algorithms quickly outperformed the original
-hand-coded walk designed for this robot. The fastest walk, for example, was
-4 times faster.
-
-\subsection{Results}
 % Present the quantitative results of your experiments. Graphical data
@@
 exploration are shown in \figref{explore_dim_1} through
-\figref{explore_dim_5}.
+\figref{explore_dim_5}.  As is shown in the figures, some dimensions
+are smoother than others, and some measurements are fairly noisy.
 
-\figvarp{explore_dim_1}{.5}{Motion for varying just dimension 1}{}
-
-\figvarp{explore_dim_2}{.5}{Motion for varying just dimension 2}{}
-
-\figvarp{explore_dim_3}{.5}{Motion for varying just dimension 3}{}
-
-\figvarp{explore_dim_4}{.5}{Motion for varying just dimension 4}{}
-
-\figvarp{explore_dim_5}{.5}{Motion for varying just dimension 5}{}
-
-\edit{Jason: add SVM and HyperNEAT results here!}
+\figvarp{explore_dim_1}{.65}{Fitness mean and standard deviation
+  vs. dimension 1.  The circle is a common point in
+  \figref{explore_dim_1} through \figref{explore_dim_5}}{}
+\figvarp{explore_dim_2}{.65}{Fitness mean and standard deviation
+  vs. dimension 2.  The circle is a common point in
+  \figref{explore_dim_1} through \figref{explore_dim_5}}{}
+\figvarp{explore_dim_3}{.65}{Fitness mean and standard deviation
+  vs. dimension 3.  The circle is a common point in
+  \figref{explore_dim_1} through \figref{explore_dim_5}}{}
+\figvarp{explore_dim_4}{.65}{Fitness mean and standard deviation
+  vs. dimension 4.  The circle is a common point in
+  \figref{explore_dim_1} through \figref{explore_dim_5}}{}
+\figvarp{explore_dim_5}{.65}{Fitness mean and standard deviation
+  vs. dimension 5.  The circle is a common point in
+  \figref{explore_dim_1} through \figref{explore_dim_5}}{}
 
@@
 used.
-
-\section{Future work}
-% What are the major shortcomings of your current method? For each 
-% shortcoming, propose additions or enhancements that would help overcome it.
-
-There are several directions in which we could continue our
-work. First, the error margins for our runs were large, so reducing
-them by running more trials would lead to more conclusive data. It
-would also be insightful to run our algorithms on different motion
-models.  We suspect that our choice of motion model influenced the
-results greatly, as even random choices in the space produced gaits
-that moved a significant fraction of the time (> 5\%).  It would be
-interesting to see how learning methods would perform using a model
-that included a much higher percentage of unproductive gaits. We also
-intend to experiment more with evolutionary algorithms/HyperNEAT.
-Some parameter vectors resulted in the robot turning, as opposed to it
-walking long distances. This could be an interesting learning goal in
-future projects. To these ends, we propose the following additions and
-enhancements:
-
-\begin{itemize}
-\item More runs and/or longer runs
-\item Different motion representations
-\item Evolutionary algorithms/HyperNEAT\cite{clune}
-\item Learning how to turn
-\end{itemize}
-
-\section{Conclusion}
-% Briefly summarize the important results and conclusions presented in 
-% the paper. What are the most important points illustrated by your 
-% work? How will your results improve future research and applications 
-% in the area?
-
-We have presented an array of approaches to optimizing a quadrupedal
-gait for forward speed.  We have implemented and tested different
-learning strategies, including uniform and Gaussian random hill
-climbing, policy gradient reinforcement learning, Nelder-Mead simplex,
-several new predictive methods based on linear and support vector
-regression, and an evolved neural network (HyperNEAT).  We have also
-compared these approaches to random search as a baseline. Many of the
-methods resulted in walks significantly faster than previously
-hand-tuned gaits.
-
-Because only three trials were tested with each algorithm and no
-algorithm consistently outperformed the others, there was a large
-standard error for each method, as shown in \figref{std_error}. Thus
-it was unclear if any algorithm was superior to another in this
-application. More trials with each algorithm would be necessary to
-reach a definitive ranking. Because each algorithm discovered at least one gait of
-over 10 body lengths/minute, including random search, we also
-conjectured that the motion representation for the robot is more
-integral to forward speed than the learning algorithm.  How to learn
-the motion representation, in addition to its parameters, remains an
-open problem.
-
-\section{Acknowledgments}
-% List any people not on the team who helped you with your project: Your
-% TA, other people you consulted with or had useful discussions (say in 
-% a word or two what they did), people who proofread your report, and 
-% any external code you used (libraries etc).
-\begin{itemize}
-\item Hod Lipson, Cornell Computational Synthesis Lab: adviser.
-\item Jim T\o rreson, University of Oslo: adviser.
-\item Juan Zagal, University of Chile: designed and printed robot and provided code for hand tuned gait.
-\item Jeff Clune, CCSL Red Couch: collaborated on HyperNEAT implementation/testing.
-\item Cooper Bills, Cornell University: assisted with Wii tracker development.
-\item Anshumali Srivastava, Cornell University: Teaching Assistant.
-\end{itemize}
-
-\section{The Team}
-% A table listing the names of the people who worked on the project, and who
-% did what.
-
-\begin{itemize}
-
-\item Diana Hidalgo: 
-\begin{itemize}
-\item Simplex (Nelder-Mead)
-\item \code{WiiTrackClient} and \code{WiiTrackServer} classes and hardware
-\item Plotting and evaluating results
-\item Investigation into using Optitrack IR system for tracking robot
-\end{itemize}
-
-\item Sarah Nguyen
-\begin{itemize}
-\item Random search
-\item Uniform random hill climbing
-\item Gaussian random hill climbing
-\item N-dimensional policy gradient descent
-\item Linear regression and prediction
-\item \code{RunManager} and \code{Strategy} classes
-\item Video editing
-\end{itemize}
-
-\item Jason Yosinski
-\begin{itemize}
-\item SVM regression and prediction
-\item Evolutionary Neural Network (HyperNEAT)
-\item \code{WiiTrackClient} and \code{WiiTrackServer} classes and hardware
-\item \code{Robot}, \code{RunManager}, and \code{Strategy} classes
-\item \code{Parameterized Motion Model} abstract base class
-\item \code{SineModel} classes
-\item \code{explore\_dimensions}
-\item Plotting and evaluating results
-\end{itemize}
-
-\end{itemize}
-
-\section{Appendix}
-% Describe the code files you uploaded into CMS
-
-A brief description of the code uploaded to the CMS follows:
-
-\begin{itemize}
-\item \code{optimize.py}: Main program, determines a strategy to try and runs the robot
-  with that strategy.
-\item \code{RunManager.py}: Deals with all the details of running the robot,
-  including choosing an initial parameter, running the robot multiple times, 
-  tracking distance walked, and writing to the log file. Also includes the
-  \code{explore\_dimensions} method.
-\item \code{Strategy.py}: Contains all the different possible strategies, which
-  will be passed as objects in \code{optimize.py}. 
-\item \code{Robot.py}: Implements the \code{Robot} class, described in
-  \secref{implement}.
-\item \code{SineModel.py}: Implements a sine based motion model,
-  described in \secref{implement}.
-\item \code{Motion.py}: Motion helper functions.
-\item \code{WiiTrackServer.py}: Broadcasts the position of the infrared LED.
-\item \code{WiiTrackClient.py}: Connects to the WiiTrackServer to get the current position information.
-\end{itemize}
-
-% LocalWords:  vertices

### 2011-01-20 paper/approach.tex
Added terms, cleaned edit ugliness

@@
 
-\edit{write this.  Old quadratot System Architecture and Implementation section below.}
+\editbox{write this, old quadratot System Architecture and Implementation section below.}
 
@@
   climbing, Gaussian random hill climbing, N-dimensional policy
-  gradient descent, Nelder-Mead simplex, linear regression/prediction,
+  gradient descent, Nelder-Mead simplex, linear regression / prediction,
   SVM regression/prediction, and HyperNEAT evolution.  Each strategy
@@
 
-\edit{write this.  Old quadratot Experimental Evaluation section below.}
+\editbox{write this.  Old quadratot Experimental Evaluation section below.}
 

### 2011-01-20 paper/approach.tex
Fixed figure ugliness; figs now use acmFig and acmFigDouble

@@
 
-\figvarp{svm_sim_results}{.6}{Results for the SVM regression strategy
+\acmFig{svm_sim_results}{1}{Results for the SVM regression strategy
   in simulation.  This simulation was used to tune the SVM strategy's
   parameters before trying it on the physical robot.  The strategy
-  quickly approaches the maximum simulated fitness of 100.}{}
+  quickly approaches the maximum simulated fitness of 100.}
 

### 2011-01-20 paper/results.tex
Fixed figure ugliness; figs now use acmFig and acmFigDouble

@@
 
-\figvarp{vectorA}{1}{Results for runs beginning with vector A.
+\acmFig{vectorA}{1}{Results for runs beginning with vector A.
   Linear regression worked significantly better than the other
@@
   hill climbing at 11.37 body lengths/minute -- and 5.3 times better
-  than the previous hand-coded gait.}{}
+  than the previous hand-coded gait.}
 
-\figvarp{vectorB}{1}{Results for run starting with vector B.  All
+\acmFig{vectorB}{1}{Results for run starting with vector B.  All
   algorithms performed similarly.  The random method actually got
@@
   (9.44 body lengths/minute). The remaining algorithms each produced a
-  gait around 13 body lengths/minute.}{}
+  gait around 13 body lengths/minute.}
 
-\figvarp{vectorC}{1}{Results for runs starting with vector C.
+\acmFig{vectorC}{1}{Results for runs starting with vector C.
   Simplex and Gaussian random hill climbing each produced a gait that
@@
   other algorithms returned gaits of less than 5 body
-  lengths/minute.}{}
+  lengths/minute.}
 
@@
 
-\figvarp{std_error}{1}{Average results for each method starting with
+\acmFigDouble{std_error}{.6}{Average results for each method starting with
   parameter vectors A, B, and C.  Error bars are plotted showing the
@@
   gradient descent performed similarly to each other on average, but
-  differed greatly on individual runs.}{}
+  differed greatly on individual runs.}
 
@@
 
-\figvarp{explore_dim_1}{.65}{Fitness mean and standard deviation
+\acmFig{explore_dim_1}{1}{Fitness mean and standard deviation
   vs. dimension 1.  The circle is a common point in
-  \figref{explore_dim_1} through \figref{explore_dim_5}}{}
-\figvarp{explore_dim_2}{.65}{Fitness mean and standard deviation
+  \figref{explore_dim_1} through \figref{explore_dim_5}}
+\acmFig{explore_dim_2}{1}{Fitness mean and standard deviation
   vs. dimension 2.  The circle is a common point in
-  \figref{explore_dim_1} through \figref{explore_dim_5}}{}
-\figvarp{explore_dim_3}{.65}{Fitness mean and standard deviation
+  \figref{explore_dim_1} through \figref{explore_dim_5}}
+\acmFig{explore_dim_3}{1}{Fitness mean and standard deviation
   vs. dimension 3.  The circle is a common point in
-  \figref{explore_dim_1} through \figref{explore_dim_5}}{}
-\figvarp{explore_dim_4}{.65}{Fitness mean and standard deviation
+  \figref{explore_dim_1} through \figref{explore_dim_5}}
+\acmFig{explore_dim_4}{1}{Fitness mean and standard deviation
   vs. dimension 4.  The circle is a common point in
-  \figref{explore_dim_1} through \figref{explore_dim_5}}{}
-\figvarp{explore_dim_5}{.65}{Fitness mean and standard deviation
+  \figref{explore_dim_1} through \figref{explore_dim_5}}
+\acmFig{explore_dim_5}{1}{Fitness mean and standard deviation
   vs. dimension 5.  The circle is a common point in
-  \figref{explore_dim_1} through \figref{explore_dim_5}}{}
+  \figref{explore_dim_1} through \figref{explore_dim_5}}
 

### 2011-01-20 paper/problemDefinition.tex
working

@@
 work on this challenging multidimensional control problem.
+
+Given our physical robot, the gait learning problem aims to find a
+\emph{gait} that maximizes some performance metric.  Gaits may loosely
+be defined either with or without feedback from the robot.  Gaits
+without feedback simply specify the motor commands to be given at any
+time, and
+
+\edit{Jason writing this now}
+
+can best be
+formulated as follows: find functions

### 2011-01-20 paper/problemDefinition.tex
Wrote most of Problem Definition

--- 2011-01-20 paper/problemDefinition.tex
@@
 work on this challenging multidimensional control problem.
+
+\edit{Jason writing this now}
+
+The gait learning problem aims to find a \emph{gait} that maximizes
+some performance metric.
+
+Mathematically, we define a gait as a function that specifies
+commanded motor positions for a robot over time.  Using this
+definition, we require one function per motor.  We can write gaits
+without feedback --- also called feed forward gaits --- as
+
+\be
+x_i = f_i(t)
+\ee
+
+\noindent for commanded position $x_i$ for motor $i$.  The
+function depends only on time.  Using the same framework, a gait
+including measured motor position, force, or other values could be
+written as
+
+\be
+x_i = f_i(t, z)
+\ee
+
+\noindent for some measurement $z$.  Though perhaps obvious, it is
+worth noting that feedforward gaits are deterministic (unless they
+include random elements), producing the same command pattern each time
+they are run.  The output of a gait that includes feedback, on the
+other hand, depends on the actual measurements, and noise in these
+measurements may result in different $x_i$ values each time they are
+run.
+
+For the system evaluated in this paper, we chose to use only
+feedforward gaits.  An interesting extension would be to allow gaits
+to depend on the measured servo positions, loads, voltage drops, or
+other quantities.
+
+The ultimate goal was to design gaits that were as fast as possible,
+so to optimize for speed, our performance metric was displacement over
+a period of motion lasting approximately 10 seconds.  For details of
+exactly how this displacement was measured, see \secref{foo}.

### 2011-01-20 paper/approach.tex
Split approach into three files

--- 2011-01-20 paper/approach.tex
@@
 
-\subsection{Learning Methods}
+\input{learningMethods}
 
-\edit{write this.  Old quadratot Method section below}
+\input{platformDetails}
 
-%\section{Method}
-\seclabel{method}
-
-% Describe in reasonable detail the algorithm you are using to address
-% this problem. A pseudo-code description of the algorithm you are
-% using is frequently useful. If it makes sense for your project,
-% trace through a concrete example, showing how your algorithm
-% processes this example. The example should be complex enough to
-% illustrate all of the important aspects of the problem but simple
-% enough to be easily understood. If possible, an intuitively
-% meaningful example is better than one with meaningless symbols.
-
-We use several parameterized motion models that command motors to
-positions based on a sine wave, creating a periodic pattern.  While we
-investigated several models, for the bulk of our experiments, we used
-a model whose five parameters are: amplitude, wavelength, scale inner
-vs outer motors, scale left vs right motors, scale back vs front
-motors. Each strategy below attempts to choose the best possible
-parameters for this motion model.
-
-We implemented and tested 8 different learning strategies.  All
-strategies except for the HyperNEAT method\cite{clune} were
-constrained to pick parameters from within predetermined ranges.
-
-\begin{itemize}
-
-\item \emph{Random}: This method randomly generates parameter vectors
-  in the allowable range. This strategy is used only as baseline.
-
-\item \emph{Uniform random hill climbing}: This method begins by
-  selecting a single random parameter vector.  Subsequent iterations
-  generate a neighbor by randomly choosing one parameter to adjust and
-  replacing it with a new value chosen with uniform probability in the
-  allowable range for that parameter. The neighbor is evaluated by
-  running the robot with the newly chosen parameters. If this neighbor
-  results in a longer distance walked than the previous best gait, it
-  is saved as the new best gait. The process is then repeated, always
-  starting with the best gait.
-
-\item \emph{Gaussian random hill climbing}: This method works
-  similarly to Uniform random hill climbing, except neighbors are
-  generated by adding random Gaussian noise to the current best gait.
-  This results in all parameters being changed at once, but the
-  resulting vector is always fairly close to the previous best gait.
-  We used independently selected noise in each dimension, scaled such
-  that the standard deviation of the noise was 5\% of the range of
-  that dimension.
-
-\item \emph{N-dimensional policy gradient descent}: As opposed to the
-  previous methods, this method explicitly estimates the gradient for
-  the objective function. It does this by first evaluating \emph{t}
-  randomly generated parameter vectors near the initial vector, each
-  dimension of these vectors being perturbed by either $-\epsilon$,
-  $0$, or $\epsilon$. Then, for each dimension, it groups vectors into
-  three groups: $-\epsilon$, $0$, and $\epsilon$.  The gradient along
-  this dimension is then estimated as the average score for the
-  $\epsilon$ group minus the average score for the $-\epsilon$
-  group. Finally, the method creates a new vector by changing all
-  parameters by a fixed-size step in the direction of the gradient.
-
-\item \emph{Nelder-Mead simplex method}\cite{nm}: The Nelder-Mead
-  simplex method creates an initial simplex with 6 vertices. The
-  initial parameter vector is stored as the first vertex and the other
-  five vertices are created by adding to one dimension at a time one
-  tenth of the allowable range for that parameter. It then tests the
-  fitness of each vertex and based on these fitnesses, it reflects the
-  worst point over the centroid in an attempt to improve it.  However,
-  to prevent cycles and becoming stuck in local minima, several other
-  rules are used.  In general, the worst vertex is reflected over the
-  centroid. If the reflected point is better than the second worst
-  point and worse than the best point, then the reflected point
-  replaces the worst. If the reflected point is better than the best
-  point, the simplex is expanded in the direction of the reflected
-  point. The better of the reflected and the expanded point replaces
-  the worst point. If the reflected point is worse than the second
-  worst point, then the simplex is contracted away from the reflected
-  point. If the contracted point is better than the reflected point,
-  the contracted point replaces the worst point. If the contracted
-  point is worse than the reflected point, the entire simplex is
-  shrunk \cite{nm}.
-
-\item \emph{Linear regression}: To initialize, this method chooses and
-  evaluates five random parameter vectors. It then fits a linear model
-  from parameter vector to fitness. In a loop, the method chooses and
-  evaluates a new parameter vector generated by taking a fixed-size
-  step in the direction of the gradient for each parameter, and fits a
-  new linear model to all vectors evaluated so far, choosing the model
-  to minimize the sum of squared errors.
-
-\item \emph{SVM regression}: Similarly to linear regression, this
-  model starts with several random vectors, but this time they are
-  chosen in a small neighborhood about some initial random vector.
-  These vectors (generally 8) are evaluated, and a support vector
-  regression model is fit to the observed fitnesses.  To choose the
-  next vector for evaluation, we randomly generate some number
-  (typically 100) of vectors in the neighborhood of the best observed
-  gait, and select for evaluation the vector with the best predicted
-  performance.  We suspected that if we always chose the best
-  predicted point out of 100, we may end up progressing along a narrow
-  subspace, prohibiting learning of the true local fitness function.
-  Put another way, we would always choose exploitation of knowledge
-  vs. exploration of the space.  To address this concern, we added a
-  parameter dubbed \code{bumpBy} that added noise to the final
-  selected point before it was evaluated.
-
-  Such a method naturally has many tunable parameters, and we
-  endeavored to select these parameters by tuning the method in
-  simulation.  To estimate the performance of the algorithm, we ran it
-  against a simulation with a known optimum.  The simulated function
-  was in the same five dimensional parameter space, and simply
-  returned a fitness determined as the height of a Gaussian with a
-  random mean.  The width of the Gaussian in each dimension was 20\%
-  of the range of each dimension, and the maximum value at the peak
-  was 100.  \figref{svm_sim_results} shows the learning results on
-  this simulated model using the ultimately selected SVM parameters.
-  Interestingly, a non-zero value of \code{bumpBy} resulted in better
-  learning than noise free (exploration free) learning.
-
-  Ultimately, however, the version of SVM tuned for simulation still
-  did not show competitive performance on the real robot.  We tried
-  tuning some parameters on the real robot, but after some amount of
-  tuning, the method still exhibited too little exploration and easily
-  became stuck in local minima.
-
-\item \emph{Evolutionary Neural Network (HyperNEAT)\cite{clune}}: We 
-  put together an interface between HyperNEAT
-  -- an implementation of a method for evolving neural networks -- and
-  the robot, requiring a slightly modified strategy interface.
-
-  % Preliminary HyperNEAT runs were promising and resulted in several
-  % interesting gaits.
-
-  % Unfortunately, the gaits generated by HyperNEAT
-  % also tended to stress the robot more than typical gaits had before,
-  % and the servos would often overheat and malfunction, requiring
-  % restarts.  We think these issues may be addressed by adding a small
-  % layer between the HyperNEAT strategy and the robot that disallows
-  % quickly shifting commanded positions, and we hope to be able to test
-  % these methods further once this filter is in place.
-
-\end{itemize}
-
-\figvarp{svm_sim_results}{.6}{Results for the SVM regression strategy
-  in simulation.  This simulation was used to tune the SVM strategy's
-  parameters before trying it on the physical robot.  The strategy
-  quickly approaches the maximum simulated fitness of 100.}{}
-
-\subsection{Platform details}
-
-\edit{write this.  Old quadratot System Architecture and Implementation section below.}
-
-%\section{System Architecture and Implementation}
-\seclabel{implement}
-
-% Describe how you implemented your system and how you structured it. 
-% This should give an overview of the system, not a detailed 
-% documentation of the code. The documentation of the code is part of 
-% the code you hand in. You might want to comment on high-level design 
-% decisions that you made. Also explain how you obtained your
-% data and any pre-processing you did to it.
-
-The quadruped robot has an on-board computer running Linux. The lower
-level drivers are in C and the system is implemented in
-Python. Feedback about distance travelled is provided via an infrared
-LED mounted on the robot and a Wii remote fixed to the ceiling.
-
-An overview of the code follows.
-
-\begin{itemize}
-
-\item \code{Robot} class: Class wrapper for commanding
-  motion of the robot.  The \code{Robot} class takes care of the robot
-  initialization, communication with the servos, and timing of the
-  runs.  In addition, it prevents the servos from ever being commanded
-  to a point outside their normal range (0 - 1023) as well as beyond
-  points where limbs would collide with parts of the robot body.  The
-  main class function, \code{run}, accepts a motion model (any
-  function that takes a time argument and outputs a 9 dimensional
-  position) and will run the robot using this motion model, including,
-  if desired, smooth interpolation over time for the beginning and end
-  of the run.
-
-\item \code{RunManager} class: Deals with all the details of running
-  the robot, including running the robot multiple times, tracking
-  distance walked via a \code{WiiTrackClient} member object, and
-  writing results to the log file. Also includes the
-  \code{explore\_dimensions} method to generate plots by varying each
-  parameter independently.
-
-\item \code{Strategy} class: The user has a choice between eight
-  different learning strategies: Random search, uniform random hill
-  climbing, Gaussian random hill climbing, N-dimensional policy
-  gradient descent, Nelder-Mead simplex, linear regression/prediction,
-  SVM regression/prediction, and HyperNEAT evolution.  Each strategy
-  must derive from the base \code{Strategy} class and must implement
-  two methods: \code{getNext}, for getting the next parameter vector
-  to try, and \code{updateResults}, for communicating results of a run
-  back to the strategy.
-
-\item \code{MotionModel} class: We implemented several motion models,
-  the main being the \code{SineModel5} class.  All motion models take
-  as input a parameter vector and produce as output a motion model,
-  which is simply a function mapping from time to nine motor
-  positions.  The \code{SineModel5} model commands positions based on
-  a sine wave, creating a periodic pattern. The parameters are:
-  amplitude, wavelength, a multiplier for the inner vs outer motors,
-  multiplier for left vs right motors, and multiplier for back vs
-  front motors.  Other similar models were tested, including a seven
-  parameter model which allowed sine waves shifted in time, but these
-  were not used as extensively in our experiments as
-  \code{SineModel5}.
-
-\item \code{WiiTrackClient} and \code{WiiTrackServer} classes and
-  hardware: A Wii remote tracks the location of the robot through an
-  infrared LED mounted on top of the robot. A \code{WiiTrackServer} is
-  run on the robot and continuously tracks its position using the
-  CWiid library\cite{cwiid} to interface with the remote via
-  bluetooth.  The \code{RunManager} then makes a
-  \code{WiiTrackClient}, which connects via socket to the tracking
-  server and requests position updates periodically.
-  \code{RunManager} currently gets the robot's position at the
-  beginning of each run and then again at the end and uses this to
-  calculate the net change in position.
-
-\end{itemize}
-
-\subsection{Fitness evaluation details}
-
-\edit{write this.  Old quadratot Experimental Evaluation section below.}
-
-%\section{Experimental Evaluation}
-
-%\subsection{Methodology}
-% What are the criteria you are using to evaluate your method? What
-% specific hypotheses does your experiment test? Describe the 
-% experimental methodology that you used. What are the dependent and 
-% independent variables? For projects in machine learning, what is 
-% the training/test data that was used, and why is it realistic or
-% interesting? Exactly what performance data did you collect and how 
-% are you presenting and analyzing it? Comparisons to competing methods 
-% that address the same problem or to variations of your own algorithm 
-% are particularly useful.
-
-The metric for evaluation of the designed gait is speed. We stop each 
-run after plateauing results (no improvement for one third of the 
-policies seen so far). The standard length of a run designates that it
-should be stopped after there is no improvement for one half of the policies
-seen so far, but since all runs took place on the actual robot, without use
-of a simulator, certain time limitations were imposed on the learning process.
-
-We controlled our experiments from a computer that was connected via a
-wireless ethernet to the robot. The robot collected data about
-distance walked automatically on its own. If it walked outside of the
-Wii remote's viewable area, it informed the user, so the only human
-intervention required during an experiment was to move the robot back
-inside the viewable area and to resume the run, which did not
-interrupt the learning process or result in the loss of data.
-
-We evaluated the efficacy of a set of parameters by sending these
-parameters to the robot and instructing it to walk for a certain
-length of time. The robot always began from the same position and
-returned to the starting position at the end of the run in order to
-measure true displacement without giving credit for ending in a leaned
-position. More efficient parameters resulted in a faster gait, which
-translated into a longer distance walked and a better score. After completing
-an evaluation, the robot sent the resulting distance walked back to the
-host computer and prepared itself for a new set of parameters to evaluate.
-
-Each algorithm was run on 3 different initial parameter vectors on the
-physical robot. We decided to evaluate all methods starting at the
-same three vector in order to allow for the fair comparison of each
-algorithm.  We evaluate each method based on the amount of improvement
-seen from the initial parameter vectors, and on the fastest speeds
-achieved during runs.
-
-The resulting gaits from our algorithms quickly outperformed the original
-hand-coded walk designed for this robot. The fastest walk, for example, was
-4 times faster.
+\input{fitnessEvaluation}

### 2011-01-20 paper/fitnessEvaluation.tex
Split approach into three files

--- 2010-12-07 report/report.tex
@@
-\section{Introduction}
+\subsection{Fitness evaluation details}
 
-% Motivate and abstractly describe the problem you are addressing and
-% how you are addressing it. What is the problem? Why is it important?
-% What is your basic approach? A short discussion of how it fits into
-% % related work in the area is also desirable. Summarize the basic
-% results and conclusions that you will present.
+\editbox{write this.  Old quadratot Experimental Evaluation section below.}
 
-Applications of walking robots often call for the ability to walk as
-quickly, efficiently, or with as little power as possible.  Often
-these optimizations are done manually by an expert who designs and
-tweaks a gait specifically for a given objective.  Other groups have
-used learning methods to generate gaits optimized for some metric.
-Approaches differ in their starting assumptions, some essentially
-tweaking the parameters of a hand-tuned model \cite{chernova}, others
-exploring a reasonably compact parameter space \cite{kohl}, and still
-others beginning with few assumptions besides periodicity
-\cite{zykov}.
+%\section{Experimental Evaluation}
 
-We aimed to strike a middle ground between these approaches.  Our
-motion generator did not rely on hand-tweaked gaits, but it did use
-parameterized gaits which, by their nature, impose some assumptions on
-the answers produced.  We then used machine learning to design gaits
-for a quadruped robot with these models.  This paper presents a
-comparison of the different learning methods implemented.  Most
-methods created walks that are several times faster than the original
-hand-tuned gait.  We invite readers with short attention spans to view
-a video of some of our results online here:
-
-\url{http://www.youtube.com/watch?v=ODoiOj9DdGg}
-
-%The code so far consists of the robot class, optimization class,
-%parameterized model, sine model, and camera feedback.
-
-\edit{proofread}
-
-\section{Problem definition}
-% Precisely define the problem you are addressing (i.e. formally specify
-% the inputs and outputs). Elaborate on why this is an interesting and
-% important problem.
-
-We are testing several different learning methods to design a
-parametrized gait for a quadruped robot from the Cornell Computational
-Synthesis Lab.
-
-The output each of the learning algorithms is a function of time,
-$f(t)$, that outputs a vector of commanded motor positions.  This
-function is generated using a parametrized motion model, described in
-\secref{motion}.
-
-The robot executes these commands and measures its change in location
-using the tracking system described in \secref{implement}.  The input
-to the learning algorithms is this measured displacement, which the
-algorithms attempt to maximize. This displacement is measured for each
-gait over a constant length run, usually 12 seconds.
-
-A comparison and evaluation of the many different methods available
-for optimizing the gait of legged robots will be useful for future
-work on this challenging multidimensional control problem.
-
-\section{Method}
-\seclabel{method}
-
-% Describe in reasonable detail the algorithm you are using to address
-% this problem. A pseudo-code description of the algorithm you are
-% using is frequently useful. If it makes sense for your project,
-% trace through a concrete example, showing how your algorithm
-% processes this example. The example should be complex enough to
-% illustrate all of the important aspects of the problem but simple
-% enough to be easily understood. If possible, an intuitively
-% meaningful example is better than one with meaningless symbols.
-
-We use several parameterized motion models that command motors to
-positions based on a sine wave, creating a periodic pattern.  While we
-investigated several models, for the bulk of our experiments, we used
-a model whose five parameters are: amplitude, wavelength, scale inner
-vs outer motors, scale left vs right motors, scale back vs front
-motors. Each strategy below attempts to choose the best possible
-parameters for this motion model.
-
-We implemented and tested 8 different learning strategies.  All
-strategies except for the HyperNEAT method\cite{clune} were
-constrained to pick parameters from within predetermined ranges.
-
-\begin{itemize}
-
-\item \emph{Random}: This method randomly generates parameter vectors
-  in the allowable range. This strategy is used only as baseline.
-
-\item \emph{Uniform random hill climbing}: This method begins by
-  selecting a single random parameter vector.  Subsequent iterations
-  generate a neighbor by randomly choosing one parameter to adjust and
-  replacing it with a new value chosen with uniform probability in the
-  allowable range for that parameter. The neighbor is evaluated by
-  running the robot with the newly chosen parameters. If this neighbor
-  results in a longer distance walked than the previous best gait, it
-  is saved as the new best gait. The process is then repeated, always
-  starting with the best gait.
-
-\item \emph{Gaussian random hill climbing}: This method works
-  similarly to Uniform random hill climbing, except neighbors are
-  generated by adding random Gaussian noise to the current best gait.
-  This results in all parameters being changed at once, but the
-  resulting vector is always fairly close to the previous best gait.
-  We used independently selected noise in each dimension, scaled such
-  that the standard deviation of the noise was 5\% of the range of
-  that dimension.
-
-\item \emph{N-dimensional policy gradient descent}: As opposed to the
-  previous methods, this method explicitly estimates the gradient for
-  the objective function. It does this by first evaluating \emph{t}
-  randomly generated parameter vectors near the initial vector, each
-  dimension of these vectors being perturbed by either $-\epsilon$,
-  $0$, or $\epsilon$. Then, for each dimension, it groups vectors into
-  three groups: $-\epsilon$, $0$, and $\epsilon$.  The gradient along
-  this dimension is then estimated as the average score for the
-  $\epsilon$ group minus the average score for the $-\epsilon$
-  group. Finally, the method creates a new vector by changing all
-  parameters by a fixed-size step in the direction of the gradient.
-
-\item \emph{Nelder-Mead simplex method}\cite{nm}: The Nelder-Mead
-  simplex method creates an initial simplex with 6 vertices. The
-  initial parameter vector is stored as the first vertex and the other
-  five vertices are created by adding to one dimension at a time one
-  tenth of the allowable range for that parameter. It then tests the
-  fitness of each vertex and based on these fitnesses, it reflects the
-  worst point over the centroid in an attempt to improve it.  However,
-  to prevent cycles and becoming stuck in local minima, several other
-  rules are used.  In general, the worst vertex is reflected over the
-  centroid. If the reflected point is better than the second worst
-  point and worse than the best point, then the reflected point
-  replaces the worst. If the reflected point is better than the best
-  point, the simplex is expanded in the direction of the reflected
-  point. The better of the reflected and the expanded point replaces
-  the worst point. If the reflected point is worse than the second
-  worst point, then the simplex is contracted away from the reflected
-  point. If the contracted point is better than the reflected point,
-  the contracted point replaces the worst point. If the contracted
-  point is worse than the reflected point, the entire simplex is
-  shrunk \cite{NMWebsite}.
-
-\item \emph{Linear regression}: To initialize, this method chooses and
-  evaluates five random parameter vectors. It then fits a linear model
-  from parameter vector to fitness. In a loop, the method chooses and
-  evaluates a new parameter vector generated by taking a fixed-size
-  step in the direction of the gradient for each parameter, and fits a
-  new linear model to all vectors evaluated so far, choosing the model
-  to minimize the sum of squared errors.
-
-\item \emph{SVM regression}: Similarly to linear regression, this
-  model starts with several random vectors, but this time they are
-  chosen in a small neighborhood about some initial random vector.
-  These vectors (generally 8) are evaluated, and a support vector
-  regression model is fit to the observed fitnesses.  To choose the
-  next vector for evaluation, we randomly generate some number
-  (typically 100) of vectors in the neighborhood of the best observed
-  gait, and select for evaluation the vector with the best predicted
-  performance.
-
-  \edit{Has many parameters... in simulation... we found that it's
-    better to bump points...}
-
-\item \emph{Evolutionary Neural Network (HyperNEAT)\cite{clune}}
-
-  \edit{do this}
-
-\end{itemize}
-
-\section{Related work}
-% Briefly explain who else worked on related problems in the past and what
-% methods they used. Explain if you are using similar methods, or if your
-% approach is different and if so - how (either is ok).
-
-Various maching learning techniques have proven to be useful in
-finding control policies for a wide variety of robots. Kohl and
-Stone\cite{kohl} presented a policy gradient reinforcement learning
-approach for generating a fast walk on legged robots. We experimented
-with this method to create a walk for our robot (Policy Gradient
-Descent). Chernova and Velosa\cite{chernova} took an evolutionary
-approach to this problem which we did not implement.  Zykov, Bongard,
-and Lipson\cite{zykov} describe the evolution of dynamic gaits on a
-physical robot requiring no prior assumptions about the locomotion
-pattern beyond the fact that it should be rhythmic.
-
-\section{System Architecture and Implementation}
-\seclabel{implement}
-% Describe how you implemented your system and how you structured it. 
-% This should give an overview of the system, not a detailed 
-% documentation of the code. The documentation of the code is part of 
-% the code you hand in. You might want to comment on high-level design 
-% decisions that you made. Also explain how you obtained your
-% data and any pre-processing you did to it.
-
-The quadruped robot has an on-board computer running Linux. The lower
-level drivers are in C and the system is implemented in
-Python. Feedback about distance travelled is provided via an infrared
-LED mounted on the robot and a Wii remote fixed to the ceiling.
-
-An overview of the code follows.
-
-\begin{itemize}
-
-\item \code{Robot} class: Class wrapper for commanding
-  motion of the robot.  The \code{Robot} class takes care of the robot
-  initialization, communication with the servos, and timing of the
-  runs.  In addition, it prevents the servos from ever being commanded
-  to a point outside their normal range (0 - 1023) as well as beyond
-  points where limbs would collide with parts of the robot body.  The
-  main class function, \code{run}, accepts a motion model (any
-  function that takes a time argument and outputs a 9 dimensional
-  position) and will run the robot using this motion model, including,
-  if desired, smooth interpolation over time for the beginning and end
-  of the run.
-
-\item \code{RunManager} class: Deals with all the details of running
-  the robot, including running the robot multiple times, tracking
-  distance walked via a \code{WiiTrackClient} member object, and
-  writing results to the log file. Also includes the
-  \code{explore\_dimensions} method to generate plots by varying each
-  parameter independently.
-
-\item \code{Strategy} class: The user has a choice between eight
-  different learning strategies: Random search, uniform random hill
-  climbing, Gaussian random hill climbing, N-dimensional policy
-  gradient descent, Nelder-Mead simplex, linear regression/prediction,
-  SVM regression/prediction, and HyperNEAT evolution.  Each strategy
-  must derive from the base \code{Strategy} class and must implement
-  two methods: \code{getNext}, for getting the next parameter vector
-  to try, and \code{updateResults}, for communicating results of a run
-  back to the strategy.
-
-\item \code{MotionModel} class: We implemented several motion models,
-  the main being the \code{SineModel5} class.  All motion models take
-  as input a parameter vector and produce as output a motion model,
-  which is simply a function mapping from time to nine motor
-  positions.  The \code{SineModel5} model commands positions based on
-  a sine wave, creating a periodic pattern. The parameters are:
-  amplitude, wavelength, a multiplier for the inner vs outer motors,
-  multiplier for left vs right motors, and multiplier for back vs
-  front motors.  Other similar models were tested, including a seven
-  parameter model which allowed sine waves shifted in time, but these
-  were not used as extensively in our experiments as
-  \code{SineModel5}.
-
-\item \code{WiiTrackClient} and \code{WiiTrackServer} classes and
-  hardware: A Wii remote tracks the location of the robot through an
-  infrared LED mounted on top of the robot. A \code{WiiTrackServer} is
-  run on the robot and continuously tracks its position using the
-  CWiid library\cite{cwiid} to interface with the remote via
-  bluetooth.  The \code{RunManager} then makes a
-  \code{WiiTrackClient}, which connects via socket to the tracking
-  server and requests position updates periodically.
-  \code{RunManager} currently gets the robot's position at the
-  beginning of each run and then again at the end and uses this to
-  calculate the net change in position.
-
-\end{itemize}
-
-\section{Experimental Evaluation}
-
-\subsection{Methodology}
+%\subsection{Methodology}
 % What are the criteria you are using to evaluate your method? What
@@
 4 times faster.
-
-\subsection{Results}
-% Present the quantitative results of your experiments. Graphical data
-% presentation such as graphs and histograms are frequently better 
-% than tables. Explain what are the basic differences revealed in the 
-% data. If your method is stochastic or you analyze multiple datasets, 
-% calculate mean and standard error for every quantitative
-% performance result you show based on multiple runs.
-
-We have done complete runs of 3 different initial parameter vectors
-for random search, uniform random hill climbing, Gaussian random hill
-climbing, policy gradient descent, Nelder-Mead simplex, and linear
-regression.  We also evaluated the SVM regression and HyperNEAT
-methods, but these methods were more experimental, and thus we do not
-yet have the same volume of data for these runs.  We developed several
-gaits that were about 4 times faster than the original hand-coded
-gait.
-
-\edit{Convert to Table}
-
-\begin{tabular}{|r|c|c|c||c|}
-\hline
-                                         & A       & B      & C      &  Average \\
-\hline                                                               
-\hline                                                               
-Previous hand-coded gait                 & --      & --     & --     &  5.16 \\
-\hline                                                                 
-Random search                            & 6.04    & 17.26  & 4.90   &  9.40 \\
-\hline                                                                 
-Uniform Random Hill Climbing             & 11.37   & 9.44   & 2.69   &  7.83 \\
-\hline                                                                 
-Gaussian Random Hill Climbing            & 3.10    & 13.59  & 13.40  &  10.03 \\
-\hline                                                                 
-Policy Gradient Descent                  & 0.68    & 14.69  & 3.60   &  6.32 \\
-\hline                                                                 
-Nelder-Mead simplex                      & 8.51    & 13.62  & 14.83  &  12.32 \\
-\hline                                                                 
-Linear Regression                        & 27.58   & 12.51  & 1.95   &  14.01 \\
-%\hline                                                                
-%SVM Regression                           & ???     & ???    & ???   &   \\
-%\hline                                                                
-%Evolutionary Neural Network (HyperNEAT)  & ???     & ???    & ???   &   \\
-\hline
-\end{tabular}
-
-\edit{Caption: The best gaits found for each starting vector and
-  algorithm, in body lengths per minute.}
-
-\begin{itemize}
-
-\item For vector A, shown in \figref{vectorA}, linear regression
-  worked significantly better than the other algorithms, resulting in
-  a gait (27.58 body lengths/minute) that walked over twice as fast as
-  the next best gait -- uniform random hill climbing at 11.37 body
-  lengths/minute -- and 5.3 times better than the previous hand-coded
-  gait.
-
-\item Vector B, depicted in \figref{vectorB}, resulted in similarly
-  performing gaits with all the algorithms. The random method got
-  lucky and produced the best gait (17.26 body lengths/minute) and
-  uniform random hill climbing produced the worst (9.44 body
-  lengths/minute). The remaining algorithms each produced a gait
-  around 13 body lengths/minute.
-
-\item For vector C, shown in \figref{vectorC}, simplex and Gaussian
-  random hill climbing each produced a gait that substantially
-  outperformed the other algorithms. Simplex resulted in a gait of
-  nearly 15 body lengths/minute and Gaussian random hill climbing
-  produced a gait of just over 13 body lengths/minute, whereas the
-  other algorithms returned gaits of less than 5 body lengths/minute.
-
-\end{itemize}
-
-\figvarp{vectorA}{1}{Results for runs beginning with vector A.
-  Linear regression worked significantly better than the other
-  algorithms, resulting in a gait (27.58 body lengths/minute) that
-  walked over twice as fast as the next best gait -- uniform random
-  hill climbing at 11.37 body lengths/minute -- and 5.3 times better
-  than the previous hand-coded gait.}{}
-
-\figvarp{vectorB}{1}{Results for run starting with vector B.  All
-  algorithms performed similarly.  The random method actually got
-  lucky this time and produced the best gait (17.26 body
-  lengths/minute) and uniform random hill climbing produced the worst
-  (9.44 body lengths/minute). The remaining algorithms each produced a
-  gait around 13 body lengths/minute.}{}
-
-\figvarp{vectorC}{1}{Results for runs starting with vector C.
-  Simplex and Gaussian random hill climbing each produced a gait that
-  substantially outperformed the other algorithms, with around 15 body
-  lengths/minute and 13 body lengths/minute, respectively, whereas the
-  other algorithms returned gaits of less than 5 body
-  lengths/minute.}{}
-
-Based on the three trials, as shown in \figref{std_error}, linear
-regression worked the best, followed by simplex. Gaussian random hill
-climbing performed about as well as random, but it is unclear whether
-Gaussian would continue to improve and become much better than random
-search if allowed to run for more iterations. Uniform random hill
-climbing and policy gradient descent performed similarly to each other
-on average, but differed greatly on individual runs.
-
-\figvarp{std_error}{1}{Average results for each method starting with
-  parameter vectors A, B, and C.  Error bars are plotted showing the
-  standard error.  Linear regression outperformed other methods,
-  followed by simplex. Gaussian random hill climbing performed about
-  as well as random, but it is unclear whether Gaussian would continue
-  to improve and become much better than random search if allowed to
-  run for more iterations. Uniform random hill climbing and policy
-  gradient descent performed similarly to each other on average, but
-  differed greatly on individual runs.}{}
-
-In addition to running strategies to optimize parameterized gaits, we
-also ......
-
-We also coded \code{explore\_dimensions} to collect data for dimension
-varying plots. We intend to do further explore this by running
-\code{explore\_dimension} around a parameter that produces a medium-good gait.
-
-\edit{Jason: add exploration runs}
-
-\edit{Jason: add SVM and HyperNEAT results here!}
-
-\subsection{Discussion}
-% Is your hypothesis supported? What conclusions do the results support
-% about the strengths and weaknesses of your method compared to other 
-% methods? How can the results be explained in terms of the underlying 
-% properties of the algorithm and/or the data.
-
-Because only three trials were tested with each algorithm and no
-algorithm consistently outperformed the others, there is a large
-standard error for each algorithm, as show in \figref{std_error}. For
-this reason, it is unclear if any algorithm is superior to another in
-this application. More trials with each algorithm would be necessary
-to reach a definitive ranking of the performance of the algorithms.
-
-However, each algorithm discovered at least one gait of over 10 body
-lengths/minute, including random. For this reason, we speculate that
-the motion model is at least as critical as the learning algorithm
-used.
-
-\section{Future work}
-% What are the major shortcomings of your current method? For each 
-% shortcoming, propose additions or enhancements that would help overcome it.
-
-There are several directions in which we could continue our
-work. First, the error margins for our runs were large, so reducing
-them by running more trials would lead to more conclusive data. It
-would also be insightful to run our algorithms on different motion
-models.  We suspect that our choice of motion model influenced the
-results greatly, as even random choices in the space produced gaits
-that moved a significant fraction of the time (> 5\%).  It would be
-interesting to see how learning methods would perform using a model
-that included a much higher percentage of unproductive gaits. We also
-intend to experiment more with evolutionary algorithms/HyperNEAT.
-Some parameter vectors resulted in the robot turning, as opposed to it
-walking long distances. This could be an interesting learning goal in
-future projects. To these ends, we propose the following additions and
-enhancements:
-
-\begin{itemize}
-\item More runs and/or longer runs
-\item Different motion representations
-\item Evolutionary algorithms/HyperNEAT\cite{clune}
-\item Learning how to turn
-\end{itemize}
-
-\section{Conclusion}
-% Briefly summarize the important results and conclusions presented in 
-% the paper. What are the most important points illustrated by your 
-% work? How will your results improve future research and applications 
-% in the area?
-
-We have presented an array of approaches to optimizing a quadrupedal
-gait for forward speed.  We have implemented and tested different
-learning strategies, including uniform and Gaussian random hill
-climbing, policy gradient reinforcement learning, Nelder-Mead simplex,
-several new predictive methods based on linear and support vector
-regression, and an evolved neural network (HyperNEAT).  We have also
-compared these approaches to random search as a baseline. Many of the
-methods resulted in walks significantly faster than previously
-hand-tuned gaits.
-
-Because only three trials were tested with each algorithm and no
-algorithm consistently outperformed the others, there was a large
-standard error for each method, as shown in \figref{std_error}. Thus
-it was unclear if any algorithm was superior to another in this
-application. More trials with each algorithm would be necessary to
-reach a definitive ranking. Because each algorithm discovered at least one gait of
-over 10 body lengths/minute, including random search, we also
-conjectured that the motion representation for the robot is more
-integral to forward speed than the learning algorithm.  How to learn
-the motion representation, in addition to its parameters, remains an
-open problem.
-
-\section{Acknowledgments}
-% List any people not on the team who helped you with your project: Your
-% TA, other people you consulted with or had useful discussions (say in 
-% a word or two what they did), people who proofread your report, and 
-% any external code you used (libraries etc).
-\begin{itemize}
-\item Hod Lipson, Cornell Computational Synthesis Lab: adviser.
-\item Jim T\o rreson, University of Oslo: adviser.
-\item Juan Zagal, University of Chile: designed and printed robot and provided code for hand tuned gait.
-\item Jeff Clune, CCSL Red Couch: collaborated on HyperNEAT implementation/testing.
-\item Cooper Bills, Cornell University: assisted with Wii tracker development.
-\item Anshumali Srivastava, Cornell University: Teaching Assistant.
-\end{itemize}
-
-\section{The Team}
-% A table listing the names of the people who worked on the project, and who
-% did what.
-
-\begin{itemize}
-
-\item Diana Hidalgo: 
-\begin{itemize}
-\item Simplex (Nelder-Mead)
-\item \code{WiiTrackClient} and \code{WiiTrackServer} classes and hardware
-\item Plotting and evaluating results
-\item Investigation into using Optitrack IR system for tracking robot
-\end{itemize}
-
-\item Sarah Nguyen
-\begin{itemize}
-\item Random search
-\item Uniform random hill climbing
-\item Gaussian random hill climbing
-\item N-dimensional policy gradient descent
-\item Linear regression and prediction
-\item \code{RunManager} and \code{Strategy} classes
-\item Video editing
-\end{itemize}
-
-\item Jason Yosinski
-\begin{itemize}
-\item SVM regression and prediction
-\item Evolutionary Neural Network (HyperNEAT)
-\item \code{WiiTrackClient} and \code{WiiTrackServer} classes and hardware
-\item \code{Robot}, \code{RunManager}, and \code{Strategy} classes
-\item \code{Parameterized Motion Model} abstract base class
-\item \code{SineModel} classes
-\item \code{explore\_dimensions}
-\item Plotting and evaluating results
-\end{itemize}
-
-\end{itemize}
-
-\section{Appendix}
-% Describe the code files you uploaded into CMS
-
-A brief description of the code uploaded to the CMS follows:
-
-\begin{itemize}
-\item \code{optimize.py}: Main program, determines a strategy to try and runs the robot
-  with that strategy.
-\item \code{RunManager.py}: Deals with all the details of running the robot,
-  including choosing an initial parameter, running the robot multiple times, 
-  tracking distance walked, and writing to the log file. Also includes the
-  \code{explore\_dimensions} method.
-\item \code{Strategy.py}: Contains all the different possible strategies, which
-  will be passed as objects in \code{optimize.py}. 
-\item \code{Robot.py}: Implements the \code{Robot} class, described in
-  \secref{implement}.
-\item \code{SineModel.py}: Implements a sine based motion model,
-  described in \secref{implement}.
-\item \code{Motion.py}: Motion helper functions.
-\item \code{WiiTrackServer.py}: Broadcasts the position of the infrared LED.
-\item \code{WiiTrackClient.py}: Connects to the WiiTrackServer to get the current position information.
-\end{itemize}
-
-% LocalWords:  vertices

### 2011-01-20 paper/learningMethods.tex
Split approach into three files

--- 2010-12-07 report/report.tex
@@
-\section{Introduction}
+\subsection{Learning Methods}
 
-% Motivate and abstractly describe the problem you are addressing and
-% how you are addressing it. What is the problem? Why is it important?
-% What is your basic approach? A short discussion of how it fits into
-% % related work in the area is also desirable. Summarize the basic
-% results and conclusions that you will present.
+\edit{write this.  Old quadratot Method section below}
 
-Applications of walking robots often call for the ability to walk as
-quickly, efficiently, or with as little power as possible.  Often
-these optimizations are done manually by an expert who designs and
-tweaks a gait specifically for a given objective.  Other groups have
-used learning methods to generate gaits optimized for some metric.
-Approaches differ in their starting assumptions, some essentially
-tweaking the parameters of a hand-tuned model \cite{chernova}, others
-exploring a reasonably compact parameter space \cite{kohl}, and still
-others beginning with few assumptions besides periodicity
-\cite{zykov}.
-
-We aimed to strike a middle ground between these approaches.  Our
-motion generator did not rely on hand-tweaked gaits, but it did use
-parameterized gaits which, by their nature, impose some assumptions on
-the answers produced.  We then used machine learning to design gaits
-for a quadruped robot with these models.  This paper presents a
-comparison of the different learning methods implemented.  Most
-methods created walks that are several times faster than the original
-hand-tuned gait.  We invite readers with short attention spans to view
-a video of some of our results online here:
-
-\url{http://www.youtube.com/watch?v=ODoiOj9DdGg}
-
-%The code so far consists of the robot class, optimization class,
-%parameterized model, sine model, and camera feedback.
-
-\edit{proofread}
-
-\section{Problem definition}
-% Precisely define the problem you are addressing (i.e. formally specify
-% the inputs and outputs). Elaborate on why this is an interesting and
-% important problem.
-
-We are testing several different learning methods to design a
-parametrized gait for a quadruped robot from the Cornell Computational
-Synthesis Lab.
-
-The output each of the learning algorithms is a function of time,
-$f(t)$, that outputs a vector of commanded motor positions.  This
-function is generated using a parametrized motion model, described in
-\secref{motion}.
-
-The robot executes these commands and measures its change in location
-using the tracking system described in \secref{implement}.  The input
-to the learning algorithms is this measured displacement, which the
-algorithms attempt to maximize. This displacement is measured for each
-gait over a constant length run, usually 12 seconds.
-
-A comparison and evaluation of the many different methods available
-for optimizing the gait of legged robots will be useful for future
-work on this challenging multidimensional control problem.
-
-\section{Method}
+%\section{Method}
 \seclabel{method}
@@
   gait, and select for evaluation the vector with the best predicted
-  performance.
+  performance.  We suspected that if we always chose the best
+  predicted point out of 100, we may end up progressing along a narrow
+  subspace, prohibiting learning of the true local fitness function.
+  Put another way, we would always choose exploitation of knowledge
+  vs. exploration of the space.  To address this concern, we added a
+  parameter dubbed \code{bumpBy} that added noise to the final
+  selected point before it was evaluated.
 
-  \edit{Has many parameters... in simulation... we found that it's
-    better to bump points...}
+  Such a method naturally has many tunable parameters, and we
+  endeavored to select these parameters by tuning the method in
+  simulation.  To estimate the performance of the algorithm, we ran it
+  against a simulation with a known optimum.  The simulated function
+  was in the same five dimensional parameter space, and simply
+  returned a fitness determined as the height of a Gaussian with a
+  random mean.  The width of the Gaussian in each dimension was 20\%
+  of the range of each dimension, and the maximum value at the peak
+  was 100.  \figref{svm_sim_results} shows the learning results on
+  this simulated model using the ultimately selected SVM parameters.
+  Interestingly, a non-zero value of \code{bumpBy} resulted in better
+  learning than noise free (exploration free) learning.
 
-\item \emph{Evolutionary Neural Network (HyperNEAT)\cite{clune}}
+  Ultimately, however, the version of SVM tuned for simulation still
+  did not show competitive performance on the real robot.  We tried
+  tuning some parameters on the real robot, but after some amount of
+  tuning, the method still exhibited too little exploration and easily
+  became stuck in local minima.
 
-  \edit{do this}
+\item \emph{Evolutionary Neural Network (HyperNEAT)\cite{clune}}: We 
+  put together an interface between HyperNEAT
+  -- an implementation of a method for evolving neural networks -- and
+  the robot, requiring a slightly modified strategy interface.
+
+  % Preliminary HyperNEAT runs were promising and resulted in several
+  % interesting gaits.
+
+  % Unfortunately, the gaits generated by HyperNEAT
+  % also tended to stress the robot more than typical gaits had before,
+  % and the servos would often overheat and malfunction, requiring
+  % restarts.  We think these issues may be addressed by adding a small
+  % layer between the HyperNEAT strategy and the robot that disallows
+  % quickly shifting commanded positions, and we hope to be able to test
+  % these methods further once this filter is in place.
 
@@
 
-\section{Related work}
-% Briefly explain who else worked on related problems in the past and what
-% methods they used. Explain if you are using similar methods, or if your
-% approach is different and if so - how (either is ok).
-
-Various maching learning techniques have proven to be useful in
-finding control policies for a wide variety of robots. Kohl and
-Stone\cite{kohl} presented a policy gradient reinforcement learning
-approach for generating a fast walk on legged robots. We experimented
-with this method to create a walk for our robot (Policy Gradient
-Descent). Chernova and Velosa\cite{chernova} took an evolutionary
-approach to this problem which we did not implement.  Zykov, Bongard,
-and Lipson\cite{zykov} describe the evolution of dynamic gaits on a
-physical robot requiring no prior assumptions about the locomotion
-pattern beyond the fact that it should be rhythmic.
-
-\section{System Architecture and Implementation}
-\seclabel{implement}
-% Describe how you implemented your system and how you structured it. 
-% This should give an overview of the system, not a detailed 
-% documentation of the code. The documentation of the code is part of 
-% the code you hand in. You might want to comment on high-level design 
-% decisions that you made. Also explain how you obtained your
-% data and any pre-processing you did to it.
-
-The quadruped robot has an on-board computer running Linux. The lower
-level drivers are in C and the system is implemented in
-Python. Feedback about distance travelled is provided via an infrared
-LED mounted on the robot and a Wii remote fixed to the ceiling.
-
-An overview of the code follows.
-
-\begin{itemize}
-
-\item \code{Robot} class: Class wrapper for commanding
-  motion of the robot.  The \code{Robot} class takes care of the robot
-  initialization, communication with the servos, and timing of the
-  runs.  In addition, it prevents the servos from ever being commanded
-  to a point outside their normal range (0 - 1023) as well as beyond
-  points where limbs would collide with parts of the robot body.  The
-  main class function, \code{run}, accepts a motion model (any
-  function that takes a time argument and outputs a 9 dimensional
-  position) and will run the robot using this motion model, including,
-  if desired, smooth interpolation over time for the beginning and end
-  of the run.
-
-\item \code{RunManager} class: Deals with all the details of running
-  the robot, including running the robot multiple times, tracking
-  distance walked via a \code{WiiTrackClient} member object, and
-  writing results to the log file. Also includes the
-  \code{explore\_dimensions} method to generate plots by varying each
-  parameter independently.
-
-\item \code{Strategy} class: The user has a choice between eight
-  different learning strategies: Random search, uniform random hill
-  climbing, Gaussian random hill climbing, N-dimensional policy
-  gradient descent, Nelder-Mead simplex, linear regression/prediction,
-  SVM regression/prediction, and HyperNEAT evolution.  Each strategy
-  must derive from the base \code{Strategy} class and must implement
-  two methods: \code{getNext}, for getting the next parameter vector
-  to try, and \code{updateResults}, for communicating results of a run
-  back to the strategy.
-
-\item \code{MotionModel} class: We implemented several motion models,
-  the main being the \code{SineModel5} class.  All motion models take
-  as input a parameter vector and produce as output a motion model,
-  which is simply a function mapping from time to nine motor
-  positions.  The \code{SineModel5} model commands positions based on
-  a sine wave, creating a periodic pattern. The parameters are:
-  amplitude, wavelength, a multiplier for the inner vs outer motors,
-  multiplier for left vs right motors, and multiplier for back vs
-  front motors.  Other similar models were tested, including a seven
-  parameter model which allowed sine waves shifted in time, but these
-  were not used as extensively in our experiments as
-  \code{SineModel5}.
-
-\item \code{WiiTrackClient} and \code{WiiTrackServer} classes and
-  hardware: A Wii remote tracks the location of the robot through an
-  infrared LED mounted on top of the robot. A \code{WiiTrackServer} is
-  run on the robot and continuously tracks its position using the
-  CWiid library\cite{cwiid} to interface with the remote via
-  bluetooth.  The \code{RunManager} then makes a
-  \code{WiiTrackClient}, which connects via socket to the tracking
-  server and requests position updates periodically.
-  \code{RunManager} currently gets the robot's position at the
-  beginning of each run and then again at the end and uses this to
-  calculate the net change in position.
-
-\end{itemize}
-
-\section{Experimental Evaluation}
-
-\subsection{Methodology}
-% What are the criteria you are using to evaluate your method? What
-% specific hypotheses does your experiment test? Describe the 
-% experimental methodology that you used. What are the dependent and 
-% independent variables? For projects in machine learning, what is 
-% the training/test data that was used, and why is it realistic or
-% interesting? Exactly what performance data did you collect and how 
-% are you presenting and analyzing it? Comparisons to competing methods 
-% that address the same problem or to variations of your own algorithm 
-% are particularly useful.
-
-The metric for evaluation of the designed gait is speed. We stop each 
-run after plateauing results (no improvement for one third of the 
-policies seen so far). The standard length of a run designates that it
-should be stopped after there is no improvement for one half of the policies
-seen so far, but since all runs took place on the actual robot, without use
-of a simulator, certain time limitations were imposed on the learning process.
-
-We controlled our experiments from a computer that was connected via a
-wireless ethernet to the robot. The robot collected data about
-distance walked automatically on its own. If it walked outside of the
-Wii remote's viewable area, it informed the user, so the only human
-intervention required during an experiment was to move the robot back
-inside the viewable area and to resume the run, which did not
-interrupt the learning process or result in the loss of data.
-
-We evaluated the efficacy of a set of parameters by sending these
-parameters to the robot and instructing it to walk for a certain
-length of time. The robot always began from the same position and
-returned to the starting position at the end of the run in order to
-measure true displacement without giving credit for ending in a leaned
-position. More efficient parameters resulted in a faster gait, which
-translated into a longer distance walked and a better score. After completing
-an evaluation, the robot sent the resulting distance walked back to the
-host computer and prepared itself for a new set of parameters to evaluate.
-
-Each algorithm was run on 3 different initial parameter vectors on the
-physical robot. We decided to evaluate all methods starting at the
-same three vector in order to allow for the fair comparison of each
-algorithm.  We evaluate each method based on the amount of improvement
-seen from the initial parameter vectors, and on the fastest speeds
-achieved during runs.
-
-The resulting gaits from our algorithms quickly outperformed the original
-hand-coded walk designed for this robot. The fastest walk, for example, was
-4 times faster.
-
-\subsection{Results}
-% Present the quantitative results of your experiments. Graphical data
-% presentation such as graphs and histograms are frequently better 
-% than tables. Explain what are the basic differences revealed in the 
-% data. If your method is stochastic or you analyze multiple datasets, 
-% calculate mean and standard error for every quantitative
-% performance result you show based on multiple runs.
-
-We have done complete runs of 3 different initial parameter vectors
-for random search, uniform random hill climbing, Gaussian random hill
-climbing, policy gradient descent, Nelder-Mead simplex, and linear
-regression.  We also evaluated the SVM regression and HyperNEAT
-methods, but these methods were more experimental, and thus we do not
-yet have the same volume of data for these runs.  We developed several
-gaits that were about 4 times faster than the original hand-coded
-gait.
-
-\edit{Convert to Table}
-
-\begin{tabular}{|r|c|c|c||c|}
-\hline
-                                         & A       & B      & C      &  Average \\
-\hline                                                               
-\hline                                                               
-Previous hand-coded gait                 & --      & --     & --     &  5.16 \\
-\hline                                                                 
-Random search                            & 6.04    & 17.26  & 4.90   &  9.40 \\
-\hline                                                                 
-Uniform Random Hill Climbing             & 11.37   & 9.44   & 2.69   &  7.83 \\
-\hline                                                                 
-Gaussian Random Hill Climbing            & 3.10    & 13.59  & 13.40  &  10.03 \\
-\hline                                                                 
-Policy Gradient Descent                  & 0.68    & 14.69  & 3.60   &  6.32 \\
-\hline                                                                 
-Nelder-Mead simplex                      & 8.51    & 13.62  & 14.83  &  12.32 \\
-\hline                                                                 
-Linear Regression                        & 27.58   & 12.51  & 1.95   &  14.01 \\
-%\hline                                                                
-%SVM Regression                           & ???     & ???    & ???   &   \\
-%\hline                                                                
-%Evolutionary Neural Network (HyperNEAT)  & ???     & ???    & ???   &   \\
-\hline
-\end{tabular}
-
-\edit{Caption: The best gaits found for each starting vector and
-  algorithm, in body lengths per minute.}
-
-\begin{itemize}
-
-\item For vector A, shown in \figref{vectorA}, linear regression
-  worked significantly better than the other algorithms, resulting in
-  a gait (27.58 body lengths/minute) that walked over twice as fast as
-  the next best gait -- uniform random hill climbing at 11.37 body
-  lengths/minute -- and 5.3 times better than the previous hand-coded
-  gait.
-
-\item Vector B, depicted in \figref{vectorB}, resulted in similarly
-  performing gaits with all the algorithms. The random method got
-  lucky and produced the best gait (17.26 body lengths/minute) and
-  uniform random hill climbing produced the worst (9.44 body
-  lengths/minute). The remaining algorithms each produced a gait
-  around 13 body lengths/minute.
-
-\item For vector C, shown in \figref{vectorC}, simplex and Gaussian
-  random hill climbing each produced a gait that substantially
-  outperformed the other algorithms. Simplex resulted in a gait of
-  nearly 15 body lengths/minute and Gaussian random hill climbing
-  produced a gait of just over 13 body lengths/minute, whereas the
-  other algorithms returned gaits of less than 5 body lengths/minute.
-
-\end{itemize}
-
-\figvarp{vectorA}{1}{Results for runs beginning with vector A.
-  Linear regression worked significantly better than the other
-  algorithms, resulting in a gait (27.58 body lengths/minute) that
-  walked over twice as fast as the next best gait -- uniform random
-  hill climbing at 11.37 body lengths/minute -- and 5.3 times better
-  than the previous hand-coded gait.}{}
-
-\figvarp{vectorB}{1}{Results for run starting with vector B.  All
-  algorithms performed similarly.  The random method actually got
-  lucky this time and produced the best gait (17.26 body
-  lengths/minute) and uniform random hill climbing produced the worst
-  (9.44 body lengths/minute). The remaining algorithms each produced a
-  gait around 13 body lengths/minute.}{}
-
-\figvarp{vectorC}{1}{Results for runs starting with vector C.
-  Simplex and Gaussian random hill climbing each produced a gait that
-  substantially outperformed the other algorithms, with around 15 body
-  lengths/minute and 13 body lengths/minute, respectively, whereas the
-  other algorithms returned gaits of less than 5 body
-  lengths/minute.}{}
-
-Based on the three trials, as shown in \figref{std_error}, linear
-regression worked the best, followed by simplex. Gaussian random hill
-climbing performed about as well as random, but it is unclear whether
-Gaussian would continue to improve and become much better than random
-search if allowed to run for more iterations. Uniform random hill
-climbing and policy gradient descent performed similarly to each other
-on average, but differed greatly on individual runs.
-
-\figvarp{std_error}{1}{Average results for each method starting with
-  parameter vectors A, B, and C.  Error bars are plotted showing the
-  standard error.  Linear regression outperformed other methods,
-  followed by simplex. Gaussian random hill climbing performed about
-  as well as random, but it is unclear whether Gaussian would continue
-  to improve and become much better than random search if allowed to
-  run for more iterations. Uniform random hill climbing and policy
-  gradient descent performed similarly to each other on average, but
-  differed greatly on individual runs.}{}
-
-In addition to running strategies to optimize parameterized gaits, we
-also ......
-
-We also coded \code{explore\_dimensions} to collect data for dimension
-varying plots. We intend to do further explore this by running
-\code{explore\_dimension} around a parameter that produces a medium-good gait.
-
-\edit{Jason: add exploration runs}
-
-\edit{Jason: add SVM and HyperNEAT results here!}
-
-\subsection{Discussion}
-% Is your hypothesis supported? What conclusions do the results support
-% about the strengths and weaknesses of your method compared to other 
-% methods? How can the results be explained in terms of the underlying 
-% properties of the algorithm and/or the data.
-
-Because only three trials were tested with each algorithm and no
-algorithm consistently outperformed the others, there is a large
-standard error for each algorithm, as show in \figref{std_error}. For
-this reason, it is unclear if any algorithm is superior to another in
-this application. More trials with each algorithm would be necessary
-to reach a definitive ranking of the performance of the algorithms.
-
-However, each algorithm discovered at least one gait of over 10 body
-lengths/minute, including random. For this reason, we speculate that
-the motion model is at least as critical as the learning algorithm
-used.
-
-\section{Future work}
-% What are the major shortcomings of your current method? For each 
-% shortcoming, propose additions or enhancements that would help overcome it.
-
-There are several directions in which we could continue our
-work. First, the error margins for our runs were large, so reducing
-them by running more trials would lead to more conclusive data. It
-would also be insightful to run our algorithms on different motion
-models.  We suspect that our choice of motion model influenced the
-results greatly, as even random choices in the space produced gaits
-that moved a significant fraction of the time (> 5\%).  It would be
-interesting to see how learning methods would perform using a model
-that included a much higher percentage of unproductive gaits. We also
-intend to experiment more with evolutionary algorithms/HyperNEAT.
-Some parameter vectors resulted in the robot turning, as opposed to it
-walking long distances. This could be an interesting learning goal in
-future projects. To these ends, we propose the following additions and
-enhancements:
-
-\begin{itemize}
-\item More runs and/or longer runs
-\item Different motion representations
-\item Evolutionary algorithms/HyperNEAT\cite{clune}
-\item Learning how to turn
-\end{itemize}
-
-\section{Conclusion}
-% Briefly summarize the important results and conclusions presented in 
-% the paper. What are the most important points illustrated by your 
-% work? How will your results improve future research and applications 
-% in the area?
-
-We have presented an array of approaches to optimizing a quadrupedal
-gait for forward speed.  We have implemented and tested different
-learning strategies, including uniform and Gaussian random hill
-climbing, policy gradient reinforcement learning, Nelder-Mead simplex,
-several new predictive methods based on linear and support vector
-regression, and an evolved neural network (HyperNEAT).  We have also
-compared these approaches to random search as a baseline. Many of the
-methods resulted in walks significantly faster than previously
-hand-tuned gaits.
-
-Because only three trials were tested with each algorithm and no
-algorithm consistently outperformed the others, there was a large
-standard error for each method, as shown in \figref{std_error}. Thus
-it was unclear if any algorithm was superior to another in this
-application. More trials with each algorithm would be necessary to
-reach a definitive ranking. Because each algorithm discovered at least one gait of
-over 10 body lengths/minute, including random search, we also
-conjectured that the motion representation for the robot is more
-integral to forward speed than the learning algorithm.  How to learn
-the motion representation, in addition to its parameters, remains an
-open problem.
-
-\section{Acknowledgments}
-% List any people not on the team who helped you with your project: Your
-% TA, other people you consulted with or had useful discussions (say in 
-% a word or two what they did), people who proofread your report, and 
-% any external code you used (libraries etc).
-\begin{itemize}
-\item Hod Lipson, Cornell Computational Synthesis Lab: adviser.
-\item Jim T\o rreson, University of Oslo: adviser.
-\item Juan Zagal, University of Chile: designed and printed robot and provided code for hand tuned gait.
-\item Jeff Clune, CCSL Red Couch: collaborated on HyperNEAT implementation/testing.
-\item Cooper Bills, Cornell University: assisted with Wii tracker development.
-\item Anshumali Srivastava, Cornell University: Teaching Assistant.
-\end{itemize}
-
-\section{The Team}
-% A table listing the names of the people who worked on the project, and who
-% did what.
-
-\begin{itemize}
-
-\item Diana Hidalgo: 
-\begin{itemize}
-\item Simplex (Nelder-Mead)
-\item \code{WiiTrackClient} and \code{WiiTrackServer} classes and hardware
-\item Plotting and evaluating results
-\item Investigation into using Optitrack IR system for tracking robot
-\end{itemize}
-
-\item Sarah Nguyen
-\begin{itemize}
-\item Random search
-\item Uniform random hill climbing
-\item Gaussian random hill climbing
-\item N-dimensional policy gradient descent
-\item Linear regression and prediction
-\item \code{RunManager} and \code{Strategy} classes
-\item Video editing
-\end{itemize}
-
-\item Jason Yosinski
-\begin{itemize}
-\item SVM regression and prediction
-\item Evolutionary Neural Network (HyperNEAT)
-\item \code{WiiTrackClient} and \code{WiiTrackServer} classes and hardware
-\item \code{Robot}, \code{RunManager}, and \code{Strategy} classes
-\item \code{Parameterized Motion Model} abstract base class
-\item \code{SineModel} classes
-\item \code{explore\_dimensions}
-\item Plotting and evaluating results
-\end{itemize}
-
-\end{itemize}
-
-\section{Appendix}
-% Describe the code files you uploaded into CMS
-
-A brief description of the code uploaded to the CMS follows:
-
-\begin{itemize}
-\item \code{optimize.py}: Main program, determines a strategy to try and runs the robot
-  with that strategy.
-\item \code{RunManager.py}: Deals with all the details of running the robot,
-  including choosing an initial parameter, running the robot multiple times, 
-  tracking distance walked, and writing to the log file. Also includes the
-  \code{explore\_dimensions} method.
-\item \code{Strategy.py}: Contains all the different possible strategies, which
-  will be passed as objects in \code{optimize.py}. 
-\item \code{Robot.py}: Implements the \code{Robot} class, described in
-  \secref{implement}.
-\item \code{SineModel.py}: Implements a sine based motion model,
-  described in \secref{implement}.
-\item \code{Motion.py}: Motion helper functions.
-\item \code{WiiTrackServer.py}: Broadcasts the position of the infrared LED.
-\item \code{WiiTrackClient.py}: Connects to the WiiTrackServer to get the current position information.
-\end{itemize}
-
-% LocalWords:  vertices
+\acmFig{svm_sim_results}{1}{Results for the SVM regression strategy
+  in simulation.  This simulation was used to tune the SVM strategy's
+  parameters before trying it on the physical robot.  The strategy
+  quickly approaches the maximum simulated fitness of 100.}

### 2011-01-20 paper/platformDetails.tex
Split approach into three files

--- 2010-12-07 report/report.tex
@@
-\section{Introduction}
+\subsection{Platform details}
 
-% Motivate and abstractly describe the problem you are addressing and
-% how you are addressing it. What is the problem? Why is it important?
-% What is your basic approach? A short discussion of how it fits into
-% % related work in the area is also desirable. Summarize the basic
-% results and conclusions that you will present.
+\editbox{write this, old quadratot System Architecture and Implementation section below.}
 
-Applications of walking robots often call for the ability to walk as
-quickly, efficiently, or with as little power as possible.  Often
-these optimizations are done manually by an expert who designs and
-tweaks a gait specifically for a given objective.  Other groups have
-used learning methods to generate gaits optimized for some metric.
-Approaches differ in their starting assumptions, some essentially
-tweaking the parameters of a hand-tuned model \cite{chernova}, others
-exploring a reasonably compact parameter space \cite{kohl}, and still
-others beginning with few assumptions besides periodicity
-\cite{zykov}.
+%\section{System Architecture and Implementation}
+\seclabel{implement}
 
-We aimed to strike a middle ground between these approaches.  Our
-motion generator did not rely on hand-tweaked gaits, but it did use
-parameterized gaits which, by their nature, impose some assumptions on
-the answers produced.  We then used machine learning to design gaits
-for a quadruped robot with these models.  This paper presents a
-comparison of the different learning methods implemented.  Most
-methods created walks that are several times faster than the original
-hand-tuned gait.  We invite readers with short attention spans to view
-a video of some of our results online here:
-
-\url{http://www.youtube.com/watch?v=ODoiOj9DdGg}
-
-%The code so far consists of the robot class, optimization class,
-%parameterized model, sine model, and camera feedback.
-
-\edit{proofread}
-
-\section{Problem definition}
-% Precisely define the problem you are addressing (i.e. formally specify
-% the inputs and outputs). Elaborate on why this is an interesting and
-% important problem.
-
-We are testing several different learning methods to design a
-parametrized gait for a quadruped robot from the Cornell Computational
-Synthesis Lab.
-
-The output each of the learning algorithms is a function of time,
-$f(t)$, that outputs a vector of commanded motor positions.  This
-function is generated using a parametrized motion model, described in
-\secref{motion}.
-
-The robot executes these commands and measures its change in location
-using the tracking system described in \secref{implement}.  The input
-to the learning algorithms is this measured displacement, which the
-algorithms attempt to maximize. This displacement is measured for each
-gait over a constant length run, usually 12 seconds.
-
-A comparison and evaluation of the many different methods available
-for optimizing the gait of legged robots will be useful for future
-work on this challenging multidimensional control problem.
-
-\section{Method}
-\seclabel{method}
-
-% Describe in reasonable detail the algorithm you are using to address
-% this problem. A pseudo-code description of the algorithm you are
-% using is frequently useful. If it makes sense for your project,
-% trace through a concrete example, showing how your algorithm
-% processes this example. The example should be complex enough to
-% illustrate all of the important aspects of the problem but simple
-% enough to be easily understood. If possible, an intuitively
-% meaningful example is better than one with meaningless symbols.
-
-We use several parameterized motion models that command motors to
-positions based on a sine wave, creating a periodic pattern.  While we
-investigated several models, for the bulk of our experiments, we used
-a model whose five parameters are: amplitude, wavelength, scale inner
-vs outer motors, scale left vs right motors, scale back vs front
-motors. Each strategy below attempts to choose the best possible
-parameters for this motion model.
-
-We implemented and tested 8 different learning strategies.  All
-strategies except for the HyperNEAT method\cite{clune} were
-constrained to pick parameters from within predetermined ranges.
-
-\begin{itemize}
-
-\item \emph{Random}: This method randomly generates parameter vectors
-  in the allowable range. This strategy is used only as baseline.
-
-\item \emph{Uniform random hill climbing}: This method begins by
-  selecting a single random parameter vector.  Subsequent iterations
-  generate a neighbor by randomly choosing one parameter to adjust and
-  replacing it with a new value chosen with uniform probability in the
-  allowable range for that parameter. The neighbor is evaluated by
-  running the robot with the newly chosen parameters. If this neighbor
-  results in a longer distance walked than the previous best gait, it
-  is saved as the new best gait. The process is then repeated, always
-  starting with the best gait.
-
-\item \emph{Gaussian random hill climbing}: This method works
-  similarly to Uniform random hill climbing, except neighbors are
-  generated by adding random Gaussian noise to the current best gait.
-  This results in all parameters being changed at once, but the
-  resulting vector is always fairly close to the previous best gait.
-  We used independently selected noise in each dimension, scaled such
-  that the standard deviation of the noise was 5\% of the range of
-  that dimension.
-
-\item \emph{N-dimensional policy gradient descent}: As opposed to the
-  previous methods, this method explicitly estimates the gradient for
-  the objective function. It does this by first evaluating \emph{t}
-  randomly generated parameter vectors near the initial vector, each
-  dimension of these vectors being perturbed by either $-\epsilon$,
-  $0$, or $\epsilon$. Then, for each dimension, it groups vectors into
-  three groups: $-\epsilon$, $0$, and $\epsilon$.  The gradient along
-  this dimension is then estimated as the average score for the
-  $\epsilon$ group minus the average score for the $-\epsilon$
-  group. Finally, the method creates a new vector by changing all
-  parameters by a fixed-size step in the direction of the gradient.
-
-\item \emph{Nelder-Mead simplex method}\cite{nm}: The Nelder-Mead
-  simplex method creates an initial simplex with 6 vertices. The
-  initial parameter vector is stored as the first vertex and the other
-  five vertices are created by adding to one dimension at a time one
-  tenth of the allowable range for that parameter. It then tests the
-  fitness of each vertex and based on these fitnesses, it reflects the
-  worst point over the centroid in an attempt to improve it.  However,
-  to prevent cycles and becoming stuck in local minima, several other
-  rules are used.  In general, the worst vertex is reflected over the
-  centroid. If the reflected point is better than the second worst
-  point and worse than the best point, then the reflected point
-  replaces the worst. If the reflected point is better than the best
-  point, the simplex is expanded in the direction of the reflected
-  point. The better of the reflected and the expanded point replaces
-  the worst point. If the reflected point is worse than the second
-  worst point, then the simplex is contracted away from the reflected
-  point. If the contracted point is better than the reflected point,
-  the contracted point replaces the worst point. If the contracted
-  point is worse than the reflected point, the entire simplex is
-  shrunk \cite{NMWebsite}.
-
-\item \emph{Linear regression}: To initialize, this method chooses and
-  evaluates five random parameter vectors. It then fits a linear model
-  from parameter vector to fitness. In a loop, the method chooses and
-  evaluates a new parameter vector generated by taking a fixed-size
-  step in the direction of the gradient for each parameter, and fits a
-  new linear model to all vectors evaluated so far, choosing the model
-  to minimize the sum of squared errors.
-
-\item \emph{SVM regression}: Similarly to linear regression, this
-  model starts with several random vectors, but this time they are
-  chosen in a small neighborhood about some initial random vector.
-  These vectors (generally 8) are evaluated, and a support vector
-  regression model is fit to the observed fitnesses.  To choose the
-  next vector for evaluation, we randomly generate some number
-  (typically 100) of vectors in the neighborhood of the best observed
-  gait, and select for evaluation the vector with the best predicted
-  performance.
-
-  \edit{Has many parameters... in simulation... we found that it's
-    better to bump points...}
-
-\item \emph{Evolutionary Neural Network (HyperNEAT)\cite{clune}}
-
-  \edit{do this}
-
-\end{itemize}
-
-\section{Related work}
-% Briefly explain who else worked on related problems in the past and what
-% methods they used. Explain if you are using similar methods, or if your
-% approach is different and if so - how (either is ok).
-
-Various maching learning techniques have proven to be useful in
-finding control policies for a wide variety of robots. Kohl and
-Stone\cite{kohl} presented a policy gradient reinforcement learning
-approach for generating a fast walk on legged robots. We experimented
-with this method to create a walk for our robot (Policy Gradient
-Descent). Chernova and Velosa\cite{chernova} took an evolutionary
-approach to this problem which we did not implement.  Zykov, Bongard,
-and Lipson\cite{zykov} describe the evolution of dynamic gaits on a
-physical robot requiring no prior assumptions about the locomotion
-pattern beyond the fact that it should be rhythmic.
-
-\section{System Architecture and Implementation}
-\seclabel{implement}
 % Describe how you implemented your system and how you structured it. 
@@
 \end{itemize}
-
-\section{Experimental Evaluation}
-
-\subsection{Methodology}
-% What are the criteria you are using to evaluate your method? What
-% specific hypotheses does your experiment test? Describe the 
-% experimental methodology that you used. What are the dependent and 
-% independent variables? For projects in machine learning, what is 
-% the training/test data that was used, and why is it realistic or
-% interesting? Exactly what performance data did you collect and how 
-% are you presenting and analyzing it? Comparisons to competing methods 
-% that address the same problem or to variations of your own algorithm 
-% are particularly useful.
-
-The metric for evaluation of the designed gait is speed. We stop each 
-run after plateauing results (no improvement for one third of the 
-policies seen so far). The standard length of a run designates that it
-should be stopped after there is no improvement for one half of the policies
-seen so far, but since all runs took place on the actual robot, without use
-of a simulator, certain time limitations were imposed on the learning process.
-
-We controlled our experiments from a computer that was connected via a
-wireless ethernet to the robot. The robot collected data about
-distance walked automatically on its own. If it walked outside of the
-Wii remote's viewable area, it informed the user, so the only human
-intervention required during an experiment was to move the robot back
-inside the viewable area and to resume the run, which did not
-interrupt the learning process or result in the loss of data.
-
-We evaluated the efficacy of a set of parameters by sending these
-parameters to the robot and instructing it to walk for a certain
-length of time. The robot always began from the same position and
-returned to the starting position at the end of the run in order to
-measure true displacement without giving credit for ending in a leaned
-position. More efficient parameters resulted in a faster gait, which
-translated into a longer distance walked and a better score. After completing
-an evaluation, the robot sent the resulting distance walked back to the
-host computer and prepared itself for a new set of parameters to evaluate.
-
-Each algorithm was run on 3 different initial parameter vectors on the
-physical robot. We decided to evaluate all methods starting at the
-same three vector in order to allow for the fair comparison of each
-algorithm.  We evaluate each method based on the amount of improvement
-seen from the initial parameter vectors, and on the fastest speeds
-achieved during runs.
-
-The resulting gaits from our algorithms quickly outperformed the original
-hand-coded walk designed for this robot. The fastest walk, for example, was
-4 times faster.
-
-\subsection{Results}
-% Present the quantitative results of your experiments. Graphical data
-% presentation such as graphs and histograms are frequently better 
-% than tables. Explain what are the basic differences revealed in the 
-% data. If your method is stochastic or you analyze multiple datasets, 
-% calculate mean and standard error for every quantitative
-% performance result you show based on multiple runs.
-
-We have done complete runs of 3 different initial parameter vectors
-for random search, uniform random hill climbing, Gaussian random hill
-climbing, policy gradient descent, Nelder-Mead simplex, and linear
-regression.  We also evaluated the SVM regression and HyperNEAT
-methods, but these methods were more experimental, and thus we do not
-yet have the same volume of data for these runs.  We developed several
-gaits that were about 4 times faster than the original hand-coded
-gait.
-
-\edit{Convert to Table}
-
-\begin{tabular}{|r|c|c|c||c|}
-\hline
-                                         & A       & B      & C      &  Average \\
-\hline                                                               
-\hline                                                               
-Previous hand-coded gait                 & --      & --     & --     &  5.16 \\
-\hline                                                                 
-Random search                            & 6.04    & 17.26  & 4.90   &  9.40 \\
-\hline                                                                 
-Uniform Random Hill Climbing             & 11.37   & 9.44   & 2.69   &  7.83 \\
-\hline                                                                 
-Gaussian Random Hill Climbing            & 3.10    & 13.59  & 13.40  &  10.03 \\
-\hline                                                                 
-Policy Gradient Descent                  & 0.68    & 14.69  & 3.60   &  6.32 \\
-\hline                                                                 
-Nelder-Mead simplex                      & 8.51    & 13.62  & 14.83  &  12.32 \\
-\hline                                                                 
-Linear Regression                        & 27.58   & 12.51  & 1.95   &  14.01 \\
-%\hline                                                                
-%SVM Regression                           & ???     & ???    & ???   &   \\
-%\hline                                                                
-%Evolutionary Neural Network (HyperNEAT)  & ???     & ???    & ???   &   \\
-\hline
-\end{tabular}
-
-\edit{Caption: The best gaits found for each starting vector and
-  algorithm, in body lengths per minute.}
-
-\begin{itemize}
-
-\item For vector A, shown in \figref{vectorA}, linear regression
-  worked significantly better than the other algorithms, resulting in
-  a gait (27.58 body lengths/minute) that walked over twice as fast as
-  the next best gait -- uniform random hill climbing at 11.37 body
-  lengths/minute -- and 5.3 times better than the previous hand-coded
-  gait.
-
-\item Vector B, depicted in \figref{vectorB}, resulted in similarly
-  performing gaits with all the algorithms. The random method got
-  lucky and produced the best gait (17.26 body lengths/minute) and
-  uniform random hill climbing produced the worst (9.44 body
-  lengths/minute). The remaining algorithms each produced a gait
-  around 13 body lengths/minute.
-
-\item For vector C, shown in \figref{vectorC}, simplex and Gaussian
-  random hill climbing each produced a gait that substantially
-  outperformed the other algorithms. Simplex resulted in a gait of
-  nearly 15 body lengths/minute and Gaussian random hill climbing
-  produced a gait of just over 13 body lengths/minute, whereas the
-  other algorithms returned gaits of less than 5 body lengths/minute.
-
-\end{itemize}
-
-\figvarp{vectorA}{1}{Results for runs beginning with vector A.
-  Linear regression worked significantly better than the other
-  algorithms, resulting in a gait (27.58 body lengths/minute) that
-  walked over twice as fast as the next best gait -- uniform random
-  hill climbing at 11.37 body lengths/minute -- and 5.3 times better
-  than the previous hand-coded gait.}{}
-
-\figvarp{vectorB}{1}{Results for run starting with vector B.  All
-  algorithms performed similarly.  The random method actually got
-  lucky this time and produced the best gait (17.26 body
-  lengths/minute) and uniform random hill climbing produced the worst
-  (9.44 body lengths/minute). The remaining algorithms each produced a
-  gait around 13 body lengths/minute.}{}
-
-\figvarp{vectorC}{1}{Results for runs starting with vector C.
-  Simplex and Gaussian random hill climbing each produced a gait that
-  substantially outperformed the other algorithms, with around 15 body
-  lengths/minute and 13 body lengths/minute, respectively, whereas the
-  other algorithms returned gaits of less than 5 body
-  lengths/minute.}{}
-
-Based on the three trials, as shown in \figref{std_error}, linear
-regression worked the best, followed by simplex. Gaussian random hill
-climbing performed about as well as random, but it is unclear whether
-Gaussian would continue to improve and become much better than random
-search if allowed to run for more iterations. Uniform random hill
-climbing and policy gradient descent performed similarly to each other
-on average, but differed greatly on individual runs.
-
-\figvarp{std_error}{1}{Average results for each method starting with
-  parameter vectors A, B, and C.  Error bars are plotted showing the
-  standard error.  Linear regression outperformed other methods,
-  followed by simplex. Gaussian random hill climbing performed about
-  as well as random, but it is unclear whether Gaussian would continue
-  to improve and become much better than random search if allowed to
-  run for more iterations. Uniform random hill climbing and policy
-  gradient descent performed similarly to each other on average, but
-  differed greatly on individual runs.}{}
-
-In addition to running strategies to optimize parameterized gaits, we
-also ......
-
-We also coded \code{explore\_dimensions} to collect data for dimension
-varying plots. We intend to do further explore this by running
-\code{explore\_dimension} around a parameter that produces a medium-good gait.
-
-\edit{Jason: add exploration runs}
-
-\edit{Jason: add SVM and HyperNEAT results here!}
-
-\subsection{Discussion}
-% Is your hypothesis supported? What conclusions do the results support
-% about the strengths and weaknesses of your method compared to other 
-% methods? How can the results be explained in terms of the underlying 
-% properties of the algorithm and/or the data.
-
-Because only three trials were tested with each algorithm and no
-algorithm consistently outperformed the others, there is a large
-standard error for each algorithm, as show in \figref{std_error}. For
-this reason, it is unclear if any algorithm is superior to another in
-this application. More trials with each algorithm would be necessary
-to reach a definitive ranking of the performance of the algorithms.
-
-However, each algorithm discovered at least one gait of over 10 body
-lengths/minute, including random. For this reason, we speculate that
-the motion model is at least as critical as the learning algorithm
-used.
-
-\section{Future work}
-% What are the major shortcomings of your current method? For each 
-% shortcoming, propose additions or enhancements that would help overcome it.
-
-There are several directions in which we could continue our
-work. First, the error margins for our runs were large, so reducing
-them by running more trials would lead to more conclusive data. It
-would also be insightful to run our algorithms on different motion
-models.  We suspect that our choice of motion model influenced the
-results greatly, as even random choices in the space produced gaits
-that moved a significant fraction of the time (> 5\%).  It would be
-interesting to see how learning methods would perform using a model
-that included a much higher percentage of unproductive gaits. We also
-intend to experiment more with evolutionary algorithms/HyperNEAT.
-Some parameter vectors resulted in the robot turning, as opposed to it
-walking long distances. This could be an interesting learning goal in
-future projects. To these ends, we propose the following additions and
-enhancements:
-
-\begin{itemize}
-\item More runs and/or longer runs
-\item Different motion representations
-\item Evolutionary algorithms/HyperNEAT\cite{clune}
-\item Learning how to turn
-\end{itemize}
-
-\section{Conclusion}
-% Briefly summarize the important results and conclusions presented in 
-% the paper. What are the most important points illustrated by your 
-% work? How will your results improve future research and applications 
-% in the area?
-
-We have presented an array of approaches to optimizing a quadrupedal
-gait for forward speed.  We have implemented and tested different
-learning strategies, including uniform and Gaussian random hill
-climbing, policy gradient reinforcement learning, Nelder-Mead simplex,
-several new predictive methods based on linear and support vector
-regression, and an evolved neural network (HyperNEAT).  We have also
-compared these approaches to random search as a baseline. Many of the
-methods resulted in walks significantly faster than previously
-hand-tuned gaits.
-
-Because only three trials were tested with each algorithm and no
-algorithm consistently outperformed the others, there was a large
-standard error for each method, as shown in \figref{std_error}. Thus
-it was unclear if any algorithm was superior to another in this
-application. More trials with each algorithm would be necessary to
-reach a definitive ranking. Because each algorithm discovered at least one gait of
-over 10 body lengths/minute, including random search, we also
-conjectured that the motion representation for the robot is more
-integral to forward speed than the learning algorithm.  How to learn
-the motion representation, in addition to its parameters, remains an
-open problem.
-
-\section{Acknowledgments}
-% List any people not on the team who helped you with your project: Your
-% TA, other people you consulted with or had useful discussions (say in 
-% a word or two what they did), people who proofread your report, and 
-% any external code you used (libraries etc).
-\begin{itemize}
-\item Hod Lipson, Cornell Computational Synthesis Lab: adviser.
-\item Jim T\o rreson, University of Oslo: adviser.
-\item Juan Zagal, University of Chile: designed and printed robot and provided code for hand tuned gait.
-\item Jeff Clune, CCSL Red Couch: collaborated on HyperNEAT implementation/testing.
-\item Cooper Bills, Cornell University: assisted with Wii tracker development.
-\item Anshumali Srivastava, Cornell University: Teaching Assistant.
-\end{itemize}
-
-\section{The Team}
-% A table listing the names of the people who worked on the project, and who
-% did what.
-
-\begin{itemize}
-
-\item Diana Hidalgo: 
-\begin{itemize}
-\item Simplex (Nelder-Mead)
-\item \code{WiiTrackClient} and \code{WiiTrackServer} classes and hardware
-\item Plotting and evaluating results
-\item Investigation into using Optitrack IR system for tracking robot
-\end{itemize}
-
-\item Sarah Nguyen
-\begin{itemize}
-\item Random search
-\item Uniform random hill climbing
-\item Gaussian random hill climbing
-\item N-dimensional policy gradient descent
-\item Linear regression and prediction
-\item \code{RunManager} and \code{Strategy} classes
-\item Video editing
-\end{itemize}
-
-\item Jason Yosinski
-\begin{itemize}
-\item SVM regression and prediction
-\item Evolutionary Neural Network (HyperNEAT)
-\item \code{WiiTrackClient} and \code{WiiTrackServer} classes and hardware
-\item \code{Robot}, \code{RunManager}, and \code{Strategy} classes
-\item \code{Parameterized Motion Model} abstract base class
-\item \code{SineModel} classes
-\item \code{explore\_dimensions}
-\item Plotting and evaluating results
-\end{itemize}
-
-\end{itemize}
-
-\section{Appendix}
-% Describe the code files you uploaded into CMS
-
-A brief description of the code uploaded to the CMS follows:
-
-\begin{itemize}
-\item \code{optimize.py}: Main program, determines a strategy to try and runs the robot
-  with that strategy.
-\item \code{RunManager.py}: Deals with all the details of running the robot,
-  including choosing an initial parameter, running the robot multiple times, 
-  tracking distance walked, and writing to the log file. Also includes the
-  \code{explore\_dimensions} method.
-\item \code{Strategy.py}: Contains all the different possible strategies, which
-  will be passed as objects in \code{optimize.py}. 
-\item \code{Robot.py}: Implements the \code{Robot} class, described in
-  \secref{implement}.
-\item \code{SineModel.py}: Implements a sine based motion model,
-  described in \secref{implement}.
-\item \code{Motion.py}: Motion helper functions.
-\item \code{WiiTrackServer.py}: Broadcasts the position of the infrared LED.
-\item \code{WiiTrackClient.py}: Connects to the WiiTrackServer to get the current position information.
-\end{itemize}
-
-% LocalWords:  vertices

### 2011-01-21 paper/problemDefinition.tex
Done with Problem Definition for now

@@
 % important problem.
-
-\edit{write this.  Quadratot report section below}
-
-We are testing several different learning methods to design a
-parametrized gait for a quadruped robot from the Cornell Computational
-Synthesis Lab.
-
-The output each of the learning algorithms is a function of time,
-$f(t)$, that outputs a vector of commanded motor positions.  This
-function is generated using a parametrized motion model, described in
-\secref{implement}.
-
-The robot executes these commands and measures its change in location
-using the tracking system described in \secref{implement}.  The input
-to the learning algorithms is this measured displacement, which the
-algorithms attempt to maximize. This displacement is measured for each
-gait over a constant length run, usually 12 seconds.
-
-A comparison and evaluation of the many different methods available
-for optimizing the gait of legged robots will be useful for future
-work on this challenging multidimensional control problem.
-
-\edit{Jason writing this now}
+%
+%\edit{write this.  Quadratot report section below}
+%
+%We are testing several different learning methods to design a
+%parametrized gait for a quadruped robot from the Cornell Computational
+%Synthesis Lab.
+%
+%The output each of the learning algorithms is a function of time,
+%$f(t)$, that outputs a vector of commanded motor positions.  This
+%function is generated using a parametrized motion model, described in
+%\secref{implement}.
+%
+%The robot executes these commands and measures its change in location
+%using the tracking system described in \secref{implement}.  The input
+%to the learning algorithms is this measured displacement, which the
+%algorithms attempt to maximize. This displacement is measured for each
+%gait over a constant length run, usually 12 seconds.
+%
+%A comparison and evaluation of the many different methods available
+%for optimizing the gait of legged robots will be useful for future
+%work on this challenging multidimensional control problem.
 
@@
 a period of motion lasting approximately 10 seconds.  For details of
-exactly how this displacement was measured, see \secref{foo}.
+exactly how this displacement was measured, see
+\secref{fitnessEvaluation}.

### 2011-01-21 paper/approach.tex
More work on Experimental Approach and Learning Methods

@@
 \section{Experimental Approach}
+
+Given the goal of producing a deterministic, feedforward motion
+function $g(t)$, we now turn to several important considerations.
+\secref{learningMethods} talks about how this function is generated, \secref{platformDetails} discusses the hardware platform on which it is run, and \secref{fitnessEvaluation} details exactly how its fitness is measured.
 

### 2011-01-21 paper/learningMethods.tex
More work on Experimental Approach and Learning Methods

@@
 
-\edit{write this.  Old quadratot Method section below}
+There are a plethora of methods available for the creation of a vector motion generating function $g(t)$, and we've chosen two for comparison in this study.
+
+The first method is to choose a family of \emph{parametrized functions}.
+By fixing the parameters to a particular set of values, we obtain a deterministic motion function over time.
+We tried several parametrizations on the robot and, obtaining reasonble early success, settled on one particular parametrization.
+We dubbed this the \emph{SineModel5}, as it has five parameters and employs a sine wave as its root pattern.
+
+The five parameters are as follows:
+
+ - amp - Amplitude
+ - tau   - sine wave period
+ - multIO - multiplier scaling inner joings from outer joints
+ - multFB - multiplier scaling front joints from back joints
+ - multLR - multiplier scaling left joints from right joints
+(verify this)
+
+Inuitively, SineModel5 starts with a sine wave of amplitude 1
+
+ used is a function parametrized by five numbers, shown in \figref{foo}
+
+\edit{make figure to go here, show drawing with inner/outer, l/r, and f/b motors.}
+
+\edit{write this.  Old quadratot Method section below alsdkfj }
 
@@
 
-We use several parameterized motion models that command motors to
+We use several parametrized motion models that command motors to
 positions based on a sine wave, creating a periodic pattern.  While we

### 2011-01-21 paper/problemDefinition.tex
More work on Experimental Approach and Learning Methods

@@
 %The output each of the learning algorithms is a function of time,
-%$f(t)$, that outputs a vector of commanded motor positions.  This
+%$g(t)$, that outputs a vector of commanded motor positions.  This
 %function is generated using a parametrized motion model, described in
@@
 Mathematically, we define a gait as a function that specifies
-commanded motor positions for a robot over time.  Using this
-definition, we require one function per motor.  We can write gaits
-without feedback --- also called feed forward gaits --- as
+a vector of commanded motor positions for a robot over time.
+%Using this definition, we require one function per motor.
+We can write gaits without feedback --- also called feed forward gaits
+--- as
 
 \be
-x_i = f_i(t)
+\vec{x} = g(t)
 \ee
 
-\noindent for commanded position $x_i$ for motor $i$.  The
+%\noindent for commanded position $x_i$ for motor $i$.  The
+\noindent for commanded position vector $\vec{x}$.  The
 function depends only on time.  Using the same framework, a gait
@@
 \be
-x_i = f_i(t, z)
+\vec{x} = g(t, \vec{z})
 \ee
 
-\noindent for some measurement $z$.  Though perhaps obvious, it is
-worth noting that feedforward gaits are deterministic (unless they
+\noindent for some measurement vector $\vec{z}$.  Though perhaps obvious, it
+is worth noting that feedforward gaits are deterministic (unless they
 include random elements), producing the same command pattern each time
@@
 other hand, depends on the actual measurements, and noise in these
-measurements may result in different $x_i$ values each time they are
+measurements may result in different $\vec{x}$ values each time they are
 run.

### 2011-01-21 paper/learningMethods.tex
More learning methods work (mostly)

--- 2011-01-20 paper/learningMethods.tex
@@
-\subsection{Learning Methods}
+\subsection{Parametrized Motion Models}
+\seclabel{motionModel}
 
-\edit{write this.  Old quadratot Method section below}
+\edit{this might no longer flow...fix} There are a plethora of methods available for the creation of a vector motion generating function $g(t)$, and we've chosen two for comparison in this study.
+
+The first method is to choose a family of \emph{parametrized functions}.
+By fixing the parameters to a particular set of values, we obtain a deterministic motion function over time.
+We tried several parametrizations on the robot and, obtaining reasonble early success, settled on one particular parametrization.
+We dubbed this the \emph{SineModel5}, as it has five parameters and employs a sine wave as its root pattern.
+
+The five parameters in $\vec{\theta}$ are as follows:
+
+ - amp - Amplitude       - allowable range
+ - tau   - sine wave period   - allowable range
+ - multIO - multiplier scaling inner joings from outer joints  - allowable range
+ - multFB - multiplier scaling front joints from back joints  - allowable range
+ - multLR - multiplier scaling left joints from right joints  - allowable range
+(verify this)
+
+\newcommand{\amp}{\mathrm{amp}}
+
+Inuitively, SineModel5 starts with 8 identicaly sine waves of amplitude $\amp$ and period $\tau$, multiplies the waves for all inner motors by multIO, multiplies the waves for all front motors by multFB, and multiplies the waves for all left motors by multLR.
+To obtain the actual motor position commands, these waves are offset by an appropriate fixed constant so that the base position (when the sine waves are at 0) is approximately a crouch (the position shown in \figref{crouchingRobot}).
+Finally, in this motion model, the ninth "hip" motor is kept at a neutral position.
+Thus, the commanded position for all motors, as a vector function of time, is:
+
+\begin{verbatim}
+        [g_0(t) ]    [ \amp * sin(\tau t) ]
+        [g_1(t) ]    [ \amp * sin(\tau t) ]
+        [g_2(t) ]    [ \amp * sin(\tau t) ]
+g(t)   [g_3(t) ] = [ \amp * sin(\tau t) ]
+        [g_4(t) ]    [ \amp * sin(\tau t) ]
+        [g_5(t) ]    [ \amp * sin(\tau t) ]
+        [g_6(t..) ]    [ \amp * sin(\tau t) ]
+\end{verbatim}
+
+The nine subscripted functions correspond to the nine labeled motors in \figref{robotDiagram}
+
+% used is a function parametrized by five numbers, shown in \figref{foo}
+
+\subsection{Learning Methods for Parametrized Motion Models}
+\seclabel{learningMethods}
+
+Using the SineModel5 parametrized motion model described in the previous section, along with the allowable ranges for each of the five parameters (shown in \tabref{parameterTable}), the task becomes how to choose the combination of five parameters that results in the fastest motion, per the evaluation methods in \secref{fitnessEvaluation}.
+
+If we choose a value for the five dimensional parameter $\vec{\theta}$, then a given physical trial gives us one measurement of the fitness of that parameter vector, or $f(\vec{\theta})$.
+Two things make learnign difficult.
+First, each evaluation of $f(\vec{\theta})$ is expensive, taking 15-20 seconds of time on average.
+Second, the fitness returned by such evaluations has proven to be very noisy, with the standard deviation of the noise often being roughly equivalent to the size of the measurement.
+
+For both of these reasons, an intelligent learning algorithm is needed.
+
+In this context, by learning algorithm we simply mean a method for choosing:
+
+1) The value for $\vec{\theta}$ for the initial trial, and
+2) The next value of $\vec{\theta}$ to try, given the $\vec{\theta}$ and $f(\vec{\theta})$ for all previous trials
+
+We evaluated a total of eight subtly or not so subtly different learning methods.
+All employed a simple random sampling method for requirement (1); that is, all methods picked their initial $\vec{\theta}$ value via uniform random sampling within the allowed parameter ranges.
+\edit{make sure this is true}.
+Thus, the differences in the algorithms was how the selected new $\vec{\theta}$ values to try from their past experience.
+
+The eight methods used, and their methods for choosing the next $\vec{\theta}$ are:
+
+\edit{write this.  Old quadratot Method section below alsdkfj }
 
 %\section{Method}
-\seclabel{method}
+%\seclabel{method}
 
@@
 
-We use several parameterized motion models that command motors to
-positions based on a sine wave, creating a periodic pattern.  While we
-investigated several models, for the bulk of our experiments, we used
-a model whose five parameters are: amplitude, wavelength, scale inner
-vs outer motors, scale left vs right motors, scale back vs front
-motors. Each strategy below attempts to choose the best possible
-parameters for this motion model.
+%We use several parametrized motion models that command motors to
+%positions based on a sine wave, creating a periodic pattern.  While we
+%investigated several models, for the bulk of our experiments, we used
+%a model whose five parameters are: amplitude, wavelength, scale inner
+%vs outer motors, scale left vs right motors, scale back vs front
+%motors. Each strategy below attempts to choose the best possible
+%parameters for this motion model.
 
-We implemented and tested 8 different learning strategies.  All
-strategies except for the HyperNEAT method\cite{clune} were
-constrained to pick parameters from within predetermined ranges.
+%We implemented and tested 8 different learning strategies.  All
+%strategies except for the HyperNEAT method\cite{clune} were
+%constrained to pick parameters from within predetermined ranges.
 
@@
 \item \emph{Random}: This method randomly generates parameter vectors
-  in the allowable range. This strategy is used only as baseline.
+  in the allowable range for every trial. This strategy was used only as baseline for comparison.
 
@@
 
-\item \emph{N-dimensional policy gradient descent}: As opposed to the
-  previous methods, this method explicitly estimates the gradient for
-  the objective function. It does this by first evaluating \emph{t}
-  randomly generated parameter vectors near the initial vector, each
-  dimension of these vectors being perturbed by either $-\epsilon$,
-  $0$, or $\epsilon$. Then, for each dimension, it groups vectors into
-  three groups: $-\epsilon$, $0$, and $\epsilon$.  The gradient along
-  this dimension is then estimated as the average score for the
-  $\epsilon$ group minus the average score for the $-\epsilon$
-  group. Finally, the method creates a new vector by changing all
-  parameters by a fixed-size step in the direction of the gradient.
+\item \emph{N-dimensional policy gradient descent}: In \cite{kohl},
+  Kohl and Stone document a method for local gradient ascent for gait
+  learning with noisy fitness evaluations, and we have included this
+  method in our evaluation.  This strategy explicitly estimates the
+  gradient for the objective function. It does this by first
+  evaluating \emph{n} randomly generated parameter vectors near the
+  initial vector, each dimension of these vectors being perturbed by
+  either $-\epsilon$, $0$, or $\epsilon$. Then, for each dimension, it
+  groups vectors into three groups: $-\epsilon$, $0$, and $\epsilon$.
+  The gradient along this dimension is then estimated as the average
+  score for the $\epsilon$ group minus the average score for the
+  $-\epsilon$ group. Finally, the method creates a new vector by
+  changing all parameters by a fixed-size step in the direction of the
+  gradient.
 
-\item \emph{Nelder-Mead simplex method}\cite{nm}: The Nelder-Mead
-  simplex method creates an initial simplex with 6 vertices. The
-  initial parameter vector is stored as the first vertex and the other
-  five vertices are created by adding to one dimension at a time one
-  tenth of the allowable range for that parameter. It then tests the
-  fitness of each vertex and based on these fitnesses, it reflects the
-  worst point over the centroid in an attempt to improve it.  However,
-  to prevent cycles and becoming stuck in local minima, several other
-  rules are used.  In general, the worst vertex is reflected over the
-  centroid. If the reflected point is better than the second worst
-  point and worse than the best point, then the reflected point
-  replaces the worst. If the reflected point is better than the best
-  point, the simplex is expanded in the direction of the reflected
-  point. The better of the reflected and the expanded point replaces
-  the worst point. If the reflected point is worse than the second
-  worst point, then the simplex is contracted away from the reflected
-  point. If the contracted point is better than the reflected point,
-  the contracted point replaces the worst point. If the contracted
-  point is worse than the reflected point, the entire simplex is
-  shrunk \cite{nm}.
+\item \emph{Nelder-Mead simplex method}: The Nelder-Mead simplex
+  method \cite{nm} creates an initial simplex with $d+1$ vertices,
+  where $d$ is the dimension of the parameter space. The initial
+  parameter vector is stored as the first vertex and the other five
+  vertices are created by adding to one dimension at a time one tenth
+  of the allowable range for that parameter. It then tests the fitness
+  of each vertex and based on these fitnesses, it reflects the worst
+  point over the centroid in an attempt to improve it.  In general,
+  the worst vertex is reflected; however, to prevent cycles and
+  becoming stuck in local minima, several other rules are used.  If
+  the reflected point is better than the second worst point and worse
+  than the best point, then the reflected point replaces the worst. If
+  the reflected point is better than the best point, the simplex is
+  expanded in the direction of the reflected point. The better of the
+  reflected and the expanded point replaces the worst point. If the
+  reflected point is worse than the second worst point, then the
+  simplex is contracted away from the reflected point. If the
+  contracted point is better than the reflected point, the contracted
+  point replaces the worst point. If the contracted point is worse
+  than the reflected point, the entire simplex is shrunk \cite{nm}.
 

### 2011-01-21 paper/introduction.tex
Hello all. Good: added HyperNEAT description and lots of comments. Bad, its long, and I don't have HyperNEAT figures in there yet. I'll keep working at it, which will  mean it will likely get longer.
Length will be our enemy. Mark my words <ar har har har har...evil laugh...>
 Good stuff so far Jason!

@@
  - Gait learning has been done before, in simulation and on hardware (cite, cite, cite) \\
- - HyperNEAT has been used to evolve gaits before in sim (cite, ...) \\
+ - HyperNEAT has been used to evolve gaits before in simulation~{clune2009evolving, clune2011performance, clune2009hybrid}, and it worked well in part because the gaits were regular, raising the question of whether this high performance was just because the gaits were regular?
+We test this hypothesis by comparing HyperNEAT to traditional machine learning methods biased towards regular gaits.
+\\
  - Here, we do two cool things for the first time: \\

### 2011-01-22 paper/results.tex
first draft of the results section

@@
 
-\edit{write this.}
+\edit{proofread this.}
+\subsection{Maximizing Strategies}
+The maximizing strategies implemented include uniform random hill climbing, Gaussian random hill climbing, policy gradient descent, and Nelder-Mead simplex.
+As a group, these strategies produced results similar to gaits randomly generated using the SineModel5 model.
+None of these strategies consistently outperformed the others in the three trials; A different strategy generated the best gait in each trial.
+Uniform random hill climbing produced the best gait (11.37 body lengths/minute) in the first trial, with the next best performing strategy (Nelder-Mead simplex) resulting in a gait of 8.51 body lengths/minute.
+The best gait in the second trial was actually generated randomly (17.26 body-lengths/minute).
+Gaussian random hill climbing, policy gradient descent, and Nelder-Mead simplex all produced similar best gaits in the second trial, each around 14 body lengths/minute.
+In the third trial, Nelder-Mead simplex and Gaussian random hill climbing significantly outperformed the other maximizing strategies, resulting in gaits of 14.83 and 13.40 body lengths/minute respectively.
+Overall, Nelder-Mead simplex performed the best, producing an average best gait of 12.32 body lengths/minute.
+Gaussian random hill climbing followed, with an average best gait of 10.03 body lengths/minute.
+
+\subsection{Predictive Strategies}
+The predictive strategies used were linear regression and support vector regression.
+In the first trial, linear regression produced a gait (27.58 body lengths/minute) that was over three times faster than the best gait produced by a maximizing strategy (uniform random hill climbing -- 11.37 body lengths/minute).
+The average of the best gait for the three trials of linear regression was 14.01 body lengths/minute, which was better than the average best gait for the best performing maximizing strategy (Nelder-Mead simplex -- 12.32 body lengths/minute).
+
+As mentioned above, the SVM strategy involved many tunable parameters.
+The version of SVM tuned for simulation did not show competitive performance on the real robot.
+We tried tuning some parameters on the real robot, but after some amount of tuning, the method still exhibited too little exploration and easily became stuck in local minima.
+
+\subsection{Evolved Neural Network Strategy}
+The evolved neural network strategy used was HyperNEAT.
+HyperNEAT produced the fastest gaits by far.
+Linear regression produced a gait in the first trial (27.58 body lengths/minute) that was on par with the best gaits generated by HyperNEAT, by no other maximizing or predictive strategy produced gaits close to those of HyperNEAT.
+Linear regression was the best performing maximizing or predictive strategy, but it resulted in an average best gait of only 14.01 body lengths/minute.
+HyperNEAT, on the other hand, resulted in an average best gait of 29.26 body lengths/minute.
+One issue encountered with HyperNEAT was that it generated gaits that were rougher on the robot than gaits generated by the other strategies.
+The gaits sometimes caused the motor servos to overheat, which causes the motor to behave differently than expected.
+Since an overheated motor produces a different walk that expected, a test was performed after each run to make sure all motors were working correctly.
+If a motor was not, the distance walked would be halved to penalize walks that were harmful to the robot.
+
+\begin{table}
+\begin{center}
+\begin{tabular}{|r|c|c|c||c|}
+\hline
+                                         & A       & B      & C      &  Average \\
+\hline                                                               
+\hline                                                               
+Previous hand-coded gait                 & --      & --     & --     &  5.16 \\
+\hline                                                                 
+Random search                            & 6.04    & 17.26  & 4.90   &  9.40 \\
+\hline                                                                 
+Uniform Random Hill Climbing             & 11.37   & 9.44   & 2.69   &  7.83 \\
+\hline                                                                 
+Gaussian Random Hill Climbing            & 3.10    & 13.59  & 13.40  &  10.03 \\
+\hline                                                                 
+Policy Gradient Descent                  & 0.68    & 14.69  & 3.60   &  6.32 \\
+\hline                                                                 
+Nelder-Mead simplex                      & 8.51    & 13.62  & 14.83  &  12.32 \\
+\hline                                                                 
+Linear Regression                        & 27.58   & 12.51  & 1.95   &  14.01 \\
+\hline                                                                
+Evolutionary Neural Network (HyperNEAT)  & 24.27     & 36.44    & 27.07   & 29.26  \\
+\hline
+\end{tabular}
+\caption{The best gaits found for each starting vector and algorithm,
+  in body lengths per minute.}
+\tablabel{results_table}
+\end{center}
+\end{table}
 
@@
 Linear Regression                        & 27.58   & 12.51  & 1.95   &  14.01 \\
-%\hline                                                                
-%SVM Regression                           & ???     & ???    & ???   &   \\
-%\hline                                                                
-%Evolutionary Neural Network (HyperNEAT)  & ???     & ???    & ???   &   \\
+\hline                                                                
+Evolutionary Neural Network (HyperNEAT)  & 24.27     & 36.44    & 27.07   & 29.26  \\
 \hline

### 2011-01-26 paper/problemDefinition.tex
Few edits

@@
 %Using this definition, we require one function per motor.
-We can write gaits without feedback --- also called feed forward gaits
+We can write gaits without feedback --- also called feedforward gaits
 --- as
@@
 
-\noindent for some measurement vector $\vec{z}$.  Though perhaps obvious, it
-is worth noting that feedforward gaits are deterministic (unless they
-include random elements), producing the same command pattern each time
-they are run.  The output of a gait that includes feedback, on the
-other hand, depends on the actual measurements, and noise in these
-measurements may result in different $\vec{x}$ values each time they are
-run.
+\noindent for some measurement vector $\vec{z}$.  Feedforward gaits
+are deterministic (unless they include random elements), so they
+produce the same command pattern each time they are run.  While the
+commanded positions will be the same from trial to trial, the actual
+robot motion and measured fitness will vary due to the noisiness of
+trials in the real world.
+
+The output of a gait that includes feedback, on the other hand,
+depends on the actual measurements, and noise in these measurements
+may result in different $\vec{x}$ values each time they are run.
 

### 2011-01-26 paper/learningMethods.tex
Updated HyperNEAT description

@@
 
+\edit{make figure to go here, show drawing with inner/outer, l/r, and f/b motors.}
+\edit{JMC: Use photo of physical robot in this image, and add labels in photoshop?}
 The nine subscripted functions correspond to the nine labeled motors in \figref{robotDiagram}
@@
 %constrained to pick parameters from within predetermined ranges.
+
+\edit{JMC:Bullets=narrower column=wasted space. subsubsection instead?}
 
@@
 
+  \edit{JMC: This next paragraph sounds like it belongs in Results?}
   Ultimately, however, the version of SVM tuned for simulation still
@@
 
-\item \emph{Evolutionary Neural Network (HyperNEAT)\cite{clune}}: We 
-  put together an interface between HyperNEAT
-  -- an implementation of a method for evolving neural networks -- and
-  the robot, requiring a slightly modified strategy interface.
+\item \emph{Neuroevolution (Evolving Artificial Neural Networks with HyperNEAT)}:
+
+  HyperNEAT is an indirect encoding for evolving artificial neural networks (ANNs) that is inspired by the way natural organisms develop~\cite{stanley2009hypercube}.
+It evolves Compositional Pattern Producing Networks (CPPNs)~\cite{stanley2007compositional}, each of which is a genome that encodes an ANN phenotype~\cite{stanley2009hypercube}.
+Each CPPN is itself a directed graph, where each node is a mathematical function, such as sine or Gaussian.
+The nature of these functions can facilitate the evolution of properties such as symmetry (e.g., an absolute value or Gaussian function) and repetition (e.g., a sine function)~\cite{stanley2009hypercube, stanley2007compositional}.
+The signal on each link in the CPPN is multiplied by that link's weight, which can alter its effect.
+
+A CPPN is queried once for each link in the ANN substrate to determine that link's weight.
+The inputs to the CPPN are the Cartesian coordinates of both the source (e.g., $x = 2$, $y = 4$) and target (e.g., $x = 3$, $y = 5$) nodes of a link and a constant bias value.
+The CPPN takes these five values as inputs and produces one or two output values, depending on the substrate topology.
+If there is no hidden layer in the substrate, the single output is the weight of the link between a source node on the input layer and a target node on the output layer.
+If there is a hidden layer, the first output value determines the weight of the link between the associated input (source) and hidden layer (target) nodes, and the second output value determines the weight of the link between the associated hidden (source) and output (target) layer nodes.
+All pairwise combinations of source and target nodes are iteratively passed as inputs to a CPPN to determine the weight of each substrate link.
+
+\edit{following paragraph is figure caption for file named: hyperneatExplanation.png}
+HyperNEAT Produces ANNs from CPPNs.
+ANN weights are specified as a function of the geometric coordinates of the source node and the target node for each connection.
+The coordinates of these nodes and a constant bias are iteratively passed to the CPPN to determine each connection weight.
+The CPPN has two output values, which specify the weights for each connection layer as shown.
+
+HyperNEAT is capable of exploiting the geometry of a problem: because the link values between nodes in the final ANN substrate are a function of the geometric positions of those nodes, HyperNEAT can exploit such information when solving a problem~\cite{stanley2009hypercube, clune2009sensitivity, clune2011performance}.
+In the case of quadruped locomotion, this property helped Hyper-NEAT produce gaits with front-back, left-right, and four-way symmetries~\cite{clune2009evolving, clune2011performance}.
+
+The evolution of the population of CPPNs occurs according to the principles of the NeuroEvolution of Augmenting Topologies (NEAT) algorithm~\cite{stanley2002evolving}, which was originally designed to evolve ANNs.
+NEAT can be fruitfully applied to CPPNs because of their structural similarity to ANNs.
+For example, mutations can add a node, and thus a function, to a CPPN graph, or change its link weights.
+The NEAT algorithm is unique in three main ways~\cite{stanley2002evolving}.
+Initially, it starts with small genomes that encode simple networks and slowly complexifies them via mutations that add nodes and links to the network, enabling the algorithm to evolve the network topology in addition to its weights.
+Secondly, NEAT has a fitness sharing mechanism that preserves diversity in the system and gives time for new innovations to be tuned by evolution before competing them against more adapted rivals.
+Finally, NEAT tracks historical information to perform intelligent crossover while avoiding the need for expensive topological analysis.
+A full explanation of NEAT can be found in~\cite{stanley2002evolving}].
+
+The ANN configuration follows previous studies that evolved quadruped gaits with HyperNEAT in simulation~\cite{clune2011performance, clune2009evolving}, but was adapted to accommodate differences in the robot.
+Specifically, the ANNs consists of three $3 \times 4$ Cartesian grids of nodes forming input, hidden, and output layers.
+Adjacent layers were completely connected, meaning that there were $(3 \times 4)^2= 288$ links in each substrate.
+The inputs to the substrate were the current angles of each of the 9 joints of the robot (described below) and a sine and cosine wave (to facilitate the production of periodic behaviors).
+The sine and cosine waves were the following function of time (t) in milli-seconds: $sin(t/12.0)\pi$.
+This function produces numbers in the range [$-\pi, \pi$], which were then JASON FILL ME IN TO DESCRIBE HOW JOINT COMMANDS WERE SPECIFIED.
+
+\edit{are we mentioning that we gave HyperNEAT the center joint here or earlier?}
+\edit{jason: edit this section to reflect the post-processing...}
+
+\edit{add the following paragraph as a figure caption to SpiderANN.jpg}
+
+ANN Configuration for HyperNEAT Runs.
+The first two columns of each row of the input layer receive information about a single leg (the current angle of its two joints).
+The final column provides a sine and cosine wave to enable periodic movements and the angle of the center joint.
+Evolution determines the function of the hidden-layer nodes.
+The nodes in the output layer specify new joint angles for each respective joint.
+The unlabeled node in the input and output layer is ignored.
+
+\acmFig{explore_dim_1}{1}{Fitness mean and standard deviation
+  vs. dimension 1.  The circle is a common point in
+  \figref{explore_dim_1} through \figref{explore_dim_5}}
+
+The population size for HyperNEAT was 9 and runs lasted 20 generations.
+These numbers are small, but were necesary given how much time it took to conduct evolution directly on a real robot.
+The remaining parameters were identical to Clune et al.
+~\cite{clune2011performance}.
 

### 2011-01-26 paper/learningMethods.tex

@@
 
-\edit{this might no longer flow...fix} There are a plethora of methods available for the creation of a vector motion generating function $g(t)$, and we've chosen two for comparison in this study.
+\edit{this might no longer flow...fix} There are a plethora of methods
+available for the creation of a vector motion generating function
+$g(t)$, and we've chosen two for comparison in this study.
 
-The first method is to choose a family of \emph{parametrized functions}.
-By fixing the parameters to a particular set of values, we obtain a deterministic motion function over time.
-We tried several parametrizations on the robot and, obtaining reasonble early success, settled on one particular parametrization.
-We dubbed this the \emph{SineModel5}, as it has five parameters and employs a sine wave as its root pattern.
+The first method is to choose a family of \emph{parametrized
+  functions}.  By fixing the parameters to a particular set of values,
+we obtain a deterministic motion function over time.  We tried several
+parametrizations on the robot and, upon obtaining reasonable early
+success, settled on one particular parametrization.  We dubbed this
+the \emph{SineModel5}, as it has five parameters and employs a sine
+wave as its root pattern.
 
@@
 
- - amp - Amplitude       - allowable range
- - tau   - sine wave period   - allowable range
- - multIO - multiplier scaling inner joings from outer joints  - allowable range
- - multFB - multiplier scaling front joints from back joints  - allowable range
- - multLR - multiplier scaling left joints from right joints  - allowable range
-(verify this)
+%\newcommand{\amp}{\ensuremath{\mathrm{amp}}}
+\newcommand{\amp}{\ensuremath{\alpha}}
 
-\newcommand{\amp}{\mathrm{amp}}
+\begin{table}[h]
+\begin{center}
+\begin{tabular}{|c|c|c|}
+\hline
+Parameter    & Description               & Range \\
+\hline
+\hline
+\amp         & Amplitude                 & [0, 400] \\
+\hline
+$\tau$       & Period                    & [.5, 8] \\
+\hline
+$m_O$        & Outer motor multiplier    & [-2, 2] \\
+\hline
+$m_F$        & Front motor multiplier    & [-1, 1] \\
+\hline
+$m_R$        & Right motor multiplier    & [-1, 1] \\
+\hline
+\end{tabular}
+\caption{The parameters used for the \emph{SineModel5} motion model.}
+\tablabel{parameters}
+\end{center}
+\end{table}
 
-Inuitively, SineModel5 starts with 8 identicaly sine waves of amplitude $\amp$ and period $\tau$, multiplies the waves for all inner motors by multIO, multiplies the waves for all front motors by multFB, and multiplies the waves for all left motors by multLR.
-To obtain the actual motor position commands, these waves are offset by an appropriate fixed constant so that the base position (when the sine waves are at 0) is approximately a crouch (the position shown in \figref{crouchingRobot}).
-Finally, in this motion model, the ninth "hip" motor is kept at a neutral position.
-Thus, the commanded position for all motors, as a vector function of time, is:
+Intuitively, SineModel5 starts with 8 identically sine waves of
+amplitude $\amp$ and period $\tau$, multiplies the waves for all inner
+motors by multIO, multiplies the waves for all front motors by multFB,
+and multiplies the waves for all left motors by multLR.  To obtain the
+actual motor position commands, these waves are offset by
+appropriate fixed constants so that the base position (when the sine
+waves are at 0) is approximately a crouch (the position shown in
+\figref{crouchingRobot}).  Finally, in this motion model, the ninth
+"hip" motor is kept at a neutral position.  Thus, the commanded
+position for all motors, as a vector function of time, is:
 
-\begin{verbatim}
-        [g_0(t) ]    [ \amp * sin(\tau t) ]
-        [g_1(t) ]    [ \amp * sin(\tau t) ]
-        [g_2(t) ]    [ \amp * sin(\tau t) ]
-g(t)   [g_3(t) ] = [ \amp * sin(\tau t) ]
-        [g_4(t) ]    [ \amp * sin(\tau t) ]
-        [g_5(t) ]    [ \amp * sin(\tau t) ]
-        [g_6(t..) ]    [ \amp * sin(\tau t) ]
-\end{verbatim}
+\[
+\vec{g}(t) =
+\left[ {\begin{array}{c@{ }l@{ }l}
+\amp \cdot \sin(2\pi t / \tau) & \ \cdot \           m_F            & + C_I \\ % 0
+\amp \cdot \sin(2\pi t / \tau) & \ \cdot \ m_O \cdot m_F            & + C_O \\ % 1
+\amp \cdot \sin(2\pi t / \tau) & \                                  & + C_I \\ % 2
+\amp \cdot \sin(2\pi t / \tau) & \ \cdot \ m_O                      & + C_O \\ % 3
+\amp \cdot \sin(2\pi t / \tau) & \ \cdot \                     m_R  & + C_I \\ % 4
+\amp \cdot \sin(2\pi t / \tau) & \ \cdot \ m_O           \cdot m_R  & + C_O \\ % 5
+\amp \cdot \sin(2\pi t / \tau) & \ \cdot \           m_F \cdot m_R  & + C_I \\ % 6
+\amp \cdot \sin(2\pi t / \tau) & \ \cdot \ m_O \cdot m_F \cdot m_R  & + C_O \\ % 7
+0                              & \                                  & + C_C \\ % 8
+\end{array} } \right]
+\]
 
-\edit{make figure to go here, show drawing with inner/outer, l/r, and f/b motors.}
-\edit{JMC: Use photo of physical robot in this image, and add labels in photoshop?}
-The nine subscripted functions correspond to the nine labeled motors in \figref{robotDiagram}
+\noindent Here the fixed constants that determine the crouched
+position are $C_I$ for inner motors, $C_O$ for outer motors, and $C_C$
+for the center hip motor.
+
+\edit{make figure to go here (or another section), show drawing with
+  inner/outer, l/r, and f/b motors.}
+
+The nine subscripted functions correspond to the nine labeled motors
+in \figref{robotDiagram}
 
@@
 
-Using the SineModel5 parametrized motion model described in the previous section, along with the allowable ranges for each of the five parameters (shown in \tabref{parameterTable}), the task becomes how to choose the combination of five parameters that results in the fastest motion, per the evaluation methods in \secref{fitnessEvaluation}.
+Using the SineModel5 parametrized motion model described in the
+previous section, along with the allowable ranges for each of the five
+parameters (shown in \tabref{parameters}), the task becomes how to
+choose the combination of five parameters that results in the fastest
+motion, per the evaluation methods in \secref{fitnessEvaluation}.
 
-If we choose a value for the five dimensional parameter $\vec{\theta}$, then a given physical trial gives us one measurement of the fitness of that parameter vector, or $f(\vec{\theta})$.
-Two things make learnign difficult.
-First, each evaluation of $f(\vec{\theta})$ is expensive, taking 15-20 seconds of time on average.
-Second, the fitness returned by such evaluations has proven to be very noisy, with the standard deviation of the noise often being roughly equivalent to the size of the measurement.
+If we choose a value for the five dimensional parameter
+$\vec{\theta}$, then a given physical trial gives us one measurement
+of the fitness of that parameter vector, or $f(\vec{\theta})$.  Two
+things make learning difficult.  First, each evaluation of
+$f(\vec{\theta})$ is expensive, taking 15-20 seconds of time on
+average.  Second, the fitness returned by such evaluations has proved
+to be very noisy, with the standard deviation of the noise often being
+roughly equivalent to the size of the measurement.
 
-For both of these reasons, an intelligent learning algorithm is needed.
+For both of these reasons, an intelligent learning algorithm is
+needed.
 
-In this context, by learning algorithm we simply mean a method for choosing:
+In this context, by learning algorithm we mean a method for choosing:
 
-1) The value for $\vec{\theta}$ for the initial trial, and
-2) The next value of $\vec{\theta}$ to try, given the $\vec{\theta}$ and $f(\vec{\theta})$ for all previous trials
+\begin{enumerate}
+\item The value for $\vec{\theta}$ for the initial trial, and
+\item The next value of $\vec{\theta}$ to try, given the
+  $\vec{\theta}$ and $f(\vec{\theta})$ for all previous trials.
+\end{enumerate}
 
-We evaluated a total of eight subtly or not so subtly different learning methods.
-All employed a simple random sampling method for requirement (1); that is, all methods picked their initial $\vec{\theta}$ value via uniform random sampling within the allowed parameter ranges.
-\edit{make sure this is true}.
-Thus, the differences in the algorithms was how the selected new $\vec{\theta}$ values to try from their past experience.
+We evaluated a total of eight subtly or not so subtly different
+learning methods.  All employed a simple random sampling method for
+requirement (1); that is, all methods picked their initial
+$\vec{\theta}$ value via uniform random sampling within the allowed
+parameter ranges. \edit{make sure this is true. Isn't really, should say that we started with A, B, C}.  Thus, the
+differences in the algorithms was how they selected new $\vec{\theta}$
+values to try from their past experience.
 
-The eight methods used, and their methods for choosing the next $\vec{\theta}$ are:
+The eight methods used, and their methods for choosing the next
+$\vec{\theta}$ are:
 

### 2011-01-26 paper/platformDetails.tex
More learning methods work (mostly)
updated platform details by taking out the old stuff CS 4701 and reorganizing

@@
 \subsection{Platform details}
+
+\edit{figure:robotDiagram make figure to go somewhere here, show drawing with inner/outer, l/r, and f/b motors.}
 
@@
 level drivers are in C and the system is implemented in
-Python. Feedback about distance travelled is provided via an infrared
-LED mounted on the robot and a Wii remote fixed to the ceiling.
+Python.
 
-An overview of the code follows.
+A Wii remote tracks the location of the robot, providing feedback
+about distance traveled, through an infrared LED mounted on top of 
+the robot. A server is run on the robot and continuously tracks its
+position using the CWiid library\cite{cwiid} to interface with the remote
+via bluetooth. A client then connects via a socket to the tracking
+server and requests position updates periodically. If the robot moves 
+beyond the viewable range of the Wii remote, the system pauses and
+directs the user to move the robot back within the viewable range 
+before running again.
 
-\begin{itemize}
-
-\item \code{Robot} class: Class wrapper for commanding
-  motion of the robot.  The \code{Robot} class takes care of the robot
-  initialization, communication with the servos, and timing of the
-  runs.  In addition, it prevents the servos from ever being commanded
-  to a point outside their normal range (0 - 1023) as well as beyond
-  points where limbs would collide with parts of the robot body.  The
-  main class function, \code{run}, accepts a motion model (any
-  function that takes a time argument and outputs a 9 dimensional
-  position) and will run the robot using this motion model, including,
-  if desired, smooth interpolation over time for the beginning and end
-  of the run.
-
-\item \code{RunManager} class: Deals with all the details of running
-  the robot, including running the robot multiple times, tracking
-  distance walked via a \code{WiiTrackClient} member object, and
-  writing results to the log file. Also includes the
-  \code{explore\_dimensions} method to generate plots by varying each
-  parameter independently.
-
-\item \code{Strategy} class: The user has a choice between eight
-  different learning strategies: Random search, uniform random hill
-  climbing, Gaussian random hill climbing, N-dimensional policy
-  gradient descent, Nelder-Mead simplex, linear regression / prediction,
-  SVM regression/prediction, and HyperNEAT evolution.  Each strategy
-  must derive from the base \code{Strategy} class and must implement
-  two methods: \code{getNext}, for getting the next parameter vector
-  to try, and \code{updateResults}, for communicating results of a run
-  back to the strategy.
-
-\item \code{MotionModel} class: We implemented several motion models,
-  the main being the \code{SineModel5} class.  All motion models take
-  as input a parameter vector and produce as output a motion model,
-  which is simply a function mapping from time to nine motor
-  positions.  The \code{SineModel5} model commands positions based on
-  a sine wave, creating a periodic pattern. The parameters are:
-  amplitude, wavelength, a multiplier for the inner vs outer motors,
-  multiplier for left vs right motors, and multiplier for back vs
-  front motors.  Other similar models were tested, including a seven
-  parameter model which allowed sine waves shifted in time, but these
-  were not used as extensively in our experiments as
-  \code{SineModel5}.
-
-\item \code{WiiTrackClient} and \code{WiiTrackServer} classes and
-  hardware: A Wii remote tracks the location of the robot through an
-  infrared LED mounted on top of the robot. A \code{WiiTrackServer} is
-  run on the robot and continuously tracks its position using the
-  CWiid library\cite{cwiid} to interface with the remote via
-  bluetooth.  The \code{RunManager} then makes a
-  \code{WiiTrackClient}, which connects via socket to the tracking
-  server and requests position updates periodically.
-  \code{RunManager} currently gets the robot's position at the
-  beginning of each run and then again at the end and uses this to
-  calculate the net change in position.
-
-\end{itemize}
+The robot is run using a given motion model, including, if desired, 
+smooth interpolation over time for the beginning and end of the run. 
+The servos are prevented from being commanded to a point outside their
+normal range (0 - 1023) as well as beyond points where limbs would collide

### 2011-01-26 paper/fitnessEvaluation.tex
Done with Problem Definition for now
wrote fitness evaluation section

@@
 \subsection{Fitness evaluation details}
+\seclabel{fitnessEvaluation}
 
-\editbox{write this.  Old quadratot Experimental Evaluation section below.}
+\editbox{proofread.}
 
@@
 
-%\subsection{Methodology}
-% What are the criteria you are using to evaluate your method? What
-% specific hypotheses does your experiment test? Describe the 
-% experimental methodology that you used. What are the dependent and 
-% independent variables? For projects in machine learning, what is 
-% the training/test data that was used, and why is it realistic or
-% interesting? Exactly what performance data did you collect and how 
-% are you presenting and analyzing it? Comparisons to competing methods 
-% that address the same problem or to variations of your own algorithm 
-% are particularly useful.
+We controlled the experiments from a computer that was connected via a wireless Ethernet to the robot.
+An infrared LED was mounted to the antenna on top of the robot and a Wii remote was attached to the ceiling.
+The Wii remote tracked the location of the LED.
+The robot tracked its position via Bluetooth by using the CWiid library\cite{cwiid} to communicate with the Wii remote.
+If the robot walked outside of the Wii remote’s viewable area, a prompt informed the experimenter.
+The only human intervention required during an experiment was to move the robot back inside the viewable area and resume the run.
+This did not interrupt the learning process or result in the loss of data.
 
-The metric for evaluation of the designed gait is speed. We stop each 
-run after plateauing results (no improvement for one third of the 
-policies seen so far). The standard length of a run designates that it
-should be stopped after there is no improvement for one half of the policies
-seen so far, but since all runs took place on the actual robot, without use
-of a simulator, certain time limitations were imposed on the learning process.
+The metric for evaluation of the designed gait was speed.
+To evaluate a set of parameters, the robot was sent the parameters and instructed to walk for a certain length of time.
+For each evaluation, the robot always started and ended in the same position in order to measure true displacement and not reward gaits the ended in a lean, since the LED would have moved a different distance the robot.
+More efficient parameters resulted in a faster gait, which translated into a longer distance walked and a better score.
 
-We controlled our experiments from a computer that was connected via a
-wireless ethernet to the robot. The robot collected data about
-distance walked automatically on its own. If it walked outside of the
-Wii remote's viewable area, it informed the user, so the only human
-intervention required during an experiment was to move the robot back
-inside the viewable area and to resume the run, which did not
-interrupt the learning process or result in the loss of data.
-
-We evaluated the efficacy of a set of parameters by sending these
-parameters to the robot and instructing it to walk for a certain
-length of time. The robot always began from the same position and
-returned to the starting position at the end of the run in order to
-measure true displacement without giving credit for ending in a leaned
-position. More efficient parameters resulted in a faster gait, which
-translated into a longer distance walked and a better score. After completing
-an evaluation, the robot sent the resulting distance walked back to the
-host computer and prepared itself for a new set of parameters to evaluate.
-
-Each algorithm was run on 3 different initial parameter vectors on the
-physical robot. We decided to evaluate all methods starting at the
-same three vector in order to allow for the fair comparison of each
-algorithm.  We evaluate each method based on the amount of improvement
-seen from the initial parameter vectors, and on the fastest speeds
-achieved during runs.
-
-The resulting gaits from our algorithms quickly outperformed the original
-hand-coded walk designed for this robot. The fastest walk, for example, was
-4 times faster.
+Each of the parameter ???
+methods was run on 3 different initial parameter vectors, in order to allow for the fair comparison of the algorithms.
+We allowed each run for the parameter ???
+methods to continue until the results plateaued (no improvement for one third of the policies seen so far).
+Three runs of HyperNEAT were completed, each with a different initial seed and run for 20 generations.

### 2011-01-26 paper/platformDetails.tex
added period.

@@
 \edit{figure:robotDiagram make figure to go somewhere here, show drawing with inner/outer, l/r, and f/b motors.}
-
-\editbox{write this, old quadratot System Architecture and Implementation section below.}
 
@@
 The servos are prevented from being commanded to a point outside their
-normal range (0 - 1023) as well as beyond points where limbs would collide
+normal range (0 - 1023) as well as beyond points where limbs would collide.

### 2011-01-26 paper/platformDetails.tex

@@
 % data and any pre-processing you did to it.
+\acmFig{quadratot.jpg}{1}{The robot}
+\acmFig{topdown.png}{1}{A figure of the robot from a top-down perspective,
+with motors labeled}
 
@@
 Python.
+
+\acmFig{robot.JPG}{1}{The infrared LED mounted on top of the robot, which
+provides feedback about distance traveled in conjunction with a Wii
+remote}
+\acmFig{wiiMote.JPG}{1}{A Wii remote tracks the location of the robot,
+providing feedback about distance traveled}
 

### 2011-01-26 paper/problemDefinition.tex
Incorporated some of Jeff's suggestions

@@
 %Using this definition, we require one function per motor.
-We can write gaits without feedback --- also called feedforward gaits
---- as
+We can write gaits without feedback --- also called open-loop or
+feedforward gaits --- as
 
@@
 \noindent for commanded position vector $\vec{x}$.  The
-function depends only on time.  Using the same framework, a gait
-including measured motor position, force, or other values could be
-written as
+function depends only on time.
 
-\be
-\vec{x} = g(t, \vec{z})
-\ee
+%Using the same framework, a gait
+%including measured motor position, force, or other values could be
+%written as
+%
+%\be
+%\vec{x} = g(t, \vec{z})
+%\ee
+%
+%\noindent for some measurement vector $\vec{z}$.
 
-\noindent for some measurement vector $\vec{z}$.  Feedforward gaits
-are deterministic (unless they include random elements), so they
-produce the same command pattern each time they are run.  While the
-commanded positions will be the same from trial to trial, the actual
-robot motion and measured fitness will vary due to the noisiness of
-trials in the real world.
+It follows that feedforward gaits are deterministic, producing the
+same command pattern each time they are run.  While the commanded
+positions will be the same from trial to trial, the actual robot
+motion and measured fitness will vary due to the noisiness of trials
+in the real world.
 
-The output of a gait that includes feedback, on the other hand,
-depends on the actual measurements, and noise in these measurements
-may result in different $\vec{x}$ values each time they are run.
+%The output of a gait that includes feedback, on the other hand,
+%depends on the actual measurements, and noise in these measurements
+%may result in different $\vec{x}$ values each time they are run.
 

### 2011-01-26 code/plot/plot_error_new.m
new plot error function added

@@ -1,0 +1,79 @@
+function plot_error_new(files, saveFileName)
+
+clr = {};
+  clr{end+1} = 'k';
+  clr{end+1} = 'r';
+  clr{end+1} = 'g';
+  clr{end+1} = 'b';
+  clr{end+1} = 'c';
+  clr{end+1} = 'm';
+%  clr{end+1} = 'y';
+
+figure(2); clf; hold on;
+ax = zeros(4, 1);
+ax(1) = 0; ax(2) = 80; ax(3) = 0; ax(4) = 20;
+axis(ax);
+fs = 30; %font size
+xlabel("Iteration", "fontsize", fs);
+ylabel("Body lengths/minute", "fontsize", fs);
+
+leg = cell(1, (((length(files))/3)-1));%set to correct size
+
+for ii = 0:(((length(files))/3)-1) %set to correct size
+  a = files{(ii * 3) + 1};
+  b = files{(ii * 3) + 2};
+  c = files{(ii * 3) + 3};
+  [aN,aiteration,aparameters,afitness,abest] = load_run_data(a);
+  [bN,biteration,bparameters,bfitness,bbest] = load_run_data(b);
+  [cN,citeration,cparameters,cfitness,cbest] = load_run_data(c);
+  
+  leg{1, ii+1} = strrep (a, "_", "\\_");
+  
+  %find shortest
+  N = min(aN, bN);
+  N = min(N, cN);
+  iteration = zeros(0, 0);
+  if N == length(aiteration)
+    iteration = aiteration;
+  elseif N == length(biteration)
+    iteration = biteration;
+  else
+    iteration = citeration;
+  endif
+  
+  %find mean
+  means = zeros(N, 1);
+  stdev = zeros(N, 1);
+  x = zeros(N, 1);
+  for i = 1:N
+    v = zeros(3, 1);
+    v(1) = abest(i);
+    v(2) = bbest(i);
+    v(3) = cbest(i);
+    means(i) = (abest(i) + bbest(i) + cbest(i)) / 3;
+    stdev(i) = std(v) / sqrt(3);
+    x(i) = i + (.03 * ii);
+  end
+  
+  %plot(iteration, best, '-', 'color', clr{mod(ii,6)+1});
+  e = errorbar(x, means, stdev);
+  set(e, "color", clr{mod(ii,6)+1});
+
+  %e = plot(x, means, 'linewidth', 2);
+  %set(e, "color", clr{mod(ii,6)+1});
[…]

### 2011-01-26 paper/platformDetails.tex
added robot_close to platform details
edited platform details

@@
 % data and any pre-processing you did to it.
-\acmFig{quadratot.jpg}{1}{The robot}
+\acmFig{robot_close.jpg}{1}{The robot}
 \acmFig{topdown.png}{1}{A figure of the robot from a top-down perspective,
@@
 
-A Wii remote tracks the location of the robot, providing feedback
-about distance traveled, through an infrared LED mounted on top of 
+A Nintendo Wii remote tracks the location of the robot, providing feedback
+about distance traveled, in pixels, through an infrared LED mounted on top of 
 the robot. A server is run on the robot and continuously tracks its
@@
 via bluetooth. A client then connects via a socket to the tracking
-server and requests position updates periodically. If the robot moves 
-beyond the viewable range of the Wii remote, the system pauses and
-directs the user to move the robot back within the viewable range 
-before running again.
+server and requests position updates periodically. If the robot walks 
+beyond the viewable range of the Wii remote, a prompt informed the
+user. The only human intervention required during the
+experiment was to move the robot back inside the viewable area and resume
+the run, and to handle any mechnanical failures that arose. This did not
+interrupt the learning process or result in the loss of data.
 

### 2011-01-26 paper/fitnessEvaluation.tex
Fixed minor bug
Wrap
took out hardware stuff and minor edits

@@
 
-We controlled the experiments from a computer that was connected via a wireless Ethernet to the robot.
-An infrared LED was mounted to the antenna on top of the robot and a Wii remote was attached to the ceiling.
-The Wii remote tracked the location of the LED.
-The robot tracked its position via Bluetooth by using the CWiid library\cite{cwiid} to communicate with the Wii remote.
-If the robot walked outside of the Wii remote’s viewable area, a prompt informed the experimenter.
-The only human intervention required during an experiment was to move the robot back inside the viewable area and resume the run.
-This did not interrupt the learning process or result in the loss of data.
+The metric for evaluation of the designed gait was speed. To evaluate
+a set of parameters, the robot was sent the parameters and instructed
+to walk for a certain length of time. For each evaluation, the robot
+always started and ended in the same position in order to measure true
+displacement and not reward gaits the ended in a lean, since the LED
+would have moved a different distance the robot. More efficient
+parameters resulted in a faster gait, which translated into a longer
+distance walked, measured in pixels, and a better score.
 
-The metric for evaluation of the designed gait was speed.
-To evaluate a set of parameters, the robot was sent the parameters and instructed to walk for a certain length of time.
-For each evaluation, the robot always started and ended in the same position in order to measure true displacement and not reward gaits the ended in a lean, since the LED would have moved a different distance the robot.
-More efficient parameters resulted in a faster gait, which translated into a longer distance walked and a better score.
-
-Each of the parameter ???
-methods was run on 3 different initial parameter vectors, in order to allow for the fair comparison of the algorithms.
-We allowed each run for the parameter ???
-methods to continue until the results plateaued (no improvement for one third of the policies seen so far).
-Three runs of HyperNEAT were completed, each with a different initial seed and run for 20 generations.
+Each of the parameter ??? methods was run on 3 different initial
+parameter vectors, in order to fairly compare the algorithms. We allowed 
+each run for the parameter ??? methods to
+continue until the results plateaued (no improvement for one third of
+the policies seen so far). Three runs of HyperNEAT were completed,
+each with a different initial seed and run for 20 generations.

### 2011-01-26 paper/fitnessEvaluation.tex
Added sentence from other section

--- 2011-01-26 paper/fitnessEvaluation.tex
@@
 \seclabel{fitnessEvaluation}
+
+\edit{Put this in there somewhere:}
+
+To be as fair as possible when obtaining the results presented in
+\secref{results}, we picked three random points --- $\vec{\theta}_A$,
+$\vec{\theta}_B$, and $\vec{\theta}_C$ --- and started all seven
+methods at these points...
 
@@
 
-We controlled the experiments from a computer that was connected via a wireless Ethernet to the robot.
-An infrared LED was mounted to the antenna on top of the robot and a Wii remote was attached to the ceiling.
-The Wii remote tracked the location of the LED.
-The robot tracked its position via Bluetooth by using the CWiid library\cite{cwiid} to communicate with the Wii remote.
-If the robot walked outside of the Wii remote’s viewable area, a prompt informed the experimenter.
-The only human intervention required during an experiment was to move the robot back inside the viewable area and resume the run.
-This did not interrupt the learning process or result in the loss of data.
+We controlled the experiments from a computer that was connected via a
+wireless Ethernet to the robot. An infrared LED was mounted to the
+antenna on top of the robot and a Wii remote was attached to the
+ceiling. The Wii remote tracked the location of the LED. The robot
+tracked its position via Bluetooth by using the CWiid
+library\cite{cwiid} to communicate with the Wii remote. If the robot
+walked outside of the Wii remote's viewable area, a prompt informed
+the experimenter. The only human intervention required during an
+experiment was to move the robot back inside the viewable area and
+resume the run. This did not interrupt the learning process or result
+in the loss of data.
 
-The metric for evaluation of the designed gait was speed.
-To evaluate a set of parameters, the robot was sent the parameters and instructed to walk for a certain length of time.
-For each evaluation, the robot always started and ended in the same position in order to measure true displacement and not reward gaits the ended in a lean, since the LED would have moved a different distance the robot.
-More efficient parameters resulted in a faster gait, which translated into a longer distance walked and a better score.
+The metric for evaluation of the designed gait was speed. To evaluate
+a set of parameters, the robot was sent the parameters and instructed
+to walk for a certain length of time. For each evaluation, the robot
+always started and ended in the same position in order to measure true
+displacement and not reward gaits the ended in a lean, since the LED
+would have moved a different distance the robot. More efficient
+parameters resulted in a faster gait, which translated into a longer
+distance walked and a better score.
 
-Each of the parameter ???
-methods was run on 3 different initial parameter vectors, in order to allow for the fair comparison of the algorithms.
-We allowed each run for the parameter ???
-methods to continue until the results plateaued (no improvement for one third of the policies seen so far).
-Three runs of HyperNEAT were completed, each with a different initial seed and run for 20 generations.
+Each of the parameter ??? methods was run on 3 different initial
+parameter vectors, in order to allow for the fair comparison of the
+algorithms. We allowed each run for the parameter ??? methods to
+continue until the results plateaued (no improvement for one third of
+the policies seen so far). Three runs of HyperNEAT were completed,
+each with a different initial seed and run for 20 generations.

### 2011-01-26 paper/platformDetails.tex
added sentence about the wireless ethernet
Made wii/LEd light photos smaller, put them together.

@@
 
-The quadruped robot has an on-board computer running Linux. The lower
+The quadruped robot (\figref{robot_close.jpg}) has an on-board computer running Linux. The lower
 level drivers are in C and the system is implemented in
@@
 
-\acmFig{robot.JPG}{1}{The infrared LED mounted on top of the robot, which
-provides feedback about distance traveled in conjunction with a Wii
-remote}
-\acmFig{wiiMote.JPG}{1}{A Wii remote tracks the location of the robot,
-providing feedback about distance traveled}
+\acmFigg{wiiMote.JPG}{robot.JPG}{1}{A Nintendo Wii remote tracks the location
+of the robot, providing feedback about distance traveled, in pixels, through
+an infrared LED mounted on top of the robot.}
 
-A Nintendo Wii remote tracks the location of the robot, providing feedback
+We controlled the experiments from a computer that was connected via a
+wireless Ethernet to the robot. A Nintendo Wii remote tracks the location 
+of the robot, providing feedback
 about distance traveled, in pixels, through an infrared LED mounted on top of 
-the robot. A server is run on the robot and continuously tracks its
+the robot (\figref{wiiMote.JPG}). A server is run on the robot and continuously tracks its
 position using the CWiid library\cite{cwiid} to interface with the remote

### 2011-01-26 paper/learningMethods.tex
Wrap
Reworked a bit of learningMethods

@@
 
-For both of these reasons, an intelligent learning algorithm is
+For both of these reasons, an intelligent \emph{learning algorithm} is
 needed.
+In this context, a learning algorithm is a method for choosing
+%\begin{enumerate}
+%\item The value for $\vec{\theta}$ for the initial trial, and
+the next value of $\vec{\theta}$ to try, given a list of the
+$\vec{\theta}$ values tried so far and the fitness $f(\vec{\theta})$
+measured for each.
+%\end{enumerate}
 
-In this context, by learning algorithm we mean a method for choosing:
+We evaluated a total of seven subtly or not so subtly different
+learning methods for the parametrized motion models.  All methods
+needed a way of picking a $\vec{\theta}$ for the first trial.  In
+general this is done by selecting a random starting vector within the
+allowable domain.  However, as discussed in
+\secref{fitnessEvaluation}, we decided to evaluate the different
+methods starting at one of several common starting points to reduce
+one possible contribution of noise.
 
-\begin{enumerate}
-\item The value for $\vec{\theta}$ for the initial trial, and
-\item The next value of $\vec{\theta}$ to try, given the
-  $\vec{\theta}$ and $f(\vec{\theta})$ for all previous trials.
-\end{enumerate}
-
-We evaluated a total of eight subtly or not so subtly different
-learning methods.  All employed a simple random sampling method for
-requirement (1); that is, all methods picked their initial
-$\vec{\theta}$ value via uniform random sampling within the allowed
-parameter ranges. \edit{make sure this is true. Isn't really, should say that we started with A, B, C}.  Thus, the
-differences in the algorithms was how they selected new $\vec{\theta}$
-values to try from their past experience.
+%All employed a simple random sampling method for
+%requirement (1); that is, all methods picked their initial
+%$\vec{\theta}$ value via uniform random sampling within the allowed
+%parameter ranges. \edit{make sure this is true. Isn't really, should say that we started with A, B, C}.  Thus, the
+%differences in the algorithms was how they selected new $\vec{\theta}$
+%values to try from their past experience.
 
@@
 
-  HyperNEAT is an indirect encoding for evolving artificial neural networks (ANNs) that is inspired by the way natural organisms develop~\cite{stanley2009hypercube}.
-It evolves Compositional Pattern Producing Networks (CPPNs)~\cite{stanley2007compositional}, each of which is a genome that encodes an ANN phenotype~\cite{stanley2009hypercube}.
-Each CPPN is itself a directed graph, where each node is a mathematical function, such as sine or Gaussian.
-The nature of these functions can facilitate the evolution of properties such as symmetry (e.g., an absolute value or Gaussian function) and repetition (e.g., a sine function)~\cite{stanley2009hypercube, stanley2007compositional}.
-The signal on each link in the CPPN is multiplied by that link's weight, which can alter its effect.
+HyperNEAT is an indirect encoding for evolving artificial neural
+networks (ANNs) that is inspired by the way natural organisms
+develop~\cite{stanley2009hypercube}. It evolves Compositional Pattern
+Producing Networks (CPPNs)~\cite{stanley2007compositional}, each of
+which is a genome that encodes an ANN
+phenotype~\cite{stanley2009hypercube}. Each CPPN is itself a directed
+graph, where each node is a mathematical function, such as sine or
+Gaussian. The nature of these functions can facilitate the evolution
+of properties such as symmetry (e.g., an absolute value or Gaussian
+function) and repetition (e.g., a sine
+function)~\cite{stanley2009hypercube, stanley2007compositional}. The
+signal on each link in the CPPN is multiplied by that link's weight,
+which can alter its effect.
 
-A CPPN is queried once for each link in the ANN substrate to determine that link's weight.
-The inputs to the CPPN are the Cartesian coordinates of both the source (e.g., $x = 2$, $y = 4$) and target (e.g., $x = 3$, $y = 5$) nodes of a link and a constant bias value.
-The CPPN takes these five values as inputs and produces one or two output values, depending on the substrate topology.
-If there is no hidden layer in the substrate, the single output is the weight of the link between a source node on the input layer and a target node on the output layer.
-If there is a hidden layer, the first output value determines the weight of the link between the associated input (source) and hidden layer (target) nodes, and the second output value determines the weight of the link between the associated hidden (source) and output (target) layer nodes.
-All pairwise combinations of source and target nodes are iteratively passed as inputs to a CPPN to determine the weight of each substrate link.
+A CPPN is queried once for each link in the ANN substrate to determine
+that link's weight. The inputs to the CPPN are the Cartesian
+coordinates of both the source (e.g., $x = 2$, $y = 4$) and target
+(e.g., $x = 3$, $y = 5$) nodes of a link and a constant bias
+value. The CPPN takes these five values as inputs and produces one or
+two output values, depending on the substrate topology. If there is no
+hidden layer in the substrate, the single output is the weight of the
+link between a source node on the input layer and a target node on the
+output layer. If there is a hidden layer, the first output value
+determines the weight of the link between the associated input
+(source) and hidden layer (target) nodes, and the second output value
+determines the weight of the link between the associated hidden
+(source) and output (target) layer nodes. All pairwise combinations of
+source and target nodes are iteratively passed as inputs to a CPPN to
+determine the weight of each substrate link.
 
-\edit{following paragraph is figure caption for file named: hyperneatExplanation.png}
-HyperNEAT Produces ANNs from CPPNs.
-ANN weights are specified as a function of the geometric coordinates of the source node and the target node for each connection.
-The coordinates of these nodes and a constant bias are iteratively passed to the CPPN to determine each connection weight.
-The CPPN has two output values, which specify the weights for each connection layer as shown.
+\edit{following paragraph is figure caption for file named:
+  hyperneatExplanation.png} HyperNEAT Produces ANNs from CPPNs. ANN
+weights are specified as a function of the geometric coordinates of
+the source node and the target node for each connection. The
+coordinates of these nodes and a constant bias are iteratively passed
+to the CPPN to determine each connection weight. The CPPN has two
+output values, which specify the weights for each connection layer as
+shown.
 
-HyperNEAT is capable of exploiting the geometry of a problem: because the link values between nodes in the final ANN substrate are a function of the geometric positions of those nodes, HyperNEAT can exploit such information when solving a problem~\cite{stanley2009hypercube, clune2009sensitivity, clune2011performance}.
-In the case of quadruped locomotion, this property helped Hyper-NEAT produce gaits with front-back, left-right, and four-way symmetries~\cite{clune2009evolving, clune2011performance}.
+HyperNEAT is capable of exploiting the geometry of a problem: because
+the link values between nodes in the final ANN substrate are a
+function of the geometric positions of those nodes, HyperNEAT can
+exploit such information when solving a
+problem~\cite{stanley2009hypercube, clune2009sensitivity,
+  clune2011performance}. In the case of quadruped locomotion, this
+property helped Hyper-NEAT produce gaits with front-back, left-right,
+and four-way symmetries~\cite{clune2009evolving,
+  clune2011performance}.
 
-The evolution of the population of CPPNs occurs according to the principles of the NeuroEvolution of Augmenting Topologies (NEAT) algorithm~\cite{stanley2002evolving}, which was originally designed to evolve ANNs.
-NEAT can be fruitfully applied to CPPNs because of their structural similarity to ANNs.
-For example, mutations can add a node, and thus a function, to a CPPN graph, or change its link weights.
-The NEAT algorithm is unique in three main ways~\cite{stanley2002evolving}.
-Initially, it starts with small genomes that encode simple networks and slowly complexifies them via mutations that add nodes and links to the network, enabling the algorithm to evolve the network topology in addition to its weights.
-Secondly, NEAT has a fitness sharing mechanism that preserves diversity in the system and gives time for new innovations to be tuned by evolution before competing them against more adapted rivals.
-Finally, NEAT tracks historical information to perform intelligent crossover while avoiding the need for expensive topological analysis.
-A full explanation of NEAT can be found in~\cite{stanley2002evolving}].
+The evolution of the population of CPPNs occurs according to the
+principles of the NeuroEvolution of Augmenting Topologies (NEAT)
+algorithm~\cite{stanley2002evolving}, which was originally designed to
+evolve ANNs. NEAT can be fruitfully applied to CPPNs because of their
+structural similarity to ANNs. For example, mutations can add a node,
+and thus a function, to a CPPN graph, or change its link weights. The
+NEAT algorithm is unique in three main
+ways~\cite{stanley2002evolving}. Initially, it starts with small
+genomes that encode simple networks and slowly complexifies them via
+mutations that add nodes and links to the network, enabling the
+algorithm to evolve the network topology in addition to its
+weights. Secondly, NEAT has a fitness sharing mechanism that preserves
+diversity in the system and gives time for new innovations to be tuned
+by evolution before competing them against more adapted
+rivals. Finally, NEAT tracks historical information to perform
+intelligent crossover while avoiding the need for expensive
+topological analysis. A full explanation of NEAT can be found
+in~\cite{stanley2002evolving}].
 
-The ANN configuration follows previous studies that evolved quadruped gaits with HyperNEAT in simulation~\cite{clune2011performance, clune2009evolving}, but was adapted to accommodate differences in the robot.
-Specifically, the ANNs consists of three $3 \times 4$ Cartesian grids of nodes forming input, hidden, and output layers.
-Adjacent layers were completely connected, meaning that there were $(3 \times 4)^2= 288$ links in each substrate.
-The inputs to the substrate were the current angles of each of the 9 joints of the robot (described below) and a sine and cosine wave (to facilitate the production of periodic behaviors).
-The sine and cosine waves were the following function of time (t) in milli-seconds: $sin(t/12.0)\pi$.
-This function produces numbers in the range [$-\pi, \pi$], which were then JASON FILL ME IN TO DESCRIBE HOW JOINT COMMANDS WERE SPECIFIED.
+The ANN configuration follows previous studies that evolved quadruped
+gaits with HyperNEAT in simulation~\cite{clune2011performance,
+  clune2009evolving}, but was adapted to accommodate differences in
+the robot. Specifically, the ANNs consists of three $3 \times 4$
+Cartesian grids of nodes forming input, hidden, and output
+layers. Adjacent layers were completely connected, meaning that there
+were $(3 \times 4)^2= 288$ links in each substrate. The inputs to the
+substrate were the current angles of each of the 9 joints of the robot
+(described below) and a sine and cosine wave (to facilitate the
+production of periodic behaviors). The sine and cosine waves were the
+following function of time (t) in milli-seconds:
+$sin(t/12.0)\pi$. This function produces numbers in the range [$-\pi,
+  \pi$], which were then JASON FILL ME IN TO DESCRIBE HOW JOINT
+COMMANDS WERE SPECIFIED.
 
@@
 
-ANN Configuration for HyperNEAT Runs.
-The first two columns of each row of the input layer receive information about a single leg (the current angle of its two joints).
-The final column provides a sine and cosine wave to enable periodic movements and the angle of the center joint.
-Evolution determines the function of the hidden-layer nodes.
-The nodes in the output layer specify new joint angles for each respective joint.
-The unlabeled node in the input and output layer is ignored.
+ANN Configuration for HyperNEAT Runs. The first two columns of each
+row of the input layer receive information about a single leg (the
+current angle of its two joints). The final column provides a sine and
+cosine wave to enable periodic movements and the angle of the center
+joint. Evolution determines the function of the hidden-layer
+nodes. The nodes in the output layer specify new joint angles for each
+respective joint. The unlabeled node in the input and output layer is
+ignored.
 
@@
 
-The population size for HyperNEAT was 9 and runs lasted 20 generations.
-These numbers are small, but were necesary given how much time it took to conduct evolution directly on a real robot.
-The remaining parameters were identical to Clune et al.
-~\cite{clune2011performance}.
+The population size for HyperNEAT was 9 and runs lasted 20
+generations. These numbers are small, but were necesary given how much
+time it took to conduct evolution directly on a real robot. The
+remaining parameters were identical to Clune et
+al. ~\cite{clune2011performance}.
 

### 2011-01-26 code/plot/plot_error_new.m
plot_error_new update

@@ -1,14 +1,5 @@
 function plot_error_new(files, saveFileName)
 
-clr = {};
-  clr{end+1} = 'k';
-  clr{end+1} = 'r';
-  clr{end+1} = 'g';
-  clr{end+1} = 'b';
-  clr{end+1} = 'c';
-  clr{end+1} = 'm';
-%  clr{end+1} = 'y';
-
 figure(2); clf; hold on;
 ax = zeros(4, 1);
 ax(1) = 0; ax(2) = 80; ax(3) = 0; ax(4) = 20;
@@ -17,9 +8,11 @@ fs = 30; %font size
 xlabel("Iteration", "fontsize", fs);
 ylabel("Body lengths/minute", "fontsize", fs);
 
-leg = cell(1, (((length(files))/3)-1));%set to correct size
+leg = cell(1, (((length(files))/3)-1));
 
-for ii = 0:(((length(files))/3)-1) %set to correct size
+frac = 5;
+
+for ii = 0:(((length(files))/3)-1)
   a = files{(ii * 3) + 1};
   b = files{(ii * 3) + 2};
   c = files{(ii * 3) + 3};
@@ -27,9 +20,11 @@ for ii = 0:(((length(files))/3)-1) %set to correct size
   [bN,biteration,bparameters,bfitness,bbest] = load_run_data(b);
   [cN,citeration,cparameters,cfitness,cbest] = load_run_data(c);
   
-  leg{1, ii+1} = strrep (a, "_", "\\_");
+  legName = strrep (a, "_", "\\_");
+  legName = substr(legName, 6, (length(legName) - 12));
+  leg{1, ii+1} = legName;
   
-  %find shortest
+  % find shortest
   N = min(aN, bN);
   N = min(N, cN);
   iteration = zeros(0, 0);
@@ -40,8 +35,9 @@ for ii = 0:(((length(files))/3)-1) %set to correct size
   else
     iteration = citeration;
   endif
+  % iteration now = 1:N
   
-  %find mean
+  % find mean
   means = zeros(N, 1);
   stdev = zeros(N, 1);
   x = zeros(N, 1);
@@ -55,9 +51,27 @@ for ii = 0:(((length(files))/3)-1) %set to correct size
     x(i) = i + (.03 * ii);
   end
   
+  
+  
+  % get every frac data point and the error for those points
+  someMeans = zeros(N/frac, 1);
+  someX = zeros(N/frac, 1);
[…]

### 2011-01-26 paper/introduction.tex

--- 2011-01-20 paper/introduction.tex
@@
 \section{Introduction and Background}
+
+\acmFig{robot_close.jpg}{1}{The 3D printed quadruped robot used for
+  this study.}
 
@@
 
-\editbox{Outline \\
- - Why gait learning is cool \\
- - Gait learning has been done before, in simulation and on hardware (cite, cite, cite) \\
- - HyperNEAT has been used to evolve gaits before in sim (cite, ...) \\
- - Here, we do two cool things for the first time: \\
- - - evolve HN in hardware \\
- - - compare it to other learning methods on the same hardware.}
+%\editbox{Outline \\
+% - Why gait learning is cool \\
+% - Gait learning has been done before, in simulation and on hardware (cite, cite, cite) \\
+% - HyperNEAT has been used to evolve gaits before in simulation~{clune2009evolving, clune2011performance, clune2009hybrid}, and it worked well in part because the gaits were regular, raising the question of whether this high performance was just because the gaits were regular?
+We test this hypothesis by comparing HyperNEAT to traditional machine learning methods biased towards regular gaits.
+\\
+% - Here, we do two cool things for the first time: \\
+% - - evolve HN in hardware \\
+% - - compare it to other learning methods on the same hardware.}
 

### 2011-01-26 paper/platformDetails.tex
Moved robot figure to intro
added information about motors, printing, length and weight

@@
 % data and any pre-processing you did to it.
-\acmFig{robot_close.jpg}{1}{The robot}
+
 \acmFig{topdown.png}{1}{A figure of the robot from a top-down perspective,
-with motors labeled}
+with servos labeled}
 
-The quadruped robot (\figref{robot_close.jpg}) has an on-board computer running Linux. The lower
-level drivers are in C and the system is implemented in
-Python.
+The quadruped robot was printed on the Objet
+Connex 500 3-D Printing System. It weights 1884 g and measures approximately
+15 centimeters from leg to leg in the crouch position depicted in
+\figref{robot_close.jpg}. There are a total of 9 AX-12+ Dynamixel
+Servomotors: 4 outer leg joint
+servos, 4 inner leg joint servos, and a final servo at the center "hip" of the
+robot (displayed in \figref{topdown.png}). The robot has an on-board computer running Linux. 
+The lower level drivers are in C and the system is implemented in Python.
 

### 2011-01-26 paper/platformDetails.tex
minor edit

@@
 \figref{robot_close.jpg}. There are a total of 9 AX-12+ Dynamixel
-Servomotors: 4 outer leg joint
+Servomotors, displayed in \figref{topdown.png}: 4 outer leg joint
 servos, 4 inner leg joint servos, and a final servo at the center "hip" of the
-robot (displayed in \figref{topdown.png}). The robot has an on-board computer running Linux. 
+robot. The robot has an on-board computer running Linux. 
 The lower level drivers are in C and the system is implemented in Python.

### 2011-01-26 paper/platformDetails.tex
Added info about min and max servo positions

@@
 The servos are prevented from being commanded to a point outside their
-normal range (0 - 1023) as well as beyond points where limbs would collide.
+normal range (0 - 1023) as well as beyond points where limbs would collide
+(150 - 770 for inner leg servos, 30 - 680 for outer leg servos, and 392 - 623
+for the center, or "hip" servo.

### 2011-01-27 code/plot/plot_error_new.m
update to plot_error_new. legend is not correct. colors may have to change depending on the order of the methods.

@@ -1,5 +1,21 @@
 function plot_error_new(files, saveFileName)
 
+mark = {};
+  mark{end+1} = '+';
+  mark{end+1} = '+';
+  mark{end+1} = '*';
+  mark{end+1} = 'o';
+  mark{end+1} = 'x';
+  mark{end+1} = 'x';
+
+clr = {};
+  clr{end+1} = 'r';
+  clr{end+1} = 'k';
+  clr{end+1} = 'b';
+  clr{end+1} = 'b';
+  clr{end+1} = 'k';
+  clr{end+1} = 'r';
+  
 figure(2); clf; hold on;
 ax = zeros(4, 1);
 ax(1) = 0; ax(2) = 80; ax(3) = 0; ax(4) = 20;
@@ -8,7 +24,8 @@ fs = 30; %font size
 xlabel("Iteration", "fontsize", fs);
 ylabel("Body lengths/minute", "fontsize", fs);
 
-leg = cell(1, (((length(files))/3)-1));
+%leg = cell(1, (((length(files))/3)-1));
+leg = cell(1, (length(files)-6));
 
 frac = 5;
 
@@ -22,7 +39,10 @@ for ii = 0:(((length(files))/3)-1)
   
   legName = strrep (a, "_", "\\_");
   legName = substr(legName, 6, (length(legName) - 12));
-  leg{1, ii+1} = legName;
+  %leg{1, ii+1} = legName;
+  leg{1, ii*2+1} = legName;
+  leg{1, ii*2+2} = '';
+  %leg{1, ii*2+3} = '';
   
   % find shortest
   N = min(aN, bN);
@@ -58,15 +78,16 @@ for ii = 0:(((length(files))/3)-1)
   someX = zeros(N/frac, 1);
   someSD = zeros(N/frac, 1);
   for i = 1:(N/frac)
-    someMeans(i) = means(i*frac);
-    someX(i) = x(i*frac);
-    someSD(i) = stdev(i*frac);
+    offset = mod(ii,3) - 1;
+    someMeans(i) = means(i*frac + offset);
+    someX(i) = x(i*frac + offset);
+    someSD(i) = stdev(i*frac + offset);
   end
   
-  %plot(someX, someMeans, 'o');
-  
-  e = errorbar(someX, someMeans, someSD);
-  plot(means);
+ % plot(someX, someMeans, mark{mod(ii,6)+1}, 'color', clr{mod(ii,6)+1});
[…]

### 2011-01-27 paper/fitnessEvaluation.tex

--- 2011-01-26 paper/fitnessEvaluation.tex
@@
 
-\editbox{proofread.}
+\edit{Put this in there somewhere:}
+
+To be as fair as possible when obtaining the results presented in
+\secref{results}, we picked three random points --- $\vec{\theta}_A$,
+$\vec{\theta}_B$, and $\vec{\theta}_C$ --- and started all seven
+methods at these points...
+
+\editbox{proofread. especially middle paragraph}
 
@@
 
+The size of the Wii remote window is approximately 69 x 47 cm. A
+concern about this is that if the robot walks further than usual,
+it walks outside the viewable range and thus does not count the run.
+This possibility could bias the final results reported on the robot.
+% probably could be better written
+
 Each of the parameter ??? methods was run on 3 different initial

### 2011-01-31 code/plot/plot_error_new.m
Changed a bit with Diana

@@ -27,7 +27,7 @@ ylabel("Body lengths/minute", "fontsize", fs);
 %leg = cell(1, (((length(files))/3)-1));
 leg = cell(1, (length(files)-6));
 
-frac = 5;
+frac = 10;
 
 for ii = 0:(((length(files))/3)-1)
   a = files{(ii * 3) + 1};
@@ -79,8 +79,10 @@ for ii = 0:(((length(files))/3)-1)
   someSD = zeros((N/frac) - 1, 1);
   for i = 1:((N/frac)-1)
     offset = mod(ii,3) - 1;
+    %offset = 0;
     someMeans(i) = means(i*frac + offset);
     someX(i) = x(i*frac + offset);
+    %someX(i) = x(i*frac + offset) + ii/10.;
     someSD(i) = stdev(i*frac + offset);
   end
   

### 2011-01-31 code/plot/error_plot.m
error_plot.m added to run plot_error_new

@@ -1,0 +1,3 @@
+function error_plot()
+
+plot_error_new({'log_uniform_a.txt', 'log_uniform_b.txt', 'log_uniform_c.txt', 'log_gaussian_a.txt', 'log_gaussian_b.txt', 'log_gaussian_c.txt', 'log_gradient_a.txt', 'log_gradient_b.txt', 'log_gradient_c.txt', 'log_random_a.txt', 'log_random_b.txt', 'log_random_c.txt',  'log_linear_regression_a.txt', 'log_linear_regression_b.txt', 'log_linear_regression_c.txt', 'log_simplex_a.txt', 'log_simplex_b.txt', 'log_simplex_c.txt'}, 'name')

### 2011-01-31 code/plot/plot_error_new.m

@@ -12,9 +12,9 @@ clr = {};
   clr{end+1} = 'r';
   clr{end+1} = 'k';
   clr{end+1} = 'b';
-  clr{end+1} = 'b';
-  clr{end+1} = 'k';
-  clr{end+1} = 'r';
+  clr{end+1} = [0.5, 0.75, 0.15];
+  clr{end+1} = [.3, 0, .5];
+  clr{end+1} = [1 128/255, 0];
   
 figure(2); clf; hold on;
 ax = zeros(4, 1);
@@ -23,11 +23,12 @@ axis(ax);
 fs = 30; %font size
 xlabel("Iteration", "fontsize", fs);
 ylabel("Body lengths/minute", "fontsize", fs);
+set(gcf, 'DefaultLineLineWidth', 2.5);
 
 %leg = cell(1, (((length(files))/3)-1));
 leg = cell(1, (length(files)-6));
 
-frac = 10;
+frac = 15;
 
 for ii = 0:(((length(files))/3)-1)
   a = files{(ii * 3) + 1};
@@ -78,7 +79,8 @@ for ii = 0:(((length(files))/3)-1)
   someX = zeros((N/frac) - 1, 1);
   someSD = zeros((N/frac) - 1, 1);
   for i = 1:((N/frac)-1)
-    offset = mod(ii,3) - 1;
+    offset = ii*3;
+    %offset = mod(ii,3) - 1;
     %offset = 0;
     someMeans(i) = means(i*frac + offset);
     someX(i) = x(i*frac + offset);
@@ -88,6 +90,7 @@ for ii = 0:(((length(files))/3)-1)
   
  % plot(someX, someMeans, mark{mod(ii,6)+1}, 'color', clr{mod(ii,6)+1});
   e = errorbar(someX, someMeans, someSD, '~');
+  %set(e, 'linestyle', 'none');
   set(e, 'color', clr{mod(ii,6)+1});
   plot(means, 'color', clr{mod(ii,6)+1});
   

### 2011-01-31 code/plot/plot_error_new.m
plot_error_new 2 error bars/line

@@ -29,6 +29,7 @@ set(gcf, 'DefaultLineLineWidth', 2.5);
 leg = cell(1, (length(files)-6));
 
 frac = 15;
+markSize = 11;
 
 for ii = 0:(((length(files))/3)-1)
   a = files{(ii * 3) + 1};
@@ -75,23 +76,27 @@ for ii = 0:(((length(files))/3)-1)
   
   
   % get every frac data point and the error for those points
-  someMeans = zeros((N/frac) - 1, 1);
-  someX = zeros((N/frac) - 1, 1);
-  someSD = zeros((N/frac) - 1, 1);
-  for i = 1:((N/frac)-1)
-    offset = ii*3;
-    %offset = mod(ii,3) - 1;
-    %offset = 0;
-    someMeans(i) = means(i*frac + offset);
-    someX(i) = x(i*frac + offset);
-    %someX(i) = x(i*frac + offset) + ii/10.;
-    someSD(i) = stdev(i*frac + offset);
+  fracLen = 2;
+  someMeans = zeros(fracLen, 1);
+  someX = zeros(fracLen, 1);
+  someSD = zeros(fracLen, 1);
+  for i = 1:fracLen
+    ind = round(i * N/2);
+    if (ii == 2 && i == 1)
+      ind = ind - 2;
+    elseif (ii==2)
+      ind = ind -1;
+    end
+    someMeans(i) = means(ind);
+    someX(i) = x(ind);
+    %someX(i) = x(ind);
+    someSD(i) = stdev(ind);
   end
   
  % plot(someX, someMeans, mark{mod(ii,6)+1}, 'color', clr{mod(ii,6)+1});
   e = errorbar(someX, someMeans, someSD, '~');
   %set(e, 'linestyle', 'none');
-  set(e, 'color', clr{mod(ii,6)+1});
+  set(e, 'color', clr{mod(ii,6)+1}, 'markerSize', markSize);
   plot(means, 'color', clr{mod(ii,6)+1});
   
   %plot(iteration, best, '-', 'color', clr{mod(ii,6)+1});

### 2011-01-31 paper/learningMethods.tex
Many edits...

@@
 
-\edit{this might no longer flow...fix} There are a plethora of methods
-available for the creation of a vector motion generating function
-$g(t)$, and we've chosen two for comparison in this study.
+There are a plethora of methods available for the creation of the
+vector motion generating function $g(t)$ introduced in
+\secref{problemDefinition}, and we've chosen two for comparison in
+this study.  The first method we evaluated was to use a family of
+parametrized gaits, described in this section and the next, and the
+second method was to use an evolved non-parametric gait, detailed in
+\secref{hyperNeatMethod}.
 
-The first method is to choose a family of \emph{parametrized
-  functions}.  By fixing the parameters to a particular set of values,
-we obtain a deterministic motion function over time.  We tried several
-parametrizations on the robot and, upon obtaining reasonable early
-success, settled on one particular parametrization.  We dubbed this
-the \emph{SineModel5}, as it has five parameters and employs a sine
-wave as its root pattern.
+By a \emph{parametrized gait}, we mean a gait produced by a
+parametrized function $g(t; \vec{\theta})$. Fixing the parameter
+$\vec{\theta}$ yields a deterministic motion function over time.  We
+tried several parametrizations on the robot and, upon obtaining
+reasonable early success, settled on one particular parametrization.
+We dubbed this the \emph{SineModel5}, as it has five parameters and
+employs a sine wave as its root pattern.
 
 The five parameters in $\vec{\theta}$ are as follows:
+
+\edit{make sure this text flows into the table}
 
@@
 
-\begin{table}[h]
+\begin{table}[h!]
 \begin{center}
@@
 amplitude $\amp$ and period $\tau$, multiplies the waves for all inner
-motors by multIO, multiplies the waves for all front motors by multFB,
-and multiplies the waves for all left motors by multLR.  To obtain the
+motors by $m_o$, multiplies the waves for all front motors by $m_F$,
+and multiplies the waves for all left motors by $m_R$.  To obtain the
 actual motor position commands, these waves are offset by
@@
 waves are at 0) is approximately a crouch (the position shown in
-\figref{crouchingRobot}).  Finally, in this motion model, the ninth
-"hip" motor is kept at a neutral position.  Thus, the commanded
+\figref{robot_close.jpg}).  Finally, in this motion model, the ninth
+``hip'' motor is kept at a neutral position.  Thus, the commanded
 position for all motors, as a vector function of time, is:
@@
 position are $C_I$ for inner motors, $C_O$ for outer motors, and $C_C$
-for the center hip motor.
-
-\edit{make figure to go here (or another section), show drawing with
-  inner/outer, l/r, and f/b motors.}
-
-The nine subscripted functions correspond to the nine labeled motors
-in \figref{robotDiagram}
-
-% used is a function parametrized by five numbers, shown in \figref{foo}
+for the center hip motor.  The nine subscripted functions correspond
+to the nine labeled motors in \figref{topdown.png}.
 
@@
 
-The eight methods used, and their methods for choosing the next
+The seven methods used, and their methods for choosing the next
 $\vec{\theta}$ are:
-
-\edit{write this.  Old quadratot Method section below alsdkfj }
 
@@
 
-\edit{JMC:Bullets=narrower column=wasted space. subsubsection instead?}
+%\edit{JMC:Bullets=narrower column=wasted space. subsubsection instead?}
 
@@
 \item \emph{Random}: This method randomly generates parameter vectors
-  in the allowable range for every trial. This strategy was used only as baseline for comparison.
+  in the allowable range for every trial. This strategy was used only
+  as baseline for comparison.
 
@@
 
-\item \emph{Neuroevolution (Evolving Artificial Neural Networks with HyperNEAT)}:
-
-HyperNEAT is an indirect encoding for evolving artificial neural
-networks (ANNs) that is inspired by the way natural organisms
-develop~\cite{stanley2009hypercube}. It evolves Compositional Pattern
-Producing Networks (CPPNs)~\cite{stanley2007compositional}, each of
-which is a genome that encodes an ANN
-phenotype~\cite{stanley2009hypercube}. Each CPPN is itself a directed
-graph, where each node is a mathematical function, such as sine or
-Gaussian. The nature of these functions can facilitate the evolution
-of properties such as symmetry (e.g., an absolute value or Gaussian
-function) and repetition (e.g., a sine
-function)~\cite{stanley2009hypercube, stanley2007compositional}. The
-signal on each link in the CPPN is multiplied by that link's weight,
-which can alter its effect.
-
-A CPPN is queried once for each link in the ANN substrate to determine
-that link's weight. The inputs to the CPPN are the Cartesian
-coordinates of both the source (e.g., $x = 2$, $y = 4$) and target
-(e.g., $x = 3$, $y = 5$) nodes of a link and a constant bias
-value. The CPPN takes these five values as inputs and produces one or
-two output values, depending on the substrate topology. If there is no
-hidden layer in the substrate, the single output is the weight of the
-link between a source node on the input layer and a target node on the
-output layer. If there is a hidden layer, the first output value
-determines the weight of the link between the associated input
-(source) and hidden layer (target) nodes, and the second output value
-determines the weight of the link between the associated hidden
-(source) and output (target) layer nodes. All pairwise combinations of
-source and target nodes are iteratively passed as inputs to a CPPN to
-determine the weight of each substrate link.
-
-\edit{following paragraph is figure caption for file named:
-  hyperneatExplanation.png} HyperNEAT Produces ANNs from CPPNs. ANN
-weights are specified as a function of the geometric coordinates of
-the source node and the target node for each connection. The
-coordinates of these nodes and a constant bias are iteratively passed
-to the CPPN to determine each connection weight. The CPPN has two
-output values, which specify the weights for each connection layer as
-shown.
-
-HyperNEAT is capable of exploiting the geometry of a problem: because
-the link values between nodes in the final ANN substrate are a
-function of the geometric positions of those nodes, HyperNEAT can
-exploit such information when solving a
-problem~\cite{stanley2009hypercube, clune2009sensitivity,
-  clune2011performance}. In the case of quadruped locomotion, this
-property helped Hyper-NEAT produce gaits with front-back, left-right,
-and four-way symmetries~\cite{clune2009evolving,
-  clune2011performance}.
-
-The evolution of the population of CPPNs occurs according to the
-principles of the NeuroEvolution of Augmenting Topologies (NEAT)
-algorithm~\cite{stanley2002evolving}, which was originally designed to
-evolve ANNs. NEAT can be fruitfully applied to CPPNs because of their
-structural similarity to ANNs. For example, mutations can add a node,
-and thus a function, to a CPPN graph, or change its link weights. The
-NEAT algorithm is unique in three main
-ways~\cite{stanley2002evolving}. Initially, it starts with small
-genomes that encode simple networks and slowly complexifies them via
-mutations that add nodes and links to the network, enabling the
-algorithm to evolve the network topology in addition to its
-weights. Secondly, NEAT has a fitness sharing mechanism that preserves
-diversity in the system and gives time for new innovations to be tuned
-by evolution before competing them against more adapted
-rivals. Finally, NEAT tracks historical information to perform
-intelligent crossover while avoiding the need for expensive
-topological analysis. A full explanation of NEAT can be found
-in~\cite{stanley2002evolving}].
-
-The ANN configuration follows previous studies that evolved quadruped
-gaits with HyperNEAT in simulation~\cite{clune2011performance,
-  clune2009evolving}, but was adapted to accommodate differences in
-the robot. Specifically, the ANNs consists of three $3 \times 4$
-Cartesian grids of nodes forming input, hidden, and output
-layers. Adjacent layers were completely connected, meaning that there
-were $(3 \times 4)^2= 288$ links in each substrate. The inputs to the
-substrate were the current angles of each of the 9 joints of the robot
-(described below) and a sine and cosine wave (to facilitate the
-production of periodic behaviors). The sine and cosine waves were the
-following function of time (t) in milli-seconds:
-$sin(t/12.0)\pi$. This function produces numbers in the range [$-\pi,
-  \pi$], which were then JASON FILL ME IN TO DESCRIBE HOW JOINT
-COMMANDS WERE SPECIFIED.
-
-\edit{are we mentioning that we gave HyperNEAT the center joint here or earlier?}
-\edit{jason: edit this section to reflect the post-processing...}
-
-\edit{add the following paragraph as a figure caption to SpiderANN.jpg}
-
-ANN Configuration for HyperNEAT Runs. The first two columns of each
-row of the input layer receive information about a single leg (the
-current angle of its two joints). The final column provides a sine and
-cosine wave to enable periodic movements and the angle of the center
-joint. Evolution determines the function of the hidden-layer
-nodes. The nodes in the output layer specify new joint angles for each
-respective joint. The unlabeled node in the input and output layer is
-ignored.
-
-\acmFig{explore_dim_1}{1}{Fitness mean and standard deviation
-  vs. dimension 1.  The circle is a common point in
-  \figref{explore_dim_1} through \figref{explore_dim_5}}
-
-The population size for HyperNEAT was 9 and runs lasted 20
-generations. These numbers are small, but were necesary given how much
-time it took to conduct evolution directly on a real robot. The
-remaining parameters were identical to Clune et
-al. ~\cite{clune2011performance}.
-
-  % Preliminary HyperNEAT runs were promising and resulted in several
-  % interesting gaits.
-
-  % Unfortunately, the gaits generated by HyperNEAT
-  % also tended to stress the robot more than typical gaits had before,
-  % and the servos would often overheat and malfunction, requiring
-  % restarts.  We think these issues may be addressed by adding a small
-  % layer between the HyperNEAT strategy and the robot that disallows
-  % quickly shifting commanded positions, and we hope to be able to test
-  % these methods further once this filter is in place.
-
 \end{itemize}
-
-\acmFig{svm_sim_results}{1}{Results for the SVM regression strategy
-  in simulation.  This simulation was used to tune the SVM strategy's
-  parameters before trying it on the physical robot.  The strategy
-  quickly approaches the maximum simulated fitness of 100.}

### 2011-01-31 paper/results.tex
Many edits...

@@
 \section{Results and Discussion}
+\seclabel{results}
+
+% Present the quantitative results of your experiments. Graphical data
+% presentation such as graphs and histograms are frequently better 
+% than tables. Explain what are the basic differences revealed in the 
+% data. If your method is stochastic or you analyze multiple datasets, 
+% calculate mean and standard error for every quantitative
+% performance result you show based on multiple runs.
 
 \edit{proofread this.}
+
+\subsection{Exploration of Parameter Space}
+
+In addition to running strategies to optimize parameterized gaits, we
+also wanted to investigate the space of possible gaits.  To accomplish
+this, we selected a parameter vector that resulted in motion, but not
+an exceptional gait, and plotted performance along each dimension
+individually.  In addition, we duplicated each measurement to be able
+to estimate the measurement noise at each point.  Results for this
+exploration are shown in \figref{explore_dim_1} through
+\figref{explore_dim_5}.  As is shown in the figures, some dimensions
+are smoother than others, and some measurements are fairly noisy.
+
+\acmFig{explore_dim_1}{1}{Fitness mean and standard deviation
+  vs. dimension 1.  The circle is a common point in
+  \figref{explore_dim_1} through \figref{explore_dim_5}}
+\acmFig{explore_dim_2}{1}{Fitness mean and standard deviation
+  vs. dimension 2.  The circle is a common point in
+  \figref{explore_dim_1} through \figref{explore_dim_5}}
+\acmFig{explore_dim_3}{1}{Fitness mean and standard deviation
+  vs. dimension 3.  The circle is a common point in
+  \figref{explore_dim_1} through \figref{explore_dim_5}}
+\acmFig{explore_dim_4}{1}{Fitness mean and standard deviation
+  vs. dimension 4.  The circle is a common point in
+  \figref{explore_dim_1} through \figref{explore_dim_5}}
+\acmFig{explore_dim_5}{1}{Fitness mean and standard deviation
+  vs. dimension 5.  The circle is a common point in
+  \figref{explore_dim_1} through \figref{explore_dim_5}}
+
 \subsection{Maximizing Strategies}
-The maximizing strategies implemented include uniform random hill climbing, Gaussian random hill climbing, policy gradient descent, and Nelder-Mead simplex.
-As a group, these strategies produced results similar to gaits randomly generated using the SineModel5 model.
-None of these strategies consistently outperformed the others in the three trials; A different strategy generated the best gait in each trial.
-Uniform random hill climbing produced the best gait (11.37 body lengths/minute) in the first trial, with the next best performing strategy (Nelder-Mead simplex) resulting in a gait of 8.51 body lengths/minute.
-The best gait in the second trial was actually generated randomly (17.26 body-lengths/minute).
-Gaussian random hill climbing, policy gradient descent, and Nelder-Mead simplex all produced similar best gaits in the second trial, each around 14 body lengths/minute.
-In the third trial, Nelder-Mead simplex and Gaussian random hill climbing significantly outperformed the other maximizing strategies, resulting in gaits of 14.83 and 13.40 body lengths/minute respectively.
-Overall, Nelder-Mead simplex performed the best, producing an average best gait of 12.32 body lengths/minute.
-Gaussian random hill climbing followed, with an average best gait of 10.03 body lengths/minute.
+
+The maximizing strategies implemented include uniform random hill
+climbing, Gaussian random hill climbing, policy gradient descent, and
+Nelder-Mead simplex. As a group, these strategies produced results
+similar to gaits randomly generated using the SineModel5 model. None
+of these strategies consistently outperformed the others in the three
+trials; A different strategy generated the best gait in each
+trial. Uniform random hill climbing produced the best gait (11.37 body
+lengths/minute) in the first trial, with the next best performing
+strategy (Nelder-Mead simplex) resulting in a gait of 8.51 body
+lengths/minute. The best gait in the second trial was actually
+generated randomly (17.26 body-lengths/minute). Gaussian random hill
+climbing, policy gradient descent, and Nelder-Mead simplex all
+produced similar best gaits in the second trial, each around 14 body
+lengths/minute. In the third trial, Nelder-Mead simplex and Gaussian
+random hill climbing significantly outperformed the other maximizing
+strategies, resulting in gaits of 14.83 and 13.40 body lengths/minute
+respectively. Overall, Nelder-Mead simplex performed the best,
+producing an average best gait of 12.32 body lengths/minute. Gaussian
+random hill climbing followed, with an average best gait of 10.03 body
+lengths/minute.
 
 \subsection{Predictive Strategies}
-The predictive strategies used were linear regression and support vector regression.
-In the first trial, linear regression produced a gait (27.58 body lengths/minute) that was over three times faster than the best gait produced by a maximizing strategy (uniform random hill climbing -- 11.37 body lengths/minute).
-The average of the best gait for the three trials of linear regression was 14.01 body lengths/minute, which was better than the average best gait for the best performing maximizing strategy (Nelder-Mead simplex -- 12.32 body lengths/minute).
 
-As mentioned above, the SVM strategy involved many tunable parameters.
-The version of SVM tuned for simulation did not show competitive performance on the real robot.
-We tried tuning some parameters on the real robot, but after some amount of tuning, the method still exhibited too little exploration and easily became stuck in local minima.
+The predictive strategies used were linear regression and support
+vector regression. In the first trial, linear regression produced a
+gait (27.58 body lengths/minute) that was over three times faster than
+the best gait produced by a maximizing strategy (uniform random hill
+climbing -- 11.37 body lengths/minute). The average of the best gait
+for the three trials of linear regression was 14.01 body
+lengths/minute, which was better than the average best gait for the
+best performing maximizing strategy (Nelder-Mead simplex -- 12.32 body
+lengths/minute).
+
+As mentioned above, the SVM strategy involved many tunable
+parameters. The version of SVM tuned for simulation did not show
+competitive performance on the real robot. We tried tuning some
+parameters on the real robot, but after some amount of tuning, the
+method still exhibited too little exploration and easily became stuck
+in local minima.
+
+\edit{say something about tuning SVM and the sim figure}
+
+\acmFig{svm_sim_results}{1}{Results for the SVM regression strategy
+  in simulation.  This simulation was used to tune the SVM strategy's
+  parameters before trying it on the physical robot.  The strategy
+  quickly approaches the maximum simulated fitness of 100.}
 
 \subsection{Evolved Neural Network Strategy}
-The evolved neural network strategy used was HyperNEAT.
-HyperNEAT produced the fastest gaits by far.
-Linear regression produced a gait in the first trial (27.58 body lengths/minute) that was on par with the best gaits generated by HyperNEAT, by no other maximizing or predictive strategy produced gaits close to those of HyperNEAT.
-Linear regression was the best performing maximizing or predictive strategy, but it resulted in an average best gait of only 14.01 body lengths/minute.
-HyperNEAT, on the other hand, resulted in an average best gait of 29.26 body lengths/minute.
-One issue encountered with HyperNEAT was that it generated gaits that were rougher on the robot than gaits generated by the other strategies.
-The gaits sometimes caused the motor servos to overheat, which causes the motor to behave differently than expected.
-Since an overheated motor produces a different walk that expected, a test was performed after each run to make sure all motors were working correctly.
-If a motor was not, the distance walked would be halved to penalize walks that were harmful to the robot.
 
-\begin{table}
+The evolved neural network strategy used was HyperNEAT. HyperNEAT
+produced the fastest gaits by far. Linear regression produced a gait
+in the first trial (27.58 body lengths/minute) that was on par with
+the best gaits generated by HyperNEAT, by no other maximizing or
+predictive strategy produced gaits close to those of HyperNEAT. Linear
+regression was the best performing maximizing or predictive strategy,
+but it resulted in an average best gait of only 14.01 body
+lengths/minute. HyperNEAT, on the other hand, resulted in an average
+best gait of 29.26 body lengths/minute. One issue encountered with
+HyperNEAT was that it generated gaits that were rougher on the robot
+than gaits generated by the other strategies. The gaits sometimes
+caused the motor servos to overheat, which causes the motor to behave
+differently than expected. Since an overheated motor produces a
+different walk that expected, a test was performed after each run to
+make sure all motors were working correctly. If a motor was not, the
+distance walked would be halved to penalize walks that were harmful to
+the robot.
+
+\edit{take this out?}
+
+We have done complete runs of 3 different initial parameter vectors
+for random search, uniform random hill climbing, Gaussian random hill
+climbing, policy gradient descent, Nelder-Mead simplex, and linear
+regression.  We also evaluated the SVM regression and HyperNEAT
+methods, but these methods were more experimental, and thus we do not
+yet have the same volume of data for these runs.  We developed several
+gaits that were about 4 times faster than the original hand-coded
+gait.  Results are shown in \tabref{results_table}.
+
+\begin{table*}
 \begin{center}
@@
 \end{center}
-\end{table}
-
-% Present the quantitative results of your experiments. Graphical data
-% presentation such as graphs and histograms are frequently better 
-% than tables. Explain what are the basic differences revealed in the 
-% data. If your method is stochastic or you analyze multiple datasets, 
-% calculate mean and standard error for every quantitative
-% performance result you show based on multiple runs.
-
-We have done complete runs of 3 different initial parameter vectors
-for random search, uniform random hill climbing, Gaussian random hill
-climbing, policy gradient descent, Nelder-Mead simplex, and linear
-regression.  We also evaluated the SVM regression and HyperNEAT
-methods, but these methods were more experimental, and thus we do not
-yet have the same volume of data for these runs.  We developed several
-gaits that were about 4 times faster than the original hand-coded
-gait.  Results are shown in \tabref{results_table}.
-
-\begin{table}
-\begin{center}
-\begin{tabular}{|r|c|c|c||c|}
-\hline
-                                         & A       & B      & C      &  Average \\
-\hline                                                               
-\hline                                                               
-Previous hand-coded gait                 & --      & --     & --     &  5.16 \\
-\hline                                                                 
-Random search                            & 6.04    & 17.26  & 4.90   &  9.40 \\
-\hline                                                                 
-Uniform Random Hill Climbing             & 11.37   & 9.44   & 2.69   &  7.83 \\
-\hline                                                                 
-Gaussian Random Hill Climbing            & 3.10    & 13.59  & 13.40  &  10.03 \\
-\hline                                                                 
-Policy Gradient Descent                  & 0.68    & 14.69  & 3.60   &  6.32 \\
-\hline                                                                 
-Nelder-Mead simplex                      & 8.51    & 13.62  & 14.83  &  12.32 \\
-\hline                                                                 
-Linear Regression                        & 27.58   & 12.51  & 1.95   &  14.01 \\
-\hline                                                                
-Evolutionary Neural Network (HyperNEAT)  & 24.27     & 36.44    & 27.07   & 29.26  \\
-\hline
-\end{tabular}
-\caption{The best gaits found for each starting vector and algorithm,
-  in body lengths per minute.}
-\tablabel{results_table}
-\end{center}
-\end{table}
+\end{table*}
 
@@
 
-In addition to running strategies to optimize parameterized gaits, we
-also wanted to investigate the space of possible gaits.  To accomplish
-this, we selected a parameter vector that resulted in motion, but not
-an exceptional gait, and plotted performance along each dimension
-individually.  In addition, we duplicated each measurement to be able
-to estimate the measurement noise at each point.  Results for this
-exploration are shown in \figref{explore_dim_1} through
-\figref{explore_dim_5}.  As is shown in the figures, some dimensions
-are smoother than others, and some measurements are fairly noisy.
+\subsection{Discussion}
 
-\acmFig{explore_dim_1}{1}{Fitness mean and standard deviation
-  vs. dimension 1.  The circle is a common point in
-  \figref{explore_dim_1} through \figref{explore_dim_5}}
-\acmFig{explore_dim_2}{1}{Fitness mean and standard deviation
-  vs. dimension 2.  The circle is a common point in
-  \figref{explore_dim_1} through \figref{explore_dim_5}}
-\acmFig{explore_dim_3}{1}{Fitness mean and standard deviation
-  vs. dimension 3.  The circle is a common point in
-  \figref{explore_dim_1} through \figref{explore_dim_5}}
-\acmFig{explore_dim_4}{1}{Fitness mean and standard deviation
-  vs. dimension 4.  The circle is a common point in
-  \figref{explore_dim_1} through \figref{explore_dim_5}}
-\acmFig{explore_dim_5}{1}{Fitness mean and standard deviation
-  vs. dimension 5.  The circle is a common point in
-  \figref{explore_dim_1} through \figref{explore_dim_5}}
-
-\subsection{Discussion}
 % Is your hypothesis supported? What conclusions do the results support

### 2011-01-31 paper/introduction.tex
Many edits...
Added some introduction work

@@
   this study.}
+
+Gaits for walking robots are often designed with some explicit or
+implicit goal in mind.  For some applications, the design criteria may
+be obvious: perhaps the robot needs to move as quickly or as
+efficiently as possible, or with minimal power usage or strain on
+actuators.  In many cases, however, the objective is more complicated,
+requiring simultaneous optimization of several desired traits, each
+with its own relative importance.  For example, Honda's Asimo
+\cite{_asmio} and Boston Dynamic's Big Dog \cite{_bigdog} both require
+gaits that are relatively quick, power efficient, and robust to
+changing terrain, but with they vary widely in the importance placed
+on each attribute.  A primary concern for Big Dog's gait is that it is
+robust to steep terrain, whereas
+
+In other cases, the objective may be more complicated; perhaps it is
+desired that the gait be fairly [robust]
+
+\_asimo and \_bigDog very different (and some hexapod?)
+
+Many times gait will be designed and tweaked by hand \cite{_asimo}
+
+here we describe a study, a comparison of few methods. started with
+sped, but then not-breaking as well.
+
+When designing gaits for walking robots, often it is desirable that
+this gait enable the robot to walk as
+
+Old: design by hand
+Newer: Learn gait
+  - parmametrized or more free form (HN, ???)
+  - in sim or hardware
+Us: first time HN in hardware, first time compared to other methods aiming for symmetry on same hardware
+
+Often
+these optimizations are done manually by an expert who designs and
+tweaks a gait specifically for a given objective.  Other groups have
+used learning methods to generate gaits optimized for some metric.
+Approaches differ in their starting assumptions, some essentially
+tweaking the parameters of a hand-tuned model \cite{chernova}, others
+exploring a reasonably compact parameter space \cite{kohl}, and still
+others beginning with few assumptions besides periodicity
+\cite{zykov}.
 

### 2011-01-31 code/plot/plot_error_new.m
more updates to plot_error_new

@@ -1,20 +1,28 @@
 function plot_error_new(files, saveFileName)
 
 mark = {};
-  mark{end+1} = '+';
-  mark{end+1} = '+';
-  mark{end+1} = '*';
+  mark{end+1} = 'v';
+  mark{end+1} = 's';
+  mark{end+1} = 'd';%<
   mark{end+1} = 'o';
-  mark{end+1} = 'x';
-  mark{end+1} = 'x';
+  mark{end+1} = '.';%>
+  mark{end+1} = '^';
+
+mark2 = {};
+  mark2{end+1} = '-v';
+  mark2{end+1} = '-s';
+  mark2{end+1} = '-d';
+  mark2{end+1} = '-o';
+  mark2{end+1} = '-.';
+  mark2{end+1} = '-^';
 
 clr = {};
   clr{end+1} = 'r';
   clr{end+1} = 'k';
   clr{end+1} = 'b';
-  clr{end+1} = [0.5, 0.75, 0.15];
-  clr{end+1} = [.3, 0, .5];
-  clr{end+1} = [1 128/255, 0];
+  clr{end+1} = [0.5, 0.75, 0.15];%green
+  clr{end+1} = [.3, 0, .5]; %purple
+  clr{end+1} = [1 128/255, 0];%pumpkin
   
 figure(2); clf; hold on;
 ax = zeros(4, 1);
@@ -22,11 +30,11 @@ ax(1) = 0; ax(2) = 80; ax(3) = 0; ax(4) = 20;
 axis(ax);
 fs = 30; %font size
 xlabel("Iteration", "fontsize", fs);
-ylabel("Body lengths/minute", "fontsize", fs);
+ylabel("Body lengths per minute", "fontsize", fs);
 set(gcf, 'DefaultLineLineWidth', 2.5);
 
 %leg = cell(1, (((length(files))/3)-1));
-leg = cell(1, (length(files)-6));
+leg = cell(1, 6);
 
 frac = 15;
 markSize = 11;
@@ -41,10 +49,19 @@ for ii = 0:(((length(files))/3)-1)
   
   legName = strrep (a, "_", "\\_");
   legName = substr(legName, 6, (length(legName) - 12));
-  %leg{1, ii+1} = legName;
-  leg{1, ii*2+1} = legName;
-  leg{1, ii*2+2} = '';
-  %leg{1, ii*2+3} = '';
+  leg{1, ii+1} = legName;
+  
[…]

### 2011-01-31 paper/platformDetails.tex
fixed centimeter measurements

--- 2011-01-26 paper/platformDetails.tex
@@
 Connex 500 3-D Printing System. It weights 1884 g and measures approximately
-15 centimeters from leg to leg in the crouch position depicted in
+38 centimeters from leg to leg in the crouch position depicted in
 \figref{robot_close.jpg}. There are a total of 9 AX-12+ Dynamixel
 Servomotors, displayed in \figref{topdown.png}: 4 outer leg joint
-servos, 4 inner leg joint servos, and a final servo at the center "hip" of the
+servos, 4 inner leg joint servos, and a final servo at the center ``hip'' of the
 robot. The robot has an on-board computer running Linux. 
@@
 The servos are prevented from being commanded to a point outside their
-normal range (0 - 1023) as well as beyond points where limbs would collide.
+normal range (0 - 1023) as well as beyond points where limbs would collide
+(150 - 770 for inner leg servos, 30 - 680 for outer leg servos, and 392 - 623
+for the center, or ``hip'' servo.

### 2011-01-31 paper/results.tex

@@
 in the first trial (27.58 body lengths/minute) that was on par with
-the best gaits generated by HyperNEAT, by no other maximizing or
+the best gaits generated by HyperNEAT, but no other maximizing or
 predictive strategy produced gaits close to those of HyperNEAT. Linear
@@
 than gaits generated by the other strategies. The gaits sometimes
-caused the motor servos to overheat, which causes the motor to behave
+caused the motor servos to overheat, which caused the motor to behave
 differently than expected. Since an overheated motor produces a
 different walk that expected, a test was performed after each run to
-make sure all motors were working correctly. If a motor was not, the
+make sure all motors were working correctly. If a motor failed, the
 distance walked would be halved to penalize walks that were harmful to

### 2011-01-31 code/plot/plot_error_new.m
updates

@@ -115,7 +115,7 @@ for ii = 0:(((length(files))/3)-1)
   %set(e, 'color', clr{mod(ii,6)+1}, 'markerSize', markSize);
   set(e, 'color', clr{mod(ii,6)+1}, 'marker', '.');
   plot(means, 'color', clr{mod(ii,6)+1});
-  plot(someX, someMeans, mark{mod(ii,6)+1}, 'color', clr{mod(ii,6)+1}, 'MarkerFaceColor', 'k');
+  plot(someX, someMeans, mark{mod(ii,6)+1}, 'color', clr{mod(ii,6)+1}, 'MarkerFaceColor', 'k', 'markerSize', markSize);
 end
 
 disp(leg);

### 2011-02-04 paper/hyperNeatMethod.tex
More edits to intro, hyperNeatMethod

\subsection{HyperNEAT Motion Model}
\seclabel{hyperNeatMethod}

\emph{Neuroevolution (Evolving Artificial Neural Networks with HyperNEAT)}:

HyperNEAT is an indirect encoding for evolving artificial neural
networks (ANNs) that is inspired by the way natural organisms
develop~\cite{stanley2009hypercube}. It evolves Compositional Pattern
Producing Networks (CPPNs)~\cite{stanley2007compositional}, each of
which is a genome that encodes an ANN
phenotype~\cite{stanley2009hypercube}. Each CPPN is itself a directed
graph, where each node is a mathematical function, such as sine or
Gaussian. The nature of these functions can facilitate the evolution
of properties such as symmetry (e.g., an absolute value or Gaussian
function) and repetition (e.g., a sine
function)~\cite{stanley2009hypercube, stanley2007compositional}. The
signal on each link in the CPPN is multiplied by that link's weight,
which can alter its effect.

A CPPN is queried once for each link in the ANN substrate to determine
that link's weight. The inputs to the CPPN are the Cartesian
coordinates of both the source (e.g., $x = 2$, $y = 4$) and target
(e.g., $x = 3$, $y = 5$) nodes of a link and a constant bias
value. The CPPN takes these five values as inputs and produces one or
two output values, depending on the substrate topology. If there is no
hidden layer in the substrate, the single output is the weight of the
link between a source node on the input layer and a target node on the
output layer. If there is a hidden layer, the first output value
determines the weight of the link between the associated input
(source) and hidden layer (target) nodes, and the second output value
determines the weight of the link between the associated hidden
(source) and output (target) layer nodes. All pairwise combinations of
source and target nodes are iteratively passed as inputs to a CPPN to
determine the weight of each substrate link.

\acmFig{hyperneatExplanation.png}{1}{HyperNEAT Produces ANNs from
  CPPNs. ANN weights are specified as a function of the geometric
  coordinates of the source node and the target node for each
  connection. The coordinates of these nodes and a constant bias are
  iteratively passed to the CPPN to determine each connection
  weight. The CPPN has two output values, which specify the weights
  for each connection layer as shown.}

HyperNEAT is capable of exploiting the geometry of a problem: because
the link values between nodes in the final ANN substrate are a
function of the geometric positions of those nodes, HyperNEAT can
exploit such information when solving a
problem~\cite{stanley2009hypercube, clune2009sensitivity,
  clune2011performance}. In the case of quadruped locomotion, this
property helped Hyper-NEAT produce gaits with front-back, left-right,
and four-way symmetries~\cite{clune2009evolving,
  clune2011performance}.

The evolution of the population of CPPNs occurs according to the
principles of the NeuroEvolution of Augmenting Topologies (NEAT)
algorithm~\cite{stanley2002evolving}, which was originally designed to
evolve ANNs. NEAT can be fruitfully applied to CPPNs because of their
structural similarity to ANNs. For example, mutations can add a node,
and thus a function, to a CPPN graph, or change its link weights. The
NEAT algorithm is unique in three main
ways~\cite{stanley2002evolving}. Initially, it starts with small
genomes that encode simple networks and slowly complexifies them via
mutations that add nodes and links to the network, enabling the
algorithm to evolve the network topology in addition to its
weights. Secondly, NEAT has a fitness sharing mechanism that preserves
diversity in the system and gives time for new innovations to be tuned
by evolution before competing them against more adapted
rivals. Finally, NEAT tracks historical information to perform
intelligent crossover while avoiding the need for expensive
topological analysis. A full explanation of NEAT can be found
in~\cite{stanley2002evolving}.

The ANN configuration follows previous studies that evolved quadruped
gaits with HyperNEAT in simulation~\cite{clune2011performance,
  clune2009evolving}, but was adapted to accommodate differences in
the physical robot. Specifically, the ANNs consists of three $3 \times 4$
Cartesian grids of nodes forming input, hidden, and output
layers. Adjacent layers were completely connected, meaning that there
were $(3 \times 4)^2= 288$ links in each substrate. The inputs to the
substrate were the current angles of each of the 9 joints of the robot
(described below) and a sine and cosine wave (to facilitate the
production of periodic behaviors). The sine and cosine waves were the
following function of time ($t$) in milli-seconds:
$\sin(t/12)\pi$. This function produces numbers in the range [$-\pi,
  \pi$], which were then JASON FILL ME IN TO DESCRIBE HOW JOINT
COMMANDS WERE SPECIFIED.

\editbox{are we mentioning that we gave HyperNEAT the center joint
  here or earlier?}

\edit{jason: edit this section to reflect the post-processing...}

\acmFig{SpiderANN.jpg}{1}{ANN Configuration for HyperNEAT Runs. The
  first two columns of each row of the input layer receive information
  about a single leg (the current angle of its two joints). The final
  column provides a sine and cosine wave to enable periodic movements
  and the angle of the center joint. Evolution determines the function
  of the hidden-layer nodes. The nodes in the output layer specify new
  joint angles for each respective joint. The unlabeled node in the
  input and output layer is ignored.}

The population size for HyperNEAT was 9 and runs lasted 20
generations. These numbers are small, but were necessarily constrained
given how much time it took to conduct evolution directly on a real
robot. The remaining parameters were identical to Clune et
al. ~\cite{clune2011performance}.

% Preliminary HyperNEAT runs were promising and resulted in several
% interesting gaits.

% Unfortunately, the gaits generated by HyperNEAT also tended to
% stress the robot more than typical gaits had before, and the servos
% would often overheat and malfunction, requiring restarts.  We think
% these issues may be addressed by adding a small layer between the
% HyperNEAT strategy and the robot that disallows quickly shifting
% commanded positions, and we hope to be able to test these methods
% further once this filter is in place.

### 2011-02-04 paper/introduction.tex
More edits to intro, hyperNeatMethod

@@
 implicit goal in mind.  For some applications, the design criteria may
-be obvious: perhaps the robot needs to move as quickly or as
-efficiently as possible, or with minimal power usage or strain on
-actuators.  In many cases, however, the objective is more complicated,
+be obvious --- perhaps the robot needs to move as quickly or as
+efficiently as possible --- but other times the objective is more complicated,
 requiring simultaneous optimization of several desired traits, each
-with its own relative importance.  For example, Honda's Asimo
+with its own relative importance.  The different combinations of desired traits and the relative weight placed on each can produce drastically different gaits.    For example, Honda's Asimo
 \cite{_asmio} and Boston Dynamic's Big Dog \cite{_bigdog} both require
 gaits that are relatively quick, power efficient, and robust to
-changing terrain, but with they vary widely in the importance placed
-on each attribute.  A primary concern for Big Dog's gait is that it is
-robust to steep terrain, whereas
+changing terrain, but they vary widely in the importance placed
+on each attribute.  Big Dog's gait is optimzed for much more difficult terrain than Asimo's, resulting in a gait of a completely different form.
 
-In other cases, the objective may be more complicated; perhaps it is
-desired that the gait be fairly [robust]
+Whatever the design goals, gaits may be obtained by one of several methods.
+Gaits may be designed by hand by an expert, as in \cite{_asmio, _bigdog}, or they may be learned by repeated trial and evaluation.
+Learned gaits offer several advantages over manually designed gaits.
+Automatically learning gaits can save valuable engineering time and can allow customization of gaits to a particular robot and its unique actuators.
+Most importantly, learned gaits can outperform manually designed gaits
+
+save programmer time
+customize gaits per model
+design better gaits than people
+
+In many cases additional objectives may be
+
+, or with minimal power usage or strain on
+actuators.  In many cases, however,
+
+In other cases, the objective may be more complicated; perhaps it is desired that the gait be fairly [robust]
 

### 2011-02-04 code/plot/plot_all.m
Added plot of commanded angles for HyperNEAT runs

@@ -51,3 +51,6 @@ files = {'log_uniform_a.txt', 'log_uniform_b.txt', 'log_uniform_c.txt', ...
          'log_linear_regression_a.txt', 'log_linear_regression_b.txt', 'log_linear_regression_c.txt', ...
          'log_simplex_a.txt', 'log_simplex_b.txt', 'log_simplex_c.txt'};
 plot_error(files, 'std_error');
+
+
+plot_angles();

### 2011-02-04 code/plot/plot_angles.m

@@ -1,0 +1,16 @@
+function plot_angles()
+dat = load('../../hyperneat_runs/hyperneatTo20gens_101/neat_110115_211410_00000_002_filt');
+tt = linspace(0, 12, size(dat,1));
+
+figure(1);
+plot(tt, dat, 'linewidth', 2, 'linewidth', 4);
+xlabel('Time (s)',       'fontsize', 22);
+ylabel('Motor Position', 'fontsize', 22);
+print('-depsc2', '-F:18', 'neat_110115_211410_00000_002_filt.eps');
+
+figure(2);
+plot(tt(1:85), dat(1:85,:), 'linewidth', 4);
+xlabel('Time (s)',       'fontsize', 22);
+ylabel('Motor Position', 'fontsize', 22);
+print('-depsc2', '-F:18', 'neat_110115_211410_00000_002_filt_zoom.eps');
+

### 2011-02-05 paper/introduction.tex
More intro...

--- 2011-01-26 paper/introduction.tex
@@
   this study.}
+
+Gaits for walking robots are often designed with some explicit or
+implicit goal in mind.  For some applications, the design criteria may
+be obvious --- perhaps the robot needs to move as quickly or as
+efficiently as possible --- but other times the objective is more complicated,
+requiring simultaneous optimization of several desired traits, each
+with its own relative importance.  The different combinations of desired traits and the relative weight placed on each can produce drastically different gaits.    For example, Honda's Asimo
+\cite{_asmio} and Boston Dynamic's Big Dog \cite{_bigdog} both require
+gaits that are relatively quick, power efficient, and robust to
+changing terrain, but they vary widely in the importance placed
+on each attribute.  Big Dog's gait is optimzed for much more difficult terrain than Asimo's, resulting in a gait of a completely different form.
+
+Once the design goals are decided upon, gaits may be obtained by one of several methods.
+Gaits may be designed by hand by an expert, as in \cite{_asmio, _bigdog}.
+Alternately, they may be learned through repeated trial and evaluation.
+Learned gaits offer several advantages over manually designed gaits.
+Automatically learning gaits can save valuable engineering time and can allow customization of gaits to a particular robot and its unique actuators.
+Most importantly, in some cases learned gaits can outperform manually designed gaits.
+
+In this paper we compare the performance of three different methods of designing gaits: manual design, parametrized gait optimized using one of seven learnign methods, and gaits generated and evolved by HyperNEAT.
+All experiments were done in hardware using the platform shown in \figref{robot_close.jpg}.
+To our knowledge this is the first time HyperNEAT has been used to directly evolve gaits in hardware.
+
+In simulation HyperNEAT has been shown to produce the highest performing gaits when using an encoding that respects the geometric regularity of a problem \cite{jeffSomething}.
+To take advantage of this fact, we used an encoding for the physical robot that respects the symmetry of its form; this encoding is shown in \figref{hyperEncoding???}.
+
+In addition, past research has suggested \cite{jeffSomething} one advantage the HyperNEAT encoding has over the NEAT encoding is in the bias toward regularity of its output.
+That is, biasing gaits toward regularity appears to be beneficial.
+Thus, to make our comparison as fair as possible, for the parametrized family of gaits, we used a construction that also biases toward regularity.
+
+\edit{more past work in here???}
+
+%Old: design by hand
+%Newer: Learn gait
+%  - parmametrized or more free form (HN, ???)
+%  - in sim or hardware
+%Us: first time HN in hardware, first time compared to other methods aiming for %symmetry on same hardware
+
+Often
+these optimizations are done manually by an expert who designs and
+tweaks a gait specifically for a given objective.  Other groups have
+used learning methods to generate gaits optimized for some metric.
+Approaches differ in their starting assumptions, some essentially
+tweaking the parameters of a hand-tuned model \cite{chernova}, others
+exploring a reasonably compact parameter space \cite{kohl}, and still
+others beginning with few assumptions besides periodicity
+\cite{zykov}.
 

### 2011-02-05 paper/platformDetails.tex
Slight approach restructuring, platform details ~done, fixed wiimote fig

--- 2011-01-26 paper/platformDetails.tex
@@
 \subsection{Platform details}
+\seclabel{platformDetails}
 
-\edit{figure:robotDiagram make figure to go somewhere here, show drawing with inner/outer, l/r, and f/b motors.}
-
-%\section{System Architecture and Implementation}
-\seclabel{implement}
+%\seclabel{implement}
 
@@
 
-The quadruped robot was printed on the Objet
-Connex 500 3-D Printing System. It weights 1884 g and measures approximately
-15 centimeters from leg to leg in the crouch position depicted in
-\figref{robot_close.jpg}. There are a total of 9 AX-12+ Dynamixel
-Servomotors, displayed in \figref{topdown.png}: 4 outer leg joint
-servos, 4 inner leg joint servos, and a final servo at the center "hip" of the
-robot. The robot has an on-board computer running Linux. 
-The lower level drivers are in C and the system is implemented in Python.
+The quadruped robot used in this study was assembled from parts
+printed on the Objet Connex 500 3-D Printing System. It weights 1.88
+kg with the on-board computer and measures approximately 38
+centimeters from leg to opposite leg in the crouch position depicted
+in \figref{robot_close.jpg}. The robot is actuated by 9 AX-12+
+Dynamixel servos: one inner joint and one outer joint servo in each of
+the four legs, and one servo at the center ``hip'' joint.  This final
+unique servo allows the two halves of the robot to rotate with respect
+to each other. \figref{topdown.png} shows the positions and numerical
+designations of all nine servoes.
 
-\acmFigg{wiiMote.JPG}{robot.JPG}{1}{A Nintendo Wii remote tracks the location
-of the robot, providing feedback about distance traveled, in pixels, through
-an infrared LED mounted on top of the robot.}
+Computation for gait learning, fitness evaluation, and robot control
+is performed on the compact on-board CompuLab Fit-PC2, running Ubuntu
+Linux.  All gait generation, learning, and fitness evaluation code,
+except HyperNEAT, is written in Python and is available on our website
+\cite{quadraWeb}.  HyperNEAT is written in C++.  We made use of the
+excellent \code{pydynamixel} library \cite{pydynamixel} for
+controlling the servos, available under the GPLv2 license.  The robot
+was set up to connect to a wireless network on boot, and thus we were
+able to control all experiments via ssh.
 
-We controlled the experiments from a computer that was connected via a
-wireless Ethernet to the robot. A Nintendo Wii remote tracks the location 
-of the robot, providing feedback
-about distance traveled, in pixels, through an infrared LED mounted on top of 
-the robot (\figref{wiiMote.JPG}). A server is run on the robot and continuously tracks its
-position using the CWiid library\cite{cwiid} to interface with the remote
-via bluetooth. A client then connects via a socket to the tracking
-server and requests position updates periodically. If the robot walks 
-beyond the viewable range of the Wii remote, a prompt informed the
-user. The only human intervention required during the
-experiment was to move the robot back inside the viewable area and resume
-the run, and to handle any mechnanical failures that arose. This did not
-interrupt the learning process or result in the loss of data.
+%\acmFigg{wiiMote.jpg}{robot_led.jpg}{1}{A Nintendo Wii remote tracks the
+%  location of the robot, providing feedback about distance traveled,
+%  in pixels, through an infrared LED mounted on top of the robot.}
 
-The robot is run using a given motion model, including, if desired, 
-smooth interpolation over time for the beginning and end of the run. 
-The servos are prevented from being commanded to a point outside their
-normal range (0 - 1023) as well as beyond points where limbs would collide.
+\acmFigg{wiiMote_crop.jpg}{robot_led_crop.jpg}{1}{A Nintendo Wii remote tracks the
+  location of the robot, providing feedback about distance traveled,
+  in pixels, through an infrared LED mounted on top of the robot.}
+
+To track the position of the robot and determine gait fitness, we
+mounted a Nintendo Wii remote on the ceiling and an infrared LED on
+top of the robot (\figref{wiiMote.JPG}).  The Wii remote contains an
+IR camera that can track and report the position of any IR sources in
+its image frame.  The resolution of the camera was 1024 by 768 pixels,
+which produced a resolution of 1.7mm per pixel when mounted at a
+height of \edit{???}m.
+
+A separate tracking server, written in Python, ran on the robot and
+interfaced with the Wii remote via bluetooth using the CWiid
+library\cite{cwiid}.  Our fitness testing code connected to this
+server via socket and requested position updates at the beginning and
+end of each run (see \secref{fitnessEvaluation for more details}).
+
+%to interface with the remote via bluetooth. A client then connects via
+%a socket to the tracking server and requests position updates
+%periodically. If the robot walks beyond the viewable range of the Wii
+%remote, a prompt informed the user. The only human intervention
+%required during the experiment was to move the robot back inside the
+%viewable area and resume the run, and to handle any mechnanical
+%failures that arose. This did not interrupt the learning process or
+%result in the loss of data.
+
+To run a gait on the robot, one simply provides a \emph{gait
+  function}, any Python function that accepts a single input --- time
+starting at 0 --- and outputs a list of nine commanded positions, one
+for each servo.  To safeguard against limb collision with the robot
+body, the robot control code cropped the commands to a safe range.
+This range was [150, 770] for the inner leg servos, [30, 680] for the
+outer leg servos, and [392, 623] for the center hip servo.
+
+%The robot was set up to run when given any giat
+
+%is run using a given motion model, including, if desired, 
+%smooth interpolation over time for the beginning and end of the run.
+
+%The servos are prevented from being commanded to a point outside their
+%normal range (0 - 1023) as well as beyond points where limbs would collide

### 2011-02-05 paper/paperOutline.tex
Added paperOutline

@@
 
-\edit{write this}
-
-The remainder of the paper is organized as follows. In \secref{foo}, we
-introduce,  \secref{bar} 3 describes.  Finally, we describe future work
-in \secref{future}
+The remainder of the paper is organized as follows. In
+\secref{problemDefinition} we define more rigorously the problem of
+gait learning. \secref{experimentalSetup} describes our experimental
+setup, including the hardware we used and our methods for evaluating
+fitness of a gait. \secref{gaitGenLearn} discusses the different gait
+generation and learning methods we tested, and \secref{results}
+presents and discusses performance results.  Finally,
+\secref{conclusion} concludes with possible future extensions to this
+work.

### 2011-02-09 paper/abstract.tex
First ugly version of report in acm style.  Figures ugly, bibliography commented out (not yet working)
Restructuring of report into multiple files to minimize conflicts, references working, dealt old sections to appropriate locations in paper framework.
lots of editing of the intro/motivation/previouswork, and some in the robot platform

@@
 learning strategies including uniform and Gaussian random hill
-climbing, policy gradient reinforcement learning\cite{kohl},
-Nelder-Mead simplex\cite{nm}, new predictive methods based on linear
+climbing, policy gradient reinforcement learning,
+Nelder-Mead simplex, new predictive methods based on linear
 and support vector regression, and an evolved neural network
-(HyperNEAT)\cite{clune}.  We compare results to a baseline random
+(HyperNEAT).  We compare results to a baseline random
 search method.  Many of the methods resulted in walks significantly
@@
 
-Keywords: Learning Control, Walking Robots, Gait Learning, Quadruped
+\edit{Write the abstract}

### 2011-02-09 paper/acknowledgments.tex
lots of editing of the intro/motivation/previouswork, and some in the robot platform

@@
 
-Thanks to ..., some grant maybe, Jim T\o rreson, ...and the reviewers.
-
-\edit{write this, old below}
+Left blank for anonymity.
 
@@
 % any external code you used (libraries etc).
-\begin{itemize}
-\item Hod Lipson, Cornell Computational Synthesis Lab: adviser.
-\item Jim T\o rreson, University of Oslo: adviser.
-\item Juan Zagal, University of Chile: designed and printed robot and provided code for hand tuned gait.
-\item Jeff Clune, CCSL Red Couch: collaborated on HyperNEAT implementation/testing.
-\item Cooper Bills, Cornell University: assisted with Wii tracker development.
-\item Anshumali Srivastava, Cornell University: Teaching Assistant.
-\end{itemize}
+%\begin{itemize}
+%\item Hod Lipson, Cornell Computational Synthesis Lab: adviser.
+%\item Jim T\o rreson, University of Oslo: adviser.
+%\item Juan Zagal, University of Chile: designed and printed robot and provided code for hand tuned gait.
+%\item Jeff Clune, CCSL Red Couch: collaborated on HyperNEAT implementation/testing.
+%\item Cooper Bills, Cornell University: assisted with Wii tracker development.
+%\item Anshumali Srivastava, Cornell University: Teaching Assistant.
+%\end{itemize}

### 2011-02-09 paper/introduction.tex
lots of editing of the intro/motivation/previouswork, and some in the robot platform

--- 2011-02-04 paper/introduction.tex
@@
 
-\acmFig{robot_close.jpg}{1}{The 3D printed quadruped robot used for
-  this study.}
+\acmFig{robot_close.jpg}{1}{The quadruped robot. The translucent parts were created with 3D-printing technology.}
 
@@
 with its own relative importance.  The different combinations of desired traits and the relative weight placed on each can produce drastically different gaits.    For example, Honda's Asimo
-\cite{_asmio} and Boston Dynamic's Big Dog \cite{_bigdog} both require
+\cite{chestnutt2006footstep} and Boston Dynamic's Big Dog \cite{raibert2008bigdog} both require
 gaits that are relatively quick, power efficient, and robust to
 changing terrain, but they vary widely in the importance placed
-on each attribute.  Big Dog's gait is optimzed for much more difficult terrain than Asimo's, resulting in a gait of a completely different form.
+on each attribute.  Big Dog's gait is optimized for much more difficult terrain than Asimo's, resulting in a gait of a completely different form.
 
-Whatever the design goals, gaits may be obtained by one of several methods.
-Gaits may be designed by hand by an expert, as in \cite{_asmio, _bigdog}, or they may be learned by repeated trial and evaluation.
+Once the design goals are decided upon, gaits may be obtained by one of several methods.
+Gaits may be designed manually by an expert, as for Asimo and Big Dog \cite{chestnutt2006footstep, raibert2008bigdog}.
+Alternately, they may be learned through repeated trial and evaluation.
 Learned gaits offer several advantages over manually designed gaits.
 Automatically learning gaits can save valuable engineering time and can allow customization of gaits to a particular robot and its unique actuators.
-Most importantly, learned gaits can outperform manually designed gaits
+Most importantly, in some cases learned gaits can outperform manually designed gaits~\cite{hornby2005autonomous, valsalam2008modular}.
 
-save programmer time
-customize gaits per model
-design better gaits than people
+In this paper we compare the performance of three different methods of designing gaits: manual design, parametrized gaits optimized with seven different learning methods, and gaits generated by evolving neural networks with the HyperNEAT generative encoding~\cite{stanley2009hypercube}.
+While some of these methods, such as HyperNEAT, have been tested in simulation~\cite{clune2009evolving, clune2011performance}, we investigate how they perform when evolving on a physical robot (\figref{robot_close.jpg}).
 
-In many cases additional objectives may be
+Previous work has shown that quadruped gaits perform better when they are \emph{regular} (i.e.\ when the legs are coordinated)~\cite{clune2009evolving, clune2011performance,valsalam2008modular}.
+For example, HyperNEAT produced fast, natural gaits in part because its bias towards regular gaits created coordinated movements that outperformed gaits evolved by an encoding not biased towards regularity~\cite{clune2009evolving, clune2011performance}.
+One of the motivations of this paper is to investigate whether any learning method biased towards regularity would perform well at producing quadruped gaits, or whether HyperNEAT's high performance is due to additional factors, such as its abstraction of biological development (described below).
+We test this hypothesis by comparing HyperNEAT to seven machine learning algorithms biased toward regularity.
 
-, or with minimal power usage or strain on
-actuators.  In many cases, however,
-
-In other cases, the objective may be more complicated; perhaps it is desired that the gait be fairly [robust]
-
-\_asimo and \_bigDog very different (and some hexapod?)
-
-Many times gait will be designed and tweaked by hand \cite{_asimo}
-
-here we describe a study, a comparison of few methods. started with
-sped, but then not-breaking as well.
-
-When designing gaits for walking robots, often it is desirable that
-this gait enable the robot to walk as
-
-Old: design by hand
-Newer: Learn gait
-  - parmametrized or more free form (HN, ???)
-  - in sim or hardware
-Us: first time HN in hardware, first time compared to other methods aiming for symmetry on same hardware
-
-Often
-these optimizations are done manually by an expert who designs and
-tweaks a gait specifically for a given objective.  Other groups have
-used learning methods to generate gaits optimized for some metric.
-Approaches differ in their starting assumptions, some essentially
-tweaking the parameters of a hand-tuned model \cite{chernova}, others
-exploring a reasonably compact parameter space \cite{kohl}, and still
-others beginning with few assumptions besides periodicity
-\cite{zykov}.
-
-\edit{Write this}
-
-%\editbox{Outline \\
-% - Why gait learning is cool \\
-% - Gait learning has been done before, in simulation and on hardware (cite, cite, cite) \\
-% - HyperNEAT has been used to evolve gaits before in simulation~{clune2009evolving, clune2011performance, clune2009hybrid}, and it worked well in part because the gaits were regular, raising the question of whether this high performance was just because the gaits were regular?
-We test this hypothesis by comparing HyperNEAT to traditional machine learning methods biased towards regular gaits.
-\\
-% - Here, we do two cool things for the first time: \\
-% - - evolve HN in hardware \\
-% - - compare it to other learning methods on the same hardware.}
-
-\edit{Old Quadratot intro follows}
-
-% Motivate and abstractly describe the problem you are addressing and
-% how you are addressing it. What is the problem? Why is it important?
-% What is your basic approach? A short discussion of how it fits into
-% % related work in the area is also desirable. Summarize the basic
-% results and conclusions that you will present.
-
-Applications of walking robots often call for the ability to walk as
-quickly, efficiently, or with as little power as possible.  Often
-these optimizations are done manually by an expert who designs and
-tweaks a gait specifically for a given objective.  Other groups have
-used learning methods to generate gaits optimized for some metric.
-Approaches differ in their starting assumptions, some essentially
-tweaking the parameters of a hand-tuned model \cite{chernova}, others
-exploring a reasonably compact parameter space \cite{kohl}, and still
-others beginning with few assumptions besides periodicity
-\cite{zykov}.
-
-We aimed to strike a middle ground between these approaches.  Our
-motion generator did not rely on hand-tweaked gaits, but it did use
-parameterized gaits which, by their nature, impose some assumptions on
-the answers produced.  We then used machine learning to design gaits
-for a quadruped robot with these models.  This paper presents a
-comparison of the different learning methods implemented.  Most
-methods created walks that are several times faster than the original
-hand-tuned gait.  We invite readers with short attention spans to view
-a video of some of our results online here:
-
-%\url{http://www.youtube.com/watch?v=ODoiOj9DdGg}
-\texttt{http://www.youtube.com/watch?v=ODoiOj9DdGg}
-
-%The code so far consists of the robot class, optimization class,
-%parameterized model, sine model, and camera feedback.
+A final motivation of this problem is to simply evolve effective gaits for a physical robot.
+Because HyperNEAT performed well in simulation, it is interesting to test whether it can produce a fast gait for a physical robot.
+It is additionally interesting to test how more traditional machine learning techniques compete with evolutionary algorithms when evolving in hardware.

### 2011-02-09 paper/platformDetails.tex
lots of editing of the intro/motivation/previouswork, and some in the robot platform

@@
 
-Computation for gait learning, fitness evaluation, and robot control
-is performed on the compact on-board CompuLab Fit-PC2, running Ubuntu
+All of the computation for gait learning, fitness evaluation, and robot control
+was performed on the compact on-board CompuLab Fit-PC2, running Ubuntu
 Linux.  All gait generation, learning, and fitness evaluation code,
-except HyperNEAT, is written in Python and is available on our website
-\cite{quadraWeb}.  HyperNEAT is written in C++.  We made use of the
-excellent \code{pydynamixel} library \cite{pydynamixel} for
-controlling the servos, available under the GPLv2 license.  The robot
-was set up to connect to a wireless network on boot, and thus we were
-able to control all experiments via ssh.
+except HyperNEAT, is in Python and is available on our website
+\cite{quadraWeb}.  HyperNEAT is written in C++.  We controlled the servos with the \code{pydynamixel} library \cite{pydynamixel}.  The robot
+connects to a wireless network on boot, which enabled us to control it via SSH.
 
@@
 
-To track the position of the robot and determine gait fitness, we
+To track the position of the robot and thus determine gait fitness, we
 mounted a Nintendo Wii remote on the ceiling and an infrared LED on
-top of the robot (\figref{wiiMote.JPG}).  The Wii remote contains an
+top of the robot (\figref{wiiMote_crop.jpg}).  The Wii remote contains an
 IR camera that can track and report the position of any IR sources in
@@
 server via socket and requested position updates at the beginning and
-end of each run (see \secref{fitnessEvaluation for more details}).
+end of each run (see \secref{fitnessEvaluation}).
 
@@
 
-To run a gait on the robot, one simply provides a \emph{gait
-  function}, any Python function that accepts a single input --- time
+To run a gait on the robot, a \emph{gait
+  function} is provided, any Python function that accepts a single input --- time
 starting at 0 --- and outputs a list of nine commanded positions, one

### 2011-02-09 paper/relatedWork.tex
lots of editing of the intro/motivation/previouswork, and some in the robot platform

@@
 
-\editbox{If we talk about past gait-learning work in here, is it a bit redundant?}
-
-\edit{write this.  Old quadratot section below}
-
-% Briefly explain who else worked on related problems in the past and what
-% methods they used. Explain if you are using similar methods, or if your
-% approach is different and if so - how (either is ok).
-
-Various maching learning techniques have proven to be useful in
-finding control policies for a wide variety of robots. Kohl and
-Stone\cite{kohl} presented a policy gradient reinforcement learning
-approach for generating a fast walk on legged robots. We experimented
-with this method to create a walk for our robot (Policy Gradient
-Descent). Chernova and Velosa\cite{chernova} took an evolutionary
-approach to this problem which we did not implement.  Zykov, Bongard,
-and Lipson\cite{zykov} describe the evolution of dynamic gaits on a
-physical robot requiring no prior assumptions about the locomotion
-pattern beyond the fact that it should be rhythmic.
+Various machine learning techniques have proven to be effective at generating gaits for legged robots.
+Kohl and Stone presented a policy gradient reinforcement learning approach for generating a fast walk on legged robots\cite{kohl}.
+We experiment with this method to create a walk for our robot (called Policy Gradient Descent, described below).
+Others have evolved gaits for legged robots, producing competitive results~\cite{chernova2005evolutionary, hornby2005autonomous, zykov, clune2009evolving, clune2011performance, clune2009hybrid, clune2009sensitivity, tellez2006evolving, valsalam2008modular}.
+In fact, an evolved gait was used in the first commercially-available version of Sony's AIBO robot~\cite{hornby2005autonomous}.
+Except for work with HyperNEAT~\cite{clune2009evolving, clune2011performance, clune2009hybrid, clune2009sensitivity}, the previous evolutionary approaches have helped evolution 'see' the regularity of the problem by manually decomposing the task.
+In other words, experimenters have to choose which legs should be coordinated, or otherwise facilitate the coordination of robotic legs.
+Part of the motivation of this paper is to see if the space of regularities can be explored automatically, which has previously performed well~\cite{valsalam2008modular}, and to perform a direct comparison on the same robot to HyperNEAT.

### 2011-02-09 paper/fitnessEvaluation.tex
Slight approach restructuring, platform details ~done, fixed wiimote fig
Some half completed fitnessEvaluation edits

@@
-\subsection{Fitness evaluation details}
+\subsection{Fitness Evaluation Details}
 \seclabel{fitnessEvaluation}
+
+% Points to make:
+% - linear interpolation
+% - fitness reported in pixels
+% - parametrized models
+%   - started all on three thetas
+%   - plateued at 1/3 of lifetime
+% - HyperNEAT
+%   - broke things a lot... so guided evolution (or in HN section?)
+%   - eventually included ``shimmy''
+% - finite sized window led to unfortunate bias
+%
+% Given earlier section, now have way of evaluating any gait g(t).
+%
+% Want to measure average speed, so we measure distance displaced over
+% a fixed lenght run. Because we're measureing a single point on
+% robot, need to start and end in same position (crouch)
+% So we linearly interpolate begining (1s) and end (2s).
+%
+% Fitness is measured and reported directly as euclidian distance of pixels moved: equation
+%
+% For the parametrized models, the starting location may be important. To be as fair as possible, we picked three random points and starred all models there.
+%   generally ran until 1/3 plateau
+%
+% For HN, couldn't do a,b,c thing, so just did three runs of length <foo>
+% Things started breaking, so included shimmy to test brokenness
+% penalized runs that broke servos
+%
+% Finite sized window kinda hurt stuff.
+
+As mentioned earlier, the metric for evaluating gaits was their
+average speed.  In reality this was measured as the distance the robot
+moved over a 12 second run, measured by the Wii remote tracking system
+(\secref{platformDetails}).  The Euclidean distance between the
+starting position and the ending position, in pixels, was reported
+directly as the fitness (save for a tweak for HyperNEAT; see later in
+this section).
+
+Since only a single point on the robot --- the IR LED --- was measured
+for the purposes of computing fitness, it was important that the
+position of the IR LED accurately reflect the position of the robot as
+a whole.  To enforce this constraint, the robot was always measured
+while in the \emph{ready} position (see \figref{robot_close.jpg}).  If
+this was not done, it would have resulted in giving extra fitness to,
+for example, gaits that ended with the robot leaning into the
+direction of travel.  While this might mean that a part of the robot
+has traveled farther, this extra distance would not generalize to a
+longer run.  Thus, we chose not to reward this behavior.
+
+Wishing to measure start and end position in the same pose,
+
+In order to ensure we could
+start and end each run in the same position,
+
+\begin{tabular}{ccccc}
+Ready position & a & b & c & d\\
+\hline
+t = 0, measure  & $t \in [0,1]$ & $t \in [1,10]$ & $t \in [10,12]$ & t = 12
+\end{tabular}
 
@@
 
-The size of the Wii remote window is approximately 69 x 47 cm. A
+The size of the Wii remote window is approximately 175 x 120 cm. A
 concern about this is that if the robot walks further than usual,
@@
 % probably could be better written
+
+The only human intervention required during most learning was to
+occasionally move the robot back into the viewable area of the w
 

### 2011-02-09 paper/experimentalSetup.tex
Cleaned up main.tex

\section{Experimental Setup}
\seclabel{experimentalSetup}

\secref{platformDetails} discusses the hardware platform used for the
experiments presented in this paper, and \secref{fitnessEvaluation}
describes the method we used to measure a gait's fitness.

% Given the goal of producing a deterministic, feedforward motion
% function $g(t)$, we now turn to several important considerations.
% \secref{learningMethods} talks about how this function is generated,
% \secref{platformDetails} discusses the hardware platform on which it
% is run, and \secref{fitnessEvaluation} details exactly how its
% fitness is measured.

\input{platformDetails}

\input{fitnessEvaluation}

### 2011-02-09 paper/experimentalSetup.tex
Fitness cleanup and rewording, shuffled some points between sections

@@
 
-\secref{platformDetails} discusses the hardware platform used for the
-experiments presented in this paper, and \secref{fitnessEvaluation}
-describes the method we used to measure a gait's fitness.
+An important aspect of this study was to evaluate gait generation and
+learning algorithms in the noisy, unforgiving real world, instead of
+using a simulator.  In this section we give details of the hardware
+setup we used: \secref{platformDetails} discusses the robotic
+platform, and \secref{fitnessEvaluation} describes the tracking system
+we used to measure each gait's fitness.
 

### 2011-02-09 paper/fitnessEvaluation.tex
Fitness cleanup and rewording, shuffled some points between sections

@@
 
+\acmFigg{wiiMote_crop.jpg}{robot_led_crop.jpg}{1}{A Nintendo Wii
+  remote provided the location of the robot by tracking the infrared
+  LED mounted on the robot's antenna.  The LED position was measured
+  in pixels and transmitted from the Wii remote to the robot via
+  bluetooth.}
+
+To track the position of the robot and thus determine gait fitness, we
+mounted a Nintendo Wii remote on the ceiling and an infrared LED on
+top of the robot (\figref{wiiMote_crop.jpg}).  The Wii remote contains
+an IR camera that can track and report the position of up to four IR
+sources in its image frame.  The resolution of the camera was 1024 by
+768 pixels with view angles of about 40$^\circ$ by 30$^\circ$, which
+produced a resolution of 1.7mm per pixel when mounted at a height of
+2.63m.  At this height, the viewable window on the floor was
+approximately 175 x 120 cm.
+
+A separate tracking server, written in Python, ran on the robot and
+interfaced with the Wii remote via bluetooth using the CWiid
+library\cite{cwiid}.  Our fitness testing code communicated with this
+server via socket connection and requested position updates at the
+beginning and end of each run.
+
 As mentioned earlier, the metric for evaluating gaits was their
 average speed.  In reality this was measured as the distance the robot
-moved over a 12 second run, measured by the Wii remote tracking system
-(\secref{platformDetails}).  The Euclidean distance between the
+moved over a 12 second run.  The Euclidean distance between the
 starting position and the ending position, in pixels, was reported
-directly as the fitness (save for a tweak for HyperNEAT; see later in
-this section).
+directly as the fitness (save for a tweak for HyperNEAT; see below).
 
@@
 a whole.  To enforce this constraint, the robot was always measured
-while in the \emph{ready} position (see \figref{robot_close.jpg}).  If
-this was not done, it would have resulted in giving extra fitness to,
-for example, gaits that ended with the robot leaning into the
-direction of travel.  While this might mean that a part of the robot
-has traveled farther, this extra distance would not generalize to a
-longer run.  Thus, we chose not to reward this behavior.
+while in the \emph{ready} position (the position shown in
+\figref{robot_close.jpg}).  This was done to prevent assigning extra
+fitness to, for example, gaits that ended with the robot leaning into
+the direction of travel.  While this would mean that a part of the
+robot has traveled farther, this extra distance would not generalize
+to a longer run.  Thus, we chose not to reward this behavior.
 
-Wishing to measure start and end position in the same pose,
+In order to measure start and end position in the same pose,
 
-In order to ensure we could
+ensure we could
 start and end each run in the same position,

### 2011-02-09 paper/platformDetails.tex
Fitness cleanup and rewording, shuffled some points between sections

@@
 to each other. \figref{topdown.png} shows the positions and numerical
-designations of all nine servoes.
+designations of all nine servos.  Each servo could be commanded to a
+position in the range [0,~1023], corresponding to a physical range
+[-120$^{\circ}$,~+120$^{\circ}$].
 
-All of the computation for gait learning, fitness evaluation, and robot control
-was performed on the compact on-board CompuLab Fit-PC2, running Ubuntu
-Linux.  All gait generation, learning, and fitness evaluation code,
-except HyperNEAT, is in Python and is available on our website
-\cite{quadraWeb}.  HyperNEAT is written in C++.  We controlled the servos with the \code{pydynamixel} library \cite{pydynamixel}.  The robot
-connects to a wireless network on boot, which enabled us to control it via SSH.
+All of the computation for gait learning, fitness evaluation, and
+robot control was performed on the compact on-board CompuLab Fit-PC2,
+running Ubuntu Linux.  All gait generation, learning, and fitness
+evaluation code, except HyperNEAT, was written in Python and is
+available on our website \cite{quadratotWeb}.  HyperNEAT is written in
+C++.  We controlled the servos with the \code{pydynamixel} library
+\cite{pydynamixel}.  The robot connected to a wireless network on
+boot, which enabled us to control it via SSH.
 
-%\acmFigg{wiiMote.jpg}{robot_led.jpg}{1}{A Nintendo Wii remote tracks the
-%  location of the robot, providing feedback about distance traveled,
-%  in pixels, through an infrared LED mounted on top of the robot.}
-
-\acmFigg{wiiMote_crop.jpg}{robot_led_crop.jpg}{1}{A Nintendo Wii remote tracks the
-  location of the robot, providing feedback about distance traveled,
-  in pixels, through an infrared LED mounted on top of the robot.}
-
-To track the position of the robot and thus determine gait fitness, we
-mounted a Nintendo Wii remote on the ceiling and an infrared LED on
-top of the robot (\figref{wiiMote_crop.jpg}).  The Wii remote contains an
-IR camera that can track and report the position of any IR sources in
-its image frame.  The resolution of the camera was 1024 by 768 pixels,
-which produced a resolution of 1.7mm per pixel when mounted at a
-height of \edit{???}m.
-
-A separate tracking server, written in Python, ran on the robot and
-interfaced with the Wii remote via bluetooth using the CWiid
-library\cite{cwiid}.  Our fitness testing code connected to this
-server via socket and requested position updates at the beginning and
-end of each run (see \secref{fitnessEvaluation}).
-
-%to interface with the remote via bluetooth. A client then connects via
-%a socket to the tracking server and requests position updates
-%periodically. If the robot walks beyond the viewable range of the Wii
-%remote, a prompt informed the user. The only human intervention
-%required during the experiment was to move the robot back inside the
-%viewable area and resume the run, and to handle any mechnanical
-%failures that arose. This did not interrupt the learning process or
-%result in the loss of data.
-
-To run a gait on the robot, a \emph{gait
-  function} is provided, any Python function that accepts a single input --- time
-starting at 0 --- and outputs a list of nine commanded positions, one
-for each servo.  To safeguard against limb collision with the robot
-body, the robot control code cropped the commands to a safe range.
-This range was [150, 770] for the inner leg servos, [30, 680] for the
-outer leg servos, and [392, 623] for the center hip servo.
+To run a gait on the robot, one needed only to provide a \emph{gait
+  function} --- any Python function that accepts a single input, time
+starting at 0, and outputs a list of nine commanded positions, one for
+each servo.  To safeguard against limb collision with the robot body,
+the robot control code cropped the commands to a safe range.  This
+range was [150,~770] for the inner leg servos, [30,~680] for the outer
+leg servos, and [392,~623] for the center hip servo.
 

### 2011-02-09 paper/problemDefinition.tex
Fitness cleanup and rewording, shuffled some points between sections

@@
 \section{Problem definition}
+\seclabel{problemDefinition}
+
 % Precisely define the problem you are addressing (i.e. formally specify
@@
 so to optimize for speed, our performance metric was displacement over
-a period of motion lasting approximately 10 seconds.  For details of
+a period of motion lasting approximately 12 seconds.  For details of
 exactly how this displacement was measured, see

### 2011-02-09 paper/results.tex
Added plots
Fitness cleanup and rewording, shuffled some points between sections

@@
 \subsection{Maximizing Strategies}
+
+\edit{maybe useful vvvvv}
+
+To be as fair as possible when obtaining the results presented in
+\secref{results}, we picked three random points --- $\vec{\theta}_A$,
+$\vec{\theta}_B$, and $\vec{\theta}_C$ --- and started all seven
+methods at these points...
+
+Each of the parameter ??? methods was run on 3 different initial
+parameter vectors, in order to fairly compare the algorithms. We allowed 
+each run for the parameter ??? methods to
+continue until the results plateaued (no improvement for one third of
+the policies seen so far). Three runs of HyperNEAT were completed,
+each with a different initial seed and run for 20 generations.
+
+\edit{$\wedge\wedge\wedge\wedge\wedge\wedge\wedge\wedge$}
 
@@
 used.
+
+\acmFig{neat_110115_211410_00000_002_filt}{1}{Caption here...???}
+\acmFig{neat_110115_211410_00000_002_filt_zoom}{1}{Caption here...???}

### 2011-02-09 paper/fitnessEvaluation.tex
Finished linearInterp figure

@@
 
+\acmFig{linearInterp}{1}{To ensure fair fitness measurements, the
+  position of the robot was measured at the beginning and end of each
+  run in the same ``ready pose'' (\figref{robot_close.jpg}).  Motion
+  was interpolated linearly between this stationary pose and the
+  commanded gait g(t) for one second at the beginning of the run and
+  two seconds at the end, as shown above.}
+
 \begin{tabular}{ccccc}
@@
 \end{tabular}
-
-\edit{Put this in there somewhere:}
-
-To be as fair as possible when obtaining the results presented in
-\secref{results}, we picked three random points --- $\vec{\theta}_A$,
-$\vec{\theta}_B$, and $\vec{\theta}_C$ --- and started all seven
-methods at these points...
 
@@
 occasionally move the robot back into the viewable area of the w
-
-Each of the parameter ??? methods was run on 3 different initial
-parameter vectors, in order to fairly compare the algorithms. We allowed 
-each run for the parameter ??? methods to
-continue until the results plateaued (no improvement for one third of
-the policies seen so far). Three runs of HyperNEAT were completed,
-each with a different initial seed and run for 20 generations.

### 2011-02-09 paper/fitnessEvaluation.tex
More fitness, explanation of bias due to sensor area

@@
 
-In order to measure start and end position in the same pose,
+In order to measure start and end position in the same pose, we
+linearly interpolated the motion of the robot between the ready
+position and the commanded gait, $g(t)$.  As shown in
+\figref{linearInterp}, motion for the first second was interpolated
+between the ready position and $g(t)$, for the next nine seconds the
+robot moved exactly as specified by $g(t)$, and for the final two
+seconds its motion was interpolated smoothly back to the ready
+position.
 
-ensure we could
-start and end each run in the same position,
+\acmFig{linearInterp}{1}{To ensure fair fitness evaluations with as
+  little noise as possible, the position of the robot was measured at
+  the beginning and end of each run (red circles) in the same
+  \emph{ready} pose (\figref{robot_close.jpg}).  Motion was
+  interpolated linearly between this stationary pose and the commanded
+  gait g(t) for one second at the beginning of the run and two seconds
+  at the end, as shown above.}
 
-\acmFig{linearInterp}{1}{To ensure fair fitness measurements, the
-  position of the robot was measured at the beginning and end of each
-  run in the same ``ready pose'' (\figref{robot_close.jpg}).  Motion
-  was interpolated linearly between this stationary pose and the
-  commanded gait g(t) for one second at the beginning of the run and
-  two seconds at the end, as shown above.}
+% finite size
+% shimmy
 
-\begin{tabular}{ccccc}
-Ready position & a & b & c & d\\
-\hline
-t = 0, measure  & $t \in [0,1]$ & $t \in [1,10]$ & $t \in [10,12]$ & t = 12
-\end{tabular}
+The only human intervention required during most learning trials was
+to occasionally move the robot back into the viewable area of the Wii
+remote if it had walked outside.  Initially this was a rare
+occurrence, as the gaits did not typically produce motion as large as
+the size of the window (roughly 175 x 120 cm).  However, as gaits
+improved, particularly when using HyperNEAT, the robot began to walk
+out of the measurement area a non-negligible fraction of the time.
+Whenever it did so, we would discard the trial and repeat it until the
+gait was entirely within the window.
 
-\editbox{proofread. especially middle paragraph}
+While this process guaranteed that we always obtained a measurement
+for a given gait before proceeding, it also biased the measurements
+downward.  Because the performance of the robot on a given gait varied
+from trial to trial, a successful measurement was more likely to be
+obtained when the gait happened to perform poorly.  This phenomenon
+was negligible at first, but became significantly more pronounced when
+evaluating several gaits that produced motion across nearly the entire
+area.  For future studies we will likely use a 2x2 array of Wii
+remotes to double the length and width of the measurement arena.
 
-%\section{Experimental Evaluation}
-
-The metric for evaluation of the designed gait was speed. To evaluate
-a set of parameters, the robot was sent the parameters and instructed
-to walk for a certain length of time. For each evaluation, the robot
-always started and ended in the same position in order to measure true
-displacement and not reward gaits the ended in a lean, since the LED
-would have moved a different distance the robot. More efficient
-parameters resulted in a faster gait, which translated into a longer
-distance walked, measured in pixels, and a better score.
-
-The size of the Wii remote window is approximately 175 x 120 cm. A
-concern about this is that if the robot walks further than usual,
-it walks outside the viewable range and thus does not count the run.
-This possibility could bias the final results reported on the robot.
-% probably could be better written
-
-The only human intervention required during most learning was to
-occasionally move the robot back into the viewable area of the w
+%We made one final tweak to the fitness measurement process.  HN..... 1/2

### 2011-02-09 paper/fitnessEvaluation.tex
bit for jeff
edited methods section

@@
 library\cite{cwiid}.  Our fitness testing code communicated with this
-server via socket connection and requested position updates at the
+server via a socket connection and requested position updates at the
 beginning and end of each run.
@@
 As mentioned earlier, the metric for evaluating gaits was their
-average speed.  In reality this was measured as the distance the robot
-moved over a 12 second run.  The Euclidean distance between the
-starting position and the ending position, in pixels, was reported
-directly as the fitness (save for a tweak for HyperNEAT; see below).
+average speed, which was measured as the Euclidian distance the robot
+moved during a 12-second run.
+For the manual and parameterized gaits, the fitness was this value.
+The HyperNEAT gaits stressed the motors more than the other gaits, so to encourage gaits that did not tax the motors we penalized gaits that caused the servos to stop responding.
+The penalty was to set the fitness to half of the distance the robot actually traveled.
+We tested whether the servos were responding after each gait by commanding them to specific positions and checking whether they actually moved to those positions.
+This test had the additional benefit of rewarding those gaits that did not flip the robot into a position where it could not move its legs, which HyperNEAT also did more than the other learning methods.
+Because the fitness of HyperNEAT gaits were often halved, in results we compare actual distance traveled in addition to fitness for the best gaits produced by each class of gait-generating algorithms.
 
@@
 \figref{robot_close.jpg}).  This was done to prevent assigning extra
-fitness to, for example, gaits that ended with the robot leaning into
-the direction of travel.  While this would mean that a part of the
-robot has traveled farther, this extra distance would not generalize
-to a longer run.  Thus, we chose not to reward this behavior.
+fitness to, for example, gaits that ended with the robot leaning toward
+the direction of travel (this extra distance would not likely generalize
+in a longer run, which is why we did not want to reward this behavior).
 
-In order to measure start and end position in the same pose, we
+In order to measure the start and end position in the same pose, we
 linearly interpolated the motion of the robot between the ready
@@
 to occasionally move the robot back into the viewable area of the Wii
-remote if it had walked outside.  Initially this was a rare
+remote.  Initially this was a rare
 occurrence, as the gaits did not typically produce motion as large as
@@
 evaluating several gaits that produced motion across nearly the entire
-area.  For future studies we will likely use a 2x2 array of Wii
+area.
+HyperNEAT gaits were especially likely to require additional trials, meaning that the reported performance for HyperNEAT is worse than it would have been otherwise.
+For future studies we will likely use a 2x2 array of Wii
 remotes to double the length and width of the measurement arena.
-
-%We made one final tweak to the fitness measurement process.  HN..... 1/2

### 2011-02-09 paper/learningMethods.tex
Edits to learningMethods, took out SVM for now

@@
 The five parameters in $\vec{\theta}$ are as follows:
-
-\edit{make sure this text flows into the table}
 
@@
 \noindent Here the fixed constants that determine the crouched
-position are $C_I$ for inner motors, $C_O$ for outer motors, and $C_C$
-for the center hip motor.  The nine subscripted functions correspond
-to the nine labeled motors in \figref{topdown.png}.
+position are $C_I = 800$ for inner motors, $C_O = 40$ for outer
+motors, and $C_C = 512$ for the center hip motor.  The nine functions
+correspond to the nine motors in the same order as listed in
+\figref{topdown.png}.
 
@@
 $\vec{\theta}$, then a given physical trial gives us one measurement
-of the fitness of that parameter vector, or $f(\vec{\theta})$.  Two
+of the fitness $f(\vec{\theta})$ of that parameter vector.  Two
 things make learning difficult.  First, each evaluation of
@@
 general this is done by selecting a random starting vector within the
-allowable domain.  However, as discussed in
-\secref{fitnessEvaluation}, we decided to evaluate the different
-methods starting at one of several common starting points to reduce
-one possible contribution of noise.
+allowable domain.  However, to most directly compare learning methods
+by their ability to improve on the initially chosen $\vec{\theta}$, we
+decided to evaluate the different methods by starting them at the same
+initial $\vec{\theta}$ vectors.
 
@@
 
-\item \emph{Uniform random hill climbing}: This method begins by
-  selecting a single random parameter vector.  Subsequent iterations
-  generate a neighbor by randomly choosing one parameter to adjust and
-  replacing it with a new value chosen with uniform probability in the
-  allowable range for that parameter. The neighbor is evaluated by
-  running the robot with the newly chosen parameters. If this neighbor
-  results in a longer distance walked than the previous best gait, it
-  is saved as the new best gait. The process is then repeated, always
-  starting with the best gait.
+\item \emph{Uniform random hill climbing}: This method always starts
+  with the current best gait, and then chooses next $\vec{\theta}$ by
+  randomly choosing one parameter to adjust and replacing it with a
+  new value chosen with uniform probability in the allowable range for
+  that parameter. This new point is evaluated, and if it results in a
+  longer distance walked than the previous best gait, it is saved as
+  the new best gait. The process is then repeated, always starting
+  with the best gait.
 
 \item \emph{Gaussian random hill climbing}: This method works
-  similarly to Uniform random hill climbing, except neighbors are
-  generated by adding random Gaussian noise to the current best gait.
-  This results in all parameters being changed at once, but the
-  resulting vector is always fairly close to the previous best gait.
-  We used independently selected noise in each dimension, scaled such
-  that the standard deviation of the noise was 5\% of the range of
-  that dimension.
+  similarly to Uniform random hill climbing, except the next
+  $\vec{\theta}$ is generated by adding random Gaussian noise to the
+  current best gait.  This results in all parameters being changed at
+  once, but the resulting vector is always fairly close to the
+  previous best gait.  We used independently selected noise in each
+  dimension, scaled such that the standard deviation of the noise was
+  5\% of the range of that dimension.
 
-\item \emph{N-dimensional policy gradient descent}: In \cite{kohl},
-  Kohl and Stone document a method for local gradient ascent for gait
-  learning with noisy fitness evaluations, and we have included this
-  method in our evaluation.  This strategy explicitly estimates the
-  gradient for the objective function. It does this by first
-  evaluating \emph{n} randomly generated parameter vectors near the
-  initial vector, each dimension of these vectors being perturbed by
-  either $-\epsilon$, $0$, or $\epsilon$. Then, for each dimension, it
-  groups vectors into three groups: $-\epsilon$, $0$, and $\epsilon$.
-  The gradient along this dimension is then estimated as the average
-  score for the $\epsilon$ group minus the average score for the
-  $-\epsilon$ group. Finally, the method creates a new vector by
-  changing all parameters by a fixed-size step in the direction of the
-  gradient.
+\item \emph{N-dimensional policy gradient ascent}: We implemented Kohl
+  and Stone's \cite{kohl} method for local gradient ascent for gait
+  learning with noisy fitness evaluations. This strategy explicitly
+  estimates the gradient of the objective function. It does this by
+  first generating $n$ parameter vectors near the initial vector by
+  perturbing each dimension of each vector randomly by either
+  $-\epsilon$, $0$, or $\epsilon$. Then each vector is run on the
+  robot, and for each dimension we group the results into three
+  groups: $-\epsilon$, $0$, and $\epsilon$.  The gradient along this
+  dimension is then estimated as the average score for the $\epsilon$
+  group minus the average score for the $-\epsilon$ group. Finally,
+  the method creates the next $\vec{\theta}$ by changing all
+  parameters by a fixed-size step in the direction of the gradient.
+  For this study we used values of $\epsilon$ equal to 5\% of the
+  allowable range in each dimension (ranges listed in
+  \tabref{parameters}), and a step size scaled such that if all dimensions
+  were in the range [0,~1], the norm of the step size would be 0.1.
 
@@
   of the allowable range for that parameter. It then tests the fitness
-  of each vertex and based on these fitnesses, it reflects the worst
-  point over the centroid in an attempt to improve it.  In general,
-  the worst vertex is reflected; however, to prevent cycles and
-  becoming stuck in local minima, several other rules are used.  If
-  the reflected point is better than the second worst point and worse
-  than the best point, then the reflected point replaces the worst. If
-  the reflected point is better than the best point, the simplex is
-  expanded in the direction of the reflected point. The better of the
-  reflected and the expanded point replaces the worst point. If the
-  reflected point is worse than the second worst point, then the
-  simplex is contracted away from the reflected point. If the
+  of each vertex, and, in general, it reflects the worst point over
+  the centroid in an attempt to improve it.  However, to prevent
+  cycles and becoming stuck in local minima, several other rules are
+  used.  If the reflected point is better than the second worst point
+  and worse than the best point, then the reflected point replaces the
+  worst. If the reflected point is better than the best point, the
+  simplex is expanded in the direction of the reflected point. The
+  better of the reflected and the expanded point replaces the worst
+  point. If the reflected point is worse than the second worst point,
+  then the simplex is contracted away from the reflected point. If the
   contracted point is better than the reflected point, the contracted
@@
   new linear model to all vectors evaluated so far, choosing the model
-  to minimize the sum of squared errors.
+  to minimize the sum of squared errors. Step size is the same as in
+  \emph{N-dimensional policy gradient ascent}.
 
-\item \emph{SVM regression}: Similarly to linear regression, this
-  model starts with several random vectors, but this time they are
-  chosen in a small neighborhood about some initial random vector.
-  These vectors (generally 8) are evaluated, and a support vector
-  regression model is fit to the observed fitnesses.  To choose the
-  next vector for evaluation, we randomly generate some number
-  (typically 100) of vectors in the neighborhood of the best observed
-  gait, and select for evaluation the vector with the best predicted
-  performance.  We suspected that if we always chose the best
-  predicted point out of 100, we may end up progressing along a narrow
-  subspace, prohibiting learning of the true local fitness function.
-  Put another way, we would always choose exploitation of knowledge
-  vs. exploration of the space.  To address this concern, we added a
-  parameter dubbed \code{bumpBy} that added noise to the final
-  selected point before it was evaluated.
-
-  Such a method naturally has many tunable parameters, and we
-  endeavored to select these parameters by tuning the method in
-  simulation.  To estimate the performance of the algorithm, we ran it
-  against a simulation with a known optimum.  The simulated function
-  was in the same five dimensional parameter space, and simply
-  returned a fitness determined as the height of a Gaussian with a
-  random mean.  The width of the Gaussian in each dimension was 20\%
-  of the range of each dimension, and the maximum value at the peak
-  was 100.  \figref{svm_sim_results} shows the learning results on
-  this simulated model using the ultimately selected SVM parameters.
-  Interestingly, a non-zero value of \code{bumpBy} resulted in better
-  learning than noise free (exploration free) learning.
-
-  \edit{JMC: This next paragraph sounds like it belongs in Results?}
-  Ultimately, however, the version of SVM tuned for simulation still
-  did not show competitive performance on the real robot.  We tried
-  tuning some parameters on the real robot, but after some amount of
-  tuning, the method still exhibited too little exploration and easily
-  became stuck in local minima.
+%\item \emph{SVM regression}: Similarly to linear regression, this
+%  model starts with several random vectors, but this time they are
+%  chosen in a small neighborhood about some initial random vector.
+%  These vectors (generally 8) are evaluated, and a support vector
+%  regression model is fit to the observed fitnesses.  To choose the
+%  next vector for evaluation, we randomly generate 100 vectors in the
+%  neighborhood of the best observed gait, and select for evaluation
+%  the vector with the best predicted performance in the learned model.
+%  We suspected that if we always chose the best predicted point out of
+%  100, we may end up progressing along a narrow subspace, prohibiting
+%  learning of the true local fitness function.  Put another way, we
+%  would always choose exploitation of knowledge vs. exploration of the
+%  space.  To address this concern, we added a small amount of noise to
+%  the new chosen $\vec{\theta}$ before continuing with the next phase
+%  of evaluation and modeling.
+%
+%  Such a method naturally has many tunable parameters, and we
+%  endeavored to select these parameters by tuning the method in
+%  simulation.  To estimate the performance of the algorithm, we ran it
+%  against a simulation with a known optimum.  The simulated function
+%  was in the same five dimensional parameter space, and simply
+%  returned a fitness determined as the height of a Gaussian with a
+%  random mean.  The width of the Gaussian in each dimension was 20\%
+%  of the range of each dimension, and the maximum value at the peak
+%  was 100.  \figref{svm_sim_results} shows the learning results on
+%  this simulated model using the ultimately selected SVM parameters.
+%  Interestingly, a non-zero value of \code{bumpBy} resulted in better
+%  learning than noise free (exploration free) learning.
+%
+%  \edit{JMC: This next paragraph sounds like it belongs in Results?}
+%  Ultimately, however, the version of SVM tuned for simulation still
+%  did not show competitive performance on the real robot.  We tried
+%  tuning some parameters on the real robot, but after some amount of
+%  tuning, the method still exhibited too little exploration and easily
+%  became stuck in local minima.
 

### 2011-02-09 paper/learningMethods.tex
changes

--- 2011-01-31 paper/learningMethods.tex
@@
 
-There are a plethora of methods available for the creation of the
-vector motion generating function $g(t)$ introduced in
-\secref{problemDefinition}, and we've chosen two for comparison in
-this study.  The first method we evaluated was to use a family of
-parametrized gaits, described in this section and the next, and the
-second method was to use an evolved non-parametric gait, detailed in
+We have chosen to implement two classes of algorithms to automatically create the motion-generating function $g(t)$ (described in
+\secref{problemDefinition}).  The first method uses a family of
+parametrized gaits, described in this section and the next. The
+second method evolves a non-parametric gait controlled by neural networks, and is detailed in
 \secref{hyperNeatMethod}.
@@
 tried several parametrizations on the robot and, upon obtaining
-reasonable early success, settled on one particular parametrization.
-We dubbed this the \emph{SineModel5}, as it has five parameters and
-employs a sine wave as its root pattern.
-
-The five parameters in $\vec{\theta}$ are as follows:
-
-\edit{make sure this text flows into the table}
+reasonable early success, settled on one particular parametrization, which we call  \emph{SineModel5}. Its root pattern is a sine wave and it has five parameters (Table~\ref{tab:params}) .
 
@@
 \hline
-$m_O$        & Outer motor multiplier    & [-2, 2] \\
+$m_O$        & Outer-motor multiplier    & [-2, 2] \\
 \hline
-$m_F$        & Front motor multiplier    & [-1, 1] \\
+$m_F$        & Front-motor multiplier    & [-1, 1] \\
 \hline
-$m_R$        & Right motor multiplier    & [-1, 1] \\
+$m_R$        & Right-motor multiplier    & [-1, 1] \\
 \hline
 \end{tabular}
-\caption{The parameters used for the \emph{SineModel5} motion model.}
+\caption{The \emph{SineModel5} motion model parameters.}
 \tablabel{parameters}
+\label{tab:params}
 \end{center}
@@
 
-Intuitively, SineModel5 starts with 8 identically sine waves of
+\edit{called outer multiplier, but text says we multiply by inner. same for left}
+
+Intuitively, SineModel5 starts with 8 identical sine waves of
 amplitude $\amp$ and period $\tau$, multiplies the waves for all inner
-motors by $m_o$, multiplies the waves for all front motors by $m_F$,
+motors by $m_O$, multiplies the waves for all front motors by $m_F$,
 and multiplies the waves for all left motors by $m_R$.  To obtain the
@@
 waves are at 0) is approximately a crouch (the position shown in
-\figref{robot_close.jpg}).  Finally, in this motion model, the ninth
-``hip'' motor is kept at a neutral position.  Thus, the commanded
+\figref{robot_close.jpg}).  In this motion model, the ninth
+(center) motor is kept at a neutral position.  Thus, the commanded
 position for all motors, as a vector function of time, is:
@@
 
+\edit{what subscripts?}
+
 \subsection{Learning Methods for Parametrized Motion Models}
@@
 things make learning difficult.  First, each evaluation of
-$f(\vec{\theta})$ is expensive, taking 15-20 seconds of time on
+$f(\vec{\theta})$ is expensive, taking 15-20 seconds on
 average.  Second, the fitness returned by such evaluations has proved
@@
 
-For both of these reasons, an intelligent \emph{learning algorithm} is
-needed.
-In this context, a learning algorithm is a method for choosing
-%\begin{enumerate}
-%\item The value for $\vec{\theta}$ for the initial trial, and
-the next value of $\vec{\theta}$ to try, given a list of the
-$\vec{\theta}$ values tried so far and the fitness $f(\vec{\theta})$
-measured for each.
-%\end{enumerate}
+We test the ability of different \emph{learning algorithms} to choose the next value of $\vec{\theta}$ to try, given a list of the
+$\vec{\theta}$ values already evaluated and their fitness measurements $f(\vec{\theta})$.
 
-We evaluated a total of seven subtly or not so subtly different
-learning methods for the parametrized motion models.  All methods
-needed a way of picking a $\vec{\theta}$ for the first trial.  In
-general this is done by selecting a random starting vector within the
+We evaluated a total of seven different
+learning algorithms for the parametrized motion models.  All methods
+needed a way of picking a $\vec{\theta}$ for the first trial of each run.  This could be done by selecting a random starting vector within the
 allowable domain.  However, as discussed in
-\secref{fitnessEvaluation}, we decided to evaluate the different
-methods starting at one of several common starting points to reduce
-one possible contribution of noise.
+\secref{fitnessEvaluation}, we started all of the methods at the same three starting points in the three different runts to make comparisons fair.
 
@@
 
-The seven methods used, and their methods for choosing the next
-$\vec{\theta}$ are:
+The seven learning algorithms for the parametrized motion models are as follows:
 
@@
 \item \emph{Random}: This method randomly generates parameter vectors
-  in the allowable range for every trial. This strategy was used only
+  in the allowable range for every trial. This strategy serves
   as baseline for comparison.
 
-\item \emph{Uniform random hill climbing}: This method begins by
+\item \emph{Uniform random hill climbing}: \edit{this contradicts what we say earlier, where we say that we use the same initial vector for all runs in the 3 different treatments} This method begins by
   selecting a single random parameter vector.  Subsequent iterations
@@
 
-\item \emph{N-dimensional policy gradient descent}: In \cite{kohl},
-  Kohl and Stone document a method for local gradient ascent for gait
-  learning with noisy fitness evaluations, and we have included this
-  method in our evaluation.  This strategy explicitly estimates the
-  gradient for the objective function. It does this by first
-  evaluating \emph{n} randomly generated parameter vectors near the
-  initial vector, each dimension of these vectors being perturbed by
+\item \emph{N-dimensional policy gradient descent}: This method, presented in \cite{kohl},
+  features local gradient \edit{descent?} ascent for gait
+  learning with noisy fitness evaluations.  It explicitly estimates the
+  gradient for the objective function by first
+  evaluating \emph{n} randomly-generated parameter vectors near the
+  initial vector, with each dimension of the vector being perturbed by
   either $-\epsilon$, $0$, or $\epsilon$. Then, for each dimension, it

### 2011-02-09 paper/learningMethods.tex

--- 2011-02-09 paper/learningMethods.tex
@@
 tried several parametrizations on the robot and, upon obtaining
-reasonable early success, settled on one particular parametrization.
-We dubbed this the \emph{SineModel5}, as it has five parameters and
-employs a sine wave as its root pattern.
-
-The five parameters in $\vec{\theta}$ are as follows:
+reasonable early success, settled on one particular parametrization, which we call  \emph{SineModel5}. Its root pattern is a sine wave and it has five parameters (Table~\ref{tab:params}) .
 
@@
 
-For both of these reasons, an intelligent \emph{learning algorithm} is
-needed.
-In this context, a learning algorithm is a method for choosing
-%\begin{enumerate}
-%\item The value for $\vec{\theta}$ for the initial trial, and
-the next value of $\vec{\theta}$ to try, given a list of the
-$\vec{\theta}$ values tried so far and the fitness $f(\vec{\theta})$
-measured for each.
-%\end{enumerate}
+We test the ability of different \emph{learning algorithms} to choose the next value of $\vec{\theta}$ to try, given a list of the
+$\vec{\theta}$ values already evaluated and their fitness measurements $f(\vec{\theta})$.
 
-We evaluated a total of seven subtly or not so subtly different
-learning methods for the parametrized motion models.  All methods
-needed a way of picking a $\vec{\theta}$ for the first trial.  In
-general this is done by selecting a random starting vector within the
+We evaluated a total of seven different
+learning algorithms for the parametrized motion models.  All methods
+needed a way of picking a $\vec{\theta}$ for the first trial of each run.  This could be done by selecting a random starting vector within the
 allowable domain.  However, to most directly compare learning methods
 by their ability to improve on the initially chosen $\vec{\theta}$, we
-decided to evaluate the different methods by starting them at the same
-initial $\vec{\theta}$ vectors.
+decided to evaluate the different methods by starting them at the same three
+initial $\vec{\theta}$ vectors in the three runs.
 
@@
 \item \emph{Random}: This method randomly generates parameter vectors
+
   in the allowable range for every trial.  This strategy was used only

### 2011-02-09 paper/results.tex

@@
 
-As mentioned above, the SVM strategy involved many tunable
-parameters. The version of SVM tuned for simulation did not show
-competitive performance on the real robot. We tried tuning some
-parameters on the real robot, but after some amount of tuning, the
-method still exhibited too little exploration and easily became stuck
-in local minima.
-
-\edit{say something about tuning SVM and the sim figure}
-
-\acmFig{svm_sim_results}{1}{Results for the SVM regression strategy
-  in simulation.  This simulation was used to tune the SVM strategy's
-  parameters before trying it on the physical robot.  The strategy
-  quickly approaches the maximum simulated fitness of 100.}
+%As mentioned above, the SVM strategy involved many tunable
+%parameters. The version of SVM tuned for simulation did not show
+%competitive performance on the real robot. We tried tuning some
+%parameters on the real robot, but after some amount of tuning, the
+%method still exhibited too little exploration and easily became stuck
+%in local minima.
+%
+%\edit{say something about tuning SVM and the sim figure}
+%
+%\acmFig{svm_sim_results}{1}{Results for the SVM regression strategy
+%  in simulation.  This simulation was used to tune the SVM strategy's
+%  parameters before trying it on the physical robot.  The strategy
+%  quickly approaches the maximum simulated fitness of 100.}
 

### 2011-02-09 code/Strategy.py
minor cleanup

@@ -628,8 +628,8 @@ class LinearRegressionStrategy(LearningStrategy):
     def predict_distance_walked(self, inputs):
         '''
         Given a weight vector and an input vector, predicts the distance
-        walked by the robot.        
-<        '''
+        walked by the robot.
+        '''
         return sum([self.theta[i]*inputs[i] for i in range(len(self.theta))])
 
     def calculate_weights(self, training_params, distances_walked):

### 2011-02-09 paper/hyperNeatMethod.tex
hyperneat edits

@@
 \seclabel{hyperNeatMethod}
-
-\emph{Neuroevolution (Evolving Artificial Neural Networks with HyperNEAT)}:
 
@@
 phenotype~\cite{stanley2009hypercube}. Each CPPN is itself a directed
-graph, where each node is a mathematical function, such as sine or
+graph, where the nodes in the graph are mathematical functions, such as sine or
 Gaussian. The nature of these functions can facilitate the evolution
-of properties such as symmetry (e.g., an absolute value or Gaussian
-function) and repetition (e.g., a sine
+of properties such as symmetry (e.g.\ a Gaussian function) and repetition (e.g.\ a sine
 function)~\cite{stanley2009hypercube, stanley2007compositional}. The
 signal on each link in the CPPN is multiplied by that link's weight,
-which can alter its effect.
+which can magnify or diminish its effect.
 
-A CPPN is queried once for each link in the ANN substrate to determine
+A CPPN is queried once for each link in the ANN phenotype to determine
 that link's weight. The inputs to the CPPN are the Cartesian
-coordinates of both the source (e.g., $x = 2$, $y = 4$) and target
-(e.g., $x = 3$, $y = 5$) nodes of a link and a constant bias
-value. The CPPN takes these five values as inputs and produces one or
-two output values, depending on the substrate topology. If there is no
-hidden layer in the substrate, the single output is the weight of the
-link between a source node on the input layer and a target node on the
-output layer. If there is a hidden layer, the first output value
+coordinates of both the source (e.g.\ $x = 2$, $y = 4$) and target
+(e.g.\ $x = 3$, $y = 5$) nodes of a link and a constant bias
+value. The CPPN takes these five values as inputs and produces two output values. The first output value
 determines the weight of the link between the associated input
@@
 source and target nodes are iteratively passed as inputs to a CPPN to
-determine the weight of each substrate link.
+determine the weight of each ANN link.
 
@@
 
-HyperNEAT is capable of exploiting the geometry of a problem: because
-the link values between nodes in the final ANN substrate are a
-function of the geometric positions of those nodes, HyperNEAT can
-exploit such information when solving a
-problem~\cite{stanley2009hypercube, clune2009sensitivity,
+HyperNEAT is capable of exploiting the geometry of a problem because
+the link values between nodes in the final ANN phenotype are a
+function of the geometric positions of those nodes~\cite{stanley2009hypercube, clune2009sensitivity,
   clune2011performance}. In the case of quadruped locomotion, this
-property helped Hyper-NEAT produce gaits with front-back, left-right,
+property has previously been shown to help Hyper-NEAT produce gaits with front-back, left-right,
 and four-way symmetries~\cite{clune2009evolving,
@@
 mutations that add nodes and links to the network, enabling the
-algorithm to evolve the network topology in addition to its
-weights. Secondly, NEAT has a fitness sharing mechanism that preserves
+algorithm to evolve the topology of an ANN in addition to its
+weights. Secondly, NEAT has a fitness-sharing mechanism that preserves
 diversity in the system and gives time for new innovations to be tuned
@@
 gaits with HyperNEAT in simulation~\cite{clune2011performance,
-  clune2009evolving}, but was adapted to accommodate differences in
-the physical robot. Specifically, the ANNs consists of three $3 \times 4$
+  clune2009evolving}, but was adapted to accommodate the physical robot used in this paper. Specifically, the ANNs consists of three $3 \times 4$
 Cartesian grids of nodes forming input, hidden, and output
-layers. Adjacent layers were completely connected, meaning that there
-were $(3 \times 4)^2= 288$ links in each substrate. The inputs to the
+layers (\figref{SpiderANN.jpg}). Adjacent layers were allowed to be completely connected, meaning that there
+could be $(3 \times 4)^2= 288$ links in each ANN (although evolution can set weights to 0, functionally eliminating the connection). The inputs to the
 substrate were the current angles of each of the 9 joints of the robot
@@
   joint angles for each respective joint. The unlabeled node in the
-  input and output layer is ignored.}
+  input and output layer is ignored.}\label{hyperneatLayout}
 
@@
 robot. The remaining parameters were identical to Clune et
-al. ~\cite{clune2011performance}.
+al.~\cite{clune2011performance}.
 

### 2011-02-09 paper/results.tex

@@
 
-% Present the quantitative results of your experiments. Graphical data
-% presentation such as graphs and histograms are frequently better 
-% than tables. Explain what are the basic differences revealed in the 
-% data. If your method is stochastic or you analyze multiple datasets, 
-% calculate mean and standard error for every quantitative
-% performance result you show based on multiple runs.
+% Outline:
+% - exploration of parameter space
+% - parameterized optimization results, discussion
+% - HyperNEAT results, discussion
 
-\edit{proofread this.}
+\subsection{Exploration of Parameterized Gait Space}
 
-\subsection{Exploration of Parameter Space}
+%\edit{Jeff: proofread}
 
-In addition to running strategies to optimize parameterized gaits, we
-also wanted to investigate the space of possible gaits.  To accomplish
-this, we selected a parameter vector that resulted in motion, but not
-an exceptional gait, and plotted performance along each dimension
-individually.  In addition, we duplicated each measurement to be able
-to estimate the measurement noise at each point.  Results for this
-exploration are shown in \figref{explore_dim_1} through
+Before attempting to use the learning methods
+(\secref{learningMethods}) to optimize the chosen family of
+parameterized gaits (SineModel5, \secref{motionModel}), we did a brief
+experiment to explore the five dimensions of the parameter space. To
+accomplish this, we first selected a random parameter vector that resulted
+in motion, but not an exceptional gait.
+
+Then, in five different experiments, we varied each parameter
+individually and plotted performance versus each changing parameter.
+In addition, we duplicated each measurement to be able to estimate
+standard deviation of the measurement noise at each point.  Results
+for this exploration are shown in \figref{explore_dim_1} through
 \figref{explore_dim_5}.  As is shown in the figures, some dimensions

### 2011-02-10 paper/fitnessEvaluation.tex
edits

@@
 
-In order to measure the start and end position in the same pose, we
+In order to measure the start and end position in the same pose, and to ensure fair fitness evaluations with as little noise as possible, we
 linearly interpolated the motion of the robot between the ready
@@
 
-\acmFig{linearInterp}{1}{To ensure fair fitness evaluations with as
-  little noise as possible, the position of the robot was measured at
+\acmFig{linearInterp}{1}{Linear interpolation from the ready position to the first and last commanded gait positions. The position of the robot was measured at
   the beginning and end of each run (red circles) in the same
-  \emph{ready} pose (\figref{robot_close.jpg}).  Motion was
+ ready pose.  Motion was
   interpolated linearly between this stationary pose and the commanded

### 2011-02-10 paper/results.tex
edits

@@
 
-Before attempting to use the learning methods
-(\secref{learningMethods}) to optimize the chosen family of
-parameterized gaits (SineModel5, \secref{motionModel}), we did a brief
-experiment to explore the five dimensions of the parameter space. To
-accomplish this, we first selected a random parameter vector that resulted
-in motion, but not an exceptional gait.
-
-Then, in five different experiments, we varied each parameter
-individually and plotted performance versus each changing parameter.
-In addition, we duplicated each measurement to be able to estimate
-standard deviation of the measurement noise at each point.  Results
-for this exploration are shown in \figref{explore_dim_1} through
-\figref{explore_dim_5}.  As is shown in the figures, some dimensions
-are smoother than others, and some measurements are fairly noisy.
+Before optimizing the chosen family of
+parameterized gaits (\secref{motionModel}) with learning methods, we performed an experiment to explore the five dimensions of the parameter space.
+Specifically, we first selected a random parameter vector that resulted
+in motion, but not an exceptional gait. In five different experiments, we then varied each parameter
+individually and measured performance.
+Each measurement was repeated to get a rough estimate
+of the measurement noise at each point.  The results
+of this exploration (Figures \ref{fig:explore_dim_1} through
+\ref{fig:explore_dim_5}) reveal that some dimensions
+are smoother than others and that the measurement noise is variable and can be quite high.
 
@@
 
-\edit{maybe useful vvvvv}
-
-To be as fair as possible when obtaining the results presented in
-\secref{results}, we picked three random points --- $\vec{\theta}_A$,
-$\vec{\theta}_B$, and $\vec{\theta}_C$ --- and started all seven
-methods at these points...
-
-Each of the parameter ??? methods was run on 3 different initial
-parameter vectors, in order to fairly compare the algorithms. We allowed 
-each run for the parameter ??? methods to
-continue until the results plateaued (no improvement for one third of
-the policies seen so far). Three runs of HyperNEAT were completed,
-each with a different initial seed and run for 20 generations.
-
-\edit{$\wedge\wedge\wedge\wedge\wedge\wedge\wedge\wedge$}
-
 The maximizing strategies implemented include uniform random hill
@@
 Nelder-Mead simplex. As a group, these strategies produced results
-similar to gaits randomly generated using the SineModel5 model. None
+similar to gaits randomly-generated with the SineModel5 model. None
 of these strategies consistently outperformed the others in the three
@@
 
-\acmFig{vectorA}{1}{Results for runs beginning with vector A.
-  Linear regression worked significantly better than the other
-  algorithms, resulting in a gait (27.58 body lengths/minute) that
-  walked over twice as fast as the next best gait -- uniform random
-  hill climbing at 11.37 body lengths/minute -- and 5.3 times better
-  than the previous hand-coded gait.}
+%\acmFig{vectorA}{1}{Results for runs beginning with vector A.
+%  Linear regression worked significantly better than the other
+%  algorithms, resulting in a gait (27.58 body lengths/minute) that
+%  walked over twice as fast as the next best gait -- uniform random
+%  hill climbing at 11.37 body lengths/minute -- and 5.3 times better
+%  than the previous hand-coded gait.}
+%
+%\acmFig{vectorB}{1}{Results for run starting with vector B.  All
+%  algorithms performed similarly.  The random method actually got
+%  lucky this time and produced the best gait (17.26 body
+%  lengths/minute) and uniform random hill climbing produced the worst
+%  (9.44 body lengths/minute). The remaining algorithms each produced a
+%  gait around 13 body lengths/minute.}
+%
+%\acmFig{vectorC}{1}{Results for runs starting with vector C.
+%  Simplex and Gaussian random hill climbing each produced a gait that
+%  substantially outperformed the other algorithms, with around 15 body
+%  lengths/minute and 13 body lengths/minute, respectively, whereas the
+%  other algorithms returned gaits of less than 5 body
+%  lengths/minute.}
 
-\acmFig{vectorB}{1}{Results for run starting with vector B.  All
-  algorithms performed similarly.  The random method actually got
-  lucky this time and produced the best gait (17.26 body
-  lengths/minute) and uniform random hill climbing produced the worst
-  (9.44 body lengths/minute). The remaining algorithms each produced a
-  gait around 13 body lengths/minute.}
-
-\acmFig{vectorC}{1}{Results for runs starting with vector C.
-  Simplex and Gaussian random hill climbing each produced a gait that
-  substantially outperformed the other algorithms, with around 15 body
-  lengths/minute and 13 body lengths/minute, respectively, whereas the
-  other algorithms returned gaits of less than 5 body
-  lengths/minute.}
-
-Based on the three trials, as shown in \figref{std_error}, linear
-regression worked the best, followed by simplex. Gaussian random hill
-climbing performed about as well as random, but it is unclear whether
-Gaussian would continue to improve and become much better than random
-search if allowed to run for more iterations. Uniform random hill
+\acmFigDouble{std_error}{.6}{Average results ($\pm $ SEM) for each of the parametrized learning methods.  Linear regression performed the best,
+  followed by Nelder-Mead simplex. Gaussian random hill
+climbing performed about as well as random search, but may have surpassed random
+with additional iterations. Uniform random hill
 climbing and policy gradient descent performed similarly to each other
-on average, but differed greatly on individual runs.
-
-\acmFigDouble{std_error}{.6}{Average results for each method starting with
-  parameter vectors A, B, and C.  Error bars are plotted showing the
-  standard error.  Linear regression outperformed other methods,
-  followed by simplex. Gaussian random hill climbing performed about
-  as well as random, but it is unclear whether Gaussian would continue
-  to improve and become much better than random search if allowed to
-  run for more iterations. Uniform random hill climbing and policy
-  gradient descent performed similarly to each other on average, but
-  differed greatly on individual runs.}
+and underperformed random search.}
 

### 2011-02-10 paper/results.tex
Stacked plot_explore plots

--- 2011-02-09 paper/results.tex
@@
 
-\acmFig{explore_dim_1}{1}{Fitness mean and standard deviation
-  vs. dimension 1.  The circle is a common point in
-  \figref{explore_dim_1} through \figref{explore_dim_5}}
-\acmFig{explore_dim_2}{1}{Fitness mean and standard deviation
-  vs. dimension 2.  The circle is a common point in
-  \figref{explore_dim_1} through \figref{explore_dim_5}}
-\acmFig{explore_dim_3}{1}{Fitness mean and standard deviation
-  vs. dimension 3.  The circle is a common point in
-  \figref{explore_dim_1} through \figref{explore_dim_5}}
-\acmFig{explore_dim_4}{1}{Fitness mean and standard deviation
-  vs. dimension 4.  The circle is a common point in
-  \figref{explore_dim_1} through \figref{explore_dim_5}}
-\acmFig{explore_dim_5}{1}{Fitness mean and standard deviation
-  vs. dimension 5.  The circle is a common point in
-  \figref{explore_dim_1} through \figref{explore_dim_5}}
+%\acmFig{explore_dim_1}{1}{Fitness mean and standard deviation
+%  vs. dimension 1.  The circle is a common point in
+%  \figref{explore_dim_1} through \figref{explore_dim_5}}
+%\acmFig{explore_dim_2}{1}{Fitness mean and standard deviation
+%  vs. dimension 2.  The circle is a common point in
+%  \figref{explore_dim_1} through \figref{explore_dim_5}}
+%\acmFig{explore_dim_3}{1}{Fitness mean and standard deviation
+%  vs. dimension 3.  The circle is a common point in
+%  \figref{explore_dim_1} through \figref{explore_dim_5}}
+%\acmFig{explore_dim_4}{1}{Fitness mean and standard deviation
+%  vs. dimension 4.  The circle is a common point in
+%  \figref{explore_dim_1} through \figref{explore_dim_5}}
+%\acmFig{explore_dim_5}{1}{Fitness mean and standard deviation
+%  vs. dimension 5.  The circle is a common point in
+%  \figref{explore_dim_1} through \figref{explore_dim_5}}
+
+\acmFiggggg{explore_dim_1}{explore_dim_2}{explore_dim_3}{explore_dim_4}{explore_dim_5}{1}{.5}{Fitness
+  mean and standard deviation vs. dimension 5.  The circle is a common
+  point in \figref{explore_dim_1} through \figref{explore_dim_5}}
 

### 2011-02-10 paper/hyperNeatMethod.tex

@@
 A CPPN is queried once for each link in the ANN phenotype to determine
-that link's weight. The inputs to the CPPN are the Cartesian
+that link's weight~(\figref{hyperneatExplanation.png}). The inputs to the CPPN are the Cartesian
 coordinates of both the source (e.g.\ $x = 2$, $y = 4$) and target
@@
   of the hidden-layer nodes. The nodes in the output layer specify new
-  joint angles for each respective joint. The unlabeled node in the
-  input and output layer is ignored.}\label{hyperneatLayout}
+  joint angles for each respective joint. The unlabeled nodes in the
+  input and output layers are ignored.}\label{hyperneatLayout}
 
-The population size for HyperNEAT was 9 and runs lasted 20
+As with the parameterized learning methods, three runs of HyperNEAT were performed. The population size for HyperNEAT was 9 and runs lasted 20
 generations. These numbers are small, but were necessarily constrained

### 2011-02-10 paper/learningMethods.tex
minor
hyperneat edits

@@
 tried several parametrizations on the robot and, upon obtaining
-reasonable early success, settled on one particular parametrization, which we call  \emph{SineModel5}. Its root pattern is a sine wave and it has five parameters (Table~\ref{tab:params}) .
+reasonable early success, settled on one particular parametrization, which we call  \emph{SineModel5}. Its root pattern is a sine wave and it has five parameters (\tabref{params}).
 
@@
 
-\edit{called outer multiplier, but text says we multiply by inner. same for left}
-
 Intuitively, SineModel5 starts with 8 identical sine waves of
-amplitude $\amp$ and period $\tau$, multiplies the waves for all inner
+amplitude $\amp$ and period $\tau$, multiplies the waves for all outer
 motors by $m_O$, multiplies the waves for all front motors by $m_F$,
-and multiplies the waves for all left motors by $m_R$.  To obtain the
+and multiplies the waves for all right motors by $m_R$.  To obtain the
 actual motor position commands, these waves are offset by
@@
 position are $C_I = 800$ for inner motors, $C_O = 40$ for outer
-motors, and $C_C = 512$ for the center hip motor.  The nine functions
-correspond to the nine motors in the same order as listed in
+motors, and $C_C = 512$ for the center hip motor.  The nine functions (in order)
+correspond to the nine motors listed in
 \figref{topdown.png}.
-
-\edit{what subscripts?}
 
@@
   new linear model to all vectors evaluated so far, choosing the model
-  to minimize the sum of squared errors. Step size is the same as in
+  to minimize the sum of squared errors. The step size is the same as in
   \emph{N-dimensional policy gradient ascent}.
+
+\end{itemize}
+
+Each of the seven learning methods was run on 3 different initial
+parameter vectors ($\vec{\theta}_A$,
+$\vec{\theta}_B$, and $\vec{\theta}_C$) in order to fairly compare the algorithms.
+Three runs were performed per learning method.
+Runs continued until the performance plateaued, which we defined as when there was no improvement during the last third of
+a run.
 
@@
 %  became stuck in local minima.
-
-\end{itemize}

### 2011-02-10 paper/results.tex
Fixed ref

@@
 
-Before optimizing the chosen family of
-parameterized gaits (\secref{motionModel}) with learning methods, we performed an experiment to explore the five dimensions of the parameter space.
-Specifically, we first selected a random parameter vector that resulted
-in motion, but not an exceptional gait. In five different experiments, we then varied each parameter
-individually and measured performance.
-Each measurement was repeated to get a rough estimate
-of the measurement noise at each point.  The results
-of this exploration (Figures \ref{fig:explore_dim_1} through
-\ref{fig:explore_dim_5}) reveal that some dimensions
-are smoother than others and that the measurement noise is variable and can be quite high.
+Before optimizing the chosen family of parameterized gaits
+(\secref{motionModel}) with learning methods, we performed an
+experiment to explore the five dimensions of the parameter
+space. Specifically, we first selected a random parameter vector that
+resulted in motion, but not an exceptional gait. In five different
+experiments, we then varied each parameter individually and measured
+performance.  Each measurement was repeated to get a rough estimate of
+the measurement noise at each point.  The results of this exploration
+(\figref{explore_dim_1}) reveal that some dimensions are smoother than
+others and that the measurement noise is variable and can be quite
+high.
 

### 2011-02-10 paper/results.tex
change

--- 2011-02-10 paper/results.tex
@@
 
-%\acmFig{explore_dim_1}{1}{Fitness mean and standard deviation
-%  vs. dimension 1.  The circle is a common point in
-%  \figref{explore_dim_1} through \figref{explore_dim_5}}
-%\acmFig{explore_dim_2}{1}{Fitness mean and standard deviation
-%  vs. dimension 2.  The circle is a common point in
-%  \figref{explore_dim_1} through \figref{explore_dim_5}}
-%\acmFig{explore_dim_3}{1}{Fitness mean and standard deviation
-%  vs. dimension 3.  The circle is a common point in
-%  \figref{explore_dim_1} through \figref{explore_dim_5}}
-%\acmFig{explore_dim_4}{1}{Fitness mean and standard deviation
-%  vs. dimension 4.  The circle is a common point in
-%  \figref{explore_dim_1} through \figref{explore_dim_5}}
-%\acmFig{explore_dim_5}{1}{Fitness mean and standard deviation
-%  vs. dimension 5.  The circle is a common point in
-%  \figref{explore_dim_1} through \figref{explore_dim_5}}
-
-\acmFiggggg{explore_dim_1}{explore_dim_2}{explore_dim_3}{explore_dim_4}{explore_dim_5}{1}{.5}{Fitness
-  mean and standard deviation vs. dimension 5.  The circle is a common
-  point in \figref{explore_dim_1} through \figref{explore_dim_5}}
+\acmFig{explore_dim_1}{1}{Fitness mean and standard deviation
+  vs. dimension 1.  The circle is a common point in
+  \figref{explore_dim_1} through \figref{explore_dim_5}}
+\acmFig{explore_dim_2}{1}{Fitness mean and standard deviation
+  vs. dimension 2.  The circle is a common point in
+  \figref{explore_dim_1} through \figref{explore_dim_5}}
+\acmFig{explore_dim_3}{1}{Fitness mean and standard deviation
+  vs. dimension 3.  The circle is a common point in
+  \figref{explore_dim_1} through \figref{explore_dim_5}}
+\acmFig{explore_dim_4}{1}{Fitness mean and standard deviation
+  vs. dimension 4.  The circle is a common point in
+  \figref{explore_dim_1} through \figref{explore_dim_5}}
+\acmFig{explore_dim_5}{1}{Fitness mean and standard deviation
+  vs. dimension 5.  The circle is a common point in
+  \figref{explore_dim_1} through \figref{explore_dim_5}}
 
@@
 Nelder-Mead simplex. As a group, these strategies produced results
-similar to gaits randomly-generated with the SineModel5 model. None
-of these strategies consistently outperformed the others in the three
+similar to or worse than gaits randomly-generated with the SineModel5 model (\figref{std_error}). None
+of these strategies consistently outperformed the other maximizing strategies in the three
 trials; A different strategy generated the best gait in each

### 2011-02-10 paper/hyperNeatMethod.tex
HN edits, other

@@
   clune2011performance}. In the case of quadruped locomotion, this
-property has previously been shown to help Hyper-NEAT produce gaits with front-back, left-right,
+property has previously been shown to help HyperNEAT produce gaits with front-back, left-right,
 and four-way symmetries~\cite{clune2009evolving,
@@
 substrate were the current angles of each of the 9 joints of the robot
-(described below) and a sine and cosine wave (to facilitate the
+(described below \edit{Jeff: is it?}) and a sine and cosine wave (to facilitate the
 production of periodic behaviors). The sine and cosine waves were the
 following function of time ($t$) in milli-seconds:
-$\sin(t/12)\pi$. This function produces numbers in the range [$-\pi,
-  \pi$], which were then JASON FILL ME IN TO DESCRIBE HOW JOINT
+$\sin(t/12)\pi$ \editbox{Jeff: this can't be right, because it would imply a period of only 12*2*pi=75 ms.
+Perhaps this was as a function of time in units of timesteps?}.
+This function produces numbers in the range [$-\pi,
+  \pi$] \edit{Jeff: this is actually [-1,1], right?}, which were then JASON FILL ME IN TO DESCRIBE HOW JOINT
 COMMANDS WERE SPECIFIED.
 
-\editbox{are we mentioning that we gave HyperNEAT the center joint
-  here or earlier?}
+\editbox{Jeff: proofread this paragraph to make sure I didn't lie (too
+  much)}
 
-\edit{jason: edit this section to reflect the post-processing...}
+The outputs of the substrate at each time step were nine numbers in
+the range $[-1,1]$, which were scaled according to the allowable
+ranges for each of the nine motors and then used as commanded
+positions.  Occasionally HyperNEAT would produce networks that
+exhibited rapid oscillatory behavior, switching between nearly -1 and
+nearly 1 from one time step to the next.  This resulted in motor
+commands to alternate extremes every 25ms (using a command rate of
+40Hz), which tended to damage and overheat the motors.  To ameliorate
+this problem, we simply requested four times as many commanded
+positions from the ANN and averaged over four commands at a time to
+obtain the actual $g(t)$ to be used as a gait.  This proved to work
+well and did not restrict the expressiveness of HyperNEAT.
+
+%\editbox{are we mentioning that we gave HyperNEAT the center joint
+%  here or earlier?}
 
@@
 
-As with the parameterized learning methods, three runs of HyperNEAT were performed. The population size for HyperNEAT was 9 and runs lasted 20
-generations. These numbers are small, but were necessarily constrained
-given how much time it took to conduct evolution directly on a real
-robot. The remaining parameters were identical to Clune et
+As with the parameterized learning methods, three runs of HyperNEAT
+were performed. The population size for HyperNEAT was 9 and runs
+lasted 20 generations. These numbers are small, but were necessarily
+constrained given how much time it took to conduct evolution directly
+on a real robot. The remaining parameters were identical to Clune et
 al.~\cite{clune2011performance}.

### 2011-02-10 paper/learningMethods.tex
HN edits, other

@@
 
-We evaluated a total of seven different
+We evaluated a total of six different
 learning algorithms for the parametrized motion models.  All methods
@@
 
-The seven learning algorithms for the parametrized motion models are as follows:
+The six learning algorithms for the parametrized motion models are as follows:
 
@@
 
-Each of the seven learning methods was run on 3 different initial
+Each of the six learning methods was run on 3 different initial
 parameter vectors ($\vec{\theta}_A$,

### 2011-02-10 paper/results.tex
Edits to results/exploration and results table

@@
 
-%\edit{Jeff: proofread}
+\edit{Jeff: proofread again}
 
@@
 (\secref{motionModel}) with learning methods, we performed an
-experiment to explore the five dimensions of the parameter
-space. Specifically, we first selected a random parameter vector that
-resulted in motion, but not an exceptional gait. In five different
-experiments, we then varied each parameter individually and measured
-performance.  Each measurement was repeated to get a rough estimate of
-the measurement noise at each point.  The results of this exploration
-(\figref{explore_dim_1}) reveal that some dimensions are smoother than
-others and that the measurement noise is variable and can be quite
-high.
+experiment to explore the five dimensions of the SineModel5 parameter
+space. Specifically, we selected a random parameter vector that
+resulted in some motion, but not an exceptional gait. We then varied
+each of the five parameters individually and measured performance,
+repeating each measurement twice to get a rough estimate of the
+measurement noise at each point.  The results of this exploration,
+shown in \figref{explore_dim_1}, reveal that some dimensions ($\amp$,
+$\tau$, $m_F$) are fairly smooth and exhibit global structure across
+the allowable parameter range, while others ($m_O$, $m_R$) exhibit
+more complex behavior.  In addition, it gives a rough indication that
+measurement noise is often significant and is more likely to be large
+for larger measurements.  Of course, this is only a slice in each
+dimension through a single point, and slices through a different point
+would produce different behavior.  The common point at the
+intersection of all slices is shown as a red triangle in
+each plot of \figref{explore_dim_1}.
 
@@
 
-\subsection{Maximizing Strategies}
+\subsection{Learning Methods for Parametrized Gaits}
 
@@
 Nelder-Mead simplex. As a group, these strategies produced results
-similar to or worse than gaits randomly-generated with the SineModel5 model (\figref{std_error}). None
-of these strategies consistently outperformed the other maximizing strategies in the three
-trials; A different strategy generated the best gait in each
-trial. Uniform random hill climbing produced the best gait (11.37 body
+similar to or worse than gaits randomly-generated with the SineModel5
+model (\figref{std_error}). None of these strategies consistently
+outperformed the other maximizing strategies in the three trials; A
+different strategy generated the best gait in each trial. Uniform
+random hill climbing produced the best gait (11.37 body
 lengths/minute) in the first trial, with the next best performing
@@
 
-\begin{table*}
+% octave:15> fit = [6.04 17.26 4.90;  11.37 9.44 2.69;  3.10 13.59 13.40;  0.68 14.69 3.60;  8.51 13.62 14.83;  27.58 12.51 1.95;  24.27  36.44  27.07];
+% octave:16> mean(fit,2)
+% ans =
+% 
+%     9.4000
+%     7.8333
+%    10.0300
+%     6.3233
+%    12.3200
+%    14.0133
+%    29.2600
+% 
+% octave:17> std(fit,0,2)
+% ans =
+% 
+%     6.8308
+%     4.5576
+%     6.0023
+%     7.3914
+%     3.3546
+%    12.8810
+%     6.3737
+
+%%%%%% OLD TABLE
+% \begin{table*}
+% \begin{center}
+% \begin{tabular}{|r|c|c|c||c|}
+% \hline
+%                                          & A       & B      & C      &  Average \\
+% \hline                                                               
+% \hline                                                               
+% Previous hand-coded gait                 & --      & --     & --     &  5.16 \\
+% \hline                                                                 
+% Random search                            & 6.04    & 17.26  & 4.90   &  9.40 \\
+% \hline                                                                 
+% Uniform Random Hill Climbing             & 11.37   & 9.44   & 2.69   &  7.83 \\
+% \hline                                                                 
+% Gaussian Random Hill Climbing            & 3.10    & 13.59  & 13.40  &  10.03 \\
+% \hline                                                                 
+% Policy Gradient Descent                  & 0.68    & 14.69  & 3.60   &  6.32 \\
+% \hline                                                                 
+% Nelder-Mead simplex                      & 8.51    & 13.62  & 14.83  &  12.32 \\
+% \hline                                                                 
+% Linear Regression                        & 27.58   & 12.51  & 1.95   &  14.01 \\
+% \hline                                                                
+% Evolutionary Neural Network (HyperNEAT)  & 24.27     & 36.44    & 27.07   & 29.26  \\
+% \hline
+% \end{tabular}
+% \caption{The best gaits found for each starting vector and algorithm,
+%   in body lengths per minute.}
+% \tablabel{results_table}
+% \end{center}
+% \end{table*}
+
+%
+
+%
+
+%
+
+%
+
+%
+
+%
+
+%
+
+%%%%%%% NEW TABLE
+\begin{table}
 \begin{center}
@@
 \hline
-                                         & A       & B      & C      &  Average \\
+                                         & Average & Std. Dev. \\
 \hline                                    
 \hline                                    
-Previous hand-coded gait                 & --      & --     & --     &  5.16 \\
+Previous hand-coded gait                 & 5.16   &   --     \\
 \hline
-Random search                            & 6.04    & 17.26  & 4.90   &  9.40 \\
+Random search                            & 9.40   &   6.83   \\
 \hline
-Uniform Random Hill Climbing             & 11.37   & 9.44   & 2.69   &  7.83 \\
+Uniform Random Hill Climbing             & 7.83   &   4.56   \\
 \hline
-Gaussian Random Hill Climbing            & 3.10    & 13.59  & 13.40  &  10.03 \\
+Gaussian Random Hill Climbing            & 10.03  &   6.00   \\
 \hline
-Policy Gradient Descent                  & 0.68    & 14.69  & 3.60   &  6.32 \\
+Policy Gradient Descent                  & 6.32   &   7.39   \\
 \hline
-Nelder-Mead simplex                      & 8.51    & 13.62  & 14.83  &  12.32 \\
+Nelder-Mead simplex                      & 12.32  &   3.35   \\
 \hline
-Linear Regression                        & 27.58   & 12.51  & 1.95   &  14.01 \\
+Linear Regression                        & 14.01  &  12.88   \\
 \hline
-Evolutionary Neural Network (HyperNEAT)  & 24.27     & 36.44    & 27.07   & 29.26  \\
+Evolutionary Neural Network              &        &          \\
+(HyperNEAT)                              & 29.26  &   6.37   \\
 \hline
@@
 \end{center}
-\end{table*}
+\end{table}
 

### 2011-02-10 paper/introduction.tex
Added some potentially useful intro text back in
changes
parametrized -> parameterized

@@
 
-\acmFig{robot_close.jpg}{1}{The quadruped robot. The translucent parts were created with 3D-printing technology.}
+\acmFig{robot_close.jpg}{1}{The quadruped robot.
+The translucent parts were created with 3D-printing technology.
+Videos of the gaits we evolved for this robot can be viewed online~\cite{quadratotWeb}.}
 
@@
 
-In this paper we compare the performance of three different methods of designing gaits: manual design, parametrized gaits optimized with seven different learning methods, and gaits generated by evolving neural networks with the HyperNEAT generative encoding~\cite{stanley2009hypercube}.
+In this paper we compare the performance of three different methods of designing gaits: manual design, parameterized gaits optimized with seven different learning methods, and gaits generated by evolving neural networks with the HyperNEAT generative encoding~\cite{stanley2009hypercube}.
 While some of these methods, such as HyperNEAT, have been tested in simulation~\cite{clune2009evolving, clune2011performance}, we investigate how they perform when evolving on a physical robot (\figref{robot_close.jpg}).
@@
 It is additionally interesting to test how more traditional machine learning techniques compete with evolutionary algorithms when evolving in hardware.
+
+% Put back in for reference... some good text here I think
+
+% Applications of walking robots often call for the ability to walk as 
+% quickly, efficiently, or with as little power as possible.  Often 
+% these optimizations are done manually by an expert who designs and 
+% tweaks a gait specifically for a given objective.  Other groups have 
+% used learning methods to generate gaits optimized for some metric. 
+% Approaches differ in their starting assumptions, some essentially 
+% tweaking the parameters of a hand-tuned model \cite{chernova}, others 
+% exploring a reasonably compact parameter space \cite{kohl}, and still 
+% others beginning with few assumptions besides periodicity 
+% \cite{zykov}. 
+%  
+% We aimed to strike a middle ground between these approaches.  Our 
+% motion generator did not rely on hand-tweaked gaits, but it did use 
+% parameterized gaits which, by their nature, impose some assumptions on 
+% the answers produced.  We then used machine learning to design gaits 
+% for a quadruped robot with these models.  This paper presents a 
+% comparison of the different learning methods implemented.  Most 
+% methods created walks that are several times faster than the original 
+% hand-tuned gait.  We invite readers with short attention spans to view 
+% a video of some of our results online here: 
+%  
+% 
+% %\url{http://www.youtube.com/watch?v=ODoiOj9DdGg} 
+% \texttt{http://www.youtube.com/watch?v=ODoiOj9DdGg}

### 2011-02-10 paper/learningMethods.tex
Edits to results/exploration and results table

@@
-\subsection{Parametrized Motion Models}
+\subsection{Parametrized Gaits}
 \seclabel{motionModel}
@@
 \secref{problemDefinition}).  The first method uses a family of
-parametrized gaits, described in this section and the next. The
+parameterized gaits, described in this section and the next. The
 second method evolves a non-parametric gait controlled by neural networks, and is detailed in
@@
 
-By a \emph{parametrized gait}, we mean a gait produced by a
-parametrized function $g(t; \vec{\theta})$. Fixing the parameter
+By a \emph{parameterized gait}, we mean a gait produced by a
+parameterized function $g(t; \vec{\theta})$. Fixing the parameter
 $\vec{\theta}$ yields a deterministic motion function over time.  We
@@
 
-\subsection{Learning Methods for Parametrized Motion Models}
+\subsection{Learning Methods for Parametrized Gaits}
 \seclabel{learningMethods}
 
-Using the SineModel5 parametrized motion model described in the
+Using the SineModel5 parameterized motion model described in the
 previous section, along with the allowable ranges for each of the five
@@
 We evaluated a total of six different
-learning algorithms for the parametrized motion models.  All methods
+learning algorithms for the parameterized motion models.  All methods
 needed a way of picking a $\vec{\theta}$ for the first trial of each run.  This could be done by selecting a random starting vector within the
@@
 
-The six learning algorithms for the parametrized motion models are as follows:
+The six learning algorithms for the parameterized motion models are as follows:
 
@@
 
-%We use several parametrized motion models that command motors to
+%We use several parameterized motion models that command motors to
 %positions based on a sine wave, creating a periodic pattern.  While we

### 2011-02-10 paper/results.tex
Results cleanup/cuts/edits

@@
 
-\subsection{Learning Methods for Parametrized Gaits}
+\subsection{Learning Methods for Parameterized Gaits}
 
-The maximizing strategies implemented include uniform random hill
-climbing, Gaussian random hill climbing, policy gradient descent, and
-Nelder-Mead simplex. As a group, these strategies produced results
-similar to or worse than gaits randomly-generated with the SineModel5
-model (\figref{std_error}). None of these strategies consistently
-outperformed the other maximizing strategies in the three trials; A
-different strategy generated the best gait in each trial. Uniform
-random hill climbing produced the best gait (11.37 body
-lengths/minute) in the first trial, with the next best performing
-strategy (Nelder-Mead simplex) resulting in a gait of 8.51 body
-lengths/minute. The best gait in the second trial was actually
-generated randomly (17.26 body-lengths/minute). Gaussian random hill
-climbing, policy gradient descent, and Nelder-Mead simplex all
-produced similar best gaits in the second trial, each around 14 body
-lengths/minute. In the third trial, Nelder-Mead simplex and Gaussian
-random hill climbing significantly outperformed the other maximizing
-strategies, resulting in gaits of 14.83 and 13.40 body lengths/minute
-respectively. Overall, Nelder-Mead simplex performed the best,
-producing an average best gait of 12.32 body lengths/minute. Gaussian
-random hill climbing followed, with an average best gait of 10.03 body
-lengths/minute.
+% Outline
+% - beat hand gait
+% - no learning strategy outperformed another by that much
+% - real world noise yay
 
-\subsection{Predictive Strategies}
-
-The predictive strategies used were linear regression and support
-vector regression. In the first trial, linear regression produced a
-gait (27.58 body lengths/minute) that was over three times faster than
-the best gait produced by a maximizing strategy (uniform random hill
-climbing -- 11.37 body lengths/minute). The average of the best gait
-for the three trials of linear regression was 14.01 body
-lengths/minute, which was better than the average best gait for the
-best performing maximizing strategy (Nelder-Mead simplex -- 12.32 body
-lengths/minute).
-
-%As mentioned above, the SVM strategy involved many tunable
-%parameters. The version of SVM tuned for simulation did not show
-%competitive performance on the real robot. We tried tuning some
-%parameters on the real robot, but after some amount of tuning, the
-%method still exhibited too little exploration and easily became stuck
-%in local minima.
-%
-%\edit{say something about tuning SVM and the sim figure}
-%
-%\acmFig{svm_sim_results}{1}{Results for the SVM regression strategy
-%  in simulation.  This simulation was used to tune the SVM strategy's
-%  parameters before trying it on the physical robot.  The strategy
-%  quickly approaches the maximum simulated fitness of 100.}
-
-\subsection{Evolved Neural Network Strategy}
-
-The evolved neural network strategy used was HyperNEAT. HyperNEAT
-produced the fastest gaits by far. Linear regression produced a gait
-in the first trial (27.58 body lengths/minute) that was on par with
-the best gaits generated by HyperNEAT, but no other maximizing or
-predictive strategy produced gaits close to those of HyperNEAT. Linear
-regression was the best performing maximizing or predictive strategy,
-but it resulted in an average best gait of only 14.01 body
-lengths/minute. HyperNEAT, on the other hand, resulted in an average
-best gait of 29.26 body lengths/minute. One issue encountered with
-HyperNEAT was that it generated gaits that were rougher on the robot
-than gaits generated by the other strategies. The gaits sometimes
-caused the motor servos to overheat, which caused the motor to behave
-differently than expected. Since an overheated motor produces a
-different walk that expected, a test was performed after each run to
-make sure all motors were working correctly. If a motor failed, the
-distance walked would be halved to penalize walks that were harmful to
-the robot.
-
-\edit{take this out?}
-
-We have done complete runs of 3 different initial parameter vectors
-for random search, uniform random hill climbing, Gaussian random hill
-climbing, policy gradient descent, Nelder-Mead simplex, and linear
-regression.  We also evaluated the SVM regression and HyperNEAT
-methods, but these methods were more experimental, and thus we do not
-yet have the same volume of data for these runs.  We developed several
-gaits that were about 4 times faster than the original hand-coded
-gait.  Results are shown in \tabref{results_table}.
+The results for the parameterized gaits are shown in
+\figref{std_error} and \tabref{results}.  As a group, these strategies
+produced results not significantly better than the randomly-generated
+gaits, however, all methods managed to explore the space enough to
+significantly improve on the previous hand coded gait in at least one
+of the three runs.  No single strategy consistently outperformed the
+others: for the first trial Linear Regression produced the fastest
+gait at 27.58 body lengths per minute, for the second a random gait
+actually won with 17.26 body lengths / minute, and for the third trial
+the Nelder-Mead simplex method attained the fastest gait at 14.83 body
+lengths / minute.
 
@@
 %   in body lengths per minute.}
-% \tablabel{results_table}
+% \tablabel{results}
 % \end{center}
 % \end{table*}
-
-%
-
-%
-
-%
-
-%
-
-%
-
-%
-
-%
 
@@
   in body lengths per minute.}
-\tablabel{results_table}
+\tablabel{results}
 \end{center}
@@
 
-\begin{itemize}
+\acmFig{std_error}{1}{Average results ($\pm $ standard error) for each
+  of the parameterized learning methods.  Linear regression found the
+  fastest overall gait and had the highest average, followed by
+  Nelder-Mead simplex. Many methods were beaten by the simple random
+  strategy.}
 
-\item For vector A, shown in \figref{vectorA}, linear regression
-  worked significantly better than the other algorithms, resulting in
-  a gait (27.58 body lengths/minute) that walked over twice as fast as
-  the next best gait -- uniform random hill climbing at 11.37 body
-  lengths/minute -- and 5.3 times better than the previous hand-coded
-  gait.
+\subsection{Evolved Neural Network Strategy}
 
-\item Vector B, depicted in \figref{vectorB}, resulted in similarly
-  performing gaits with all the algorithms. The random method got
-  lucky and produced the best gait (17.26 body lengths/minute) and
-  uniform random hill climbing produced the worst (9.44 body
-  lengths/minute). The remaining algorithms each produced a gait
-  around 13 body lengths/minute.
+The evolved neural network strategy used was HyperNEAT. HyperNEAT
+produced the fastest gaits by far. Linear regression produced a gait
+in the first trial (27.58 body lengths/minute) that was on par with
+the best gaits generated by HyperNEAT, but no other maximizing or
+predictive strategy produced gaits close to those of HyperNEAT. Linear
+regression was the best performing maximizing or predictive strategy,
+but it resulted in an average best gait of only 14.01 body
+lengths/minute. HyperNEAT, on the other hand, resulted in an average
+best gait of 29.26 body lengths/minute. One issue encountered with
+HyperNEAT was that it generated gaits that were rougher on the robot
+than gaits generated by the other strategies. The gaits sometimes
+caused the motor servos to overheat, which caused the motor to behave
+differently than expected. Since an overheated motor produces a
+different walk that expected, a test was performed after each run to
+make sure all motors were working correctly. If a motor failed, the
+distance walked would be halved to penalize walks that were harmful to
+the robot.
 
-\item For vector C, shown in \figref{vectorC}, simplex and Gaussian
-  random hill climbing each produced a gait that substantially
-  outperformed the other algorithms. Simplex resulted in a gait of
-  nearly 15 body lengths/minute and Gaussian random hill climbing
-  produced a gait of just over 13 body lengths/minute, whereas the
-  other algorithms returned gaits of less than 5 body lengths/minute.
+\edit{take this out?}
 
-\end{itemize}
-
-%\acmFig{vectorA}{1}{Results for runs beginning with vector A.
-%  Linear regression worked significantly better than the other
-%  algorithms, resulting in a gait (27.58 body lengths/minute) that
-%  walked over twice as fast as the next best gait -- uniform random
-%  hill climbing at 11.37 body lengths/minute -- and 5.3 times better
-%  than the previous hand-coded gait.}
-%
-%\acmFig{vectorB}{1}{Results for run starting with vector B.  All
-%  algorithms performed similarly.  The random method actually got
-%  lucky this time and produced the best gait (17.26 body
-%  lengths/minute) and uniform random hill climbing produced the worst
-%  (9.44 body lengths/minute). The remaining algorithms each produced a
-%  gait around 13 body lengths/minute.}
-%
-%\acmFig{vectorC}{1}{Results for runs starting with vector C.
-%  Simplex and Gaussian random hill climbing each produced a gait that
-%  substantially outperformed the other algorithms, with around 15 body
-%  lengths/minute and 13 body lengths/minute, respectively, whereas the
-%  other algorithms returned gaits of less than 5 body
-%  lengths/minute.}
-
-\acmFigDouble{std_error}{.6}{Average results ($\pm $ SEM) for each of the parametrized learning methods.  Linear regression performed the best,
-  followed by Nelder-Mead simplex. Gaussian random hill
-climbing performed about as well as random search, but may have surpassed random
-with additional iterations. Uniform random hill
-climbing and policy gradient descent performed similarly to each other
-and underperformed random search.}
+We have done complete runs of 3 different initial parameter vectors
+for random search, uniform random hill climbing, Gaussian random hill
+climbing, policy gradient descent, Nelder-Mead simplex, and linear
+regression.  We also evaluated the SVM regression and HyperNEAT
+methods, but these methods were more experimental, and thus we do not
+yet have the same volume of data for these runs.  We developed several
+gaits that were about 4 times faster than the original hand-coded
+gait.  Results are shown in \tabref{results}.
 

### 2011-02-10 paper/hyperNeatMethod.tex
Edits to results/exploration and results table

--- 2011-02-10 paper/hyperNeatMethod.tex
@@
-\subsection{HyperNEAT Motion Model}
+\subsection{HyperNEAT Gait Generation and Learning}
 \seclabel{hyperNeatMethod}
@@
 could be $(3 \times 4)^2= 288$ links in each ANN (although evolution can set weights to 0, functionally eliminating the connection). The inputs to the
-substrate were the current angles of each of the 9 joints of the robot
-(described below) and a sine and cosine wave (to facilitate the
+substrate were the current angles of each of the 9 joints of the robot and a sine and cosine wave (to facilitate the
 production of periodic behaviors). The sine and cosine waves were the
 following function of time ($t$) in milli-seconds:
-$\sin(t/12)\pi$. This function produces numbers in the range [$-\pi,
-  \pi$], which were then JASON FILL ME IN TO DESCRIBE HOW JOINT
-COMMANDS WERE SPECIFIED.
+$\sin(t/12)\pi$, which produces numbers in the range [$-\pi, \pi$].
 
-\editbox{are we mentioning that we gave HyperNEAT the center joint
-  here or earlier?}
+\editbox{Jeff: this can't be right, because it would imply a period of only 12*2*pi=75 ms. Perhaps this was as a function of time in units of timesteps?}
+\editbox{Jeff: proofread this paragraph to make sure I didn't lie (too
+  much)}
 
-\edit{jason: edit this section to reflect the post-processing...}
+The outputs of the substrate at each time step were nine numbers in
+the range $[-1,1]$, which were scaled according to the allowable
+ranges for each of the nine motors and then commanded
+the positions for each motor.  Occasionally HyperNEAT would produce networks that
+exhibited rapid oscillatory behaviors, switching from extreme negative to extreme positive numbers each time step.  This resulted in motor
+commands to alternate extremes every 25ms (given the command rate of
+40Hz), which tended to damage and overheat the motors.  To ameliorate
+this problem, we requested four times as many commanded
+positions from HyperNEAT ANN's and averaged over four commands at a time to
+obtain the actual gait $g(t)$.  This solution worked well and did not restrict the expressiveness of HyperNEAT.
+
+%\editbox{are we mentioning that we gave HyperNEAT the center joint
+%  here or earlier?}
 

### 2011-02-10 paper/results.tex
minor fix

@@
 
-\acmFiggggg{explore_dim_1}{explore_dim_2}{explore_dim_3}{explore_dim_4}{explore_dim_5}{1}{.5}{Fitness
-  mean and standard deviation vs. dimension 5.  The circle is a common
-  point in \figref{explore_dim_1} through \figref{explore_dim_5}}
+\acmFiggggg{explore_dim_1}{explore_dim_2}{explore_dim_3}{explore_dim_4}{explore_dim_5}{1}{.53}{Fitness
+  mean and standard deviation when each parameter dimension is varied
+  independently.  The red triangle is the intersection of the slices
+  along each dimension and is common to all plots.}
 

### 2011-02-10 paper/hyperNeatMethod.tex
changes

@@
 substrate were the current angles of each of the 9 joints of the robot and a sine and cosine wave (to facilitate the
-production of periodic behaviors). The sine and cosine waves were the
-following function of time ($t$) in milli-seconds:
-$\sin(t/12)\pi$, which produces numbers in the range [$-\pi, \pi$].
-
-\editbox{Jeff: this can't be right, because it would imply a period of only 12*2*pi=75 ms. Perhaps this was as a function of time in units of timesteps?}
-\editbox{Jeff: proofread this paragraph to make sure I didn't lie (too
-  much)}
+production of periodic behaviors). The sine and cosine waves had a period of about half a second.
 

### 2011-02-10 paper/results.tex
changes
slight change to red triangle description
HN results

@@
   mean and standard deviation when each parameter dimension is varied
-  independently.  The red triangle is the intersection of the slices
-  along each dimension and is common to all plots.}
+  independently.  The red triangle in each plot represents the same point in the 5-dimensional parameter space.}
 
@@
 others: for the first trial Linear Regression produced the fastest
-gait at 27.58 body lengths per minute, for the second a random gait
-actually won with 17.26 body lengths / minute, and for the third trial
+gait at 27.58 body lengths/minute, for the second a random gait
+actually won with 17.26 body lengths/minute, and for the third trial
 the Nelder-Mead simplex method attained the fastest gait at 14.83 body
-lengths / minute.
+lengths/minute.
 
@@
 \hline
-Evolutionary Neural Network              &        &          \\
+Evolved Neural Network              &        &          \\
 (HyperNEAT)                              & 29.26  &   6.37   \\
@@
 
-\subsection{Evolved Neural Network Strategy}
+\subsection{HyperNEAT Gaits}
 
-The evolved neural network strategy used was HyperNEAT. HyperNEAT
-produced the fastest gaits by far. Linear regression produced a gait
-in the first trial (27.58 body lengths/minute) that was on par with
-the best gaits generated by HyperNEAT, but no other maximizing or
-predictive strategy produced gaits close to those of HyperNEAT. Linear
-regression was the best performing maximizing or predictive strategy,
-but it resulted in an average best gait of only 14.01 body
-lengths/minute. HyperNEAT, on the other hand, resulted in an average
-best gait of 29.26 body lengths/minute. One issue encountered with
-HyperNEAT was that it generated gaits that were rougher on the robot
-than gaits generated by the other strategies. The gaits sometimes
-caused the motor servos to overheat, which caused the motor to behave
-differently than expected. Since an overheated motor produces a
-different walk that expected, a test was performed after each run to
-make sure all motors were working correctly. If a motor failed, the
-distance walked would be halved to penalize walks that were harmful to
-the robot.
+The results for the gaits produced and evolved by HyperNEAT are shown
+in \figref{hnResults} and \tabref{results}.  Overall the evolved gaits
+were the fastest by far, beating all the parameterized models when
+comparing either the average or the best gaits of each.
 
-\edit{take this out?}
+\acmFig{hnResults}{1}{Average fitness ($\pm $ standard error) of
+  generation champion (highest performing individual among the
+  population of nine individuals) for each generation of the three
+  HyperNEAT runs.}
 
-We have done complete runs of 3 different initial parameter vectors
-for random search, uniform random hill climbing, Gaussian random hill
-climbing, policy gradient descent, Nelder-Mead simplex, and linear
-regression.  We also evaluated the SVM regression and HyperNEAT
-methods, but these methods were more experimental, and thus we do not
-yet have the same volume of data for these runs.  We developed several
-gaits that were about 4 times faster than the original hand-coded
-gait.  Results are shown in \tabref{results}.
+%\acmFig{neat_110115_211410_00000_002_filt}{1}{Caption here...???}
 
-\subsection{Discussion}
-
-% Is your hypothesis supported? What conclusions do the results support
-% about the strengths and weaknesses of your method compared to other 
-% methods? How can the results be explained in terms of the underlying 
-% properties of the algorithm and/or the data.
-
-Because only three trials were tested with each algorithm and no
-algorithm consistently outperformed the others, there is a large
-standard error for each algorithm, as show in \figref{std_error}. For
-this reason, it is unclear if any algorithm is superior to another in
-this application. More trials with each algorithm would be necessary
-to reach a definitive ranking of the performance of the algorithms.
-
-However, each algorithm discovered at least one gait of over 10 body
-lengths/minute, including random. For this reason, we speculate that
-the motion model is at least as critical as the learning algorithm
-used.
-
-\acmFig{neat_110115_211410_00000_002_filt}{1}{Caption here...???}
-\acmFig{neat_110115_211410_00000_002_filt_zoom}{1}{Caption here...???}
+\acmFig{neat_110115_211410_00000_002_filt_zoom}{1}{Example of one
+  particular high-performance gait produced by HyperNEAT showing
+  commands for each of nine motors.  Note the complexity of the motion
+  pattern. Such patterns were not possible with the parameterized
+  SineModel5, nor would they likely be contrived by a human designing
+  a different low-dimensional parameterized motion model.}

### 2011-02-10 paper/results.tex
changes

--- 2011-02-10 paper/results.tex
@@
 \subsection{Exploration of Parameterized Gait Space}
-
-\edit{Jeff: proofread again}
 
@@
 measurement noise is often significant and is more likely to be large
-for larger measurements.  Of course, this is only a slice in each
+for gaits that move more.  Of course, this is only a slice in each
 dimension through a single point, and slices through a different point
-would produce different behavior.  The common point at the
+could reveal different behavior.  The common point at the
 intersection of all slices is shown as a red triangle in

### 2011-02-10 paper/results.tex
5.3

--- 2011-02-10 paper/results.tex
@@
 were the fastest by far, beating all the parameterized models when
-comparing either the average or the best gaits of each.
+comparing either the average or the best gaits of each.  We suspect
+that this is due to the fact that HyperNEAT was allowed to explore a
+much richer space of motions, but did so while still utilizing
+symmetries when advantageous.
+\figref{neat_110115_211410_00000_002_filt_zoom} shows a typical
+HyperNEAT gait that had high fitness.  The pattern of motion is both
+complex --- containing multiple frequencies and repeating patterns ---
+and reuses patterns for multiple motors.
 

### 2011-02-10 paper/results.tex
foo

--- 2011-02-10 paper/results.tex
@@
 \figref{std_error} and \tabref{results}.  As a group, these strategies
-produced results not significantly better than the randomly-generated
+produced results equal to or worse not significantly better than the randomly-generated
 gaits, however, all methods managed to explore the space enough to

### 2011-02-10 paper/futureWork.tex
Added paperOutline
lots of editing of the intro/motivation/previouswork, and some in the robot platform
Minor
More fitness, explanation of bias due to sensor area
Conclusion/Future Work

@@
 %\section{Future work}
+\seclabel{futureWork}
 
-\edit{write this, old section below}
+Because all trials were done in hardware, it was difficult to gather
+the thousands of trials that would be necessary to establish a more
+rigorous statistical ranking of the methods' performance, or their
+asymptotic behavior as longer and longer training runs are allowed.
+Thus, one extension of the current work would be to obtain these
+thousands more trials.
 
-% What are the major shortcomings of your current method? For each 
-% shortcoming, propose additions or enhancements that would help overcome it.
+However, a more effective extension might be to combine trials in
+simulation and in hardware, perhaps doing only every tenth or
+hundredth trial in hardware.  The simulation would produce the
+necessary volume of trials to allow the learning methods to be
+effective, and the hardware trials would serve to continuously ground
+and refine the simulator.  One could also guide evolution to the most
+fertile territory by penalizing gaits that produced large
+discrepancies between simulation and reality.
 
-There are several directions in which we could continue our
-work. First, the error margins for our runs were large, so reducing
-them by running more trials would lead to more conclusive data. It
-would also be insightful to run our algorithms on different motion
-models.  We suspect that our choice of motion model influenced the
-results greatly, as even random choices in the space produced gaits
-that moved a significant fraction of the time (> 5\%).  It would be
-interesting to see how learning methods would perform using a model
-that included a much higher percentage of unproductive gaits. We also
-intend to experiment further with SVM regression and evolutionary
-algorithms/HyperNEAT.  Some parameter vectors resulted in the robot
-turning, as opposed to it walking long distances. This could be an
-interesting learning goal in future projects. To these ends, we
-propose the following additions and enhancements:
-
-\begin{itemize}
-\item More runs and/or longer runs
-\item Different motion representations
-\item Better tuning of SVM regression
-\item Evolutionary algorithms/HyperNEAT\cite{clune}
-\item Learning how to turn
-\end{itemize}
+Another natural extension would be to allow gaits with feedback;
+i.e. gaits that depend not only on time, but also the measured
+position of the legs, voltage drop across the servos, etc.

### 2011-02-10 paper/results.tex
changes

--- 2011-02-10 paper/results.tex
@@
 
-\acmFig{std_error}{1}{Average results ($\pm $ standard error) for each
+\acmFig{std_error}{1}{Average results ($\pm $ SE) for each
   of the parameterized learning methods.  Linear regression found the
   fastest overall gait and had the highest average, followed by
-  Nelder-Mead simplex. Many methods were beaten by the simple random
+  Nelder-Mead simplex. Many methods were beaten by the random
   strategy.}
@@
 
-The results for the gaits produced and evolved by HyperNEAT are shown
-in \figref{hnResults} and \tabref{results}.  Overall the evolved gaits
+The results for the gaits evolved by HyperNEAT are shown
+in \figref{hnResults} and \tabref{results}.  Overall the HyperNEAT gaits
 were the fastest by far, beating all the parameterized models when
-comparing either the average or the best gaits of each.
+comparing either average or best gaits.  We believe
+that this is because HyperNEAT was allowed to explore a
+much richer space of motions, but did so while still utilizing
+symmetries when advantageous.
+\figref{neat_110115_211410_00000_002_filt_zoom} shows a typical
+HyperNEAT gait that had high fitness.  The pattern of motion is both
+complex (containing multiple frequencies and repeating patterns across time)
+and regular, in that patterns of multiple motors are coordinated.
 
-\acmFig{hnResults}{1}{Average fitness ($\pm $ standard error) of
-  generation champion (highest performing individual among the
-  population of nine individuals) for each generation of the three
-  HyperNEAT runs.}
+The evaluation of the gaits produced by HyperNEAT was more noisy than for the parameterized gaits, which made learning difficult.
+For example, we tested an example HyperNEAT generation-champion gait 11 times and found that its mean performance was 26 body lengths per minute ($\pm$ 13 SD), but it had a max of 38 and a min of 3.
+Many effective HyperNEAT gaits were not preserved across generations because if performance in one trial was poor, the genome was unlikely to be selected for.
+We believe the slope of the HyperNEAT learning curve would be steeper if the noise in the evaluations could be reduced.
+
+\acmFig{hnResults}{1}{Average fitness ($\pm $ SE) of
+  the highest performing individual in the population for each generation of HyperNEAT runs.
+The fitness of many high-performing HyperNEAT gaits were halved if the gait overly stressed the motors (see text), meaning that HyperNEAT's true performance without this penalty would be much higher.}
 

### 2011-02-10 paper/conclusion.tex
Added paperOutline

@@
 \section{Conclusion and Future Work}
-
-\edit{write this, old section below}
-
-% Briefly summarize the important results and conclusions presented in 
-% the paper. What are the most important points illustrated by your 
-% work? How will your results improve future research and applications 
-% in the area?
+\seclabel{conclusion}
 
 We have presented an array of approaches to optimizing a quadrupedal
-gait for forward speed.  We have implemented and tested different
-learning strategies, including uniform and Gaussian random hill
-climbing, policy gradient reinforcement learning, Nelder-Mead simplex,
-several new predictive methods based on linear and support vector
-regression, and an evolved neural network (HyperNEAT).  We have also
-compared these approaches to random search as a baseline. Many of the
-methods resulted in walks significantly faster than previously
-hand-tuned gaits.
+gait for forward speed.  We implemented and tested six learning
+strategies for parameterized gaits and compared them with gaits
+encoded and evolved by HyperNEAT.
 
-Because only three trials were tested with each algorithm and no
-algorithm consistently outperformed the others, there was a large
-standard error for each method, as shown in \figref{std_error}. Thus
-it was unclear if any algorithm was superior to another in this
-application. More trials with each algorithm would be necessary to
-reach a definitive ranking. Because each algorithm discovered at least one gait of
-over 10 body lengths/minute, including random search, we also
-conjectured that the motion representation for the robot is more
-integral to forward speed than the learning algorithm.  How to learn
-the motion representation, in addition to its parameters, remains an
-open problem.
+All methods resulted in an improvement over the robot's previous
+hand-coded gait.  Using linear regression to build a model of gait
+performance and predict promising directions for further exploration
+seemed to work well, producing at best a gait 27.58 of body
+lengths/minute.  The Nelder-Mead simplex method also worked fairly
+well, likely due to its robustness to noise.  The other parameterized
+methods did not outperform a random gait generation method.  Because
+randomly choosing SineModel5 gaits performed so well, we concluded
+that the motion representation for the robot may be more critical to
+parameterized gait development than the learning algorithm itself.
+
+The HyperNEAT gaits performed much better than all parameterized
+methods, considering both the average and best gaits for each.  We
+conjecture that this was because the HyperNEAT encoding allowed it to
+take advantage of the symmetries of the problem, but otherwise left
+the evolution to explore a much larger space of possibilities than the
+more restrictive 5-dimensional parameterized space.  The typical
+higher-performance HyperNEAT gait pictured in
+\figref{neat_110115_211410_00000_002_filt_zoom} shows the fairly
+complex pattern of motion that was obtained.  Whereas the
+parameterized gaits were restricted to scaled and shifted sine waves
+of a single frequency, this evolved gait contains some patterns with
+multiple frequencies.  Despite this larger space of possibilities,
+this gait also reused the same pattern for multiple motors.

### 2011-02-10 paper/results.tex
minor

--- 2011-02-10 paper/results.tex
@@
 \subsection{Exploration of Parameterized Gait Space}
+
+\edit{Jeff: proofread again}
 
@@
 measurement noise is often significant and is more likely to be large
-for gaits that move more.  Of course, this is only a slice in each
+for larger measurements.  Of course, this is only a slice in each
 dimension through a single point, and slices through a different point
-could reveal different behavior.  The common point at the
+would produce different behavior.  The common point at the
 intersection of all slices is shown as a red triangle in
@@
 \figref{std_error} and \tabref{results}.  As a group, these strategies
-produced results equal to or worse not significantly better than the randomly-generated
+produced results not significantly better than the randomly-generated
 gaits, however, all methods managed to explore the space enough to
@@
   fastest overall gait and had the highest average, followed by
-  Nelder-Mead simplex. Many methods were beaten by the simple random
+  Nelder-Mead simplex. Other methods did not outperform a random
   strategy.}

### 2011-02-10 paper/abstract.tex
Working on abstract

@@
+Applications of walking robots often call for the ability to walk as
+quickly, efficiently, or with as little power as possible.  Gaits to
+achieve these objectives may be manually designed by an expert or
+learned by repeated trial and error.
+
+Learning approaches differ in their starting assumptions, some
+essentially tweaking the parameters of a hand-tuned model, others
+exploring a reasonably compact parameter space, and still others
+beginning with few assumptions besides periodicity.
+
+This study extensively compares three methods of designing gaits:
+manually design, local search of parameterized motion models, and
+evolution of artificial neural networks using the HyperNEAT encoding.
+
+All tests....
+
+While these three methods have been used individually for designing
+gaits, to the authors' knowledge they have never been compared
+
+LR works, NM works
+
+% In this study we make an ex
+% We aimed to strike a middle ground between these approaches.  Our 
+% motion generator did not rely on hand-tweaked gaits, but it did use 
+% parameterized gaits which, by their nature, impose some assumptions on 
+% the answers produced.  We then used machine learning to design gaits 
+% for a quadruped robot with these models.  This paper presents a 
+% comparison of the different learning methods implemented.  Most 
+% methods created walks that are several times faster than the original 
+% hand-tuned gait.  We invite readers with short attention spans to view 
+% a video of some of our results online here:
+
+We have presented an array of approaches to optimizing a quadrupedal
+gait for forward speed.  We implemented and tested six learning
+strategies for parameterized gaits and compared them with gaits
+encoded and evolved by HyperNEAT.
+
+All methods resulted in an improvement over the robot's previous
+hand-coded gait.  Using linear regression to build a model of gait
+performance and predict promising directions for further exploration
+seemed to work well, producing at best a gait 27.58 of body
+lengths/minute.  The Nelder-Mead simplex method also worked fairly
+well, likely due to its robustness to noise.  The other parameterized
+methods did not outperform a random gait generation method.  Because
+randomly choosing SineModel5 gaits performed so well, we concluded
+that the motion representation for the robot may be more critical to
+parameterized gait development than the learning algorithm itself.
+
+The HyperNEAT gaits performed much better than all parameterized
+methods, considering both the average and best gaits for each.  We
+conjecture that this was because the HyperNEAT encoding allowed it to
+take advantage of the symmetries of the problem, but otherwise left
+the evolution to explore a much larger space of possibilities than the
+more restrictive 5-dimensional parameterized space.  The typical
+higher-performance HyperNEAT gait pictured in
+\figref{neat_110115_211410_00000_002_filt_zoom} shows the fairly
+complex pattern of motion that was obtained.  Whereas the
+parameterized gaits were restricted to scaled and shifted sine waves
+of a single frequency, this evolved gait contains some patterns with
+multiple frequencies.  Despite this larger space of possibilities,
+this gait also reused the same pattern for multiple motors.
+
 % A brief summary of the report. This is not an introduction -

### 2011-02-10 paper/abstract.tex
abstract draft 1 finished

--- 2011-02-09 paper/abstract.tex
@@
-% A brief summary of the report. This is not an introduction -
-% it should be complete. Include 1-2 sentences for each of the items
-% below, up to and including conclusions.
+Applications of walking robots often call for the ability to walk as
+quickly, efficiently, or with as little power as possible.  Gaits to
+achieve these objectives may designed manually or learned by repeated
+trial and error.
+Learning approaches differ in their starting assumptions, some
+tweaking the parameters of a hand-tuned model, others exploring a
+reasonably compact parameter space, and still others beginning with
+few assumptions besides periodicity.
 
-This paper presents an array of approaches to optimizing a quadrupedal
-gait for forward speed.  We implement, test, and compare different
-learning strategies including uniform and Gaussian random hill
-climbing, policy gradient reinforcement learning,
-Nelder-Mead simplex, new predictive methods based on linear
-and support vector regression, and an evolved neural network
-(HyperNEAT).  We compare results to a baseline random
-search method.  Many of the methods resulted in walks significantly
-faster than previously hand-tuned gaits.
+This study compares the performance of two methods of learning gaits:
+local search of parameterized motion models and evolution of
+artificial neural networks using the HyperNEAT encoding.
 
-%Because the fastest learned walk
-%was not significantly faster than the fastest randomly generated walk,
-%we conjecture that the motion representation for the robot is more
-%integral to forward speed than the learning algorithm.
+We tested six different learning strategies for parameterized gaits,
+including uniform and Gaussian random hill climbing, policy gradient
+reinforcement learning, Nelder-Mead simplex, and a new method that
+uses linear regression to build a model of the fitness landscape and
+predict promising areas of parameter space for further exploration.
+While all parameter search methods outperformed a manually designed
+gait, only the Nelder-Mead simplex and linear regression strategies
+beat a random baseline strategy.
 
-\edit{Write the abstract}
+The HyperNEAT gaits performed considerably better than all
+parameterized local search methods.  Successful evolved gaits showed
+complex motion patterns containing multiple fundamental frequencies,
+but they also demonstrated reuse of patterns among several motors.
+Both served to produce quick gaits.
+
+All tests were performed directly in hardware on a quadruped robot
+with nine degrees of freedom.  To the authors' knowledge this is the
+first time HyperNEAT gaits have been evolved or tested in hardware.

### 2011-02-10 paper/introduction.tex
intro edits

@@
 
-\acmFig{robot_close.jpg}{1}{The quadruped robot.
-The translucent parts were created with 3D-printing technology.
-Videos of the gaits we evolved for this robot can be viewed online~\cite{quadratotWeb}.}
+\acmFig{robot_close.jpg}{1}{The quadruped robot. The translucent parts
+  were created with 3D-printing technology. Videos of the gaits
+  evolved for this robot can be viewed online~\cite{quadratotWeb}.}
 
@@
 
-Once the design goals are decided upon, gaits may be obtained by one of several methods.
-Gaits may be designed manually by an expert, as for Asimo and Big Dog \cite{chestnutt2006footstep, raibert2008bigdog}.
-Alternately, they may be learned through repeated trial and evaluation.
-Learned gaits offer several advantages over manually designed gaits.
-Automatically learning gaits can save valuable engineering time and can allow customization of gaits to a particular robot and its unique actuators.
-Most importantly, in some cases learned gaits can outperform manually designed gaits~\cite{hornby2005autonomous, valsalam2008modular}.
+Once the design goals are decided upon, gaits may be obtained by one
+of several methods.  Gaits may be designed manually by an expert, as
+for Asimo and Big Dog \cite{chestnutt2006footstep, raibert2008bigdog}.
+Alternately, they may be learned through repeated trial and
+evaluation.  Learned gaits offer several advantages over manually
+designed gaits.  Automatically learning gaits can save valuable
+engineering time and can allow customization of gaits to a particular
+robot and its unique actuators.  Most importantly, in many cases
+learned gaits can outperform manually designed
+gaits~\cite{hornby2005autonomous, valsalam2008modular}.
 
-In this paper we compare the performance of three different methods of designing gaits: manual design, parameterized gaits optimized with seven different learning methods, and gaits generated by evolving neural networks with the HyperNEAT generative encoding~\cite{stanley2009hypercube}.
-While some of these methods, such as HyperNEAT, have been tested in simulation~\cite{clune2009evolving, clune2011performance}, we investigate how they perform when evolving on a physical robot (\figref{robot_close.jpg}).
+In this paper we compare the performance of two different methods of
+learning gaits: parameterized gaits optimized with six different
+learning methods, and gaits generated by evolving neural networks with
+the HyperNEAT generative encoding~\cite{stanley2009hypercube}. While
+some of these methods, such as HyperNEAT, have been tested in
+simulation~\cite{clune2009evolving, clune2011performance}, we
+investigate how they perform when evolving on a physical robot
+(\figref{robot_close.jpg}).
 
-Previous work has shown that quadruped gaits perform better when they are \emph{regular} (i.e.\ when the legs are coordinated)~\cite{clune2009evolving, clune2011performance,valsalam2008modular}.
-For example, HyperNEAT produced fast, natural gaits in part because its bias towards regular gaits created coordinated movements that outperformed gaits evolved by an encoding not biased towards regularity~\cite{clune2009evolving, clune2011performance}.
-One of the motivations of this paper is to investigate whether any learning method biased towards regularity would perform well at producing quadruped gaits, or whether HyperNEAT's high performance is due to additional factors, such as its abstraction of biological development (described below).
-We test this hypothesis by comparing HyperNEAT to seven machine learning algorithms biased toward regularity.
+Previous work has shown that quadruped gaits perform better when they
+are \emph{regular} (i.e.\ when the legs are
+coordinated)~\cite{clune2009evolving,
+  clune2011performance,valsalam2008modular}. For example, HyperNEAT
+produced fast, natural gaits in part because its bias towards regular
+gaits created coordinated movements that outperformed gaits evolved by
+an encoding not biased towards regularity~\cite{clune2009evolving,
+  clune2011performance}. One of the motivations of this paper is to
+investigate whether any learning method biased towards regularity
+would perform well at producing quadruped gaits, or whether
+HyperNEAT's high performance is due to additional factors, such as its
+abstraction of biological development (described below). We test this
+hypothesis by comparing HyperNEAT to six local search algorithms with
+a parametrization biased toward regularity.
 
-A final motivation of this problem is to simply evolve effective gaits for a physical robot.
-Because HyperNEAT performed well in simulation, it is interesting to test whether it can produce a fast gait for a physical robot.
-It is additionally interesting to test how more traditional machine learning techniques compete with evolutionary algorithms when evolving in hardware.
+A final motivation of this problem is to simply evolve effective gaits
+for a physical robot. Because HyperNEAT performed well in simulation,
+it is interesting to test whether it can produce a fast gait for a
+physical robot. It is additionally interesting to test how more
+traditional gait optimization techniques compete with evolutionary
+algorithms when evolving in hardware.
 

### 2011-02-10 paper/conclusion.tex

--- 2011-01-20 paper/conclusion.tex
@@
 \section{Conclusion and Future Work}
+\seclabel{conclusion}
 
-\edit{write this, old section below}
+We have presented an array of approaches to optimizing a fast quadrupedal
+gait.  We implemented and tested six learning
+strategies for parameterized gaits and compared them to gaits produced by neural networks
+evolved with the HyperNEAT generative encoding.
 
-% Briefly summarize the important results and conclusions presented in 
-% the paper. What are the most important points illustrated by your 
-% work? How will your results improve future research and applications 
-% in the area?
+All methods resulted in an improvement over the robot's previous
+hand-coded gait.  Building a model of gait
+performance with linear regression to predict promising directions for further exploration
+worked well, producing a gait of 27.58 body
+lengths/minute.  The Nelder-Mead simplex method performed nearly as well, likely due to its robustness to noise.  The other parameterized
+methods did not outperform random search.
+One reason the randomly-generated SineModel5 gaits performed so well could be because the gait representation was biased towards effective, regular gaits, making the highly exploratory random strategy more effective than more exploitative learning algorithms.
 
-We have presented an array of approaches to optimizing a quadrupedal
-gait for forward speed.  We have implemented and tested different
-learning strategies, including uniform and Gaussian random hill
-climbing, policy gradient reinforcement learning, Nelder-Mead simplex,
-several new predictive methods based on linear and support vector
-regression, and an evolved neural network (HyperNEAT).  We have also
-compared these approaches to random search as a baseline. Many of the
-methods resulted in walks significantly faster than previously
-hand-tuned gaits.
-
-Because only three trials were tested with each algorithm and no
-algorithm consistently outperformed the others, there was a large
-standard error for each method, as shown in \figref{std_error}. Thus
-it was unclear if any algorithm was superior to another in this
-application. More trials with each algorithm would be necessary to
-reach a definitive ranking. Because each algorithm discovered at least one gait of
-over 10 body lengths/minute, including random search, we also
-conjectured that the motion representation for the robot is more
-integral to forward speed than the learning algorithm.  How to learn
-the motion representation, in addition to its parameters, remains an
-open problem.
+HyperNEAT produced higher-performing gaits than all of the parameterized
+methods.  This could be because HyperNEAT tends to generate coordinated gaits~\cite{clune2011performance, clune2009evolving}, allowing it to
+take advantage of the symmetries of the problem, but can also explore a much larger space of possibilities than the
+more restrictive 5-dimensional parameterized space.  HyperNEAT gaits tended to produce more complex sequences of motor commands, with different frequencies and degrees of coordination, whereas the
+parameterized gaits were restricted to scaling single-frequency sine waves and could only produce certain types of motor regularities.

### 2011-02-10 paper/futureWork.tex

@@
 Because all trials were done in hardware, it was difficult to gather
-the thousands of trials that would be necessary to establish a more
-rigorous statistical ranking of the methods' performance, or their
-asymptotic behavior as longer and longer training runs are allowed.
-Thus, one extension of the current work would be to obtain these
-thousands more trials.
+the many trials that would be necessary to properly rank the methods statistically.
+One extension of the current work could be to obtain many more trials.
 
-However, a more effective extension might be to combine trials in
-simulation and in hardware, perhaps doing only every tenth or
-hundredth trial in hardware.  The simulation would produce the
+However, a more effective extension might be to combine frequent trials in
+simulation with infrequent trials in hardware~\cite{bongard}.  The simulation would produce the
 necessary volume of trials to allow the learning methods to be
@@
 fertile territory by penalizing gaits that produced large
-discrepancies between simulation and reality.
+discrepancies between simulation and reality~{koos2010crossing}.
 
-Another natural extension would be to allow gaits with feedback;
-i.e. gaits that depend not only on time, but also the measured
-position of the legs, voltage drop across the servos, etc.
+Another natural extension would be to allow gaits that sensed the position of the robot and other variables to enable the robot to adjust to its physical state, instead of providing an open-loop sequence of motor commands.

### 2011-02-10 paper/results.tex
5.2 done + minor

--- 2011-02-10 paper/results.tex
@@
 The results for the parameterized gaits are shown in
-\figref{std_error} and \tabref{results}.  As a group, these strategies
-produced results equal to or worse not significantly better than the randomly-generated
-gaits, however, all methods managed to explore the space enough to
-significantly improve on the previous hand coded gait in at least one
-of the three runs.  No single strategy consistently outperformed the
-others: for the first trial Linear Regression produced the fastest
-gait at 27.58 body lengths/minute, for the second a random gait
-actually won with 17.26 body lengths/minute, and for the third trial
-the Nelder-Mead simplex method attained the fastest gait at 14.83 body
-lengths/minute.
+\figref{std_error} and \tabref{results}.  The best overall gait for
+the parameterized methods was found by linear regression, which also
+had the highest average performance. The Nelder-Mead simplex also
+performed quite well on average.  The other local search methods did not
+outperform random search; however, all methods did manage to explore
+enough of the parameter space to significantly improve on the previous
+hand-coded gait in at least one of the three runs.  No single strategy
+consistently beat the others: for the first trial Linear Regression
+produced the fastest gait at 27.58 body lengths/minute, for the second
+a random gait actually won with 17.26 body lengths/minute, and for the
+third trial the Nelder-Mead simplex method attained the fastest gait
+at 14.83 body lengths/minute.
+
+One reason the randomly-generated SineModel5 gaits were so effective
+may have been due to the SineModel5's bias toward regular, symmetric
+gaits.  This may have allowed the random strategy --- focusing on
+exploration --- to be competitive with the more directed strategies
+that exploit information from past runs.
 
@@
 
-\acmFig{std_error}{1}{Average results ($\pm $ SE) for each
-  of the parameterized learning methods.  Linear regression found the
-  fastest overall gait and had the highest average, followed by
-  Nelder-Mead simplex. Many methods were beaten by the random
-  strategy.}
+\acmFig{std_error}{1}{Average results ($\pm $ SE) for each of the
+  parameterized learning methods.  Linear regression found the fastest
+  overall gait and had the highest average, followed by Nelder-Mead
+  simplex. Other methods did not outperform a random strategy.}
 

### 2011-02-10 paper/abstract.tex
abstract

--- 2011-02-09 paper/abstract.tex
@@
-% A brief summary of the report. This is not an introduction -
-% it should be complete. Include 1-2 sentences for each of the items
-% below, up to and including conclusions.
+Creating gaits for legged robots is an important task to enable robots to access rugged terrain, yet designing such gaits by hand is a challenging and time-consuming process.
+In this paper we investigate various algorithms for automating the creation of quadruped gaits.
+Because many robots do not have accurate simulators, we test gait learning algorithms entirely on a physical robot.
+	We compare the performance of two classes of learning gaits:
+locally searching parameterized motion models and evolving artificial neural networks with the HyperNEAT generative encoding.
+Specifically, we test six different parameterized learning strategies: uniform and Gaussian random hill climbing, policy gradient
+reinforcement learning, Nelder-Mead simplex, and a new method that
+builds a model of the fitness landscape with linear regression to guide further exploration.
+While all parameter search methods outperformed a manually designed
+gait, only the linear regression and Nelder-Mead simplex strategies
+outperform a random baseline strategy. Gaits evolved by HyperNEAT perform considerably better than all
+parameterized local search methods.  The best HyperNEAT gaits exhibit
+complex motion patterns containing multiple fundamental frequencies,
+yet are regular in that the leg movements are coordinated. Overall, this research illuminates various approaches to evolving quadruped gaits directly in hardware.
 
-This paper presents an array of approaches to optimizing a quadrupedal
-gait for forward speed.  We implement, test, and compare different
-learning strategies including uniform and Gaussian random hill
-climbing, policy gradient reinforcement learning,
-Nelder-Mead simplex, new predictive methods based on linear
-and support vector regression, and an evolved neural network
-(HyperNEAT).  We compare results to a baseline random
-search method.  Many of the methods resulted in walks significantly
-faster than previously hand-tuned gaits.
-
-%Because the fastest learned walk
-%was not significantly faster than the fastest randomly generated walk,
-%we conjecture that the motion representation for the robot is more
-%integral to forward speed than the learning algorithm.
-
-\edit{Write the abstract}
+%Learning approaches differ in their starting assumptions, some
+%tweaking the parameters of a hand-tuned model, others exploring a
+%reasonably compact parameter space, and still others beginning with
+%few assumptions besides periodicity.

### 2011-02-10 paper/learningMethods.tex
spelling
NM shrink

@@
-\subsection{Parametrized Gaits}
+\subsection{Parameterized Gaits}
 \seclabel{motionModel}
@@
 
-\subsection{Learning Methods for Parametrized Gaits}
+\subsection{Learning Methods for Parameterized Gaits}
 \seclabel{learningMethods}
@@
 
+%\item \emph{Nelder-Mead simplex method}: The Nelder-Mead simplex
+%  method \cite{nm} creates an initial simplex with $d+1$ vertices,
+%  where $d$ is the dimension of the parameter space. The initial
+%  parameter vector is stored as the first vertex and the other five
+%  vertices are created by adding to one dimension at a time one tenth
+%  of the allowable range for that parameter. It then tests the fitness
+%  of each vertex, and, in general, it reflects the worst point over
+%  the centroid in an attempt to improve it.  However, to prevent
+%  cycles and becoming stuck in local minima, several other rules are
+%  used.  If the reflected point is better than the second worst point
+%  and worse than the best point, then the reflected point replaces the
+%  worst. If the reflected point is better than the best point, the
+%  simplex is expanded in the direction of the reflected point. The
+%  better of the reflected and the expanded point replaces the worst
+%  point. If the reflected point is worse than the second worst point,
+%  then the simplex is contracted away from the reflected point. If the
+%  contracted point is better than the reflected point, the contracted
+%  point replaces the worst point. If the contracted point is worse
+%  than the reflected point, the entire simplex is shrunk \cite{nm}.
+
 \item \emph{Nelder-Mead simplex method}: The Nelder-Mead simplex
   method \cite{nm} creates an initial simplex with $d+1$ vertices,
-  where $d$ is the dimension of the parameter space. The initial
-  parameter vector is stored as the first vertex and the other five
-  vertices are created by adding to one dimension at a time one tenth
-  of the allowable range for that parameter. It then tests the fitness
-  of each vertex, and, in general, it reflects the worst point over
-  the centroid in an attempt to improve it.  However, to prevent
-  cycles and becoming stuck in local minima, several other rules are
-  used.  If the reflected point is better than the second worst point
-  and worse than the best point, then the reflected point replaces the
-  worst. If the reflected point is better than the best point, the
-  simplex is expanded in the direction of the reflected point. The
-  better of the reflected and the expanded point replaces the worst
-  point. If the reflected point is worse than the second worst point,
-  then the simplex is contracted away from the reflected point. If the
-  contracted point is better than the reflected point, the contracted
-  point replaces the worst point. If the contracted point is worse
-  than the reflected point, the entire simplex is shrunk \cite{nm}.
+  where $d$ is the dimension of the parameter space. It then tests the
+  fitness of each vertex, and, in general, it reflects the worst point
+  over the centroid in an attempt to improve it.  Several additional
+  rules are used to prevent cycles and local minima \cite{nm}.
 

### 2011-02-10 paper/results.tex
shrinking figs

@@
 
-\acmFig{hnResults}{1}{Average fitness ($\pm $ SE) of
-  the highest performing individual in the population for each generation of HyperNEAT runs.
-The fitness of many high-performing HyperNEAT gaits were halved if the gait overly stressed the motors (see text), meaning that HyperNEAT's true performance without this penalty would be much higher.}
+\acmFig{hnResults}{.8}{Average fitness ($\pm $ SE) of the highest
+  performing individual in the population for each generation of
+  HyperNEAT runs. The fitness of many high-performing HyperNEAT gaits
+  were halved if the gait overly stressed the motors (see text),
+  meaning that HyperNEAT's true performance without this penalty would
+  be much higher.}
 
@@
 
-\acmFig{neat_110115_211410_00000_002_filt_zoom}{1}{Example of one
-  particular high-performance gait produced by HyperNEAT showing
+\acmFig{neat_110115_211410_00000_002_filt_zoom}{.8}{Example of one
+  high-performance gait produced by HyperNEAT showing
   commands for each of nine motors.  Note the complexity of the motion

### 2011-02-10 paper/futureWork.tex
foo

@@
 fertile territory by penalizing gaits that produced large
-discrepancies between simulation and reality~{koos2010crossing}.
-
+discrepancies between simulation and reality~\cite{koos2010crossing}.
 Another natural extension would be to allow gaits that sensed the position of the robot and other variables to enable the robot to adjust to its physical state, instead of providing an open-loop sequence of motor commands.

### 2011-02-10 paper/learningMethods.tex
foo

@@
 
-\begin{itemize}
-
-\item \emph{Random}: This method randomly generates parameter vectors
-
-  in the allowable range for every trial.  This strategy was used only
+\emph{Random}: This method randomly generates parameter vectors in the allowable range for every trial.  This strategy was used only
   as baseline for comparison.
 
-\item \emph{Uniform random hill climbing}: This method always starts
+\emph{Uniform random hill climbing}: This method always starts
   with the current best gait, and then chooses next $\vec{\theta}$ by
@@
 
-\item \emph{Gaussian random hill climbing}: This method works
+\emph{Gaussian random hill climbing}: This method works
   similarly to Uniform random hill climbing, except the next
@@
 
-\item \emph{N-dimensional policy gradient ascent}: We implemented Kohl
+\emph{N-dimensional policy gradient ascent}: We implemented Kohl
   and Stone's \cite{kohl} method for local gradient ascent for gait
@@
 
-\item \emph{Nelder-Mead simplex method}: The Nelder-Mead simplex
+\emph{Nelder-Mead simplex method}: The Nelder-Mead simplex
   method \cite{nm} creates an initial simplex with $d+1$ vertices,
@@
 
-\item \emph{Linear regression}: To initialize, this method chooses and
+\emph{Linear regression}: To initialize, this method chooses and
   evaluates five random parameter vectors. It then fits a linear model
@@
   \emph{N-dimensional policy gradient ascent}.
-
-\end{itemize}
 

### 2011-02-10 paper/abstract.tex
minor + shrink

@@
 Specifically, we test six different parameterized learning strategies: uniform and Gaussian random hill climbing, policy gradient
-reinforcement learning, Nelder-Mead simplex, and a new method that
+reinforcement learning, Nelder-Mead simplex, a random baseline, and a new method that
 builds a model of the fitness landscape with linear regression to guide further exploration.
@@
 parameterized local search methods.  The best HyperNEAT gaits exhibit
-complex motion patterns containing multiple fundamental frequencies,
+complex motion patterns containing multiple frequencies,
 yet are regular in that the leg movements are coordinated. Overall, this research illuminates various approaches to evolving quadruped gaits directly in hardware.

### 2011-02-10 paper/hyperNeatMethod.tex
shrink
foo
minor + shrink

@@
-\subsection{HyperNEAT Gait Generation and Learning}
+\subsection{HyperNEAT Gait Generation \& Learning}
 \seclabel{hyperNeatMethod}
@@
 
-\acmFig{hyperneatExplanation.png}{1}{HyperNEAT Produces ANNs from
+\acmFig{hyperneatExplanation.png}{.8}{HyperNEAT produces ANNs from
   CPPNs. ANN weights are specified as a function of the geometric
@@
 
-\acmFig{SpiderANN.jpg}{1}{ANN Configuration for HyperNEAT Runs. The
+\acmFig{SpiderANN.jpg}{.7}{ANN configuration for HyperNEAT Runs. The
   first two columns of each row of the input layer receive information
@@
 
-As with the parameterized learning methods, three runs of HyperNEAT
+As with the parameterized methods, three runs of HyperNEAT
 were performed. The population size for HyperNEAT was 9 and runs

### 2011-02-10 paper/introduction.tex
shrinking figs
foo
cut intro

@@
 
-\acmFig{robot_close.jpg}{1}{The quadruped robot. The translucent parts
+\acmFig{robot_close.jpg}{.8}{The quadruped robot. The translucent parts
   were created with 3D-printing technology. Videos of the gaits
-  evolved for this robot can be viewed online~\cite{quadratotWeb}.}
+  evolved for this robot can be viewed at www.anonymous.url.}
 
-Gaits for walking robots are often designed with some explicit or
-implicit goal in mind.  For some applications, the design criteria may
-be obvious --- perhaps the robot needs to move as quickly or as
-efficiently as possible --- but other times the objective is more complicated,
-requiring simultaneous optimization of several desired traits, each
-with its own relative importance.  The different combinations of desired traits and the relative weight placed on each can produce drastically different gaits.    For example, Honda's Asimo
-\cite{chestnutt2006footstep} and Boston Dynamic's Big Dog \cite{raibert2008bigdog} both require
-gaits that are relatively quick, power efficient, and robust to
-changing terrain, but they vary widely in the importance placed
-on each attribute.  Big Dog's gait is optimized for much more difficult terrain than Asimo's, resulting in a gait of a completely different form.
+%Gaits for walking robots are often designed with some explicit or
+%implicit goal in mind.  For some applications, the design criteria may
+%be obvious --- perhaps the robot needs to move as quickly or as
+%efficiently as possible --- but other times the objective is more complicated,
+%requiring simultaneous optimization of several desired traits, each
+%with its own relative importance.  The different combinations of desired traits and the relative weight placed on each can produce drastically different gaits.    For example, Honda's Asimo
+%\cite{chestnutt2006footstep} and Boston Dynamic's Big Dog \cite{raibert2008bigdog} both require
+%gaits that are relatively quick, power efficient, and robust to
+%changing terrain, but they vary widely in the importance placed
+%on each attribute.  Big Dog's gait is optimized for much more difficult terrain than Asimo's, resulting in a gait of a completely different form.
 
-Once the design goals are decided upon, gaits may be obtained by one
-of several methods.  Gaits may be designed manually by an expert, as
-for Asimo and Big Dog \cite{chestnutt2006footstep, raibert2008bigdog}.
-Alternately, they may be learned through repeated trial and
-evaluation.  Learned gaits offer several advantages over manually
-designed gaits.  Automatically learning gaits can save valuable
-engineering time and can allow customization of gaits to a particular
-robot and its unique actuators.  Most importantly, in many cases
-learned gaits can outperform manually designed
-gaits~\cite{hornby2005autonomous, valsalam2008modular}.
+%Once the design goals are decided upon, gaits may be obtained by one
+
+Legged robots have the potential to access many types of terrain
+unsuitable for wheeled robots, but doing so requires the creation of a
+gait specifying how the robot is to walk.  Gaits may be designed
+by one of several methods: either manually by an expert or
+automatically through repeated trial and evaluation.  Learned gaits
+offer several advantages over manually designed gaits.  Automatically
+learning gaits can save valuable engineering time and can allow
+customization of gaits to a particular robot and its unique actuators.
+Most importantly, in many cases learned gaits can outperform manually
+designed gaits~\cite{hornby2005autonomous, valsalam2008modular}.
 

### 2011-02-10 paper/conclusion.tex

@@
 performance with linear regression to predict promising directions for further exploration
-worked well, producing a gait of 27.58 body
+worked well, producing a gait of 27.5 body
 lengths/minute.  The Nelder-Mead simplex method performed nearly as well, likely due to its robustness to noise.  The other parameterized
@@
 HyperNEAT produced higher-performing gaits than all of the parameterized
-methods.  This could be because HyperNEAT tends to generate coordinated gaits~\cite{clune2011performance, clune2009evolving}, allowing it to
+methods, with a max of 45.7 body lengths per minute, which is nearly 9 times the speed of the hand-coded gait.
+This could be because HyperNEAT tends to generate coordinated gaits~\cite{clune2011performance, clune2009evolving}, allowing it to
 take advantage of the symmetries of the problem, but can also explore a much larger space of possibilities than the

### 2011-02-10 paper/fitnessEvaluation.tex
parametrized -> parameterized

@@
 % - fitness reported in pixels
-% - parametrized models
+% - parameterized models
 %   - started all on three thetas
@@
 %
-% For the parametrized models, the starting location may be important. To be as fair as possible, we picked three random points and starred all models there.
+% For the parameterized models, the starting location may be important. To be as fair as possible, we picked three random points and starred all models there.
 %   generally ran until 1/3 plateau
@@
 
-\acmFigg{wiiMote_crop.jpg}{robot_led_crop.jpg}{1}{A Nintendo Wii
+\acmFigg{wiiMote_crop.jpg}{robot_led_crop.jpg}{.7}{A Nintendo Wii
   remote provided the location of the robot by tracking the infrared

### 2011-02-10 paper/results.tex

@@
 
-The results for the gaits evolved by HyperNEAT are shown
-in \figref{hnResults} and \tabref{results}.  Overall the HyperNEAT gaits
+The results for the gaits evolved by HyperNEAT are shown in
+\figref{hnResults} and \tabref{results}.  Overall the HyperNEAT gaits
 were the fastest by far, beating all the parameterized models when
-comparing either average or best gaits.  We believe
-that this is because HyperNEAT was allowed to explore a
-much richer space of motions, but did so while still utilizing
-symmetries when advantageous.
+comparing either average or best gaits.  We believe that this is
+because HyperNEAT was allowed to explore a much richer space of
+motions, but did so while still utilizing symmetries when
+advantageous.  The single best gait found during this study had a
+speed of 45.72 body lengths / minute, 66\% better than the best
+non-HyperNEAT gait and 8.9 times faster than the hand-coded gait.
 \figref{neat_110115_211410_00000_002_filt_zoom} shows a typical
 HyperNEAT gait that had high fitness.  The pattern of motion is both
-complex (containing multiple frequencies and repeating patterns across time)
-and regular, in that patterns of multiple motors are coordinated.
+complex (containing multiple frequencies and repeating patterns across
+time) and regular, in that patterns of multiple motors are
+coordinated.
 

### 2011-02-10 paper/abstract.tex
abstract minor, title change

@@
-Creating gaits for legged robots is an important task to enable robots to access rugged terrain, yet designing such gaits by hand is a challenging and time-consuming process.
-In this paper we investigate various algorithms for automating the creation of quadruped gaits.
-Because many robots do not have accurate simulators, we test gait learning algorithms entirely on a physical robot.
-	We compare the performance of two classes of learning gaits:
-locally searching parameterized motion models and evolving artificial neural networks with the HyperNEAT generative encoding.
-Specifically, we test six different parameterized learning strategies: uniform and Gaussian random hill climbing, policy gradient
-reinforcement learning, Nelder-Mead simplex, a random baseline, and a new method that
-builds a model of the fitness landscape with linear regression to guide further exploration.
-While all parameter search methods outperformed a manually designed
-gait, only the linear regression and Nelder-Mead simplex strategies
-outperform a random baseline strategy. Gaits evolved by HyperNEAT perform considerably better than all
-parameterized local search methods.  The best HyperNEAT gaits exhibit
-complex motion patterns containing multiple frequencies,
-yet are regular in that the leg movements are coordinated. Overall, this research illuminates various approaches to evolving quadruped gaits directly in hardware.
+Creating gaits for legged robots is an important task to enable robots
+to access rugged terrain, yet designing such gaits by hand is a
+challenging and time-consuming process. In this paper we investigate
+various algorithms for automating the creation of quadruped
+gaits. Because many robots do not have accurate simulators, we test
+gait learning algorithms entirely on a physical robot.  We compare the
+performance of two classes of learning gaits: locally searching
+parameterized motion models and evolving artificial neural networks
+with the HyperNEAT generative encoding. Specifically, we test six
+different parameterized learning strategies: uniform and Gaussian
+random hill climbing, policy gradient reinforcement learning,
+Nelder-Mead simplex, a random baseline, and a new method that builds a
+model of the fitness landscape with linear regression to guide further
+exploration.  While all parameter search methods outperformed a
+manually designed gait, only the linear regression and Nelder-Mead
+simplex strategies outperform a random baseline strategy. Gaits
+evolved by HyperNEAT perform considerably better than all
+parameterized local search methods.  The best evolved gaits exhibit
+complex motion patterns that contain multiple frequencies, yet are
+regular in that the leg movements are coordinated.
+
+%Overall, this
+%research illuminates various approaches to evolving quadruped gaits
+%directly in hardware.
 

### 2011-02-10 paper/conclusion.tex

@@
 HyperNEAT produced higher-performing gaits than all of the parameterized
-methods, with a max of 45.7 body lengths per minute, which is nearly 9 times the speed of the hand-coded gait.
+methods.
+Its best-performing gait traveled 45.7 body lengths per minute, which is nearly 9 times the speed of the hand-coded gait.
 This could be because HyperNEAT tends to generate coordinated gaits~\cite{clune2011performance, clune2009evolving}, allowing it to

### 2011-02-10 paper/results.tex

@@
 exploration --- to be competitive with the more directed strategies
-that exploit information from past runs.
+that exploit information from past evaluations.
 

### 2011-02-10 paper/futureWork.tex
done!

@@
 the many trials that would be necessary to properly rank the methods statistically.
-One extension of the current work could be to obtain many more trials.
-
-However, a more effective extension might be to combine frequent trials in
+One direction for future work could be to obtain many more trials. However, a more effective extension might be to combine frequent trials in
 simulation with infrequent trials in hardware~\cite{bongard}.  The simulation would produce the
@@
 Another natural extension would be to allow gaits that sensed the position of the robot and other variables to enable the robot to adjust to its physical state, instead of providing an open-loop sequence of motor commands.
+All of these approaches would likely improve the quality of automatically generated gaits for legged robots, which will hasten the day that humanity can benefit from their vast potential.

### 2011-02-10 paper/results.tex
Clarified table caption
done!

@@
 produced the fastest gait at 27.58 body lengths/minute, for the second
-a random gait actually won with 17.26 body lengths/minute, and for the
+a random gait actually won with 17.26, and for the
 third trial the Nelder-Mead simplex method attained the fastest gait
-at 14.83 body lengths/minute.
+with 14.83.
 

### 2011-02-10 paper/abstract.tex
abstract change
ff
foo

@@
 model of the fitness landscape with linear regression to guide further
-exploration.  While all parameter search methods outperformed a
-manually designed gait, only the linear regression and Nelder-Mead
+exploration.  While all parameter search methods outperform a
+manually-designed gait, only the linear regression and Nelder-Mead
 simplex strategies outperform a random baseline strategy. Gaits
-evolved by HyperNEAT perform considerably better than all
-parameterized local search methods.  The best evolved gaits exhibit
+evolved with HyperNEAT perform considerably better than all
+parameterized local search methods, and produce gaits nearly 9 times faster than a hand-designed gait.  The best HyperNEAT gaits exhibit
 complex motion patterns that contain multiple frequencies, yet are

### 2011-02-10 paper/abstract.tex
minor

--- 2011-02-10 paper/abstract.tex
@@
 evolved by HyperNEAT perform considerably better than all
-parameterized local search methods.  The best evolved gaits exhibit
+parameterized local search methods and produced gaits nearly 9 times faster than a hand-designed gait.  The best evolved HyperNEAT gaits exhibit
 complex motion patterns that contain multiple frequencies, yet are

### 2011-04-05 paper_ecal/relatedWork.tex
Moved GECCO paper to its own directory, copied to new directory for ECAL paper

--- 2011-02-09 paper/relatedWork.tex
@@
 
-Various machine learning techniques have proven to be effective at generating gaits for legged robots.
-Kohl and Stone presented a policy gradient reinforcement learning approach for generating a fast walk on legged robots\cite{kohl}.
-We experiment with this method to create a walk for our robot (called Policy Gradient Descent, described below).
-Others have evolved gaits for legged robots, producing competitive results~\cite{chernova2005evolutionary, hornby2005autonomous, zykov, clune2009evolving, clune2011performance, clune2009hybrid, clune2009sensitivity, tellez2006evolving, valsalam2008modular}.
-In fact, an evolved gait was used in the first commercially-available version of Sony's AIBO robot~\cite{hornby2005autonomous}.
-Except for work with HyperNEAT~\cite{clune2009evolving, clune2011performance, clune2009hybrid, clune2009sensitivity}, the previous evolutionary approaches have helped evolution 'see' the regularity of the problem by manually decomposing the task.
-In other words, experimenters have to choose which legs should be coordinated, or otherwise facilitate the coordination of robotic legs.
-Part of the motivation of this paper is to see if the space of regularities can be explored automatically, which has previously performed well~\cite{valsalam2008modular}, and to perform a direct comparison on the same robot to HyperNEAT.
+Various machine learning techniques have proved to be effective at
+generating gaits for legged robots. Kohl and Stone presented a policy
+gradient reinforcement learning approach for generating a fast walk on
+legged robots\cite{kohl}, which we implemented for comparison. Others
+have evolved gaits for legged robots, producing competitive
+results~\cite{chernova2005evolutionary, hornby2005autonomous, zykov,
+  clune2009evolving, clune2011performance, clune2009hybrid,
+  clune2009sensitivity, tellez2006evolving, valsalam2008modular}. In
+fact, an evolved gait was used in the first commercially-available
+version of Sony's AIBO robot~\cite{hornby2005autonomous}. Except for
+work with HyperNEAT~\cite{clune2009evolving, clune2011performance,
+  clune2009hybrid, clune2009sensitivity}, the previous evolutionary
+approaches have helped evolution exploit the regularity of the problem
+by manually decomposing the task. Experimenters have
+to choose which legs should be coordinated, or otherwise facilitate
+the coordination of motion. Part of the motivation of this paper
+is to see if the space of regularities can be explored automatically,
+which has previously performed well~\cite{valsalam2008modular}, and to
+perform a direct comparison on the same robot to HyperNEAT.

### 2011-04-05 paper_gecco_8/sample_acm_sig/sig-alternate.tex
Moved GECCO paper to its own directory, copied to new directory for ECAL paper

% This is "sig-alternate.tex" V1.9 April 2009
% This file should be compiled with V2.4 of "sig-alternate.cls" April 2009
%
% This example file demonstrates the use of the 'sig-alternate.cls'
% V2.4 LaTeX2e document class file. It is for those submitting
% articles to ACM Conference Proceedings WHO DO NOT WISH TO
% STRICTLY ADHERE TO THE SIGS (PUBS-BOARD-ENDORSED) STYLE.
% The 'sig-alternate.cls' file will produce a similar-looking,
% albeit, 'tighter' paper resulting in, invariably, fewer pages.
%
% ----------------------------------------------------------------------------------------------------------------
% This .tex file (and associated .cls V2.4) produces:
%       1) The Permission Statement
%       2) The Conference (location) Info information
%       3) The Copyright Line with ACM data
%       4) NO page numbers
%
% as against the acm_proc_article-sp.cls file which
% DOES NOT produce 1) thru' 3) above.
%
% Using 'sig-alternate.cls' you have control, however, from within
% the source .tex file, over both the CopyrightYear
% (defaulted to 200X) and the ACM Copyright Data
% (defaulted to X-XXXXX-XX-X/XX/XX).
% e.g.
% \CopyrightYear{2007} will cause 2007 to appear in the copyright line.
% \crdata{0-12345-67-8/90/12} will cause 0-12345-67-8/90/12 to appear in the copyright line.
%
% ---------------------------------------------------------------------------------------------------------------
% This .tex source is an example which *does* use
% the .bib file (from which the .bbl file % is produced).
% REMEMBER HOWEVER: After having produced the .bbl file,
% and prior to final submission, you *NEED* to 'insert'
% your .bbl file into your source .tex file so as to provide
% ONE 'self-contained' source file.
%
% ================= IF YOU HAVE QUESTIONS =======================
% Questions regarding the SIGS styles, SIGS policies and
% procedures, Conferences etc. should be sent to
% Adrienne Griscti (griscti@acm.org)
%
% Technical questions _only_ to
% Gerald Murray (murray@hq.acm.org)
% ===============================================================
%
% For tracking purposes - this is V1.9 - April 2009

\documentclass{sig-alternate}

\begin{document}
%
% --- Author Metadata here ---
\conferenceinfo{WOODSTOCK}{'97 El Paso, Texas USA}
%\CopyrightYear{2007} % Allows default copyright year (20XX) to be over-ridden - IF NEED BE.
%\crdata{0-12345-67-8/90/01}  % Allows default copyright data (0-89791-88-6/97/05) to be over-ridden - IF NEED BE.
% --- End of Author Metadata ---

\title{Alternate {\ttlit ACM} SIG Proceedings Paper in LaTeX
Format\titlenote{(Produces the permission block, and
copyright information). For use with
SIG-ALTERNATE.CLS. Supported by ACM.}}
\subtitle{[Extended Abstract]
\titlenote{A full version of this paper is available as
\textit{Author's Guide to Preparing ACM SIG Proceedings Using
\LaTeX$2_\epsilon$\ and BibTeX} at
\texttt{www.acm.org/eaddress.htm}}}
%
% You need the command \numberofauthors to handle the 'placement
% and alignment' of the authors beneath the title.
%
% For aesthetic reasons, we recommend 'three authors at a time'
% i.e. three 'name/affiliation blocks' be placed beneath the title.
%
% NOTE: You are NOT restricted in how many 'rows' of
% "name/affiliations" may appear. We just ask that you restrict
% the number of 'columns' to three.
%
% Because of the available 'opening page real-estate'
% we ask you to refrain from putting more than six authors
% (two rows with three columns) beneath the article title.
% More than six makes the first-page appear very cluttered indeed.
%
% Use the \alignauthor commands to handle the names
% and affiliations for an 'aesthetic maximum' of six authors.
% Add names, affiliations, addresses for
% the seventh etc. author(s) as the argument for the
% \additionalauthors command.
% These 'additional authors' will be output/set for you
% without further effort on your part as the last section in
% the body of your article BEFORE References or any Appendices.

\numberofauthors{8} %  in this sample file, there are a *total*
% of EIGHT authors. SIX appear on the 'first-page' (for formatting
% reasons) and the remaining two appear in the \additionalauthors section.
%
\author{
% You can go ahead and credit any number of authors here,
% e.g. one 'row of three' or two rows (consisting of one row of three
% and a second row of one, two or three).
%
% The command \alignauthor (no curly braces needed) should
% precede each author name, affiliation/snail-mail address and
% e-mail address. Additionally, tag each line of
% affiliation/address with \affaddr, and tag the
% e-mail address with \email.
%
% 1st. author
\alignauthor
Ben Trovato\titlenote{Dr.~Trovato insisted his name be first.}\\
       \affaddr{Institute for Clarity in Documentation}\\
       \affaddr{1932 Wallamaloo Lane}\\
       \affaddr{Wallamaloo, New Zealand}\\
       \email{trovato@corporation.com}
% 2nd. author
\alignauthor
G.K.M. Tobin\titlenote{The secretary disavows
any knowledge of this author's actions.}\\
       \affaddr{Institute for Clarity in Documentation}\\
       \affaddr{P.O. Box 1212}\\
       \affaddr{Dublin, Ohio 43017-6221}\\
       \email{webmaster@marysville-ohio.com}
% 3rd. author
\alignauthor Lars Th{\o}rv{\"a}ld\titlenote{This author is the
one who did all the really hard work.}\\
       \affaddr{The Th{\o}rv{\"a}ld Group}\\
       \affaddr{1 Th{\o}rv{\"a}ld Circle}\\
       \affaddr{Hekla, Iceland}\\
       \email{larst@affiliation.org}
\and  % use '\and' if you need 'another row' of author names
% 4th. author
\alignauthor Lawrence P. Leipuner\\
       \affaddr{Brookhaven Laboratories}\\
       \affaddr{Brookhaven National Lab}\\
       \affaddr{P.O. Box 5000}\\
       \email{lleipuner@researchlabs.org}
% 5th. author
\alignauthor Sean Fogarty\\
       \affaddr{NASA Ames Research Center}\\
       \affaddr{Moffett Field}\\
       \affaddr{California 94035}\\
       \email{fogartys@amesres.org}
% 6th. author
\alignauthor Charles Palmer\\
       \affaddr{Palmer Research Laboratories}\\
       \affaddr{8600 Datapoint Drive}\\
       \affaddr{San Antonio, Texas 78229}\\
       \email{cpalmer@prl.com}
}
% There's nothing stopping you putting the seventh, eighth, etc.
% author on the opening page (as the 'third row') but we ask,
% for aesthetic reasons that you place these 'additional authors'
% in the \additional authors block, viz.
\additionalauthors{Additional authors: John Smith (The Th{\o}rv{\"a}ld Group,
email: {\texttt{jsmith@affiliation.org}}) and Julius P.~Kumquat
(The Kumquat Consortium, email: {\texttt{jpkumquat@consortium.net}}).}
\date{30 July 1999}
% Just remember to make sure that the TOTAL number of authors
% is the number that will appear on the first page PLUS the
% number that will appear in the \additionalauthors section.

\maketitle
\begin{abstract}
This paper provides a sample of a \LaTeX\ document which conforms,
somewhat loosely, to the formatting guidelines for
ACM SIG Proceedings. It is an {\em alternate} style which produces
a {\em tighter-looking} paper and was designed in response to
concerns expressed, by authors, over page-budgets.
It complements the document \textit{Author's (Alternate) Guide to
Preparing ACM SIG Proceedings Using \LaTeX$2_\epsilon$\ and Bib\TeX}.
This source file has been written with the intention of being
compiled under \LaTeX$2_\epsilon$\ and BibTeX.

The developers have tried to include every imaginable sort
of ``bells and whistles", such as a subtitle, footnotes on
title, subtitle and authors, as well as in the text, and
every optional component (e.g. Acknowledgments, Additional
Authors, Appendices), not to mention examples of
equations, theorems, tables and figures.

To make best use of this sample document, run it through \LaTeX\
and BibTeX, and compare this source code with the printed
output produced by the dvi file. A compiled PDF version
is available on the web page to help you with the
`look and feel'.
\end{abstract}

% A category with the (minimum) three required fields
\category{H.4}{Information Systems Applications}{Miscellaneous}
%A category including the fourth, optional field follows...
\category{D.2.8}{Software Engineering}{Metrics}[complexity measures, performance measures]

\terms{Theory}

\keywords{ACM proceedings, \LaTeX, text tagging}

\section{Introduction}
The \textit{proceedings} are the records of a conference.
ACM seeks to give these conference by-products a uniform,
high-quality appearance.  To do this, ACM has some rigid
requirements for the format of the proceedings documents: there
is a specified format (balanced  double columns), a specified
set of fonts (Arial or Helvetica and Times Roman) in
certain specified sizes (for instance, 9 point for body copy),
a specified live area (18 $\times$ 23.5 cm [7" $\times$ 9.25"]) centered on
the page, specified size of margins (2.54cm [1"] top and
bottom and 1.9cm [.75"] left and right; specified column width
(8.45cm [3.33"]) and gutter size (.083cm [.33"]).

The good news is, with only a handful of manual
settings\footnote{Two of these, the {\texttt{\char'134 numberofauthors}}
and {\texttt{\char'134 alignauthor}} commands, you have
already used; another, {\texttt{\char'134 balancecolumns}}, will
be used in your very last run of \LaTeX\ to ensure
balanced column heights on the last page.}, the \LaTeX\ document
class file handles all of this for you.

The remainder of this document is concerned with showing, in
the context of an ``actual'' document, the \LaTeX\ commands
specifically available for denoting the structure of a
proceedings paper, rather than with giving rigorous descriptions
or explanations of such commands.

\section{The {\secit Body} of The Paper}
Typically, the body of a paper is organized
into a hierarchical structure, with numbered or unnumbered
headings for sections, subsections, sub-subsections, and even
smaller sections.  The command \texttt{{\char'134}section} that
precedes this paragraph is part of such a
hierarchy.\footnote{This is the second footnote.  It
starts a series of three footnotes that add nothing
informational, but just give an idea of how footnotes work
and look. It is a wordy one, just so you see
how a longish one plays out.} \LaTeX\ handles the numbering
and placement of these headings for you, when you use
the appropriate heading commands around the titles
of the headings.  If you want a sub-subsection or
smaller part to be unnumbered in your output, simply append an
asterisk to the command name.  Examples of both
numbered and unnumbered headings will appear throughout the
balance of this sample document.

Because the entire article is contained in
the \textbf{document} environment, you can indicate the
start of a new paragraph with a blank line in your
input file; that is why this sentence forms a separate paragraph.

\subsection{Type Changes and {\subsecit Special} Characters}
We have already seen several typeface changes in this sample.  You
can indicate italicized words or phrases in your text with
the command \texttt{{\char'134}textit}; emboldening with the
command \texttt{{\char'134}textbf}
and typewriter-style (for instance, for computer code) with
\texttt{{\char'134}texttt}.  But remember, you do not
have to indicate typestyle changes when such changes are
part of the \textit{structural} elements of your
article; for instance, the heading of this subsection will
be in a sans serif\footnote{A third footnote, here.
Let's make this a rather short one to
see how it looks.} typeface, but that is handled by the
document class file. Take care with the use
of\footnote{A fourth, and last, footnote.}
the curly braces in typeface changes; they mark
the beginning and end of
the text that is to be in the different typeface.

You can use whatever symbols, accented characters, or
non-English characters you need anywhere in your document;
you can find a complete list of what is
available in the \textit{\LaTeX\
User's Guide}\cite{Lamport:LaTeX}.

\subsection{Math Equations}
You may want to display math equations in three distinct styles:
inline, numbered or non-numbered display.  Each of
the three are discussed in the next sections.

\subsubsection{Inline (In-text) Equations}
A formula that appears in the running text is called an
inline or in-text formula.  It is produced by the
\textbf{math} environment, which can be
invoked with the usual \texttt{{\char'134}begin. . .{\char'134}end}
construction or with the short form \texttt{\$. . .\$}. You
can use any of the symbols and structures,
from $\alpha$ to $\omega$, available in
\LaTeX\cite{Lamport:LaTeX}; this section will simply show a
few examples of in-text equations in context. Notice how
this equation: \begin{math}\lim_{n\rightarrow \infty}x=0\end{math},
set here in in-line math style, looks slightly different when
set in display style.  (See next section).

\subsubsection{Display Equations}
A numbered display equation -- one set off by vertical space
from the text and centered horizontally -- is produced
by the \textbf{equation} environment. An unnumbered display
equation is produced by the \textbf{displaymath} environment.

Again, in either environment, you can use any of the symbols
and structures available in \LaTeX; this section will just
give a couple of examples of display equations in context.
First, consider the equation, shown as an inline equation above:
\begin{equation}\lim_{n\rightarrow \infty}x=0\end{equation}
Notice how it is formatted somewhat differently in
the \textbf{displaymath}
environment.  Now, we'll enter an unnumbered equation:
\begin{displaymath}\sum_{i=0}^{\infty} x + 1\end{displaymath}
and follow it with another numbered equation:
\begin{equation}\sum_{i=0}^{\infty}x_i=\int_{0}^{\pi+2} f\end{equation}
just to demonstrate \LaTeX's able handling of numbering.

\subsection{Citations}
Citations to articles \cite{bowman:reasoning,
clark:pct, braams:babel, herlihy:methodology},
conference proceedings \cite{clark:pct} or
books \cite{salas:calculus, Lamport:LaTeX} listed
in the Bibliography section of your
article will occur throughout the text of your article.
You should use BibTeX to automatically produce this bibliography;
you simply need to insert one of several citation commands with
a key of the item cited in the proper location in
the \texttt{.tex} file \cite{Lamport:LaTeX}.
The key is a short reference you invent to uniquely
identify each work; in this sample document, the key is
the first author's surname and a
word from the title.  This identifying key is included
with each item in the \texttt{.bib} file for your article.

The details of the construction of the \texttt{.bib} file
are beyond the scope of this sample document, but more
information can be found in the \textit{Author's Guide},
and exhaustive details in the \textit{\LaTeX\ User's
Guide}\cite{Lamport:LaTeX}.

This article shows only the plainest form
of the citation command, using \texttt{{\char'134}cite}.
This is what is stipulated in the SIGS style specifications.
No other citation format is endorsed or supported.

\subsection{Tables}
Because tables cannot be split across pages, the best
placement for them is typically the top of the page
nearest their initial cite.  To
ensure this proper ``floating'' placement of tables, use the
environment \textbf{table} to enclose the table's contents and
the table caption.  The contents of the table itself must go
in the \textbf{tabular} environment, to
be aligned properly in rows and columns, with the desired
horizontal and vertical rules.  Again, detailed instructions
on \textbf{tabular} material
is found in the \textit{\LaTeX\ User's Guide}.

Immediately following this sentence is the point at which
Table 1 is included in the input file; compare the
placement of the table here with the table in the printed
dvi output of this document.

\begin{table}
\centering
\caption{Frequency of Special Characters}
\begin{tabular}{|c|c|l|} \hline
Non-English or Math&Frequency&Comments\\ \hline
\O & 1 in 1,000& For Swedish names\\ \hline
$\pi$ & 1 in 5& Common in math\\ \hline
\$ & 4 in 5 & Used in business\\ \hline
$\Psi^2_1$ & 1 in 40,000& Unexplained usage\\
\hline\end{tabular}
\end{table}

To set a wider table, which takes up the whole width of
the page's live area, use the environment
\textbf{table*} to enclose the table's contents and
the table caption.  As with a single-column table, this wide
table will ``float" to a location deemed more desirable.
Immediately following this sentence is the point at which
Table 2 is included in the input file; again, it is
instructive to compare the placement of the
table here with the table in the printed dvi
output of this document.

\begin{table*}
\centering
\caption{Some Typical Commands}
\begin{tabular}{|c|c|l|} \hline
Command&A Number&Comments\\ \hline
\texttt{{\char'134}alignauthor} & 100& Author alignment\\ \hline
\texttt{{\char'134}numberofauthors}& 200& Author enumeration\\ \hline
\texttt{{\char'134}table}& 300 & For tables\\ \hline
\texttt{{\char'134}table*}& 400& For wider tables\\ \hline\end{tabular}
\end{table*}
% end the environment with {table*}, NOTE not {table}!

\subsection{Figures}
Like tables, figures cannot be split across pages; the
best placement for them
is typically the top or the bottom of the page nearest
their initial cite.  To ensure this proper ``floating'' placement
of figures, use the environment
\textbf{figure} to enclose the figure and its caption.

This sample document contains examples of \textbf{.eps}
and \textbf{.ps} files to be displayable with \LaTeX.  More
details on each of these is found in the \textit{Author's Guide}.

\begin{figure}
\centering
\epsfig{file=fly.eps}
\caption{A sample black and white graphic (.eps format).}
\end{figure}

\begin{figure}
\centering
\epsfig{file=fly.eps, height=1in, width=1in}
\caption{A sample black and white graphic (.eps format)
that has been resized with the \texttt{epsfig} command.}
\end{figure}

As was the case with tables, you may want a figure
that spans two columns.  To do this, and still to
ensure proper ``floating'' placement of tables, use the environment
\textbf{figure*} to enclose the figure and its caption.
and don't forget to end the environment with
{figure*}, not {figure}!

\begin{figure*}
\centering
\epsfig{file=flies.eps}
\caption{A sample black and white graphic (.eps format)
that needs to span two columns of text.}
\end{figure*}

Note that either {\textbf{.ps}} or {\textbf{.eps}} formats are
used; use
the \texttt{{\char'134}epsfig} or \texttt{{\char'134}psfig}
commands as appropriate for the different file types.

\begin{figure}
\centering
\psfig{file=rosette.ps, height=1in, width=1in,}
\caption{A sample black and white graphic (.ps format) that has
been resized with the \texttt{psfig} command.}
\vskip -6pt
\end{figure}

\subsection{Theorem-like Constructs}
Other common constructs that may occur in your article are
the forms for logical constructs like theorems, axioms,
corollaries and proofs.  There are
two forms, one produced by the
command \texttt{{\char'134}newtheorem} and the
other by the command \texttt{{\char'134}newdef}; perhaps
the clearest and easiest way to distinguish them is
to compare the two in the output of this sample document:

This uses the \textbf{theorem} environment, created by
the\linebreak\texttt{{\char'134}newtheorem} command:
\newtheorem{theorem}{Theorem}
\begin{theorem}
Let $f$ be continuous on $[a,b]$.  If $G$ is
an antiderivative for $f$ on $[a,b]$, then
\begin{displaymath}\int^b_af(t)dt = G(b) - G(a).\end{displaymath}
\end{theorem}

The other uses the \textbf{definition} environment, created
by the \texttt{{\char'134}newdef} command:
\newdef{definition}{Definition}
\begin{definition}
If $z$ is irrational, then by $e^z$ we mean the
unique number which has
logarithm $z$: \begin{displaymath}{\log e^z = z}\end{displaymath}
\end{definition}

Two lists of constructs that use one of these
forms is given in the
\textit{Author's  Guidelines}.

There is one other similar construct environment, which is
already set up
for you; i.e. you must \textit{not} use
a \texttt{{\char'134}newdef} command to
create it: the \textbf{proof} environment.  Here
is a example of its use:
\begin{proof}
Suppose on the contrary there exists a real number $L$ such that
\begin{displaymath}
\lim_{x\rightarrow\infty} \frac{f(x)}{g(x)} = L.
\end{displaymath}
Then
\begin{displaymath}
l=\lim_{x\rightarrow c} f(x)
= \lim_{x\rightarrow c}
\left[ g{x} \cdot \frac{f(x)}{g(x)} \right ]
= \lim_{x\rightarrow c} g(x) \cdot \lim_{x\rightarrow c}
\frac{f(x)}{g(x)} = 0\cdot L = 0,
\end{displaymath}
which contradicts our assumption that $l\neq 0$.
\end{proof}

Complete rules about using these environments and using the
two different creation commands are in the
\textit{Author's Guide}; please consult it for more
detailed instructions.  If you need to use another construct,
not listed therein, which you want to have the same
formatting as the Theorem
or the Definition\cite{salas:calculus} shown above,
use the \texttt{{\char'134}newtheorem} or the
\texttt{{\char'134}newdef} command,
respectively, to create it.

\subsection*{A {\secit Caveat} for the \TeX\ Expert}
Because you have just been given permission to
use the \texttt{{\char'134}newdef} command to create a
new form, you might think you can
use \TeX's \texttt{{\char'134}def} to create a
new command: \textit{Please refrain from doing this!}
Remember that your \LaTeX\ source code is primarily intended
to create camera-ready copy, but may be converted
to other forms -- e.g. HTML. If you inadvertently omit
some or all of the \texttt{{\char'134}def}s recompilation will
be, to say the least, problematic.

\section{Conclusions}
This paragraph will end the body of this sample document.
Remember that you might still have Acknowledgments or
Appendices; brief samples of these
follow.  There is still the Bibliography to deal with; and
we will make a disclaimer about that here: with the exception
of the reference to the \LaTeX\ book, the citations in
this paper are to articles which have nothing to
do with the present subject and are used as
examples only.
%\end{document}  % This is where a 'short' article might terminate

%ACKNOWLEDGMENTS are optional
\section{Acknowledgments}
This section is optional; it is a location for you
to acknowledge grants, funding, editing assistance and
what have you.  In the present case, for example, the
authors would like to thank Gerald Murray of ACM for
his help in codifying this \textit{Author's Guide}
and the \textbf{.cls} and \textbf{.tex} files that it describes.

%
% The following two commands are all you need in the
% initial runs of your .tex file to
% produce the bibliography for the citations in your paper.
\bibliographystyle{abbrv}
\bibliography{sigproc}  % sigproc.bib is the name of the Bibliography in this case
% You must have a proper ".bib" file
%  and remember to run:
% latex bibtex latex latex
% to resolve all references
%
% ACM needs 'a single self-contained file'!
%
%APPENDICES are optional
%\balancecolumns
\appendix
%Appendix A
\section{Headings in Appendices}
The rules about hierarchical headings discussed above for
the body of the article are different in the appendices.
In the \textbf{appendix} environment, the command
\textbf{section} is used to
indicate the start of each Appendix, with alphabetic order
designation (i.e. the first is A, the second B, etc.) and
a title (if you include one).  So, if you need
hierarchical structure
\textit{within} an Appendix, start with \textbf{subsection} as the
highest level. Here is an outline of the body of this
document in Appendix-appropriate form:
\subsection{Introduction}
\subsection{The Body of the Paper}
\subsubsection{Type Changes and  Special Characters}
\subsubsection{Math Equations}
\paragraph{Inline (In-text) Equations}
\paragraph{Display Equations}
\subsubsection{Citations}
\subsubsection{Tables}
\subsubsection{Figures}
\subsubsection{Theorem-like Constructs}
\subsubsection*{A Caveat for the \TeX\ Expert}
\subsection{Conclusions}
\subsection{Acknowledgments}
\subsection{Additional Authors}
This section is inserted by \LaTeX; you do not insert it.
You just add the names and information in the
\texttt{{\char'134}additionalauthors} command at the start
of the document.
\subsection{References}
Generated by bibtex from your ~.bib file.  Run latex,
then bibtex, then latex twice (to resolve references)
to create the ~.bbl file.  Insert that ~.bbl file into
the .tex source file and comment out
the command \texttt{{\char'134}thebibliography}.
% This next section command marks the start of
% Appendix B, and does not continue the present hierarchy
\section{More Help for the Hardy}
The sig-alternate.cls file itself is chock-full of succinct
and helpful comments.  If you consider yourself a moderately
experienced to expert user of \LaTeX, you may find reading
it useful but please remember not to change it.
%\balancecolumns % GM June 2007
% That's all folks!
\end{document}

### 2011-04-05 paper_gecco_2/abstract.tex
Added directory for two page version of GECCO paper

--- 2011-02-10 paper/abstract.tex
@@
 evolved with HyperNEAT perform considerably better than all
-parameterized local search methods, and produce gaits nearly 9 times faster than a hand-designed gait.  The best HyperNEAT gaits exhibit
+parameterized local search methods and produce gaits nearly 9 times faster than a hand-designed gait.  The best HyperNEAT gaits exhibit
 complex motion patterns that contain multiple frequencies, yet are

### 2011-04-06 paper_ecal/main.tex
Converted paper to ECAL format, still need to fix section references

%
% Paper for ECAL conference, formatted per alifexi.sty
% JBY 4/6/11
%

\documentclass[letterpaper]{article}
\usepackage{natbib,alifexi}   %% The order is important
\usepackage{graphicx}
\usepackage{fancyhdr}
\usepackage{color}
%\usepackage{ulem}
\usepackage{ownstyles}
%\usepackage{hyperref}

\graphicspath{{../figures/}}

\title{Generating gaits for physical quadruped robots: evolved neural networks vs. local parameterized search}
\author {Jason Yosinski$^{1}$,
Diana Hidalgo$^{1}$,
Sarah Nguyen$^{1}$,
Jeff Clune$^{1}$,
Juan Cristobal Zagal$^{2}$,
\and Hod Lipson$^{1}$\\
\mbox{} \\
$^1$ Cornell University, 239 Upson Hall, Ithaca, NY  14853, USA\\
$^2$ University of Chile, Beauchef 850, Santiago 8370448, Chile\\
jy495@cornell.edu}

\begin{document}

\maketitle

\begin{abstract}
\input{abstract}
\end{abstract}

%%%%%%%%%%%%%%%%%%%%%%

% Section 1
\input{introduction}
\input{relatedWork}
\input{paperOutline}

% Section 2
\input{problemDefinition}

% Section 3
\section{Experimental Setup}
\seclabel{experimentalSetup}

An important aspect of this study was to evaluate gait generation and
learning algorithms in the noisy, unforgiving real world, instead of
using a simulator.  In this section we give details of the hardware
setup we used: \secref{platformDetails} discusses the robotic
platform, and \secref{fitnessEvaluation} describes the tracking system
we used to measure each gait's fitness.

\input{platformDetails}
\input{fitnessEvaluation}

% Section 4
\section{Gait Generation and Learning}
\seclabel{gaitGenLearn}
\input{learningMethods}
\input{hyperNeatMethod}

% Section 5
\input{results}

% Section 6
\input{conclusion}
\input{futureWork}

% Section 7
\input{acknowledgments}

%%%%%%%%%%%%%%%%%%%%%%

\footnotesize
\bibliography{references.bib}
\bibliographystyle{apalike}

\end{document}

### 2011-04-12 paper_ecal/introduction.tex
Added another figure from Juan to replace figure 1 in intro

--- 2011-02-10 paper/introduction.tex
@@
 
-\acmFig{robot_close.jpg}{.8}{The quadruped robot. The translucent parts
-  were created with 3D-printing technology. Videos of the gaits
+\acmFig{robot_whitebg.jpg}{.85}{The quadruped robot. The translucent
+  parts were created with 3D-printing technology. Videos of the gaits
   evolved for this robot can be viewed at www.anonymous.url.}
@@
 by one of several methods: either manually by an expert or
-automatically through repeated trial and evaluation.  Learned gaits
+via computer learning algorithms.  Learned gaits
 offer several advantages over manually designed gaits.  Automatically
@@
 investigate how they perform when evolving on a physical robot
-(\figref{robot_close.jpg}).
+(\figref{robot_whitebg.jpg}).
 

### 2011-04-12 paper_ecal/platformDetails.tex
Added Juan's better topdown figure
Added another figure from Juan to replace figure 1 in intro

--- 2011-02-09 paper/platformDetails.tex
@@
 
-\acmFig{topdown.png}{1}{A figure of the robot from a top-down perspective,
-with servos labeled}
+\acmFig{topdown2.jpg}{1}{A figure of the robot from a top-down perspective,
+with the nine joints and associated servos labeled.}
 
@@
 centimeters from leg to opposite leg in the crouch position depicted
-in \figref{robot_close.jpg}. The robot is actuated by 9 AX-12+
+in \figref{robot_whitebg.jpg}. The robot is actuated by 9 AX-12+
 Dynamixel servos: one inner joint and one outer joint servo in each of

### 2011-04-12 paper_ecal/platformDetails.tex
Reworded a few sections

@@
 The quadruped robot used in this study was assembled from parts
-printed on the Objet Connex 500 3-D Printing System. It weights 1.88
+printed on the Objet Connex 500 3-D Printing System. It weighs 1.88
 kg with the on-board computer and measures approximately 38
@@
 All of the computation for gait learning, fitness evaluation, and
-robot control was performed on the compact on-board CompuLab Fit-PC2,
+robot control was performed on the compact, on-board CompuLab Fit-PC2,
 running Ubuntu Linux.  All gait generation, learning, and fitness

### 2011-04-12 paper_ecal/experimentalSetup.tex
More edits

--- 2011-02-09 paper/experimentalSetup.tex
@@
 An important aspect of this study was to evaluate gait generation and
-learning algorithms in the noisy, unforgiving real world, instead of
+learning algorithms in the noisy, unforgiving real world instead of
 using a simulator.  In this section we give details of the hardware

### 2011-04-12 paper_ecal/hyperNeatMethod.tex
More edits

--- 2011-02-10 paper/hyperNeatMethod.tex
@@
 could be $(3 \times 4)^2= 288$ links in each ANN (although evolution can set weights to 0, functionally eliminating the connection). The inputs to the
-substrate were the current angles of each of the 9 joints of the robot and a sine and cosine wave (to facilitate the
+substrate were the current \emph{commanded} angles of each of the 9 joints of the robot and a sine and cosine wave (to facilitate the
 production of periodic behaviors).  The sine and cosine waves had a period of about half a second.
+
+While measured angles could have been used as input instead of
+commanded angles, this may have given HyperNEAT gaits an unfair
+advantage by making them closed-loop. Thus, we restricted inputs to
+the commanded angles only, which allowed us to produce completely
+deterministic gaits using HyperNEAT.
 

### 2011-04-12 paper_ecal/introduction.tex
More edits

@@
 investigate how they perform when evolving on a physical robot
-(\figref{robot_whitebg.jpg}).
+(\figref{robot_whitebg.jpg}) on flat terrain.
 

### 2011-04-12 paper_ecal/platformDetails.tex
More edits

--- 2011-02-09 paper/platformDetails.tex
@@
 
-\acmFig{topdown.png}{1}{A figure of the robot from a top-down perspective,
-with servos labeled}
+\acmFig{topdown2.jpg}{1}{(a) Top-down perspective of the robot with
+  the nine joints and associated servos labeled. (b) The robot in a
+  flat pose with the hip joint centered. (c,d,e) Various views of a
+  pose in which the hip joint is rotated.}
 
 The quadruped robot used in this study was assembled from parts
-printed on the Objet Connex 500 3-D Printing System. It weights 1.88
-kg with the on-board computer and measures approximately 38
-centimeters from leg to opposite leg in the crouch position depicted
-in \figref{robot_close.jpg}. The robot is actuated by 9 AX-12+
+printed on the Objet Connex 500 3-D Printing System. It weighs 1.88 kg
+with the on-board computer and measures approximately 38 centimeters
+from leg to opposite leg in the crouch position depicted in
+\figref{robot_whitebg.jpg}. The robot is actuated by 9 AX-12+
 Dynamixel servos: one inner joint and one outer joint servo in each of
@@
 unique servo allows the two halves of the robot to rotate with respect
-to each other. \figref{topdown.png} shows the positions and numerical
-designations of all nine servos.  Each servo could be commanded to a
-position in the range [0,~1023], corresponding to a physical range
-[-120$^{\circ}$,~+120$^{\circ}$].
+to each other. \figref{topdown2.jpg} shows this unique motion, as well
+as the positions and numerical designations of all nine servos.  Each
+servo could be commanded to a position in the range [0,~1023],
+corresponding to a physical range [-120$^{\circ}$,~+120$^{\circ}$].
 
 All of the computation for gait learning, fitness evaluation, and
-robot control was performed on the compact on-board CompuLab Fit-PC2,
-running Ubuntu Linux.  All gait generation, learning, and fitness
+robot control was performed on the compact, on-board CompuLab Fit-PC2,
+running Ubuntu Linux 10.10. 
+The most computationally expensive portion of code was HyperNEAT, but even
+this took less than one second to run on the on-board computer. Thus,
+we chose not to offload any computation.
+All gait generation, learning, and fitness
 evaluation code, except HyperNEAT, was written in Python and is
@@
 C++.  We controlled the servos with the \code{pydynamixel} library
-\cite{pydynamixel}.  The robot connected to a wireless network on
-boot, which enabled us to control it via SSH.
+\cite{pydynamixel}, sending commanded positions at 40Hz.  The robot
+connected to a wireless network on boot, which enabled us to control
+it via SSH.
 

### 2011-04-12 paper_ecal/hyperNeatMethod.tex
changes to hyperneat section

@@
 gaits with HyperNEAT in simulation~\cite{clune2011performance,
-  clune2009evolving}, but was adapted to accommodate the physical robot used in this paper. Specifically, the ANNs consists of three $3 \times 4$
+  clune2009evolving}, but was adapted to accommodate the physical robot used in this paper.
+Specifically, the ANN has a fixed topology (i.e.
+the number of nodes does not evolve) that consists of three $3 \times 4$
 Cartesian grids of nodes forming input, hidden, and output
@@
 obtain the actual gait $g(t)$.  This solution worked well and did not restrict the expressiveness of HyperNEAT.
-
 %\editbox{are we mentioning that we gave HyperNEAT the center joint
@@
 As with the parameterized methods, three runs of HyperNEAT
-were performed. The population size for HyperNEAT was 9 and runs
-lasted 20 generations. These numbers are small, but were necessarily
+were performed. Runs lasted 20 generations and the population size was 9 (which allowed a bare minimum of diversity within and between NEAT species).  These numbers are small, but were necessarily
 constrained given how much time it took to conduct evolution directly
 on a real robot. The remaining parameters were identical to Clune et
-al.~\cite{clune2011performance}.
+al.~(2011).
 

### 2011-04-13 paper_ecal/conclusion.tex
Fixed refs

--- 2011-02-10 paper/conclusion.tex
@@
 Its best-performing gait traveled 45.7 body lengths per minute, which is nearly 9 times the speed of the hand-coded gait.
-This could be because HyperNEAT tends to generate coordinated gaits~\cite{clune2011performance, clune2009evolving}, allowing it to
+This could be because HyperNEAT tends to generate coordinated gaits~\citep{clune2011performance, clune2009evolving}, allowing it to
 take advantage of the symmetries of the problem, but can also explore a much larger space of possibilities than the

### 2011-04-13 paper_ecal/fitnessEvaluation.tex
Moved GECCO paper to its own directory, copied to new directory for ECAL paper
Added another figure from Juan to replace figure 1 in intro
Fixed refs

--- 2011-02-10 paper/fitnessEvaluation.tex
@@
 top of the robot (\figref{wiiMote_crop.jpg}).  The Wii remote contains
-an IR camera that can track and report the position of up to four IR
-sources in its image frame.  The resolution of the camera was 1024 by
+an IR camera that tracks and reports the position of IR sources.  The resolution of the camera was 1024 by
 768 pixels with view angles of about 40$^\circ$ by 30$^\circ$, which
@@
 interfaced with the Wii remote via bluetooth using the CWiid
-library\cite{cwiid}.  Our fitness testing code communicated with this
+library\citep{cwiid}.  Our fitness testing code communicated with this
 server via a socket connection and requested position updates at the
@@
 while in the \emph{ready} position (the position shown in
-\figref{robot_close.jpg}).  This was done to prevent assigning extra
+\figref{robot_whitebg.jpg}).  This was done to prevent assigning extra
 fitness to, for example, gaits that ended with the robot leaning toward

### 2011-04-13 paper_ecal/futureWork.tex
Fixed refs

--- 2011-02-10 paper/futureWork.tex
@@
 One direction for future work could be to obtain many more trials. However, a more effective extension might be to combine frequent trials in
-simulation with infrequent trials in hardware~\cite{bongard}.  The simulation would produce the
+simulation with infrequent trials in hardware~\citep{bongard}.  The simulation would produce the
 necessary volume of trials to allow the learning methods to be
@@
 fertile territory by penalizing gaits that produced large
-discrepancies between simulation and reality~\cite{koos2010crossing}.
+discrepancies between simulation and reality~\citep{koos2010crossing}.
 Another natural extension would be to allow gaits that sensed the position of the robot and other variables to enable the robot to adjust to its physical state, instead of providing an open-loop sequence of motor commands.

### 2011-04-13 paper_ecal/hyperNeatMethod.tex
Fixed refs

@@
 networks (ANNs) that is inspired by the way natural organisms
-develop~\cite{stanley2009hypercube}. It evolves Compositional Pattern
-Producing Networks (CPPNs)~\cite{stanley2007compositional}, each of
+develop~\citep{stanley2009hypercube}. It evolves Compositional Pattern
+Producing Networks (CPPNs)~\citep{stanley2007compositional}, each of
 which is a genome that encodes an ANN
-phenotype~\cite{stanley2009hypercube}. Each CPPN is itself a directed
+phenotype~\citep{stanley2009hypercube}. Each CPPN is itself a directed
 graph, where the nodes in the graph are mathematical functions, such as sine or
@@
 of properties such as symmetry (e.g.\ a Gaussian function) and repetition (e.g.\ a sine
-function)~\cite{stanley2009hypercube, stanley2007compositional}. The
+function)~\citep{stanley2009hypercube, stanley2007compositional}. The
 signal on each link in the CPPN is multiplied by that link's weight,
@@
 the link values between nodes in the final ANN phenotype are a
-function of the geometric positions of those nodes~\cite{stanley2009hypercube, clune2009sensitivity,
+function of the geometric positions of those nodes~\citep{stanley2009hypercube, clune2009sensitivity,
   clune2011performance}. In the case of quadruped locomotion, this
 property has previously been shown to help HyperNEAT produce gaits with front-back, left-right,
-and four-way symmetries~\cite{clune2009evolving,
+and four-way symmetries~\citep{clune2009evolving,
   clune2011performance}.
@@
 principles of the NeuroEvolution of Augmenting Topologies (NEAT)
-algorithm~\cite{stanley2002evolving}, which was originally designed to
+algorithm~\citep{stanley2002evolving}, which was originally designed to
 evolve ANNs. NEAT can be fruitfully applied to CPPNs because of their
@@
 NEAT algorithm is unique in three main
-ways~\cite{stanley2002evolving}. Initially, it starts with small
+ways~\citep{stanley2002evolving}. Initially, it starts with small
 genomes that encode simple networks and slowly complexifies them via
@@
 topological analysis. A full explanation of NEAT can be found
-in~\cite{stanley2002evolving}.
+in~\citep{stanley2002evolving}.
 
 The ANN configuration follows previous studies that evolved quadruped
-gaits with HyperNEAT in simulation~\cite{clune2011performance,
+gaits with HyperNEAT in simulation~\citep{clune2011performance,
   clune2009evolving}, but was adapted to accommodate the physical robot used in this paper.

### 2011-04-13 paper_ecal/introduction.tex
Fixed refs

--- 2011-04-12 paper_ecal/introduction.tex
@@
 %with its own relative importance.  The different combinations of desired traits and the relative weight placed on each can produce drastically different gaits.    For example, Honda's Asimo
-%\cite{chestnutt2006footstep} and Boston Dynamic's Big Dog \cite{raibert2008bigdog} both require
+%\citep{chestnutt2006footstep} and Boston Dynamic's Big Dog \citep{raibert2008bigdog} both require
 %gaits that are relatively quick, power efficient, and robust to
@@
 Most importantly, in many cases learned gaits can outperform manually
-designed gaits~\cite{hornby2005autonomous, valsalam2008modular}.
+designed gaits~\citep{hornby2005autonomous, valsalam2008modular}.
 
@@
 learning methods, and gaits generated by evolving neural networks with
-the HyperNEAT generative encoding~\cite{stanley2009hypercube}. While
+the HyperNEAT generative encoding~\citep{stanley2009hypercube}. While
 some of these methods, such as HyperNEAT, have been tested in
-simulation~\cite{clune2009evolving, clune2011performance}, we
+simulation~\citep{clune2009evolving, clune2011performance}, we
 investigate how they perform when evolving on a physical robot
-(\figref{robot_whitebg.jpg}).
+(\figref{robot_whitebg.jpg}) on flat terrain.
 
@@
 are \emph{regular} (i.e.\ when the legs are
-coordinated)~\cite{clune2009evolving,
+coordinated)~\citep{clune2009evolving,
   clune2011performance,valsalam2008modular}. For example, HyperNEAT
@@
 gaits created coordinated movements that outperformed gaits evolved by
-an encoding not biased towards regularity~\cite{clune2009evolving,
+an encoding not biased towards regularity~\citep{clune2009evolving,
   clune2011performance}. One of the motivations of this paper is to
@@
 % Approaches differ in their starting assumptions, some essentially 
-% tweaking the parameters of a hand-tuned model \cite{chernova}, others 
-% exploring a reasonably compact parameter space \cite{kohl}, and still 
+% tweaking the parameters of a hand-tuned model \citep{chernova}, others 
+% exploring a reasonably compact parameter space \citep{kohl}, and still 
 % others beginning with few assumptions besides periodicity 
-% \cite{zykov}. 
+% \citep{zykov}. 
 %  

### 2011-04-13 paper_ecal/learningMethods.tex
Added another figure from Juan to replace figure 1 in intro
More edits
Fixed refs

--- 2011-02-10 paper/learningMethods.tex
@@
 waves are at 0) is approximately a crouch (the position shown in
-\figref{robot_close.jpg}).  In this motion model, the ninth
+\figref{robot_whitebg.jpg}).  In this motion model, the ninth
 (center) motor is kept at a neutral position.  Thus, the commanded
@@
 correspond to the nine motors listed in
-\figref{topdown.png}.
+\figref{topdown2.jpg}.
 
@@
 %We implemented and tested 8 different learning strategies.  All
-%strategies except for the HyperNEAT method\cite{clune} were
+%strategies except for the HyperNEAT method\citep{clune} were
 %constrained to pick parameters from within predetermined ranges.
@@
 \emph{N-dimensional policy gradient ascent}: We implemented Kohl
-  and Stone's \cite{kohl} method for local gradient ascent for gait
+  and Stone's \citep{kohl} method for local gradient ascent for gait
   learning with noisy fitness evaluations. This strategy explicitly
@@
 %\item \emph{Nelder-Mead simplex method}: The Nelder-Mead simplex
-%  method \cite{nm} creates an initial simplex with $d+1$ vertices,
+%  method \citep{nm} creates an initial simplex with $d+1$ vertices,
 %  where $d$ is the dimension of the parameter space. The initial
@@
 %  point replaces the worst point. If the contracted point is worse
-%  than the reflected point, the entire simplex is shrunk \cite{nm}.
+%  than the reflected point, the entire simplex is shrunk \citep{nm}.
 
 \emph{Nelder-Mead simplex method}: The Nelder-Mead simplex
-  method \cite{nm} creates an initial simplex with $d+1$ vertices,
+  method \citep{nm} creates an initial simplex with $d+1$ vertices,
   where $d$ is the dimension of the parameter space. It then tests the
@@
   over the centroid in an attempt to improve it.  Several additional
-  rules are used to prevent cycles and local minima \cite{nm}.
+  rules are used to prevent cycles and local minima \citep{nm}.
 

### 2011-04-13 paper_ecal/platformDetails.tex
Fixed refs

@@
 evaluation code, except HyperNEAT, was written in Python and is
-available on our website \cite{quadratotWeb}.  HyperNEAT is written in
+available on our website \citep{quadratotWeb}.  HyperNEAT is written in
 C++.  We controlled the servos with the \code{pydynamixel} library
-\cite{pydynamixel}, sending commanded positions at 40Hz.  The robot
+\citep{pydynamixel}, sending commanded positions at 40Hz.  The robot
 connected to a wireless network on boot, which enabled us to control

### 2011-04-13 paper_ecal/relatedWork.tex
Fixed refs

@@
 gradient reinforcement learning approach for generating a fast walk on
-legged robots\cite{kohl}, which we implemented for comparison. Others
+legged robots\citep{kohl}, which we implemented for comparison. Others
 have evolved gaits for legged robots, producing competitive
-results~\cite{chernova2005evolutionary, hornby2005autonomous, zykov,
+results~\citep{chernova2005evolutionary, hornby2005autonomous, zykov,
   clune2009evolving, clune2011performance, clune2009hybrid,
@@
 fact, an evolved gait was used in the first commercially-available
-version of Sony's AIBO robot~\cite{hornby2005autonomous}. Except for
-work with HyperNEAT~\cite{clune2009evolving, clune2011performance,
+version of Sony's AIBO robot~\citep{hornby2005autonomous}. Except for
+work with HyperNEAT~\citep{clune2009evolving, clune2011performance,
   clune2009hybrid, clune2009sensitivity}, the previous evolutionary
@@
 is to see if the space of regularities can be explored automatically,
-which has previously performed well~\cite{valsalam2008modular}, and to
+which has previously performed well~\citep{valsalam2008modular}, and to
 perform a direct comparison on the same robot to HyperNEAT.

### 2011-04-13 paper_ecal/experimentalSetup.tex
More edits, 75% done :)

@@
 using a simulator.  In this section we give details of the hardware
-setup we used: \secref{platformDetails} discusses the robotic
-platform, and \secref{fitnessEvaluation} describes the tracking system
-we used to measure each gait's fitness.
+setup we used.  First we discuss the robotic platform on which gaits
+were run, and then we describe the tracking system used to quantify
+the fitness of those gaits.
+
+%: \secref{platformDetails} discusses the robotic
+%platform, and \secref{fitnessEvaluation} describes the tracking system
+%we used to measure each gait's fitness.
 
@@
 
-\input{platformDetails}
+%\input{platformDetails}
 
-\input{fitnessEvaluation}
+%\input{fitnessEvaluation}

### 2011-04-13 paper_ecal/hyperNeatMethod.tex
More edits, 75% done :)

@@
   weight. The CPPN has two output values, which specify the weights
-  for each connection layer as shown.}
+  for each connection layer as shown.  Figure from
+  \cite{clune2011performance}.}
 
@@
   joint angles for each respective joint. The unlabeled nodes in the
-  input and output layers are ignored.}\label{hyperneatLayout}
+  input and output layers are ignored. Figure adapted from
+  \cite{clune2011performance}.}\label{hyperneatLayout}
 

### 2011-04-13 paper_ecal/learningMethods.tex
More edits, 75% done :)

@@
+We have chosen to implement two classes of algorithms to produce
+the motion-generating function $g(t)$ (described earlier).  The
+first method uses a family of parameterized gaits, described in this
+section and the next. The second method, HyperNEAT, evolves a non-parametric gait
+controlled by neural networks and is described last.
+
 \subsection{Parameterized Gaits}
 \seclabel{motionModel}
-
-We have chosen to implement two classes of algorithms to automatically create the motion-generating function $g(t)$ (described in
-\secref{problemDefinition}).  The first method uses a family of
-parameterized gaits, described in this section and the next. The
-second method evolves a non-parametric gait controlled by neural networks, and is detailed in
-\secref{hyperNeatMethod}.
 
@@
 choose the combination of five parameters that results in the fastest
-motion, per the evaluation methods in \secref{fitnessEvaluation}.
+motion, per the chosen evaluation methods described earlier.
 

### 2011-04-13 paper_ecal/main.tex
More edits, 75% done :)

@@
 % Section 3
-\section{Experimental Setup}
-\seclabel{experimentalSetup}
-
-An important aspect of this study was to evaluate gait generation and
-learning algorithms in the noisy, unforgiving real world instead of
-using a simulator.  In this section we give details of the hardware
-setup we used: \secref{platformDetails} discusses the robotic
-platform, and \secref{fitnessEvaluation} describes the tracking system
-we used to measure each gait's fitness.
-
+\input{experimentalSetup}
 \input{platformDetails}

### 2011-04-13 paper_ecal/problemDefinition.tex
Moved GECCO paper to its own directory, copied to new directory for ECAL paper
Reworded a few sections
More edits, 75% done :)

--- 2011-02-09 paper/problemDefinition.tex
@@
-\section{Problem definition}
+\section{Problem Definition}
 \seclabel{problemDefinition}
@@
 %We are testing several different learning methods to design a
-%parametrized gait for a quadruped robot from the Cornell Computational
+%parameterized gait for a quadruped robot from the Cornell Computational
 %Synthesis Lab.
@@
 %$g(t)$, that outputs a vector of commanded motor positions.  This
-%function is generated using a parametrized motion model, described in
+%function is generated using a parameterized motion model, described in
 %\secref{implement}.
@@
 
-For the system evaluated in this paper, we chose to use only
-feedforward gaits.  An interesting extension would be to allow gaits
-to depend on the measured servo positions, loads, voltage drops, or
-other quantities.
+For the system evaluated in this paper, we chose to compare only
+feedforward gaits generated by both the parameterized methods and
+HyperNEAT.  An interesting extension would be to allow gaits to depend
+on the measured servo positions, loads, voltage drops, or other
+quantities.
 
@@
 so to optimize for speed, our performance metric was displacement over
-a period of motion lasting approximately 12 seconds.  For details of
-exactly how this displacement was measured, see
-\secref{fitnessEvaluation}.
+a period of motion lasting approximately 12 seconds. Details of
+exactly how this displacement was measured are given below.

### 2011-04-13 paper_ecal/results.tex
More edits, 75% done :)

--- 2011-02-10 paper/results.tex
@@
 
-\subsection{Exploration of Parameterized Gait Space}
-
-Before optimizing the chosen family of parameterized gaits
-(\secref{motionModel}) with learning methods, we performed an
-experiment to explore the five dimensions of the SineModel5 parameter
-space. Specifically, we selected a random parameter vector that
-resulted in some motion, but not an exceptional gait. We then varied
-each of the five parameters individually and measured performance,
-repeating each measurement twice to get a rough estimate of the
-measurement noise at each point.  The results of this exploration,
-shown in \figref{explore_dim_1}, reveal that some dimensions ($\amp$,
-$\tau$, $m_F$) are fairly smooth and exhibit global structure across
-the allowable parameter range, while others ($m_O$, $m_R$) exhibit
-more complex behavior.  In addition, it gives a rough indication that
-measurement noise is often significant and is more likely to be large
-for gaits that move more.  Of course, this is only a slice in each
-dimension through a single point, and slices through a different point
-could reveal different behavior.  The common point at the
-intersection of all slices is shown as a red triangle in
-each plot of \figref{explore_dim_1}.
+% \subsection{Exploration of Parameterized Gait Space}
+% 
+% Before optimizing the chosen family of parameterized gaits
+% (\secref{motionModel}) with learning methods, we performed an
+% experiment to explore the five dimensions of the SineModel5 parameter
+% space. Specifically, we selected a random parameter vector that
+% resulted in some motion, but not an exceptional gait. We then varied
+% each of the five parameters individually and measured performance,
+% repeating each measurement twice to get a rough estimate of the
+% measurement noise at each point.  The results of this exploration,
+% shown in \figref{explore_dim_1}, reveal that some dimensions ($\amp$,
+% $\tau$, $m_F$) are fairly smooth and exhibit global structure across
+% the allowable parameter range, while others ($m_O$, $m_R$) exhibit
+% more complex behavior.  In addition, it gives a rough indication that
+% measurement noise is often significant and is more likely to be large
+% for gaits that move more.  Of course, this is only a slice in each
+% dimension through a single point, and slices through a different point
+% could reveal different behavior.  The common point at the
+% intersection of all slices is shown as a red triangle in
+% each plot of \figref{explore_dim_1}.
 
@@
 
-\acmFiggggg{explore_dim_1}{explore_dim_2}{explore_dim_3}{explore_dim_4}{explore_dim_5}{1}{.53}{Fitness
-  mean and standard deviation when each parameter dimension is varied
-  independently.  The red triangle in each plot represents the same point in the 5-dimensional parameter space.}
+% \acmFiggggg{explore_dim_1}{explore_dim_2}{explore_dim_3}{explore_dim_4}{explore_dim_5}{1}{.53}{Fitness
+%   mean and standard deviation when each parameter dimension is varied
+%   independently.  The red triangle in each plot represents the same point in the 5-dimensional parameter space.}
 
@@
 \end{tabular}
-\caption{The best gaits found for each starting vector and algorithm,
-  in body lengths per minute.}
-\tablabel{results}
+\caption{The average and standard deviation of the best gaits found
+  for each algorithm during each of three runs, in body lengths per
+  minute.}  \tablabel{results}
 \end{center}

### 2011-04-13 paper_ecal/fitnessEvaluation.tex
Cleaned a few things up

@@
 A separate tracking server, written in Python, ran on the robot and
-interfaced with the Wii remote via bluetooth using the CWiid
-library\citep{cwiid}.  Our fitness testing code communicated with this
+interfaced with the Wii remote via bluetooth using the \code{CWiid}
+library \citep{cwiid}.  Our fitness testing code communicated with this
 server via a socket connection and requested position updates at the
@@
 to occasionally move the robot back into the viewable area of the Wii
-remote.  Initially this was a rare
+remote whenever it wandered outside this window.  Initially this was a rare
 occurrence, as the gaits did not typically produce motion as large as

### 2011-04-13 paper_ecal/fitnessEvaluation.tex
More edits from reviews, made figures larger

@@
 
-\acmFigg{wiiMote_crop.jpg}{robot_led_crop.jpg}{.7}{A Nintendo Wii
+\acmFigg{wiiMote_crop.jpg}{robot_led_crop.jpg}{1}{A Nintendo Wii
   remote provided the location of the robot by tracking the infrared
@@
 top of the robot (\figref{wiiMote_crop.jpg}).  The Wii remote contains
-an IR camera that tracks and reports the position of IR sources.  The resolution of the camera was 1024 by
-768 pixels with view angles of about 40$^\circ$ by 30$^\circ$, which
-produced a resolution of 1.7mm per pixel when mounted at a height of
-2.63m.  At this height, the viewable window on the floor was
-approximately 175 x 120 cm.
+an IR camera that tracks and reports the position of IR sources.  The
+resolution of the camera was 1024 by 768 pixels with view angles of
+about 40$^\circ$ by 30$^\circ$, which produced a resolution of 1.7mm
+per pixel when mounted at a height of 2.63m.  At this height, the
+viewable window on the floor was approximately 175 x 120 cm.
 
@@
 average speed, which was measured as the Euclidian distance the robot
-moved during a 12-second run.
-For the manual and parameterized gaits, the fitness was this value.
-The HyperNEAT gaits stressed the motors more than the other gaits, so to encourage gaits that did not tax the motors we penalized gaits that caused the servos to stop responding.
-The penalty was to set the fitness to half of the distance the robot actually traveled.
-We tested whether the servos were responding after each gait by commanding them to specific positions and checking whether they actually moved to those positions.
-This test had the additional benefit of rewarding those gaits that did not flip the robot into a position where it could not move its legs, which HyperNEAT also did more than the other learning methods.
-Because the fitness of HyperNEAT gaits were often halved, in results we compare actual distance traveled in addition to fitness for the best gaits produced by each class of gait-generating algorithms.
+moved during a 12-second run. For the manual and parameterized gaits,
+the fitness was this value. The HyperNEAT gaits stressed the motors
+more than the other gaits, so to encourage gaits that did not tax the
+motors we penalized gaits that caused the servos to stop
+responding. When the servos stopped responding they could, in nearly
+all cases, be restarted by cycling power, though over the course of
+this study we did have to replace four servos that were damaged.
+The penalty was to set the fitness to half of the distance
+the robot actually traveled. We tested whether the servos were
+responding after each gait by commanding them to specific positions
+and checking whether they actually moved to those positions. This test
+had the additional benefit of rewarding those gaits that did not flip
+the robot into a position where it could not move its legs, which
+HyperNEAT also did more than the other learning methods. Because the
+fitness of HyperNEAT gaits were often halved, in results we compare
+actual distance traveled in addition to fitness for the best gaits
+produced by each class of gait-generating algorithms.
 

### 2011-04-13 paper_ecal/introduction.tex
More edits, 75% done :)
More edits from reviews, made figures larger

@@
 
-\acmFig{robot_whitebg.jpg}{.85}{The quadruped robot. The translucent
+\acmFig{robot_whitebg.jpg}{.75}{The quadruped robot. The translucent
   parts were created with 3D-printing technology. Videos of the gaits
@@
 A final motivation of this problem is to simply evolve effective gaits
-for a physical robot. Because HyperNEAT performed well in simulation,
-it is interesting to test whether it can produce a fast gait for a
-physical robot. It is additionally interesting to test how more
-traditional gait optimization techniques compete with evolutionary
-algorithms when evolving in hardware.
+for a physical robot. Because HyperNEAT has performed well in
+simulation, it is interesting to test whether it can produce a fast
+gait for a physical robot. It is additionally interesting to test how
+more traditional gait optimization techniques compete with
+evolutionary algorithms when evolving in hardware.
 

### 2011-04-13 paper_ecal/relatedWork.tex
More edits from reviews, made figures larger

@@
 gradient reinforcement learning approach for generating a fast walk on
-legged robots\citep{kohl}, which we implemented for comparison. Others
+legged robots \citep{kohl}, which we implemented for comparison. Others
 have evolved gaits for legged robots, producing competitive
-results~\citep{chernova2005evolutionary, hornby2005autonomous, zykov,
+results \citep{chernova2005evolutionary, hornby2005autonomous, zykov,
   clune2009evolving, clune2011performance, clune2009hybrid,
@@
 version of Sony's AIBO robot~\citep{hornby2005autonomous}. Except for
-work with HyperNEAT~\citep{clune2009evolving, clune2011performance,
+work with HyperNEAT \citep{clune2009evolving, clune2011performance,
   clune2009hybrid, clune2009sensitivity}, the previous evolutionary
@@
 is to see if the space of regularities can be explored automatically,
-which has previously performed well~\citep{valsalam2008modular}, and to
+which has previously performed well \citep{valsalam2008modular}, and to
 perform a direct comparison on the same robot to HyperNEAT.

### 2011-04-13 paper_ecal/results.tex
More edits from reviews, made figures larger

@@
 The results for the parameterized gaits are shown in
-\figref{std_error} and \tabref{results}.  The best overall gait for
+\figref{std_error} and \tabref{results}.  A total of 1217 hardware
+fitness evaluations were performed during the learning of
+parameterized gaits, with the following distribution by learning
+method: 200 random, 234 uniform, 284 Gaussian, 174 gradient, 172
+simplex, 153 linear regression.  The number of runs varies because
+each run plateaued at its own pace. The best overall gait for
 the parameterized methods was found by linear regression, which also
@@
 \acmFig{std_error}{1}{Average results ($\pm $ SE) for each of the
-  parameterized learning methods.  Linear regression found the fastest
-  overall gait and had the highest average, followed by Nelder-Mead
-  simplex. Other methods did not outperform a random strategy.}
+  parameterized learning methods, computed over three separately
+  initialized runs.  Linear regression found the fastest overall gait
+  and had the highest average, followed by Nelder-Mead simplex. Other
+  methods did not outperform a random strategy.}
 
@@
 The results for the gaits evolved by HyperNEAT are shown in
-\figref{hnResults} and \tabref{results}.  Overall the HyperNEAT gaits
+\figref{hnResults} and \tabref{results}.  A total of 540 evaluations
+were performed for HyperNEAT, 180 in each of three runs.
+Overall the HyperNEAT gaits
 were the fastest by far, beating all the parameterized models when
@@
 
-The evaluation of the gaits produced by HyperNEAT was more noisy than for the parameterized gaits, which made learning difficult.
-For example, we tested an example HyperNEAT generation-champion gait 11 times and found that its mean performance was 26 body lengths per minute ($\pm$ 13 SD), but it had a max of 38 and a min of 3.
-Many effective HyperNEAT gaits were not preserved across generations because if performance in one trial was poor, the genome was unlikely to be selected for.
-We believe the slope of the HyperNEAT learning curve would be steeper if the noise in the evaluations could be reduced.
+The evaluation of the gaits produced by HyperNEAT was more noisy than
+for the parameterized gaits, which made learning difficult. For
+example, we tested an example HyperNEAT generation-champion gait 11
+times and found that its mean performance was 26 body lengths per
+minute ($\pm$ 13 SD), but it had a max of 38 and a min of 3.  Many
+effective HyperNEAT gaits were not preserved across generations
+because if performance in one trial was poor, the genome was unlikely
+to be selected for. The HyperNEAT learning curve would be smoother if
+the noise in the evaluations could be reduced or more than one
+evaluation per individual could be afforded.
 
-\acmFig{hnResults}{.8}{Average fitness ($\pm $ SE) of the highest
+\acmFig{hnResults}{1}{Average fitness ($\pm $ SE) of the highest
   performing individual in the population for each generation of
@@
   meaning that HyperNEAT's true performance without this penalty would
-  be much higher.}
+  be much higher.  Each generation involved nine evaluations, one for
+  each of the nine individuals, so the total evaluations per HyperNEAT
+  run was 180.}
 
@@
 
-\acmFig{neat_110115_211410_00000_002_filt_zoom}{.8}{Example of one
+\acmFig{neat_110115_211410_00000_002_filt_zoom}{1}{Example of one
   high-performance gait produced by HyperNEAT showing

### 2011-04-13 paper_ecal/introduction.tex
forging

@@
 unsuitable for wheeled robots, but doing so requires the creation of a
-gait specifying how the robot is to walk.  Gaits may be designed
-by one of several methods: either manually by an expert or
-via computer learning algorithms.  Learned gaits
-offer several advantages over manually designed gaits.  Automatically
-learning gaits can save valuable engineering time and can allow
-customization of gaits to a particular robot and its unique actuators.
-Most importantly, in many cases learned gaits can outperform manually
-designed gaits~\citep{hornby2005autonomous, valsalam2008modular}.
+gait specifying how the robot walks.  Such gaits may be designed either manually by an expert or
+via computer learning algorithms.  It is advantageous to automatically
+learn gaits because doing so can save valuable engineering time and allows
+gaits to be customized to the idiosyncrasies of different robots.
+Additionally, learned gaits have outperformed engineered gaits in some cases~\citep{hornby2005autonomous, valsalam2008modular}.
 
@@
 investigate how they perform when evolving on a physical robot
-(\figref{robot_whitebg.jpg}) on flat terrain.
+(\figref{robot_whitebg.jpg}).
 
@@
 
-A final motivation of this problem is to simply evolve effective gaits
-for a physical robot. Because HyperNEAT has performed well in
-simulation, it is interesting to test whether it can produce a fast
-gait for a physical robot. It is additionally interesting to test how
+An additional motivation is to test whether gaits evolved in simulation transfer to reality well, especially those evolved with cutting-edge evolutionary algorithms.
+Because HyperNEAT gaits performed well in
+simulation, it is interesting to test whether HyperNEAT can produce fast
+gaits for a physical robot. Such tests help us better understand the real world implications of results reported only in simulation. It is additionally interesting to test how
 more traditional gait optimization techniques compete with
-evolutionary algorithms when evolving in hardware.
+evolutionary algorithms when evolving in hardware. A final motivation of this research is simply to evolve effective gaits for a physical robot.
 

### 2011-04-13 paper_ecal/problemDefinition.tex
forging

@@
 
-For the system evaluated in this paper, we chose to compare only
-feedforward gaits generated by both the parameterized methods and
-HyperNEAT.  An interesting extension would be to allow gaits to depend
+For the system evaluated in this paper, we chose to compare open-loop gaits generated by both the parameterized methods and
+HyperNEAT.  An interesting extension would be to allow closed-loop gaits that depend
 on the measured servo positions, loads, voltage drops, or other

### 2011-04-13 paper_ecal/relatedWork.tex
forging

@@
 the coordination of motion. Part of the motivation of this paper
-is to see if the space of regularities can be explored automatically,
+is to see if the space of regularities can be explored more systematically,
 which has previously performed well \citep{valsalam2008modular}, and to
-perform a direct comparison on the same robot to HyperNEAT.
+compare such an approach to HyperNEAT.

### 2011-04-14 paper_ecal/experimentalSetup.tex
more forging

@@
 
-An important aspect of this study was to evaluate gait generation and
-learning algorithms in the noisy, unforgiving real world instead of
-using a simulator.  In this section we give details of the hardware
-setup we used.  First we discuss the robotic platform on which gaits
-were run, and then we describe the tracking system used to quantify
-the fitness of those gaits.
+We next describe the details of the experimental system.
 

### 2011-04-14 paper_ecal/fitnessEvaluation.tex
more forging

@@
   remote provided the location of the robot by tracking the infrared
-  LED mounted on the robot's antenna.  The LED position was measured
+  LED mounted on the robot's antenna.  The position was measured
   in pixels and transmitted from the Wii remote to the robot via
@@
 
-A separate tracking server, written in Python, ran on the robot and
+A separate Python tracking server ran on the robot and
 interfaced with the Wii remote via bluetooth using the \code{CWiid}
-library \citep{cwiid}.  Our fitness testing code communicated with this
+library \citep{cwiid}.  Our fitness-testing code communicated with this
 server via a socket connection and requested position updates at the
@@
 
-As mentioned earlier, the metric for evaluating gaits was their
-average speed, which was measured as the Euclidian distance the robot
+As mentioned earlier, the metric for evaluating gaits was the Euclidian distance the robot
 moved during a 12-second run. For the manual and parameterized gaits,
@@
 to occasionally move the robot back into the viewable area of the Wii
-remote whenever it wandered outside this window.  Initially this was a rare
+remote whenever it left this window.  Initially this was a rare
 occurrence, as the gaits did not typically produce motion as large as
@@
 Whenever it did so, we would discard the trial and repeat it until the
-gait was entirely within the window.
-
-While this process guaranteed that we always obtained a measurement
-for a given gait before proceeding, it also biased the measurements
+gait finished within the window. While this process guaranteed that we always obtained a measurement
+for a given gait before proceeding, it also biased some measurements
 downward.  Because the performance of the robot on a given gait varied
@@
 obtained when the gait happened to perform poorly.  This phenomenon
-was negligible at first, but became significantly more pronounced when
-evaluating several gaits that produced motion across nearly the entire
+was negligible at first, but became more pronounced as gaits began traversing the entire
 area.
 HyperNEAT gaits were especially likely to require additional trials, meaning that the reported performance for HyperNEAT is worse than it would have been otherwise.
-For future studies we will likely use a 2x2 array of Wii
-remotes to double the length and width of the measurement arena.
+For future studies we will use a $2\times2$ array of Wii
+remotes to double the size of the measurement arena.

### 2011-04-14 paper_ecal/introduction.tex
more forging

@@
 simulation, it is interesting to test whether HyperNEAT can produce fast
-gaits for a physical robot. Such tests help us better understand the real world implications of results reported only in simulation. It is additionally interesting to test how
+gaits for a physical robot, including handling the noisy, unforgiving nature of the real world.
+Such tests help us better understand the real world implications of results reported only in simulation.
+It is additionally interesting to test how
 more traditional gait optimization techniques compete with

### 2011-04-14 paper_ecal/learningMethods.tex
More edits from reviews, made figures larger
more forging

@@
-We have chosen to implement two classes of algorithms to produce
+We implemented two classes of algorithms to produce
 the motion-generating function $g(t)$ (described earlier).  The
@@
 By a \emph{parameterized gait}, we mean a gait produced by a
-parameterized function $g(t; \vec{\theta})$. Fixing the parameter
+parameterized function $g(t; \vec{\theta})$. Fixing the parameters
 $\vec{\theta}$ yields a deterministic motion function over time.  We
@@
 \hline
-Parameter    & Description               & Range \\
+Parameters        &                           &       \\
+in $\vec{\theta}$ & Description               & Range \\
 \hline
@@
 actual motor position commands, these waves are offset by
-appropriate fixed constants so that the base position (when the sine
+appropriate fixed constants ($C_O = 40$ for outer
+motors, $C_I = 800$ for inner motors, and $C_C = 512$ for the center hip motor) so that the base position (when the sine
 waves are at 0) is approximately a crouch (the position shown in
@@
 (center) motor is kept at a neutral position.  Thus, the commanded
-position for all motors, as a vector function of time, is:
+position for each motor (numbered in \figref{topdown2.jpg}), as a vector function of time, are produced by the following nine functions (listed in the order):
 
@@
 
-\noindent Here the fixed constants that determine the crouched
-position are $C_I = 800$ for inner motors, $C_O = 40$ for outer
-motors, and $C_C = 512$ for the center hip motor.  The nine functions (in order)
-correspond to the nine motors listed in
-\figref{topdown2.jpg}.
-
 \subsection{Learning Methods for Parameterized Gaits}
@@
 
-Using the SineModel5 parameterized motion model described in the
-previous section, along with the allowable ranges for each of the five
-parameters (shown in \tabref{parameters}), the task becomes how to
-choose the combination of five parameters that results in the fastest
-motion, per the chosen evaluation methods described earlier.
+Given the SineModel5 parameterized motion model (see previous section) and the allowable ranges for its five
+parameters (\tabref{parameters}), the task is discovering
+values for the five parameters that result in fast 
+gaits.
 
@@
 
-We evaluated a total of six different
-learning algorithms for the parameterized motion models.  All methods
-needed a way of picking a $\vec{\theta}$ for the first trial of each run.  This could be done by selecting a random starting vector within the
-allowable domain.  However, to most directly compare learning methods
-by their ability to improve on the initially chosen $\vec{\theta}$, we
-decided to evaluate the different methods by starting them at the same three
-initial $\vec{\theta}$ vectors in the three runs.
+We evaluated the following six different
+learning algorithms for the parameterized motion models:
 
@@
 %values to try from their past experience.
-
-The six learning algorithms for the parameterized motion models are as follows:
 
@@
 
-\emph{Random}: This method randomly generates parameter vectors in the allowable range for every trial.  This strategy was used only
-  as baseline for comparison.
+\emph{Random}: This method randomly generates parameter vectors in the allowable range for every trial.  This strategy serves as as baseline for comparison.
 
 \emph{Uniform random hill climbing}: This method always starts
-  with the current best gait, and then chooses next $\vec{\theta}$ by
+  with the current best gait, and then chooses the next $\vec{\theta}$ by
   randomly choosing one parameter to adjust and replacing it with a
@@
   longer distance walked than the previous best gait, it is saved as
-  the new best gait. The process is then repeated, always starting
-  with the best gait.
+  the new best gait. The process is then repeated.
 
@@
   $-\epsilon$, $0$, or $\epsilon$. Then each vector is run on the
-  robot, and for each dimension we group the results into three
+  robot, and for each dimension we segment the results into three
   groups: $-\epsilon$, $0$, and $\epsilon$.  The gradient along this
@@
 
-Each of the six learning methods was run on 3 different initial
-parameter vectors ($\vec{\theta}_A$,
-$\vec{\theta}_B$, and $\vec{\theta}_C$) in order to fairly compare the algorithms.
-Three runs were performed per learning method.
-Runs continued until the performance plateaued, which we defined as when there was no improvement during the last third of
+Three runs were performed per learning method. To most directly compare learning methods, we
+evaluated the different methods by starting each of their three runs, respectively, with the same three randomly-chosen initial parameter vectors~($\vec{\theta}_A$,
+$\vec{\theta}_B$, and $\vec{\theta}_C$). Runs continued until the performance plateaued, which we defined as when there was no improvement during the last third of
 a run.

### 2011-04-14 paper_ecal/platformDetails.tex
More edits, 75% done :)
Cleaned a few things up
more forging

@@
-\subsection{Platform details}
+\subsection{Platform Details}
 \seclabel{platformDetails}
@@
 
-The quadruped robot used in this study was assembled from parts
+The quadruped robot in this study was assembled from off-the-shelf components and parts
 printed on the Objet Connex 500 3-D Printing System. It weighs 1.88 kg
@@
 corresponding to a physical range [-120$^{\circ}$,~+120$^{\circ}$].
+The computer and servos can be powered by two on-board batteries, but
+for the tests presented in this paper power was provided by a tethered
+cable.
 
@@
 available on our website \citep{quadratotWeb}.  HyperNEAT is written in
-C++.  We controlled the servos with the \code{pydynamixel} library
+C++.  We controlled the servos with the Pydynamixel library
 \citep{pydynamixel}, sending commanded positions at 40Hz.  The robot
@@
 
-To run a gait on the robot, one needed only to provide a \emph{gait
-  function} --- any Python function that accepts a single input, time
-starting at 0, and outputs a list of nine commanded positions, one for
-each servo.  To safeguard against limb collision with the robot body,
+Robot gaits are executed via a Python \emph{gait function} that takes time (starting at 0) as a single input and outputs a list of nine commanded positions (one for each servo).
+To safeguard against limb collision with the robot body,
 the robot control code cropped the commands to a safe range.  This

### 2011-04-14 paper_ecal/problemDefinition.tex
more forging

@@
 The gait learning problem aims to find a \emph{gait} that maximizes
-some performance metric.
-
-Mathematically, we define a gait as a function that specifies
+some performance metric. Mathematically, we define a gait as a function that specifies
 a vector of commanded motor positions for a robot over time.
 %Using this definition, we require one function per motor.
-We can write gaits without feedback --- also called open-loop or
-feedforward gaits --- as
+We can write gaits without feedback --- also called open-loop gaits --- as
 
@@
 
-It follows that feedforward gaits are deterministic, producing the
+It follows that open-loop gaits are deterministic, producing the
 same command pattern each time they are run.  While the commanded
@@
 
-The ultimate goal was to design gaits that were as fast as possible,
-so to optimize for speed, our performance metric was displacement over
-a period of motion lasting approximately 12 seconds. Details of
-exactly how this displacement was measured are given below.
+The ultimate goal was to design gaits that were as fast as possible.
+Our performance metric was thus displacement over
+the evaluation period of 12 seconds. Details of
+how this displacement was measured are given below.

### 2011-04-14 paper_ecal/conclusion.tex
more more forging

--- 2011-02-10 paper/conclusion.tex
@@
 
-We have presented an array of approaches to optimizing a fast quadrupedal
-gait.  We implemented and tested six learning
+We have presented an array of approaches for optimizing a quadrupedal
+gaits for speed.  We implemented and tested six learning
 strategies for parameterized gaits and compared them to gaits produced by neural networks
@@
 Its best-performing gait traveled 45.7 body lengths per minute, which is nearly 9 times the speed of the hand-coded gait.
-This could be because HyperNEAT tends to generate coordinated gaits~\cite{clune2011performance, clune2009evolving}, allowing it to
-take advantage of the symmetries of the problem, but can also explore a much larger space of possibilities than the
+This could be because HyperNEAT tends to generate coordinated gaits~\citep{clune2011performance, clune2009evolving}, allowing it to
+take advantage of the symmetries of the problem. HyperNEAT can also explore a much larger space of possibilities than the
 more restrictive 5-dimensional parameterized space.  HyperNEAT gaits tended to produce more complex sequences of motor commands, with different frequencies and degrees of coordination, whereas the

### 2011-04-14 paper_ecal/futureWork.tex
more more forging

@@
 discrepancies between simulation and reality~\citep{koos2010crossing}.
-Another natural extension would be to allow gaits that sensed the position of the robot and other variables to enable the robot to adjust to its physical state, instead of providing an open-loop sequence of motor commands.
+Another extension would be to allow gaits that sensed the position of the robot and other variables to enable the robot to adjust to its physical state, instead of providing an open-loop sequence of motor commands.
 All of these approaches would likely improve the quality of automatically generated gaits for legged robots, which will hasten the day that humanity can benefit from their vast potential.

### 2011-04-14 paper_ecal/hyperNeatMethod.tex
Cleaned a few things up
More edits from reviews, made figures larger
more forging
more more forging

@@
-\subsection{HyperNEAT Gait Generation \& Learning}
+\subsection{HyperNEAT Gait Generation and Learning}
 \seclabel{hyperNeatMethod}
@@
 
-\acmFig{hyperneatExplanation.png}{.8}{HyperNEAT produces ANNs from
+\acmFig{hyperneatExplanation.png}{1}{HyperNEAT produces ANNs from
   CPPNs. ANN weights are specified as a function of the geometric
@@
 
-HyperNEAT is capable of exploiting the geometry of a problem because
-the link values between nodes in the final ANN phenotype are a
+HyperNEAT can exploit the geometry of a problem because
+the link values between nodes in the ANN phenotype are a
 function of the geometric positions of those nodes~\citep{stanley2009hypercube, clune2009sensitivity,
-  clune2011performance}. In the case of quadruped locomotion, this
-property has previously been shown to help HyperNEAT produce gaits with front-back, left-right,
+  clune2011performance}. For quadruped locomotion, this
+property has been shown to help HyperNEAT produce gaits with front-back, left-right,
 and four-way symmetries~\citep{clune2009evolving,
@@
 gaits with HyperNEAT in simulation~\citep{clune2011performance,
-  clune2009evolving}, but was adapted to accommodate the physical robot used in this paper.
+  clune2009evolving}, but was adapted to accommodate the physical robot in this paper.
 Specifically, the ANN has a fixed topology (i.e.
@@
 could be $(3 \times 4)^2= 288$ links in each ANN (although evolution can set weights to 0, functionally eliminating the connection). The inputs to the
-substrate were the current \emph{commanded} angles of each of the 9 joints of the robot and a sine and cosine wave (to facilitate the
+substrate were the angles \emph{requested} in the previous time step for each of the 9 joints of the robot (recall that gaits are open-loop, so actual joint angles are unknown) and a sine and cosine wave (to facilitate the
 production of periodic behaviors).  The sine and cosine waves had a period of about half a second.
@@
   first two columns of each row of the input layer receive information
-  about a single leg (the current angle of its two joints). The final
-  column provides a sine and cosine wave to enable periodic movements
-  and the angle of the center joint. Evolution determines the function
+  about a single leg (the angles requested in the previous time step for its two joints). The final
+  column provides the previously requested angle of the center joint and, to enable periodic movements, a sine and cosine wave. Evolution determines the function
   of the hidden-layer nodes. The nodes in the output layer specify new

### 2011-04-14 paper_ecal/results.tex
more more forging

@@
 
-\acmFig{std_error}{1}{Average results ($\pm $ SE) for each of the
+\acmFig{std_error}{1}{Average results ($\pm $ SE) for the
   parameterized learning methods, computed over three separately
@@
 \figref{hnResults} and \tabref{results}.  A total of 540 evaluations
-were performed for HyperNEAT, 180 in each of three runs.
+were performed for HyperNEAT (180 in each of three runs).
 Overall the HyperNEAT gaits
@@
 effective HyperNEAT gaits were not preserved across generations
-because if performance in one trial was poor, the genome was unlikely
-to be selected for. The HyperNEAT learning curve would be smoother if
+because a single poor-performing trial could prevent their selection. The HyperNEAT learning curve would be smoother if
 the noise in the evaluations could be reduced or more than one
@@
   meaning that HyperNEAT's true performance without this penalty would
-  be much higher.  Each generation involved nine evaluations, one for
-  each of the nine individuals, so the total evaluations per HyperNEAT
-  run was 180.}
+  be even higher.}
 
@@
   pattern. Such patterns were not possible with the parameterized
-  SineModel5, nor would they likely be contrived by a human designing
+  SineModel5, nor would they likely result from a human designing
   a different low-dimensional parameterized motion model.}

### 2011-04-15 code/Robot.py
Almost ready...

@@ -6,7 +6,7 @@ from copy import copy
 from numpy import array
 
 import dynamixel
-from Motion import lInterp
+from Motion import lInterp, scaleTime
 
 '''
 Much inspiration taken from http://code.google.com/p/pydynamixel/
@@ -127,7 +127,8 @@ class Robot():
 
         
     def run(self, motionFunction, runSeconds = 10, resetFirst = True,
-            interpBegin = 0, interpEnd = 0):
+            interpBegin = 0, interpEnd = 0, timeScale = 1, logFile = None,
+            extraLogInfoFn = None):
         '''Run the robot with a given motion generating function.
 
         Positional arguments:
@@ -158,20 +159,29 @@ class Robot():
                       current position to that specified by
                       motionFunction.  This should probably be used
                       for motion models which do not return POS_READY
-                      at time 0.  Default: None
+                      at time 0.  Affected by timeScale. Default: None
 
         interpEnd -- Same as interpBegin, but at the end of motion.
                       If interpEnd is not None, interpolation is
                       performed from final commanded position to
-                      POS_READY, over the given number of seconds.
-                      Default: None
+                      POS_READY, over the given number of
+                      seconds. Affected by timeScale.  Default: None
+                      
+        timeScale -- Factor by which time should be scaled during this
+                      run, higher is slower. Default: 1
+                      
+        logFile -- File to log time/positions to, should already be
+                      opened. Default: None
+
+        extraLogInfoFn -- Function to call and append info to every
+                      line the log file. Should return a
+                      string. Default: None
         '''
 
         #net, actuators = initialize()
 
         #def run(self, motionFunction, runSeconds = 10, resetFirst = True
         #    interpBegin = 0, interpEnd = 0):
-
 
         if self.loud:
             print 'Starting motion.'
@@ -186,6 +196,9 @@ class Robot():
         self.resetClock()
         self.currentPos = self.readCurrentPosition()
 
+        if logFile:
+            print >>logFile, '# time, goal(9 pos)'
+
         # Reset the robot position, if desired
         if resetFirst:
             self.interpMove(self.readCurrentPosition(), POS_FLAT, 3)
@@ -197,17 +210,26 @@ class Robot():
         # Begin with a segment smoothly interpolated between the
         # current position and the motion model.
         if interpBegin is not None:
-            self.interpMove(self.currentPos, motionFunction, interpBegin)
[…]

### 2011-04-15 code/RunManager.py

@@ -8,6 +8,17 @@ from Neighbor import Neighbor
 from util import prettyVec
 from wii.WiiTrackClient import WiiTrackClient
 
+
+
+def getLogPosString(wiiTrack):
+    pos = wiiTrack.getPosition()
+    if pos is None:
+        return '-1 -1'
+    else:
+        return ' '.join(pos)
+
+
+
 class RunManager:
     '''Manage runs..'''
 
@@ -84,7 +95,7 @@ class RunManager:
         else:
             return distance_walked
             
-    def run_function_and_log(self, motionFunction, runSeconds, logFileName):
+    def run_function_and_log(self, motionFunction, runSeconds, timeScale = 1, logFileName = None):
         '''
         Runs the robot with the given motion function from 0 to
         runSeconds, logging time and position to logFileName
@@ -104,14 +115,19 @@ class RunManager:
             self.manual_reset('Shimmy failed.  Fix and push enter to retry.')
             raise Exception
 
+        ff = open(logFileName, 'a')
+
         try:
             self.robot.run(motionFunction, runSeconds = runSeconds, resetFirst = False,
-                           interpBegin = 1, interpEnd = 2)
+                           interpBegin = 1, interpEnd = 2, logFile = ff,
+                           extraLogInfoFn = lambda: getLogPosString(wiiTrack))
         except RobotFailure as ee:
+            ff.close()
             print ee
             override = self.manual_reset('Robot run failure.  Fix something and push enter to retry.')
             raise Exception
-
+        ff.close()
+        
         endPos = wiiTrack.getPosition()
         if endPos is None:
             # Robot walked out of sensor view

### 2011-04-15 code/jimRuns.py

@@ -7,8 +7,11 @@ from Motion import scaleTime
 
 
 
-USAGE = '''
-'''
+USAGE = '''Usage:
+./jimRuns.py <gait type> <gait specifier>
+
+Example:
+  ./jimRuns.py -SineModel5 "400 2.0936276321011063 0.40692057680126537 0.70168527110547441 0.29694496659796576"'''
 
 
 
@@ -44,7 +47,7 @@ def main():
         print
         print 'Iteration', ii
         try:
-            runman.run_function_and_log(motionFunctionScaled, 10, 'loggy.txt')
+            runman.run_function_and_log(motionFunction, runSeconds = 10, timeScale = 1, logFilename = 'loggy.txt')
         except:
             print '*** Something failed, skipping'
 

### 2011-04-15 code/Robot.py
Works, just needs to sleep a bit less

@@ -197,7 +197,7 @@ class Robot():
         self.currentPos = self.readCurrentPosition()
 
         if logFile:
-            print >>logFile, '# time, goal(9 pos)'
+            print >>logFile, '# time, goal(9 pos), .........'
 
         # Reset the robot position, if desired
         if resetFirst:
@@ -259,7 +259,7 @@ class Robot():
                 extraInfo = ''
                 if extraLogInfoFn:
                     extraInfo = extraLogInfoFn()
-                print >>logFile, self.time, goal, ' '.join(ac.position for ac in self.actuators), extraInfo
+                print >>logFile, self.time, ' '.join([str(x) for x in goal]), ' '.join(str(ac.current_position) for ac in self.actuators), extraInfo
             sleep(self.sleep)
             self.updateClock()
 

### 2011-04-15 code/RunManager.py

@@ -15,7 +15,7 @@ def getLogPosString(wiiTrack):
     if pos is None:
         return '-1 -1'
     else:
-        return ' '.join(pos)
+        return ' '.join([str(x) for x in pos])
 
 
 
@@ -95,7 +95,7 @@ class RunManager:
         else:
             return distance_walked
             
-    def run_function_and_log(self, motionFunction, runSeconds, timeScale = 1, logFileName = None):
+    def run_function_and_log(self, motionFunction, runSeconds, timeScale = 1, logFilename = None):
         '''
         Runs the robot with the given motion function from 0 to
         runSeconds, logging time and position to logFileName
@@ -115,7 +115,7 @@ class RunManager:
             self.manual_reset('Shimmy failed.  Fix and push enter to retry.')
             raise Exception
 
-        ff = open(logFileName, 'a')
+        ff = open(logFilename, 'a')
 
         try:
             self.robot.run(motionFunction, runSeconds = runSeconds, resetFirst = False,

### 2011-04-15 code/jimRuns.py

@@ -28,7 +28,7 @@ def main():
             print 'Using SineModel5 with params: ', sineModel5Params
             
             motionFunction = lambda time: SineModel5().model(time,
-                                                             parameters = currentState)
+                                                             parameters = sineModel5Params)
         elif len(sys.argv) > 2 and sys.argv[1] == '-neatFiltFile':
             raise Exception('not yet')
             filtFile = sys.argv[2]
@@ -43,13 +43,13 @@ def main():
 
     runman = RunManager()
     
-    for ii in range(10):
+    for ii in range(1):
         print
         print 'Iteration', ii
-        try:
-            runman.run_function_and_log(motionFunction, runSeconds = 10, timeScale = 1, logFilename = 'loggy.txt')
-        except:
-            print '*** Something failed, skipping'
+        runman.run_function_and_log(motionFunction, runSeconds = 10, timeScale = 1, logFilename = 'loggy.txt')
+        #try:
+        #except:
+        #    print '*** Something failed, skipping'
 
 
 

### 2011-04-15 paper_ecal/fitnessEvaluation.tex
more more forging
Just about ready to submit

@@
 A separate Python tracking server ran on the robot and
-interfaced with the Wii remote via bluetooth using the \code{CWiid}
-library \citep{cwiid}.  Our fitness-testing code communicated with this
+interfaced with the Wii remote via bluetooth using the CWiid
+library.  Our fitness-testing code communicated with this
 server via a socket connection and requested position updates at the
@@
 
-\acmFig{linearInterp}{1}{Linear interpolation from the ready position to the first and last commanded gait positions. The position of the robot was measured at
-  the beginning and end of each run (red circles) in the same
- ready pose.  Motion was
-  interpolated linearly between this stationary pose and the commanded
-  gait g(t) for one second at the beginning of the run and two seconds
-  at the end, as shown above.}
+\acmFig{linearInterp}{1}{Motion was interpolated linearly between a
+  stationary pose and the commanded gait $g(t)$ for one second at the
+  beginning of each run and two seconds at the end, as shown above.
+  The position of the robot was measured at the beginning and end of
+  each run (red circles) in the ready pose.}
 

### 2011-04-15 paper_ecal/hyperNeatMethod.tex
Just about ready to submit

@@
   CPPNs. ANN weights are specified as a function of the geometric
-  coordinates of the source node and the target node for each
-  connection. The coordinates of these nodes and a constant bias are
+  coordinates of each connection's source and target nodes.
+ These coordinates and a constant bias are
   iteratively passed to the CPPN to determine each connection
@@
 
+\acmFig{SpiderANN.jpg}{.7}{ANN configuration for HyperNEAT runs. The
+  first two columns of each row of the input layer receive information
+  about a single leg (the angles requested in the previous time step for its two joints). The final
+  column provides the previously requested angle of the center joint and, to enable periodic movements, a sine and cosine wave. Evolution determines the function
+  of the hidden-layer nodes. The nodes in the output layer specify new
+  joint angles for each respective joint. The unlabeled nodes in the
+  input and output layers are ignored. Figure adapted from
+  \cite{clune2011performance}.}\label{hyperneatLayout}
+
 The ANN configuration follows previous studies that evolved quadruped
@@
 
-While measured angles could have been used as input instead of
-commanded angles, this may have given HyperNEAT gaits an unfair
-advantage by making them closed-loop. Thus, we restricted inputs to
-the commanded angles only, which allowed us to produce completely
-deterministic gaits using HyperNEAT.
+%While measured angles could have been used as input instead of
+%commanded angles, this may have given HyperNEAT gaits an unfair
+%advantage by making them closed-loop. Thus, we restricted inputs to
+%the commanded angles only, which allowed us to produce completely
+%deterministic gaits using HyperNEAT.
 
@@
 %  here or earlier?}
-
-\acmFig{SpiderANN.jpg}{.7}{ANN configuration for HyperNEAT Runs. The
-  first two columns of each row of the input layer receive information
-  about a single leg (the angles requested in the previous time step for its two joints). The final
-  column provides the previously requested angle of the center joint and, to enable periodic movements, a sine and cosine wave. Evolution determines the function
-  of the hidden-layer nodes. The nodes in the output layer specify new
-  joint angles for each respective joint. The unlabeled nodes in the
-  input and output layers are ignored. Figure adapted from
-  \cite{clune2011performance}.}\label{hyperneatLayout}
 

### 2011-04-15 paper_ecal/introduction.tex
Just about ready to submit

@@
 
-\acmFig{robot_whitebg.jpg}{.75}{The quadruped robot. The translucent
-  parts were created with 3D-printing technology. Videos of the gaits
-  evolved for this robot can be viewed at www.anonymous.url.}
+\acmFig{robot_whitebg.jpg}{.75}{The quadruped robot for which gaits
+  were evolved. The translucent parts were produced by a 3D
+  printer. Videos of the gaits can be viewed at http://bit.ly/hLHBYq}
 

### 2011-04-15 paper_ecal/learningMethods.tex
Just about ready to submit

@@
-We implemented two classes of algorithms to produce
-the motion-generating function $g(t)$ (described earlier).  The
-first method uses a family of parameterized gaits, described in this
-section and the next. The second method, HyperNEAT, evolves a non-parametric gait
-controlled by neural networks and is described last.
+%We implemented two classes of algorithms to produce the
+%motion-generating function $g(t)$ (described earlier).  The first
+%method uses a family of parameterized gaits, described in this section
+%and the next. The second method, HyperNEAT, evolves a non-parametric
+%gait controlled by neural networks and is described last.
+
+We now describe the gait-generating algorithm classes.
 
@@
 
-\begin{table}[h!]
+\begin{table}[b]
 \begin{center}
@@
 
-\emph{Uniform random hill climbing}: This method always starts
-  with the current best gait, and then chooses the next $\vec{\theta}$ by
+\emph{Uniform random hill climbing}: This method repeatedly starts
+  with the current best gait and then selects the next $\vec{\theta}$ by
   randomly choosing one parameter to adjust and replacing it with a
@@
   longer distance walked than the previous best gait, it is saved as
-  the new best gait. The process is then repeated.
+  the new best gait.
 

### 2011-04-15 paper_ecal/platformDetails.tex
more more forging
Just about ready to submit

@@
 evaluation code, except HyperNEAT, was written in Python and is
-available on our website \citep{quadratotWeb}.  HyperNEAT is written in
-C++.  We controlled the servos with the Pydynamixel library
-\citep{pydynamixel}, sending commanded positions at 40Hz.  The robot
+available on our website (http://bit.ly/hLHBYq).  HyperNEAT is written in
+C++.  We controlled the servos with the Pydynamixel library, sending commanded positions at 40Hz.  The robot
 connected to a wireless network on boot, which enabled us to control

### 2011-04-15 paper_ecal/relatedWork.tex
Just about ready to submit

@@
 gradient reinforcement learning approach for generating a fast walk on
-legged robots \citep{kohl}, which we implemented for comparison. Others
-have evolved gaits for legged robots, producing competitive
-results \citep{chernova2005evolutionary, hornby2005autonomous, zykov,
-  clune2009evolving, clune2011performance, clune2009hybrid,
-  clune2009sensitivity, tellez2006evolving, valsalam2008modular}. In
-fact, an evolved gait was used in the first commercially-available
-version of Sony's AIBO robot~\citep{hornby2005autonomous}. Except for
-work with HyperNEAT \citep{clune2009evolving, clune2011performance,
-  clune2009hybrid, clune2009sensitivity}, the previous evolutionary
-approaches have helped evolution exploit the regularity of the problem
-by manually decomposing the task. Experimenters have
-to choose which legs should be coordinated, or otherwise facilitate
-the coordination of motion. Part of the motivation of this paper
-is to see if the space of regularities can be explored more systematically,
-which has previously performed well \citep{valsalam2008modular}, and to
-compare such an approach to HyperNEAT.
+legged robots \citep{kohl}, which we implemented for
+comparison. Others have evolved gaits for legged robots, producing
+competitive results \citep{chernova2005evolutionary,
+  hornby2005autonomous, zykov, clune2009evolving,
+  clune2011performance, clune2009hybrid, clune2009sensitivity,
+  tellez2006evolving, valsalam2008modular}. In fact, an evolved gait
+was used in the first commercially-available version of Sony's AIBO
+robot~\citep{hornby2005autonomous}. Except for work with HyperNEAT
+\citep{clune2009evolving, clune2011performance, clune2009hybrid,
+  clune2009sensitivity}, the previous evolutionary approaches have
+helped evolution exploit the regularity of the problem by manually
+decomposing the task. Experimenters have to choose which legs should
+be coordinated, or otherwise facilitate the coordination of
+motion. Part of the motivation of this paper is to compare the
+regularities produced by HyperNEAT to those generated by a more
+systematic exploration of regularities via a parameterized model.

### 2011-04-15 paper_ecal/platformDetails.tex
Changed to degrees

@@
 
-Robot gaits are executed via a Python \emph{gait function} that takes time (starting at 0) as a single input and outputs a list of nine commanded positions (one for each servo).
-To safeguard against limb collision with the robot body,
-the robot control code cropped the commands to a safe range.  This
-range was [150,~770] for the inner leg servos, [30,~680] for the outer
-leg servos, and [392,~623] for the center hip servo.
+Robot gaits are defined by a Python \emph{gait function} that takes
+time (starting at 0) as a single input and outputs a list of nine
+commanded positions (one for each servo).  To safeguard against limb
+collision with the robot body, the control code cropped the commands
+to a safe range.  This range was [-85$^{\circ}$, +60$^{\circ}$] for
+the inner leg servos, [-113$^{\circ}$, +39$^{\circ}$] for the outer
+leg servos, and [-28$^{\circ}$, +28$^{\circ}$] for the center hip
+servo.
+
+%This range was [150,~770] for the inner leg servos, [30,~680] for the outer
+%leg servos, and [392,~623] for the center hip servo.
 

### 2011-04-17 paper_ecal/platformDetails.tex
Changed URL to correct bit.ly link

@@
 evaluation code, except HyperNEAT, was written in Python and is
-available on our website (http://bit.ly/hLHBYq).  HyperNEAT is written in
+available on our website (http://bit.ly/ecalgait).  HyperNEAT is written in
 C++.  We controlled the servos with the Pydynamixel library, sending commanded positions at 40Hz.  The robot

### 2011-04-18 paper_gecco_2/abstract.tex
Cut 8 page version to 2 pages. 1:42:49 :)

@@
 parameterized motion models and evolving artificial neural networks
-with the HyperNEAT generative encoding. Specifically, we test six
-different parameterized learning strategies: uniform and Gaussian
-random hill climbing, policy gradient reinforcement learning,
-Nelder-Mead simplex, a random baseline, and a new method that builds a
-model of the fitness landscape with linear regression to guide further
-exploration.  While all parameter search methods outperform a
-manually-designed gait, only the linear regression and Nelder-Mead
-simplex strategies outperform a random baseline strategy. Gaits
-evolved with HyperNEAT perform considerably better than all
-parameterized local search methods and produce gaits nearly 9 times faster than a hand-designed gait.  The best HyperNEAT gaits exhibit
-complex motion patterns that contain multiple frequencies, yet are
-regular in that the leg movements are coordinated.
+with the HyperNEAT generative encoding.  All parameter search methods
+outperform a manually-designed reference gait, but HyperNEAT performs
+better still, producing gaits nearly 9 times faster than the reference
+gait.
 

### 2011-04-18 paper_gecco_2/futureWork.tex
Cut 8 page version to 2 pages. 1:42:49 :)

%\section{Future work}
\seclabel{futureWork}

Because all trials were done in hardware, it was difficult to gather
the many trials that would be necessary to properly rank the methods
statistically.  One direction for future work could be to obtain many
more trials. However, a more effective extension might be to combine
frequent trials in simulation with infrequent trials in
hardware.
%The simulation would produce the necessary
%volume of trials to allow the learning methods to be effective, and
%the hardware trials would serve to continuously ground and refine the
%simulator.  One could also guide evolution to the most fertile
%territory by penalizing gaits that produced large discrepancies
%between simulation and reality~\cite{koos2010crossing}.
%Another
%natural extension would be to allow gaits that sensed the position of
%the robot and other variables to enable the robot to adjust to its
%physical state, instead of providing an open-loop sequence of motor
%commands.
%All of these approaches would likely improve the quality of
%automatically generated gaits for legged robots, which will hasten the
%day that humanity can benefit from their vast potential.

### 2011-04-18 paper_gecco_2/platformDetails.tex
Cut 8 page version to 2 pages. 1:42:49 :)

--- 2011-02-09 paper/platformDetails.tex
@@
 
-\acmFig{topdown.png}{1}{A figure of the robot from a top-down perspective,
+\acmFig{topdown.png}{.7}{A figure of the robot from a top-down perspective,
 with servos labeled}
@@
 evaluation code, except HyperNEAT, was written in Python and is
-available on our website \cite{quadratotWeb}.  HyperNEAT is written in
+available on our website (http://bit.ly/geccogait).  HyperNEAT is written in
 C++.  We controlled the servos with the \code{pydynamixel} library

### 2011-04-18 paper_gecco_2/relatedWork.tex
Cut 8 page version to 2 pages. 1:42:49 :)

%\subsection{Related Work}

Various machine learning techniques have proved to be effective at
generating gaits for legged robots. Kohl and Stone presented a policy
gradient reinforcement learning approach for generating a fast walk on
legged robots\cite{kohl}, which we implemented for comparison. Others
have evolved gaits for legged robots, producing competitive
%results~\cite{chernova2005evolutionary, hornby2005autonomous, zykov,
%  clune2009evolving, clune2011performance, clune2009hybrid,
%  clune2009sensitivity, tellez2006evolving, valsalam2008modular}. In
results~\cite{clune2009evolving, clune2009sensitivity, hornby2005autonomous}. In
fact, an evolved gait was used in the first commercially-available
version of Sony's AIBO robot~\cite{hornby2005autonomous}.

%Except for
%work with HyperNEAT~\cite{clune2009evolving, clune2009sensitivity}, the previous evolutionary
%approaches have helped evolution exploit the regularity of the problem
%by manually decomposing the task. Experimenters have
%to choose which legs should be coordinated, or otherwise facilitate
%the coordination of motion. Part of the motivation of this paper
%is to see if the space of regularities can be explored automatically,
%which has previously performed well~\cite{valsalam2008modular}, and to
%perform a direct comparison on the same robot to HyperNEAT.

In this paper we compare the performance of two different methods of
learning gaits: parameterized gaits optimized with six different
learning methods, and gaits generated by evolving neural networks with
the HyperNEAT generative encoding~\cite{stanley2009hypercube}. While
some of these methods, such as HyperNEAT, have been tested in
simulation~\cite{clune2009evolving}, we
investigate how they perform when evolving on a physical robot.

### 2011-04-18 paper_ecal/main.tex
More edits from reviews, made figures larger
more forging
Just about ready to submit
Updated (and resubmitted) ECAL version, changed author order

@@
 
-\title{Generating gaits for physical quadruped robots: evolved neural networks vs. local parameterized search}
+\title{Evolving Robot Gaits in Hardware: the HyperNEAT Generative Encoding Vs. Parameter Optimization}
+%\title{Generating gaits for physical quadruped robots: evolved neural networks vs. local parameterized search}
+%\title{Using HyperNEAT vs. Parameter Optimization to Create Quadruped Robot Gaits}
 \author {Jason Yosinski$^{1}$,
+Jeff Clune$^{1}$,
 Diana Hidalgo$^{1}$,
 Sarah Nguyen$^{1}$,
-Jeff Clune$^{1}$,
 Juan Cristobal Zagal$^{2}$,
@@
 \input{relatedWork}
-\input{paperOutline}
+%\input{paperOutline}
 

### 2011-06-10 paper_ecal/hyperNeatMethod.tex
Changes per three ECAL reviewers. Package for chairs started.

@@
 
-As with the parameterized methods, three runs of HyperNEAT
-were performed. Runs lasted 20 generations and the population size was 9 (which allowed a bare minimum of diversity within and between NEAT species).  These numbers are small, but were necessarily
-constrained given how much time it took to conduct evolution directly
-on a real robot. The remaining parameters were identical to Clune et
-al.~(2011).
+As with the parameterized methods, three runs of HyperNEAT were
+performed. Runs lasted 20 generations with a population size of 9
+organisms in 3 species, allowing a bare minimum of diversity
+within and between NEAT species.  These numbers were
+necessarily small given how much time it took to conduct
+evolution directly on a real robot. The remaining parameters were
+identical to Clune et al.~(2011).
 

### 2011-06-10 paper_ecal/learningMethods.tex
Changes per three ECAL reviewers. Package for chairs started.

@@
 
-We now describe the gait-generating algorithm classes.
+We now describe the classes of gait-generating algorithms.
 
@@
 actual motor position commands, these waves are offset by
-appropriate fixed constants ($C_O = 40$ for outer
+fixed constants ($C_O = 40$ for outer
 motors, $C_I = 800$ for inner motors, and $C_C = 512$ for the center hip motor) so that the base position (when the sine
 waves are at 0) is approximately a crouch (the position shown in
-\figref{robot_whitebg.jpg}).  In this motion model, the ninth
-(center) motor is kept at a neutral position.  Thus, the commanded
-position for each motor (numbered in \figref{topdown2.jpg}), as a vector function of time, are produced by the following nine functions (listed in the order):
+\figref{robot_whitebg.jpg}).
+To keep the size of the model search space as
+small as possible, we decided to keep the ninth (center) motor at a fixed neutral
+position.
+Thus, the commanded position for each motor as a vector function of
+time is as follows (numbered as in \figref{topdown2.jpg}):
 
@@
 \vec{g}(t) =
-\left[ {\begin{array}{c@{ }l@{ }l}
-\amp \cdot \sin(2\pi t / \tau) & \ \cdot \           m_F            & + C_I \\ % 0
-\amp \cdot \sin(2\pi t / \tau) & \ \cdot \ m_O \cdot m_F            & + C_O \\ % 1
-\amp \cdot \sin(2\pi t / \tau) & \                                  & + C_I \\ % 2
-\amp \cdot \sin(2\pi t / \tau) & \ \cdot \ m_O                      & + C_O \\ % 3
-\amp \cdot \sin(2\pi t / \tau) & \ \cdot \                     m_R  & + C_I \\ % 4
-\amp \cdot \sin(2\pi t / \tau) & \ \cdot \ m_O           \cdot m_R  & + C_O \\ % 5
-\amp \cdot \sin(2\pi t / \tau) & \ \cdot \           m_F \cdot m_R  & + C_I \\ % 6
-\amp \cdot \sin(2\pi t / \tau) & \ \cdot \ m_O \cdot m_F \cdot m_R  & + C_O \\ % 7
-0                              & \                                  & + C_C \\ % 8
+\left[ {\begin{array}{c@{ }c@{ }c@{ }l@{ }l}
+\amp \cdot \sin(2\pi t / \tau) & \             & \ \cdot \ m_F & \            & + C_I \\ % 0
+\amp \cdot \sin(2\pi t / \tau) & \ \cdot \ m_O & \ \cdot \ m_F & \            & + C_O \\ % 1
+\amp \cdot \sin(2\pi t / \tau) & \             & \             & \            & + C_I \\ % 2
+\amp \cdot \sin(2\pi t / \tau) & \ \cdot \ m_O & \             & \            & + C_O \\ % 3
+\amp \cdot \sin(2\pi t / \tau) & \             & \             & \ \cdot \ m_R  & + C_I \\ % 4
+\amp \cdot \sin(2\pi t / \tau) & \ \cdot \ m_O & \             & \ \cdot \ m_R  & + C_O \\ % 5
+\amp \cdot \sin(2\pi t / \tau) & \             & \ \cdot \ m_F & \ \cdot \ m_R  & + C_I \\ % 6
+\amp \cdot \sin(2\pi t / \tau) & \ \cdot \ m_O & \ \cdot \ m_F & \ \cdot \ m_R  & + C_O \\ % 7
+0                              & \             & \             & \            & + C_C \\ % 8
 \end{array} } \right]
@@
 \emph{Nelder-Mead simplex method}: The Nelder-Mead simplex
-  method \citep{nm} creates an initial simplex with $d+1$ vertices,
-  where $d$ is the dimension of the parameter space. It then tests the
+  method creates an initial simplex with $d+1$ vertices,
+  for a $d$ dimensional parameter space. It then tests the
   fitness of each vertex, and, in general, it reflects the worst point
-  over the centroid in an attempt to improve it.  Several additional
-  rules are used to prevent cycles and local minima \citep{nm}.
+  over the simplex's centroid in an attempt to improve it.  Several additional
+  rules are used to prevent cycles and local minima; see \cite{nm} for more information.
 

### 2011-06-13 paper_ecal/learningMethods.tex
few small tweaks, and the necessary shortening elsewhere to keep them space neutral

@@
 \emph{Nelder-Mead simplex method}: The Nelder-Mead simplex
-  method creates an initial simplex with $d+1$ vertices,
+  method creates an initial simplex with $d+1$ vertices
   for a $d$ dimensional parameter space. It then tests the
-  fitness of each vertex, and, in general, it reflects the worst point
+  fitness of each vertex and, in general, it reflects the worst point
   over the simplex's centroid in an attempt to improve it.  Several additional

### 2011-07-13 poster_gecco/poster.tex
Working on poster, 70% finished.

\documentclass[portrait,final]{baposter}

\usepackage{times}
\usepackage{calc}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{relsize}
\usepackage{multirow}
\usepackage{ownstyles}
\usepackage{bm}

\usepackage{wrapfig}
\usepackage{graphicx}
\usepackage{multicol}

\usepackage{pgfbaselayers}
\pgfdeclarelayer{background}
\pgfdeclarelayer{foreground}
\pgfsetlayers{background,main,foreground}

\usepackage{helvet}
%\usepackage{bookman}
\usepackage{palatino}

\newcommand{\captionfont}{\footnotesize}

\renewcommand\refname{}
%\renewcommand\bibname{New Title2}

\selectcolormodel{cmyk}

\graphicspath{{../figures/}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%% Some math symbols used in the text
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Format 
\newcommand{\Matrix}[1]{\begin{bmatrix} #1 \end{bmatrix}}
\newcommand{\Vector}[1]{\Matrix{#1}}
\newcommand*{\SET}[1]  {\ensuremath{\mathcal{#1}}}
\newcommand*{\MAT}[1]  {\ensuremath{\mathbf{#1}}}
\newcommand*{\VEC}[1]  {\ensuremath{\bm{#1}}}
\newcommand*{\CONST}[1]{\ensuremath{\mathit{#1}}}
\newcommand*{\norm}[1]{\mathopen\| #1 \mathclose\|}% use instead of $\|x\|$
\newcommand*{\abs}[1]{\mathopen| #1 \mathclose|}% use instead of $\|x\|$
\newcommand*{\absLR}[1]{\left| #1 \right|}% use instead of $\|x\|$

\def\norm#1{\mathopen\| #1 \mathclose\|}% use instead of $\|x\|$
\newcommand{\normLR}[1]{\left\| #1 \right\|}% use instead of $\|x\|$

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Multicol Settings
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\setlength{\columnsep}{0.7em}
\setlength{\columnseprule}{0mm}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Save space in lists. Use this after the opening of the list
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\compresslist}{%
\setlength{\itemsep}{1pt}%
\setlength{\parskip}{0pt}%
\setlength{\parsep}{0pt}%
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% Begin of Document
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% Here starts the poster
%%%---------------------------------------------------------------------------
%%% Format it to your taste with the options
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\typeout{Poster Starts}
\background{
  \begin{tikzpicture}[remember picture,overlay]%
    \draw (current page.north west)+(-2em,-0em) node[anchor=north west] {\hspace{-2em}\includegraphics[height=1.1\textheight]{silhouettes_background}};
  \end{tikzpicture}%
}
\definecolor{silver}{cmyk}{0,0,0,0.3}
\definecolor{blue1}{cmyk}{0.3,0.05,0.05,0.0}
\definecolor{darkblue1}{cmyk}{1.0,0.5,0.5,0.0}
\definecolor{black}{cmyk}{0,0,0.0,1.0}
\definecolor{darkYellow}{cmyk}{0,0,1.0,0.5}
\definecolor{darkSilver}{cmyk}{0,0,0,0.1}

\definecolor{lightorange}{rgb}{1.0, 0.9, 0.8}
\definecolor{lighteryellow}{cmyk}{0,0,0.1,0.0}
\definecolor{lightestyellow}{cmyk}{0,0,0.05,0.0}
\begin{poster}{
  columns = 3,
  % Show grid to help with alignment
  grid=no,
  % Column spacing
  colspacing=1em,
  % Color style
  bgColorOne=white,
  bgColorTwo=white,
  borderColor=darkblue1,
  headerColorOne=blue1,
  headerColorTwo=darkblue1,
  headerFontColor=black,
  boxColorOne=white,
  boxColorTwo=white,
  % Format of textbox
  textborder=roundedleft,
  % Format of text header
  eyecatcher=no,
  headerborder=open,
  headerheight=0.09\textheight,
  headershape=roundedright,
  headershade=plain,
  headerfont=\normalsize\textsf, %Sans Serif
  boxshade=plain,
%  background=shade-tb,
  background=plain,
  linewidth=2pt
  }
  % Eye Catcher
  {} % No eye catcher for this poster. If an eye catcher is present, the title is centered between eye-catcher and logo.
  % Title
  {\sf %Sans Serif
  %\bf% Serif
  \vspace{0em}{\huge Generating Gaits for Physical Quadruped Robots:} \\{\LARGE Evolved Neural Networks Vs. Local Parameterized Search}}
  % Authors
  {\sf %Sans Serif
  % Serif
  \vspace{.3em}\\{\normalsize Jason Yosinski, Jeff Clune, Diana Hidalgo, Sarah Nguyen, Juan Cristobal Zagal, Hod Lipson}\\\hspace{0.3em}
  \code{\footnotesize \{jy495, jeffclune, djh283, smn64\}@cornell.edu, jczagal@ing.uchile.cl, hod.lipson@cornell.edu}\hspace{3em}
  }
  % University logo
  {{\begin{minipage}{.25\textwidth}
    \hfill
    \includegraphics[height=5.0em]{cornellLogo}
  \end{minipage}}
  }

  \tikzstyle{light shaded}=[top color=baposterBGtwo!30!white,bottom color=baposterBGone!30!white,shading=axis,shading angle=30]

  % Width of left inset image
     \newlength{\leftimgwidth}
     \setlength{\leftimgwidth}{0.78em+8.0em}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% Now define the boxes that make up the poster
%%%---------------------------------------------------------------------------
%%% Each box has a name and can be placed absolutely or relatively.
%%% The only inconvenience is that you can only specify a relative position 
%%% towards an already declared box. So if you have a box attached to the 
%%% bottom, one to the top and a third one which should be in between, you 
%%% have to specify the top and bottom boxes before you specify the middle 
%%% box.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    %
    % A coloured circle useful as a bullet with an adjustably strong filling
    \newcommand{\colouredcircle}[1]{%
      \tikz{\useasboundingbox (-0.2em,-0.32em) rectangle(0.2em,0.32em); \draw[draw=black,fill=baposterBGone!80!black!#1!white,line width=0.03em] (0,0) circle(0.18em);}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
  \headerbox{Abstract}{name=abstract,column=0,row=0}{
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \input{abstract}
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
  \headerbox{Platform}{name=platform,column=0,below=abstract}{
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

    %\pwfig[l]{robot_whitebg.jpg}{.5}{The quadruped robot for which
    %  gaits were evolved. The translucent parts were produced by a 3D
    %  printer. Videos of the gaits can be viewed at
    %  http://bit.ly/geccogait}
    %
    %\pwfig[r]{topdown2.jpg}{.3}{(a) Top-down perspective of the robot
    %  with the nine joints and associated servos labeled. (b) The
    %  robot in a flat pose with the hip joint centered. (c,d,e)
    %  Various views of a pose in which the hip joint is rotated.}

    \pfigg{robot_whitebg.jpg}{topdown2.jpg}{.49}{The 3D printed quadruped robot. Videos of the gaits can be viewed at
      http://bit.ly/geccogait}

The quadruped robot in this study was assembled from off-the-shelf components and parts
printed on the Objet Connex 500 3-D Printing System. It weighs 1.88 kg
with the on-board computer and measures approximately 38 centimeters
from leg to opposite leg in the crouch position depicted above.
The robot is actuated by 9 AX-12+
Dynamixel servos: one inner joint and one outer joint servo in each of
the four legs, and one servo at the center ``hip'' joint.  This final
unique servo allows the two halves of the robot to rotate with respect
to each other.

All of the computation for gait learning, fitness evaluation, and
robot control was performed on the compact, on-board CompuLab Fit-PC2,
running Ubuntu Linux 10.10.
All gait generation, learning, and fitness
evaluation code, except HyperNEAT, was written in Python and is
available on our website (http://bit.ly/geccogait).

  }

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
  \headerbox{Experimental Setup}{name=experimentalSetup,column=1}{
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%\pfigg{wiiMote_crop.jpg}{robot_led_crop.jpg}{1}{A Nintendo Wii
%  remote provided the location of the robot by tracking the infrared
%  LED mounted on the robot's antenna.  The position was measured
%  in pixels and transmitted from the Wii remote to the robot via
%  bluetooth.}

\pwfigg{wiiMote_crop.jpg}{robot_led_crop.jpg}{.6}{A Nintendo Wii
  remote tracked an IR LED on the robot, transmitting the (x,y)
  location via bluetooth.}

%To track the position of the robot and thus determine gait fitness, we
%mounted a Nintendo Wii remote on the ceiling and an infrared LED on
%top of the robot.  The Wii remote contains
%an IR camera that tracks and reports the position of IR sources.
The
resolution of the camera was 1024 by 768 pixels with view angles of
about 40$^\circ$ by 30$^\circ$, which produced a resolution of 1.7mm
per pixel when mounted at a height of 2.63m.  At this height, the
viewable window on the floor was approximately 175 x 120 cm.

\pwfig[r]{linearInterp}{.6}{Motion was interpolated linearly between a
  stationary pose and the commanded gait $g(t)$.}

\vspace{.5em}

The fitness was the Euclidian distance the robot
moved during a 12-second run on flat terrain.
The HyperNEAT gaits stressed the motors
more than the other gaits, so to encourage gaits that did not tax the
motors, we penalized gaits that caused the servos to stop
responding by a factor of 1/2.

The robot was always measured while in the \emph{ready} position
(shown in the first figure) to ensure fair fitness measurements. As
shown below the instantaneous robot limb configuration during the
first and last portions of the evaluation was an interpolation between
the initial ready position and $g(t)$; during the rest of the
evaluation, the robot followed the commanded gait exactly.

  }

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
  \headerbox{Parameterized Gaits}{name=parameterizedGaits,column=1,span=1,below=experimentalSetup}{
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Parameterized gaits were produced by the following gait function $g(t; \vec{\theta})$:

%\newcommand{\amp}{\ensuremath{\mathrm{amp}}}
\newcommand{\amp}{\ensuremath{\alpha}}

{
\footnotesize
\[
\vec{g}(t) =
\left[ {\begin{array}{c@{ }c@{ }c@{ }l@{ }l}
\amp \cdot \sin(2\pi t / \tau) & \             & \ \cdot \ m_F & \            & + C_I \\ % 0
\amp \cdot \sin(2\pi t / \tau) & \ \cdot \ m_O & \ \cdot \ m_F & \            & + C_O \\ % 1
\amp \cdot \sin(2\pi t / \tau) & \             & \             & \            & + C_I \\ % 2
\amp \cdot \sin(2\pi t / \tau) & \ \cdot \ m_O & \             & \            & + C_O \\ % 3
\amp \cdot \sin(2\pi t / \tau) & \             & \             & \ \cdot \ m_R  & + C_I \\ % 4
\amp \cdot \sin(2\pi t / \tau) & \ \cdot \ m_O & \             & \ \cdot \ m_R  & + C_O \\ % 5
\amp \cdot \sin(2\pi t / \tau) & \             & \ \cdot \ m_F & \ \cdot \ m_R  & + C_I \\ % 6
\amp \cdot \sin(2\pi t / \tau) & \ \cdot \ m_O & \ \cdot \ m_F & \ \cdot \ m_R  & + C_O \\ % 7
0                              & \             & \             & \            & + C_C \\ % 8
\end{array} } \right]
\]
}

\noindent with parameters in the following ranges:

{
\footnotesize
\begin{center}
\begin{tabular}{|c|c|c|}
\hline
Parameters        &                           &       \\
in $\vec{\theta}$ & Description               & Range \\
\hline
\hline
\amp         & Amplitude                 & [0, 400] \\
\hline
$\tau$       & Period                    & [.5, 8] \\
\hline
$m_O$        & Outer-motor multiplier    & [-2, 2] \\
\hline
$m_F$        & Front-motor multiplier    & [-1, 1] \\
\hline
$m_R$        & Right-motor multiplier    & [-1, 1] \\
\hline
\end{tabular}
\tablabel{parameters}
\label{tab:params}
\end{center}
}

Intuitively, SineModel5 starts with 8 identical sine waves of
amplitude $\amp$ and period $\tau$, multiplies the waves for all outer
motors by $m_O$, multiplies the waves for all front motors by $m_F$,
and multiplies the waves for all right motors by $m_R$.  To obtain the
actual motor position commands, these waves are offset by
fixed constants ($C_O = 40$ for outer
motors, $C_I = 800$ for inner motors, and $C_C = 512$ for the center hip motor) so that the base position (when the sine
waves are at 0) is approximately a crouch.
To keep the size of the model search space as
small as possible, we decided to keep the ninth (center) motor at a fixed neutral
position.

\vspace{.5em}

The \textbf{learning methods} used to search the parameterized gait
space were: \textbf{Random}, \textbf{Uniform random hill climbing},
\textbf{Gaussian random hill climbing}, \textbf{N-dimensional policy
  gradient ascent}, \textbf{Nelder-Mead simplex method}, \textbf{Linear
  regression}, and \textbf{SVM regression}.

  }

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
  \headerbox{HyperNEAT Gaits}{name=hyperneatGaits,column=1,span=1}{
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Gaits generated via HyperNEAT were compared to the parameterized gaits.

\pfig{hyperneatExplanation.png}{1}{HyperNEAT produces ANNs from
  CPPNs \cite{clune2011performance}.}

HyperNEAT is an indirect encoding for evolving artificial neural
networks (ANNs) that is inspired by the way natural organisms
develop~\cite{stanley2009hypercube}. It evolves Compositional Pattern
Producing Networks (CPPNs)~\cite{stanley2007compositional}, each of
which is a genome that encodes an ANN
\pwfig{SpiderANN.jpg}{.6}{The ANN configuration used to generate gaits
  \cite{clune2011performance}.}\label{hyperneatLayout}
phenotype~\cite{stanley2009hypercube}. Each CPPN is itself a directed
graph, where the nodes in the graph are mathematical functions, such as sine or
Gaussian. The nature of these functions can facilitate the evolution
of properties such as symmetry (e.g.\ a Gaussian function) and repetition (e.g.\ a sine
function)~\cite{stanley2009hypercube, stanley2007compositional}.

For quadruped locomotion, HyperNEAT has previously produced gaits with
front-back, left-right, and four-way
symmetries~\cite{clune2009evolving, clune2011performance}.

\pfig{neat_110115_211410_00000_002_filt_zoom}{1}{Joint angles for a
  high-performance HyperNEAT gait.}

  }

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
  \headerbox{Results}{name=results,column=2,span=1,row=0}{
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

    By training on data collected in the lab, we were able to predict
    quadrotor drift quite accurately.  The below images show the
    actual versus predicted drift, including the directly predicted
    derivatives as well as the inferred (integrated) position.

    %\begin{multicols}{2}

      %\begin{center}
      %  \includegraphics[width=0.9\linewidth]{dec17_1_DataLog-1292574599_07_int}
      %\end{center}

    %\end{multicols}\vspace{-1em}

    %%%%%%%%%%%%%%%%%%%%%%
    \mbox{\hspace{0.3\linewidth}\rule{0.4\linewidth}{1pt}\hspace{0.3\linewidth}}\\

    Drift predictions under normal conditions, trained on data
    collected in similar conditions.
    \vspace{-.7em}

    \begin{multicols}{3}

      \begin{tabular}{ccc}
        \includegraphics[width=.9\linewidth]{robot_whitebg.jpg}
        &
        \includegraphics[width=.9\linewidth]{robot_whitebg.jpg}
        &
        \includegraphics[width=.9\linewidth]{robot_whitebg.jpg}
        \\

        \capt{Predicted versus actual x drift at each timestep in
          pixels per second.} & \capt{Predicted versus actual y drift
          at each timestep in pixels per second.} & \capt{Actual
          position versus position inferred from integrated
          predictions}
      \end{tabular}\\

    \end{multicols}\vspace{-1em}

    %%%%%%%%%%%%%%%%%%%%%%
    \mbox{\hspace{0.3\linewidth}\rule{0.4\linewidth}{1pt}\hspace{0.3\linewidth}}\\

    Predictions under artificially windy conditions, trained on data
    collected under normal conditions as well as windy conditions.
    \vspace{-.7em}

    \begin{multicols}{3}

      \begin{tabular}{ccc}
        \includegraphics[width=.9\linewidth]{robot_whitebg.jpg}
        &
        \includegraphics[width=.9\linewidth]{robot_whitebg.jpg}
        &
        \includegraphics[width=.9\linewidth]{robot_whitebg.jpg}
        \\

        \capt{Predicted vs. actual x drift} & \capt{Predicted
          vs. actual y drift} & \capt{Integrated position}
      \end{tabular}\\

      \end{multicols}\vspace{-1em}

    %%%%%%%%%%%%%%%%%%%%%%
    \mbox{\hspace{0.3\linewidth}\rule{0.4\linewidth}{1pt}\hspace{0.3\linewidth}}\\

    Although our algorithm was able to accurately predict drift, we
    are still working on getting the prediction code to be fast enough
    to control the quadrotor in live flight.  The SVM regression code
    runs quickly, but we're currently facing difficulties with delays
    due to I/O as well as a suboptimal configuration.\\

  }%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
  \headerbox{Conclusion}{name=conclusion,column=2,span=1,below=results}{
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    We are making strong progress towards predicting and removing
    quadrotor drift.  The drift is predicted fairly accurately via a
    60 dimensional sensor data vector from the quadrotor, and we hope
    to soon demonstrate a controller enhanced by the predicted drift. \vspace{.5em}
  }%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
  \headerbox{Abbreviated References}{name=references,column=0,above=bottom}{
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

    \vspace{-1em}

    \footnotesize
    \bibliography{references.bib}
    \bibliographystyle{is-plain}
  }%
\end{poster}%
%
\end{document}

### 2011-07-13 poster_gecco/poster.tex
Everything fits!

@@
 
-\renewcommand\refname{}
-%\renewcommand\bibname{New Title2}
+\renewcommand\refname{}   % get rid of ``References'' title
 
@@
 \begin{poster}{
-  columns = 3,
+  columns = 4,
   % Show grid to help with alignment
@@
   headerborder=open,
-  headerheight=0.09\textheight,
+  headerheight=0.075\textheight,
   headershape=roundedright,
@@
   %\bf% Serif
-  \vspace{0em}{\huge Generating Gaits for Physical Quadruped Robots:} \\{\LARGE Evolved Neural Networks Vs. Local Parameterized Search}}
+  \vspace{.4em}{\huge Generating Gaits for Physical Quadruped Robots:} \\{\LARGE Evolved Neural Networks Vs. Local Parameterized Search}\vspace{.3em}}
   % Authors
@@
   % Serif
-  \vspace{.3em}\\{\normalsize Jason Yosinski, Jeff Clune, Diana Hidalgo, Sarah Nguyen, Juan Cristobal Zagal, Hod Lipson}\\\hspace{0.3em}
-  \code{\footnotesize \{jy495, jeffclune, djh283, smn64\}@cornell.edu, jczagal@ing.uchile.cl, hod.lipson@cornell.edu}\hspace{3em}
+  \vspace{-.25em}{\normalsize Jason Yosinski, Jeff Clune, Diana Hidalgo, Sarah Nguyen, Juan Cristobal Zagal, Hod Lipson}\\\hspace{0.3em}
+  \code{\footnotesize yosinski@cs.cornell.edu}\hspace{3em}
   }
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+    {\small
     \input{abstract}
+    }
+    \vspace{.5em}
 }
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-  \headerbox{Platform}{name=platform,column=0,below=abstract}{
+  \headerbox{Platform and Experimental Setup}{name=platform,column=0,below=abstract,above=bottom}{
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+    \small
 
@@
 
-The quadruped robot in this study was assembled from off-the-shelf components and parts
-printed on the Objet Connex 500 3-D Printing System. It weighs 1.88 kg
-with the on-board computer and measures approximately 38 centimeters
-from leg to opposite leg in the crouch position depicted above.
-The robot is actuated by 9 AX-12+
-Dynamixel servos: one inner joint and one outer joint servo in each of
-the four legs, and one servo at the center ``hip'' joint.  This final
-unique servo allows the two halves of the robot to rotate with respect
-to each other.
+    The quadruped robot was assembled from off-the-shelf
+    components and parts printed on the Objet Connex 500 3-D Printing
+    System.
+    % It weighs 1.88 kg with the on-board computer and measures 38
+    % centimeters from leg to opposite leg in the crouch position
+    % depicted above.
+    The robot is actuated by 9 AX-12+ Dynamixel servos: 8 for the legs and one at the center ``hip'' joint.
+    %This final unique servo allows
+    %the two halves of the robot to rotate with respect to each other.
 
-All of the computation for gait learning, fitness evaluation, and
-robot control was performed on the compact, on-board CompuLab Fit-PC2,
-running Ubuntu Linux 10.10.
-All gait generation, learning, and fitness
-evaluation code, except HyperNEAT, was written in Python and is
-available on our website (http://bit.ly/geccogait).
+    \vspace{.5em}
 
-  }
+    All computation for gait learning, fitness evaluation, and
+    robot control was performed on the compact, on-board CompuLab
+    Fit-PC2, running Ubuntu Linux 10.10.  All gait generation,
+    learning, and fitness evaluation code, except HyperNEAT, was
+    written in Python and is available on our website
+    (http://bit.ly/geccogait).
 
+%  }
+%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-  \headerbox{Experimental Setup}{name=experimentalSetup,column=1}{
+%  \headerbox{Experimental Setup}{name=experimentalSetup,column=0}{
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-\pwfigg{wiiMote_crop.jpg}{robot_led_crop.jpg}{.6}{A Nintendo Wii
+\pfigg{wiiMote_crop.jpg}{robot_led_crop.jpg}{1}{A Nintendo Wii
   remote tracked an IR LED on the robot, transmitting the (x,y)
@@
 
-\pwfig[r]{linearInterp}{.6}{Motion was interpolated linearly between a
-  stationary pose and the commanded gait $g(t)$.}
-
     \vspace{.5em}
-
-The fitness was the Euclidian distance the robot
-moved during a 12-second run on flat terrain.
-The HyperNEAT gaits stressed the motors
-more than the other gaits, so to encourage gaits that did not tax the
-motors, we penalized gaits that caused the servos to stop
-responding by a factor of 1/2.
-
-The robot was always measured while in the \emph{ready} position
-(shown in the first figure) to ensure fair fitness measurements. As
-shown below the instantaneous robot limb configuration during the
-first and last portions of the evaluation was an interpolation between
-the initial ready position and $g(t)$; during the rest of the
-evaluation, the robot followed the commanded gait exactly.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-  \headerbox{Parameterized Gaits}{name=parameterizedGaits,column=1,span=1,below=experimentalSetup}{
+  \headerbox{Fitness Evaluation}{name=fitnessEvaluation,column=1}{
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+The fitness was the Euclidian distance the robot moved during a
+12-second run on flat terrain.  To encourage gaits that did not tax
+the motors, we penalized gaits that caused the servos to stop
+responding by a factor of 1/2.
+
+\pfig{linearInterp}{1}{Motion was interpolated linearly between a
+  stationary pose and the gait $g(t)$.}
+
+The robot was always measured while in the \emph{ready} position
+to ensure fair fitness measurements. As
+shown above the instantaneous robot limb configuration during the
+first and last portions of the evaluation was an interpolation between
+the initial ready position and $g(t)$.
+
+  }
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+  \headerbox{Parameterized Gaits}{name=parameterizedGaits,column=1,span=1,below=fitnessEvaluation,above=bottom}{
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
+\vspace{-.75em}
+
 {
-\footnotesize
+\tiny
 \[
@@
 
-\noindent with parameters in the following ranges:
+\vspace{-.75em}
+
+\noindent with parameters in the ranges:
+
+\vspace{-.75em}
 
 {
-\footnotesize
+\scriptsize
 \begin{center}
@@
 \hline
-$m_O$        & Outer-motor multiplier    & [-2, 2] \\
+$m_O$        & Outer multiplier    & [-2, 2] \\
 \hline
-$m_F$        & Front-motor multiplier    & [-1, 1] \\
+$m_F$        & Front multiplier    & [-1, 1] \\
 \hline
-$m_R$        & Right-motor multiplier    & [-1, 1] \\
+$m_R$        & Right multiplier    & [-1, 1] \\
 \hline
@@
 
-Intuitively, SineModel5 starts with 8 identical sine waves of
-amplitude $\amp$ and period $\tau$, multiplies the waves for all outer
-motors by $m_O$, multiplies the waves for all front motors by $m_F$,
-and multiplies the waves for all right motors by $m_R$.  To obtain the
+Intuitively, we start with 8 identical sine waves of
+amplitude $\amp$ and period $\tau$, multiply the waves for all outer
+motors by $m_O$, multiply the waves for all front motors by $m_F$,
+and multiply the waves for all right motors by $m_R$.  To obtain the
 actual motor position commands, these waves are offset by
@@
 \textbf{Gaussian random hill climbing}, \textbf{N-dimensional policy
-  gradient ascent}, \textbf{Nelder-Mead simplex method}, \textbf{Linear
-  regression}, and \textbf{SVM regression}.
+  gradient ascent}, \textbf{Nelder-Mead simplex method},
+\textbf{Linear regression}, and \textbf{SVM regression}. A discussion
+of each is in the paper.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-  \headerbox{HyperNEAT Gaits}{name=hyperneatGaits,column=1,span=1}{
+  \headerbox{HyperNEAT Gaits}{name=hyperneatGaits,column=2,span=1}{
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+    \small
 
@@
     networks (ANNs) that is inspired by the way natural organisms
-develop~\cite{stanley2009hypercube}. It evolves Compositional Pattern
-Producing Networks (CPPNs)~\cite{stanley2007compositional}, each of
-which is a genome that encodes an ANN
-\pwfig{SpiderANN.jpg}{.6}{The ANN configuration used to generate gaits
+    develop~\cite{stanley2009hypercube}. It evolves Compositional
+    Pattern Producing Networks
+    (CPPNs)~\cite{stanley2007compositional}, each of which is a genome
+    that encodes an ANN \pwfig{SpiderANN.jpg}{.7}{The ANN
+      configuration used to generate gaits
       \cite{clune2011performance}.}\label{hyperneatLayout}
-phenotype~\cite{stanley2009hypercube}. Each CPPN is itself a directed
-graph, where the nodes in the graph are mathematical functions, such as sine or
-Gaussian. The nature of these functions can facilitate the evolution
-of properties such as symmetry (e.g.\ a Gaussian function) and repetition (e.g.\ a sine
+    phenotype~\cite{stanley2009hypercube}. Each CPPN is itself a
+    directed graph, where the nodes in the graph are mathematical
+    functions, such as sine or Gaussian. The nature of these functions
+    can facilitate the evolution of properties such as symmetry
+    (e.g.\ a Gaussian function) and repetition (e.g.\ a sine
     function)~\cite{stanley2009hypercube, stanley2007compositional}.
 
-For quadruped locomotion, HyperNEAT has previously produced gaits with
-front-back, left-right, and four-way
+    For quadruped locomotion, HyperNEAT has previously produced gaits
+    with front-back, left-right, and four-way
     symmetries~\cite{clune2009evolving, clune2011performance}.
 
-\pfig{neat_110115_211410_00000_002_filt_zoom}{1}{Joint angles for a
-  high-performance HyperNEAT gait.}
+    \pfig{neat_110115_211410_00000_002_filt_zoom}{1}{Joint angles for
+      a high-performance HyperNEAT gait.}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-  \headerbox{Results}{name=results,column=2,span=1,row=0}{
+  \headerbox{Results}{name=results,column=3,span=1,row=0}{
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-    By training on data collected in the lab, we were able to predict
-    quadrotor drift quite accurately.  The below images show the
-    actual versus predicted drift, including the directly predicted
-    derivatives as well as the inferred (integrated) position.
+\small
+\textbf{Parameterized Gaits}
+\vspace{.5em}
 
-    %\begin{multicols}{2}
+\noindent A total of 1217 hardware fitness evaluations were performed to learn
+parameterized gaits in three separate runs (200 random, 234 uniform, 284 Gaussian, 174
+gradient, 172 simplex, 153 linear regression). The SVM-modeled method
+was never successful.
 
-      %\begin{center}
-      %  \includegraphics[width=0.9\linewidth]{dec17_1_DataLog-1292574599_07_int}
-      %\end{center}
+%One reason the randomly-generated SineModel5 gaits were so effective
+%may have been due to the SineModel5's bias toward regular, symmetric
+%gaits.  This may have allowed the random strategy --- focusing on
+%exploration --- to be competitive with the more directed strategies
+%that exploit information from past evaluations.
 
-    %\end{multicols}\vspace{-1em}
+\pfig{std_error}{1}{Average results ($\pm $ SE) for parameterized
+  learning methods over three runs.  Linear regression found the
+  fastest overall gait, followed by
+  Nelder-Mead simplex.}
 
-    %%%%%%%%%%%%%%%%%%%%%%
-    \mbox{\hspace{0.3\linewidth}\rule{0.4\linewidth}{1pt}\hspace{0.3\linewidth}}\\
+\textbf{HyperNEAT Gaits}
+\vspace{.5em}
 
-    Drift predictions under normal conditions, trained on data
-    collected in similar conditions.
-    \vspace{-.7em}
+\noindent A total of 540 evaluations were performed for HyperNEAT, whose gaits
+were the faster by far than the parameterized models.  The single best
+gait found had a speed of 45.72 body lengths/minute, 66\% better than
+the best non-HyperNEAT gait and 8.9 times faster than the hand-coded
+gait.
 
-    \begin{multicols}{3}
+Evaluation of HyperNEAT gaits was very noisy, which made learning
+difficult. We tested one gait 11 times and found performance of 26
+$\pm$ 13 body lengths/minute, with a max of 38 and a min of 3.
 
-      \begin{tabular}{ccc}
-        \includegraphics[width=.9\linewidth]{robot_whitebg.jpg}
-        &
-        \includegraphics[width=.9\linewidth]{robot_whitebg.jpg}
-        &
-        \includegraphics[width=.9\linewidth]{robot_whitebg.jpg}
-        \\
+%Many effective HyperNEAT gaits were not preserved across generations
+%because a single poor-performing trial could prevent their
+%selection. The HyperNEAT learning curve would be smoother if the noise
+%in the evaluations could be reduced or more than one evaluation per
+%individual could be afforded.
 
-        \capt{Predicted versus actual x drift at each timestep in
-          pixels per second.} & \capt{Predicted versus actual y drift
-          at each timestep in pixels per second.} & \capt{Actual
-          position versus position inferred from integrated
-          predictions}
-      \end{tabular}\\
+\pfig{hnResults}{.8}{Average fitness ($\pm $ SE) of the highest
+  performing individual at each generation.}
 
-    \end{multicols}\vspace{-1em}
+{\scriptsize
+\begin{center}
+\begin{tabular}{|r|c|c|c||c|}
+\hline
+                                         & Avg & SD \\
+\hline                                    
+\hline                                    
+Previous hand-coded gait                 & 5.16   &   --     \\
+\hline
+Random search                            & 9.40   &   6.83   \\
+\hline
+Uniform Hill Climbing             & 7.83   &   4.56   \\
+\hline
+Gaussian Hill Climbing            & 10.03  &   6.00   \\
+\hline
+Policy Gradient Descent                  & 6.32   &   7.39   \\
+\hline
+Nelder-Mead simplex                      & 12.32  &   3.35   \\
+\hline
+Linear Regression                        & 14.01  &  12.88   \\
+\hline
+Evolved Neural Network              &        &          \\
+(HyperNEAT)                              & 29.26  &   6.37   \\
+\hline
+\end{tabular}
+\capt{Average and SD of the best gaits found
+  during each of three runs, in body
+  lengths/minute.}  \tablabel{results}
+\end{center}
+}
 
-    %%%%%%%%%%%%%%%%%%%%%%
-    \mbox{\hspace{0.3\linewidth}\rule{0.4\linewidth}{1pt}\hspace{0.3\linewidth}}\\
-
-    Predictions under artificially windy conditions, trained on data
-    collected under normal conditions as well as windy conditions.
-    \vspace{-.7em}
-
-    \begin{multicols}{3}
-
-      \begin{tabular}{ccc}
-        \includegraphics[width=.9\linewidth]{robot_whitebg.jpg}
-        &
-        \includegraphics[width=.9\linewidth]{robot_whitebg.jpg}
-        &
-        \includegraphics[width=.9\linewidth]{robot_whitebg.jpg}
-        \\
-
-        \capt{Predicted vs. actual x drift} & \capt{Predicted
-          vs. actual y drift} & \capt{Integrated position}
-      \end{tabular}\\
-
-      \end{multicols}\vspace{-1em}
-
-    %%%%%%%%%%%%%%%%%%%%%%
-    \mbox{\hspace{0.3\linewidth}\rule{0.4\linewidth}{1pt}\hspace{0.3\linewidth}}\\
-
-    Although our algorithm was able to accurately predict drift, we
-    are still working on getting the prediction code to be fast enough
-    to control the quadrotor in live flight.  The SVM regression code
-    runs quickly, but we're currently facing difficulties with delays
-    due to I/O as well as a suboptimal configuration.\\
-
-  }%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-  \headerbox{Conclusion}{name=conclusion,column=2,span=1,below=results}{
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-    We are making strong progress towards predicting and removing
-    quadrotor drift.  The drift is predicted fairly accurately via a
-    60 dimensional sensor data vector from the quadrotor, and we hope
-    to soon demonstrate a controller enhanced by the predicted drift. \vspace{.5em}
   }%
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-  \headerbox{Abbreviated References}{name=references,column=0,above=bottom}{
+  \headerbox{Abbreviated References}{name=references,column=2,below=hyperneatGaits,above=bottom}{
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-    \footnotesize
+    \scriptsize
     \bibliography{references.bib}

### 2011-07-13 poster_gecco/abstract.tex
Added abstract

--- 2011-04-18 paper_gecco_2/abstract.tex
@@
 to access rugged terrain, yet designing such gaits by hand is a
-challenging and time-consuming process. In this paper we investigate
+time-consuming process. We investigate
 various algorithms for automating the creation of quadruped
 gaits. Because many robots do not have accurate simulators, we test
-gait learning algorithms entirely on a physical robot.  We compare the
+gait learning algorithms entirely on a physical robot, comparing the
 performance of two classes of learning gaits: locally searching

### 2011-07-13 poster_gecco/poster.tex
Beautified!

@@
     (CPPNs)~\cite{stanley2007compositional}, each of which is a genome
-    that encodes an ANN \pwfig{SpiderANN.jpg}{.7}{The ANN
-      configuration used to generate gaits
-      \cite{clune2011performance}.}\label{hyperneatLayout}
+    that encodes an ANN 
     phenotype~\cite{stanley2009hypercube}. Each CPPN is itself a
@@
 
+    \pfig{SpiderANN.jpg}{.8}{ANN
+      configuration to generate gaits
+      \cite{clune2011performance}.}\label{hyperneatLayout}
     For quadruped locomotion, HyperNEAT has previously produced gaits
@@
 
-%Many effective HyperNEAT gaits were not preserved across generations
-%because a single poor-performing trial could prevent their
-%selection. The HyperNEAT learning curve would be smoother if the noise
-%in the evaluations could be reduced or more than one evaluation per
-%individual could be afforded.
+\vspace{.5em}
 
-\pfig{hnResults}{.8}{Average fitness ($\pm $ SE) of the highest
+Many effective HyperNEAT gaits were not preserved across generations
+because a single poor-performing trial could prevent their
+selection. The HyperNEAT learning curve would be smoother if the noise
+in the evaluations could be reduced or more than one evaluation per
+individual could be afforded.
+
+\pfig{hnResults}{1}{Average fitness ($\pm $ SE) of the highest
   performing individual at each generation.}
+
+\textbf{Overall}
+\vspace{.5em}
 
@@
 \end{tabular}
+\vspace{.5em}\\
 \capt{Average and SD of the best gaits found
@@
     \scriptsize
+
     \bibliography{references.bib}
-    \bibliographystyle{is-plain}
+    %\bibliographystyle{is-plain}
+    \bibliographystyle{plain}
   }%

### 2011-07-26 code/Robot.py
Many updates grabbed from robot

@@ -281,6 +281,7 @@ class Robot():
             #[ac.read_all() for ac in self.actuators]
             #positions = ['%d: %s' % (ii,ac.cache[dynamixel.defs.REGISTER['CurrentPosition']]) for ii,ac in enumerate(self.actuators)]
             #print ' '.join(positions)
+            #print ''.join(['x' if ac.led else ' ' for ac in self.actuators])
                 
             #sleep(self.sleep)
             #sleep(float(1)/100)
@@ -345,9 +346,16 @@ class Robot():
         timeDiff  = datetime.now() - self.time0
         self.time = timeDiff.seconds + timeDiff.microseconds/1e6
 
-    def readyPosition(self):
-        self.commandPosition(POS_READY)
-        sleep(2)
+    def readyPosition(self, persist = False):
+        if persist:
+            self.resetClock()
+            while self.time < 2.0:
+                self.commandPosition(POS_READY)
+                sleep(.1)
+                self.updateClock()
+        else:
+            self.commandPosition(POS_READY)
+            sleep(2)
 
     def commandPosition(self, position, crop = True, cropWarning = False):
         '''Command the given position
@@ -380,6 +388,13 @@ class Robot():
         for ii,actuator in enumerate(self.actuators):
             actuator.goal_position = goalPosition[ii]
         self.net.synchronize()
+
+        #[ac.read_all() for ac in self.actuators]
+        #positions = ['%d: %s' % (ii,ac.cache[dynamixel.defs.REGISTER['CurrentPosition']]) for ii,ac in enumerate(self.actuators)]
+        #print ' '.join(positions)
+        #print ''.join(['x' if ac.led else ' ' for ac in self.actuators]) + '  ' ,
+        #print ' '.join(['%.1f' % ac.current_voltage for ac in self.actuators])
+
         return goalPosition
     
 

### 2011-07-26 code/justWalk.py

@@ -13,19 +13,28 @@ from Strategy import *
 
 
 def main():
-    if len(sys.argv) > 1:
-        filtFile = sys.argv[1]
+    if len(sys.argv) > 2 and sys.argv[1] == '-filt':
+        filtFile = sys.argv[2]
+        strategy = FileStrategy(filtFile = filtFile)
+        motionFunction, logInfo = strategy.getNext()
+    elif len(sys.argv) > 2 and sys.argv[1] == '-sine':
+        sineModel5Params = [eval(xx) for xx in sys.argv[2].split()]
+        print 'Using SineModel5 with params: ', sineModel5Params
+        motionFunction = lambda time: SineModel5().model(time,
+                                                         parameters = sineModel5Params)
     else:
         #filtFile = 'out/hyperneatTo20gens_101/neat_110115_175446_00014_008_filt'
         filtFile = 'out/hyperneatTo20gens_101/neat_110115_175446_00004_007_filt'
-
+        strategy = FileStrategy(filtFile = filtFile)
+        motionFunction, logInfo = strategy.getNext()
+    
 
     
-    strategy = FileStrategy(filtFile = filtFile)
     #runman = RunManager()
     #runman.do_many_runs(strategy, SineModel5.typicalRanges)
 
-    motionFunction, logInfo = strategy.getNext()
+    timeScale = .3
+    motionFunctionScaled = scaleTime(motionFunction, timeScale)
 
     robot = Robot()
     robot.run(motionFunction, runSeconds = 8, resetFirst = False,

### 2011-07-26 paper_ecal/fitnessEvaluation.tex

@@
 As mentioned earlier, the metric for evaluating gaits was the Euclidian distance the robot
-moved during a 12-second run. For the manual and parameterized gaits,
+moved during a 12-second run on flat terrain. For the manual and parameterized gaits,
 the fitness was this value. The HyperNEAT gaits stressed the motors
@@
 position and the commanded gait, $g(t)$.  As shown in
-\figref{linearInterp}, motion for the first second was interpolated
-between the ready position and $g(t)$, for the next nine seconds the
-robot moved exactly as specified by $g(t)$, and for the final two
-seconds its motion was interpolated smoothly back to the ready
-position.
+%\figref{linearInterp}, motion for the first second was interpolated
+%between the ready position and $g(t)$, for the next nine seconds the
+%robot moved exactly as specified by $g(t)$, and for the final two
+%seconds its motion was interpolated smoothly back to the ready
+%position.
+\figref{linearInterp}, the instantaneous robot limb configuration
+during the first and last portions of the evaluation was an
+interpolation between the initial ready position and $g(t)$;
+during the rest of the evaluation, the robot followed the commanded
+gait exactly.
 
@@
 was negligible at first, but became more pronounced as gaits began traversing the entire
-area.
-HyperNEAT gaits were especially likely to require additional trials, meaning that the reported performance for HyperNEAT is worse than it would have been otherwise.
-For future studies we will use a $2\times2$ array of Wii
-remotes to double the size of the measurement arena.
+area.  HyperNEAT gaits were especially likely to require additional
+trials, meaning that the reported performance for HyperNEAT is worse
+than it would have been otherwise. Future studies could employ an
+array of Wii remotes to increase the size of the measurement arena.

### 2011-07-26 poster_gecco/poster.tex

@@
     % depicted above.
-    The robot is actuated by 9 AX-12+ Dynamixel servos: 8 for the legs and one at the center ``hip'' joint.
+    The robot is actuated by nine AX-12+ Dynamixel servos: eight for the legs and one at the center ``hip'' joint.
     %This final unique servo allows
@@
 
-\pfigg{wiiMote_crop.jpg}{robot_led_crop.jpg}{1}{A Nintendo Wii
+\pfigg{wiiMote_crop.jpg}{robot_led_crop.jpg}{1}{\\A Nintendo Wii
   remote tracked an IR LED on the robot, transmitting the (x,y)
@@
 
-The fitness was the Euclidian distance the robot moved during a
-12-second run on flat terrain.  To encourage gaits that did not tax
-the motors, we penalized gaits that caused the servos to stop
+    Gait fitness was the Euclidian distance the robot moved during a
+    12-second run on flat terrain.  To encourage gaits that did not
+    tax the motors, we penalized gaits that caused the servos to stop
     responding by a factor of 1/2.
@@
     The robot was always measured while in the \emph{ready} position
-to ensure fair fitness measurements. As
-shown above the instantaneous robot limb configuration during the
-first and last portions of the evaluation was an interpolation between
-the initial ready position and $g(t)$.
+    to ensure fair fitness measurements. As shown above, the
+    instantaneous robot limb configuration during the first and last
+    portions of the evaluation was an interpolation between the
+    initial ready position and $g(t)$.
 
@@
 
-    \pfig{neat_110115_211410_00000_002_filt_zoom}{1}{Joint angles for
+    \pfig{neat_110115_211410_00000_002_filt_zoom}{1}{Joint angles over time for
       a high-performance HyperNEAT gait.}
@@
 parameterized gaits in three separate runs (200 random, 234 uniform, 284 Gaussian, 174
-gradient, 172 simplex, 153 linear regression). The SVM-modeled method
+gradient, 172 simplex, 153 linear regression). The SVM-based self-modeling method
 was never successful.
@@
 Evaluation of HyperNEAT gaits was very noisy, which made learning
-difficult. We tested one gait 11 times and found performance of 26
-$\pm$ 13 body lengths/minute, with a max of 38 and a min of 3.
+difficult. For example, we tested one gait 11 times and found an average performance of 26
+ body lengths/minute, $\pm$13 std. dev., with a max of 38 and a min of 3.
 
@@
 
-\textbf{Overall}
+\textbf{Summary of Results}
 \vspace{.5em}

### 2011-07-26 code/Robot.py
Pulling changes from robot

@@ -483,3 +483,4 @@ class Robot():
             print bb - posBB
         return success
         
+

### 2011-07-26 code/optimize.py

@@ -53,14 +53,14 @@ def doRun():
         strategy
     except:
         #strategy = UniformStrategy(currentState, SineModel5.typicalRanges)
-        #strategy = GaussianStrategy(currentState, SineModel5.typicalRanges)
+        strategy = GaussianStrategy(currentState, SineModel5.typicalRanges)
         #strategy = GradientSampleStrategy(currentState)
         #strategy = LinearRegressionStrategy(currentState)
         #strategy = SimplexStrategy(currentState, SineModel5.typicalRanges)
         #strategy = RandomStrategy(currentState, SineModel5.typicalRanges)
         #strategy = SVMLearningStrategy(currentState, SineModel5.typicalRanges)
         #strategy = NEATStrategy(currentState, SineModel5.typicalRanges, neatFile = neatFile)   # these args aren't used
-        strategy = FileStrategy(filtFile = filtFile)
+        #strategy = FileStrategy(filtFile = filtFile)
         
     #runman.do_many_runs(currentState, lambda state: Neighbor.gaussian(SineModel5.typicalRanges, state))
     #runman.do_many_runs(currentState, lambda state: gradient_search(SineModel5.typicalRanges, state))

### 2011-07-26 code/Robot.py
minor, working with Kyrre

@@ -392,8 +392,8 @@ class Robot():
         #[ac.read_all() for ac in self.actuators]
         #positions = ['%d: %s' % (ii,ac.cache[dynamixel.defs.REGISTER['CurrentPosition']]) for ii,ac in enumerate(self.actuators)]
         #print ' '.join(positions)
-        #print ''.join(['x' if ac.led else ' ' for ac in self.actuators]) + '  ' ,
-        #print ' '.join(['%.1f' % ac.current_voltage for ac in self.actuators])
+        print ''.join(['x' if ac.led else ' ' for ac in self.actuators]) + '  ' ,
+        print ' '.join(['%.1f' % ac.current_voltage for ac in self.actuators])
 
         return goalPosition
     

### 2011-08-11 paper_scratch/abstract.tex
Added a scratch version to produced reordered table for ECAL presentation

--- 2011-02-10 paper/abstract.tex
@@
 gaits. Because many robots do not have accurate simulators, we test
-gait learning algorithms entirely on a physical robot.  We compare the
-performance of two classes of learning gaits: locally searching
+gait-learning algorithms entirely on a physical robot.  We compare the
+performance of two classes of gait-learning algorithms: locally searching
 parameterized motion models and evolving artificial neural networks

### 2011-08-11 paper_scratch/introduction.tex
Added a scratch version to produced reordered table for ECAL presentation

--- 2011-04-15 paper_ecal/introduction.tex
@@
   were evolved. The translucent parts were produced by a 3D
-  printer. Videos of the gaits can be viewed at http://bit.ly/hLHBYq}
+  printer. Videos of the gaits can be viewed at http://bit.ly/ecalgait}
 
@@
 
-An additional motivation is to test whether gaits evolved in simulation transfer to reality well, especially those evolved with cutting-edge evolutionary algorithms.
-Because HyperNEAT gaits performed well in
+An additional motivation is to test whether techniques for evolving
+gaits in simulation, especially cutting-edge evolutionary algorithms,
+transfer to reality well. Because HyperNEAT gaits performed well in
 simulation, it is interesting to test whether HyperNEAT can produce fast

### 2011-08-11 paper_scratch/paperOutline.tex
Added a scratch version to produced reordered table for ECAL presentation

\subsection{Outline of Sections}

% The remainder of the paper is organized as follows. In
% \secref{problemDefinition} we define more rigorously the problem of
% gait learning. \secref{experimentalSetup} describes our experimental
% setup, including the hardware we used and our methods for evaluating
% fitness of a gait. \secref{gaitGenLearn} discusses the different gait
% generation and learning methods we tested, and \secref{results}
% presents and discusses performance results.  Finally,
% \secref{conclusion} concludes and offers possible future extensions to this
% work.

The remainder of the paper is organized as follows. We begin by giving
a more rigorous \emph{Problem Definition}, then describe our
\emph{Experimental Setup}, including the hardware we used and our
methods for evaluating fitness of a gait. We then discuss methods used
for \emph{Gait Generation and Learning} and then present and discuss
\emph{Results}.  Finally, we offer several \emph{Conclusions and
  Future Work}.

### 2011-08-11 paper_scratch/results.tex
Added a scratch version to produced reordered table for ECAL presentation

--- 2011-04-14 paper_ecal/results.tex
@@
 \caption{The average and standard deviation of the best gaits found
-  for each algorithm during each of three runs, in body lengths per
-  minute.}  \tablabel{results}
+  for each algorithm during each of three runs, in body
+  lengths/minute.}  \tablabel{results}
 \end{center}
@@
 advantageous.  The single best gait found during this study had a
-speed of 45.72 body lengths / minute, 66\% better than the best
+speed of 45.72 body lengths/minute, 66\% better than the best
 non-HyperNEAT gait and 8.9 times faster than the hand-coded gait.
@@
 example, we tested an example HyperNEAT generation-champion gait 11
-times and found that its mean performance was 26 body lengths per
-minute ($\pm$ 13 SD), but it had a max of 38 and a min of 3.  Many
+times and found that its mean performance was 26 body lengths/minute
+($\pm$ 13 SD), but it had a max of 38 and a min of 3.  Many
 effective HyperNEAT gaits were not preserved across generations
@@
 
-\acmFig{hnResults}{1}{Average fitness ($\pm $ SE) of the highest
+\acmFig{hnResults}{.8}{Average fitness ($\pm $ SE) of the highest
   performing individual in the population for each generation of

### 2012-02-16 code/plot/plot_angles.m
Plotting changes from a while back

@@ -14,3 +14,64 @@ xlabel('Time (s)',       'fontsize', 22);
 ylabel('Motor Position', 'fontsize', 22);
 print('-depsc2', '-F:18', 'neat_110115_211410_00000_002_filt_zoom.eps');
 
+
+
+
+figure(3);
+tt = linspace(0, 12, size(dat,1) * .6);
+synth = (ones(9, 1) * 150*sin(tt*1))';
+synth(:,[1,3,5,7]) = synth(:,[1,3,5,7]) * 1.6;
+synth(:,[1,2,3,4]) = synth(:,[1,2,3,4]) * -.78;
+synth(:,[1,2,5,6]) = synth(:,[1,2,5,6]) * 1.9;
+synth(:,[1,3,5,7]) = synth(:,[1,3,5,7]) + 800;   % inner
+synth(:,[2,4,6,8]) = synth(:,[2,4,6,8]) + 40;    % outer
+synth(:,[9]) = synth(:,[9]) * 0 + 512;           % center
+
+%synth(:,[1,3,5,7]) = min(synth(:,[1,3,5,7]), (60+120)*1024/240);   % inner
+%synth(:,[1,3,5,7]) = max(synth(:,[1,3,5,7]), (-85+120)*1024/240);  % inner
+%synth(:,[2,4,6,8]) = min(synth(:,[2,4,6,8]), (39+120)*1024/240);   % outer
+%synth(:,[2,4,6,8]) = max(synth(:,[2,4,6,8]), (-113+120)*1024/240); % outer
+%synth(:,[9]) = min(synth(:,[9]), (39+120)*1024/240);               % center
+%synth(:,[9]) = max(synth(:,[9]), (-113+120)*1024/240);             % center
+
+synth(:,[1,3,5,7]) = min(synth(:,[1,3,5,7]), 770);   % inner
+synth(:,[1,3,5,7]) = max(synth(:,[1,3,5,7]), 150);  % inner
+synth(:,[2,4,6,8]) = min(synth(:,[2,4,6,8]), 680);   % outer
+synth(:,[2,4,6,8]) = max(synth(:,[2,4,6,8]), 30); % outer
+synth(:,[9]) = min(synth(:,[9]), 623);               % center
+synth(:,[9]) = max(synth(:,[9]), 392);             % center
+
+
+plot(tt, synth, 'linewidth', 4);
+xlabel('Time (s)',       'fontsize', 22);
+ylabel('Motor Position', 'fontsize', 22);
[…]

### 2012-03-12 paper_charlotte/main.tex
Pre-edit version of rex/charlotte paper

\documentclass[letterpaper]{article}
\usepackage{natbib,alife13}

\title{An Open-Source Quadruped Platform for Evolutionary Robotics}
\author{Sara Lohmann$^{1}$, Eric Gold$^{1}$, Jeremy Blum$^{1}$, Jason Yosinski$^{1}$ \and Hod Lipson$^{1}$ \\
\mbox{}\\
$^1$Cornell University, 239 Upson Hall, Ithaca, NY 14853 \\
sml253@cornell.edu}

\begin{document}
\maketitle
\begin{figure}[t]
\begin{center}
\includegraphics[width=2.25in,angle=0]{fig1.jpg}
\caption{QuadraRex: an open-sourced quadruped robot platform. All instructions and downloads are publicly available at http://creativemachines.cornell.edu/quadrarex \citep{WEB}.}
\label{fig1}
\end{center}
\end{figure}

\begin{abstract}
We report on a new open-source quadruped robotic platform, QuadraRex, for evolutionary robotics experiments. The robot comprises four legs with two joints each, for a total of eight kinematic degrees of freedom. Four-bar linkage mechanisms in each leg drive the
pitch of the knee joint and the hip joint remotely, allowing the motors to remain in the robot center, thereby reducing the inertia of each leg. Due to these non-conventional kinematics, the robot requires non intuitive movements for locomotion that are challenging for learning. We suggest that this platform can provide a low-cost, accessible platform for testing and comparing physical evolutionary robotics algorithms.
\end{abstract}

\section{Introduction}
We address the need for a low-cost platform with non-intuitive walking kinematics to explore evolutionary robotics. QuadraRex is the third evolutionary learning quadruped robot developed by the Creative Machines Lab \citep{HL, JY}. A common feature among the quadruped robots is two actuators that drive the flexion/extension of a knee and hip joint around parallel axes in each leg. The original Creative Machines Lab quadruped robot favored starfish-like movements \citep{HL}. The second quadruped robot, QuadraTot \citep{JY}, developed spider-like movements, but was limited by its weight. When creating the QuadraRex, we designed the hardware to compliment fast spider-like movements.

\section{Hardware}
\begin{figure}[t]
\begin{center}
\includegraphics[width=2.25in,angle=0]{fig3.pdf}
\caption{Crank-rocker 4 bar linkage controlling flexion/extension of the knee joint.}
\label{fig3}
\end{center}
\end{figure}
\begin{figure}[t]
\begin{center}
\includegraphics[width=2.25in,angle=0]{fig4.pdf}
\caption{Crank-rocker 4 bar linkage controlling flexion/extension of the hip joint.}
\label{fig4}
\end{center}
\end{figure}
The hardware of QuadraRex evolved from the previous Creative Machines Lab quadruped robots. We kept the same two degree of freedom pitch joint scheme but decreased the weight and constrained the movement of the joints to create faster spider-like movement. To prevent starfish-like movement, the legs were constrained. We designed the robot with two four bar mechanisms to drive the joints in each leg. With a four bar mechanism in place, the leg moves at a fraction of the output angle of the actuator. Figure~\ref{fig3} and Figure~\ref{fig4} show the crank-rocker system, where the input crank (link OA) is actuated by a servo, the rocker is the leg (link CB), and the fixed link is OC. In this configuration, we keep the servo motors contained in the center of the robot, thereby reducing the inertia of each leg. The four bar mechanisms satisfied the design goal of making a robot that had non-traditional movements. We avoided designing a robot similar to an off-the-shelf robot to create something unique. To make the robot accessible as an open-source platform, it was designed to be 3D printed. The 3D print files and CAD files can be accessed by the public to be modified and printed.
\\
We designed the robot to be lightweight to focus on the goal of evolutionary robotics. We use a single LiPo 11.1V battery to power 8 Dynamixel AX-18 servo motors and an ArbotiX microcontroller. The main processing occurs on an external computer, which allows the robot to have cheaper and lighter components on-board. We use the space in the center of the body to house the battery to reduce material weight and cost. QuadraRex weighs less than 1 kg.
\begin{figure}[t]
\begin{center}
\includegraphics[width=2.25in,angle=0]{fig5.jpg}
\caption{Renderd CAD model.}
\label{fi52}
\end{center}
\end{figure}

\section{Software}
The software is open-sourced code written in Python based on the evolutionary learning algorithms of the QuadraTot \citep{JY}. We provide the controller code on our website \citep{WEB}. We use an external computer for processing. We use an infrared light emitting diode on the robot along with an external Wii remote to provide feedback of the distance traveled. The external computer receives feedback from the Wii remote and internal servo sensors and then processes the information to send the next command to position the servos. The communication between the external computer and the ArbotiX on-board microcontroller occurs over wireless XBee.

\section{Specifications}
\begin{figure}[t]
\begin{center}
\includegraphics[width=2.25in,angle=0]{fig2.jpg}
\caption{Robot printed as one piece with support material intact.}
\label{fig2}
\end{center}
\end{figure}
We 3D printed the robot as one part (Figure~\ref{fig2}). The robot was completed after approximately 26 hours, and used approximately 1000 g of model material and 1500 g of support material. We estimated the cost of the printed parts to be \$410 by using the estimated model material cost to be \$1/4.5g and the support material cost to be \$1/8g. This cost estimate is highly variable depending on the type of material used. The robot has no removable printed parts except for the battery cover. We will make the robot available to print as separate parts for smaller 3D printer tray sizes. Printing the robot as separate parts will reduce the amount of support material and cost to print. We estimated the total cost to be less than \$1500 for the custom robot platform. We made these estimates based on the material calculations above and prices from trossenrobotics.com.

\begin{table}[h]
\center{
\begin{tabular}{|c|c|}
\hline
Part & Cost\\
\hline\hline
3D Print Materials & \$410.00\\
ArbotiX Robocontroller Kit & \$189.00\\
Dynamixel AX-18A Robot Actuator (x8) & \$721.28\\
3S 11.1V 2000mAh Pro Lite LiPo Battery & \$72.99\\
LiPo Battery Balance Charger Kit & \$69.99\\
Cables, Connectors, Misc & \$27.78\\
\hline\hline
\bf Total & \bf \$1491.04\\
 \hline
\end{tabular}
}
\vskip 0.25cm
\caption{Estimated total cost. A specifiic parts list is on our website \citep{WEB}}
\end{table}

\section{Acknowledgements}
This work was supported by

\footnotesize
\bibliographystyle{apalike}
\bibliography{QuadraRex}
\end{document}

### 2012-03-12 paper_charlotte/main.tex
chagnes

@@
 
-\title{An Open-Source Quadruped Platform for Evolutionary Robotics}
-\author{Sara Lohmann$^{1}$, Eric Gold$^{1}$, Jeremy Blum$^{1}$, Jason Yosinski$^{1}$ \and Hod Lipson$^{1}$ \\
+\title{QuadraRex: An Open-Source Quadruped Platform for Evolutionary Robotics}
+\author{Sara Lohmann, Eric Gold, Jeremy Blum, Jason Yosinski \and Hod Lipson \\
 \mbox{}\\
-$^1$Cornell University, 239 Upson Hall, Ithaca, NY 14853 \\
-sml253@cornell.edu}
+Cornell University, 239 Upson Hall, Ithaca, NY 14853 \\
+\texttt{sml253@cornell.edu}}
 
@@
 \maketitle
-\begin{figure}[t]
-\begin{center}
-\includegraphics[width=2.25in,angle=0]{fig1.jpg}
-\caption{QuadraRex: an open-sourced quadruped robot platform. All instructions and downloads are publicly available at http://creativemachines.cornell.edu/quadrarex \citep{WEB}.}
-\label{fig1}
-\end{center}
-\end{figure}
 
@@
 \section{Introduction}
+
 We address the need for a low-cost platform with non-intuitive walking kinematics to explore evolutionary robotics.
@@
 
-\section{Hardware}
 \begin{figure}[t]
 \begin{center}
-\includegraphics[width=2.25in,angle=0]{fig3.pdf}
-\caption{Crank-rocker 4 bar linkage controlling flexion/extension of the knee joint.}
+\includegraphics[width=.45\textwidth]{fig1.jpg}
+\caption{QuadraRex: an open-sourced quadruped robot platform. All instructions and downloads are publicly available at http://creativemachines.cornell.edu/quadrarex \citep{WEB}.}
+\label{fig1}
+\end{center}
+\end{figure}
+
+\section{Hardware}
+
+\begin{figure}[t]
+\begin{center}
+\includegraphics[width=.23\textwidth]{fig3.pdf}
+\includegraphics[width=.23\textwidth]{fig4.pdf}
+\caption{Crank-rocker 4 bar linkage controlling flexion/extension of the knee and hip joints.}
 \label{fig3}
@@
 \end{figure}
-\begin{figure}[t]
-\begin{center}
-\includegraphics[width=2.25in,angle=0]{fig4.pdf}
-\caption{Crank-rocker 4 bar linkage controlling flexion/extension of the hip joint.}
-\label{fig4}
-\end{center}
-\end{figure}
+
+%\begin{figure}[t]
+%\begin{center}
+%\includegraphics[width=2.25in,angle=0]{fig4.pdf}
+%\caption{Crank-rocker 4 bar linkage controlling flexion/extension of the hip joint.}
+%\label{fig4}
+%\end{center}
+%\end{figure}
+
 The hardware of QuadraRex evolved from the previous Creative Machines Lab quadruped robots.
@@
 With a four bar mechanism in place, the leg moves at a fraction of the output angle of the actuator.
-Figure~\ref{fig3} and Figure~\ref{fig4} show the crank-rocker system, where the input crank (link OA) is actuated by a servo, the rocker is the leg (link CB), and the fixed link is OC.
+Figure~\ref{fig3} shows the crank-rocker system, where the input crank (link OA) is actuated by a servo, the rocker is the leg (link CB), and the fixed link is OC.
 In this configuration, we keep the servo motors contained in the center of the robot, thereby reducing the inertia of each leg.
@@
 \vskip 0.25cm
-\caption{Estimated total cost. A specifiic parts list is on our website \citep{WEB}}
+\caption{Estimated total cost. A specifiic parts list is on our website \citep{WEB}.}
 \end{table}
@@
 \section{Acknowledgements}
+
 This work was supported by
+%NSF Creative-IT grant 0757478.
+the National Science Foundation's Office of Emerging Frontiers in Research and Innovation (grant number 0735953).
 

### 2012-03-13 paper_charlotte/main.tex
spacing

@@
 \begin{abstract}
-We report on a new open-source quadruped robotic platform, QuadraRex, for evolutionary robotics experiments.
-The robot comprises four legs with two joints each, for a total of eight kinematic degrees of freedom.
-Four-bar linkage mechanisms in each leg drive the
-pitch of the knee joint and the hip joint remotely, allowing the motors to remain in the robot center, thereby reducing the inertia of each leg.
-Due to these non-conventional kinematics, the robot requires non intuitive movements for locomotion that are challenging for learning.
-We suggest that this platform can provide a low-cost, accessible platform for testing and comparing physical evolutionary robotics algorithms.
+We report on a new open-source quadruped robotic platform, QuadraRex,
+for evolutionary robotics experiments. The robot comprises four legs
+with two joints each, for a total of eight kinematic degrees of
+freedom. Four-bar linkage mechanisms in each leg drive the pitch of
+the knee joint and the hip joint remotely, allowing the motors to
+remain in the robot center, thereby reducing the inertia of each
+leg. Due to these non-conventional kinematics, the robot requires non
+intuitive movements for locomotion that are challenging for
+learning. We suggest that this platform can provide a low-cost,
+accessible platform for testing and comparing physical evolutionary
+robotics algorithms.
 \end{abstract}
@@
 
-We address the need for a low-cost platform with non-intuitive walking kinematics to explore evolutionary robotics.
-QuadraRex is the third evolutionary learning quadruped robot developed by the Creative Machines Lab \citep{HL, JY}.
-A common feature among the quadruped robots is two actuators that drive the flexion/extension of a knee and hip joint around parallel axes in each leg.
-The original Creative Machines Lab quadruped robot favored starfish-like movements \citep{HL}.
-The second quadruped robot, QuadraTot \citep{JY}, developed spider-like movements, but was limited by its weight.
-When creating the QuadraRex, we designed the hardware to compliment fast spider-like movements.
+We address the need for a low-cost platform with non-intuitive walking
+kinematics to explore evolutionary robotics. QuadraRex is the third
+evolutionary learning quadruped robot developed by the Creative
+Machines Lab \citep{HL, JY}. A common feature among the quadruped
+robots is two actuators that drive the flexion/extension of a knee and
+hip joint around parallel axes in each leg. The original Creative
+Machines Lab quadruped robot favored starfish-like movements
+\citep{HL}. The second quadruped robot, QuadraTot \citep{JY},
+developed spider-like movements, but was limited by its weight. When
+creating the QuadraRex, we designed the hardware to compliment fast
+spider-like movements.
 
@@
 \includegraphics[width=.45\textwidth]{fig1.jpg}
-\caption{QuadraRex: an open-sourced quadruped robot platform. All instructions and downloads are publicly available at http://creativemachines.cornell.edu/quadrarex \citep{WEB}.}
+\caption{QuadraRex: an open-sourced quadruped robot platform. All
+  instructions and downloads are publicly available at
+  http://creativemachines.cornell.edu/quadrarex \citep{WEB}.}
 \label{fig1}
@@
 \includegraphics[width=.23\textwidth]{fig4.pdf}
-\caption{Crank-rocker 4 bar linkage controlling flexion/extension of the knee and hip joints.}
+\caption{Crank-rocker 4 bar linkage controlling flexion/extension of
+  the knee and hip joints.}
 \label{fig3}
@@
 
-The hardware of QuadraRex evolved from the previous Creative Machines Lab quadruped robots.
-We kept the same two degree of freedom pitch joint scheme but decreased the weight and constrained the movement of the joints to create faster spider-like movement.
-To prevent starfish-like movement, the legs were constrained.
-We designed the robot with two four bar mechanisms to drive the joints in each leg.
-With a four bar mechanism in place, the leg moves at a fraction of the output angle of the actuator.
-Figure~\ref{fig3} shows the crank-rocker system, where the input crank (link OA) is actuated by a servo, the rocker is the leg (link CB), and the fixed link is OC.
-In this configuration, we keep the servo motors contained in the center of the robot, thereby reducing the inertia of each leg.
-The four bar mechanisms satisfied the design goal of making a robot that had non-traditional movements.
-We avoided designing a robot similar to an off-the-shelf robot to create something unique.
-To make the robot accessible as an open-source platform, it was designed to be 3D printed.
-The 3D print files and CAD files can be accessed by the public to be modified and printed.
-\\
-We designed the robot to be lightweight to focus on the goal of evolutionary robotics.
-We use a single LiPo 11.1V battery to power 8 Dynamixel AX-18 servo motors and an ArbotiX microcontroller.
-The main processing occurs on an external computer, which allows the robot to have cheaper and lighter components on-board.
-We use the space in the center of the body to house the battery to reduce material weight and cost.
-QuadraRex weighs less than 1 kg.
+The hardware of QuadraRex evolved from the previous Creative Machines
+Lab quadruped robots. We kept the same two degree of freedom pitch
+joint scheme but decreased the weight and constrained the movement of
+the joints to create faster spider-like movement. To prevent
+starfish-like movement, the legs were constrained. We designed the
+robot with two four bar mechanisms to drive the joints in each
+leg. With a four bar mechanism in place, the leg moves at a fraction
+of the output angle of the actuator. Figure~\ref{fig3} shows the
+crank-rocker system, where the input crank (link OA) is actuated by a
+servo, the rocker is the leg (link CB), and the fixed link is OC. In
+this configuration, we keep the servo motors contained in the center
+of the robot, thereby reducing the inertia of each leg. The four bar
+mechanisms satisfied the design goal of making a robot that had
+non-traditional movements. We avoided designing a robot similar to an
+off-the-shelf robot to create something unique. To make the robot
+accessible as an open-source platform, it was designed to be 3D
+printed. The 3D print files and CAD files can be accessed by the
+public to be modified and printed.  \\ We designed the robot to be
+lightweight to focus on the goal of evolutionary robotics. We use a
+single LiPo 11.1V battery to power 8 Dynamixel AX-18 servo motors and
+an ArbotiX microcontroller. The main processing occurs on an external
+computer, which allows the robot to have cheaper and lighter
+components on-board. We use the space in the center of the body to
+house the battery to reduce material weight and cost. QuadraRex weighs
+less than 1 kg.
+
 \begin{figure}[t]
@@
 \section{Software}
-The software is open-sourced code written in Python based on the evolutionary learning algorithms of the QuadraTot \citep{JY}.
-We provide the controller code on our website \citep{WEB}.
-We use an external computer for processing.
-We use an infrared light emitting diode on the robot along with an external Wii remote to provide feedback of the distance traveled.
-The external computer receives feedback from the Wii remote and internal servo sensors and then processes the information to send the next command to position the servos.
-The communication between the external computer and the ArbotiX on-board microcontroller occurs over wireless XBee.
+
+The software is open-sourced code written in Python based on the
+evolutionary learning algorithms of the QuadraTot \citep{JY}. We
+provide the controller code on our website \citep{WEB}. We use an
+external computer for processing. We use an infrared light emitting
+diode on the robot along with an external Wii remote to provide
+feedback of the distance traveled. The external computer receives
+feedback from the Wii remote and internal servo sensors and then
+processes the information to send the next command to position the
+servos. The communication between the external computer and the
+ArbotiX on-board microcontroller occurs over wireless XBee.
 
 \section{Specifications}
+
 \begin{figure}[t]
@@
 \end{figure}
-We 3D printed the robot as one part (Figure~\ref{fig2}).
-The robot was completed after approximately 26 hours, and used approximately 1000 g of model material and 1500 g of support material.
-We estimated the cost of the printed parts to be \$410 by using the estimated model material cost to be \$1/4.5g and the support material cost to be \$1/8g.
-This cost estimate is highly variable depending on the type of material used.
-The robot has no removable printed parts except for the battery cover.
-We will make the robot available to print as separate parts for smaller 3D printer tray sizes.
-Printing the robot as separate parts will reduce the amount of support material and cost to print.
-We estimated the total cost to be less than \$1500 for the custom robot platform.
-We made these estimates based on the material calculations above and prices from trossenrobotics.com.
+
+We 3D printed the robot as one part (Figure~\ref{fig2}). The robot was
+completed after approximately 26 hours, and used approximately 1000 g
+of model material and 1500 g of support material. We estimated the
+cost of the printed parts to be \$410 by using the estimated model
+material cost to be \$1/4.5g and the support material cost to be
+\$1/8g. This cost estimate is highly variable depending on the type of
+material used. The robot has no removable printed parts except for the
+battery cover. We will make the robot available to print as separate
+parts for smaller 3D printer tray sizes. Printing the robot as
+separate parts will reduce the amount of support material and cost to
+print. We estimated the total cost to be less than \$1500 for the
+custom robot platform. We made these estimates based on the material
+calculations above and prices from trossenrobotics.com.
 
@@
 
-This work was supported by
-%NSF Creative-IT grant 0757478.
-the National Science Foundation's Office of Emerging Frontiers in Research and Innovation (grant number 0735953).
+This work was supported by the National Science Foundation's Office of
+Emerging Frontiers in Research and Innovation (grant number 0735953).
 

### 2012-03-13 paper_charlotte/main.tex
name change, ref fix, etc.

@@
 
-\title{QuadraRex: An Open-Source Quadruped Platform for Evolutionary Robotics}
+\title{Aracna: An Open-Source Quadruped Platform for Evolutionary Robotics}
 \author{Sara Lohmann, Eric Gold, Jeremy Blum, Jason Yosinski \and Hod Lipson \\
@@
 \begin{abstract}
-We report on a new open-source quadruped robotic platform, QuadraRex,
+We report on a new open-source quadruped robotic platform, Aracna,
 for evolutionary robotics experiments. The robot comprises four legs
@@
 We address the need for a low-cost platform with non-intuitive walking
-kinematics to explore evolutionary robotics. QuadraRex is the third
+kinematics to explore evolutionary robotics. Aracna is the third
 evolutionary learning quadruped robot developed by the Creative
@@
 developed spider-like movements, but was limited by its weight. When
-creating the QuadraRex, we designed the hardware to compliment fast
+creating the Aracna, we designed the hardware to compliment fast
 spider-like movements.
@@
 \includegraphics[width=.45\textwidth]{fig1.jpg}
-\caption{QuadraRex: an open-sourced quadruped robot platform. All
+\caption{Aracna: an open-sourced quadruped robot platform. All
   instructions and downloads are publicly available at
@@
 
-The hardware of QuadraRex evolved from the previous Creative Machines
+The hardware of Aracna evolved from the previous Creative Machines
 Lab quadruped robots. We kept the same two degree of freedom pitch
@@
 components on-board. We use the space in the center of the body to
-house the battery to reduce material weight and cost. QuadraRex weighs
+house the battery to reduce material weight and cost. Aracna weighs
 less than 1 kg.
@@
 \vskip 0.25cm
-\caption{Estimated total cost. A specifiic parts list is on our website \citep{WEB}.}
+\caption{Estimated total cost. A specific parts list is on our website \citep{WEB}.}
 \end{table}
@@
 \bibliographystyle{apalike}
-\bibliography{QuadraRex}
+\bibliography{references}
 \end{document}

### 2012-03-13 paper_aracna/main.tex
moved

\documentclass[letterpaper]{article}
\usepackage{natbib,alife13}

\title{Aracna: An Open-Source Quadruped Platform for Evolutionary Robotics}
\author{Sara Lohmann, Eric Gold, Jeremy Blum, Jason Yosinski \and Hod Lipson \\
\mbox{}\\
Cornell University, 239 Upson Hall, Ithaca, NY 14853 \\
\texttt{sml253@cornell.edu}}

\begin{document}
\maketitle

\begin{abstract}
We report on a new open-source quadruped robotic platform, Aracna,
developed at Cornell's Creative Machines Lab for robotics
experiments. The robot comprises four legs with two joints each for a
total of eight kinematic degrees of freedom. Four-bar linkage
mechanisms in each leg drive the pitch of the knee joint and the hip
joint remotely, allowing the motors to remain in the robot center,
which reduces the inertia of each leg. Due to these non-conventional
kinematics, the robot requires non-intuitive movements for locomotion
and provides an interesting challenge for gait learning
algorithms. Aracna is a low-cost, accessible robotic platform for
testing in the real world.
\end{abstract}

\section{Introduction}

We address the need for a low-cost platform with non-intuitive walking
kinematics to explore evolutionary robotics. Aracna is the third
evolutionary learning quadruped robot developed by the Creative
Machines Lab \citep{HL, JY}. A common feature among the quadruped
robots is two actuators that drive the flexion/extension of a knee and
hip joint around parallel axes in each leg. The original Creative
Machines Lab quadruped robot favored starfish-like movements
\citep{HL}. The second quadruped robot --- QuadraTot ---
developed spider-like movements but was found to be limited by its
weight \citep{JY}. When creating the Aracna, we designed the hardware to
complement fast spider-like movements.

\begin{figure}[t]
\begin{center}
\includegraphics[width=.45\textwidth]{fig1.jpg}
\caption{Aracna: an open-sourced quadruped robot platform. All
  instructions and downloads are publicly available at
  http://creativemachines.cornell.edu/aracna .}
\label{fig1}
\end{center}
\end{figure}

\section{Hardware}

\begin{figure}[t]
\begin{center}
\includegraphics[width=.23\textwidth]{fig3.pdf}
\includegraphics[width=.23\textwidth]{fig4.pdf}
\caption{Crank-rocker four-bar linkage controlling flexion/extension of
  the knee and hip joints. In both cases above, the input crank (link
  OA) is actuated by a servo, the rocker is the leg (link CB), and the
  fixed link is OC.}
\label{fig3}
\end{center}
\end{figure}

%\begin{figure}[t]
%\begin{center}
%\includegraphics[width=2.25in,angle=0]{fig4.pdf}
%\caption{Crank-rocker four-bar linkage controlling flexion/extension of the hip joint.}
%\label{fig4}
%\end{center}
%\end{figure}

The hardware of Aracna evolved from the previous Creative Machines Lab
quadruped robots. We kept the same two degree of freedom pitch joint
scheme but decreased the weight and constrained the movement of the
joints toward the goal of creating faster spider-like movement. To
prevent starfish-like movement, the legs were constrained. We designed
the robot with two four-bar mechanisms to drive the joints in each
leg. With a four-bar mechanism in place, the leg moves at a fraction
of the output angle of the actuator, giving the motor a relatively
larger mechanical advantage over the position of each leg.
Figure~\ref{fig3} shows the crank-rocker system, where the input crank
is actuated by a servo and the rocker is the leg. In this
configuration, we keep the servo motors contained in the center of the
robot, reducing both the inertia and mass of each leg. The four-bar
mechanisms satisfied the design goal of making a robot that had
non-traditional movements.  To make the robot
accessible as an open-source platform, it was designed to be 3D
printed. The 3D print files and CAD files can be accessed by the
public to be printed and, if desired, modified.

The robot is designed to be lightweight, because previous similar
robots may have been too heavy and underpowered to generate
interesting gaits. We use a single LiPo 11.1V battery to
power 8 Dynamixel AX-18 servo motors and an ArbotiX
microcontroller. The main processing occurs on an external computer,
which allows the robot to have cheaper and lighter components
on-board. The space in the center of the body houses the
battery to reduce material weight and cost. Aracna weighs 1.03 kg.
% 1034g

\begin{figure}[t]
\begin{center}
\includegraphics[width=.37\textwidth]{fig5.jpg}
\caption{Rendered CAD model of Aracna.}
\label{fi52}
\end{center}
\end{figure}

\section{Software}

The software is open-sourced Python based on the
code developed for the QuadraTot platform \citep{JY}. All code is
available on the project website \citep{WEB}. We use an infrared light
emitting diode (LED) on the robot along with an external Nintendo Wii
remote to provide feedback of the distance traveled. The robot
receives feedback from the Wii remote and internal servo sensors and
then processes the information to send the next command to
the servos. The communication between the external control computer
and the ArbotiX on-board microcontroller occurs over wireless XBee.

\section{Specifications}

\begin{figure}[t]
\begin{center}
\includegraphics[width=.37\textwidth]{fig2.jpg}
\caption{Robot printed as one piece with support material intact.}
\label{fig2}
\end{center}
\end{figure}

The Aracna may be 3D printed as a single part (Figure~\ref{fig2}). On
an Object Connex500 printer, the robot was completed in 26 hours,
using approximately 1000 g of model material and 1500 g of support
material. We estimated the cost of the printed parts to be \$410 by
using an estimated model material cost of \$1/4.5g and support
material cost of \$1/8g. This cost estimate is variable depending on
the type of material used. The robot has no removable printed parts
except for the battery cover. In the near future, we will make STL files
available for printing the robot as separate parts on
3D printers with smaller tray sizes. Printing the robot as separate parts
will also reduce the amount of support material needed, and thus the
overall cost. As shown in Table~\ref{tab:cost}, the total cost of the
Aracna is just under \$1500; these estimates are based on the material
calculations above and prices from trossenrobotics.com in March 2012.

\begin{table}[h]
\center{
\begin{tabular}{|c|c|}
\hline
Part & Cost\\
\hline\hline
3D Print Materials & \$410\\
ArbotiX Robocontroller Kit & \$189\\
Dynamixel AX-18A Robot Actuator (x8) & \$721\\
3S 11.1V 2000mAh Pro Lite LiPo Battery & \$73\\
LiPo Battery Balance Charger Kit & \$70\\
Cables, Connectors, Misc & \$28\\
\hline\hline
\bf Total & \bf \$1491\\
 \hline
\end{tabular}
}
\vskip 0.25cm
\caption{Estimated total cost. A complete parts list is on our website \citep{WEB}.}
\label{tab:cost}
\end{table}

\section{Acknowledgements}

This work was supported by the National Science Foundation's Office of
Emerging Frontiers in Research and Innovation (grant number 0735953).

\footnotesize
\bibliographystyle{apalike}
\bibliography{references}
\end{document}

### 2012-05-14 code/RobotQuadratot.py
Ready to sync with Jason's old code.
Signed-off-by: Eric Gold <elg77@cornell.edu>

@@ -265,7 +265,7 @@ class RobotQuadratot():
             posS = start(self.time) if isinstance(start, FunctionType) else start
             posE =   end(self.time) if isinstance(end,   FunctionType) else end
             goal = lInterp(self.time, [timeStart, timeEnd], posS, posE)
-
+            print goal
             cmdPos = self.commandPosition(goal)
             if logFile:
                 extraInfo = ''

### 2012-05-14 code/ax12.py

@@ -87,3 +87,7 @@ AX_ACTION = 5
 AX_RESET = 6
 AX_SYNC_WRITE = 131
 
+
+
+'''My additions for Aracana'''
+COMMAND_OVER_TIME = 20

### 2012-05-14 code/driver.py

@@ -43,15 +43,87 @@ class Driver:
     def execute(self, index, ins, params):
         """ Send an instruction to a device. """
         self.ser.flushInput()
-        self.ser.flushOutput()
         length = 2 + len(params)
         checksum = 255 - ((index + length + ins + sum(params))%256)
-        res = self.ser.write(chr(0xFF)+chr(0xFF)+chr(index)+chr(length)+chr(ins))
+        self.ser.write(chr(0xFF)+chr(0xFF)+chr(index)+chr(length)+chr(ins))
+        for val in params:
+            self.ser.write(chr(val))
+        self.ser.write(chr(checksum))
+        return self.getPacket(0)
+        
+    def execute2(self, ins, params):
+        """ Send an instruction to a device. """
+        self.ser.flushInput()
+        #self.ser.flushOutput()
+        length = len(params)
+        checksum = (ins + sum(params) + length) % 256
+        #checksum = 255 - ((index + length + ins + sum(params))%256)
+        res = self.ser.write(chr(0xFF)+chr(length)+chr(ins))
         for val in params:
             res += self.ser.write(chr(val))
         res += self.ser.write(chr(checksum))
         #print "bytes sent: " + str(res)
-        return self.getPacket(0) #TODO: uncomment this line, get rid of res = ...
+        #return self.getPacket2(0) #TODO: uncomment this line, get rid of res = ...
+    
+    def getPacket2(self, mode, id=-1, leng=-1, error=-1, params = None):
+        """ Read a return packet, iterative attempt """
+        # need a positive byte
+        d = self.ser.read()
+        if d == '': 
+            #TODO: Uncomment me
+            print "Fail Read"
+            return None
+
+        # now process our byte
+        if mode == 0:           # get our first 0xFF
+            if ord(d) == 0xff:   
+                print "Oxff found"
+                return self.getPacket2(1)
+            else:
+                print "Oxff NOT found 1, restart: " + str(ord(d))
+                return self.getPacket2(0)
+        elif mode == 1:         # get our second 0xFF
+            if ord(d) == 0xff:
+                print "Oxff found"
+                return self.getPacket2(2)
+            else:
+                print "Oxff NOT found 2, restart: " + str(ord(d))
+                return self.getPacket2(0)
+        elif mode == 2:         # get id
+            if d != 0xff:
+                print "ID found: " + str(ord(d))
+                return self.getPacket2(3, ord(d))
+            else:              
+                print "0xff is not ID, restart"
[…]

### 2012-05-14 code/Aracna/obj/pypose.pde.cpp
Removed code/Aracna/obj; updated .gitignore

@@ -1,274 +1,0 @@
-#include <WProgram.h>
-/* 
-  ArbotiX Test Program for use with PyPose 0015
-  Copyright (c) 2008-2011 Michael E. Ferguson.  All right reserved.
-
-  This library is free software; you can redistribute it and/or
-  modify it under the terms of the GNU Lesser General Public
-  License as published by the Free Software Foundation; either
-  version 2.1 of the License, or (at your option) any later version.
-
-  This library is distributed in the hope that it will be useful,
-  but WITHOUT ANY WARRANTY; without even the implied warranty of
-  MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the GNU
-  Lesser General Public License for more details.
-
-  You should have received a copy of the GNU Lesser General Public
-  License along with this library; if not, write to the Free Software
-  Foundation, Inc., 51 Franklin St, Fifth Floor, Boston, MA  02110-1301  USA
-*/
-
-//TODO: remove the absolute path and set up avr-gcc to know to look in the current directory
-//TODO: fix the absolute path in HardwareSerial.cpp for Arduino.h
-#include <C:\Users\Eric Gold\Documents\School Work\Research\Aracna\quadratotGit\code\Aracna\ax12.h>
-#include <C:\Users\Eric Gold\Documents\School Work\Research\Aracna\quadratotGit\code\Aracna\BioloidController.h>
-#include <C:\Users\Eric Gold\Documents\School Work\Research\Aracna\quadratotGit\code\Aracna\Motors2.h>
-#include <C:\Users\Eric Gold\Documents\School Work\Research\Aracna\quadratotGit\code\Aracna\HardwareSerial.cpp>
-#include "Arduino.h"
-
-BioloidController bioloid = BioloidController(1000000);
-Motors2 drive = Motors2();
-
-#define ARB_SIZE_POSE   7  // also initializes
-#define ARB_LOAD_POSE   8
-#define ARB_LOAD_SEQ    9
-#define ARB_PLAY_SEQ    10
-#define ARB_LOOP_SEQ    11
-#define ARB_TEST        25
-
-int mode = 0;                   // where we are in the frame
-
-unsigned char id = 0;           // id of this frame
-unsigned char length = 0;       // length of this frame
-unsigned char ins = 0;          // instruction of this frame
-
-unsigned char params[143];      // parameters (match RX-64 buffer size)
-unsigned char index = 0;        // index in param buffer
-
-int checksum;                   // checksum
-
[…]

### 2012-05-14 code/RobotQuadratot.py
fixed up a previous commit
Signed-off-by: Eric Gold <elg77@cornell.edu>

@@ -24,19 +24,6 @@ certain vectors of motor position.
 from ConstantsQuadratot import *
 
 
-
-import inspect
-def fileLine():
-    """Returns the current line number in our program."""
-    #print dir(inspect.currentframe().f_back.f_back)
-    #print inspect.currentframe().f_back.f_back.f_code.co_filename
-
-    return inspect.currentframe().f_back.f_back.f_code.co_filename, inspect.currentframe().f_back.f_back.f_lineno
-    #return inspect.getouterframes( inspect.currentframe() )[1].f_back.f_lineno
-
-
-
-
 class RobotFailure(Exception):
     pass
 
@@ -273,12 +260,12 @@ class RobotQuadratot():
         ii = 0
         tlast = self.time
         while self.time < timeEnd:
-            print 'time:', self.time
+            #print 'time:', self.time
             ii += 1
             posS = start(self.time) if isinstance(start, FunctionType) else start
             posE =   end(self.time) if isinstance(end,   FunctionType) else end
             goal = lInterp(self.time, [timeStart, timeEnd], posS, posE)
-            print goal
+
             cmdPos = self.commandPosition(goal)
             if logFile:
                 extraInfo = ''
@@ -358,14 +345,6 @@ class RobotQuadratot():
         '''Updates the Robots clock to the current time'''
         timeDiff  = datetime.now() - self.time0
         self.time = timeDiff.seconds + timeDiff.microseconds/1e6
-
-    def tic(self):
-        self.updateClock()
-        self.tictime = self.time
-
-    def toc(self):
-        self.updateClock()
-        print 'Elapsed (%s:%s):' % fileLine(), self.time - self.tictime
 
     def readyPosition(self, persist = False):
         if persist:

### 2012-05-14 code/RobotRex.py

@@ -1,13 +1,10 @@
-#! /usr/bin/env python
-
 #Notes: I'm going to want to make a new RunManager for Rex
 
 
 
 import math
-import pdb
 from math import *
-from datetime import *
+from datetime import datetime
 from time import sleep
 from types import FunctionType
 from copy import copy
@@ -23,11 +20,10 @@ from ConstantsRex import *
 from MotionHandler import SmoothMotionFunction
 
 
-
 class RobotRex(RobotQuadratot):
     ''''''
 
-    def __init__(self, nServos, portName="", cmdRate=40):
+    def __init__(self, nServos, portName="", commandRate = 40):
         '''Initialize the robot.
         
         Keyword arguments:
@@ -43,22 +39,19 @@ class RobotRex(RobotQuadratot):
                    in Hertz.  Default: 40.
         '''
         
-        RobotQuadratot.__init__(self, commandRate=cmdRate, skipInit=True)
+        RobotQuadratot.__init__(self, commandRate, skipInit=True)
         self.commandsSent = 0
         self.nServos = nServos
         
         #Find a valid port.
         if os.name == "posix":
-            possibilities = ['/dev/ttyUSB0', '/dev/ttyUSB1', '/dev/cu.usbserial-A800KDV8']
+            possibilities = ['/dev/ttyUSB0', '/dev/ttyUSB1']
             for pos in possibilities:
                 if os.path.exists(pos):
                     portName = pos
             if portName is None:
                 raise Exception('Could not find any of %s' % repr(possibilities))
-            self.port = Driver(portName, 38400, True)
         else:
-            if portName is None:
-                portName = 'COM6'
             self.port = Driver(portName, 38400, True)
 #For some reason, this code gets caught trying to load COM3 instead of COM6.
 #My PC usually links to COM6. Just let the user choose the port for now.
@@ -109,7 +102,6 @@ class RobotRex(RobotQuadratot):
             raise Exception('Expected postion vector of length %d, got %s instead'
                             % (self.nServos, repr(position)))
         
-        goalPosition = None
         if crop:
             goalPosition = self.cropPosition([int(xx) for xx in position], cropWarning)
         else:
@@ -119,41 +111,19 @@ class RobotRex(RobotQuadratot):
             posstr = ', '.join(['%4d' % xx for xx in goalPosition])
[…]

### 2012-05-14 code/ax12.py

@@ -87,7 +87,3 @@ AX_ACTION = 5
 AX_RESET = 6
 AX_SYNC_WRITE = 131
 
-
-
-'''My additions for Aracana'''
-COMMAND_OVER_TIME = 20

### 2012-05-14 code/driver.py

@@ -46,85 +46,12 @@ class Driver:
         self.ser.flushOutput()
         length = 2 + len(params)
         checksum = 255 - ((index + length + ins + sum(params))%256)
-        self.ser.write(chr(0xFF)+chr(0xFF)+chr(index)+chr(length)+chr(ins))
-        for val in params:
-            self.ser.write(chr(val))
-        self.ser.write(chr(checksum))
-        return self.getPacket(0)
-        
-    def execute2(self, ins, params):
-        """ Send an instruction to a device. """
-        self.ser.flushInput()
-        #self.ser.flushOutput()
-        length = len(params)
-        checksum = (ins + sum(params) + length) % 256
-        #checksum = 255 - ((index + length + ins + sum(params))%256)
-        res = self.ser.write(chr(0xFF)+chr(length)+chr(ins))
+        res = self.ser.write(chr(0xFF)+chr(0xFF)+chr(index)+chr(length)+chr(ins))
         for val in params:
             res += self.ser.write(chr(val))
         res += self.ser.write(chr(checksum))
         #print "bytes sent: " + str(res)
-        #return self.getPacket2(0) #TODO: uncomment this line, get rid of res = ...
-    
-    def getPacket2(self, mode, id=-1, leng=-1, error=-1, params = None):
-        """ Read a return packet, iterative attempt """
-        # need a positive byte
-        d = self.ser.read()
-        if d == '': 
-            #TODO: Uncomment me
-            print "Fail Read"
-            return None
-
-        # now process our byte
-        if mode == 0:           # get our first 0xFF
-            if ord(d) == 0xff:   
-                print "Oxff found"
-                return self.getPacket2(1)
-            else:
-                print "Oxff NOT found 1, restart: " + str(ord(d))
-                return self.getPacket2(0)
-        elif mode == 1:         # get our second 0xFF
-            if ord(d) == 0xff:
-                print "Oxff found"
-                return self.getPacket2(2)
-            else:
-                print "Oxff NOT found 2, restart: " + str(ord(d))
-                return self.getPacket2(0)
-        elif mode == 2:         # get id
-            if d != 0xff:
-                print "ID found: " + str(ord(d))
-                return self.getPacket2(3, ord(d))
-            else:              
-                print "0xff is not ID, restart"
-                return self.getPacket2(0)
-        elif mode == 3:         # get length
-            print "Length found: " + str(ord(d))
-            return self.getPacket2(4, id, ord(d))
[…]

### 2012-05-14 code/RobotQuadratot.py
Reverted changes.

@@ -24,6 +24,19 @@ certain vectors of motor position.
 from ConstantsQuadratot import *
 
 
+
+import inspect
+def fileLine():
+    """Returns the current line number in our program."""
+    #print dir(inspect.currentframe().f_back.f_back)
+    #print inspect.currentframe().f_back.f_back.f_code.co_filename
+
+    return inspect.currentframe().f_back.f_back.f_code.co_filename, inspect.currentframe().f_back.f_back.f_lineno
+    #return inspect.getouterframes( inspect.currentframe() )[1].f_back.f_lineno
+
+
+
+
 class RobotFailure(Exception):
     pass
 
@@ -260,12 +273,12 @@ class RobotQuadratot():
         ii = 0
         tlast = self.time
         while self.time < timeEnd:
-            #print 'time:', self.time
+            print 'time:', self.time
             ii += 1
             posS = start(self.time) if isinstance(start, FunctionType) else start
             posE =   end(self.time) if isinstance(end,   FunctionType) else end
             goal = lInterp(self.time, [timeStart, timeEnd], posS, posE)
-
+            print goal
             cmdPos = self.commandPosition(goal)
             if logFile:
                 extraInfo = ''
@@ -345,6 +358,14 @@ class RobotQuadratot():
         '''Updates the Robots clock to the current time'''
         timeDiff  = datetime.now() - self.time0
         self.time = timeDiff.seconds + timeDiff.microseconds/1e6
+
+    def tic(self):
+        self.updateClock()
+        self.tictime = self.time
+
+    def toc(self):
+        self.updateClock()
+        print 'Elapsed (%s:%s):' % fileLine(), self.time - self.tictime
 
     def readyPosition(self, persist = False):
         if persist:

### 2012-05-14 code/RobotRex.py

@@ -1,10 +1,13 @@
+#! /usr/bin/env python
+
 #Notes: I'm going to want to make a new RunManager for Rex
 
 
 
 import math
+import pdb
 from math import *
-from datetime import datetime
+from datetime import *
 from time import sleep
 from types import FunctionType
 from copy import copy
@@ -20,10 +23,11 @@ from ConstantsRex import *
 from MotionHandler import SmoothMotionFunction
 
 
+
 class RobotRex(RobotQuadratot):
     ''''''
 
-    def __init__(self, nServos, portName="", commandRate = 40):
+    def __init__(self, nServos, portName="", cmdRate=40):
         '''Initialize the robot.
         
         Keyword arguments:
@@ -39,19 +43,22 @@ class RobotRex(RobotQuadratot):
                    in Hertz.  Default: 40.
         '''
         
-        RobotQuadratot.__init__(self, commandRate, skipInit=True)
+        RobotQuadratot.__init__(self, commandRate=cmdRate, skipInit=True)
         self.commandsSent = 0
         self.nServos = nServos
         
         #Find a valid port.
         if os.name == "posix":
-            possibilities = ['/dev/ttyUSB0', '/dev/ttyUSB1']
+            possibilities = ['/dev/ttyUSB0', '/dev/ttyUSB1', '/dev/cu.usbserial-A800KDV8']
             for pos in possibilities:
                 if os.path.exists(pos):
                     portName = pos
             if portName is None:
                 raise Exception('Could not find any of %s' % repr(possibilities))
+            self.port = Driver(portName, 38400, True)
         else:
+            if portName is None:
+                portName = 'COM6'
             self.port = Driver(portName, 38400, True)
 #For some reason, this code gets caught trying to load COM3 instead of COM6.
 #My PC usually links to COM6. Just let the user choose the port for now.
@@ -102,6 +109,7 @@ class RobotRex(RobotQuadratot):
             raise Exception('Expected postion vector of length %d, got %s instead'
                             % (self.nServos, repr(position)))
         
+        goalPosition = None
         if crop:
             goalPosition = self.cropPosition([int(xx) for xx in position], cropWarning)
         else:
@@ -111,19 +119,41 @@ class RobotRex(RobotQuadratot):
             posstr = ', '.join(['%4d' % xx for xx in goalPosition])
[…]

### 2012-05-14 code/driver.py

@@ -46,12 +46,85 @@ class Driver:
         self.ser.flushOutput()
         length = 2 + len(params)
         checksum = 255 - ((index + length + ins + sum(params))%256)
-        res = self.ser.write(chr(0xFF)+chr(0xFF)+chr(index)+chr(length)+chr(ins))
+        self.ser.write(chr(0xFF)+chr(0xFF)+chr(index)+chr(length)+chr(ins))
+        for val in params:
+            self.ser.write(chr(val))
+        self.ser.write(chr(checksum))
+        return self.getPacket(0)
+        
+    def execute2(self, ins, params):
+        """ Send an instruction to a device. """
+        self.ser.flushInput()
+        #self.ser.flushOutput()
+        length = len(params)
+        checksum = (ins + sum(params) + length) % 256
+        #checksum = 255 - ((index + length + ins + sum(params))%256)
+        res = self.ser.write(chr(0xFF)+chr(length)+chr(ins))
         for val in params:
             res += self.ser.write(chr(val))
         res += self.ser.write(chr(checksum))
         #print "bytes sent: " + str(res)
-        return self.getPacket(0) #TODO: uncomment this line, get rid of res = ...
+        #return self.getPacket2(0) #TODO: uncomment this line, get rid of res = ...
+    
+    def getPacket2(self, mode, id=-1, leng=-1, error=-1, params = None):
+        """ Read a return packet, iterative attempt """
+        # need a positive byte
+        d = self.ser.read()
+        if d == '': 
+            #TODO: Uncomment me
+            print "Fail Read"
+            return None
+
+        # now process our byte
+        if mode == 0:           # get our first 0xFF
+            if ord(d) == 0xff:   
+                print "Oxff found"
+                return self.getPacket2(1)
+            else:
+                print "Oxff NOT found 1, restart: " + str(ord(d))
+                return self.getPacket2(0)
+        elif mode == 1:         # get our second 0xFF
+            if ord(d) == 0xff:
+                print "Oxff found"
+                return self.getPacket2(2)
+            else:
+                print "Oxff NOT found 2, restart: " + str(ord(d))
+                return self.getPacket2(0)
+        elif mode == 2:         # get id
+            if d != 0xff:
+                print "ID found: " + str(ord(d))
+                return self.getPacket2(3, ord(d))
+            else:              
+                print "0xff is not ID, restart"
+                return self.getPacket2(0)
+        elif mode == 3:         # get length
[…]

### 2012-05-14 code/ExternalStrategy.py
may 14 2012

@@ -5,6 +5,8 @@ from numpy import array, ix_, vstack, linspace, hstack, ones, mean
 from time import sleep
 from datetime import datetime
 import os
+import sys
+import pdb
 
 from RobotConstants import MIN_INNER, MAX_INNER, MIN_OUTER, MAX_OUTER, NORM_CENTER, MIN_CENTER, MAX_CENTER
 from Strategy import Strategy, OneStepStrategy
@@ -30,22 +32,27 @@ class NEATStrategy(OneStepStrategy):
         #self.identifier = ''.join([random.choice('abcdef1234567890') for x in range(8)])
         self.identifier = now.strftime('neat_%y%m%d_%H%M%S')
         
-        self.executable = '/home/team/s/h2_synced/HyperNEAT_v2_5/out/hyperneatTo20gens_102/Hypercube_NEAT'
-        self.motionFile = '/home/team/s/h2_synced/HyperNEAT_v2_5/out/hyperneatTo20gens_102/spiderJointAngles.txt'
-        self.fitnessFile = '/home/team/s/h2_synced/HyperNEAT_v2_5/out/hyperneatTo20gens_102/fitness'
-        self.datFile    = '/home/team/s/h2_synced/HyperNEAT_v2_5/out/hyperneatTo20gens_102/SpiderRobotExperiment.dat'
+#        self.executable = '/home/team/s/h2_synced/HyperNEAT_v2_5/out/hyperneatTo20gens_102/Hypercube_NEAT'
+        self.executable = '/home/sean/hyperneat/HyperNEAT_v2_5/out/Hypercube_NEAT'
+#        self.motionFile = '/home/team/s/h2_synced/HyperNEAT_v2_5/out/hyperneatTo20gens_102/spiderJointAngles.txt'
+        self.motionFile = '/home/sean/quadratot/code/spiderJointAngles.txt'
+#        self.fitnessFile = '/home/team/s/h2_synced/HyperNEAT_v2_5/out/hyperneatTo20gens_102/fitness'
+        self.fitnessFile = '/home/sean/quadratot/code/fitness'
+#        self.datFile    = '/home/team/s/h2_synced/HyperNEAT_v2_5/out/hyperneatTo20gens_102/SpiderRobotExperiment.dat'
+        self.datFile    = '/home/sean/hyperneat/HyperNEAT_v2_5/out/SpiderRobotExperiment.dat'
 
         self.avgPoints       = 4                         # Average over this many points
         self.junkPoints      = 1000
[…]

### 2012-05-14 code/SimOptimize.py

@@ -1,36 +1,35 @@
-import sys, pickle
+import sys, pickle, pdb
 from SineModel import SineModel5
-import simrun
+from simrun import simrun
 from ExternalStrategy import NEATStrategy
 from util import randUniformPoint
 
 
-def SimOptimize(sim):
+def SimOptimize():
     strategy = NEATStrategy(SineModel5.typicalRanges)
-
-
-    for ii in range(10):
-        nextGait = strategy.getNext()
-
-        # Run the gait
-        dist = simrun.runSim(nextGait)
-        strategy.updateResults(dist)
-
-
     print 'NEATStrategy'
-
-    
+    sim = simrun()
+    trial = 0
+    for ii in range(10000):
+        nextGait = strategy.getNext()
+        # Run the gait
+        dist = sim.runSim(nextGait)
+        strategy.updateResults(dist)
+        trial += 1
+        print 'Trial', trial, 'moved', dist
+        print
 
 def main(limit = 10):
-    sim = simrun.simrun()
-    if limit is None:
-        limit = 10000
-    
-    for ii in xrange(limit):
-        SimOptimize(sim)
-        sim.getDist()
+#    sim = simrun.simrun()
+ #   if limit is None:
+  #      limit = 10000
+   # 
+   # for ii in xrange(limit):
+    #    SimOptimize(sim)
+     #   sim.getDist()
 #        print currentState
-        print sim.getDist()
+      #  print sim.getDist()
+      SimOptimize()
 
 if __name__ == '__main__':
     main()

### 2012-05-14 code/optimize.py

@@ -53,13 +53,13 @@ def doRun():
         strategy
     except:
         #strategy = UniformStrategy(currentState, SineModel5.typicalRanges)
-        strategy = GaussianStrategy(currentState, SineModel5.typicalRanges)
+        #strategy = GaussianStrategy(currentState, SineModel5.typicalRanges)
         #strategy = GradientSampleStrategy(currentState)
         #strategy = LinearRegressionStrategy(currentState)
         #strategy = SimplexStrategy(currentState, SineModel5.typicalRanges)
         #strategy = RandomStrategy(currentState, SineModel5.typicalRanges)
         #strategy = SVMLearningStrategy(currentState, SineModel5.typicalRanges)
-        #strategy = NEATStrategy(currentState, SineModel5.typicalRanges, neatFile = neatFile)   # these args aren't used
+        strategy = NEATStrategy(currentState, SineModel5.typicalRanges, neatFile = neatFile)   # these args aren't used
         #strategy = FileStrategy(filtFile = filtFile)
         
     #runman.do_many_runs(currentState, lambda state: Neighbor.gaussian(SineModel5.typicalRanges, state))

### 2012-05-14 code/simrun.py

@@ -14,28 +14,40 @@ class simrun:
         self.filename = "output.txt"
  
     def runSim(self, gaitFunction):
+
+        os.remove('/home/sean/quadratot/code/input.txt')
+        
         ff = file('input.txt', "w")
         timeMax = 12.0
-        timeDiv = 1.0/12.0
-        divs = timeMax/timeDiv
+        timeDiv = 1.0/60.0
+        divs = int(timeMax/timeDiv)
         t = 0.0
 
-        for i in xrange(0,limit): 
-            gait = gaitFunction(t)
-            t = t+div
+        for i in xrange(0,divs):
+            gait = gaitFunction[0](t)
+            length = len(gait)
+            t = t+timeDiv
             if t == timeMax:
-                file.write(str(t) + ' ' + str(gait))
+                ff.write(str(t) + ' ')
+
+                for ii in xrange(0, length):
+                    ff.write(str(gait[ii]) + ' ')
+
             else:
-                file.write(str(t) + ' ' + str(gait) + '\n')
-            
-            
-        os.system('./physXsimulator -i input.txt -o output.txt')
+                ff.write(str(t) + ' ')
+                for ii in xrange(0,length):
+                    ff.write(str(gait[ii]) + ' ')
+                ff.write('\n')
+                           
+        ff.close()    
+        os.system('./crossSim -i input.txt -o output.txt -n')
         dist = self.getDist()      
-                
+        return dist
+    
     def getDist(self):
 
-	outputRaw= file(self.filename,"r")
-	outputData = [line.split() for line in outputRaw]
+        outputRaw= file(self.filename,"r")
+        outputData = [line.split() for line in outputRaw]
         outputColLen = len(outputData)-1
         outputRowLen = len(outputData[1])-1
         posBeg = [outputData[1][outputRowLen-2], outputData[1][outputRowLen-1], outputData[1][outputRowLen]]
@@ -43,9 +55,9 @@ class simrun:
 
         # finds the distance the robot travelled using x and y '
         xdist = float(posEnd[0])-float(posBeg[0])
-        ydist = float(posEnd[1])-float(posBeg[1])
+        zdist = float(posEnd[2])-float(posBeg[2])
 #        zdist = float(self.posEnd[2])-float(self.posBeg[2])
-        return math.sqrt(math.pow(xdist,2)+math.pow(ydist,2))
+        return math.sqrt(math.pow(xdist,2)+math.pow(zdist,2))
 
[…]

### 2012-05-16 paper_aracna/main.tex
test line
Deleted empty lines!
Jeff entering his edits to abstract and intro

@@
 \title{Aracna: An Open-Source Quadruped Platform for Evolutionary Robotics}
-\author{Sara Lohmann, Eric Gold, Jeremy Blum, Jason Yosinski \and Hod Lipson \\
+\author{Sara Lohmann, Eric Gold, Jason Yosinski, Jeremy Blum \and Hod Lipson \\
 \mbox{}\\
@@
 \begin{abstract}
-We report on a new open-source quadruped robotic platform, Aracna,
-developed at Cornell's Creative Machines Lab for robotics
-experiments. The robot comprises four legs with two joints each for a
-total of eight kinematic degrees of freedom. Four-bar linkage
-mechanisms in each leg drive the pitch of the knee joint and the hip
-joint remotely, allowing the motors to remain in the robot center,
-which reduces the inertia of each leg. Due to these non-conventional
-kinematics, the robot requires non-intuitive movements for locomotion
-and provides an interesting challenge for gait learning
-algorithms. Aracna is a low-cost, accessible robotic platform for
-testing in the real world.
+We describe a new, quadruped robot platform, Aracna,
+which requires non-intuitive motor commands in order to locomote and thus provides an interesting challenge for gait learning algorithms, such as those frequently developed in the Evolutionary Computation and Artificial Life communities.
+Aracna is an open-source hardware project composed of off-the-shelf and 3D-printed parts, enabling other research teams to modify its design according to their scientific needs.
+Aracna was designed to avoid the Achilles Heel of a previous quadruped robot platform, whose legs were so heavy that the motors could not reliably execute the commands sent to them.
+Aracna minimizes leg inertia to avoid this problem by having all motors located in the body core instead of on the legs.
+Specifically, each of the four legs has two joints separately controlled by separate four-bar linkage
+mechanisms that drive the pitch of the hip joint and knee joint remotely from motors in the robot's core. 
+%The eight kinematic degrees of freedom in Aracna are controlled by its eight servos. 
+This novel design causes unconventional kinematics, however, creating an opportunity for gait-learning algorithms, which excel in counter-intuitive design spaces where human engineers tend to underperform.
+Because it is low-cost, flexible, kinematically interesting, and and improvement over a previous design, Aracna provides a useful new hardware platform for testing algorithms that automatically generate robotic behaviors.
 \end{abstract}
@@
 \section{Introduction}
+
+There is a long history in the Artificial Life and Evolutionary Robotics community of automatically generating behaviors for robots~\citep{nolfi2000evolutionary, pfeifer2007body, sims1994evolving, hornby2005autonomous, lipson2000automatic}.
+Much work has focused on evolving gaits for legged
 

### 2012-05-16 code/Robot.py
Haocheng and Jason: python side works

@@ -1,4 +1,5 @@
 import os
+import pdb
 from datetime import datetime
 from time import sleep
 from types import FunctionType
@@ -33,7 +34,7 @@ class Robot():
     ''''''
 
     def __init__(self, silentNetFail = False, expectedIds = None, commandRate = 40,
-                 loud = False, skipInit = False, cropMethod = 'square'):
+                 loud = False, skipInit = False, cropMethod = 'smart'):
         '''Initialize the robot.
         
         Keyword arguments:
@@ -431,18 +432,23 @@ class Robot():
         return ret
 	
     def project(self,threshold, a1, a2):
-        if threshold - MIN_INNER < MIN_OUTER:
-                raise Exception('threshold too small to execute!')
-	
-        m1 = a1 - (a1+a2-780)/2
-        m2 = a2 - (a1+a2-780)/2
-        if m2 < MIN_OUTER:
-                return m1, MIN_OUTER
-        if m1 < MIN_INNER:
-                return MIN_INNER, m2
-        return m1,m2
+        #if threshold - MIN_INNER < MIN_OUTER:
+        #        raise Exception('threshold too small to execute!')
+	    #
+        #m1 = a1 - (a1+a2-780)/2
+        #m2 = a2 - (a1+a2-780)/2
+        #if m2 < MIN_OUTER:
+        #        return m1, MIN_OUTER
+        #if m1 < MIN_INNER:
+        #        return MIN_INNER, m2
+        #return m1,m2
+        if a1 + a2 < threshold:
+            delta = (threshold-a1-a2)/2
+            a1 += delta
+            a2 += delta
+        return a1, a2
 
-    def cropPositionSmart(self, position, cropWarning = False):
+    def cropPositionSmart(self, position, cropWarning = True):
         '''Crops the given positions to their appropriate values.
         Requires a vector of length 9 to be sure the IDs are in the
         assumed order.'''
@@ -452,10 +458,11 @@ class Robot():
 
         ret = self.cropPositionSquare(position)
         for ii in [0, 2, 4, 6]:
-            duple =self.project(THRESH, ret[ii], ret[ii+1])
+            duple =self.project(DIAG_CROP_THRESH, ret[ii], ret[ii+1])
             ret[ii] = int(duple[0])
-	    ret[ii+1] = int(duple[1])
+            ret[ii+1] = int(duple[1])
 
+        cropWarning = True
         if cropWarning and ret != position:
             print 'Warning: cropped %s to %s' % (repr(position), repr(ret))
 

### 2012-05-16 code/RobotConstants.py

@@ -11,7 +11,7 @@ MAX_OUTER = 680  # changing because screws still interfere
 MIN_CENTER = 512 - 120
 MAX_CENTER = 512 + 120
 NORM_CENTER = 512
-THRESH = 600
+DIAG_CROP_THRESH = 730
 POS_FLAT      = [512] * 9
 POS_READY     = [770,  40] * 4 + [512]
 POS_HALFSTAND = [700, 100] * 4 + [512]

### 2012-05-16 code/util.py

@@ -87,19 +87,21 @@ def writeArray(ff, arr):
 
 def readArray(ff):
     '''Read array from file object ff in writeArray format.'''
-    
+
+    linesSeen = 0
     for ii,line in enumerate(ff):
         #print 'line', ii, 'is', line
         line = line.strip()
         if len(line) == 0 or line[0] == '#':
             continue
         nums = [float(xx) for xx in line.split()]
-        if ii == 0:
-            ll = len(nums)
+        if linesSeen == 0:
+            firstLineSize = len(nums)
             ret = array(nums)
         else:
-            if len(nums) != ll:
+            if len(nums) != firstLineSize:
                 raise Exception('Row %s contained unexpected number of fields' % line)
             ret = vstack((ret, array(nums)))
+        linesSeen += 1
 
     return ret

### 2012-05-17 paper_aracna/main.tex
fleshed out intro and lit review..and made figures larger

@@
 There is a long history in the Artificial Life and Evolutionary Robotics community of automatically generating behaviors for robots~\citep{nolfi2000evolutionary, pfeifer2007body, sims1994evolving, hornby2005autonomous, lipson2000automatic}.
-Much work has focused on evolving gaits for legged
+Much work has focused on evolving gaits for legged robots~\citep{clune2009evolving, clune2011performance, hornby2005autonomous, hornby2003generative, kodjabachian1998evolution, Koos2012, bongard2006resilient, yosinski2011gaits, gallagher1996application}.
+While some of this previous work involved evolution directly on a physical robot~\citep{yosinski2011gaits, zykov2004evolving}, more often a gait was evolved in simulation and then transferred to the physical robot~\citep{lipson2006evolutionary, Koos2012, hornby2005autonomous, bongard2006resilient}.
+Many of these studies report that evolutionary algorithms produced gaits that outperformed those designed by a human engineer~\citep{yosinski2011gaits, hornby2005autonomous}, which is not surprising given that evolutionary algorithms routinely create solutions that are superior to manually created solutions~\citep{koza2003genetic}.
 
-We address the need for a low-cost platform with non-intuitive walking
-kinematics to explore evolutionary robotics. Aracna is the third
-evolutionary learning quadruped robot developed by the Creative
-Machines Lab \citep{HL, JY}. A common feature among the quadruped
+\begin{figure}[t]
+\begin{center}
+%\includegraphics[width=.45\textwidth]{fig1.jpg}
+\includegraphics[width=\columnwidth]{fig1.jpg}
+\caption{Aracna: an open-sourced quadruped robot platform. All
+  instructions and downloads are publicly available at
+  http://creativemachines.cornell.edu/aracna .}
+\label{fig1}
+\end{center}
+\end{figure}
+
+The results just mentioned suggest that evolutionary algorithms are a promising approach for generating gaits and other behaviors for physical robots.
+Despite this promise, the field remains small, partly because robots are expensive and they are difficult to modify.
+Access to cheap, customizable robots could increase the number of researchers able to participate in the field.
+Moreover, in nearly all of the papers mentioned previously, the robots were custom-made, preventing teams at other universities from reproducing the results of other groups and or testing new algorithms on a robotic platform used in a previous study.
+That, in turn, slows the progress of science because it is difficult to interpret whether the variance in results between different studies was due to the algorithms used or the robotic platform those algorithms were tested on.
+
+Some robot platforms are emerging, but they tend to be wheeled robots without complex kinematics, such as the ePuck~\citep{mondada2009puck}.
+Wheeled robots are interesting testbeds for many robotic behaviors, but they do not allow gait evolution and are unable to traverse rugged terrains.
+Legged robotic platforms exist, but they tend to be extremely expensive, such as the Aldebaran Nao, which costs more than \$10,000 USD.
+Another drawback to these commercial platforms is that it is hard, if not impossible, to modify the hardware design because they are not open-source hardware projects, and complex manufacturing tools would be required to manufacture new parts.
+
+In this paper we address these needs by introducing \emph{Aracna}, a low-cost, easily customizable robot platform with non-intuitive walking
+kinematics~(Figure 1). Aracna is the third quadruped robot developed for evolutionary learning algorithms by the Creative Machines Lab at Cornell University~\citep{HL, JY}.
+
+A common feature among the quadruped
 robots is two actuators that drive the flexion/extension of a knee and
@@
 complement fast spider-like movements.
-
-\begin{figure}[t]
-\begin{center}
-\includegraphics[width=.45\textwidth]{fig1.jpg}
-\caption{Aracna: an open-sourced quadruped robot platform. All
-  instructions and downloads are publicly available at
-  http://creativemachines.cornell.edu/aracna .}
-\label{fig1}
-\end{center}
-\end{figure}
 
@@
 \begin{center}
-\includegraphics[width=.37\textwidth]{fig5.jpg}
+\includegraphics[width=\columnwidth]{fig5.jpg}
 \caption{Rendered CAD model of Aracna.}

### 2012-05-17 paper_aracna/main.tex
small updates

@@
 Legged robotic platforms exist, but they tend to be extremely expensive, such as the Aldebaran Nao, which costs more than \$10,000 USD.
-Another drawback to these commercial platforms is that it is hard, if not impossible, to modify the hardware design because they are not open-source hardware projects, and complex manufacturing tools would be required to manufacture new parts.
+Another drawback to these commercial platforms is that it is hard, if not impossible, to modify the hardware design because they are not open-source hardware projects, and do not take advantage of off-the-shelf components and 3D printing, meaning that complex manufacturing tools are required to manufacture newly designed parts.
 
-In this paper we address these needs by introducing \emph{Aracna}, a low-cost, easily customizable robot platform with non-intuitive walking
-kinematics~(Figure 1). Aracna is the third quadruped robot developed for evolutionary learning algorithms by the Creative Machines Lab at Cornell University~\citep{HL, JY}.
-
-A common feature among the quadruped
-robots is two actuators that drive the flexion/extension of a knee and
-hip joint around parallel axes in each leg. The original Creative
-Machines Lab quadruped robot favored starfish-like movements
-\citep{HL}. The second quadruped robot --- QuadraTot ---
-developed spider-like movements but was found to be limited by its
-weight \citep{JY}. When creating the Aracna, we designed the hardware to
-complement fast spider-like movements.
+In this paper we address these needs by introducing \emph{Aracna}, a low-cost, open-source hardware, easily customizable robot platform with non-intuitive walking
+kinematics~(Figure 1).
+Aracna is the third quadruped robot developed for evolutionary learning algorithms by the Creative Machines Lab at Cornell University~\citep{HL, JY}.
+Like the previous QuadraTot robot design~\citep{yosinski2011gaits}, the body of Aracna is 3D printed, and the STL files are available online
+%TODO: INSERT LINK
+, meaning that other teams can easily customize the body's design. As in the previous two designs, each leg has a hip and knee joint controlled by two actuators. The original Creative
+Machines Lab quadruped robot favored starfish-like movements~\citep{HL}. The second quadruped robot --- QuadraTot ---
+developed spider-like movements, but was found to be limited by its
+weight, such that the motors would overheat and timeout when trying to execute many commands~\citep{JY, Glette2012Evolution}.
+We designed Aracna to be able to produce fast, spider-like movements, yet be lightweight enough that the motors would not overheat.
+We also designed Aracna to be inexpensive: as described below, its overall price is under \$1,500  USD.
+In the following sections we describe the Aracna platform in more detail.
 

### 2012-05-18 paper_aracna/main.tex
more writing...

@@
 kinematics~(Figure 1).
-Aracna is the third quadruped robot developed for evolutionary learning algorithms by the Creative Machines Lab at Cornell University~\citep{HL, JY}.
-Like the previous QuadraTot robot design~\citep{yosinski2011gaits}, the body of Aracna is 3D printed, and the STL files are available online
+Aracna is the third quadruped robot developed for evolutionary learning algorithms by the Creative Machines Lab at Cornell University~\citep{bongard2006resilient, yosinski2011gaits}.
+Like the most recent of the two previous designs, called QuadraTot~\citep{yosinski2011gaits}, the body of Aracna is 3D printed and the STL files are available online,
 %TODO: INSERT LINK
-, meaning that other teams can easily customize the body's design. As in the previous two designs, each leg has a hip and knee joint controlled by two actuators. The original Creative
-Machines Lab quadruped robot favored starfish-like movements~\citep{HL}. The second quadruped robot --- QuadraTot ---
+meaning that other researchers can easily customize the body's design. As in the both previous designs, each leg has a hip and knee joint controlled by two actuators. The original Creative
+Machines Lab quadruped robot favored starfish-like movements~\citep{bongard2006resilient}. The second quadruped robot --- QuadraTot ---
 developed spider-like movements, but was found to be limited by its
-weight, such that the motors would overheat and timeout when trying to execute many commands~\citep{JY, Glette2012Evolution}.
+weight, such that the motors would overheat and timeout when trying to execute many commands~\citep{yosinski2011gaits, Glette2012Evolution}.
 We designed Aracna to be able to produce fast, spider-like movements, yet be lightweight enough that the motors would not overheat.
@@
 
-\section{Hardware}
+\section{Overall Hardware Design}
 
@@
 \begin{center}
-\includegraphics[width=.23\textwidth]{fig3.pdf}
-\includegraphics[width=.23\textwidth]{fig4.pdf}
-\caption{Crank-rocker four-bar linkage controlling flexion/extension of
-  the knee and hip joints. In both cases above, the input crank (link
-  OA) is actuated by a servo, the rocker is the leg (link CB), and the
-  fixed link is OC.}
-\label{fig3}
+\includegraphics[width=\columnwidth]{fig5.jpg}
+\caption{Rendered CAD model of Aracna.}
+\label{cadModelOfRobot}
 \end{center}
@@
 
-The hardware of Aracna evolved from the previous Creative Machines Lab
-quadruped robots. We kept the same two degree of freedom pitch joint
-scheme but decreased the weight and constrained the movement of the
-joints toward the goal of creating faster spider-like movement. To
-prevent starfish-like movement, the legs were constrained. We designed
-the robot with two four-bar mechanisms to drive the joints in each
-leg. With a four-bar mechanism in place, the leg moves at a fraction
+The hardware of Aracna was designed to improve upon the previous Creative Machines Lab
+quadruped robots~\citep{bongard2006resilient, yosinski2011gaits}, while still qualitatively resembling those robots.
+Aracna is similar in that it has a body and four legs, with each leg having two joints that can pitch forward and back like knees~(Figure~\ref{cadModelOfRobot}).
+
+One change was to constrain the movement of the
+joints toward the goal of creating faster, spider-like movements. To
+prevent starfish-like movements and instead encourage a walking gait with the robot body permanently off the ground, the legs were constrained such that they cannot straighten out and the knee cannot hyperextend.
+
+Another change was to reduce the both the overall weight of the robot and the weight of each leg.
+Two previous studies that used the QuadraTot robot report that the motors quickly wore out and could not reliably execute the commands sent to them, likely because of both the overall weight of the QuadraTot and the fact that housing servos on the legs made them heavy~\citep{bongard2006resilient, yosinski2011gaits}.
+The weight of the robot's core was reduced in a number of ways.
+
+Initially, we eliminated the QuadraTot's fit-PC, an onboard Linux computer weighing 370g, and replaced it with an ArbotiX microcontroller that weights only 47g.
+As a result, the main processing occurs on an external computer and commands are sent to Aracna wirelessly.
+This change, of course, comes of the cost of the robot being truly autonomous.
+
+A second means of eliminating weight involved switching to a lighter battery.
+The QuadraTot had two 12V lithium-ion battery packs that weighted 140g each for a total of 280g.
+Aracna has a single lithium-polymer 11.1V battery that weighs SARA-TO-INSERT-WEIGHT.
+INSERT SENTENCE ON THE COST OF USING THE SMALLER BATTERY.
+
+A major modification, targeted at reducing the weight of legs, was the use of two four-bar mechanisms to drive the joints in each
+leg. This mechanism causes the controlled joint to move at a fraction
 of the output angle of the actuator, giving the motor a relatively
 larger mechanical advantage over the position of each leg.
-Figure~\ref{fig3} shows the crank-rocker system, where the input crank
-is actuated by a servo and the rocker is the leg. In this
-configuration, we keep the servo motors contained in the center of the
-robot, reducing both the inertia and mass of each leg. The four-bar
-mechanisms satisfied the design goal of making a robot that had
-non-traditional movements.  To make the robot
-accessible as an open-source platform, it was designed to be 3D
-printed. The 3D print files and CAD files can be accessed by the
-public to be printed and, if desired, modified.
+Figure~\ref{crankRocker} shows the crank-rocker system, where the input crank
+is actuated by a servo and the rocker is the leg. This
+configuration allows the servo motors to be contained in the robot core, reducing both the inertia and mass of each leg.
+The weight of an Aracna leg is SARA-TO-INSERT compared to the 217g for a QuadraTot leg.
 
-The robot is designed to be lightweight, because previous similar
-robots may have been too heavy and underpowered to generate
-interesting gaits. We use a single LiPo 11.1V battery to
-power 8 Dynamixel AX-18 servo motors and an ArbotiX
-microcontroller. The main processing occurs on an external computer,
-which allows the robot to have cheaper and lighter components
-on-board. The space in the center of the body houses the
-battery to reduce material weight and cost. Aracna weighs 1.03 kg.
-% 1034g
+Combined, these changes to minimize weight led to a SARA-TO-INSERT percent reduction in weight of the robot. The QuadraTot weighs 1.88kg whereas Aracna weighs SARA-TO-INSERTkg
+
+A final change was to upgrade the power of the servo motors in order to increase the ability of the robot to strike whichever configurations are specified by the learning algorithms.
+Specifically, we upgraded from Dynamixel AX 12+ motors to AX-18 motors, which provide INSERT-SENTENCE-ON-DIFFERENCE-IN-SERVO-POWER.
+
+\section{3D Printed Body}
+
+The body of Arcana takes advantage of 3D printing technology, also known as additive manufacturing, which generates physical objects from digital designs by building them up layer by layer~\citep{gibson2009additive, lipson2010factory}.
+The use of 3D printing means that other Aracna users can easily make copies of Aracna, either by having access to a 3D printer or via online 3D printing services such as Shapeways.com.
+Either option requires the 3D design files in the STereoLithography (STL) format, which we are publishing in the online support material for this paper~\citep{WEB}.
+Moreover, to catalyze innovation in this open-source hardware project, we are also providing the source files for the SolidWorks computer-aided-design program, because it is easier to modify a design with the original source files than with STL files.
+It should thus be easy for future Aracna users to improve or alter the design and quickly obtain a physical instantiation of the design.
+Importantly, the use of 3D printing eliminates the need to know how to machine parts, allowing many more researchers to participate in using physical robot morphologies that they design themselves.
+These ideas are in line with a broader trend toward enabling non-technical users to design and manufacture physical objects~\citep{clune2011objects, clune2011endless, lipson2010factory}.
+
+An initial version of Aracna was designed to be printed in one piece~(Figure~\ref{notYetCleanedOnePieceRobot}).
+However, if one part of the robot became damaged, an entire new robot had to be reprinted, which took over 26 hours and cost roughly \$410 USD on an Object Connex500 printer.
+To make repairing the robot easier, cheaper, and quicker, Aracna was redesigned to be modular.
+It consists of INSERT pieces--four legs and the core--that can be separately 3D printed~(Figure~REFERENCE FIGURES OF THE NEW DESIGN PRINTED).
+Printing a leg takes INSERT hours and costs INSERT.
+Printing the core takes INSERT hours and costs INSERT.
+All INSERT Aracna pieces can still be printed as one print job, with an overall time of INSERT and cost of INSERT.
+These figures are based on Aracna's use of  approximately INSERT g of rigid material and INSERT g of support
+material, and current material costs of 4.5g per USD and 8g per USD for rigid and support material, respectively. This cost estimate is variable depending on
+the type of material used. Table~\ref{tab:cost} outlines the total estimated cost of Aracna, which is just under \$1500, including its off-the-shelf electronic components.
+
+\section{Specifications}
+
+\begin{figure}[htb]
+\begin{center}
+\includegraphics[width=.37\textwidth]{fig2.jpg}
+\caption{Robot printed as one piece with support material intact.}
+\label{notYetCleanedOnePieceRobot}
+\end{center}
+\end{figure}
+
+\section{Control}
+
+The four-bar
+mechanisms also satisfied the design goal of making a robot that had
+non-traditional movements.
+There is not a one-to-one mapping between the servo output and the movement of the joint that servo controls.
+Figure INSERT displays the resulting joint angle as a function of the angle the servo is commanded to.
 
@@
 \begin{center}
-\includegraphics[width=\columnwidth]{fig5.jpg}
-\caption{Rendered CAD model of Aracna.}
-\label{fi52}
+\includegraphics[width=.23\textwidth]{fig3.pdf}
+\includegraphics[width=.23\textwidth]{fig4.pdf}
+\caption{Crank-rocker four-bar linkage controlling flexion/extension of
+  the knee and hip joints. In both cases above, the input crank (link
+  OA) is actuated by a servo, the rocker is the leg (link CB), and the
+  fixed link is OC.}
+\label{crankRocker}
 \end{center}
 \end{figure}
+
+% 1034g
 
@@
 The software is open-sourced Python based on the
-code developed for the QuadraTot platform \citep{JY}. All code is
+code developed for the QuadraTot platform \citep{yosinski2011gaits}. All code is
 available on the project website \citep{WEB}. We use an infrared light
@@
 and the ArbotiX on-board microcontroller occurs over wireless XBee.
-
-\section{Specifications}
-
-\begin{figure}[t]
-\begin{center}
-\includegraphics[width=.37\textwidth]{fig2.jpg}
-\caption{Robot printed as one piece with support material intact.}
-\label{fig2}
-\end{center}
-\end{figure}
-
-The Aracna may be 3D printed as a single part (Figure~\ref{fig2}). On
-an Object Connex500 printer, the robot was completed in 26 hours,
-using approximately 1000 g of model material and 1500 g of support
-material. We estimated the cost of the printed parts to be \$410 by
-using an estimated model material cost of \$1/4.5g and support
-material cost of \$1/8g. This cost estimate is variable depending on
-the type of material used. The robot has no removable printed parts
-except for the battery cover. In the near future, we will make STL files
-available for printing the robot as separate parts on
-3D printers with smaller tray sizes. Printing the robot as separate parts
-will also reduce the amount of support material needed, and thus the
-overall cost. As shown in Table~\ref{tab:cost}, the total cost of the
-Aracna is just under \$1500; these estimates are based on the material
-calculations above and prices from trossenrobotics.com in March 2012.
 
@@
 \vskip 0.25cm
-\caption{Estimated total cost. A complete parts list is on our website \citep{WEB}.}
+\caption{Estimated total cost. The cost of components and printing material reflect market prices from March 2012. A complete parts list is on our website \citep{WEB}.}
 \label{tab:cost}

### 2012-05-19 paper_aracna/main.tex
first pass through the paper done...now waiting for data and images from new robot

@@
 \includegraphics[width=\columnwidth]{fig5.jpg}
-\caption{Rendered CAD model of Aracna.}
+\caption{A rendered CAD model of Aracna. Note the lack of heavy servos on the legs themselves, which are instead controlled via four-bar linkages by servos in the robot's core.}
 \label{cadModelOfRobot}
@@
 
-Initially, we eliminated the QuadraTot's fit-PC, an onboard Linux computer weighing 370g, and replaced it with an ArbotiX microcontroller that weights only 47g.
+Initially, we eliminated the QuadraTot's fit-PC, an onboard Linux computer weighing 370g, and replaced it with an on-board ArbotiX microcontroller that weights only 47g.
 As a result, the main processing occurs on an external computer and commands are sent to Aracna wirelessly.
 This change, of course, comes of the cost of the robot being truly autonomous.
+Wireless communication between the external control computer
+and the ArbotiX  microcontroller occurs over wireless XBee.
 
@@
 A final change was to upgrade the power of the servo motors in order to increase the ability of the robot to strike whichever configurations are specified by the learning algorithms.
-Specifically, we upgraded from Dynamixel AX 12+ motors to AX-18 motors, which provide INSERT-SENTENCE-ON-DIFFERENCE-IN-SERVO-POWER.
+Specifically, we upgraded from Dynamixel AX 12+ motors to AX-18A motors, which have a higher stall torque (1.8Nm vs.
+1.5Nm at 12V), a higher stall current (2.2A vs 1.5A), and a higher no load speed (97 vs.
+59 RPM).
+
+\begin{figure}[t]
+\begin{center}
+\includegraphics[width=.23\textwidth]{fig3.pdf}
+\includegraphics[width=.23\textwidth]{fig4.pdf}
+\caption{Crank-rocker four-bar linkage controlling flexion/extension of
+  the knee and hip joints. In both cases above, the input crank (link
+  OA) is actuated by a servo, the rocker is the leg (link CB), and the
+  fixed link is OC.}
+\label{crankRocker}
+\end{center}
+\end{figure}
 
@@
 The use of 3D printing means that other Aracna users can easily make copies of Aracna, either by having access to a 3D printer or via online 3D printing services such as Shapeways.com.
-Either option requires the 3D design files in the STereoLithography (STL) format, which we are publishing in the online support material for this paper~\citep{WEB}.
+Either option requires the 3D design files in the stereolithography (STL) format, which we are publishing in the online support material for this paper~\citep{WEB}.
 Moreover, to catalyze innovation in this open-source hardware project, we are also providing the source files for the SolidWorks computer-aided-design program, because it is easier to modify a design with the original source files than with STL files.
@@
 
-\section{Specifications}
-
-\begin{figure}[htb]
+\begin{figure}[t]
 \begin{center}
 \includegraphics[width=.37\textwidth]{fig2.jpg}
-\caption{Robot printed as one piece with support material intact.}
+\caption{A draft version of Aracna that was printed in one piece.
+This non-modular design proved expensive to maintain if part of the robot was damaged, and was replaced in a later version with a modular design.
+This image is of the printed body with support material still present.}
 \label{notYetCleanedOnePieceRobot}
@@
 \end{figure}
-
-\section{Control}
-
-The four-bar
-mechanisms also satisfied the design goal of making a robot that had
-non-traditional movements.
-There is not a one-to-one mapping between the servo output and the movement of the joint that servo controls.
-Figure INSERT displays the resulting joint angle as a function of the angle the servo is commanded to.
-
-\begin{figure}[t]
-\begin{center}
-\includegraphics[width=.23\textwidth]{fig3.pdf}
-\includegraphics[width=.23\textwidth]{fig4.pdf}
-\caption{Crank-rocker four-bar linkage controlling flexion/extension of
-  the knee and hip joints. In both cases above, the input crank (link
-  OA) is actuated by a servo, the rocker is the leg (link CB), and the
-  fixed link is OC.}
-\label{crankRocker}
-\end{center}
-\end{figure}
-
-% 1034g
-
-\section{Software}
-
-The software is open-sourced Python based on the
-code developed for the QuadraTot platform \citep{yosinski2011gaits}. All code is
-available on the project website \citep{WEB}. We use an infrared light
-emitting diode (LED) on the robot along with an external Nintendo Wii
-remote to provide feedback of the distance traveled. The robot
-receives feedback from the Wii remote and internal servo sensors and
-then processes the information to send the next command to
-the servos. The communication between the external control computer
-and the ArbotiX on-board microcontroller occurs over wireless XBee.
 
@@
 
+\section{Control}
+
+In addition to reducing the weight of the legs, the four-bar
+mechanisms also satisfied the design goal of making a robot that had
+non-traditional movements.
+Unusual kinematics make for a more effective evolutionary robotics test platform, since evolutionary algorithms are most helpful in domains that humans find hard to program solutions for.
+The reason Aracna's kinematics are counter-intuitive is because there is a complicated mapping between each servo's output and the movement of the joint controlled by that servo~(Figure~INSERTjointAnglesAsAFunctionOfServoPositions).
+INSERT-OTHER-REASONS-THE-KINEMATICS-ARE-STRANGE.
+
+The range of motion for the hip joint is INSERT degrees and that of the knee joint is INSERT degrees~(Figure~INSERT).
+The range of the knee joint technically depends on the position of the hip joint, but the effect is slight enough (less than INSERT) degrees, that it can be effectively ignored.
+
+There are two different ways to encode movements for Aracna, increasing its flexibility as a testing platform.
+The first method is to specify a sequence of positions over time for all eight servos.
+In this method the joint angles would be determined as shown in Figure~INSERTjointAnglesAsAFunctionOfServoPositions).
+The second method is to set the speed of each servo (and possibly its initial position).
+This method is possible because with the four bar mechanism a servo constantly rotating in one direction will move the joint back and forth between its minimum and maximum opening angle.
+This latter method provides a much smaller search space and would encourage regular gaits, which have been shown to be beneficial when evolving gaits for legged robots~\citep{clune2011performance, hornby2005autonomous}.
+
+\section{Software}
+
+The software, which is also open-source and freely available~\citep{WEB}, is written in Python and based on the
+code developed for the QuadraTot platform~\citep{yosinski2011gaits}.
+The software translates a series of requested joint angles from the learning algorithm into servo movements.
+Additionally, it returns information to the learning algorithm, such as the distance traveled or the specific trajectory the robot took, so the learning algorithm can assess the quality of the gait.
+To provide this information, an infrared light
+emitting diode (LED) was placed on the robot and a Nintendo Wii
+remote was attached overhead. The software uses the combination of the two to ascertain the X, Y position of the robot. The software is interoperable with any gait or behavior learning algorithm.
+
+\section{Example Gaits}
+
+Evolutionary algorithms work best when they have a gradient to follow, instead of searching for a needle in a haystack.
+To get a sense of how randomly-generated gaits would perform, we randomly chose a few parameters....INSERT DESCRIPTION OF WHAT YOU DID, HOW IT WORKED, AND A LINK TO THE VIDEO.
+
+\section{Conclusion}
+
+I think we should wait until the rest of the paper is fleshed out before writing this.
+
 \section{Acknowledgements}
@@
 This work was supported by the National Science Foundation's Office of
-Emerging Frontiers in Research and Innovation (grant number 0735953).
+Emerging Frontiers in Research and Innovation (grant number 0735953) and an NSF Postdoctoral Research Fellowship in Biology to Jeff Clune (DBI-1003220).
 

### 2012-05-20 paper_aracna/main.tex
added weight of leg
updated missing numbers

@@
 We designed Aracna to be able to produce fast, spider-like movements, yet be lightweight enough that the motors would not overheat.
-We also designed Aracna to be inexpensive: as described below, its overall price is under \$1,500  USD.
+We also designed Aracna to be inexpensive: as described below, its overall price is under \$1,400  USD.
 In the following sections we describe the Aracna platform in more detail.
@@
 The QuadraTot had two 12V lithium-ion battery packs that weighted 140g each for a total of 280g.
-Aracna has a single lithium-polymer 11.1V battery that weighs SARA-TO-INSERT-WEIGHT.
-INSERT SENTENCE ON THE COST OF USING THE SMALLER BATTERY.
+Aracna has a single lithium-polymer 11.1V battery that weighs 122 g.
+INSERT SENTENCE ON THE COST/PENALTY OF USING THE SMALLER BATTERY.
 
@@
 is actuated by a servo and the rocker is the leg. This
-configuration allows the servo motors to be contained in the robot core, reducing both the inertia and mass of each leg.
-The weight of an Aracna leg is SARA-TO-INSERT compared to the 217g for a QuadraTot leg.
+configuration allows the servo motors to be contained in the robot core, reducing both the inertia and mass of each leg. The weight of an Aracna leg is 105g compared to the 217g for a QuadraTot leg.
 
-Combined, these changes to minimize weight led to a SARA-TO-INSERT percent reduction in weight of the robot. The QuadraTot weighs 1.88kg whereas Aracna weighs SARA-TO-INSERTkg
+Combined, these changes to minimize weight led to a 31.4 percent reduction in weight of the robot. The QuadraTot weighs 1.88kg whereas Aracna weighs 1.29kg.
 
@@
 Either option requires the 3D design files in the stereolithography (STL) format, which we are publishing in the online support material for this paper~\citep{WEB}.
-Moreover, to catalyze innovation in this open-source hardware project, we are also providing the source files for the SolidWorks computer-aided-design program, because it is easier to modify a design with the original source files than with STL files.
+Moreover, to catalyze innovation in this open-source hardware project, we are also providing the source files for the SolidWorks computer-aided-design program, to enable others to modify the design.
 It should thus be easy for future Aracna users to improve or alter the design and quickly obtain a physical instantiation of the design.
@@
 An initial version of Aracna was designed to be printed in one piece~(Figure~\ref{notYetCleanedOnePieceRobot}).
-However, if one part of the robot became damaged, an entire new robot had to be reprinted, which took over 26 hours and cost roughly \$410 USD on an Object Connex500 printer.
+However, if one part of the robot became damaged, an entire new robot had to be reprinted, which took over 26 hours and costs roughly \$355 USD on an Objet Connex500 printer.
 To make repairing the robot easier, cheaper, and quicker, Aracna was redesigned to be modular.
-It consists of INSERT pieces--four legs and the core--that can be separately 3D printed~(Figure~REFERENCE FIGURES OF THE NEW DESIGN PRINTED).
-Printing a leg takes INSERT hours and costs INSERT.
-Printing the core takes INSERT hours and costs INSERT.
-All INSERT Aracna pieces can still be printed as one print job, with an overall time of INSERT and cost of INSERT.
-These figures are based on Aracna's use of  approximately INSERT g of rigid material and INSERT g of support
-material, and current material costs of 4.5g per USD and 8g per USD for rigid and support material, respectively. This cost estimate is variable depending on
-the type of material used. Table~\ref{tab:cost} outlines the total estimated cost of Aracna, which is just under \$1500, including its off-the-shelf electronic components.
+It consists of 15 pieces--four legs and the core--that can be separately 3D printed~(Figure~\ref{notYetCleanedMulitPieceRobot}).
+Printing a leg takes 3.3 hours and costs roughly \$64 USD.
+Printing the core takes 3 hours and costs \$101 USD.
+All 15 Aracna pieces can still be printed as one print job, with an overall time of approximately 10 hours and cost of \$308.
+These figures are based on Aracna's use of  approximately 967g of model material and 746g of support material, and current material costs of 4.5g per USD and 8g per USD for rigid and support material, respectively.
+This cost estimate is variable depending on the type of material used.
+The print times are estimates calculated by the Objet's software and are meant to be used as a relative comparison of print times.
+Table~\ref{tab:cost} outlines the total estimated cost of Aracna, which is just under \$1,400, including its off-the-shelf electronic components.
 
@@
 
+\begin{figure}[t]
+\begin{center}
+\includegraphics[width=.37\textwidth]{fig6.jpg}
+\caption{A final version of Aracna printed in multiple pieces.The body is printed as two pieces, with 9 smaller parts within the top piece to reduce support material and print time.
+These parts can be printed individually if replacements are necessary.
+The body consists of a total of 11 parts.
+This image shows a set of 12 printed parts (the complete body and a single leg) with support material still present.}
+\label{notYetCleanedMulitPieceRobot}
+\end{center}
+\end{figure}
+
 \begin{table}[h]
@@
 \hline\hline
-3D Print Materials & \$410\\
+3D Print Materials & \$308\\
 ArbotiX Robocontroller Kit & \$189\\
@@
 \hline\hline
-\bf Total & \bf \$1491\\
+\bf Total & \bf \$1389\\
  \hline

### 2012-05-20 paper_aracna/main.tex
Finished cleaning up figures and references.

@@
 \usepackage{natbib,alife13}
+\usepackage{ownstyles}
+
+\graphicspath{{../figures/}}
 
@@
 
-\begin{figure}[t]
-\begin{center}
-%\includegraphics[width=.45\textwidth]{fig1.jpg}
-\includegraphics[width=\columnwidth]{fig1.jpg}
-\caption{Aracna: an open-sourced quadruped robot platform. All
-  instructions and downloads are publicly available at
-  http://creativemachines.cornell.edu/aracna .}
-\label{fig1}
-\end{center}
-\end{figure}
+%\begin{figure}[t]
+%\begin{center}
+%%\includegraphics[width=.45\textwidth]{aracna_black_1.jpg}
+%\includegraphics[width=\columnwidth]{aracna_black_1.jpg}
+%\caption{Aracna: an open-sourced quadruped robot platform. All
+%  instructions and downloads are publicly available at
+%  http://creativemachines.cornell.edu/aracna .}
+%\label{fig:aracna_black_1}
+%\end{center}
+%\end{figure}
+
+\figp{aracna_black_1}{1}{Aracna: an open-sourced quadruped robot
+  platform. All instructions and downloads are publicly available at
+  http://creativemachines.cornell.edu/aracna.}{}
 
@@
 In this paper we address these needs by introducing \emph{Aracna}, a low-cost, open-source hardware, easily customizable robot platform with non-intuitive walking
-kinematics~(Figure 1).
+kinematics~(\figref{aracna_black_1}).
 Aracna is the third quadruped robot developed for evolutionary learning algorithms by the Creative Machines Lab at Cornell University~\citep{bongard2006resilient, yosinski2011gaits}.
@@
 
-\begin{figure}[t]
-\begin{center}
-\includegraphics[width=\columnwidth]{fig5.jpg}
-\caption{A rendered CAD model of Aracna. Note the lack of heavy servos on the legs themselves, which are instead controlled via four-bar linkages by servos in the robot's core.}
-\label{cadModelOfRobot}
-\end{center}
-\end{figure}
+%\begin{figure}[t]
+%\begin{center}
+%\includegraphics[width=\columnwidth]{aracna_cad_ver1.jpg}
+%\caption{A rendered CAD model of Aracna. Note the lack of heavy servos on the legs themselves, which are instead controlled via four-bar linkages by servos in the robot's core.}
+%\label{fig:aracna_cad_ver1.jpg}
+%\end{center}
+%\end{figure}
+
+\figp{aracna_cad_ver1}{1}{A rendered CAD model of Aracna. Note the
+  lack of heavy servos on the legs themselves, which are instead
+  controlled via four-bar linkages by servos in the robot's core.}
 
@@
 quadruped robots~\citep{bongard2006resilient, yosinski2011gaits}, while still qualitatively resembling those robots.
-Aracna is similar in that it has a body and four legs, with each leg having two joints that can pitch forward and back like knees~(Figure~\ref{cadModelOfRobot}).
+Aracna is similar in that it has a body and four legs, with each leg having two joints that can pitch forward and back like knees~(\figref{aracna_cad_ver1}).
 
@@
 larger mechanical advantage over the position of each leg.
-Figure~\ref{crankRocker} shows the crank-rocker system, where the input crank
+\figref{aracna_leg_drawing_outer} shows the crank-rocker system, where the input crank
 is actuated by a servo and the rocker is the leg. This
@@
 
-\begin{figure}[t]
-\begin{center}
-\includegraphics[width=.23\textwidth]{fig3.pdf}
-\includegraphics[width=.23\textwidth]{fig4.pdf}
-\caption{Crank-rocker four-bar linkage controlling flexion/extension of
-  the knee and hip joints. In both cases above, the input crank (link
-  OA) is actuated by a servo, the rocker is the leg (link CB), and the
-  fixed link is OC.}
-\label{crankRocker}
-\end{center}
-\end{figure}
+%\begin{figure}[t]
+%\begin{center}
+%\includegraphics[width=.23\textwidth]{aracna_leg_drawing_outer}
+%\includegraphics[width=.23\textwidth]{aracna_leg_drawing_inner}
+%\caption{Crank-rocker four-bar linkage controlling flexion/extension of
+%  the knee and hip joints. In both cases above, the input crank (link
+%  OA) is actuated by a servo, the rocker is the leg (link CB), and the
+%  fixed link is OC.}
+%\label{fig:crankRocker}
+%\end{center}
+%\end{figure}
+
+\figgp{aracna_leg_drawing_outer}{.46}{aracna_leg_drawing_inner}{.46}{Crank-rocker
+  four-bar linkage controlling flexion/extension of the knee and hip
+  joints. In both cases above, the input crank (link OA) is actuated
+  by a servo, the rocker is the leg (link CB), and the fixed link is
+  OC.}
 
@@
 
-An initial version of Aracna was designed to be printed in one piece~(Figure~\ref{notYetCleanedOnePieceRobot}).
+An initial version of Aracna was designed to be printed in one piece~(\figref{aracna_notYetCleanedOnePieceRobot}).
 However, if one part of the robot became damaged, an entire new robot had to be reprinted, which took over 26 hours and costs roughly \$355 USD on an Objet Connex500 printer.
 To make repairing the robot easier, cheaper, and quicker, Aracna was redesigned to be modular.
-It consists of 15 pieces--four legs and the core--that can be separately 3D printed~(Figure~\ref{notYetCleanedMulitPieceRobot}).
+It consists of 15 pieces--four legs and the core--that can be separately 3D printed~(\figref{aracna_notYetCleanedMultiPieceRobot}).
 Printing a leg takes 3.3 hours and costs roughly \$64 USD.
@@
 
-\begin{figure}[t]
-\begin{center}
-\includegraphics[width=.37\textwidth]{fig2.jpg}
-\caption{A draft version of Aracna that was printed in one piece.
+%\begin{figure}[t]
+%\begin{center}
+%\includegraphics[width=.37\textwidth]{aracna_notYetCleanedOnePieceRobot.jpg}
+%\caption{A draft version of Aracna that was printed in one piece.
 This non-modular design proved expensive to maintain if part of the robot was damaged, and was replaced in a later version with a modular design.
 This image is of the printed body with support material still present.}
-\label{notYetCleanedOnePieceRobot}
-\end{center}
-\end{figure}
+%\label{fig:aracna_notYetCleanedOnePieceRobot}
+%\end{center}
+%\end{figure}
 
-\begin{figure}[t]
-\begin{center}
-\includegraphics[width=.37\textwidth]{fig6.jpg}
-\caption{A final version of Aracna printed in multiple pieces.The body is printed as two pieces, with 9 smaller parts within the top piece to reduce support material and print time.
+\figp{aracna_notYetCleanedOnePieceRobot}{.74}{A draft version of
+  Aracna that was printed in one piece. This non-modular design proved
+  expensive to maintain if part of the robot was damaged, and was
+  replaced in a later version with a modular design. This image is of
+  the printed body with support material still present.}
+
+%\begin{figure}[t]
+%\begin{center}
+%\includegraphics[width=.37\textwidth]{aracna_notYetCleanedMultiPieceRobot.jpg}
+%\caption{A final version of Aracna printed in multiple pieces.The body is printed as two pieces, with 9 smaller parts within the top piece to reduce support material and print time.
 These parts can be printed individually if replacements are necessary.
@@
 This image shows a set of 12 printed parts (the complete body and a single leg) with support material still present.}
-\label{notYetCleanedMulitPieceRobot}
-\end{center}
-\end{figure}
+%\label{fig:aracna_notYetCleanedMultiPieceRobot}
+%\end{center}
+%\end{figure}
+
+\figp{aracna_notYetCleanedMultiPieceRobot}{.74}{A final version of
+  Aracna printed in multiple pieces.The body is printed as two pieces,
+  with 9 smaller parts within the top piece to reduce support material
+  and print time. These parts can be printed individually if
+  replacements are necessary. The body consists of a total of 11
+  parts. This image shows a set of 12 printed parts (the complete body
+  and a single leg) with support material still present.}
 
@@
 \caption{Estimated total cost. The cost of components and printing material reflect market prices from March 2012. A complete parts list is on our website \citep{WEB}.}
-\label{tab:cost}
+\tablabel{cost}
 \end{table}

### 2012-05-21 code/RobotRex.py
Keeping track of changes to RobotRex.py
Signed-off-by: Eric Gold <elg77@cornell.edu>

@@ -5,7 +5,6 @@
 
 
 import math
-import pdb
 from math import *
 from datetime import *
 from time import sleep
@@ -20,7 +19,6 @@ from driver import Driver
 
 from RobotQuadratot import *
 from ConstantsRex import *
-from MotionHandler import SmoothMotionFunction
 
 
 
@@ -354,18 +352,9 @@ class RobotRex(RobotQuadratot):
 
 if __name__ == "__main__":
     robot = RobotRex(8, "COM6", cmdRate = 14)
-    pos0 = [0] * 8
-    pos1 = [270] * 8
-    pos2 = [0,0,0,0,270,270,270,270]
-    
     pi = math.pi
     
     dur = 100.0
-    start = lambda t: (abs(200.0*sin(pi*t)),0.0,
-                       abs(270.0/dur*t*sin(pi*t)), 125.0,
-                       t*270.0/dur, (1-t/dur)*270.0,
-                       270.0*(t/dur)**3,25.0)
-    end = lambda t: ([135.0] * 8)
     
     f1 = lambda t: (abs(270.0*sin(10*pi*t/dur)),abs(270.0*sin(10*pi*t/dur)),
                     abs(270.0*sin(10*pi*t/dur)),abs(270.0*sin(10*pi*t/dur)),
@@ -373,33 +362,4 @@ if __name__ == "__main__":
                     abs(270.0*sin(10*pi*t/dur)),abs(270.0*sin(10*pi*t/dur)))
     sleep(2)
 
-#    print "sending first"
-#    robot.interpMove2(pos0,pos0,2)
-#    print "sent first"
-#    sleep(3)
-#    print "sending second"
-#    robot.commandPosition(pos1)
-#    print "sent second"
-#    print "starting function"
-#    robot.interpMove(start,end,4)
-#    print "Commands sent: " + str(robot.commandsSent)
-    print robot.sleep
-    print "starting position"
     robot.interpMove2(f1,f1,dur)
-    print "Commands sent: " + str(robot.commandsSent)
-
-    pdb.set_trace()
-
-    #robot.commandFunction(myFunction)
-#    sleep(1)
-#    print "relaxing"
-#    robot.relax()
-#    robot.commandPosition(pos0)
-#    sleep(2)
-#    print robot.readCurrentPosition()
-#    sleep(.5)
-#    print "About to interpMove"
-#    robot.interpMove(pos0, pos1, 10)
-#    sleep(3)
-#    print robot.readCurrentPosition()
-#    robot.relax()

### 2012-05-21 code/RobotRex.py
cleaned code and removed unnecessary imports
Signed-off-by: Eric Gold <elg77@cornell.edu>

@@ -9,10 +9,7 @@ from math import *
 from datetime import *
 from time import sleep
 from types import FunctionType
-from copy import copy
-from numpy import array
-
-from Motion import lInterp, scaleTime
+from numpy import array, interp
 
 from ax12 import *
 from driver import Driver
@@ -21,6 +18,11 @@ from RobotQuadratot import *
 from ConstantsRex import *
 
 
+def lInterp(time, theDomain, val1, val2):
+    ret = []
+    for ii in range(len(val1)):
+        ret.append(interp([time], theDomain, [val1[ii], val2[ii]])[0])
+    return ret
 
 class RobotRex(RobotQuadratot):
     ''''''
@@ -57,19 +59,11 @@ class RobotRex(RobotQuadratot):
         else:
             if portName is None:
                 portName = 'COM6'
-            self.port = Driver(portName, 38400, True)
-#For some reason, this code gets caught trying to load COM3 instead of COM6.
-#My PC usually links to COM6. Just let the user choose the port for now.
-#            for i in range(12):
-#                portName = "COM" + str(i)
-#                try:
-#                    self.port = Driver(portName, 38400, True)
-#                    break
-#                except:
-#                    if self.loud: print portName + " not found"
+            #self.port = Driver(portName, 38400, True)
+            self.port = None
         
-        if self.port is None:
-            raise Exception("Failed  to open any Serial/COM port")
+#        if self.port is None:
+#            raise Exception("Failed  to open any Serial/COM port")
 
         self.currentPos = None
         self.resetClock()
@@ -86,80 +80,6 @@ class RobotRex(RobotQuadratot):
             self.commandPosition(POS_READY)
             sleep(2)
     
-    def commandPosition(self, position, crop = True, cropWarning = False, allAtOnce = True):
-        '''Set Rex to the given postion as quickly as possible.
-
-        commandPosition will command the robot to move its servos to
-        the given position vector.  This vector is cropped to
-        the physical limits of the robot and converted to an integer.
-
-        Positional arguments:
-        position -- A length 8 vector of desired positions.
-
-        Keyword arguments:
-        cropWarning -- Whether or not to print a warning if the
-                       positions are cropped.  Default: False.
-        allAtOnce -- Whether or not to move the servos using smooth interpolated
-                     movements. If false, one servo is adjusted at a time.
-        '''
-                
[…]

### 2012-05-21 paper_aracna/main.tex
stuff

@@
 
-The range of motion for the hip joint is INSERT degrees and that of the knee joint is INSERT degrees~(Figure~INSERT).
-The range of the knee joint technically depends on the position of the hip joint, but the effect is slight enough (less than INSERT) degrees, that it can be effectively ignored.
+The range of motion for the hip joint is 21 degrees and that of the knee joint is 40 degrees~(Figure~INSERT).
 
@@
 
-I think we should wait until the rest of the paper is fleshed out before writing this.
+Aracna is a relatively low-cost open source platform for evolutionary robotics.
+The complex kinematics along with the open source nature of the robot will provide an interesting and challenging universal platform for comparing walking gait algorithms.
+The updated platform is modular, allowing for low-cost replacement parts and varied leg designs.
+Future work can include modifying a single leg to have different linkage lengths, or to replace a kinematic joint with a compliant, or flexible, joint.
+This platform enables testing an experimantal leg such as this without having to print a complete robot.
+Aracna will enable mutiple users to compare data across a single lightweight, low-cost evolutionary robotic platform.
 

### 2012-05-21 paper_aracna/main.tex
merged

@@
 
-\graphicspath{{../figures/}}
+\graphicspath{
+  {../figures/}
+  {../figures/aracna_legs/}
+}
 
 \title{Aracna: An Open-Source Quadruped Platform for Evolutionary Robotics}
-\author{Sara Lohmann, Eric Gold, Jason Yosinski, Jeremy Blum \and Hod Lipson \\
+\author{Sara Lohmann\thanks{~~Authors contributed equally to this work.}~, Jason Yosinski$^{*}$, Eric Gold, Jeff Clune, Jeremy Blum \and Hod Lipson  \\
   \mbox{}\\
 Cornell University, 239 Upson Hall, Ithaca, NY 14853 \\
-\texttt{sml253@cornell.edu}}
+\texttt{sml253@cornell.edu, yosinski@cs.cornell.edu}}
 
@@
 \begin{abstract}
-We describe a new, quadruped robot platform, Aracna,
-which requires non-intuitive motor commands in order to locomote and thus provides an interesting challenge for gait learning algorithms, such as those frequently developed in the Evolutionary Computation and Artificial Life communities.
-Aracna is an open-source hardware project composed of off-the-shelf and 3D-printed parts, enabling other research teams to modify its design according to their scientific needs.
-Aracna was designed to avoid the Achilles Heel of a previous quadruped robot platform, whose legs were so heavy that the motors could not reliably execute the commands sent to them.
-Aracna minimizes leg inertia to avoid this problem by having all motors located in the body core instead of on the legs.
-Specifically, each of the four legs has two joints separately controlled by separate four-bar linkage
-mechanisms that drive the pitch of the hip joint and knee joint remotely from motors in the robot's core. 
+
+We describe a new, quadruped robot platform, Aracna, which requires
+non-intuitive motor commands in order to locomote and thus provides an
+interesting challenge for gait learning algorithms, such as those
+frequently developed in the Evolutionary Computation and Artificial
+Life communities. Aracna is an open-source hardware project composed
+of off-the-shelf and 3D-printed parts, enabling other research teams
+to modify its design according to their scientific needs.  Aracna was
+designed to overcome the shortcomings of a previous quadruped robot
+platform, whose legs were so heavy that the motors could not reliably
+execute the commands sent to them. We avoid this problem by
+locating all motors in the body core instead of on the legs and
+through a design which enables the servos to have a greater mechanical
+advantage.  Specifically, each of the four legs has two joints
+controlled by separate four-bar linkage mechanisms that
+drive the pitch of the hip joint and knee joint.
 %The eight kinematic degrees of freedom in Aracna are controlled by its eight servos. 
-This novel design causes unconventional kinematics, however, creating an opportunity for gait-learning algorithms, which excel in counter-intuitive design spaces where human engineers tend to underperform.
-Because it is low-cost, flexible, kinematically interesting, and and improvement over a previous design, Aracna provides a useful new hardware platform for testing algorithms that automatically generate robotic behaviors.
+This novel design causes unconventional kinematics, creating
+an opportunity for gait-learning algorithms, which excel in
+counter-intuitive design spaces where human engineers tend to
+underperform.  Because it is low-cost, flexible, kinematically
+interesting, and and improvement over a previous design, Aracna
+provides a useful new hardware platform for testing algorithms that
+automatically generate robotic behaviors.
+
 \end{abstract}
@@
 %\includegraphics[width=\columnwidth]{aracna_black_1.jpg}
-%\caption{Aracna: an open-sourced quadruped robot platform. All
+%\caption{Aracna: an open-source quadruped robot platform. All
 %  instructions and downloads are publicly available at
@@
 
-\figp{aracna_black_1}{1}{Aracna: an open-sourced quadruped robot
-  platform. All instructions and downloads are publicly available at
-  http://creativemachines.cornell.edu/aracna.}{}
+\figp{aracna_black_2}{1}{Aracna: an open-source 3D printed quadruped
+  robot platform, here printed in a black rapid prototyping
+  polymer. STL files for 3D printing the robot and drivers for the
+  Arbotix board and servos are publicly available at
+  http://creativemachines.cornell.edu/aracna.}
 
 The results just mentioned suggest that evolutionary algorithms are a promising approach for generating gaits and other behaviors for physical robots.
-Despite this promise, the field remains small, partly because robots are expensive and they are difficult to modify.
+Despite this promise, the field remains small, partly because robots are expensive, and they are difficult to modify.
 Access to cheap, customizable robots could increase the number of researchers able to participate in the field.
@@
 
-In this paper we address these needs by introducing \emph{Aracna}, a low-cost, open-source hardware, easily customizable robot platform with non-intuitive walking
-kinematics~(\figref{aracna_black_1}).
+In this paper we address these needs by introducing \emph{Aracna}, a low-cost, open-source, easily customizable robot platform with non-intuitive walking
+kinematics~(\figref{aracna_black_2}).
 Aracna is the third quadruped robot developed for evolutionary learning algorithms by the Creative Machines Lab at Cornell University~\citep{bongard2006resilient, yosinski2011gaits}.
@@
 developed spider-like movements, but was found to be limited by its
-weight, such that the motors would overheat and timeout when trying to execute many commands~\citep{yosinski2011gaits, Glette2012Evolution}.
+weight and lack of mechanical advantage, such that the motors would overheat and timeout when trying to execute many commands~\citep{yosinski2011gaits, Glette2012Evolution}.
 We designed Aracna to be able to produce fast, spider-like movements, yet be lightweight enough that the motors would not overheat.
@@
 %\begin{center}
-%\includegraphics[width=\columnwidth]{aracna_cad_ver1.jpg}
+%\includegraphics[width=\columnwidth]{aracna_cad_ver2.jpg}
 %\caption{A rendered CAD model of Aracna. Note the lack of heavy servos on the legs themselves, which are instead controlled via four-bar linkages by servos in the robot's core.}
@@
 
-\figp{aracna_cad_ver1}{1}{A rendered CAD model of Aracna. Note the
+\figsp{aracna_cad_ver2}{.8}{A rendered CAD model of Aracna. Note the
   lack of heavy servos on the legs themselves, which are instead
@@
 quadruped robots~\citep{bongard2006resilient, yosinski2011gaits}, while still qualitatively resembling those robots.
-Aracna is similar in that it has a body and four legs, with each leg having two joints that can pitch forward and back like knees~(\figref{aracna_cad_ver1}).
+Aracna is similar in that it has a body and four legs, with each leg having two joints that can pitch forward and back like knees~(\figref{aracna_cad_ver2}~and~\ref{fig:crankRocker}).
 
@@
 
-Initially, we eliminated the QuadraTot's fit-PC, an onboard Linux computer weighing 370g, and replaced it with an on-board ArbotiX microcontroller that weights only 47g.
-As a result, the main processing occurs on an external computer and commands are sent to Aracna wirelessly.
-This change, of course, comes of the cost of the robot being truly autonomous.
-Wireless communication between the external control computer
-and the ArbotiX  microcontroller occurs over wireless XBee.
+Initially, we eliminated the QuadraTot's fit-PC, an onboard Linux
+computer weighing 370g, and replaced it with an on-board ArbotiX
+microcontroller that weights only 47g.
+%As a result, the main
+%processing occurs on an external computer and commands are sent to
+%Aracna wirelessly. This change, of course, comes of the cost of the
+%robot being truly autonomous.
+Wireless communication between the
+external control computer and the ArbotiX microcontroller occurs over
+wireless XBee.
 
-A second means of eliminating weight involved switching to a lighter battery.
-The QuadraTot had two 12V lithium-ion battery packs that weighted 140g each for a total of 280g.
-Aracna has a single lithium-polymer 11.1V battery that weighs 122 g.
-INSERT SENTENCE ON THE COST/PENALTY OF USING THE SMALLER BATTERY.
+A second means of eliminating weight involved switching to a lighter
+battery. The QuadraTot had two 12V lithium-ion battery packs that
+weighted 140g each for a total of 280g. Aracna has a single
+lithium-polymer 11.1V battery that weighs 122 g. As with the
+QuadraTot, Aracna can also run tethered to power, if desired, to avoid
+the need to run from battery power. This may be helpful for extended
+experiments.
 
@@
 larger mechanical advantage over the position of each leg.
-\figref{aracna_leg_drawing_outer} shows the crank-rocker system, where the input crank
+\figref{crankRocker} shows the crank-rocker system, where the input crank
 is actuated by a servo and the rocker is the leg. This
@@
 Specifically, we upgraded from Dynamixel AX 12+ motors to AX-18A motors, which have a higher stall torque (1.8Nm vs.
-1.5Nm at 12V), a higher stall current (2.2A vs 1.5A), and a higher no load speed (97 vs.
+1.5Nm at 12V), a higher stall current (2.2A vs 1.5A), and a higher no-load speed (97 vs.
 59 RPM).
@@
 
-\figgp{aracna_leg_drawing_outer}{.46}{aracna_leg_drawing_inner}{.46}{Crank-rocker
+\figp{crankRocker}{1}{Crank-rocker
   four-bar linkage controlling flexion/extension of the knee and hip
@@
 
+\figp{aracna_legs_all_vert}{.8}{The range of motion of the hip and
+  knee joints in each leg of Aracna. The hip joint rotates by
+  $21.3^\circ$ and the knee joint by $40.2^\circ$.}
+
 \section{3D Printed Body}
@@
 The body of Arcana takes advantage of 3D printing technology, also known as additive manufacturing, which generates physical objects from digital designs by building them up layer by layer~\citep{gibson2009additive, lipson2010factory}.
-The use of 3D printing means that other Aracna users can easily make copies of Aracna, either by having access to a 3D printer or via online 3D printing services such as Shapeways.com.
-Either option requires the 3D design files in the stereolithography (STL) format, which we are publishing in the online support material for this paper~\citep{WEB}.
+The use of 3D printing means that other Aracna users can easily make copies of Aracna, either by having access to a 3D printer or via online 3D printing services such as Shapeways, Sculpteo, or other online vendors.
+Either option requires the 3D design files in the stereolithography (STL) format, which are published in the online support material for this paper~\citep{WEB}.
 Moreover, to catalyze innovation in this open-source hardware project, we are also providing the source files for the SolidWorks computer-aided-design program, to enable others to modify the design.
-It should thus be easy for future Aracna users to improve or alter the design and quickly obtain a physical instantiation of the design.
+It is thus possible for future Aracna users to improve or alter the design and quickly obtain a physical instantiation of the design.
 Importantly, the use of 3D printing eliminates the need to know how to machine parts, allowing many more researchers to participate in using physical robot morphologies that they design themselves.
@@
 To make repairing the robot easier, cheaper, and quicker, Aracna was redesigned to be modular.
-It consists of 15 pieces--four legs and the core--that can be separately 3D printed~(\figref{aracna_notYetCleanedMultiPieceRobot}).
+It consists of 15 pieces--four legs and the core--that can be separately 3D printed~(\figref{aracna_notYetCleanedMultiPieceRobot_edit}).
 Printing a leg takes 3.3 hours and costs roughly \$64 USD.
@@
 The print times are estimates calculated by the Objet's software and are meant to be used as a relative comparison of print times.
-Table~\ref{tab:cost} outlines the total estimated cost of Aracna, which is just under \$1,400, including its off-the-shelf electronic components.
+\tabref{cost} outlines the total estimated cost of Aracna, which is just under \$1,400, including its off-the-shelf electronic components.
 
@@
 
-\figp{aracna_notYetCleanedOnePieceRobot}{.74}{A draft version of
-  Aracna that was printed in one piece. This non-modular design proved
+\figp{aracna_notYetCleanedOnePieceRobot}{1}{A draft version of Aracna
+  that was printed in one piece. This monolithic design proved
   expensive to maintain if part of the robot was damaged, and was
-  replaced in a later version with a modular design. This image is of
+  replaced in a later version with a modular design. This figure shows
   the printed body with support material still present.}
@@
 
-\figp{aracna_notYetCleanedMultiPieceRobot}{.74}{A final version of
-  Aracna printed in multiple pieces.The body is printed as two pieces,
+\figp{aracna_notYetCleanedMultiPieceRobot_edit}{1}{A final version of
+  Aracna printed in multiple pieces. The body is printed as two pieces,
   with 9 smaller parts within the top piece to reduce support material
@@
   and a single leg) with support material still present.}
+
+\figp{aracna_black_1_edit}{1}{Aracna with optional top cover.}
 
@@
 mechanisms also satisfied the design goal of making a robot that had
-non-traditional movements.
-Unusual kinematics make for a more effective evolutionary robotics test platform, since evolutionary algorithms are most helpful in domains that humans find hard to program solutions for.
-The reason Aracna's kinematics are counter-intuitive is because there is a complicated mapping between each servo's output and the movement of the joint controlled by that servo~(Figure~INSERTjointAnglesAsAFunctionOfServoPositions).
-INSERT-OTHER-REASONS-THE-KINEMATICS-ARE-STRANGE.
+non-traditional movements. Unusual kinematics make for a more
+effective algorithmic test platform, since gait learning algorithms
+are most helpful in domains that humans find hard to program solutions
+for. The reason Aracna's kinematics are counter-intuitive is because
+there is a nonlinear mapping between each servo's output and the
+movement of the joint controlled by that
+servo.
 
-The range of motion for the hip joint is 21 degrees and that of the knee joint is 40 degrees~(Figure~INSERT).
+The range of motion for the hip joint is $21.3^\circ$ and that of the
+knee joint is $40.2^\circ$.
+% Nope.
+%The range of the knee joint technically
+%depends on the position of the hip joint, but the effect is small
+%enough that it can be effectively ignored.
+These ranges are realized
+over a servo motion of $184^\circ$ and $192^\circ$ for the hip and
+knee joints, respectively. These amounts are notably larger than the
+corresponding joint motions, which produces the desired mechanical
+advantage.
 
-There are two different ways to encode movements for Aracna, increasing its flexibility as a testing platform.
-The first method is to specify a sequence of positions over time for all eight servos.
-In this method the joint angles would be determined as shown in Figure~INSERTjointAnglesAsAFunctionOfServoPositions).
-The second method is to set the speed of each servo (and possibly its initial position).
-This method is possible because with the four bar mechanism a servo constantly rotating in one direction will move the joint back and forth between its minimum and maximum opening angle.
-This latter method provides a much smaller search space and would encourage regular gaits, which have been shown to be beneficial when evolving gaits for legged robots~\citep{clune2011performance, hornby2005autonomous}.
+There are two paradigms for encoding movements for Aracna, increasing
+its flexibility as a testing platform. The first method is to specify
+explicitly a sequence of positions over time for all eight servos.
+The second method is to set the speed of each servo. This method is
+possible because with the four bar mechanism a servo constantly
+rotating in one direction will move the joint back and forth between
+its minimum and maximum opening angle. This latter method provides a
+much smaller search space and would encourage regular gaits, which
+have been shown to be beneficial when evolving gaits for legged
+robots~\citep{clune2011performance, hornby2005autonomous}.
 
@@
 
-Evolutionary algorithms work best when they have a gradient to follow, instead of searching for a needle in a haystack.
-To get a sense of how randomly-generated gaits would perform, we randomly chose a few parameters....INSERT DESCRIPTION OF WHAT YOU DID, HOW IT WORKED, AND A LINK TO THE VIDEO.
+Evolutionary algorithms work best when they have a gradient to follow
+through a space rich with partial solutions. To get a sense of how
+randomly-generated gaits would perform, we chose a few gaits by
+setting random positions and having the robot interpolate between them
+in a repeated pattern. We found that many such patterns resulted in
+motion. Videos of several gaits are available on the
+website~\citep{WEB}.
 
@@
 
-Aracna is a relatively low-cost open source platform for evolutionary robotics.
-The complex kinematics along with the open source nature of the robot will provide an interesting and challenging universal platform for comparing walking gait algorithms.
-The updated platform is modular, allowing for low-cost replacement parts and varied leg designs.
-Future work can include modifying a single leg to have different linkage lengths, or to replace a kinematic joint with a compliant, or flexible, joint.
-This platform enables testing an experimantal leg such as this without having to print a complete robot.
-Aracna will enable mutiple users to compare data across a single lightweight, low-cost evolutionary robotic platform.
+Here we have introduced Aracna, a low-cost open source platform for
+evolutionary robotics. The complex kinematics along with the open
+source nature of the robot will provide an interesting and challenging
+platform for comparing walking gait algorithms. The updated
+platform is modular, allowing for low-cost replacement parts and
+varied leg designs. Future work can include modifying a single leg to
+have different linkage lengths, or to replace a kinematic joint with a
+compliant, or flexible, joint.  Aracna will enable multiple users to
+compare data across a single lightweight, low-cost evolutionary
+robotic platform.
 

### 2012-05-24 paper_aimsa/introduction.tex
Skeleton version of AIMSA paper

\section{Introduction}

What the problem is...

\figp{aracna_black_2}{.75}{Example figure...}

### 2012-05-24 paper_aimsa/main.tex
Skeleton version of AIMSA paper

--- 2012-05-20 paper_aracna/main.tex
@@
 \documentclass[letterpaper]{article}
-\usepackage{natbib,alife13}
+\usepackage{natbib,alife13mod}
 \usepackage{ownstyles}
 
-\graphicspath{{../figures/}}
+\graphicspath{
+  {../figures/}
+  {../figures/aracna_legs/}
+}
 
 \title{Aracna: An Open-Source Quadruped Platform for Evolutionary Robotics}
-\author{Sara Lohmann, Eric Gold, Jason Yosinski, Jeremy Blum \and Hod Lipson \\
+\author{Sara Lohmann\thanks{~~Authors contributed equally to this work.}~, Jason Yosinski$^{*}$, Eric Gold, Jeff Clune, Jeremy Blum \and Hod Lipson  \\
   \mbox{}\\
 Cornell University, 239 Upson Hall, Ithaca, NY 14853 \\
-\texttt{sml253@cornell.edu}}
+\texttt{sml253@cornell.edu, yosinski@cs.cornell.edu}}
 
@@
 
+%%%%%%%%%%%%%
+% Abstract
+%%%%%%%%%%%%%
+
 \begin{abstract}
-We describe a new, quadruped robot platform, Aracna,
-which requires non-intuitive motor commands in order to locomote and thus provides an interesting challenge for gait learning algorithms, such as those frequently developed in the Evolutionary Computation and Artificial Life communities.
-Aracna is an open-source hardware project composed of off-the-shelf and 3D-printed parts, enabling other research teams to modify its design according to their scientific needs.
-Aracna was designed to avoid the Achilles Heel of a previous quadruped robot platform, whose legs were so heavy that the motors could not reliably execute the commands sent to them.
-Aracna minimizes leg inertia to avoid this problem by having all motors located in the body core instead of on the legs.
-Specifically, each of the four legs has two joints separately controlled by separate four-bar linkage
-mechanisms that drive the pitch of the hip joint and knee joint remotely from motors in the robot's core. 
-%The eight kinematic degrees of freedom in Aracna are controlled by its eight servos. 
-This novel design causes unconventional kinematics, however, creating an opportunity for gait-learning algorithms, which excel in counter-intuitive design spaces where human engineers tend to underperform.
-Because it is low-cost, flexible, kinematically interesting, and and improvement over a previous design, Aracna provides a useful new hardware platform for testing algorithms that automatically generate robotic behaviors.
+\input{abstract}
 \end{abstract}
 
-\section{Introduction}
+%%%%%%%%%%%%%
+% Main Paper
+%%%%%%%%%%%%%
 
-There is a long history in the Artificial Life and Evolutionary Robotics community of automatically generating behaviors for robots~\citep{nolfi2000evolutionary, pfeifer2007body, sims1994evolving, hornby2005autonomous, lipson2000automatic}.
-Much work has focused on evolving gaits for legged robots~\citep{clune2009evolving, clune2011performance, hornby2005autonomous, hornby2003generative, kodjabachian1998evolution, Koos2012, bongard2006resilient, yosinski2011gaits, gallagher1996application}.
-While some of this previous work involved evolution directly on a physical robot~\citep{yosinski2011gaits, zykov2004evolving}, more often a gait was evolved in simulation and then transferred to the physical robot~\citep{lipson2006evolutionary, Koos2012, hornby2005autonomous, bongard2006resilient}.
-Many of these studies report that evolutionary algorithms produced gaits that outperformed those designed by a human engineer~\citep{yosinski2011gaits, hornby2005autonomous}, which is not surprising given that evolutionary algorithms routinely create solutions that are superior to manually created solutions~\citep{koza2003genetic}.
+\input{introduction}
 
-%\begin{figure}[t]
-%\begin{center}
-%%\includegraphics[width=.45\textwidth]{aracna_black_1.jpg}
-%\includegraphics[width=\columnwidth]{aracna_black_1.jpg}
-%\caption{Aracna: an open-sourced quadruped robot platform. All
-%  instructions and downloads are publicly available at
-%  http://creativemachines.cornell.edu/aracna .}
-%\label{fig:aracna_black_1}
-%\end{center}
-%\end{figure}
+\input{relatedWork}
 
-\figp{aracna_black_1}{1}{Aracna: an open-sourced quadruped robot
-  platform. All instructions and downloads are publicly available at
-  http://creativemachines.cornell.edu/aracna.}{}
+\input{methods}
 
-The results just mentioned suggest that evolutionary algorithms are a promising approach for generating gaits and other behaviors for physical robots.
-Despite this promise, the field remains small, partly because robots are expensive and they are difficult to modify.
-Access to cheap, customizable robots could increase the number of researchers able to participate in the field.
-Moreover, in nearly all of the papers mentioned previously, the robots were custom-made, preventing teams at other universities from reproducing the results of other groups and or testing new algorithms on a robotic platform used in a previous study.
-That, in turn, slows the progress of science because it is difficult to interpret whether the variance in results between different studies was due to the algorithms used or the robotic platform those algorithms were tested on.
+\input{experimentalSetup}
 
-Some robot platforms are emerging, but they tend to be wheeled robots without complex kinematics, such as the ePuck~\citep{mondada2009puck}.
-Wheeled robots are interesting testbeds for many robotic behaviors, but they do not allow gait evolution and are unable to traverse rugged terrains.
-Legged robotic platforms exist, but they tend to be extremely expensive, such as the Aldebaran Nao, which costs more than \$10,000 USD.
-Another drawback to these commercial platforms is that it is hard, if not impossible, to modify the hardware design because they are not open-source hardware projects, and do not take advantage of off-the-shelf components and 3D printing, meaning that complex manufacturing tools are required to manufacture newly designed parts.
+\input{results}
 
-In this paper we address these needs by introducing \emph{Aracna}, a low-cost, open-source hardware, easily customizable robot platform with non-intuitive walking
-kinematics~(\figref{aracna_black_1}).
-Aracna is the third quadruped robot developed for evolutionary learning algorithms by the Creative Machines Lab at Cornell University~\citep{bongard2006resilient, yosinski2011gaits}.
-Like the most recent of the two previous designs, called QuadraTot~\citep{yosinski2011gaits}, the body of Aracna is 3D printed and the STL files are available online,
-%TODO: INSERT LINK
-meaning that other researchers can easily customize the body's design. As in the both previous designs, each leg has a hip and knee joint controlled by two actuators. The original Creative
-Machines Lab quadruped robot favored starfish-like movements~\citep{bongard2006resilient}. The second quadruped robot --- QuadraTot ---
-developed spider-like movements, but was found to be limited by its
-weight, such that the motors would overheat and timeout when trying to execute many commands~\citep{yosinski2011gaits, Glette2012Evolution}.
-We designed Aracna to be able to produce fast, spider-like movements, yet be lightweight enough that the motors would not overheat.
-We also designed Aracna to be inexpensive: as described below, its overall price is under \$1,400  USD.
-In the following sections we describe the Aracna platform in more detail.
+\input{conclusion}
 
-\section{Overall Hardware Design}
+\input{acknowledgements}
 
-%\begin{figure}[t]
-%\begin{center}
-%\includegraphics[width=\columnwidth]{aracna_cad_ver1.jpg}
-%\caption{A rendered CAD model of Aracna. Note the lack of heavy servos on the legs themselves, which are instead controlled via four-bar linkages by servos in the robot's core.}
-%\label{fig:aracna_cad_ver1.jpg}
-%\end{center}
-%\end{figure}
-
-\figp{aracna_cad_ver1}{1}{A rendered CAD model of Aracna. Note the
-  lack of heavy servos on the legs themselves, which are instead
-  controlled via four-bar linkages by servos in the robot's core.}
-
-%\begin{figure}[t]
-%\begin{center}
-%\includegraphics[width=2.25in,angle=0]{fig4.pdf}
-%\caption{Crank-rocker four-bar linkage controlling flexion/extension of the hip joint.}
-%\label{fig4}
-%\end{center}
-%\end{figure}
-
-The hardware of Aracna was designed to improve upon the previous Creative Machines Lab
-quadruped robots~\citep{bongard2006resilient, yosinski2011gaits}, while still qualitatively resembling those robots.
-Aracna is similar in that it has a body and four legs, with each leg having two joints that can pitch forward and back like knees~(\figref{aracna_cad_ver1}).
-
-One change was to constrain the movement of the
-joints toward the goal of creating faster, spider-like movements. To
-prevent starfish-like movements and instead encourage a walking gait with the robot body permanently off the ground, the legs were constrained such that they cannot straighten out and the knee cannot hyperextend.
-
-Another change was to reduce the both the overall weight of the robot and the weight of each leg.
-Two previous studies that used the QuadraTot robot report that the motors quickly wore out and could not reliably execute the commands sent to them, likely because of both the overall weight of the QuadraTot and the fact that housing servos on the legs made them heavy~\citep{bongard2006resilient, yosinski2011gaits}.
-The weight of the robot's core was reduced in a number of ways.
-
-Initially, we eliminated the QuadraTot's fit-PC, an onboard Linux computer weighing 370g, and replaced it with an on-board ArbotiX microcontroller that weights only 47g.
-As a result, the main processing occurs on an external computer and commands are sent to Aracna wirelessly.
-This change, of course, comes of the cost of the robot being truly autonomous.
-Wireless communication between the external control computer
-and the ArbotiX  microcontroller occurs over wireless XBee.
-
-A second means of eliminating weight involved switching to a lighter battery.
-The QuadraTot had two 12V lithium-ion battery packs that weighted 140g each for a total of 280g.
-Aracna has a single lithium-polymer 11.1V battery that weighs 122 g.
-INSERT SENTENCE ON THE COST/PENALTY OF USING THE SMALLER BATTERY.
-
-A major modification, targeted at reducing the weight of legs, was the use of two four-bar mechanisms to drive the joints in each
-leg. This mechanism causes the controlled joint to move at a fraction
-of the output angle of the actuator, giving the motor a relatively
-larger mechanical advantage over the position of each leg.
-\figref{aracna_leg_drawing_outer} shows the crank-rocker system, where the input crank
-is actuated by a servo and the rocker is the leg. This
-configuration allows the servo motors to be contained in the robot core, reducing both the inertia and mass of each leg. The weight of an Aracna leg is 105g compared to the 217g for a QuadraTot leg.
-
-Combined, these changes to minimize weight led to a 31.4 percent reduction in weight of the robot. The QuadraTot weighs 1.88kg whereas Aracna weighs 1.29kg.
-
-A final change was to upgrade the power of the servo motors in order to increase the ability of the robot to strike whichever configurations are specified by the learning algorithms.
-Specifically, we upgraded from Dynamixel AX 12+ motors to AX-18A motors, which have a higher stall torque (1.8Nm vs.
-1.5Nm at 12V), a higher stall current (2.2A vs 1.5A), and a higher no load speed (97 vs.
-59 RPM).
-
-%\begin{figure}[t]
-%\begin{center}
-%\includegraphics[width=.23\textwidth]{aracna_leg_drawing_outer}
-%\includegraphics[width=.23\textwidth]{aracna_leg_drawing_inner}
-%\caption{Crank-rocker four-bar linkage controlling flexion/extension of
-%  the knee and hip joints. In both cases above, the input crank (link
-%  OA) is actuated by a servo, the rocker is the leg (link CB), and the
-%  fixed link is OC.}
-%\label{fig:crankRocker}
-%\end{center}
-%\end{figure}
-
-\figgp{aracna_leg_drawing_outer}{.46}{aracna_leg_drawing_inner}{.46}{Crank-rocker
-  four-bar linkage controlling flexion/extension of the knee and hip
-  joints. In both cases above, the input crank (link OA) is actuated
-  by a servo, the rocker is the leg (link CB), and the fixed link is
-  OC.}
-
-\section{3D Printed Body}
-
-The body of Arcana takes advantage of 3D printing technology, also known as additive manufacturing, which generates physical objects from digital designs by building them up layer by layer~\citep{gibson2009additive, lipson2010factory}.
-The use of 3D printing means that other Aracna users can easily make copies of Aracna, either by having access to a 3D printer or via online 3D printing services such as Shapeways.com.
-Either option requires the 3D design files in the stereolithography (STL) format, which we are publishing in the online support material for this paper~\citep{WEB}.
-Moreover, to catalyze innovation in this open-source hardware project, we are also providing the source files for the SolidWorks computer-aided-design program, to enable others to modify the design.
-It should thus be easy for future Aracna users to improve or alter the design and quickly obtain a physical instantiation of the design.
-Importantly, the use of 3D printing eliminates the need to know how to machine parts, allowing many more researchers to participate in using physical robot morphologies that they design themselves.
-These ideas are in line with a broader trend toward enabling non-technical users to design and manufacture physical objects~\citep{clune2011objects, clune2011endless, lipson2010factory}.
-
-An initial version of Aracna was designed to be printed in one piece~(\figref{aracna_notYetCleanedOnePieceRobot}).
-However, if one part of the robot became damaged, an entire new robot had to be reprinted, which took over 26 hours and costs roughly \$355 USD on an Objet Connex500 printer.
-To make repairing the robot easier, cheaper, and quicker, Aracna was redesigned to be modular.
-It consists of 15 pieces--four legs and the core--that can be separately 3D printed~(\figref{aracna_notYetCleanedMultiPieceRobot}).
-Printing a leg takes 3.3 hours and costs roughly \$64 USD.
-Printing the core takes 3 hours and costs \$101 USD.
-All 15 Aracna pieces can still be printed as one print job, with an overall time of approximately 10 hours and cost of \$308.
-These figures are based on Aracna's use of  approximately 967g of model material and 746g of support material, and current material costs of 4.5g per USD and 8g per USD for rigid and support material, respectively.
-This cost estimate is variable depending on the type of material used.
-The print times are estimates calculated by the Objet's software and are meant to be used as a relative comparison of print times.
-Table~\ref{tab:cost} outlines the total estimated cost of Aracna, which is just under \$1,400, including its off-the-shelf electronic components.
-
-%\begin{figure}[t]
-%\begin{center}
-%\includegraphics[width=.37\textwidth]{aracna_notYetCleanedOnePieceRobot.jpg}
-%\caption{A draft version of Aracna that was printed in one piece.
-This non-modular design proved expensive to maintain if part of the robot was damaged, and was replaced in a later version with a modular design.
-This image is of the printed body with support material still present.}
-%\label{fig:aracna_notYetCleanedOnePieceRobot}
-%\end{center}
-%\end{figure}
-
-\figp{aracna_notYetCleanedOnePieceRobot}{.74}{A draft version of
-  Aracna that was printed in one piece. This non-modular design proved
-  expensive to maintain if part of the robot was damaged, and was
-  replaced in a later version with a modular design. This image is of
-  the printed body with support material still present.}
-
-%\begin{figure}[t]
-%\begin{center}
-%\includegraphics[width=.37\textwidth]{aracna_notYetCleanedMultiPieceRobot.jpg}
-%\caption{A final version of Aracna printed in multiple pieces.The body is printed as two pieces, with 9 smaller parts within the top piece to reduce support material and print time.
-These parts can be printed individually if replacements are necessary.
-The body consists of a total of 11 parts.
-This image shows a set of 12 printed parts (the complete body and a single leg) with support material still present.}
-%\label{fig:aracna_notYetCleanedMultiPieceRobot}
-%\end{center}
-%\end{figure}
-
-\figp{aracna_notYetCleanedMultiPieceRobot}{.74}{A final version of
-  Aracna printed in multiple pieces.The body is printed as two pieces,
-  with 9 smaller parts within the top piece to reduce support material
-  and print time. These parts can be printed individually if
-  replacements are necessary. The body consists of a total of 11
-  parts. This image shows a set of 12 printed parts (the complete body
-  and a single leg) with support material still present.}
-
-\begin{table}[h]
-\center{
-\begin{tabular}{|c|c|}
-\hline
-Part & Cost\\
-\hline\hline
-3D Print Materials & \$308\\
-ArbotiX Robocontroller Kit & \$189\\
-Dynamixel AX-18A Robot Actuator (x8) & \$721\\
-3S 11.1V 2000mAh Pro Lite LiPo Battery & \$73\\
-LiPo Battery Balance Charger Kit & \$70\\
-Cables, Connectors, Misc & \$28\\
-\hline\hline
-\bf Total & \bf \$1389\\
- \hline
-\end{tabular}
-}
-\vskip 0.25cm
-\caption{Estimated total cost. The cost of components and printing material reflect market prices from March 2012. A complete parts list is on our website \citep{WEB}.}
-\tablabel{cost}
-\end{table}
-
-\section{Control}
-
-In addition to reducing the weight of the legs, the four-bar
-mechanisms also satisfied the design goal of making a robot that had
-non-traditional movements.
-Unusual kinematics make for a more effective evolutionary robotics test platform, since evolutionary algorithms are most helpful in domains that humans find hard to program solutions for.
-The reason Aracna's kinematics are counter-intuitive is because there is a complicated mapping between each servo's output and the movement of the joint controlled by that servo~(Figure~INSERTjointAnglesAsAFunctionOfServoPositions).
-INSERT-OTHER-REASONS-THE-KINEMATICS-ARE-STRANGE.
-
-The range of motion for the hip joint is INSERT degrees and that of the knee joint is INSERT degrees~(Figure~INSERT).
-The range of the knee joint technically depends on the position of the hip joint, but the effect is slight enough (less than INSERT) degrees, that it can be effectively ignored.
-
-There are two different ways to encode movements for Aracna, increasing its flexibility as a testing platform.
-The first method is to specify a sequence of positions over time for all eight servos.
-In this method the joint angles would be determined as shown in Figure~INSERTjointAnglesAsAFunctionOfServoPositions).
-The second method is to set the speed of each servo (and possibly its initial position).
-This method is possible because with the four bar mechanism a servo constantly rotating in one direction will move the joint back and forth between its minimum and maximum opening angle.
-This latter method provides a much smaller search space and would encourage regular gaits, which have been shown to be beneficial when evolving gaits for legged robots~\citep{clune2011performance, hornby2005autonomous}.
-
-\section{Software}
-
-The software, which is also open-source and freely available~\citep{WEB}, is written in Python and based on the
-code developed for the QuadraTot platform~\citep{yosinski2011gaits}.
-The software translates a series of requested joint angles from the learning algorithm into servo movements.
-Additionally, it returns information to the learning algorithm, such as the distance traveled or the specific trajectory the robot took, so the learning algorithm can assess the quality of the gait.
-To provide this information, an infrared light
-emitting diode (LED) was placed on the robot and a Nintendo Wii
-remote was attached overhead. The software uses the combination of the two to ascertain the X, Y position of the robot. The software is interoperable with any gait or behavior learning algorithm.
-
-\section{Example Gaits}
-
-Evolutionary algorithms work best when they have a gradient to follow, instead of searching for a needle in a haystack.
-To get a sense of how randomly-generated gaits would perform, we randomly chose a few parameters....INSERT DESCRIPTION OF WHAT YOU DID, HOW IT WORKED, AND A LINK TO THE VIDEO.
-
-\section{Conclusion}
-
-I think we should wait until the rest of the paper is fleshed out before writing this.
-
-\section{Acknowledgements}
-
-This work was supported by the National Science Foundation's Office of
-Emerging Frontiers in Research and Innovation (grant number 0735953) and an NSF Postdoctoral Research Fellowship in Biology to Jeff Clune (DBI-1003220).
+%%%%%%%%%%%%%
+% References
+%%%%%%%%%%%%%
 

### 2012-05-24 paper_aimsa/results.tex
Skeleton version of AIMSA paper

\section{Results}

how well it worked, plots, ...

### 2012-05-24 code/simrun.py
added the frequency reducer hack in simrun.py

@@ -13,14 +13,15 @@ class simrun:
 
     def __init__(self):
         self.filename = "output.txt"
- 
+        self.freq_threshold = 216 # 9 servos, 12 seconds at 2Hz
+        
     def runSim(self, gaitFunction, filename):
 
         try:
             os.remove('/home/sean/simtest/input.txt')
         except:
             pass
-        
+
         ff = file('input.txt', "w")
         timeMax = 12.0
         timeDiv = 1.0/60.0
@@ -45,15 +46,23 @@ class simrun:
                            
         ff.close()    
         os.system('./crossSim -i input.txt -o output.txt -n')
-        #os.system('./crossSim -i input.txt -o output.txt')
-        dist = self.getDist()      
+        outputData = self.getOutput()
+        dist = self.getDist(outputData)
+        counter = self.freqcheck(outputData)
+        if counter > self.freq_threshold:
+            dist = dist*0.5
+        elif counter > self.freq_threshold*0.8:
+            dist = dist*0.8
+        
         shutil.move('/home/sean/simtest/input.txt', '/home/sean/simtest/%s' % filename)
         return dist
-    
-    def getDist(self):
 
+    def getOutput(self):
         outputRaw= file(self.filename,"r")
         outputData = [line.split() for line in outputRaw]
+        return outputData
+
+    def getDist(self, outputData):
         outputColLen = len(outputData)-1
         outputRowLen = len(outputData[1])-1
         posBeg = [outputData[1][outputRowLen-2], outputData[1][outputRowLen-1], outputData[1][outputRowLen]]
@@ -64,6 +73,35 @@ class simrun:
         zdist = float(posEnd[2])-float(posBeg[2])
 #        zdist = float(self.posEnd[2])-float(self.posBeg[2])
         return math.sqrt(math.pow(xdist,2)+math.pow(zdist,2))
+
+    def freqcheck(self, data):
+	'''checks for the number of inflection points in the servo positions'''
+	pos = numpy.zeros(shape=(18,10))
+        for i in range(18):
[…]

### 2012-05-24 code/simrun.py
fully debugged version of simrun.py with frequency limiter

@@ -3,6 +3,7 @@ import random
 import os
 import shutil
 import string
+import numpy
 from datetime import datetime
 from copy import copy
 from time import sleep
@@ -75,13 +76,13 @@ class simrun:
         return math.sqrt(math.pow(xdist,2)+math.pow(zdist,2))
 
     def freqcheck(self, data):
-	'''checks for the number of inflection points in the servo positions'''
-	pos = numpy.zeros(shape=(18,10))
-        for i in range(18):
+      	'''checks for the number of inflection points in the servo positions'''
+        pos = numpy.zeros(shape=(720,10))
+        for i in range(720):
             for ii in range(10):
-                pos[i][ii] = float(data[i][ii])
+                pos[i][ii] = float(data[i+1][ii])
         counter = 0
-        for i in range(18): # 720 lines in the data file (12 seconds at 60Hz)
+        for i in range(720): # 720 lines in the data file (12 seconds at 60Hz)
             # [time servo servo servo servo servo servo servo servo servo]
             if  (i != 0 and i != 1) and ((pos[i][1]-pos[i-1][1] > 0 and pos[i-1][1]-pos[i-2][1] < 0) or (pos[i][1]-pos[i-1][1] < 0 and pos[i-1][1]-pos[i-2][1] > 0)):
                 counter += 1

### 2012-05-24 code/Load_trajectory.m
5/23 (rebased and edited by Jason to remove core and octave-core)

@@ -24,7 +24,6 @@ function traj = Load_trajectory(filename) % E.g. 'data/001/output_001.txt'
   hFig = figure('Name', 'Individual axis trajectories', 'position', [1260, 200, 600, 900]); hold on; box on; grid on;
   subplot(3,1,1);
   disp('111');
-  keyboard;
   plot(traj(:,1), traj(:,2), 'linewidth', 2);
   disp('222');
   xlabel('time'); ylabel('x');

### 2012-05-24 code/PoWER_QuadraTot_23.m

@@ -10,7 +10,7 @@ clear all;
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
 for i=1:1
-    relearn = 0;
+    relearn = 1;
     [Return, s_Return, param, rl] = RL_PoWER(4, 4, relearn);
 end
 

### 2012-05-24 code/RL_PoWER.m

@@ -4,11 +4,11 @@ function [Return, s_Return, param, rl] = RL_PoWER(fromKnots, toKnots, relearn, R
 disp('Starting RL...');
 %  rl_m = m; % make a copy of m
 
-hFig = figure('Name', 'Rollouts', 'position', [50, 100, 800, 600]);
-hFigResults = figure('Name', 'Results', 'position',[1000,600,800,400]); axis on; grid on; hold on;
+%hFig = figure('Name', 'Rollouts', 'position', [50, 100, 800, 600]);
+%hFigResults = figure('Name', 'Results', 'position',[1000,600,800,400]); axis on; grid on; hold on;
 hFigExported = figure('Name', 'Exported trajectory', 'position', [850, 100, 600, 400]); hold on; box on; grid on;
 
-plotRollouts(hFig, 0, 0, 0, 0);
+% plotRollouts(hFig, 0, 0, 0, 0);
 
 settings = getSettings();
 
@@ -16,7 +16,7 @@ settings = getSettings();
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
 % number of iterations
-n_iter = 60;
+n_iter = 300;
 % number of parameters of the policy = number of spline knots here
 n_splines = 8;
 n_rfs = n_splines*(fromKnots - 1);  % because the last knot of the spline y(n) = y(1)
@@ -59,7 +59,7 @@ for si=1:n_splines
   policy(1).s(si).y = initialPos(si)*ones(1,policy(1).s(si).n); % set initial policy y values to the middle 0.5
   % now turn it into a cyclic spline
   [policy(1).s(si).pp, policy(1).s(si).x, policy(1).s(si).y, policy(1).s(si).n] = getCyclicSplinePlus6e(policy(1).s(si).x, policy(1).s(si).y);
-  plotSpline(policy(1).s(si).pp, policy(1).s(si).n, policy(1).s(si).x, policy(1).s(si).y, 'green', 3); % visualize the spline
+  % plotSpline(policy(1).s(si).pp, policy(1).s(si).n, policy(1).s(si).x, policy(1).s(si).y, 'green', 3); % visualize the spline
 end
 %pause
 
@@ -84,9 +84,11 @@ if relearn==1 % resume learning of a previous session
 else
     iter = 1;
 end
+best = 1;
+counter = 0;
[…]

### 2012-05-24 code/ReturnOfRollout.m

@@ -12,11 +12,13 @@ function r = ReturnOfRollout(traj)
   % avgHeight = mean(traj(:,4))
   % r = exp(avgHeight);
   % return
-
-  avgHeight = mean(traj(:,4))
-  distTraveled = norm(traj(end,2:4) - traj(1,2:4)) % in [cm]
+  %disp(traj(:,2))
+  %disp(traj(:,3))
+  %disp(traj(:,4))
+  avgHeight = mean(traj(:,3))
+  distTraveled = norm((traj(end,2:4) - traj(1,2:4))/5.6351) % in [cm]
   duration = traj(end,1) - traj(1,1) % in [s]
-  avgSpeed = distTraveled / duration % in [cm/s]
+  avgSpeed = distTraveled / (duration) % in [cm/s]
 
   r = avgSpeed;
 %  r = (10*avgSpeed + avgHeight)^10; % combining speed and height

### 2012-05-24 code/Robot.py

@@ -462,7 +462,6 @@ class Robot():
             ret[ii] = int(duple[0])
             ret[ii+1] = int(duple[1])
 
-        cropWarning = True
         if cropWarning and ret != position:
             print 'Warning: cropped %s to %s' % (repr(position), repr(ret))
 

### 2012-05-24 code/exportTrajectoryWithoutRescaling.m

@@ -36,10 +36,10 @@ function exportTrajectoryWithoutRescaling(policy, hFig, filepath);
   % plot the 'Exported trajectory' velocity and acceleration
   figure(hFig); clf; hold on; box on; grid on;
   clrmap = colormap(jet(policy.n_splines));
-  for si=1:policy.n_splines
-    plot(totalTime, s(si).totalPos, 'color', clrmap(si,:), 'linewidth', 2); % all cycles
-    plot(cycleTime, s(si).cyclePos, 'color', clrmap(si,:), 'linewidth', 4); % just the first cycle is thicker, to highlight it
-  end
+  %for si=1:policy.n_splines
+    %plot(totalTime, s(si).totalPos, 'color', clrmap(si,:), 'linewidth', 2); % all cycles
+    %plot(cycleTime, s(si).cyclePos, 'color', clrmap(si,:), 'linewidth', 4); % just the first cycle is thicker, to highlight it
+  %end
   
   % Generate file input.txt for the simulator
   inputData = [];

### 2012-05-24 code/extract.py

@@ -1,0 +1,40 @@
+#! /usr/bin/env python
+
+from os import listdir
+import matplotlib.pyplot as plt
+import re
+
+def main():
+    path = '/home/team/results/20120523_snd250Run/'
+    trials = sorted(listdir (path))[4:]
+    speeds = []
+    pattern = re.compile("[0-9]{3}")
+    for trial in trials:
+        if pattern.match(trial):
+            output = open(path + trial + '/output.txt' )
+            print path, trial
+            print ("processing " + trial)
+            lines = output.readlines()
+            length =len(lines)
+            ss = []
+            for ii, line in enumerate(lines):
+                if ii == 1 or ii == length-1 :
+                    t = line.split()
+                    xy = [float(t[19]),float(t[20])]
+                    ss.append(xy)
+            speed = ((( (ss[1][0]-ss[0][0])/5.6)**2+( (ss[1][1]-ss[0][1])/5.6 )**2)**0.5)/12
+            print speed
+            if len(speeds) > 0 and speeds[len(speeds)-1]-speed > speed*5:
+                speeds.append(speeds[len(speeds)-1]*0.85)
+            else:
+                speeds.append(speed)
+            print("processing complete: "+trial)
+
+    print len(speeds)
+    plt.plot(range(1,len(speeds)+1), speeds)
+    plt.ylabel('SPEED')
+    plt.xlabel('Trial')
+    plt.show()
+
+if __name__ == '__main__':
+	main();

### 2012-05-24 code/runGaitAndLogPos.py

@@ -13,7 +13,7 @@ from Robot import Robot
 from Strategy import TimedFileStrategy
 from wii.WiiTrackFastClient import WiiTrackFastClient
 from datetime import datetime, timedelta
-
+from Robot import RobotFailure
 
 def savePosition(wiiTrack, posAgeList):
     position,age = wiiTrack.getPosAge()
@@ -23,6 +23,7 @@ def savePosition(wiiTrack, posAgeList):
         return '-1 -1 %f' % age
     else:
         timeOfPosition = datetime.now() - timedelta(seconds = age)
+
         posAgeList.append((timeOfPosition, position, age))
         #print len(posAgeList)
         return ' '.join([str(x) for x in [timeOfPosition] + position + [age]])
@@ -53,24 +54,30 @@ def main():
     if age is None:
         raise Exception('Could not get position from wiiTrack.')
 
-    robot = Robot(loud = True)
+    #robot = Robot(loud = True)
+    robot = Robot()
+
+
+    ##############
     bucket = []
 
     def foo():
         savePosition(wiiTrack, bucket)
 
     robot.run(motionFunction, runSeconds = 10, resetFirst = True,
-              interpBegin = 2, interpEnd = 2, extraLogInfoFn = foo)
-
-    print 'Positions:'
-    print len(bucket)
+                               interpBegin = 2, interpEnd = 2, extraLogInfoFn = foo)
 
     relTimeBucket = []
     for ii, line in enumerate(bucket):
         delta = line[0] - bucket[0][0]
-        relTime = delta.seconds + delta.microseconds/1e6
-        relTimeBucket.append((relTime, line[1], line[2]))
-
+        if delta.microseconds/1e6 < 1:
+                    relTime = delta.seconds + delta.microseconds/1e6
+                    relTimeBucket.append((relTime, line[1], line[2]))
+        #pdb.set_trace()
+    print relTimeBucket[0]
+    if list(relTimeBucket[-1])[0]-list(relTimeBucket[0])[0] < 13.7:
+        exit(2)
+        ##############
 
     ff = open (posFile, 'w')
     ff.write('# time (junk junk)x9 pos.x pos.y 0 age\n')

### 2012-05-24 code/runRobot.m

@@ -1,6 +1,28 @@
-function runRobot()
+function runRobot(filepath)
   % TODO: define this function. It should just call the python executable.
 
   settings = getSettings();
+
+  disp(settings.robotDir);
+  disp(settings.robotExecutable);
+
+  completeInputFilename = [filepath 'input.txt'];
+  completeOutputFilename = [filepath 'output.txt'];
+
+  if isunix
+    command = [settings.robotDir settings.robotExecutable ' ' completeInputFilename ' ' completeOutputFilename];
+    disp(sprintf('Running command: %s', command));
+    
+    % Takes ~12 seconds, at least
+ 
+    [status, out] = system(command);
+    while status ~= 0
+      warning(sprintf('Exit code was %d, press Enter to redo the run', status));
+      pause();
+      [status, out] = system(command);
+    end
+    disp('Done running robot!');
+  else
+    error('No windows implementation for this yet');
 
 end

### 2012-05-24 code/wii/WiiTrackFastClient.py

@@ -72,8 +72,8 @@ class WiiTrackFastClient(object):
 
             if len(data) < 1:
                 print 'Could not contact server'
-                None
-
+                continue
+            
             if data[0] == 'E':
                 # No target, or too many targets
                 #print 'Server reported error:', data

### 2012-05-30 paper_aimsa/relatedWork.tex

\section{Related Work}

Other attempts to solve it

### 2012-05-30 code/Robot.py
Fixed Robot bug for non-sequential expectedIds

@@ -101,16 +101,19 @@ class Robot():
 
         self.actuators   = []
         self.actuatorIds = []
-
+        self.actuatorIDsOnNetwork = []
+        
         for dyn in self.net.get_dynamixels():
-            print dyn.id,
-            self.actuatorIds.append(dyn.id)
-            self.actuators.append(self.net[dyn.id])
+            self.actuatorIDsOnNetwork.append(dyn.id)
+            if dyn.id in self.expectedIds:
+                print dyn.id,
+                self.actuatorIds.append(dyn.id)
+                self.actuators.append(self.net[dyn.id])
         print "...Done"
 
         if len(self.actuators) != self.nServos and not self.silentNetFail:
-            raise RobotFailure('Expected to find %d servos on network, but only got %d (%s)'
-                               % (self.nServos, len(self.actuators), repr(self.actuatorIds)))
+            raise RobotFailure('Expected to find servos %s on network, but only got %s (scanned from %d to %d)'
+                               % (self.expectedIds, repr(self.actuatorIDsOnNetwork), min(self.expectedIds), max(self.expectedIds)))
 
         for actuator in self.actuators:
             #actuator.moving_speed = 90

### 2012-05-30 code/extract.py
Added lines to save figures

@@ -34,6 +34,9 @@ def main():
     plt.plot(range(1,len(speeds)+1), speeds)
     plt.ylabel('SPEED')
     plt.xlabel('Trial')
+    plt.savefig('savedPlot.eps')
+    plt.savefig('savedPlot.pdf')
+    plt.savefig('savedPlot.png')
     plt.show()
 
 if __name__ == '__main__':

### 2012-05-31 paper_aimsa/extract.py
5/31/2012

@@ -1,0 +1,51 @@
+#! /usr/bin/env python-32
+
+from os import listdir
+import matplotlib as mpl 
+mpl.use('cocoaagg') 
+import matplotlib.pyplot as plt
+import re
+
+def main():
+    path = '/Users/hs454/Desktop/tmp/'
+    runs = sorted(listdir (path))
+    total = []
+    avg = []
+    avg_all=[]
+    pattern = re.compile("return_[0-9]*")
+    bucket = []
+    best = 0
+    for run in runs:
+        if pattern.match(run):
+            speeds=[]
+            f = open(run)
+            lines = f.readlines()
+            for line in lines:
+                spd_tr= line.split()
+                ts= (int(spd_tr[1]),float(spd_tr[0])*(14.0/12))
+                speeds.append(ts)
+            speeds=sorted(list(set(sorted(speeds))))
+            total.append(speeds)
+            print speeds[0]
+            max_p = speeds[-1]
+            print max_p
+    for i in range(0,300):
+        mean = (list(total[0][i])[1]+list(total[1][i])[1]+list(total[2][i])[1])/3
+        avg.append(mean)
+        avg_all.append((mean,i))
+        
+    best = 0
+    best_index=0
+    for x,ii in avg_all:
+        if best < x:
+            best = x
+            best_index = ii
+    
+    plt.plot(range(1,len(avg)+1), avg)
+    plt.ylabel('AVERAGE SPEED')
+    plt.xlabel('Trial')
+    plt.annotate('global max: '+str(best), xy=(best_index,best), xytext=(best_index+15, best+0.25),arrowprops=dict(facecolor='black', shrink=0.05))
+    plt.show()
+
+if __name__ == '__main__':
+	main();

### 2012-05-31 paper_aimsa/extractGait.py

@@ -1,0 +1,41 @@
+#! /usr/bin/env python-32
+
+from os import listdir
+import matplotlib as mpl 
+mpl.use('cocoaagg') 
+import matplotlib.pyplot as plt
+import re
+
+def main():
+    path = '/Users/hs454/Desktop/tmp/input.txt'
+    f= open(path)
+    lines = f.readlines()
+    t = []
+    j1 = []
+    j2 = []
+    j3 = []
+    j4 = []
+    j5 = []
+    j6 = []
+    j7 = []
+    j8 = []
+    for line in lines:
+        joints = line.split()
+        t.append(float(joints[0])-1)
+        j1.append(float(joints[1]))
+        j2.append(float(joints[2]))
+        j3.append(float(joints[3]))
+        j4.append(float(joints[4]))
+        j5.append(float(joints[5]))
+        j6.append(float(joints[6]))
+        j7.append(float(joints[7]))
+        j8.append(float(joints[8]))
+    
+    
+    plt.plot(t,j1,'r', t,j2,'b', t,j3,'g', t,j4,'k', t,j5,'c', t,j6,'m', t,j7,'y', t,j8, 'g^')
+    plt.ylabel('Motor Position')
+    plt.xlabel('Time(s)')
+    plt.show()
+    
+if __name__ == '__main__':
+	main();

### 2012-05-31 paper_aimsa/experimentalSetup.tex
Skeleton version of AIMSA paper
5/31/2012

--- 2011-04-14 paper_ecal/experimentalSetup.tex
@@
 \section{Experimental Setup}
-\seclabel{experimentalSetup}
 
-We next describe the details of the experimental system.
+hardware details, ...
+Setup:
+The quadruped robot used was assembled from parts purchased online and parts printed by the Objet Connex 500 3-D Printing System.
+The robot actuation system consists of 5 AX-18+ Dynamixel servos and 4 AX-12+ Dynamixel servos: one inner joint with one AX-18+ servo and one outer joint with AX-12+servo in each of the four legs, and one AX-18+ servo at the center.
+To avoid the formerly reported AX-18+ servos are used in this robot because of their stronger actuation power than that of AX-12+.
+Each servo could be set to a position in the range [0, 1023] by using PyDynamixel library, corresponding to a physical range [-120, +120].
+Also, to prevent collisions with the robot body, the control module filter out the commands to a safe range.
+This range was [-85, +60 ] for the inner leg servos, [-113 , +39 ] for the outer leg servos, and [-28 , +28] for the center hip servo.
+In the studies of this paper, tethered cables powered both the computer and the servos.
+It measures approximately 39.5 centimeters from leg to opposite leg in the crouch position showed in [pic].
 
-%: \secref{platformDetails} discusses the robotic
-%platform, and \secref{fitnessEvaluation} describes the tracking system
-%we used to measure each gait's fitness.
+Our performance metric was the displacement over the evaluation period of 12 seconds for each.
+Same as Jason et al described in [xx], the displacement was measured using a Wii remote that was placed on the ceiling.
+Different from the original model described in [xx], the quadruped robot was equipped with a three-infrared-LED cluster on top rather than just one.
+The reason for this setup is that when fierce gaits were executed, the Wii remote loses tracking of the robot position due to the limited visible angle of a single LED.
+These three LEDs were place tightly together to act as one signal emitter.
+Each LED was tilted outwards in order to maximize the visible range.
+[pic]
 
-% Given the goal of producing a deterministic, feedforward motion
-% function $g(t)$, we now turn to several important considerations.
-% \secref{learningMethods} talks about how this function is generated,
-% \secref{platformDetails} discusses the hardware platform on which it
-% is run, and \secref{fitnessEvaluation} details exactly how its
-% fitness is measured.
+A separate tracking server ran on the robot PC interacted with the Wii remote via bluetooth by using the CWiid library.
+A corresponding client communicates with this server via a socket connection.
+Our fitness evaluation code talks with the Wii remote by using this client and updates the position from beginning to end during each run.
 
-%\input{platformDetails}
+The metric for evaluating gaits was the Euclidian distance the robot traveled during a 12-second run on flat surface.
+Previous work done in [xx] and [xx1(k)] suggested that extremely fierce are not viable in general.
+These gaits tend to either overburden the servos or flip the robot.
+Thus, RL_POWER(?) were tested after carefully cropping out the extremely fierce gaits each time.
+(As described earlier, each leg has two joints, inner joint j1 and outer j2.
+The position of one leg is determined by the sum of the values of j1 and j2.
 
-%\input{fitnessEvaluation}
+Extremely fierce gaits usually have relatively small values of g.
+Thus cropping works by mapping the wild gaits onto the boundary of a triangular space over the two dimensional space of j1 and j2.
+[pic of j1 j2 space] [g = j1 + j2] caption

### 2012-05-31 paper_aimsa/introduction.tex
5/31/2012

@@
 \figp{aracna_black_2}{.75}{Example figure...}
+
+General learning:
+Learning gaits 
+Various learning algorithms have been tested to be effective for legged robots. 
+Their studies tested the viability of applying RL method to legged robot gaits learning.
+Competitive performance has been verified on methods such as HyperNEAT (Yosinski et al, 2011) and GA(Chernova et al, 2004) and other  methods(Hornby et al., 2005; Zykov et al., 2004; T´ellez et al., 2006; Valsalam and Miikkulainen, 2008).
+Despite the excellent performance of these algorithms, a major part of achieving good performance relies in tuning the parameters for the evolutionary algorithms(Kormushev et al, 2006).
+Here we present a different way for learning gaits using RL(Hill climbing).
+In our experiment, it turned out this method achieved competitive performance with
+Problem Definition:
+The gait-learning problem is defined to find a gait that maximizes some specific metric.
+Mathematically, we define a gait as a function that specifies a vector of commanded motor positions for a robot over time.
+Gaits without feedback --- also called open-loop gaits --- can be defined as
+
+\be
+\vec{x} = g(t)
+\ee 
+According to the definition, open-loop gaits are deterministic.
+One particular gait should behave exactly the same when it is run from trial to trial.
+However, the actual robot motion and fitness measured will vary due to the noisiness of the real world physics.
+In our trials, one gait pattern was generated and sent to the robot and executed in an open loop manner, as defined.
+In particular, in this paper, we chose to compare open-loop gaits generated specifically by the RL_POWER and HyperNEAT.
+The particular metric used will be described later.

### 2012-05-31 paper_aimsa/methods.tex
5/31/2012

\section{Methods}

how we tried to solve it...
Parameterized Gaits by RL_PoWER:

Here we used an RL approach to change the complexity of the policy representation dynamically while the trial is running. In Petar et al’ s studies on reducing energy consumption for bipedal robots [xx], a mechanism that can evolve the policy parameterization was used. The method starts from a very simple parameterization and gradually increases its representational capability. The method tested to be able to generate an adaptive policy parameterization that can accommodate increasingly more complex policies. In the work described in [xx], the policy generated by this approach can reach the global optimum at a faster rate. Also found is that the chance of converging to a suboptimal solution is reduced, because in the lower-dimensional representation the effect is less exhibited.

Spline policy representation
The simplest model with back-compatibility is geometric splines. For a given model f(x) with K knots, we can preserve the exact shape of the generated curve while adding extra knots to the original spline. Say, if we put one additional knot between every two consecutive knots of the original spline, we end up with a 2K - 1 knots and a spline that has the same shape as the original one. In order to do this, we need to define an algorithm for evolving the parameterization from K to L knots (L > K ), which is formulated in Algorithm 1.  Without loss of generality, the policy parameters are normalized into [0, 1], and appropriately scaled/shifted as necessary later upon use.

[pic]
Fig. 2 illustrates the process of using spline representation for the evolving policy parameterization. Fig. 1 shows an example for a reinforcement learning process using evolving policy parameterization to approximate an unknown function. Petar et al

Integrating evolving policy parameterization with Hill-climbing(RL?)
Kober et al  proposed (Kober et al) a RL algorithm called Policy learning by Weighting Exploration with the Returns(PoWER), which is based on Expectation-Maximization  algorithm (EM). Our proposed technique for evolving the policy parameterization are combined with this EM-based RL algorithm PoWER [xx], due to its relatively fewer parameters that need tuning. Also because PoWER demonstrated high performance in tasks learned directly on real robots, such as pendulum swing-up and pancake flipping task [9].

We evolved the policy parameterization only on those past trials ranked the highest by the importance sampling technique used by the PoWER algorithm(). The intuition behind is that high-ranked parameterizations have more potential to perform even better in the future. Besides, evolving all the parameterizations requires significantly more efforts. (Bongard et al.,2006)

Since our experiment is done on physical robot, explore all the variations of every parameterization is not practical. A promising extension can be, using physical simulator for the robot, to further explore the parameterization space by evolving more parameterizations.

Gaits of RL_POWER:
For the experiment, we have 3 knots for each spline and 8 splines in total for the servos. The servo in the hip is not used in our experiment. Previous work has verified that quadruped gaits perform better when they are coordinated (Clune et al., 2009a, 2011; Valsalam and Miikkulainen, 2008). So the gaits For each set of splines, we calculate its corresponding parameterized gait within a time cycle. Given the pattern of one unit cycle, then apply the same pattern to every cycle throughout the whole period of trial. 
Specifically, each spline is stretched to servo positions as following:
   rescale for one cycle:
 	Unit Cycle :
s(si).cyclePos = s(si).YY .* posRange + posMin;
    		s(si).cyclePos = s(si).cyclePos(:);
    	Repeat cycle,  
    	Cycle[i] = UnitCycle
% the last pos in a cycle is the same as the first pos in the next cycle, from 
[pic for one gait]

### 2012-05-31 paper_aimsa/introduction.tex
Fix line endings, updated ownstyles, fixed underscores

@@
 
-General learning:
-Learning gaits 
-Various learning algorithms have been tested to be effective for legged robots. 
+General learning:Learning gaits Various learning algorithms have been tested to be effective for legged robots.
 Their studies tested the viability of applying RL method to legged robot gaits learning.
@@
 Here we present a different way for learning gaits using RL(Hill climbing).
-In our experiment, it turned out this method achieved competitive performance with
-Problem Definition:
-The gait-learning problem is defined to find a gait that maximizes some specific metric.
+In our experiment, it turned out this method achieved competitive performance with Problem Definition:The gait-learning problem is defined to find a gait that maximizes some specific metric.
 Mathematically, we define a gait as a function that specifies a vector of commanded motor positions for a robot over time.
-Gaits without feedback --- also called open-loop gaits --- can be defined as
-
-\be
-\vec{x} = g(t)
-\ee 
-According to the definition, open-loop gaits are deterministic.
+Gaits without feedback --- also called open-loop gaits --- can be defined as \be\vec{x} = g(t)\ee According to the definition, open-loop gaits are deterministic.
 One particular gait should behave exactly the same when it is run from trial to trial.
-However, the actual robot motion and fitness measured will vary due to the noisiness of the real world physics.
-In our trials, one gait pattern was generated and sent to the robot and executed in an open loop manner, as defined.
-In particular, in this paper, we chose to compare open-loop gaits generated specifically by the RL_POWER and HyperNEAT.
+However, the actual robot motion and fitness measured will vary due to the noisiness of the real world physics.In our trials, one gait pattern was generated and sent to the robot and executed in an open loop manner, as defined.
+In particular, in this paper, we chose to compare open-loop gaits generated specifically by the RL\_POWER and HyperNEAT.
 The particular metric used will be described later.

### 2012-05-31 paper_aimsa/methods.tex
Fix line endings, updated ownstyles, fixed underscores

@@
 how we tried to solve it...
-Parameterized Gaits by RL_PoWER:
 
-Here we used an RL approach to change the complexity of the policy representation dynamically while the trial is running.
+Parameterized Gaits by RL\_PoWER: Here we used an RL approach to change the complexity of the policy representation dynamically while the trial is running.
 In Petar et al’ s studies on reducing energy consumption for bipedal robots [xx], a mechanism that can evolve the policy parameterization was used.
@@
 Also found is that the chance of converging to a suboptimal solution is reduced, because in the lower-dimensional representation the effect is less exhibited.
-
-Spline policy representation
-The simplest model with back-compatibility is geometric splines.
+Spline policy representationThe simplest model with back-compatibility is geometric splines.
 For a given model f(x) with K knots, we can preserve the exact shape of the generated curve while adding extra knots to the original spline.
@@
 In order to do this, we need to define an algorithm for evolving the parameterization from K to L knots (L > K ), which is formulated in Algorithm 1.
-Without loss of generality, the policy parameters are normalized into [0, 1], and appropriately scaled/shifted as necessary later upon use.
-
-[pic]
-Fig.
+Without loss of generality, the policy parameters are normalized into [0, 1], and appropriately scaled/shifted as necessary later upon use.[pic]Fig.
 2 illustrates the process of using spline representation for the evolving policy parameterization.
@@
 1 shows an example for a reinforcement learning process using evolving policy parameterization to approximate an unknown function.
-Petar et al
-
-Integrating evolving policy parameterization with Hill-climbing(RL?)
-Kober et al  proposed (Kober et al) a RL algorithm called Policy learning by Weighting Exploration with the Returns(PoWER), which is based on Expectation-Maximization  algorithm (EM).
+Petar et alIntegrating evolving policy parameterization with Hill-climbing(RL?)Kober et al  proposed (Kober et al) a RL algorithm called Policy learning by Weighting Exploration with the Returns(PoWER), which is based on Expectation-Maximization  algorithm (EM).
 Our proposed technique for evolving the policy parameterization are combined with this EM-based RL algorithm PoWER [xx], due to its relatively fewer parameters that need tuning.
-Also because PoWER demonstrated high performance in tasks learned directly on real robots, such as pendulum swing-up and pancake flipping task [9].
-
-We evolved the policy parameterization only on those past trials ranked the highest by the importance sampling technique used by the PoWER algorithm().
+Also because PoWER demonstrated high performance in tasks learned directly on real robots, such as pendulum swing-up and pancake flipping task [9].We evolved the policy parameterization only on those past trials ranked the highest by the importance sampling technique used by the PoWER algorithm().
 The intuition behind is that high-ranked parameterizations have more potential to perform even better in the future.
 Besides, evolving all the parameterizations requires significantly more efforts.
-(Bongard et al.,2006)
-
-Since our experiment is done on physical robot, explore all the variations of every parameterization is not practical.
+(Bongard et al.,2006) Since our experiment is done on physical robot, explore all the variations of every parameterization is not practical.
 A promising extension can be, using physical simulator for the robot, to further explore the parameterization space by evolving more parameterizations.
-
-Gaits of RL_POWER:
-For the experiment, we have 3 knots for each spline and 8 splines in total for the servos.
+Gaits of RL\_POWER:For the experiment, we have 3 knots for each spline and 8 splines in total for the servos.
 The servo in the hip is not used in our experiment.
@@
 Given the pattern of one unit cycle, then apply the same pattern to every cycle throughout the whole period of trial.
-Specifically, each spline is stretched to servo positions as following:
-   rescale for one cycle:
- 	Unit Cycle :
-s(si).cyclePos = s(si).YY .* posRange + posMin;
-    		s(si).cyclePos = s(si).cyclePos(:);
-    	Repeat cycle,  
-    	Cycle[i] = UnitCycle
-% the last pos in a cycle is the same as the first pos in the next cycle, from 
-[pic for one gait]
+Specifically, each spline is stretched to servo positions as following:   rescale for one cycle: 	Unit Cycle :s(si).cyclePos = s(si).YY .* posRange + posMin;    		s(si).cyclePos = s(si).cyclePos(:);    	Repeat cycle,      	Cycle[i] = UnitCycle% the last pos in a cycle is the same as the first pos in the next cycle, from [pic for one gait]

### 2012-06-01 paper_aimsa/conclusion.tex
finally

\section{Conclusion}

Conclusion and Future Work
We have presented a new reinforcement-learning-based algorithm for optimizing a quadrupedal gait for linear speed. We implemented and tested six learning strategies for parameterized gaits and compared them to gaits produced by neural networks evolved with the HyperNEAT generative encoding. Though both methods resulted in an improvement over the robot’s previous naïve gaits, based on the statistics collected, RL_PoWER has a more elegant and consistent performance.

Based on previous studies on evolving gait on quadruped robots[xx], RL_PoWER produced higher performing gaits than all of the parameterized methods. Its best-performing gait traveled XX body lengths per minute.  HyperNEAT, showed in the previous studies, achieved the competitive best performance with RL_PoWER. HyperNEAT gaits

Though over 900 trials have been made to investigate the applicability of RL_PoWER to quadruped robots. It is difficult to gather the enough trials that would be necessary to properly rank the methods. One direction for future work could be to obtain many more trials. But due to the physical limitations, obtaining  One solution to this is simulation. Because of the low cost of simulation, it would produce the necessary volume of trials to allow the learning methods to be effective, and the hardware trials would serve to continuously ground and refine the simulator. [KYRREG XX]

One guess led by this study is that for feedback oriented tasks, reinforcement learning methods are more fit in natural. Despite the complexities of HyperNEAT, a simpler algorithm on the code level delivered better performance in general.  Also, evolvable spline interpolation is shown to be simple and representationally powerful at the same time. Can serve as a general representation for various other learning problems.

### 2012-06-01 paper_aimsa/results.tex
finally

@@
 
-how well it worked, plots, ...
+RL_POWER Gaits
+The results for the gaits evolved by RL_POWER are shown
+in Figure 8 and Table 2.
+A total of 900 evaluations were performed for HyperNEAT (300 in each of three runs).
+Overall the RL_POWER gaits were the faster and more stable by far, beating the HyperNEAT when comparing either average or best gaits
+
+(pic)RL_POWER AND HyperNEAT performance chart
+
+or best gaits. We believe that this is because RL_POWER uses evolvable splines to represent gaits As explained earlier,  
+The single best gait found during this study had a speed of
+XXXX body lengths/minute, XX%
+better than the best HyperNEAT gait.
+Figure 9 shows a typical RL_POWER gait that had high fitness.
+The pattern of motion is less complex and more regular than the HyperNEAT, in that patterns of multiple motors are coordinated.
+A corresponding observation of this property is that the noisiness of RL_POWER is significantly lower than that of HyperNEAT.
+One reason of this is that the when Expectation Maximization methodology is used, explorations happened in the vicinity space of the highest ranked patterns.
+Even some individual HyperNEAT gaits turned out very effective, the average noisiness of HyperNEAT is very large.
+For example, as found in [xx], the mean of HyperNEAT generation-champion’s performance was XX body lengths/minute (+-XX SD), but it had a max of 38 and a min of 3.
+RL_POWER, on the other hand,  proved to be stable and much less noisy throughout the entire experiment.
+The maximum SD over all three runs is +- XX SD.
+This may also be due to a much larger space of possibilities of HyperNEAT than the more heuristically guided reinforcement learning.
+
+[pic][speed chart of two algos]

### 2012-06-01 paper_aimsa/experimentalSetup.tex

@@
 hardware details, ...
-Setup:
-The quadruped robot used was assembled from parts purchased online and parts printed by the Objet Connex 500 3-D Printing System.
+
+Setup: The quadruped robot used was assembled from parts purchased online and parts printed by the Objet Connex 500 3-D Printing System.
 The robot actuation system consists of 5 AX-18+ Dynamixel servos and 4 AX-12+ Dynamixel servos: one inner joint with one AX-18+ servo and one outer joint with AX-12+servo in each of the four legs, and one AX-18+ servo at the center.
@@
 In the studies of this paper, tethered cables powered both the computer and the servos.
-It measures approximately 39.5 centimeters from leg to opposite leg in the crouch position showed in [pic].
-
-Our performance metric was the displacement over the evaluation period of 12 seconds for each.
+It measures approximately 39.5 centimeters from leg to opposite leg in the crouch position showed in [pic].Our performance metric was the displacement over the evaluation period of 12 seconds for each.
 Same as Jason et al described in [xx], the displacement was measured using a Wii remote that was placed on the ceiling.
@@
 Each LED was tilted outwards in order to maximize the visible range.
-[pic]
-
-A separate tracking server ran on the robot PC interacted with the Wii remote via bluetooth by using the CWiid library.
+[pic] A separate tracking server ran on the robot PC interacted with the Wii remote via bluetooth by using the CWiid library.
 A corresponding client communicates with this server via a socket connection.
 Our fitness evaluation code talks with the Wii remote by using this client and updates the position from beginning to end during each run.
-
 The metric for evaluating gaits was the Euclidian distance the robot traveled during a 12-second run on flat surface.
@@
 These gaits tend to either overburden the servos or flip the robot.
-Thus, RL_POWER(?) were tested after carefully cropping out the extremely fierce gaits each time.
+Thus, RL\_POWER(?) were tested after carefully cropping out the extremely fierce gaits each time.
 (As described earlier, each leg has two joints, inner joint j1 and outer j2.
 The position of one leg is determined by the sum of the values of j1 and j2.
-
 Extremely fierce gaits usually have relatively small values of g.

### 2012-06-01 paper_aimsa/conclusion.tex
Added splines example figure, wrapped lines of text for easier merging

@@
-\section{Conclusion}
+\section{Conclusion and Future Work}
 
-Conclusion and Future Work
-We have presented a new reinforcement-learning-based algorithm for optimizing a quadrupedal gait for linear speed.
-We implemented and tested six learning strategies for parameterized gaits and compared them to gaits produced by neural networks evolved with the HyperNEAT generative encoding.
-Though both methods resulted in an improvement over the robot’s previous naïve gaits, based on the statistics collected, RL_PoWER has a more elegant and consistent performance.
+We have presented a new reinforcement-learning-based algorithm for
+optimizing a quadrupedal gait for linear speed. We implemented and
+tested six learning strategies for parameterized gaits and compared
+them to gaits produced by neural networks evolved with the HyperNEAT
+generative encoding. Though both methods resulted in an improvement
+over the robot’s previous \naive gaits, based on the statistics
+collected, RL\_PoWER has a more elegant and consistent performance.
 
-Based on previous studies on evolving gait on quadruped robots[xx], RL_PoWER produced higher performing gaits than all of the parameterized methods.
-Its best-performing gait traveled XX body lengths per minute.
-HyperNEAT, showed in the previous studies, achieved the competitive best performance with RL_PoWER.
-HyperNEAT gaits
+Based on previous studies on evolving gait on quadruped robots[xx],
+RL\_PoWER produced higher performing gaits than all of the
+parameterized methods. Its best-performing gait traveled XX body
+lengths per minute.  HyperNEAT, showed in the previous studies,
+achieved the competitive best performance with RL\_PoWER. HyperNEAT
+gaits
 
-Though over 900 trials have been made to investigate the applicability of RL_PoWER to quadruped robots.
-It is difficult to gather the enough trials that would be necessary to properly rank the methods.
-One direction for future work could be to obtain many more trials.
-But due to the physical limitations, obtaining  One solution to this is simulation.
-Because of the low cost of simulation, it would produce the necessary volume of trials to allow the learning methods to be effective, and the hardware trials would serve to continuously ground and refine the simulator.
-[KYRREG XX]
+Though over 900 trials have been made to investigate the applicability
+of RL\_PoWER to quadruped robots. It is difficult to gather the enough
+trials that would be necessary to properly rank the methods. One
+direction for future work could be to obtain many more trials. But due
+to the physical limitations, obtaining One solution to this is
+simulation. Because of the low cost of simulation, it would produce
+the necessary volume of trials to allow the learning methods to be
+effective, and the hardware trials would serve to continuously ground
+and refine the simulator. [KYRREG XX]
 
-One guess led by this study is that for feedback oriented tasks, reinforcement learning methods are more fit in natural.
-Despite the complexities of HyperNEAT, a simpler algorithm on the code level delivered better performance in general.
-Also, evolvable spline interpolation is shown to be simple and representationally powerful at the same time.
-Can serve as a general representation for various other learning problems.
+One guess led by this study is that for feedback oriented tasks,
+reinforcement learning methods are more fit in natural. Despite the
+complexities of HyperNEAT, a simpler algorithm on the code level
+delivered better performance in general.  Also, evolvable spline
+interpolation is shown to be simple and representationally powerful at
+the same time. Can serve as a general representation for various other
+learning problems.

### 2012-06-01 paper_aimsa/introduction.tex
Added splines example figure, wrapped lines of text for easier merging

@@
 \section{Introduction}
-
-What the problem is...
 
@@
 
-General learning:Learning gaits Various learning algorithms have been tested to be effective for legged robots.
-Their studies tested the viability of applying RL method to legged robot gaits learning.
-Competitive performance has been verified on methods such as HyperNEAT (Yosinski et al, 2011) and GA(Chernova et al, 2004) and other  methods(Hornby et al., 2005; Zykov et al., 2004; T´ellez et al., 2006; Valsalam and Miikkulainen, 2008).
-Despite the excellent performance of these algorithms, a major part of achieving good performance relies in tuning the parameters for the evolutionary algorithms(Kormushev et al, 2006).
-Here we present a different way for learning gaits using RL(Hill climbing).
-In our experiment, it turned out this method achieved competitive performance with Problem Definition:The gait-learning problem is defined to find a gait that maximizes some specific metric.
-Mathematically, we define a gait as a function that specifies a vector of commanded motor positions for a robot over time.
-Gaits without feedback --- also called open-loop gaits --- can be defined as \be\vec{x} = g(t)\ee According to the definition, open-loop gaits are deterministic.
-One particular gait should behave exactly the same when it is run from trial to trial.
-However, the actual robot motion and fitness measured will vary due to the noisiness of the real world physics.In our trials, one gait pattern was generated and sent to the robot and executed in an open loop manner, as defined.
-In particular, in this paper, we chose to compare open-loop gaits generated specifically by the RL\_POWER and HyperNEAT.
-The particular metric used will be described later.
+Various learning algorithms have been tested to be effective for
+legged robots. Their studies tested the viability of applying RL
+method to legged robot gaits learning. Competitive performance has
+been verified on methods such as HyperNEAT (Yosinski et al, 2011) and
+GA(Chernova et al, 2004) and other methods(Hornby et al., 2005; Zykov
+et al., 2004; T´ellez et al., 2006; Valsalam and Miikkulainen,
+2008). Despite the excellent performance of these algorithms, a major
+part of achieving good performance relies in tuning the parameters for
+the evolutionary algorithms(Kormushev et al, 2006). Here we present a
+different way for learning gaits using RL(Hill climbing). In our
+experiment, it turned out this method achieved competitive performance
+with Problem Definition:The gait-learning problem is defined to find a
+gait that maximizes some specific metric. Mathematically, we define a
+gait as a function that specifies a vector of commanded motor
+positions for a robot over time. Gaits without feedback --- also
+called open-loop gaits --- can be defined as
+
+\be\vec{x} = g(t)\ee
+
+According to the definition, open-loop gaits are deterministic. One
+particular gait should behave exactly the same when it is run from
+trial to trial. However, the actual robot motion and fitness measured
+will vary due to the noisiness of the real world physics.In our
+trials, one gait pattern was generated and sent to the robot and
+executed in an open loop manner, as defined. In particular, in this
+paper, we chose to compare open-loop gaits generated specifically by
+the RL\_POWER and HyperNEAT. The particular metric used will be
+described later.

### 2012-06-01 paper_aimsa/methods.tex
Added splines example figure, wrapped lines of text for easier merging

@@
 
-how we tried to solve it...
+\subsection{Parameterized Gaits by RL\_PoWER}
 
-Parameterized Gaits by RL\_PoWER: Here we used an RL approach to change the complexity of the policy representation dynamically while the trial is running.
-In Petar et al’ s studies on reducing energy consumption for bipedal robots [xx], a mechanism that can evolve the policy parameterization was used.
-The method starts from a very simple parameterization and gradually increases its representational capability.
-The method tested to be able to generate an adaptive policy parameterization that can accommodate increasingly more complex policies.
-In the work described in [xx], the policy generated by this approach can reach the global optimum at a faster rate.
-Also found is that the chance of converging to a suboptimal solution is reduced, because in the lower-dimensional representation the effect is less exhibited.
-Spline policy representationThe simplest model with back-compatibility is geometric splines.
-For a given model f(x) with K knots, we can preserve the exact shape of the generated curve while adding extra knots to the original spline.
-Say, if we put one additional knot between every two consecutive knots of the original spline, we end up with a 2K - 1 knots and a spline that has the same shape as the original one.
-In order to do this, we need to define an algorithm for evolving the parameterization from K to L knots (L > K ), which is formulated in Algorithm 1.
-Without loss of generality, the policy parameters are normalized into [0, 1], and appropriately scaled/shifted as necessary later upon use.[pic]Fig.
-2 illustrates the process of using spline representation for the evolving policy parameterization.
-Fig.
-1 shows an example for a reinforcement learning process using evolving policy parameterization to approximate an unknown function.
-Petar et alIntegrating evolving policy parameterization with Hill-climbing(RL?)Kober et al  proposed (Kober et al) a RL algorithm called Policy learning by Weighting Exploration with the Returns(PoWER), which is based on Expectation-Maximization  algorithm (EM).
-Our proposed technique for evolving the policy parameterization are combined with this EM-based RL algorithm PoWER [xx], due to its relatively fewer parameters that need tuning.
-Also because PoWER demonstrated high performance in tasks learned directly on real robots, such as pendulum swing-up and pancake flipping task [9].We evolved the policy parameterization only on those past trials ranked the highest by the importance sampling technique used by the PoWER algorithm().
-The intuition behind is that high-ranked parameterizations have more potential to perform even better in the future.
-Besides, evolving all the parameterizations requires significantly more efforts.
-(Bongard et al.,2006) Since our experiment is done on physical robot, explore all the variations of every parameterization is not practical.
-A promising extension can be, using physical simulator for the robot, to further explore the parameterization space by evolving more parameterizations.
-Gaits of RL\_POWER:For the experiment, we have 3 knots for each spline and 8 splines in total for the servos.
-The servo in the hip is not used in our experiment.
-Previous work has verified that quadruped gaits perform better when they are coordinated (Clune et al., 2009a, 2011; Valsalam and Miikkulainen, 2008).
-So the gaits For each set of splines, we calculate its corresponding parameterized gait within a time cycle.
-Given the pattern of one unit cycle, then apply the same pattern to every cycle throughout the whole period of trial.
-Specifically, each spline is stretched to servo positions as following:   rescale for one cycle: 	Unit Cycle :s(si).cyclePos = s(si).YY .* posRange + posMin;    		s(si).cyclePos = s(si).cyclePos(:);    	Repeat cycle,      	Cycle[i] = UnitCycle% the last pos in a cycle is the same as the first pos in the next cycle, from [pic for one gait]
+Here we used an RL approach to change the complexity of the policy
+representation dynamically while the trial is running. In Petar et al’
+s studies on reducing energy consumption for bipedal robots [xx], a
+mechanism that can evolve the policy parameterization was used. The
+method starts from a very simple parameterization and gradually
+increases its representational capability. The method tested to be
+able to generate an adaptive policy parameterization that can
+accommodate increasingly more complex policies. In the work described
+in [xx], the policy generated by this approach can reach the global
+optimum at a faster rate. Also found is that the chance of converging
+to a suboptimal solution is reduced, because in the lower-dimensional
+representation the effect is less exhibited.
+
+\subsection{Spline policy representation}
+
+The simplest model with back-compatibility is geometric
+splines. For a given model f(x) with K knots, we can preserve the
+exact shape of the generated curve while adding extra knots to the
+original spline. Say, if we put one additional knot between every two
+consecutive knots of the original spline, we end up with a 2K - 1
+knots and a spline that has the same shape as the original one. In
+order to do this, we need to define an algorithm for evolving the
+parameterization from $K$ to $L$ knots ($L > K$), which is formulated in
+Algorithm \edit{proper reference}.  Without loss of generality, the policy parameters are
+normalized into $[0, 1]$, and appropriately scaled/shifted as necessary
+later upon use.
+
+\figp{powerSplinesExample}{.75}{Example figure... from Petar's paper...}
+
+[pic]Fig. 2 illustrates the process of using spline
+representation for the evolving policy parameterization. Fig. 1 shows
+an example for a reinforcement learning process using evolving policy
+parameterization to approximate an unknown function. Petar et
+alIntegrating evolving policy parameterization with
+Hill-climbing(RL?)Kober et al proposed (Kober et al) a RL algorithm
+called Policy learning by Weighting Exploration with the
+Returns(PoWER), which is based on Expectation-Maximization algorithm
+(EM). Our proposed technique for evolving the policy parameterization
+are combined with this EM-based RL algorithm PoWER [xx], due to its
+relatively fewer parameters that need tuning. Also because PoWER
+demonstrated high performance in tasks learned directly on real
+robots, such as pendulum swing-up and pancake flipping task [9].We
+evolved the policy parameterization only on those past trials ranked
+the highest by the importance sampling technique used by the PoWER
+algorithm(). The intuition behind is that high-ranked
+parameterizations have more potential to perform even better in the
+future. Besides, evolving all the parameterizations requires
+significantly more efforts. (Bongard et al.,2006) Since our experiment
+is done on physical robot, explore all the variations of every
+parameterization is not practical. A promising extension can be, using
+physical simulator for the robot, to further explore the
+parameterization space by evolving more parameterizations. Gaits of
+RL\_POWER:For the experiment, we have 3 knots for each spline and 8
+splines in total for the servos. The servo in the hip is not used in
+our experiment. Previous work has verified that quadruped gaits
+perform better when they are coordinated (Clune et al., 2009a, 2011;
+Valsalam and Miikkulainen, 2008). So the gaits For each set of
+splines, we calculate its corresponding parameterized gait within a
+time cycle. Given the pattern of one unit cycle, then apply the same
+pattern to every cycle throughout the whole period of
+trial. Specifically, each spline is stretched to servo positions as
+following: rescale for one cycle: Unit Cycle :s(si).cyclePos =
+s(si).YY .* posRange + posMin; s(si).cyclePos = s(si).cyclePos(:);
+Repeat cycle, Cycle[i] = UnitCycle \% the last pos in a cycle is the
+same as the first pos in the next cycle, from [pic for one gait]

### 2012-06-01 paper_aimsa/results.tex
Added splines example figure, wrapped lines of text for easier merging

@@
 
-RL_POWER Gaits
-The results for the gaits evolved by RL_POWER are shown
-in Figure 8 and Table 2.
-A total of 900 evaluations were performed for HyperNEAT (300 in each of three runs).
-Overall the RL_POWER gaits were the faster and more stable by far, beating the HyperNEAT when comparing either average or best gaits
+RL\_POWER Gaits
 
-(pic)RL_POWER AND HyperNEAT performance chart
+The results for the gaits evolved by RL\_POWER are shown in Figure 8
+and Table 2. A total of 900 evaluations were performed for HyperNEAT
+(300 in each of three runs). Overall the RL\_POWER gaits were the
+faster and more stable by far, beating the HyperNEAT when comparing
+either average or best gaits
 
-or best gaits. We believe that this is because RL_POWER uses evolvable splines to represent gaits As explained earlier,  
+(pic)RL\_POWER AND HyperNEAT performance chart
+
+or best gaits. We believe that this is because RL\_POWER uses evolvable splines to represent gaits As explained earlier,  
 The single best gait found during this study had a speed of
-XXXX body lengths/minute, XX%
-better than the best HyperNEAT gait.
-Figure 9 shows a typical RL_POWER gait that had high fitness.
-The pattern of motion is less complex and more regular than the HyperNEAT, in that patterns of multiple motors are coordinated.
-A corresponding observation of this property is that the noisiness of RL_POWER is significantly lower than that of HyperNEAT.
-One reason of this is that the when Expectation Maximization methodology is used, explorations happened in the vicinity space of the highest ranked patterns.
-Even some individual HyperNEAT gaits turned out very effective, the average noisiness of HyperNEAT is very large.
-For example, as found in [xx], the mean of HyperNEAT generation-champion’s performance was XX body lengths/minute (+-XX SD), but it had a max of 38 and a min of 3.
-RL_POWER, on the other hand,  proved to be stable and much less noisy throughout the entire experiment.
-The maximum SD over all three runs is +- XX SD.
-This may also be due to a much larger space of possibilities of HyperNEAT than the more heuristically guided reinforcement learning.
+
+XXXX body lengths/minute, XX\%
+
+better than the best HyperNEAT gait. Figure 9 shows a typical
+RL\_POWER gait that had high fitness. The pattern of motion is less
+complex and more regular than the HyperNEAT, in that patterns of
+multiple motors are coordinated.
+
+A corresponding observation of this property is that the noisiness of
+RL\_POWER is significantly lower than that of HyperNEAT. One reason of
+this is that the when Expectation Maximization methodology is used,
+explorations happened in the vicinity space of the highest ranked
+patterns. Even some individual HyperNEAT gaits turned out very
+effective, the average noisiness of HyperNEAT is very large. For
+example, as found in [xx], the mean of HyperNEAT generation-champion’s
+performance was XX body lengths/minute (+-XX SD), but it had a max of
+38 and a min of 3. RL\_POWER, on the other hand, proved to be stable
+and much less noisy throughout the entire experiment. The maximum SD
+over all three runs is +- XX SD. This may also be due to a much larger
+space of possibilities of HyperNEAT than the more heuristically guided
+reinforcement learning.
 

### 2012-06-01 paper_aimsa/experimentalSetup.tex
word wrap, added fig
Added links to many references

@@
 
-hardware details, ...
+The quadruped robot used was assembled from parts purchased online and
+parts printed by the Objet Connex 500 3-D Printing System. The robot
+actuation system consists of 5 AX-18+ Dynamixel servos and 4 AX-12+
+Dynamixel servos: one inner joint with one AX-18+ servo and one outer
+joint with AX-12+servo in each of the four legs, and one AX-18+ servo
+at the center. To avoid the formerly reported AX-18+ servos are used
+in this robot because of their stronger actuation power than that of
+AX-12+. Each servo could be set to a position in the range [0, 1023]
+by using PyDynamixel library, corresponding to a physical range [-120,
+  +120]. Also, to prevent collisions with the robot body, the control
+module filter out the commands to a safe range. This range was [-85,
+  +60 ] for the inner leg servos, [-113 , +39 ] for the outer leg
+servos, and [-28 , +28] for the center hip servo. In the studies of
+this paper, tethered cables powered both the computer and the
+servos. It measures approximately 39.5 centimeters from leg to
+opposite leg in the crouch position showed in [pic].Our performance
+metric was the displacement over the evaluation period of 12 seconds
+for each. Same as \cite{yosinski2011evolving-robot-gaits}, the displacement was
+measured using a Wii remote that was placed on the ceiling. Different
+from the original model described in \cite{yosinski2011evolving-robot-gaits}, the quadruped robot was
+equipped with a three-infrared-LED cluster on top rather than just
+one. The reason for this setup is that when fierce gaits were
+executed, the Wii remote loses tracking of the robot position due to
+the limited visible angle of a single LED.  These three LEDs were
+place tightly together to act as one signal emitter. Each LED was
+tilted outwards in order to maximize the visible range. [pic] A
+separate tracking server ran on the robot PC interacted with the Wii
+remote via bluetooth by using the CWiid library.  A corresponding
+client communicates with this server via a socket connection. Our
+fitness evaluation code talks with the Wii remote by using this client
+and updates the position from beginning to end during each run. The
+metric for evaluating gaits was the Euclidian distance the robot
+traveled during a 12-second run on flat surface. Previous work done in
+\cite{yosinski2011evolving-robot-gaits} and \cite{clune2009evolving-coordinated-quadruped} suggested that extremely fierce gaits are not viable in
+general. These gaits tend to either overburden the servos or flip the
+robot. Thus, RL\_POWER(?) were tested after carefully cropping out the
+extremely fierce gaits each time. (As described earlier, each leg has
+two joints, inner joint j1 and outer j2. The position of one leg is
+determined by the sum of the values of j1 and j2. Extremely fierce
+gaits usually have relatively small values of g. Thus cropping works
+by mapping the wild gaits onto the boundary of a triangular space over
+the two dimensional space of j1 and j2. \figref{smartCrop}....
 
-Setup: The quadruped robot used was assembled from parts purchased online and parts printed by the Objet Connex 500 3-D Printing System.
-The robot actuation system consists of 5 AX-18+ Dynamixel servos and 4 AX-12+ Dynamixel servos: one inner joint with one AX-18+ servo and one outer joint with AX-12+servo in each of the four legs, and one AX-18+ servo at the center.
-To avoid the formerly reported AX-18+ servos are used in this robot because of their stronger actuation power than that of AX-12+.
-Each servo could be set to a position in the range [0, 1023] by using PyDynamixel library, corresponding to a physical range [-120, +120].
-Also, to prevent collisions with the robot body, the control module filter out the commands to a safe range.
-This range was [-85, +60 ] for the inner leg servos, [-113 , +39 ] for the outer leg servos, and [-28 , +28] for the center hip servo.
-In the studies of this paper, tethered cables powered both the computer and the servos.
-It measures approximately 39.5 centimeters from leg to opposite leg in the crouch position showed in [pic].Our performance metric was the displacement over the evaluation period of 12 seconds for each.
-Same as Jason et al described in [xx], the displacement was measured using a Wii remote that was placed on the ceiling.
-Different from the original model described in [xx], the quadruped robot was equipped with a three-infrared-LED cluster on top rather than just one.
-The reason for this setup is that when fierce gaits were executed, the Wii remote loses tracking of the robot position due to the limited visible angle of a single LED.
-These three LEDs were place tightly together to act as one signal emitter.
-Each LED was tilted outwards in order to maximize the visible range.
-[pic] A separate tracking server ran on the robot PC interacted with the Wii remote via bluetooth by using the CWiid library.
-A corresponding client communicates with this server via a socket connection.
-Our fitness evaluation code talks with the Wii remote by using this client and updates the position from beginning to end during each run.
-The metric for evaluating gaits was the Euclidian distance the robot traveled during a 12-second run on flat surface.
-Previous work done in [xx] and [xx1(k)] suggested that extremely fierce are not viable in general.
-These gaits tend to either overburden the servos or flip the robot.
-Thus, RL\_POWER(?) were tested after carefully cropping out the extremely fierce gaits each time.
-(As described earlier, each leg has two joints, inner joint j1 and outer j2.
-The position of one leg is determined by the sum of the values of j1 and j2.
-Extremely fierce gaits usually have relatively small values of g.
-Thus cropping works by mapping the wild gaits onto the boundary of a triangular space over the two dimensional space of j1 and j2.
-[pic of j1 j2 space] [g = j1 + j2] caption
+\figp{smartCrop}{1}{$g = j1 +  j2$ caption}
+
+\figp{sixLegPositions}{1}{caption... the left two are allowed, middle
+  two are on the red border line from \figref{smartCrop}, right two
+  are disallowed.}

### 2012-06-01 paper_aimsa/introduction.tex
Added picture of quadratot with booties
Added links to many references

@@
 
-\figp{aracna_black_2}{.75}{Example figure...}
+\figp{quadratotWhiteBooties}{.75}{Example figure...}
 
@@
 legged robots. Their studies tested the viability of applying RL
-method to legged robot gaits learning. Competitive performance has
-been verified on methods such as HyperNEAT (Yosinski et al, 2011) and
-GA(Chernova et al, 2004) and other methods(Hornby et al., 2005; Zykov
-et al., 2004; T´ellez et al., 2006; Valsalam and Miikkulainen,
-2008). Despite the excellent performance of these algorithms, a major
+cite{yosinski2011evolvingmethod to legged robot gaits learning. Competitive performance has
+been verified on methods such as HyperNEAT \cite{yosinski2011evolving-robot-gaits} and
+GA \cite{chernova2004an-evolutionary-approach-to-gait} and other methods \cite{hornby2005autonomous-evolution-of-dynamic} \cite{zykov2004evolving-dynamic-gaits} \cite{tellez2006evolving-the-walking-behaviour} \cite{valsalam2008modular-neuroevolution-for-multilegged}.
+Despite the excellent performance of these algorithms, a major
 part of achieving good performance relies in tuning the parameters for
-the evolutionary algorithms(Kormushev et al, 2006). Here we present a
-different way for learning gaits using RL(Hill climbing). In our
+the evolutionary algorithms \cite{kormushev2011bipedal-walking-energya}. Here we present a
+different way for learning gaits using RL (Hill climbing). In our
 experiment, it turned out this method achieved competitive performance

### 2012-06-01 paper_aimsa/results.tex
Added links to many references

@@
 effective, the average noisiness of HyperNEAT is very large. For
-example, as found in [xx], the mean of HyperNEAT generation-champion’s
+example, as found in \cite{yosinski2011evolving-robot-gaits}, the mean of HyperNEAT generation-champion’s
 performance was XX body lengths/minute (+-XX SD), but it had a max of

### 2012-06-01 paper_aimsa/extractGait.py
minor

@@ -7,7 +7,7 @@ import matplotlib.pyplot as plt
 import re
 
 def main():
-    path = '/Users/hs454/Desktop/tmp/input.txt'
+    path = 'input.txt'
     f= open(path)
     lines = f.readlines()
     t = []

### 2012-06-01 paper_aimsa/results.tex
results.tex

--- 2012-06-01 paper_aimsa/results.tex
@@
 
-The results for the gaits evolved by RL\_POWER are shown in Figure 8
+The results for the gaits evolved by RL PoWER are shown in Figure 8
 and Table 2. A total of 900 evaluations were performed for HyperNEAT
-(300 in each of three runs). Overall the RL\_POWER gaits were the
+(300 in each of three runs). Overall the RL PoWER gaits were the
 faster and more stable by far, beating the HyperNEAT when comparing
-either average or best gaits
+either average or best gaits.
 
-(pic)RL\_POWER AND HyperNEAT performance chart
+(pic)RL POWER AND HyperNEAT performance chart
 
-or best gaits. We believe that this is because RL\_POWER uses evolvable splines to represent gaits As explained earlier,  
+We believe that this is because RL PoWER uses evolvable splines to represent gaits As explained earlier,  
 The single best gait found during this study had a speed of
 
-XXXX body lengths/minute, XX\%
+11.0855923766 centimeters/sec, XX\%
 

### 2012-06-01 paper_aimsa/extractGait.py
Better colors, better axis, thicker lines

@@ -1,8 +1,9 @@
-#! /usr/bin/env python-32
+#! /usr/bin/env python
 
+import pdb
 from os import listdir
 import matplotlib as mpl 
-mpl.use('cocoaagg') 
+#mpl.use('cocoaagg') 
 import matplotlib.pyplot as plt
 import re
 
@@ -21,7 +22,7 @@ def main():
     j8 = []
     for line in lines:
         joints = line.split()
-        t.append(float(joints[0])-1)
+        t.append(float(joints[0])-2)
         j1.append(float(joints[1]))
         j2.append(float(joints[2]))
         j3.append(float(joints[3]))
@@ -32,10 +33,15 @@ def main():
         j8.append(float(joints[8]))
     
     
-    plt.plot(t,j1,'r', t,j2,'b', t,j3,'g', t,j4,'k', t,j5,'c', t,j6,'m', t,j7,'y', t,j8, 'g^')
+    #plt.plot(t,j1,'r', t,j2,'b', t,j3,'g', t,j4,'k', t,j5,'c', t,j6,'m', t,j7,'y', t,j8, 'g^')
+    plt.plot(t,j1,t,j2, t,j3, t,j4, t,j5, t,j6, t,j7, t,j8, linewidth=2)
     plt.ylabel('Motor Position')
     plt.xlabel('Time(s)')
+    ax = plt.axis()
+    plt.axis((min(t), max(t), ax[2], ax[3]))
+    plt.savefig('gait.pdf')
     plt.show()
+    
     
 if __name__ == '__main__':
 	main();

### 2012-06-01 paper_aimsa/methods.tex

@@
 Here we used an RL approach to change the complexity of the policy
-representation dynamically while the trial is running. In Petar et al’
-s studies on reducing energy consumption for bipedal robots [xx], a
-mechanism that can evolve the policy parameterization was used. The
-method starts from a very simple parameterization and gradually
-increases its representational capability. The method tested to be
-able to generate an adaptive policy parameterization that can
-accommodate increasingly more complex policies. In the work described
-in [xx], the policy generated by this approach can reach the global
-optimum at a faster rate. Also found is that the chance of converging
-to a suboptimal solution is reduced, because in the lower-dimensional
-representation the effect is less exhibited.
+representation dynamically while the trial is running. In
+\cite{kormushev2011bipedal-walking-energya} studies on reducing energy
+consumption for bipedal robots, a mechanism that can
+evolve the policy parameterization was used. The method starts from a
+very simple parameterization and gradually increases its
+representational capability. The method tested to be able to generate
+an adaptive policy parameterization that can accommodate increasingly
+more complex policies. In \cite{kormushev2011bipedal-walking-energya}, the policy
+generated by this approach can reach the global optimum at a faster
+rate. Also found is that the chance of converging to a suboptimal
+solution is reduced, because in the lower-dimensional representation
+the effect is less exhibited.
 
@@
 alIntegrating evolving policy parameterization with
-Hill-climbing(RL?)Kober et al proposed (Kober et al) a RL algorithm
+Hill-climbing(RL?) \cite{kober2009learning-motor-primitives} proposed a RL algorithm
 called Policy learning by Weighting Exploration with the
@@
 (EM). Our proposed technique for evolving the policy parameterization
-are combined with this EM-based RL algorithm PoWER [xx], due to its
+are combined with this EM-based RL algorithm PoWER \cite{kober2009learning-motor-primitives}, due to its
 relatively fewer parameters that need tuning. Also because PoWER
 demonstrated high performance in tasks learned directly on real
-robots, such as pendulum swing-up and pancake flipping task [9].We
+robots, such as pendulum swing-up and pancake flipping task \cite{kormushev2010robot-motor-skill}. We
 evolved the policy parameterization only on those past trials ranked
@@
 future. Besides, evolving all the parameterizations requires
-significantly more efforts. (Bongard et al.,2006) Since our experiment
+significantly more efforts, as shown in \cite{bongard2006resilient-machines-through}. Since our experiment
 is done on physical robot, explore all the variations of every
@@
 our experiment. Previous work has verified that quadruped gaits
-perform better when they are coordinated (Clune et al., 2009a, 2011;
-Valsalam and Miikkulainen, 2008). So the gaits For each set of
+perform better when they are coordinated \cite{clune2009evolving-coordinated-quadruped} \cite{clune2011on-the-performance-of-indirect-encoding}
+\cite{valsalam2008modular-neuroevolution-for-multilegged}. For each set of
 splines, we calculate its corresponding parameterized gait within a
@@
 trial. Specifically, each spline is stretched to servo positions as
-following: rescale for one cycle: Unit Cycle :s(si).cyclePos =
-s(si).YY .* posRange + posMin; s(si).cyclePos = s(si).cyclePos(:);
-Repeat cycle, Cycle[i] = UnitCycle \% the last pos in a cycle is the
+following:
+
+\begin{graycode}
+rescale for one cycle:
+Unit Cycle :
+s(si).cyclePos = s(si).YY .* posRange + posMin;
+s(si).cyclePos = s(si).cyclePos(:);
+Repeat cycle,
+Cycle[i] = UnitCycle \% the last pos in a cycle is the
 same as the first pos in the next cycle, from [pic for one gait]
+\end{graycode}

### 2012-06-01 paper_aimsa/introduction.tex
minor ref fix

@@
 part of achieving good performance relies in tuning the parameters for
-the evolutionary algorithms \cite{kormushev2011bipedal-walking-energya}. Here we present a
+the evolutionary algorithms \cite{kormushev2011bipedal-walking-energy}. Here we present a
 different way for learning gaits using RL (Hill climbing). In our

### 2012-06-01 paper_aimsa/methods.tex
minor ref fix

@@
 representation dynamically while the trial is running. In
-\cite{kormushev2011bipedal-walking-energya} studies on reducing energy
+\cite{kormushev2011bipedal-walking-energy} studies on reducing energy
 consumption for bipedal robots, a mechanism that can
@@
 an adaptive policy parameterization that can accommodate increasingly
-more complex policies. In \cite{kormushev2011bipedal-walking-energya}, the policy
+more complex policies. In \cite{kormushev2011bipedal-walking-energy}, the policy
 generated by this approach can reach the global optimum at a faster

### 2012-06-01 code/plot/plot_hnperf.m
Added script to regenerate HN performance plot

@@ -1,0 +1,83 @@
+function plot_hnperf()
+  % Plots the original fitness fig from \cite{yosinski2011evolving-robot-gaits}
+
+  hn1 = [1 135.833 7 33.6269 -1;
+         2 97.3872 7 55.1697 -1;
+         3 119.309 7 39.8728 -1;
+         4 71.4006 7 33.8382 -1;
+         5 90.0056 7 41.2928 -1;
+         6 29.1114 8 14.4977 -1;
+         7 60.1714 7 23.9829 -1;
+         8 80.2465 7 39.0903 -1;
+         9 209.581 8 76.6233 -1;
+         10 172.856 8 69.086 -1;
+         11 127.67 8 44.8484 -1;
+         12 279.805 8 79.1913 -1;
+         13 165.659 8 57.8455 -1;
+         14 166.17 8 42.8344 -1;
+         15 109.005 8 45.3226 -1;
+         16 237.578 8 61.1262 -1;
+         17 289.669 8 87.3376 -1;
+         18 242.033 8 123.579 -1;
+         19 164.248 8 77.4146 -1;
+         20 117.726 9 61.0878 -1];
+
+  hn2 = [1 374.761 7 93.2818 -1;
+         2 88.5129 7 34.4679 -1;
+         3 208.251 8 69.4669 -1;
+         4 92.0055 8 37.9347 -1;
+         5 147.455 8 56.3117 -1;
+         6 99.1899 8 48.662 -1;
+         8 434.803 9 130.461 -1;
+         9 245.99 9 132.422 -1;
+         10 314.383 7 128.53 -1;
+         11 372.162 9 148.273 -1;
+         12 350.288 9 79.1271 -1;
+         13 272.221 9 87.958 -1;
+         14 250.618 9 114.752 -1;
+         15 236.213 9 100.141 -1;
+         16 209.331 8 75.3523 -1;
+         17 95.2019 8 44.2824 -1;
+         18 255.294 7 67.4262 -1;
+         19 317.172 7 108.348 -1;
+         20 254.518 7 103.594 -1;
+         21 343.794 7 155.953 -1];
+
+  hn3 = [1 202.137 7 97.1735 -1;
+         2 211.231 7 98.7642 -1;
+         3 277.827 7 100.933 -1;
+         4 98.6281 7 50.5121 -1;
+         5 145.86 7 75.5518 -1;
[…]

### 2012-06-01 code/plot/plot_hnperf.m
minor

@@ -76,6 +76,9 @@ function plot_hnperf()
   plot(gg, gcMean-gcStd/sqrt(3), 'k', 'linewidth', 1);
   plot(gg, gcMean+gcStd/sqrt(3), 'k', 'linewidth', 1);
 
+  xlabel('generations');
+  ylabel('body lengths per minute');
+
   % Save fig
   print('-dpdf', 'hyperneatPerformance.pdf');
 

### 2012-06-01 paper_aimsa/methods.tex
updated methods.tex

@@
 
-\subsection{Parameterized Gaits by RL\_PoWER}
-
-Here we used an RL approach to change the complexity of the policy
-representation dynamically while the trial is running. In
-\cite{kormushev2011bipedal-walking-energy} studies on reducing energy
-consumption for bipedal robots, a mechanism that can
-evolve the policy parameterization was used. The method starts from a
-very simple parameterization and gradually increases its
-representational capability. The method tested to be able to generate
-an adaptive policy parameterization that can accommodate increasingly
-more complex policies. In \cite{kormushev2011bipedal-walking-energy}, the policy
-generated by this approach can reach the global optimum at a faster
-rate. Also found is that the chance of converging to a suboptimal
-solution is reduced, because in the lower-dimensional representation
-the effect is less exhibited.
-
-\subsection{Spline policy representation}
+\subsection{Policy Representation by Splines}
 
@@
 
-\figp{powerSplinesExample}{.75}{Example figure... from Petar's paper...}
+\subsection{Parameterized Gaits by RL PoWER}
 
-[pic]Fig. 2 illustrates the process of using spline
-representation for the evolving policy parameterization. Fig. 1 shows
-an example for a reinforcement learning process using evolving policy
-parameterization to approximate an unknown function. Petar et
-alIntegrating evolving policy parameterization with
-Hill-climbing(RL?) \cite{kober2009learning-motor-primitives} proposed a RL algorithm
+Here we used an RL approach to change the complexity of the policy
+representation dynamically while the trial is running. In
+\cite{kormushev2011bipedal-walking-energya}'s studies on reducing energy
+consumption for bipedal robots, a mechanism that can
+evolve the policy parameterization was used. The method starts from a
+very simple parameterization and gradually increases its
+representational capability. The method was tested to be capable of generating
+an adaptive policy parameterization that can accommodate increasingly
+more complex policies. Presented in the studies of \cite{kormushev2011bipedal-walking-energya}, the policy
+generated by this approach can reach the global optimum at a fast
+rate when applied to the energy reduction problem.
+Another property found about this method is its chance of converging to a suboptimal solution is reduced, because in the lower-dimensional representation the effect is less exhibited.
+
+\figp{powerSplinesExample}{.75}{An example for an evolving policy parameterization based on
+spline representation of the policy. The set of spline knots is the policy
+parameterization. The spline knots are the actual policy parameter
+values. This original parameterization starts from 4 knots and grows up to 32 knots}
+
+\cite{kober2009learning-motor-primitives} proposed a RL algorithm
 called Policy learning by Weighting Exploration with the
 Returns(PoWER), which is based on Expectation-Maximization algorithm
-(EM). Our proposed technique for evolving the policy parameterization
-are combined with this EM-based RL algorithm PoWER \cite{kober2009learning-motor-primitives}, due to its
-relatively fewer parameters that need tuning. Also because PoWER
+(EM). The proposed technique for evolving the policy parameterization
+is a combination with this EM-based RL algorithm, named PoWER \cite{kober2009learning-motor-primitives}.
+The reason for using this is its relatively fewer parameters that need tuning.
+Also because PoWER
 demonstrated high performance in tasks learned directly on real
@@
 the highest by the importance sampling technique used by the PoWER
-algorithm(). The intuition behind is that high-ranked
-parameterizations have more potential to perform even better in the
+algorithm(). The intuition behind is that highly ranked
+parameterizations have more potential to evolve even better in the
 future. Besides, evolving all the parameterizations requires
@@
 is done on physical robot, explore all the variations of every
-parameterization is not practical. A promising extension can be, using
-physical simulator for the robot, to further explore the
-parameterization space by evolving more parameterizations. Gaits of
-RL\_POWER:For the experiment, we have 3 knots for each spline and 8
-splines in total for the servos. The servo in the hip is not used in
+parameterization is not practical.
+
+For the experiment, we have 3 knots for each servo and 8
+servos in total. The servo in the hip is not used in
 our experiment. Previous work has verified that quadruped gaits
@@
 pattern to every cycle throughout the whole period of
-trial. Specifically, each spline is stretched to servo positions as
+trial. Specifically, each spline(a set of 3 knots) is interpreted to corresponding servo positions as
 following:
 
-\begin{graycode}
-rescale for one cycle:
-Unit Cycle :
-s(si).cyclePos = s(si).YY .* posRange + posMin;
-s(si).cyclePos = s(si).cyclePos(:);
-Repeat cycle,
-Cycle[i] = UnitCycle \% the last pos in a cycle is the
-same as the first pos in the next cycle, from [pic for one gait]
-\end{graycode}
+\begin{table}[b]
+\begin{center}
+\begin{tabular}{|c|c|c|}
+\hline
+Parameters        &                           &       \\
+in $\vec{\theta}$ & Description               & Range \\
+\hline
+\hline
+$f(s1,s2,s3)$        & Spline function           & [0,1] \\  %subject to change
+\hline
+$R$          & Position multiplier                & [256, 768] \\
+\hline
+\end{tabular}
+\caption{The \emph{RL PoWER} motion model parameters.}
+\tablabel{parameters}
+\label{tab:params}
+\end{center}
+\end{table}
+
+\[
+\vec{g}(t) =
+\left[ {\begin{array}{c@{ }c@{ }c@{ }l@{ }l}
+R \cdot f(s1, s2, s3) & \ \cdot \ m_O         & \             & \            & + C \\ % 1
+0                              & \             & \             & \            & + C_C \\ % 8
+\end{array} } \right]
+\]

### 2012-06-01 paper_aimsa/introduction.tex
?

--- 2012-06-01 paper_aimsa/introduction.tex
@@
 
-\figp{quadratotWhiteBooties}{.75}{Example figure...}
+\figp{quadratotWhiteBooties}{.75}{}
 
 Various learning algorithms have been tested to be effective for
-legged robots. Their studies tested the viability of applying RL
-cite{yosinski2011evolvingmethod to legged robot gaits learning. Competitive performance has
+legged robots. These studies tested the viability of applying RL to legged robot gaits learning. Competitive performance has
 been verified on methods such as HyperNEAT \cite{yosinski2011evolving-robot-gaits} and
@@
 the evolutionary algorithms \cite{kormushev2011bipedal-walking-energya}. Here we present a
-different way for learning gaits using RL (Hill climbing). In our
+different way for learning gaits using reinforcement learning and expectation maximization. In our
 experiment, it turned out this method achieved competitive performance
-with Problem Definition:The gait-learning problem is defined to find a
+with
+
+\subsection{Problem Definition}
+The gait-learning problem is defined to find a
 gait that maximizes some specific metric. Mathematically, we define a
@@
 
-According to the definition, open-loop gaits are deterministic. One
+According to its definition, open-loop gaits are deterministic. One
 particular gait should behave exactly the same when it is run from
 trial to trial. However, the actual robot motion and fitness measured
-will vary due to the noisiness of the real world physics.In our
-trials, one gait pattern was generated and sent to the robot and
-executed in an open loop manner, as defined. In particular, in this
-paper, we chose to compare open-loop gaits generated specifically by
-the RL\_POWER and HyperNEAT. The particular metric used will be
+will vary due to the errors and uncertainty of the real world physics. In our
+trials, the gaits generated were sent to the robot and
+executed in an open loop manner. We will present the performance comparison between
+RL POWER and HyperNEAT, the latter of which have the best in details. The metric used will be
 described later.

### 2012-06-02 paper_aimsa/introduction.tex
with minor update

@@
 
-Various learning algorithms have been tested to be effective for
-legged robots. These studies tested the viability of applying RL to legged robot gaits learning. Competitive performance has
-been verified on methods such as HyperNEAT \cite{yosinski2011evolving-robot-gaits} and
-GA \cite{chernova2004an-evolutionary-approach-to-gait} and other methods \cite{hornby2005autonomous-evolution-of-dynamic} \cite{zykov2004evolving-dynamic-gaits} \cite{tellez2006evolving-the-walking-behaviour} \cite{valsalam2008modular-neuroevolution-for-multilegged}.
-Despite the excellent performance of these algorithms, a major
-part of achieving good performance relies in tuning the parameters for
-the evolutionary algorithms \cite{kormushev2011bipedal-walking-energya}. Here we present a
-different way for learning gaits using reinforcement learning and expectation maximization. In our
-experiment, it turned out this method achieved competitive performance
-with
+Various learning algorithms have been proved to be effective for
+legged robots. Algorithms such as HyperNEAT \cite{yosinski2011evolving-robot-gaits},
+GA \cite{chernova2004an-evolutionary-approach-to-gait} and others \cite{hornby2005autonomous-evolution-of-dynamic} \cite{zykov2004evolving-dynamic-gaits} \cite{tellez2006evolving-the-walking-behaviour} \cite{valsalam2008modular-neuroevolution-for-multilegged} have been tested to be effective for automatic learning gaits for robots.
+Despite of the competetive performance, a major
+task lies in tuning the parameters for
+these evolutionary algorithms \cite{kormushev2011bipedal-walking-energya}. Here we present a
+different way for learning gaits using reinforcement learning and expectation maximization, called RL PoWER. In our
+experiment, the main focus of the research is on the applicability of RL PoWER to quadruped robot gait learning.
+Another motivation is to compare the state-of-art neural network algorithm, HyperNEAT, with the performance of our proposed method.
 
 \subsection{Problem Definition}
-The gait-learning problem is defined to find a
+The gait learning problem is defined to find a
 gait that maximizes some specific metric. Mathematically, we define a
@@
 
-According to its definition, open-loop gaits are deterministic. One
+According to this definition, open-loop gaits are deterministic. One
 particular gait should behave exactly the same when it is run from
@@
 executed in an open loop manner. We will present the performance comparison between
-RL POWER and HyperNEAT, the latter of which have the best in details. The metric used will be
+RL PoWER and HyperNEAT.  the latter of which have the best in details. The metric used will be
 described later.

### 2012-06-02 paper_aimsa/relatedWork.tex
with minor update

@@
 \section{Related Work}
-
-Other attempts to solve it
+Many attempts have been made on robot gaits learning using machine learning algorithms, producing competitive results (Chernova and Veloso, 2005; Hornby et al., 2005; Zykov et al., 2004; Clune et al., 2009a, 2011, 2009b,c; T´ellez et al., 2006; Valsalam and Miikkulainen, 2008).
+In fact, Jason et al.
+tested and compared six different algorithms for quadruped robot gait learning in their studies.
+HyperNEAT, a generative encoding neural network algorithm, achieved the best performance among all six methods.
+In Kormushev et al.
+(2011)'s bipedal robot energy reduction research, a reinforcment learning method was used to optimize the performance.
+The results from Kormushev's research showed this algorithm's poetential for walking problems.
+This reinforcement learning method still needs more tests for further explorations.

### 2012-06-02 paper_aimsa/extract.py

@@ -5,16 +5,25 @@ import matplotlib as mpl
 mpl.use('cocoaagg') 
 import matplotlib.pyplot as plt
 import re
+import numpy as np
 
 def main():
     path = '/Users/hs454/Desktop/tmp/'
     runs = sorted(listdir (path))
     total = []
-    avg = []
+    best1_all = []
+    best2_all = []
+    best3_all = []
     avg_all=[]
+    avg=[]
+    stds=[]
+    upper=[]
+    lower=[]
     pattern = re.compile("return_[0-9]*")
     bucket = []
-    best = 0
+    best1 = 0
+    best2 = 0
+    best3 = 0
     for run in runs:
         if pattern.match(run):
             speeds=[]
@@ -27,25 +36,37 @@ def main():
             speeds=sorted(list(set(sorted(speeds))))
             total.append(speeds)
             print speeds[0]
-            max_p = speeds[-1]
-            print max_p
+
     for i in range(0,300):
-        mean = (list(total[0][i])[1]+list(total[1][i])[1]+list(total[2][i])[1])/3
+        e1 = list(total[0][i])[1]
+        e2 = list(total[1][i])[1]
+        e3 = list(total[2][i])[1]
+        if best1 < e1:
+            best1=e1
+        if best2 < e2:
+            best2=e2
+        if best3 < e3:
+            best3=e3
+        std = np.std([best1,best2,best3])
+        stds.append(std)
+        best1_all.append(best1)
+        best2_all.append(best2)
+        best3_all.append(best3)
+        mean = (best1+best2+best3)/3
         avg.append(mean)
+        upper.append(mean+std)
+        lower.append(mean-std)
         avg_all.append((mean,i))
         
-    best = 0
-    best_index=0
-    for x,ii in avg_all:
-        if best < x:
-            best = x
-            best_index = ii
-    
-    plt.plot(range(1,len(avg)+1), avg)
-    plt.ylabel('AVERAGE SPEED')
+    maxi = max([best1,best2,best3])
+    print 'global max ',maxi
+    std = np.std(avg)
+    print 'standard deviation: ', std
+    plt.plot(range(1,len(avg)+1), avg, 'b', range(1,len(avg)+1),upper,'g', range(1,len(avg)+1), lower,'r')
+    plt.ylabel('Speed(cm/sec)')
     plt.xlabel('Trial')
[…]

### 2012-06-05 paper_aimsa/abstract.tex
fix 6/5 version (Jason: trimmed generated files out)

A reinforcement learning approach which can evolve the policy parameterization dynamically during
the learning process for quadruped robot gait . By gradually increasing the representational
power of the policy parameterization, it manages to find better
policies faster than HyperNEAT, a state-of-the-art evolutionary ANN algorithm that achieved the best performance in previous studies.

Expectation-Maximization-based reinforcement learning algorithm
is used. The test was done on a physical quadrupedal robot(QuadraTot). The results show that the evolving policy parameterization combined with expectation-maximization (RL PoWER)
outperforms substantially the HyperNEAT, by having
faster convergence and higher final
reward. According to the statistics collected, the RL PoWER achieves 16.3\%
advantage over the HyperNEAT, the best method tested on QuadraTot in previous studies.

### 2012-06-05 paper_aimsa/introduction.tex
fix 6/5 version (Jason: trimmed generated files out)

@@
 
-\figp{quadratotWhiteBooties}{.75}{}
+\figp{quadratotWhiteBooties}{.75}{The quadratot robot used in this research. The white booties are newly added to prevent sliding on the surface thus minimize measure error.}
 
@@
 Despite of the competetive performance, a major
-task lies in tuning the parameters for
+task that are usually hidden in the published results lies in tuning the parameters for
 these evolutionary algorithms \cite{kormushev2011bipedal-walking-energya}. Here we present a
@@
 experiment, the main focus of the research is on the applicability of RL PoWER to quadruped robot gait learning.
-Another motivation is to compare the state-of-art neural network algorithm, HyperNEAT, with the performance of our proposed method.
+Another motivation is to compare the state-of-the-art neural network algorithm, HyperNEAT, with our proposed method in quadruped robot gait learning.
 
@@
 gait as a function that specifies a vector of commanded motor
-positions for a robot over time. Gaits without feedback --- also
-called open-loop gaits --- can be defined as
+positions for a robot over time. Gaits without feedback---also
+called open-loop gaits---can be defined as
 
@@
 trials, the gaits generated were sent to the robot and
-executed in an open loop manner. We will present the performance comparison between
-RL PoWER and HyperNEAT.  the latter of which have the best in details. The metric used will be
+executed in an open loop manner. We will measure and analyze the performance between
+RL PoWER and HyperNEAT, the latter of which will be the focus of discussion in this paper. The metric used for the fitness in this paper will be
 described later.

### 2012-06-05 paper_aimsa/methods.tex
fix 6/5 version (Jason: trimmed generated files out)

@@
 \subsection{Policy Representation by Splines}
-
 The simplest model with back-compatibility is geometric
@@
 order to do this, we need to define an algorithm for evolving the
-parameterization from $K$ to $L$ knots ($L > K$), which is formulated in
-Algorithm \edit{proper reference}.  Without loss of generality, the policy parameters are
+parameterization from $K$ to $L$ knots ($L > K$), which is formulated as 
+Algorithm 1 in \cite{kormushev2011bipedal-walking-energy}.  Without loss of generality, the policy parameters are
 normalized into $[0, 1]$, and appropriately scaled/shifted as necessary
@@
 representation dynamically while the trial is running. In
-\cite{kormushev2011bipedal-walking-energya}'s studies on reducing energy
+\cite{kormushev2011bipedal-walking-energy}'s studies on reducing energy
 consumption for bipedal robots, a mechanism that can
@@
 an adaptive policy parameterization that can accommodate increasingly
-more complex policies. Presented in the studies of \cite{kormushev2011bipedal-walking-energya}, the policy
+more complex policies. Presented in the studies of \cite{kormushev2011bipedal-walking-energy}, the policy
 generated by this approach can reach the global optimum at a fast
 rate when applied to the energy reduction problem.
-Another property found about this method is its chance of converging to a suboptimal solution is reduced, because in the lower-dimensional representation the effect is less exhibited.
+Another property found about this method is its chance of converging to a suboptimal solution is reduced, because in the lower-dimensional representation this effect is less exhibited.
 
-\figp{powerSplinesExample}{.75}{An example for an evolving policy parameterization based on
+\figp{powerSplinesExample}{.6}{An example for an evolving policy parameterization based on
 spline representation of the policy. The set of spline knots is the policy
@@
 (EM). The proposed technique for evolving the policy parameterization
-is a combination with this EM-based RL algorithm, named PoWER \cite{kober2009learning-motor-primitives}.
-The reason for using this is its relatively fewer parameters that need tuning.
-Also because PoWER
-demonstrated high performance in tasks learned directly on real
-robots, such as pendulum swing-up and pancake flipping task \cite{kormushev2010robot-motor-skill}. We
+is a combination with this EM-based RL algorithm, named PoWER \cite{kober2009learning-motor-primitives}. The reason for using this is its relatively fewer parameters that need tuning. We
 evolved the policy parameterization only on those past trials ranked
 the highest by the importance sampling technique used by the PoWER
-algorithm(). The intuition behind is that highly ranked
+algorithm. The intuition behind is that highly ranked
 parameterizations have more potential to evolve even better in the
-future. Besides, evolving all the parameterizations requires
-significantly more efforts, as shown in \cite{bongard2006resilient-machines-through}. Since our experiment
-is done on physical robot, explore all the variations of every
-parameterization is not practical.
+future. Besides, evolving all the parameterizations increases the exploring space. Since our experiment
+is done on a physical robot, explore all the variations of every
+parameterization is not practical. Future work may incorporate simulations into the studies, as illustrated in \cite{bongard2006resilient-machines-through}.
 
-For the experiment, we have 3 knots for each servo and 8
+For the experiment, we set 3 knots for each servo and there are 8
 servos in total. The servo in the hip is not used in
@@
 perform better when they are coordinated \cite{clune2009evolving-coordinated-quadruped} \cite{clune2011on-the-performance-of-indirect-encoding}
-\cite{valsalam2008modular-neuroevolution-for-multilegged}. For each set of
-splines, we calculate its corresponding parameterized gait within a
-time cycle. Given the pattern of one unit cycle, then apply the same
-pattern to every cycle throughout the whole period of
-trial. Specifically, each spline(a set of 3 knots) is interpreted to corresponding servo positions as
+\cite{valsalam2008modular-neuroevolution-for-multilegged}. For each spline, we calculate its corresponding parameterized gait for one unit time cycle. Given that, then apply the same
+pattern to every cycle throughout the 12 seconds of one
+trial. Specifically, each spline(a set of 3 knots) is interpreted to its corresponding servo positions as
 following:
@@
 \left[ {\begin{array}{c@{ }c@{ }c@{ }l@{ }l}
-R \cdot f(s1, s2, s3) & \ \cdot \ m_O         & \             & \            & + C \\ % 1
+R \cdot f(s1, s2, s3) & \ \          & \             & \            & + C \\ % 1
 0                              & \             & \             & \            & + C_C \\ % 8

### 2012-06-05 paper_aimsa/report/report.tex
fix 6/5 version (Jason: trimmed generated files out)

\section{Introduction}

% Motivate and abstractly describe the problem you are addressing and
% how you are addressing it. What is the problem? Why is it important?
% What is your basic approach? A short discussion of how it fits into
% % related work in the area is also desirable. Summarize the basic
% results and conclusions that you will present.

Applications of walking robots often call for the ability to walk as
quickly, smartly or with as little power as possible. Some groups have
used learning methods to generate gaits optimized for some metric.
Approaches differ in their starting assumptions, some essentially
tweaking the parameters of a hand-tuned model others
exploring a reasonably compact parameter space.
Here we present a new method, RL PoWER, reinforcement learning with EM(Expectation Maximization) as it was tested effective on related robotics problems. 
This paper shows a comparison of the two different learning methods 
implemented, HyperNEAT and RL PoWER.  RL PoWER
method created walks that are substantially faster than the previous
HyperNEAT gait. I invite readers with short attention spans to view
two videos to compare the previous gaits and the current stable gaits:

\url{http://www.youtube.com/watch?v=ODoiOj9DdGg}

\url{http://www.youtube.com/watch?v=LI76mQepEjc&feature=youtu.be}

%The code so far consists of the robot class, optimization class,
%parameterized model, sine model, and camera feedback.

\section{Problem definition}
% Precisely define the problem you are addressing (i.e. formally specify
% the inputs and outputs). Elaborate on why this is an interesting and
% important problem.

I am testing and comparing two learning methods, HyperNEAT and RL PoWER
, to design a parametrized gait for a quadruped robot from the Cornell
Computational Synthesis Lab that maximize forward walking speed. 
Subproblems include improving the robostness of the robot so that it 
fails less frequently and make the measurement more reliable by minimizing the leg sliding.

\section{Implementation}
\seclabel{implement}
% Describe how you implemented your system and how you structured it. 
% This should give an overview of the system, not a detailed 
% documentation of the code. The documentation of the code is part of 
% the code you hand in. You might want to comment on high-level design 
% decisions that you made. Also explain how you obtained your
% data and any pre-processing you did to it.
\begin{itemize}
\item \code{RL\_PoWER.m*} The RL PoWER algorithm implementation defined in matlab code. Output a gait sequence file for 9 motors. Supports restart from previously interrupted trial. 
\item \code{runRobot.m*} Communication code for reading the generated file of motor actions and execute on robot by calling the python code. 
\item \code{localSettings.m*} Execution settings file. 
\item \code{load\_trajectory.m*} Load the generated gait file and do the conversion.
\item \code{exportTrajectoryWithoutRescaling.m*} Export the spline function to actual robot gaits
\item \code{paramToPolicy.m*} Transforming spline representation to actual robot gaits parameters.
\item \code{policyToParam.m*} Transforming gait parameters to corresponding spline representation.
\item \code{PoWER\_QuadraTot\_23.m*} main method for system call
\end{itemize}

\footnote{*Contributed equally by Haocheng Shen, Petar Kormushev and Jason Yosinski}

\section{Method}
\seclabel{method}

% Describe in reasonable detail the algorithm you are using to address
% this problem. A pseudo-code description of the algorithm you are
% using is frequently useful. If it makes sense for your project,
% trace through a concrete example, showing how your algorithm
% processes this example. The example should be complex enough to
% illustrate all of the important aspects of the problem but simple
% enough to be easily understood. If possible, an intuitively
% meaningful example is better than one with meaningless symbols.
\subsection{Policy Representation by Splines}
The simplest model with back-compatibility is geometric
splines. For a given model f(x) with K knots, we can preserve the
exact shape of the generated curve while adding extra knots to the
original spline. Say, if we put one additional knot between every two
consecutive knots of the original spline, we end up with a 2K - 1
knots and a spline that has the same shape as the original one. In
order to do this, we need to define an algorithm for evolving the
parameterization from $K$ to $L$ knots ($L > K$), which is formulated as 
Algorithm 1. Without loss of generality, the policy parameters are
normalized into $[0, 1]$, and appropriately scaled/shifted as necessary
later upon use.

\begin{algorithm}
\caption{EvolvePolicy-Spline ($P_{current}$: current policy, $L$: desired new number of parameters)}
\label{alg_Evolve_Spline}
\begin{algorithmic}[1]

\STATE $K$ $\leftarrow$ $P_{current}.numberOfParameters$

\STATE $X_{current}$ $\leftarrow$ $[0, \frac{1}{K-1}, \frac{2}{K-1}, ... , 1]$ % Knots
\STATE $Y_{current}$ $\leftarrow$ $P_{current}.parameterValues$ % Values
\STATE $S_{current}$ $\leftarrow$ ConstructSpline($X_{current}$, $Y_{current}$) % Spline
%\COMMENT{ Optional: make the spline cyclic}

\STATE $X_{new}$ $\leftarrow$ $[0, \frac{1}{L-1}, \frac{2}{L-1}, ... , 1]$
\STATE $Y_{new}$ $\leftarrow$ EvaluateSplineAtKnots($S_{current}$, $X_{new}$)
\STATE $S_{new}$ $\leftarrow$ ConstructSpline($X_{new}$, $Y_{new}$)
%\COMMENT{ Optional: make the spline cyclic}

\STATE $P_{new}.numberOfParameters$ $\leftarrow$ $L$
\STATE $P_{new}.parameterValues$ $\leftarrow$ $S_{new}$.$Y_{new}$

\STATE return $P_{new}$

\end{algorithmic}
\end{algorithm}

For the experiment, we set 3 knots for each servo and there are 8
servos in total. The servo in the hip is not used in
our experiment. Previous work has verified that quadruped gaits
perform better when they are coordinated \cite{clune2009evolving-coordinated-quadruped} \cite{clune2011on-the-performance-of-indirect-encoding}
\cite{valsalam2008modular-neuroevolution-for-multilegged}. For each spline, we calculate its corresponding parameterized gait for one unit time cycle. Given that, then apply the same
pattern to every cycle throughout the 12 seconds of one
trial. Specifically, each spline(a set of 3 knots) is interpreted to its corresponding servo positions as
following:
\[
\vec{g}(t) =
\left[ {\begin{array}{c@{ }c@{ }c@{ }l@{ }l}
R \cdot f(s1, s2, s3) & \ \          & \             & \            & + C \\ % 1
0                              & \             & \             & \            & + C_C \\ % 8
\end{array} } \right]
\]
And the meaning of its parameters are listed in \tabref{parameters}
\begin{table}[b]
\begin{center}
\begin{tabular}{|c|c|c|}
\hline
Parameters        &                           &       \\
in $\vec{\theta}$ & Description               & Range \\
\hline
\hline
$f(s1,s2,s3)$        & Spline function           & [0,1] \\  %subject to change
\hline
$R$          & Position multiplier                & [256, 768] \\
\hline
\end{tabular}
\caption{The \emph{RL PoWER} motion model parameters.}
\tablabel{parameters}
\label{tab:params}
\end{center}
\end{table}

\section{Related work}
% Briefly explain who else worked on related problems in the past and what
% methods they used. Explain if you are using similar methods, or if your
% approach is different and if so - how (either is ok).
Many attempts have been made on robot gaits learning using machine learning algorithms, producing competitive results (Chernova and Veloso, 2005; Hornby et al., 2005; Zykov et al., 2004; Clune et al., 2009a, 2011, 2009b,c; T´ellez et al., 2006; Valsalam and Miikkulainen, 2008). In fact, Jason et al. tested and compared six different algorithms for quadruped robot gait learning in their studies. HyperNEAT, a generative encoding neural network algorithm, achieved the best performance among all six methods. In Kormushev et al. (2011)'s bipedal robot energy reduction research, a reinforcment learning method was used to optimize the performance. The results from Kormushev's research showed this algorithm's poetential for walking problems. This reinforcement learning method still needs more tests for further explorations.

\section{Experimental Evaluation}
Our performance metric was the displacement over the evaluation period of 12 seconds
for each. Same as Yosinski et al. (2011), the displacement was measured using a
Wii remote that was placed on the ceiling. Different from the original model described
in Yosinski et al. (2011), the quadruped robot was equipped with a three-infrared-LED
cluster on top rather than just one. The reason for this setup is that when fierce gaits
were executed, the Wii remote loses tracking of the robot position due to the limited
visible angle of a single LED. These three LEDs were place tightly together to act
as one signal emitter. Each LED was tilted outwards in order to maximize the visible
range. A separate tracking server ran on the robot PC interacted with theWii remote via
bluetooth by using the CWiid library. A corresponding client communicates with this
server via socket connection. Our fitness evaluation code talks with the Wii remote by
using this client and updates the position from beginning to end during each run. The
metric for evaluating gaits was the Euclidian distance the robot traveled during a 12-
second run on flat surface. Previous work done in Yosinski et al. (2011) and Clune et al.
(2009) suggested that extremely fierce gaits are not viable in general. These gaits tend
to either overburden the servos or flip the robot. Thus, RL PoWER were tested after
carefully cropping out the extremely fierce gaits each time. As described earlier, each
leg has two joints, inner joint j1 and outer j2. The position of one leg is determined by
the sum of the values of j1 and j2. Extremely fierce gaits usually have relatively small
values of g. As shown in Figure 3, this cropping method works by mapping the wild
gaits onto the boundary of a quasi-triangular area in the two-dimensional space of j1
and j2.

\section{Results}

\subsection{Hardware and interface Improvement}
The robot has been verified to fail under extreme fierce motions. So, one change I made to this is preventing these actions to happen from the first hand. Thus the servo failure chance is solved for most of the time. The threshold of cropping was carefully chosen so that most gaits are maintained.

\subsection{Quantitative}
The results for the gaits evolved by RL PoWER and HyperNEAT are shown in \tabref{results}. A total of 900 evaluations were performed for RL PoWER (300 for each of three runs). The reason these two algorithms have different number of tirals is that runs continued until the performance plateaued, which we defined as when there was no improvement during the last third of a run. Overall, the RL PoWER gaits were
faster by far, beating the HyperNEAT when comparing
either average or best gaits.

\begin{table}
\begin{center}
\begin{tabular}{|r|c|c|c||c|}
\hline
                                         & Average & Std. Dev \\
\hline                                    
\hline                                    
RL PoWER                       & 7.62  &    2.1   \\
\hline
HyperNEAT                      & 6.28   &   1.26   \\
\hline
\end{tabular}
\caption{The average and standard deviation of the best gaits found
  for both algorithm during each of three runs, in cm/sec.}  \tablabel{results}
\end{center}
\end{table}

\subsection{Discussion}
% Is your hypothesis supported? What conclusions do the results support
% about the strengths and weaknesses of your method compared to other 
% methods? How can the results be explained in terms of the underlying 
% properties of the algorithm and/or the data.
Overall, the RL PoWER gaits were
faster by far, beating the HyperNEAT when comparing
either average or best gaits. I believe that this is because RL PoWER uses evolvable splines to represent gaits As explained earlier,  
The single best gait found during this study had a speed of 11.09 centimeters/sec,  14.7\% better than the best HyperNEAT gait. \figref{rlPattern} shows a typical RL PoWER gait that had high fitness. Compared with \figref{neat_110115_211410_00000_002_filt_zoom}, the pattern of RL PoWER's motion is less
complex but more regular than the HyperNEAT, but producing higher speed. As shown in \figref{rlPattern}
, multiple motors are more coordinated than the HyperNEAT gait. 
\figp{rlPattern}{0.75}{The motor positions of a typical well performing gait over one cycle. It matches exactly the shape of its corresponding spline representation. This pattern shows the simplicity and power of geometric splines representations.}

\figp{neat_110115_211410_00000_002_filt_zoom}{0.75}{A well performing HyperNEAT gait's pattern of motor positions}

A corresponding observation from the graph on this property is that the noisiness of
RL PoWER is higher than HyperNEAT. One reason for this is that the RL PoWER intentionally adds noise when exploring the space near the best ranked policies. Another phenomenon seen from the \figref{rlPerform} is that one run of RL PoWER's general performance largely depends on initial gaits, which is a trait of hill climbing algorithms. The larger standard deviation in RL PoWER for space exploring can be easily fixed by tuning the noise parameter. RL PoWER converges to optimality at a higher rate. This is due to the more heuristically guided reinforcement learning of RL PoWER. HyperNEAT, on the other hand, has a larger dimension to explore due to its generative encoding method. While the evolvable spline representation used by RL PoWER has lower dimensions. The convergence rate of HyperNEAT is thus lower.

\section{Future work}
% What are the major shortcomings of your current method? For each 
% shortcoming, propose additions or enhancements that would help overcome it.
Though 900 trials for RL PoWER and 540 trials for HyperNEAT has been used to test the applicability of RL PoWER and 
compare the performance of the two, still, much more runs are needed to statisstically compare the two algorithms.

So, here are some work that may be achieved in future
\begin{itemize}
\item Improve hardware connection, lower the chance of servo failure
\item More runs and/or longer runs
\item Tuning simulator for good approximation 
\item Using simulation to statistically compare the algorithms on quadrupedal robots
\end{itemize}

\section{Conclusion}
% Briefly summarize the important results and conclusions presented in 
% the paper. What are the most important points illustrated by your 
% work? How will your results improve future research and applications 
% in the area?
The previously encountered hardware problems about using QuadraTot have been solved by the newly added booties.
The measurement reliability has been enchanced by the fixed Robot class in python, preventing extremely fierce gaits.

We have presented a new reinforcement-learning-based algorithm for
optimizing a quadrupedal gait for linear speed. We implemented and
tested six learning strategies for parameterized gaits and compared
them to gaits produced by neural networks evolved with the HyperNEAT
generative encoding. Though both methods resulted in an improvement
over the robot’s previous \naive gaits, based on the statistics
collected, RL PoWER has a more elegant and consistent performance.

One guess led by this study is that for feedback oriented tasks,
reinforcement learning methods are more fit in natural. Despite the
complexities of HyperNEAT, a simpler algorithm on the code level,
delivered better performance in general. Also, evolvable spline
interpolation is shown to be simple and representationally powerful at
the same time. Evolvable splines can serve as a general representation for various other
learning problems.

\section{Acknowledgments}
% List any people not on the team who helped you with your project: Your
% TA, other people you consulted with or had useful discussions (say in 
% a word or two what they did), people who proofread your report, and 
% any external code you used (libraries etc).
\begin{itemize}
\item Hod Lipson, Cornell Computational Synthesis Lab: adviser.
\item Jason Yosinski, Cornell Computational Syntheis Lab: adviser
\item Petar Kormushev, Italy Institute of Technology: inventor of RL PoWER algorithm
\item Kyrrehg Glette, University of Oslo: collaborated on the design of anti-sliding booties.
\end{itemize}

\section{Team}
% A table listing the names of the people who worked on the project, and who
% did what.

\begin{itemize}
\item Haocheng Shen
\begin{itemize}
\item Robot platform reconstructing and improvement
\item RL PoWER algorithm integration and testing
\item \code{RL\_PoWER.m} and \code{runRobot.m} matlab code and python code integration for running trials on the reconstructed robot platform
\item \code{Robot.py} classes, with new cropping method and bug fix
\item \code{extract.py} scripts, for processing collected gait data and plotting
\item \code{extractGait.py} scripts, for processing collected gait data and plotting
\item Plotting and evaluating results
\end{itemize}

\item Jason Yosinski
\begin{itemize}
\item {RL\_PoWER}\code{exportTrajectoryWithoutRescaling.m} and \code{getDataPath.m} \code{paramToPolicy}, \code{LoadTrajectory}, and \code{PoWER\_QuadraTot.m} matlab code
\item mentoring \& evaluating results
\end{itemize}

\item Petar Kormushev
\begin{itemize}
\item The inventor of combining reinforcemen learning and EM algoorithms for quadrupedal robot gait learning
\item \code{RL\_PoWER.m}\code{exportTrajectoryWithoutRescaling.m} and \code{getDataPath.m} \code{paramToPolicy}, \code{LoadTrajectory}, and \code{PoWER\_QuadraTot.m} matlab code
\item implementation of the original algorithm
\end{itemize}

\section{Appendix}
% Describe the code files you uploaded into CMS
\begin{itemize}
\item Publication paper draft for AIMSA conference
\item RL PoWER performance video online 
\item matlab data workspace files for recording all three runs' parameters
\item python code for processing data
\end{itemize}
\end{itemize}

% LocalWords:  vertices

### 2012-06-05 paper_aimsa/extract.py

@@ -27,13 +27,17 @@ def main():
     for run in runs:
         if pattern.match(run):
             speeds=[]
+            speeds2=[]
             f = open(run)
             lines = f.readlines()
             for line in lines:
                 spd_tr= line.split()
                 ts= (int(spd_tr[1]),float(spd_tr[0])*(14.0/12))
                 speeds.append(ts)
+                speeds2.append(float(spd_tr[0])*(14.0/12))
             speeds=sorted(list(set(sorted(speeds))))
+            print "mean: ", np.mean(speeds2)
+            print "std: ", np.std(speeds2)
             total.append(speeds)
             print speeds[0]
 
@@ -62,6 +66,7 @@ def main():
     print 'global max ',maxi
     std = np.std(avg)
     print 'standard deviation: ', std
+    
     plt.plot(range(1,len(avg)+1), avg, 'b', range(1,len(avg)+1),upper,'g', range(1,len(avg)+1), lower,'r')
     plt.ylabel('Speed(cm/sec)')
     plt.xlabel('Trial')

### 2012-06-05 paper_aimsa/extractGait.app/Contents/Resources/extractGait.py

@@ -1,9 +1,9 @@
-#! /usr/bin/env python-32
+#! /usr/bin/env python
 
 import pdb
 from os import listdir
 import matplotlib as mpl 
-mpl.use('cocoaagg') 
+#mpl.use('cocoaagg') 
 import matplotlib.pyplot as plt
 import re
 
@@ -40,7 +40,7 @@ def main():
     ax = plt.axis()
     plt.axis((min(t), max(t), ax[2], ax[3]))
     plt.savefig('gait.pdf')
-    #plt.show()
+    plt.show()
     
     
 if __name__ == '__main__':

### 2012-06-08 paper_aimsa/conclusion.tex
Added links to many references
fix2

@@
 over the robot’s previous \naive gaits, based on the statistics
-collected, RL\_PoWER has a more elegant and consistent performance.
-
-Based on previous studies on evolving gait on quadruped robots[xx],
-RL\_PoWER produced higher performing gaits than all of the
-parameterized methods. Its best-performing gait traveled XX body
-lengths per minute.  HyperNEAT, showed in the previous studies,
-achieved the competitive best performance with RL\_PoWER. HyperNEAT
-gaits
+collected, RL PoWER has a more elegant and consistent performance.
 
 Though over 900 trials have been made to investigate the applicability
-of RL\_PoWER to quadruped robots. It is difficult to gather the enough
+of RL PoWER to quadruped robots. It is difficult to gather the enough
 trials that would be necessary to properly rank the methods. One
@@
 effective, and the hardware trials would serve to continuously ground
-and refine the simulator. [KYRREG XX]
+and refine the simulator. \cite{glette2012evolution-of-locomotion-in-a-simulated}
 
@@
 interpolation is shown to be simple and representationally powerful at
-the same time. Can serve as a general representation for various other
+the same time. Evolvable splines can serve as a general representation for various other
 learning problems.

### 2012-06-08 paper_aimsa/main.tex
fix2

@@
 
-\title{Aracna: An Open-Source Quadruped Platform for Evolutionary Robotics}
-\author{Sara Lohmann\thanks{~~Authors contributed equally to this work.}~, Jason Yosinski$^{*}$, Eric Gold, Jeff Clune, Jeremy Blum \and Hod Lipson  \\
+\title{RL-PoWER, Evolving Policy Parameterization with Reinforcement Learning and its Application to Quadrupedal Walking Optimization}
+\author{Haocheng Shen, Jason Yosinski, Petar Kormushev, \and Hod Lipson  \\
   \mbox{}\\
 Cornell University, 239 Upson Hall, Ithaca, NY 14853 \\
-\texttt{sml253@cornell.edu, yosinski@cs.cornell.edu}}
+\texttt{hs454@cornell.edu, yosinski@cs.cornell.edu}}
 

### 2012-06-08 paper_aimsa/extract.py

@@ -1,0 +1,77 @@
+#! /usr/bin/env python-32
+
+from os import listdir
+import matplotlib as mpl 
+mpl.use('cocoaagg') 
+import matplotlib.pyplot as plt
+import re
+import numpy as np
+
+def main():
+    path = '/Users/hs454/Desktop/tmp/'
+    runs = sorted(listdir (path))
+    total = []
+    best1_all = []
+    best2_all = []
+    best3_all = []
+    avg_all=[]
+    avg=[]
+    stds=[]
+    upper=[]
+    lower=[]
+    pattern = re.compile("return_[0-9]*")
+    bucket = []
+    best1 = 0
+    best2 = 0
+    best3 = 0
+    for run in runs:
+        if pattern.match(run):
+            speeds=[]
+            speeds2=[]
+            f = open(run)
+            lines = f.readlines()
+            for line in lines:
+                spd_tr= line.split()
+                ts= (int(spd_tr[1]),float(spd_tr[0])*(14.0/12))
+                speeds.append(ts)
+                speeds2.append(float(spd_tr[0])*(14.0/12))
+            speeds=sorted(list(set(sorted(speeds))))
+            print "mean: ", np.mean(speeds2)
+            print "std: ", np.std(speeds2)
+            total.append(speeds)
+            print speeds[0]
+
+    for i in range(0,300):
+        e1 = list(total[0][i])[1]
+        e2 = list(total[1][i])[1]
+        e3 = list(total[2][i])[1]
+        if best1 < e1:
+            best1=e1
+        if best2 < e2:
+            best2=e2
+        if best3 < e3:
+            best3=e3
+        std = np.std([best1,best2,best3])
+        stds.append(std)
+        best1_all.append(best1)
+        best2_all.append(best2)
+        best3_all.append(best3)
+        mean = (best1+best2+best3)/3
+        avg.append(mean)
+        upper.append(mean+std)
+        lower.append(mean-std)
+        avg_all.append((mean,i))
+        
+    maxi = max([best1,best2,best3])
+    print 'global max ',maxi
+    std = np.std(avg)
+    print 'standard deviation: ', std
+    
+    plt.plot(range(1,len(avg)+1), avg, 'b', range(1,len(avg)+1),upper,'g', range(1,len(avg)+1), lower,'r')
+    plt.ylabel('Speed(cm/sec)')
+    plt.xlabel('Trial')
+    plt.savefig('average.pdf')
+    #plt.show()
+
+if __name__ == '__main__':
+	main();

### 2012-06-08 paper_aimsa/extractGait.app/Contents/Resources/__argvemulator_extractGait.py

@@ -1,0 +1,4 @@
+import argvemulator, os
+
+argvemulator.ArgvCollector().mainloop()
+execfile(os.path.join(os.path.split(__file__)[0], "extractGait.py"))

### 2012-06-08 paper_aimsa_report/abstract.tex
fixe4

--- 2012-06-05 paper_aimsa/abstract.tex
@@
-A reinforcement learning approach which can evolve the policy parameterization dynamically during
-the learning process for quadruped robot gait . By gradually increasing the representational
+A reinforcement learning approach is proposed
+which can evolve the policy parameterization dynamically during
+the learning process for quadruped robot gait learning. By gradually increasing the representational
 power of the policy parameterization, it manages to find better
-policies faster than HyperNEAT, a state-of-the-art evolutionary ANN algorithm that achieved the best performance in previous studies.
+policies faster than by using HyperNEAT, a state-of-art evolutionary ANN algorithm that achieved the best performance in previous studies.
 
 Expectation-Maximization-based reinforcement learning algorithm
-is used. The test was done on a physical quadrupedal robot(QuadraTot). The results show that the evolving policy parameterization combined with expectation-maximization (RL PoWER)
+is used. The results show that the evolving policy parameterization combined with expectation-maximization (RL PoWER)
 outperforms substantially the HyperNEAT, by having
 faster convergence and higher final
-reward. According to the statistics collected, the RL PoWER achieves 16.3\%
-advantage over the HyperNEAT, the best method tested on QuadraTot in previous studies.
+reward.

### 2012-06-10 paper_aimsa/abstract.tex
Edits toward a more polished paper

@@
 A reinforcement learning approach which can evolve the policy parameterization dynamically during
-the learning process for quadruped robot gait . By gradually increasing the representational
+the learning process for quadruped robot gait \edit{???}. By gradually increasing the representational
 power of the policy parameterization, it manages to find better
@@
 Expectation-Maximization-based reinforcement learning algorithm
-is used. The test was done on a physical quadrupedal robot(QuadraTot). The results show that the evolving policy parameterization combined with expectation-maximization (RL PoWER)
+is used. The test was done on a physical quadrupedal robot (QuadraTot). The results show that the evolving policy parameterization combined with expectation-maximization (RL PoWER)
 outperforms substantially the HyperNEAT, by having

### 2012-06-10 paper_aimsa/experimentalSetup.tex
Edits toward a more polished paper

@@
 joint with AX-12+servo in each of the four legs, and one AX-18+ servo
-at the center. To avoid the formerly reported AX-18+ servos are used
+at the center. To avoid the formerly reported \edit{???} AX-18+ servos are used
 in this robot because of their stronger actuation power than that of
 AX-12+. Each servo could be set to a position in the range [0, 1023]
-by using PyDynamixel library, corresponding to a physical range [-120,
-  +120]. Also, to prevent collisions with the robot body, the control
-module filter out the commands to a safe range. This range was [-85,
-  +60 ] for the inner leg servos, [-113 , +39 ] for the outer leg
-servos, and [-28 , +28] for the center hip servo. In the studies of
+by using pydynamixel library, corresponding to a physical range [-120$^\circ$,
+  +120$^\circ$]. Also, to prevent collisions with the robot body, the control
+module filter out the commands to a safe range. This range was [-85$^\circ$,
+  +60$^\circ$] for the inner leg servos, [-113$^\circ$, +39$^\circ$] for the outer leg
+servos, and [-28$^\circ$, +28$^\circ$] for the center hip servo. In the studies of
 this paper, tethered cables powered both the computer and the
 servos. It measures approximately 39.5 centimeters from leg to
-opposite leg in the crouch position showed in [pic].Our performance
-metric was the displacement over the evaluation period of 12 seconds
+opposite leg in the crouch position showed in \figref{quadratotWhiteBooties}.
+
+Our performance metric was the displacement over the evaluation period of 12 seconds
 for each. Same as \cite{yosinski2011evolving-robot-gaits}, the displacement was
@@
 place tightly together to act as one signal emitter. Each LED was
-tilted outwards in order to maximize the visible range. [pic] A
+tilted outwards in order to maximize the visible range. A
 separate tracking server ran on the robot PC interacted with the Wii
 remote via bluetooth by using the CWiid library.  A corresponding
-client communicates with this server via a socket connection. Our
+client communicates with this server via socket connection. Our
 fitness evaluation code talks with the Wii remote by using this client
@@
 general. These gaits tend to either overburden the servos or flip the
-robot. Thus, RL\_POWER(?) were tested after carefully cropping out the
-extremely fierce gaits each time. (As described earlier, each leg has
+robot. Thus, RL PoWER were tested after carefully cropping out the
+extremely fierce gaits each time. As described earlier, each leg has
 two joints, inner joint j1 and outer j2. The position of one leg is
 determined by the sum of the values of j1 and j2. Extremely fierce
-gaits usually have relatively small values of g. Thus cropping works
-by mapping the wild gaits onto the boundary of a triangular space over
-the two dimensional space of j1 and j2. \figref{smartCrop}....
+gaits usually have relatively small values of $g$. As shown in \figref{smartCrop}, this cropping method works
+by mapping the wild gaits onto the boundary of a quasi-triangular area in
+the two-dimensional space of j1 and j2.
 
-\figp{smartCrop}{1}{$g = j1 +  j2$ caption}
+\figp{smartCrop}{1}{The red line is the cropping border, $g = j1 +  j2$. As shown in \figref{sixLegPositions}}
 
-\figp{sixLegPositions}{1}{caption... the left two are allowed, middle
-  two are on the red border line from \figref{smartCrop}, right two
-  are disallowed.}
+\figp{sixLegPositions}{1}{The left two positions are allowed; middle
+  two positions are on the red border line from \figref{smartCrop}; right two positions are deemed as ``extremely fierce'', thus are disallowed.}

### 2012-06-10 paper_aimsa/introduction.tex
Edits toward a more polished paper

@@
 GA \cite{chernova2004an-evolutionary-approach-to-gait} and others \cite{hornby2005autonomous-evolution-of-dynamic} \cite{zykov2004evolving-dynamic-gaits} \cite{tellez2006evolving-the-walking-behaviour} \cite{valsalam2008modular-neuroevolution-for-multilegged} have been tested to be effective for automatic learning gaits for robots.
-Despite of the competetive performance, a major
+Despite competetive performance, a major
 task that are usually hidden in the published results lies in tuning the parameters for
-these evolutionary algorithms \cite{kormushev2011bipedal-walking-energya}. Here we present a
+these evolutionary algorithms \cite{kormushev2011bipedal-walking-energy}. Here we present a
 different way for learning gaits using reinforcement learning and expectation maximization, called RL PoWER. In our
@@
 \subsection{Problem Definition}
-The gait learning problem is defined to find a
+
+As in \cite{yosinski2011evolving-robot-gaits}, the gait learning problem is defined to find a
 gait that maximizes some specific metric. Mathematically, we define a
@@
 trials, the gaits generated were sent to the robot and
-executed in an open loop manner. We will measure and analyze the performance between
-RL PoWER and HyperNEAT, the latter of which will be the focus of discussion in this paper. The metric used for the fitness in this paper will be
+executed in an open loop manner. We will measure and analyze the performance between HyperNEAT and 
+RL PoWER, the latter of which will be the focus of discussion in this paper. The metric used for the fitness in this paper will be
 described later.

### 2012-06-10 paper_aimsa/methods.tex
Edits toward a more polished paper

@@
 The simplest model with back-compatibility is geometric
-splines. For a given model f(x) with K knots, we can preserve the
+splines. \editbox{Too fast... add a sentence or two on what the methods are trying to do, and what back-compatibility is.} For a given model f(x) with K knots, we can preserve the
 exact shape of the generated curve while adding extra knots to the
@@
 parameterization from $K$ to $L$ knots ($L > K$), which is formulated as 
-Algorithm 1 in \cite{kormushev2011bipedal-walking-energy}.  Without loss of generality, the policy parameters are
+Algorithm 1 in \cite{kormushev2011bipedal-walking-energy} \edit{reproduce that algorithm here?}.  Without loss of generality, the policy parameters are
 normalized into $[0, 1]$, and appropriately scaled/shifted as necessary
@@
 our experiment. Previous work has verified that quadruped gaits
-perform better when they are coordinated \cite{clune2009evolving-coordinated-quadruped} \cite{clune2011on-the-performance-of-indirect-encoding}
-\cite{valsalam2008modular-neuroevolution-for-multilegged}. For each spline, we calculate its corresponding parameterized gait for one unit time cycle. Given that, then apply the same
+perform better when they are coordinated \cite{clune2009evolving-coordinated-quadruped, clune2011on-the-performance-of-indirect-encoding, valsalam2008modular-neuroevolution-for-multilegged}.
+For each spline, we calculate its corresponding parameterized gait for one unit time cycle.
+Given that, then apply the same
 pattern to every cycle throughout the 12 seconds of one
-trial. Specifically, each spline(a set of 3 knots) is interpreted to its corresponding servo positions as
-following:
+trial. Specifically, each spline (a set of 3 knots) is interpreted to its corresponding servo positions as shown in \eqnref{rl} and \tabref{params}.
 
-\begin{table}[b]
+\begin{table}[h]
 \begin{center}
@@
 \tablabel{parameters}
-\label{tab:params}
+\tablabel{params}
 \end{center}
@@
 
-\[
+\be
 \vec{g}(t) =
@@
 \end{array} } \right]
-\]
+\eqnlabel{rl}
+\ee
+
+\editbox{Add a new equation showing all servos (all nine) with their
+  actual values of the parameters $R$ and $C$ plugged in.}

### 2012-06-10 paper_aimsa/relatedWork.tex
Edits toward a more polished paper

--- 2012-05-30 paper_aimsa/relatedWork.tex
@@
 
-Other attempts to solve it
+Many attempts have been made on robot gaits learning using machine
+learning algorithms, producing competitive results \edit{Change these
+  all to citations} (Chernova and Veloso, 2005; Hornby et al., 2005;
+Zykov et al., 2004; Clune et al., 2009a, 2011, 2009b,c; T´ellez et
+al., 2006; Valsalam and Miikkulainen, 2008). In fact, Jason et
+al. tested and compared six different algorithms for quadruped robot
+gait learning in their studies. HyperNEAT, a generative encoding
+neural network algorithm, achieved the best performance among all six
+methods. In Kormushev et al. (2011)'s bipedal robot energy reduction
+research, a reinforcement learning method was used to optimize the
+performance. The results from Kormushev's research showed this
+algorithm's potential for walking problems. This reinforcement
+learning method still needs more tests for further explorations.

### 2012-06-10 paper_aimsa/results.tex
Edits toward a more polished paper

--- 2012-06-01 paper_aimsa/results.tex
@@
 \section{Results}
-
-RL\_POWER Gaits
-
-The results for the gaits evolved by RL PoWER are shown in Figure 8
-and Table 2. A total of 900 evaluations were performed for HyperNEAT
-(300 in each of three runs). Overall the RL PoWER gaits were the
-faster and more stable by far, beating the HyperNEAT when comparing
+The results for the gaits evolved by RL PoWER are shown in \figref{rlPerform} and \tabref{results}.The HyperNEAT performance is shown in \figref{hyperPerform}.
+A total of 900 evaluations were performed for RL PoWER (300 for each of three runs).
+The reason these two algorithms have different number of trials is that runs continued until the performance plateaued, which we defined as when there was no improvement during the last third of a run.
+Overall, the RL PoWER gaits were
+faster by far, beating the HyperNEAT when comparing
 either average or best gaits.
 
-(pic)RL POWER AND HyperNEAT performance chart
+\figp{rlPerform}{0.95}{Average fitness ($\pm$ SE) in cm/sec of the highest performing run of RL PoWER's three runs. }
+\figp{hyperPerform}{0.75}{Average fitness ($\pm$ SE) in cm/sec of the highest performing
+individual in one HyperNEAT run.}
+
+\begin{table}
+\begin{center}
+\begin{tabular}{|r|c|c|c||c|}
+\hline
+                                         & Average & Std. Dev \\
+\hline                                    
+\hline                                    
+RL PoWER                       & 7.62  &    2.1   \\
+\hline
+HyperNEAT                      & 6.28   &   1.26   \\
+\hline
+\end{tabular}
+\caption{The average and standard deviation of the best gaits found
+  for both algorithm during each of three runs, in cm/sec.}  \tablabel{results}
+\end{center}
+\end{table}
 
 We believe that this is because RL PoWER uses evolvable splines to represent gaits As explained earlier,  
-The single best gait found during this study had a speed of
+The single best gait found during this study had a speed of 11.09 centimeters/sec,  14.7\% better than the best HyperNEAT gait.
+\figref{rlPattern} shows a typical RL PoWER gait that had high fitness.
+Compared with \figref{neat_110115_211410_00000_002_filt_zoom}, the pattern of RL PoWER's motion is less
+complex but more regular than the HyperNEAT, but producing higher speed. As shown in \figref{rlPattern}
+, multiple motors are more coordinated than the HyperNEAT gait. 
+\figp{rlPattern}{0.75}{The motor positions of a typical well performing gait over one cycle.
+It matches exactly the shape of its corresponding spline representation.
+This pattern shows the simplicity and power of geometric splines representations.}
 
-11.0855923766 centimeters/sec, XX\%
+\figp{neat_110115_211410_00000_002_filt_zoom}{0.75}{A typical well performing HyperNEAT gait's pattern of motor positions}
 
-better than the best HyperNEAT gait. Figure 9 shows a typical
-RL\_POWER gait that had high fitness. The pattern of motion is less
-complex and more regular than the HyperNEAT, in that patterns of
-multiple motors are coordinated.
-
-A corresponding observation of this property is that the noisiness of
-RL\_POWER is significantly lower than that of HyperNEAT. One reason of
-this is that the when Expectation Maximization methodology is used,
-explorations happened in the vicinity space of the highest ranked
-patterns. Even some individual HyperNEAT gaits turned out very
-effective, the average noisiness of HyperNEAT is very large. For
-example, as found in [xx], the mean of HyperNEAT generation-champion’s
-performance was XX body lengths/minute (+-XX SD), but it had a max of
-38 and a min of 3. RL\_POWER, on the other hand, proved to be stable
-and much less noisy throughout the entire experiment. The maximum SD
-over all three runs is +- XX SD. This may also be due to a much larger
-space of possibilities of HyperNEAT than the more heuristically guided
-reinforcement learning.
-
-[pic][speed chart of two algos]
+A corresponding observation from the graph on this property is that the noisiness of
+RL PoWER is higher than HyperNEAT.
+One reason for this is that the RL PoWER intentionally adds noise when exploring the space near the best ranked policies.
+Another phenomenon seen from the \figref{rlPerform} is that one run of RL PoWER's general performance largely depends on initial gaits, which is a trait of hill climbing algorithms.
+The larger standard deviation in RL PoWER for space exploring can be easily fixed by tuning the noise parameter.
+RL PoWER converges to optimality at a higher rate.
+This is due to the more heuristically guided reinforcement learning of RL PoWER.
+HyperNEAT, on the other hand, has a larger dimension to explore due to its generative encoding method.
+While the evolvable spline representation used by RL PoWER has lower dimensions.
+The convergence of HyperNEAT is thus slower.

### 2012-06-10 paper_aimsa/abstract.tex
Abstract rewrite, many edits and suggestions for improvement.

--- 2012-06-05 paper_aimsa/abstract.tex
@@
-A reinforcement learning approach which can evolve the policy parameterization dynamically during
-the learning process for quadruped robot gait . By gradually increasing the representational
-power of the policy parameterization, it manages to find better
-policies faster than HyperNEAT, a state-of-the-art evolutionary ANN algorithm that achieved the best performance in previous studies.
+%A reinforcement learning approach which can evolve the policy
+%parameterization dynamically during the learning process for quadruped
+%robot gait \edit{???}. By gradually increasing the representational
+%power of the policy parameterization, it manages to find better
+%policies faster than HyperNEAT, a state-of-the-art evolutionary ANN
+%algorithm that achieved the best performance in previous studies.
+%
+%Expectation-Maximization-based reinforcement learning algorithm is
+%used. The test was done on a physical quadrupedal robot
+%(QuadraTot). The results show that the evolving policy
+%parameterization combined with expectation-maximization (RL PoWER)
+%outperforms substantially the HyperNEAT, by having faster convergence
+%and higher final reward. According to the statistics collected, the RL
+%PoWER achieves 16.3\% advantage over the HyperNEAT, the best method
+%tested on QuadraTot in previous studies.
 
-Expectation-Maximization-based reinforcement learning algorithm
-is used. The test was done on a physical quadrupedal robot(QuadraTot). The results show that the evolving policy parameterization combined with expectation-maximization (RL PoWER)
-outperforms substantially the HyperNEAT, by having
-faster convergence and higher final
-reward. According to the statistics collected, the RL PoWER achieves 16.3\%
-advantage over the HyperNEAT, the best method tested on QuadraTot in previous studies.
+Legged robots are uniquely privileged over their wheeled counterparts
+in their potential to access rugged terrain. However, designing
+walking gaits by hand for legged robots is a difficult and
+time-consuming process, so we seek algorithms for learning such gaits
+automatically using real world experimentation. Numerous previous
+studies have examined a variety algorithms for learning gaits, using
+an assortment of different robots. It is often difficult to compare
+the algorithmic results from one study to the next, because the
+conditions and robots used vary. With this in mind, we have used an
+open-source, 3D printed quadruped robot called QuadraTot, so the results may be
+verified, and hopefully improved upon, by any group so
+desiring. Because many robots do not have accurate simulators, we test
+gait-learning algorithms entirely on the physical robot. Previous
+studies using the QuadraTot have compared parameterized splines, the
+HyperNEAT generative encoding, and (Kyrre's algorithm). Here we
+compare these results to an algorithm called Policy learning by
+Weighting Exploration with the Returns, or RL PoWER. We report that
+this algorithm has learned the fastest gait yet reported in the
+literature, 16.3\% faster \edit{different number than in results?} than reported for HyperNEAT and \edit{???
+compared to Kyrre's?}. In addition, the learned gaits are less taxing
+on the robot and more repeatable than previous record-breaking gaits.

### 2012-06-10 paper_aimsa/conclusion.tex
Abstract rewrite, many edits and suggestions for improvement.

@@
 
-We have presented a new reinforcement-learning-based algorithm for
-optimizing a quadrupedal gait for linear speed. We implemented and
+We have presented results from a recently introduced reinforcement-learning-based algorithm for
+optimizing a quadrupedal gait for linear speed, tested on a physical robot. \edit{Next sentence mistakenly copied?} We implemented and
 tested six learning strategies for parameterized gaits and compared
@@
 
-Though over 900 trials have been made to investigate the applicability
+Over 900 trials have been made to investigate the applicability
 of RL PoWER to quadruped robots. It is difficult to gather the enough
@@
 direction for future work could be to obtain many more trials. But due
-to the physical limitations, obtaining One solution to this is
+to the physical limitations, obtaining one solution to this is
 simulation. Because of the low cost of simulation, it would produce
@@
 effective, and the hardware trials would serve to continuously ground
-and refine the simulator. \cite{glette2012evolution-of-locomotion-in-a-simulated}
+and refine the simulator. \cite{glette2012evolution-of-locomotion-in-a-simulated} \edit{what is this citation for?}
 
-One guess led by this study is that for feedback oriented tasks,
-reinforcement learning methods are more fit in natural. Despite the
-complexities of HyperNEAT, a simpler algorithm on the code level
+One view supported by this study is that for feedback oriented tasks,
+reinforcement learning methods are more fit and natural. Despite the
+complexities of HyperNEAT, a simpler algorithm on the code level \edit{???}
 delivered better performance in general.  Also, evolvable spline
@@
 the same time. Evolvable splines can serve as a general representation for various other
-learning problems.
+learning problems.\edit{probably rewrite this last paragraph}

### 2012-06-10 paper_aimsa/experimentalSetup.tex
Abstract rewrite, many edits and suggestions for improvement.

@@
 
-The quadruped robot used was assembled from parts purchased online and
+The quadruped robot used, QuadraTot, was assembled from parts purchased online and
 parts printed by the Objet Connex 500 3-D Printing System. The robot
@@
 AX-12+. Each servo could be set to a position in the range [0, 1023]
-by using pydynamixel library, corresponding to a physical range [-120$^\circ$,
+by using pydynamixel library, corresponding roughly to a physical range [-120$^\circ$,
   +120$^\circ$]. Also, to prevent collisions with the robot body, the control
@@
 general. These gaits tend to either overburden the servos or flip the
-robot. Thus, RL PoWER were tested after carefully cropping out the
+robot. \edit{Explain this next part on smart cropping better.} Thus, RL PoWER were tested after carefully cropping out the
 extremely fierce gaits each time. As described earlier, each leg has
@@
 
-\figp{smartCrop}{1}{The red line is the cropping border, $g = j1 +  j2$. As shown in \figref{sixLegPositions}}
+\figp{smartCrop}{1}{The red line is the cropping border, $g = j1 +  j2$. As shown in \figref{sixLegPositions} \edit{???}}
 
 \figp{sixLegPositions}{1}{The left two positions are allowed; middle
-  two positions are on the red border line from \figref{smartCrop}; right two positions are deemed as ``extremely fierce'', thus are disallowed.}
+  two positions are on the red border line from \figref{smartCrop};
+  right two positions are deemed as ``extremely fierce'', thus are
+  disallowed.   \\\editbox{Smart crop needs to be explained somewhere, probably briefly in this caption and more extensively in the text.}}

### 2012-06-10 paper_aimsa/introduction.tex
Abstract rewrite, many edits and suggestions for improvement.

@@
 
-\figp{quadratotWhiteBooties}{.75}{The quadratot robot used in this research. The white booties are newly added to prevent sliding on the surface thus minimize measure error.}
+\figp{quadratotWhiteBooties}{.75}{The open source, 3D printed
+  QuadraTot robot used in this research. The white printed booties are
+  new addition to prevent sliding on surfaces and to minimize measurement
+  error.}
 
 Various learning algorithms have been proved to be effective for
-legged robots. Algorithms such as HyperNEAT \cite{yosinski2011evolving-robot-gaits},
-GA \cite{chernova2004an-evolutionary-approach-to-gait} and others \cite{hornby2005autonomous-evolution-of-dynamic} \cite{zykov2004evolving-dynamic-gaits} \cite{tellez2006evolving-the-walking-behaviour} \cite{valsalam2008modular-neuroevolution-for-multilegged} have been tested to be effective for automatic learning gaits for robots.
-Despite competetive performance, a major
-task that are usually hidden in the published results lies in tuning the parameters for
-these evolutionary algorithms \cite{kormushev2011bipedal-walking-energy}. Here we present a
-different way for learning gaits using reinforcement learning and expectation maximization, called RL PoWER. In our
-experiment, the main focus of the research is on the applicability of RL PoWER to quadruped robot gait learning.
-Another motivation is to compare the state-of-the-art neural network algorithm, HyperNEAT, with our proposed method in quadruped robot gait learning.
+legged robots. Algorithms such as HyperNEAT
+\citep{yosinski2011evolving-robot-gaits}, Genetic Algorithms
+\citep{chernova2004an-evolutionary-approach-to-gait} and others
+\citep{hornby2005autonomous-evolution-of-dynamic,
+  zykov2004evolving-dynamic-gaits,
+  tellez2006evolving-the-walking-behaviour,
+  valsalam2008modular-neuroevolution-for-multilegged} have been tested
+to be effective for automatic learning gaits for robots. Despite
+competitive performance, a major task is usually hidden in the
+published results: tuning the parameters for these evolutionary
+algorithms \citep{kormushev2011bipedal-walking-energy}. Here we
+present results using a different way of learning gaits: a
+Reinforcement Learning algorithm called Policy learning by Weighting
+Exploration with the Returns, or RL PoWER, proposed by
+\cite{kober2009learning-motor-primitives}. In our experiment, the main
+focus of the research is on the applicability of RL PoWER to quadruped
+robot gait learning. Another motivation is to compare the
+state-of-the-art neural network algorithm, HyperNEAT, with our
+proposed method in quadruped robot gait learning.
 
@@
 
-As in \cite{yosinski2011evolving-robot-gaits}, the gait learning problem is defined to find a
+As in \cite{yosinski2011evolving-robot-gaits}, we define the gait learning problem to be the search for a
 gait that maximizes some specific metric. Mathematically, we define a

### 2012-06-10 paper_aimsa/main.tex
Abstract rewrite, many edits and suggestions for improvement.

@@
 
-\title{RL-PoWER, Evolving Policy Parameterization with Reinforcement Learning and its Application to Quadrupedal Walking Optimization}
+%\title{RL-PoWER, Evolving Policy Parameterization with Reinforcement Learning and its Application to Quadrupedal Walking Optimization}
+\title{Learning Fast Quadruped Robot Gaits with the RL PoWER Spline Parameterization}
 \author{Haocheng Shen, Jason Yosinski, Petar Kormushev, \and Hod Lipson  \\
@@
 Cornell University, 239 Upson Hall, Ithaca, NY 14853 \\
-\texttt{hs454@cornell.edu, yosinski@cs.cornell.edu}}
+\texttt{hs454@cornell.edu,yosinski@cs.cornell.edu}}
 

### 2012-06-10 paper_aimsa/methods.tex
Abstract rewrite, many edits and suggestions for improvement.

@@
 representation dynamically while the trial is running. In
-\cite{kormushev2011bipedal-walking-energy}'s studies on reducing energy
-consumption for bipedal robots, a mechanism that can
+earlier studies on reducing energy
+consumption for bipedal robots \citep{kormushev2011bipedal-walking-energy}, a mechanism that can
 evolve the policy parameterization was used. The method starts from a
@@
 called Policy learning by Weighting Exploration with the
-Returns(PoWER), which is based on Expectation-Maximization algorithm
+Returns (RL PoWER), which is based on Expectation-Maximization algorithm
 (EM). The proposed technique for evolving the policy parameterization
-is a combination with this EM-based RL algorithm, named PoWER \cite{kober2009learning-motor-primitives}. The reason for using this is its relatively fewer parameters that need tuning. We
+is a combination with this EM-based RL algorithm, named PoWER \cite{kober2009learning-motor-primitives}.
+\edit{Check the prev sentence...
+I don't get it} The reason for using this is its relatively fewer parameters that need tuning.
+We
 evolved the policy parameterization only on those past trials ranked
@@
 future. Besides, evolving all the parameterizations increases the exploring space. Since our experiment
-is done on a physical robot, explore all the variations of every
+is done on a physical robot, exploring all the variations of every
 parameterization is not practical. Future work may incorporate simulations into the studies, as illustrated in \cite{bongard2006resilient-machines-through}.
 
-For the experiment, we set 3 knots for each servo and there are 8
+For the experiment, we set the splines to have 3 knots for each servo, and there are 8
 servos in total. The servo in the hip is not used in
 our experiment. Previous work has verified that quadruped gaits
-perform better when they are coordinated \cite{clune2009evolving-coordinated-quadruped, clune2011on-the-performance-of-indirect-encoding, valsalam2008modular-neuroevolution-for-multilegged}.
-For each spline, we calculate its corresponding parameterized gait for one unit time cycle.
-Given that, then apply the same
+perform better when they are coordinated \citep{clune2009evolving-coordinated-quadruped, clune2011on-the-performance-of-indirect-encoding, valsalam2008modular-neuroevolution-for-multilegged}.
+For each spline, we calculate its corresponding parameterized gait for one unit time cycle \edit{how long is this?} and then apply the same
 pattern to every cycle throughout the 12 seconds of one

### 2012-06-10 paper_aimsa/results.tex
Abstract rewrite, many edits and suggestions for improvement.

@@
 \hline                                    
-RL PoWER                       & 7.62  &    2.1   \\
+RL PoWER                       & {\bf 7.62}  &    2.1   \\
 \hline
@@
 \caption{The average and standard deviation of the best gaits found
-  for both algorithm during each of three runs, in cm/sec.}  \tablabel{results}
+  for both algorithm during each of three runs, in cm/sec. \edit{Add the best gait each algorithm found as well.}}
+\tablabel{results}
 \end{center}
@@
 We believe that this is because RL PoWER uses evolvable splines to represent gaits As explained earlier,  
-The single best gait found during this study had a speed of 11.09 centimeters/sec,  14.7\% better than the best HyperNEAT gait.
+The single best gait found during this study had a speed of 11.09 centimeters/sec,  14.7\% better \edit{different number than in abstract?} than the best HyperNEAT gait.
 \figref{rlPattern} shows a typical RL PoWER gait that had high fitness.
@@
 , multiple motors are more coordinated than the HyperNEAT gait.
-\figp{rlPattern}{0.75}{The motor positions of a typical well performing gait over one cycle.
-It matches exactly the shape of its corresponding spline representation.
-This pattern shows the simplicity and power of geometric splines representations.}
 
-\figp{neat_110115_211410_00000_002_filt_zoom}{0.75}{A typical well performing HyperNEAT gait's pattern of motor positions}
+\figp{rlPattern}{0.75}{The motor positions of a typical well
+  performing gait over one cycle. It matches exactly the shape of its
+  corresponding spline representation \edit{what does this sentence mean?}. This pattern shows the
+  simplicity and power of geometric splines representations \edit{how does it show this?}.}
+
+\figp{neat_110115_211410_00000_002_filt_zoom}{0.75}{A typical well
+  performing HyperNEAT gait's pattern of motor positions, reproduced
+  from \cite{yosinski2011evolving-robot-gaits}.}
 

### 2012-06-15 paper_aimsa/abstract.tex
submission to Petar

@@
 studies using the QuadraTot have compared parameterized splines, the
-HyperNEAT generative encoding, and (Kyrre's algorithm). Here we
+HyperNEAT generative encoding and genetic algorithm.
+Among these, the research on the genetic algorithm was conducted by \citep{glette2012evolution-of-locomotion-in-a-simulated} in a simulator and tested on a real robot.
+Here we
 compare these results to an algorithm called Policy learning by
 Weighting Exploration with the Returns, or RL PoWER. We report that
-this algorithm has learned the fastest gait yet reported in the
-literature, 16.3\% faster \edit{different number than in results?} than reported for HyperNEAT and \edit{???
-compared to Kyrre's?}. In addition, the learned gaits are less taxing
+this algorithm has learned the fastest gait through phsysical experiments yet reported in the
+literature, 16.3\% faster than reported for HyperNEAT. In addition, the learned gaits are less taxing
 on the robot and more repeatable than previous record-breaking gaits.

### 2012-06-15 paper_aimsa/conclusion.tex
submission to Petar

@@
 We have presented results from a recently introduced reinforcement-learning-based algorithm for
-optimizing a quadrupedal gait for linear speed, tested on a physical robot. \edit{Next sentence mistakenly copied?} We implemented and
-tested six learning strategies for parameterized gaits and compared
-them to gaits produced by neural networks evolved with the HyperNEAT
+optimizing a quadrupedal gait for linear speed, tested on a physical robot.  We implemented the algorithm, Policy learning by Weighting
+Exploration with the Returns (a.k.a RL PoWER), for parameterized gaits and compared
+it to gaits produced by neural networks evolved with the HyperNEAT
 generative encoding. Though both methods resulted in an improvement
@@
 effective, and the hardware trials would serve to continuously ground
-and refine the simulator. \cite{glette2012evolution-of-locomotion-in-a-simulated} \edit{what is this citation for?}
+and refine the simulator.
 
@@
 reinforcement learning methods are more fit and natural. Despite the
-complexities of HyperNEAT, a simpler algorithm on the code level \edit{???}
-delivered better performance in general.  Also, evolvable spline
+complexities of HyperNEAT, a structurally simpler algorithm delivered better performance in general.  Also, evolvable spline
 interpolation is shown to be simple and representationally powerful at
 the same time. Evolvable splines can serve as a general representation for various other
-learning problems.\edit{probably rewrite this last paragraph}
+learning problems.

### 2012-06-15 paper_aimsa/experimentalSetup.tex
submission to Petar

@@
 joint with AX-12+servo in each of the four legs, and one AX-18+ servo
-at the center. To avoid the formerly reported \edit{???} AX-18+ servos are used
+at the center. To avoid the formerly reported problem with  AX-18+ servos are used
 in this robot because of their stronger actuation power than that of
@@
 general. These gaits tend to either overburden the servos or flip the
-robot. \edit{Explain this next part on smart cropping better.} Thus, RL PoWER were tested after carefully cropping out the
+robot. Thus, RL PoWER were tested after carefully cropping out the
 extremely fierce gaits each time. As described earlier, each leg has
-two joints, inner joint j1 and outer j2. The position of one leg is
-determined by the sum of the values of j1 and j2. Extremely fierce
+two joints, inner joint $j1$ and outer $j2$. The position of one leg is
+determined by the sum of the values of $j1$ and $j2$, in the range of [0, 1023]. Extremely fierce
 gaits usually have relatively small values of $g$. As shown in \figref{smartCrop}, this cropping method works
-by mapping the wild gaits onto the boundary of a quasi-triangular area in
-the two-dimensional space of j1 and j2.
+by mapping the wild gaits right onto the boundary of a quasi-triangular area in
+the two-dimensional space of $j1$ and $j2$.
+In our experimentation, we set the threshhold to 730 to balance between gait performance and safety.
+However, an obvious trade-off to this method is the reduced exploratory region of the parameters.
+A extension would be experiments using smaller $g$, cropping less gaits, to possibly enhance the performance.
 
-\figp{smartCrop}{1}{The red line is the cropping border, $g = j1 +  j2$. As shown in \figref{sixLegPositions} \edit{???}}
+\figp{smartCrop}{0.7}{The red line is the smart cropping border, $g = j1 +  j2$.
+The dashed grey lines are the physical limits of the joints.
+As shown in \figref{sixLegPositions}, the fierce movements are prevented by mapping these out-of-range positions to the border line as shown in the graph.
+The value of $g$ cannot drop below 730 otherwise $j1$ and $j2$ are projected right to the red line.
+}
 
@@
   two positions are on the red border line from \figref{smartCrop};
-  right two positions are deemed as ``extremely fierce'', thus are
-  disallowed.   \\\editbox{Smart crop needs to be explained somewhere, probably briefly in this caption and more extensively in the text.}}
+  right two positions are deemed as ''extremely fierce'', thus are
+  disallowed. Future work may involve studying a larger range of $g$ for increased performance.}

### 2012-06-15 paper_aimsa/main.tex
submission to Petar

@@
 \title{Learning Fast Quadruped Robot Gaits with the RL PoWER Spline Parameterization}
-\author{Haocheng Shen, Jason Yosinski, Petar Kormushev, \and Hod Lipson  \\
+\author{Haocheng Shen, Jason Yosinski, Petar Kormushev, Darwin G. Caldwell \and Hod Lipson  \\
   \mbox{}\\
-Cornell University, 239 Upson Hall, Ithaca, NY 14853 \\
-\texttt{hs454@cornell.edu,yosinski@cs.cornell.edu}}
+Cornell University, 239 Upson Hall, Ithaca, NY 14853, USA\\
+Department of Advanced Robotics, Istituto Italiano di Tecnologia, via Morego 30, 16163 Genova, Italy \\
+\texttt{hs454@cornell.edu, Petar.Kormushev@iit.it}}
 

### 2012-06-15 paper_aimsa/methods.tex
submission to Petar

@@
 The simplest model with back-compatibility is geometric
-splines. \editbox{Too fast... add a sentence or two on what the methods are trying to do, and what back-compatibility is.} For a given model f(x) with K knots, we can preserve the
+splines. For example, for a given model f(x) with K knots, we can preserve the
 exact shape of the generated curve while adding extra knots to the
-original spline. Say, if we put one additional knot between every two
+original spline. If we put one additional knot between every two
 consecutive knots of the original spline, we end up with a 2K - 1
@@
 order to do this, we need to define an algorithm for evolving the
-parameterization from $K$ to $L$ knots ($L > K$), which is formulated as 
-Algorithm 1 in \cite{kormushev2011bipedal-walking-energy} \edit{reproduce that algorithm here?}.  Without loss of generality, the policy parameters are
-normalized into $[0, 1]$, and appropriately scaled/shifted as necessary
-later upon use.
+parameterization from $K$ to $L$ knots ($L > K$), which is formulated as in
+\ref{alg_Evolve_Spline}.  Without loss of generality, the policy parameters are
+normalized into $[0, 1]$, and appropriately scaled and shifted as necessary upon use.
+
+\begin{algorithm}
+\caption{EvolvePolicy-Spline ($P_{current}$: current policy, $L$: desired new number of parameters)}
+\label{alg_Evolve_Spline}
+\begin{algorithmic}[1]
+
+\STATE $K$ $\leftarrow$ $P_{current}.numberOfParameters$
+
+\STATE $X_{current}$ $\leftarrow$ $[0, \frac{1}{K-1}, \frac{2}{K-1}, ... , 1]$ % Knots
+\STATE $Y_{current}$ $\leftarrow$ $P_{current}.parameterValues$ % Values
+\STATE $S_{current}$ $\leftarrow$ ConstructSpline($X_{current}$, $Y_{current}$) % Spline
+%\COMMENT{ Optional: make the spline cyclic}
+
+\STATE $X_{new}$ $\leftarrow$ $[0, \frac{1}{L-1}, \frac{2}{L-1}, ... , 1]$
+\STATE $Y_{new}$ $\leftarrow$ EvaluateSplineAtKnots($S_{current}$, $X_{new}$)
+\STATE $S_{new}$ $\leftarrow$ ConstructSpline($X_{new}$, $Y_{new}$)
+%\COMMENT{ Optional: make the spline cyclic}
+
+\STATE $P_{new}.numberOfParameters$ $\leftarrow$ $L$
+\STATE $P_{new}.parameterValues$ $\leftarrow$ $S_{new}$.$Y_{new}$
+
+\STATE return $P_{new}$
+
+\end{algorithmic}
+\end{algorithm}
 
@@
 \cite{kober2009learning-motor-primitives} proposed a RL algorithm
-called Policy learning by Weighting Exploration with the
+named Policy learning by Weighting Exploration with the
 Returns (RL PoWER), which is based on Expectation-Maximization algorithm
-(EM). The proposed technique for evolving the policy parameterization
-is a combination with this EM-based RL algorithm, named PoWER \cite{kober2009learning-motor-primitives}.
-\edit{Check the prev sentence...
-I don't get it} The reason for using this is its relatively fewer parameters that need tuning.
-We
+(EM). The reason for using this is its relatively fewer parameters that need tuning. We
 evolved the policy parameterization only on those past trials ranked
@@
 perform better when they are coordinated \citep{clune2009evolving-coordinated-quadruped, clune2011on-the-performance-of-indirect-encoding, valsalam2008modular-neuroevolution-for-multilegged}.
-For each spline, we calculate its corresponding parameterized gait for one unit time cycle \edit{how long is this?} and then apply the same
+For each spline, we calculate its corresponding parameterized gait for one unit time cycle of 1.8 seconds and then apply the same
 pattern to every cycle throughout the 12 seconds of one
-trial. Specifically, each spline (a set of 3 knots) is interpreted to its corresponding servo positions as shown in \eqnref{rl} and \tabref{params}.
+trial. Specifically, each spline (a set of 3 knots) is interpreted to its corresponding servo positions as in the following equation and in \tabref{params}.
 
@@
 
-\be
+\[
 \vec{g}(t) =
@@
 \end{array} } \right]
-\eqnlabel{rl}
-\ee
-
-\editbox{Add a new equation showing all servos (all nine) with their
-  actual values of the parameters $R$ and $C$ plugged in.}
+\]

### 2012-06-15 paper_aimsa/relatedWork.tex
submission to Petar

@@
 Many attempts have been made on robot gaits learning using machine
-learning algorithms, producing competitive results \edit{Change these
-  all to citations} (Chernova and Veloso, 2005; Hornby et al., 2005;
+learning algorithms, producing competitive results (Chernova and Veloso, 2005; Hornby et al., 2005;
 Zykov et al., 2004; Clune et al., 2009a, 2011, 2009b,c; T´ellez et

### 2012-06-15 paper_aimsa/results.tex
submission to Petar

@@
 \hline
-                                         & Average & Std. Dev \\
+                                & Average & Best Single & Std. Dev \\
 \hline                                    
 \hline                                    
-RL PoWER                       & {\bf 7.62}  &    2.1   \\
+RL PoWER                       & {\bf 7.62}  & 11.05  &    2.1   \\
 \hline
-HyperNEAT                      & 6.28   &   1.26   \\
+HyperNEAT                      & 6.28   & 9.7 & 1.26   \\
 \hline
@@
 \caption{The average and standard deviation of the best gaits found
-  for both algorithm during each of three runs, in cm/sec. \edit{Add the best gait each algorithm found as well.}}
+  for both algorithm during each of three runs, in cm/sec.}
 \tablabel{results}
@@
 We believe that this is because RL PoWER uses evolvable splines to represent gaits As explained earlier,  
-The single best gait found during this study had a speed of 11.09 centimeters/sec,  14.7\% better \edit{different number than in abstract?} than the best HyperNEAT gait.
+The single best gait found during this study had a speed of 11.09 cm/sec,  16.3\% better than the best HyperNEAT gait.
 \figref{rlPattern} shows a typical RL PoWER gait that had high fitness.
@@
 \figp{rlPattern}{0.75}{The motor positions of a typical well
-  performing gait over one cycle. It matches exactly the shape of its
-  corresponding spline representation \edit{what does this sentence mean?}. This pattern shows the
-  simplicity and power of geometric splines representations \edit{how does it show this?}.}
+performing gait over one cycle.
+According to the spline transformation, the curves of joint positions are merely linear transformations of the spline curves.
+Compared with the pattern of a well-performing HyperNEAT gait, the spline representation of the parameterized gaits are simple yet powerful.}
 

### 2012-06-18 paper_aimsa/experimentalSetup.tex
Next round of edits.

@@
 
+\editbox{Much better, but I would add an intuitive explanation of what the
+  smart cropping accomplishes. Old method: crop each servo
+  individually (square). New: individual + triangular chop to prevent
+  legs going under robot.}
+
 \figp{smartCrop}{0.7}{The red line is the smart cropping border, $g = j1 +  j2$.
@@
 As shown in \figref{sixLegPositions}, the fierce movements are prevented by mapping these out-of-range positions to the border line as shown in the graph.
-The value of $g$ cannot drop below 730 otherwise $j1$ and $j2$ are projected right to the red line.
+The value of $g$ cannot drop below 730 otherwise $j1$ and $j2$ are projected perpendicularly to the red line.
 }
@@
   two positions are on the red border line from \figref{smartCrop};
-  right two positions are deemed as ''extremely fierce'', thus are
+  right two positions are deemed as ``extremely fierce'', thus are
   disallowed. Future work may involve studying a larger range of $g$ for increased performance.}

### 2012-06-18 paper_aimsa/relatedWork.tex
Next round of edits.

--- 2012-06-10 paper_aimsa/relatedWork.tex
@@
 neural network algorithm, achieved the best performance among all six
-methods. In Kormushev et al. (2011)'s bipedal robot energy reduction
-research, a reinforcement learning method was used to optimize the
-performance. The results from Kormushev's research showed this
+methods. In previous bipedal robot energy reduction
+research \citep{kormushev2011bipedal-walking-energy}, a reinforcement learning method was used to optimize the
+performance. The results from this research showed the
 algorithm's potential for walking problems. This reinforcement

### 2012-06-18 paper_aimsa/results.tex
Next round of edits.

@@
 \section{Results}
+
 The results for the gaits evolved by RL PoWER are shown in \figref{rlPerform} and \tabref{results}.The HyperNEAT performance is shown in \figref{hyperPerform}.
@@
 \begin{center}
-\begin{tabular}{|r|c|c|c||c|}
+\begin{tabular}{|r|c|c|c|}
 \hline
-                                & Average & Best Single & Std. Dev \\
+                                & Best Single & Average  & Std. Dev \\
 \hline                                    
 \hline                                    
-RL PoWER                       & {\bf 7.62}  & 11.05  &    2.1   \\
+RL PoWER                       & {\bf 11.05} & {\bf 7.62}   &    2.1   \\
 \hline
-HyperNEAT                      & 6.28   & 9.7 & 1.26   \\
+HyperNEAT                       & 9.7 & 6.28   & 1.26   \\
 \hline
@@
 
+%\edit{Add Kyrre's results to table (one line for tested on his robot (better) and one for tested on our robot (not as good).}
+
 We believe that this is because RL PoWER uses evolvable splines to represent gaits As explained earlier,  
-The single best gait found during this study had a speed of 11.09 cm/sec,  16.3\% better than the best HyperNEAT gait.
+The single best gait found during this study had a speed of 11.09 cm/sec \edit{11.05 or 11.09?
+See table 2},  16.3\% better than the best HyperNEAT gait.
 \figref{rlPattern} shows a typical RL PoWER gait that had high fitness.

### 2012-06-25 code/Load_trajectory.m
changes in local

@@ -6,34 +6,34 @@ function traj = Load_trajectory(filename) % E.g. 'data/001/output_001.txt'
   data = dlmread(filename, '', 14, 0); % skip the first 13 lines
   traj = [data(:,1) data(:,20) data(:,22) data(:,21)]; % t, x, y, z
 
-  plot3(traj(:,2), traj(:,3), traj(:,4), 'linewidth', 2);
+  %plot3(traj(:,2), traj(:,3), traj(:,4), 'linewidth', 2);
   % plot traj starting pos
-  plot3(traj(1,2), traj(1,3), traj(1,4),'.','markersize',20,'color',[0 1 0]);
+  %plot3(traj(1,2), traj(1,3), traj(1,4),'.','markersize',20,'color',[0 1 0]);
   % plot traj end pos
-  plot3(traj(end,2), traj(end,3), traj(end,4),'x','markersize',20,'color',[1 0 0]);
+  %plot3(traj(end,2), traj(end,3), traj(end,4),'x','markersize',20,'color',[1 0 0]);
 
-  view(3); axis equal;
-  mn = min(min(traj(:,2:4)));
-  mx = max(max(traj(:,2:4)));
-  range = mx - mn;
-  mn = mn - 0.1*range;
-  mx = mx + 0.1*range;
-  axis([mn mx mn mx mn mx]);
-  xlabel('x'); ylabel('y'); zlabel('z');
-
-  hFig = figure('Name', 'Individual axis trajectories', 'position', [1260, 200, 600, 900]); hold on; box on; grid on;
-  subplot(3,1,1);
-  disp('111');
-  plot(traj(:,1), traj(:,2), 'linewidth', 2);
-  disp('222');
-  xlabel('time'); ylabel('x');
-  ylim([mn mx]);
-  subplot(3,1,2);
-  plot(traj(:,1), traj(:,3), 'linewidth', 2);
-  xlabel('time'); ylabel('y');
-  ylim([mn mx]);
-  subplot(3,1,3);
-  plot(traj(:,1), traj(:,4), 'linewidth', 2);
-  xlabel('time'); ylabel('z');
-  ylim([mn mx]);
+  %view(3); axis equal;
+  %mn = min(min(traj(:,2:4)));
+  %mx = max(max(traj(:,2:4)));
+  %range = mx - mn;
+  %mn = mn - 0.1*range;
+  %mx = mx + 0.1*range;
[…]

### 2012-06-25 code/PoWER_QuadraTot_23.m

@@ -3,8 +3,8 @@ function PoWER_QuadraTot_23()
 % Author: Petar Kormushev
 % April 2012
 %
-close all;
-clear all;
+%close all;
+%clear all;
 
 %% Parameters
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2012-06-25 code/RL_PoWER.m

@@ -6,7 +6,7 @@ disp('Starting RL...');
 
 %hFig = figure('Name', 'Rollouts', 'position', [50, 100, 800, 600]);
 %hFigResults = figure('Name', 'Results', 'position',[1000,600,800,400]); axis on; grid on; hold on;
-hFigExported = figure('Name', 'Exported trajectory', 'position', [850, 100, 600, 400]); hold on; box on; grid on;
+%hFigExported = figure('Name', 'Exported trajectory', 'position', [850, 100, 600, 400]); hold on; box on; grid on;
 
 % plotRollouts(hFig, 0, 0, 0, 0);
 
@@ -59,7 +59,7 @@ for si=1:n_splines
   policy(1).s(si).y = initialPos(si)*ones(1,policy(1).s(si).n); % set initial policy y values to the middle 0.5
   % now turn it into a cyclic spline
   [policy(1).s(si).pp, policy(1).s(si).x, policy(1).s(si).y, policy(1).s(si).n] = getCyclicSplinePlus6e(policy(1).s(si).x, policy(1).s(si).y);
-  % plotSpline(policy(1).s(si).pp, policy(1).s(si).n, policy(1).s(si).x, policy(1).s(si).y, 'green', 3); % visualize the spline
+  %plotSpline(policy(1).s(si).pp, policy(1).s(si).n, policy(1).s(si).x, policy(1).s(si).y, 'green', 3); % visualize the spline
 end
 %pause
 
@@ -76,7 +76,9 @@ disp('Running PoWER algorithm...');
 if relearn==1 % resume learning of a previous session
     % load iteration number
     iter_matrix = dlmread([getDataPath() 'iter.txt']);
+    disp(['n_iter: ', n_iter])
     iter = iter_matrix(1,1);
+    disp(['iter: ', iter])
     load([getDataPath() num2str(iter,'%03d') '/workspace.mat']); % loads ALL workspace variables!!
     % iter = iter + 1; ?? no, let it repeat it, if necessary I can just reuse the previous reward
     % here make changes to parameters, if necessary, e.g. variability, etc.
@@ -93,8 +95,8 @@ while (iter <= n_iter)
         disp(['Iteration: ', num2str(iter)]);
     %end
 
-%     disp('param:');
-%     param(:,iter)'
+     disp('param:');
+     param(:,iter)'
 
     if fromKnots < toKnots
         % regularly increase the number of knots
@@ -152,7 +154,7 @@ while (iter <= n_iter)
     end
[…]

### 2012-06-25 code/exportTrajectoryWithoutRescaling.m

@@ -13,17 +13,25 @@ function exportTrajectoryWithoutRescaling(policy, hFig, filepath);
 
   cycleTime = (0: dt: cycle_duration)';
   totalTime = (0: dt: total_duration)';
-
+  
+  
   % stretch the spline over the duration of one cycle
   for si=1:policy.n_splines
     s(si).XX = policy.s(si).x(1+3) : (policy.s(si).x(end-3)-policy.s(si).x(1+3))/(nbCycleSamples-1) : policy.s(si).x(end-3);
     s(si).YY = ppval(policy.s(si).pp, s(si).XX);
+    keyboard;
   end
+
+  disp('PASSED THIS POINT')
+  disp('PASSED THIS POINT')
+  disp('PASSED THIS POINT')
+  disp('PASSED THIS POINT')
   
   % rescale the trajectory y from [0,1] to stretch inside [zmin,zmax]
   posMin = 256;
   posMax = posMin + 512;
   posRange = posMax - posMin;
+
   for si=1:policy.n_splines
     % rescale
     s(si).cyclePos = s(si).YY .* posRange + posMin;
@@ -32,14 +40,14 @@ function exportTrajectoryWithoutRescaling(policy, hFig, filepath);
     s(si).totalPos = [repmat(s(si).cyclePos(1:end-1), ceil(nbCycles), 1) ; s(si).cyclePos(end)]; % the last pos in a cycle is the same as the first pos in the next cycle
     s(si).totalPos = s(si).totalPos(1:nbTotalSamples); % the last cycle will probably be interrupted before its end
   end
-  
+   
   % plot the 'Exported trajectory' velocity and acceleration
   figure(hFig); clf; hold on; box on; grid on;
   clrmap = colormap(jet(policy.n_splines));
-  %for si=1:policy.n_splines
-    %plot(totalTime, s(si).totalPos, 'color', clrmap(si,:), 'linewidth', 2); % all cycles
-    %plot(cycleTime, s(si).cyclePos, 'color', clrmap(si,:), 'linewidth', 4); % just the first cycle is thicker, to highlight it
-  %end
+ % for si=1:policy.n_splines
+  %  plot(totalTime, s(si).totalPos, 'color', clrmap(si,:), 'linewidth', 2); % all cycles
+   % plot(cycleTime, s(si).cyclePos, 'color', clrmap(si,:), 'linewidth', 4); % just the first cycle is thicker, to highlight it
+ % end
   
   % Generate file input.txt for the simulator
   inputData = [];

### 2012-06-25 code/extract.py

@@ -1,39 +1,49 @@
-#! /usr/bin/env python
+#! /usr/bin/env python-32
 
 from os import listdir
+import matplotlib as mpl 
+mpl.use('cocoaagg') 
 import matplotlib.pyplot as plt
 import re
 
 def main():
-    path = '/home/team/results/20120523_snd250Run/'
-    trials = sorted(listdir (path))[4:]
-    speeds = []
-    pattern = re.compile("[0-9]{3}")
-    for trial in trials:
-        if pattern.match(trial):
-            output = open(path + trial + '/output.txt' )
-            print path, trial
-            print ("processing " + trial)
-            lines = output.readlines()
-            length =len(lines)
-            ss = []
-            for ii, line in enumerate(lines):
-                if ii == 1 or ii == length-1 :
-                    t = line.split()
-                    xy = [float(t[19]),float(t[20])]
-                    ss.append(xy)
-            speed = ((( (ss[1][0]-ss[0][0])/5.6)**2+( (ss[1][1]-ss[0][1])/5.6 )**2)**0.5)/12
-            print speed
-            if len(speeds) > 0 and speeds[len(speeds)-1]-speed > speed*5:
-                speeds.append(speeds[len(speeds)-1]*0.85)
-            else:
-                speeds.append(speed)
-            print("processing complete: "+trial)
-
-    print len(speeds)
-    plt.plot(range(1,len(speeds)+1), speeds)
-    plt.ylabel('SPEED')
+    path = '/Users/hs454/Desktop/tmp/'
+    runs = sorted(listdir (path))
+    total = []
+    avg = []
+    avg_all=[]
+    pattern = re.compile("return_[0-9]*")
+    bucket = []
+    for run in runs:
+        if pattern.match(run):
+            speeds=[]
+            f = open(run)
+            lines = f.readlines()
+            for line in lines:
+                spd_tr= line.split()
+                ts= (int(spd_tr[1]),float(spd_tr[0])*(14.0/12))
+                speeds.append(ts)
+            speeds=sorted(list(set(sorted(speeds))))
+            total.append(speeds)
+            print speeds[0]
+            max_p = speeds[-1]
+            print max_p
+    for i in range(0,300):
+        mean = (list(total[0][i])[1]+list(total[1][i])[1]+list(total[2][i])[1])/3
+        avg.append(mean)
+        avg_all.append((mean,i))
+        
+    best = 0
[…]

### 2012-06-25 code/runGaitAndLogPos.py

@@ -18,7 +18,7 @@ from Robot import RobotFailure
 def savePosition(wiiTrack, posAgeList):
     position,age = wiiTrack.getPosAge()
     #print 'savePosition:', position, age
-
+#    pdb.set_trace()
     if position is None:
         return '-1 -1 %f' % age
     else:
@@ -75,8 +75,8 @@ def main():
                     relTimeBucket.append((relTime, line[1], line[2]))
         #pdb.set_trace()
     print relTimeBucket[0]
-    if list(relTimeBucket[-1])[0]-list(relTimeBucket[0])[0] < 13.7:
-        exit(2)
+#    if list(relTimeBucket[-1])[0]-list(relTimeBucket[0])[0] < 13.7:
+ #       exit(2)
         ##############
 
     ff = open (posFile, 'w')
@@ -90,7 +90,9 @@ def main():
         ff.write(line + '\n')
     ff.close()
     print 'Wrote position file:', posFile
-
+    lines = open(posFile, 'r').readlines()
+    speed = ((float(lines[1].split()[19])-float(lines[-1].split()[19]))**2+(float(lines[1].split()[20])-float(lines[-1].split()[20]))**2)**0.5/(6.25*12)
+    print 'SPEED: ', speed
 
 
 if __name__ == '__main__':

### 2012-06-25 paper_aimsa/abstract.tex

@@
 studies using the QuadraTot have compared parameterized splines, the
-HyperNEAT generative encoding and genetic algorithm.
-Among these, the research on the genetic algorithm was conducted by \citep{glette2012evolution-of-locomotion-in-a-simulated} in a simulator and tested on a real robot.
-Here we
+HyperNEAT generative encoding and custom spline-based genetic algorithms. Some of the most promising results 
+have been learned in a simulator and tested on a real robot \citep{glette2012evolution-of-locomotion-in-a-simulated}. Here we
 compare these results to an algorithm called Policy learning by

### 2012-06-25 paper_aimsa/conclusion.tex

@@
 optimizing a quadrupedal gait for linear speed, tested on a physical robot.  We implemented the algorithm, Policy learning by Weighting
-Exploration with the Returns (a.k.a RL PoWER), for parameterized gaits and compared
+Exploration with the Returns (RL PoWER), for parameterized gaits and compared
 it to gaits produced by neural networks evolved with the HyperNEAT
 generative encoding. Though both methods resulted in an improvement
-over the robot’s previous \naive gaits, based on the statistics
+over the robot's previous \naive gaits, based on the statistics
 collected, RL PoWER has a more elegant and consistent performance.
@@
 the same time. Evolvable splines can serve as a general representation for various other
-learning problems.
+learning problems. \editbox{This last paragraph is better, but it still needs to be rewritten to be more readable. It's the last thing people will read in the paper, so it should be strong.}

### 2012-06-25 paper_aimsa/methods.tex
line endings

@@
 The simplest model with back-compatibility is geometric
-splines. For example, for a given model f(x) with K knots, we can preserve the
+splines. \editbox{Too fast... add a sentence or two on what the methods are trying to do, and what back-compatibility is.} For example, for a given model f(x) with K knots, we can preserve the
 exact shape of the generated curve while adding extra knots to the
@@
 order to do this, we need to define an algorithm for evolving the
-parameterization from $K$ to $L$ knots ($L > K$), which is formulated as in
-\ref{alg_Evolve_Spline}.  Without loss of generality, the policy parameters are
+parameterization from $K$ to $L$ knots ($L > K$). One such algorithm from
+\cite{kormushev2011bipedal-walking-energy} is reproduced below as \algref{Evolve_Spline}.
+Without loss of generality, the policy parameters are
 normalized into $[0, 1]$, and appropriately scaled and shifted as necessary upon use.
@@
 \caption{EvolvePolicy-Spline ($P_{current}$: current policy, $L$: desired new number of parameters)}
-\label{alg_Evolve_Spline}
+\alglabel{Evolve_Spline}
 \begin{algorithmic}[1]
@@
 parameterization. The spline knots are the actual policy parameter
-values. This original parameterization starts from 4 knots and grows up to 32 knots}
+values. This original parameterization starts from 4 knots and grows up to 32 knots. Figure reproduced from \cite{kormushev2011bipedal-walking-energy}.}
 
@@
 \]
+
+\editbox{Add a new equation showing all servos (all nine) with their
+  actual values of the parameters $R$ and $C$ plugged in.}

### 2012-06-25 paper_aimsa/report/abstract.tex

% A brief summary of the report. This is not an introduction -
% it should be complete. Include 1-2 sentences for each of the items
% below, up to and including conclusions.
An evolvable spline representation has been tested on bipedal walking energy reduction. 
Combined with expectation-maximization, we proposed a new reinforcement way for quadupedal robot gait learning for walking speed maximization. 
We compare results with HyperNEAT, a previously tested efficient 
method for this problem. The proposed new method is tested to be better on average and best gait than the HyperNEAT. In the meanwhile, we tried to solve some existing issue with the testing platform---QuadraTot. New methods was added to the existing 
robot control library and fixed over-fierce action issue. 
%Because the fastest learned walk
%was not significantly faster than the fastest randomly generated walk,
%we conjecture that the motion representation for the robot is more
%integral to forward speed than the learning algorithm.

Keywords: Platform Improvement, Learning Control, Walking Robots, Gait Learning, Quadruped

### 2012-06-25 paper_aimsa/results.tex

@@
 \figp{rlPerform}{0.95}{Average fitness ($\pm$ SE) in cm/sec of the highest performing run of RL PoWER's three runs. }
+
 \figp{hyperPerform}{0.75}{Average fitness ($\pm$ SE) in cm/sec of the highest performing
-individual in one HyperNEAT run.}
+individual in one HyperNEAT run. \edit{Make this a vector graphic, not raster (raster == blurry)}.}
 

### 2012-06-25 paper_aimsa_report/experimentalSetup.tex

--- 2012-06-10 paper_aimsa/experimentalSetup.tex
@@
 
-The quadruped robot used, QuadraTot, was assembled from parts purchased online and
+The quadruped robot used was assembled from parts purchased online and
 parts printed by the Objet Connex 500 3-D Printing System. The robot
@@
 joint with AX-12+servo in each of the four legs, and one AX-18+ servo
-at the center. To avoid the formerly reported \edit{???} AX-18+ servos are used
+at the center. To avoid the formerly reported AX-18+ servos are used
 in this robot because of their stronger actuation power than that of
 AX-12+. Each servo could be set to a position in the range [0, 1023]
-by using pydynamixel library, corresponding roughly to a physical range [-120$^\circ$,
+by using PyDynamixel library, corresponding to a physical range [-120$^\circ$,
   +120$^\circ$]. Also, to prevent collisions with the robot body, the control
@@
 general. These gaits tend to either overburden the servos or flip the
-robot. \edit{Explain this next part on smart cropping better.} Thus, RL PoWER were tested after carefully cropping out the
+robot. Thus, RL PoWER were tested after carefully cropping out the
 extremely fierce gaits each time. As described earlier, each leg has
@@
 
-\figp{smartCrop}{1}{The red line is the cropping border, $g = j1 +  j2$. As shown in \figref{sixLegPositions} \edit{???}}
+\figp{smartCrop}{1}{The red line is the cropping border, $g = j1 +  j2$. As shown in \figref{sixLegPositions}}
 
 \figp{sixLegPositions}{1}{The left two positions are allowed; middle
-  two positions are on the red border line from \figref{smartCrop};
-  right two positions are deemed as ``extremely fierce'', thus are
-  disallowed.   \\\editbox{Smart crop needs to be explained somewhere, probably briefly in this caption and more extensively in the text.}}
+  two positions are on the red border line from \figref{smartCrop}; right two positions are deemed as "extremely fierce", thus are disallowed.}

### 2012-06-25 paper_aimsa_report/results.tex

--- 2012-06-05 paper_aimsa/report/report.tex
@@
-\section{Introduction}
-
-% Motivate and abstractly describe the problem you are addressing and
-% how you are addressing it. What is the problem? Why is it important?
-% What is your basic approach? A short discussion of how it fits into
-% % related work in the area is also desirable. Summarize the basic
-% results and conclusions that you will present.
-
-Applications of walking robots often call for the ability to walk as
-quickly, smartly or with as little power as possible. Some groups have
-used learning methods to generate gaits optimized for some metric.
-Approaches differ in their starting assumptions, some essentially
-tweaking the parameters of a hand-tuned model others
-exploring a reasonably compact parameter space.
-Here we present a new method, RL PoWER, reinforcement learning with EM(Expectation Maximization) as it was tested effective on related robotics problems. 
-This paper shows a comparison of the two different learning methods 
-implemented, HyperNEAT and RL PoWER.  RL PoWER
-method created walks that are substantially faster than the previous
-HyperNEAT gait. I invite readers with short attention spans to view
-two videos to compare the previous gaits and the current stable gaits:
-
-\url{http://www.youtube.com/watch?v=ODoiOj9DdGg}
-
-\url{http://www.youtube.com/watch?v=LI76mQepEjc&feature=youtu.be}
-
-%The code so far consists of the robot class, optimization class,
-%parameterized model, sine model, and camera feedback.
-
-\section{Problem definition}
-% Precisely define the problem you are addressing (i.e. formally specify
-% the inputs and outputs). Elaborate on why this is an interesting and
-% important problem.
-
-I am testing and comparing two learning methods, HyperNEAT and RL PoWER
-, to design a parametrized gait for a quadruped robot from the Cornell
-Computational Synthesis Lab that maximize forward walking speed. 
-Subproblems include improving the robostness of the robot so that it 
-fails less frequently and make the measurement more reliable by minimizing the leg sliding.
-
-\section{Implementation}
-\seclabel{implement}
-% Describe how you implemented your system and how you structured it. 
-% This should give an overview of the system, not a detailed 
-% documentation of the code. The documentation of the code is part of 
-% the code you hand in. You might want to comment on high-level design 
-% decisions that you made. Also explain how you obtained your
-% data and any pre-processing you did to it.
-\begin{itemize}
-\item \code{RL\_PoWER.m*} The RL PoWER algorithm implementation defined in matlab code. Output a gait sequence file for 9 motors. Supports restart from previously interrupted trial. 
-\item \code{runRobot.m*} Communication code for reading the generated file of motor actions and execute on robot by calling the python code. 
-\item \code{localSettings.m*} Execution settings file. 
-\item \code{load\_trajectory.m*} Load the generated gait file and do the conversion.
-\item \code{exportTrajectoryWithoutRescaling.m*} Export the spline function to actual robot gaits
-\item \code{paramToPolicy.m*} Transforming spline representation to actual robot gaits parameters.
-\item \code{policyToParam.m*} Transforming gait parameters to corresponding spline representation.
-\item \code{PoWER\_QuadraTot\_23.m*} main method for system call
-\end{itemize}
-
-\footnote{*Contributed equally by Haocheng Shen, Petar Kormushev and Jason Yosinski}
-
-\section{Method}
-\seclabel{method}
-
-% Describe in reasonable detail the algorithm you are using to address
-% this problem. A pseudo-code description of the algorithm you are
-% using is frequently useful. If it makes sense for your project,
-% trace through a concrete example, showing how your algorithm
-% processes this example. The example should be complex enough to
-% illustrate all of the important aspects of the problem but simple
-% enough to be easily understood. If possible, an intuitively
-% meaningful example is better than one with meaningless symbols.
-\subsection{Policy Representation by Splines}
-The simplest model with back-compatibility is geometric
-splines. For a given model f(x) with K knots, we can preserve the
-exact shape of the generated curve while adding extra knots to the
-original spline. Say, if we put one additional knot between every two
-consecutive knots of the original spline, we end up with a 2K - 1
-knots and a spline that has the same shape as the original one. In
-order to do this, we need to define an algorithm for evolving the
-parameterization from $K$ to $L$ knots ($L > K$), which is formulated as 
-Algorithm 1. Without loss of generality, the policy parameters are
-normalized into $[0, 1]$, and appropriately scaled/shifted as necessary
-later upon use.
-
-\begin{algorithm}
-\caption{EvolvePolicy-Spline ($P_{current}$: current policy, $L$: desired new number of parameters)}
-\label{alg_Evolve_Spline}
-\begin{algorithmic}[1]
-
-\STATE $K$ $\leftarrow$ $P_{current}.numberOfParameters$
-
-\STATE $X_{current}$ $\leftarrow$ $[0, \frac{1}{K-1}, \frac{2}{K-1}, ... , 1]$ % Knots
-\STATE $Y_{current}$ $\leftarrow$ $P_{current}.parameterValues$ % Values
-\STATE $S_{current}$ $\leftarrow$ ConstructSpline($X_{current}$, $Y_{current}$) % Spline
-%\COMMENT{ Optional: make the spline cyclic}
-
-\STATE $X_{new}$ $\leftarrow$ $[0, \frac{1}{L-1}, \frac{2}{L-1}, ... , 1]$
-\STATE $Y_{new}$ $\leftarrow$ EvaluateSplineAtKnots($S_{current}$, $X_{new}$)
-\STATE $S_{new}$ $\leftarrow$ ConstructSpline($X_{new}$, $Y_{new}$)
-%\COMMENT{ Optional: make the spline cyclic}
-
-\STATE $P_{new}.numberOfParameters$ $\leftarrow$ $L$
-\STATE $P_{new}.parameterValues$ $\leftarrow$ $S_{new}$.$Y_{new}$
-
-\STATE return $P_{new}$
-
-\end{algorithmic}
-\end{algorithm}
-
-For the experiment, we set 3 knots for each servo and there are 8
-servos in total. The servo in the hip is not used in
-our experiment. Previous work has verified that quadruped gaits
-perform better when they are coordinated \cite{clune2009evolving-coordinated-quadruped} \cite{clune2011on-the-performance-of-indirect-encoding}
-\cite{valsalam2008modular-neuroevolution-for-multilegged}. For each spline, we calculate its corresponding parameterized gait for one unit time cycle. Given that, then apply the same
-pattern to every cycle throughout the 12 seconds of one
-trial. Specifically, each spline(a set of 3 knots) is interpreted to its corresponding servo positions as
-following:
-\[
-\vec{g}(t) =
-\left[ {\begin{array}{c@{ }c@{ }c@{ }l@{ }l}
-R \cdot f(s1, s2, s3) & \ \          & \             & \            & + C \\ % 1
-0                              & \             & \             & \            & + C_C \\ % 8
-\end{array} } \right]
-\]
-And the meaning of its parameters are listed in \tabref{parameters}
-\begin{table}[b]
-\begin{center}
-\begin{tabular}{|c|c|c|}
-\hline
-Parameters        &                           &       \\
-in $\vec{\theta}$ & Description               & Range \\
-\hline
-\hline
-$f(s1,s2,s3)$        & Spline function           & [0,1] \\  %subject to change
-\hline
-$R$          & Position multiplier                & [256, 768] \\
-\hline
-\end{tabular}
-\caption{The \emph{RL PoWER} motion model parameters.}
-\tablabel{parameters}
-\label{tab:params}
-\end{center}
-\end{table}
-
-\section{Related work}
-% Briefly explain who else worked on related problems in the past and what
-% methods they used. Explain if you are using similar methods, or if your
-% approach is different and if so - how (either is ok).
-Many attempts have been made on robot gaits learning using machine learning algorithms, producing competitive results (Chernova and Veloso, 2005; Hornby et al., 2005; Zykov et al., 2004; Clune et al., 2009a, 2011, 2009b,c; T´ellez et al., 2006; Valsalam and Miikkulainen, 2008).
-In fact, Jason et al.
-tested and compared six different algorithms for quadruped robot gait learning in their studies.
-HyperNEAT, a generative encoding neural network algorithm, achieved the best performance among all six methods.
-In Kormushev et al.
-(2011)'s bipedal robot energy reduction research, a reinforcment learning method was used to optimize the performance.
-The results from Kormushev's research showed this algorithm's poetential for walking problems.
-This reinforcement learning method still needs more tests for further explorations.
-
-\section{Experimental Evaluation}
-Our performance metric was the displacement over the evaluation period of 12 seconds
-for each. Same as Yosinski et al. (2011), the displacement was measured using a
-Wii remote that was placed on the ceiling. Different from the original model described
-in Yosinski et al. (2011), the quadruped robot was equipped with a three-infrared-LED
-cluster on top rather than just one. The reason for this setup is that when fierce gaits
-were executed, the Wii remote loses tracking of the robot position due to the limited
-visible angle of a single LED. These three LEDs were place tightly together to act
-as one signal emitter. Each LED was tilted outwards in order to maximize the visible
-range. A separate tracking server ran on the robot PC interacted with theWii remote via
-bluetooth by using the CWiid library. A corresponding client communicates with this
-server via socket connection. Our fitness evaluation code talks with the Wii remote by
-using this client and updates the position from beginning to end during each run. The
-metric for evaluating gaits was the Euclidian distance the robot traveled during a 12-
-second run on flat surface. Previous work done in Yosinski et al. (2011) and Clune et al.
-(2009) suggested that extremely fierce gaits are not viable in general. These gaits tend
-to either overburden the servos or flip the robot. Thus, RL PoWER were tested after
-carefully cropping out the extremely fierce gaits each time. As described earlier, each
-leg has two joints, inner joint j1 and outer j2. The position of one leg is determined by
-the sum of the values of j1 and j2. Extremely fierce gaits usually have relatively small
-values of g. As shown in Figure 3, this cropping method works by mapping the wild
-gaits onto the boundary of a quasi-triangular area in the two-dimensional space of j1
-and j2.
-
 \section{Results}
-
-\subsection{Hardware and interface Improvement}
-The robot has been verified to fail under extreme fierce motions.
-So, one change I made to this is preventing these actions to happen from the first hand.
-Thus the servo failure chance is solved for most of the time.
-The threshold of cropping was carefully chosen so that most gaits are maintained.
-
-\subsection{Quantitative}
-The results for the gaits evolved by RL PoWER and HyperNEAT are shown in \tabref{results}.
+The results for the gaits evolved by RL PoWER are shown in \figref{rlPerform} and \tabref{results}.The HyperNEAT performance is shown in \figref{hyperPerform}.
 A total of 900 evaluations were performed for RL PoWER (300 for each of three runs).
@@
 either average or best gaits.
+
+\figp{rlPerform}{0.95}{Average fitness ($\pm$ SE) in cm/sec of the highest performing run of RL PoWER's three runs. }
+\figp{hyperPerform}{0.75}{Average fitness ($\pm$ SE) in cm/sec of the highest performing
+individual in one HyperNEAT run.}
 
@@
 
-\subsection{Discussion}
-% Is your hypothesis supported? What conclusions do the results support
-% about the strengths and weaknesses of your method compared to other 
-% methods? How can the results be explained in terms of the underlying 
-% properties of the algorithm and/or the data.
-Overall, the RL PoWER gaits were
-faster by far, beating the HyperNEAT when comparing
-either average or best gaits. I believe that this is because RL PoWER uses evolvable splines to represent gaits As explained earlier,  
+We believe that this is because RL PoWER uses evolvable splines to represent gaits As explained earlier,  
 The single best gait found during this study had a speed of 11.09 centimeters/sec,  14.7\% better than the best HyperNEAT gait.
@@
 
-\figp{neat_110115_211410_00000_002_filt_zoom}{0.75}{A well performing HyperNEAT gait's pattern of motor positions}
+\figp{neat_110115_211410_00000_002_filt_zoom}{0.75}{A typical well performing HyperNEAT gait's pattern of motor positions}
 
@@
 While the evolvable spline representation used by RL PoWER has lower dimensions.
-The convergence rate of HyperNEAT is thus lower.
-
-\section{Future work}
-% What are the major shortcomings of your current method? For each 
-% shortcoming, propose additions or enhancements that would help overcome it.
-Though 900 trials for RL PoWER and 540 trials for HyperNEAT has been used to test the applicability of RL PoWER and 
-compare the performance of the two, still, much more runs are needed to statisstically compare the two algorithms.
-
-So, here are some work that may be achieved in future
-\begin{itemize}
-\item Improve hardware connection, lower the chance of servo failure
-\item More runs and/or longer runs
-\item Tuning simulator for good approximation 
-\item Using simulation to statistically compare the algorithms on quadrupedal robots
-\end{itemize}
-
-\section{Conclusion}
-% Briefly summarize the important results and conclusions presented in 
-% the paper. What are the most important points illustrated by your 
-% work? How will your results improve future research and applications 
-% in the area?
-The previously encountered hardware problems about using QuadraTot have been solved by the newly added booties.
-The measurement reliability has been enchanced by the fixed Robot class in python, preventing extremely fierce gaits.
-
-We have presented a new reinforcement-learning-based algorithm for
-optimizing a quadrupedal gait for linear speed. We implemented and
-tested six learning strategies for parameterized gaits and compared
-them to gaits produced by neural networks evolved with the HyperNEAT
-generative encoding. Though both methods resulted in an improvement
-over the robot’s previous \naive gaits, based on the statistics
-collected, RL PoWER has a more elegant and consistent performance.
-
-One guess led by this study is that for feedback oriented tasks,
-reinforcement learning methods are more fit in natural. Despite the
-complexities of HyperNEAT, a simpler algorithm on the code level,
-delivered better performance in general. Also, evolvable spline
-interpolation is shown to be simple and representationally powerful at
-the same time. Evolvable splines can serve as a general representation for various other
-learning problems.
-
-\section{Acknowledgments}
-% List any people not on the team who helped you with your project: Your
-% TA, other people you consulted with or had useful discussions (say in 
-% a word or two what they did), people who proofread your report, and 
-% any external code you used (libraries etc).
-\begin{itemize}
-\item Hod Lipson, Cornell Computational Synthesis Lab: adviser.
-\item Jason Yosinski, Cornell Computational Syntheis Lab: adviser
-\item Petar Kormushev, Italy Institute of Technology: inventor of RL PoWER algorithm
-\item Kyrrehg Glette, University of Oslo: collaborated on the design of anti-sliding booties.
-\end{itemize}
-
-\section{Team}
-% A table listing the names of the people who worked on the project, and who
-% did what.
-
-\begin{itemize}
-\item Haocheng Shen
-\begin{itemize}
-\item Robot platform reconstructing and improvement
-\item RL PoWER algorithm integration and testing
-\item \code{RL\_PoWER.m} and \code{runRobot.m} matlab code and python code integration for running trials on the reconstructed robot platform
-\item \code{Robot.py} classes, with new cropping method and bug fix
-\item \code{extract.py} scripts, for processing collected gait data and plotting
-\item \code{extractGait.py} scripts, for processing collected gait data and plotting
-\item Plotting and evaluating results
-\end{itemize}
-
-\item Jason Yosinski
-\begin{itemize}
-\item {RL\_PoWER}\code{exportTrajectoryWithoutRescaling.m} and \code{getDataPath.m} \code{paramToPolicy}, \code{LoadTrajectory}, and \code{PoWER\_QuadraTot.m} matlab code
-\item mentoring \& evaluating results
-\end{itemize}
-
-\item Petar Kormushev
-\begin{itemize}
-\item The inventor of combining reinforcemen learning and EM algoorithms for quadrupedal robot gait learning
-\item \code{RL\_PoWER.m}\code{exportTrajectoryWithoutRescaling.m} and \code{getDataPath.m} \code{paramToPolicy}, \code{LoadTrajectory}, and \code{PoWER\_QuadraTot.m} matlab code
-\item implementation of the original algorithm
-\end{itemize}
-
-\section{Appendix}
-% Describe the code files you uploaded into CMS
-\begin{itemize}
-\item Publication paper draft for AIMSA conference
-\item RL PoWER performance video online 
-\item matlab data workspace files for recording all three runs' parameters
-\item python code for processing data
-\end{itemize}
-\end{itemize}
-
-% LocalWords:  vertices
+The convergence of HyperNEAT is thus slower.

### 2012-06-25 paper_aimsa_report/extractGait.py
extractGait.py

@@ -3,7 +3,7 @@
 import pdb
 from os import listdir
 import matplotlib as mpl 
-mpl.use('cocoaagg') 
+#mpl.use('cocoaagg') 
 import matplotlib.pyplot as plt
 import re
 
@@ -39,7 +39,7 @@ def main():
     plt.xlabel('Time(s)')
     ax = plt.axis()
     plt.axis((min(t), max(t), ax[2], ax[3]))
-    plt.savefig('gait.pdf')
+    plt.savefig('gait2.pdf')
     #plt.show()
     
     

### 2012-08-07 extabstract_alife/abstract.tex
Submitted to John Rieffel

Legged robots are uniquely privileged over their wheeled counterparts
in their ability to access rugged terrain, yet manually designing gaits for
such robots can be a tedious process. Thus, we
investigate methods by which gaits may be learned autonomously by a
robot, either using physical real world trials or virtual trials in a
simulator. Each of these two methods has its advantages and
disadvantages. Physical trials are slow and often require expensive
human supervision. Simulated trials can be run quickly, in parallel,
and without supervision, which in practice allows orders of magnitude
increases in the number of trials per experiment. However, the
unavoidable shortcoming of simulated trials is that the simulation
inevitably fails to match reality exactly, resulting in what is often
called the simulation-reality gap. This gap frequently results in
learning methods overfitting to the idiosyncrasies of the simulated
world and producing gaits that do not transfer well to the physical
robot.  Thus, the chief benefit of learning gaits directly on a
physical system is the ability to avoid this gap. Herein we summarize
recent gait learning results using both types of fitness evaluation
--- physical and simulated --- on the open-source QuadraTot robot.

### 2012-08-07 extabstract_alife/introduction.tex
Submitted to John Rieffel

\section{Introduction and Previous Work}

\figgp{robot_whitebg}{.44}{simulator_screenshot_crop}{.55}{The
  open-source, 3D printed QuadraTot robot used for this study. Left:
  physical robot \citep{yosinski2011evolving-robot-gaits}. Right:
  simulator \citep{Glette2012Evolution}.}

There has been much previous work in the Artificial Life and
Evolutionary Robotics community on automatically generating behaviors
for robots~\citep{nolfi2000evolutionary,
  sims1994evolving, hornby2005autonomous, lipson2000automatic}. Much
of this work has focused on learning gaits for legged
robots~\citep{clune2009evolving, clune2011performance,
  hornby2005autonomous, hornby2003generative,
  Koos2012, bongard2006resilient,
  yosinski2011evolving-robot-gaits}.  Some previous work
has focused both on evolution directly on a physical
system~\citep{yosinski2011evolving-robot-gaits, zykov2004evolving}, but more
frequently gaits have been evolved in simulation and then transferred
to the physical robot~\citep{lipson2006evolutionary, Koos2012,
  hornby2005autonomous, bongard2006resilient}.  Whether simulation or
reality is used gait evaluation, a common thread in these
studies is that learning algorithms are able to produce gaits that
outperform those designed by a human
engineer~\citep{yosinski2011evolving-robot-gaits, hornby2005autonomous}.

%\subsection{Problem Definition}
%\seclabel{problemDefinition}
%
%Before proceeding, we briefly pause to more concretely formulate the
%gait learning problem to avoid any ambiguity. The gait learning
%problem aims to find a \emph{gait} that maximizes some performance
%metric. Mathematically, we define a gait as a function that specifies
%a vector of commanded motor positions for a robot over time.  We can
%write gaits without feedback --- also called open-loop gaits --- as
%
%\be
%\vec{x} = g(t)
%\ee
%
%\noindent for commanded position vector $\vec{x}$.  The function
%depends only on time, and thus it follows that open-loop gaits are
%deterministic, producing the same command pattern each time they are
%run.  While the commanded positions will be the same from trial to
%trial, the actual robot motion and measured fitness will vary due to
%the noisiness of trials in the real world.

### 2012-08-07 extabstract_alife/results.tex
Submitted to John Rieffel

\section{Summary of Results and Discussion}
\seclabel{results}

In the interest of space, we present only a brief summary of gait
learning results in \tabref{results}.  All results were obtained using
the QuadraTot robot, with learning performed using either physical or
simulated trials.  These promising results suggest that evolutionary and
learning algorithms may enable more rapid
exploration of the space of robotic gaits and other behaviors than
would be possible manually.

%%%%%%% NEW TABLE in cm/s
\begin{table*}
\begin{center}
\begin{tabular}{|r|c|c|c|c|c|}
\hline
Learning Algorithm                       &   & Type of  &         &           & Single \\
(number of trials)                       & Paper  & Trials   & Average & Std. Dev. & Best Gait \\
\hline                                        
\hline                                        
Hand-coded gait                          & \cite{yosinski2011evolving-robot-gaits}  & --       & 1.09   &   --   &       \\
\hline                                        
Policy Gradient Descent (174)            & \textquotedbl  & physical & 1.34   &   1.56 &       \\
\hline                                                         
Uniform Random Hill Climbing (234)       & \textquotedbl  & physical & 1.66   &   0.97 &       \\
\hline                                                          
Random search (200)                      & \textquotedbl  & physical & 1.99   &   1.45 &       \\
\hline                                                          
Gaussian Random Hill Climbing (284)      & \textquotedbl  & physical & 2.12   &   1.27 &       \\
\hline                                                          
Nelder-Mead simplex (172)                & \textquotedbl  & physical & 2.61   &   0.71 &       \\
\hline                                                          
Linear Regression (153)                  & \textquotedbl  & physical & 2.97   &  2.73  &       \\
\hline                                                          
HyperNEAT (540)                          & \textquotedbl  & physical & 5.56   &   1.35 &  9.68 \\
\hline                                                      
RL PoWER (900) & \cite{shen2012learning-fast-quadruped}        & physical   & 7.62   & 2.1   &  11.05 \\
\hline                                                      
Alternate Genetic Algorithm (60,000) & \cite{Glette2012Evolution} & simulated & (14.2)   &  (5.2) &  (17.8) 13.0$^*$ \\
\hline                                                      
HyperNEAT (800,000) & \cite{lee2012INPREP-evolving-quadruped-gaits} & simulated &    &    &  14.5 \\
\hline
\end{tabular}
\caption{The average and standard deviation of the best gait speeds,
  in cm/sec, found using various encodings and learning
  algorithms. The single best observed gait is also included where
  available. All reported speeds are those observed in the physical
  world; i.e. for simulated trials we report only the speed of the
  gait when transferred to reality, not the speed observed in
  simulation.  Speeds in parentheses are reported in
  \cite{Glette2012Evolution} and were observed on a QuadraTot
  configured similarly, though not identically, to
  that in \cite{yosinski2011evolving-robot-gaits}. The speed with an asterisk
  is the measurement of the same gait on the QuadraTot from
  \cite{yosinski2011evolving-robot-gaits}.}  \tablabel{results}
\end{center}
\end{table*}

\subsection{Open Robotics}

Despite this promise, the field remains small, partly because robots
are expensive and difficult to modify. Access to cheap, customizable
robots could increase the number of researchers able to participate in
the field. Moreover, in nearly all of the papers mentioned previously,
the robots were custom-made, preventing teams at other universities
from reproducing the results of other groups and or testing new
algorithms on a robotic platform used in a previous study. That, in
turn, slows the progress of science because it is difficult to
interpret whether the variance in results between different studies
was due to the algorithms used or the robotic platform those
algorithms were tested on.

Some robot platforms are emerging, but they tend to be wheeled robots
without complex kinematics, such as the ePuck~\citep{mondada2009puck}.
For this reason we have introduced several low-cost, open-source, and
easily customizable robotic platforms, such as the QuadraTot
\citep{yosinski2011evolving-robot-gaits} and Aracna
\citep{lohmann2012aracna-an-open-source-quadruped}. These robots can
be 3D printed from design files available online, allowing other
researchers to customize the body's design.  These robots will enable
multiple users to compare data across a small number of lightweight,
low-cost evolutionary robotic platforms.

### 2012-08-14 paper_sean/abstract.tex
First draft of Sean's paper, text only

Evolving gaits for physical robots is a longstanding and still open challenge. 
While previous work has shown that evolutionary algorithms can outperform human engineers in generating gaits, in all of that work the experimenter has had to manually decide the regularities of the gaits. 
Such manual intervention is time consuming, requires expert knowledge, and adds constraints that may hurt performance. 
In this paper, we tested whether HyperNEAT -- a generative encoding that has shown to be able to automatically evolve a variety of high-performing regular gaits -- could generate gaits that could outperform manually designed, regularity-constrained, directly-encoded gaits. 
For our experiments, gaits were evolved in simulation and the final gait transferred to the real-world to evaluate the gait's actual performance.
The platform used was the QuadraTot, an open-source, 3D printed quadrupedal robot on which previous studies in gait evolution has already been done several times. 
In this study, we produced the fastest gait ever tried on this platform, outperforming the fastest published gait by 5.4\% with 20000 fewer evaluations run during gait generation. 
The gaits produced showed exploitation in regularity in simulation and real-world as was expected.

### 2012-08-14 paper_sean/bibliography.tex
First draft of Sean's paper, text only

\bibitem {bongard:lipson}
Bongard, J., Zykov, V., and Lipson, H.:
Resilient Machines Through Continuous Self-Modeling.
Science. 314(5802):1118-1121 (2006)
%
\bibitem {valsalam:mii}
Valsalam, V. and Miikkulainen, R.:
Modular Neuroevolution for Multilegged Locomotion.
In: Proceedings of the Genetic and Evolutionary Computation Conference, pp. 265--272 (2008)
%
\bibitem {kohl:stone}
Kohl, N. and Stone, P.:
Machine Learning for Fast Quadrupedal Locomotion.
In: The Ninetheenth National Conference on Artificial Intelligence (AAAI 2004), pp. 611--616. San Francisco (2004)
%
\bibitem {clune1}
Clune, J., Beckmann, B., Ofria, C., and Pennock, R.:
Evolving Coordinated Quadruped Gaits with the HyperNEAT Generative Encoding.
In: Proceedings of the IEEE Congress on Evolutionary Computation, pp. 2764--2771 (2009)
%
\bibitem {clune3}
Clune, J., Ofria, C., and Pennock, R.:
The Sensitivity of HyperNEAT to Different Geometric Representations of a Problem.
In: Proceedings of the Genetic and Evolutionary Commuptation Conference, pp. 675--682 (2009)
%
\bibitem {stanley1}
Stanley, K. O., D'Ambrosio, D. B., and Gauci, J.:
A Hypercube-Based Encoding for Evolving Large-Scale Neural Networks.
Artificial Life. 15(2), 185-212 (2009)
%
\bibitem {clune2}
Clune, J., Stanley, K. O., Pennock, R., and Ofria, C.:
On the Performance of Indirect Encoding Across the Continuum of Regularity.
IEEE Transactions on Evolutionary Computation. 15, 346-367 (2011)
%
\bibitem {yos:clune}
Yosinski, J., Clune, J., Hidalgo, D., Nguyen, S., Zagal, J. C.:
Evolving Robot Gaits in Hardware: the Hyperneat Generative Encoding vs. Parameter Optimization.
In: Proceedings of the 20th European Conference on Artificial Life, pp. 11--18 (2011)
%
\bibitem {glette}
Glette, K., Kalus, G., Zagal, J. C., and Torresen, J.:
Evolution of Locomotion in a Simulated Quadruped Robot and Transferral to Reality.
In: Proceedings of the Seventeenth International Symposium on Aritifical Life and Robotics, (2012)
%
\bibitem {haocheng}
Shen, H., Yosinski, J., Kormushev, P., Caldwell, D. G., Lipson, H.:
Learning Fast Quadruped Robot Gaits with the RL PoWER Spline Parameterization.
In: AIMSA 2012 Workshop on Advances in Robot Learning and Human-Robot Interaction, (2012)
%
\bibitem {robotis}
Robotis:
User's Manual, Dynamixel AX-12.
URL: http://www.trossenrobotics.com/images/productdownloads/AX-12(English).pdf
%
\bibitem {clune:lipson}
Clune, J., Lipson, H.:
Evolving Three-Dimensional Objects with a Generative Encoding Inspired by Developmental Biology.
In: Proceedings of the European Conference on Artificial Life, pp. 144--148 (2011)
%
\bibitem {stanley2}
Stanley, K. O.:
Compositional Pattern Producing Networks: A Novel Abstraction of Development.
Genetic Programming and Evolvable Matter. 8(2), 131--152 (2007)
%
\bibitem {stanley3}
Stanley, K. O.:
Exploiting Regularity Without Development.
In: Proceedings of the AAAI Fall Symposium on Developmental Systems, (2006)
%
\bibitem {stanley4}
Stanley, K. O., Miikkulainen, R.:
Evolving Neural Networks Through Augmenting Topologies.
Evolutionary Computation, 10(2):99-127 (2002)
%
\bibitem {jakobi}
Jakobi, N., Husbands, P., and Harvey, I.:
Noise and the Reality Gap: The Use of Simulation in Evolutionary Robotics.
In: Advances in Artificial Life: Proceedigns of the 3rd European Conference on Artificial Life, pp. 704--720, Springer-Verlang, (1995)
%%
\bibitem {koos1}
Koos, S., Mouret, J.-B., Doncieux, S.:
Crossing the Reality Gap in Evolutionary Robotics by Promoting Transferable Controllers.
In: Proceedings of the 12th Annual Conference on Genetic and Evoutionary Computation, pp. 119--126 (2010)
%
\bibitem {carroll}
Carroll, S.:
Endless Forms Most Beautiful: The New Science of Evo Devo and the Making of the Animal Kingdom.
Norton, New York (2005)
%
\bibitem {bongard}
Bongard, J.:
Synthesizing Physically-Realistic Environmental Models from Robot Explorations.
In: Proceedings of the 9th European Confrence on Advances in Artificial Life, pp. 806--815 (2007)
%
\bibitem {zagal}
Zagal, J., Ruiz-del Solar, J., and Vallejos, P.:
Back to Reality: Crossing the Reality Gap in Evolutionary Robotics.
In: Proceedings of IAV 2004, the 5th IFAC Symposium on Intelligent Autonomous Vehicles, (2004)
%
\bibitem {koos2}
Koos, S., Mouret, J.-B., Doncieux, S.:
The Transferability Approach: Crossing the Reality Gap in Evolutionary Robotics.
IEEE Transactions on Evolutionary Computation, (2011)

### 2012-08-14 paper_sean/conclusion.tex
First draft of Sean's paper, text only

HyperNEAT has proven to be an effective encoding in gait evolution and is able to find gait regularities without human intervention \cite{clune1}, \cite{clune2}, \cite{yos:clune}. 
Glette et al. \cite{glette} has shown that evolving gaits in simulation using a simple algorithm with a large number of evaluations could outperform gait evolution using a more sophisticated algorithm in hardware.  
This was possible because the simulator afford a much larger number of evaluations to be performed during the evolution process. 
In this study, we took the two -- a sophisticated algorithm (HyperNEAT) and a physical simulator -- to see how HyperNEAT would perform with a many evaluations in the evolution process. 
We showed that HyperNEAT, when used with a simulator, could evolve very impressive gaits. 
In simulation, HyperNEAT gaits outperformed Genetic Algorithm gaits \cite{glette} by 54.3\%. 
In the real life trials, HyperNEAT's best gait performed 5.4\% better than the GA gait. 
This study also showed that HyperNEAT will exploit the characteristics of the problem in order to maximize its fitness.
Without frequency limiters, it would make the legs move really fast in order to increase fitness. 
When that was no longer allowed, it pushed the servo positions to their boundaries by increasing the stride lengths as wide as possible. 
However, because of the physical limitations of the robot, these exploitations could not be transferred well to reality. 
Moreover, the inevitable faults with the simulator (it cannot model the robot or the real world perfectly) added to the difficulty in the transfer from simulation to reality. 
Thus, many of the gaits performed much worse than it did in simulation and had difficulty in achieving consistent results.

Next steps for this research heads towards improving the transferabilities of the gaits from simulation to reality. 
The performance gap between simulation and reality is called the \emph{reality gap} \cite{jakobi}. 
Many different approaches to reduce the reality gap has been proposed \cite{koos1}, \cite{bongard}, \cite{zagal}.
To improve upon the results from this study, one or more of these approaches could be used to aid in the transfer to reality. 
The one easiest and most promising to implement is the Transferability Approach \cite{koos2} which has shown to be successful in quadrapedal gait evolution. 
HyperNEAT, used with a simulator and a transfer approach, shows much hope in producing yet a better, more coordinated, and higher performing gait on the QuadraTot platform.

### 2012-08-14 paper_sean/hyperneat.tex
First draft of Sean's paper, text only

HyperNEAT is a neuroevolution method which evolves artificial neuron networks (ANNs) using evolutionary algorithms.
It indirectly encodes ANNs using compositional pattern producing networks (CPPNs) which represent genomes encoding ANN phenotypes \cite{stanley1}, \cite{stanley2}. 
HyperNEAT has shown that it could evolve large neural networks that represent the networks and features present in brains such as repetitions and regularities \cite{stanley3}. 
This encoding has previously been applied for gait-learning \cite{yos:clune} and for evolving 2-D and 3-D objects \cite{clune:lipson}. 
It has been shown that objects evolved using HyperNEAT look complex, natural and feature characteristics such as symmetries and repetition of themes which are often found in nature. % figure from jeff paper

The motivation for CPPNs is that complex patterns can be produced by determining attributes of their phenotypic components as a function of their geometric location. 
This is based on the belief that in nature, cells in a body differentiates and specializes in different things with respect to their location. 
A cell in natural organisms cannot determine its location in space by itself and requires chemical gradients are used to signal its location \cite{carroll}. 
In contrast, \emph{in silico}, cells can be given their geometric coordinates and CPPNs use this to their advantage.

Each CPPN is a directed graph in which every node is a mathematical function, such as sine or Gaussian. % figure!
Moreover, a CPPN genome allows functions to be made of other functions, allowing coordinate frames to be combined and hierarchies to develop.
For example, a sine function could develop a repeating theme which is passed onto a Gaussian function to create a repeating series of symmetrical motifs. 
These functions allow evolution to exploit various properties such as symmetry (e.g. Gaussian) and repetition (e.g. sine function) \cite{stanley1}. 
Because of evolution is able to exploit such properties, things like gaits are able to feature repetition and symmetry on its own without human intervention \cite{clune1}.

There are links between nodes in CPPNs each of which has a weight used to multiply the signal in each link to determine the magnitude of effect from each node. 
In HyperNEAT, a CPPN genome takes in Cartesian coordinates (X,Y) of the source and the target nodes, and a constant bias value and outputs the weight between the input and hidden layer and a weight between the hidden layer and output (assuming that there is a hidden layer, as in the case for this study). 
All pairwise combinations of source and target nodes are passed into a CPPN to determine the weight of each ANN link.

HyperNEAT is first neuroevolutionary algorithm that has shown capabilities in exploiting the geometry of the problem \cite{stanley1}, \cite{clune3}.
This is possible because the connection weights are a function of geometric positions of these nodes. 
If positions inputed to CPPNs represent aspects of the problem relevant to the solution, HyperNEAT could use that information to its advantage.

HyperNEAT evolves the CPPNs according to the principles of the Neuroevolution of Augmenting Topologies (NEAT) algorithm \cite{stanley4}. 
The NEAT algorithm has three major components \cite{stanley4}. 
First, it starts with small genomes that encode simple networks and complexifies them via mutations and add nodes and links to the network which allows the alogithm to evolve the network topology and weight. 
Second, NEAT has a fitness-sharing system that preserves diversity in the system and allows for new innovations to be tuned by evolution before competing them against more adapted rivals. 
Third, the historical information stored in genes helps to perform crossover in a way that is effective, yet avoids the need for expensive topological analysis. A full explanation of NEAT can be found in Stanley and Miikkulainen \cite{stanley4}.

In this study, the ANN configuration from evolving gaits with HyperNEAT in hardware \cite{yos:clune} was used. 
In this ANN configuration, the ANN has a fixed topology that consists of three 3 X 4 Cartesian grids of nodes for the input, hidden, and output layers. 
The inputs to the substrate were the angles requested in the prevous time step for each of the 9 joints of the robot and a sine and cosine wave to facilitate periodic motion. 
The outputs of the substrate at each time step were nine numbers (for each joint) in the range [-1, 1] which were scaled to the range [0, 1023], the allowable ranges for the servos. 
As in Yosinski et al. \cite{yos:clune}, we requested four times as many commanded positions from HyperNEAT ANNs and averaged over four commands at a time in order to reduce the possibility of switching from extreme negative to extreme positive numbers at a very high frequency.

%%% FIGURES %%%

\begin{figure}
\begin{center}
\vspace{1cm}
\epsfig{file=CPPN.eps, width=8cm}
\caption [ ]{Compositional Pattern Producing Networks (CPPNs). CPPN  genomes use mathematical functions to generate regularities such as symmetry and repetitions. CPPN genomes allow functions to be made of other functions, allowing multiple regularity motifs to be present. Figure taken from Stanley \cite{stanley2}.}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\vspace{1cm}
\epsfig{file=hyperneat.eps, width=8cm}
\caption [ ]{Producing ANNs from CPPNs. In HyperNEAT, weights are specified as a function of the Cartesian coordinates for source and input nodes and a constant bias for each connection between the source and input. All pairwise combinations of source and target node coordinates are iteratively passed into CPPNs to determine the weight of each ANN link. Figure taken from Clune et al. \cite{clune2}}.
\end{center}
\end{figure}

### 2012-08-14 paper_sean/introduction.tex
First draft of Sean's paper, text only

Legged locomotion provides several benefits over wheeled locomotion. 
A few examples of this are in flexibility in the type of terrain a robot could travel and in adapability to changes in a robot's physiology \cite{bongard:lipson}.
But designing a gait for a robot is often difficult and time consuming because the engineer must take into consideration many factors which must all fall into place for a successful gait pattern. 
Because of setbacks present in hand-designing gaits, evolving gaits using machine learning algorithms is an attractive and flexible approach that could be used for many different robot platforms with only a few tweaks.

Much research has been done in this subject, and previous work has shown that learned gaits can outperform designed gaits \cite{valsalam:mii}, \cite{kohl:stone}. 
It has also been shown that gaits with regularity (coordination in leg movement, such as left-right symmetry or front-back symmetry) perform better than gaits without regularities \cite{valsalam:mii}, \cite{clune1}, \cite{clune3}.
However, in most previous works, the experimenter has had to manually decide the regularities of the gaits. 
Such manual intervention is time consuming, requires expert knowledge, and adds constraints that may hurt performance. 
Previous work has shown that the Hypercube-based NeuroEvolution of Augmenting Topologies (HyperNEAT) generative encoding \cite{stanley1} can automatically generate a variety of regular gaits and that it outperforms direct encoding controls on this task \cite{clune1}, \cite{clune2}. 
However, that work only verified these claims in simulation. 
A follow-up paper evolved gaits with HyperNEAT directly in hardware on the QuadraTot robot platform and found that HyperNEAT's gaits outperformed manually designed parameterized learning algorithms \cite{yos:clune}. % insert (figure 1 -- pic of quadratot) 
A further follow-up study \cite{glette} built a simulator for the QuadraTot robot platform to test whether the inclusion of a simulator would improve results and found that it did:  an evolutionary algorithm with direct encoding in simulation produced gaits that outperformed those gaits evolved by HyperNEAT in hardware. 
The simulator reduced the time required to evaluate gaits and afforded much larger population sizes and generations than was possible when evolving in hardware, ultimately performing 333 times as many evaluations as HyperNEAT had in hardware per run (60000 vs. 180) \cite{yos:clune}, \cite{glette}. 
The evolutionary algorithm used in Glette et al. \cite{glette} was manually constrained to evolve specific, hand-chosen regularities. 
Another study evolved gaits in hardware on the QuadraTot platform using RL PoWER, a spline-based algorithm, and was successful in evolving faster and more repeatable gaits than HyperNEAT evolved in hardware \cite{haocheng}.

In this paper we test whether the benefits of HyperNEAT -- higher performance and the automatic discovery of effective regularities \cite{clune2} -- will outperform the manually designed, regularity-constrained, direct encoding once HyperNEAT is combined with a simulator. 
We test this by evolving gaits using HyperNEAT using a physical simulator and then comparing the performance of these gaits in the real-world to those that were evaluated on the QuadraTot robotic platform in previous publications.  
%Our experiments confirm the hypothesis: gaits evolved with HyperNEAT and then transferred to reality were the highest performing observed to date on the QuadraTot platform. 
%That said, gaits still transferred poorly, revealing that substantial further performance gains are likely via future work that combines a simulator, HyperNEAT, and techniques that reduce the simulator-to-reality gap. 
%More broadly, this work supports previous work that evolving with a simulator and then transferring to reality is a more powerful means of achieving results than evolving directly in hardware. 
%Our work also confirms that generative encodings such as HyperNEAT are a powerful way to evolve gaits and, more generally, automatically exploit regularities in challenging engineering domains.

% figure of the quadratot in its natural stance
%\begin{figure}
%\begin{center}
%\vspace{1.5cm}
% just using a photo from Haocheng's paper! Should replace if needed be
%\epsfig{file=quadratot_fig.eps, width=11cm}
%\caption[ ]{The QuadraTot robotic platform on which experiments on this study was performed. It is open-sourced and 3-D printable, which has allowed for other researchers to use the robot for their research around the world. The robot has 9 DOF, 2 per leg and a hip joint which connects the two symmetrical halves of the robot. The robot uses Dynamixel AX-18 and AX-12 servos for actuation.}
%\label{fig:quadratot}
%\end{center}
%\end{figure}

### 2012-08-14 paper_sean/methods.tex
First draft of Sean's paper, text only

Gaits were first evolved in simulation, as previous done in Glette et al. \cite{glette}. 
During the evolution process, we evolved 200 organisms for 200 generations in 20 runs with different initial conditions specified by randomly seeding the run. 
To keep in line with previous studies that have been performed on the QuadraTot platform, each gait ran for 12 seconds. 
All the HyperNEAT settings used for gait generation was kept the same as in Yosinski et al. \cite{yos:clune} except for the frequency of the sine wave function used. 
Through experimentation, we found that high frequency gaits, gaits that require servos to change their directions often, were much more prone to servo shutdowns than low frequency gaits (at around 1.5Hz per servo). 
HyperNEAT often evolved high frequency gaits (over 8Hz per servo) which performed poorly in reality due to servo constraints. 
This led to gaits performing dramatically different in real-life compared to simulations. 
To address this issue, two changes were made:
First, the frequency of the sine wave function used in HyperNEAT was reduced by a factor of 6.63 by changing the sine function to

\begin{equation}
sin(2\pi t/250)\pi
\end{equation}

from

\begin{equation}
sin(t/6)\pi
\end{equation}

so that HyperNEAT would lean towards evolving low frequency gaits. Sine frequency of 0.004  was used because we experimentally found that HyperNEAT was able to produce natural and high-performing gaits with low frequencies with this frequency. 
Second, a policy that punished high frequency gaits was implemented during the evolution process. 
If the average frequency of each for the gait was over the threshold frequency, the fitness was punished by exponentially reducing the distance travelled by the robot:

\begin{equation}
distance = distance*e^{(freq_{thr}-freq)} \\
\end{equation}

Through experimentation, we chose the threshold frequency to be 1.67 Hz/servo. After implementing the frequency punishment policy, the servos shutdown much less frequently.

The fitness was evaluated using the following equation:

\begin{equation}
fitness = 2^{d^{2}}
\end{equation}

where \emph{d} is the distance the center of mass of the robot travelled during the given time as was done in Clune et al. \cite{clune1}.
The distance travelled was calculated as the Euclidean distance travelled by the robot in the XY plane during the alotted time, in centimeters.

In simulation, the distance travelled was calculated using the output file generated by the simulator.
In real-life, the distance travelled was calculated using the log file recorded using the Wiimote system described earlier in the paper. For real-life trials, the actual fitnesses were not calculated and the gaits were evaluated using only the velocity of the robot.

To evolve gaits using the physical simulator, we wrote a Python script\footnote{All scripts and drivers available at http://quadratot.yosinski.com} that interfaced between the simulator and HyperNEAT which allowed us to evolve gaits fast and efficiently with no human intervention.
The runs were performed on three separate machines all running the same version of Linux (Ubuntu 11.04) using the same HyperNEAT settings and files to speed up the simulation process.

After evolving the gaits in simulation, the champion gait of the last generation of each of the 20 runs was transferred onto the real robot and evaluated for its performance on the real robot.
In real-life trials, the motion of the robot was interpolated linearly between a stationary pose and the commanded gait for two seconds at the beginning and the end. % figure!! 
The position of the robot was measured at the beginning and end of each run, starting from the stationary pose and ending with the stationary pose. 
If a gait performed extremely poorly due to servo shutdowns, the gait was tried again until we could conclude that poor performance is because of the gait and not because of the condition of the robot.

### 2012-08-14 paper_sean/results.tex
First draft of Sean's paper, text only

HyperNEAT, used with a physical simulator, produced gaits that outperformed all previous gaits tried and recorded on the QuadraTot. 
The fastest gait evolved by HyperNEAT in simulation and transfered onto the real robot recorded a speed of 14.5 cm/s, faster than the previously fastest gait published in Glette et al. \cite{glette}. 
In their paper, Kyrre et al. \cite{glette} recorded 17.8 cm/s in their robot in Norway \cite{glette}, but to make sure that there were no differences in robots, the same gait was also tested on the robot used for our experimental testing.
The same gait, ran by using the gait file sent to us by Kyrre Glette, recorded an average speed of 12.951 cm/s and a maximum of 13.76 cm/s on the same QuadraTot platform in the Cornell Creative Machines Lab where Yosinski et al. \cite{yos:clune} and this study was done.
HyperNEAT, used with a simulator, outperformed Kyrre et al's \cite{glette} gait by 5.4\% with much fewer evaluations performed per run, at 40000 evaluations per run with 200 organisms and 200 generations as opposed to Kyrre et al's \cite{glette} 60000 evaluations per run with 200 organisms and 300 generations.
The best gait evolved using HyperNEAT with simulator performed 49.5\% better than the best gait evolved using HyperNEAT in hardware \cite{yos:clune}. 
This is most likely due to the difference in the number of evaluations between the two studies. 
The previous study, HyperNEAT in hardware, only had 180 evaluations while in this study, a simulator allowed a much higher number -- 20000 -- of evaluations. 
Another minor reason for this may be that the implementation placed to reduce gait frequency may have helped in producing gaits that were less susceptible to shutdowns, as Yosinski et al. \cite{yos:clune} did not have such things.
Although most gaits started off its evolution process by producing fast and high frequency gaits, with more generations, the gaits dramatically reduced its frequencies while continually increasing their fitness. The average frequency of the final gaits was 1.29 Hz per servo.

Gaits produced in this study confirmed that with the use of a physical simulator, HyperNEAT is able to fully exploit the geometry of the problem without human intervention, which was not evident in gait evolution using HyperNEAT in hardware \cite{yos:clune}. 
HyperNEAT took full advantage of the regularity of the problem and produced gaits more regular, coordinated, and effective than all other gaits evolved for the QuadraTot platform in the past.
This is evident in the servo position plots where the gaits produced in this study have smooth and repeating curves while gaits produced by HyperNEAT in hardware produced much more ragged and random curves.
The difference in regularity between the two studies is most likely due to the number of evaluations that were done in gait evolution. Additionally, the noise in the real world was extremely high, making learning in hardware difficult, which may have further impeded the search for higher-performing, regular gaits.

While this study was able to produce the highest fitness gait that has ever been tried on the QuadraTot, it is worth noting that most of the gaits in simulation did not transfer well to reality. 
Many gaits that performed very well in simulation performed poorly on the real robot because of the physical limitations and differences between simulation and real-world, such as limited servo power, frictional differences, and other minor differences in the robot physique, all of which accumulates to have a large impact in performance.
The most oft-seen problem with real trials, regardless of the actions taken to reduce it, was the servo shutdowns. 
Because we limited gait frequencies, HyperNEAT was forced to turn to other means for maximizing distance. 
So it pushed the servo positions to their limits by extending and retracting from one extreme to the other at a pace which was difficult for the servos to handle. 
Another reason for the shutdowns was that the servos could not easily support the robot's weight well, which often led to the servos shutting down after a few seconds of walking off-ground.

These problems led to difficulties in repeatability. 
The robot is weak and the power of the motors changes between and even within trials. 
And since the gaits evolved in this study stressed the servos a lot, after one successful run, there was no guarantee that the robot would be able to perform that same gait succesfully next time. 
In that sense, the gaits produced in this study is perhaps worse than the gaits produced by RL PoWER Spline \cite{haocheng} or by genetic algorithm with a physical simulator \cite{glette} whose gaits were repeatable and less taxing on the robot.

% table!
\begin{table}
\caption{The number of evaluations per run (gen x orgs), simulated and real trial velocities in [cm/s]. Data used from Yosinksi et al. \cite{yos:clune}, Glette et al. \cite{glette}, and Shen et al. \cite{haocheng}.
*17.8 cm/s was recorded on Glette et al. \cite{glette} but the same gait tried on our robot averaged over ten runs was 12.951 cm/s}  %\tablabel{results}
\begin{center}
\begin{tabular}{|l|c|c|c|c|}
\hline
                                         & Num. Evaluations  & Sim. Velocity  & Real. Velocity \\
\hline
Best parameter optimization learning    &153    & --    & 4.64 \\
\hline
Hardware HyperNEAT                  & 180         & --         &   9.7     \\
\hline
Simulated GA              & 60000       & 16.4       &   12.951*     \\
\hline
RL PoWER Spline                          & 300         & --         &   11.05 \\
\hline
Simulated HyperNEAT                      & 40000       & 25.4       &   14.5 \\
\hline
\end{tabular}
\end{center}
\end{table}

%
%

\begin{figure}
\begin{center}
\vspace{1cm}
\epsfig{file=fit_gen.eps, width=8cm}
\caption[ ]{Fitnesses of the champion gaits averaged over 20 runs, measured in distance travelled over 12 seconds [cm/s]. These gaits outperformed gaits evolved in simulation using Genetic Algorithm by 54.3\% in simulation and 5.4\% better in reality while outperforming gaits evolved using HyperNEAT in hardware by 49.5\%.}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\vspace{1cm}
\epsfig{file=freq_fitness.eps, width=8cm}
\caption[ ]{Average frequencies of the gaits averaged over 20 runs. After implementing a policy which punished gaits with high frequencies, HyperNEAT was able to find high-performing, low frequency gaits. This policy was implemented to reduce servo shutdowns which heavily affected the reality performance of the robot.}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\vspace{1cm}
\epsfig{file=servo_pos.eps, width=8cm}
\caption[ ]{Top: A plot of servo positions for a gait produced in this study. Bottom: A plot of servo positions for a gait produced by HyperNEAT in hardware. Figure taken from Yosinski et al. \cite{yos:clune}. The smoothness, symmetry, and regularities of the top plot shows that by using a simulator, HyperNEAT was able to fully exploit the geometry of the problem.}
\end{center}
\end{figure}

### 2012-08-14 paper_sean/robot.tex
First draft of Sean's paper, text only

The robot platform used for this project was a quadrupedal robot named the QuadraTot\footnote{Robot available at http://quadratot.yosinski.com} designed in the Cornell Creative Machines Lab by Juan Zagal.
It has 9 degrees-of-freedom (DOF), with 2 DOF for the inner joint and the outer joint of each leg and 1 DOF for the hip joint of the robot. 
The QuadraTot is an open-sourced and 3-D printable robot, which allows it to be used in other labs for research, providing a ground on which people could compare the results of different gait learning algorithms. 
Currently, there are published data and results from 9 different learning algorithms on the platform.

The body of the robot was printed using an Objet Connex 500 3-D Printing System. 
It was actuated using 9 Robotis Dynamixel servos: 5 AX-18A servos and 4 AX-12A servos. 
The stronger AX-18A servos were used to actuate the inner joints of the robot and the hip joint while the AX-12A servos were used to actuate the outer joints. 
The servos were attached to the robot using off-the-shelf components and parts.
The servos have a range of motion of [0, 1023], roughly equivalent to a physical range of [-12\degree, +120\degree]. 
The Dynamixel AX-12A and AX-18A servos have a safety feature called \emph{Alarm Shutdown} which would turn off the servo if the servo's current, range, temperature, or torque was too high \cite{robotis}. 
The ones most relevant to this study were shutdowns due to high torque and range. 
Often, many of the gaits that were tested on the QuadraTot forced the servos to exert too much torque or would command positions which was out of range for the robot \cite{yos:clune} and cause the servos to shutdown.

In order to address this problem and avoid collisions with its own body, a motion cropping system was implemented on the QuadraTot. 
In addition to placing limits on the range of motions for the inner joints, outer joints, and the hip joint by [-85\degree, +60\degree], [-113\degree, +39\degree], and [-28\degree, +28\degree] respectively, an extra cropping system named the smart cropping system was adopted from Shen et al. \cite{haocheng} to further prevent servo shutdowns from fierce gaits -- gaits that commanded servo positions to either extremes. 
A leg position is defined by the sum of servo position values from the inner and outer joints of a leg, \emph{\textbf{g}} (\emph{\textbf{g}} = $j_{inner}$ + $j_{outer}$) as represented in Haocheng et al. \cite{haocheng}.
Smart cropping works by modifying the servo positions from fierce gaits, which usually have relatively low \emph{\textbf{g}}s, to remap them onto the boundary of a quasi-trangular area in a 2-D space defined by $j_{inner}$ and $j_{outer}$.
Through experimentation, we found that a \emph{\textbf{g}} of 730 provides a good balance between performance and safety.

The robot was controlled by CompuLab Fit-PC2, a compact computer running Ubuntu Linux 10.10. 
In previous study \cite{yos:clune}, this computer was placed on the robot to keep the robot untethered but for this study the computer was taken off the robot to reduce weight, because the weight of the robot seemed to be one of the factors contributing to servo shutdowns. 
The servos were commanded positions using the Pydynamixel library at 40Hz from the computer.

We tracked the motion of the robot using infrared (IR) LEDs and a Wiimote (from the Nintendo Wii gaming system).
Three IR LEDs were placed at the center of the QuadraTot for the Wiimote to detect. 
A Wiimote contains an IR camera which could track positions of IR light. 
For this study, the Wiimote was velcroed onto the ceiling pointing down to maximize its viewable window. 
The Wiimote communicated with the computer tethered to the robot via bluetooth using the CWiid library and a separate Python script recorded the position of the robot reported by the Wiimote during the duration of the run.

% figures
% figure of the quadratot in its natural stance
\begin{figure}
\begin{center}
\vspace{1cm}
% just using a photo from Haocheng's paper! Should replace if needed be           
\epsfig{file=quadratot_fig.eps, width=8cm}
\caption[ ]{The QuadraTot robotic platform. It is open-sourced and 3-D printable, which has allowed for other researchers to use the robot for their research around the world. The robot has 9 DOF, 2 per leg and a hip joint which connects the two symmetrical halves of the robot. The robot uses Dynamixel AX-18 and AX-12 servos for actuation. Figure taken from Shen et al. \cite{haocheng}.}
\label{fig:quadratot}
\end{center}
\end{figure}
%

\begin{figure}
\begin{center}
\vspace{1cm}
\epsfig{file=legpos.eps, width=8cm}
\caption[ ]{The figure shows an example of leg positions allowed and not allowed on the robot. The twoleft positions are allowed. The middle two positions are on the borderline. The right two positions are considered as ``fierce'' and are disallowed. Instead, these gaits are cropped to the middle two positions during gait performance by the cropping mechanism placed to prevent servo shutdowns. Figure taken from Shen et al. \cite{haocheng}}
\end{center}
\end{figure}
%

\begin{figure}
\begin{center}
\vspace{1cm}
\epsfig{file=gaitspace.eps, width=8cm}
\caption[ ]{A map of possible leg positions. Manually placed position cropping systems limited possible leg positions to avoid servo shutdowns and the robot hitting itself. The sum of the inner and outer servo positions cannot drop below 730, the smart cropping border, marked by the red line. Figure taken from Shen et al. \cite{haocheng}.}
\end{center}
\end{figure}
%

\begin{figure}
\begin{center}
\vspace{1cm}
\epsfig{file=wiimote.eps, width=8cm}
\caption[ ]{The Wiimote used to track the motion of ther robot. The Wiimote was used to track the IR LED placed on top of the robot and send the location data to a nearby computer which recorded the position of ther robot with respect to time. Figure taken from Yosinski et al. \cite{yos:clune}.}
\end{center}
\end{figure}

### 2012-08-14 paper_sean/simulator.tex
First draft of Sean's paper, text only

The simulations were done using a simulator\footnote{Simulator available for Linux and Windows in headless and visual versions at http://quadratot.yosinski.com} designed and written by Kyrre Glette and Jim Torresen in the University of Oslo, Norway\footnote{{kyrrehg, jimtoer}@ifi.uio.no} \cite{glette}. 
It was written in C++ using the NVIDIA PhysX physics software library to provide an accurate physical representation of the behaviors of the robot. 
A physical representation of the QuadraTot was built in simulation to capture the most important parts of the robot like the number and type of joints, mass, and rough shapes and lengths of the robot. 
The simulator was visualized by OpenGL library to allow the user to have a more intuitive grasp of what is going on. 
The simulator visualization included a recording feature which allows the user to create videos of the robot in simulation.
The simulator also features a command line function which allows the users to input a gait file and run it on the simulator. 
The simulator then outputs a file containing the X, Y, Z coordinates of the center of mass of the robot over time. 
In simulation, the robot traverses in the XY plane with the Z axis representing the height off the ground. 
In an attempt to more accurately model the robot behavior, we implemented the same cropping function in the simulator as in the actual robot. 
Although the simulator was originally written in Windows for the Windows operating platform we ported and ran it on Linux (Ubuntu 11.04) to seamlessly link the simulator and the already exising libraries together. 
The friction values used for this study was the same as that used in Glette et al. \cite{glette}. 
%
% and uses OpenGL and GLFW libraries for visualization. The simula- tor was originally written for the Windows operating system but was ported and run on Linux (Ubuntu 11.04) for this ex- periment. It features a commandline function which allows the user to specify a gait input file and extract an output file with the X, Y, Z coordinate information over time. The Eu- clidean distance travelled by the robot in the XY plane was used to calculate the speed of the robot in simulation and was used as the fitness value used for evolution.
%

\begin{figure}
\begin{center}
\vspace{1cm}
\epsfig{file=simulator.eps, width=8cm}
\caption[ ]{Physical representation of the QuadraTot in simulation. It captures the important parts of the robot, such as the number and types of the joints, mass, and the rough shape and lengths. The simulator was written by Kyrre Glette in C++ using NVIDIA PhysX software library and visualized using OpenGL. Figure taken from Glette et al. \cite{glette}.}
\end{center}
\end{figure}

### 2012-08-14 paper_sean/results.tex
Updated figure commands, added as many figures as possible.

@@
 
-\begin{figure}
-\begin{center}
-\vspace{1cm}
-\epsfig{file=fit_gen.eps, width=8cm}
-\caption[ ]{Fitnesses of the champion gaits averaged over 20 runs, measured in distance travelled over 12 seconds [cm/s].
+% \begin{figure}
+% \begin{center}
+% \vspace{1cm}
+% \epsfig{file=fit_gen.eps, width=8cm}
+% \caption[ ]{Fitnesses of the champion gaits averaged over 20 runs, measured in distance travelled over 12 seconds [cm/s].
 These gaits outperformed gaits evolved in simulation using Genetic Algorithm by 54.3\% in simulation and 5.4\% better in reality while outperforming gaits evolved using HyperNEAT in hardware by 49.5\%.}
-\end{center}
-\end{figure}
+% \end{center}
+% \end{figure}
+\figp{aracna_notYetCleanedMultiPieceRobot_edit}{.6}{Fitnesses of the champion gaits averaged over 20 runs, measured in distance travelled over 12 seconds [cm/s].
+These gaits outperformed gaits evolved in simulation using Genetic Algorithm by 54.3\% in simulation and 5.4\% better in reality while outperforming gaits evolved using HyperNEAT in hardware by 49.5\%.
+\edit{wrong fig, regenerate as pdf and add.}}
 
-\begin{figure}
-\begin{center}
-\vspace{1cm}
-\epsfig{file=freq_fitness.eps, width=8cm}
-\caption[ ]{Average frequencies of the gaits averaged over 20 runs.
+% \begin{figure}
+% \begin{center}
+% \vspace{1cm}
+% \epsfig{file=freq_fitness.eps, width=8cm}
+% \caption[ ]{Average frequencies of the gaits averaged over 20 runs.
 After implementing a policy which punished gaits with high frequencies, HyperNEAT was able to find high-performing, low frequency gaits.
 This policy was implemented to reduce servo shutdowns which heavily affected the reality performance of the robot.}
-\end{center}
-\end{figure}
+% \end{center}
+% \end{figure}
+\figp{aracna_notYetCleanedMultiPieceRobot_edit}{.6}{Average frequencies of the gaits averaged over 20 runs.
+After implementing a policy which punished gaits with high frequencies, HyperNEAT was able to find high-performing, low frequency gaits.
+This policy was implemented to reduce servo shutdowns which heavily affected the reality performance of the robot.
+\edit{wrong fig, regenerate as pdf and add.}}
 
-\begin{figure}
-\begin{center}
-\vspace{1cm}
-\epsfig{file=servo_pos.eps, width=8cm}
-\caption[ ]{Top: A plot of servo positions for a gait produced in this study.
+% \begin{figure}
+% \begin{center}
+% \vspace{1cm}
+% \epsfig{file=servo_pos.eps, width=8cm}
+% \caption[ ]{Top: A plot of servo positions for a gait produced in this study.
 Bottom: A plot of servo positions for a gait produced by HyperNEAT in hardware.
@@
 The smoothness, symmetry, and regularities of the top plot shows that by using a simulator, HyperNEAT was able to fully exploit the geometry of the problem.}
-\end{center}
-\end{figure}
+% \end{center}
+% \end{figure}
+\figgp{aracna_notYetCleanedMultiPieceRobot_edit}{.6}{neat_110115_211410_00000_002_filt_zoom}{.6}{Top: A plot of servo positions for a gait produced in this study.
+Bottom: A plot of servo positions for a gait produced by HyperNEAT in hardware.
+Figure taken from Yosinski et al.
+\cite{yos:clune}.
+The smoothness, symmetry, and regularities of the top plot shows that by using a simulator, HyperNEAT was able to fully exploit the geometry of the problem.
+\edit{wrong fig on top, regenerate top fig as pdf and add.}}

### 2012-08-14 paper_sean/robot.tex
Updated figure commands, added as many figures as possible.

@@
 
-% figures
-% figure of the quadratot in its natural stance
-\begin{figure}
-\begin{center}
-\vspace{1cm}
-% just using a photo from Haocheng's paper! Should replace if needed be           
-\epsfig{file=quadratot_fig.eps, width=8cm}
-\caption[ ]{The QuadraTot robotic platform.
+% % figures
+% % figure of the quadratot in its natural stance
+% \begin{figure}
+% \begin{center}
+% \vspace{1cm}
+% % just using a photo from Haocheng's paper! Should replace if needed be           
+% \epsfig{file=quadratot_fig.eps, width=8cm}
+% \caption[ ]{The QuadraTot robotic platform.
 It is open-sourced and 3-D printable, which has allowed for other researchers to use the robot for their research around the world.
@@
 \cite{haocheng}.}
-\label{fig:quadratot}
-\end{center}
-\end{figure}
-%
+% \label{fig:quadratot}
+% \end{center}
+% \end{figure}
+% %
+\figp{robot_whitebg}{.6}{The QuadraTot robotic platform.
+It is open-sourced and 3-D printable, which has allowed for other  researchers to use the robot for their research around the  world.
+The robot has 9 DOF, 2 per leg and a hip joint which connects  the two symmetrical halves of the robot.
+The robot uses Dynamixel  AX-18 and AX-12 servos for actuation.
+Figure taken from Shen et  al.
+\cite{haocheng}.}
 
-\begin{figure}
-\begin{center}
-\vspace{1cm}
-\epsfig{file=legpos.eps, width=8cm}
-\caption[ ]{The figure shows an example of leg positions allowed and not allowed on the robot.
+% \begin{figure}
+% \begin{center}
+% \vspace{1cm}
+% \epsfig{file=legpos.eps, width=8cm}
+% \caption[ ]{The figure shows an example of leg positions allowed and not allowed on the robot.
 The twoleft positions are allowed.
@@
 \cite{haocheng}}
-\end{center}
-\end{figure}
-%
+% \end{center}
+% \end{figure}
+\figp{sixLegPositions}{.6}{The figure shows an example of leg positions allowed and not allowed on the robot.
+The twoleft positions are allowed.
+The middle two positions are on the borderline.
+The right two positions are considered as ``fierce'' and are disallowed.
+Instead, these gaits are cropped to the middle two positions during gait performance by the cropping mechanism placed to prevent servo shutdowns.
+Figure taken from Shen et al.
+\cite{haocheng}}
 
-\begin{figure}
-\begin{center}
-\vspace{1cm}
-\epsfig{file=gaitspace.eps, width=8cm}
-\caption[ ]{A map of possible leg positions.
+% \begin{figure}
+% \begin{center}
+% \vspace{1cm}
+% \epsfig{file=gaitspace.eps, width=8cm}
+% \caption[ ]{A map of possible leg positions.
 Manually placed position cropping systems limited possible leg positions to avoid servo shutdowns and the robot hitting itself.
@@
 \cite{haocheng}.}
-\end{center}
-\end{figure}
-%
+% \end{center}
+% \end{figure}
+\figp{smartCrop}{.6}{A map of possible leg positions.
+Manually placed position cropping systems limited possible leg positions to avoid servo shutdowns and the robot hitting itself.
+The sum of the inner and outer servo positions cannot drop below 730, the smart cropping border, marked by the red line.
+Figure taken from Shen et al.
+\cite{haocheng}.}
 
-\begin{figure}
-\begin{center}
-\vspace{1cm}
-\epsfig{file=wiimote.eps, width=8cm}
-\caption[ ]{The Wiimote used to track the motion of ther robot.
+% \begin{figure}
+% \begin{center}
+% \vspace{1cm}
+% \epsfig{file=wiimote.eps, width=8cm}
+% \caption[ ]{The Wiimote used to track the motion of ther robot.
 The Wiimote was used to track the IR LED placed on top of the robot and send the location data to a nearby computer which recorded the position of ther robot with respect to time.
@@
 \cite{yos:clune}.}
-\end{center}
-\end{figure}
+% \end{center}
+% \end{figure}
+\figgp{wiiMote_crop.jpg}{.6}{robot_led_crop.jpg}{.6}{The Wiimote used to track the motion of ther robot.
+The Wiimote was used to track the IR LED placed on top of the robot and send the location data to a nearby computer which recorded the position of ther robot with respect to time.
+Figure taken from Yosinski et al.
+\cite{yos:clune}.}

### 2012-08-21 paper_sean/abstract.tex
JMC: reworked abstract...also some edits to intro:

@@
-Evolving gaits for physical robots is a longstanding and still open challenge. 
-While previous work has shown that evolutionary algorithms can outperform human engineers in generating gaits, in all of that work the experimenter has had to manually decide the regularities of the gaits.
+% !TEX root =main.tex
+Creating gaits for physical robots is a longstanding and still open challenge. 
+%Evolutionary algorithms can outperform human engineers in generating such gaits, but this approach historically required engineers to manually decide which types of gaits would evolve.
 Such manual intervention is time consuming, requires expert knowledge, and adds constraints that may hurt performance.
-In this paper, we tested whether HyperNEAT -- a generative encoding that has shown to be able to automatically evolve a variety of high-performing regular gaits -- could generate gaits that could outperform manually designed, regularity-constrained, directly-encoded gaits.
-For our experiments, gaits were evolved in simulation and the final gait transferred to the real-world to evaluate the gait's actual performance.
-The platform used was the QuadraTot, an open-source, 3D printed quadrupedal robot on which previous studies in gait evolution has already been done several times. 
-In this study, we produced the fastest gait ever tried on this platform, outperforming the fastest published gait by 5.4\% with 20000 fewer evaluations run during gait generation. 
-The gaits produced showed exploitation in regularity in simulation and real-world as was expected.
+Recently, the HyperNEAT generative encoding was shown to automatically discover a variety of gait regularities, producing fast, coordinated gaits, but only for simulated robots.
+A follow-up study found that HyperNEAT did not produce impressive gaits when they evolved directly on a physical robot.
+A simpler encoding hand-tuned to produce regular gaits was tried on the same robot, and outperformed HyperNEAT, but these gaits were first evolved in simulation before being transferred to the robot.
+In this paper, we tested the hypothesis that the beneficial properties of HyperNEAT would outperform the simpler encoding if HyperNEAT gaits are first evolved in simulation before being transferred to reality.
+That hypothesis was confirmed, resulting in the fasted gaits yet observed for this robot.
+This result is important because it confirms that the early promise shown by generative encodings, specifically HyperNEAT, were not limited to evolving gaits in simulation, but indeed work on real robots.
+
+%We further test whether gaits created by generative encodings such as HyperNEAT tend to be transferable from simulation to reality, which is an important question because the regularity-producing property of generative encodings should make the gap smaller.
+could generate gaits that could outperform manually designed, regularity-constrained, directly-encoded gaits.
+%
+%For our experiments, gaits were evolved in simulation and the final gait transferred to the real-world to evaluate the gait's actual performance.
+%The platform used was the QuadraTot, an open-source, 3D printed quadrupedal robot on which previous studies in gait evolution has already been done several times. 
+%In this study, we produced the fastest gait ever tried on this platform, outperforming the fastest published gait by 5.4\% with 20000 fewer evaluations run during gait generation. 
+%The gaits produced showed exploitation in regularity in simulation and real-world as was expected.

### 2012-08-21 paper_sean/introduction.tex
JMC: reworked abstract...also some edits to intro:

@@
+% !TEX root =main.tex
 Legged locomotion provides several benefits over wheeled locomotion. 
@@
 But designing a gait for a robot is often difficult and time consuming because the engineer must take into consideration many factors which must all fall into place for a successful gait pattern. 
-Because of setbacks present in hand-designing gaits, evolving gaits using machine learning algorithms is an attractive and flexible approach that could be used for many different robot platforms with only a few tweaks.
+Because of setbacks present in hand-designing gaits, evolving gaits with machine learning algorithms is an attractive and flexible approach that can automatically produce effective gaits for many different robot platforms with minimal human intervention.
 

### 2012-08-21 paper_sean/introduction.tex
JMC edits

@@
 Much research has been done in this subject, and previous work has shown that learned gaits can outperform designed gaits \cite{valsalam:mii}, \cite{kohl:stone}.
-It has also been shown that gaits with regularity (coordination in leg movement, such as left-right symmetry or front-back symmetry) perform better than gaits without regularities \cite{valsalam:mii}, \cite{clune1}, \cite{clune3}.
-However, in most previous works, the experimenter has had to manually decide the regularities of the gaits. 
+%JMC - I have a similar claim in my CEC 2009 paper.
+Grab those cites for here (especially the Hornby ones).
+It has also been shown that gaits with regularity (coordination in leg movement, such as left-right symmetry or front-back symmetry) perform better than gaits without regularities 
+\cite{valsalam:mii,clune2,clune1,clune3}. %JMC-note how you can put multiple cites in one bracket...but no spaces after the comma!
+However, in most previous works, the experimenter has had to manually decide the regularities of the gaits. %we need cites for this claim. you can get them from Clune1
 Such manual intervention is time consuming, requires expert knowledge, and adds constraints that may hurt performance. 
@@
 However, that work only verified these claims in simulation. 
-A follow-up paper evolved gaits with HyperNEAT directly in hardware on the QuadraTot robot platform and found that HyperNEAT's gaits outperformed manually designed parameterized learning algorithms \cite{yos:clune}.
+A follow-up paper by Glette et al.
+evolved gaits with HyperNEAT directly in hardware on the QuadraTot robot platform and found that HyperNEAT's gaits outperformed manually designed parameterized learning algorithms, but still did not produce impressive, natural gaits \cite{yos:clune}.
 % insert (figure 1 -- pic of quadratot)
-A further follow-up study \cite{glette} built a simulator for the QuadraTot robot platform to test whether the inclusion of a simulator would improve results and found that it did:  an evolutionary algorithm with direct encoding in simulation produced gaits that outperformed those gaits evolved by HyperNEAT in hardware.
-The simulator reduced the time required to evaluate gaits and afforded much larger population sizes and generations than was possible when evolving in hardware, ultimately performing 333 times as many evaluations as HyperNEAT had in hardware per run (60000 vs.
+A further follow-up study built a simulator for the QuadraTot robot platform to test whether the inclusion of a simulator would improve results and found that it did:  an evolutionary algorithm with direct encoding in simulation produced gaits that outperformed those gaits evolved by HyperNEAT in hardware~\cite{glette}.
+%JMC-Should there be a space between the end of the sentence and the cite?
+Whatever the answer, we should be consistent.
+If so, I have added here with the ~.
+The ~ tells Latex not to split that space from the character before and after it.
+The simulator reduced the time required to evaluate gaits and afforded much larger population sizes and more generations than was possible when evolving in hardware, ultimately performing 333 times as many evaluations as HyperNEAT had in hardware per run (60000 vs.
 180) \cite{yos:clune}, \cite{glette}.
 The evolutionary algorithm used in Glette et al. \cite{glette} was manually constrained to evolve specific, hand-chosen regularities. 
-Another study evolved gaits in hardware on the QuadraTot platform using RL PoWER, a spline-based algorithm, and was successful in evolving faster and more repeatable gaits than HyperNEAT evolved in hardware \cite{haocheng}.
+%Another study evolved gaits in hardware on the QuadraTot platform using RL PoWER, a spline-based algorithm, and was successful in evolving faster and more repeatable gaits than HyperNEAT evolved in hardware \cite{haocheng}.
+%JMC- I don't think we need to cite this paper...it is a non-evolutionary method, and it doesn't outperform Glette...and the sentence just sticks out as a detail that isn't necessary to our story.
 
-In this paper we test whether the benefits of HyperNEAT -- higher performance and the automatic discovery of effective regularities \cite{clune2} -- will outperform the manually designed, regularity-constrained, direct encoding once HyperNEAT is combined with a simulator.
+In this paper we test whether the benefits of HyperNEAT -- higher performance and the automatic discovery of complex, effective regularities \cite{clune2,yos:clune} -- will outperform the manually designed, regularity-constrained, direct encoding once HyperNEAT is combined with a simulator.
+%JMC-I don't think they used a direct encoding.
+I prefer describing it as simple and manually designed to produce specific regularities.
 We test this by evolving gaits using HyperNEAT using a physical simulator and then comparing the performance of these gaits in the real-world to those that were evaluated on the QuadraTot robotic platform in previous publications.
+%JMC - Try to write without ever saying ``use'' or ``using.'' If you do, you'll find how much it is overused in writing, as evidenced by this last sentence.
+Often the word is a stand in for a different, clearer verb...and its use is ambiguous.
+Please search for this word and try to swap it out throughout the rest of the paper.
+Often times it requires restructuring a sentence, but the sentence usually ends up better.
+Similarly, make sure to always follow the word ``this'' with referent.
+This is extremely ambiguous.
+You know what you refer to when you write the sentence, but readers (often including your future self!) do not know precisely what you mean, and in science precision is critical.
 %Our experiments confirm the hypothesis: gaits evolved with HyperNEAT and then transferred to reality were the highest performing observed to date on the QuadraTot platform. 

### 2012-08-21 paper_sean/robot.tex
JMC edits

@@
-The robot platform used for this project was a quadrupedal robot named the QuadraTot\footnote{Robot available at http://quadratot.yosinski.com} designed in the Cornell Creative Machines Lab by Juan Zagal.
+% !TEX root =main.tex
+The robot platform used for this project was a quadrupedal robot named the QuadraTot\footnote{Robot plans and part files for 3D printing are freely available at http://quadratot.yosinski.com} designed in the Cornell Creative Machines Lab by Juan Zagal~\cite{yos:clune}.
 It has 9 degrees-of-freedom (DOF), with 2 DOF for the inner joint and the outer joint of each leg and 1 DOF for the hip joint of the robot. 
-The QuadraTot is an open-sourced and 3-D printable robot, which allows it to be used in other labs for research, providing a ground on which people could compare the results of different gait learning algorithms.
-Currently, there are published data and results from 9 different learning algorithms on the platform.
+The QuadraTot is an open-sourced and 3-D printable robot, which allows it to be used in other labs for research, providing a platform with which people could compare the results of different gait learning algorithms.
+Currently, there are published data and results from 9 different learning algorithms on the platform from INSERT different publications.
+%JMC fill in INSERT, and dump all cites here.
+This is a good place to cite haochengs paper, which is fun for him and Jason, but does not distract from our story.
 
@@
 The servos have a range of motion of [0, 1023], roughly equivalent to a physical range of [-12\degree, +120\degree].
-The Dynamixel AX-12A and AX-18A servos have a safety feature called \emph{Alarm Shutdown} which would turn off the servo if the servo's current, range, temperature, or torque was too high \cite{robotis}.
+
+The Dynamixel AX-12A and AX-18A servos have a safety feature called \emph{Alarm Shutdown} that would turn off the servo if the servo's current, range, temperature, or torque was too high \cite{robotis}.
 The ones most relevant to this study were shutdowns due to high torque and range. 
-Often, many of the gaits that were tested on the QuadraTot forced the servos to exert too much torque or would command positions which was out of range for the robot \cite{yos:clune} and cause the servos to shutdown.
-
+As has been reported previously with QuadraTot\cite{yos:clune}, many of the gaits that were tested on the QuadraTot forced the servos to exert too much torque or would command out of range positions and cause the servos to shutdown.
 In order to address this problem and avoid collisions with its own body, a motion cropping system was implemented on the QuadraTot.
-In addition to placing limits on the range of motions for the inner joints, outer joints, and the hip joint by [-85\degree, +60\degree], [-113\degree, +39\degree], and [-28\degree, +28\degree] respectively, an extra cropping system named the smart cropping system was adopted from Shen et al.
-\cite{haocheng} to further prevent servo shutdowns from fierce gaits -- gaits that commanded servo positions to either extremes.
-A leg position is defined by the sum of servo position values from the inner and outer joints of a leg, \emph{\textbf{g}} (\emph{\textbf{g}} = $j_{inner}$ + $j_{outer}$) as represented in Haocheng et al.
-\cite{haocheng}.
-Smart cropping works by modifying the servo positions from fierce gaits, which usually have relatively low \emph{\textbf{g}}s, to remap them onto the boundary of a quasi-trangular area in a 2-D space defined by $j_{inner}$ and $j_{outer}$.
+In addition to placing limits on the range of motions for the inner joints, outer joints, and the hip joint of [-85\degree, +60\degree], [-113\degree, +39\degree], and [-28\degree, +28\degree] respectively, an extra cropping system named the smart cropping system was adopted from Shen et al.
+\cite{haocheng} to further prevent servo shutdowns from leg positions where the combination of the two angles interact to produce an extreme amount of torque.
+This smart cropping algorithm calculates a leg position by summing the servo position values from the inner and outer joints of a leg, \emph{\textbf{g}} (\emph{\textbf{g}} = $j_{inner}$ + $j_{outer}$) %JMC - do they sum the position values, or the joint angles?
+Since all the previous extremes are expressed in angles, the writing makes me expect angles...
+and modifying the servo positions from fierce gaits, which usually have relatively low \emph{\textbf{g}}s, to remap them onto the boundary of a quasi-trangular area in a 2-D space defined by $j_{inner}$ and $j_{outer}$.
+%the term fierce sounds silly and non-professional (sorry!) I took out the place where you defined it, and we only use it in two other places.
+No need to invent a definition only to use it 2-3 times...can you swap in different, more descriptive words?
 Through experimentation, we found that a \emph{\textbf{g}} of 730 provides a good balance between performance and safety.
+%JMC - This description is not great.
+The last sentence's use of g doesn't make sense with the equation...is g a threshold or the sum?
+Please try to clean up this description.
 
 The robot was controlled by CompuLab Fit-PC2, a compact computer running Ubuntu Linux 10.10. 
-In previous study \cite{yos:clune}, this computer was placed on the robot to keep the robot untethered but for this study the computer was taken off the robot to reduce weight, because the weight of the robot seemed to be one of the factors contributing to servo shutdowns.
+In previous study \cite{yos:clune}, this computer was placed on the robot to keep the robot untethered, but for this study the computer was taken off the robot to reduce weight, because the weight of the robot seemed to be one of the factors contributing to servo shutdowns.
 The servos were commanded positions using the Pydynamixel library at 40Hz from the computer.
@@
 For this study, the Wiimote was velcroed onto the ceiling pointing down to maximize its viewable window. 
-The Wiimote communicated with the computer tethered to the robot via bluetooth using the CWiid library and a separate Python script recorded the position of the robot reported by the Wiimote during the duration of the run.
+The Wiimote communicated with the computer tethered to the robot via bluetooth using the CWiid library and a separate Python script recorded the position of the robot reported by the Wiimote during the duration of the run\~cite{yos:clune}.
 

### 2012-08-21 paper_sean/simulator.tex
Updated figure commands, added as many figures as possible.
JMC edits

@@
-The simulations were done using a simulator\footnote{Simulator available for Linux and Windows in headless and visual versions at http://quadratot.yosinski.com} designed and written by Kyrre Glette and Jim Torresen in the University of Oslo, Norway\footnote{{kyrrehg, jimtoer}@ifi.uio.no} \cite{glette}.
+% !TEX root =main.tex
+The simulations were done using a simulator\footnote{Available for Linux and Windows at http://quadratot.yosinski.com} designed and written by Kyrre Glette and Jim Torresen in the University of Oslo, Norway\footnote{{kyrrehg, jimtoer}@ifi.uio.no} \cite{glette}.
 It was written in C++ using the NVIDIA PhysX physics software library to provide an accurate physical representation of the behaviors of the robot. 
@@
 
-\begin{figure}
-\begin{center}
-\vspace{1cm}
-\epsfig{file=simulator.eps, width=8cm}
-\caption[ ]{Physical representation of the QuadraTot in simulation.
+% \begin{figure}
+% \begin{center}
+% \vspace{1cm}
+% \epsfig{file=simulator.eps, width=8cm}
+% \caption[ ]{Physical representation of the QuadraTot in simulation.
 It captures the important parts of the robot, such as the number and types of the joints, mass, and the rough shape and lengths.
@@
 \cite{glette}.}
-\end{center}
-\end{figure}
+% \end{center}
+% \end{figure}
+\figp{simulator_outline}{.6}{Physical representation of the QuadraTot in simulation.
+It captures the important parts of the robot, such as the number and types of the joints, mass, and the rough shape and lengths.
+The simulator was written by Kyrre Glette in C++ using NVIDIA PhysX software library and visualized using OpenGL.
+Figure taken from Glette et al.
+\cite{glette}.}

### 2012-08-24 paper_sean/robot.tex
jmc edits

@@
 The robot was controlled by CompuLab Fit-PC2, a compact computer running Ubuntu Linux 10.10. 
-In previous study \cite{yos:clune}, this computer was placed on the robot to keep the robot untethered, but for this study the computer was taken off the robot to reduce weight, because the weight of the robot seemed to be one of the factors contributing to servo shutdowns.
-The servos were commanded positions using the Pydynamixel library at 40Hz from the computer.
+In a previous study \cite{yos:clune}, this computer was placed on the robot to keep the robot untethered, but for this study the computer was taken off the robot to reduce weight, because the weight of the robot seemed to be one of the factors contributing to servo shutdowns.
+The servos were sent positions from the computer at 40Hz via the Pydynamixel library.
 
-We tracked the motion of the robot using infrared (IR) LEDs and a Wiimote (from the Nintendo Wii gaming system).
-Three IR LEDs were placed at the center of the QuadraTot for the Wiimote to detect. 
-A Wiimote contains an IR camera which could track positions of IR light. 
-For this study, the Wiimote was velcroed onto the ceiling pointing down to maximize its viewable window. 
-The Wiimote communicated with the computer tethered to the robot via bluetooth using the CWiid library and a separate Python script recorded the position of the robot reported by the Wiimote during the duration of the run\~cite{yos:clune}.
+We identified the robot's position via infrared LEDs on the robot that could be tracked by a Nintendo Wii remote.
+%JMC you sure there was more than one IR LED?
+I thought it was just one IR and one visible for us to tell if there is power to the LEDs.
+If it is just one, switch to singular instead of plural in the previous sentence
+To maximize its viewable window, the Wii remote was fixed to the ceiling pointing down. It communicated with the robot computer via bluetooth using the CWiid library~\cite{yos:clune}. 
+%JMC The paragraph about the Wii mote was very wordy.
+I think I trimmed it by half.
+You should try to be similarly compact the rest of the paper.
+Compare your version to mine and try to see how I condensed things.
 
@@
 % %
-\figp{robot_whitebg}{.6}{The QuadraTot robotic platform.
-It is open-sourced and 3-D printable, which has allowed for other  researchers to use the robot for their research around the  world.
-The robot has 9 DOF, 2 per leg and a hip joint which connects  the two symmetrical halves of the robot.
-The robot uses Dynamixel  AX-18 and AX-12 servos for actuation.
-Figure taken from Shen et  al.
-\cite{haocheng}.}
+\figp{robot_whitebg}{.6}{The QuadraTot robotic platform, on which numerous gait evolution papers are based.
+Our method of evolving gaits in simulation with the HyperNEAT generative encoding and transferring them to the robot produced faster gaits than all previously published techniques.
+%It is open-sourced and 3-D printable, which has allowed for other  researchers to use the robot for their research around the  world. 
+%JMC The previous sentence (which I commented out) is alright for the main text, but figures are very expensive real estate.
+The goal of our paper is not to sell this robot, but to sell the work we did.
+There are 9 degrees of freedom: 2 per leg, and a single `hip joint' that connects  the two symmetrical halves of the robot.
+%JMC: If you use which instead of that, you need a comma before which.
+For example, this is wrong: There are 9 degrees of freedom: 2 per leg, and a single `hip joint' which connects  the two symmetrical halves of the robot.
+This is right.
+There are 9 degrees of freedom: 2 per leg, and a single `hip joint' that connects  the two symmetrical halves of the robot.
+This is also right: There are 9 degrees of freedom: 2 per leg, and a single `hip joint,' which connects  the two symmetrical halves of the robot.
+% The robot uses Dynamixel  AX-18 and AX-12 servos for actuation. %JMC We said this already in methods
+% Figure taken from Shen et  al. \cite{haocheng}.%JMC Actually it's from yos:clune, but I think it is ok to leave this out, since this is just a stock photo of the QuadraTot.
+ }
+%JMC Let's put this figure on the first page (before the introduction), unless you can get a series of pictures that show a gait.
+(This might be hard to do).
+A new trend in papers is to have a sexy image on the first page, and Hod will want to see this.
+I also think it makes people want to read the paper more.
 

### 2012-08-24 paper_sean/simulator.tex

@@
 % !TEX root =main.tex
-The simulations were done using a simulator\footnote{Available for Linux and Windows at http://quadratot.yosinski.com} designed and written by Kyrre Glette and Jim Torresen in the University of Oslo, Norway\footnote{{kyrrehg, jimtoer}@ifi.uio.no} \cite{glette}.
-It was written in C++ using the NVIDIA PhysX physics software library to provide an accurate physical representation of the behaviors of the robot. 
-A physical representation of the QuadraTot was built in simulation to capture the most important parts of the robot like the number and type of joints, mass, and rough shapes and lengths of the robot.
-The simulator was visualized by OpenGL library to allow the user to have a more intuitive grasp of what is going on. 
-The simulator visualization included a recording feature which allows the user to create videos of the robot in simulation.
-The simulator also features a command line function which allows the users to input a gait file and run it on the simulator. 
-The simulator then outputs a file containing the X, Y, Z coordinates of the center of mass of the robot over time. 
-In simulation, the robot traverses in the XY plane with the Z axis representing the height off the ground. 
-In an attempt to more accurately model the robot behavior, we implemented the same cropping function in the simulator as in the actual robot. 
+Simulations occurred in C++ software customized for the QuadraTot \footnote{Available for Linux and Windows at http://quadratot.yosinski.com} in the NVIDIA PhysX physics simulator, as described previously~\cite{glette}.
+Specifically, the simulation captured the number and type of joints, mass, and rough shapes and lengths of the QuadraTot components. 
+%The simulator was visualized by OpenGL library to allow the user to have a more intuitive grasp of what is going on. %JMC Unnecessary detail since this paper is not about the simulator. 
+%The simulator visualization included a recording feature which allows the user to create videos of the robot in simulation.%JMC: Assumed
+%The simulator also features a command line function which allows the users to input a gait file and run it on the simulator. %JMC: Assumed
+%The simulator then outputs a file containing the X, Y, Z coordinates of the center of mass of the robot over time. %JMC: Assumed
+%In simulation, the robot traverses in the XY plane with the Z axis representing the height off the ground. %JMC Unnecessary detail
+In an attempt to more accurately model the robot behavior, we implemented the same cropping function in the simulator as in the actual robot. %Do all results in this paper reflect this?
 Although the simulator was originally written in Windows for the Windows operating platform we ported and ran it on Linux (Ubuntu 11.04) to seamlessly link the simulator and the already exising libraries together.

### 2012-08-26 paper_sean/introduction.tex
jmc edits
added changes to references.bib made changes to introduction.tex

@@
 
-Much research has been done in this subject, and previous work has shown that learned gaits can outperform designed gaits \cite{valsalam:mii}, \cite{kohl:stone}.
-%JMC - I have a similar claim in my CEC 2009 paper.
+Much research has been done in this subject, and previous work has shown that learned gaits can outperform designed gaits \cite{valsalam:mii,kohl:stone,hornby1,hornby2}.
+%JMC/SL- I have a similar claim in my CEC 2009 paper.
 Grab those cites for here (especially the Hornby ones).
-It has also been shown that gaits with regularity (coordination in leg movement, such as left-right symmetry or front-back symmetry) perform better than gaits without regularities 
-\cite{valsalam:mii,clune2,clune1,clune3}. %JMC-note how you can put multiple cites in one bracket...but no spaces after the comma!
-However, in most previous works, the experimenter has had to manually decide the regularities of the gaits. %we need cites for this claim. you can get them from Clune1
+/ have added hornby citations
+It has also been shown that gaits with regularity (coordination in leg movement, such as left-right symmetry or front-back symmetry) perform better than gaits without regularities \cite{valsalam:mii,clune2,clune1,clune3}.
+%JMC/SL-note how you can put multiple cites in one bracket...but no spaces after the comma!
+/ gotcha!
+However, in most previous works, the experimenter has had to manually decide the regularities of the gaits \cite{valsalam:mii,tellez,beer,raibert} .
+%JMC/SL - we need cites for this claim.
+you can get them from Clune1 / done
 Such manual intervention is time consuming, requires expert knowledge, and adds constraints that may hurt performance. 
@@
 However, that work only verified these claims in simulation. 
-A follow-up paper by Glette et al.
+A follow-up paper by Yosinski et al.
 evolved gaits with HyperNEAT directly in hardware on the QuadraTot robot platform and found that HyperNEAT's gaits outperformed manually designed parameterized learning algorithms, but still did not produce impressive, natural gaits \cite{yos:clune}.
-% insert (figure 1 -- pic of quadratot)
-A further follow-up study built a simulator for the QuadraTot robot platform to test whether the inclusion of a simulator would improve results and found that it did:  an evolutionary algorithm with direct encoding in simulation produced gaits that outperformed those gaits evolved by HyperNEAT in hardware~\cite{glette}.
-%JMC-Should there be a space between the end of the sentence and the cite?
+A further follow-up study built a simulator for the QuadraTot robot platform to test whether the inclusion of a simulator would improve results and found that it did:  an evolutionary algorithm with direct encoding in simulation produced gaits that outperformed those gaits evolved by HyperNEAT in hardware \cite{glette}.
+%JMC/SL-Should there be a space between the end of the sentence and the cite?
 Whatever the answer, we should be consistent.
@@
 The ~ tells Latex not to split that space from the character before and after it.
+/ There should have been and will make changes accordingly
 The simulator reduced the time required to evaluate gaits and afforded much larger population sizes and more generations than was possible when evolving in hardware, ultimately performing 333 times as many evaluations as HyperNEAT had in hardware per run (60000 vs.
@@
 
-In this paper we test whether the benefits of HyperNEAT -- higher performance and the automatic discovery of complex, effective regularities \cite{clune2,yos:clune} -- will outperform the manually designed, regularity-constrained, direct encoding once HyperNEAT is combined with a simulator.
-%JMC-I don't think they used a direct encoding.
+In this paper we test whether the benefits of HyperNEAT -- higher performance and the automatic discovery of complex, effective regularities \cite{clune2,yos:clune} -- will outperform a simple genetic algorithm manually designed to produce regularities once HyperNEAT is combined with a simulator.
+%JMC/SL-I don't think they used a direct encoding.
 I prefer describing it as simple and manually designed to produce specific regularities.
-We test this by evolving gaits using HyperNEAT using a physical simulator and then comparing the performance of these gaits in the real-world to those that were evaluated on the QuadraTot robotic platform in previous publications.
+/ changed to reflect that
+We test this by evolving gaits using HyperNEAT with a physical simulator and then comparing the performance of these gaits in the real-world to those that were evaluated on the QuadraTot robotic platform in previous publications.
 %JMC - Try to write without ever saying ``use'' or ``using.'' If you do, you'll find how much it is overused in writing, as evidenced by this last sentence.
@@
 %Our work also confirms that generative encodings such as HyperNEAT are a powerful way to evolve gaits and, more generally, automatically exploit regularities in challenging engineering domains.
-
-% figure of the quadratot in its natural stance
-%\begin{figure}
-%\begin{center}
-%\vspace{1.5cm}
-% just using a photo from Haocheng's paper! Should replace if needed be
-%\epsfig{file=quadratot_fig.eps, width=11cm}
-%\caption[ ]{The QuadraTot robotic platform on which experiments on this study was performed.
-It is open-sourced and 3-D printable, which has allowed for other researchers to use the robot for their research around the world.
-The robot has 9 DOF, 2 per leg and a hip joint which connects the two symmetrical halves of the robot.
-The robot uses Dynamixel AX-18 and AX-12 servos for actuation.}
-%\label{fig:quadratot}
-%\end{center}
-%\end{figure}

### 2012-08-27 paper_sean/robot.tex
finished editing robot.tex moved the quadratot photo to before the introduction fixed bug in references.bib

@@
 % !TEX root =main.tex
-The robot platform used for this project was a quadrupedal robot named the QuadraTot\footnote{Robot plans and part files for 3D printing are freely available at http://quadratot.yosinski.com} designed in the Cornell Creative Machines Lab by Juan Zagal~\cite{yos:clune}.
+The robot platform used for this project was a quadrupedal robot named the QuadraTot\footnote{Robot plans and part files for 3D printing are freely available at http://quadratot.yosinski.com} designed in the Cornell Creative Machines Lab by Juan Zagal \cite{yos:clune}.
 It has 9 degrees-of-freedom (DOF), with 2 DOF for the inner joint and the outer joint of each leg and 1 DOF for the hip joint of the robot. 
 The QuadraTot is an open-sourced and 3-D printable robot, which allows it to be used in other labs for research, providing a platform with which people could compare the results of different gait learning algorithms.
-Currently, there are published data and results from 9 different learning algorithms on the platform from INSERT different publications.
-%JMC fill in INSERT, and dump all cites here.
-This is a good place to cite haochengs paper, which is fun for him and Jason, but does not distract from our story.
+Currently, there are published data and results from 9 different learning algorithms on the platform from three different publications \cite{yos:clune,glette,haocheng}.
+%JMC/SL fill in INSERT, and dump all cites here.
+This is a good place to cite haochengs paper, which is fun for him and Jason, but does not distract from our story./ DONE
 
-The body of the robot was printed using an Objet Connex 500 3-D Printing System. 
-It was actuated using 9 Robotis Dynamixel servos: 5 AX-18A servos and 4 AX-12A servos. 
+The body of the robot was printed using an Objet Connex 500 3-D Printing System and actuated using 9 Robotis Dynamixel servos: 5 AX-18A servos and 4 AX-12A servos. 
 The stronger AX-18A servos were used to actuate the inner joints of the robot and the hip joint while the AX-12A servos were used to actuate the outer joints. 
@@
 The ones most relevant to this study were shutdowns due to high torque and range. 
-As has been reported previously with QuadraTot\cite{yos:clune}, many of the gaits that were tested on the QuadraTot forced the servos to exert too much torque or would command out of range positions and cause the servos to shutdown.
+As has been reported previously with QuadraTot \cite{yos:clune}, many of the gaits that were tested on the QuadraTot forced the servos to exert too much torque or would command out of range positions and cause the servos to shutdown.
 In order to address this problem and avoid collisions with its own body, a motion cropping system was implemented on the QuadraTot.
@@
 \cite{haocheng} to further prevent servo shutdowns from leg positions where the combination of the two angles interact to produce an extreme amount of torque.
-This smart cropping algorithm calculates a leg position by summing the servo position values from the inner and outer joints of a leg, \emph{\textbf{g}} (\emph{\textbf{g}} = $j_{inner}$ + $j_{outer}$) %JMC - do they sum the position values, or the joint angles?
+This smart cropping algorithm calculates a leg position by summing the servo position values from the inner and outer joints of a leg, \emph{\textbf{g}} (\emph{\textbf{g}} = $j_{inner}$ + $j_{outer}$) %JMC/SL - do they sum the position values, or the joint angles?
 Since all the previous extremes are expressed in angles, the writing makes me expect angles...
-and modifying the servo positions from fierce gaits, which usually have relatively low \emph{\textbf{g}}s, to remap them onto the boundary of a quasi-trangular area in a 2-D space defined by $j_{inner}$ and $j_{outer}$.
-%the term fierce sounds silly and non-professional (sorry!) I took out the place where you defined it, and we only use it in two other places.
+/ They do use the position values and I think that it would be best to be consistent with what Jason/Haocheng did.
+This is also the way g is calculated in the simulator and the python drivers as well and more accurately reflects what we actually did.
+and modifying the servo positions from ``fierce gaits'', which usually have relatively low \emph{\textbf{g}}s, to remap them onto the boundary of a quasi-trangular area in a 2-D space defined by $j_{inner}$ and $j_{outer}$.
+%JMC/SL - the term fierce sounds silly and non-professional (sorry!) I took out the place where you defined it, and we only use it in two other places.
 No need to invent a definition only to use it 2-3 times...can you swap in different, more descriptive words?
-Through experimentation, we found that a \emph{\textbf{g}} of 730 provides a good balance between performance and safety.
-%JMC - This description is not great.
+/ Again, I am trying to be consistent with what Jason and Haocheng used.
+I agree that 'fierce' is not the best word but if it is different from what J/H did, then there are inconsistencies between the publications...
+We set the minimum threshold \emph{\textbf{g}} to be 730, found through experimentation to provide a good balance between performance and safety.
+%JMC/SL - This description is not great.
 The last sentence's use of g doesn't make sense with the equation...is g a threshold or the sum?
 Please try to clean up this description.
+/ Think this clarified things
 
@@
 We identified the robot's position via infrared LEDs on the robot that could be tracked by a Nintendo Wii remote.
-%JMC you sure there was more than one IR LED?
+%JMC/SL you sure there was more than one IR LED?
 I thought it was just one IR and one visible for us to tell if there is power to the LEDs.
-If it is just one, switch to singular instead of plural in the previous sentence
-To maximize its viewable window, the Wii remote was fixed to the ceiling pointing down. It communicated with the robot computer via bluetooth using the CWiid library~\cite{yos:clune}. 
-%JMC The paragraph about the Wii mote was very wordy.
+If it is just one, switch to singular instead of plural in the previous sentence / Yep there are like 3 IR LEDs being used.
+To maximize its viewable window, the Wii remote was fixed to the ceiling pointing down. It communicated with the robot computer via bluetooth using the CWiid library \cite{yos:clune}. 
+%JMC/SL The paragraph about the Wii mote was very wordy.
 I think I trimmed it by half.
@@
 Compare your version to mine and try to see how I condensed things.
+/ Will do boss.
 
@@
 % %
-\figp{robot_whitebg}{.6}{The QuadraTot robotic platform, on which numerous gait evolution papers are based.
-Our method of evolving gaits in simulation with the HyperNEAT generative encoding and transferring them to the robot produced faster gaits than all previously published techniques.
+
 %It is open-sourced and 3-D printable, which has allowed for other  researchers to use the robot for their research around the  world. 
-%JMC The previous sentence (which I commented out) is alright for the main text, but figures are very expensive real estate.
+%JMC/SL The previous sentence (which I commented out) is alright for the main text, but figures are very expensive real estate.
 The goal of our paper is not to sell this robot, but to sell the work we did.
-There are 9 degrees of freedom: 2 per leg, and a single `hip joint' that connects  the two symmetrical halves of the robot.
-%JMC: If you use which instead of that, you need a comma before which.
+/ Okay I understand
+
+%JMC/SL: If you use which instead of that, you need a comma before which.
 For example, this is wrong: There are 9 degrees of freedom: 2 per leg, and a single `hip joint' which connects  the two symmetrical halves of the robot.
@@
 This is also right: There are 9 degrees of freedom: 2 per leg, and a single `hip joint,' which connects  the two symmetrical halves of the robot.
+/ Ok
 % The robot uses Dynamixel  AX-18 and AX-12 servos for actuation. %JMC We said this already in methods
 % Figure taken from Shen et  al. \cite{haocheng}.%JMC Actually it's from yos:clune, but I think it is ok to leave this out, since this is just a stock photo of the QuadraTot.
- }
  %JMC Let's put this figure on the first page (before the introduction), unless you can get a series of pictures that show a gait.

### 2012-09-03 paper_sean/conclusion.tex
JMC edits
editting, cutting paper

@@
-HyperNEAT has proven to be an effective encoding in gait evolution and is able to find gait regularities without human intervention \cite{clune1}, \cite{clune2}, \cite{yos:clune}. 
+% !TEX root =main.tex
+HyperNEAT has proven to be an effective encoding in gait evolution which is able to find gait regularities without human intervention \cite{clune1,clune2,yos:clune}. 
 Glette et al.
@@
 In the real life trials, HyperNEAT's best gait performed 5.4\% better than the GA gait. 
-This study also showed that HyperNEAT will exploit the characteristics of the problem in order to maximize its fitness.
-Without frequency limiters, it would make the legs move really fast in order to increase fitness. 
-When that was no longer allowed, it pushed the servo positions to their boundaries by increasing the stride lengths as wide as possible. 
-However, because of the physical limitations of the robot, these exploitations could not be transferred well to reality. 
+This study also showed that HyperNEAT will exploit the characteristics of the problem, such as ferquency and regularity, in order to maximize its fitness.
+%Without frequency limiters, it would make the legs move really fast in order to increase fitness. 
+%When that was no longer allowed, it pushed the servo positions to their boundaries by increasing the stride lengths as wide as possible. 
+However, because of the physical limitations of the robot, these exploitations did not transfer well to reality. %could not be transferred well to reality.
 Moreover, the inevitable faults with the simulator (it cannot model the robot or the real world perfectly) added to the difficulty in the transfer from simulation to reality. 
@@
 
-Next steps for this research heads towards improving the transferabilities of the gaits from simulation to reality. 
+Next steps for this research heads towards improving the transfer process of the gaits from simulation to reality. 
 The performance gap between simulation and reality is called the \emph{reality gap} \cite{jakobi}. 
-Many different approaches to reduce the reality gap has been proposed \cite{koos1}, \cite{bongard}, \cite{zagal}.
+Many different approaches to reduce the reality gap has been proposed \cite{koos1,bongard,zagal}. 
 To improve upon the results from this study, one or more of these approaches could be used to aid in the transfer to reality. 

### 2012-09-03 paper_sean/hyperneat.tex
Updated figure commands, added as many figures as possible.
Added hyperneat_bug_example figure to Sean's paper
JMC edits
editting, cutting paper

@@
+% !TEX root =main.tex
 HyperNEAT is a neuroevolution method which evolves artificial neuron networks (ANNs) using evolutionary algorithms.
-It indirectly encodes ANNs using compositional pattern producing networks (CPPNs) which represent genomes encoding ANN phenotypes \cite{stanley1}, \cite{stanley2}. 
-HyperNEAT has shown that it could evolve large neural networks that represent the networks and features present in brains such as repetitions and regularities \cite{stanley3}. 
-This encoding has previously been applied for gait-learning \cite{yos:clune} and for evolving 2-D and 3-D objects \cite{clune:lipson}. 
-It has been shown that objects evolved using HyperNEAT look complex, natural and feature characteristics such as symmetries and repetition of themes which are often found in nature.
-% figure from jeff paper
+HyperNEAT has shown to evolve large neural networks that represent many features present in brains such as repetitions and regularities \cite{stanley3}. 
+This encoding has previously been applied for gait-learning \cite{yos:clune} and for evolving 2-D and 3-D objects \cite{clune:lipson} and has been successful in displaying complex, natural characteristics such as symmetries and repetition of themes often found in nature.
 
+HyperNEAT indirectly encodes ANNs using compositional pattern producing networks (CPPNs) which represent genomes encoding ANN phenotypes \cite{stanley1}, \cite{stanley2}. 
 The motivation for CPPNs is that complex patterns can be produced by determining attributes of their phenotypic components as a function of their geometric location. 
@@
 
-Each CPPN is a directed graph in which every node is a mathematical function, such as sine or Gaussian. % figure!
+Each CPPN is a directed graph in which every node is a mathematical function, such as sine or Gaussian. 
 Moreover, a CPPN genome allows functions to be made of other functions, allowing coordinate frames to be combined and hierarchies to develop.
@@
 
-\begin{figure}
-\begin{center}
-\vspace{1cm}
-\epsfig{file=CPPN.eps, width=8cm}
-\caption [ ]{Compositional Pattern Producing Networks (CPPNs).
+% \begin{figure}
+% \begin{center}
+% \vspace{1cm}
+% \epsfig{file=CPPN.eps, width=8cm}
+% \caption [ ]{Compositional Pattern Producing Networks (CPPNs).
 CPPN  genomes use mathematical functions to generate regularities such as symmetry and repetitions.
@@
 Figure taken from Stanley \cite{stanley2}.}
-\end{center}
-\end{figure}
+% \end{center}
+% \end{figure}
+\figp{hyperneat_bug_example}{.6}{Compositional Pattern Producing Networks (CPPNs).
+CPPN  genomes use mathematical functions to generate regularities such as symmetry and repetitions.
+CPPN genomes allow functions to be made of other functions, allowing multiple regularity motifs to be present.
+Figure taken from Stanley \cite{stanley2}.}
 
-\begin{figure}
-\begin{center}
-\vspace{1cm}
-\epsfig{file=hyperneat.eps, width=8cm}
-\caption [ ]{Producing ANNs from CPPNs.
+% \begin{figure}
+% \begin{center}
+% \vspace{1cm}
+% \epsfig{file=hyperneat.eps, width=8cm}
+% \caption [ ]{Producing ANNs from CPPNs.
 In HyperNEAT, weights are specified as a function of the Cartesian coordinates for source and input nodes and a constant bias for each connection between the source and input.
@@
 \cite{clune2}}.
-\end{center}
-\end{figure}
+% \end{center}
+% \end{figure}
+\figp{hyperneatExplanation}{.6}{Producing ANNs from CPPNs.
+In HyperNEAT, weights are specified as a function of the Cartesian coordinates for source and input nodes and a constant bias for each connection between the source and input.
+All pairwise combinations of source and target node coordinates are iteratively passed into CPPNs to determine the weight of each ANN link.
+Figure taken from Clune et al.
+\cite{clune2}}

### 2012-09-03 paper_sean/methods.tex
JMC edits
editting, cutting paper

@@
-Gaits were first evolved in simulation, as previous done in Glette et al. \cite{glette}. 
+% !TEX root =main.tex
+Gaits were first evolved in simulation and then transferred onto the real robot, as previous done in Glette et al. \cite{glette}. 
 During the evolution process, we evolved 200 organisms for 200 generations in 20 runs with different initial conditions specified by randomly seeding the run. 
-To keep in line with previous studies that have been performed on the QuadraTot platform, each gait ran for 12 seconds. 
-All the HyperNEAT settings used for gait generation was kept the same as in Yosinski et al. \cite{yos:clune} except for the frequency of the sine wave function used. 
-Through experimentation, we found that high frequency gaits, gaits that require servos to change their directions often, were much more prone to servo shutdowns than low frequency gaits (at around 1.5Hz per servo).
-HyperNEAT often evolved high frequency gaits (over 8Hz per servo) which performed poorly in reality due to servo constraints. 
+Each gait was 12 seconds long, both in simulation and in reality. 
+All the HyperNEAT settings used for gait generation was kept the same as in Yosinski et al.
+\cite{yos:clune} except for the frequency of the sine wave function used, which was changed in order to address the servo shutdown problem mentioned previously.
+We found that gaits that require servos to change their directions often were more prone to servo shutdowns than low frequency gaits. %(at around 1.5Hz per servo). 
+But with the sine wave frequency used for Yosinski et al.
+\cite{yos:clune}, HyperNEAT often evolved high frequency gaits (over 8Hz per servo) which performed poorly in reality due to servo constraints.
 This led to gaits performing dramatically different in real-life compared to simulations. 
-To address this issue, two changes were made:
-First, the frequency of the sine wave function used in HyperNEAT was reduced by a factor of 6.63 by changing the sine function to
+%To address this issue, two changes were made:
+To address thsi, we reduced by a factor of 6.63 by changing the sine function to
 
@@
 
-from
-
-\begin{equation}
-sin(t/6)\pi
-\end{equation}
+%from
+%
+%\begin{equation}
+%sin(t/6)\pi
+%\end{equation}
+%
 
 so that HyperNEAT would lean towards evolving low frequency gaits.
-Sine frequency of 0.004  was used because we experimentally found that HyperNEAT was able to produce natural and high-performing gaits with low frequencies with this frequency.
-Second, a policy that punished high frequency gaits was implemented during the evolution process. 
-If the average frequency of each for the gait was over the threshold frequency, the fitness was punished by exponentially reducing the distance travelled by the robot:
+%Sine frequency of 0.004 was used because we experimentally found that HyperNEAT was able to produce natural and high-performing gaits with low frequencies with this frequency.
+We further added a policy that punished high frequency gaits while evolving in simulation. 
+If the average frequency of each leg for the gait was higher than the threshold frequency, the fitness was punished by exponentially reducing the distance travelled by the robot:
 
@@
 
-Through experimentation, we chose the threshold frequency to be 1.67 Hz/servo. After implementing the frequency punishment policy, the servos shutdown much less frequently.
+We chose the threshold frequency for punishment to be 1.67 Hz/servo. 
+After implementing the frequency punishment policy, the servos shutdown much less frequently.
 
@@
 
-where \emph{d} is the distance the center of mass of the robot travelled during the given time as was done in Clune et al. \cite{clune1}.
-The distance travelled was calculated as the Euclidean distance travelled by the robot in the XY plane during the alotted time, in centimeters.
+where \emph{d} is the Euclidean distance the center of mass of the robot travelled during the gait duration. %as was done in Clune et al. \cite{clune1}.
 
-In simulation, the distance travelled was calculated using the output file generated by the simulator.
-In real-life, the distance travelled was calculated using the log file recorded using the Wiimote system described earlier in the paper.
+%In simulation, the distance travelled was calculated using the output file generated by the simulator.
+%In real-life, the distance travelled was calculated using the log file recorded using the Wiimote system described earlier in the paper.
 For real-life trials, the actual fitnesses were not calculated and the gaits were evaluated using only the velocity of the robot.
 
-To evolve gaits using the physical simulator, we wrote a Python script\footnote{All scripts and drivers available at http://quadratot.yosinski.com} that interfaced between the simulator and HyperNEAT which allowed us to evolve gaits fast and efficiently with no human intervention.
-The runs were performed on three separate machines all running the same version of Linux (Ubuntu 11.04) using the same HyperNEAT settings and files to speed up the simulation process.
+To evolve gaits using the physical simulator, we wrote a Python script\footnote{All scripts and drivers available at http://quadratot.yosinski.com} that interfaced between the simulator and HyperNEAT.
+% which allowed us to evolve gaits fast and efficiently with no human intervention.
+The runs were performed on three separate machines all running Ubuntu 11.04 using the same HyperNEAT settings and files to speed up the simulation process.
 

### 2012-09-03 paper_sean/results.tex
JMC edits
editting, cutting paper

@@
-HyperNEAT, used with a physical simulator, produced gaits that outperformed all previous gaits tried and recorded on the QuadraTot. 
-The fastest gait evolved by HyperNEAT in simulation and transfered onto the real robot recorded a speed of 14.5 cm/s, faster than the previously fastest gait published in Glette et al. \cite{glette}. 
-In their paper, Kyrre et al.
-\cite{glette} recorded 17.8 cm/s in their robot in Norway \cite{glette}, but to make sure that there were no differences in robots, the same gait was also tested on the robot used for our experimental testing.
-The same gait, ran by using the gait file sent to us by Kyrre Glette, recorded an average speed of 12.951 cm/s and a maximum of 13.76 cm/s on the same QuadraTot platform in the Cornell Creative Machines Lab where Yosinski et al.
-\cite{yos:clune} and this study was done.
-HyperNEAT, used with a simulator, outperformed Kyrre et al's \cite{glette} gait by 5.4\% with much fewer evaluations performed per run, at 40000 evaluations per run with 200 organisms and 200 generations as opposed to Kyrre et al's \cite{glette} 60000 evaluations per run with 200 organisms and 300 generations.
-The best gait evolved using HyperNEAT with simulator performed 49.5\% better than the best gait evolved using HyperNEAT in hardware \cite{yos:clune}. 
-This is most likely due to the difference in the number of evaluations between the two studies. 
-The previous study, HyperNEAT in hardware, only had 180 evaluations while in this study, a simulator allowed a much higher number -- 20000 -- of evaluations. 
-Another minor reason for this may be that the implementation placed to reduce gait frequency may have helped in producing gaits that were less susceptible to shutdowns, as Yosinski et al.
+% !TEX root =main.tex
+HyperNEAT, used with a physical simulator, produced gaits that outperformed all previous gaits tried and recorded on the QuadraTot.\footnote{Videos of the gaits in simulation and on robot available at http://quadratot.yosinski.com}
+The speed of the fastest gait recorded in this study was 14.5 cm/s, faster than the previously fastest gait published in Glette et al. \cite{glette}. 
+Kyrre et al.
+\cite{glette} recorded 17.8 cm/s on their robot, but to make sure that there were no differences in robots, the same gait was also tested on the robot used for this study (which was the one used by Yosinski et al.
+\cite{yos:clune} and Shen et al.
+\cite{haocheng}.
+The same gait recorded an average speed of 12.951 cm/s and a maximum of 13.76 cm/s on our robot. We used these numbers in comparing results. 
+The gait from this study outperformed the previous Kyrre et al's \cite{glette} gait by 5.4\% with much fewer evaluations performed per run, at 40000 evaluations per run as opposed to Kyrre et al's \cite{glette} 60000 evaluations per run.
+It also outperformed the best gait evolved using HyperNEAT in hardware \cite{yos:clune} by 49.5\%, most likely due to the difference in the number of evaluations between the two studies -- 40000 vs 180.
+Another minor reason for superior performance may be that frequency limiter policy that we placed may have helped in producing gaits that were less susceptible to shutdown while Yosinski et al.
 \cite{yos:clune} did not have such things.
@@
 
-Gaits produced in this study confirmed that with the use of a physical simulator, HyperNEAT is able to fully exploit the geometry of the problem without human intervention, which was not evident in gait evolution using HyperNEAT in hardware \cite{yos:clune}.
-HyperNEAT took full advantage of the regularity of the problem and produced gaits more regular, coordinated, and effective than all other gaits evolved for the QuadraTot platform in the past.
-This is evident in the servo position plots where the gaits produced in this study have smooth and repeating curves while gaits produced by HyperNEAT in hardware produced much more ragged and random curves.
-The difference in regularity between the two studies is most likely due to the number of evaluations that were done in gait evolution.
+%Gaits produced in this study confirmed that with the use of a physical simulator, HyperNEAT is able to fully exploit the geometry of the problem without human intervention, which was not evident in gait evolution using HyperNEAT in hardware \cite{yos:clune}.
+HyperNEAT took advantage of the regularity of the problem and produced gaits more regular, coordinated, and effective than all other gaits evolved for the QuadraTot platform in the past.
+This is evident in the servo position plots where the gaits produced in this study are smooth and are similar between the servos. 
+%The difference in regularity between the two studies is most likely due to the number of evaluations that were done in gait evolution.
 Additionally, the noise in the real world was extremely high, making learning in hardware difficult, which may have further impeded the search for higher-performing, regular gaits.
@@
 \end{table}
+%JMC All numbers in the same column should be to the same precision (e.g.
+for the real velocity column, one decimal place).
+Also, can you right-justify the numbers (while still keeping them roughly centered)?
+Comparing numbers that are not right-justified is hard...so if you can't center and right-justify, then I let's see if just having it be right justified is better.
 

### 2012-09-03 paper_sean/simulator.tex
editting, cutting paper

@@
 % !TEX root =main.tex
-Simulations occurred in C++ software customized for the QuadraTot \footnote{Available for Linux and Windows at http://quadratot.yosinski.com} in the NVIDIA PhysX physics simulator, as described previously~\cite{glette}.
+Simulations occurred in C++ software customized for the QuadraTot\footnote{Available for Linux and Windows at http://quadratot.yosinski.com} in the NVIDIA PhysX physics simulator, as described previously \cite{glette}.
 Specifically, the simulation captured the number and type of joints, mass, and rough shapes and lengths of the QuadraTot components. 
@@
 %In simulation, the robot traverses in the XY plane with the Z axis representing the height off the ground. %JMC Unnecessary detail
-In an attempt to more accurately model the robot behavior, we implemented the same cropping function in the simulator as in the actual robot. %Do all results in this paper reflect this?
+In an attempt to more accurately model the robot behavior, we implemented the same cropping function in the simulator as in the actual robot.
+%JMC/SL - Do all results in this paper reflect this?
+/ They will soon.
 Although the simulator was originally written in Windows for the Windows operating platform we ported and ran it on Linux (Ubuntu 11.04) to seamlessly link the simulator and the already exising libraries together.

### 2012-09-12 paper_sean/conclusion.tex
lots of cuts, down to 13 pages.

@@
 % !TEX root =main.tex
-HyperNEAT has proven to be an effective encoding in gait evolution which is able to find gait regularities without human intervention \cite{clune1,clune2,yos:clune}. 
+HyperNEAT has proven to be an effective encoding in gait evolution with an ability to find gait regularities without human intervention \cite{clune1,clune2,yos:clune}. 
 Glette et al.
-\cite{glette} has shown that evolving gaits in simulation using a simple algorithm with a large number of evaluations could outperform gait evolution using a more sophisticated algorithm in hardware.
-This was possible because the simulator afford a much larger number of evaluations to be performed during the evolution process. 
-In this study, we took the two -- a sophisticated algorithm (HyperNEAT) and a physical simulator -- to see how HyperNEAT would perform with a many evaluations in the evolution process. 
-We showed that HyperNEAT, when used with a simulator, could evolve very impressive gaits. 
-In simulation, HyperNEAT gaits outperformed Genetic Algorithm gaits \cite{glette} by 54.3\%. 
-In the real life trials, HyperNEAT's best gait performed 5.4\% better than the GA gait. 
+\cite{glette} has shown that evolving gaits in simulation using a simple algorithm with a large number of evaluations could outperform gait evolution using a more sophisticated algorithm in hardware thanks to the simulator affording a larger number of evaluations to be performed for evolution.
+In this study, we took the two -- HyperNEAT and a physical simulator -- and showed that HyperNEAT, when it could afford many evaluations, could evolve impressive gaits. 
+In simulation, HyperNEAT gaits outperformed Genetic Algorithm gaits \cite{glette} by 54.3\% and by 5.4\% in the real life trials. 
 This study also showed that HyperNEAT will exploit the characteristics of the problem, such as ferquency and regularity, in order to maximize its fitness.

### 2012-09-12 paper_sean/results.tex
lots of cuts, down to 13 pages.

@@
 HyperNEAT, used with a physical simulator, produced gaits that outperformed all previous gaits tried and recorded on the QuadraTot.\footnote{Videos of the gaits in simulation and on robot available at http://quadratot.yosinski.com}
-The speed of the fastest gait recorded in this study was 14.5 cm/s, faster than the previously fastest gait published in Glette et al. \cite{glette}. 
-Kyrre et al.
-\cite{glette} recorded 17.8 cm/s on their robot, but to make sure that there were no differences in robots, the same gait was also tested on the robot used for this study (which was the one used by Yosinski et al.
-\cite{yos:clune} and Shen et al.
-\cite{haocheng}.
-The same gait recorded an average speed of 12.951 cm/s and a maximum of 13.76 cm/s on our robot. We used these numbers in comparing results. 
-The gait from this study outperformed the previous Kyrre et al's \cite{glette} gait by 5.4\% with much fewer evaluations performed per run, at 40000 evaluations per run as opposed to Kyrre et al's \cite{glette} 60000 evaluations per run.
+The fastest gait published was from Kyrre et al. \cite{glette}, which recorded 17.8 cm/s on their robot. 
+For consistency, we took that gait and ran it ten times on our robot, and found that on our robot, the gait had an average speed of 12.95 cm/s and a maximum speed of 13.76 cm/s.
+The speed of the fastest gait recorded in this study was 14.5 cm/s, which is 5.4\% faster than Kyrre et al.'s \cite{glette} gait with 20000 less evaluations performed per run.
+%Kyrre et al.
+\cite{glette} published 17.8 cm/s on their robot, but to make sure that there were no differences in robots, the same gait was also tested on our robot which then recorded an average speed of 12.951 cm/s and a maximum speed of 13.76 cm/s.
+%The gait from this study outperformed the previous Kyrre et al's \cite{glette} gait by 5.4\% with much fewer evaluations performed per run, at 40000 evaluations per run as opposed to Kyrre et al's \cite{glette} 60000 evaluations per run.
 It also outperformed the best gait evolved using HyperNEAT in hardware \cite{yos:clune} by 49.5\%, most likely due to the difference in the number of evaluations between the two studies -- 40000 vs 180.
-Another minor reason for superior performance may be that frequency limiter policy that we placed may have helped in producing gaits that were less susceptible to shutdown while Yosinski et al.
-\cite{yos:clune} did not have such things.
-Although most gaits started off its evolution process by producing fast and high frequency gaits, with more generations, the gaits dramatically reduced its frequencies while continually increasing their fitness.
+Another minor reason for superior performance may be that frequency limiter policy that we placed may have helped in producing gaits less susceptible to shutdown while Yosinski et al.
+\cite{yos:clune} had no such implementation.
+Although most gaits started its evolution process by producing fast and high frequency gaits, with more generations, the gaits dramatically reduced its frequencies while continually increasing their fitness.
 The average frequency of the final gaits was 1.29 Hz per servo.
@@
 % \end{figure}
-\figp{aracna_notYetCleanedMultiPieceRobot_edit}{.6}{Fitnesses of the champion gaits averaged over 20 runs, measured in distance travelled over 12 seconds [cm/s].
-These gaits outperformed gaits evolved in simulation using Genetic Algorithm by 54.3\% in simulation and 5.4\% better in reality while outperforming gaits evolved using HyperNEAT in hardware by 49.5\%.
-\edit{wrong fig, regenerate as pdf and add.}}
+\figp{avg_fit_vs_gen}{.6}{Fitnesses of the champion gaits averaged over 20 runs, measured in distance travelled over 12 seconds [cm/s].
+These gaits outperformed gaits evolved in simulation using Genetic Algorithm by 54.3\% in simulation and 5.4\% better in reality while outperforming gaits evolved using HyperNEAT in hardware by 49.5\%.}% \edit{wrong fig, regenerate as pdf and add.}}
 
@@
 % \end{figure}
-\figp{aracna_notYetCleanedMultiPieceRobot_edit}{.6}{Average frequencies of the gaits averaged over 20 runs.
+\figp{avg_freq_vs_gen}{.6}{Average frequencies of the gaits averaged over 20 runs.
 After implementing a policy which punished gaits with high frequencies, HyperNEAT was able to find high-performing, low frequency gaits.
-This policy was implemented to reduce servo shutdowns which heavily affected the reality performance of the robot.
-\edit{wrong fig, regenerate as pdf and add.}}
+This policy was implemented to reduce servo shutdowns which heavily affected the reality performance of the robot.} %\edit{wrong fig, regenerate as pdf and add.}}
 
@@
 % \end{figure}
-\figgp{aracna_notYetCleanedMultiPieceRobot_edit}{.6}{neat_110115_211410_00000_002_filt_zoom}{.6}{Top: A plot of servo positions for a gait produced in this study.
+%\figgp{servo_plot_111}{.6}{neat_110115_211410_00000_002_filt_zoom}{.6}{Top: A plot of servo positions for a gait produced in this study.
 Bottom: A plot of servo positions for a gait produced by HyperNEAT in hardware.
@@
 \cite{yos:clune}.
-The smoothness, symmetry, and regularities of the top plot shows that by using a simulator, HyperNEAT was able to fully exploit the geometry of the problem.
-\edit{wrong fig on top, regenerate top fig as pdf and add.}}
+The smoothness, symmetry, and regularities of the top plot shows that by using a simulator, HyperNEAT was able to fully exploit the geometry of the problem.}%\edit{wrong fig on top, regenerate top fig as pdf and add.}}
+\figp{servo_plot_111}{.6}{A plot of servo positions for a gait produced in this study.
+The smoothness, symmetry, and regularities of the top plot shows that by using a simulator, HyperNEAT was able to fully exploit the geometry of the problem.}%\edit{wrong fig on top, regenerate top fig as pdf and add.}}

### 2012-09-12 paper_sean/robot.tex
lots of cuts, down to 13 pages.

@@
 \cite{haocheng} to further prevent servo shutdowns from leg positions where the combination of the two angles interact to produce an extreme amount of torque.
-This smart cropping algorithm calculates a leg position by summing the servo position values from the inner and outer joints of a leg, \emph{\textbf{g}} (\emph{\textbf{g}} = $j_{inner}$ + $j_{outer}$) %JMC/SL - do they sum the position values, or the joint angles?
+%This smart cropping algorithm calculates a leg position by summing the servo position values from the inner and outer joints of a leg, \emph{\textbf{g}} (\emph{\textbf{g}} = $j_{inner}$ + $j_{outer}$) %JMC/SL - do they sum the position values, or the joint angles?
 Since all the previous extremes are expressed in angles, the writing makes me expect angles...
@@
 This is also the way g is calculated in the simulator and the python drivers as well and more accurately reflects what we actually did.
-and modifying the servo positions from ``fierce gaits'', which usually have relatively low \emph{\textbf{g}}s, to remap them onto the boundary of a quasi-trangular area in a 2-D space defined by $j_{inner}$ and $j_{outer}$.
+%and modifying the servo positions from ``fierce gaits'', which usually have relatively low \emph{\textbf{g}}s, to remap them onto the boundary of a quasi-trangular area in a 2-D space defined by $j_{inner}$ and $j_{outer}$.
 %JMC/SL - the term fierce sounds silly and non-professional (sorry!) I took out the place where you defined it, and we only use it in two other places.
@@
 I agree that 'fierce' is not the best word but if it is different from what J/H did, then there are inconsistencies between the publications...
-We set the minimum threshold \emph{\textbf{g}} to be 730, found through experimentation to provide a good balance between performance and safety.
+%We set the minimum threshold \emph{\textbf{g}} to be 730, found through experimentation to provide a good balance between performance and safety.
 %JMC/SL - This description is not great.
@@
 % \end{figure}
-\figp{smartCrop}{.6}{A map of possible leg positions.
+%\figp{smartCrop}{.6}{A map of possible leg positions.
 Manually placed position cropping systems limited possible leg positions to avoid servo shutdowns and the robot hitting itself.

### 2012-09-12 paper_sean/simulator.tex
lots of cuts, down to 13 pages.

@@
 %In simulation, the robot traverses in the XY plane with the Z axis representing the height off the ground. %JMC Unnecessary detail
-In an attempt to more accurately model the robot behavior, we implemented the same cropping function in the simulator as in the actual robot.
+We implemented the initial cropping function in the simulator as in the actual robot but not the smart cropping function because with the smart cropping function implemented in simulation, there was too much limitations in the movements which drastically limited the possible gaitspace HyperNEAT could explore.
 %JMC/SL - Do all results in this paper reflect this?

### 2012-09-20 paper_sean/hyperneat.tex
latest change before jeff's read

@@
 % !TEX root =main.tex
-HyperNEAT is a neuroevolution method which evolves artificial neuron networks (ANNs) using evolutionary algorithms.
+HyperNEAT is a neuroevolution method that evolves artificial neuron networks (ANNs) using evolutionary algorithms.
 HyperNEAT has shown to evolve large neural networks that represent many features present in brains such as repetitions and regularities \cite{stanley3}. 
@@
 HyperNEAT indirectly encodes ANNs using compositional pattern producing networks (CPPNs) which represent genomes encoding ANN phenotypes \cite{stanley1}, \cite{stanley2}. 
-The motivation for CPPNs is that complex patterns can be produced by determining attributes of their phenotypic components as a function of their geometric location. 
+The motivation for CPPNs is that complex patterns can be produced by determining the attributes of their phenotypic components as a function of their geometric location. 
 This is based on the belief that in nature, cells in a body differentiates and specializes in different things with respect to their location. 
-A cell in natural organisms cannot determine its location in space by itself and requires chemical gradients are used to signal its location \cite{carroll}. 
-In contrast, \emph{in silico}, cells can be given their geometric coordinates and CPPNs use this to their advantage.
+In real life, a cell in natural organisms cannot determine its location in space without the help of chemical gradients but \emph{in silico}, cells can be given their geometric coordinates, which CPPNs exploit.
+%A cell in natural organisms cannot determine its location in space by itself and requires chemical gradients are used to signal its location \cite{carroll}. 
+%In contrast, \emph{in silico}, cells can be given their geometric coordinates and CPPNs use this to their advantage.
 
 Each CPPN is a directed graph in which every node is a mathematical function, such as sine or Gaussian. 
-Moreover, a CPPN genome allows functions to be made of other functions, allowing coordinate frames to be combined and hierarchies to develop.
-For example, a sine function could develop a repeating theme which is passed onto a Gaussian function to create a repeating series of symmetrical motifs. 
-These functions allow evolution to exploit various properties such as symmetry (e.g. Gaussian) and repetition (e.g. sine function) \cite{stanley1}. 
-Because of evolution is able to exploit such properties, things like gaits are able to feature repetition and symmetry on its own without human intervention \cite{clune1}.
+Moreover, a CPPN genome allows functions to be made of other functions, allowing coordinate frames to be combined and hierarchies to develop, allowing for several different themes to appear such as a repeating theme (sine) with symmetry (Gaussian).
+%For example, a sine function could develop a repeating theme which is passed onto a Gaussian function to create a repeating series of symmetrical motifs. 
+These functions allow evolution to exploit various properties that allow HyperNEAT to find regularities without further human intervention \cite{stanley1,clune1}.
+%such as symmetry (e.g.
+Gaussian) and repetition (e.g.
+sine function) \cite{stanley1}.
+%Because evolution is able to exploit such properties, things like gaits are able to feature repetition and symmetry on its own without human intervention \cite{clune1}.
 
-There are links between nodes in CPPNs each of which has a weight used to multiply the signal in each link to determine the magnitude of effect from each node. 
+There are weighted links between nodes in CPPNs that are used to multiply the signal in each link and determine the magnitude of effect from each node. 
 In HyperNEAT, a CPPN genome takes in Cartesian coordinates (X,Y) of the source and the target nodes, and a constant bias value and outputs the weight between the input and hidden layer and a weight between the hidden layer and output (assuming that there is a hidden layer, as in the case for this study).

### 2012-09-20 paper_sean/methods.tex
latest change before jeff's read

@@
 All the HyperNEAT settings used for gait generation was kept the same as in Yosinski et al.
-\cite{yos:clune} except for the frequency of the sine wave function used, which was changed in order to address the servo shutdown problem mentioned previously.
-We found that gaits that require servos to change their directions often were more prone to servo shutdowns than low frequency gaits. %(at around 1.5Hz per servo). 
-But with the sine wave frequency used for Yosinski et al.
-\cite{yos:clune}, HyperNEAT often evolved high frequency gaits (over 8Hz per servo) which performed poorly in reality due to servo constraints.
-This led to gaits performing dramatically different in real-life compared to simulations. 
-%To address this issue, two changes were made:
-To address thsi, we reduced by a factor of 6.63 by changing the sine function to
+\cite{yos:clune} except for the frequency of the sine wave function used, which was changed in order to address the servo shutdown problem mentioned previously after discovering that gaits that require servos to change their directions often were more prone to servo shutdowns than low frequency gaits.
+%(at around 1.5Hz per servo).
+With the sine wave frequency used for Yosinski et al.
+\cite{yos:clune} (0.0265Hz), HyperNEAT often evolved high frequency gaits (over 8Hz per servo) that performed poorly in reality due to servo constraints.
+This led to dramatically different gait performance in real-life compared to simulations. 
+To address this, we changed the frequency of sine function to 0.004Hz so that HyperNEAT would lean towards evolving low frequency gaits.
 
-\begin{equation}
-sin(2\pi t/250)\pi
-\end{equation}
+%\begin{equation}
+%sin(2\pi t/250)\pi
+%\end{equation}
 
@@
 
-so that HyperNEAT would lean towards evolving low frequency gaits.
  %Sine frequency of 0.004 was used because we experimentally found that HyperNEAT was able to produce natural and high-performing gaits with low frequencies with this frequency. 
-We further added a policy that punished high frequency gaits while evolving in simulation. 
+We also added a policy that punished high frequency gaits while evolving in simulation. 
 If the average frequency of each leg for the gait was higher than the threshold frequency, the fitness was punished by exponentially reducing the distance travelled by the robot:
@@
 
-We chose the threshold frequency for punishment to be 1.67 Hz/servo. 
-After implementing the frequency punishment policy, the servos shutdown much less frequently.
+We chose the threshold frequency for punishment to be 1.67 Hz/servo. After implementing the frequency punishment policy, the servos shutdown much less frequently.
 
@@
 After evolving the gaits in simulation, the champion gait of the last generation of each of the 20 runs was transferred onto the real robot and evaluated for its performance on the real robot.
-In real-life trials, the motion of the robot was interpolated linearly between a stationary pose and the commanded gait for two seconds at the beginning and the end. % figure!! 
-The position of the robot was measured at the beginning and end of each run, starting from the stationary pose and ending with the stationary pose. 
-If a gait performed extremely poorly due to servo shutdowns, the gait was tried again until we could conclude that poor performance is because of the gait and not because of the condition of the robot.
+In real-life trials, the motion of the robot was interpolated linearly between a stationary pose and the commanded gait for two seconds at the beginning and the end. 
+The position of the robot was measured at the beginning and the end of each run, starting from the stationary pose and ending with the stationary pose. 
+If a gait performed extremely poorly due to servo shutdowns, the gait was tried again until we could conclude that poor performance is because of the gait and not due to the condition of the robot such as overheated servos.

### 2012-09-20 paper_sean/robot.tex
latest change before jeff's read

@@
 
-The body of the robot was printed using an Objet Connex 500 3-D Printing System and actuated using 9 Robotis Dynamixel servos: 5 AX-18A servos and 4 AX-12A servos. 
+The body of the robot was printed using an Objet Connex 500 3-D Printing System and is actuated using 9 Robotis Dynamixel servos: 5 AX-18A servos and 4 AX-12A servos. 
 The stronger AX-18A servos were used to actuate the inner joints of the robot and the hip joint while the AX-12A servos were used to actuate the outer joints. 
-The servos were attached to the robot using off-the-shelf components and parts.
 The servos have a range of motion of [0, 1023], roughly equivalent to a physical range of [-12\degree, +120\degree].
@@
 The ones most relevant to this study were shutdowns due to high torque and range. 
-As has been reported previously with QuadraTot \cite{yos:clune}, many of the gaits that were tested on the QuadraTot forced the servos to exert too much torque or would command out of range positions and cause the servos to shutdown.
-In order to address this problem and avoid collisions with its own body, a motion cropping system was implemented on the QuadraTot.
-In addition to placing limits on the range of motions for the inner joints, outer joints, and the hip joint of [-85\degree, +60\degree], [-113\degree, +39\degree], and [-28\degree, +28\degree] respectively, an extra cropping system named the smart cropping system was adopted from Shen et al.
+It has been reported previously Yosinski et al.
+\cite{yos:clune}, that many gait tested on the QuadraTot forced the servos to exert too much torque or command positions that are out of range and cause the servos to shutdown.
+In order to address this problem and avoid collisions with its own body, we implemented a motion cropping system by limiting the possible range of movement of the QuadraTot by [-85\degree, +60\degree], [-113\degree, +39\degree], and [-28\degree, +28\degree] for the inner, outer, and hip joints respectively.
+In addition to this, we implemented an extra cropping system named the Smart Cropping System from Shen et al.
 \cite{haocheng} to further prevent servo shutdowns from leg positions where the combination of the two angles interact to produce an extreme amount of torque.
@@
 
-The robot was controlled by CompuLab Fit-PC2, a compact computer running Ubuntu Linux 10.10. 
-In a previous study \cite{yos:clune}, this computer was placed on the robot to keep the robot untethered, but for this study the computer was taken off the robot to reduce weight, because the weight of the robot seemed to be one of the factors contributing to servo shutdowns.
-The servos were sent positions from the computer at 40Hz via the Pydynamixel library.
+We controlled the robot using a CompuLab Fit-PC2, a compact computer running Ubuntu Linux 10.10 and commanded the servos at 40Hz via the Pydynamixel library.
 
-We identified the robot's position via infrared LEDs on the robot that could be tracked by a Nintendo Wii remote.
+%In a previous study \cite{yos:clune}, this computer was placed on the robot to keep the robot untethered, but for this study the computer was taken off the robot to reduce weight, because the weight of the robot seemed to be one of the factors contributing to servo shutdowns.
+
+We identified the robot's position using infrared LEDs on the robot which was tracked using a Wiimote as in Yosinski et al. \cite{yos:clune}.
 %JMC/SL you sure there was more than one IR LED?
@@
 If it is just one, switch to singular instead of plural in the previous sentence / Yep there are like 3 IR LEDs being used.
-To maximize its viewable window, the Wii remote was fixed to the ceiling pointing down. It communicated with the robot computer via bluetooth using the CWiid library \cite{yos:clune}. 
+%To maximize its viewable window, the Wii remote was fixed to the ceiling pointing down. It communicated with the robot computer via bluetooth using the CWiid library \cite{yos:clune}. 
 %JMC/SL The paragraph about the Wii mote was very wordy.

### 2012-09-20 paper_sean/simulator.tex
latest change before jeff's read

--- 2012-09-03 paper_sean/simulator.tex
@@
 % !TEX root =main.tex
-Simulations occurred in C++ software customized for the QuadraTot\footnote{Available for Linux and Windows at http://quadratot.yosinski.com} in the NVIDIA PhysX physics simulator, as described previously \cite{glette}.
-Specifically, the simulation captured the number and type of joints, mass, and rough shapes and lengths of the QuadraTot components. 
+The simulator\footnote{Available for Linux and Windows at http://quadratot.yosinski.com} used for this study was the same as the one used in Glette et al.
+\cite{glette}.
+The simulator is written in C++ using the Nvidia PhysX engine.
+The simulation captured the number and type of joints, mass, and rough shapes and lengths of the QuadraTot components to model the robot.
+Variables such as the friction coefficient were kept the same for this study.
+%Simulations occurred in C++ software customized for the in the NVIDIA PhysX physics simulator, as described previously \cite{glette}. 
 %The simulator was visualized by OpenGL library to allow the user to have a more intuitive grasp of what is going on. %JMC Unnecessary detail since this paper is not about the simulator. 
@@
 %In simulation, the robot traverses in the XY plane with the Z axis representing the height off the ground. %JMC Unnecessary detail
-In an attempt to more accurately model the robot behavior, we implemented the same cropping function in the simulator as in the actual robot.
+We implemented the initial cropping function in the simulator as in the actual robot but not the smart cropping function because with the smart cropping function implemented in simulation, there was too much limitation in the movements which drastically limited the possible gaits HyperNEAT could explore.
 %JMC/SL - Do all results in this paper reflect this?
 / They will soon.
-Although the simulator was originally written in Windows for the Windows operating platform we ported and ran it on Linux (Ubuntu 11.04) to seamlessly link the simulator and the already exising libraries together.
-The friction values used for this study was the same as that used in Glette et al. \cite{glette}. 
+%Although the simulator was originally written in Windows for the Windows operating platform we ported and ran it on Linux (Ubuntu 11.04) to seamlessly link the simulator and the already exising libraries together.
 %

### 2012-09-21 paper_sean/abstract.tex
editting, cutting paper
small changes

@@
 Recently, the HyperNEAT generative encoding was shown to automatically discover a variety of gait regularities, producing fast, coordinated gaits, but only for simulated robots.
-A follow-up study found that HyperNEAT did not produce impressive gaits when they evolved directly on a physical robot.
+A follow-up study found that HyperNEAT did not produce impressive gaits when they were evolved directly on a physical robot.
 A simpler encoding hand-tuned to produce regular gaits was tried on the same robot, and outperformed HyperNEAT, but these gaits were first evolved in simulation before being transferred to the robot.
 In this paper, we tested the hypothesis that the beneficial properties of HyperNEAT would outperform the simpler encoding if HyperNEAT gaits are first evolved in simulation before being transferred to reality.
-That hypothesis was confirmed, resulting in the fasted gaits yet observed for this robot.
+That hypothesis was confirmed, resulting in the fastest gaits yet observed for this robot, including those produced by nine different algorithms from three previous papers describing techniques for generating gaits for this robot.
 This result is important because it confirms that the early promise shown by generative encodings, specifically HyperNEAT, were not limited to evolving gaits in simulation, but indeed work on real robots.

### 2012-09-21 paper_sean/introduction.tex
small changes

@@
 % !TEX root =main.tex
-Legged locomotion provides several benefits over wheeled locomotion. 
-A few examples of this are in flexibility in the type of terrain a robot could travel and in adapability to changes in a robot's physiology \cite{bongard:lipson}.
+Robots that can move around on legs can operate in a much wider range of environments than their wheeled counterparts.
+A downside is that designing gaits for legged robots is a difficult and time-consuming process~\cite{ridder storm, wetter green}.
 But designing a gait for a robot is often difficult and time consuming because the engineer must take into consideration many factors which must all fall into place for a successful gait pattern. 

### 2012-09-23 paper_sean/conclusion.tex

@@
 % !TEX root =main.tex
-HyperNEAT has proven to be an effective encoding in gait evolution with an ability to find gait regularities without human intervention \cite{clune1,clune2,yos:clune}. 
+HyperNEAT has proven to be an effective encoding in gait evolution with an ability to find gait regularities without human intervention \cite{clune2009evolving,clune2011performance,yos:clune}. 
 Glette et al.
@@
 The performance gap between simulation and reality is called the \emph{reality gap} \cite{jakobi}. 
-Many different approaches to reduce the reality gap has been proposed \cite{koos1,bongard,zagal}. 
+Many different approaches to reduce the reality gap has been proposed \cite{koos2010crossing,bongard,zagal}. 
 To improve upon the results from this study, one or more of these approaches could be used to aid in the transfer to reality. 
-The one easiest and most promising to implement is the Transferability Approach \cite{koos2} which has shown to be successful in quadrapedal gait evolution. 
+The one easiest and most promising to implement is the Transferability Approach \cite{koos2011transferability} which has shown to be successful in quadrapedal gait evolution. 
 HyperNEAT, used with a simulator and a transfer approach, shows much hope in producing yet a better, more coordinated, and higher performing gait on the QuadraTot platform.

### 2012-09-23 paper_sean/hyperneat.tex

@@
 HyperNEAT is a neuroevolution method that evolves artificial neuron networks (ANNs) using evolutionary algorithms.
-HyperNEAT has shown to evolve large neural networks that represent many features present in brains such as repetitions and regularities \cite{stanley3}. 
-This encoding has previously been applied for gait-learning \cite{yos:clune} and for evolving 2-D and 3-D objects \cite{clune:lipson} and has been successful in displaying complex, natural characteristics such as symmetries and repetition of themes often found in nature.
+HyperNEAT has shown to evolve large neural networks that represent many features present in brains such as repetitions and regularities \cite{stanley2002evolving}. 
+This encoding has previously been applied for gait-learning \cite{yos:clune} and for evolving 2-D and 3-D objects \cite{clune:lipson20113d} and has been successful in displaying complex, natural characteristics such as symmetries and repetition of themes often found in nature.
 
-HyperNEAT indirectly encodes ANNs using compositional pattern producing networks (CPPNs) which represent genomes encoding ANN phenotypes \cite{stanley1}, \cite{stanley2}. 
+HyperNEAT indirectly encodes ANNs using compositional pattern producing networks (CPPNs) which represent genomes encoding ANN phenotypes \cite{stanley2009hypercube}, \cite{stanley2007CPPN}. 
 The motivation for CPPNs is that complex patterns can be produced by determining the attributes of their phenotypic components as a function of their geometric location. 
@@
 %For example, a sine function could develop a repeating theme which is passed onto a Gaussian function to create a repeating series of symmetrical motifs. 
-These functions allow evolution to exploit various properties that allow HyperNEAT to find regularities without further human intervention \cite{stanley1,clune1}.
+These functions allow evolution to exploit various properties that allow HyperNEAT to find regularities without further human intervention \cite{stanley2009hypercube,clune2009evolving}.
 %such as symmetry (e.g.
@@
 
-HyperNEAT is first neuroevolutionary algorithm that has shown capabilities in exploiting the geometry of the problem \cite{stanley1}, \cite{clune3}.
+HyperNEAT is first neuroevolutionary algorithm that has shown capabilities in exploiting the geometry of the problem \cite{stanley2009hypercube}, \cite{clune2009sensitivity}.
 This is possible because the connection weights are a function of geometric positions of these nodes. 
@@
 
-HyperNEAT evolves the CPPNs according to the principles of the Neuroevolution of Augmenting Topologies (NEAT) algorithm \cite{stanley4}. 
-The NEAT algorithm has three major components \cite{stanley4}. 
+HyperNEAT evolves the CPPNs according to the principles of the Neuroevolution of Augmenting Topologies (NEAT) algorithm \cite{stanley2006exploiting}. 
+The NEAT algorithm has three major components \cite{stanley2006exploiting}. 
 First, it starts with small genomes that encode simple networks and complexifies them via mutations and add nodes and links to the network which allows the alogithm to evolve the network topology and weight.
@@
 Third, the historical information stored in genes helps to perform crossover in a way that is effective, yet avoids the need for expensive topological analysis.
-A full explanation of NEAT can be found in Stanley and Miikkulainen \cite{stanley4}.
+A full explanation of NEAT can be found in Stanley and Miikkulainen \cite{stanley2006exploiting}.
 
@@
 CPPN genomes allow functions to be made of other functions, allowing multiple regularity motifs to be present.
-Figure taken from Stanley \cite{stanley2}.}
+Figure taken from Stanley \cite{stanley2007CPPN}.}
 
@@
 Figure taken from Clune et al.
-\cite{clune2}}
+\cite{clune2011performance}}.

### 2012-09-23 paper_sean/introduction.tex
small changes
ridder storm bib add

@@
 Robots that can move around on legs can operate in a much wider range of environments than their wheeled counterparts.
-A downside is that designing gaits for legged robots is a difficult and time-consuming process~\cite{ridder storm, wetter green}.
+A downside is that designing gaits for legged robots is a difficult and time-consuming process for human engineers~\cite{strom1999legged,wettergreen1992gait}.
 But designing a gait for a robot is often difficult and time consuming because the engineer must take into consideration many factors which must all fall into place for a successful gait pattern. 
@@
 
-Much research has been done in this subject, and previous work has shown that learned gaits can outperform designed gaits \cite{valsalam:mii,kohl:stone,hornby1,hornby2}.
-%JMC/SL- I have a similar claim in my CEC 2009 paper.
-Grab those cites for here (especially the Hornby ones).
-/ have added hornby citations
-It has also been shown that gaits with regularity (coordination in leg movement, such as left-right symmetry or front-back symmetry) perform better than gaits without regularities \cite{valsalam:mii,clune2,clune1,clune3}.
+Much research has been done in this subject, and previous work has shown that learned gaits can outperform designed gaits \cite{valsalam:mii,kohl:stone,hornby2005autonomous,hornby2003generative}. 
+It has also been shown that gaits with regularity (coordination in leg movement, such as left-right symmetry or front-back symmetry) perform better than gaits without regularities \cite{valsalam:mii,clune2011performance,clune2009evolving,clune2009sensitivity}.
 %JMC/SL-note how you can put multiple cites in one bracket...but no spaces after the comma!
@@
 Such manual intervention is time consuming, requires expert knowledge, and adds constraints that may hurt performance. 
-Previous work has shown that the Hypercube-based NeuroEvolution of Augmenting Topologies (HyperNEAT) generative encoding \cite{stanley1} can automatically generate a variety of regular gaits and that it outperforms direct encoding controls on this task \cite{clune1}, \cite{clune2}.
+Previous work has shown that the Hypercube-based NeuroEvolution of Augmenting Topologies (HyperNEAT) generative encoding \cite{stanley2009hypercube} can automatically generate a variety of regular gaits and that it outperforms direct encoding controls on this task \cite{clune2009evolving}, \cite{clune2011performance}.
 However, that work only verified these claims in simulation. 
@@
 
-In this paper we test whether the benefits of HyperNEAT -- higher performance and the automatic discovery of complex, effective regularities \cite{clune2,yos:clune} -- will outperform a simple genetic algorithm manually designed to produce regularities once HyperNEAT is combined with a simulator.
+In this paper we test whether the benefits of HyperNEAT -- higher performance and the automatic discovery of complex, effective regularities \cite{clune2011performance,yos:clune} -- will outperform a simple genetic algorithm manually designed to produce regularities once HyperNEAT is combined with a simulator.
 %JMC/SL-I don't think they used a direct encoding.

### 2012-09-24 paper_sean/hyperneat.tex
made changes up to, but not including, the hyperneat section

--- 2012-09-20 paper_sean/hyperneat.tex
@@
 % !TEX root =main.tex
-HyperNEAT is a neuroevolution method that evolves artificial neuron networks (ANNs) using evolutionary algorithms.
-HyperNEAT has shown to evolve large neural networks that represent many features present in brains such as repetitions and regularities \cite{stanley3}. 
-This encoding has previously been applied for gait-learning \cite{yos:clune} and for evolving 2-D and 3-D objects \cite{clune:lipson} and has been successful in displaying complex, natural characteristics such as symmetries and repetition of themes often found in nature.
+HyperNEAT is an algorithm for evolving artificial neural networks (ANNs)~\cite{stanley3}.
 
@@
 If positions inputed to CPPNs represent aspects of the problem relevant to the solution, HyperNEAT could use that information to its advantage.
+
+Previous work has shown that Hyper
+HyperNEAT has shown to evolve large neural networks that represent many features present in brains such as repetitions and regularities . 
+This encoding has previously been applied for gait-learning \cite{yos:clune} and for evolving 2-D and 3-D objects \cite{clune:lipson} and has been successful in displaying complex, natural characteristics such as symmetries and repetition of themes often found in nature.
 

### 2012-09-24 paper_sean/introduction.tex
made changes up to, but not including, the hyperneat section

--- 2012-09-21 paper_sean/introduction.tex
@@
 Robots that can move around on legs can operate in a much wider range of environments than their wheeled counterparts.
-A downside is that designing gaits for legged robots is a difficult and time-consuming process~\cite{ridder storm, wetter green}.
-But designing a gait for a robot is often difficult and time consuming because the engineer must take into consideration many factors which must all fall into place for a successful gait pattern. 
-Because of setbacks present in hand-designing gaits, evolving gaits with machine learning algorithms is an attractive and flexible approach that can automatically produce effective gaits for many different robot platforms with minimal human intervention.
+A downside is that designing gaits for legged robots is a difficult and time-consuming process for human engineers~\cite{ridder storm, wetter green}, and must be repeated every time the design of a robot is changed or a new robot is created~\cite{hornby1}.
+Because of these difficulties in hand-designing gaits, scientists have investigated automatically producing gaits with machine learning or evolutionary algorithms, an approach that often produces better gaits than those created by human engineers~\cite{valsalam:mii,kohl:stone,hornby1,hornby2,yos:clune}.
+Both machine-created and human-created gaits perform better if they are \emph{regular}, in that they have coordinated movements, such as left-right symmetry or front-back symmetry~\cite{valsalam:mii,clune2,clune1,clune3}.
+However, in most previous works, the experimenter has had to manually decide the regularities of the evolved gaits \cite{valsalam:mii,tellez,beer,raibert} . 
+Such manual intervention is time consuming, requires expert knowledge, and adds constraints that may hurt performance.
 
-Much research has been done in this subject, and previous work has shown that learned gaits can outperform designed gaits \cite{valsalam:mii,kohl:stone,hornby1,hornby2}.
-%JMC/SL- I have a similar claim in my CEC 2009 paper.
-Grab those cites for here (especially the Hornby ones).
-/ have added hornby citations
-It has also been shown that gaits with regularity (coordination in leg movement, such as left-right symmetry or front-back symmetry) perform better than gaits without regularities \cite{valsalam:mii,clune2,clune1,clune3}.
-%JMC/SL-note how you can put multiple cites in one bracket...but no spaces after the comma!
-/ gotcha!
-However, in most previous works, the experimenter has had to manually decide the regularities of the gaits \cite{valsalam:mii,tellez,beer,raibert} .
-%JMC/SL - we need cites for this claim.
-you can get them from Clune1 / done
-Such manual intervention is time consuming, requires expert knowledge, and adds constraints that may hurt performance. 
 Previous work has shown that the Hypercube-based NeuroEvolution of Augmenting Topologies (HyperNEAT) generative encoding \cite{stanley1} can automatically generate a variety of regular gaits and that it outperforms direct encoding controls on this task \cite{clune1}, \cite{clune2}.
@@
 evolved gaits with HyperNEAT directly in hardware on the QuadraTot robot platform and found that HyperNEAT's gaits outperformed manually designed parameterized learning algorithms, but still did not produce impressive, natural gaits \cite{yos:clune}.
-A further follow-up study built a simulator for the QuadraTot robot platform to test whether the inclusion of a simulator would improve results and found that it did:  an evolutionary algorithm with direct encoding in simulation produced gaits that outperformed those gaits evolved by HyperNEAT in hardware \cite{glette}.
-%JMC/SL-Should there be a space between the end of the sentence and the cite?
-Whatever the answer, we should be consistent.
-If so, I have added here with the ~.
-The ~ tells Latex not to split that space from the character before and after it.
-/ There should have been and will make changes accordingly
-The simulator reduced the time required to evaluate gaits and afforded much larger population sizes and more generations than was possible when evolving in hardware, ultimately performing 333 times as many evaluations as HyperNEAT had in hardware per run (60000 vs.
-180) \cite{yos:clune}, \cite{glette}.
-The evolutionary algorithm used in Glette et al. \cite{glette} was manually constrained to evolve specific, hand-chosen regularities. 
-%Another study evolved gaits in hardware on the QuadraTot platform using RL PoWER, a spline-based algorithm, and was successful in evolving faster and more repeatable gaits than HyperNEAT evolved in hardware \cite{haocheng}.
-%JMC- I don't think we need to cite this paper...it is a non-evolutionary method, and it doesn't outperform Glette...and the sentence just sticks out as a detail that isn't necessary to our story.
+A further follow-up study built a simulator for the QuadraTot robot platform to test whether the inclusion of a simulator would improve results and found that it did:  when gaits were evolved in this simulator with a simple direct encoding and then transferred to the QuadraTot robot, the resulting gaits were faster than the gaits produced by evolving gaits with HyperNEAT directly on the robot \cite{glette}.
+The simulator helped because it afforded much larger population sizes and more generations than were possible when evolving directly in hardware.
+Specifically, the simulator allowed 333 times as many evaluations per run as HyperNEAT had in hardware (60000 vs.
+180)~\cite{yos:clune}, \cite{glette}.
 
-In this paper we test whether the benefits of HyperNEAT -- higher performance and the automatic discovery of complex, effective regularities \cite{clune2,yos:clune} -- will outperform a simple genetic algorithm manually designed to produce regularities once HyperNEAT is combined with a simulator.
-%JMC/SL-I don't think they used a direct encoding.
-I prefer describing it as simple and manually designed to produce specific regularities.
-/ changed to reflect that
-We test this by evolving gaits using HyperNEAT with a physical simulator and then comparing the performance of these gaits in the real-world to those that were evaluated on the QuadraTot robotic platform in previous publications.
-%JMC - Try to write without ever saying ``use'' or ``using.'' If you do, you'll find how much it is overused in writing, as evidenced by this last sentence.
-Often the word is a stand in for a different, clearer verb...and its use is ambiguous.
-Please search for this word and try to swap it out throughout the rest of the paper.
-Often times it requires restructuring a sentence, but the sentence usually ends up better.
-Similarly, make sure to always follow the word ``this'' with referent.
-This is extremely ambiguous.
-You know what you refer to when you write the sentence, but readers (often including your future self!) do not know precisely what you mean, and in science precision is critical.
-%Our experiments confirm the hypothesis: gaits evolved with HyperNEAT and then transferred to reality were the highest performing observed to date on the QuadraTot platform. 
-%That said, gaits still transferred poorly, revealing that substantial further performance gains are likely via future work that combines a simulator, HyperNEAT, and techniques that reduce the simulator-to-reality gap.
-%More broadly, this work supports previous work that evolving with a simulator and then transferring to reality is a more powerful means of achieving results than evolving directly in hardware. 
+The work with the simulator (Glette et al.~\cite{glette}) evolved gaits with a simple encoding that was manually constrained to produce specific regularities.
+The success of that work raises the question of whether the performance gains were due to the use of a simulator, or the use of a simple, manually-constrained encoding.
+We hypothesized that HyperNEAT, which has been previously shown to automatically discover complex regularities to produce high-performing gaits~\cite{clune2,yos:clune}, would outperform the simpler encoding from Glette et al.~\cite{glette} if HyperNEAT is combined with a simulator.
+Here we test that hypothesis by evolving gaits with HyperNEAT in the same simulator from Glette et al.
+and then transferring those gaits to the QuadraTot robot.
+Our experiments confirm the hypothesis: gaits evolved with HyperNEAT and then transferred to reality were the highest performing observed to date on the QuadraTot robot platform.
+
+%The results supports previous work that evolving with a simulator and then transferring to reality is a more powerful means of achieving results than evolving directly in hardware.
+%JMC: Can you add a few cites here, hopefully from our current cite list?
+I think some of the JBM papers show this.
 %Our work also confirms that generative encodings such as HyperNEAT are a powerful way to evolve gaits and, more generally, automatically exploit regularities in challenging engineering domains.

### 2012-09-24 paper_sean/robot.tex
made changes up to, but not including, the hyperneat section

@@
 % !TEX root =main.tex
-The robot platform used for this project was a quadrupedal robot named the QuadraTot\footnote{Robot plans and part files for 3D printing are freely available at http://quadratot.yosinski.com} designed in the Cornell Creative Machines Lab by Juan Zagal \cite{yos:clune}.
-It has 9 degrees-of-freedom (DOF), with 2 DOF for the inner joint and the outer joint of each leg and 1 DOF for the hip joint of the robot. 
-The QuadraTot is an open-sourced and 3-D printable robot, which allows it to be used in other labs for research, providing a platform with which people could compare the results of different gait learning algorithms.
-Currently, there are published data and results from 9 different learning algorithms on the platform from three different publications \cite{yos:clune,glette,haocheng}.
-%JMC/SL fill in INSERT, and dump all cites here.
-This is a good place to cite haochengs paper, which is fun for him and Jason, but does not distract from our story./ DONE
+We performed the experiments on the QuadraTot quadrupedal robot platform~\cite{yos:clune}.
+It has 9 degrees-of-freedom (DOF): two joints per leg and one joint that allowed rotation along the robot's midline~(Fig.\ref{pictureOfRobot}).
+%we def want a picture of the cool robot.
+I propose putting the picture of the robot on the left and the picture of the simulated robot on the right of the same figure.
+That also does a nice job of highlighting the theme of the paper (robot+sim) in one figure.
+Try to get this figure to show up as early as possible in the paper (e.g top of the 2nd page).
+QuadraTot has open source software and hardware\footnote{A list of all of the parts and all software, including the simulator, is available for free at http://quadratot.yosinski.com.} %JMC: This should really be a Creative Machines/Cornell web address.
+Can you ask Jason if he agrees and, if so, take the lead on making sure it gets made?
+, including many parts that can be 3D printed. These traits enable other labs to perform research with it, providing a platform for comparing different gait-learning algorithms. 
+To date there are results on the platform from nine different learning algorithms from three previous publications \cite{yos:clune,glette,haocheng}.
+We identified the robot's position using infrared LEDs on the robot which was tracked using a Wiimote~\cite{yos:clune}.
 
-The body of the robot was printed using an Objet Connex 500 3-D Printing System and is actuated using 9 Robotis Dynamixel servos: 5 AX-18A servos and 4 AX-12A servos. 
-The stronger AX-18A servos were used to actuate the inner joints of the robot and the hip joint while the AX-12A servos were used to actuate the outer joints. 
-The servos have a range of motion of [0, 1023], roughly equivalent to a physical range of [-12\degree, +120\degree].
+Both in this work and that of Glette et al.~\cite{glette}, the joints are powered by Robotis Dynamixel servos; five AX-18A servos for the inner joint of each leg and the single midline joint, and four AX-12A servos for the the outer joint of each leg, which requires less power and can thus have less expensive motors.
+The servos were sent new positions at 40Hz via the Pydynamixel library.
 
-The Dynamixel AX-12A and AX-18A servos have a safety feature called \emph{Alarm Shutdown} that would turn off the servo if the servo's current, range, temperature, or torque was too high \cite{robotis}.
-The ones most relevant to this study were shutdowns due to high torque and range. 
-It has been reported previously Yosinski et al.
-\cite{yos:clune}, that many gait tested on the QuadraTot forced the servos to exert too much torque or command positions that are out of range and cause the servos to shutdown.
-In order to address this problem and avoid collisions with its own body, we implemented a motion cropping system by limiting the possible range of movement of the QuadraTot by [-85\degree, +60\degree], [-113\degree, +39\degree], and [-28\degree, +28\degree] for the inner, outer, and hip joints respectively.
-In addition to this, we implemented an extra cropping system named the Smart Cropping System from Shen et al.
-\cite{haocheng} to further prevent servo shutdowns from leg positions where the combination of the two angles interact to produce an extreme amount of torque.
-%This smart cropping algorithm calculates a leg position by summing the servo position values from the inner and outer joints of a leg, \emph{\textbf{g}} (\emph{\textbf{g}} = $j_{inner}$ + $j_{outer}$) %JMC/SL - do they sum the position values, or the joint angles?
-Since all the previous extremes are expressed in angles, the writing makes me expect angles...
-/ They do use the position values and I think that it would be best to be consistent with what Jason/Haocheng did.
-This is also the way g is calculated in the simulator and the python drivers as well and more accurately reflects what we actually did.
-%and modifying the servo positions from ``fierce gaits'', which usually have relatively low \emph{\textbf{g}}s, to remap them onto the boundary of a quasi-trangular area in a 2-D space defined by $j_{inner}$ and $j_{outer}$.
-%JMC/SL - the term fierce sounds silly and non-professional (sorry!) I took out the place where you defined it, and we only use it in two other places.
-No need to invent a definition only to use it 2-3 times...can you swap in different, more descriptive words?
-/ Again, I am trying to be consistent with what Jason and Haocheng used.
-I agree that 'fierce' is not the best word but if it is different from what J/H did, then there are inconsistencies between the publications...
-%We set the minimum threshold \emph{\textbf{g}} to be 730, found through experimentation to provide a good balance between performance and safety.
-%JMC/SL - This description is not great.
-The last sentence's use of g doesn't make sense with the equation...is g a threshold or the sum?
-Please try to clean up this description.
-/ Think this clarified things
+Each servo has a built-in safety mechanism that shuts itself off to prevent damage if the servo's current, range, temperature, or torque was too high \cite{robotis}.
+During evolution, this safety mechanism frequently activated, and did so inconsistently, adding significant noise to the evaluation process.
+As pointed out in Yosinski et al.
+\cite{yos:clune}, gaits generated during evolution are highly variable and generate many shutdowns because they forced the servos to exert too much torque or command positions that are out of range.
+%JMC: Does requesting a value out of range really cause the servo to shut down?
+That seems like something we could easily fix (by never allowing an out-of-range value).
+Now that we do not allow values out of range, we still see lots of shutdowns...making me think that the problem was really torque.
+What do you think?
+To minimize shutdowns, we limited the allowable range of movement for the inner, outer, and hip joints to [-85\degree, +60\degree], [-113\degree, +39\degree], and [-28\degree, +28\degree], respectively.
+We also implemented the Smart Cropping System from Shen et al.
+\cite{haocheng}, which prevents combinations of joint positions for the inner and outer joint of each leg that generate extreme amounts of torque.
+A final method of reducing torque was to reduce the weight of the robot.
+Yosinski et al.~\cite{yos:clune} had the small Linux computer that performed all computation on the robot, but we removed it and sent commands to the robot via a cable.
 
-We controlled the robot using a CompuLab Fit-PC2, a compact computer running Ubuntu Linux 10.10 and commanded the servos at 40Hz via the Pydynamixel library.
-
-%In a previous study \cite{yos:clune}, this computer was placed on the robot to keep the robot untethered, but for this study the computer was taken off the robot to reduce weight, because the weight of the robot seemed to be one of the factors contributing to servo shutdowns.
-
-We identified the robot's position using infrared LEDs on the robot which was tracked using a Wiimote as in Yosinski et al. \cite{yos:clune}.
-%JMC/SL you sure there was more than one IR LED?
-I thought it was just one IR and one visible for us to tell if there is power to the LEDs.
-If it is just one, switch to singular instead of plural in the previous sentence / Yep there are like 3 IR LEDs being used.
-%To maximize its viewable window, the Wii remote was fixed to the ceiling pointing down. It communicated with the robot computer via bluetooth using the CWiid library \cite{yos:clune}. 
-%JMC/SL The paragraph about the Wii mote was very wordy.
-I think I trimmed it by half.
-You should try to be similarly compact the rest of the paper.
-Compare your version to mine and try to see how I condensed things.
-/ Will do boss.
-
-% % figures
 % % figure of the quadratot in its natural stance
@@
 
-%It is open-sourced and 3-D printable, which has allowed for other  researchers to use the robot for their research around the  world. 
-%JMC/SL The previous sentence (which I commented out) is alright for the main text, but figures are very expensive real estate.
-The goal of our paper is not to sell this robot, but to sell the work we did.
-/ Okay I understand
-
-%JMC/SL: If you use which instead of that, you need a comma before which.
-For example, this is wrong: There are 9 degrees of freedom: 2 per leg, and a single `hip joint' which connects  the two symmetrical halves of the robot.
-This is right.
-There are 9 degrees of freedom: 2 per leg, and a single `hip joint' that connects  the two symmetrical halves of the robot.
-This is also right: There are 9 degrees of freedom: 2 per leg, and a single `hip joint,' which connects  the two symmetrical halves of the robot.
-/ Ok
-% The robot uses Dynamixel  AX-18 and AX-12 servos for actuation. %JMC We said this already in methods
-% Figure taken from Shen et  al. \cite{haocheng}.%JMC Actually it's from yos:clune, but I think it is ok to leave this out, since this is just a stock photo of the QuadraTot.
- %JMC Let's put this figure on the first page (before the introduction), unless you can get a series of pictures that show a gait.
-(This might be hard to do).
-A new trend in papers is to have a sexy image on the first page, and Hod will want to see this.
-I also think it makes people want to read the paper more.
-
-% \begin{figure}
-% \begin{center}
-% \vspace{1cm}
-% \epsfig{file=legpos.eps, width=8cm}
-% \caption[ ]{The figure shows an example of leg positions allowed and not allowed on the robot.
-The twoleft positions are allowed.
+\figp{sixLegPositions}{.6}{Allowed and disallowed leg position angles.
+%JMC: Isn't this described in Haocheng's paper?
+I don't think we need to show this here, because it is not a new thing in our paper and is not central to our story.
+If we do keep it, instead of 3 sentences of text saying what the columns are, just put the headings ``Allowed'', ``End of Allowable Range'', and ``Disallowed''
+%The figure shows an example of leg positions allowed and not allowed on the robot.
+The two left positions are allowed.
 The middle two positions are on the borderline.
 The right two positions are considered as ``fierce'' and are disallowed.
-Instead, these gaits are cropped to the middle two positions during gait performance by the cropping mechanism placed to prevent servo shutdowns.
-Figure taken from Shen et al.
-\cite{haocheng}}
-% \end{center}
-% \end{figure}
-\figp{sixLegPositions}{.6}{The figure shows an example of leg positions allowed and not allowed on the robot.
-The twoleft positions are allowed.
-The middle two positions are on the borderline.
-The right two positions are considered as ``fierce'' and are disallowed.
-Instead, these gaits are cropped to the middle two positions during gait performance by the cropping mechanism placed to prevent servo shutdowns.
-Figure taken from Shen et al.
-\cite{haocheng}}
+Instead, these
+Disallowed gaits are cropped to the middle two positions to prevent servo shutdowns. Figure taken from Shen et al. \cite{haocheng}}
 

### 2012-09-24 paper_sean/simulator.tex
made changes up to, but not including, the hyperneat section

--- 2012-09-03 paper_sean/simulator.tex
@@
 % !TEX root =main.tex
-Simulations occurred in C++ software customized for the QuadraTot\footnote{Available for Linux and Windows at http://quadratot.yosinski.com} in the NVIDIA PhysX physics simulator, as described previously \cite{glette}.
-Specifically, the simulation captured the number and type of joints, mass, and rough shapes and lengths of the QuadraTot components. 
-%The simulator was visualized by OpenGL library to allow the user to have a more intuitive grasp of what is going on. %JMC Unnecessary detail since this paper is not about the simulator. 
-%The simulator visualization included a recording feature which allows the user to create videos of the robot in simulation.%JMC: Assumed
-%The simulator also features a command line function which allows the users to input a gait file and run it on the simulator. %JMC: Assumed
-%The simulator then outputs a file containing the X, Y, Z coordinates of the center of mass of the robot over time. %JMC: Assumed
-%In simulation, the robot traverses in the XY plane with the Z axis representing the height off the ground. %JMC Unnecessary detail
-In an attempt to more accurately model the robot behavior, we implemented the same cropping function in the simulator as in the actual robot.
-%JMC/SL - Do all results in this paper reflect this?
-/ They will soon.
-Although the simulator was originally written in Windows for the Windows operating platform we ported and ran it on Linux (Ubuntu 11.04) to seamlessly link the simulator and the already exising libraries together.
-The friction values used for this study was the same as that used in Glette et al. \cite{glette}. 
-%
-% and uses OpenGL and GLFW libraries for visualization.
-The simula- tor was originally written for the Windows operating system but was ported and run on Linux (Ubuntu 11.04) for this ex- periment.
-It features a commandline function which allows the user to specify a gait input file and extract an output file with the X, Y, Z coordinate information over time.
-The Eu- clidean distance travelled by the robot in the XY plane was used to calculate the speed of the robot in simulation and was used as the fitness value used for evolution.
-%
+The simulator\footnote{Available for Linux and Windows at http://quadratot.yosinski.com} was the same one used in Glette et al.
+\cite{glette}.
+C++ code for the the Nvidia PhysX physics engine approximately describes the QuadraTot, including the mass and size of the robot components and its degrees of freedom.
+In the simulator, each individual joint range was limit as described above, but Smart Cropping was not included because it hindered performance by limiting the types of gaits evolution could explore in early generations.
 
-% \begin{figure}
-% \begin{center}
-% \vspace{1cm}
-% \epsfig{file=simulator.eps, width=8cm}
-% \caption[ ]{Physical representation of the QuadraTot in simulation.
-It captures the important parts of the robot, such as the number and types of the joints, mass, and the rough shape and lengths.
-The simulator was written by Kyrre Glette in C++ using NVIDIA PhysX software library and visualized using OpenGL.
-Figure taken from Glette et al.
-\cite{glette}.}
-% \end{center}
-% \end{figure}
 \figp{simulator_outline}{.6}{Physical representation of the QuadraTot in simulation.

### 2012-09-24 paper_sean/introduction.tex

@@
 Robots that can move around on legs can operate in a much wider range of environments than their wheeled counterparts.
-A downside is that designing gaits for legged robots is a difficult and time-consuming process for human engineers~\cite{ridder storm, wetter green}, and must be repeated every time the design of a robot is changed or a new robot is created~\cite{hornby1}.
+A downside is that designing gaits for legged robots is a difficult and time-consuming process for human engineers~\cite{strom1999legged,wettergreen1992gait}, and must be repeated every time the design of a robot is changed or a new robot is created~\cite{hornby1}.
 Because of these difficulties in hand-designing gaits, scientists have investigated automatically producing gaits with machine learning or evolutionary algorithms, an approach that often produces better gaits than those created by human engineers~\cite{valsalam:mii,kohl:stone,hornby1,hornby2,yos:clune}.
@@
 I think some of the JBM papers show this.
+%JMC: Right now I have this commented out.
+No need to add the cites if we leave it commented out.
+
 %Our work also confirms that generative encodings such as HyperNEAT are a powerful way to evolve gaits and, more generally, automatically exploit regularities in challenging engineering domains.

### 2012-09-24 paper_sean/hyperneat.tex
Fixed comitted mis-merge as I thought best given the commit history.

--- 2012-09-23 paper_sean/hyperneat.tex
@@
 % !TEX root =main.tex
-HyperNEAT is a neuroevolution method that evolves artificial neuron networks (ANNs) using evolutionary algorithms.
-HyperNEAT has shown to evolve large neural networks that represent many features present in brains such as repetitions and regularities \cite{stanley2002evolving}. 
-This encoding has previously been applied for gait-learning \cite{yos:clune} and for evolving 2-D and 3-D objects \cite{clune:lipson20113d} and has been successful in displaying complex, natural characteristics such as symmetries and repetition of themes often found in nature.
+HyperNEAT is an algorithm for evolving artificial neural networks (ANNs)~\cite{stanley2002evolving}.
 

### 2012-09-25 paper_sean/hyperneat.tex
updated references in introduction and hyperneat.

@@
 HyperNEAT has shown to evolve large neural networks that represent many features present in brains such as repetitions and regularities . 
-This encoding has previously been applied for gait-learning \cite{yos:clune} and for evolving 2-D and 3-D objects \cite{clune:lipson} and has been successful in displaying complex, natural characteristics such as symmetries and repetition of themes often found in nature.
+This encoding has previously been applied for gait-learning \cite{yos:clune} and for evolving 2-D and 3-D objects \cite{clune:lipson20113d} and has been successful in displaying complex, natural characteristics such as symmetries and repetition of themes often found in nature.
 

### 2012-09-25 paper_sean/introduction.tex
updated references in introduction and hyperneat.

@@
 Robots that can move around on legs can operate in a much wider range of environments than their wheeled counterparts.
-A downside is that designing gaits for legged robots is a difficult and time-consuming process for human engineers~\cite{strom1999legged,wettergreen1992gait}, and must be repeated every time the design of a robot is changed or a new robot is created~\cite{hornby1}.
-Because of these difficulties in hand-designing gaits, scientists have investigated automatically producing gaits with machine learning or evolutionary algorithms, an approach that often produces better gaits than those created by human engineers~\cite{valsalam:mii,kohl:stone,hornby1,hornby2,yos:clune}.
-Both machine-created and human-created gaits perform better if they are \emph{regular}, in that they have coordinated movements, such as left-right symmetry or front-back symmetry~\cite{valsalam:mii,clune2,clune1,clune3}.
-However, in most previous works, the experimenter has had to manually decide the regularities of the evolved gaits \cite{valsalam:mii,tellez,beer,raibert} . 
+A downside is that designing gaits for legged robots is a difficult and time-consuming process for human engineers~\cite{strom1999legged,wettergreen1992gait}, and must be repeated every time the design of a robot is changed or a new robot is created~\cite{hornby2005autonomous}.
+Because of these difficulties in hand-designing gaits, scientists have investigated automatically producing gaits with machine learning or evolutionary algorithms, an approach that often produces better gaits than those created by human engineers~\cite{valsalam:mii,kohl:stone,hornby2005autonomous,hornby2003generative,yos:clune}.
+Both machine-created and human-created gaits perform better if they are \emph{regular}, in that they have coordinated movements, such as left-right symmetry or front-back symmetry~\cite{valsalam:mii,clune2011performance,clune2009evolving,clune2009sensitivity}.
+However, in most previous works, the experimenter has had to manually decide the regularities of the evolved gaits \cite{valsalam:mii,tellez,beer,raibert}. 
 Such manual intervention is time consuming, requires expert knowledge, and adds constraints that may hurt performance.
 
-Previous work has shown that the Hypercube-based NeuroEvolution of Augmenting Topologies (HyperNEAT) generative encoding \cite{stanley1} can automatically generate a variety of regular gaits and that it outperforms direct encoding controls on this task \cite{clune1}, \cite{clune2}.
+Previous work has shown that the Hypercube-based NeuroEvolution of Augmenting Topologies (HyperNEAT) generative encoding \cite{stanley2009hypercube} can automatically generate a variety of regular gaits and that it outperforms direct encoding controls on this task \cite{clune2009evolving,clune2011performance}.
 However, that work only verified these claims in simulation. 
@@
 Specifically, the simulator allowed 333 times as many evaluations per run as HyperNEAT had in hardware (60000 vs.
-180)~\cite{yos:clune}, \cite{glette}.
+180)~\cite{yos:clune,glette}.
 
@@
 The success of that work raises the question of whether the performance gains were due to the use of a simulator, or the use of a simple, manually-constrained encoding.
-We hypothesized that HyperNEAT, which has been previously shown to automatically discover complex regularities to produce high-performing gaits~\cite{clune2,yos:clune}, would outperform the simpler encoding from Glette et al.~\cite{glette} if HyperNEAT is combined with a simulator.
+We hypothesized that HyperNEAT, which has been previously shown to automatically discover complex regularities to produce high-performing gaits~\cite{clune2011performance,yos:clune}, would outperform the simpler encoding from Glette et al.~\cite{glette} if HyperNEAT is combined with a simulator.
 Here we test that hypothesis by evolving gaits with HyperNEAT in the same simulator from Glette et al.

### 2012-09-25 paper_sean/introduction.tex
minor wording changes

@@
 Robots that can move around on legs can operate in a much wider range of environments than their wheeled counterparts.
-A downside is that designing gaits for legged robots is a difficult and time-consuming process for human engineers~\cite{strom1999legged,wettergreen1992gait}, and must be repeated every time the design of a robot is changed or a new robot is created~\cite{hornby2005autonomous}.
+A downside is that designing gaits for legged robots is a difficult and time-consuming process for human engineers~\cite{strom1999legged,wettergreen1992gait}, and must be repeated every time a new robot is created or the design of an existing robot is changed~\cite{hornby2005autonomous}.
 Because of these difficulties in hand-designing gaits, scientists have investigated automatically producing gaits with machine learning or evolutionary algorithms, an approach that often produces better gaits than those created by human engineers~\cite{valsalam:mii,kohl:stone,hornby2005autonomous,hornby2003generative,yos:clune}.
-Both machine-created and human-created gaits perform better if they are \emph{regular}, in that they have coordinated movements, such as left-right symmetry or front-back symmetry~\cite{valsalam:mii,clune2011performance,clune2009evolving,clune2009sensitivity}.
+It has often been found that both machine-created and human-created gaits perform better if they are \emph{regular}, in that they have coordinated movements, such as left-right symmetry or front-back symmetry~\cite{valsalam:mii,clune2011performance,clune2009evolving,clune2009sensitivity}.
 However, in most previous works, the experimenter has had to manually decide the regularities of the evolved gaits \cite{valsalam:mii,tellez,beer,raibert}. 

### 2012-09-25 paper_sean/robot.tex
minor edits and comments

@@
 We performed the experiments on the QuadraTot quadrupedal robot platform~\cite{yos:clune}.
-It has 9 degrees-of-freedom (DOF): two joints per leg and one joint that allowed rotation along the robot's midline~(Fig.\ref{pictureOfRobot}).
+It has 9 degrees-of-freedom (DOF): two joints per leg and one joint that allowed rotation along the robot's midline~(Fig.\ref{pictureOfRobot}) \edit{fix this}.
 %we def want a picture of the cool robot.
@@
 \figp{sixLegPositions}{.6}{Allowed and disallowed leg position angles.
-%JMC: Isn't this described in Haocheng's paper?
+\edit{which are which?
+see JMC comment too} %JMC: Isn't this described in Haocheng's paper?
 I don't think we need to show this here, because it is not a new thing in our paper and is not central to our story.
@@
 % \epsfig{file=wiimote.eps, width=8cm}
-% \caption[ ]{The Wiimote used to track the motion of ther robot.
-The Wiimote was used to track the IR LED placed on top of the robot and send the location data to a nearby computer which recorded the position of ther robot with respect to time.
+% \caption[ ]{The Wiimote used to track the motion of the robot.
+The Wiimote tracked an IR LED placed on top of the robot and sent the location data to a nearby computer which recorded the position of the robot with respect to time.
 Figure taken from Yosinski et al.

### 2012-09-25 paper_sean/results.tex
changed figure placement so that two are in one line

@@
 % \end{figure}
-\figp{avg_fit_vs_gen}{.6}{Fitnesses of the champion gaits averaged over 20 runs, measured in distance travelled over 12 seconds [cm/s].
+\figgp{avg_fit_vs_gen}{.48}{avg_freq_vs_gen}{.48}{Fitnesses of the champion gaits averaged over 20 runs, measured in distance travelled over 12 seconds [cm/s].
 These gaits outperformed gaits evolved in simulation using Genetic Algorithm by 54.3\% in simulation and 5.4\% better in reality while outperforming gaits evolved using HyperNEAT in hardware by 49.5\%.}% \edit{wrong fig, regenerate as pdf and add.}}
@@
 % \end{figure}
-\figp{avg_freq_vs_gen}{.6}{Average frequencies of the gaits averaged over 20 runs.
+%\figp{avg_freq_vs_gen}{.6}{Average frequencies of the gaits averaged over 20 runs.
 After implementing a policy which punished gaits with high frequencies, HyperNEAT was able to find high-performing, low frequency gaits.

### 2012-09-25 paper_sean/hyperneat.tex
- changed image placement so that there are two figures side-by-side - removed one bug in robot.tex that said: [Fig. ???]

@@
 
-Previous work has shown that Hyper
-HyperNEAT has shown to evolve large neural networks that represent many features present in brains such as repetitions and regularities . 
+%Previous work has shown that Hyper
+HyperNEAT has shown to evolve large neural networks that represent many features present in brains such as repetitions and regularities. 
 This encoding has previously been applied for gait-learning \cite{yos:clune} and for evolving 2-D and 3-D objects \cite{clune:lipson20113d} and has been successful in displaying complex, natural characteristics such as symmetries and repetition of themes often found in nature.
@@
 Figure taken from Clune et al.
-\cite{clune2011performance}}.
+\cite{clune2011performance}.}

### 2012-09-25 paper_sean/introduction.tex
- changed image placement so that there are two figures side-by-side - removed one bug in robot.tex that said: [Fig. ???]

@@
 % !TEX root =main.tex
+\figgp{quadratotWhiteBooties}{.45}{quadratot_simulator}{.45}{\textbf{Left: }The QuadraTot robotic platform.
+The robot is a 3-D printable and serves as a good platform to compare different gait learning algorithms, with 9 algorithms tested on it thus far.
+\textbf{Right: }Physical representation of the QuadraTot in simulation.
+It captures the important parts of the robot, such as the number and types of the joints, mass, and the rough shape and lengths.
+The simulator was written by Kyrre Glette in C++ using NVIDIA PhysX software library and visualized using OpenGL.
+Figure taken from Glette et al.
+\cite{glette}.}
+
 Robots that can move around on legs can operate in a much wider range of environments than their wheeled counterparts.

### 2012-09-25 paper_sean/results.tex
- changed image placement so that there are two figures side-by-side - removed one bug in robot.tex that said: [Fig. ???]

@@
 % \end{figure}
-\figgp{avg_fit_vs_gen}{.48}{avg_freq_vs_gen}{.48}{Fitnesses of the champion gaits averaged over 20 runs, measured in distance travelled over 12 seconds [cm/s].
-These gaits outperformed gaits evolved in simulation using Genetic Algorithm by 54.3\% in simulation and 5.4\% better in reality while outperforming gaits evolved using HyperNEAT in hardware by 49.5\%.}% \edit{wrong fig, regenerate as pdf and add.}}
+\figgp{avg_fit_vs_gen}{.48}{avg_freq_vs_gen}{.48}{\textbf{Left: }Fitnesses of the champion gaits averaged over 20 runs, measured in distance travelled over 12 seconds [cm/s].
+These gaits outperformed gaits evolved in simulation using Genetic Algorithm by 54.3\% in simulation and 5.4\% better in reality while outperforming gaits evolved using HyperNEAT in hardware by 49.5\%.
+\textbf{Right: }Average frequencies of the gaits averaged over 20 runs.
+After implementing a policy which punished gaits with high frequencies, HyperNEAT was able to find high-performing, low frequency gaits.
+This policy was implemented to reduce servo shutdowns which heavily affected the reality performance of the robot.}
 

### 2012-09-25 paper_sean/robot.tex
- changed image placement so that there are two figures side-by-side - removed one bug in robot.tex that said: [Fig. ???]

@@
 % !TEX root =main.tex
+%\figgp{quadratotWhiteBooties}{.45}{quadratot_simulator}{.45}{captioncaption}
+
 We performed the experiments on the QuadraTot quadrupedal robot platform~\cite{yos:clune}.
-It has 9 degrees-of-freedom (DOF): two joints per leg and one joint that allowed rotation along the robot's midline~(Fig.\ref{pictureOfRobot}) \edit{fix this}.
+It has 9 degrees-of-freedom (DOF): two joints per leg and one joint that allowed rotation along the robot's midline.
 %we def want a picture of the cool robot.
@@
 
-\figp{sixLegPositions}{.6}{Allowed and disallowed leg position angles.
+%\figp{sixLegPositions}{.6}{Allowed and disallowed leg position angles.
 \edit{which are which?
@@
 Instead, these
-Disallowed gaits are cropped to the middle two positions to prevent servo shutdowns. Figure taken from Shen et al. \cite{haocheng}}
+%Disallowed gaits are cropped to the middle two positions to prevent servo shutdowns. Figure taken from Shen et al. \cite{haocheng}}
 

### 2012-09-25 paper_sean/simulator.tex
- changed image placement so that there are two figures side-by-side - removed one bug in robot.tex that said: [Fig. ???]

@@
 
-\figp{simulator_outline}{.6}{Physical representation of the QuadraTot in simulation.
+%\figp{simulator_outline}{.6}{Physical representation of the QuadraTot in simulation.
 It captures the important parts of the robot, such as the number and types of the joints, mass, and the rough shape and lengths.

### 2012-09-25 paper_sean/introduction.tex
changed figure caption for figure 1

--- 2012-09-25 paper_sean/introduction.tex
@@
 % !TEX root =main.tex
+\figgp{quadratotWhiteBooties}{.45}{quadratot_simulator}{.45}{The QuadraTot robot platform on which gaits were evolved.
+\textbf{Left: } The physical robot, which is composed of 3-D printable and off-the-shelf components.
+\textbf{Right: } The robot in the simulator. Figure adapted from from Glette et al. \cite{glette}.}
+
 Robots that can move around on legs can operate in a much wider range of environments than their wheeled counterparts.

### 2012-09-27 paper_sean/hyperneat.tex
finished the HyperNEAT section. some small changes to the methods section

@@
 % !TEX root =main.tex
-HyperNEAT is an algorithm for evolving artificial neural networks (ANNs)~\cite{stanley2002evolving}.
+HyperNEAT is an algorithm for evolving artificial neural networks (ANNs)~\cite{stanley2009hypercube}.
+It has been repeatedly described in detail~\cite{stanley2009hypercube, gauci2007generating, clune2011performance},  so here we provide only a summary.
+Instead of directly encoding the weights of the ANN individually on the genome, in HyperNEAT the genome is a compositional pattern producing network (CPPN)~\cite{stanley2007compositional}.
+The CPPN genome specifies the weights in a similar way to how natural organisms develop.
+In nature, phenotypic attributes are specified as a function of their geometric location, and such positional information is conveyed through chemical morphogen gradients~\cite{carroll2005endless}.
+For example, the concentration of one chemical could indicate the position along the anterior-posterior (head-to-tail) axis, and another chemical in bands could indicate if a cell is in an odd-numbered or even-numbered segment.
+Based on the relative concentrations of these chemicals, a cell can know where it is geometrically and, thus, what type of cell to become~\cite{carroll2005endless}.
 
-HyperNEAT indirectly encodes ANNs using compositional pattern producing networks (CPPNs) which represent genomes encoding ANN phenotypes \cite{stanley2009hypercube}, \cite{stanley2007CPPN}. 
-The motivation for CPPNs is that complex patterns can be produced by determining the attributes of their phenotypic components as a function of their geometric location. 
-This is based on the belief that in nature, cells in a body differentiates and specializes in different things with respect to their location. 
-In real life, a cell in natural organisms cannot determine its location in space without the help of chemical gradients but \emph{in silico}, cells can be given their geometric coordinates, which CPPNs exploit.
-%A cell in natural organisms cannot determine its location in space by itself and requires chemical gradients are used to signal its location \cite{carroll}. 
-%In contrast, \emph{in silico}, cells can be given their geometric coordinates and CPPNs use this to their advantage.
+With CPPNs this process is abstracted as a network of math functions that operate in a Cartesian geometric space.
+The coordinates of phenotypic elements are provided as inputs to the CPPN and the output specifies a phenotypic trait.
+For example, when CPPNs encode 2D pictures, the coordinates of each pixel are iteratively input into the genome and the output is the grayscale value at that coordinate~\cite{secretan2011picbreeder}.
+Because a CPPN network is composed of math functions, those functions can create geometric regularities in the phenotype.
+For example, a Gaussian function of the x-axis can provide left-right symmetry, and a repeating function (e.g.
+sine) of the y-axis could provide a related motif such as segmentation~(Fig.~\figref{hyperneat_bug_example}).
+Both 2D pictures and 3D objects evolved with CPPNs look like natural and engineered objects, and contain complex, regularities like symmetries and repeated motifs, with and without variation~\cite{secretan2011picbreeder, clune2011objects}.
 
-Each CPPN is a directed graph in which every node is a mathematical function, such as sine or Gaussian. 
-Moreover, a CPPN genome allows functions to be made of other functions, allowing coordinate frames to be combined and hierarchies to develop, allowing for several different themes to appear such as a repeating theme (sine) with symmetry (Gaussian).
-%For example, a sine function could develop a repeating theme which is passed onto a Gaussian function to create a repeating series of symmetrical motifs. 
-These functions allow evolution to exploit various properties that allow HyperNEAT to find regularities without further human intervention \cite{stanley2009hypercube,clune2009evolving}.
-%such as symmetry (e.g.
-Gaussian) and repetition (e.g.
-sine function) \cite{stanley1}.
-%Because evolution is able to exploit such properties, things like gaits are able to feature repetition and symmetry on its own without human intervention \cite{clune1}.
+\figp{hyperneat_bug_example}{.8}{Compositional Pattern Producing Networks (CPPNs).
+CPPN  genomes use mathematical functions to generate regularities such as symmetry and repetitions.
+CPPN genomes allow functions to be made of other functions, allowing multiple regularity motifs to be present.
+Figure taken from Stanley \cite{stanley2007CPPN}.}
 
-There are weighted links between nodes in CPPNs that are used to multiply the signal in each link and determine the magnitude of effect from each node. 
-In HyperNEAT, a CPPN genome takes in Cartesian coordinates (X,Y) of the source and the target nodes, and a constant bias value and outputs the weight between the input and hidden layer and a weight between the hidden layer and output (assuming that there is a hidden layer, as in the case for this study).
-All pairwise combinations of source and target nodes are passed into a CPPN to determine the weight of each ANN link.
+In HyperNEAT, CPPNs encode the weights of the connections between neurons as a function of the geometric locations of those neurons~(\figref{hyperneatExplanation}).
+The Cartesian coordinates of the two neurons at the end of each connection are input into the CPPN, and the output is the weight of that connection.
+If the output is smaller than a threshold, the weight is set to zero, functionally removing the connection.
+The process is repeated for each possible connection.
+Just as in 2D pictures and 3D objects, the CPPN can create complex, regular, geometric patterns (e.g.
+left-right symmetry or repeated modules), but in this case the patterns are in the weights of the neural network~\cite{clune2011performance, }.
+The neural regularities produced by HyperNEAT enable significantly improved performance on problems that are regular~\cite{clune2011performance, stanley2009hypercube}, including evolving quadruped gaits~\cite{clune2011performance, clune2009evolving, clune2009sensitivity}.
 
-HyperNEAT is first neuroevolutionary algorithm that has shown capabilities in exploiting the geometry of the problem \cite{stanley2009hypercube}, \cite{clune2009sensitivity}.
-This is possible because the connection weights are a function of geometric positions of these nodes. 
-If positions inputed to CPPNs represent aspects of the problem relevant to the solution, HyperNEAT could use that information to its advantage.
-
-%Previous work has shown that Hyper
-HyperNEAT has shown to evolve large neural networks that represent many features present in brains such as repetitions and regularities. 
-This encoding has previously been applied for gait-learning \cite{yos:clune} and for evolving 2-D and 3-D objects \cite{clune:lipson20113d} and has been successful in displaying complex, natural characteristics such as symmetries and repetition of themes often found in nature.
-
-HyperNEAT evolves the CPPNs according to the principles of the Neuroevolution of Augmenting Topologies (NEAT) algorithm \cite{stanley2006exploiting}. 
+In HyperNEAT, the CPPN genomes evolve via the Neuroevolution of Augmenting Topologies (NEAT) algorithm \cite{stanley2006exploiting}. 
 The NEAT algorithm has three major components \cite{stanley2006exploiting}. 
-First, it starts with small genomes that encode simple networks and complexifies them via mutations and add nodes and links to the network which allows the alogithm to evolve the network topology and weight.
-Second, NEAT has a fitness-sharing system that preserves diversity in the system and allows for new innovations to be tuned by evolution before competing them against more adapted rivals. 
-Third, the historical information stored in genes helps to perform crossover in a way that is effective, yet avoids the need for expensive topological analysis.
+First, it starts with small genomes that encode simple networks and complexifies them by mutations that add nodes and links to the network.
+Second, NEAT has a fitness-sharing system that preserves diversity and allows for new innovations to be tuned by evolution before competing them against more adapted rivals.
+Third, historical information is recorded that facilitates crossover in a way that is effective, yet avoids the need for expensive topological analysis.
 A full explanation of NEAT can be found in Stanley and Miikkulainen \cite{stanley2006exploiting}.
 
-In this study, the ANN configuration from evolving gaits with HyperNEAT in hardware \cite{yos:clune} was used. 
-In this ANN configuration, the ANN has a fixed topology that consists of three 3 X 4 Cartesian grids of nodes for the input, hidden, and output layers. 
+In this study, the ANN inputs, outputs, and the size of the hidden layer are the same as in Yosinski et al. 2011 \cite{yos:clune}. 
+The ANN has a fixed topology of three \begin{math}3\times4\end{math} Cartesian grids of nodes for the input, hidden, and output layers. 
 The inputs to the substrate were the angles requested in the prevous time step for each of the 9 joints of the robot and a sine and cosine wave to facilitate periodic motion. 
@@
 As in Yosinski et al.
-\cite{yos:clune}, we requested four times as many commanded positions from HyperNEAT ANNs and averaged over four commands at a time in order to reduce the possibility of switching from extreme negative to extreme positive numbers at a very high frequency.
+\cite{yos:clune}, to obtain each commanded position we averaged over four outputs of the network to reduce the possibility of switching from extreme negative to extreme positive numbers at a very high frequency.
 
 %%% FIGURES %%%
-
-% \begin{figure}
-% \begin{center}
-% \vspace{1cm}
-% \epsfig{file=CPPN.eps, width=8cm}
-% \caption [ ]{Compositional Pattern Producing Networks (CPPNs).
-CPPN  genomes use mathematical functions to generate regularities such as symmetry and repetitions.
-CPPN genomes allow functions to be made of other functions, allowing multiple regularity motifs to be present.
-Figure taken from Stanley \cite{stanley2}.}
-% \end{center}
-% \end{figure}
-\figp{hyperneat_bug_example}{.6}{Compositional Pattern Producing Networks (CPPNs).
-CPPN  genomes use mathematical functions to generate regularities such as symmetry and repetitions.
-CPPN genomes allow functions to be made of other functions, allowing multiple regularity motifs to be present.
-Figure taken from Stanley \cite{stanley2007CPPN}.}
 
@@
 % \end{figure}
-\figp{hyperneatExplanation}{.6}{Producing ANNs from CPPNs.
-In HyperNEAT, weights are specified as a function of the Cartesian coordinates for source and input nodes and a constant bias for each connection between the source and input.
-All pairwise combinations of source and target node coordinates are iteratively passed into CPPNs to determine the weight of each ANN link.
-Figure taken from Clune et al.
+\figp{hyperneatExplanation}{1}{A CPPN specifying a neural network.
+In HyperNEAT, weights are a function of the Cartesian coordinates of the source and target node for each connection.
+All pairwise combinations of source and target node coordinates are iteratively passed into a CPPN to determine the weight of each ANN link.
+Figure from Clune et al.
 \cite{clune2011performance}.}

### 2012-09-27 paper_sean/methods.tex
finished the HyperNEAT section. some small changes to the methods section

@@
 % !TEX root =main.tex
-Gaits were first evolved in simulation and then transferred onto the real robot, as previous done in Glette et al. \cite{glette}. 
-During the evolution process, we evolved 200 organisms for 200 generations in 20 runs with different initial conditions specified by randomly seeding the run. 
-Each gait was 12 seconds long, both in simulation and in reality. 
-All the HyperNEAT settings used for gait generation was kept the same as in Yosinski et al.
-\cite{yos:clune} except for the frequency of the sine wave function used, which was changed in order to address the servo shutdown problem mentioned previously after discovering that gaits that require servos to change their directions often were more prone to servo shutdowns than low frequency gaits.
-%(at around 1.5Hz per servo).
+We evolved gaits first in simulation and then transferred them to the physical QuadraTot robot, as in Glette et al. \cite{glette}. 
+Each run had a population size of 200 and lasted 200 generations.
+20 runs were performed, each of which was identical except for the seed provided to the random number generator, which affected stochastic events such as mutation.
+Each gait was evaluated for 12 seconds, both in simulation and in reality. 
+All HyperNEAT parameters were identical to those in Yosinski et al.
+\cite{yos:clune} except for the frequency of the sine wave input to the ANN, which was lowered to decrease the number of servo shutdowns do to high-frequency gaits.
 With the sine wave frequency used for Yosinski et al.

### 2012-09-27 paper_sean/methods.tex
JMC pass through...first forage as far as the distant plains of results. cowered in fear at the conclusion and camped for the night to gather courage

@@
 All HyperNEAT parameters were identical to those in Yosinski et al.
-\cite{yos:clune} except for the frequency of the sine wave input to the ANN, which was lowered to decrease the number of servo shutdowns do to high-frequency gaits.
-With the sine wave frequency used for Yosinski et al.
-\cite{yos:clune} (0.0265Hz), HyperNEAT often evolved high frequency gaits (over 8Hz per servo) that performed poorly in reality due to servo constraints.
-This led to dramatically different gait performance in real-life compared to simulations. 
-To address this, we changed the frequency of sine function to 0.004Hz so that HyperNEAT would lean towards evolving low frequency gaits.
+2011~\cite{yos:clune} except for the frequency of the sine wave input to the ANN, which was lowered from 0.0265Hz to 0.004Hz to decrease the number of servo shutdowns do to high-frequency gaits.
 
@@
  %Sine frequency of 0.004 was used because we experimentally found that HyperNEAT was able to produce natural and high-performing gaits with low frequencies with this frequency. 
-We also added a policy that punished high frequency gaits while evolving in simulation. 
-If the average frequency of each leg for the gait was higher than the threshold frequency, the fitness was punished by exponentially reducing the distance travelled by the robot:
+We also punished high-frequency gaits during evolving in simulation, which was effective at reducing the number of servo shutdowns observed when evolved gaits were transferred to the robot. 
+If the average frequency of each leg for the gait was higher than the threshold frequency (1.67 Hz), the fitness was punished by exponentially reducing the distance travelled by the robot:
 
@@
 
-We chose the threshold frequency for punishment to be 1.67 Hz/servo. After implementing the frequency punishment policy, the servos shutdown much less frequently.
-
-The fitness was evaluated using the following equation:
+\noindent where \emph{d} is the Euclidean distance the center of mass of the robot travelled during the gait duration. Following Clune et al. 2009~\cite{clune2009evolving}, the fitness equation was
 
 \begin{equation}
-fitness = 2^{d^{2}}
+fitness = 2^{distance^{2}}
 \end{equation}
 
-where \emph{d} is the Euclidean distance the center of mass of the robot travelled during the gait duration. %as was done in Clune et al. \cite{clune1}.
-
-%In simulation, the distance travelled was calculated using the output file generated by the simulator.
-%In real-life, the distance travelled was calculated using the log file recorded using the Wiimote system described earlier in the paper.
-For real-life trials, the actual fitnesses were not calculated and the gaits were evaluated using only the velocity of the robot.
-
-To evolve gaits using the physical simulator, we wrote a Python script\footnote{All scripts and drivers available at http://quadratot.yosinski.com} that interfaced between the simulator and HyperNEAT.
-% which allowed us to evolve gaits fast and efficiently with no human intervention.
-The runs were performed on three separate machines all running Ubuntu 11.04 using the same HyperNEAT settings and files to speed up the simulation process.
-
-After evolving the gaits in simulation, the champion gait of the last generation of each of the 20 runs was transferred onto the real robot and evaluated for its performance on the real robot.
-In real-life trials, the motion of the robot was interpolated linearly between a stationary pose and the commanded gait for two seconds at the beginning and the end. 
-The position of the robot was measured at the beginning and the end of each run, starting from the stationary pose and ending with the stationary pose. 
-If a gait performed extremely poorly due to servo shutdowns, the gait was tried again until we could conclude that poor performance is because of the gait and not due to the condition of the robot such as overheated servos.
+After evolving the gaits in simulation, the champion gait of the last generation of each of the 20 runs was transferred onto the real robot and evaluated.
+The robot's position was measured at the beginning and the end of each run, starting from the stationary pose and ending with the stationary pose. 
+If a gait was obviously affected by servo shutdowns, the gait was tried again after letting the servos cool down.

### 2012-09-27 paper_sean/results.tex
JMC pass through...first forage as far as the distant plains of results. cowered in fear at the conclusion and camped for the night to gather courage

@@
 % !TEX root =main.tex
-HyperNEAT, used with a physical simulator, produced gaits that outperformed all previous gaits tried and recorded on the QuadraTot.\footnote{Videos of the gaits in simulation and on robot available at http://quadratot.yosinski.com}
-The fastest gait published was from Kyrre et al. \cite{glette}, which recorded 17.8 cm/s on their robot. 
-For consistency, we took that gait and ran it ten times on our robot, and found that on our robot, the gait had an average speed of 12.95 cm/s and a maximum speed of 13.76 cm/s.
-The speed of the fastest gait recorded in this study was 14.5 cm/s, which is 5.4\% faster than Kyrre et al.'s \cite{glette} gait with 20000 less evaluations performed per run.
-%Kyrre et al.
-\cite{glette} published 17.8 cm/s on their robot, but to make sure that there were no differences in robots, the same gait was also tested on our robot which then recorded an average speed of 12.951 cm/s and a maximum speed of 13.76 cm/s.
-%The gait from this study outperformed the previous Kyrre et al's \cite{glette} gait by 5.4\% with much fewer evaluations performed per run, at 40000 evaluations per run as opposed to Kyrre et al's \cite{glette} 60000 evaluations per run.
-It also outperformed the best gait evolved using HyperNEAT in hardware \cite{yos:clune} by 49.5\%, most likely due to the difference in the number of evaluations between the two studies -- 40000 vs 180.
-Another minor reason for superior performance may be that frequency limiter policy that we placed may have helped in producing gaits less susceptible to shutdown while Yosinski et al.
-\cite{yos:clune} had no such implementation.
-Although most gaits started its evolution process by producing fast and high frequency gaits, with more generations, the gaits dramatically reduced its frequencies while continually increasing their fitness.
-The average frequency of the final gaits was 1.29 Hz per servo.
+Because of the noise inherent in all physical robots, especially different copies of robots, the best way to compare two algorithms is in simulation.
+The gaits evolved with HyperNEAT in simulation significantly outperformed the gaits evolved with a GA by Glette et al.
+2012~(Table~\ref{resultsTable}).
+INSERT-RES-OF-SIM-PARAGRAPH.
 
-%Gaits produced in this study confirmed that with the use of a physical simulator, HyperNEAT is able to fully exploit the geometry of the problem without human intervention, which was not evident in gait evolution using HyperNEAT in hardware \cite{yos:clune}.
-HyperNEAT took advantage of the regularity of the problem and produced gaits more regular, coordinated, and effective than all other gaits evolved for the QuadraTot platform in the past.
-This is evident in the servo position plots where the gaits produced in this study are smooth and are similar between the servos. 
-%The difference in regularity between the two studies is most likely due to the number of evaluations that were done in gait evolution.
-Additionally, the noise in the real world was extremely high, making learning in hardware difficult, which may have further impeded the search for higher-performing, regular gaits.
+Gaits evolved in the simulator with HyperNEAT and then transferred to reality (HyperNEAT+Simulator) outperformed all previous gaits reported for the QuadraTot robot platform~(Table~\ref{resultsTable}).
+Videos of the gaits in simulation and on robot can be viewed at http://MAKEMEACREATIVEMACHINESLINK.
+The fastest previously reported gait moved 17.8 cm/s, but was evaluated on a copy of the robot in Norway~\cite{glette}.
+To control for the possibility that the robot itself may be different, we ran that gait 10 times on our copy of the robot.
+We found that our robot is inherently slower, and the average speed of the Glette et al.
+2012 gait was 12.95 cm/s on our copy of the robot, with a maximum speed of 13.76 cm/s.
+The speed of the fastest gait produced in this paper was 14.5 cm/s on our copy of the robot, which is 5.4\% faster than the best gait from Glette et al.
+2012, once robot itself is controlled for.
+This result is even more significant when one considers that the gait produced by HyperNEAT with a simulator evolved with 20000 (33\%) fewer evaluations per run than Glette et al.
+2010, a decision mandated by the available computational resources.
 
-While this study was able to produce the highest fitness gait that has ever been tried on the QuadraTot, it is worth noting that most of the gaits in simulation did not transfer well to reality. 
-Many gaits that performed very well in simulation performed poorly on the real robot because of the physical limitations and differences between simulation and real-world, such as limited servo power, frictional differences, and other minor differences in the robot physique, all of which accumulates to have a large impact in performance.
-The most oft-seen problem with real trials, regardless of the actions taken to reduce it, was the servo shutdowns. 
-Because we limited gait frequencies, HyperNEAT was forced to turn to other means for maximizing distance. 
-So it pushed the servo positions to their limits by extending and retracting from one extreme to the other at a pace which was difficult for the servos to handle. 
-Another reason for the shutdowns was that the servos could not easily support the robot's weight well, which often led to the servos shutting down after a few seconds of walking off-ground.
+Glette et al.
+2012 did not report the average speed of all transferred gaits.
+Further hindering a comparison of averages is the fact that servo shutdowns were so frequent that we were unable to reliably gather data on all 20 HyperNEAT+Simulator gaits.
 
-These problems led to difficulties in repeatability. 
-The robot is weak and the power of the motors changes between and even within trials. 
-And since the gaits evolved in this study stressed the servos a lot, after one successful run, there was no guarantee that the robot would be able to perform that same gait succesfully next time. 
-In that sense, the gaits produced in this study is perhaps worse than the gaits produced by RL PoWER Spline \cite{haocheng} or by genetic algorithm with a physical simulator \cite{glette} whose gaits were repeatable and less taxing on the robot.
+HyperNEAT+Simulator gaits also outperformed the best gait evolved with HyperNEAT directly in hardware~\cite{yos:clune} by 49.5\%, most likely due to the difference in the population size (200 vs.
+9) and the number of generations (200 vs.
+20), leading to a total difference in the number of evaluations between the two studies of 40000 vs 180~\cite{yos:clune}.
+Another potential cause of improved performance is that with a simulator HyperNEAT had much less noise in the evaluation process during early generations: it could thus find coordinated, regular, gaits, which are both higher-performing and more likely to transfer.
+On the physical robot, the noise in the evaluation was substantial, preventing effective learning and the discovery of effective gaits~\cite{yos:clune}.
+Additionally, we cannot rule out that the addition in this study of a punishment for high-frequency gaits aided performance, especially since in simulation the gaits were high-frequency in a few early generations before rapidly settling at a lower frequency due to the penalty.
+It is possible that quickly leaving the high-frequency area of the search space caused a significant performance boost.
 
@@
 \hline
-                                         & Num. Evaluations  & Sim. Velocity  & Real. Velocity \\
+                                         & Evaluations  & Simulated Velocity  & Real Velocity \\
 \hline
@@
 \end{center}
+\label{resultsTable}
 \end{table}
@@
 Comparing numbers that are not right-justified is hard...so if you can't center and right-justify, then I let's see if just having it be right justified is better.
+
+Qualitatively, the HyperNEAT gaits are regular and coordinated, as has been found in simulation~\cite{clune2009evolving, clune2011performance}.
+This is evident in the servo position plots where the gaits produced in this study are smooth and are similar between the servos~\figref{servo_plot_111}.
+This result is important because it confirms that HyperNEAT can produce the important property of regularity in physically realizable applications, instead of just producing those behaviors in simulation.
+This conclusion was not obvious after HyperNEAT failed to produce substantially coordinated, regular, effective gaits on this robot platform when evolving directly on the hardware~\cite{yos:clune}.
+
+While this study was able to produce the fastest QuadraTot gait to date, most of the gaits in simulation did not transfer well to reality.
+Many gaits that performed very well in simulation performed poorly on the real robot, largely due to servos that were too weak and timed out, or because of frictional differences between simulation and reality.
+Repeated attempts to minimize these problems were unsuccessful.
+A major, unavoidable cause of the servo shutdowns was the design of the robot: the force required to lift the robot in many poses was too much for the servos.
+In the future we will switch to a robot that has more mechanical advantage, requiring less torque out of each servo, such as the Aracna platform~\cite{lohmann2012aracna}, which was designed in response to the challenges experienced with the QuadraTot platform.
 

### 2012-09-27 paper_sean/robot.tex
JMC pass through...first forage as far as the distant plains of results. cowered in fear at the conclusion and camped for the night to gather courage

@@
 % \end{figure}
-\figgp{wiiMote_crop.jpg}{.6}{robot_led_crop.jpg}{.6}{The Wiimote used to track the motion of ther robot.
+
+%\figgp{wiiMote_crop.jpg}{.6}{robot_led_crop.jpg}{.6}{The Wiimote used to track the motion of ther robot.
 The Wiimote was used to track the IR LED placed on top of the robot and send the location data to a nearby computer which recorded the position of ther robot with respect to time.

### 2012-09-28 paper_sean/conclusion.tex
big commit before meeting Jeff/Jason today. Changes: - Shorten - Citation problems - Figure updates (fonts larger) - Figure caption updates - Table right justify - Table sigfigs - Table specified whether max or avg - Adding Kyrre's data to authorlist

--- 2012-09-12 paper_sean/conclusion.tex
@@
 % !TEX root =main.tex
-HyperNEAT has proven to be an effective encoding in gait evolution with an ability to find gait regularities without human intervention \cite{clune1,clune2,yos:clune}. 
+HyperNEAT has proven to be an effective encoding in gait evolution with an ability to find gait regularities without human intervention~\cite{clune2009evolving,clune2011performance,yos:clune}.
 Glette et al.
-\cite{glette} has shown that evolving gaits in simulation using a simple algorithm with a large number of evaluations could outperform gait evolution using a more sophisticated algorithm in hardware thanks to the simulator affording a larger number of evaluations to be performed for evolution.
-In this study, we took the two -- HyperNEAT and a physical simulator -- and showed that HyperNEAT, when it could afford many evaluations, could evolve impressive gaits. 
-In simulation, HyperNEAT gaits outperformed Genetic Algorithm gaits \cite{glette} by 54.3\% and by 5.4\% in the real life trials. 
-This study also showed that HyperNEAT will exploit the characteristics of the problem, such as ferquency and regularity, in order to maximize its fitness.
+2012~\cite{glette} showed the benefits of a physical simulator by evolving, with a simple genetic algorithm, a gait that outperformed those HyperNEAT evolved in hardware.
+In this study, we took the two -- HyperNEAT and a physical simulator -- and showed that HyperNEAT's performance significantly increases with longer evolution periods.
+The gaits produced in this study outperformed the previous best gait on the QuadraTot by 54.3\% in simulation and by 5.4\% in real life trials.
+This study also showed that HyperNEAT will exploit the characteristics of the problem, such as frequency and regularity, in order to maximize its fitness.
+%In simulation, HyperNEAT gaits outperformed Genetic Algorithm gaits~\cite{glette} by 54.3\% and by 5.4\% in the real life trials.
+
 %Without frequency limiters, it would make the legs move really fast in order to increase fitness. 
 %When that was no longer allowed, it pushed the servo positions to their boundaries by increasing the stride lengths as wide as possible. 
-However, because of the physical limitations of the robot, these exploitations did not transfer well to reality. %could not be transferred well to reality.
-Moreover, the inevitable faults with the simulator (it cannot model the robot or the real world perfectly) added to the difficulty in the transfer from simulation to reality. 
-Thus, many of the gaits performed much worse than it did in simulation and had difficulty in achieving consistent results.
+However, we faced a problem in the transfer from simulation to reality.
+%because of the physical limitations of the robot, these exploitations did not transfer well to reality.
+%could not be transferred well to reality.
+One of the reasons was that the robot was greatly underpowered and could not perform the transferred gaits.
+The other reasons was the inevitable faults with the simulator (as it cannot model the robot or the real world perfectly).
+Because of these factors, many gaits performed much worse than it did in simulation and had difficulty in achieving consistent results.
 
-Next steps for this research heads towards improving the transfer process of the gaits from simulation to reality. 
-The performance gap between simulation and reality is called the \emph{reality gap} \cite{jakobi}. 
-Many different approaches to reduce the reality gap has been proposed \cite{koos1,bongard,zagal}. 
-To improve upon the results from this study, one or more of these approaches could be used to aid in the transfer to reality. 
-The one easiest and most promising to implement is the Transferability Approach \cite{koos2} which has shown to be successful in quadrapedal gait evolution. 
-HyperNEAT, used with a simulator and a transfer approach, shows much hope in producing yet a better, more coordinated, and higher performing gait on the QuadraTot platform.
+Next steps for this research heads towards improving the transfer process of the gaits from simulation to reality by reducing the \emph{reality gap}~\cite{jakobi}.
+Many different approaches to reduce the reality gap has been proposed~\cite{koos2010crossing,bongard,zagal}.
+One of the easiest and most promising of these methods is the Transferability Approach~\cite{koos2011transferability}, which has already proven to be successful in quadrapedal gait evolution.
+Another method that we could take is to evolve the simulator itself to adapt itself to the real world.
+HyperNEAT, used with a simulator and a transfer approach, shows much hope in producing yet a better, more coordinated, and higher performing gait for many robotic platforms.

### 2012-09-28 paper_sean/hyperneat.tex
big commit before meeting Jeff/Jason today. Changes: - Shorten - Citation problems - Figure updates (fonts larger) - Figure caption updates - Table right justify - Table sigfigs - Table specified whether max or avg - Adding Kyrre's data to authorlist

@@
 CPPN genomes allow functions to be made of other functions, allowing multiple regularity motifs to be present.
-Figure taken from Stanley \cite{stanley2007CPPN}.}
+Figure taken from Stanley 2007~\cite{stanley2007CPPN}.}
 
@@
 
-In HyperNEAT, the CPPN genomes evolve via the Neuroevolution of Augmenting Topologies (NEAT) algorithm \cite{stanley2006exploiting}. 
-The NEAT algorithm has three major components \cite{stanley2006exploiting}. 
+In HyperNEAT, the CPPN genomes evolve via the Neuroevolution of Augmenting Topologies (NEAT) algorithm~\cite{stanley2006exploiting}. 
+The NEAT algorithm has three major components~\cite{stanley2006exploiting}. 
 First, it starts with small genomes that encode simple networks and complexifies them by mutations that add nodes and links to the network.
@@
 Third, historical information is recorded that facilitates crossover in a way that is effective, yet avoids the need for expensive topological analysis.
-A full explanation of NEAT can be found in Stanley and Miikkulainen \cite{stanley2006exploiting}.
+A full explanation of NEAT can be found in Stanley and Miikkulainen 2006~\cite{stanley2006exploiting}.
 
-In this study, the ANN inputs, outputs, and the size of the hidden layer are the same as in Yosinski et al. 2011 \cite{yos:clune}. 
+In this study, the ANN inputs, outputs, and the size of the hidden layer are the same as in Yosinski et al. 2011~\cite{yos:clune}. 
 The ANN has a fixed topology of three \begin{math}3\times4\end{math} Cartesian grids of nodes for the input, hidden, and output layers. 
@@
 As in Yosinski et al.
-\cite{yos:clune}, to obtain each commanded position we averaged over four outputs of the network to reduce the possibility of switching from extreme negative to extreme positive numbers at a very high frequency.
+2011~\cite{yos:clune}, to obtain each commanded position we averaged over four outputs of the network to reduce the possibility of switching from extreme negative to extreme positive numbers at a very high frequency.
 
@@
 All pairwise combinations of source and target node coordinates are iteratively passed into CPPNs to determine the weight of each ANN link.
-Figure taken from Clune et al.
-\cite{clune2}}.
+Figure taken from Clune et al.~\cite{clune2}}.
 % \end{center}
@@
 Figure from Clune et al.
-\cite{clune2011performance}.}
+2011~\cite{clune2011performance}.}

### 2012-09-28 paper_sean/introduction.tex
big commit before meeting Jeff/Jason today. Changes: - Shorten - Citation problems - Figure updates (fonts larger) - Figure caption updates - Table right justify - Table sigfigs - Table specified whether max or avg - Adding Kyrre's data to authorlist

@@
 \textbf{Left: } The physical robot, which is composed of 3-D printable and off-the-shelf components.
-\textbf{Right: } The robot in the simulator. Figure adapted from from Glette et al. \cite{glette}.}
+\textbf{Right: } The robot in the simulator. Figure adapted from from Glette et al. 2012~\cite{glette}.}
 
@@
 It has often been found that both machine-created and human-created gaits perform better if they are \emph{regular}, in that they have coordinated movements, such as left-right symmetry or front-back symmetry~\cite{valsalam:mii,clune2011performance,clune2009evolving,clune2009sensitivity}.
-However, in most previous works, the experimenter has had to manually decide the regularities of the evolved gaits \cite{valsalam:mii,tellez,beer,raibert}. 
+However, in most previous works, the experimenter has had to manually decide the regularities of the evolved gaits~\cite{valsalam:mii,tellez,beer,raibert}. 
 Such manual intervention is time consuming, requires expert knowledge, and adds constraints that may hurt performance.
 
-Previous work has shown that the Hypercube-based NeuroEvolution of Augmenting Topologies (HyperNEAT) generative encoding \cite{stanley2009hypercube} can automatically generate a variety of regular gaits and that it outperforms direct encoding controls on this task \cite{clune2009evolving,clune2011performance}.
+Previous work has shown that the Hypercube-based NeuroEvolution of Augmenting Topologies (HyperNEAT) generative encoding~\cite{stanley2009hypercube} can automatically generate a variety of regular gaits and that it outperforms direct encoding controls on this task~\cite{clune2009evolving,clune2011performance}.
 However, that work only verified these claims in simulation. 
 A follow-up paper by Yosinski et al.
-evolved gaits with HyperNEAT directly in hardware on the QuadraTot robot platform and found that HyperNEAT's gaits outperformed manually designed parameterized learning algorithms, but still did not produce impressive, natural gaits \cite{yos:clune}.
-A further follow-up study built a simulator for the QuadraTot robot platform to test whether the inclusion of a simulator would improve results and found that it did:  when gaits were evolved in this simulator with a simple direct encoding and then transferred to the QuadraTot robot, the resulting gaits were faster than the gaits produced by evolving gaits with HyperNEAT directly on the robot \cite{glette}.
+evolved gaits with HyperNEAT directly in hardware on the QuadraTot robot platform and found that HyperNEAT's gaits outperformed manually designed parameterized learning algorithms, but still did not produce impressive, natural gaits~\cite{yos:clune}.
+A further follow-up study built a simulator for the QuadraTot robot platform to test whether the inclusion of a simulator would improve results and found that it did:  when gaits were evolved in this simulator with a simple direct encoding and then transferred to the QuadraTot robot, the resulting gaits were faster than the gaits produced by evolving gaits with HyperNEAT directly on the robot~\cite{glette}.
 The simulator helped because it afforded much larger population sizes and more generations than were possible when evolving directly in hardware.
@@
 
-The work with the simulator (Glette et al.~\cite{glette}) evolved gaits with a simple encoding that was manually constrained to produce specific regularities.
+The work with the simulator (Glette et al.
+2012~\cite{glette}) evolved gaits with a simple encoding that was manually constrained to produce specific regularities.
 The success of that work raises the question of whether the performance gains were due to the use of a simulator, or the use of a simple, manually-constrained encoding.
-We hypothesized that HyperNEAT, which has been previously shown to automatically discover complex regularities to produce high-performing gaits~\cite{clune2011performance,yos:clune}, would outperform the simpler encoding from Glette et al.~\cite{glette} if HyperNEAT is combined with a simulator.
+We hypothesized that HyperNEAT, which has been previously shown to automatically discover complex regularities to produce high-performing gaits~\cite{clune2011performance,yos:clune}, would outperform the simpler encoding from Glette et al.
+2012~\cite{glette} if HyperNEAT is combined with a simulator.
 Here we test that hypothesis by evolving gaits with HyperNEAT in the same simulator from Glette et al.

### 2012-09-28 paper_sean/results.tex
big commit before meeting Jeff/Jason today. Changes: - Shorten - Citation problems - Figure updates (fonts larger) - Figure caption updates - Table right justify - Table sigfigs - Table specified whether max or avg - Adding Kyrre's data to authorlist

@@
 % !TEX root =main.tex
-Because of the noise inherent in all physical robots, especially different copies of robots, the best way to compare two algorithms is in simulation.
+Because of the noise inherent in all physical robots, especially with different copies of robots, the best way to compare two algorithms is in simulation.
 The gaits evolved with HyperNEAT in simulation significantly outperformed the gaits evolved with a GA by Glette et al.
@@
 \begin{table}
-\caption{The number of evaluations per run (gen x orgs), simulated and real trial velocities in [cm/s].
+\caption{The number of evaluations per run, simulated (average) and real trial (max) velocities in [cm/s].
 Data used from Yosinksi et al.
-\cite{yos:clune}, Glette et al.
-\cite{glette}, and Shen et al.
-\cite{haocheng}.
-*17.8 cm/s was recorded on Glette et al. \cite{glette} but the same gait tried on our robot averaged over ten runs was 12.951 cm/s}  %\tablabel{results}
+2011~\cite{yos:clune}, Glette et al.
+2012~\cite{glette}, and Shen et al.
+2012~\cite{haocheng}.
+*17.8 cm/s was recorded on Glette et al. 2012~\cite{glette} but the same gait tried on our robot averaged over ten runs was 12.951 cm/s}  %\tablabel{results}
 \begin{center}
-\begin{tabular}{|l|c|c|c|c|}
+\begin{tabular}{|l|r|r|r|r|}
 \hline
@@
 \hline
-Best parameter optimization learning    &153    & --    & 4.64 \\
+Best parameter optimization learning    &153    & --    & 4.6 \\
 \hline
@@
 \hline
-Simulated GA              & 60000       & 16.4       &   12.951*     \\
+Simulated GA              & 60000       & 16.4       &   13.0*     \\
 \hline
-RL PoWER Spline                          & 300         & --         &   11.05 \\
+RL PoWER Spline                          & 300         & --         &   11.1 \\
 \hline
@@
 % \end{figure}
-\figgp{avg_fit_vs_gen}{.48}{avg_freq_vs_gen}{.48}{\textbf{Left: }Fitnesses of the champion gaits averaged over 20 runs, measured in distance travelled over 12 seconds [cm/s].
+\figgp{avg_fit_vs_gen}{.48}{avg_freq_vs_gen}{.48}{Performance plots.
+\textbf{Left: }Fitnesses of the champion gaits averaged over 20 runs, measured in distance travelled over 12 seconds [cm/s].
 These gaits outperformed gaits evolved in simulation using Genetic Algorithm by 54.3\% in simulation and 5.4\% better in reality while outperforming gaits evolved using HyperNEAT in hardware by 49.5\%.
@@
 Bottom: A plot of servo positions for a gait produced by HyperNEAT in hardware.
-Figure taken from Yosinski et al.
-\cite{yos:clune}.
+Figure taken from Yosinski et al.~\cite{yos:clune}.
 The smoothness, symmetry, and regularities of the top plot shows that by using a simulator, HyperNEAT was able to fully exploit the geometry of the problem.}
@@
 Bottom: A plot of servo positions for a gait produced by HyperNEAT in hardware.
-Figure taken from Yosinski et al.
-\cite{yos:clune}.
+Figure taken from Yosinski et al.~\cite{yos:clune}.
 The smoothness, symmetry, and regularities of the top plot shows that by using a simulator, HyperNEAT was able to fully exploit the geometry of the problem.}%\edit{wrong fig on top, regenerate top fig as pdf and add.}}
-\figp{servo_plot_111}{.6}{A plot of servo positions for a gait produced in this study.
-The smoothness, symmetry, and regularities of the top plot shows that by using a simulator, HyperNEAT was able to fully exploit the geometry of the problem.}%\edit{wrong fig on top, regenerate top fig as pdf and add.}}
+\figp{servo_plot_111}{0.95}{A plot of servo positions for a gait produced in this study.
+The smoothness, symmetry, and regularities of the top plot shows that by using a simulator, HyperNEAT was able to fully exploit the geometry of the problem.}

### 2012-09-28 paper_sean/robot.tex
big commit before meeting Jeff/Jason today. Changes: - Shorten - Citation problems - Figure updates (fonts larger) - Figure caption updates - Table right justify - Table sigfigs - Table specified whether max or avg - Adding Kyrre's data to authorlist

@@
 , including many parts that can be 3D printed. These traits enable other labs to perform research with it, providing a platform for comparing different gait-learning algorithms. 
-To date there are results on the platform from nine different learning algorithms from three previous publications \cite{yos:clune,glette,haocheng}.
+To date there are results on the platform from nine different learning algorithms from three previous publications~\cite{yos:clune,glette,haocheng}.
 We identified the robot's position using infrared LEDs on the robot which was tracked using a Wiimote~\cite{yos:clune}.
 
-Both in this work and that of Glette et al.~\cite{glette}, the joints are powered by Robotis Dynamixel servos; five AX-18A servos for the inner joint of each leg and the single midline joint, and four AX-12A servos for the the outer joint of each leg, which requires less power and can thus have less expensive motors.
+Both in this work and that of Glette et al.
+2012~\cite{glette}, the joints are powered by Robotis Dynamixel servos; five AX-18A servos for the inner joint of each leg and the single midline joint, and four AX-12A servos for the the outer joint of each leg, which requires less power and can thus have less expensive motors.
 The servos were sent new positions at 40Hz via the Pydynamixel library.
 
-Each servo has a built-in safety mechanism that shuts itself off to prevent damage if the servo's current, range, temperature, or torque was too high \cite{robotis}.
+Each servo has a built-in safety mechanism that shuts itself off to prevent damage if the servo's current, range, temperature, or torque was too high~\cite{robotis}.
 During evolution, this safety mechanism frequently activated, and did so inconsistently, adding significant noise to the evaluation process.
 As pointed out in Yosinski et al.
-\cite{yos:clune}, gaits generated during evolution are highly variable and generate many shutdowns because they forced the servos to exert too much torque or command positions that are out of range.
+2011~\cite{yos:clune}, gaits generated during evolution are highly variable and generate many shutdowns because they forced the servos to exert too much torque or command positions that are out of range.
 %JMC: Does requesting a value out of range really cause the servo to shut down?
@@
 We also implemented the Smart Cropping System from Shen et al.
-\cite{haocheng}, which prevents combinations of joint positions for the inner and outer joint of each leg that generate extreme amounts of torque.
+2012~\cite{haocheng}, which prevents combinations of joint positions for the inner and outer joint of each leg that generate extreme amounts of torque.
 A final method of reducing torque was to reduce the weight of the robot.
-Yosinski et al.~\cite{yos:clune} had the small Linux computer that performed all computation on the robot, but we removed it and sent commands to the robot via a cable.
+Yosinski et al.
+2011~\cite{yos:clune} had the small Linux computer that performed all computation on the robot, but we removed it and sent commands to the robot via a cable.
 

### 2012-10-12 paper_sean/hyperneat.tex
made every change as requested by Jeff, except for p from Wilcoxon rank sum test

@@
 HyperNEAT is an algorithm for evolving artificial neural networks (ANNs)~\cite{stanley2009hypercube}.
-It has been repeatedly described in detail~\cite{stanley2009hypercube, gauci2007generating, clune2011performance},  so here we provide only a summary.
-Instead of directly encoding the weights of the ANN individually on the genome, in HyperNEAT the genome is a compositional pattern producing network (CPPN)~\cite{stanley2007compositional}.
+It has been repeatedly described in detail~\cite{stanley2009hypercube,gauci2007generating,clune2011performance},  so here we provide only a summary.
+Instead of directly encoding the weights of the ANN individually on the genome, in HyperNEAT the genome is a compositional pattern producing network (CPPN)~\cite{stanley2007CPPN}.
 The CPPN genome specifies the weights in a similar way to how natural organisms develop.
@@
 For example, a Gaussian function of the x-axis can provide left-right symmetry, and a repeating function (e.g.
-sine) of the y-axis could provide a related motif such as segmentation~(Fig.~\figref{hyperneat_bug_example}).
-Both 2D pictures and 3D objects evolved with CPPNs look like natural and engineered objects, and contain complex, regularities like symmetries and repeated motifs, with and without variation~\cite{secretan2011picbreeder, clune2011objects}.
+sine) of the y-axis could provide a related motif such as segmentation~(\figref{hyperneat_bug_example}).
+Both 2D pictures and 3D objects evolved with CPPNs look like natural and engineered objects, and contain complex, regularities like symmetries and repeated motifs, with and without variation~\cite{secretan2011picbreeder,clune2011objects}.
 
@@
 Just as in 2D pictures and 3D objects, the CPPN can create complex, regular, geometric patterns (e.g.
-left-right symmetry or repeated modules), but in this case the patterns are in the weights of the neural network~\cite{clune2011performance, }.
-The neural regularities produced by HyperNEAT enable significantly improved performance on problems that are regular~\cite{clune2011performance, stanley2009hypercube}, including evolving quadruped gaits~\cite{clune2011performance, clune2009evolving, clune2009sensitivity}.
+left-right symmetry or repeated modules), but in this case the patterns are in the weights of the neural network~\cite{clune2011performance}.
+The neural regularities produced by HyperNEAT enable significantly improved performance on problems that are regular~\cite{clune2011performance,stanley2009hypercube}, including evolving quadruped gaits~\cite{clune2011performance,clune2009evolving,clune2009sensitivity}.
 

### 2012-10-12 paper_sean/results.tex
changed a few things, including table..and cut out some of my comments that sean has addressed
made every change as requested by Jeff, except for p from Wilcoxon rank sum test

@@
 2012~(Table~\ref{resultsTable}).
-INSERT-RES-OF-SIM-PARAGRAPH.
+We compared Kyrre et al 2012's best simulated run to the average of our 20 runs and found that our gait is, in average, greater in fitness by 7.67 cm/s, with 2.0 cm/s standard deviation, for every generation of evolution during the runs.
+Using the same data, we recorded a fitness of 21.0 cm/s with 4.29 cm/s standard deviation as the mean fitness throughout the evolutionary process while Kyrre et al 2012's data recorded a fitness of 13.35 cm/s with 4.06 cm/s standard deviation.
+These results show that in simulation, HyperNEAT's gaits evolve faster and have higher fitnesses than those of GA's.
 
-Gaits evolved in the simulator with HyperNEAT and then transferred to reality (HyperNEAT+Simulator) outperformed all previous gaits reported for the QuadraTot robot platform~(Table~\ref{resultsTable}).
-Videos of the gaits in simulation and on robot can be viewed at http://MAKEMEACREATIVEMACHINESLINK.
+Gaits evolved in the simulator with HyperNEAT and then transferred to reality (HyperNEAT+Simulator) outperformed all previous gaits reported for the QuadraTot robot platform~(Table~\ref{resultsTable}).\footnote{Videos of the gaits in simulation and on the robot can be viewed at http://creativemachines.cornell.edu/HyperNEAT-simulation-quadrupedal-gaits}
 The fastest previously reported gait moved 17.8 cm/s, but was evaluated on a copy of the robot in Norway~\cite{glette}.
@@
 The speed of the fastest gait produced in this paper was 14.5 cm/s on our copy of the robot, which is 5.4\% faster than the best gait from Glette et al.
-2012, once robot itself is controlled for.
+2012 performed on our robot.
 This result is even more significant when one considers that the gait produced by HyperNEAT with a simulator evolved with 20000 (33\%) fewer evaluations per run than Glette et al.
-2010, a decision mandated by the available computational resources.
+2012, a decision mandated by the available computational resources.
 
@@
 On the physical robot, the noise in the evaluation was substantial, preventing effective learning and the discovery of effective gaits~\cite{yos:clune}.
-Additionally, we cannot rule out that the addition in this study of a punishment for high-frequency gaits aided performance, especially since in simulation the gaits were high-frequency in a few early generations before rapidly settling at a lower frequency due to the penalty.
+Additionally, we cannot rule out that the addition in this study of a punishment for high-frequency gaits aided performance, especially since in simulation the gaits were high-frequency in a few early generations before rapidly settling at a lower frequency due to the penalty~(\figref{avg_fit_vs_gen}).
 It is possible that quickly leaving the high-frequency area of the search space caused a significant performance boost.
@@
 \hline
-Best parameter optimization learning    &153    & --    & 4.6 \\
+Parameterized gaits + optimization    &153    & --    & 4.6 \\
 \hline
-Hardware HyperNEAT                  & 180         & --         &   9.7     \\
-\hline
-Simulated GA              & 60000       & 16.4       &   13.0*     \\
+HyperNEAT in hardware                  & 180         & --         &   9.7     \\
 \hline
@@
 \hline
-Simulated HyperNEAT                      & 40000       & 25.4       &   14.5 \\
+GA + simulator              & 60000       & 16.4       &   *13.0     \\
+\hline
+HyperNEAT + simulator                      & 40000       & 25.4       &   14.5 \\
 \hline
@@
 \end{table}
-%JMC All numbers in the same column should be to the same precision (e.g.
-for the real velocity column, one decimal place).
-Also, can you right-justify the numbers (while still keeping them roughly centered)?
-Comparing numbers that are not right-justified is hard...so if you can't center and right-justify, then I let's see if just having it be right justified is better.
 
-Qualitatively, the HyperNEAT gaits are regular and coordinated, as has been found in simulation~\cite{clune2009evolving, clune2011performance}.
-This is evident in the servo position plots where the gaits produced in this study are smooth and are similar between the servos~\figref{servo_plot_111}.
+Qualitatively, the HyperNEAT gaits are regular and coordinated, as has been found in simulation~\cite{clune2009evolving,clune2011performance}.
+This is evident in the servo position plots where the gaits produced in this study are smooth and are similar between the servos~(\figref{servo_plot_111}).
 This result is important because it confirms that HyperNEAT can produce the important property of regularity in physically realizable applications, instead of just producing those behaviors in simulation.

### 2012-10-12 paper_sean/simulator.tex
JMC pass through...first forage as far as the distant plains of results. cowered in fear at the conclusion and camped for the night to gather courage
big commit before meeting Jeff/Jason today. Changes: - Shorten - Citation problems - Figure updates (fonts larger) - Figure caption updates - Table right justify - Table sigfigs - Table specified whether max or avg - Adding Kyrre's data to authorlist
made every change as requested by Jeff, except for p from Wilcoxon rank sum test

@@
 % !TEX root =main.tex
-The simulator\footnote{Available for Linux and Windows at http://quadratot.yosinski.com} was the same one used in Glette et al.
-\cite{glette}.
-C++ code for the the Nvidia PhysX physics engine approximately describes the QuadraTot, including the mass and size of the robot components and its degrees of freedom.
+The simulator was the same as Glette et al.
+2012 \cite{glette}.
+C++ code for the the Nvidia PhysX physics engine approximately describes the QuadraTot, including the mass and size of the robot components and its degrees of freedom~(\figref{quadratotWhiteBooties}).
 In the simulator, each individual joint range was limit as described above, but Smart Cropping was not included because it hindered performance by limiting the types of gaits evolution could explore in early generations.

### 2012-10-12 code/Robot.py

@@ -31,7 +31,7 @@ class RobotFailure(Exception):
 
 
 class Robot():
-    ''''''
+
 
     def __init__(self, silentNetFail = False, expectedIds = None, commandRate = 40,
                  loud = False, skipInit = False, cropMethod = 'smart'):

### 2012-10-16 paper_sean/abstract.tex
JBY edits and comments.

@@
 % !TEX root =main.tex
-Creating gaits for physical robots is a longstanding and still open challenge. 
+Creating gaits for physical robots is a longstanding and open challenge. 
 %Evolutionary algorithms can outperform human engineers in generating such gaits, but this approach historically required engineers to manually decide which types of gaits would evolve.
 Such manual intervention is time consuming, requires expert knowledge, and adds constraints that may hurt performance.
-Recently, the HyperNEAT generative encoding was shown to automatically discover a variety of gait regularities, producing fast, coordinated gaits, but only for simulated robots.
+Recently, the HyperNEAT generative encoding was shown to discover automatically a variety of gait regularities, producing fast, coordinated gaits, but only for simulated robots.
 A follow-up study found that HyperNEAT did not produce impressive gaits when they were evolved directly on a physical robot.

### 2012-10-16 paper_sean/conclusion.tex
JBY edits and comments.

@@
 The other reasons was the inevitable faults with the simulator (as it cannot model the robot or the real world perfectly).
-Because of these factors, many gaits performed much worse than it did in simulation and had difficulty in achieving consistent results.
+Because of these factors, many gaits performed much worse than they did in simulation and had difficulty in achieving consistent results.
 
-Next steps for this research heads towards improving the transfer process of the gaits from simulation to reality by reducing the \emph{reality gap}~\cite{jakobi}.
-Many different approaches to reduce the reality gap has been proposed~\cite{koos2010crossing,bongard,zagal}.
+Next steps for this research heads towards improving the transfer process of the gaits from simulation to reality by reducing or managing the \emph{reality gap}~\cite{jakobi}.
+Many different approaches to reduce the reality gap have been proposed~\cite{koos2010crossing,bongard,zagal}.
 One of the easiest and most promising of these methods is the Transferability Approach~\cite{koos2011transferability}, which has already proven to be successful in quadrapedal gait evolution.

### 2012-10-16 paper_sean/hyperneat.tex
JBY edits and comments.

@@
 % !TEX root =main.tex
+
+\subsubsection{HyperNEAT.}
+
+\editbox{JBY: this whole section could be shortened if needed. At the very least, it could be rewritten to seem less like it was ripped out of another paper. Work on this last.}
+
 HyperNEAT is an algorithm for evolving artificial neural networks (ANNs)~\cite{stanley2009hypercube}.
@@
 
-With CPPNs this process is abstracted as a network of math functions that operate in a Cartesian geometric space.
+With CPPNs this process is abstracted as a network of mathematical functions that operate in a Cartesian geometric space.
 The coordinates of phenotypic elements are provided as inputs to the CPPN and the output specifies a phenotypic trait.
@@
 As in Yosinski et al.
-2011~\cite{yos:clune}, to obtain each commanded position we averaged over four outputs of the network to reduce the possibility of switching from extreme negative to extreme positive numbers at a very high frequency.
+2011~\cite{yos:clune}, we generated pseudo-positions at 160Hz and then downsampled over consecutive blocks of 4 to obtain the actual commanded positions at 40Hz; this reduced the number of gaits which commanded switches from extreme negative to extreme positive numbers at 40Hz, an undesirable artifact which only served to tax the servos.
 

### 2012-10-16 paper_sean/methods.tex
big commit before meeting Jeff/Jason today. Changes: - Shorten - Citation problems - Figure updates (fonts larger) - Figure caption updates - Table right justify - Table sigfigs - Table specified whether max or avg - Adding Kyrre's data to authorlist
JBY edits and comments.

@@
 % !TEX root =main.tex
-We evolved gaits first in simulation and then transferred them to the physical QuadraTot robot, as in Glette et al. \cite{glette}. 
+
+\subsection{Methods}
+
+We evolved gaits first in simulation and then transferred them to the physical QuadraTot robot, as in Glette et al. 2012~\cite{glette}. 
 Each run had a population size of 200 and lasted 200 generations.
 20 runs were performed, each of which was identical except for the seed provided to the random number generator, which affected stochastic events such as mutation.
-Each gait was evaluated for 12 seconds, both in simulation and in reality. 
+Each gait was evaluated for 12 seconds \edit{JBY: how? As in fig 4 of \cite{yos:clune}, or another way?}, both in simulation and in reality. 
 All HyperNEAT parameters were identical to those in Yosinski et al.
-2011~\cite{yos:clune} except for the frequency of the sine wave input to the ANN, which was lowered from 0.0265Hz to 0.004Hz to decrease the number of servo shutdowns do to high-frequency gaits.
+2011~\cite{yos:clune} except for the frequency of the sine wave input to the ANN, which was lowered from 0.0265Hz to 0.004Hz to decrease the number of servo shutdowns due to high-frequency gaits \edit{JBY: what?
+why?}.
 
@@
 We also punished high-frequency gaits during evolving in simulation, which was effective at reducing the number of servo shutdowns observed when evolved gaits were transferred to the robot. 
-If the average frequency of each leg for the gait was higher than the threshold frequency (1.67 Hz), the fitness was punished by exponentially reducing the distance travelled by the robot:
+If the average frequency of each leg \edit{JBY: any leg or all legs?} for the gait was higher than a threshold frequency (1.67 Hz), the fitness was punished by exponentially reducing the distance travelled by the robot:
 
 \begin{equation}
-distance = distance*e^{(freq_{thr}-freq)} \\
+\mathrm{distance} = \mathrm{distance} \cdot e^{(\mathrm{freq}_{\mathrm{thr}}-\mathrm{freq})} \\
 \end{equation}
 
-\noindent where \emph{d} is the Euclidean distance the center of mass of the robot travelled during the gait duration. Following Clune et al. 2009~\cite{clune2009evolving}, the fitness equation was
+\noindent where \emph{$\mathrm{distance}$} is the Euclidean distance the center of mass of the robot travelled during the gait duration.
+Following Clune et al.
+2009~\cite{clune2009evolving}, the fitness equation was
 
 \begin{equation}
-fitness = 2^{distance^{2}}
+\mathrm{fitness} = 2^{\mathrm{distance}^{2}}
 \end{equation}

### 2012-10-16 paper_sean/results.tex
JBY edits and comments.

@@
 % !TEX root =main.tex
+
+\edit{JBY: IMHO, this section could use the most work. Maybe we can attack it together, Sean?}
+
 Because of the noise inherent in all physical robots, especially with different copies of robots, the best way to compare two algorithms is in simulation.
@@
 On the physical robot, the noise in the evaluation was substantial, preventing effective learning and the discovery of effective gaits~\cite{yos:clune}.
-Additionally, we cannot rule out that the addition in this study of a punishment for high-frequency gaits aided performance, especially since in simulation the gaits were high-frequency in a few early generations before rapidly settling at a lower frequency due to the penalty~(\figref{avg_fit_vs_gen}).
+Additionally, we cannot rule out that the addition in this study of a punishment for high-frequency gaits aided performance, especially since in simulation the gaits were high-frequency in a few early generations before rapidly settling at a lower frequency due to the penalty~(\figref{avg_fit_vs_gen}, right).
 It is possible that quickly leaving the high-frequency area of the search space caused a significant performance boost.
@@
 2012~\cite{haocheng}.
-*17.8 cm/s was recorded on Glette et al. 2012~\cite{glette} but the same gait tried on our robot averaged over ten runs was 12.951 cm/s}  %\tablabel{results}
+*17.8 cm/s was recorded on Glette et al. 2012~\cite{glette} but the same gait tried on our robot averaged over ten runs was 12.951 cm/s.}  %\tablabel{results}
 \begin{center}
@@
 \hline
-Parameterized gaits + optimization    &153    & --    & 4.6 \\
+Parameterized gaits + optimization \cite{yos:clune}   &153    & --    & 4.6 \\
 \hline
-HyperNEAT in hardware                  & 180         & --         &   9.7     \\
+HyperNEAT in hardware \cite{yos:clune}                 & 180         & --         &   9.7     \\
 \hline
-RL PoWER Spline                          & 300         & --         &   11.1 \\
+RL PoWER Spline \cite{haocheng}                         & 300         & --         &   11.1 \\
 \hline
-GA + simulator              & 60000       & 16.4       &   *13.0     \\
+GA + simulator \cite{glette}             & 60000       & 16.4       &   *13.0     \\
 \hline
-HyperNEAT + simulator                      & 40000       & 25.4       &   14.5 \\
+HyperNEAT + simulator [this paper]                     & 40000       & \textbf{25.4}       &   \textbf{14.5} \\
 \hline
@@
 Qualitatively, the HyperNEAT gaits are regular and coordinated, as has been found in simulation~\cite{clune2009evolving,clune2011performance}.
-This is evident in the servo position plots where the gaits produced in this study are smooth and are similar between the servos~(\figref{servo_plot_111}).
+This is evident in the servo position plots where the gaits produced in this study are smooth and are similar between the servos~(see \figref{servo_plot_111} for a representative example).
 This result is important because it confirms that HyperNEAT can produce the important property of regularity in physically realizable applications, instead of just producing those behaviors in simulation.
@@
 These gaits outperformed gaits evolved in simulation using Genetic Algorithm by 54.3\% in simulation and 5.4\% better in reality while outperforming gaits evolved using HyperNEAT in hardware by 49.5\%.
-\textbf{Right: }Average frequencies of the gaits averaged over 20 runs.
+\edit{JBY: is it really standard deviation or standard error?}\textbf{Right: }Average frequencies of the gaits averaged over 20 runs.
 After implementing a policy which punished gaits with high frequencies, HyperNEAT was able to find high-performing, low frequency gaits.
-This policy was implemented to reduce servo shutdowns which heavily affected the reality performance of the robot.}
+This policy was implemented to reduce servo shutdowns which heavily affected the reality performance of the robot.\edit{JBY: fig might be easier to read if y-axis was log scale?}}
 
@@
 The smoothness, symmetry, and regularities of the top plot shows that by using a simulator, HyperNEAT was able to fully exploit the geometry of the problem.}%\edit{wrong fig on top, regenerate top fig as pdf and add.}}
-\figp{servo_plot_111}{0.95}{A plot of servo positions for a gait produced in this study.
-The smoothness, symmetry, and regularities of the top plot shows that by using a simulator, HyperNEAT was able to fully exploit the geometry of the problem.}
+\figp{servo_plot_111}{0.95}{Servo positions over time (for nine servos) for a gait produced in this study.
+The smoothness, symmetry, and regularities of the gait show that by using a simulator, HyperNEAT was able to fully exploit the geometry of the problem.}

### 2012-10-16 paper_sean/robot.tex
changed a few things, including table..and cut out some of my comments that sean has addressed
made every change as requested by Jeff, except for p from Wilcoxon rank sum test
JBY edits and comments.

@@
 % !TEX root =main.tex
+\subsubsection{Robot Hardware.}
 %\figgp{quadratotWhiteBooties}{.45}{quadratot_simulator}{.45}{captioncaption}
 
-We performed the experiments on the QuadraTot quadrupedal robot platform~\cite{yos:clune}.
+We performed the experiments on the QuadraTot quadrupedal robot platform~(\figref{quadratotWhiteBooties}).
 It has 9 degrees-of-freedom (DOF): two joints per leg and one joint that allowed rotation along the robot's midline.
@@
 Try to get this figure to show up as early as possible in the paper (e.g top of the 2nd page).
-QuadraTot has open source software and hardware\footnote{A list of all of the parts and all software, including the simulator, is available for free at http://quadratot.yosinski.com.} %JMC: This should really be a Creative Machines/Cornell web address.
-Can you ask Jason if he agrees and, if so, take the lead on making sure it gets made?
-, including many parts that can be 3D printed. These traits enable other labs to perform research with it, providing a platform for comparing different gait-learning algorithms. 
+The QuadraTot hardware designs and software for this project are open source\footnote{A list of all of the parts and all software, including the simulator, is available for free at http://creativemachines.cornell.edu/evolved-quadruped-gaits}, and all hardware components are either off-the-shelf or 3D printed parts.
+These traits enable other labs to perform research using the same platform, which eases comparison of different gait-learning algorithms.
 To date there are results on the platform from nine different learning algorithms from three previous publications~\cite{yos:clune,glette,haocheng}.
-We identified the robot's position using infrared LEDs on the robot which was tracked using a Wiimote~\cite{yos:clune}.
+We identified the robot's position using an infrared LED on the robot which was tracked using a Wiimote~\cite{yos:clune}.
 
@@
 As pointed out in Yosinski et al.
-2011~\cite{yos:clune}, gaits generated during evolution are highly variable and generate many shutdowns because they forced the servos to exert too much torque or command positions that are out of range.
+2011~\cite{yos:clune}, \edit{is it better to say ``As pointed out in a previous study \cite{yos:clune}, gaits...''} gaits generated during evolution are highly variable and generate many shutdowns because they forced the servos to exert too much torque or command positions that are out of range.
 %JMC: Does requesting a value out of range really cause the servo to shut down?
@@
 What do you think?
-To minimize shutdowns, we limited the allowable range of movement for the inner, outer, and hip joints to [-85\degree, +60\degree], [-113\degree, +39\degree], and [-28\degree, +28\degree], respectively.
+To prevent collisions between different pieces of the robot's body, we limited the allowable range of movement for the inner, outer, and hip joints to [-85\degree, +60\degree], [-113\degree, +39\degree], and [-28\degree, +28\degree], respectively.
 We also implemented the Smart Cropping System from Shen et al.
@@
 2011~\cite{yos:clune} had the small Linux computer that performed all computation on the robot, but we removed it and sent commands to the robot via a cable.
-
-% % figure of the quadratot in its natural stance
-% \begin{figure}
-% \begin{center}
-% \vspace{1cm}
-% % just using a photo from Haocheng's paper! Should replace if needed be           
-% \epsfig{file=quadratot_fig.eps, width=8cm}
-% \caption[ ]{The QuadraTot robotic platform.
-It is open-sourced and 3-D printable, which has allowed for other researchers to use the robot for their research around the world.
-The robot has 9 DOF, 2 per leg and a hip joint which connects the two symmetrical halves of the robot.
-The robot uses Dynamixel AX-18 and AX-12 servos for actuation.
-Figure taken from Shen et al.
-\cite{haocheng}.}
-% \label{fig:quadratot}
-% \end{center}
-% \end{figure}
-% %
-
-%\figp{sixLegPositions}{.6}{Allowed and disallowed leg position angles.
-\edit{which are which?
-see JMC comment too} %JMC: Isn't this described in Haocheng's paper?
-I don't think we need to show this here, because it is not a new thing in our paper and is not central to our story.
-If we do keep it, instead of 3 sentences of text saying what the columns are, just put the headings ``Allowed'', ``End of Allowable Range'', and ``Disallowed''
-%The figure shows an example of leg positions allowed and not allowed on the robot.
-The two left positions are allowed.
-The middle two positions are on the borderline.
-The right two positions are considered as ``fierce'' and are disallowed.
-Instead, these
-%Disallowed gaits are cropped to the middle two positions to prevent servo shutdowns. Figure taken from Shen et al. \cite{haocheng}}
-
-% \begin{figure}
-% \begin{center}
-% \vspace{1cm}
-% \epsfig{file=gaitspace.eps, width=8cm}
-% \caption[ ]{A map of possible leg positions.
-Manually placed position cropping systems limited possible leg positions to avoid servo shutdowns and the robot hitting itself.
-The sum of the inner and outer servo positions cannot drop below 730, the smart cropping border, marked by the red line.
-Figure taken from Shen et al.
-\cite{haocheng}.}
-% \end{center}
-% \end{figure}
-%\figp{smartCrop}{.6}{A map of possible leg positions.
-Manually placed position cropping systems limited possible leg positions to avoid servo shutdowns and the robot hitting itself.
-The sum of the inner and outer servo positions cannot drop below 730, the smart cropping border, marked by the red line.
-Figure taken from Shen et al.
-\cite{haocheng}.}
-
-% \begin{figure}
-% \begin{center}
-% \vspace{1cm}
-% \epsfig{file=wiimote.eps, width=8cm}
-% \caption[ ]{The Wiimote used to track the motion of the robot.
-The Wiimote tracked an IR LED placed on top of the robot and sent the location data to a nearby computer which recorded the position of the robot with respect to time.
-Figure taken from Yosinski et al.
-\cite{yos:clune}.}
-% \end{center}
-% \end{figure}
-
-%\figgp{wiiMote_crop.jpg}{.6}{robot_led_crop.jpg}{.6}{The Wiimote used to track the motion of ther robot.
-The Wiimote was used to track the IR LED placed on top of the robot and send the location data to a nearby computer which recorded the position of ther robot with respect to time.
-Figure taken from Yosinski et al.
-\cite{yos:clune}.}

### 2012-10-16 paper_sean/simulator.tex
JBY edits and comments.

@@
 % !TEX root =main.tex
-The simulator was the same as Glette et al.
-2012 \cite{glette}.
-C++ code for the the Nvidia PhysX physics engine approximately describes the QuadraTot, including the mass and size of the robot components and its degrees of freedom~(\figref{quadratotWhiteBooties}).
+
+\subsubsection{Simulator.}
+The simulator was the same as Glette et al. 2012 \cite{glette}. C++ code for the the Nvidia PhysX physics engine\footnotemark[\value{footnote}]
+ approximately describes the QuadraTot, including the mass and size of the robot components and its degrees of freedom~(\figref{quadratotWhiteBooties}, right).
 In the simulator, each individual joint range was limit as described above, but Smart Cropping was not included because it hindered performance by limiting the types of gaits evolution could explore in early generations.

### 2012-10-21 paper_sean/methods.tex
little changes to make on sections: robot methods results
after Jason's review, before meeting J for major paper dissection.

@@
 20 runs were performed, each of which was identical except for the seed provided to the random number generator, which affected stochastic events such as mutation.
-Each gait was evaluated for 12 seconds \edit{JBY: how? As in fig 4 of \cite{yos:clune}, or another way?}, both in simulation and in reality. 
+Each gait was evaluated for 12 seconds \edit{JBY: how?
+As in fig 4 of \cite{yos:clune}, or another way?
+SSL: reality: fig 4 of \cite{yos:clune}.
+simulation: not the same, with interpolation in start but not in end}, both in simulation and in reality.
 All HyperNEAT parameters were identical to those in Yosinski et al.
-2011~\cite{yos:clune} except for the frequency of the sine wave input to the ANN, which was lowered from 0.0265Hz to 0.004Hz to decrease the number of servo shutdowns due to high-frequency gaits \edit{JBY: what?
-why?}.
+2011~\cite{yos:clune} except for the frequency of the sine wave input to the ANN, which was lowered from 0.0265Hz to 0.004Hz.
+It was necessary to lower the sine wave frequency because after experimentation, we found that the previous settings for the sine wave resulted in high-frequency gaits that the servos could not run without shutting down.
+\edit{JBY: what?
+why?
+SSL: It was necessary to lower the sine wave frequency because after experimentation, we found that the previous settings for the sine wave resulted in high-frequency gaits that the servos could not run without shutting down}.
 
@@
 We also punished high-frequency gaits during evolving in simulation, which was effective at reducing the number of servo shutdowns observed when evolved gaits were transferred to the robot.
-If the average frequency of each leg \edit{JBY: any leg or all legs?} for the gait was higher than a threshold frequency (1.67 Hz), the fitness was punished by exponentially reducing the distance travelled by the robot:
+We calculated for the average frequency of the gait per leg by summing up the number of servo direction changes in all the legs throughout a gait and dividing that number by the length of the run and the number of legs.
+If the average frequency per leg \edit{JBY: any leg or all legs?
+SSL: see change -- too lengthy?} for the gait was higher than a threshold frequency (1.67 Hz), the fitness was punished by exponentially reducing the distance travelled by the robot:
 

### 2012-10-21 paper_sean/results.tex
little changes to make on sections: robot methods results
after Jason's review, before meeting J for major paper dissection.

@@
 These gaits outperformed gaits evolved in simulation using Genetic Algorithm by 54.3\% in simulation and 5.4\% better in reality while outperforming gaits evolved using HyperNEAT in hardware by 49.5\%.
-\edit{JBY: is it really standard deviation or standard error?}\textbf{Right: }Average frequencies of the gaits averaged over 20 runs.
+\edit{JBY: is it really standard deviation or standard error?
+SSL: yes..
+believe so.
+Used python package numpy's std function to calculate for this}\textbf{Right: }Average frequencies of the gaits averaged over 20 runs.
 After implementing a policy which punished gaits with high frequencies, HyperNEAT was able to find high-performing, low frequency gaits.
-This policy was implemented to reduce servo shutdowns which heavily affected the reality performance of the robot.\edit{JBY: fig might be easier to read if y-axis was log scale?}}
+This policy was implemented to reduce servo shutdowns which heavily affected the reality performance of the robot.\edit{JBY: fig might be easier to read if y-axis was log scale?
+SSL: we chose to abandon this right?}}
 

### 2012-10-21 paper_sean/robot.tex
little changes to make on sections: robot methods results
after Jason's review, before meeting J for major paper dissection.

@@
 During evolution, this safety mechanism frequently activated, and did so inconsistently, adding significant noise to the evaluation process.
-As pointed out in Yosinski et al.
-2011~\cite{yos:clune}, \edit{is it better to say ``As pointed out in a previous study \cite{yos:clune}, gaits...''} gaits generated during evolution are highly variable and generate many shutdowns because they forced the servos to exert too much torque or command positions that are out of range.
+As pointed out in a previous study~\cite{yos:clune}, \edit{is it better to say ``As pointed out in a previous study \cite{yos:clune}, gaits...'' SSL: Change made} gaits generated during evolution are highly variable and generate many shutdowns because they forced the servos to exert too much torque or command positions that are out of range.
 %JMC: Does requesting a value out of range really cause the servo to shut down?

### 2012-10-21 paper_sean/methods.tex
minor changes with Sean

--- 2012-10-16 paper_sean/methods.tex
@@
 20 runs were performed, each of which was identical except for the seed provided to the random number generator, which affected stochastic events such as mutation.
-Each gait was evaluated for 12 seconds \edit{JBY: how? As in fig 4 of \cite{yos:clune}, or another way?}, both in simulation and in reality. 
+Each gait was evaluated for 14 seconds in reality, with the first two
+seconds being interpolated from a stationary position into the gait
+and the last two seconds from the gait back into a stationary
+position, respectively. This process was similar to that used in
+\cite{yos:clune} and resulted in effectively 12 seconds of full speed
+motion. In simulation gaits were simply evaluated for 12 seconds of
+full speed motion. 
 All HyperNEAT parameters were identical to those in Yosinski et al.
-2011~\cite{yos:clune} except for the frequency of the sine wave input to the ANN, which was lowered from 0.0265Hz to 0.004Hz to decrease the number of servo shutdowns due to high-frequency gaits \edit{JBY: what?
-why?}.
+2011~\cite{yos:clune} except for the frequency of the sine wave input to the ANN, which was lowered from 4.24Hz to 0.640Hz.
+We lowered the sine wave frequency because after experimentation, we found that the previous settings for the sine wave resulted in high-frequency gaits that the servos could not run without shutting down.
 
@@
 We also punished high-frequency gaits during evolving in simulation, which was effective at reducing the number of servo shutdowns observed when evolved gaits were transferred to the robot.
-If the average frequency of each leg \edit{JBY: any leg or all legs?} for the gait was higher than a threshold frequency (1.67 Hz), the fitness was punished by exponentially reducing the distance travelled by the robot:
+We calculated the average frequency of the gait per leg by summing up the number of servo direction changes in all the legs throughout a gait and dividing by the length of the run and the number of legs.
+If the average frequency per leg for the gait was higher than a threshold frequency (1.67 Hz), the fitness was punished by exponentially reducing the distance travelled by the robot:
 

### 2012-10-21 paper_sean/results.tex
minor changes with Sean

--- 2012-10-16 paper_sean/results.tex
@@
 \begin{table}
-\caption{The number of evaluations per run, simulated (average) and real trial (max) velocities in [cm/s].
-Data used from Yosinksi et al.
-2011~\cite{yos:clune}, Glette et al.
-2012~\cite{glette}, and Shen et al.
-2012~\cite{haocheng}.
-*17.8 cm/s was recorded on Glette et al. 2012~\cite{glette} but the same gait tried on our robot averaged over ten runs was 12.951 cm/s.}  %\tablabel{results}
 \begin{center}
@@
 \hline
-Parameterized gaits + optimization \cite{yos:clune}   &153    & --    & 4.6 \\
+Parameterized gaits + optimization \cite{yos:clune}   &153    & --    & 5.8 \\
 \hline
@@
 \label{resultsTable}
+\caption{The number of evaluations per run, simulated (average) and real trial (max) velocities in [cm/s].
+Data used from Yosinksi et al.
+2011~\cite{yos:clune}, Glette et al.
+2012~\cite{glette}, and Shen et al.
+2012~\cite{haocheng}.
+*17.8 cm/s was recorded on Glette et al. 2012~\cite{glette} but the same gait tried on our robot averaged over ten runs was 12.951 cm/s.}  %\tablabel{results}
 \end{table}
@@
 These gaits outperformed gaits evolved in simulation using Genetic Algorithm by 54.3\% in simulation and 5.4\% better in reality while outperforming gaits evolved using HyperNEAT in hardware by 49.5\%.
-\edit{JBY: is it really standard deviation or standard error?}\textbf{Right: }Average frequencies of the gaits averaged over 20 runs.
+\edit{Add Kyrre's runs, switch from STD to SE if necessary.}\textbf{Right: }Average frequencies of the gaits averaged over 20 runs.
 After implementing a policy which punished gaits with high frequencies, HyperNEAT was able to find high-performing, low frequency gaits.
-This policy was implemented to reduce servo shutdowns which heavily affected the reality performance of the robot.\edit{JBY: fig might be easier to read if y-axis was log scale?}}
+This policy was implemented to reduce servo shutdowns which heavily affected the reality performance of the robot.
+\edit{add line at 1.67 Hz}}
 

### 2012-10-21 paper_sean/robot.tex
minor changes with Sean

--- 2012-10-16 paper_sean/robot.tex
@@
 During evolution, this safety mechanism frequently activated, and did so inconsistently, adding significant noise to the evaluation process.
-As pointed out in Yosinski et al.
-2011~\cite{yos:clune}, \edit{is it better to say ``As pointed out in a previous study \cite{yos:clune}, gaits...''} gaits generated during evolution are highly variable and generate many shutdowns because they forced the servos to exert too much torque or command positions that are out of range.
+As pointed out in a previous study~\cite{yos:clune}, gaits generated during evolution are highly variable and generate many shutdowns because they forced the servos to exert too much torque or command positions that are out of range.
 %JMC: Does requesting a value out of range really cause the servo to shut down?

### 2012-10-21 paper_sean/robot.tex
updated footnote

@@
 Try to get this figure to show up as early as possible in the paper (e.g top of the 2nd page).
-The QuadraTot hardware designs and software for this project are open source\footnote{A list of all of the parts and all software, including the simulator, is available for free at http://creativemachines.cornell.edu/evolved-quadruped-gaits}, and all hardware components are either off-the-shelf or 3D printed parts.
+The QuadraTot hardware designs and software for this project are open source\footnote{A parts list, hardware CAD files, software (including the simulator), and gait videos are available at http://creativemachines.cornell.edu/evolved-quadruped-gaits}, and all hardware components are either off-the-shelf or 3D printed parts.
 These traits enable other labs to perform research using the same platform, which eases comparison of different gait-learning algorithms.

### 2012-10-22 paper_sean/robot.tex
changes to wiimote mention and suggestion of quadratot being a good platform for gait learning

@@
 The QuadraTot hardware designs and software for this project are open source\footnote{A parts list, hardware CAD files, software (including the simulator), and gait videos are available at http://creativemachines.cornell.edu/evolved-quadruped-gaits}, and all hardware components are either off-the-shelf or 3D printed parts.
-These traits enable other labs to perform research using the same platform, which eases comparison of different gait-learning algorithms.
+%These traits enable other labs to perform research using the same platform, which eases comparison of different gait-learning algorithms.
 To date there are results on the platform from nine different learning algorithms from three previous publications~\cite{yos:clune,glette,haocheng}.
-We identified the robot's position using an infrared LED on the robot which was tracked using a Wiimote~\cite{yos:clune}.
 
@@
 The servos were sent new positions at 40Hz via the Pydynamixel library.
-
 Each servo has a built-in safety mechanism that shuts itself off to prevent damage if the servo's current, range, temperature, or torque was too high~\cite{robotis}.
@@
 2011~\cite{yos:clune} had the small Linux computer that performed all computation on the robot, but we removed it and sent commands to the robot via a cable.
+
+We tracked the robot's position using an infrared LED on the robot and tracked it with a Wiimote.

### 2012-10-22 paper_sean/results.tex
Rewrote Results with Sean, plus other changes. All remaining needed changes are reflected in edit blocks.

@@
 
-\edit{JBY: IMHO, this section could use the most work. Maybe we can attack it together, Sean?}
+All learning in this study took place in simulation, but the ultimate
+goal of our evolved gaits was to perform well in reality. Thus, we
+compare our learning performance both in simulation and in hardware
+(reality) to previous studies where data is available.
 
-Because of the noise inherent in all physical robots, especially with different copies of robots, the best way to compare two algorithms is in simulation.
-The gaits evolved with HyperNEAT in simulation significantly outperformed the gaits evolved with a GA by Glette et al.
-2012~(Table~\ref{resultsTable}).
-We compared Kyrre et al 2012's best simulated run to the average of our 20 runs and found that our gait is, in average, greater in fitness by 7.67 cm/s, with 2.0 cm/s standard deviation, for every generation of evolution during the runs.
-Using the same data, we recorded a fitness of 21.0 cm/s with 4.29 cm/s standard deviation as the mean fitness throughout the evolutionary process while Kyrre et al 2012's data recorded a fitness of 13.35 cm/s with 4.06 cm/s standard deviation.
-These results show that in simulation, HyperNEAT's gaits evolve faster and have higher fitnesses than those of GA's.
+\subsubsection{Simulation}
 
-Gaits evolved in the simulator with HyperNEAT and then transferred to reality (HyperNEAT+Simulator) outperformed all previous gaits reported for the QuadraTot robot platform~(Table~\ref{resultsTable}).\footnote{Videos of the gaits in simulation and on the robot can be viewed at http://creativemachines.cornell.edu/HyperNEAT-simulation-quadrupedal-gaits}
-The fastest previously reported gait moved 17.8 cm/s, but was evaluated on a copy of the robot in Norway~\cite{glette}.
-To control for the possibility that the robot itself may be different, we ran that gait 10 times on our copy of the robot.
-We found that our robot is inherently slower, and the average speed of the Glette et al.
-2012 gait was 12.95 cm/s on our copy of the robot, with a maximum speed of 13.76 cm/s.
-The speed of the fastest gait produced in this paper was 14.5 cm/s on our copy of the robot, which is 5.4\% faster than the best gait from Glette et al.
-2012 performed on our robot.
-This result is even more significant when one considers that the gait produced by HyperNEAT with a simulator evolved with 20000 (33\%) fewer evaluations per run than Glette et al.
-2012, a decision mandated by the available computational resources.
+\edit{Punt for now, update this once we have results from Kyrre} 
+The gaits evolved with HyperNEAT in simulation
+significantly outperformed the gaits evolved with a GA by Glette et
+al. 2012~(Table~\ref{results}). We compared Kyrre et al 2012's
+best simulated run to the average of our 20 runs and found that our
+gait is, in average, greater in fitness by 7.67 cm/s, with 2.0 cm/s
+standard deviation, for every generation of evolution during the
+runs. Using the same data, we recorded a fitness of 21.0 cm/s with
+4.29 cm/s standard deviation as the mean fitness throughout the
+evolutionary process while Kyrre et al 2012's data recorded a fitness
+of 13.35 cm/s with 4.06 cm/s standard deviation. These results show
+that in simulation, HyperNEAT's gaits evolve faster and have higher
+fitnesses than those of GA's.
 
-Glette et al.
-2012 did not report the average speed of all transferred gaits.
-Further hindering a comparison of averages is the fact that servo shutdowns were so frequent that we were unable to reliably gather data on all 20 HyperNEAT+Simulator gaits.
+Also add: This result
+is even more significant when one considers that the gait produced by
+HyperNEAT with a simulator evolved with 20000 (33\%) fewer evaluations
+per run than Glette et al. 2012, a decision mandated by the available
+computational resources.
 
-HyperNEAT+Simulator gaits also outperformed the best gait evolved with HyperNEAT directly in hardware~\cite{yos:clune} by 49.5\%, most likely due to the difference in the population size (200 vs.
-9) and the number of generations (200 vs.
-20), leading to a total difference in the number of evaluations between the two studies of 40000 vs 180~\cite{yos:clune}.
-Another potential cause of improved performance is that with a simulator HyperNEAT had much less noise in the evaluation process during early generations: it could thus find coordinated, regular, gaits, which are both higher-performing and more likely to transfer.
-On the physical robot, the noise in the evaluation was substantial, preventing effective learning and the discovery of effective gaits~\cite{yos:clune}.
-Additionally, we cannot rule out that the addition in this study of a punishment for high-frequency gaits aided performance, especially since in simulation the gaits were high-frequency in a few early generations before rapidly settling at a lower frequency due to the penalty~(\figref{avg_fit_vs_gen}, right).
-It is possible that quickly leaving the high-frequency area of the search space caused a significant performance boost.
+\subsubsection{Hardware}
 
-% table!
+Gaits evolved in simulation in this study and then transferred to a
+physical robot outperformed gaits from all previous studies that were
+evolved using only hardware evaluations~\cite{yos:clune,haocheng}
+(\tabref{results}). The first important result is the confirmation of
+the hypothesis proposed earlier: that using a simulator, although
+imperfect, can benefit the evolution of gaits for the real world.
+
+The most likely explanation for this benefit is the large difference
+in the population size afforded by a simulator (200 for simulation
+vs. 9 in hardware) and the number of generations (200 vs. 20), leading
+to a total difference in the number of evaluations between the two
+studies of 40000 vs 180 per evolutionary run~\cite{yos:clune}. Another
+potential cause of improved performance is that with a simulator
+HyperNEAT had much less noise in the evaluation process during early
+generations: it could thus find coordinated, regular, gaits, which are
+both higher-performing and more likely to transfer \edit{Jeff, what do you think?}. On the physical
+robot, the noise in the evaluation was substantial, preventing
+effective learning and the discovery of effective
+gaits~\cite{yos:clune}. Moreover, we cannot rule out that the
+addition in this study of a punishment for high-frequency gaits aided
+performance, especially since in simulation the gaits were
+high-frequency in a few early generations before rapidly settling to a
+lower frequency due to the penalty~(\figref{avg_fit_vs_gen},
+right). It is possible that quickly leaving the high-frequency area of
+the search space caused a significant performance boost.
+
+In addition to showing the benefits of using a simulator, the second
+important result is that HyperNEAT is able to find faster real world gaits
+than a simpler encoding~\cite{glette} when both are evolved in simulation.
+
+However, comparing performance in hardware between studies performed in
+different laboratories is difficult, not only because reality is
+inherently noisy, but because even two copies of the same robot are
+not identical and may produce different gait speeds for the same input
+gait. In particular, one copy of the QuadraTot robot ---
+CCML\footnote{Cornell Creative Machines Lab} --- was used to evaluate
+gaits for this study and previous studies~\cite{yos:clune,haocheng},
+but a different copy of the same robot --- ROBIN\footnote{Robotics and
+Intelligent Systems, U. Oslo} --- was used for another
+study~\cite{glette}.
+
+The previous fastest gait on any copy of a QuadraTot was from Glette
+et al. 2012~\cite{glette} at 17.8 cm/s, as measured on ROBIN. The
+fastest gait measured in our study on our robot, CCML, was 14.5
+cm/s. It was unclear whether this difference in performance was due to
+the differences in the gaits themselves or dissimilarities between
+hardware. To control for this possibility, we ran the fastest gait
+from \cite{glette} 10 times on CCML and measured an average speed of
+only 12.95 cm/s (vs. 17.8 cm/s as measured on ROBIN) and a maximum of
+13.76 cm/s (\tabref{results}). Thus, it appears that our robot,
+CCML, is slightly slower than ROBIN.
+
+Further, we note that all learning for both this paper
+and \cite{glette} was done in simulation, without any influence from
+either hardware copy of the robot. Because gaits produced in this
+study performed better than all previously recorded gaits on CCML (the only
+platform on which they have been compared), we tentatively
+conclude that this higher performance would generalize to new copies
+of the QuadraTot platform.
+
+\subsubsection{Discussion}
+
 \begin{table}
@@
 \hline
-                                         & Evaluations  & Simulated Velocity  & Real Velocity \\
+                                         &              & Simulated  & Real Vel. & Real Vel.  \\
+                                         & Evaluations  & Velocity &    (CCML)     & (ROBIN) \\
 \hline
-Parameterized gaits + optimization \cite{yos:clune}   &153    & --    & 5.8 \\
+Parameterized gaits + optimization \cite{yos:clune}   &153    & --    & 5.8 & --\\
 \hline
-HyperNEAT in hardware \cite{yos:clune}                 & 180         & --         &   9.7     \\
+HyperNEAT in hardware \cite{yos:clune}                 & 180         & --         &   9.7  & --   \\
 \hline
-RL PoWER Spline \cite{haocheng}                         & 300         & --         &   11.1 \\
+RL PoWER Spline \cite{haocheng}                         & 300         & --         &   11.1 & --\\
 \hline
-GA + simulator \cite{glette}             & 60000       & 16.4       &   *13.0     \\
+GA + simulator \cite{glette}             & 60000       & 16.4       &   13.0   & \textbf{17.8}  \\
 \hline
-HyperNEAT + simulator [this paper]                     & 40000       & \textbf{25.4}       &   \textbf{14.5} \\
+HyperNEAT + simulator [this paper]                     & 40000       & \textbf{25.4}       &   \textbf{14.5} & --\\
 \hline
@@
 \end{center}
-\label{resultsTable}
-\caption{The number of evaluations per run, simulated (average) and real trial (max) velocities in [cm/s].
-Data used from Yosinksi et al.
-2011~\cite{yos:clune}, Glette et al.
-2012~\cite{glette}, and Shen et al.
-2012~\cite{haocheng}.
-*17.8 cm/s was recorded on Glette et al. 2012~\cite{glette} but the same gait tried on our robot averaged over ten runs was 12.951 cm/s.}  %\tablabel{results}
+\tablabel{results}
+\caption{Gait speeds in simulation and on each of two QuadraTot hardware platforms for this paper and three previous studies.
+We report the number of evaluations per run, simulated (average) and real trial (max) velocities in cm/s.}  %\tablabel{results}
 \end{table}
 
-Qualitatively, the HyperNEAT gaits are regular and coordinated, as has been found in simulation~\cite{clune2009evolving,clune2011performance}.
-This is evident in the servo position plots where the gaits produced in this study are smooth and are similar between the servos~(see \figref{servo_plot_111} for a representative example).
-This result is important because it confirms that HyperNEAT can produce the important property of regularity in physically realizable applications, instead of just producing those behaviors in simulation.
-This conclusion was not obvious after HyperNEAT failed to produce substantially coordinated, regular, effective gaits on this robot platform when evolving directly on the hardware~\cite{yos:clune}.
+Qualitatively, the HyperNEAT gaits are regular and coordinated, as has
+been found in previous simulation
+studies~\cite{clune2009evolving,clune2011performance}. This is evident
+in the servo position plots where the gaits produced in this study are
+smooth and are similar between the servos~(see \figref{servo_plot_111}
+for a representative example). This result is important because it
+confirms that HyperNEAT can produce the important property of
+regularity in simulation, which was not previously observed when
+evolving directly on the hardware~\cite{yos:clune}.
 
-While this study was able to produce the fastest QuadraTot gait to date, most of the gaits in simulation did not transfer well to reality.
-Many gaits that performed very well in simulation performed poorly on the real robot, largely due to servos that were too weak and timed out, or because of frictional differences between simulation and reality.
-Repeated attempts to minimize these problems were unsuccessful.
-A major, unavoidable cause of the servo shutdowns was the design of the robot: the force required to lift the robot in many poses was too much for the servos.
-In the future we will switch to a robot that has more mechanical advantage, requiring less torque out of each servo, such as the Aracna platform~\cite{lohmann2012aracna}, which was designed in response to the challenges experienced with the QuadraTot platform.
+While this study was able to produce the fastest QuadraTot gait to
+date, most of the gaits in simulation did not transfer well to
+reality. Many gaits that performed very well in simulation performed
+poorly on the real robot, largely due to servos that were too weak and
+timed out, or because of frictional differences between simulation and
+reality. Repeated attempts to minimize these problems were
+unsuccessful.  A major, unavoidable cause of the servo shutdowns was
+the design of the robot: the force required to lift the robot in many
+poses was too much for the servos. Future studies may use a
+robot that has more mechanical advantage and requires less torque from
+each servo, such as the Aracna platform~\cite{lohmann2012aracna},
+which was designed in response to the challenges experienced with the
+QuadraTot platform.
 
-%
-%
+\figgp{avg_fit_vs_gen}{.48}{avg_freq_vs_gen}{.48}
+{Performance plots. \textbf{Left: }Fitnesses of the champion gaits
+averaged over 20 runs, measured in distance travelled over 12 seconds
+in cm/s. These gaits outperformed gaits evolved in simulation using a
+genetic algorithm~\cite{glette} by 54.3\% in simulation and 5.4\% in
+reality, and they outperformed gaits evolved using HyperNEAT in
+hardware by 49.5\% (\tabref{results}). \edit{Add Kyrre's runs (+ Haocheng?), switch
+from STD to SE if necessary.}\textbf{Right: }Average frequencies of
+the gaits averaged over 20 runs. After implementing a policy which
+punished gaits with high frequencies to reduce servo shutdowns,
+HyperNEAT was able to find high-performing, low frequency
+gaits. \edit{add line at 1.67 Hz}}
 
-% \begin{figure}
-% \begin{center}
-% \vspace{1cm}
-% \epsfig{file=fit_gen.eps, width=8cm}
-% \caption[ ]{Fitnesses of the champion gaits averaged over 20 runs, measured in distance travelled over 12 seconds [cm/s].
-These gaits outperformed gaits evolved in simulation using Genetic Algorithm by 54.3\% in simulation and 5.4\% better in reality while outperforming gaits evolved using HyperNEAT in hardware by 49.5\%.}
-% \end{center}
-% \end{figure}
-\figgp{avg_fit_vs_gen}{.48}{avg_freq_vs_gen}{.48}{Performance plots.
-\textbf{Left: }Fitnesses of the champion gaits averaged over 20 runs, measured in distance travelled over 12 seconds [cm/s].
-These gaits outperformed gaits evolved in simulation using Genetic Algorithm by 54.3\% in simulation and 5.4\% better in reality while outperforming gaits evolved using HyperNEAT in hardware by 49.5\%.
-\edit{Add Kyrre's runs, switch from STD to SE if necessary.}\textbf{Right: }Average frequencies of the gaits averaged over 20 runs.
-After implementing a policy which punished gaits with high frequencies, HyperNEAT was able to find high-performing, low frequency gaits.
-This policy was implemented to reduce servo shutdowns which heavily affected the reality performance of the robot.
-\edit{add line at 1.67 Hz}}
-
-% \begin{figure}
-% \begin{center}
-% \vspace{1cm}
-% \epsfig{file=freq_fitness.eps, width=8cm}
-% \caption[ ]{Average frequencies of the gaits averaged over 20 runs.
-After implementing a policy which punished gaits with high frequencies, HyperNEAT was able to find high-performing, low frequency gaits.
-This policy was implemented to reduce servo shutdowns which heavily affected the reality performance of the robot.}
-% \end{center}
-% \end{figure}
-%\figp{avg_freq_vs_gen}{.6}{Average frequencies of the gaits averaged over 20 runs.
-After implementing a policy which punished gaits with high frequencies, HyperNEAT was able to find high-performing, low frequency gaits.
-This policy was implemented to reduce servo shutdowns which heavily affected the reality performance of the robot.} %\edit{wrong fig, regenerate as pdf and add.}}
-
-% \begin{figure}
-% \begin{center}
-% \vspace{1cm}
-% \epsfig{file=servo_pos.eps, width=8cm}
-% \caption[ ]{Top: A plot of servo positions for a gait produced in this study.
-Bottom: A plot of servo positions for a gait produced by HyperNEAT in hardware.
-Figure taken from Yosinski et al.~\cite{yos:clune}.
-The smoothness, symmetry, and regularities of the top plot shows that by using a simulator, HyperNEAT was able to fully exploit the geometry of the problem.}
-% \end{center}
-% \end{figure}
-%\figgp{servo_plot_111}{.6}{neat_110115_211410_00000_002_filt_zoom}{.6}{Top: A plot of servo positions for a gait produced in this study.
-Bottom: A plot of servo positions for a gait produced by HyperNEAT in hardware.
-Figure taken from Yosinski et al.~\cite{yos:clune}.
-The smoothness, symmetry, and regularities of the top plot shows that by using a simulator, HyperNEAT was able to fully exploit the geometry of the problem.}%\edit{wrong fig on top, regenerate top fig as pdf and add.}}
-\figp{servo_plot_111}{0.95}{Servo positions over time (for nine servos) for a gait produced in this study.
-The smoothness, symmetry, and regularities of the gait show that by using a simulator, HyperNEAT was able to fully exploit the geometry of the problem.}
+\figp{servo_plot_111}{0.95}
+{Servo positions over time (for nine servos) for a gait produced in
+this study. The smoothness, symmetry, and regularities of the gait
+show that by using a simulator, HyperNEAT was able to fully exploit
+the geometry of the problem.}

### 2012-10-24 paper_sean/results.tex
changes in results before Jeff's read - swapped figure positions: fitness and servo pos - changed figures for fitness and frequency

@@
 
-\edit{Punt for now, update this once we have results from Kyrre} 
-The gaits evolved with HyperNEAT in simulation
-significantly outperformed the gaits evolved with a GA by Glette et
-al. 2012~(Table~\ref{results}). We compared Kyrre et al 2012's
-best simulated run to the average of our 20 runs and found that our
-gait is, in average, greater in fitness by 7.67 cm/s, with 2.0 cm/s
-standard deviation, for every generation of evolution during the
-runs. Using the same data, we recorded a fitness of 21.0 cm/s with
-4.29 cm/s standard deviation as the mean fitness throughout the
-evolutionary process while Kyrre et al 2012's data recorded a fitness
-of 13.35 cm/s with 4.06 cm/s standard deviation. These results show
-that in simulation, HyperNEAT's gaits evolve faster and have higher
-fitnesses than those of GA's.
+A simulated physical environment provides a noise-free setting for
+testing learning algorithms. The resulting gaits in simulation after
+learning with HyperNEAT reveal that HyperNEAT is capable of evolving
+high fitness and highly organic-looking gaits when many evolutionary
+evaluations can be afforded.
 
-Also add: This result
-is even more significant when one considers that the gait produced by
-HyperNEAT with a simulator evolved with 20000 (33\%) fewer evaluations
-per run than Glette et al. 2012, a decision mandated by the available
-computational resources.
+We compared the average fitnesses of 20 learning runs performed in
+simulation using HyperNEAT and a GA\footnote{For a more
+robust comparison to our 20 runs, we ran 19 more runs with the same
+learning settings as used in Kyrre et al. 2012~\cite{glette}, in
+addition to the 1 run published.}, with settings used in Glette et
+al. 2012~\cite{glette} and found that HyperNEAT outperformed the
+latter: $\emph{p} < 6.8\cdot10^{-8}$, comparing the
+20 learning runs from both learning methods.\footnote{\emph{p} value generated
+with MATLAB's non-parametric Mann-Whitney \emph{U}-test} 
+HyperNEAT's gaits are in average of
+55\% faster at the final learning generation --- HyperNEAT: mean 25.4
+cm/s $\pm$ 3.4 SD, and Genetic Algorithm: mean 16.7 cm/s $\pm$ 1.9 SD
+(\tabref{results}).
+
+%Also add: This result
+%is even more significant when one considers that the gait produced by
+%HyperNEAT with a simulator evolved with 20000 (33\%) fewer evaluations
+%per run than Glette et al. 2012, a decision mandated by the available
+%computational resources.
 
@@
 generations: it could thus find coordinated, regular, gaits, which are
-both higher-performing and more likely to transfer \edit{Jeff, what do you think?}. On the physical
-robot, the noise in the evaluation was substantial, preventing
-effective learning and the discovery of effective
-gaits~\cite{yos:clune}. Moreover, we cannot rule out that the
-addition in this study of a punishment for high-frequency gaits aided
-performance, especially since in simulation the gaits were
+both higher-performing and more likely to transfer \edit{Jeff, what do
+you think?}. On the physical robot, the noise in the evaluation was
+substantial, preventing effective learning and the discovery of
+effective gaits~\cite{yos:clune}. Moreover, we cannot rule out that
+the addition in this study of a punishment for high-frequency gaits
+aided performance, especially since in simulation the gaits were
 high-frequency in a few early generations before rapidly settling to a
@@
 \hline
-GA + simulator \cite{glette}             & 60000       & 16.4       &   13.0   & \textbf{17.8}  \\
+GA + simulator \cite{glette}             & 60000       & *16.7       &   13.0   & \textbf{17.8}  \\
 \hline
@@
 \caption{Gait speeds in simulation and on each of two QuadraTot hardware platforms for this paper and three previous studies.
-We report the number of evaluations per run, simulated (average) and real trial (max) velocities in cm/s.}  %\tablabel{results}
+We report the number of evaluations per run, simulated (average) and real trial (max) velocities in cm/s.
+*Instead of using the single fitness value reported in~\cite{glette}, we ran 19 additional runs and used the average fitness of those runs.}  %\tablabel{results}
 \end{table}
@@
 regularity in simulation, which was not previously observed when
-evolving directly on the hardware~\cite{yos:clune}.
+evolving directly on the hardware~\cite{yos:clune}. HyperNEAT's higher
+performance, as well as its higher standard deviation, is likely due
+to HyperNEAT exploring and exploiting a greater variation of gaits,
+some of which are better than others.
 
@@
 
-\figgp{avg_fit_vs_gen}{.48}{avg_freq_vs_gen}{.48}
-{Performance plots. \textbf{Left: }Fitnesses of the champion gaits
+\figp{final_sean_kyrre_fitness}{0.92}
+{Fitnesses of the champion gaits evolved with HyperNEAT and a genetic algorithm
 averaged over 20 runs, measured in distance travelled over 12 seconds
-in cm/s. These gaits outperformed gaits evolved in simulation using a
+in cm/s. The dashed lines are standard deviations. HyperNEAT gaits outperformed gaits evolved in simulation using a
 genetic algorithm~\cite{glette} by 54.3\% in simulation and 5.4\% in
 reality, and they outperformed gaits evolved using HyperNEAT in
-hardware by 49.5\% (\tabref{results}). \edit{Add Kyrre's runs (+ Haocheng?), switch
-from STD to SE if necessary.}\textbf{Right: }Average frequencies of
-the gaits averaged over 20 runs. After implementing a policy which
+hardware by 49.5\% (\tabref{results}).}
+
+\figgp{servo_plot_111}{.48}{frequencyPlotThreshFinal}{.48}
+{\textbf{Left: }
+Servo positions over time (for nine servos) for a gait produced in
+this study. The smoothness, symmetry, and regularities of the gait
+show that by using a simulator, HyperNEAT was able to fully exploit
+the geometry of the problem.
+\textbf{Right: }Average frequencies of the gaits averaged over 20 runs. 
+After implementing a policy which
 punished gaits with high frequencies to reduce servo shutdowns,
 HyperNEAT was able to find high-performing, low frequency
-gaits. \edit{add line at 1.67 Hz}}
-
-\figp{servo_plot_111}{0.95}
-{Servo positions over time (for nine servos) for a gait produced in
-this study. The smoothness, symmetry, and regularities of the gait
-show that by using a simulator, HyperNEAT was able to fully exploit
-the geometry of the problem.}
+gaits.}

### 2012-10-24 paper_sean/results.tex
final changes in figure before handing it over to jeff

@@
 
-\figp{final_sean_kyrre_fitness}{0.92}
-{Fitnesses of the champion gaits evolved with HyperNEAT and a genetic algorithm
-averaged over 20 runs, measured in distance travelled over 12 seconds
-in cm/s. The dashed lines are standard deviations. HyperNEAT gaits outperformed gaits evolved in simulation using a
-genetic algorithm~\cite{glette} by 54.3\% in simulation and 5.4\% in
-reality, and they outperformed gaits evolved using HyperNEAT in
-hardware by 49.5\% (\tabref{results}).}
+\figp{FINAL_kyrre_sean_fitness}{0.92}
+{Fitnesses of the champion gaits evolved with HyperNEAT and a genetic
+algorithm~\cite{glete} averaged over 20 runs, measured in simulation
+as distance travelled over 12 seconds in cm/s. The dashed lines are
+standard deviations. HyperNEAT gaits outperformed gaits evolved in
+simulation using a genetic algorithm~\cite{glette} by 54.3\% in
+simulation and 5.4\% in reality, and they outperformed gaits evolved
+using HyperNEAT in hardware by 49.5\% (\tabref{results}).}
 

### 2012-10-29 paper_sean/hyperneat.tex
lots o editing

@@
 
-\subsubsection{HyperNEAT.}
-
-\editbox{JBY: this whole section could be shortened if needed. At the very least, it could be rewritten to seem less like it was ripped out of another paper. Work on this last.}
+\subsubsection{HyperNEAT:}
 
@@
 In nature, phenotypic attributes are specified as a function of their geometric location, and such positional information is conveyed through chemical morphogen gradients~\cite{carroll2005endless}.
-For example, the concentration of one chemical could indicate the position along the anterior-posterior (head-to-tail) axis, and another chemical in bands could indicate if a cell is in an odd-numbered or even-numbered segment.
+For example, the concentration of one chemical could indicate the position along the anterior-posterior (head-to-tail) axis, and another chemical in bands could indicate if a cell is in an odd- or even-numbered segment.
 Based on the relative concentrations of these chemicals, a cell can know where it is geometrically and, thus, what type of cell to become~\cite{carroll2005endless}.
 
-With CPPNs this process is abstracted as a network of mathematical functions that operate in a Cartesian geometric space.
+With CPPNs this process is abstracted as a network of math functions that operate in a Cartesian geometric space.
 The coordinates of phenotypic elements are provided as inputs to the CPPN and the output specifies a phenotypic trait.
@@
 sine) of the y-axis could provide a related motif such as segmentation~(\figref{hyperneat_bug_example}).
-Both 2D pictures and 3D objects evolved with CPPNs look like natural and engineered objects, and contain complex, regularities like symmetries and repeated motifs, with and without variation~\cite{secretan2011picbreeder,clune2011objects}.
+Both 2D pictures and 3D objects evolved with CPPNs look like natural and engineered objects, and contain complex, regularities such as symmetries and repeated motifs, with and without variation~\cite{secretan2011picbreeder,clune2011objects}.
 
@@
 
-%%% FIGURES %%%
-
-% \begin{figure}
-% \begin{center}
-% \vspace{1cm}
-% \epsfig{file=hyperneat.eps, width=8cm}
-% \caption [ ]{Producing ANNs from CPPNs.
-In HyperNEAT, weights are specified as a function of the Cartesian coordinates for source and input nodes and a constant bias for each connection between the source and input.
-All pairwise combinations of source and target node coordinates are iteratively passed into CPPNs to determine the weight of each ANN link.
-Figure taken from Clune et al.~\cite{clune2}}.
-% \end{center}
-% \end{figure}
 \figp{hyperneatExplanation}{1}{A CPPN specifying a neural network.

### 2012-10-29 paper_sean/results.tex
lots o editing

--- 2012-10-22 paper_sean/results.tex
@@
 
-All learning in this study took place in simulation, but the ultimate
-goal of our evolved gaits was to perform well in reality. Thus, we
-compare our learning performance both in simulation and in hardware
-(reality) to previous studies where data is available.
+%We
+%compare performance in simulation and hardware
+%to available data from previous studies.
+%\subsubsection{Simulation:}
 
-\subsubsection{Simulation}
+The simulator enabled HyperNEAT to evolve
+fast gaits that look like those of natural animals.
+In simulation, HyperNEAT gaits were significantly faster than those from Glette et al.
+(\figref{FINAL_kyrre_sean_fitness}, $\emph{p} < 6.8\times10^{-8}$, comparing the best gaits in the final generation of each run via MATLAB's Wilcoxon rank sum test).
+Specifically, HyperNEAT gaits were 55\% faster on average in simulation (25.4
+cm/s $\pm$ 3.4 SD versus 16.7 cm/s $\pm$ 1.9 SD, \tabref{results}).\edit{Table 1, not 3} 
+Plots of servo positions over time reveal that the evolved HyperNEAT gaits are regular, smooth and coordinated~(\figref{servo_plot_111}), confirming previous results with HyperNEAT in simulation~\cite{clune2009evolving,clune2011performance}.
 
-\edit{Punt for now, update this once we have results from Kyrre} 
-The gaits evolved with HyperNEAT in simulation
-significantly outperformed the gaits evolved with a GA by Glette et
-al. 2012~(Table~\ref{results}). We compared Kyrre et al 2012's
-best simulated run to the average of our 20 runs and found that our
-gait is, in average, greater in fitness by 7.67 cm/s, with 2.0 cm/s
-standard deviation, for every generation of evolution during the
-runs. Using the same data, we recorded a fitness of 21.0 cm/s with
-4.29 cm/s standard deviation as the mean fitness throughout the
-evolutionary process while Kyrre et al 2012's data recorded a fitness
-of 13.35 cm/s with 4.06 cm/s standard deviation. These results show
-that in simulation, HyperNEAT's gaits evolve faster and have higher
-fitnesses than those of GA's.
+\figp{FINAL_kyrre_sean_fitness}{0.62}
+{HyperNEAT outperforms a genetic algorithm with a simpler encoding, when both algorithms are combined with a simulator.
+Plotted are means over 20 runs in simulation (solid lines) $\pm$ STD (dashed lines).
+HyperNEAT gaits are 54.3\% faster in
+simulation and 5.4\% faster in reality~(\tabref{results}).}
 
-Also add: This result
-is even more significant when one considers that the gait produced by
-HyperNEAT with a simulator evolved with 20000 (33\%) fewer evaluations
-per run than Glette et al. 2012, a decision mandated by the available
-computational resources.
+\figgp{servo_plot_111}{.48}{frequencyPlotThreshFinal}{.48}
+{\textbf{Left: }
+Servo positions over time (for nine servos) for a simulated HyperNEAT gait. HyperNEAT produces gaits that are smooth, symmetrical, and contain other regularities. 
+\textbf{Right: }Mean gait frequency averaged over 20 runs. 
+Gaits with frequencies above a threshold (horizontal line) receive a fitness penalty. HyperNEAT quickly learns to produce gaits with frequencies low enough to avoid this penalty. }
 
-\subsubsection{Hardware}
-
-Gaits evolved in simulation in this study and then transferred to a
-physical robot outperformed gaits from all previous studies that were
-evolved using only hardware evaluations~\cite{yos:clune,haocheng}
-(\tabref{results}). The first important result is the confirmation of
-the hypothesis proposed earlier: that using a simulator, although
-imperfect, can benefit the evolution of gaits for the real world.
-
-The most likely explanation for this benefit is the large difference
-in the population size afforded by a simulator (200 for simulation
-vs. 9 in hardware) and the number of generations (200 vs. 20), leading
-to a total difference in the number of evaluations between the two
-studies of 40000 vs 180 per evolutionary run~\cite{yos:clune}. Another
-potential cause of improved performance is that with a simulator
-HyperNEAT had much less noise in the evaluation process during early
-generations: it could thus find coordinated, regular, gaits, which are
-both higher-performing and more likely to transfer \edit{Jeff, what do you think?}. On the physical
-robot, the noise in the evaluation was substantial, preventing
-effective learning and the discovery of effective
-gaits~\cite{yos:clune}. Moreover, we cannot rule out that the
-addition in this study of a punishment for high-frequency gaits aided
-performance, especially since in simulation the gaits were
-high-frequency in a few early generations before rapidly settling to a
-lower frequency due to the penalty~(\figref{avg_fit_vs_gen},
-right). It is possible that quickly leaving the high-frequency area of
-the search space caused a significant performance boost.
-
-In addition to showing the benefits of using a simulator, the second
-important result is that HyperNEAT is able to find faster real world gaits
-than a simpler encoding~\cite{glette} when both are evolved in simulation.
-
+On the physical robot, HyperNEAT gaits from this study outperformed gaits from all previous Quadratot studies~(\tabref{results})~\cite{yos:clune,haocheng,glette}, including those of Glette et al. 
  However, comparing performance in hardware between studies performed in
@@
 not identical and may produce different gait speeds for the same input
-gait. In particular, one copy of the QuadraTot robot ---
-CCML\footnote{Cornell Creative Machines Lab} --- was used to evaluate
-gaits for this study and previous studies~\cite{yos:clune,haocheng},
-but a different copy of the same robot --- ROBIN\footnote{Robotics and
-Intelligent Systems, U. Oslo} --- was used for another
-study~\cite{glette}.
+gait.
+Gaits for this study and two previous studies~\cite{yos:clune,haocheng} were evaluated on a copy of the QuadraTot robot in the Cornell Creative Machines Lab (CCML), while the gaits in Glette et al.
+were evaluated on a different copy of the same robot in the Robotics and
+Intelligent Systems (ROBIN) lab at the University of Oslo.
 
 The previous fastest gait on any copy of a QuadraTot was from Glette
-et al. 2012~\cite{glette} at 17.8 cm/s, as measured on ROBIN. The
-fastest gait measured in our study on our robot, CCML, was 14.5
-cm/s. It was unclear whether this difference in performance was due to
+et al., and traveled 17.8 cm/s on the ROBIN QuadraTot. The
+fastest gait produced by HyperNEAT with a simulator in the experiments for this paper traveled 14.5
+cm/s on the CCML QuadraTot. It was unclear whether this difference in performance was due to
 the differences in the gaits themselves or dissimilarities between
 hardware. To control for this possibility, we ran the fastest gait
-from \cite{glette} 10 times on CCML and measured an average speed of
-only 12.95 cm/s (vs. 17.8 cm/s as measured on ROBIN) and a maximum of
-13.76 cm/s (\tabref{results}). Thus, it appears that our robot,
-CCML, is slightly slower than ROBIN.
-
-Further, we note that all learning for both this paper
-and \cite{glette} was done in simulation, without any influence from
-either hardware copy of the robot. Because gaits produced in this
-study performed better than all previously recorded gaits on CCML (the only
-platform on which they have been compared), we tentatively
-conclude that this higher performance would generalize to new copies
-of the QuadraTot platform.
-
-\subsubsection{Discussion}
+from Glette et al. 10 times on CCML and measured an average speed of
+only 12.95 cm/s (vs. 17.8 cm/s measured on ROBIN) and a maximum of
+13.76 cm/s (\tabref{results}).
+It thus appears that the CCML version of the robot is slower.
+Moreover, the HyperNEAT gait (14.5 cm/s) is faster than the best gait of Glette et al.
+on the CCML QuadraTot.
+Unfortunately, it was not possible to test the best HyperNEAT gait on the ROBIN QuadraTot.
+Because HyperNEAT outperformed the simple encoding from Glette et al.
+on the same robotic hardware, we tentatively
+conclude that HyperNEAT produces faster gaits for the QuadraTot robot. This conclusion is supported by the fact that HyperNEAT outperformed the encoding from Glette et al. in simulation.
 
@@
 \tablabel{results}
-\caption{Gait speeds in simulation and on each of two QuadraTot hardware platforms for this paper and three previous studies.
-We report the number of evaluations per run, simulated (average) and real trial (max) velocities in cm/s.}  %\tablabel{results}
+\caption{Velocities of learned gaits in simulation and on two different copies of the QuadraTot robot.
+Subject to availability, data are reported from the experiments for this paper and three previous studies.
+Reported are the total number of evaluations per run, the average of the fastest gaits produced in each run in simulation, and the fastest gait produced on a physical robot, either the QuadraTot in the Cornell Creative Machines Lab (CCML) or the QuadraTot in the Robotics and Intelligent Systems (ROBIN) lab at the University of Oslo.
+*Instead of using the single fitness value reported in~\cite{glette}, we ran 19 additional runs and used the average fitness of those 20 runs.
+Velocities are in cm/s, and bold indicates the best performance.}
 \end{table}
 
-Qualitatively, the HyperNEAT gaits are regular and coordinated, as has
-been found in previous simulation
-studies~\cite{clune2009evolving,clune2011performance}. This is evident
-in the servo position plots where the gaits produced in this study are
-smooth and are similar between the servos~(see \figref{servo_plot_111}
-for a representative example). This result is important because it
-confirms that HyperNEAT can produce the important property of
-regularity in simulation, which was not previously observed when
-evolving directly on the hardware~\cite{yos:clune}.
+We now discuss the differences between the gaits evolved by HyperNEAT directly on the physical robot, and those first evolved in a simulator and then transferred to the robot. 
+On the physical robot, the gaits produced by HyperNEAT with a simulator were faster, more natural, and more repeatable than those evolved directly on the Quadratot robot~\cite{yos:clune}.
+The gaits were also more regular, as they were in simulation~(\figref{servo_plot_111}).
+This result is important because it confirms that HyperNEAT can produce the important property of
+regularity in a challenging, real-world domain, which was not previously observed when
+evolving directly on the hardware~\cite{yos:clune}. Producing regular solutions is a key to exploiting regularity in difficult engineering problems~\cite{clune2011performance}.
+
+The simulator likely improved performance because of the number of evaluations it enabled, both in terms of the population (200 vs. 9) and the number of generations (200 vs. 20), leading
+to a total difference of 40000 vs 180 per evolutionary run~\cite{yos:clune}. Another
+potential cause of improved performance is that with a simulator
+HyperNEAT had much less noise in the evaluation process during early
+generations and could thus find coordinated, regular, gaits, which are
+both higher-performing and more likely to transfer. On the physical robot, the noise in the evaluation was
+substantial, preventing effective learning and the discovery of
+effective gaits~\cite{yos:clune}. It could also be the case that 
+the addition in this study of a punishment for high-frequency gaits
+aided performance, especially since in simulation the gaits were
+high-frequency in a few early generations before rapidly settling to a
+frequency range below the penalized threshold~(\figref{servo_plot_111},
+right).
 
@@
 date, most of the gaits in simulation did not transfer well to
-reality. Many gaits that performed very well in simulation performed
+reality. Many gaits that were fast in simulation performed
 poorly on the real robot, largely due to servos that were too weak and
-timed out, or because of frictional differences between simulation and
+shut down, or because of differences between simulation and
 reality. Repeated attempts to minimize these problems were
@@
 the design of the robot: the force required to lift the robot in many
-poses was too much for the servos. Future studies may use a
+poses was too much for the servos. Future studies will use a
 robot that has more mechanical advantage and requires less torque from
@@
 QuadraTot platform.
-
-\figgp{avg_fit_vs_gen}{.48}{avg_freq_vs_gen}{.48}
-{Performance plots. \textbf{Left: }Fitnesses of the champion gaits
-averaged over 20 runs, measured in distance travelled over 12 seconds
-in cm/s. These gaits outperformed gaits evolved in simulation using a
-genetic algorithm~\cite{glette} by 54.3\% in simulation and 5.4\% in
-reality, and they outperformed gaits evolved using HyperNEAT in
-hardware by 49.5\% (\tabref{results}). \edit{Add Kyrre's runs (+ Haocheng?), switch
-from STD to SE if necessary.}\textbf{Right: }Average frequencies of
-the gaits averaged over 20 runs. After implementing a policy which
-punished gaits with high frequencies to reduce servo shutdowns,
-HyperNEAT was able to find high-performing, low frequency
-gaits. \edit{add line at 1.67 Hz}}
-
-\figp{servo_plot_111}{0.95}
-{Servo positions over time (for nine servos) for a gait produced in
-this study. The smoothness, symmetry, and regularities of the gait
-show that by using a simulator, HyperNEAT was able to fully exploit
-the geometry of the problem.}

### 2012-10-29 paper_sean/robot.tex
lots o editing

@@
 % !TEX root =main.tex
-\subsubsection{Robot Hardware.}
+\subsubsection{Robot Hardware:}
 %\figgp{quadratotWhiteBooties}{.45}{quadratot_simulator}{.45}{captioncaption}
 
-We performed the experiments on the QuadraTot quadrupedal robot platform~(\figref{quadratotWhiteBooties}).
-It has 9 degrees-of-freedom (DOF): two joints per leg and one joint that allowed rotation along the robot's midline.
+We performed the experiments on the QuadraTot quadrupedal robot platform~(\figref{quadratotWhiteBooties})~\cite{yos:clune}.
+It has 9 degrees of freedom: two joints per leg and one joint that allows rotation along the robot's midline.
 %we def want a picture of the cool robot.
@@
 Try to get this figure to show up as early as possible in the paper (e.g top of the 2nd page).
-The QuadraTot hardware designs and software for this project are open source\footnote{A parts list, hardware CAD files, software (including the simulator), and gait videos are available at http://creativemachines.cornell.edu/evolved-quadruped-gaits}, and all hardware components are either off-the-shelf or 3D printed parts.
+The QuadraTot hardware designs and the software for this project are open source\footnote{A parts list, hardware CAD files, software (including the simulator), and gait videos are available at http://creativemachines.cornell.edu/evolved-quadruped-gaits}, and all hardware components are either off-the-shelf or 3D printed parts.
 %These traits enable other labs to perform research using the same platform, which eases comparison of different gait-learning algorithms.
-To date there are results on the platform from nine different learning algorithms from three previous publications~\cite{yos:clune,glette,haocheng}.
+To date there are results on the platform for nine different learning algorithms from three previous publications~\cite{yos:clune,glette,haocheng}.
 
-Both in this work and that of Glette et al.
-2012~\cite{glette}, the joints are powered by Robotis Dynamixel servos; five AX-18A servos for the inner joint of each leg and the single midline joint, and four AX-12A servos for the the outer joint of each leg, which requires less power and can thus have less expensive motors.
+The joints are powered by Robotis Dynamixel servos; five AX-18A servos for the inner joint of each leg and the single midline joint, and four AX-12A servos for the the outer joint of each leg, which requires less power and can thus have less expensive motors.
 The servos were sent new positions at 40Hz via the Pydynamixel library.
-Each servo has a built-in safety mechanism that shuts itself off to prevent damage if the servo's current, range, temperature, or torque was too high~\cite{robotis}.
+Each servo has a built-in safety mechanism that shuts itself off to prevent damage if the servo's current, range, temperature, or torque is too high.
 During evolution, this safety mechanism frequently activated, and did so inconsistently, adding significant noise to the evaluation process.
-As pointed out in a previous study~\cite{yos:clune}, gaits generated during evolution are highly variable and generate many shutdowns because they forced the servos to exert too much torque or command positions that are out of range.
-%JMC: Does requesting a value out of range really cause the servo to shut down?
-That seems like something we could easily fix (by never allowing an out-of-range value).
-Now that we do not allow values out of range, we still see lots of shutdowns...making me think that the problem was really torque.
-What do you think?
+As pointed out in a previous study~\cite{yos:clune}, gaits generated during evolution on QuadraTot are highly variable and produce many shutdowns because they force the servos to exert too much torque.
 To prevent collisions between different pieces of the robot's body, we limited the allowable range of movement for the inner, outer, and hip joints to [-85\degree, +60\degree], [-113\degree, +39\degree], and [-28\degree, +28\degree], respectively.
@@
 Yosinski et al.
-2011~\cite{yos:clune} had the small Linux computer that performed all computation on the robot, but we removed it and sent commands to the robot via a cable.
-
-We tracked the robot's position using an infrared LED on the robot and tracked it with a Wiimote.
+2011 had the small Linux computer that performed all computation on the robot, but we removed it and sent commands to the robot via a cable.
+We tracked the robot's position using an infrared LED observed by a Wiimote.

### 2012-10-30 paper_sean/results.tex
before adding the 3 headed figure monster

@@
 Subject to availability, data are reported from the experiments for this paper and three previous studies.
-Reported are the total number of evaluations per run, the average of the fastest gaits produced in each run in simulation, and the fastest gait produced on a physical robot, either the QuadraTot in the Cornell Creative Machines Lab (CCML) or the QuadraTot in the Robotics and Intelligent Systems (ROBIN) lab at the University of Oslo.
+Reported are the total number of evaluations per run, the average of the fastest gaits produced in each run in simulation, and the fastest gait produced on two different copies of the QuadraTot robot (CCML and ROBIN, see text).
 *Instead of using the single fitness value reported in~\cite{glette}, we ran 19 additional runs and used the average fitness of those 20 runs.

### 2012-10-30 paper_sean/conclusion.tex
Rewrote Results with Sean, plus other changes. All remaining needed changes are reflected in edit blocks.

@@
 % !TEX root =main.tex
+
+\edit{Jeff: use me} Overall, these results provide important confirmation that the HyperNEAT generative encoding can evolve state-of-the-art results for challenging engineering problems, such as evolving gaits for real, legged robots.
+The results also reaffirm the benefits of using simulators when solving real-world challenges with evolutionary algorithms.
+
+\edit{Last: check and edit this section.}
+
 HyperNEAT has proven to be an effective encoding in gait evolution with an ability to find gait regularities without human intervention~\cite{clune2009evolving,clune2011performance,yos:clune}.

### 2012-10-30 paper_sean/introduction.tex

@@
 
-Robots that can move around on legs can operate in a much wider range of environments than their wheeled counterparts.
-A downside is that designing gaits for legged robots is a difficult and time-consuming process for human engineers~\cite{strom1999legged,wettergreen1992gait}, and must be repeated every time a new robot is created or the design of an existing robot is changed~\cite{hornby2005autonomous}.
-Because of these difficulties in hand-designing gaits, scientists have investigated automatically producing gaits with machine learning or evolutionary algorithms, an approach that often produces better gaits than those created by human engineers~\cite{valsalam:mii,kohl:stone,hornby2005autonomous,hornby2003generative,yos:clune}.
-It has often been found that both machine-created and human-created gaits perform better if they are \emph{regular}, in that they have coordinated movements, such as left-right symmetry or front-back symmetry~\cite{valsalam:mii,clune2011performance,clune2009evolving,clune2009sensitivity}.
-However, in most previous works, the experimenter has had to manually decide the regularities of the evolved gaits~\cite{valsalam:mii,tellez,beer,raibert}. 
+Legged robots can operate in a much wider range of environments than their wheeled counterparts.
+However, designing gaits for legged robots is a difficult and time-consuming process for human engineers~\cite{strom1999legged,wettergreen1992gait}, and must be repeated every time a robot is created or modified~\cite{hornby2005autonomous}.
+Scientists thus investigate how to automatically produce gaits via machine learning and evolutionary algorithms, which often produces better gaits than those created by human engineers~\cite{valsalam:mii,kohl:stone,hornby2005autonomous,hornby2003generative,yos:clune}.
+While it has been shown that gaits perform better if they are \emph{regular}, in that they have coordinated movements, such as left-right symmetry or front-back symmetry~\cite{valsalam:mii,clune2011performance,clune2009evolving,clune2009sensitivity}, experimenters usually have to manually decide and specify these regularities~\cite{valsalam:mii,tellez,beer,raibert}.
 Such manual intervention is time consuming, requires expert knowledge, and adds constraints that may hurt performance.
 
-Previous work has shown that the Hypercube-based NeuroEvolution of Augmenting Topologies (HyperNEAT) generative encoding~\cite{stanley2009hypercube} can automatically generate a variety of regular gaits and that it outperforms direct encoding controls on this task~\cite{clune2009evolving,clune2011performance}.
+Previous work has shown that the Hypercube-based NeuroEvolution of Augmenting Topologies (HyperNEAT) generative encoding~\cite{stanley2009hypercube} can automatically generate a variety of regular gaits that outperform gaits evolved with direct encodings~\cite{clune2009evolving,clune2011performance}.
 However, that work only verified these claims in simulation. 
-A follow-up paper by Yosinski et al.
-evolved gaits with HyperNEAT directly in hardware on the QuadraTot robot platform and found that HyperNEAT's gaits outperformed manually designed parameterized learning algorithms, but still did not produce impressive, natural gaits~\cite{yos:clune}.
-A further follow-up study built a simulator for the QuadraTot robot platform to test whether the inclusion of a simulator would improve results and found that it did:  when gaits were evolved in this simulator with a simple direct encoding and then transferred to the QuadraTot robot, the resulting gaits were faster than the gaits produced by evolving gaits with HyperNEAT directly on the robot~\cite{glette}.
-The simulator helped because it afforded much larger population sizes and more generations than were possible when evolving directly in hardware.
-Specifically, the simulator allowed 333 times as many evaluations per run as HyperNEAT had in hardware (60000 vs.
+Yosinski et al.
+evolved gaits with HyperNEAT directly in hardware on the QuadraTot robot platform~(\figref{quadratotWhiteBooties}).
+They found that HyperNEAT's gaits outperformed manually designed, parameterized learning algorithms, but still did not produce impressive, natural gaits~\cite{yos:clune}.
+A follow-up study built a simulator for QuadraTot to test whether the inclusion of a simulator would improve results and found that it did:  when gaits were evolved in this simulator with a simple direct encoding and then transferred to the QuadraTot robot, the resulting gaits were faster than the gaits produced by evolving gaits with HyperNEAT directly on the robot~\cite{glette}.
+The simulator helped because it afforded much larger population sizes and more generations than were possible when evolving directly in hardware, resulting in 333 times more evaluations per run (60000 vs.
 180)~\cite{yos:clune,glette}.
@@
 The work with the simulator (Glette et al.
-2012~\cite{glette}) evolved gaits with a simple encoding that was manually constrained to produce specific regularities.
-The success of that work raises the question of whether the performance gains were due to the use of a simulator, or the use of a simple, manually-constrained encoding.
+2012) evolved gaits with a simple encoding that was manually constrained to produce specific regularities.
+The success of that work raises the question of whether the performance gains were due to the added benefit of a simulator or the use of a simple, hand-designed encoding.
 We hypothesized that HyperNEAT, which has been previously shown to automatically discover complex regularities to produce high-performing gaits~\cite{clune2011performance,yos:clune}, would outperform the simpler encoding from Glette et al.
-2012~\cite{glette} if HyperNEAT is combined with a simulator.
+if HyperNEAT is combined with a simulator.
 Here we test that hypothesis by evolving gaits with HyperNEAT in the same simulator from Glette et al.
 and then transferring those gaits to the QuadraTot robot.
-Our experiments confirm the hypothesis: gaits evolved with HyperNEAT and then transferred to reality were the highest performing observed to date on the QuadraTot robot platform.
+Our experiments confirmed the hypothesis: with the simulator, HyperNEAT evolved the highest-performing gaits observed to date for the QuadraTot robot platform.
 

### 2012-10-30 paper_sean/methods.tex

--- 2012-10-16 paper_sean/methods.tex
@@
 
-\subsection{Methods}
+\subsubsection{Evolutionary Process and Parameters:}
 
-We evolved gaits first in simulation and then transferred them to the physical QuadraTot robot, as in Glette et al. 2012~\cite{glette}. 
 Each run had a population size of 200 and lasted 200 generations.
-20 runs were performed, each of which was identical except for the seed provided to the random number generator, which affected stochastic events such as mutation.
-Each gait was evaluated for 12 seconds \edit{JBY: how? As in fig 4 of \cite{yos:clune}, or another way?}, both in simulation and in reality. 
-All HyperNEAT parameters were identical to those in Yosinski et al.
-2011~\cite{yos:clune} except for the frequency of the sine wave input to the ANN, which was lowered from 0.0265Hz to 0.004Hz to decrease the number of servo shutdowns due to high-frequency gaits \edit{JBY: what?
-why?}.
-
-%\begin{equation}
-%sin(2\pi t/250)\pi
-%\end{equation}
-
-%from
-%
-%\begin{equation}
-%sin(t/6)\pi
-%\end{equation}
-%
-
- %Sine frequency of 0.004 was used because we experimentally found that HyperNEAT was able to produce natural and high-performing gaits with low frequencies with this frequency. 
-We also punished high-frequency gaits during evolving in simulation, which was effective at reducing the number of servo shutdowns observed when evolved gaits were transferred to the robot. 
-If the average frequency of each leg \edit{JBY: any leg or all legs?} for the gait was higher than a threshold frequency (1.67 Hz), the fitness was punished by exponentially reducing the distance travelled by the robot:
-
-\begin{equation}
-\mathrm{distance} = \mathrm{distance} \cdot e^{(\mathrm{freq}_{\mathrm{thr}}-\mathrm{freq})} \\
-\end{equation}
-
-\noindent where \emph{$\mathrm{distance}$} is the Euclidean distance the center of mass of the robot travelled during the gait duration.
+We performed twenty HyperNEAT runs that differed only in the seed provided to the random number generator, which affected stochastic events such as mutation.
+To make a statistical comparison to the encoding from Glette et al., we conducted 19 runs using the original code to supplement the one run performed for that paper.
+Each gait was evaluated for fourteen seconds in reality, with interpolation from and to a stationary position in the first and last two seconds, respectively, as done in Yosinski et al.
+The result was effectively 12 seconds of full-speed
+motion. In simulation gaits were evaluated for 12 seconds. 
+All HyperNEAT parameters were identical to those in Yosinski et al. except for the frequency of the sine wave input to the ANN, which was lowered from 4.2Hz to 0.64Hz to reduce servo shutdowns. 
+To further reduce servo shutdowns, we punished high-frequency gaits during evolution.
+We calculated a gait frequency as the average number of servo direction changes per leg per second.
+If this frequency was higher than an experimentally-selected threshold of 1.67 Hz, the measure of distance traveled by the center of mass during the gait was reduced exponentially by multiplying it by a discount factor, which was $e^{(1.67-\mathrm{freq})}$.
 Following Clune et al.
-2009~\cite{clune2009evolving}, the fitness equation was
-
-\begin{equation}
-\mathrm{fitness} = 2^{\mathrm{distance}^{2}}
-\end{equation}
-
-After evolving the gaits in simulation, the champion gait of the last generation of each of the 20 runs was transferred onto the real robot and evaluated.
-The robot's position was measured at the beginning and the end of each run, starting from the stationary pose and ending with the stationary pose. 
-If a gait was obviously affected by servo shutdowns, the gait was tried again after letting the servos cool down.
+2009~\cite{clune2009evolving}, fitness equation was $2^{\mathrm{distance}^{2}}$.
+After evolving the gaits in simulation, the champion gait of the last generation of each of the 20 runs was transferred onto the real robot and the distance traveled was measured.

### 2012-10-30 paper_sean/simulator.tex

--- 2012-10-12 paper_sean/simulator.tex
@@
 % !TEX root =main.tex
-The simulator was the same as Glette et al.
-2012 \cite{glette}.
-C++ code for the the Nvidia PhysX physics engine approximately describes the QuadraTot, including the mass and size of the robot components and its degrees of freedom~(\figref{quadratotWhiteBooties}).
-In the simulator, each individual joint range was limit as described above, but Smart Cropping was not included because it hindered performance by limiting the types of gaits evolution could explore in early generations.
+
+\subsubsection{Simulator:}
+Gaits evolved in the simulator from Glette et al., which represents QuadraTot in the Nvidia PhysX physics engine\footnotemark[\value{footnote}], including the mass and size of the QuadraTot components and its degrees of freedom~(\figref{quadratotWhiteBooties}, right).
+In the simulator, each individual joint range was limit as described above, but Smart Cropping was not included because we found that it hindered performance by limiting the types of gaits evolution could explore in early generations.
 

### 2012-10-30 paper_sean/conclusion.tex
I was almost to 10 pages, then sean interrupted me

--- 2012-09-28 paper_sean/conclusion.tex
@@
 % !TEX root =main.tex
-HyperNEAT has proven to be an effective encoding in gait evolution with an ability to find gait regularities without human intervention~\cite{clune2009evolving,clune2011performance,yos:clune}.
-Glette et al.
-2012~\cite{glette} showed the benefits of a physical simulator by evolving, with a simple genetic algorithm, a gait that outperformed those HyperNEAT evolved in hardware.
-In this study, we took the two -- HyperNEAT and a physical simulator -- and showed that HyperNEAT's performance significantly increases with longer evolution periods.
-The gaits produced in this study outperformed the previous best gait on the QuadraTot by 54.3\% in simulation and by 5.4\% in real life trials.
-This study also showed that HyperNEAT will exploit the characteristics of the problem, such as frequency and regularity, in order to maximize its fitness.
-%In simulation, HyperNEAT gaits outperformed Genetic Algorithm gaits~\cite{glette} by 54.3\% and by 5.4\% in the real life trials.
 
-%Without frequency limiters, it would make the legs move really fast in order to increase fitness. 
-%When that was no longer allowed, it pushed the servo positions to their boundaries by increasing the stride lengths as wide as possible. 
-However, we faced a problem in the transfer from simulation to reality.
-%because of the physical limitations of the robot, these exploitations did not transfer well to reality.
-%could not be transferred well to reality.
-One of the reasons was that the robot was greatly underpowered and could not perform the transferred gaits.
-The other reasons was the inevitable faults with the simulator (as it cannot model the robot or the real world perfectly).
-Because of these factors, many gaits performed much worse than it did in simulation and had difficulty in achieving consistent results.
-
-Next steps for this research heads towards improving the transfer process of the gaits from simulation to reality by reducing the \emph{reality gap}~\cite{jakobi}.
-Many different approaches to reduce the reality gap has been proposed~\cite{koos2010crossing,bongard,zagal}.
-One of the easiest and most promising of these methods is the Transferability Approach~\cite{koos2011transferability}, which has already proven to be successful in quadrapedal gait evolution.
-Another method that we could take is to evolve the simulator itself to adapt itself to the real world.
-HyperNEAT, used with a simulator and a transfer approach, shows much hope in producing yet a better, more coordinated, and higher performing gait for many robotic platforms.
+With a simulator, HyperNEAT evolved the fasted gait yet recorded for the QuadraTot robot, outperforming nine machine learning algorithms from three previous publications~\cite{yos:clune,glette,haocheng}, including an improvement of 54.3\% in simulation and 5.4\% in reality over the previous best QuadraTot gait by Glette et al.
+These results provide an important demonstration that the HyperNEAT generative encoding can evolve state-of-the-art results for challenging engineering problems, in this case evolving gaits for legged robots.
+The results further reaffirm the benefits of simulators when solving real-world challenges with evolutionary algorithms.
+Our results additionally confirm--with a different robot and simulator--previous work that has shown HyperNEAT to be an effective encoding for automatically evolving coordinated, regular gaits in simulation~\cite{clune2009evolving,clune2011performance,yos:clune}.
+That HyperNEAT outperformed the encoding hand-designed by Glette et al.
+shows that HyperNEAT can even outperform evolutionary algorithms manually designed to incorporate human domain knowledge regarding which regularities are thought to be beneficial, demonstrating that automatically discovering regularities to specifying them even for problems that are relatively well understood.

### 2012-10-30 paper_sean/hyperneat.tex

@@
 It has been repeatedly described in detail~\cite{stanley2009hypercube,gauci2007generating,clune2011performance},  so here we provide only a summary.
-Instead of directly encoding the weights of the ANN individually on the genome, in HyperNEAT the genome is a compositional pattern producing network (CPPN)~\cite{stanley2007CPPN}.
-The CPPN genome specifies the weights in a similar way to how natural organisms develop.
+Instead of directly encoding each ANN weight individually on the genome, in HyperNEAT the genome is a compositional pattern producing network (CPPN)~\cite{stanley2007CPPN}.
+The CPPN specifies the weights in a similar way to how natural organisms develop.
 In nature, phenotypic attributes are specified as a function of their geometric location, and such positional information is conveyed through chemical morphogen gradients~\cite{carroll2005endless}.
-For example, the concentration of one chemical could indicate the position along the anterior-posterior (head-to-tail) axis, and another chemical in bands could indicate if a cell is in an odd- or even-numbered segment.
+For example, the concentration of one chemical could indicate the position along the head-to-tail axis and another chemical in bands could indicate if a cell is in an odd- or even-numbered segment.
 Based on the relative concentrations of these chemicals, a cell can know where it is geometrically and, thus, what type of cell to become~\cite{carroll2005endless}.
@@
 With CPPNs this process is abstracted as a network of math functions that operate in a Cartesian geometric space.
-The coordinates of phenotypic elements are provided as inputs to the CPPN and the output specifies a phenotypic trait.
+The coordinates of phenotypic elements are provided as inputs to the CPPN and the outputs specify phenotypic traits.
 For example, when CPPNs encode 2D pictures, the coordinates of each pixel are iteratively input into the genome and the output is the grayscale value at that coordinate~\cite{secretan2011picbreeder}.
 Because a CPPN network is composed of math functions, those functions can create geometric regularities in the phenotype.
-For example, a Gaussian function of the x-axis can provide left-right symmetry, and a repeating function (e.g.
-sine) of the y-axis could provide a related motif such as segmentation~(\figref{hyperneat_bug_example}).
+For example, a Gaussian function an axis can provide symmetry (e.g.
+left-right), and a repeating function (e.g.
+sine) of an axis could provide repetition (e.g segmentation).
 Both 2D pictures and 3D objects evolved with CPPNs look like natural and engineered objects, and contain complex, regularities such as symmetries and repeated motifs, with and without variation~\cite{secretan2011picbreeder,clune2011objects}.
 
-\figp{hyperneat_bug_example}{.8}{Compositional Pattern Producing Networks (CPPNs).
-CPPN  genomes use mathematical functions to generate regularities such as symmetry and repetitions.
-CPPN genomes allow functions to be made of other functions, allowing multiple regularity motifs to be present.
-Figure taken from Stanley 2007~\cite{stanley2007CPPN}.}
+%\figp{hyperneat_bug_example}{.8}{Compositional Pattern Producing Networks (CPPNs).
+CPPN  genomes use mathematical functions to generate phenotypic regularities, such as symmetry and repetitions.
+Figure adapted from Stanley 2007~\cite{stanley2007CPPN}.}
 

### 2012-10-30 paper_sean/results.tex

@@
 In simulation, HyperNEAT gaits were significantly faster than those from Glette et al.
-(\figref{FINAL_kyrre_sean_fitness}, $\emph{p} < 6.8\times10^{-8}$, comparing the best gaits in the final generation of each run via MATLAB's Wilcoxon rank sum test).
+(\figref{threeplots}-Top, $\emph{p} < 6.8\times10^{-8}$, comparing the best gaits in the final generation of each run via MATLAB's Wilcoxon rank sum test).
 Specifically, HyperNEAT gaits were 55\% faster on average in simulation (25.4
 cm/s $\pm$ 3.4 SD versus 16.7 cm/s $\pm$ 1.9 SD, \tabref{results}).\edit{Table 1, not 3} 
-Plots of servo positions over time reveal that the evolved HyperNEAT gaits are regular, smooth and coordinated~(\figref{servo_plot_111}), confirming previous results with HyperNEAT in simulation~\cite{clune2009evolving,clune2011performance}.
+Plots of servo positions over time reveal that the evolved HyperNEAT gaits are regular, smooth and coordinated~(\figref{threeplots}-Left), confirming previous results with HyperNEAT in simulation~\cite{clune2009evolving,clune2011performance}.
 
-\figp{FINAL_kyrre_sean_fitness}{0.62}
+\figOneOnTwo[h!]{threeplots}{FINAL_kyrre_sean_fitness}{0.62}{servo_plot_111}{.48}{frequencyPlotThreshFinal}{.48}
 {HyperNEAT outperforms a genetic algorithm with a simpler encoding, when both algorithms are combined with a simulator.
@@
 HyperNEAT gaits are 54.3\% faster in
-simulation and 5.4\% faster in reality~(\tabref{results}).}
-
-\figgp{servo_plot_111}{.48}{frequencyPlotThreshFinal}{.48}
-{\textbf{Left: }
+simulation and 5.4\% faster in reality~(\tabref{results}).
+\textbf{Left: }
 Servo positions over time (for nine servos) for a simulated HyperNEAT gait. HyperNEAT produces gaits that are smooth, symmetrical, and contain other regularities. 
 \textbf{Right: }Mean gait frequency averaged over 20 runs. 
-Gaits with frequencies above a threshold (horizontal line) receive a fitness penalty. HyperNEAT quickly learns to produce gaits with frequencies low enough to avoid this penalty. }
+Gaits with frequencies above a threshold (horizontal line) receive a fitness penalty. HyperNEAT quickly learns to produce gaits with frequencies low enough to avoid this penalty.
+}
 
@@
 only 12.95 cm/s (vs. 17.8 cm/s measured on ROBIN) and a maximum of
-13.76 cm/s (\tabref{results}).
+13.8 cm/s (\tabref{results}).
 It thus appears that the CCML version of the robot is slower.
@@
 \hline
-GA + simulator \cite{glette}             & 60000       & *16.7       &   13.0   & \textbf{17.8}  \\
+GA + simulator \cite{glette}             & 60000       & *16.7       &   13.8   & \textbf{17.8}  \\
 \hline
@@
 On the physical robot, the gaits produced by HyperNEAT with a simulator were faster, more natural, and more repeatable than those evolved directly on the Quadratot robot~\cite{yos:clune}.
-The gaits were also more regular, as they were in simulation~(\figref{servo_plot_111}).
+The gaits were also more regular, as they were in simulation~(\figref{threeplots}-Left).
 This result is important because it confirms that HyperNEAT can produce the important property of
@@
 high-frequency in a few early generations before rapidly settling to a
-frequency range below the penalized threshold~(\figref{servo_plot_111},
-right).
+frequency range below the penalized threshold~(\figref{threeplots}-Right).
 
@@
 which was designed in response to the challenges experienced with the
-QuadraTot platform.
+QuadraTot platform. Future studies will also incorporate techniques to minimize the gap between the simulator and reality~\cite{koos2010crossing,bongard,zagal}.

### 2012-10-30 paper_sean/conclusion.tex
fixed the table numbering, and got 10 pages

@@
 With a simulator, HyperNEAT evolved the fasted gait yet recorded for the QuadraTot robot, outperforming nine machine learning algorithms from three previous publications~\cite{yos:clune,glette,haocheng}, including an improvement of 54.3\% in simulation and 5.4\% in reality over the previous best QuadraTot gait by Glette et al.
-These results provide an important demonstration that the HyperNEAT generative encoding can evolve state-of-the-art results for challenging engineering problems, in this case evolving gaits for legged robots.
+These results provide an important demonstration that the HyperNEAT generative encoding can evolve state-of-the-art results for challenging engineering problems, in this case evolving gaits for a legged robot.
 The results further reaffirm the benefits of simulators when solving real-world challenges with evolutionary algorithms.
@@
 That HyperNEAT outperformed the encoding hand-designed by Glette et al.
-shows that HyperNEAT can even outperform evolutionary algorithms manually designed to incorporate human domain knowledge regarding which regularities are thought to be beneficial, demonstrating that automatically discovering regularities to specifying them even for problems that are relatively well understood.
+shows that HyperNEAT can even outperform evolutionary algorithms manually designed to incorporate human domain knowledge regarding which regularities are thought to be beneficial, demonstrating that automatically discovering regularities can be a superior approach to specifying them even for problems that are relatively well understood.

### 2012-10-30 paper_sean/results.tex
fixed the table numbering, and got 10 pages

@@
 Specifically, HyperNEAT gaits were 55\% faster on average in simulation (25.4
-cm/s $\pm$ 3.4 SD versus 16.7 cm/s $\pm$ 1.9 SD, \tabref{results}).\edit{Table 1, not 3} 
+cm/s $\pm$ 3.4 SD versus 16.7 cm/s $\pm$ 1.9 SD, \tabref{results}).
 Plots of servo positions over time reveal that the evolved HyperNEAT gaits are regular, smooth and coordinated~(\figref{threeplots}-Left), confirming previous results with HyperNEAT in simulation~\cite{clune2009evolving,clune2011performance}.
@@
 \end{tabular}
-\end{center}
-\tablabel{results}
 \caption{Velocities of learned gaits in simulation and on two different copies of the QuadraTot robot.
@@
 Velocities are in cm/s, and bold indicates the best performance.}
+\tablabel{results}
+\end{center}
 \end{table}

### 2012-10-30 paper_sean/conclusion.tex
jeff done editing...pending feedback

@@
 These results provide an important demonstration that the HyperNEAT generative encoding can evolve state-of-the-art results for challenging engineering problems, in this case evolving gaits for a legged robot.
-The results further reaffirm the benefits of simulators when solving real-world challenges with evolutionary algorithms.
-Our results additionally confirm--with a different robot and simulator--previous work that has shown HyperNEAT to be an effective encoding for automatically evolving coordinated, regular gaits in simulation~\cite{clune2009evolving,clune2011performance,yos:clune}.
+The results further reaffirm the benefits of using simulators when solving real-world challenges with evolutionary algorithms.
+Our results additionally confirm--with a different robot and simulator--previous work that has shown that HyperNEAT is an effective encoding for automatically evolving coordinated, regular gaits in simulation~\cite{clune2009evolving,clune2011performance,yos:clune}.
 That HyperNEAT outperformed the encoding hand-designed by Glette et al.
-shows that HyperNEAT can even outperform evolutionary algorithms manually designed to incorporate human domain knowledge regarding which regularities are thought to be beneficial, demonstrating that automatically discovering regularities can be a superior approach to specifying them even for problems that are relatively well understood.
+shows that HyperNEAT can even outperform evolutionary algorithms manually designed to incorporate human domain knowledge regarding which regularities are thought to be beneficial for a problem, demonstrating that automatically discovering regularities can be a superior approach to specifying them even for problems that are relatively well understood.

### 2012-10-30 paper_sean/results.tex

@@
 
-\figOneOnTwo[h!]{threeplots}{FINAL_kyrre_sean_fitness}{0.62}{servo_plot_111}{.48}{frequencyPlotThreshFinal}{.48}
+\figOneOnTwo[h!]{threeplots}{FINAL_sean_kyrre_fitness}{0.62}{servo_plot_111}{.48}{frequencyPlotThreshFinal}{.48}
 {HyperNEAT outperforms a genetic algorithm with a simpler encoding, when both algorithms are combined with a simulator.
@@
 \end{tabular}
+\vspace{.35cm}
 \caption{Velocities of learned gaits in simulation and on two different copies of the QuadraTot robot.

### 2012-10-30 paper_sean/conclusion.tex
minor

--- 2012-10-30 paper_sean/conclusion.tex
@@
 The results further reaffirm the benefits of simulators when solving real-world challenges with evolutionary algorithms.
-Our results additionally confirm--with a different robot and simulator--previous work that has shown HyperNEAT to be an effective encoding for automatically evolving coordinated, regular gaits in simulation~\cite{clune2009evolving,clune2011performance,yos:clune}.
+Our results additionally confirm---with a different robot and simulator---previous work that has shown HyperNEAT to be an effective encoding for automatically evolving coordinated, regular gaits in simulation~\cite{clune2009evolving,clune2011performance,yos:clune}.
 That HyperNEAT outperformed the encoding hand-designed by Glette et al.
-shows that HyperNEAT can even outperform evolutionary algorithms manually designed to incorporate human domain knowledge regarding which regularities are thought to be beneficial, demonstrating that automatically discovering regularities can be a superior approach to specifying them even for problems that are relatively well understood.
+shows that HyperNEAT can outperform even evolutionary algorithms manually designed to incorporate human domain knowledge regarding which regularities are thought to be beneficial, demonstrating that automatically discovering regularities can be a superior approach to specifying them, even on problems that are relatively well understood.

### 2012-10-30 paper_sean/conclusion.tex

--- 2012-10-30 paper_sean/conclusion.tex
@@
 The results further reaffirm the benefits of using simulators when solving real-world challenges with evolutionary algorithms.
-Our results additionally confirm--with a different robot and simulator--previous work that has shown that HyperNEAT is an effective encoding for automatically evolving coordinated, regular gaits in simulation~\cite{clune2009evolving,clune2011performance,yos:clune}.
+Our results additionally confirm---with a different robot and simulator---previous work that has shown that HyperNEAT is an effective encoding for automatically evolving coordinated, regular gaits in simulation~\cite{clune2009evolving,clune2011performance,yos:clune}.
 That HyperNEAT outperformed the encoding hand-designed by Glette et al.
-shows that HyperNEAT can even outperform evolutionary algorithms manually designed to incorporate human domain knowledge regarding which regularities are thought to be beneficial for a problem, demonstrating that automatically discovering regularities can be a superior approach to specifying them even for problems that are relatively well understood.
+shows that HyperNEAT can outperform even evolutionary algorithms manually designed to incorporate human domain knowledge regarding which regularities are thought to be beneficial for a problem, demonstrating that automatically discovering regularities can be a superior approach to specifying them, even on problems that are relatively well understood.

### 2012-10-31 paper_sean/conclusion.tex
some minor changes

@@
 
-With a simulator, HyperNEAT evolved the fasted gait yet recorded for the QuadraTot robot, outperforming nine machine learning algorithms from three previous publications~\cite{yos:clune,glette,haocheng}, including an improvement of 54.3\% in simulation and 5.4\% in reality over the previous best QuadraTot gait by Glette et al.
+With a simulator, HyperNEAT evolved the fasted gait yet recorded for the QuadraTot robot, outperforming nine machine learning algorithms from three previous publications~\cite{yos:clune,glette,haocheng}, including an improvement of 54.3\% in simulation and 5.1\% in reality over the previous best QuadraTot gait by Glette et al.
 These results provide an important demonstration that the HyperNEAT generative encoding can evolve state-of-the-art results for challenging engineering problems, in this case evolving gaits for a legged robot.

### 2012-10-31 paper_sean/results.tex
some minor changes

@@
 HyperNEAT gaits are 54.3\% faster in
-simulation and 5.4\% faster in reality~(\tabref{results}).
+simulation and 5.1\% faster in reality~(\tabref{results}).
 \textbf{Left: }
@@
 
-On the physical robot, HyperNEAT gaits from this study outperformed gaits from all previous Quadratot studies~(\tabref{results})~\cite{yos:clune,haocheng,glette}, including those of Glette et al. 
+On the physical robot, HyperNEAT gaits from this study outperformed gaits from all previous QuadraTot studies~(\tabref{results})~\cite{yos:clune,haocheng,glette}, including those of Glette et al. 
  However, comparing performance in hardware between studies performed in
@@
 We now discuss the differences between the gaits evolved by HyperNEAT directly on the physical robot, and those first evolved in a simulator and then transferred to the robot. 
-On the physical robot, the gaits produced by HyperNEAT with a simulator were faster, more natural, and more repeatable than those evolved directly on the Quadratot robot~\cite{yos:clune}.
+On the physical robot, the gaits produced by HyperNEAT with a simulator were faster, more natural, and more repeatable than those evolved directly on the QuadraTot robot~\cite{yos:clune}.
 The gaits were also more regular, as they were in simulation~(\figref{threeplots}-Left).

### 2012-11-01 paper_sean/hyperneat.tex
removed the [year] from *et al. [year]

@@
 
-In this study, the ANN inputs, outputs, and the size of the hidden layer are the same as in Yosinski et al. 2011~\cite{yos:clune}. 
+In this study, the ANN inputs, outputs, and the size of the hidden layer are the same as in Yosinski et al.~\cite{yos:clune}. 
 The ANN has a fixed topology of three \begin{math}3\times4\end{math} Cartesian grids of nodes for the input, hidden, and output layers. 
@@
 The outputs of the substrate at each time step were nine numbers (for each joint) in the range [-1, 1] which were scaled to the range [0, 1023], the allowable ranges for the servos. 
-As in Yosinski et al.
-2011~\cite{yos:clune}, we generated pseudo-positions at 160Hz and then downsampled over consecutive blocks of 4 to obtain the actual commanded positions at 40Hz; this reduced the number of gaits which commanded switches from extreme negative to extreme positive numbers at 40Hz, an undesirable artifact which only served to tax the servos.
+As in Yosinski et al.~\cite{yos:clune}, we generated pseudo-positions at 160Hz and then downsampled over consecutive blocks of 4 to obtain the actual commanded positions at 40Hz; this reduced the number of gaits which commanded switches from extreme negative to extreme positive numbers at 40Hz, an undesirable artifact which only served to tax the servos.
 
@@
 All pairwise combinations of source and target node coordinates are iteratively passed into a CPPN to determine the weight of each ANN link.
-Figure from Clune et al.
-2011~\cite{clune2011performance}.}
+Figure from Clune et al.~\cite{clune2011performance}.}

### 2012-11-01 paper_sean/introduction.tex
jeff done editing...pending feedback
removed the [year] from *et al. [year]

@@
 \textbf{Left: } The physical robot, which is composed of 3-D printable and off-the-shelf components.
-\textbf{Right: } The robot in the simulator. Figure adapted from from Glette et al. 2012~\cite{glette}.}
+\textbf{Right: } The robot in the simulator.}
 
@@
 
-The work with the simulator (Glette et al.
-2012) evolved gaits with a simple encoding that was manually constrained to produce specific regularities.
+The work with the simulator (Glette et al.) evolved gaits with a simple encoding that was manually constrained to produce specific regularities.
 The success of that work raises the question of whether the performance gains were due to the added benefit of a simulator or the use of a simple, hand-designed encoding.

### 2012-11-01 paper_sean/robot.tex
removed the [year] from *et al. [year]

@@
 To prevent collisions between different pieces of the robot's body, we limited the allowable range of movement for the inner, outer, and hip joints to [-85\degree, +60\degree], [-113\degree, +39\degree], and [-28\degree, +28\degree], respectively.
-We also implemented the Smart Cropping System from Shen et al.
-2012~\cite{haocheng}, which prevents combinations of joint positions for the inner and outer joint of each leg that generate extreme amounts of torque.
+We also implemented the Smart Cropping System from Shen et al.~\cite{haocheng}, which prevents combinations of joint positions for the inner and outer joint of each leg that generate extreme amounts of torque.
 A final method of reducing torque was to reduce the weight of the robot.
-Yosinski et al.
-2011 had the small Linux computer that performed all computation on the robot, but we removed it and sent commands to the robot via a cable.
+Yosinski et al.~\cite{yosinski} had the small Linux computer that performed all computation on the robot, but we removed it and sent commands to the robot via a cable.
 We tracked the robot's position using an infrared LED observed by a Wiimote.

### 2012-11-03 paper_sean/hyperneat.tex
changes in response to kyrres second round of comments

@@
 Because a CPPN network is composed of math functions, those functions can create geometric regularities in the phenotype.
-For example, a Gaussian function an axis can provide symmetry (e.g.
+For example, a Gaussian function of an axis can provide symmetry (e.g.
 left-right), and a repeating function (e.g.
@@
 
-In HyperNEAT, the CPPN genomes evolve via the Neuroevolution of Augmenting Topologies (NEAT) algorithm~\cite{stanley2006exploiting}. 
-The NEAT algorithm has three major components~\cite{stanley2006exploiting}. 
+In HyperNEAT, the CPPN genomes evolve via the Neuroevolution of Augmenting Topologies (NEAT) algorithm~\cite{stanley2006exploiting}, which has three major components \cite{stanley2006exploiting}. 
 First, it starts with small genomes that encode simple networks and complexifies them by mutations that add nodes and links to the network.
@@
 
-In this study, the ANN inputs, outputs, and the size of the hidden layer are the same as in Yosinski et al.~\cite{yos:clune}. 
+In this study, the ANN inputs, outputs, activation functions, and the size of the hidden layer are the same as in Yosinski et al.~\cite{yos:clune}. 
 The ANN has a fixed topology of three \begin{math}3\times4\end{math} Cartesian grids of nodes for the input, hidden, and output layers. 

### 2012-11-03 paper_sean/results.tex
changes in response to kyrres second round of comments

@@
 \figOneOnTwo[h!]{threeplots}{FINAL_sean_kyrre_fitness}{0.62}{servo_plot_111}{.48}{frequencyPlotThreshFinal}{.48}
-{HyperNEAT outperforms a genetic algorithm with a simpler encoding, when both algorithms are combined with a simulator.
+{
+\textbf{Top: }
+HyperNEAT outperforms a genetic algorithm with a simpler encoding, when both algorithms are combined with a simulator.
 Plotted are means over 20 runs in simulation (solid lines) $\pm$ STD (dashed lines).

### 2012-11-03 paper_sean/robot.tex
changes in response to kyrres second round of comments

@@
 A final method of reducing torque was to reduce the weight of the robot.
-Yosinski et al.~\cite{yosinski} had the small Linux computer that performed all computation on the robot, but we removed it and sent commands to the robot via a cable.
+Yosinski et al.
+had the small Linux computer that performed all computation on the robot, but we removed it and sent commands to the robot via a cable.
 We tracked the robot's position using an infrared LED observed by a Wiimote.

### 2012-11-05 paper_sean/abstract.tex
lots o editing
final JMC pass. lots of small, but important, changes

--- 2012-09-21 paper_sean/abstract.tex
@@
 % !TEX root =main.tex
-Creating gaits for physical robots is a longstanding and still open challenge. 
+Creating gaits for physical robots is a longstanding and open challenge. 
 %Evolutionary algorithms can outperform human engineers in generating such gaits, but this approach historically required engineers to manually decide which types of gaits would evolve.
@@
 That hypothesis was confirmed, resulting in the fastest gaits yet observed for this robot, including those produced by nine different algorithms from three previous papers describing techniques for generating gaits for this robot.
-This result is important because it confirms that the early promise shown by generative encodings, specifically HyperNEAT, were not limited to evolving gaits in simulation, but indeed work on real robots.
+This result is important because it confirms that the early promise shown by generative encodings, specifically HyperNEAT, are not limited to simulation, but indeed work on challenging real-world engineering challenges such as evolving gaits for real robots.
 

### 2012-11-05 paper_sean/conclusion.tex
final JMC pass. lots of small, but important, changes

@@
 That HyperNEAT outperformed the encoding hand-designed by Glette et al.
-shows that HyperNEAT can outperform even evolutionary algorithms manually designed to incorporate human domain knowledge regarding which regularities are thought to be beneficial for a problem, demonstrating that automatically discovering regularities can be a superior approach to specifying them, even on problems that are relatively well understood.
+shows that HyperNEAT can outperform even evolutionary algorithms manually designed to incorporate human domain knowledge regarding which regularities are thought to be beneficial for a problem.
+The results thus demonstrate that automatically discovering regularities can be a superior approach to specifying them, even on problems that are relatively well-understood.

### 2012-11-05 paper_sean/hyperneat.tex
final JMC pass. lots of small, but important, changes

@@
 sine) of an axis could provide repetition (e.g segmentation).
-Both 2D pictures and 3D objects evolved with CPPNs look like natural and engineered objects, and contain complex, regularities such as symmetries and repeated motifs, with and without variation~\cite{secretan2011picbreeder,clune2011objects}.
+Both 2D pictures and 3D objects evolved with CPPNs look like natural and engineered objects, and contain complex regularities, such as symmetries and repeated motifs, with and without variation~\cite{secretan2011picbreeder,clune2011objects}.
 
@@
 left-right symmetry or repeated modules), but in this case the patterns are in the weights of the neural network~\cite{clune2011performance}.
-The neural regularities produced by HyperNEAT enable significantly improved performance on problems that are regular~\cite{clune2011performance,stanley2009hypercube}, including evolving quadruped gaits~\cite{clune2011performance,clune2009evolving,clune2009sensitivity}.
+The neural regularities produced by HyperNEAT enable significantly improved performance on problems that are regular~\cite{clune2011performance,stanley2009hypercube}, including evolving quadruped gaits in simulation~\cite{clune2011performance,clune2009evolving,clune2009sensitivity}.
 
-In HyperNEAT, the CPPN genomes evolve via the Neuroevolution of Augmenting Topologies (NEAT) algorithm~\cite{stanley2006exploiting}, which has three major components \cite{stanley2006exploiting}. 
-First, it starts with small genomes that encode simple networks and complexifies them by mutations that add nodes and links to the network.
+In HyperNEAT, the CPPN genomes evolve via the Neuroevolution of Augmenting Topologies (NEAT) algorithm~\cite{stanley2002evolving}, which has three major components. 
+First, NEAT starts with small genomes that encode simple networks and complexifies them via mutations that add nodes and links to the networks.
 Second, NEAT has a fitness-sharing system that preserves diversity and allows for new innovations to be tuned by evolution before competing them against more adapted rivals.
 Third, historical information is recorded that facilitates crossover in a way that is effective, yet avoids the need for expensive topological analysis.
-A full explanation of NEAT can be found in Stanley and Miikkulainen 2006~\cite{stanley2006exploiting}.
+A full explanation of NEAT can be found in Stanley and Miikkulainen 2002~\cite{stanley2002evolving}.
 
 In this study, the ANN inputs, outputs, activation functions, and the size of the hidden layer are the same as in Yosinski et al.~\cite{yos:clune}. 
-The ANN has a fixed topology of three \begin{math}3\times4\end{math} Cartesian grids of nodes for the input, hidden, and output layers. 
+The ANN had a fixed topology of three \begin{math}3\times4\end{math} Cartesian grids of nodes for the input, hidden, and output layers. 
 The inputs to the substrate were the angles requested in the prevous time step for each of the 9 joints of the robot and a sine and cosine wave to facilitate periodic motion. 
 The outputs of the substrate at each time step were nine numbers (for each joint) in the range [-1, 1] which were scaled to the range [0, 1023], the allowable ranges for the servos. 
-As in Yosinski et al.~\cite{yos:clune}, we generated pseudo-positions at 160Hz and then downsampled over consecutive blocks of 4 to obtain the actual commanded positions at 40Hz; this reduced the number of gaits which commanded switches from extreme negative to extreme positive numbers at 40Hz, an undesirable artifact which only served to tax the servos.
+As in Yosinski et al.~\cite{yos:clune}, we generated pseudo-positions at 160Hz and then downsampled over consecutive blocks of four time steps to obtain the actual commanded positions at 40Hz; this reduced the number of gaits which commanded switches from extreme negative to extreme positive numbers at 40Hz, which overly taxed the servos.
 

### 2012-11-05 paper_sean/introduction.tex
final JMC pass. lots of small, but important, changes

@@
 However, designing gaits for legged robots is a difficult and time-consuming process for human engineers~\cite{strom1999legged,wettergreen1992gait}, and must be repeated every time a robot is created or modified~\cite{hornby2005autonomous}.
-Scientists thus investigate how to automatically produce gaits via machine learning and evolutionary algorithms, which often produces better gaits than those created by human engineers~\cite{valsalam:mii,kohl:stone,hornby2005autonomous,hornby2003generative,yos:clune}.
-While it has been shown that gaits perform better if they are \emph{regular}, in that they have coordinated movements, such as left-right symmetry or front-back symmetry~\cite{valsalam:mii,clune2011performance,clune2009evolving,clune2009sensitivity}, experimenters usually have to manually decide and specify these regularities~\cite{valsalam:mii,tellez,beer,raibert}.
+Scientists thus investigate how to automatically produce gaits via machine learning and evolutionary algorithms, and the result is often a better gait than those created by human engineers~\cite{valsalam:mii,kohl:stone,hornby2005autonomous,hornby2003generative,yos:clune}.
+While it has been shown that gaits perform better if they are \emph{regular}---i.e., that they have coordinated movements, such as left-right symmetry or front-back symmetry~\cite{valsalam:mii,clune2011performance,clune2009evolving,clune2009sensitivity}---experimenters usually have to explicitly decide and specify these regularities~\cite{valsalam:mii,tellez,beer,raibert}.
 Such manual intervention is time consuming, requires expert knowledge, and adds constraints that may hurt performance.
@@
 evolved gaits with HyperNEAT directly in hardware on the QuadraTot robot platform~(\figref{quadratotWhiteBooties}).
-They found that HyperNEAT's gaits outperformed manually designed, parameterized learning algorithms, but still did not produce impressive, natural gaits~\cite{yos:clune}.
+They found that HyperNEAT's gaits outperformed manually designed, parameterized learning algorithms, but still did not produce impressive, natural gaits\footnote{Videos available at http://creativemachines.cornell.edu/evolved-quadruped-gaits}~\cite{yos:clune}.
 A follow-up study built a simulator for QuadraTot to test whether the inclusion of a simulator would improve results and found that it did:  when gaits were evolved in this simulator with a simple direct encoding and then transferred to the QuadraTot robot, the resulting gaits were faster than the gaits produced by evolving gaits with HyperNEAT directly on the robot~\cite{glette}.

### 2012-11-05 paper_sean/methods.tex
removed the [year] from *et al. [year]
final JMC pass. lots of small, but important, changes

@@
 To further reduce servo shutdowns, we punished high-frequency gaits during evolution.
-We calculated a gait frequency as the average number of servo direction changes per leg per second.
-If this frequency was higher than an experimentally-selected threshold of 1.67 Hz, the measure of distance traveled by the center of mass during the gait was reduced exponentially by multiplying it by a discount factor, which was $e^{(1.67-\mathrm{freq})}$.
-Following Clune et al.
-2009~\cite{clune2009evolving}, fitness equation was $2^{\mathrm{distance}^{2}}$.
+We calculated the frequency of a gait as the average number of servo direction changes per leg per second.
+If this frequency was higher than an experimentally-selected threshold of 1.67 Hz, the measure of distance traveled by the center of mass during the gait was reduced exponentially by multiplying it by a discount factor of $e^{(1.67-\mathrm{freq})}$.
+Following Clune et al.~\cite{clune2009evolving}, the fitness equation was $2^{\mathrm{distance}^{2}}$.
 After evolving the gaits in simulation, the champion gait of the last generation of each of the 20 runs was transferred onto the real robot and the distance traveled was measured.

### 2012-11-05 paper_sean/results.tex
final JMC pass. lots of small, but important, changes

@@
 cm/s $\pm$ 3.4 SD versus 16.7 cm/s $\pm$ 1.9 SD, \tabref{results}).
-Plots of servo positions over time reveal that the evolved HyperNEAT gaits are regular, smooth and coordinated~(\figref{threeplots}-Left), confirming previous results with HyperNEAT in simulation~\cite{clune2009evolving,clune2011performance}.
+Plots of servo positions over time reveal that the evolved HyperNEAT gaits are regular, smooth and coordinated~(\figref{threeplots}-Left), confirming previous results with HyperNEAT in simulation~\cite{clune2011performance,clune2009evolving}.
 
@@
 \textbf{Top: }
-HyperNEAT outperforms a genetic algorithm with a simpler encoding, when both algorithms are combined with a simulator.
-Plotted are means over 20 runs in simulation (solid lines) $\pm$ STD (dashed lines).
+HyperNEAT outperforms a genetic algorithm with a simple encoding when both algorithms are combined with a simulator.
+Plotted are means over 20 runs in simulation (solid lines) $\pm$ SD (dashed lines).
 HyperNEAT gaits are 54.3\% faster in
@@
 \textbf{Left: }
-Servo positions over time (for nine servos) for a simulated HyperNEAT gait. HyperNEAT produces gaits that are smooth, symmetrical, and contain other regularities. 
+Servo positions over time (for nine servos) for a representative simulated HyperNEAT gait. HyperNEAT produced gaits that are smooth, symmetrical, and contained complex regularities. 
 \textbf{Right: }Mean gait frequency averaged over 20 runs. 
-Gaits with frequencies above a threshold (horizontal line) receive a fitness penalty. HyperNEAT quickly learns to produce gaits with frequencies low enough to avoid this penalty.
+Gaits with frequencies above a threshold (horizontal line) receive a fitness penalty. HyperNEAT quickly learned to produce gaits with frequencies low enough to avoid this penalty.
 }
@@
 inherently noisy, but because even two copies of the same robot are
-not identical and may produce different gait speeds for the same input
+not identical and may produce different speeds for the same input
 gait.
@@
 Intelligent Systems (ROBIN) lab at the University of Oslo.
+The two robots were evaluated on two different surfaces, and had different material enveloping their feet to increase friction (compare \figref{quadratotWhiteBooties}-Left to Figure 2 of Glette et al).
 
@@
 on the same robotic hardware, we tentatively
-conclude that HyperNEAT produces faster gaits for the QuadraTot robot. This conclusion is supported by the fact that HyperNEAT outperformed the encoding from Glette et al. in simulation.
+conclude that HyperNEAT produces faster gaits for the QuadraTot robot. This conclusion is supported by the fact that HyperNEAT also outperformed the encoding from Glette et al. in simulation.
 
@@
 \vspace{.35cm}
-\caption{Velocities of learned gaits in simulation and on two different copies of the QuadraTot robot.
+\caption{Velocities of evolved gaits in simulation and on two different copies of the QuadraTot robot.
 Subject to availability, data are reported from the experiments for this paper and three previous studies.
-Reported are the total number of evaluations per run, the average of the fastest gaits produced in each run in simulation, and the fastest gait produced on two different copies of the QuadraTot robot (CCML and ROBIN, see text).
+Reported are the total number of evaluations per run, the average of the fastest gaits produced in each run in simulation, and the fastest gait produced on the CCML and ROBIN copies of the QuadraTot robot (see text).
 *Instead of using the single fitness value reported in~\cite{glette}, we ran 19 additional runs and used the average fitness of those 20 runs.
@@
 
-We now discuss the differences between the gaits evolved by HyperNEAT directly on the physical robot, and those first evolved in a simulator and then transferred to the robot. 
+We now discuss the differences between the gaits evolved by HyperNEAT directly on the physical robot (Yosinski et al.), and those first evolved in a simulator and then transferred to the robot (this study).
 On the physical robot, the gaits produced by HyperNEAT with a simulator were faster, more natural, and more repeatable than those evolved directly on the QuadraTot robot~\cite{yos:clune}.
@@
 
-The simulator likely improved performance because of the number of evaluations it enabled, both in terms of the population (200 vs. 9) and the number of generations (200 vs. 20), leading
+The simulator likely improved performance because of the number of evaluations it enabled, both in terms of the population size (200 vs. 9) and the number of generations (200 vs. 20), leading
 to a total difference of 40000 vs 180 per evolutionary run~\cite{yos:clune}. Another
 potential cause of improved performance is that with a simulator
-HyperNEAT had much less noise in the evaluation process during early
-generations and could thus find coordinated, regular, gaits, which are
-both higher-performing and more likely to transfer. On the physical robot, the noise in the evaluation was
-substantial, preventing effective learning and the discovery of
-effective gaits~\cite{yos:clune}. It could also be the case that 
-the addition in this study of a punishment for high-frequency gaits
+HyperNEAT had much less noise in the evaluation process and could thus find coordinated, regular, gaits, which perform better. On the physical robot, the noise in the evaluation was
+substantial, preventing effective learning~\cite{yos:clune}. It could also be the case that 
+the encouragement of low-frequency gaits in this study
 aided performance, especially since in simulation the gaits were
@@
 the design of the robot: the force required to lift the robot in many
-poses was too much for the servos. Future studies will use a
+poses was too much for the servos. In future studies we will use a
 robot that has more mechanical advantage and requires less torque from
@@
 which was designed in response to the challenges experienced with the
-QuadraTot platform. Future studies will also incorporate techniques to minimize the gap between the simulator and reality~\cite{koos2010crossing,bongard,zagal}.
+QuadraTot platform. In future work we will also incorporate techniques to minimize the gap between the simulator and reality~\cite{koos2010crossing,bongard,zagal}.

### 2012-11-05 paper_sean/robot.tex
final JMC pass. lots of small, but important, changes

@@
 
-We performed the experiments on the QuadraTot quadrupedal robot platform~(\figref{quadratotWhiteBooties})~\cite{yos:clune}.
-It has 9 degrees of freedom: two joints per leg and one joint that allows rotation along the robot's midline.
+We performed experiments on the QuadraTot quadrupedal robot platform~(\figref{quadratotWhiteBooties})~\cite{yos:clune}.
+It has 9 degrees of freedom: two joints per leg and one joint that rotates along the robot's midline.
 %we def want a picture of the cool robot.
@@
 Try to get this figure to show up as early as possible in the paper (e.g top of the 2nd page).
-The QuadraTot hardware designs and the software for this project are open source\footnote{A parts list, hardware CAD files, software (including the simulator), and gait videos are available at http://creativemachines.cornell.edu/evolved-quadruped-gaits}, and all hardware components are either off-the-shelf or 3D printed parts.
+The QuadraTot hardware designs and the software for this project are open source\footnote{A parts list, hardware CAD files, software (including the simulator), and gait videos are available at http://creativemachines.cornell.edu/evolved-quadruped-gaits}, and all hardware components are either off-the-shelf or 3D-printed.
 %These traits enable other labs to perform research using the same platform, which eases comparison of different gait-learning algorithms.
-To date there are results on the platform for nine different learning algorithms from three previous publications~\cite{yos:clune,glette,haocheng}.
+There are results on the platform for nine different learning algorithms from three previous publications~\cite{yos:clune,glette,haocheng}.
 
-The joints are powered by Robotis Dynamixel servos; five AX-18A servos for the inner joint of each leg and the single midline joint, and four AX-12A servos for the the outer joint of each leg, which requires less power and can thus have less expensive motors.
+The joints are powered by Robotis Dynamixel servos; five AX-18A servos for the inner joints of each leg and the single midline joint, and four AX-12A servos for the the outer joints of each leg, which require less power and can thus have less expensive motors.
 The servos were sent new positions at 40Hz via the Pydynamixel library.
@@
 Yosinski et al.
-had the small Linux computer that performed all computation on the robot, but we removed it and sent commands to the robot via a cable.
+had the small Linux computer that performed all computation on the robot, but we removed it and sent commands from it to the robot via a cable.
 We tracked the robot's position using an infrared LED observed by a Wiimote.

### 2012-11-05 paper_sean/simulator.tex
final JMC pass. lots of small, but important, changes

@@
 \subsubsection{Simulator:}
-Gaits evolved in the simulator from Glette et al., which represents QuadraTot in the Nvidia PhysX physics engine\footnotemark[\value{footnote}], including the mass and size of the QuadraTot components and its degrees of freedom~(\figref{quadratotWhiteBooties}, right).
+Gaits evolved in the simulator from Glette et al., which represents QuadraTot in the Nvidia PhysX physics engine\footnotemark[\value{footnote}], including the mass and size of the QuadraTot components and its degrees of freedom~(\figref{quadratotWhiteBooties}-Right).
 In the simulator, each individual joint range was limit as described above, but Smart Cropping was not included because we found that it hindered performance by limiting the types of gaits evolution could explore in early generations.

### 2012-11-06 paper_sean/hyperneat.tex
added changes we talked about today..near final?

@@
 The ANN had a fixed topology of three \begin{math}3\times4\end{math} Cartesian grids of nodes for the input, hidden, and output layers. 
-The inputs to the substrate were the angles requested in the prevous time step for each of the 9 joints of the robot and a sine and cosine wave to facilitate periodic motion. 
-The outputs of the substrate at each time step were nine numbers (for each joint) in the range [-1, 1] which were scaled to the range [0, 1023], the allowable ranges for the servos. 
+The inputs to the substrate were the angles requested in the previous time step for each of the 9 joints of the robot and a sine and cosine wave to facilitate periodic motion. 
+The outputs of the substrate at each time step were nine numbers (for each joint) in the range [-1, 1] which were scaled to the allowable ranges for the servos. 
 As in Yosinski et al.~\cite{yos:clune}, we generated pseudo-positions at 160Hz and then downsampled over consecutive blocks of four time steps to obtain the actual commanded positions at 40Hz; this reduced the number of gaits which commanded switches from extreme negative to extreme positive numbers at 40Hz, which overly taxed the servos.

### 2012-11-06 paper_sean/results.tex
added changes we talked about today..near final?

@@
 from Glette et al. 10 times on CCML and measured an average speed of
-only 12.95 cm/s (vs. 17.8 cm/s measured on ROBIN) and a maximum of
+only 12.95 cm/s $\pm0.93$ SD (vs. 17.8 cm/s measured on ROBIN) and a maximum of
 13.8 cm/s (\tabref{results}).
@@
 Subject to availability, data are reported from the experiments for this paper and three previous studies.
-Reported are the total number of evaluations per run, the average of the fastest gaits produced in each run in simulation, and the fastest gait produced on the CCML and ROBIN copies of the QuadraTot robot (see text).
+Reported are the total number of evaluations per run, the average of the fastest gaits produced in each run in simulation, and the single fastest gait produced on the CCML and ROBIN copies of the QuadraTot robot (see text for their differences).
 *Instead of using the single fitness value reported in~\cite{glette}, we ran 19 additional runs and used the average fitness of those 20 runs.
-Velocities are in cm/s, and bold indicates the best performance.}
+Velocities are in cm/s, and bold indicates the best performance.
+}
 \tablabel{results}

### 2012-11-07 paper_sean/abstract.tex
final read thru by Sean, looked for grammar/spelling/capitalization errors

@@
 In this paper, we tested the hypothesis that the beneficial properties of HyperNEAT would outperform the simpler encoding if HyperNEAT gaits are first evolved in simulation before being transferred to reality.
-That hypothesis was confirmed, resulting in the fastest gaits yet observed for this robot, including those produced by nine different algorithms from three previous papers describing techniques for generating gaits for this robot.
+That hypothesis was confirmed, resulting in the fastest gaits yet observed for this robot, including those produced by nine different algorithms from three previous papers describing techniques for gait generation on this robot.
 This result is important because it confirms that the early promise shown by generative encodings, specifically HyperNEAT, are not limited to simulation, but indeed work on challenging real-world engineering challenges such as evolving gaits for real robots.

### 2012-11-07 paper_sean/hyperneat.tex
final read thru by Sean, looked for grammar/spelling/capitalization errors

@@
 For example, when CPPNs encode 2D pictures, the coordinates of each pixel are iteratively input into the genome and the output is the grayscale value at that coordinate~\cite{secretan2011picbreeder}.
-Because a CPPN network is composed of math functions, those functions can create geometric regularities in the phenotype.
+Because a CPPN network is composed of math functions, these functions can create geometric regularities in the phenotype.
 For example, a Gaussian function of an axis can provide symmetry (e.g.
@@
 Just as in 2D pictures and 3D objects, the CPPN can create complex, regular, geometric patterns (e.g.
-left-right symmetry or repeated modules), but in this case the patterns are in the weights of the neural network~\cite{clune2011performance}.
+left-right symmetry or repeated modules), but in this case the patterns are in the weights of a neural network~\cite{clune2011performance}.
 The neural regularities produced by HyperNEAT enable significantly improved performance on problems that are regular~\cite{clune2011performance,stanley2009hypercube}, including evolving quadruped gaits in simulation~\cite{clune2011performance,clune2009evolving,clune2009sensitivity}.
 
-In HyperNEAT, the CPPN genomes evolve via the Neuroevolution of Augmenting Topologies (NEAT) algorithm~\cite{stanley2002evolving}, which has three major components. 
+In HyperNEAT, the CPPN genomes evolve via the NeuroEvolution of Augmenting Topologies (NEAT) algorithm~\cite{stanley2002evolving}, which has three major components. 
 First, NEAT starts with small genomes that encode simple networks and complexifies them via mutations that add nodes and links to the networks.

### 2012-11-07 paper_sean/methods.tex
final read thru by Sean, looked for grammar/spelling/capitalization errors

@@
 Each run had a population size of 200 and lasted 200 generations.
-We performed twenty HyperNEAT runs that differed only in the seed provided to the random number generator, which affected stochastic events such as mutation.
+We performed 20 HyperNEAT runs that differed only in the seed provided to the random number generator, which affected stochastic events such as mutation.
 To make a statistical comparison to the encoding from Glette et al., we conducted 19 runs using the original code to supplement the one run performed for that paper.
@@
 We calculated the frequency of a gait as the average number of servo direction changes per leg per second.
-If this frequency was higher than an experimentally-selected threshold of 1.67 Hz, the measure of distance traveled by the center of mass during the gait was reduced exponentially by multiplying it by a discount factor of $e^{(1.67-\mathrm{freq})}$.
+If this frequency was higher than the experimentally-selected threshold of 1.67 Hz, the measure of distance traveled by the center of mass during the gait was reduced exponentially by multiplying it by a discount factor of $e^{(1.67-\mathrm{freq})}$.
 Following Clune et al.~\cite{clune2009evolving}, the fitness equation was $2^{\mathrm{distance}^{2}}$.

### 2012-11-07 paper_sean/results.tex
final read thru by Sean, looked for grammar/spelling/capitalization errors

@@
 \textbf{Left: }
-Servo positions over time (for nine servos) for a representative simulated HyperNEAT gait. HyperNEAT produced gaits that are smooth, symmetrical, and contained complex regularities. 
+Servo positions over time (for nine servos) for a representative simulated HyperNEAT gait. HyperNEAT produced smooth and symmetrical gaits that contained complex regularities. 
 \textbf{Right: }Mean gait frequency averaged over 20 runs. 
@@
 The simulator likely improved performance because of the number of evaluations it enabled, both in terms of the population size (200 vs. 9) and the number of generations (200 vs. 20), leading
-to a total difference of 40000 vs 180 per evolutionary run~\cite{yos:clune}. Another
+to a total difference of 40000 vs 180 per evolutionary run compared to Yosinski et al.~\cite{yos:clune}. Another
 potential cause of improved performance is that with a simulator

### 2012-11-07 paper_sean/robot.tex
final read thru by Sean, looked for grammar/spelling/capitalization errors

@@
 
-We performed experiments on the QuadraTot quadrupedal robot platform~(\figref{quadratotWhiteBooties})~\cite{yos:clune}.
+We performed experiments on the QuadraTot quadrupedal robot platform~(\figref{quadratotWhiteBooties}-Left)~\cite{yos:clune}.
 It has 9 degrees of freedom: two joints per leg and one joint that rotates along the robot's midline.
@@
 Each servo has a built-in safety mechanism that shuts itself off to prevent damage if the servo's current, range, temperature, or torque is too high.
-During evolution, this safety mechanism frequently activated, and did so inconsistently, adding significant noise to the evaluation process.
+This safety mechanism activated frequenty and inconsistently, adding significant noise to the evaluation process.
 As pointed out in a previous study~\cite{yos:clune}, gaits generated during evolution on QuadraTot are highly variable and produce many shutdowns because they force the servos to exert too much torque.

### 2012-11-07 paper_sean/simulator.tex
final read thru by Sean, looked for grammar/spelling/capitalization errors

@@
 \subsubsection{Simulator:}
-Gaits evolved in the simulator from Glette et al., which represents QuadraTot in the Nvidia PhysX physics engine\footnotemark[\value{footnote}], including the mass and size of the QuadraTot components and its degrees of freedom~(\figref{quadratotWhiteBooties}-Right).
-In the simulator, each individual joint range was limit as described above, but Smart Cropping was not included because we found that it hindered performance by limiting the types of gaits evolution could explore in early generations.
+Gaits evolved in the simulator from Glette et al., which represents QuadraTot in the Nvidia PhysX physics engine, including the mass and size of the QuadraTot components and its degrees of freedom~(\figref{quadratotWhiteBooties}-Right).\footnotemark[\value{footnote}] In the simulator, each individual joint range was limit as described above, but Smart Cropping was not included because we found that it hindered performance by limiting the types of gaits evolution could explore in early generations.
 

### 2012-11-07 paper_sean/abstract.tex
final comit git statusgit statusgit statusgit statusgit statusgit statusgit statusgit statusgit statusgit statusgit statusgit statusgit status

@@
 In this paper, we tested the hypothesis that the beneficial properties of HyperNEAT would outperform the simpler encoding if HyperNEAT gaits are first evolved in simulation before being transferred to reality.
-That hypothesis was confirmed, resulting in the fastest gaits yet observed for this robot, including those produced by nine different algorithms from three previous papers describing techniques for gait generation on this robot.
-This result is important because it confirms that the early promise shown by generative encodings, specifically HyperNEAT, are not limited to simulation, but indeed work on challenging real-world engineering challenges such as evolving gaits for real robots.
+That hypothesis was confirmed, resulting in the fastest gaits yet observed for this robot, including those produced by nine different algorithms from three previous papers describing gait-generating techniques for this robot.
+This result is important because it confirms that the early promise shown by generative encodings, specifically HyperNEAT, are not limited to simulation, but work on challenging real-world engineering challenges such as evolving gaits for real robots.
 

### 2012-11-07 paper_sean/robot.tex
final comit git statusgit statusgit statusgit statusgit statusgit statusgit statusgit statusgit statusgit statusgit statusgit statusgit status

@@
 This safety mechanism activated frequenty and inconsistently, adding significant noise to the evaluation process.
-As pointed out in a previous study~\cite{yos:clune}, gaits generated during evolution on QuadraTot are highly variable and produce many shutdowns because they force the servos to exert too much torque.
+As pointed out in a previous study~\cite{yos:clune}, evolved gaits on QuadraTot are highly variable and produce many shutdowns because they force the servos to exert too much torque.
 To prevent collisions between different pieces of the robot's body, we limited the allowable range of movement for the inner, outer, and hip joints to [-85\degree, +60\degree], [-113\degree, +39\degree], and [-28\degree, +28\degree], respectively.

### 2012-11-07 paper_sean/introduction.tex
THIS COMMIT CONTAINS DOUBLE-BLIND REVIEW CHANGES

@@
 \textbf{Left: } The physical robot, which is composed of 3-D printable and off-the-shelf components.
-\textbf{Right: } The robot in the simulator.}
+\textbf{Right: } The representation of the robot in the simulator.}
 
@@
 evolved gaits with HyperNEAT directly in hardware on the QuadraTot robot platform~(\figref{quadratotWhiteBooties}).
-They found that HyperNEAT's gaits outperformed manually designed, parameterized learning algorithms, but still did not produce impressive, natural gaits\footnote{Videos available at http://creativemachines.cornell.edu/evolved-quadruped-gaits}~\cite{yos:clune}.
-A follow-up study built a simulator for QuadraTot to test whether the inclusion of a simulator would improve results and found that it did:  when gaits were evolved in this simulator with a simple direct encoding and then transferred to the QuadraTot robot, the resulting gaits were faster than the gaits produced by evolving gaits with HyperNEAT directly on the robot~\cite{glette}.
+They found that HyperNEAT's gaits outperformed manually designed, parameterized learning algorithms, but still did not produce impressive, natural gaits~\cite{yos:clune}.\footnote{Videos available at (Link removed for double-blind review)}
+A follow-up study built a simulator for QuadraTot to test whether the inclusion of a simulator would improve results and found that it did:  when gaits were evolved in simulation with a simple direct encoding and then transferred to the QuadraTot robot, the resulting gaits were faster than the gaits produced by evolving gaits with HyperNEAT directly on the robot~\cite{glette}.
 The simulator helped because it afforded much larger population sizes and more generations than were possible when evolving directly in hardware, resulting in 333 times more evaluations per run (60000 vs.

### 2012-11-07 paper_sean/results.tex
THIS COMMIT CONTAINS DOUBLE-BLIND REVIEW CHANGES

@@
 gait.
-Gaits for this study and two previous studies~\cite{yos:clune,haocheng} were evaluated on a copy of the QuadraTot robot in the Cornell Creative Machines Lab (CCML), while the gaits in Glette et al.
+Gaits for this study and two previous studies~\cite{yos:clune,haocheng} were evaluated on a copy of the QuadraTot robot in the (Name removed for double-blind review) Lab (---L), while the gaits in Glette et al.
 were evaluated on a different copy of the same robot in the Robotics and
@@
 fastest gait produced by HyperNEAT with a simulator in the experiments for this paper traveled 14.5
-cm/s on the CCML QuadraTot. It was unclear whether this difference in performance was due to
+cm/s on the ---L QuadraTot. It was unclear whether this difference in performance was due to
 the differences in the gaits themselves or dissimilarities between
 hardware. To control for this possibility, we ran the fastest gait
-from Glette et al. 10 times on CCML and measured an average speed of
+from Glette et al. 10 times on ---L and measured an average speed of
 only 12.95 cm/s $\pm0.93$ SD (vs. 17.8 cm/s measured on ROBIN) and a maximum of
 13.8 cm/s (\tabref{results}).
-It thus appears that the CCML version of the robot is slower.
+It thus appears that the ---L version of the robot is slower.
 Moreover, the HyperNEAT gait (14.5 cm/s) is faster than the best gait of Glette et al.
-on the CCML QuadraTot.
+on the ---L QuadraTot.
 Unfortunately, it was not possible to test the best HyperNEAT gait on the ROBIN QuadraTot.
@@
                                          &              & Simulated  & Real Vel. & Real Vel.  \\
-                                         & Evaluations  & Velocity &    (CCML)     & (ROBIN) \\
+                                         & Evaluations  & Velocity &    (---L)     & (ROBIN) \\
 \hline
@@
 Subject to availability, data are reported from the experiments for this paper and three previous studies.
-Reported are the total number of evaluations per run, the average of the fastest gaits produced in each run in simulation, and the single fastest gait produced on the CCML and ROBIN copies of the QuadraTot robot (see text for their differences).
+Reported are the total number of evaluations per run, the average of the fastest gaits produced in each run in simulation, and the single fastest gait produced on the ---L and ROBIN copies of the QuadraTot robot (see text for their differences).
 *Instead of using the single fitness value reported in~\cite{glette}, we ran 19 additional runs and used the average fitness of those 20 runs.

### 2012-11-07 paper_sean/robot.tex
THIS COMMIT CONTAINS DOUBLE-BLIND REVIEW CHANGES

@@
 Try to get this figure to show up as early as possible in the paper (e.g top of the 2nd page).
-The QuadraTot hardware designs and the software for this project are open source\footnote{A parts list, hardware CAD files, software (including the simulator), and gait videos are available at http://creativemachines.cornell.edu/evolved-quadruped-gaits}, and all hardware components are either off-the-shelf or 3D-printed.
+The QuadraTot hardware designs and the software for this project are open source\footnote{A parts list, hardware CAD files, software (including the simulator), and gait videos are available at (Link removed for double-blind review)}, and all hardware components are either off-the-shelf or 3D-printed.
 %These traits enable other labs to perform research using the same platform, which eases comparison of different gait-learning algorithms.

### 2012-11-07 paper_sean/introduction.tex
link added for anonymous youtube video

@@
 evolved gaits with HyperNEAT directly in hardware on the QuadraTot robot platform~(\figref{quadratotWhiteBooties}).
-They found that HyperNEAT's gaits outperformed manually designed, parameterized learning algorithms, but still did not produce impressive, natural gaits~\cite{yos:clune}.\footnote{Videos available at (Link removed for double-blind review)}
+They found that HyperNEAT's gaits outperformed manually designed, parameterized learning algorithms, but still did not produce impressive, natural gaits~\cite{yos:clune}.\footnote{Videos available at http://creativemachines.cornell.edu/evolved-quadruped-gaits}
 A follow-up study built a simulator for QuadraTot to test whether the inclusion of a simulator would improve results and found that it did:  when gaits were evolved in simulation with a simple direct encoding and then transferred to the QuadraTot robot, the resulting gaits were faster than the gaits produced by evolving gaits with HyperNEAT directly on the robot~\cite{glette}.

### 2012-11-07 paper_sean/results.tex
link added for anonymous youtube video

--- 2012-11-07 paper_sean/results.tex
@@
 The simulator enabled HyperNEAT to evolve
-fast gaits that look like those of natural animals.
-In simulation, HyperNEAT gaits were significantly faster than those from Glette et al.
+fast gaits that look like those of natural animals.\footnote{Videos of the evolved gaits in simulation and on hardware are available at http://youtu.be/bg90l2aCiz8} In simulation, HyperNEAT gaits were significantly faster than those from Glette et al.
 (\figref{threeplots}-Top, $\emph{p} < 6.8\times10^{-8}$, comparing the best gaits in the final generation of each run via MATLAB's Wilcoxon rank sum test).
@@
 gait.
-Gaits for this study and two previous studies~\cite{yos:clune,haocheng} were evaluated on a copy of the QuadraTot robot in the Cornell Creative Machines Lab (CCML), while the gaits in Glette et al.
+Gaits for this study and two previous studies~\cite{yos:clune,haocheng} were evaluated on a copy of the QuadraTot robot in the (Name removed for double-blind review) Lab (ANON), while the gaits in Glette et al.
 were evaluated on a different copy of the same robot in the Robotics and
@@
 fastest gait produced by HyperNEAT with a simulator in the experiments for this paper traveled 14.5
-cm/s on the CCML QuadraTot. It was unclear whether this difference in performance was due to
+cm/s on the ANON QuadraTot. It was unclear whether this difference in performance was due to
 the differences in the gaits themselves or dissimilarities between
 hardware. To control for this possibility, we ran the fastest gait
-from Glette et al. 10 times on CCML and measured an average speed of
+from Glette et al. 10 times on ANON and measured an average speed of
 only 12.95 cm/s $\pm0.93$ SD (vs. 17.8 cm/s measured on ROBIN) and a maximum of
 13.8 cm/s (\tabref{results}).
-It thus appears that the CCML version of the robot is slower.
+It thus appears that the ANON version of the robot is slower.
 Moreover, the HyperNEAT gait (14.5 cm/s) is faster than the best gait of Glette et al.
-on the CCML QuadraTot.
+on the ANON QuadraTot.
 Unfortunately, it was not possible to test the best HyperNEAT gait on the ROBIN QuadraTot.
@@
                                          &              & Simulated  & Real Vel. & Real Vel.  \\
-                                         & Evaluations  & Velocity &    (CCML)     & (ROBIN) \\
+                                         & Evaluations  & Velocity &    (ANON)     & (ROBIN) \\
 \hline
@@
 Subject to availability, data are reported from the experiments for this paper and three previous studies.
-Reported are the total number of evaluations per run, the average of the fastest gaits produced in each run in simulation, and the single fastest gait produced on the CCML and ROBIN copies of the QuadraTot robot (see text for their differences).
+Reported are the total number of evaluations per run, the average of the fastest gaits produced in each run in simulation, and the single fastest gait produced on the ANON and ROBIN copies of the QuadraTot robot (see text for their differences).
 *Instead of using the single fitness value reported in~\cite{glette}, we ran 19 additional runs and used the average fitness of those 20 runs.

### 2013-01-15 paper_sean/methods.tex
edited paper for camera ready version. not finished

@@
 To make a statistical comparison to the encoding from Glette et al., we conducted 19 runs using the original code to supplement the one run performed for that paper.
-Each gait was evaluated for fourteen seconds in reality, with interpolation from and to a stationary position in the first and last two seconds, respectively, as done in Yosinski et al.
-The result was effectively 12 seconds of full-speed
+Each gait was evaluated for fourteen seconds in reality, with interpolation from and to a stationary position in the first and last two seconds, respectively, as in Yosinski et al.
+effectively resulting in 12 seconds of full-speed
 motion. In simulation gaits were evaluated for 12 seconds. 

### 2013-01-15 paper_sean/results.tex
edited paper for camera ready version. not finished

--- 2012-11-07 paper_sean/results.tex
@@
 The simulator enabled HyperNEAT to evolve
-fast gaits that look like those of natural animals.
+fast, organic gaits.
 In simulation, HyperNEAT gaits were significantly faster than those from Glette et al.
 (\figref{threeplots}-Top, $\emph{p} < 6.8\times10^{-8}$, comparing the best gaits in the final generation of each run via MATLAB's Wilcoxon rank sum test).
-Specifically, HyperNEAT gaits were 55\% faster on average in simulation (25.4
-cm/s $\pm$ 3.4 SD versus 16.7 cm/s $\pm$ 1.9 SD, \tabref{results}).
+Specifically, HyperNEAT gaits were 55\% faster in simulation (26.9
+cm/s, 95\% CIs 23.8 cm/s and 26.750 cm/s versus 16.7 cm/s $\pm$ 1.9 SD, \tabref{results}).
 Plots of servo positions over time reveal that the evolved HyperNEAT gaits are regular, smooth and coordinated~(\figref{threeplots}-Left), confirming previous results with HyperNEAT in simulation~\cite{clune2011performance,clune2009evolving}.
 
-\figOneOnTwo[h!]{threeplots}{FINAL_sean_kyrre_fitness}{0.62}{servo_plot_111}{.48}{frequencyPlotThreshFinal}{.48}
+\figOneOnTwo[h!]{threeplots}{med_fit_gen}{0.62}{servo_plot_111}{.48}{frequencyPlotThreshFinal}{.48}
+%\figOneOnTwo[h!]{threeplots}{med_fit_gen}{0.00}{servo_plot_111}{.48}{frequencyPlotThreshFinal}{.48}
 {
@@
 HyperNEAT outperforms a genetic algorithm with a simple encoding when both algorithms are combined with a simulator.
-Plotted are means over 20 runs in simulation (solid lines) $\pm$ SD (dashed lines).
-HyperNEAT gaits are 54.3\% faster in
+Plotted are medians over 20 runs in simulation (solid lines) $\pm$ CIs (dashed lines).
+HyperNEAT gaits are 61.1\% faster in
 simulation and 5.1\% faster in reality~(\tabref{results}).
@@
 Subject to availability, data are reported from the experiments for this paper and three previous studies.
-Reported are the total number of evaluations per run, the average of the fastest gaits produced in each run in simulation, and the single fastest gait produced on the CCML and ROBIN copies of the QuadraTot robot (see text for their differences).
+Reported are the total number of evaluations per run, the median of the fastest gaits produced in each run in simulation, and the single fastest gait produced on the CCML and ROBIN copies of the QuadraTot robot (see text for their differences).
 *Instead of using the single fitness value reported in~\cite{glette}, we ran 19 additional runs and used the average fitness of those 20 runs.
@@
 HyperNEAT had much less noise in the evaluation process and could thus find coordinated, regular, gaits, which perform better. On the physical robot, the noise in the evaluation was
-substantial, preventing effective learning~\cite{yos:clune}. It could also be the case that 
+substantial, preventing effective learning~\cite{yos:clune}. 
+However, we can likely rule out the noise hypothesis because we performed 20 
+runs in simulation with the same evaluation samples as 
+in Yosinski et al.~\cite{yos:clune} and saw that the results were not 
+significantly different (mean fitness of 7.9 cm/s), making the likely explanation 
+the number of evaluations performed.
+Another factor for high-perfoming gaits may be due to
 the encouragement of low-frequency gaits in this study
@@
 high-frequency in a few early generations before rapidly settling to a
-frequency range below the penalized threshold~(\figref{threeplots}-Right).
+ range below the penalized threshold~(\figref{threeplots}-Right).
 
@@
 which was designed in response to the challenges experienced with the
-QuadraTot platform. In future work we will also incorporate techniques to minimize the gap between the simulator and reality~\cite{koos2010crossing,bongard,zagal}.
+QuadraTot platform.
+That we did not model the servos in simulation, especially with their frequent failures, suggests that even better results could be obtained via a simulator that contained or learned servo models.
+In future work we will also incorporate techniques to minimize the gap between the simulator and reality~\cite{koos2010crossing,bongard,zagal}.

### 2013-01-15 paper_sean/introduction.tex
shortened paper to 10 pages added p-value

@@
 
-The work with the simulator (Glette et al.) evolved gaits with a simple encoding that was manually constrained to produce specific regularities.
+The work with the simulator~\cite{glette} evolved gaits with a simple encoding manually constrained to produce specific regularities.
 The success of that work raises the question of whether the performance gains were due to the added benefit of a simulator or the use of a simple, hand-designed encoding.
 We hypothesized that HyperNEAT, which has been previously shown to automatically discover complex regularities to produce high-performing gaits~\cite{clune2011performance,yos:clune}, would outperform the simpler encoding from Glette et al.
-if HyperNEAT is combined with a simulator.
+if combined with a simulator.
 Here we test that hypothesis by evolving gaits with HyperNEAT in the same simulator from Glette et al.
 and then transferring those gaits to the QuadraTot robot.
-Our experiments confirmed the hypothesis: with the simulator, HyperNEAT evolved the highest-performing gaits observed to date for the QuadraTot robot platform.
+Our experiments confirmed the hypothesis: with the simulator, HyperNEAT evolved the highest-performing gaits observed to date for the QuadraTot platform.
 

### 2013-01-15 paper_sean/results.tex
shortened paper to 10 pages added p-value

@@
 (\figref{threeplots}-Top, $\emph{p} < 6.8\times10^{-8}$, comparing the best gaits in the final generation of each run via MATLAB's Wilcoxon rank sum test).
-Specifically, HyperNEAT gaits were 55\% faster in simulation (26.9
+Specifically, HyperNEAT gaits were 61.1\% faster in simulation (26.9
 cm/s, 95\% CIs 23.8 cm/s and 26.750 cm/s versus 16.7 cm/s $\pm$ 1.9 SD, \tabref{results}).
@@
 \hline
-HyperNEAT + simulator [this paper]                     & 40000       & \textbf{25.4}       &   \textbf{14.5} & --\\
+HyperNEAT + simulator [this paper]                     & 40000       & \textbf{26.9}       &   \textbf{14.5} & --\\
 \hline
@@
 
-We now discuss the differences between the gaits evolved by HyperNEAT directly on the physical robot (Yosinski et al.), and those first evolved in a simulator and then transferred to the robot (this study).
+We now discuss the differences between the gaits evolved by HyperNEAT directly on the physical robot~\cite{yos:clune}, and those first evolved in a simulator and then transferred to the robot (this study).
 On the physical robot, the gaits produced by HyperNEAT with a simulator were faster, more natural, and more repeatable than those evolved directly on the QuadraTot robot~\cite{yos:clune}.
@@
 in Yosinski et al.~\cite{yos:clune} and saw that the results were not 
-significantly different (mean fitness of 7.9 cm/s), making the likely explanation 
+significantly different ($p < 0.855$, mean fitness 7.9 cm/s), making the likely explanation 
 the number of evaluations performed.
-Another factor for high-perfoming gaits may be due to
-the encouragement of low-frequency gaits in this study
-aided performance, especially since in simulation the gaits were
+The encouragement of low-frequency gaits in this study
+also may have aided performance, especially since in simulation the gaits were
 high-frequency in a few early generations before rapidly settling to a

### 2013-01-15 paper_sean/conclusion.tex
reverted back to means/std

@@
 
-With a simulator, HyperNEAT evolved the fasted gait yet recorded for the QuadraTot robot, outperforming nine machine learning algorithms from three previous publications~\cite{yos:clune,glette,haocheng}, including an improvement of 54.3\% in simulation and 5.1\% in reality over the previous best QuadraTot gait by Glette et al.
+With a simulator, HyperNEAT evolved the fasted gait yet recorded for the QuadraTot robot, outperforming nine machine learning algorithms from three previous publications~\cite{yos:clune,glette,haocheng}, including an improvement of 52.1\% in simulation and 5.1\% in reality over the previous best QuadraTot gait by Glette et al.
 These results provide an important demonstration that the HyperNEAT generative encoding can evolve state-of-the-art results for challenging engineering problems, in this case evolving gaits for a legged robot.
 The results further reaffirm the benefits of using simulators when solving real-world challenges with evolutionary algorithms.
-Our results additionally confirm---with a different robot and simulator---previous work that has shown that HyperNEAT is an effective encoding for automatically evolving coordinated, regular gaits in simulation~\cite{clune2009evolving,clune2011performance,yos:clune}.
+Our results additionally confirm---with a different robot and simulator---previous work that has shown that HyperNEAT is an effective encoding for automatically evolving coordinated, regular gaits in simulation~\cite{yos:clune,clune2011performance,clune2009evolving}.
 That HyperNEAT outperformed the encoding hand-designed by Glette et al.

### 2013-01-15 paper_sean/introduction.tex
reverted back to means/std

@@
 However, designing gaits for legged robots is a difficult and time-consuming process for human engineers~\cite{strom1999legged,wettergreen1992gait}, and must be repeated every time a robot is created or modified~\cite{hornby2005autonomous}.
-Scientists thus investigate how to automatically produce gaits via machine learning and evolutionary algorithms, and the result is often a better gait than those created by human engineers~\cite{valsalam:mii,kohl:stone,hornby2005autonomous,hornby2003generative,yos:clune}.
+Scientists thus investigate how to automatically produce gaits via machine learning and evolutionary algorithms, and the result is often a better gait than those created by human engineers~\cite{hornby2005autonomous,valsalam:mii,kohl:stone,hornby2003generative,yos:clune}.
 While it has been shown that gaits perform better if they are \emph{regular}---i.e., that they have coordinated movements, such as left-right symmetry or front-back symmetry~\cite{valsalam:mii,clune2011performance,clune2009evolving,clune2009sensitivity}---experimenters usually have to explicitly decide and specify these regularities~\cite{valsalam:mii,tellez,beer,raibert}.
@@
 
-Previous work has shown that the Hypercube-based NeuroEvolution of Augmenting Topologies (HyperNEAT) generative encoding~\cite{stanley2009hypercube} can automatically generate a variety of regular gaits that outperform gaits evolved with direct encodings~\cite{clune2009evolving,clune2011performance}.
+Previous work has shown that the Hypercube-based NeuroEvolution of Augmenting Topologies (HyperNEAT) generative encoding~\cite{stanley2009hypercube} can automatically generate a variety of regular gaits that outperform gaits evolved with direct encodings~\cite{clune2011performance,clune2009evolving}.
 However, that work only verified these claims in simulation. 
@@
 The success of that work raises the question of whether the performance gains were due to the added benefit of a simulator or the use of a simple, hand-designed encoding.
-We hypothesized that HyperNEAT, which has been previously shown to automatically discover complex regularities to produce high-performing gaits~\cite{clune2011performance,yos:clune}, would outperform the simpler encoding from Glette et al.
+We hypothesized that HyperNEAT, which has been previously shown to automatically discover complex regularities to produce high-performing gaits~\cite{yos:clune,clune2011performance}, would outperform the simpler encoding from Glette et al.
 if combined with a simulator.

### 2013-01-15 paper_sean/results.tex
reverted back to means/std

@@
 
-%We
-%compare performance in simulation and hardware
+%We%compare performance in simulation and hardware
 %to available data from previous studies.
@@
 (\figref{threeplots}-Top, $\emph{p} < 6.8\times10^{-8}$, comparing the best gaits in the final generation of each run via MATLAB's Wilcoxon rank sum test).
-Specifically, HyperNEAT gaits were 61.1\% faster in simulation (26.9
-cm/s, 95\% CIs 23.8 cm/s and 26.750 cm/s versus 16.7 cm/s $\pm$ 1.9 SD, \tabref{results}).
+Specifically, HyperNEAT gaits were 52.1\% faster in simulation (25.4
+cm/s, $\pm$ 3.4 SD versus 16.7 cm/s $\pm$ 1.9 SD, \tabref{results} -- qualitative conclusions remain the same even if the analysis is conducted using medians, but to make it easier for comparisons with earlier works~\cite{glette}, we use means $\pm$ SD).
 Plots of servo positions over time reveal that the evolved HyperNEAT gaits are regular, smooth and coordinated~(\figref{threeplots}-Left), confirming previous results with HyperNEAT in simulation~\cite{clune2011performance,clune2009evolving}.
 
-\figOneOnTwo[h!]{threeplots}{med_fit_gen}{0.62}{servo_plot_111}{.48}{frequencyPlotThreshFinal}{.48}
+%\figOneOnTwo[h!]{threeplots}{med_fit_gen}{0.62}{servo_plot_111}{.48}{frequencyPlotThreshFinal}{.48}
+\figOneOnTwo[h!]{threeplots}{avg_fit_vs_gen}{0.62}{servo_plot_111}{.48}{frequencyPlotThreshFinal}{.48}
 %\figOneOnTwo[h!]{threeplots}{med_fit_gen}{0.00}{servo_plot_111}{.48}{frequencyPlotThreshFinal}{.48}
@@
 \textbf{Top: }
+%HyperNEAT outperforms a genetic algorithm with a simple encoding when both algorithms are combined with a simulator.
+Plotted are medians over 20 runs in simulation (solid lines) $\pm$ CIs (dashed lines).
+HyperNEAT gaits are 52.1\% faster in
 HyperNEAT outperforms a genetic algorithm with a simple encoding when both algorithms are combined with a simulator.
-Plotted are medians over 20 runs in simulation (solid lines) $\pm$ CIs (dashed lines).
-HyperNEAT gaits are 61.1\% faster in
+Plotted are means over 20 runs in simulation (solid lines) $\pm$ SD (dashed lines).
+HyperNEAT gaits are 52.1\% faster in
 simulation and 5.1\% faster in reality~(\tabref{results}).
@@
 
-On the physical robot, HyperNEAT gaits from this study outperformed gaits from all previous QuadraTot studies~(\tabref{results})~\cite{yos:clune,haocheng,glette}, including those of Glette et al. 
+On the physical robot, HyperNEAT gaits from this study outperformed gaits from all previous QuadraTot studies~(\tabref{results})~\cite{yos:clune,glette,haocheng}, including those of Glette et al. 
  However, comparing performance in hardware between studies performed in
@@
 \hline
-HyperNEAT + simulator [this paper]                     & 40000       & \textbf{26.9}       &   \textbf{14.5} & --\\
+HyperNEAT + simulator [this paper]                     & 40000       & \textbf{**25.4}       &   \textbf{14.5} & --\\
 \hline
@@
 Subject to availability, data are reported from the experiments for this paper and three previous studies.
-Reported are the total number of evaluations per run, the median of the fastest gaits produced in each run in simulation, and the single fastest gait produced on the CCML and ROBIN copies of the QuadraTot robot (see text for their differences).
+Reported are the total number of evaluations per run, the average of the fastest gaits produced in each run in simulation, and the single fastest gait produced on the CCML and ROBIN copies of the QuadraTot robot (see text for their differences).
 *Instead of using the single fitness value reported in~\cite{glette}, we ran 19 additional runs and used the average fitness of those 20 runs.
 Velocities are in cm/s, and bold indicates the best performance.
-}
+**Median fitnesses of this paper: 26.9 cm/s with 95\% confidence intervals 23.8 cm/s and 26.75 cm/s.}
 \tablabel{results}
@@
 substantial, preventing effective learning~\cite{yos:clune}. 
-However, we can likely rule out the noise hypothesis because we performed 20 
-runs in simulation with the same evaluation samples as 
-in Yosinski et al.~\cite{yos:clune} and saw that the results were not 
-significantly different ($p < 0.855$, mean fitness 7.9 cm/s), making the likely explanation 
-the number of evaluations performed.
+We performed 20 runs in simulation with the same evaluation samples as 
+in Yosinski et al.~\cite{yos:clune} and saw that simulated gaits performed
+slightly better than the hardwared learned gaits ($p < 0.1571$, mean fitness 7.9 cm/s).
+This supports the noise hypothesis, but suggests that the number of evaluations performed is a more likely explanation.
 The encouragement of low-frequency gaits in this study
@@
 reality. Repeated attempts to minimize these problems were
-unsuccessful.  A major, unavoidable cause of the servo shutdowns was
-the design of the robot: the force required to lift the robot in many
-poses was too much for the servos. In future studies we will use a
+unsuccessful. 
+%A major, unavoidable cause of the servo shutdowns was
+%the design of the robot: the force required to lift the robot in many
+%poses was too much for the servos. 
+In future studies we will use a
 robot that has more mechanical advantage and requires less torque from
-each servo, such as the Aracna platform~\cite{lohmann2012aracna},
-which was designed in response to the challenges experienced with the
-QuadraTot platform.
+each servo, such as the Aracna platform~\cite{lohmann2012aracna}.
+%which was designed in response to the challenges experienced with the
+%QuadraTot platform. 
 That we did not model the servos in simulation, especially with their frequent failures, suggests that even better results could be obtained via a simulator that contained or learned servo models.

### 2013-01-16 paper_sean/results.tex
Edits to results, Jeff's affiliation superscript.

@@
 
-The simulator enabled HyperNEAT to evolve
-fast, organic gaits.
-In simulation, HyperNEAT gaits were significantly faster than those from Glette et al.
-(\figref{threeplots}-Top, $\emph{p} < 6.8\times10^{-8}$, comparing the best gaits in the final generation of each run via MATLAB's Wilcoxon rank sum test).
+The simulator enabled HyperNEAT to evolve fast, organic gaits. In
+simulation, HyperNEAT gaits (\figref{threeplots}-Top) were
+significantly faster than those from Glette et al, $\emph{p-value} <
+6.8\times10^{-8}$ when comparing the best gaits in the final
+generation of each run via Matlab's Wilcoxon rank sum test.
 Specifically, HyperNEAT gaits were 52.1\% faster in simulation (25.4
-cm/s, $\pm$ 3.4 SD versus 16.7 cm/s $\pm$ 1.9 SD, \tabref{results} -- qualitative conclusions remain the same even if the analysis is conducted using medians, but to make it easier for comparisons with earlier works~\cite{glette}, we use means $\pm$ SD).
+cm/s $\pm$ 3.4 SD versus 16.7 cm/s $\pm$ 1.9 SD). Note that \tabref{results} reports 
+mean $\pm$ SD to make comparison with earlier work~\cite{glette} easier, but qualitative conclusions remain the same even when the analysis is conducted using medians.
 Plots of servo positions over time reveal that the evolved HyperNEAT gaits are regular, smooth and coordinated~(\figref{threeplots}-Left), confirming previous results with HyperNEAT in simulation~\cite{clune2011performance,clune2009evolving}.
@@
 HyperNEAT gaits are 52.1\% faster in
-HyperNEAT outperforms a genetic algorithm with a simple encoding when both algorithms are combined with a simulator.
+HyperNEAT outperforms a genetic algorithm with a simple encoding \cite{glette} when both algorithms are combined with a simulator.
 Plotted are means over 20 runs in simulation (solid lines) $\pm$ SD (dashed lines).
 HyperNEAT gaits are 52.1\% faster in
-simulation and 5.1\% faster in reality~(\tabref{results}).
+simulation and 5.1\% faster in reality than those from a previous study \cite{glette}~(details in \tabref{results}).
 \textbf{Left: }
@@
 Velocities are in cm/s, and bold indicates the best performance.
-**Median fitnesses of this paper: 26.9 cm/s with 95\% confidence intervals 23.8 cm/s and 26.75 cm/s.}
+**Fitness median 26.9 cm/s, 95\% confidence interval [23.8 cm/s, 26.75 cm/s].}
 \tablabel{results}

### 2013-01-16 paper_sean/results.tex
s

--- 2013-01-15 paper_sean/results.tex
@@
 In simulation, HyperNEAT gaits were significantly faster than those from Glette et al.
-(\figref{threeplots}-Top, $\emph{p} < 6.8\times10^{-8}$, comparing the best gaits in the final generation of each run via MATLAB's Wilcoxon rank sum test).
+(\figref{threeplots}-Top, $\emph{p} < 6.8\times10^{-8}$, comparing the best gaits in the final generation of each run via MATLAB's one-sided Wilcoxon rank sum test).
 Specifically, HyperNEAT gaits were 52.1\% faster in simulation (25.4

### 2013-01-16 paper_sean/results.tex

--- 2013-01-16 paper_sean/results.tex
@@
 That we did not model the servos in simulation, especially with their frequent failures, suggests that even better results could be obtained via a simulator that contained or learned servo models.
-In future work we will also incorporate techniques to minimize the gap between the simulator and reality~\cite{koos2010crossing,bongard,zagal}.
+In future work we will also incorporate techniques to minimize the gap between the simulator and reality~\cite{koos2010crossing,bongard,zagal,koos2011transferability}.

### 2013-01-16 paper_sean/results.tex
jeff last stage edits

--- 2013-01-16 paper_sean/results.tex
@@
 
-The simulator enabled HyperNEAT to evolve fast, organic gaits. In
-simulation, HyperNEAT gaits (\figref{threeplots}-Top) were
-significantly faster than those from Glette et al, $\emph{p-value} <
+The simulator enabled HyperNEAT to evolve fast, natural gaits. In
+simulation, HyperNEAT gaits were
+significantly faster than those from Glette et al. (\figref{threeplots}-Top, $p <
 6.8\times10^{-8}$ when comparing the best gaits in the final
-generation of each run via Matlab's Wilcoxon rank sum test.
+generation of each run via Matlab's Wilcoxon rank sum test).
 Specifically, HyperNEAT gaits were 52.1\% faster in simulation (25.4
-cm/s $\pm$ 3.4 SD versus 16.7 cm/s $\pm$ 1.9 SD). Note that \tabref{results} reports 
-mean $\pm$ SD to make comparison with earlier work~\cite{glette} easier, but qualitative conclusions remain the same even when the analysis is conducted using medians.
+cm/s $\pm$ 3.4 SD versus 16.7 cm/s $\pm$ 1.9 SD). To facilitate comparisons to earlier works~\cite{yos:clune,glette} we report 
+mean $\pm$ SD, but our qualitative conclusions are the same when using medians.
 Plots of servo positions over time reveal that the evolved HyperNEAT gaits are regular, smooth and coordinated~(\figref{threeplots}-Left), confirming previous results with HyperNEAT in simulation~\cite{clune2011performance,clune2009evolving}.
@@
 hardware. To control for this possibility, we ran the fastest gait
-from Glette et al. 10 times on CCML and measured an average speed of
+from Glette et al. 10 times on CCML and measured a mean speed of
 only 12.95 cm/s $\pm0.93$ SD (vs. 17.8 cm/s measured on ROBIN) and a maximum of
@@
 Subject to availability, data are reported from the experiments for this paper and three previous studies.
-Reported are the total number of evaluations per run, the average of the fastest gaits produced in each run in simulation, and the single fastest gait produced on the CCML and ROBIN copies of the QuadraTot robot (see text for their differences).
-*Instead of using the single fitness value reported in~\cite{glette}, we ran 19 additional runs and used the average fitness of those 20 runs.
+Reported are the total number of evaluations per run, the mean of the fastest gaits produced in each run in simulation, and the single fastest gait produced on the CCML and ROBIN copies of the QuadraTot robot (see text for their differences).
+*Instead of using the single fitness value reported in~\cite{glette}, we ran 19 additional runs and used the mean fitness of those 20 runs.
 Velocities are in cm/s, and bold indicates the best performance.
-**Fitness median 26.9 cm/s, 95\% confidence interval [23.8 cm/s, 26.75 cm/s].}
+**The median fitness that corresponds with this mean is 26.9 cm/s, 95\% confidence interval [23.8 cm/s, 26.75 cm/s].}
 \tablabel{results}
@@
 to a total difference of 40000 vs 180 per evolutionary run compared to Yosinski et al.~\cite{yos:clune}. Another
-potential cause of improved performance is that with a simulator
-HyperNEAT had much less noise in the evaluation process and could thus find coordinated, regular, gaits, which perform better. On the physical robot, the noise in the evaluation was
+potential cause of improved performance is the lower noise in the simulator, which could have helped HyperNEAT find coordinated, regular, gaits, which perform better.
+On the physical robot, the noise in the evaluation was
 substantial, preventing effective learning~\cite{yos:clune}. 
-We performed 20 runs in simulation with the same evaluation samples as 
-in Yosinski et al.~\cite{yos:clune} and saw that simulated gaits performed
-slightly better than the hardwared learned gaits ($p < 0.1571$, mean fitness 7.9 cm/s).
-This supports the noise hypothesis, but suggests that the number of evaluations performed is a more likely explanation.
+To investigate this hypothesis, we performed 20 runs in simulation with only 180 fitness evaluations, which was the number used in Yosinski et al.~\cite{yos:clune}. The simulated gaits performed
+slightly, but not significantly, better than those evolved in hardware ($p = 0.1571$, mean fitness 7.9 cm/s).
+Reduced noise thus may have had a small affect on performance, but the substantial performance gains that resulted from using a simulator likely came from the additional evaluations it afforded.
 The encouragement of low-frequency gaits in this study

### 2013-01-16 paper_sean/methods.tex
JMC final edits?

@@
 To make a statistical comparison to the encoding from Glette et al., we conducted 19 runs using the original code to supplement the one run performed for that paper.
-Each gait was evaluated for fourteen seconds in reality, with interpolation from and to a stationary position in the first and last two seconds, respectively, as in Yosinski et al.
-effectively resulting in 12 seconds of full-speed
+Each gait was evaluated for fourteen seconds in reality, with interpolation from and to a stationary position in the first and last two seconds, respectively, as in Yosinski et al., effectively resulting in 12 seconds of full-speed
 motion. In simulation gaits were evaluated for 12 seconds. 

### 2013-01-16 paper_sean/results.tex
JMC final edits?

@@
 The simulator likely improved performance because of the number of evaluations it enabled, both in terms of the population size (200 vs. 9) and the number of generations (200 vs. 20), leading
-to a total difference of 40000 vs 180 per evolutionary run compared to Yosinski et al.~\cite{yos:clune}. Another
+to a total difference of 40000 vs. 180 per evolutionary run compared to Yosinski et al.~\cite{yos:clune}. Another
 potential cause of improved performance is the lower noise in the simulator, which could have helped HyperNEAT find coordinated, regular, gaits, which perform better.
@@
 slightly, but not significantly, better than those evolved in hardware ($p = 0.1571$, mean fitness 7.9 cm/s).
-Reduced noise thus may have had a small affect on performance, but the substantial performance gains that resulted from using a simulator likely came from the additional evaluations it afforded.
+Reduced noise thus may have had a small affect on performance, but the substantial performance gains that resulted from using a simulator likely came from the additional evaluations the simulator afforded.
 The encouragement of low-frequency gaits in this study

### 2013-01-16 paper_sean/simulator.tex
JMC final edits?

--- 2012-11-05 paper_sean/simulator.tex
@@
 \subsubsection{Simulator:}
-Gaits evolved in the simulator from Glette et al., which represents QuadraTot in the Nvidia PhysX physics engine\footnotemark[\value{footnote}], including the mass and size of the QuadraTot components and its degrees of freedom~(\figref{quadratotWhiteBooties}-Right).
+Gaits evolved in the simulator from Glette et al.\footnotemark[\value{footnote}], which represents QuadraTot in the Nvidia PhysX physics engine, including the mass and size of the QuadraTot components and its degrees of freedom~(\figref{quadratotWhiteBooties}-Right).
 In the simulator, each individual joint range was limit as described above, but Smart Cropping was not included because we found that it hindered performance by limiting the types of gaits evolution could explore in early generations.

### 2013-03-06 paper_sean/results.tex
jmc last edit\!
Final version of paper

@@
 The simulator enabled HyperNEAT to evolve fast, natural gaits. In
-simulation, HyperNEAT gaits were
-significantly faster than those from Glette et al. (\figref{threeplots}-Top, $p <
+simulation, HyperNEAT gaits were faster than those from Glette et al. (\figref{threeplots}-Top, $p <
 6.8\times10^{-8}$ when comparing the best gaits in the final
@@
 mean $\pm$ SD, but our qualitative conclusions are the same when using medians.
-Plots of servo positions over time reveal that the evolved HyperNEAT gaits are regular, smooth and coordinated~(\figref{threeplots}-Left), confirming previous results with HyperNEAT in simulation~\cite{clune2011performance,clune2009evolving}.
+Plots of servo positions over time reveal that the evolved HyperNEAT gaits are regular and coordinated~(\figref{threeplots}-Left), confirming previous results with HyperNEAT in simulation~\cite{clune2011performance,clune2009evolving}.
 
 %\figOneOnTwo[h!]{threeplots}{med_fit_gen}{0.62}{servo_plot_111}{.48}{frequencyPlotThreshFinal}{.48}
-\figOneOnTwo[h!]{threeplots}{avg_fit_vs_gen}{0.62}{servo_plot_111}{.48}{frequencyPlotThreshFinal}{.48}
+\figOneOnTwo[h!]{threeplots}{avg_fit_vs_gen}{0.605}{servo_plot_111}{.48}{frequencyPlotThreshFinal}{.48}
 %\figOneOnTwo[h!]{threeplots}{med_fit_gen}{0.00}{servo_plot_111}{.48}{frequencyPlotThreshFinal}{.48}
@@
 To investigate this hypothesis, we performed 20 runs in simulation with only 180 fitness evaluations, which was the number used in Yosinski et al.~\cite{yos:clune}. The simulated gaits performed
-slightly, but not significantly, better than those evolved in hardware ($p = 0.1571$, mean fitness 7.9 cm/s).
+slightly, but not significantly, better than those evolved in hardware ($p = 0.1571$, mean fitness $7.9 \pm 2.14$ cm/s).
 Reduced noise thus may have had a small affect on performance, but the substantial performance gains that resulted from using a simulator likely came from the additional evaluations the simulator afforded.
@@
 That we did not model the servos in simulation, especially with their frequent failures, suggests that even better results could be obtained via a simulator that contained or learned servo models.
-In future work we will also incorporate techniques to minimize the gap between the simulator and reality~\cite{koos2010crossing,bongard,zagal,koos2011transferability}.
+In future work we will also incorporate techniques to minimize the gap between the simulator and reality~\cite{bongard,zagal,koos2011transferability}.
